---
title: "Ordinal Pairwise-Deletion Gamma"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- file.path("results", "summary.csv")
rejections_path <- file.path("results", "rejections.csv")
if (!file.exists(summary_path) || !file.exists(rejections_path)) {
  stop("Run `Rscript run_experiment.R --smoke` first.", call. = FALSE)
}
summary <- read.csv(summary_path)
rejections <- read.csv(rejections_path)
fmt_rate <- function(x, se) {
  ifelse(is.na(x), "NA", sprintf("%.3f (MCSE %.3f)", x, se))
}
wide <- reshape(
  summary[, c("n_per_group", "missing_rate", "pd_gamma",
              "rejection_rate", "mc_se", "usable", "failures")],
  idvar = c("n_per_group", "missing_rate"),
  timevar = "pd_gamma",
  direction = "wide")
wide$Nominal <- fmt_rate(wide$rejection_rate.nominal, wide$mc_se.nominal)
wide$Overlap <- fmt_rate(wide$rejection_rate.overlap, wide$mc_se.overlap)
tab <- wide[, c("n_per_group", "missing_rate", "Nominal", "Overlap",
                "usable.nominal", "failures.nominal")]
names(tab) <- c("N Per Group", "Missing Rate", "Nominal Gamma",
                "Overlap Gamma", "Usable Reps", "Failures")
```

## Question

Does the overlap-weighted ordinal pairwise-deletion NACOV behave differently from
the conventional nominal-\(N\) matrix in a WLSMV loading-invariance test with
ordinal MCAR missingness?

## Short Answer

In the smoke run, the two matrices are identical with complete data. With 50%
missingness on two group-B indicators, the overlap correction lowers the nested
test statistic in every usable replicate, which is the direction predicted by
the pairwise-deletion Gamma note. The run is intentionally tiny; use it as a
plumbing and direction check, not a stable Type I estimate.

## Evidence

```{r}
knitr::kable(tab, row.names = FALSE)
```

The replicate-level outputs are in `results/rejections.csv`; the compact
cell-level rates are in `results/summary.csv`.

## Caveats

The default run is a short probe, not a replication of Chen et al. The design
uses MCAR missingness so the experiment isolates the Gamma scaling issue rather
than MAR point-estimate bias. Some small ordinal cells can fail because the
polychoric or dense NACOV problem is numerically singular; failures are counted
instead of aborting the run.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 200 --n 300,1000 --missing-rate 0,.30,.50
quarto render report.qmd
```
