---
title: "Ordinal Pairwise-Deletion Gamma"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- file.path("results", "summary.csv")
rejections_path <- file.path("results", "rejections.csv")
if (!file.exists(summary_path) || !file.exists(rejections_path)) {
  stop("Run `Rscript run_experiment.R --smoke` first.", call. = FALSE)
}
summary <- read.csv(summary_path)
rejections <- read.csv(rejections_path)

fmt_rate <- function(x, se) {
  ifelse(is.na(x), "NA", sprintf("%.3f (MCSE %.3f)", x, se))
}
headline_methods <- c("naive", "SB", "SS", "pEBA4", "pOLS", "all",
                      "adjusted", "mixture")
head <- summary[summary$method %in% headline_methods, ]
wide <- reshape(
  head[, c("truth", "outcome", "method", "pd_gamma",
           "rejection_rate", "mc_se", "usable", "failures")],
  idvar = c("truth", "outcome", "method"),
  timevar = "pd_gamma",
  direction = "wide")
for (nm in c("rejection_rate.nominal", "mc_se.nominal",
             "usable.nominal", "failures.nominal",
             "rejection_rate.overlap", "mc_se.overlap",
             "usable.overlap", "failures.overlap")) {
  if (!nm %in% names(wide)) wide[[nm]] <- NA_real_
}
wide$Nominal <- fmt_rate(wide$rejection_rate.nominal, wide$mc_se.nominal)
wide$Overlap <- fmt_rate(wide$rejection_rate.overlap, wide$mc_se.overlap)
tab <- wide[, c("truth", "outcome", "method", "Nominal", "Overlap",
                "usable.nominal", "failures.nominal")]
names(tab) <- c("Truth", "Outcome", "Method", "Nominal Gamma",
                "Overlap Gamma", "Usable Reps", "Failures")
tab <- tab[order(tab$Truth, tab$Outcome, match(tab$Method, headline_methods)), ]
failures <- summary[summary$method == "fit_failed",
                    c("truth", "outcome", "pd_gamma", "reps", "failures")]

meta <- tryCatch(read.csv(file.path("results", "metadata.csv")),
                 error = function(e) data.frame())
```

## Question

Does overlap-weighting the ordinal pairwise-deletion NACOV change the calibration
of the standard WLSMV/FMG test battery in a simple two-factor measurement
invariance design with 20% MCAR missingness?

## Design

The simulation uses a two-group, two-factor ordinal CFA with eight indicators
and five response categories. The `fake` condition has equal loadings across
groups, so the metric invariance test is a null test. The `real` condition
changes one loading per factor in group B, so the same nested test should reject.
Both conditions also record the goodness-of-fit battery for the fitted metric
model.

The two Gamma variants share thresholds and polychorics; only the pairwise
NACOV scaling changes:

- `nominal`: conventional nominal-\(N\) pairwise-deletion Gamma.
- `overlap`: support-overlap Gamma from `docs/research/notes/ordinal_pd_gamma.tex`.

## Evidence

```{r}
knitr::kable(tab, row.names = FALSE)
```

The full replicate-level table is in `results/rejections.csv`; the aggregated
cell table is in `results/summary.csv`.

```{r}
if (nrow(failures)) {
  names(failures) <- c("Truth", "Outcome", "Gamma", "Reps", "Failures")
  knitr::kable(failures, row.names = FALSE)
}
```

## Caveats

The smoke run is a plumbing and direction check. Use a larger run before reading
Monte Carlo rates substantively. The point estimates are still ordinary
pairwise ordinal estimates; this experiment isolates the robust-reference
matrix used by WLSMV/FMG-style tests. If a Gamma variant produces a non-positive
semidefinite robust matrix, the row is retained as `fit_failed` instead of being
dropped.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 200 --n 300,1000 --missing-rate .20
quarto render report.qmd
```

```{r}
if (nrow(meta)) knitr::kable(meta, row.names = FALSE)
```
