---
title: "Ordinal Pairwise-Deletion Gamma Rejection Rates"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
rates_path <- file.path("results", "rejection_rates.csv")
p_path <- file.path("results", "p_values.csv")
if (!file.exists(rates_path) || !file.exists(p_path)) {
  stop("Run `Rscript run_experiment.R --smoke` first.", call. = FALSE)
}
rates <- read.csv(rates_path)
p_values <- read.csv(p_path)
meta <- tryCatch(read.csv(file.path("results", "metadata.csv")),
                 error = function(e) data.frame())

focus_methods <- c("naive", "SB", "SS", "pEBA4", "pOLS", "all",
                   "adjusted", "mixture")
tab <- rates[
  rates$method %in% focus_methods &
    rates$usable > 0 &
    is.finite(rates$rejection_rate),
  c("truth", "outcome", "method", "pd_gamma", "rejection_rate", "mc_se")
]
tab$rate <- sprintf("%.3f", tab$rejection_rate)
tab$mcse <- sprintf("%.3f", tab$mc_se)
tab <- tab[, c("truth", "outcome", "method", "pd_gamma", "rate", "mcse")]
names(tab) <- c("Truth", "Test", "Method", "Gamma", "Rejection Rate", "MCSE")
tab <- tab[order(tab$Truth, tab$Test, match(tab$Method, focus_methods),
                 tab$Gamma), ]

design <- if (nrow(meta)) meta[1L, ] else data.frame(
  reps = length(unique(p_values$rep)),
  n = paste(unique(p_values$n_per_group), collapse = ","),
  missing_rate = paste(unique(p_values$missing_rate), collapse = ","),
  truths = paste(unique(p_values$truth), collapse = ","))
```

## Simulated Model

This is a short rejection-rate study for the ordinal pairwise-deletion Gamma
correction. Each replication draws two equal-size groups from a two-factor
ordinal CFA:

```text
f1 =~ y1 + y2 + y3 + y4
f2 =~ y5 + y6 + y7 + y8
f1 ~~ f2
```

The latent correlation is 0.35. Group A loadings are
`0.82, 0.76, 0.70, 0.64, 0.80, 0.74, 0.68, 0.62`; residual variances are chosen
so each latent response has unit variance. Five ordinal categories are formed
with thresholds `-0.80, -0.15, 0.55, 1.15`.

The `fake` condition is the null: group B uses the same loadings as group A. The
`real` condition introduces loading non-invariance in group B by increasing the
`y4` loading and decreasing the `y8` loading. MCAR missingness is then applied
independently to all eight indicators.

For each replication, the same simulated data are analyzed twice:

- `nominal`: the conventional nominal-\(N\) pairwise-deletion Gamma.
- `overlap`: the support-overlap Gamma from `docs/research/notes/ordinal_pd_gamma.tex`.

The per-replication simulated p-values are in `results/p_values.csv`.

## Rejection Rates

```{r}
knitr::kable(tab, row.names = FALSE)
```

## Run Metadata

```{r}
knitr::kable(design, row.names = FALSE)
```

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 200 --n 500 --missing-rate .20
quarto render report.qmd
```
