---
title: "Ordinal Observed-Score Omega DWLS Probe"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- "results/simulation_summary.csv"
metadata_path <- "results/metadata.csv"
if (!file.exists(summary_path) || !file.exists(metadata_path)) {
  stop("Missing results. Run: Rscript run_experiment.R --smoke")
}
summary <- read.csv(summary_path)
metadata <- read.csv(metadata_path)
```

## Question

Does the new all-ordinal observed-score omega surface run end-to-end on a DWLS fit,
and are its first delta-method intervals at least numerically sane before we spend
effort on profile-LR or small-sample corrections?

## Short Answer

This is a smoke/probe lane, not a calibration result. In the current run
(`r metadata$reps` reps per regime, N = `r metadata$n`), the wrapper returns finite
estimates and SEs in both balanced and threshold-extreme three-category regimes.
Use the table below only to catch gross failures; coverage is too noisy until the
grid is expanded.

## Evidence

```{r}
shown <- summary
shown[] <- lapply(shown, function(x) if (is.numeric(x)) round(x, 4) else x)
knitr::kable(shown)
```

## Caveats

The target is a large Monte-Carlo category-score covariance, not an analytic
population integral. The design is a one-factor correct-model check; structural
misspecification, non-normal latent responses, multi-group pooling, and any
profile-LR/Bartlett correction are intentionally outside this first probe.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 200 --n 500 --n-pop 500000
quarto render report.qmd
```
