---
title: "Small Bias Check for Reduced-Bias SEM Estimators"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
res_dir <- "results"
needed <- file.path(res_dir, c("method_summary.csv", "parameter_summary.csv",
                               "fit_status.csv", "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/39-rbm-bias-small/run_experiment.R --smoke",
       call. = FALSE)
}
method_summary <- read.csv(file.path(res_dir, "method_summary.csv"))
parameter_summary <- read.csv(file.path(res_dir, "parameter_summary.csv"))
fit_status <- read.csv(file.path(res_dir, "fit_status.csv"))
metadata <- read.csv(file.path(res_dir, "metadata.csv"))
meta <- setNames(metadata$value, metadata$key)

method_summary$Method <- factor(method_summary$method,
                                levels = c("standard", "posthoc", "integrated"))
primary <- method_summary[order(method_summary$estimator,
                                method_summary$Method), ]
best <- do.call(rbind, lapply(split(primary, primary$estimator), function(d) {
  d[which.min(d$mean_abs_bias_primary), , drop = FALSE]
}))
fmt <- function(x, digits = 4) format(round(x, digits), nsmall = digits)
```

## Question

Do the new reduced-bias estimators move finite-sample point estimates in the
right direction in a small magmaan-owned run, rather than only in the published
result-object replication? The check compares the ordinary estimator with the
explicit post-hoc correction and the implicit integrated estimator for one GLS,
one FIML, and one ordinal DWLS setup.

## Short Answer

In this run (`r meta[["reps"]]` replications, `N = r meta[["n"]]`), the smallest
mean absolute bias over the primary loading / latent-variance parameters was:
`r paste(sprintf("%s: %s (%s)", toupper(best$estimator), best$method,
                 fmt(best$mean_abs_bias_primary)), collapse = "; ")`.

Treat this as a smoke-scale direction check, not a calibration result. The
large-sample target is a simulated reference fit with `N = r meta[["ref_n"]]`,
and the default grid is intentionally small enough to rerun while developing
the RBM machinery.

## Evidence

```{r}
tab <- primary[c("estimator", "method", "mean_abs_bias_primary",
                 "mean_rmse_primary", "mean_abs_bias_all", "acceptance_rate",
                 "median_elapsed_ms")]
names(tab) <- c("Estimator", "Method", "Mean Abs Bias", "Mean RMSE",
                "Mean Abs Bias, All Free", "Acceptance", "Median Time (ms)")
tab$Estimator <- toupper(tab$Estimator)
tab$Method <- c(standard = "Standard", posthoc = "Post-hoc",
                integrated = "Integrated")[tab$Method]
tab$`Mean Abs Bias` <- fmt(tab$`Mean Abs Bias`)
tab$`Mean RMSE` <- fmt(tab$`Mean RMSE`)
tab$`Mean Abs Bias, All Free` <- fmt(tab$`Mean Abs Bias, All Free`)
tab$Acceptance <- sprintf("%.0f%%", 100 * tab$Acceptance)
tab$`Median Time (ms)` <- round(tab$`Median Time (ms)`)
knitr::kable(tab, row.names = FALSE)
```

The primary summary uses only the free factor loadings and latent variance, so
the GLS, FIML, and DWLS rows are comparable. The full parameter summaries,
including residual variances, thresholds, and FIML intercepts where present, are
in `results/parameter_summary.csv`.

## Caveats

This is a deliberately small probe: one correctly specified CFA population, one
sample size, normal latent responses, MCAR missingness for FIML, and two
thresholds for DWLS. It can catch gross direction and stability problems in the
standard/post-hoc/integrated comparison, but it cannot establish uniform bias
reduction, coverage, or optimizer reliability.

## Reproduce

```sh
Rscript experiments/39-rbm-bias-small/run_experiment.R --smoke
Rscript experiments/39-rbm-bias-small/run_experiment.R --reps 20 --n 160 --ref-n 20000
quarto render experiments/39-rbm-bias-small/report.qmd
```
