---
title: "Small Bias Check for Reduced-Bias SEM Estimators"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
---

```{r}
res_dir <- params$results_dir
needed <- file.path(res_dir, c("method_summary.csv", "parameter_summary.csv",
                               "fit_status.csv", "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/39-rbm-bias-small/run_experiment.R --smoke",
       call. = FALSE)
}
method_summary <- read.csv(file.path(res_dir, "method_summary.csv"))
parameter_summary <- read.csv(file.path(res_dir, "parameter_summary.csv"))
fit_status <- read.csv(file.path(res_dir, "fit_status.csv"))
metadata <- read.csv(file.path(res_dir, "metadata.csv"))
meta <- setNames(metadata$value, metadata$key)
meta_value <- function(key, default = NA_character_) {
  if (key %in% names(meta)) meta[[key]] else default
}
fmt <- function(x, digits = 4) format(round(x, digits), nsmall = digits)
avg_rel <- as.numeric(meta_value("average_item_reliability"))
if (!length(avg_rel)) avg_rel <- NA_real_
avg_rel_text <- if (is.na(avg_rel)) "not recorded" else fmt(avg_rel)
estimators_text <- paste(toupper(sort(unique(method_summary$estimator))),
                         collapse = ", ")

if (!"mean_abs_error_primary" %in% names(method_summary) &&
    "mean_abs_bias_primary" %in% names(method_summary)) {
  method_summary$mean_abs_error_primary <- method_summary$mean_abs_bias_primary
}
if (!all(c("mean_bias_primary", "mean_abs_mean_bias_primary") %in%
         names(method_summary))) {
  derived <- lapply(seq_len(nrow(method_summary)), function(i) {
    key <- method_summary[i, ]
    d <- parameter_summary[parameter_summary$estimator == key$estimator &
                             parameter_summary$method == key$method &
                             parameter_summary$param_class %in%
                               c("loading", "latent_variance"), ]
    data.frame(
      mean_bias_primary = mean(d$bias, na.rm = TRUE),
      mean_abs_mean_bias_primary = mean(abs(d$bias), na.rm = TRUE)
    )
  })
  derived <- do.call(rbind, derived)
  if (!"mean_bias_primary" %in% names(method_summary)) {
    method_summary$mean_bias_primary <- derived$mean_bias_primary
  }
  if (!"mean_abs_mean_bias_primary" %in% names(method_summary)) {
    method_summary$mean_abs_mean_bias_primary <-
      derived$mean_abs_mean_bias_primary
  }
}
for (nm in c("primary_p99_abs_error", "primary_extreme_rate_abs10",
             "primary_extreme_rate_abs100")) {
  if (!nm %in% names(method_summary)) method_summary[[nm]] <- NA_real_
}

method_summary$Method <- factor(method_summary$method,
                                levels = c("standard", "posthoc", "integrated"))
primary <- method_summary[order(method_summary$estimator,
                                method_summary$Method), ]
best <- do.call(rbind, lapply(split(primary, primary$estimator), function(d) {
  d[which.min(d$mise_primary), , drop = FALSE]
}))
```

## Question

Do the new reduced-bias estimators move finite-sample point estimates in the
right direction in a small magmaan-owned run, rather than only in the published
result-object replication? The check compares the ordinary estimator with the
explicit post-hoc correction and the implicit integrated estimator for
the requested estimator set (`r estimators_text`).

## Short Answer

In this run (`r meta[["reps"]]` replications, `N = r meta[["n"]]`), the smallest
MISE over the primary loading / latent-variance vector was:
`r paste(sprintf("%s: %s (%s)", toupper(best$estimator), best$method,
                 fmt(best$mise_primary)), collapse = "; ")`.

Treat this as a smoke-scale direction check, not a calibration result. The
large-sample target is a simulated reference fit with `N = r meta[["ref_n"]]`,
the average item reliability is `r avg_rel_text`,
and the default grid is intentionally small enough to rerun while developing the
RBM machinery.

## Evidence

```{r}
tab <- primary[c("estimator", "method", "mean_bias_primary",
                 "mean_abs_mean_bias_primary", "mean_abs_error_primary",
                 "mean_rmse_primary", "mise_primary",
                 "primary_extreme_rate_abs100",
                 "relative_mise_primary", "mean_abs_pu_error_primary",
                 "acceptance_rate", "median_elapsed_ms")]
names(tab) <- c("Estimator", "Method", "Mean Bias", "Mean |Bias|", "MAE",
                "Mean RMSE", "MISE", "|Err| > 100", "MISE / Standard",
                "Mean |PU - .5|", "Acceptance", "Median Time (ms)")
tab$Estimator <- toupper(tab$Estimator)
tab$Method <- c(standard = "Standard", posthoc = "Post-hoc",
                integrated = "Integrated")[tab$Method]
tab$`Mean Bias` <- fmt(tab$`Mean Bias`)
tab$`Mean |Bias|` <- fmt(tab$`Mean |Bias|`)
tab$MAE <- fmt(tab$MAE)
tab$`Mean RMSE` <- fmt(tab$`Mean RMSE`)
tab$MISE <- fmt(tab$MISE)
tab$`|Err| > 100` <- ifelse(is.na(tab$`|Err| > 100`), "",
                            sprintf("%.2f%%", 100 * tab$`|Err| > 100`))
tab$`MISE / Standard` <- fmt(tab$`MISE / Standard`, 3)
tab$`Mean |PU - .5|` <- fmt(tab$`Mean |PU - .5|`, 3)
tab$Acceptance <- sprintf("%.0f%%", 100 * tab$Acceptance)
tab$`Median Time (ms)` <- round(tab$`Median Time (ms)`)
knitr::kable(tab, row.names = FALSE)
```

The primary summary uses only the free factor loadings and latent variance, so
the requested estimator rows are comparable. MISE is the average, over
replications, of the mean squared error across that primary parameter vector.
`MAE` is mean absolute estimation error, not absolute mean bias. The full
parameter summaries, including residual variances, thresholds, and FIML
intercepts where present, are in `r file.path(res_dir, "parameter_summary.csv")`.

## Caveats

This is still not a replication of the paper's full design: it uses one
correctly specified CFA population, one sample size, normal latent responses,
MCAR missingness for FIML, and two thresholds for DWLS. The ML branch is the
paper-relevant likelihood path, but the paper's two-factor and growth-curve
grids remain larger and harder. This probe can catch gross direction and
stability problems in the standard/post-hoc/integrated comparison, but it cannot
establish uniform bias reduction, coverage, or optimizer reliability.

## Reproduce

```sh
Rscript experiments/39-rbm-bias-small/run_experiment.R --smoke
Rscript experiments/39-rbm-bias-small/run_experiment.R --reps 250 --n 160 --ref-n 50000 --cores auto --chunk-size 120
Rscript experiments/39-rbm-bias-small/run_experiment.R --reps 5000 --n 30 --ref-n 100000 --estimators ml,gls --cores 1 --chunk-size 100 --results-dir results/tough-n30-ml-gls
quarto render experiments/39-rbm-bias-small/report.qmd
quarto render experiments/39-rbm-bias-small/report.qmd -P results_dir:results/tough-n30-ml-gls
```
