---
title: "Multi-information barrier versus the objective posterior"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
profile <- Sys.getenv("EXP86_PROFILE", "full")
res <- file.path("results", profile)
need <- file.path(res, c("fits.csv", "params.csv", "posterior.csv", "design.csv"))
if (!all(file.exists(need))) {
  stop("Missing results. Run: Rscript run_experiment.R --full", call. = FALSE)
}
fits <- read.csv(need[[1]], stringsAsFactors = FALSE)
params <- read.csv(need[[2]], stringsAsFactors = FALSE)
post <- read.csv(need[[3]], stringsAsFactors = FALSE)
design <- read.csv(need[[4]], stringsAsFactors = FALSE)

design_label <- c(
  f1_p3 = "One factor, 3 indicators",
  f1_p5 = "One factor, 5 indicators",
  f2_r90 = "Two factors, r = .90",
  f2_r97 = "Two factors, r = .97",
  path_r2_90 = "Path, R²(f3) = .90")
key_label <- c(
  f1_p3 = "residual variance, strong items", f1_p5 = "residual variance, strong items",
  f2_r90 = "factor correlation", f2_r97 = "factor correlation",
  path_r2_90 = "R² of f3")
designs <- intersect(names(design_label), unique(fits$design))
n_values <- sort(unique(fits$n))
pct <- function(x) sprintf("%.0f", 100 * x)
num <- function(x) sprintf("%.3f", x)

bar_path <- file.path(res, "params_barriers.csv")
has_bar <- file.exists(bar_path)
if (has_bar) {
  bar <- read.csv(bar_path, stringsAsFactors = FALSE)
  bar$lower <- NA_real_
  bar$upper <- NA_real_
  params <- rbind(params, bar[, names(params)])
  bar_fits <- read.csv(file.path(res, "fits_barriers.csv"), stringsAsFactors = FALSE)
}

post$pass <- with(post, !is.na(rhat_key) & rhat_key <= 1.05 & ess_key >= 400 &
                    jeffreys_ess >= 200)
key <- params[params$key, ]
key$err <- key$est - key$truth
key <- merge(key, post[, c("design", "n", "rep", "pass")], by = c("design", "n", "rep"))
keyp <- key[key$pass, ]

# Per quantity: did ML land past the admissible wall for this quantity?
ml_key <- keyp[keyp$method == "ml", c("design", "n", "rep", "quantity", "est")]
names(ml_key)[5] <- "ml_est"
ml_key$ml_past <- with(ml_key, ifelse(grepl("~~ x", quantity), ml_est < 0, ml_est > 1))
jm <- keyp[keyp$method == "jeff_mean", c("design", "n", "rep", "quantity", "est")]
names(jm)[5] <- "jeff"
tr <- merge(merge(keyp, jm, by = c("design", "n", "rep", "quantity")),
            ml_key[, c("design", "n", "rep", "quantity", "ml_past")],
            by = c("design", "n", "rep", "quantity"))
tr$gap <- tr$est - tr$jeff

rmse <- function(d, n, m) {
  s <- keyp[keyp$design == d & keyp$n == n & keyp$method == m, ]
  sqrt(mean(s$err^2, na.rm = TRUE))
}
gap_rms <- function(d, past, m, nmax = 100) {
  s <- tr[tr$design == d & tr$ml_past %in% past & tr$method == m & tr$n <= nmax, ]
  if (!nrow(s)) return(NA_real_)
  sqrt(mean(s$gap^2, na.rm = TRUE))
}
best_lambda <- function(d, past) {
  lam <- c(pen_l010 = 0.1, pen_l025 = 0.25, pen_l050 = 0.5, pen_l100 = 1)
  g <- vapply(names(lam), function(m) gap_rms(d, past, m), numeric(1))
  if (all(is.na(g))) NA_real_ else lam[[which.min(g)]]
}
cover <- function(d, n, m) {
  s <- keyp[keyp$design == d & keyp$n == n & keyp$method == m, ]
  mean(s$lower <= s$truth & s$truth <= s$upper, na.rm = TRUE)
}
pass_rate <- mean(post$pass)
sec_post <- median(post$seconds_mcmc + post$seconds_jeffreys, na.rm = TRUE)
ms_pen <- 1000 * median(fits$seconds[fits$method == "pen_l025"], na.rm = TRUE)
n_small <- min(n_values)
var_designs <- intersect(c("f1_p3", "f1_p5"), designs)
cor_designs <- intersect(c("f2_r90", "f2_r97", "path_r2_90"), designs)
lambdas <- c(pen_l010 = 0.1, pen_l025 = 0.25, pen_l050 = 0.5, pen_l100 = 1)
pooled_gap <- function(ds, m, nmax = 100) {
  s <- tr[tr$design %in% ds & tr$method == m & tr$n <= nmax, ]
  sqrt(mean(s$gap^2, na.rm = TRUE))
}
best_pooled <- function(ds) {
  g <- vapply(names(lambdas), function(m) pooled_gap(ds, m), numeric(1))
  list(lambda = lambdas[[which.min(g)]], gap = min(g), gaps = g)
}
bv <- best_pooled(var_designs)
bc <- best_pooled(cor_designs)
contenders <- c("psd", names(lambdas), "trunc_normal", "flat_mean", "jeff_mean", "jeff_median")
excess <- sapply(designs, function(d) {
  r <- sapply(contenders, function(m) rmse(d, n_small, m))
  r[["jeff_median"]] - min(r)
})
cov_range <- range(sapply(designs, function(d) sapply(n_values, function(n) cover(d, n, "jeff_mean"))))
path_done <- "path_r2_90" %in% designs
layer_gap <- function(m, layer, target = "jeff_mean", nmax = 100) {
  ref <- keyp[keyp$method == target, c("design", "n", "rep", "quantity", "est")]
  names(ref)[5] <- "ref"
  s <- merge(keyp[keyp$method == m & keyp$n <= nmax, ], ref,
             by = c("design", "n", "rep", "quantity"))
  s <- s[if (layer == "var") grepl("~~ x", s$quantity) else !grepl("~~ x", s$quantity), ]
  sqrt(mean((s$est - s$ref)^2, na.rm = TRUE))
}
if (has_bar) {
  rfit_gap <- {
    a <- keyp[keyp$method == "rfit_multi_l025", c("design", "n", "rep", "quantity", "est")]
    b <- keyp[keyp$method == "pen_l025", c("design", "n", "rep", "quantity", "est")]
    m <- merge(a, b, by = c("design", "n", "rep", "quantity"))
    max(abs(m$est.x - m$est.y))
  }
}
```

## Question

The multi-information barrier has one free number, its weight λ. There is a
natural benchmark for it: the objective Bayesian answer. Take the Jeffreys
prior, which depends on the model and not on how the latent variables are
scaled, restrict it to admissible solutions, and report the posterior mean or
median. That estimate never sits on a boundary and has nothing to tune. Does
the barrier behave like a cheap version of it, and at which λ? And is the
posterior itself a better estimator than the barrier?

## Short answer

The barrier tracks the objective posterior only if λ depends on the kind of
improper direction. On residual variances λ = `r bv$lambda` reproduces the
Jeffreys posterior mean almost exactly (root-mean-square distance
`r num(bv$gap)` at N ≤ 100). On factor correlations near one λ = `r bc$lambda`
is closest (`r num(bc$gap)`), and λ = 1 overshoots. The posterior pushes
variances well off the wall but moves correlations only a little. The default
λ = 0.25 moves both less than the posterior does.

The posterior is also the better estimator. At N = `r n_small` the Jeffreys
posterior median was the most accurate method or within
`r num(max(excess))` RMSE of it in every design. No single barrier weight
manages that. λ = 1 is the best weight on the Heywood-prone variances
(`r num(rmse("f1_p3", n_small, "pen_l100"))` against
`r num(rmse("f1_p3", n_small, "jeff_median"))` for the Jeffreys median) and the
worst on the r = .97 correlation (`r num(rmse("f2_r97", n_small, "pen_l100"))` against
`r num(rmse("f2_r97", n_small, "jeff_median"))`). Jeffreys 95% credible intervals
for the key quantity covered `r pct(cov_range[1])`–`r pct(cov_range[2])`% in every
cell. The cost is about `r sprintf("%.0f", sec_post)` s per dataset for the
sampler in R, against `r sprintf("%.0f", ms_pen)` ms for one barrier fit. The
cheap alternative that needs no λ, a normal around ML cut at the wall, fails on
exactly the samples it is meant for: its RMSE on the Heywood-prone design is
`r sprintf("%.1f", rmse("f1_p3", n_small, "trunc_normal"))`.
`r if (has_bar) paste0("A single-matrix alternative, the log-determinant of the latents' standardized covariance given the data (the latent-determinacy barrier), behaves like the multi-information barrier with slightly less shrinkage of correlations at the same weight (RMSE ", num(rmse("f2_r97", n_small, "det_l100")), " against ", num(rmse("f2_r97", n_small, "pen_l100")), " at λ = 1 and r = .97). It does not remove the tension: its best weight is still about 1 for residual variances and 0.5 for correlations. A two-weight multi-information barrier (1 on indicator terms, 0.5 on latent terms) tracks the posterior mean best overall.") else ""`

`r if (!path_done) "The path design (latent R²) has not been run yet; the correlation results above come from the two-factor designs alone." else ""`

## Evidence

### Which barrier weight tracks the Jeffreys posterior mean

Root-mean-square distance between each estimate and the Jeffreys posterior mean
of the key quantity, N ≤ 100. Rows split the key quantities by whether ML
landed past the admissible wall for that quantity (negative residual variance,
correlation or R² above one).

```{r tracking}
ms <- c(psd = "PSD-ML", pen_l010 = "λ = 0.1", pen_l025 = "λ = 0.25",
        pen_l050 = "λ = 0.5", pen_l100 = "λ = 1", trunc_normal = "Trunc. normal")
rows <- expand.grid(past = c(TRUE, FALSE), design = designs, stringsAsFactors = FALSE)
rows <- rows[mapply(function(d, p) any(tr$design == d & tr$ml_past == p & tr$n <= 100),
                    rows$design, rows$past), ]
tab <- data.frame(Design = design_label[rows$design],
                  `ML for this quantity` = ifelse(rows$past, "past the wall", "inside"),
                  Estimates = mapply(function(d, p) {
                    sum(tr$design == d & tr$ml_past == p & tr$method == "psd" & tr$n <= 100)
                  }, rows$design, rows$past),
                  check.names = FALSE, row.names = NULL)
for (m in names(ms)) {
  tab[[ms[[m]]]] <- mapply(function(d, p) num(gap_rms(d, p, m)), rows$design, rows$past)
}
knitr::kable(tab, align = paste0("llr", strrep("r", length(ms))))
```

### Accuracy against the truth

RMSE of the key quantity at N = `r n_small`. The key quantity is the residual
variance of the strong (loading .9) items in the one-factor designs, the factor
correlation in the two-factor designs, and R² of f3 in the path design.

```{r accuracy}
ms <- c(ml = "ML", psd = "PSD-ML", pen_l025 = "λ = 0.25", pen_l050 = "λ = 0.5",
        pen_l100 = "λ = 1", trunc_normal = "Trunc. normal", flat_mean = "Flat mean",
        jeff_mean = "Jeffreys mean", jeff_median = "Jeffreys median")
tab <- data.frame(Design = design_label[designs], `Key quantity` = key_label[designs],
                  check.names = FALSE, row.names = NULL)
for (m in names(ms)) tab[[ms[[m]]]] <- sapply(designs, function(d) num(rmse(d, n_small, m)))
knitr::kable(tab, align = paste0("ll", strrep("r", length(ms))))
```

```{r accuracy-large, results = "asis"}
n_big <- max(n_values)
if (n_big > n_small) {
  worst <- max(sapply(designs, function(d) {
    rmse(d, n_big, "jeff_mean") - min(sapply(c("psd", "pen_l025"), function(m) rmse(d, n_big, m)))
  }))
  cat(sprintf(paste0("At N = %d the Jeffreys mean is never worse than the better of PSD-ML ",
                     "and λ = 0.25 by more than %s on the key quantity. The full RMSE and ",
                     "bias by N are in `results/full/params.csv`."), n_big, num(max(worst, 0))))
}
```

### Credible-interval coverage

Coverage of the 95% equal-tailed posterior interval for the key quantity (%).

```{r coverage}
tab <- data.frame(Design = design_label[designs], check.names = FALSE, row.names = NULL)
for (n in n_values) {
  tab[[sprintf("Jeffreys, N = %d", n)]] <- sapply(designs, function(d) pct(cover(d, n, "jeff_mean")))
}
for (n in n_values) {
  tab[[sprintf("Flat, N = %d", n)]] <- sapply(designs, function(d) pct(cover(d, n, "flat_mean")))
}
knitr::kable(tab, align = paste0("l", strrep("r", 2 * length(n_values))))
```

```{r variants, results = "asis", eval = has_bar}
cat("### A single-matrix alternative: the latent-determinacy barrier\n\n")
cat(paste(
  "The multi-information barrier splits into a latent part, the observed",
  "correlations, and the information the indicators carry about the latents.",
  "Dropping the observed part leaves one log-determinant: the covariance of the",
  "standardized latents given all the data. It is singular exactly when some",
  "latent direction is fully determined, by the data (a Heywood case) or by",
  "other latents (a correlation of one). The table compares it with the",
  "multi-information barrier. Distances are root-mean-square distances to the",
  paste0("Jeffreys posterior mean at N ≤ 100; RMSE is against the truth at N = ",
         n_small, ".\n\n")))
ms <- c(pen_l025 = "Multi-information, λ = 0.25", pen_l050 = "Multi-information, λ = 0.5",
        pen_l100 = "Multi-information, λ = 1",
        two_weight = "Two-weight (1 indicators, 0.5 latent)",
        det_l025 = "Determinacy, λ = 0.25", det_l050 = "Determinacy, λ = 0.5",
        det_l100 = "Determinacy, λ = 1", det_l200 = "Determinacy, λ = 2",
        jeff_median = "Jeffreys median")
ms <- ms[names(ms) %in% unique(keyp$method)]
tab <- data.frame(Estimator = unname(ms),
                  `To posterior mean: variances` = sapply(names(ms), function(m) num(layer_gap(m, "var"))),
                  `To posterior mean: correlations` = sapply(names(ms), function(m) num(layer_gap(m, "cor"))),
                  check.names = FALSE, row.names = NULL)
for (d in designs) tab[[paste("RMSE,", design_label[[d]])]] <- sapply(names(ms), function(m) num(rmse(d, n_small, m)))
print(knitr::kable(tab, align = paste0("l", strrep("r", ncol(tab) - 1))))
```

## Caveats

- The sampler moves in standardized-latent coordinates with a flat prior there.
  The Jeffreys posterior is obtained by reweighting thinned draws, not by
  sampling it directly. Datasets whose chains disagreed, or whose reweighted
  sample was too thin, are excluded from every comparison
  (`r sum(!post$pass)` of `r nrow(post)`).
- Posterior summaries are reported only for quantities that do not depend on
  how the latent variables are scaled. Marker-scaled loadings can have an
  infinite posterior mean.
- "Flat" is flat in one particular coordinate system. The Jeffreys prior is the
  coordinate-free reference; the flat column shows how much the choice of prior
  matters.
- The barrier variants were fitted by a separate optimizer in standardized
  coordinates. `r if (has_bar) paste0("Refitting the default barrier this way from a different start reproduces magmaan's estimates to ", signif(rfit_gap, 2), ".") else ""`
- Data are multivariate normal and the models are correctly specified.
- `r max(fits$rep)` replications per cell. RMSE differences in the third decimal
  are within noise.

## Reproduce

```sh
just r-dev
Rscript experiments/research/86-multiinfo-jeffreys-posterior/run_experiment.R --full
Rscript experiments/research/86-multiinfo-jeffreys-posterior/scripts/determinacy.R --profile full
quarto render experiments/research/86-multiinfo-jeffreys-posterior/report.qmd
```
