---
title: "What breaks the Satorra–Bentler score test?"
subtitle: "Degrees of freedom, structure, and the eigenvalue spectrum — with a map of the test family"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({library(ggplot2); library(knitr)})
files <- file.path("results", c("df_sweep.csv", "structure.csv", "metadata.csv"))
if (!all(file.exists(files)))
  stop("missing results; run `Rscript run_experiment.R --full` (or --help)")
dfs <- read.csv(files[[1L]], stringsAsFactors = FALSE)
st  <- read.csv(files[[2L]], stringsAsFactors = FALSE)
meta <- read.csv(files[[3L]], stringsAsFactors = FALSE)
reps_df <- dfs$reps[1]; reps_st <- st$reps[1]
pretty <- c(score_chisq = "naive χ²", score_sb = "SB (mean-scaled)",
  score_mv = "mean+var", score_ss = "scaled-shifted", score_peba4 = "pEBA4",
  score_pall = "pEBA-all", score_sandwich = "sandwich", flip_basic = "flip (basic)",
  flip_effective = "flip (effective)", flip_standardized = "flip (standardized)")
```

## Question

The Satorra–Bentler mean-scaled statistic is the default robust score/LRT test
for a non-normal SEM restriction. In a companion calibration screen it was
almost suspiciously well behaved. This experiment asks **what, if anything,
actually degrades it**: the degrees of freedom of the restriction, the
structural information geometry of the model, or something else — and how it
compares with the spectrum-aware corrections and the sign-flip tests that target
the same null distribution.

## Short answer

The master variable is neither df nor structure but the **dispersion of the
test's eigenvalue spectrum**, summarized by its coefficient of variation
`CV(λ)`. Satorra–Bentler's error grows like `1 + CV(λ)²`, and under homogeneous
non-normality `CV(λ)` sits near 1 no matter what we do — so SB is remarkably
robust. **Lowering the degrees of freedom does not break it**: SB stays at the
nominal 5% across the whole range from `df = 1` to `df = 28`. **Clean structural
imbalance barely moves it** either; the one structural setting that nudges it
liberal (near-Heywood residuals) does so through boundary ill-conditioning, not
spectrum concentration, and it degrades the flip tests *more* than SB. The only
lever that would raise `CV(λ)` enough to matter is directional (heterogeneous)
kurtosis, which these clean designs deliberately exclude. Net: the corrected
score test is dependable across df and structure; the spectrum-aware and flip
variants earn their keep only in the concentrated-kurtosis corner.

## Evidence

### Degrees of freedom do not move Satorra–Bentler

Holding everything else clean (homogeneous `Φ = Θ = 1`, homogeneous excess
kurtosis 21 on every indicator), we vary only the restriction rank
`df = q · (G − 1)` — low df by testing one loading across few groups, high df by
testing several across eight. SB tracks the nominal 5% at every df; at `df = 1`
it is *exact* (a single eigenvalue leaves nothing to mean-scale), and it never
leaves the acceptable band. The flip tests run a hair liberal throughout, and
naive χ² (off the chart, rejecting 40–100%) is unusable.

```{r df-plot, fig.width = 7, fig.height = 4.2}
show <- c("score_sb", "score_peba4", "score_pall", "flip_effective")
d <- dfs[dfs$distribution == "vm", ]
long <- do.call(rbind, lapply(show, function(m)
  data.frame(df = d$df_obs, method = pretty[[m]], rate = d[[m]])))
long$method <- factor(long$method, levels = pretty[show])
ggplot(long, aes(df, rate, colour = method)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = .025, ymax = .075,
           alpha = .08, fill = "grey40") +
  geom_hline(yintercept = .05, linetype = 2, colour = "grey40") +
  geom_line(alpha = .5) + geom_point(size = 2) +
  scale_x_continuous(breaks = sort(unique(long$df))) +
  coord_cartesian(ylim = c(0, .12)) +
  labs(x = "restriction df", y = "rejection rate at α = .05", colour = NULL,
       caption = sprintf("Vale–Maurelli, homogeneous kurtosis; %s reps/cell. Band = 2.5–7.5%%.", reps_df)) +
  theme_minimal(base_size = 12) + theme(legend.position = "top")
```

### Clean structure barely moves it; near-Heywood is a fragile, shared failure

Fixing `df = 28` and pushing the model's information geometry — from the
alternating-group "geometry" regime to a concentrated near-Heywood structure —
moves the eigenvalue `CV` only modestly. SB drifts from nominal to mildly liberal
only in the concentrated regime, and there the flip tests degrade at least as much
while the pEBA corrections stay conservative. On normal data the identical
concentrated structure leaves every test at nominal, confirming there is no
structural lever without non-normality.

```{r structure-table}
cols <- c("structure", "distribution", "eigen_cv", "score_sb", "score_peba4",
          "score_pall", "flip_effective", "flip_standardized", "score_chisq")
tab <- st[, cols]
names(tab) <- c("Structure", "Data", "CV(λ)", "SB", "pEBA4", "pEBA-all",
                "flip-eff", "flip-std", "naive χ²")
kable(tab, digits = 3, caption = sprintf(
  "Rejection at α = .05, df = 28, p = 20, %s reps/cell. Target .050.", reps_st))
```

## What the tests are

*Heuristic, no proofs; the aim is to say what each statistic is and why they
differ. The lineage runs through the SEM, semiparametric M-estimation, and GEE
literatures, which are the same construction seen from three angles.*

### One object underlies all of them

With sample covariance `S`, model `Σ(θ)`, and `s = vech(S)`, `σ(θ) = vech(Σ(θ))`,
the sample moments obey $\sqrt{n}\,(s-\sigma(\theta_0)) \xrightarrow{d}
\mathcal N(0,\Gamma)$. Under normality $\Gamma = \Gamma_N =
2D^{+}(\Sigma\otimes\Sigma)D^{+\prime}$; in general $\Gamma$ carries fourth-order
cumulants (excess kurtosis) — Browne's ADF matrix. Normal-theory ML behaves near
the optimum like GLS with weight $W = \tfrac12 D'(\Sigma^{-1}\otimes\Sigma^{-1})D
\propto \Gamma_N^{-1}$, and with Jacobian $\Delta = \partial\sigma/\partial\theta$.
For a rank-$r$ restriction ($r = \text{df}$), the LRT, Lagrange-multiplier
(score), and Wald statistics are asymptotically one quadratic form, and under
$H_0$

$$
T_n \ \xrightarrow{d}\ \sum_{i=1}^{r}\lambda_i\,\chi^2_{1,i},\qquad
\lambda_i = \text{eigenvalues of } U\Gamma,\quad
U = W - W\Delta(\Delta'W\Delta)^{-1}\Delta'W .
$$

Everything is in the $\lambda_i$. If the working weight matches the truth
($W \propto \Gamma^{-1}$, e.g. normal data), $U\Gamma$ is a projector, the
$\lambda_i$ are $r$ ones and the rest zero, and $T_n \to \chi^2_r$ — the textbook
case. Non-normality keeps the normal-theory $W$ while $\Gamma \neq \Gamma_N$, so
the $\lambda_i$ spread and $\mathbb E[T_n] = \sum_i\lambda_i = \operatorname{tr}(U\Gamma) > r$:
naive χ² over-rejects. Two scalars govern the corrections:
$m_1 = \sum_i\lambda_i$, $m_2 = \sum_i\lambda_i^2$, and
$\mathrm{CV}(\lambda) = \mathrm{sd}(\lambda)/\bar\lambda$.

This is exactly a **score test under a working model**. The estimating equations
$\Delta'W(s-\sigma(\theta)) = 0$ and the sandwich law of $\hat\theta$ are the
misspecified-likelihood results of White (1982) and Godambe's estimating-function
theory; the nuisance projection in $U$ is the semiparametric **effective score**
(Bickel–Klaassen–Ritov–Wellner; Tsiatis). In GEE the normal-theory $W$ *is* the
working correlation, and the **generalized score test** (Boos 1992) with the
cluster-GLM correction of Rotnitzky & Jewell (1990) is the GEE twin of
Satorra–Bentler — same weighted sum of χ², same mean/variance adjustments.

### The corrections, by how much of the spectrum they use

| test (code) | reference | uses |
|---|---|---|
| naive χ² (`score_chisq`) | $\chi^2_r$ | nothing ($\lambda_i\equiv1$) |
| **SB** (`score_sb`) | $T_n/c \sim \chi^2_r,\ c = m_1/r$ | mean |
| mean+var / scaled-shift (`score_mv`,`score_ss`) | $\chi^2_d,\ d = m_1^2/m_2$ | first two moments |
| pEBA / Imhof (`score_peba4`,`score_pall`) | $\Pr(\sum\hat\lambda_i\chi^2_1 > T_n)$ | full spectrum |
| flip (`flip_*`) | sign-flip resample | spectrum implicitly, no $\widehat\Gamma$ |

- **Satorra–Bentler** divides by $c = \bar\lambda$ so the reference has the right
  mean. Its only error is the variance it misses:
  $\mathrm{Var}(T_n)/\mathrm{Var}(c\,\chi^2_r) = r\,m_2/m_1^2 = 1 + \mathrm{CV}(\lambda)^2$.
  Exact when the eigenvalues are equal ($\mathrm{CV}=0$; e.g. $r=1$); degrades only
  as the spectrum disperses. This identity is the whole empirical story above.
- **Mean-and-variance / scaled-shifted** (Satorra–Bentler 1994; Asparouhov–Muthén
  2010) also matches $m_2$ through a Satterthwaite fractional df — recovering the
  tail SB misses once $\mathrm{CV}$ is large.
- **pEBA / Imhof** plug in the estimated spectrum and evaluate the actual
  weighted-χ² tail (Imhof 1961; Davies 1980); pEBA uses a leading block of
  eigenvalues plus a tail penalty for stability. Most accurate — when $\widehat\Gamma$
  is well estimated.
- **Sandwich** ($V_{\text{sw}} = (\Delta'W\Delta)^{-1}\Delta'W\Gamma W\Delta(\Delta'W\Delta)^{-1}$)
  is the SEM Huber–White / Liang–Zeger robust covariance; excellent for standard
  errors, but a single sandwich χ² still treats the $\lambda_i$ as equal, so as a
  *test* it behaves like an unscaled family member.

### Flip tests: resample the null, never estimate $\Gamma$

Write the effective score as a sum of per-case contributions,
$U_n = n^{-1/2}\sum_i \psi_i$ with $\mathbb E_{H_0}[\psi_i] = 0$, where $\psi_i$ is
the case-$i$ influence for the tested constraint after projecting out the nuisance
parameters. Draw Rademacher signs $\varepsilon_i\in\{\pm1\}$ and recompute
$U_n^\varepsilon = n^{-1/2}\sum_i \varepsilon_i\hat\psi_i$. Flipping kills the mean
but preserves the covariance, so the flipped statistics reproduce the same
$\sum\lambda_i\chi^2_1$ law **without ever forming $\widehat\Gamma$**; the p-value
is $(1+\#\{T_n^\varepsilon \ge T_n\})/(1+B)$. The variants trade robustness for
accuracy: `basic` flips the raw score; `effective` flips the nuisance-orthogonalized
(Neyman-orthogonal) score; `standardized` additionally studentizes each flip for a
pivotal statistic. This is the wild / sign-flip bootstrap (Wu 1986; Liu 1988;
Mammen 1993) specialized to score contributions, exactly as in Hemerik–Goeman–Finos
(2020) for GLMs, and the SEM instance is the Foldnes–Grønneberg–Moss flip test.

### How to hold them in your head

Every test references the *same* null law $\sum_i\lambda_i\chi^2_{1,i}$ and differs
only in how much of the spectrum it uses: naive (0 moments) → SB (1) → mean+var (2)
→ pEBA/Imhof (all) → flip (all, implicitly, $\widehat\Gamma$-free). Under homogeneous
kurtosis $\mathrm{CV}(\lambda)\approx1$, SB's $1+\mathrm{CV}^2$ penalty is small, which
is why df and clean structure cannot dislodge it; only concentrated, directional
kurtosis raises $\mathrm{CV}$ enough for the richer references to matter.

## Caveats

- Reps are modest (see the caption); cell rejection rates carry visible Monte
  Carlo scatter, so read trends, not third decimals. The df result (flat) and the
  structure result (no clean lever) are the robust conclusions.
- The kurtosis here is **homogeneous** across indicators by design — that is what
  caps `CV(λ)`. The complementary study, heterogeneous/directional kurtosis, is
  the one expected to move SB and is deliberately out of scope.
- The concentrated structure leans on near-Heywood residuals; a few fits fail to
  converge there and are dropped, so that column is the least clean.
- These are calibration (Type-I) results under a true equality null; power is not
  studied here.

## Reproduce

```sh
Rscript experiments/68-score-test-calibration-drivers/run_experiment.R --help
Rscript experiments/68-score-test-calibration-drivers/run_experiment.R          # smoke
Rscript experiments/68-score-test-calibration-drivers/run_experiment.R --full   # full pilot
quarto render experiments/68-score-test-calibration-drivers/report.qmd
```

## References

*SEM.* Satorra (1989, *Psychometrika*); Satorra & Bentler (1988 ASA Proc.; 1994,
*Latent Variables Analysis*); Browne (1984, *BJMSP*); Shapiro (1986, *JASA*);
Sörbom (1989, *Psychometrika*); Asparouhov & Muthén (2010); Maydeu-Olivares (2017,
*Psychometrika*).
*Semiparametric / M-estimation.* Godambe (1960); White (1982, *Econometrica*);
Bickel, Klaassen, Ritov & Wellner (1993); Tsiatis (2006).
*GEE.* Liang & Zeger (1986, *Biometrika*); Rotnitzky & Jewell (1990, *Biometrika*);
Boos (1992, *Am. Stat.*).
*Quadratic forms in normals.* Satterthwaite (1946); Box (1954); Imhof (1961,
*Biometrika*); Davies (1980, *Appl. Stat.*).
*Sign-flip / wild bootstrap.* Wu (1986); Liu (1988); Mammen (1993);
DiCiccio & Romano (2017, *JASA*); Hemerik, Goeman & Finos (2020, *JRSS-B*).
*SEM flip test / pEBA.* Foldnes, Grønneberg & Moss — fill exact citation(s) and
the pEBA reference.
