---
title: "Robust test calibration battery: global and nested tests on textbook models"
params:
  run: frozen
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
path <- file.path("results", params$run)
has_results <- file.exists(file.path(path, "summary.csv"))
source("R/dgp.R")
source("R/populations.R")
pops <- read_populations("populations")
if (has_results) {
  s <- read.csv(file.path(path, "summary.csv"))
  fails <- read.csv(file.path(path, "failures.csv"))
  meta <- read.csv(file.path(path, "metadata.csv"))
  get_meta <- function(k) meta$value[match(k, meta$key)]
}
checks_path <- file.path(path, "checks.csv")
if (!file.exists(checks_path)) checks_path <- file.path("results", "validate", "checks.csv")
checks <- if (file.exists(checks_path)) read.csv(checks_path) else NULL
primary <- c(
  lr_own_std = "LR, chi-square", lr_own_sb = "LR, SB", lr_own_mv = "LR, MV",
  lr_unbiased_mv = "LR, MV-UG", lr_rows_cmv = "LR, corrected MV", lr_own_peba4 = "LR, pEBA4",
  score_own_sb = "Score, SB", score_own_mv = "Score, MV", score_unbiased_mv = "Score, MV-UG",
  score_rows_cmv = "Score, corrected MV", score_own_peba4 = "Score, pEBA4",
  rls_biased_sb = "RLS, SB", rls_biased_peba4 = "RLS, pEBA4")
dgp_label <- c(normal = "Normal", vm = "Vale-Maurelli (2, 7)", ig = "IG (2, 7)",
               disc = "Discretized, 5-point")
```

## Question

Across a varied set of textbook SEMs, sample sizes and non-normal data,
including discretized 5-point items analysed by ML, which calibration of which
base statistic keeps nominal Type I error, for global fit and for nested
restrictions? The candidates are the ordinary-user policy's score and
likelihood-ratio statistics and lavaan's RLS statistic, calibrated with
Satorra-Bentler mean scaling (SB), Satterthwaite mean-and-variance adjustment
(MV), MV with Du and Bentler's unbiased Gamma (MV-UG), the corrected MV of
`research/49`, and pEBA4.

## Short answer

**The score statistic beats the likelihood-ratio statistic, and nested score
tests are essentially exact.** These results cover seven textbook models,
$n$ of 100, 300 and 1000, and four data generators, with 5,000 replications
per cell.

- **Nested restrictions** (df 1-18, 48 cells). The policy's score test,
  calibrated by SB on its own null-fit spectrum, rejected between 3.6% and
  5.8% at nominal 5% in every cell, under every generator; pEBA4 gave
  2.8-5.5%. With the same calibrations the LR difference reached 6.6-17.7% at
  $n = 100$. lavaan's RLS-difference base reached 5.5-15.3%.
- **Global fit** (84 cells). Under normal and discretized data, score SB
  rejected 4.3-6.2% and score pEBA4 3.2-6.0%. LR-SB and LR-pEBA4 over-reject
  at $n = 100$ in models with many degrees of freedom, up to 20.8% and 15.8%
  (the 18-indicator longitudinal CFA, df 102); by $n = 1000$ LR-SB is within
  4.9-6.0%.
- **IG data (heterogeneous spectra) remain unsolved for global fit.** The best
  arm is score pEBA4, with a mean absolute deviation from 5% of 1.6 points and
  a worst cell of 12.3% (two-group path model, $n = 100$). Score SB reaches
  14.1%.
- **The MV family does not help the score.**
  - Plain MV and MV with the unbiased Gamma are conservative on the score
    statistic. Their median rejection is 4% under normal data and 2% under
    Vale-Maurelli and IG, with cells as low as 0.5%.
  - On LR, MV removes most of the small-sample inflation under normal and
    discretized data, but under-rejects under non-normal continuous data.
  - The corrected MV of `research/49` behaves like SB. It is slightly better
    under IG (mean deviation 2.1 against 2.6 points) and does not beat pEBA4.
- **5-point discretized data calibrate like normal data.** For every arm the
  mean deviation under Pearson-calibrated 5-point data is within 0.1 points
  of its normal-data value.

For the ordinary-user policy, this supports the score test under both SB and
pEBA4. Its LR companion is the weak half at small $n$ with many degrees of
freedom.

## Evidence

```{r results-note}
#| results: asis
if (!has_results) cat("No run in `results/", params$run, "/` yet.\n", sep = "")
```

```{r plot-fn}
plot_arms <- function(test, subset = "converged") {
  z <- s[s$subset == subset & s$test == test & s$arm %in% names(primary), ]
  arms <- rev(names(primary))
  ns <- sort(unique(z$n))
  cols <- setNames(c("#86b6ef", "#2a78d6", "#104281")[seq_along(ns)], ns)
  pchs <- setNames(c(1, 17, 15)[seq_along(ns)], ns)
  reps <- median(z$valid)
  band <- 1.96 * sqrt(.05 * .95 / reps)
  old <- par(mfrow = c(1, 4), mar = c(3.6, 0.5, 2.2, 0.6), oma = c(0, 9.5, 0, 0),
             mgp = c(2.2, .6, 0), cex = .8, col.axis = "#52514e", fg = "#c3c2b7")
  on.exit(par(old))
  xmax <- max(.12, min(.35, max(z$reject_05)))
  for (g in names(dgp_label)) {
    w <- z[z$dgp == g, ]
    plot(NA, xlim = c(0, xmax), ylim = c(.5, length(arms) + .5), yaxt = "n",
         xlab = "Rejection at 5%", ylab = "", main = dgp_label[[g]], font.main = 1,
         col.main = "#0b0b0b")
    rect(.05 - band, 0, .05 + band, length(arms) + 1, col = "#f0efec", border = NA)
    abline(v = .05, col = "#898781", lty = 2)
    abline(h = seq_along(arms), col = "#e1e0d9", lwd = .6)
    for (k in seq_along(arms)) {
      a <- w[w$arm == arms[k], ]
      jitter <- (match(a$n, ns) - 2) * .18
      points(pmin(a$reject_05, xmax), k + jitter, col = cols[as.character(a$n)],
             pch = pchs[as.character(a$n)], cex = .8)
      segments(median(a$reject_05), k - .35, median(a$reject_05), k + .35, lwd = 2, col = "#0b0b0b")
    }
    if (g == names(dgp_label)[1]) axis(2, at = seq_along(arms), labels = primary[arms],
                                       las = 1, outer = TRUE, col.axis = "#0b0b0b")
    if (g == names(dgp_label)[4]) legend("bottomright", legend = paste("n =", ns),
      col = cols, pch = pchs, bty = "n", cex = .85, text.col = "#52514e")
  }
}
```

### Global fit

LR calibrations fan out to the right at $n = 100$, while the score arms stay
near 5% except under IG. Normal and discretized panels are nearly identical.

```{r gof-plot}
#| fig-width: 12
#| fig-height: 5.2
#| fig-cap: "Global fit: rejection at nominal 5% per cell (7 cases per sample size), coloured and shaped by n. The vertical bar is the median over cells, the grey band is +/-1.96 Monte Carlo SE around 5%, and values above the axis limit are drawn at the edge."
if (has_results) plot_arms("gof")
```

### Nested tests

The score arms (SB, pEBA4) sit on 5% in every generator. LR and the RLS
difference over-reject at $n = 100$, most under IG and Vale-Maurelli in the
growth model.

```{r nested-plot}
#| fig-width: 12
#| fig-height: 5.2
#| fig-cap: "Nested tests (four cases): rejection at nominal 5% per cell, as above."
if (has_results) plot_arms("nested")
```

```{r failures}
#| results: asis
if (has_results) {
  bad <- fails[fails$outcome != "valid", ]
  total <- sum(fails$rep)
  cat(sprintf("%d of %d replicate tests were unusable (%s). Per-cell rejection rates, the admissible-only subset and per-arm summaries are in `summary.csv`, `arm_summary.csv` and `failures.csv` under `results/%s/`.\n",
              sum(bad$rep), total,
              if (nrow(bad)) paste(sprintf("%s: %d", names(tapply(bad$rep, bad$outcome, sum)), tapply(bad$rep, bad$outcome, sum)), collapse = "; ") else "none",
              params$run))
}
```

## Caveats

- Rejection rates condition on converged fits. At $n = 100$, 9.4% of
  second-order CFA fits and about 2% of Worland SEM fits failed to converge;
  elsewhere failures were negligible. Restricting to admissible fits changes
  no conclusion (per-arm summaries for both subsets are in
  `results/frozen/arm_summary.csv`).
- This is an exploratory null-size battery. It says nothing about power, and
  a final run may add generators first (see the mean-preserving discretizers
  entry in `project/backlog/speculative.md`).
- The discretized arm uses a Gaussian copula. Its non-normality is specific to
  thresholding a normal vector.
- Populations inherit the corpus models' parameter values, including high
  communalities.
- The corrected MV's moment estimators are unbiased for fixed transforms of iid
  rows, not for rows built on estimated SEM projections (as in `research/49`).

## Reproduce

From the repository root, with `magmaanlab` installed:

```sh
# once, where the textbook corpus is mounted (already done; output is tracked)
Rscript experiments/research/52-robust-calibration-battery/R/build_populations.R
# lavaan checks (needs lavaan)
OPENBLAS_NUM_THREADS=1 Rscript experiments/research/52-robust-calibration-battery/run_experiment.R --validate
# smoke, then the main run (workers are forked processes)
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 nice -n 10 Rscript experiments/research/52-robust-calibration-battery/run_experiment.R --smoke --workers 4
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 nice -n 10 Rscript experiments/research/52-robust-calibration-battery/run_experiment.R --main --workers 4
quarto render experiments/research/52-robust-calibration-battery/report.qmd
```

## Design

### Why this study

The ordinary-user policy reports a global score test and a likelihood-ratio
test against the saturated model, each calibrated by SB and pEBA4, and the same
pair for nested tests. `research/45` compared score and LR under SB and pEBA on one two-factor
population with 10 or 20 indicators. This battery widens that evidence in
four directions: model types from the textbook corpus, discretized data (the
typical shape of real item data), Satterthwaite-type calibrations, and nested
tests.

The three base statistics are closely related on complete normal-theory data.
With means absent or saturated, the policy's expected-information score
statistic equals lavaan's RLS statistic exactly. With restricted means they
differ at $O_p(n^{-1/2})$, because the score's residual is taken about the
fitted mean. The nested score statistic equals lavaan's score test. The nested
RLS base used by lavaan's single-group FMG tests
is the difference of two RLS statistics. It shares the score's first term, but
its second term is evaluated at the alternative's fit, and it is not a
quadratic form (it can be negative).

### Models

Each population is the model-implied mean and covariance of a textbook-corpus
model, fitted by lavaan ML to the book's data. Where a nested restriction is
tested, the population is the restricted model's fit, so both hypotheses hold
exactly. The populations are stored with the experiment, so a run does not
need the corpus.

```{r models}
tab <- do.call(rbind, lapply(pops, function(p) data.frame(
  Case = p$id, Source = p$corpus, Model = p$type,
  p = length(p$groups[[1]]$ov), Groups = length(p$groups),
  `GOF df` = as.integer(p$source_fit$h1$df),
  `Nested restriction` = if (is.null(p$h0)) "none" else p$restriction,
  check.names = FALSE)))
knitr::kable(tab, row.names = FALSE)
```

The nested restrictions are natural single-model hypotheses, not
measurement-invariance steps. The two-group path model's global test is itself
a test of seven cross-group equalities, because each group's path model is
saturated. Its variables are rescaled by their pooled population standard
deviation, the same factor in every group. Every tested hypothesis is invariant
to that. The rescaling only avoids a numerical failure of the nested
Satorra (2000) spectrum on badly scaled variables, logged as a bug.

### Data-generating processes

Every generator reproduces each group's population mean and covariance exactly.
A check on 200,000 draws per case put the largest standardized moment error
below 0.02, which is Monte Carlo error.

- **Normal.**
- **Vale-Maurelli (2, 7):** Fleishman margins with skewness 2 and excess
  kurtosis 7.
- **IG (2, 7):** the Foldnes-Olsson independent generator with the same
  marginal targets, Pearson marginal family, and Cholesky root. The Cholesky
  root gives heterogeneous spectra (`research/49`).
- **Discretized, 5-point:** a Gaussian copula thresholded at fixed category
  probabilities. The latent correlations are calibrated so that the Pearson
  correlations of the integer codes equal the population correlations. Each
  item is then
  mapped affinely to the population mean and SD. Items keep 5 values and are
  analysed by ML. No polychoric step is involved anywhere.
  - Default marginals alternate a symmetric shape (.10, .20, .40, .20, .10)
    and a right-skewed one (.35, .30, .18, .11, .06).
  - The CES-D growth model uses the skewed shape at every wave.

  No latent correlation matrix needed repair.

### Statistics, spectra and calibrations

Each replicate fits H1 (and H0) by ML. It then records three base
statistics:

| Base | Global | Nested |
|---|---|---|
| LR | $2N\,F_{ML}$ | difference of the two |
| Score | policy score (expected information, null fit) | policy nested score (equals lavaan's score test) |
| RLS | lavaan's RLS test | difference of the two RLS statistics |

Each statistic is calibrated against the eigenvalues of $U\Gamma$ from three
sources:

- **own:** the policy's spectrum for that statistic. This is the score's rows
  at the null fit, or the LR's centred rows.
- **biased:** empirical $\hat\Gamma$ at the fitted H1 (global), or Satorra's
  (2000) exact restriction map at H1 (nested). All three bases share it.
- **unbiased:** the same with Du and Bentler's unbiased $\hat\Gamma$.

The calibrations are:

- chi-square with the model df;
- SB mean scaling;
- MV, Satterthwaite with fractional df;
- pEBA4, penalized eigenvalue block averaging with 4 blocks;
- corrected MV (`research/49`), extended to several groups. It estimates
  $\operatorname{tr}V$ and $\operatorname{tr}V^2$ without bias from the policy
  score's casewise rows, not from a spectrum.

All combinations are stored. The report focuses on the arms listed in the
figures.

### Sample sizes, replications and seeds

Total $n \in \{100, 300, 1000\}$, split by the source group proportions in the
two-group case. Seven cases, four generators and three sample sizes give 84
cells. Each cell's seeds are `seed_base + 100000 * cell_id + rep` in the
canonical grid. The second group adds 50000. Calibration objects are built
once per cell.

### Outcomes

- The outcome is rejection at nominal 1%, 5% and 10%.
- A replicate counts for a test when the fits converged (the library verdict)
  and every arm returned a p-value, so arms are compared on common valid draws.
- Converged but inadmissible fits are kept (this is what users get), with an
  admissible-only subset reported alongside.
- Failures are tabulated by cause.

### Validation

One draw per case under normal and discretized data is compared with
lavaan. The checks cover:

- LR, RLS, and lavaan's SB and mean-and-variance adjusted p-values;
- the nested LR difference;
- the nested score, against lavaan's score test for the added constraints;
- the nested RLS difference;
- the nested Satorra (2000) SB p-value, with the exact restriction map;
- internal identities: the LR spectrum is shared, the global score equals RLS
  when means are free, and each score statistic is rebuilt from its casewise
  terms.

```{r checks}
#| results: asis
if (is.null(checks)) cat("Validation not run in this checkout.\n") else
  cat(sprintf("**%d of %d checks passed.**\n", sum(checks$passed), nrow(checks)))
```

