---
title: "Reliability coefficients and their targets"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Question

Which population quantity do covariance coefficients and fitted omega target,
and how much can misspecification change their relationship to reliability?

## Short answer

**A coefficient's probability limit is not automatically generating
reliability.** In the correct one-factor example, SG covariance omega and
fitted omega recover unit-score reliability (0.843), while alpha and lambda6
have different population values. Fitting one factor to the two-factor
example gives omega 0.713, versus generating reliability 0.820; the covariance
functionals have other limits. A confidence interval can accurately target a
functional that is not the reliability the analyst intended.

The broader Bell-inspired population examples demonstrate this distinction
across misspecified structures and plausible CFA choices. These deterministic
examples show target differences without finite-sample noise. They do not
estimate coverage, sampling bias or which model an analyst can identify from data.

## Evidence

```{r setup}
read_result <- function(lane, name) {
  p <- file.path("results", lane, name)
  if (!file.exists(p)) stop("Missing ", p, "; run the corresponding lane before rendering.")
  read.csv(p, stringsAsFactors = FALSE)
}
targets <- read_result("coefficients", "population_targets.csv")
bias <- read_result("bell", "omega_bias.csv")
sensitivity <- read_result("bell", "cfa_sensitivity_summary.csv")
```

### Covariance functionals versus a fitted one-factor target

Both six-indicator populations have unit marginal variances. The one-factor
loadings are .85/.75/.70/.65/.60/.55. In the second example loadings form two
blocks (.85/.75/.65 and .80/.70/.60), with factor correlation .35.
Generating reliability is common unit-weighted total variance divided by
observed total variance. Covariance-only coefficients are evaluated directly
at the population covariance; fitted omega uses the ML one-factor pseudo-target.

```{r}
knitr::kable(data.frame(Population = targets$population_label,
  Coefficient = targets$method, `Population value` = targets$target_functional,
  `Generating reliability` = targets$true_reliability,
  Difference = targets$target_functional-targets$true_reliability,
  check.names = FALSE), digits = 3, row.names = FALSE)
```

The demo retains the populations from former research/16, with one fitted-omega
point target instead of duplicate rows labelled by SE method. SE conventions
do not change that target. The old small sampling pilot is retained under
`results/sampling-history/` for provenance; it is not used as calibration
evidence. Formulas, gradients and delta calculations are maintained in tests.

### Bell-inspired misspecification and sensitivity

The second lane evaluates functionals at population covariance matrices for
one-factor, two-factor, bifactor and residual-dependence structures, at low
and high reliability. Its truth is reliability attributable to the generating
main factor, distinct from the preceding demo's total common-score target.
It compares one-factor ML/ULS/GLS omega, observed-total
and model-implied denominators, and SG covariance omega. The table shows worst
absolute population differences for primary ML and SG variants, by structure.

```{r}
z <- bias[(bias$estimator == "ML" & bias$omega_variant == "omega_u_model_implied") |
            bias$estimator == "closed_form", ]
tab <- do.call(rbind, lapply(split(z, interaction(z$population, z$estimator)), function(x) {
  w <- x[which.max(abs(x$bias)), ]
  data.frame(Population = w$population_label, Estimator = w$estimator,
    `Reliability setting` = w$reliability_level, `Worst absolute difference` = abs(w$bias),
    check.names = FALSE)
}))
knitr::kable(tab, digits = 3, row.names = FALSE)
```

A second set fits several CFA structures to each covariance. Their spread is
a historical model-sensitivity illustration, not a confidence interval.

```{r}
z <- sensitivity[sensitivity$reliability_level == "low", ]
knitr::kable(data.frame(Population = z$population_label,
  `Generating reliability` = z$true_reliability,
  `One-factor difference` = z$one_factor_bias,
  `Omega range across models` = z$omega_range,
  `Smallest absolute difference` = z$min_abs_bias, check.names = FALSE),
  digits = 3, row.names = FALSE)
```

Main-factor reliability, total common-score reliability and fitted one-factor
omega must be distinguished. The `congeneric_diagnostics.csv` controls inspect
rank-one/tetrad closeness; a small discrepancy or CFA spread alone does not
identify the desired reliability quantity.

## Caveats

The Bell lane is a population reconstruction and extension inspired by Bell
et al. (2024), rather than a finite-N replication. Its exact population formulas
and estimator/denominator definitions remain in the runner. Population fitting
uses a large nominal n to evaluate the ML limit, not a million-observation
Monte Carlo experiment. Different estimators' discrepancy numbers are not
interchangeable fit tests.

A fresh population smoke on the current installed package reproduced the
main one-factor/SG comparison within $10^{-6}$ and its generating targets
exactly, but some alternative CFA fits failed or reached different estimates.
The model-sensitivity table above therefore reports the frozen historical
fits, not a verified current-package model-selection result. Before using
that extension for a new claim, revalidate convergence, admissibility and
objective agreement for each alternative CFA. The compact reproduction audit
is retained under `results/verification/`; full fresh outputs stay local. Changing the model changes the functional even
with valid standard errors. Neither lane settles near-boundary interval calibration.

## Reproduce

```sh
Rscript run_experiment.R --lane coefficients --smoke
Rscript run_experiment.R --lane bell --smoke --results-dir /tmp/reliability-bell
quarto render report.qmd
```

Coefficients writes targets under `results/coefficients/`. Bell outputs came
from research/17 and remain under `results/bell/`. The Bell smoke command
computes the same deterministic population study in a fresh directory. Use
`--lane NAME --help` for commands; rendering only reads local CSVs.
