---
title: "Reliability coefficients and their targets"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Question

Which population quantity do covariance coefficients and fitted omega target,
and how much can misspecification change their relationship to reliability?

## Short answer

**A coefficient's probability limit is not automatically generating
reliability.** In the correct one-factor example, SG covariance omega and
fitted omega recover unit-score reliability (0.843), while alpha and lambda6
have different population values. Fitting one factor to the two-factor
example gives omega 0.713, versus generating reliability 0.820; the covariance
functionals have other limits. A confidence interval can accurately target a
functional that is not the reliability the analyst intended.

The broader Bell-inspired population examples demonstrate this distinction
across misspecified structures and plausible CFA choices. These deterministic
examples show target differences without finite-sample noise. They do not
estimate coverage, sampling bias or which model an analyst can identify from data.
The bounded current-package check of five low-reliability controls returned five
converged, admissible endpoints with matching re-evaluated objectives. This
validates those endpoints only; reference optima and identification remain unresolved.

## Evidence

```{r setup}
read_result <- function(lane, name) {
  p <- file.path("results", lane, name)
  if (!file.exists(p)) stop("Missing ", p, "; run the corresponding lane before rendering.")
  read.csv(p, stringsAsFactors = FALSE)
}
targets <- read_result("coefficients", "population_targets.csv")
bias <- read_result("bell", "omega_bias.csv")
sensitivity <- read_result("bell", "cfa_sensitivity_summary.csv")
```

### Covariance functionals versus a fitted one-factor target

Both six-indicator populations have unit marginal variances. The one-factor
loadings are .85/.75/.70/.65/.60/.55. In the second example loadings form two
blocks (.85/.75/.65 and .80/.70/.60), with factor correlation .35.
Generating reliability is common unit-weighted total variance divided by
observed total variance. Covariance-only coefficients are evaluated directly
at the population covariance; fitted omega uses the ML one-factor pseudo-target.

```{r}
knitr::kable(data.frame(Population = targets$population_label,
  Coefficient = targets$method, `Population value` = targets$target_functional,
  `Generating reliability` = targets$true_reliability,
  Difference = targets$target_functional-targets$true_reliability,
  check.names = FALSE), digits = 3, row.names = FALSE)
```

The demo retains the populations from former research/16, with one fitted-omega
point target instead of duplicate rows labelled by SE method. SE conventions
do not change that target. The old small sampling pilot is retained under
`results/sampling-history/` for provenance; it is not used as calibration
evidence. Formulas, gradients and delta calculations are maintained in tests.

### Bell-inspired misspecification and sensitivity

The second lane evaluates functionals at population covariance matrices for
one-factor, two-factor, bifactor and residual-dependence structures, at low
and high reliability. Its truth is reliability attributable to the generating
main factor, distinct from the preceding demo's total common-score target.
It compares one-factor ML/ULS/GLS omega, observed-total
and model-implied denominators, and SG covariance omega. The table shows worst
absolute population differences for primary ML and SG variants, by structure.

```{r}
z <- bias[(bias$estimator == "ML" & bias$omega_variant == "omega_u_model_implied") |
            bias$estimator == "closed_form", ]
tab <- do.call(rbind, lapply(split(z, interaction(z$population, z$estimator)), function(x) {
  w <- x[which.max(abs(x$bias)), ]
  data.frame(Population = w$population_label, Estimator = w$estimator,
    `Reliability setting` = w$reliability_level, `Worst absolute difference` = abs(w$bias),
    check.names = FALSE)
}))
knitr::kable(tab, digits = 3, row.names = FALSE)
```

A second set fits several CFA structures to each covariance. Their spread is
a historical model-sensitivity illustration, not a confidence interval.

```{r}
z <- sensitivity[sensitivity$reliability_level == "low", ]
knitr::kable(data.frame(Population = z$population_label,
  `Generating reliability` = z$true_reliability,
  `One-factor difference` = z$one_factor_bias,
  `Omega range across models` = z$omega_range,
  `Smallest absolute difference` = z$min_abs_bias, check.names = FALSE),
  digits = 3, row.names = FALSE)
```

Main-factor reliability, total common-score reliability and fitted one-factor
omega must be distinguished. The `congeneric_diagnostics.csv` controls inspect
rank-one/tetrad closeness; a small discrepancy or CFA spread alone does not
identify the desired reliability quantity.

### Bounded current-package control validation

The fresh `verification/task-115-population-v1` inventory fixes the existing
low-reliability higher-order population and its four CFA candidates, plus the
low-reliability one-factor population with one-factor CFA. Each uses ML,
`std_lv=TRUE`, the existing population covariance and nominal n=1,000,000;
all other fitting options are omitted. No draws, retries or optimum searches
were performed. All five attempts returned: the primitive covariance blocks
were PSD, the implied covariance was PD, and the package marked them
admissible and converged. Endpoint re-evaluation reproduced every reported
original objective exactly at recorded precision.

```{r}
v <- read_result("verification/task-115-population-v1", "attempts.csv")
knitr::kable(data.frame(Population = v$population, Model = v$model,
  Converged = v$converged, Admissible = v$admissible,
  `Original objective` = v$fmin, `Main-factor reliability` = v$omega_main_observed,
  `Generating target` = v$true_reliability, check.names = FALSE),
  digits = 6, row.names = FALSE)
```

The one-factor fit to the higher-order covariance has objective 0.474220 and
main-factor functional 0.766042, versus generating target 0.602209. The other
three higher-order controls have reported objective zero and functionals near
0.602209. The correct one-factor control gives 0.600098; its optimizer status
is `line_search_salvaged`, while its convergence verdict passes. Raw statuses,
evaluation counts, admissibility diagnostics and the frozen historical
comparison are retained separately in the CSV inventory. For these five controls,
the largest historical reliability difference is $1.11 \times 10^{-7}$ and
the largest objective difference is $3.25 \times 10^{-14}$. Fit calls totalled
0.392 seconds; the saved-endpoint diagnostic recovery adds the time recorded
in `endpoint_metadata.csv`, within the external numerical budget.

The initial diagnostic API lookup failed on all five saved endpoints. The
lookup was corrected and diagnostics completed from those endpoints without
repeating fits; `diagnostic_recovery.csv` preserves these failures. Pre-fit
syntax, covariance, source and installed-runtime hashes remain frozen. The
recovery has its own source hash. These are five selected development controls,
not a validation of every historical cell. No endpoint is asserted to be an
independently proved reference optimum, and no identification claim or
spread-of-models inference follows from this table.

## Caveats

The Bell lane is a population reconstruction and extension inspired by Bell
et al. (2024), rather than a finite-N replication. Its exact population formulas
and estimator/denominator definitions remain in the runner. Population fitting
uses a large nominal n to evaluate the ML limit, not a million-observation
Monte Carlo experiment. Different estimators' discrepancy numbers are not
interchangeable fit tests.

A fresh population smoke on the current installed package reproduced the
main one-factor/SG comparison within $10^{-6}$ and its generating targets
exactly, but some alternative CFA fits failed or reached different estimates.
The model-sensitivity table above therefore reports the frozen historical
fits, not a verified current-package model-selection result. Before using
that extension for a new claim, revalidate convergence, admissibility and
objective agreement for each alternative CFA. The compact reproduction audit
is retained under `results/verification/`; full fresh outputs stay local. Changing the model changes the functional even
with valid standard errors. Neither lane settles near-boundary interval calibration.

## Reproduce

```sh
Rscript run_experiment.R --lane coefficients --smoke
Rscript run_experiment.R --lane bell --smoke --results-dir /tmp/reliability-bell
R_LIBS=/home/jonas/.cache/magmaan-rlib/lane-e timeout 180s taskset -c 11 nice -n 10 \
  Rscript run_experiment.R --lane bell --bounded-validation --results-dir /absolute/fresh/run
quarto render report.qmd
```

Coefficients writes targets under `results/coefficients/`. Bell outputs came
from research/17 and remain under `results/bell/`. The Bell smoke command
computes the same deterministic population study in a fresh directory. Use
`--lane NAME --help` for commands; rendering only reads local CSVs.
