---
title: "What does covariance-honest NTML repair?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("full", "pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
corpus <- read_result("corpus_anatomy.csv")
changes <- read_result("corpus_parameter_changes.csv")
raw <- read_result("risk_raw.csv")
pairs <- read_result("risk_pairs.csv")
method_summary <- read_result("risk_method_summary.csv")
pair_summary <- read_result("risk_pair_summary.csv")
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)
profile <- metadata[["profile"]]
reps <- as.integer(metadata[["reps"]])
is_evidence <- profile %in% c("pilot", "full") && reps >= 200L

fmt_pct <- function(x) sprintf("%.1f%%", 100 * x)
fmt_num <- function(x, digits = 3L) formatC(
  x, format = "f", digits = digits)

ordinary_bad <- sum(!pairs$ordinary_admissible)
all_psd_usable <- all(pairs$psd_usable)
repaired <- pairs[pairs$both_analysis_ok & !pairs$ordinary_admissible, ]
mean_repaired_psi <- mean(repaired$psi_error_improvement)
mean_repaired_sigma <- mean(repaired$sigma_error_improvement)
mean_repaired_kl <- mean(repaired$kl_improvement)
rate_repaired_psi <- mean(repaired$psi_error_improvement > 1e-10)
rate_repaired_sigma <- mean(repaired$sigma_error_improvement > 1e-10)
rate_repaired_kl <- mean(repaired$kl_improvement > 1e-10)
median_time_ratio <- median(pairs$time_ratio, na.rm = TRUE)

short_answer <- if (!is_evidence) {
  sprintf(
    paste(
      "This is a %s validation run with %d replications per cell.",
      "It verifies the design and report, but it is not the prespecified",
      "risk pilot."
    ), profile, reps)
} else {
  sprintf(
    paste(
      "Across %d generated datasets, ordinary NTML left the primitive",
      "covariance domain %d times; every warm PSD refit was KKT-converged",
      "and admissible. Conditional on an ordinary inadmissible estimate,",
      "PSD-ML reduced mean relative error by %.4f for Psi and %.4f for",
      "Sigma, and reduced mean population KL loss by %.4f. It improved",
      "all three losses in %.1f%%, %.1f%%, and %.1f%% of repaired datasets,",
      "respectively. The median PSD/ordinary fit-time ratio was %.2f."
    ),
    nrow(pairs), ordinary_bad,
    mean_repaired_psi, mean_repaired_sigma, mean_repaired_kl,
    100 * rate_repaired_psi, 100 * rate_repaired_sigma,
    100 * rate_repaired_kl, median_time_ratio)
}

corpus_display <- corpus[, c(
  "set", "id", "geometry", "ordinary_psi_min", "psd_psi_min",
  "lr_scale_increase", "max_abs_parameter_change",
  "implied_sigma_relative_distance")]
names(corpus_display) <- c(
  "Set", "Case", "Geometry", "Ordinary min eig(Psi)",
  "PSD min eig(Psi)", "LR-scale cost", "Max parameter change",
  "Relative implied-Sigma change")
for (field in names(corpus_display)[4:8]) {
  corpus_display[[field]] <- format(
    corpus_display[[field]], digits = 4L, scientific = TRUE)
}

display_n <- sort(unique(pair_summary$n))
display_lambda <- sort(unique(pair_summary$lambda_min))
positive_lambda <- display_lambda[display_lambda > 0]
lambda_offset <- if (length(positive_lambda)) min(positive_lambda) / 2 else 1
lambda_coordinate <- function(x) log10(x + lambda_offset)
risk_display <- pair_summary[, c(
  "n", "lambda_min", "ordinary_inadmissible_rate",
  "psd_boundary_rate", "repaired_mean_psi_error_improvement",
  "repaired_mean_sigma_error_improvement",
  "repaired_mean_kl_improvement", "median_time_ratio")]
names(risk_display) <- c(
  "N", "lambda", "Ordinary inadmissible", "PSD boundary",
  "Psi error improvement", "Sigma error improvement",
  "KL improvement", "Time ratio")
risk_display$`Ordinary inadmissible` <- vapply(
  risk_display$`Ordinary inadmissible`, fmt_pct, character(1))
risk_display$`PSD boundary` <- vapply(
  risk_display$`PSD boundary`, fmt_pct, character(1))
for (field in c("Psi error improvement", "Sigma error improvement",
                "KL improvement", "Time ratio")) {
  risk_display[[field]] <- round(risk_display[[field]], 4L)
}
```

## Question

When ordinary complete-data normal-theory ML estimates an indefinite primitive
LISREL covariance, what does an explicit covariance-honest refit change? In a
controlled Gaussian model generated from a genuinely PSD population, does the
repair improve estimation risk near the covariance boundary?

## Short answer

`r short_answer`

This is an estimation experiment. The objective difference is descriptive and
is not assigned a p-value or likelihood-ratio reference distribution.

## Repair anatomy

```{r corpus-table}
knitr::kable(corpus_display, align = "lllrrrrr")
```

The four deterministic cases show that “an improper solution” is not a single
phenomenon. The first Little model permits a nontrivial reallocation between
residual and latent level variance with effectively unchanged likelihood and
implied observed covariance. Its linear-growth companion pays a substantial
fit cost and moves the fitted observed covariance materially. The Geiser
second-order disturbance is repaired at negligible likelihood cost, whereas
the jointly indefinite quadratic-growth covariance produces an intermediate
change despite all individual variances being positive.

The LR-scale cost is $2N(F_{PSD}-F_{ordinary})$. It is shown only to make the
size of the likelihood sacrifice legible; it is not a test statistic in this
report.

## Controlled covariance path

The risk design fixes indicator residual variance at $0.2$ and varies the
smallest eigenvalue of a three-dimensional latent covariance:

\[
  \operatorname{eig}(\Psi)=(1,0.4,\lambda_{min}),\qquad
  \Sigma=\Psi+0.2I.
\]

Consequently, the observed Gaussian likelihood remains regular even when the
primitive latent covariance is exactly singular. Ordinary NTML is saturated in
the observed covariance; PSD-ML changes it only when the implied primitive
estimate would leave the covariance domain.

```{r inadmissibility-plot, fig.width=7.5, fig.height=4.8}
colors <- grDevices::hcl.colors(length(display_n), "Dark 3")
plot(
  range(lambda_coordinate(display_lambda)), c(0, 1), type = "n",
  xlab = expression(lambda[min]),
  ylab = "Proportion",
  main = "When does ordinary NTML leave the PSD domain?",
  xaxt = "n")
axis(1, at = lambda_coordinate(display_lambda),
     labels = format(display_lambda, scientific = FALSE, trim = TRUE))
for (i in seq_along(display_n)) {
  x <- pair_summary[pair_summary$n == display_n[[i]], ]
  lines(lambda_coordinate(x$lambda_min), x$ordinary_inadmissible_rate,
        type = "b", pch = 14L + i, col = colors[[i]], lwd = 2)
}
legend("topright", legend = paste0("N = ", display_n),
       col = colors, pch = 15L + seq_along(display_n), lwd = 2,
       bty = "n")
```

The horizontal coordinate is $\log_{10}(\lambda+c)$, with $c$ equal to half
the smallest positive grid value, so the exact zero boundary remains visible.
At the exact population boundary, the constrained estimate may itself land either on the
boundary or in the interior. The fitted rank is therefore a description of the
sample solution, not a rank-selection claim.

```{r risk-table}
knitr::kable(risk_display, align = "rrrrrrrr")
```

Positive error improvements favor PSD-ML and are calculated conditional on the
ordinary estimate being inadmissible. `NA` cells indicate that no repair was
triggered in that cell. When ordinary NTML is already admissible, the paired
solutions agree to numerical precision.

## Population KL loss

```{r kl-plot, fig.width=8, fig.height=7}
old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1))
on.exit(par(old_par), add = TRUE)
method_colors <- c(ordinary = "#B23A48", psd = "#246A73")
for (n_value in display_n) {
  x <- method_summary[method_summary$n == n_value, ]
  ylim <- range(x$mean_population_kl, finite = TRUE)
  plot(
    range(lambda_coordinate(display_lambda)), ylim, type = "n", xaxt = "n",
    xlab = expression(lambda[min]), ylab = "Mean population KL",
    main = paste0("N = ", n_value))
  axis(1, at = lambda_coordinate(display_lambda),
       labels = format(display_lambda, scientific = FALSE, trim = TRUE))
  for (method in c("ordinary", "psd")) {
    z <- x[x$method == method, ]
    lines(lambda_coordinate(z$lambda_min), z$mean_population_kl, type = "b",
          pch = if (method == "ordinary") 16 else 17,
          col = method_colors[[method]], lwd = 2)
  }
  if (n_value == display_n[[1L]]) {
    legend("topright", legend = c("Ordinary", "PSD"),
           col = method_colors, pch = c(16, 17), lwd = 2, bty = "n")
  }
}
```

The constraint is beneficial where the ordinary estimate violates a true
population restriction. As sample size grows or the population moves inward,
ordinary inadmissibility becomes rare and the two estimators coincide. The
experiment therefore supports an audit-first policy, not unconditional silent
replacement of ordinary NTML.

## Numerical checks and limits

- The run contains `r nrow(pairs)` paired datasets and
  `r nrow(raw)` fitted-result rows.
- `r if (all_psd_usable) "Every PSD refit passed the KKT and covariance-admissibility audits." else "At least one PSD refit failed; inspect the raw results."`
- Ordinary and PSD estimates both returned for
  `r fmt_pct(mean(pairs$both_analysis_ok))` of datasets.
- The median warm-PSD/ordinary time ratio was
  `r fmt_num(median_time_ratio, 2L)`.

This deliberately transparent model does not represent the full difficulties
of a latent SEM with estimated loadings, structural regressions, equality
links, or competing likelihood basins. Those geometries are represented by the
four repair cases and prior validation experiments, but the risk conclusion is
limited to this controlled DGP. No inferential coverage or hypothesis-testing
claim is made.
