---
title: "FIML global-GOF stress grid"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- file.path(script_dir, "results", "pilot")
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
design <- read_result("design")
summary <- read_result("method_summary")
scorecard <- read_result("method_scorecard")
region_summary <- read_result("region_summary")
adjusted_region <- read_result("size_adjusted_power_by_region")
diagnostics <- read_result("design_diagnostics")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)
parity_path <- file.path(result_dir, "mlr_lavaan_parity_summary.csv")
parity <- if (file.exists(parity_path)) read.csv(
  parity_path, stringsAsFactors = FALSE) else NULL
sem_result_dir <- file.path(script_dir, "results", "sem-model-smoke")
sem_paths <- file.path(sem_result_dir, c(
  "design.csv", "replications.csv", "timing_projection.csv", "metadata.csv"))
sem_available <- all(file.exists(sem_paths))
if (sem_available) {
  sem_design <- read.csv(sem_paths[[1L]], stringsAsFactors = FALSE)
  sem_replications <- read.csv(sem_paths[[2L]], stringsAsFactors = FALSE)
  sem_projection <- read.csv(sem_paths[[3L]], stringsAsFactors = FALSE)
  sem_metadata <- read.csv(sem_paths[[4L]], stringsAsFactors = FALSE)
  sem_meta <- setNames(sem_metadata$value, sem_metadata$key)
}
sem1000_path <- file.path(
  script_dir, "results", "sem-model-1000", "replications.csv")
sem1000_available <- file.exists(sem1000_path)
if (sem1000_available) {
  sem1000 <- read.csv(sem1000_path, stringsAsFactors = FALSE)
}

method_labels <- c(
  lrt_naive = "ordinary LR", lrt_mlr = "Yuan--Bentler MLR",
  lrt_peba4 = "LR pEBA(4)", score_peba4 = "score pEBA(4)",
  score_sb = "score SB",
  flip_effective = "effective multiplier score",
  wald_robust = "robust Wald")
shortlist <- names(method_labels)
label_methods <- function(x) unname(method_labels[x])
format_rate <- function(x) sprintf("%.3f", x)
```

## Question and boundary

Can the usual Yuan--Bentler/Mplus FIML statistic remain adequately calibrated
across practically recognizable forms of nonnormality and missingness, and do
the eigenvalue-matched or multiplier alternatives offer a useful replacement?

The restricted model is a five-variable independence covariance model. The
less-restricted model frees all observed covariances and is saturated. The
existing affine nested-score machinery is therefore a genuine global
goodness-of-fit test *for this design*. It is a screening bridge: the result
does not establish that the score constructions already work as global tests
for nonlinear latent-variable restrictions.

Every power cell sets a latent Gaussian/copula correlation of .35 between
`x1` and `x2`; null cells set it to zero. `x1` is always observed, and missing
rates refer to the remaining four variables.

## Stress grid

Seven outcome laws are crossed factorially with five observation mechanisms
and with null versus power, for `r nrow(design)` cells:

```{r design-table}
distribution_table <- unique(design[, c(
  "distribution_label", "distribution_class", "fourth_moment")])
names(distribution_table) <- c("outcome law", "stress type", "fourth moment")
knitr::kable(distribution_table, row.names = FALSE)

missingness_table <- unique(design[, c(
  "missingness_label", "missingness_class", "missing_rate")])
missingness_table$missing_rate <- sprintf(
  "%.0f%%", 100 * missingness_table$missing_rate)
names(missingness_table) <- c("observation mechanism", "class", "target")
knitr::kable(missingness_table, row.names = FALSE)
```

The interpretation strata are intentional:

```{r region-counts}
scenarios <- unique(design[, c("scenario_id", "analysis_region",
                               "null_contract")])
region_counts <- aggregate(
  scenario_id ~ analysis_region + null_contract, scenarios, length)
names(region_counts) <- c("analysis region", "interpretation", "scenarios")
knitr::kable(region_counts, row.names = FALSE)
```

- **Gaussian reference** and **finite-moment robustness** are the primary
  calibration comparison. The latter contains non-Gaussian complete and MCAR
  cases with finite fourth moments.
- **MAR pseudo-ML stress** represents common applied use: nonnormal outcomes
  with ignorable, covariate-driven missingness. It is not treated as a blanket
  distribution-free guarantee of normal-theory FIML.
- **Moment boundary** uses a `t(3)` law, for which the fourth moment required by
  sandwich-style asymptotics does not exist.
- **Ignorability failure** is self-masked MNAR. Rejection there diagnoses
  sensitivity to a violated missingness model, not a failed promise of a robust
  statistic.

## Primary calibration scorecard

The scorecard below uses only the
`r sum(scenarios$null_contract == "primary robustness target")` primary null
scenarios. “Near nominal” means a
cell rejection rate between .025 and .075; it is a descriptive band, not a
formal equivalence test.

```{r scorecard-table}
show_scorecard <- scorecard[scorecard$method %in% shortlist, c(
  "method", "primary_mean_abs_size_error", "primary_worst_abs_size_error",
  "primary_min_size", "primary_max_size", "primary_near_nominal_fraction",
  "primary_raw_power", "primary_size_adjusted_power")]
show_scorecard$method <- label_methods(show_scorecard$method)
show_scorecard[-1L] <- lapply(show_scorecard[-1L], format_rate)
names(show_scorecard) <- c(
  "method", "mean |size-.05|", "worst |size-.05|", "min size", "max size",
  "fraction near .05", "raw power", "size-adjusted power")
knitr::kable(show_scorecard, row.names = FALSE)
```

The effective multiplier is the only shortlisted method inside the descriptive
near-nominal band in every primary cell. Its primary size ranges from
`r format_rate(scorecard$primary_min_size[scorecard$method == "flip_effective"])`
to
`r format_rate(scorecard$primary_max_size[scorecard$method == "flip_effective"])`,
with mean absolute size error
`r format_rate(scorecard$primary_mean_abs_size_error[scorecard$method == "flip_effective"])`.
The best simple analytic score correction is score SB; score pEBA(4) is more
conservative.

## Where calibration changes

Pooling is shown separately by interpretation stratum so deliberate contract
violations cannot drive the headline comparison.

```{r region-calibration}
region_null <- region_summary[
  region_summary$truth == "null" & region_summary$method %in% shortlist,
  c("analysis_region", "method", "rejection_rate")]
region_null$method <- label_methods(region_null$method)
region_wide <- reshape(region_null, idvar = "analysis_region",
                       timevar = "method", direction = "wide")
names(region_wide) <- sub("^rejection_rate\\.", "", names(region_wide))
region_wide[-1L] <- lapply(region_wide[-1L], format_rate)
names(region_wide)[1L] <- "analysis region"
knitr::kable(region_wide, row.names = FALSE)
```

In the primary finite-moment nonnormal region, MLR rejection is
`r format_rate(region_summary$rejection_rate[region_summary$truth == "null" & region_summary$analysis_region == "finite-moment robustness" & region_summary$method == "lrt_mlr"])`,
versus
`r format_rate(region_summary$rejection_rate[region_summary$truth == "null" & region_summary$analysis_region == "finite-moment robustness" & region_summary$method == "flip_effective"])`
for the effective multiplier score. The MLR contrast therefore does not rely on
the `t(3)` or MNAR cells.

## Detection after matching empirical size

For each scenario and method, the diagnostic cutoff is the empirical fifth
percentile of null p-values; that cutoff is then applied to the matched power
cell. It is unavailable in practice and should only be read as a detection
comparison after removing observed size distortion.

```{r adjusted-power}
power_show <- adjusted_region[adjusted_region$method %in% shortlist, ]
power_show$method <- label_methods(power_show$method)
power_show$raw_power <- format_rate(power_show$raw_power)
power_show$size_adjusted_power <- format_rate(
  power_show$size_adjusted_power)
names(power_show) <- c(
  "analysis region", "method", "raw power", "size-adjusted power")
knitr::kable(power_show, row.names = FALSE)
```

The alternative is held fixed on the latent Gaussian scale, not on the
observed Pearson-correlation scale. The following diagnostic makes the
resulting signal variation visible rather than attributing all power variation
to the test:

```{r signal-diagnostics}
signal <- diagnostics[diagnostics$truth == "power", ]
signal <- aggregate(
  cbind(complete_r12, observed_r12, realized_missing_eligible) ~
    distribution + missingness, signal, mean)
signal$complete_r12 <- format_rate(signal$complete_r12)
signal$observed_r12 <- format_rate(signal$observed_r12)
signal$realized_missing_eligible <- format_rate(
  signal$realized_missing_eligible)
names(signal) <- c(
  "outcome law", "mechanism", "pre-missing r12", "observed r12",
  "eligible missing fraction")
knitr::kable(signal, row.names = FALSE)
```

## Run health

This run used `r meta[["reps"]]` replications per cell and
`r meta[["flips"]]` multiplier draws. Its wall time was
`r sprintf("%.1f", as.numeric(meta[["wall_seconds"]]))` seconds. Restricted-fit,
LR, score, and Wald failures were respectively `r meta[["replication_failures"]]`,
`r meta[["lrt_failures"]]`, `r meta[["score_failures"]]`, and
`r meta[["wald_failures"]]`.

## MLR oracle

`p_lrt_mlr` comes from magmaan's dedicated
`estimate_fiml_robust_mlr()` binding. It computes the
Yuan--Bentler/Mplus FIML statistic from the model-versus-saturated LR statistic
and the H1-minus-H0 trace correction. The companion validation script fits
lavaan with `estimator = "MLR"` on the identical generated data. This checks
implementation parity; it does not validate the statistic's calibration.

```{r parity}
if (!is.null(parity)) {
  show_parity <- parity[, c("metric", "n", "max_abs_difference",
                            "mean_abs_difference")]
  knitr::kable(show_parity, digits = 7, row.names = FALSE)
} else {
  cat("Run `validate_mlr_lavaan.R` to populate the same-data oracle table.")
}
```

## Representative latent-SEM design

The next design replaces the affine independence bridge with four ordinary,
compact SEM populations. Every fitted model is exactly correct in the
population, so rejection in this section is Type-I error rather than power.

### Population models

1. **One-factor CFA, six indicators.** The population is
   $x_j=\lambda_j f+\varepsilon_j$, with
   $\lambda=(1,.8,.9,.7,1.1,.85)$, $\operatorname{Var}(f)=1$, and unique
   variances $(.5,.6,.55,.65,.5,.6)$. Indicator means are zero. The fitted CFA
   uses ordinary marker-variable identification and freely estimates indicator
   means, unique variances, the factor variance, and the remaining loadings.

2. **Correlated two-factor CFA, ten indicators.** Indicators 1--5 load only on
   factor 1 with loadings $(.4,.5,.6,.8,.4)$; indicators 6--10 load only on
   factor 2 with loadings $(.4,.7,.6,.4,.8)$. Both factor variances are one and
   their correlation is .5. Unique variances are $1-\lambda_j^2$, giving unit
   marginal variances, and all means are zero. This is the compact
   Foldnes--Moss--Grønneberg-style measurement model.

3. **Orthogonal bifactor model, twelve indicators.** Every indicator has a .6
   loading on the general factor. Three mutually orthogonal specific factors
   each cover a four-item block, with blockwise loadings .5, .4, and .45.
   General and specific factor variances are one and all factor covariances are
   zero. Unique variances are $1-.6^2-\lambda_{S,j}^2$ and indicator means are
   zero. The fitted model uses `std.lv` identification and freely estimates all
   general/specific loadings, unique variances, and means.

4. **Five-wave linear growth model.** For times $t=0,1,2,3,4$,
   $y_t=i+t s+\varepsilon_t$. The intercept/slope means are $(1,.2)$ and their
   covariance matrix is
   $\left[\begin{smallmatrix}1&.1\\.1&.1\end{smallmatrix}\right]$; each wave
   has residual variance .5. The fitted model fixes the time scores and freely
   estimates the two growth means, three latent variance/covariance terms, and
   five residual variances.

```{r sem-model-smoke}
if (sem_available) {
  model_table <- unique(sem_design[, c("model_id", "model_label", "p",
                                       "expected_df")])
  model_table$free_parameters <- c(18L, 31L, 48L, 10L)[match(
    model_table$model_id,
    c("one_factor_6", "two_factor_fmg_10", "bifactor_12", "linear_growth_5"))]
  model_table <- model_table[, c(
    "model_label", "p", "free_parameters", "expected_df")]
  names(model_table) <- c("model", "observed variables", "free parameters", "df")
  knitr::kable(model_table, row.names = FALSE)

  generator_table <- data.frame(
    code = c("normal", "vm1", "ig1", "vm2", "ig2"),
    generator = c(
      "Multivariate normal", "Vale--Maurelli / Fleishman (VM)",
      "Independent generator (IG)", "Vale--Maurelli / Fleishman (VM)",
      "Independent generator (IG)"),
    marginal_skewness = c(0, 2, 2, 3, 3),
    marginal_excess_kurtosis = c(0, 7, 7, 21, 21),
    severity = c("reference", "moderate", "moderate", "severe", "severe"),
    stringsAsFactors = FALSE)
  knitr::kable(generator_table, row.names = FALSE)

  health <- data.frame(
    quantity = c("FIML converged", "FMG call returned", "MLR finite"),
    result = sprintf("%d/%d", c(
      sum(sem_replications$fit_ok), sum(sem_replications$fmg_ok),
      sum(sem_replications$mlr_ok)), nrow(sem_replications)))
  knitr::kable(health, row.names = FALSE)

  projection_show <- sem_projection[, c(
    "panel", "reps_per_cell", "total_replications", "projected_wall_seconds")]
  projection_show$projected_wall_minutes <-
    projection_show$projected_wall_seconds / 60
  projection_show$projected_wall_seconds <- NULL
  names(projection_show) <- c(
    "panel", "reps/cell", "total fits", "projected wall minutes")
  knitr::kable(projection_show, digits = 2, row.names = FALSE)
} else {
  cat("Run `run_sem_models.R` to populate the timing smoke tables.")
}
```

**VM** is the Vale--Maurelli/Fleishman polynomial construction: magmaan solves
an intermediate Gaussian correlation matrix and applies calibrated marginal
polynomials so the generated covariance and requested marginal skewness and
excess kurtosis agree with the SEM population targets. **IG** is the
Foldnes--Olsson independent-generator construction: independent Pearson-family
nonnormal generators are calibrated and mixed through a symmetric covariance
root. VM and IG therefore share the same marginal moments and target covariance
but induce different multivariate higher-order dependence. This is the intended
comparison; `vm1` versus `ig1` and `vm2` versus `ig2` are not duplicate DGPs.

Each generator is crossed with complete data and 30% MCAR on every variable
except the first indicator/wave, which is always observed. Thus the design has
$4\times5\times2=40$ true-null cells at $n=120$.

All FIML fits converged in the retained 10-rep-per-cell smoke. The one-factor,
two-factor, and growth FMG calls were complete. The bifactor accounted for all
12 FMG failures: its strict projected-rank guard rejected eigenvalues of about
`1e-6` to `1e-4` that should be numerically projected out. Two MLR p-values
were non-finite. Thus the design runs, but the bifactor FMG rank tolerance is a
real preflight issue to resolve or prespecify before a large comparison.

The projections use the observed four-worker speedup and include only current
null-model fitting plus MLR/FMG post-processing. They do not include the general
FIML saturated-moment score projection or the effective multiplier test: that
latent-SEM extension is not yet shipped and must not be represented by the
affine bridge statistic. The doubled panel is a simple cost projection for one
matched power condition, not an implemented power DGP.

## Representative-SEM null run

The retained follow-up uses the same 40 null cells with 1,000 replications per
cell. It remains a null-calibration run: no power condition and no general
multiplier statistic have been added yet.

```{r sem-model-1000}
if (sem1000_available) {
  health1000 <- data.frame(
    quantity = c("FIML converged", "FMG call returned", "MLR finite"),
    result = sprintf("%d/%d", c(
      sum(sem1000$fit_ok), sum(sem1000$fmg_ok), sum(sem1000$mlr_ok)),
      nrow(sem1000)))
  knitr::kable(health1000, row.names = FALSE)

  p_columns <- c(
    "Yuan--Bentler MLR" = "p_lrt_mlr", "FMG SB" = "p_lrt_sb",
    "FMG SS" = "p_lrt_ss", "FMG pEBA(4)" = "p_lrt_peba4",
    "FMG all" = "p_lrt_all")
  group_vars <- c("model_id", "distribution", "missingness")
  group_id <- interaction(sem1000[group_vars], drop = TRUE, lex.order = TRUE)
  cell_size <- do.call(rbind, lapply(
    split(seq_len(nrow(sem1000)), group_id), function(ii) {
      z <- sem1000[ii, , drop = FALSE]
      do.call(rbind, lapply(names(p_columns), function(method) {
        p <- z[[p_columns[[method]]]]
        data.frame(
          z[1L, group_vars, drop = FALSE],
          method = method, finite = sum(is.finite(p)),
          rejection = mean(p < .05, na.rm = TRUE),
          stringsAsFactors = FALSE)
      }))
    }))
  calibration1000 <- do.call(rbind, lapply(
    split(cell_size, cell_size$method), function(z) {
      data.frame(
        method = z$method[[1L]], finite = sum(z$finite),
        mean_cell_rejection = mean(z$rejection),
        mean_abs_size_error = mean(abs(z$rejection - .05)),
        min_cell_rejection = min(z$rejection),
        max_cell_rejection = max(z$rejection),
        fraction_025_075 = mean(z$rejection >= .025 & z$rejection <= .075),
        stringsAsFactors = FALSE)
    }))
  names(calibration1000) <- c(
    "method", "finite p-values", "mean cell size", "mean |size-.05|",
    "minimum", "maximum", "fraction in [.025,.075]")
  knitr::kable(calibration1000, digits = 3, row.names = FALSE)

  generator_labels <- c(
    normal = "Normal", vm1 = "VM moderate", ig1 = "IG moderate",
    vm2 = "VM severe", ig2 = "IG severe")
  dgp_rows <- unique(sem1000[, c("distribution", "missingness")])
  generator_size <- do.call(rbind, lapply(seq_len(nrow(dgp_rows)), function(i) {
    z <- sem1000[
      sem1000$distribution == dgp_rows$distribution[[i]] &
        sem1000$missingness == dgp_rows$missingness[[i]], , drop = FALSE]
    summarize_p <- function(column) {
      p <- z[[column]]
      c(n = sum(is.finite(p)), rejection = mean(p < .05, na.rm = TRUE))
    }
    mlr <- summarize_p("p_lrt_mlr")
    ss <- summarize_p("p_lrt_ss")
    all <- summarize_p("p_lrt_all")
    data.frame(
      generator = generator_labels[[dgp_rows$distribution[[i]]]],
      missingness = if (dgp_rows$missingness[[i]] == "complete")
        "Complete" else "30% MCAR",
      mlr_n = mlr[["n"]], mlr_rejection = mlr[["rejection"]],
      ss_n = ss[["n"]], ss_rejection = ss[["rejection"]],
      all_n = all[["n"]], all_rejection = all[["rejection"]],
      stringsAsFactors = FALSE)
  }))
  generator_size$generator <- factor(
    generator_size$generator, levels = unname(generator_labels))
  generator_size$missingness <- factor(
    generator_size$missingness, levels = c("Complete", "30% MCAR"))
  generator_size <- generator_size[order(
    generator_size$generator, generator_size$missingness), ]
  names(generator_size) <- c(
    "generator", "observation", "MLR n", "MLR rejection",
    "SS n", "SS rejection", "all n", "all rejection")
  knitr::kable(generator_size, digits = 3, row.names = FALSE)
} else {
  cat("Run `run_sem_models.R --reps 1000` to populate this table.")
}
```

The run took 541 seconds on four workers. Fifteen FIML fits did not converge.
Among otherwise converged fits, 91 MLR p-values were non-finite and 829 FMG
calls hit the projected `U Gamma` rank guard; 797 of those FMG failures were in
the bifactor model. MLR averaged .366 rejection across the 40 cells and reached
.932 in the severe-VM bifactor MCAR cell. FMG `all` was much better at .066 on
average, but its .022--.168 range shows that it is not a uniform solution; its
largest failures were also severe nonnormal MCAR cells. All rates use finite
method-specific denominators rather than treating numerical failures as
non-rejections.

The generator-specific table makes the substantive contrast visible. MLR
rejection increases from .294/.440 under moderate VM (complete/MCAR) to
.556/.689 under severe VM. The corresponding IG rates are .232/.381 and
.318/.493. VM is harder than IG at the same marginal moments, so marginal
skewness and kurtosis alone do not explain the failure. FMG `all` is near .05
for all complete-data generators, but rises to .099/.127 under moderate/severe
VM with MCAR; IG is appreciably better at .063/.055. The new report therefore
keeps VM and IG separate rather than pooling them into a generic nonnormal arm.

## Reproduce

```sh
Rscript experiments/77-fiml-global-gof-pilot/run_experiment.R --smoke
Rscript experiments/77-fiml-global-gof-pilot/run_experiment.R --pilot
Rscript experiments/77-fiml-global-gof-pilot/run_sem_models.R
Rscript experiments/77-fiml-global-gof-pilot/run_sem_models.R --reps 1000 \
  --results-dir experiments/77-fiml-global-gof-pilot/results/sem-model-1000
Rscript experiments/77-fiml-global-gof-pilot/validate_mlr_lavaan.R
quarto render experiments/77-fiml-global-gof-pilot/report.qmd
```
