---
title: "FIML global-GOF stress grid"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- file.path(script_dir, "results", "pilot")
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
design <- read_result("design")
summary <- read_result("method_summary")
scorecard <- read_result("method_scorecard")
region_summary <- read_result("region_summary")
adjusted_region <- read_result("size_adjusted_power_by_region")
diagnostics <- read_result("design_diagnostics")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)
parity_path <- file.path(result_dir, "mlr_lavaan_parity_summary.csv")
parity <- if (file.exists(parity_path)) read.csv(
  parity_path, stringsAsFactors = FALSE) else NULL

method_labels <- c(
  lrt_naive = "ordinary LR", lrt_mlr = "Yuan--Bentler MLR",
  lrt_peba4 = "LR pEBA(4)", score_peba4 = "score pEBA(4)",
  score_sb = "score SB",
  flip_effective = "effective multiplier score",
  wald_robust = "robust Wald")
shortlist <- names(method_labels)
label_methods <- function(x) unname(method_labels[x])
format_rate <- function(x) sprintf("%.3f", x)
```

## Question and boundary

Can the usual Yuan--Bentler/Mplus FIML statistic remain adequately calibrated
across practically recognizable forms of nonnormality and missingness, and do
the eigenvalue-matched or multiplier alternatives offer a useful replacement?

The restricted model is a five-variable independence covariance model. The
less-restricted model frees all observed covariances and is saturated. The
existing affine nested-score machinery is therefore a genuine global
goodness-of-fit test *for this design*. It is a screening bridge: the result
does not establish that the score constructions already work as global tests
for nonlinear latent-variable restrictions.

Every power cell sets a latent Gaussian/copula correlation of .35 between
`x1` and `x2`; null cells set it to zero. `x1` is always observed, and missing
rates refer to the remaining four variables.

## Stress grid

Seven outcome laws are crossed factorially with five observation mechanisms
and with null versus power, for `r nrow(design)` cells:

```{r design-table}
distribution_table <- unique(design[, c(
  "distribution_label", "distribution_class", "fourth_moment")])
names(distribution_table) <- c("outcome law", "stress type", "fourth moment")
knitr::kable(distribution_table, row.names = FALSE)

missingness_table <- unique(design[, c(
  "missingness_label", "missingness_class", "missing_rate")])
missingness_table$missing_rate <- sprintf(
  "%.0f%%", 100 * missingness_table$missing_rate)
names(missingness_table) <- c("observation mechanism", "class", "target")
knitr::kable(missingness_table, row.names = FALSE)
```

The interpretation strata are intentional:

```{r region-counts}
scenarios <- unique(design[, c("scenario_id", "analysis_region",
                               "null_contract")])
region_counts <- aggregate(
  scenario_id ~ analysis_region + null_contract, scenarios, length)
names(region_counts) <- c("analysis region", "interpretation", "scenarios")
knitr::kable(region_counts, row.names = FALSE)
```

- **Gaussian reference** and **finite-moment robustness** are the primary
  calibration comparison. The latter contains non-Gaussian complete and MCAR
  cases with finite fourth moments.
- **MAR pseudo-ML stress** represents common applied use: nonnormal outcomes
  with ignorable, covariate-driven missingness. It is not treated as a blanket
  distribution-free guarantee of normal-theory FIML.
- **Moment boundary** uses a `t(3)` law, for which the fourth moment required by
  sandwich-style asymptotics does not exist.
- **Ignorability failure** is self-masked MNAR. Rejection there diagnoses
  sensitivity to a violated missingness model, not a failed promise of a robust
  statistic.

## Primary calibration scorecard

The scorecard below uses only the
`r sum(scenarios$null_contract == "primary robustness target")` primary null
scenarios. “Near nominal” means a
cell rejection rate between .025 and .075; it is a descriptive band, not a
formal equivalence test.

```{r scorecard-table}
show_scorecard <- scorecard[scorecard$method %in% shortlist, c(
  "method", "primary_mean_abs_size_error", "primary_worst_abs_size_error",
  "primary_min_size", "primary_max_size", "primary_near_nominal_fraction",
  "primary_raw_power", "primary_size_adjusted_power")]
show_scorecard$method <- label_methods(show_scorecard$method)
show_scorecard[-1L] <- lapply(show_scorecard[-1L], format_rate)
names(show_scorecard) <- c(
  "method", "mean |size-.05|", "worst |size-.05|", "min size", "max size",
  "fraction near .05", "raw power", "size-adjusted power")
knitr::kable(show_scorecard, row.names = FALSE)
```

The effective multiplier is the only shortlisted method inside the descriptive
near-nominal band in every primary cell. Its primary size ranges from
`r format_rate(scorecard$primary_min_size[scorecard$method == "flip_effective"])`
to
`r format_rate(scorecard$primary_max_size[scorecard$method == "flip_effective"])`,
with mean absolute size error
`r format_rate(scorecard$primary_mean_abs_size_error[scorecard$method == "flip_effective"])`.
The best simple analytic score correction is score SB; score pEBA(4) is more
conservative.

## Where calibration changes

Pooling is shown separately by interpretation stratum so deliberate contract
violations cannot drive the headline comparison.

```{r region-calibration}
region_null <- region_summary[
  region_summary$truth == "null" & region_summary$method %in% shortlist,
  c("analysis_region", "method", "rejection_rate")]
region_null$method <- label_methods(region_null$method)
region_wide <- reshape(region_null, idvar = "analysis_region",
                       timevar = "method", direction = "wide")
names(region_wide) <- sub("^rejection_rate\\.", "", names(region_wide))
region_wide[-1L] <- lapply(region_wide[-1L], format_rate)
names(region_wide)[1L] <- "analysis region"
knitr::kable(region_wide, row.names = FALSE)
```

In the primary finite-moment nonnormal region, MLR rejection is
`r format_rate(region_summary$rejection_rate[region_summary$truth == "null" & region_summary$analysis_region == "finite-moment robustness" & region_summary$method == "lrt_mlr"])`,
versus
`r format_rate(region_summary$rejection_rate[region_summary$truth == "null" & region_summary$analysis_region == "finite-moment robustness" & region_summary$method == "flip_effective"])`
for the effective multiplier score. The MLR contrast therefore does not rely on
the `t(3)` or MNAR cells.

## Detection after matching empirical size

For each scenario and method, the diagnostic cutoff is the empirical fifth
percentile of null p-values; that cutoff is then applied to the matched power
cell. It is unavailable in practice and should only be read as a detection
comparison after removing observed size distortion.

```{r adjusted-power}
power_show <- adjusted_region[adjusted_region$method %in% shortlist, ]
power_show$method <- label_methods(power_show$method)
power_show$raw_power <- format_rate(power_show$raw_power)
power_show$size_adjusted_power <- format_rate(
  power_show$size_adjusted_power)
names(power_show) <- c(
  "analysis region", "method", "raw power", "size-adjusted power")
knitr::kable(power_show, row.names = FALSE)
```

The alternative is held fixed on the latent Gaussian scale, not on the
observed Pearson-correlation scale. The following diagnostic makes the
resulting signal variation visible rather than attributing all power variation
to the test:

```{r signal-diagnostics}
signal <- diagnostics[diagnostics$truth == "power", ]
signal <- aggregate(
  cbind(complete_r12, observed_r12, realized_missing_eligible) ~
    distribution + missingness, signal, mean)
signal$complete_r12 <- format_rate(signal$complete_r12)
signal$observed_r12 <- format_rate(signal$observed_r12)
signal$realized_missing_eligible <- format_rate(
  signal$realized_missing_eligible)
names(signal) <- c(
  "outcome law", "mechanism", "pre-missing r12", "observed r12",
  "eligible missing fraction")
knitr::kable(signal, row.names = FALSE)
```

## Run health

This run used `r meta[["reps"]]` replications per cell and
`r meta[["flips"]]` multiplier draws. Its wall time was
`r sprintf("%.1f", as.numeric(meta[["wall_seconds"]]))` seconds. Restricted-fit,
LR, score, and Wald failures were respectively `r meta[["replication_failures"]]`,
`r meta[["lrt_failures"]]`, `r meta[["score_failures"]]`, and
`r meta[["wald_failures"]]`.

## MLR oracle

`p_lrt_mlr` comes from magmaan's dedicated
`estimate_fiml_robust_mlr()` binding. It computes the
Yuan--Bentler/Mplus FIML statistic from the model-versus-saturated LR statistic
and the H1-minus-H0 trace correction. The companion validation script fits
lavaan with `estimator = "MLR"` on the identical generated data. This checks
implementation parity; it does not validate the statistic's calibration.

```{r parity}
if (!is.null(parity)) {
  show_parity <- parity[, c("metric", "n", "max_abs_difference",
                            "mean_abs_difference")]
  knitr::kable(show_parity, digits = 7, row.names = FALSE)
} else {
  cat("Run `validate_mlr_lavaan.R` to populate the same-data oracle table.")
}
```

## Reproduce

```sh
Rscript experiments/77-fiml-global-gof-pilot/run_experiment.R --smoke
Rscript experiments/77-fiml-global-gof-pilot/run_experiment.R --pilot
Rscript experiments/77-fiml-global-gof-pilot/validate_mlr_lavaan.R
quarto render experiments/77-fiml-global-gof-pilot/report.qmd
```
