---
title: "Five-minute FIML global-GOF pilot"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- file.path(script_dir, "results", "pilot")
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
summary <- read_result("method_summary")
size_adjusted <- read_result("size_adjusted_power_pooled")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)
parity_path <- file.path(result_dir, "mlr_lavaan_parity_summary.csv")
parity <- if (file.exists(parity_path)) read.csv(
  parity_path, stringsAsFactors = FALSE) else NULL
```

## Purpose

This is a deliberately narrow bridge experiment. The restricted model is a
five-variable independence covariance model, and the less-restricted model
frees every observed covariance and is therefore saturated. Consequently the
existing affine nested-score machinery is, in this design, a genuine global
goodness-of-fit score test. This lets us screen score, multiplier, Wald, and
likelihood-ratio constructions before implementing a general nonlinear-SEM
global-score API.

The eight cells cross normal versus standardized log-normal margins, complete
data versus 30% MCAR on four variables, and a true null versus a Gaussian-copula
correlation of .35 between the first two variables. The first variable is always
observed. Each model has 10 degrees of freedom.

## Results

```{r table}
show <- summary[, c("truth", "distribution", "missing_rate", "method",
                    "valid", "rejection_rate", "ci_lower", "ci_upper")]
show$missing_rate <- sprintf("%.0f%%", 100 * show$missing_rate)
show$rejection_rate <- sprintf("%.3f", show$rejection_rate)
show$ci <- sprintf("[%.3f, %.3f]", show$ci_lower, show$ci_upper)
show$ci_lower <- show$ci_upper <- NULL
knitr::kable(show, row.names = FALSE)
```

```{r adjusted-power}
size_adjusted$size_adjusted_power <- sprintf(
  "%.3f", size_adjusted$size_adjusted_power)
knitr::kable(size_adjusted, row.names = FALSE,
             col.names = c("method", "mean size-adjusted power"))
```

This run used `r meta[["reps"]]` replications per cell and
`r meta[["flips"]]` multiplier draws. Its wall time was
`r sprintf("%.1f", as.numeric(meta[["wall_seconds"]]))` seconds. The intervals
are intentionally wide: the run is for eliminating obviously broken options
and choosing the next iteration, not for a final calibration claim.

## MLR oracle

`p_lrt_mlr` is not reconstructed in the experiment. It comes from magmaan's
dedicated `estimate_fiml_robust_mlr()` binding, which computes the
Yuan--Bentler/Mplus FIML statistic from the model-vs-saturated LRT and the H1
minus H0 trace correction. The companion validation script fits lavaan with
`estimator = "MLR", missing = "ml"` on the identical generated data.

```{r parity}
if (!is.null(parity)) {
  show_parity <- parity[, c("metric", "n", "max_abs_difference",
                            "mean_abs_difference")]
  knitr::kable(show_parity, digits = 7, row.names = FALSE)
} else {
  cat("Run `validate_mlr_lavaan.R` to populate the same-data oracle table.")
}
```

## Reproduce

```sh
Rscript experiments/77-fiml-global-gof-pilot/run_experiment.R --smoke
Rscript experiments/77-fiml-global-gof-pilot/run_experiment.R --pilot
Rscript experiments/77-fiml-global-gof-pilot/validate_mlr_lavaan.R
quarto render experiments/77-fiml-global-gof-pilot/report.qmd
```
