---
title: "FIML and two-stage goodness of fit in representative SEMs"
subtitle: "Score-spectrum, multiplier, and discrepancy calibrations under complete, MCAR, and MAR sampling"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- Sys.getenv(
  "MAGMAAN_SEM_RESULT_DIR",
  unset = file.path(
    script_dir, "results", "sem-paired-score-fmg-n200-500"
  )
)
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
replications <- read_result("replications")
design <- read_result("design")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)

model_order <- c(
  "one_factor_6", "two_factor_fmg_10", "bifactor_12", "three_factor_15",
  "linear_growth_5")
model_labels <- c(
  one_factor_6 = "One-factor CFA",
  two_factor_fmg_10 = "Two-factor CFA",
  bifactor_12 = "Bifactor",
  three_factor_15 = "Three-factor CFA",
  linear_growth_5 = "Linear growth")
estimator_order <- c("FIML", "ML2S")
generator_order <- c("normal", "vm1", "ig1", "vm2", "ig2")
generator_labels <- c(
  normal = "Normal", vm1 = "VM moderate", ig1 = "IG moderate",
  vm2 = "VM severe", ig2 = "IG severe")
missingness_labels <- c(
  complete = "Complete", mcar_30 = "30% MCAR", mar_30 = "30% MAR")
method_columns <- c(
  "Corrected multiplier" = "p_flip_corrected",
  "Legacy multiplier" = "p_flip_effective",
  "Scalar robust" = "p_lrt_mlr",
  "LR/D FMG SB" = "p_lrt_sb",
  "LR/D FMG SS" = "p_lrt_ss",
  "LR/D FMG pEBA(4)" = "p_lrt_peba4",
  "LR/D FMG all" = "p_lrt_all",
  "Score FMG SB" = "p_score_sb",
  "Score FMG SS" = "p_score_ss",
  "Score FMG pEBA(4)" = "p_score_peba4",
  "Score FMG all" = "p_score_all")

group_vars <- c("model_id", "distribution", "missingness", "estimator")
group_id <- interaction(
  replications[group_vars], drop = TRUE, lex.order = TRUE)
cell_rows <- split(seq_len(nrow(replications)), group_id)

cell_estimator_table <- do.call(rbind, lapply(cell_rows, function(ii) {
  z <- replications[ii, , drop = FALSE]
  out <- data.frame(
    model_id = z$model_id[[1L]],
    model = unname(model_labels[[z$model_id[[1L]]]]),
    distribution = z$distribution[[1L]],
    generator = unname(generator_labels[[z$distribution[[1L]]]]),
    missingness_id = z$missingness[[1L]],
    missingness = unname(missingness_labels[[z$missingness[[1L]]]]),
    estimator = z$estimator[[1L]],
    stringsAsFactors = FALSE, check.names = FALSE)
  for (method in names(method_columns)) {
    p_value <- z[[method_columns[[method]]]]
    out[[method]] <- mean(p_value < .05, na.rm = TRUE)
  }
  out
}))
cell_estimator_table$model_id <- factor(
  cell_estimator_table$model_id, levels = model_order)
cell_estimator_table$distribution <- factor(
  cell_estimator_table$distribution, levels = generator_order)
cell_estimator_table$missingness_id <- factor(
  cell_estimator_table$missingness_id,
  levels = c("complete", "mcar_30", "mar_30"))
cell_estimator_table$estimator <- factor(
  cell_estimator_table$estimator, levels = estimator_order)
cell_estimator_table <- cell_estimator_table[order(
  cell_estimator_table$model_id, cell_estimator_table$distribution,
  cell_estimator_table$missingness_id, cell_estimator_table$estimator), ]
row.names(cell_estimator_table) <- NULL

cell_key <- unique(cell_estimator_table[, c(
  "model_id", "model", "distribution", "generator", "missingness_id",
  "missingness")])
cell_table <- cell_key
wide_method_columns <- character()
for (estimator in estimator_order) {
  source <- cell_estimator_table[
    as.character(cell_estimator_table$estimator) == estimator, ]
  source_key <- interaction(
    source[c("model_id", "distribution", "missingness_id")],
    drop = TRUE, lex.order = TRUE)
  target_key <- interaction(
    cell_table[c("model_id", "distribution", "missingness_id")],
    drop = TRUE, lex.order = TRUE)
  index <- match(target_key, source_key)
  for (method in names(method_columns)) {
    if (!any(is.finite(source[[method]]))) next
    display_method <- if (method == "Scalar robust") {
      if (estimator == "FIML") "Yuan--Bentler MLR" else "Scaled ML2S"
    } else method
    column <- paste(estimator, display_method, sep = " / ")
    cell_table[[column]] <- source[[method]][index]
    wide_method_columns <- c(wide_method_columns, column)
  }
}

scorecard <- do.call(rbind, lapply(estimator_order, function(estimator) {
  z <- cell_estimator_table[
    as.character(cell_estimator_table$estimator) == estimator, ]
  raw <- replications[replications$estimator == estimator, ]
  do.call(rbind, lapply(names(method_columns), function(method) {
    rejection <- z[[method]]
    p_value <- raw[[method_columns[[method]]]]
    display_method <- if (method == "Scalar robust") {
      if (estimator == "FIML") "Yuan--Bentler MLR" else "Scaled ML2S"
    } else method
    finite_rejection <- rejection[is.finite(rejection)]
    data.frame(
      estimator = estimator, method = display_method,
      attempted = nrow(raw),
      finite = sum(is.finite(p_value)),
      mean_cell_rejection = if (length(finite_rejection)) {
        mean(finite_rejection)
      } else NA_real_,
      mean_abs_size_error = if (length(finite_rejection)) {
        mean(abs(finite_rejection - .05))
      } else NA_real_,
      min_cell_rejection = if (length(finite_rejection)) {
        min(finite_rejection)
      } else NA_real_,
      max_cell_rejection = if (length(finite_rejection)) {
        max(finite_rejection)
      } else NA_real_,
      cells_in_band = sum(
        rejection >= .025 & rejection <= .075, na.rm = TRUE),
      stringsAsFactors = FALSE)
  }))
}))

replications$nominal_geometry <- replications$flip_ok &
  replications$flip_df == replications$expected_df
replications$corrected_nominal_geometry <-
  replications$flip_corrected_ok &
  replications$flip_corrected_df == replications$expected_df
geometry <- aggregate(
  cbind(fit_ok, flip_ok, nominal_geometry, flip_corrected_ok,
        corrected_nominal_geometry) ~ estimator + model_label,
  replications, sum)
geometry$attempted <- vapply(seq_len(nrow(geometry)), function(i) {
  sum(replications$estimator == geometry$estimator[[i]] &
        replications$model_label == geometry$model_label[[i]])
}, integer(1L))
```

The question is deliberately narrow: how well calibrated are global
goodness-of-fit tests when correctly specified, ordinary latent-variable models
are fitted by FIML or normal-theory two-stage ML (ML2S) to normal,
Vale--Maurelli, and independent-generator data under complete sampling, MCAR,
and observed-predictor MAR? All entries below are Type-I rejection rates at
$\alpha=.05$. The paired design has 75 data-generating cells, $n=200$, and 500
replications per cell; FIML and ML2S are fitted to the same generated datasets.

## Headline result

```{r headline-table}
headline <- scorecard[, c(
  "estimator", "method", "mean_cell_rejection", "mean_abs_size_error",
  "min_cell_rejection", "max_cell_rejection", "cells_in_band")]
headline <- headline[scorecard$finite > 0L, ]
names(headline) <- c(
  "estimator", "method", "mean size", "mean |size-.05|", "minimum", "maximum",
  "cells in [.025,.075]")
knitr::kable(headline, digits = 3, row.names = FALSE)
```

Score pEBA(4) is the best FIML calibration in this panel. Its mean rejection is
.043, its mean absolute cellwise size error is .014, and 66/75 cells lie in the
practical [.025,.075] band. The legacy multiplier gives .058, .015, and 59/75,
respectively. Thus score pEBA(4) is modestly but consistently better here, in
line with the earlier score-FMG experiments.

For ML2S, score pEBA(4) and the multiplier are closer: their mean rejection
rates are .069 and .072, and both place 48/75 cells in the practical band. The
score calibration is better under MAR (.058 versus .068 overall) but worse
under MCAR (.102 versus .088). Neither is uniformly calibrated for ML2S in the
hard missing-data cells.

The MAR cells do not establish consistency. In the severe nonnormal MAR cells,
the legacy FIML multiplier rejects about .075--.079 while score pEBA(4) rejects
about .042--.055. That is useful finite-sample evidence for score-spectrum
calibration, but both normal-FIML and the normal-EM first stage can target
pseudo-true rather than population moments under nonnormal MAR. A reference-law
improvement cannot generally repair that estimand problem.

## All 75 design cells

Each entry is the proportion of finite p-values below .05. Method-specific
finite denominators and numerical diagnostics are reported in the appendix.
The method closest to the nominal .05 rate is bold in each row; exact ties are
all bold.

```{r all-cells-table}
all_cells <- cell_table[, c(
  "model", "generator", "missingness", wide_method_columns)]
cell_rates <- as.matrix(all_cells[, wide_method_columns])
cell_error <- abs(cell_rates - .05)
best_error <- apply(cell_error, 1L, min, na.rm = TRUE)
for (method in wide_method_columns) {
  value <- cell_rates[, method]
  rendered <- sprintf("%.3f", value)
  rendered[!is.finite(value)] <- "---"
  is_winner <- is.finite(cell_error[, method]) &
    abs(cell_error[, method] - best_error) < 1e-12
  rendered[is_winner] <- paste0("<strong>", rendered[is_winner], "</strong>")
  all_cells[[method]] <- rendered
}
knitr::kable(
  all_cells, row.names = FALSE, escape = FALSE,
  align = c("l", "l", "l", rep("r", length(wide_method_columns))))
```

The bold entry is the calibration closest to .05 in that data-generating cell,
across both estimators; exact ties are all bold. The table is intentionally
wide so the paired FIML and ML2S results can be compared without splitting the
design across panels.

## Model specifications

```{r model-table}
model_table <- unique(design[, c(
  "model_id", "model_label", "p", "expected_df")])
model_table$free_parameters <- c(18L, 31L, 48L, 48L, 10L)[match(
  model_table$model_id, model_order)]
model_table$model_id <- factor(model_table$model_id, levels = model_order)
model_table <- model_table[order(model_table$model_id), ]
model_table$model <- unname(model_labels[as.character(model_table$model_id)])
model_table <- model_table[, c(
  "model", "p", "free_parameters", "expected_df")]
names(model_table) <- c("model", "observed variables", "free parameters", "df")
knitr::kable(model_table, row.names = FALSE)
```

1. **One-factor CFA, six indicators.** The population is
   $x_j=\lambda_j f+\varepsilon_j$, with
   $\lambda=(1,.8,.9,.7,1.1,.85)$, $\operatorname{Var}(f)=1$, and unique
   variances $(.5,.6,.55,.65,.5,.6)$. Indicator means are zero. The fitted CFA
   uses marker-variable identification and freely estimates means, unique
   variances, the factor variance, and the remaining loadings.

2. **Correlated two-factor CFA, ten indicators.** Indicators 1--5 load only on
   factor 1 with loadings $(.4,.5,.6,.8,.4)$; indicators 6--10 load only on
   factor 2 with loadings $(.4,.7,.6,.4,.8)$. Factor variances are one and the
   factor correlation is .5. Unique variances are $1-\lambda_j^2$, giving unit
   marginal variances, and all means are zero.

3. **Orthogonal bifactor model, twelve indicators.** Every indicator loads .6
   on a general factor. Three orthogonal specific factors each cover four
   indicators, with blockwise loadings .5, .4, and .45. All factor variances are
   one; unique variances are $1-.6^2-\lambda_{S,j}^2$ and means are zero. The
   fitted model uses `std.lv` identification.

4. **Five-wave linear growth model.** At $t=0,1,2,3,4$,
   $y_t=i+t s+\varepsilon_t$. The intercept/slope means are $(1,.2)$, their
   covariance matrix is

   $$
   \begin{pmatrix}
   1 & 0.1 \\
   0.1 & 0.1
   \end{pmatrix},
   $$

   and each residual variance is .5. The fitted model fixes the time scores
   and freely estimates the growth means, latent variance/covariance terms,
   and residual variances.

5. **Three-factor CFA, fifteen indicators.** Five indicators
   load on each of three correlated factors. Within-factor loadings range from
   .4 to .8; factor correlations are .3, .4, and .5; residual variances give
   unit indicator variances. The fitted model uses `std.lv` identification and
   freely estimates 15 loadings, 15 residual variances, 15 means, and three
   factor covariances: 48 parameters against 135 saturated moments, hence 87
   df.

## Methods

Each generated dataset is fitted by both FIML and ML2S and passed to ten
global-fit calibrations. The current code also contains an eleventh,
FIML-specific corrected multiplier, but it landed after this production panel
started and is therefore described but not included in the tables:

- **Corrected multiplier:** the FIML projected saturated-score test described
  below, using the realized observed-data sensitivity and 199 Mammen draws.
- **Legacy multiplier:** the original expected-information projection with 199
  Rademacher draws. This remains the multiplier construction used for ML2S.
- **Scalar robust:** Yuan--Bentler/Mplus MLR for FIML and the scaled two-stage
  statistic for ML2S.
- **LR/discrepancy FMG:** SB, SS, pEBA(4), and `all`, computed from the robust
  eigenvalue spectrum of the FIML likelihood-ratio or ML2S Stage-2 discrepancy.
- **Score FMG:** the same four spectrum transforms applied to the global
  projected score statistic and its own robust eigenvalues.

### Corrected effective multiplier

This is a global score test of the fitted SEM against the local saturated
mean-and-covariance model. It is not a robust standard error and it does not
rescale the ordinary likelihood-ratio statistic. A companion
[working note](notes/multiplier_score_lineage.qmd) gives the score-test,
effective-score, wild-bootstrap, sign-randomization, and SEM lineages in detail.

Let the saturated moment vector be

$$
\eta =
\begin{pmatrix}
\operatorname{vech}(\Sigma) \\
\mu
\end{pmatrix}
\in \mathbb{R}^{m}.
$$

At the restricted estimate $\widehat\theta$, the method evaluates each case's
direct saturated observed-data score $s_i$. Under FIML, this is the likelihood
score for the variables actually observed for that case; missing values are not
filled in.

Let $\Delta=\partial\eta(\theta)/\partial\theta^{\mathsf T}$ be the local model
tangent, with affine equality constraints incorporated. If $K$ spans that
tangent, $D$ spans its Euclidean complement, and $A$ is the realized
observed-data sensitivity, define

$$
G_A=D-K(K^{\mathsf T}AK)^{-1}K^{\mathsf T}AD.
$$

Then $K^{\mathsf T}AG_A=0$, so $z_i=G_A^{\mathsf T}s_i$ retains only saturated
directions not locally explained by the SEM. The observed statistic is

$$
u=\sum_{i=1}^{n}z_i,
\qquad
V=G_A^{\mathsf T}IG_A,
\qquad
T=u^{\mathsf T}V^{-1}u.
$$

Its dimension is $d=m-\operatorname{rank}(\Delta)$, ordinarily the model df.
For multiplier draw $b$, the corrected procedure uses independent Mammen
weights with values $(1-\sqrt 5)/2$ and $(1+\sqrt 5)/2$ and respective
probabilities $(\sqrt 5+1)/(2\sqrt 5)$ and
$(\sqrt 5-1)/(2\sqrt 5)$. These weights satisfy
$E(w_{ib})=0$, $E(w_{ib}^2)=1$, and $E(w_{ib}^3)=1$. Each draw gives

$$
u_b=\sum_{i=1}^{n}w_{ib}z_i,
\qquad
T_b=u_b^{\mathsf T}V^{-1}u_b,
\qquad
\widehat p=
\frac{1+\sum_{b=1}^{B}\mathbf{1}(T_b\geq T)}{B+1}.
$$

Here $I$ is the positive-definite pattern-conditional Gaussian Fisher metric.
The realized Hessian $A$ is required for the nuisance correction, while
multiplier validity permits a separate fixed positive-definite quadratic.
Here $B=199$, so the p-value grid has width $1/200$. The fitted moments,
projection, and quadratic metric stay fixed. Every draw is only a weighted score
sum and a small quadratic form. Neither the null SEM nor a saturated alternative
is refitted. Unlike MLR's scalar adjustment of a likelihood-ratio statistic,
the multiplier reference retains casewise, direction-specific score variation
after removing the SEM tangent.

The retained legacy comparator also replaces $A$ by $I$ in the nuisance
projection and uses Rademacher weights. A separate scalar
MAR diagnostic found that the sensitivity choice accounted for the material
calibration change. Mammen and Rademacher weights were much closer once the
effective rows used the observed-data sensitivity.

For two-stage ML, the legacy gate is constructed from the two stages rather than
borrowing the FIML likelihood scores. Let $\widehat\eta_S$ be the saturated EM
moment estimate, $\phi_i$ its casewise influence rows, and
$W(\widehat\theta)$ the fitted normal-theory Stage-2 metric. The observed
saturated score is generated by
$\widehat\eta_S-\eta(\widehat\theta)$; multiplier rows propagate $\phi_i$
through the same fixed metric and tangent projection. Thus every draw still
avoids refitting, but missing-data variation enters through Stage-1 EM
influence rather than direct FIML scores. On complete data this construction
reproduces the ordinary ML observed statistic and multiplier p-value; its
robust spectrum is first-order equivalent because it retains moment influence
rather than raw likelihood-score rows.
An observed-sensitivity two-stage analogue has not been defined, so the
corrected multiplier is unavailable for ML2S.

## Paired estimators and the larger model

The runner draws each dataset once, builds the saturated EM stage once, and fits
both FIML and normal-theory two-stage ML to that shared input. The 15-indicator
CFA is included in the same factorial panel as the four smaller models. Pairing
removes Monte Carlo differences in the generated samples when the two
estimators are compared.

## VM and IG generators

```{r generator-table}
generator_table <- data.frame(
  code = c("normal", "vm1", "ig1", "vm2", "ig2"),
  generator = c(
    "Multivariate normal", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)"),
  marginal_skewness = c(0, 2, 2, 3, 3),
  marginal_excess_kurtosis = c(0, 7, 7, 21, 21),
  severity = c("reference", "moderate", "moderate", "severe", "severe"),
  stringsAsFactors = FALSE)
knitr::kable(generator_table, row.names = FALSE)
```

**VM** is the Vale--Maurelli/Fleishman construction. Magmaan solves an
intermediate Gaussian correlation matrix and applies calibrated marginal
polynomials to recover the SEM population covariance and requested marginal
moments.

**IG** is the Foldnes--Olsson independent-generator construction. Independent
Pearson-family nonnormal generators are calibrated and mixed through a
symmetric covariance root. VM and IG therefore have the same target covariance
and marginal skewness/kurtosis but different multivariate higher-order
dependence.

Each generator is observed completely, with 30% MCAR on every variable except
the first indicator or wave, or with a calibrated 30% MAR mechanism. The MAR
arm keeps the first two variables observed and makes missingness elsewhere
depend on them. Under nonnormal MAR it is deliberately a pseudo-ML stress arm:
normal-theory FIML and saturated normal EM need not target the population
moments even when the SEM covariance structure is correct.

## Appendix: numerical and computational details {.appendix}

### Method-specific denominators

Rejection rates use finite, method-specific denominators. Failures are not
counted as non-rejections.

```{r denominator-table}
denominators <- scorecard[, c("estimator", "method", "attempted", "finite")]
denominators <- denominators[denominators$finite > 0L, ]
denominators$non_finite <- denominators$attempted - denominators$finite
names(denominators) <- c(
  "estimator", "method", "attempted", "finite p-values",
  "non-finite or no fit")
knitr::kable(denominators, row.names = FALSE)
```

### Convergence and tangent rank

```{r geometry-table}
show_geometry <- geometry[, c(
  "estimator", "model_label", "attempted", "fit_ok", "flip_ok",
  "nominal_geometry")]
names(show_geometry) <- c(
  "estimator", "model", "attempted", "converged", "finite multiplier",
  "nominal tangent rank")
knitr::kable(show_geometry, row.names = FALSE)
```

The panel has four fit failures among 75,000 estimator rows: one FIML and three
ML2S. Every converged fit produced a finite legacy-multiplier p-value. Four FIML
and one ML2S call lost numerical tangent rank despite producing a finite value;
the table records these separately rather than silently counting them as
ordinary tests.

### Runtime and next experiment

The checkpointed panel took approximately 86 minutes on ten workers, including
75,000 fits, the LR/discrepancy and score-FMG batteries, and 199 multiplier
draws. The final 25-rep large-model severe-IG/MCAR chunk dominated the tail.
The corrected multiplier was not part of this timing.

### Reproduce

The archived panel above predates the corrected multiplier. To run the current
implementation into a fresh directory and render that expanded panel, point the
report at the same directory explicitly:

```sh
Rscript experiments/77-fiml-global-gof-pilot/run_sem_models.R --reps 500 --n 200 \
  --flips 199 --cores 10 --chunk-size 25 \
  --results-dir experiments/77-fiml-global-gof-pilot/results/sem-paired-corrected-score-n200-500
MAGMAAN_SEM_RESULT_DIR=experiments/77-fiml-global-gof-pilot/results/sem-paired-corrected-score-n200-500 \
  quarto render experiments/77-fiml-global-gof-pilot/sem_report.qmd
```
