---
title: "FIML goodness of fit in representative SEMs"
subtitle: "Vale--Maurelli and independent-generator null study"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- file.path(script_dir, "results", "sem-global-flip-500")
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
replications <- read_result("replications")
design <- read_result("design")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)

model_order <- c(
  "one_factor_6", "two_factor_fmg_10", "bifactor_12", "linear_growth_5")
model_labels <- c(
  one_factor_6 = "One-factor CFA",
  two_factor_fmg_10 = "Two-factor CFA",
  bifactor_12 = "Bifactor",
  linear_growth_5 = "Linear growth")
generator_order <- c("normal", "vm1", "ig1", "vm2", "ig2")
generator_labels <- c(
  normal = "Normal", vm1 = "VM moderate", ig1 = "IG moderate",
  vm2 = "VM severe", ig2 = "IG severe")
missingness_labels <- c(complete = "Complete", mcar_30 = "30% MCAR")
method_columns <- c(
  "Effective multiplier" = "p_flip_effective",
  "Yuan--Bentler MLR" = "p_lrt_mlr",
  "FMG SB" = "p_lrt_sb",
  "FMG SS" = "p_lrt_ss",
  "FMG pEBA(4)" = "p_lrt_peba4",
  "FMG all" = "p_lrt_all")

group_vars <- c("model_id", "distribution", "missingness")
group_id <- interaction(
  replications[group_vars], drop = TRUE, lex.order = TRUE)
cell_rows <- split(seq_len(nrow(replications)), group_id)

cell_table <- do.call(rbind, lapply(cell_rows, function(ii) {
  z <- replications[ii, , drop = FALSE]
  out <- data.frame(
    model_id = z$model_id[[1L]],
    model = unname(model_labels[[z$model_id[[1L]]]]),
    distribution = z$distribution[[1L]],
    generator = unname(generator_labels[[z$distribution[[1L]]]]),
    missingness_id = z$missingness[[1L]],
    missingness = unname(missingness_labels[[z$missingness[[1L]]]]),
    stringsAsFactors = FALSE, check.names = FALSE)
  for (method in names(method_columns)) {
    p_value <- z[[method_columns[[method]]]]
    out[[method]] <- mean(p_value < .05, na.rm = TRUE)
  }
  out
}))
cell_table$model_id <- factor(cell_table$model_id, levels = model_order)
cell_table$distribution <- factor(
  cell_table$distribution, levels = generator_order)
cell_table$missingness_id <- factor(
  cell_table$missingness_id, levels = c("complete", "mcar_30"))
cell_table <- cell_table[order(
  cell_table$model_id, cell_table$distribution, cell_table$missingness_id), ]
row.names(cell_table) <- NULL

scorecard <- do.call(rbind, lapply(names(method_columns), function(method) {
  rejection <- cell_table[[method]]
  p_value <- replications[[method_columns[[method]]]]
  data.frame(
    method = method,
    finite = sum(is.finite(p_value)),
    mean_cell_rejection = mean(rejection),
    mean_abs_size_error = mean(abs(rejection - .05)),
    min_cell_rejection = min(rejection),
    max_cell_rejection = max(rejection),
    cells_in_band = sum(rejection >= .025 & rejection <= .075),
    stringsAsFactors = FALSE)
}))

replications$nominal_geometry <- replications$flip_ok &
  replications$flip_df == replications$expected_df
geometry <- aggregate(
  cbind(fit_ok, flip_ok, nominal_geometry) ~ model_label,
  replications, sum)
geometry$attempted <- as.integer(table(factor(
  replications$model_label, levels = geometry$model_label)))
```

The question is deliberately narrow: how well calibrated are global
goodness-of-fit tests when correctly specified, ordinary latent-variable models
are fitted by FIML to normal, Vale--Maurelli, and independent-generator data,
with and without MCAR missingness? All entries below are Type-I rejection rates
at $\alpha=.05$. The design has 40 cells, $n=120$, and 500 replications per
cell.

## Headline result

```{r headline-table}
headline <- scorecard[, c(
  "method", "mean_cell_rejection", "mean_abs_size_error",
  "min_cell_rejection", "max_cell_rejection", "cells_in_band")]
names(headline) <- c(
  "method", "mean size", "mean |size-.05|", "minimum", "maximum",
  "cells in [.025,.075]")
knitr::kable(headline, digits = 3, row.names = FALSE)
```

The effective multiplier is the clear winner. Its mean rejection across the 40
null cells is .058, its mean absolute size error is .016, and 33/40 cells lie
in the practical [.025,.075] band. FMG `all` averages .065 but reaches .155;
FMG SS averages .072 and reaches .173. Yuan--Bentler MLR averages .364 and
reaches .930. The multiplier is therefore improving materially on the strongest
FMG corrections, not merely beating MLR.

## All 40 design cells

Each entry is the proportion of finite p-values below .05. Method-specific
finite denominators and numerical diagnostics are reported in the appendix.

```{r all-cells-table}
all_cells <- cell_table[, c(
  "model", "generator", "missingness", names(method_columns))]
knitr::kable(all_cells, digits = 3, row.names = FALSE)
```

The multiplier remains near .05 across most model and observation combinations.
Its most liberal cell is the bifactor under severe complete-data VM (.122), and
its most conservative is the bifactor under normal MCAR (.018). VM is generally
harder than IG despite matched marginal moments. MLR is liberal even under the
normal reference and deteriorates sharply with nonnormality; FMG SS and `all`
occupy the middle ground.

## Model specifications

```{r model-table}
model_table <- unique(design[, c(
  "model_id", "model_label", "p", "expected_df")])
model_table$free_parameters <- c(18L, 31L, 48L, 10L)[match(
  model_table$model_id, model_order)]
model_table$model_id <- factor(model_table$model_id, levels = model_order)
model_table <- model_table[order(model_table$model_id), ]
model_table$model <- unname(model_labels[as.character(model_table$model_id)])
model_table <- model_table[, c(
  "model", "p", "free_parameters", "expected_df")]
names(model_table) <- c("model", "observed variables", "free parameters", "df")
knitr::kable(model_table, row.names = FALSE)
```

1. **One-factor CFA, six indicators.** The population is
   $x_j=\lambda_j f+\varepsilon_j$, with
   $\lambda=(1,.8,.9,.7,1.1,.85)$, $\operatorname{Var}(f)=1$, and unique
   variances $(.5,.6,.55,.65,.5,.6)$. Indicator means are zero. The fitted CFA
   uses marker-variable identification and freely estimates means, unique
   variances, the factor variance, and the remaining loadings.

2. **Correlated two-factor CFA, ten indicators.** Indicators 1--5 load only on
   factor 1 with loadings $(.4,.5,.6,.8,.4)$; indicators 6--10 load only on
   factor 2 with loadings $(.4,.7,.6,.4,.8)$. Factor variances are one and the
   factor correlation is .5. Unique variances are $1-\lambda_j^2$, giving unit
   marginal variances, and all means are zero.

3. **Orthogonal bifactor model, twelve indicators.** Every indicator loads .6
   on a general factor. Three orthogonal specific factors each cover four
   indicators, with blockwise loadings .5, .4, and .45. All factor variances are
   one; unique variances are $1-.6^2-\lambda_{S,j}^2$ and means are zero. The
   fitted model uses `std.lv` identification.

4. **Five-wave linear growth model.** At $t=0,1,2,3,4$,
   $y_t=i+t s+\varepsilon_t$. The intercept/slope means are $(1,.2)$, their
   covariance matrix is

   $$
   \begin{pmatrix}
   1 & 0.1 \\
   0.1 & 0.1
   \end{pmatrix},
   $$

   and each residual variance is .5. The fitted model fixes the time scores
   and freely estimates the growth means, latent variance/covariance terms,
   and residual variances.

## Methods

Each dataset is fitted once by FIML and passed to six global-fit calibrations:

- **Effective multiplier:** the projected saturated-score test described below,
  using 199 Rademacher draws.
- **Yuan--Bentler/Mplus MLR:** the dedicated robust FIML likelihood-ratio
  statistic.
- **FMG corrections:** SB, SS, pEBA(4), and `all`, computed from the estimated
  eigenvalue spectrum of the null statistic.

### Effective multiplier

This is a global score test of the fitted SEM against the local saturated
mean-and-covariance model. It is not a robust standard error and it does not
rescale the ordinary likelihood-ratio statistic.

Let the saturated moment vector be

$$
\eta =
\begin{pmatrix}
\operatorname{vech}(\Sigma) \\
\mu
\end{pmatrix}
\in \mathbb{R}^{m}.
$$

At the restricted estimate $\widehat\theta$, the method evaluates each case's
direct saturated observed-data score $s_i$. Under FIML, this is the likelihood
score for the variables actually observed for that case; missing values are not
filled in.

Let $\Delta=\partial\eta(\theta)/\partial\theta^{\mathsf T}$ be the local model
tangent, with affine equality constraints incorporated. If $K$ spans that
tangent, $D$ spans its Euclidean complement, and $I$ is the saturated-moment
Fisher information summed over observed-data patterns, define

$$
G=D-K(K^{\mathsf T}IK)^{-1}K^{\mathsf T}ID.
$$

Then $K^{\mathsf T}IG=0$, so $z_i=G^{\mathsf T}s_i$ retains only saturated
directions not locally explained by the SEM. The observed statistic is

$$
u=\sum_{i=1}^{n}z_i,
\qquad
V=G^{\mathsf T}IG,
\qquad
T=u^{\mathsf T}V^{-1}u.
$$

Its dimension is $d=m-\operatorname{rank}(\Delta)$, ordinarily the model df.
For multiplier draw $b$, independent $w_{ib}\in\{-1,+1\}$ give

$$
u_b=\sum_{i=1}^{n}w_{ib}z_i,
\qquad
T_b=u_b^{\mathsf T}V^{-1}u_b,
\qquad
\widehat p=
\frac{1+\sum_{b=1}^{B}\mathbf{1}(T_b\geq T)}{B+1}.
$$

Here $B=199$, so the p-value grid has width $1/200$. The fitted moments,
projection, and Fisher metric stay fixed: every draw is only a weighted score
sum and a small quadratic form. Neither the null SEM nor a saturated alternative
is refitted. Unlike MLR's scalar adjustment of a likelihood-ratio statistic,
the multiplier reference retains casewise, direction-specific score variation
after removing the SEM tangent.

## VM and IG generators

```{r generator-table}
generator_table <- data.frame(
  code = c("normal", "vm1", "ig1", "vm2", "ig2"),
  generator = c(
    "Multivariate normal", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)"),
  marginal_skewness = c(0, 2, 2, 3, 3),
  marginal_excess_kurtosis = c(0, 7, 7, 21, 21),
  severity = c("reference", "moderate", "moderate", "severe", "severe"),
  stringsAsFactors = FALSE)
knitr::kable(generator_table, row.names = FALSE)
```

**VM** is the Vale--Maurelli/Fleishman construction. Magmaan solves an
intermediate Gaussian correlation matrix and applies calibrated marginal
polynomials to recover the SEM population covariance and requested marginal
moments.

**IG** is the Foldnes--Olsson independent-generator construction. Independent
Pearson-family nonnormal generators are calibrated and mixed through a
symmetric covariance root. VM and IG therefore have the same target covariance
and marginal skewness/kurtosis but different multivariate higher-order
dependence.

Each generator is observed completely or with 30% MCAR on every variable except
the first indicator or wave, which is always observed.

## Appendix: numerical and computational details {.appendix}

### Method-specific denominators

Rejection rates use finite, method-specific denominators. Failures are not
counted as non-rejections.

```{r denominator-table}
denominators <- scorecard[, c("method", "finite")]
denominators$non_finite <- nrow(replications) - denominators$finite
names(denominators) <- c("method", "finite p-values", "non-finite or no fit")
knitr::kable(denominators, row.names = FALSE)
```

### Convergence and tangent rank

```{r geometry-table}
show_geometry <- geometry[, c(
  "model_label", "attempted", "fit_ok", "flip_ok", "nominal_geometry")]
names(show_geometry) <- c(
  "model", "attempted", "FIML converged", "finite multiplier",
  "nominal tangent rank")
knitr::kable(show_geometry, row.names = FALSE)
```

Nine of 20,000 FIML fits did not converge. Every converged fit produced a finite
multiplier p-value. In 25 of the 19,991 finite calls, however, the model Jacobian
lost numerical rank and raised the tested df above its prespecified value.
Nineteen occurred in the two-factor CFA under MCAR, two in that model under
VM1/complete data, and four in the one-factor VM2/MCAR cell.

Sixteen of the 25 rank-anomalous calls rejected. Requiring the nominal tangent
rank leaves 19,966 calls and changes mean cell rejection only from .0584 to
.0577, with the same 33/40 cells in [.025,.075]. The anomaly does not drive the
headline result, but future power work should treat it as a separate fail-closed
denominator.

### Runtime and next experiment

The 20,000-fit panel took 446 seconds (7.4 minutes) on four workers, including
FIML, MLR, FMG, and 199 multiplier draws. The observed parallel speedup was
3.68. The next scientific experiment is a matched power panel with prespecified
sparse and diffuse SEM misspecifications, reporting raw and size-adjusted power
plus the same finite and tangent-rank denominators.

### Reproduce

```sh
Rscript experiments/77-fiml-global-gof-pilot/run_sem_models.R --reps 500 \
  --flips 199 \
  --cores 4 \
  --results-dir experiments/77-fiml-global-gof-pilot/results/sem-global-flip-500
quarto render experiments/77-fiml-global-gof-pilot/sem_report.qmd
```
