---
title: "FIML and two-stage goodness of fit in representative SEMs"
subtitle: "Score-spectrum, multiplier, and discrepancy calibrations under complete, MCAR, and MAR sampling"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- Sys.getenv(
  "MAGMAAN_SEM_RESULT_DIR",
  unset = file.path(
    script_dir, "results", "sem-paired-score-fmg-n200-500"
  )
)
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
replications <- read_result("replications")
design <- read_result("design")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)

model_order <- c(
  "one_factor_6", "two_factor_fmg_10", "bifactor_12", "three_factor_15",
  "linear_growth_5")
model_labels <- c(
  one_factor_6 = "One-factor CFA",
  two_factor_fmg_10 = "Two-factor CFA",
  bifactor_12 = "Bifactor",
  three_factor_15 = "Three-factor CFA",
  linear_growth_5 = "Linear growth")
estimator_order <- c("FIML", "ML2S")
generator_order <- c("normal", "vm1", "ig1", "vm2", "ig2")
generator_labels <- c(
  normal = "Normal", vm1 = "VM moderate", ig1 = "IG moderate",
  vm2 = "VM severe", ig2 = "IG severe")
missingness_labels <- c(
  complete = "Complete", mcar_30 = "30% MCAR", mar_30 = "30% MAR")
method_columns <- c(
  "Observed-sensitivity Mammen multiplier" = "p_flip_corrected",
  "Expected-sensitivity Mammen multiplier" = "p_flip_expected_mammen",
  "Expected-sensitivity Rademacher multiplier" = "p_flip_effective",
  "Naive LR/D" = "p_lrt_naive",
  "Scalar robust" = "p_lrt_mlr",
  "LR/D FMG SB" = "p_lrt_sb",
  "LR/D FMG SS" = "p_lrt_ss",
  "LR/D FMG pEBA(4)" = "p_lrt_peba4",
  "LR/D FMG all" = "p_lrt_all",
  "Observed-score FMG SB" = "p_score_corrected_sb",
  "Observed-score FMG SS" = "p_score_corrected_ss",
  "Observed-score FMG pEBA(4)" = "p_score_corrected_peba4",
  "Observed-score FMG all" = "p_score_corrected_all",
  "Score FMG SB" = "p_score_sb",
  "Score FMG SS" = "p_score_ss",
  "Score FMG pEBA(4)" = "p_score_peba4",
  "Score FMG all" = "p_score_all")
method_columns <- method_columns[method_columns %in% names(replications)]

group_vars <- c("model_id", "distribution", "missingness", "estimator")
group_id <- interaction(
  replications[group_vars], drop = TRUE, lex.order = TRUE)
cell_rows <- split(seq_len(nrow(replications)), group_id)

cell_estimator_table <- do.call(rbind, lapply(cell_rows, function(ii) {
  z <- replications[ii, , drop = FALSE]
  out <- data.frame(
    model_id = z$model_id[[1L]],
    model = unname(model_labels[[z$model_id[[1L]]]]),
    distribution = z$distribution[[1L]],
    generator = unname(generator_labels[[z$distribution[[1L]]]]),
    missingness_id = z$missingness[[1L]],
    missingness = unname(missingness_labels[[z$missingness[[1L]]]]),
    analysis_region = if (
      z$missingness[[1L]] == "mar_30" && z$distribution[[1L]] != "normal"
    ) "Nonnormal MAR stress" else "Identified null",
    estimator = z$estimator[[1L]],
    stringsAsFactors = FALSE, check.names = FALSE)
  for (method in names(method_columns)) {
    p_value <- z[[method_columns[[method]]]]
    out[[method]] <- mean(p_value < .05, na.rm = TRUE)
  }
  out
}))
cell_estimator_table$model_id <- factor(
  cell_estimator_table$model_id, levels = model_order)
cell_estimator_table$distribution <- factor(
  cell_estimator_table$distribution, levels = generator_order)
cell_estimator_table$missingness_id <- factor(
  cell_estimator_table$missingness_id,
  levels = c("complete", "mcar_30", "mar_30"))
cell_estimator_table$estimator <- factor(
  cell_estimator_table$estimator, levels = estimator_order)
cell_estimator_table <- cell_estimator_table[order(
  cell_estimator_table$model_id, cell_estimator_table$distribution,
  cell_estimator_table$missingness_id, cell_estimator_table$estimator), ]
row.names(cell_estimator_table) <- NULL

cell_key <- unique(cell_estimator_table[, c(
  "model_id", "model", "distribution", "generator", "missingness_id",
  "missingness")])
cell_table <- cell_key
wide_method_columns <- character()
for (estimator in estimator_order) {
  source <- cell_estimator_table[
    as.character(cell_estimator_table$estimator) == estimator, ]
  source_key <- interaction(
    source[c("model_id", "distribution", "missingness_id")],
    drop = TRUE, lex.order = TRUE)
  target_key <- interaction(
    cell_table[c("model_id", "distribution", "missingness_id")],
    drop = TRUE, lex.order = TRUE)
  index <- match(target_key, source_key)
  for (method in names(method_columns)) {
    if (!any(is.finite(source[[method]]))) next
    display_method <- if (method == "Scalar robust") {
      if (estimator == "FIML") "Yuan--Bentler MLR" else "Scaled ML2S"
    } else method
    column <- paste(estimator, display_method, sep = " / ")
    cell_table[[column]] <- source[[method]][index]
    wide_method_columns <- c(wide_method_columns, column)
  }
}

scorecard <- do.call(rbind, lapply(estimator_order, function(estimator) {
  z <- cell_estimator_table[
    as.character(cell_estimator_table$estimator) == estimator, ]
  raw <- replications[replications$estimator == estimator, ]
  do.call(rbind, lapply(names(method_columns), function(method) {
    rejection <- z[[method]]
    p_value <- raw[[method_columns[[method]]]]
    display_method <- if (method == "Scalar robust") {
      if (estimator == "FIML") "Yuan--Bentler MLR" else "Scaled ML2S"
    } else method
    finite_rejection <- rejection[is.finite(rejection)]
    data.frame(
      estimator = estimator, method = display_method,
      attempted = nrow(raw),
      finite = sum(is.finite(p_value)),
      mean_cell_rejection = if (length(finite_rejection)) {
        mean(finite_rejection)
      } else NA_real_,
      mean_abs_size_error = if (length(finite_rejection)) {
        mean(abs(finite_rejection - .05))
      } else NA_real_,
      min_cell_rejection = if (length(finite_rejection)) {
        min(finite_rejection)
      } else NA_real_,
      max_cell_rejection = if (length(finite_rejection)) {
        max(finite_rejection)
      } else NA_real_,
      cells_in_band = sum(
        rejection >= .025 & rejection <= .075, na.rm = TRUE),
      stringsAsFactors = FALSE)
  }))
}))

summarize_region <- function(region) {
  do.call(rbind, lapply(estimator_order, function(estimator) {
    z <- cell_estimator_table[
      as.character(cell_estimator_table$estimator) == estimator &
        cell_estimator_table$analysis_region == region, ]
    do.call(rbind, lapply(names(method_columns), function(method) {
      rejection <- z[[method]]
      rejection <- rejection[is.finite(rejection)]
      display_method <- if (method == "Scalar robust") {
        if (estimator == "FIML") "Yuan--Bentler MLR" else "Scaled ML2S"
      } else method
      data.frame(
        estimator = estimator, method = display_method,
        cells = length(rejection),
        mean_abs_size_error = if (length(rejection)) {
          mean(abs(rejection - .05))
        } else NA_real_,
        min_cell_rejection = if (length(rejection)) min(rejection) else NA_real_,
        max_cell_rejection = if (length(rejection)) max(rejection) else NA_real_,
        mean_cell_rejection = if (length(rejection)) mean(rejection) else NA_real_,
        stringsAsFactors = FALSE)
    }))
  }))
}
identified_scorecard <- summarize_region("Identified null")
mar_stress_scorecard <- summarize_region("Nonnormal MAR stress")

replications$nominal_geometry <- replications$flip_ok &
  replications$flip_df == replications$expected_df
replications$corrected_nominal_geometry <-
  replications$flip_corrected_ok &
  replications$flip_corrected_df == replications$expected_df
geometry <- aggregate(
  cbind(fit_ok, flip_ok, nominal_geometry, flip_corrected_ok,
        corrected_nominal_geometry) ~ estimator + model_label,
  replications, sum)
geometry$attempted <- vapply(seq_len(nrow(geometry)), function(i) {
  sum(replications$estimator == geometry$estimator[[i]] &
        replications$model_label == geometry$model_label[[i]])
}, integer(1L))
```

This is a preliminary null-calibration study of global goodness-of-fit tests in
five ordinary latent-variable models. The design crosses normal,
Vale--Maurelli, and independent-generator data with complete sampling, 30%
MCAR, and observed-predictor MAR at $n=200$, with 500 replications per cell.

The main calibration region contains 55 cells: every complete and MCAR cell,
plus normal-data MAR. These are the cells in which the generating moments are
also the relevant null target for Gaussian FIML. The remaining 20 nonnormal-MAR
cells are retained as an estimand stress test, not interpreted as ordinary
Type-I error: under nonnormal MAR, Gaussian FIML can converge to pseudo-true
moments outside the generating SEM.

## Preliminary story

### 1. Projected-score pEBA(4) is the strongest FIML result

```{r fiml-primary-table}
fiml_primary <- identified_scorecard[
  identified_scorecard$estimator == "FIML" &
    is.finite(identified_scorecard$mean_abs_size_error),
  c("method", "mean_abs_size_error", "min_cell_rejection",
    "max_cell_rejection")]
fiml_primary <- fiml_primary[order(fiml_primary$mean_abs_size_error), ]
names(fiml_primary) <- c("method", "mean |size-.05|", "minimum", "maximum")
knitr::kable(fiml_primary, digits = 3, row.names = FALSE)
```

Projected-score pEBA(4) has the smallest mean absolute size error (.013), and
its 55-cell range stops at .074. The expected-sensitivity Rademacher multiplier
from the completed panel is close on average
(.014) but reaches .094, while projected-score SB is also competitive (.015,
range .034--.092). The pEBA advantage over the multiplier is not cellwise
uniform; it comes from containing the larger errors rather than winning most
individual cells. The multiplier is therefore useful as a benchmark, but the
preliminary result does not require it as the lead method.

### 2. The correction and the base statistic cannot be separated

```{r score-lrt-crossing-table}
crossing <- data.frame(
  correction = c("SB", "SS", "pEBA(4)", "all"),
  score_method = c("Score FMG SB", "Score FMG SS",
                   "Score FMG pEBA(4)", "Score FMG all"),
  lrt_method = c("LR/D FMG SB", "LR/D FMG SS",
                 "LR/D FMG pEBA(4)", "LR/D FMG all"),
  stringsAsFactors = FALSE)
take_metric <- function(method, metric) identified_scorecard[
  identified_scorecard$estimator == "FIML" &
    identified_scorecard$method == method, metric]
crossing$score_mae <- vapply(
  crossing$score_method, take_metric, numeric(1L),
  metric = "mean_abs_size_error")
crossing$score_min <- vapply(
  crossing$score_method, take_metric, numeric(1L),
  metric = "min_cell_rejection")
crossing$score_max <- vapply(
  crossing$score_method, take_metric, numeric(1L),
  metric = "max_cell_rejection")
crossing$lrt_mae <- vapply(
  crossing$lrt_method, take_metric, numeric(1L),
  metric = "mean_abs_size_error")
crossing$lrt_min <- vapply(
  crossing$lrt_method, take_metric, numeric(1L),
  metric = "min_cell_rejection")
crossing$lrt_max <- vapply(
  crossing$lrt_method, take_metric, numeric(1L),
  metric = "max_cell_rejection")
crossing <- crossing[, c(
  "correction", "score_mae", "score_min", "score_max",
  "lrt_mae", "lrt_min", "lrt_max")]
names(crossing) <- c(
  "correction", "score MAE", "score min", "score max",
  "LR MAE", "LR min", "LR max")
knitr::kable(crossing, digits = 3, row.names = FALSE)
```

There is no generic ordering of scores above likelihood ratios. SB and pEBA(4)
work with the projected-score statistic, whereas score SS and score `all` are
strongly conservative. For the likelihood-ratio statistic the ordering
reverses: SS and `all` are the viable corrections, while SB and pEBA(4) are
liberal. In particular, LR SS does beat score SS, but the best projected-score
calibration beats the best likelihood-ratio calibration. The large reversal is
important enough to require a focused formula and implementation audit before
it becomes more than a preliminary finding.

### 3. Two-stage ML does not improve the useful methods

```{r estimator-comparison-table}
selected_methods <- c(
  "Score FMG pEBA(4)", "Score FMG SB", "LR/D FMG all", "LR/D FMG SS",
  "Expected-sensitivity Rademacher multiplier")
estimator_comparison <- do.call(rbind, lapply(selected_methods, function(method) {
  rows <- identified_scorecard[identified_scorecard$method == method, ]
  data.frame(
    method = method,
    FIML_MAE = rows$mean_abs_size_error[rows$estimator == "FIML"],
    FIML_min = rows$min_cell_rejection[rows$estimator == "FIML"],
    FIML_max = rows$max_cell_rejection[rows$estimator == "FIML"],
    ML2S_MAE = rows$mean_abs_size_error[rows$estimator == "ML2S"],
    ML2S_min = rows$min_cell_rejection[rows$estimator == "ML2S"],
    ML2S_max = rows$max_cell_rejection[rows$estimator == "ML2S"])
}))
names(estimator_comparison) <- c(
  "method", "FIML MAE", "FIML min", "FIML max",
  "ML2S MAE", "ML2S min", "ML2S max")
knitr::kable(estimator_comparison, digits = 3, row.names = FALSE)
```

ML2S is nearly indistinguishable from FIML under complete sampling, then
generally deteriorates under missingness. In the identified-null region,
projected-score pEBA(4) moves from .013 to .029 mean absolute error and its
maximum rejection rises from .074 to .162; projected-score SB deteriorates
more sharply. ML2S is therefore a secondary negative result rather than part
of the main replacement-method story.

## Next experiment: fair multiplier geometry and matched power

The completed panel predates magmaan's observed-sensitivity projection. Its
score-FMG rows and legacy multiplier use expected sensitivity. That remains the
primary geometry in the 55-cell identified-null region: there the fitted SEM
moments are the Gaussian-score target and expected and observed sensitivity
coincide in the population. A 100-rep normal audit of the new code found that
plugging the realized sensitivity into the projection can be extremely
conservative at $n=200$ in the higher-df models, so it will not silently replace
the preliminary score tests.

The revised runner instead records a three-corner multiplier ablation:
expected-sensitivity Rademacher, expected-sensitivity Mammen, and
observed-sensitivity Mammen. It also records the analytic score calibrations
under both expected and observed projection geometry. This separates weight-law
changes from sensitivity changes. The observed-sensitivity rows are diagnostics
for a genuinely defined pseudo-true null; they cannot make the present
nonnormal-MAR generating SEM into such a null.

Power will be studied under two prespecified misspecifications for each model:

- a **sparse** alternative consisting of one omitted residual covariance; and
- a **diffuse** alternative consisting of an omitted method factor on a small,
  model-specific set of indicators or waves.

```{r power-design-table}
power_design <- data.frame(
  model = unname(model_labels[model_order]),
  sparse = c("x1 ~~ x2", "x1 ~~ x6", "x1 ~~ x5", "x1 ~~ x6",
             "y2 ~~ y3"),
  sparse_effect = c(.1068, .2391, .1562, .1881, .0684),
  diffuse = c("x1, x3, x5", "x1, x3, x6, x8", "x1, x5, x9",
              "x1, x6, x11", "y1, y3, y5"),
  diffuse_effect = c(.1433, .1239, .1264, .1586, .0688),
  stringsAsFactors = FALSE)
names(power_design) <- c(
  "model", "sparse residual pair", "effect", "diffuse method set", "effect")
knitr::kable(power_design, digits = 4, row.names = FALSE)
```

The sparse effect is the added residual covariance as a fraction of the two
variables' baseline standard deviations. The diffuse effect is the variance of
a method factor whose listed loadings equal the variables' baseline standard
deviations. Thus both perturbations are scale-free, but they represent visibly
different local failures: one concentrated residual edge versus shared
cross-block dependence.

Within each model and alternative, the perturbation is calibrated by population
normal-theory ML so that the ordinary likelihood-ratio test has 50% asymptotic
power at $n=200$. Equivalently, the population discrepancy satisfies
$(n-1)F_0=\lambda_{.50}$, where $\lambda_{.50}$ solves the noncentral
$\chi^2_{df}(\lambda)$ power equation at $\alpha=.05$. This common reference
does not tune the design in favor of any robust method. Null, sparse, and
diffuse cells share random-number seeds. The primary power comparison uses each
method's own empirical-null fifth-percentile cutoff in the matching model,
generator, and missingness cell; nominal-threshold power remains descriptive.

The main run is FIML-only over the 55 identified-null mechanisms. Nonnormal MAR
is a separately labeled estimand-stress phase, and ML2S power is deferred unless
the secondary estimator story becomes independently interesting. A 100-rep
normal-complete audit put ordinary-LR rejection near the intended .50 under
both alternatives in all five models, which is sufficient to validate the
population calibration before the larger run.

## Model specifications

```{r model-table}
model_table <- unique(design[, c(
  "model_id", "model_label", "p", "expected_df")])
model_table$free_parameters <- c(18L, 31L, 48L, 48L, 10L)[match(
  model_table$model_id, model_order)]
model_table$model_id <- factor(model_table$model_id, levels = model_order)
model_table <- model_table[order(model_table$model_id), ]
model_table$model <- unname(model_labels[as.character(model_table$model_id)])
model_table <- model_table[, c(
  "model", "p", "free_parameters", "expected_df")]
names(model_table) <- c("model", "observed variables", "free parameters", "df")
knitr::kable(model_table, row.names = FALSE)
```

1. **One-factor CFA, six indicators.** The population is
   $x_j=\lambda_j f+\varepsilon_j$, with
   $\lambda=(1,.8,.9,.7,1.1,.85)$, $\operatorname{Var}(f)=1$, and unique
   variances $(.5,.6,.55,.65,.5,.6)$. Indicator means are zero. The fitted CFA
   uses marker-variable identification and freely estimates means, unique
   variances, the factor variance, and the remaining loadings.

2. **Correlated two-factor CFA, ten indicators.** Indicators 1--5 load only on
   factor 1 with loadings $(.4,.5,.6,.8,.4)$; indicators 6--10 load only on
   factor 2 with loadings $(.4,.7,.6,.4,.8)$. Factor variances are one and the
   factor correlation is .5. Unique variances are $1-\lambda_j^2$, giving unit
   marginal variances, and all means are zero.

3. **Orthogonal bifactor model, twelve indicators.** Every indicator loads .6
   on a general factor. Three orthogonal specific factors each cover four
   indicators, with blockwise loadings .5, .4, and .45. All factor variances are
   one; unique variances are $1-.6^2-\lambda_{S,j}^2$ and means are zero. The
   fitted model uses `std.lv` identification.

4. **Five-wave linear growth model.** At $t=0,1,2,3,4$,
   $y_t=i+t s+\varepsilon_t$. The intercept/slope means are $(1,.2)$, their
   covariance matrix is

   $$
   \begin{pmatrix}
   1 & 0.1 \\
   0.1 & 0.1
   \end{pmatrix},
   $$

   and each residual variance is .5. The fitted model fixes the time scores
   and freely estimates the growth means, latent variance/covariance terms,
   and residual variances.

5. **Three-factor CFA, fifteen indicators.** Five indicators
   load on each of three correlated factors. Within-factor loadings range from
   .4 to .8; factor correlations are .3, .4, and .5; residual variances give
   unit indicator variances. The fitted model uses `std.lv` identification and
   freely estimates 15 loadings, 15 residual variances, 15 means, and three
   factor covariances: 48 parameters against 135 saturated moments, hence 87
   df.

## Methods

Each generated dataset in the completed panel is fitted by both FIML and ML2S
and yields eleven finite global-fit calibrations: naive LR/discrepancy, the
scalar robust test, four LR/discrepancy FMG variants, four projected-score FMG
variants, and the expected-sensitivity Rademacher multiplier. The current code additionally
computes expected-sensitivity Mammen and observed-sensitivity Mammen
multipliers plus four score-FMG calibrations from the observed-sensitivity
geometry. Those new results landed after this production panel started and are
therefore described but not included in the preliminary tables:

- **Observed-sensitivity multiplier:** the FIML projected saturated-score test
  described below, using realized observed-data sensitivity and Mammen draws.
- **Expected-sensitivity Mammen:** the weight-law control, using the legacy
  projection with Mammen rather than Rademacher draws.
- **Expected-sensitivity Rademacher multiplier:** the completed panel's
  construction, using 199 Rademacher draws. This remains the multiplier
  construction used for ML2S.
- **Scalar robust:** Yuan--Bentler/Mplus MLR for FIML and the scaled two-stage
  statistic for ML2S.
- **LR/discrepancy FMG:** SB, SS, pEBA(4), and `all`, computed from the robust
  eigenvalue spectrum of the FIML likelihood-ratio or ML2S Stage-2 discrepancy.
- **Score FMG:** the same four spectrum transforms applied to the global
  projected score statistic and its own robust eigenvalues. The completed panel
  uses expected sensitivity. The next panel retains those rows as primary and
  records observed-sensitivity versions as a finite-sample diagnostic.

### Multiplier status

The multiplier is now a benchmark rather than the organizing method. The
legacy panel used an expected-information nuisance projection and Rademacher
weights. Current magmaan can instead use the realized observed-data
sensitivity for that projection and Mammen weights. The next run crosses these
changes through the three-corner ablation above rather than presenting them as
one indivisible correction.
The construction and the unavailable ML2S analogue are summarized in the
appendix; the full lineage remains in the companion
[working note](notes/multiplier_score_lineage.qmd).

## Paired estimators and the larger model

The runner draws each dataset once, builds the saturated EM stage once, and fits
both FIML and normal-theory two-stage ML to that shared input. The 15-indicator
CFA is included in the same factorial panel as the four smaller models. Pairing
removes Monte Carlo differences in the generated samples when the two
estimators are compared.

## VM and IG generators

```{r generator-table}
generator_table <- data.frame(
  code = c("normal", "vm1", "ig1", "vm2", "ig2"),
  generator = c(
    "Multivariate normal", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)"),
  marginal_skewness = c(0, 2, 2, 3, 3),
  marginal_excess_kurtosis = c(0, 7, 7, 21, 21),
  severity = c("reference", "moderate", "moderate", "severe", "severe"),
  stringsAsFactors = FALSE)
knitr::kable(generator_table, row.names = FALSE)
```

**VM** is the Vale--Maurelli/Fleishman construction. Magmaan solves an
intermediate Gaussian correlation matrix and applies calibrated marginal
polynomials to recover the SEM population covariance and requested marginal
moments.

**IG** is the Foldnes--Olsson independent-generator construction. Independent
Pearson-family nonnormal generators are calibrated and mixed through a
symmetric covariance root. VM and IG therefore have the same target covariance
and marginal skewness/kurtosis but different multivariate higher-order
dependence.

Each generator is observed completely, with 30% MCAR on every variable except
the first indicator or wave, or with a calibrated 30% MAR mechanism. The MAR
arm keeps the first two variables observed and makes missingness elsewhere
depend on them. Under nonnormal MAR it is deliberately a pseudo-ML stress arm:
normal-theory FIML and saturated normal EM need not target the population
moments even when the SEM covariance structure is correct.

## Appendix: numerical and computational details {.appendix}

### Complete 75-cell scorecard

This table reproduces the former headline across all 75 cells and both
estimators. It mixes the identified-null region with the 20 nonnormal-MAR
estimand-stress cells and is therefore descriptive rather than a pure Type-I
summary.

```{r headline-table}
headline <- scorecard[, c(
  "estimator", "method", "mean_cell_rejection", "mean_abs_size_error",
  "min_cell_rejection", "max_cell_rejection", "cells_in_band")]
headline <- headline[scorecard$finite > 0L, ]
names(headline) <- c(
  "estimator", "method", "mean size", "mean |size-.05|", "minimum", "maximum",
  "cells in [.025,.075]")
knitr::kable(headline, digits = 3, row.names = FALSE)
```

### Nonnormal-MAR estimand stress

```{r mar-stress-table}
mar_stress <- mar_stress_scorecard[
  is.finite(mar_stress_scorecard$mean_abs_size_error),
  c("estimator", "method", "mean_abs_size_error",
    "min_cell_rejection", "max_cell_rejection")]
names(mar_stress) <- c(
  "estimator", "method", "mean |rate-.05|", "minimum", "maximum")
knitr::kable(mar_stress, digits = 3, row.names = FALSE)
```

These rates describe rejection relative to the generating SEM moments, not
necessarily Type-I error for the Gaussian-FIML pseudo-true target.

### All 75 design cells

Each entry is the proportion of finite p-values below .05. The method closest
to .05 is bold in each row; exact ties are all bold. The table is intentionally
wide.

```{r all-cells-table}
all_cells <- cell_table[, c(
  "model", "generator", "missingness", wide_method_columns)]
cell_rates <- as.matrix(all_cells[, wide_method_columns])
cell_error <- abs(cell_rates - .05)
best_error <- apply(cell_error, 1L, min, na.rm = TRUE)
for (method in wide_method_columns) {
  value <- cell_rates[, method]
  rendered <- sprintf("%.3f", value)
  rendered[!is.finite(value)] <- "---"
  is_winner <- is.finite(cell_error[, method]) &
    abs(cell_error[, method] - best_error) < 1e-12
  rendered[is_winner] <- paste0("<strong>", rendered[is_winner], "</strong>")
  all_cells[[method]] <- rendered
}
knitr::kable(
  all_cells, row.names = FALSE, escape = FALSE,
  align = c("l", "l", "l", rep("r", length(wide_method_columns))))
```

### Multiplier constructions

Let $s_i$ be the direct saturated observed-data score at the restricted FIML
fit, $K$ span the SEM tangent, $D$ its Euclidean complement, $A$ the realized
observed-data sensitivity, and $I$ the fixed pattern-conditional Fisher metric.
The corrected projection is

$$
G_A=D-K(K^{\mathsf T}AK)^{-1}K^{\mathsf T}AD,
\qquad K^{\mathsf T}AG_A=0.
$$

With $z_i=G_A^{\mathsf T}s_i$, $u=\sum_i z_i$, and
$V=G_A^{\mathsf T}IG_A$, the observed statistic is
$T=u^{\mathsf T}V^{-1}u$. Mammen draws replace $u$ by
$u_b=\sum_iw_{ib}z_i$ while retaining the fitted moments, projection, and
quadratic metric. The Monte Carlo p-value is

$$
\widehat p=\frac{1+\sum_{b=1}^{B}\mathbf{1}(T_b\geq T)}{B+1}.
$$

Observed-geometry score-FMG diagnostics in the next run use this same $G_A$,
observed statistic, and robust spectrum. The primary identified-null score rows
retain the expected-sensitivity projection. Expected-sensitivity Mammen and
Rademacher multipliers isolate the weight law, while expected- versus
observed-sensitivity Mammen isolates the projection. No observed-sensitivity
two-stage analogue is currently defined; ML2S retains its Stage-1
influence/Stage-2 discrepancy construction.

### Method-specific denominators

Rejection rates use finite, method-specific denominators. Failures are not
counted as non-rejections.

```{r denominator-table}
denominators <- scorecard[, c("estimator", "method", "attempted", "finite")]
denominators <- denominators[denominators$finite > 0L, ]
denominators$non_finite <- denominators$attempted - denominators$finite
names(denominators) <- c(
  "estimator", "method", "attempted", "finite p-values",
  "non-finite or no fit")
knitr::kable(denominators, row.names = FALSE)
```

### Convergence and tangent rank

```{r geometry-table}
show_geometry <- geometry[, c(
  "estimator", "model_label", "attempted", "fit_ok", "flip_ok",
  "nominal_geometry")]
names(show_geometry) <- c(
  "estimator", "model", "attempted", "converged", "finite multiplier",
  "nominal tangent rank")
knitr::kable(show_geometry, row.names = FALSE)
```

The panel has four fit failures among 75,000 estimator rows: one FIML and three
ML2S. Every converged fit produced a finite legacy-multiplier p-value. Four FIML
and one ML2S call lost numerical tangent rank despite producing a finite value;
the table records these separately rather than silently counting them as
ordinary tests.

### Runtime and next experiment

The checkpointed panel took approximately 86 minutes on ten workers, including
75,000 fits, the LR/discrepancy and score-FMG batteries, and 199 multiplier
draws. The final 25-rep large-model severe-IG/MCAR chunk dominated the tail.
The corrected multiplier was not part of this timing.

### Reproduce

The archived panel above predates the corrected multiplier and remains the
report's default input. To calibrate and smoke-test the planned paired
null/sparse/diffuse continuation locally:

```sh
Rscript experiments/77-fiml-global-gof-pilot/calibrate_power.R
Rscript experiments/77-fiml-global-gof-pilot/run_sem_models.R \
  --reps 5 --n 200 --flips 19 --cores 4 --chunk-size 1 \
  --estimators FIML --truth null,sparse,diffuse --regions identified_null \
  --power-calibration-file experiments/77-fiml-global-gof-pilot/results/calibration/power_calibration_n200.csv \
  --results-dir experiments/77-fiml-global-gof-pilot/results/power-smoke
```

The Modal harness is dry-run by default. The preflight starts one two-rep cell;
the substantive focus launches 165 one-cell shards with 1,000 replications and
999 multiplier draws. `--mode all` additionally launches the separately
labeled 60-cell nonnormal-MAR stress phase at 500 replications.

```sh
modal run experiments/77-fiml-global-gof-pilot/modal/app.py
modal run experiments/77-fiml-global-gof-pilot/modal/app.py \
  --mode preflight --run-id exp77-aug21
modal run --detach experiments/77-fiml-global-gof-pilot/modal/app.py \
  --mode focus --run-id exp77-aug21
```
