---
title: "FIML goodness of fit in representative SEMs"
subtitle: "Vale--Maurelli and independent-generator null study"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- file.path(script_dir, "results", "sem-model-1000")
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
replications <- read_result("replications")
design <- read_result("design")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)
flip_gate_dir <- file.path(script_dir, "results", "sem-global-flip-500")
flip_gate <- if (file.exists(file.path(flip_gate_dir, "replications.csv"))) {
  read.csv(file.path(flip_gate_dir, "replications.csv"),
           stringsAsFactors = FALSE)
} else NULL

generator_labels <- c(
  normal = "Normal", vm1 = "VM moderate", ig1 = "IG moderate",
  vm2 = "VM severe", ig2 = "IG severe")
method_columns <- c(
  "Yuan--Bentler MLR" = "p_lrt_mlr", "FMG SB" = "p_lrt_sb",
  "FMG SS" = "p_lrt_ss", "FMG pEBA(4)" = "p_lrt_peba4",
  "FMG all" = "p_lrt_all")

summarize_p <- function(data, column) {
  p <- data[[column]]
  c(n = sum(is.finite(p)), rejection = mean(p < .05, na.rm = TRUE))
}

group_vars <- c("model_id", "distribution", "missingness")
group_id <- interaction(
  replications[group_vars], drop = TRUE, lex.order = TRUE)
cell_size <- do.call(rbind, lapply(
  split(seq_len(nrow(replications)), group_id), function(ii) {
    z <- replications[ii, , drop = FALSE]
    do.call(rbind, lapply(names(method_columns), function(method) {
      p <- z[[method_columns[[method]]]]
      data.frame(
        z[1L, group_vars, drop = FALSE], method = method,
        finite = sum(is.finite(p)), rejection = mean(p < .05, na.rm = TRUE),
        stringsAsFactors = FALSE)
    }))
  }))

scorecard <- do.call(rbind, lapply(
  split(cell_size, cell_size$method), function(z) {
    data.frame(
      method = z$method[[1L]], finite = sum(z$finite),
      mean_cell_rejection = mean(z$rejection),
      mean_abs_size_error = mean(abs(z$rejection - .05)),
      min_cell_rejection = min(z$rejection),
      max_cell_rejection = max(z$rejection),
      fraction_025_075 = mean(z$rejection >= .025 & z$rejection <= .075),
      stringsAsFactors = FALSE)
  }))

dgp_rows <- unique(replications[, c("distribution", "missingness")])
generator_size <- do.call(rbind, lapply(seq_len(nrow(dgp_rows)), function(i) {
  z <- replications[
    replications$distribution == dgp_rows$distribution[[i]] &
      replications$missingness == dgp_rows$missingness[[i]], , drop = FALSE]
  mlr <- summarize_p(z, "p_lrt_mlr")
  ss <- summarize_p(z, "p_lrt_ss")
  all <- summarize_p(z, "p_lrt_all")
  data.frame(
    generator = generator_labels[[dgp_rows$distribution[[i]]]],
    observation = if (dgp_rows$missingness[[i]] == "complete")
      "Complete" else "30% MCAR",
    mlr_n = mlr[["n"]], mlr_rejection = mlr[["rejection"]],
    ss_n = ss[["n"]], ss_rejection = ss[["rejection"]],
    all_n = all[["n"]], all_rejection = all[["rejection"]],
    stringsAsFactors = FALSE)
}))
generator_size$generator <- factor(
  generator_size$generator, levels = unname(generator_labels))
generator_size$observation <- factor(
  generator_size$observation, levels = c("Complete", "30% MCAR"))
generator_size <- generator_size[order(
  generator_size$generator, generator_size$observation), ]

model_rows <- unique(replications[, c("model_id", "model_label")])
model_size <- do.call(rbind, lapply(seq_len(nrow(model_rows)), function(i) {
  z <- replications[replications$model_id == model_rows$model_id[[i]], ]
  mlr <- summarize_p(z, "p_lrt_mlr")
  ss <- summarize_p(z, "p_lrt_ss")
  all <- summarize_p(z, "p_lrt_all")
  data.frame(
    model = model_rows$model_label[[i]],
    mlr_n = mlr[["n"]], mlr_rejection = mlr[["rejection"]],
    ss_n = ss[["n"]], ss_rejection = ss[["rejection"]],
    all_n = all[["n"]], all_rejection = all[["rejection"]],
    stringsAsFactors = FALSE)
}))

gate_method_columns <- c(
  "Effective multiplier" = "p_flip_effective",
  "FMG SS" = "p_lrt_ss", "FMG all" = "p_lrt_all",
  "Yuan--Bentler MLR" = "p_lrt_mlr", "FMG SB" = "p_lrt_sb",
  "FMG pEBA(4)" = "p_lrt_peba4")

if (!is.null(flip_gate)) {
  gate_group_id <- interaction(
    flip_gate[group_vars], drop = TRUE, lex.order = TRUE)
  gate_cell_size <- do.call(rbind, lapply(
    split(seq_len(nrow(flip_gate)), gate_group_id), function(ii) {
      z <- flip_gate[ii, , drop = FALSE]
      do.call(rbind, lapply(names(gate_method_columns), function(method) {
        p <- z[[gate_method_columns[[method]]]]
        data.frame(
          z[1L, group_vars, drop = FALSE], method = method,
          finite = sum(is.finite(p)), rejection = mean(p < .05, na.rm = TRUE),
          stringsAsFactors = FALSE)
      }))
    }))
  gate_scorecard <- do.call(rbind, lapply(
    split(gate_cell_size, gate_cell_size$method), function(z) {
      data.frame(
        method = z$method[[1L]], finite = sum(z$finite),
        mean_cell_rejection = mean(z$rejection),
        mean_abs_size_error = mean(abs(z$rejection - .05)),
        min_cell_rejection = min(z$rejection),
        max_cell_rejection = max(z$rejection),
        fraction_025_075 = mean(
          z$rejection >= .025 & z$rejection <= .075),
        stringsAsFactors = FALSE)
    }))
  gate_scorecard <- gate_scorecard[match(
    names(gate_method_columns), gate_scorecard$method), ]

  gate_dgp <- unique(flip_gate[, c("distribution", "missingness")])
  gate_generator_size <- do.call(rbind, lapply(
    seq_len(nrow(gate_dgp)), function(i) {
      z <- flip_gate[
        flip_gate$distribution == gate_dgp$distribution[[i]] &
          flip_gate$missingness == gate_dgp$missingness[[i]], , drop = FALSE]
      flip <- summarize_p(z, "p_flip_effective")
      mlr <- summarize_p(z, "p_lrt_mlr")
      ss <- summarize_p(z, "p_lrt_ss")
      all <- summarize_p(z, "p_lrt_all")
      data.frame(
        generator = generator_labels[[gate_dgp$distribution[[i]]]],
        observation = if (gate_dgp$missingness[[i]] == "complete")
          "Complete" else "30% MCAR",
        flip_n = flip[["n"]], flip_rejection = flip[["rejection"]],
        ss_rejection = ss[["rejection"]], all_rejection = all[["rejection"]],
        mlr_rejection = mlr[["rejection"]], stringsAsFactors = FALSE)
    }))
  gate_generator_size$generator <- factor(
    gate_generator_size$generator, levels = unname(generator_labels))
  gate_generator_size$observation <- factor(
    gate_generator_size$observation, levels = c("Complete", "30% MCAR"))
  gate_generator_size <- gate_generator_size[order(
    gate_generator_size$generator, gate_generator_size$observation), ]

  flip_gate$nominal_geometry <- flip_gate$flip_ok &
    flip_gate$flip_df == flip_gate$expected_df
  gate_geometry <- aggregate(
    cbind(flip_ok, nominal_geometry) ~ model_label, flip_gate, sum)
  gate_geometry$fit_ok <- as.integer(table(factor(
    flip_gate$model_label[flip_gate$fit_ok], levels = gate_geometry$model_label)))
  gate_geometry$attempted <- as.integer(table(factor(
    flip_gate$model_label, levels = gate_geometry$model_label)))
}
```

## Question

How well calibrated are Yuan--Bentler/Mplus MLR and the
Foldnes--Moss--Grønneberg (FMG) eigenvalue-tail corrections when correctly
specified, ordinary latent-variable models are fitted by FIML to normal,
Vale--Maurelli, and independent-generator data, with and without MCAR
missingness?

Every fitted model is exactly correct in its population. All rejection rates
are therefore Type-I error rates. This report contains no $t$ distributions,
lognormal margins, contamination, MAR, MNAR, or power conditions; those belong
to an earlier affine screening experiment.

## Models

1. **One-factor CFA, six indicators.** The population is
   $x_j=\lambda_j f+\varepsilon_j$, with
   $\lambda=(1,.8,.9,.7,1.1,.85)$, $\operatorname{Var}(f)=1$, and unique
   variances $(.5,.6,.55,.65,.5,.6)$. Indicator means are zero. The fitted CFA
   uses marker-variable identification and freely estimates means, unique
   variances, the factor variance, and the remaining loadings.

2. **Correlated two-factor CFA, ten indicators.** Indicators 1--5 load only on
   factor 1 with loadings $(.4,.5,.6,.8,.4)$; indicators 6--10 load only on
   factor 2 with loadings $(.4,.7,.6,.4,.8)$. Factor variances are one and the
   factor correlation is .5. Unique variances are $1-\lambda_j^2$, giving unit
   marginal variances, and all means are zero.

3. **Orthogonal bifactor model, twelve indicators.** Every indicator loads .6
   on a general factor. Three orthogonal specific factors each cover four
   indicators, with blockwise loadings .5, .4, and .45. All factor variances are
   one; unique variances are $1-.6^2-\lambda_{S,j}^2$ and means are zero. The
   fitted model uses `std.lv` identification.

4. **Five-wave linear growth model.** At $t=0,1,2,3,4$,
   $y_t=i+t s+\varepsilon_t$. The intercept/slope means are $(1,.2)$, their
   covariance matrix is

   $$
   \begin{pmatrix}
   1 & 0.1 \\
   0.1 & 0.1
   \end{pmatrix},
   $$

   and each residual variance is .5. The fitted model fixes the time scores
   and freely estimates the growth means, latent variance/covariance terms,
   and residual variances.

```{r model-table}
model_table <- unique(design[, c(
  "model_id", "model_label", "p", "expected_df")])
model_table$free_parameters <- c(18L, 31L, 48L, 10L)[match(
  model_table$model_id,
  c("one_factor_6", "two_factor_fmg_10", "bifactor_12", "linear_growth_5"))]
model_table <- model_table[, c(
  "model_label", "p", "free_parameters", "expected_df")]
names(model_table) <- c("model", "observed variables", "free parameters", "df")
knitr::kable(model_table, row.names = FALSE)
```

## VM and IG generators

```{r generator-table}
generator_table <- data.frame(
  code = c("normal", "vm1", "ig1", "vm2", "ig2"),
  generator = c(
    "Multivariate normal", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)"),
  marginal_skewness = c(0, 2, 2, 3, 3),
  marginal_excess_kurtosis = c(0, 7, 7, 21, 21),
  severity = c("reference", "moderate", "moderate", "severe", "severe"),
  stringsAsFactors = FALSE)
knitr::kable(generator_table, row.names = FALSE)
```

**VM** is the Vale--Maurelli/Fleishman construction. Magmaan solves an
intermediate Gaussian correlation matrix and applies calibrated marginal
polynomials to recover the SEM population covariance and requested marginal
moments.

**IG** is the Foldnes--Olsson independent-generator construction. Independent
Pearson-family nonnormal generators are calibrated and mixed through a
symmetric covariance root. VM and IG therefore have the same target covariance
and marginal skewness/kurtosis but different multivariate higher-order
dependence.

Each generator is observed completely or with 30% MCAR on every variable except
the first indicator/wave, which is always observed. The factorial design has
$4\times5\times2=40$ true-null cells, all at $n=120$, with 1,000 replications
per cell.

## Methods and denominators

Each dataset is fitted once by FIML. The report compares the dedicated
Yuan--Bentler/Mplus MLR statistic with FMG SB, SS, pEBA(4), and `all`
corrections. The 40,000-fit run gives the most precise MLR/FMG baseline. A
matched 20,000-fit gate adds the general effective multiplier statistic with
199 Rademacher draws per fit on the same 40-cell design.

```{r health-table}
health <- data.frame(
  quantity = c("Attempted", "FIML converged", "MLR finite", "FMG call returned"),
  result = c(
    nrow(replications), sum(replications$fit_ok), sum(replications$mlr_ok),
    sum(replications$fmg_ok)))
knitr::kable(health, row.names = FALSE)
```

Fifteen FIML fits did not converge. Among otherwise converged fits, 91 MLR
p-values were non-finite and 829 FMG calls hit the projected `U Gamma` rank
guard. The bifactor accounted for 797 of the FMG rank failures. Rejection rates
below use finite, method-specific denominators; failures are not silently
counted as non-rejections.

## Overall null calibration

```{r scorecard-table}
show_scorecard <- scorecard[, c(
  "method", "finite", "mean_cell_rejection", "mean_abs_size_error",
  "min_cell_rejection", "max_cell_rejection", "fraction_025_075")]
names(show_scorecard) <- c(
  "method", "finite p-values", "mean cell size", "mean |size-.05|",
  "minimum", "maximum", "fraction in [.025,.075]")
knitr::kable(show_scorecard, digits = 3, row.names = FALSE)
```

MLR is severely liberal: its mean rejection across the 40 cells is .366 and it
ranges from .062 to .932. FMG `all` is much better at .066 on average, but its
.022--.168 range is not uniformly satisfactory. SS behaves similarly overall;
SB and pEBA(4) remain clearly liberal.

## VM versus IG

```{r generator-results}
show_generator <- generator_size
names(show_generator) <- c(
  "generator", "observation", "MLR n", "MLR rejection",
  "SS n", "SS rejection", "all n", "all rejection")
knitr::kable(show_generator, digits = 3, row.names = FALSE)
```

At identical marginal moments, VM is consistently harder than IG. MLR rises
from .294/.440 under moderate VM (complete/MCAR) to .556/.689 under severe VM;
the corresponding IG rates are .232/.381 and .318/.493. Marginal skewness and
kurtosis alone therefore do not explain calibration.

FMG `all` is close to .05 for all complete-data generator conditions. With
MCAR it rises to .099 under moderate VM and .127 under severe VM, compared with
.063 and .055 under IG. Keeping VM and IG separate is consequently essential.

## Results by model

```{r model-results}
show_model <- model_size
names(show_model) <- c(
  "model", "MLR n", "MLR rejection", "SS n", "SS rejection",
  "all n", "all rejection")
knitr::kable(show_model, digits = 3, row.names = FALSE)
```

The MLR problem is not confined to one unusual model. Pooled rejection is .264
for the one-factor CFA, .389 for the two-factor CFA, .539 for the bifactor, and
.268 for the growth model. The bifactor is the hardest model both statistically
and numerically, but removing it would not rescue MLR.

## What the effective multiplier test does

This is a global score test of the fitted SEM against the local saturated
mean-and-covariance model. It is not a robust standard error and it does not
rescale the ordinary likelihood-ratio statistic.

Let the saturated moment vector be

$$
\eta =
\begin{pmatrix}
\operatorname{vech}(\Sigma) \\
\mu
\end{pmatrix}
\in \mathbb{R}^{m}.
$$

For $p$ observed variables with an estimated mean structure,
$m=p(p+1)/2+p$. At the restricted estimate $\widehat\theta$, the test evaluates
the direct saturated observed-data score contribution $s_i$ for every case.
Under FIML, $s_i$ is the likelihood score for the variables actually observed
for case $i$; missing values are neither filled in nor assigned artificial
residuals.

### Remove the fitted model tangent

The model occupies only a curved subspace of the saturated moment space. Its
local tangent is the column space of

$$
\Delta
=
\left.
\frac{\partial\eta(\theta)}{\partial\theta^{\mathsf T}}
\right|_{\widehat\theta},
$$

after incorporating any affine equality constraints. Write $r$ for the
numerical rank of $\Delta$. Let $K$ span the model tangent and let $D$ span its
Euclidean complement. If $I$ is the saturated-moment Fisher information,
formed by summing the appropriate conditional information over observed-data
patterns, define

$$
G
=
D-K(K^{\mathsf T}IK)^{-1}K^{\mathsf T}ID.
$$

This adjustment makes $K^{\mathsf T}IG=0$: the columns of $G$ describe the
saturated directions that remain after removing changes locally explainable by
the fitted SEM. Each case contributes the nuisance-effective score

$$
z_i=G^{\mathsf T}s_i,
$$

and the observed quadratic statistic is

$$
u=\sum_{i=1}^{n}z_i,
\qquad
V=G^{\mathsf T}IG,
\qquad
T=u^{\mathsf T}V^{-1}u.
$$

The tested dimension is

$$
d=m-r.
$$

At a regular fit this is the ordinary model df. A discrepancy is useful rather
than cosmetic: it means the fitted model Jacobian has lost numerical rank. The
rank diagnostic reported below therefore accompanies every multiplier result.

### Construct the multiplier reference

For draw $b$, generate independent Rademacher weights
$w_{ib}\in\{-1,+1\}$ and form

$$
u_b=\sum_{i=1}^{n}w_{ib}z_i,
\qquad
T_b=u_b^{\mathsf T}V^{-1}u_b.
$$

With $B$ random draws, the finite Monte Carlo p-value is

$$
\widehat p
=
\frac{1+\sum_{b=1}^{B}\mathbf{1}(T_b\geq T)}{B+1}.
$$

The experiment uses $B=199$, giving a p-value grid of width $1/200$. The fitted
moments, tangent projection, and Fisher metric are held fixed. A draw requires
only a weighted sum of the saved casewise scores and a small quadratic form;
neither the null SEM nor a saturated alternative is refitted.

The practical contrast with Yuan--Bentler MLR is therefore sharp. MLR reduces
the nonnormal correction to a scalar adjustment of a likelihood-ratio
statistic. The effective multiplier instead retains the casewise,
direction-specific score variation after the SEM tangent has been removed and
uses random reweighting to obtain its reference distribution. The null results
below evaluate whether that extra structure buys usable calibration.

## Global multiplier null gate

```{r gate-scorecard}
if (!is.null(flip_gate)) {
  show_gate_scorecard <- gate_scorecard[, c(
    "method", "finite", "mean_cell_rejection", "mean_abs_size_error",
    "min_cell_rejection", "max_cell_rejection", "fraction_025_075")]
  names(show_gate_scorecard) <- c(
    "method", "finite p-values", "mean cell size", "mean |size-.05|",
    "minimum", "maximum", "fraction in [.025,.075]")
  knitr::kable(show_gate_scorecard, digits = 3, row.names = FALSE)
} else {
  cat("Run the 500-replication global-flip command below to populate this table.")
}
```

The effective multiplier is the clear winner in the matched gate. Its mean
rejection across the 40 null cells is .058, its mean absolute size error is
.016, and 33/40 cells lie in the practical [.025,.075] band. FMG SS and `all`
average .072 and .065 but have larger cellwise errors and maxima of .173 and
.155. MLR averages .364 and reaches .930. Thus the flip is not merely beating
MLR; it also materially improves on the strongest FMG analytic corrections.

```{r gate-generator}
if (!is.null(flip_gate)) {
  show_gate_generator <- gate_generator_size
  names(show_gate_generator) <- c(
    "generator", "observation", "flip n", "effective multiplier",
    "FMG SS", "FMG all", "MLR")
  knitr::kable(show_gate_generator, digits = 3, row.names = FALSE)
}
```

The multiplier remains close to nominal under both generator families and
both observation mechanisms. Pooled over models, its rejection ranges from
.037 under normal MCAR to .085 under severe complete-data VM. Severe VM is the
remaining visible stress point; severe IG is .061 complete and .072 under
MCAR, confirming again that matching marginal skewness and kurtosis does not
match multivariate difficulty.

### Numerical geometry and failures

```{r gate-geometry}
if (!is.null(flip_gate)) {
  show_geometry <- gate_geometry[, c(
    "model_label", "attempted", "fit_ok", "flip_ok", "nominal_geometry")]
  names(show_geometry) <- c(
    "model", "attempted", "FIML converged", "finite flip",
    "nominal tangent rank")
  knitr::kable(show_geometry, row.names = FALSE)
}
```

Nine of 20,000 FIML fits did not converge. Every converged fit produced a
finite multiplier p-value. In 25 of the 19,991 finite calls, however, the local
model Jacobian lost numerical rank, raising the tested df above its
prespecified value. These cases were concentrated in the two-factor CFA under
MCAR (19 cases), with two additional two-factor VM1/complete cases and four
one-factor VM2/MCAR cases. They are reported as non-nominal geometry rather
than hidden as successful ordinary tests.

The all-finite scorecard above is deliberately conservative about this
diagnostic: 16/25 rank-anomalous calls rejected. Requiring the nominal tangent
rank leaves 19,966 calls and changes mean cell rejection only from .0584 to
.0577, with the same 33/40 cells in the practical band. Future power runs
should fail closed on a non-nominal rank and report that denominator
separately.

## Runtime and next experiment

The 20,000-fit multiplier gate took 446 seconds (7.4 minutes) on four workers,
including FIML, MLR, FMG, and 199 multiplier draws. The measured parallel
speedup was 3.68. The mean multiplier call itself was about 0.022 seconds; the
large bifactor/MCAR cells dominate the slow tail.

This null gate is strong enough to proceed. The next scientific experiment is
a matched power panel with prespecified sparse and diffuse SEM
misspecifications, retaining the same four models and VM/IG mechanisms. It
should report raw power together with size-adjusted power, finite denominators,
and tangent-rank failures. A later 1,000--2,000-replication null confirmation
can sharpen the seven edge-cell estimates; it is no longer needed to decide
whether the flip deserves the power experiment. The saturated-EM
moment-influence construction remains an independent validation route for the
missing-data projection rather than the production statistic.

## Reproduce

```sh
Rscript experiments/77-fiml-global-gof-pilot/run_sem_models.R --reps 500 \
  --flips 199 \
  --cores 4 \
  --results-dir experiments/77-fiml-global-gof-pilot/results/sem-global-flip-500
quarto render experiments/77-fiml-global-gof-pilot/sem_report.qmd
```
