---
title: "FIML goodness of fit in representative SEMs"
subtitle: "Vale--Maurelli and independent-generator null study"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- file.path(script_dir, "results", "sem-model-1000")
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
replications <- read_result("replications")
design <- read_result("design")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)
flip_smoke_dir <- file.path(script_dir, "results", "sem-global-flip-smoke")
flip_smoke <- if (file.exists(file.path(flip_smoke_dir, "replications.csv"))) {
  read.csv(file.path(flip_smoke_dir, "replications.csv"),
           stringsAsFactors = FALSE)
} else NULL

generator_labels <- c(
  normal = "Normal", vm1 = "VM moderate", ig1 = "IG moderate",
  vm2 = "VM severe", ig2 = "IG severe")
method_columns <- c(
  "Yuan--Bentler MLR" = "p_lrt_mlr", "FMG SB" = "p_lrt_sb",
  "FMG SS" = "p_lrt_ss", "FMG pEBA(4)" = "p_lrt_peba4",
  "FMG all" = "p_lrt_all")

summarize_p <- function(data, column) {
  p <- data[[column]]
  c(n = sum(is.finite(p)), rejection = mean(p < .05, na.rm = TRUE))
}

group_vars <- c("model_id", "distribution", "missingness")
group_id <- interaction(
  replications[group_vars], drop = TRUE, lex.order = TRUE)
cell_size <- do.call(rbind, lapply(
  split(seq_len(nrow(replications)), group_id), function(ii) {
    z <- replications[ii, , drop = FALSE]
    do.call(rbind, lapply(names(method_columns), function(method) {
      p <- z[[method_columns[[method]]]]
      data.frame(
        z[1L, group_vars, drop = FALSE], method = method,
        finite = sum(is.finite(p)), rejection = mean(p < .05, na.rm = TRUE),
        stringsAsFactors = FALSE)
    }))
  }))

scorecard <- do.call(rbind, lapply(
  split(cell_size, cell_size$method), function(z) {
    data.frame(
      method = z$method[[1L]], finite = sum(z$finite),
      mean_cell_rejection = mean(z$rejection),
      mean_abs_size_error = mean(abs(z$rejection - .05)),
      min_cell_rejection = min(z$rejection),
      max_cell_rejection = max(z$rejection),
      fraction_025_075 = mean(z$rejection >= .025 & z$rejection <= .075),
      stringsAsFactors = FALSE)
  }))

dgp_rows <- unique(replications[, c("distribution", "missingness")])
generator_size <- do.call(rbind, lapply(seq_len(nrow(dgp_rows)), function(i) {
  z <- replications[
    replications$distribution == dgp_rows$distribution[[i]] &
      replications$missingness == dgp_rows$missingness[[i]], , drop = FALSE]
  mlr <- summarize_p(z, "p_lrt_mlr")
  ss <- summarize_p(z, "p_lrt_ss")
  all <- summarize_p(z, "p_lrt_all")
  data.frame(
    generator = generator_labels[[dgp_rows$distribution[[i]]]],
    observation = if (dgp_rows$missingness[[i]] == "complete")
      "Complete" else "30% MCAR",
    mlr_n = mlr[["n"]], mlr_rejection = mlr[["rejection"]],
    ss_n = ss[["n"]], ss_rejection = ss[["rejection"]],
    all_n = all[["n"]], all_rejection = all[["rejection"]],
    stringsAsFactors = FALSE)
}))
generator_size$generator <- factor(
  generator_size$generator, levels = unname(generator_labels))
generator_size$observation <- factor(
  generator_size$observation, levels = c("Complete", "30% MCAR"))
generator_size <- generator_size[order(
  generator_size$generator, generator_size$observation), ]

model_rows <- unique(replications[, c("model_id", "model_label")])
model_size <- do.call(rbind, lapply(seq_len(nrow(model_rows)), function(i) {
  z <- replications[replications$model_id == model_rows$model_id[[i]], ]
  mlr <- summarize_p(z, "p_lrt_mlr")
  ss <- summarize_p(z, "p_lrt_ss")
  all <- summarize_p(z, "p_lrt_all")
  data.frame(
    model = model_rows$model_label[[i]],
    mlr_n = mlr[["n"]], mlr_rejection = mlr[["rejection"]],
    ss_n = ss[["n"]], ss_rejection = ss[["rejection"]],
    all_n = all[["n"]], all_rejection = all[["rejection"]],
    stringsAsFactors = FALSE)
}))
```

## Question

How well calibrated are Yuan--Bentler/Mplus MLR and the
Foldnes--Moss--Grønneberg (FMG) eigenvalue-tail corrections when correctly
specified, ordinary latent-variable models are fitted by FIML to normal,
Vale--Maurelli, and independent-generator data, with and without MCAR
missingness?

Every fitted model is exactly correct in its population. All rejection rates
are therefore Type-I error rates. This report contains no $t$ distributions,
lognormal margins, contamination, MAR, MNAR, or power conditions; those belong
to an earlier affine screening experiment.

## Models

1. **One-factor CFA, six indicators.** The population is
   $x_j=\lambda_j f+\varepsilon_j$, with
   $\lambda=(1,.8,.9,.7,1.1,.85)$, $\operatorname{Var}(f)=1$, and unique
   variances $(.5,.6,.55,.65,.5,.6)$. Indicator means are zero. The fitted CFA
   uses marker-variable identification and freely estimates means, unique
   variances, the factor variance, and the remaining loadings.

2. **Correlated two-factor CFA, ten indicators.** Indicators 1--5 load only on
   factor 1 with loadings $(.4,.5,.6,.8,.4)$; indicators 6--10 load only on
   factor 2 with loadings $(.4,.7,.6,.4,.8)$. Factor variances are one and the
   factor correlation is .5. Unique variances are $1-\lambda_j^2$, giving unit
   marginal variances, and all means are zero.

3. **Orthogonal bifactor model, twelve indicators.** Every indicator loads .6
   on a general factor. Three orthogonal specific factors each cover four
   indicators, with blockwise loadings .5, .4, and .45. All factor variances are
   one; unique variances are $1-.6^2-\lambda_{S,j}^2$ and means are zero. The
   fitted model uses `std.lv` identification.

4. **Five-wave linear growth model.** At $t=0,1,2,3,4$,
   $y_t=i+t s+\varepsilon_t$. The intercept/slope means are $(1,.2)$, their
   covariance matrix is
   $\left[\begin{smallmatrix}1&.1\\.1&.1\end{smallmatrix}\right]$, and each
   residual variance is .5. The fitted model fixes the time scores and freely
   estimates the growth means, latent variance/covariance terms, and residual
   variances.

```{r model-table}
model_table <- unique(design[, c(
  "model_id", "model_label", "p", "expected_df")])
model_table$free_parameters <- c(18L, 31L, 48L, 10L)[match(
  model_table$model_id,
  c("one_factor_6", "two_factor_fmg_10", "bifactor_12", "linear_growth_5"))]
model_table <- model_table[, c(
  "model_label", "p", "free_parameters", "expected_df")]
names(model_table) <- c("model", "observed variables", "free parameters", "df")
knitr::kable(model_table, row.names = FALSE)
```

## VM and IG generators

```{r generator-table}
generator_table <- data.frame(
  code = c("normal", "vm1", "ig1", "vm2", "ig2"),
  generator = c(
    "Multivariate normal", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)"),
  marginal_skewness = c(0, 2, 2, 3, 3),
  marginal_excess_kurtosis = c(0, 7, 7, 21, 21),
  severity = c("reference", "moderate", "moderate", "severe", "severe"),
  stringsAsFactors = FALSE)
knitr::kable(generator_table, row.names = FALSE)
```

**VM** is the Vale--Maurelli/Fleishman construction. Magmaan solves an
intermediate Gaussian correlation matrix and applies calibrated marginal
polynomials to recover the SEM population covariance and requested marginal
moments.

**IG** is the Foldnes--Olsson independent-generator construction. Independent
Pearson-family nonnormal generators are calibrated and mixed through a
symmetric covariance root. VM and IG therefore have the same target covariance
and marginal skewness/kurtosis but different multivariate higher-order
dependence.

Each generator is observed completely or with 30% MCAR on every variable except
the first indicator/wave, which is always observed. The factorial design has
$4\times5\times2=40$ true-null cells, all at $n=120$, with 1,000 replications
per cell.

## Methods and denominators

Each dataset is fitted once by FIML. The report compares the dedicated
Yuan--Bentler/Mplus MLR statistic with FMG SB, SS, pEBA(4), and `all`
corrections. The saved 40,000-fit run predates the general effective multiplier
statistic; a separate implementation smoke below runs that statistic on the
same 40-cell design.

```{r health-table}
health <- data.frame(
  quantity = c("Attempted", "FIML converged", "MLR finite", "FMG call returned"),
  result = c(
    nrow(replications), sum(replications$fit_ok), sum(replications$mlr_ok),
    sum(replications$fmg_ok)))
knitr::kable(health, row.names = FALSE)
```

Fifteen FIML fits did not converge. Among otherwise converged fits, 91 MLR
p-values were non-finite and 829 FMG calls hit the projected `U Gamma` rank
guard. The bifactor accounted for 797 of the FMG rank failures. Rejection rates
below use finite, method-specific denominators; failures are not silently
counted as non-rejections.

## Overall null calibration

```{r scorecard-table}
show_scorecard <- scorecard[, c(
  "method", "finite", "mean_cell_rejection", "mean_abs_size_error",
  "min_cell_rejection", "max_cell_rejection", "fraction_025_075")]
names(show_scorecard) <- c(
  "method", "finite p-values", "mean cell size", "mean |size-.05|",
  "minimum", "maximum", "fraction in [.025,.075]")
knitr::kable(show_scorecard, digits = 3, row.names = FALSE)
```

MLR is severely liberal: its mean rejection across the 40 cells is .366 and it
ranges from .062 to .932. FMG `all` is much better at .066 on average, but its
.022--.168 range is not uniformly satisfactory. SS behaves similarly overall;
SB and pEBA(4) remain clearly liberal.

## VM versus IG

```{r generator-results}
show_generator <- generator_size
names(show_generator) <- c(
  "generator", "observation", "MLR n", "MLR rejection",
  "SS n", "SS rejection", "all n", "all rejection")
knitr::kable(show_generator, digits = 3, row.names = FALSE)
```

At identical marginal moments, VM is consistently harder than IG. MLR rises
from .294/.440 under moderate VM (complete/MCAR) to .556/.689 under severe VM;
the corresponding IG rates are .232/.381 and .318/.493. Marginal skewness and
kurtosis alone therefore do not explain calibration.

FMG `all` is close to .05 for all complete-data generator conditions. With
MCAR it rises to .099 under moderate VM and .127 under severe VM, compared with
.063 and .055 under IG. Keeping VM and IG separate is consequently essential.

## Results by model

```{r model-results}
show_model <- model_size
names(show_model) <- c(
  "model", "MLR n", "MLR rejection", "SS n", "SS rejection",
  "all n", "all rejection")
knitr::kable(show_model, digits = 3, row.names = FALSE)
```

The MLR problem is not confined to one unusual model. Pooled rejection is .264
for the one-factor CFA, .389 for the two-factor CFA, .539 for the bifactor, and
.268 for the growth model. The bifactor is the hardest model both statistically
and numerically, but removing it would not rescue MLR.

## Global effective multiplier implementation

The general test is now implemented. At the fitted null moments it forms each
case's direct saturated observed-data likelihood score, builds the
pattern-conditional Fisher metric, and projects the score off the fitted SEM's
local moment tangent. The tested dimension is the saturated mean/covariance
dimension minus the tangent rank. Multiplier draws only reweight the projected
case rows: neither a saturated model nor the null SEM is refitted.

```{r flip-smoke}
if (!is.null(flip_smoke)) {
  flip_timing <- aggregate(
    flip_seconds ~ model_label + missingness, flip_smoke, mean)
  flip_timing$missingness <- ifelse(
    flip_timing$missingness == "complete", "Complete", "30% MCAR")
  names(flip_timing) <- c("model", "observation", "mean flip seconds")
  knitr::kable(flip_timing, digits = 3, row.names = FALSE)
} else {
  cat("Run the global-flip smoke command below to populate this table.")
}
```

The 3-replication, 40-cell smoke made 120 global-test calls with 199
Rademacher draws each. All 120 returned finite p-values, and every computed df
matched the model's prespecified df. The mean call cost was 0.022 seconds; by
model and observation mechanism it ranged from about 0.003 seconds for the
small CFA/growth cases to 0.053 seconds for the bifactor. The full smoke,
including fitting, MLR, FMG, and the new multiplier test, took 2.6 seconds on
four workers.

These three replications per cell are an implementation and timing gate, not a
calibration result. They project to about 7.2 minutes for 500 null replications
per cell and 14.4 minutes for 1,000; a matched null-plus-power panel doubles
those figures.

## Runtime and next experiment

The 40,000-fit run took 541 seconds (9.0 minutes) on four workers. Severe
nonnormal bifactor/MCAR cells dominate the slow tail. At the observed rate, one
matched power panel would approximately double the run to 18 minutes at 1,000
replications per cell.

The next experiment can now compare the multiplier test directly with MLR and
FMG on this panel. Before another 40,000-fit run, the useful gate is roughly
100--500 null replications per cell, followed by matched sparse and diffuse
misspecification mechanisms. The saturated-EM moment-influence construction
remains valuable as an independent validation route for the missing-data
projection rather than as the production statistic.

## Reproduce

```sh
Rscript experiments/77-fiml-global-gof-pilot/run_sem_models.R --reps 1000 \
  --flips 199 \
  --results-dir experiments/77-fiml-global-gof-pilot/results/sem-model-1000
Rscript experiments/77-fiml-global-gof-pilot/run_sem_models.R --reps 3 \
  --flips 199 \
  --results-dir experiments/77-fiml-global-gof-pilot/results/sem-global-flip-smoke
quarto render experiments/77-fiml-global-gof-pilot/sem_report.qmd
```
