---
title: "FIML and two-stage goodness of fit in representative SEMs"
subtitle: "Score-spectrum and discrepancy calibrations under complete, MCAR, and MAR sampling"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
result_dir <- Sys.getenv(
  "MAGMAAN_SEM_RESULT_DIR",
  unset = file.path(
    script_dir, "results", "sem-paired-score-fmg-n200-500"
  )
)
read_result <- function(name) read.csv(
  file.path(result_dir, paste0(name, ".csv")), stringsAsFactors = FALSE)
replications <- read_result("replications")
design <- read_result("design")
metadata <- read_result("metadata")
meta <- setNames(metadata$value, metadata$key)

model_order <- c(
  "one_factor_6", "two_factor_fmg_10", "bifactor_12", "three_factor_15",
  "linear_growth_5")
model_labels <- c(
  one_factor_6 = "One-factor CFA",
  two_factor_fmg_10 = "Two-factor CFA",
  bifactor_12 = "Bifactor",
  three_factor_15 = "Three-factor CFA",
  linear_growth_5 = "Linear growth")
estimator_order <- c("FIML", "ML2S")
generator_order <- c("normal", "vm1", "ig1", "vm2", "ig2")
generator_labels <- c(
  normal = "Normal", vm1 = "VM moderate", ig1 = "IG moderate",
  vm2 = "VM severe", ig2 = "IG severe")
missingness_labels <- c(
  complete = "Complete", mcar_30 = "30% MCAR", mar_30 = "30% MAR")
method_columns <- c(
  "Naive LR/D" = "p_lrt_naive",
  "Scalar robust" = "p_lrt_mlr",
  "LR/D FMG SB" = "p_lrt_sb",
  "LR/D FMG SS" = "p_lrt_ss",
  "LR/D FMG pEBA(4)" = "p_lrt_peba4",
  "LR/D FMG all" = "p_lrt_all",
  "Score FMG SB" = "p_score_sb",
  "Score FMG SS" = "p_score_ss",
  "Score FMG pEBA(4)" = "p_score_peba4",
  "Score FMG all" = "p_score_all")
method_columns <- method_columns[method_columns %in% names(replications)]

group_vars <- c("model_id", "distribution", "missingness", "estimator")
group_id <- interaction(
  replications[group_vars], drop = TRUE, lex.order = TRUE)
cell_rows <- split(seq_len(nrow(replications)), group_id)

cell_estimator_table <- do.call(rbind, lapply(cell_rows, function(ii) {
  z <- replications[ii, , drop = FALSE]
  out <- data.frame(
    model_id = z$model_id[[1L]],
    model = unname(model_labels[[z$model_id[[1L]]]]),
    distribution = z$distribution[[1L]],
    generator = unname(generator_labels[[z$distribution[[1L]]]]),
    missingness_id = z$missingness[[1L]],
    missingness = unname(missingness_labels[[z$missingness[[1L]]]]),
    analysis_region = if (
      z$missingness[[1L]] == "mar_30" && z$distribution[[1L]] != "normal"
    ) "Nonnormal MAR stress" else "Identified null",
    estimator = z$estimator[[1L]],
    stringsAsFactors = FALSE, check.names = FALSE)
  for (method in names(method_columns)) {
    p_value <- z[[method_columns[[method]]]]
    out[[method]] <- mean(p_value < .05, na.rm = TRUE)
  }
  out
}))
cell_estimator_table$model_id <- factor(
  cell_estimator_table$model_id, levels = model_order)
cell_estimator_table$distribution <- factor(
  cell_estimator_table$distribution, levels = generator_order)
cell_estimator_table$missingness_id <- factor(
  cell_estimator_table$missingness_id,
  levels = c("complete", "mcar_30", "mar_30"))
cell_estimator_table$estimator <- factor(
  cell_estimator_table$estimator, levels = estimator_order)
cell_estimator_table <- cell_estimator_table[order(
  cell_estimator_table$model_id, cell_estimator_table$distribution,
  cell_estimator_table$missingness_id, cell_estimator_table$estimator), ]
row.names(cell_estimator_table) <- NULL

cell_key <- unique(cell_estimator_table[, c(
  "model_id", "model", "distribution", "generator", "missingness_id",
  "missingness")])
cell_table <- cell_key
wide_method_columns <- character()
for (estimator in estimator_order) {
  source <- cell_estimator_table[
    as.character(cell_estimator_table$estimator) == estimator, ]
  source_key <- interaction(
    source[c("model_id", "distribution", "missingness_id")],
    drop = TRUE, lex.order = TRUE)
  target_key <- interaction(
    cell_table[c("model_id", "distribution", "missingness_id")],
    drop = TRUE, lex.order = TRUE)
  index <- match(target_key, source_key)
  for (method in names(method_columns)) {
    if (!any(is.finite(source[[method]]))) next
    display_method <- if (method == "Scalar robust") {
      if (estimator == "FIML") "Yuan--Bentler MLR" else "Scaled ML2S"
    } else method
    column <- paste(estimator, display_method, sep = " / ")
    cell_table[[column]] <- source[[method]][index]
    wide_method_columns <- c(wide_method_columns, column)
  }
}

scorecard <- do.call(rbind, lapply(estimator_order, function(estimator) {
  z <- cell_estimator_table[
    as.character(cell_estimator_table$estimator) == estimator, ]
  raw <- replications[replications$estimator == estimator, ]
  do.call(rbind, lapply(names(method_columns), function(method) {
    rejection <- z[[method]]
    p_value <- raw[[method_columns[[method]]]]
    display_method <- if (method == "Scalar robust") {
      if (estimator == "FIML") "Yuan--Bentler MLR" else "Scaled ML2S"
    } else method
    finite_rejection <- rejection[is.finite(rejection)]
    data.frame(
      estimator = estimator, method = display_method,
      attempted = nrow(raw),
      finite = sum(is.finite(p_value)),
      mean_cell_rejection = if (length(finite_rejection)) {
        mean(finite_rejection)
      } else NA_real_,
      mean_abs_size_error = if (length(finite_rejection)) {
        mean(abs(finite_rejection - .05))
      } else NA_real_,
      min_cell_rejection = if (length(finite_rejection)) {
        min(finite_rejection)
      } else NA_real_,
      max_cell_rejection = if (length(finite_rejection)) {
        max(finite_rejection)
      } else NA_real_,
      cells_in_band = sum(
        rejection >= .025 & rejection <= .075, na.rm = TRUE),
      stringsAsFactors = FALSE)
  }))
}))

summarize_region <- function(region) {
  do.call(rbind, lapply(estimator_order, function(estimator) {
    z <- cell_estimator_table[
      as.character(cell_estimator_table$estimator) == estimator &
        cell_estimator_table$analysis_region == region, ]
    do.call(rbind, lapply(names(method_columns), function(method) {
      rejection <- z[[method]]
      rejection <- rejection[is.finite(rejection)]
      display_method <- if (method == "Scalar robust") {
        if (estimator == "FIML") "Yuan--Bentler MLR" else "Scaled ML2S"
      } else method
      data.frame(
        estimator = estimator, method = display_method,
        cells = length(rejection),
        mean_abs_size_error = if (length(rejection)) {
          mean(abs(rejection - .05))
        } else NA_real_,
        min_cell_rejection = if (length(rejection)) min(rejection) else NA_real_,
        max_cell_rejection = if (length(rejection)) max(rejection) else NA_real_,
        mean_cell_rejection = if (length(rejection)) mean(rejection) else NA_real_,
        stringsAsFactors = FALSE)
    }))
  }))
}
identified_scorecard <- summarize_region("Identified null")
mar_stress_scorecard <- summarize_region("Nonnormal MAR stress")

convergence <- aggregate(
  fit_ok ~ estimator + model_label, replications, sum)
convergence$attempted <- vapply(seq_len(nrow(convergence)), function(i) {
  sum(replications$estimator == convergence$estimator[[i]] &
        replications$model_label == convergence$model_label[[i]])
}, integer(1L))

power_run_dir <- Sys.getenv(
  "MAGMAAN_SEM_POWER_DIR",
  unset = file.path(script_dir, "results", "power-v1-20260821"))
power_focus_dir <- file.path(power_run_dir, "focus")
power_stress_dir <- file.path(power_run_dir, "stress")
power_replications <- read.csv(
  file.path(power_focus_dir, "replications.csv"),
  stringsAsFactors = FALSE, check.names = FALSE)
power_design_data <- read.csv(
  file.path(power_focus_dir, "design.csv"), stringsAsFactors = FALSE)
power_coverage <- read.csv(
  file.path(power_focus_dir, "coverage.csv"), stringsAsFactors = FALSE)
power_metadata <- read.csv(
  file.path(power_focus_dir, "metadata.csv"), stringsAsFactors = FALSE)
power_meta <- setNames(power_metadata$value, power_metadata$key)
stress_metadata <- read.csv(
  file.path(power_stress_dir, "metadata.csv"), stringsAsFactors = FALSE)
stress_meta <- setNames(stress_metadata$value, stress_metadata$key)

power_methods <- c(
  "Score pEBA(4)" = "p_score_peba4",
  "Score SB" = "p_score_sb",
  "LR all" = "p_lrt_all",
  "LR SS" = "p_lrt_ss",
  "MLR" = "p_lrt_mlr")
power_group_vars <- c("model_id", "distribution", "missingness")
power_group_id <- interaction(
  power_replications[power_group_vars], drop = TRUE, lex.order = TRUE)
power_groups <- split(seq_len(nrow(power_replications)), power_group_id)
power_cell_summary <- do.call(rbind, lapply(power_groups, function(ii) {
  z <- power_replications[ii, , drop = FALSE]
  do.call(rbind, lapply(names(power_methods), function(method) {
    p_value <- z[[power_methods[[method]]]]
    null_p <- p_value[z$truth == "null" & is.finite(p_value)]
    cutoff <- if (length(null_p)) {
      unname(stats::quantile(null_p, .05, type = 8))
    } else NA_real_
    summarize_power <- function(truth, threshold) {
      p <- p_value[z$truth == truth & is.finite(p_value)]
      if (length(p) && is.finite(threshold)) mean(p <= threshold) else NA_real_
    }
    data.frame(
      model_id = z$model_id[[1L]],
      distribution = z$distribution[[1L]],
      missingness = z$missingness[[1L]],
      method = method,
      null_n = length(null_p),
      size = if (length(null_p)) mean(null_p <= .05) else NA_real_,
      sparse_raw = summarize_power("sparse", .05),
      sparse_matched = summarize_power("sparse", cutoff),
      diffuse_raw = summarize_power("diffuse", .05),
      diffuse_matched = summarize_power("diffuse", cutoff),
      stringsAsFactors = FALSE)
  }))
}))
power_overview <- do.call(rbind, lapply(
  split(power_cell_summary, power_cell_summary$method), function(z) {
    data.frame(
      method = z$method[[1L]],
      size_mae = mean(abs(z$size - .05), na.rm = TRUE),
      size_min = min(z$size, na.rm = TRUE),
      size_max = max(z$size, na.rm = TRUE),
      sparse_raw = mean(z$sparse_raw, na.rm = TRUE),
      sparse_matched = mean(z$sparse_matched, na.rm = TRUE),
      diffuse_raw = mean(z$diffuse_raw, na.rm = TRUE),
      diffuse_matched = mean(z$diffuse_matched, na.rm = TRUE),
      stringsAsFactors = FALSE)
  }))
power_overview <- power_overview[order(power_overview$size_mae), ]
partial_power_cells <- merge(
  power_coverage[!power_coverage$complete, ],
  power_design_data[, c(
    "cell_id", "estimator", "model_label", "distribution", "missingness",
    "truth")],
  by = c("cell_id", "estimator"), all.x = TRUE)
```

This is a preliminary null-calibration study of global goodness-of-fit tests in
five ordinary latent-variable models. The design crosses normal,
Vale--Maurelli, and independent-generator data with complete sampling, 30%
MCAR, and observed-predictor MAR at $n=200$, with 500 replications per cell.

The main calibration region contains 55 cells: every complete and MCAR cell,
plus normal-data MAR. These are the cells in which the generating moments are
also the relevant null target for Gaussian FIML. The remaining 20 nonnormal-MAR
cells are retained as an estimand stress test, not interpreted as ordinary
Type-I error: under nonnormal MAR, Gaussian FIML can converge to pseudo-true
moments outside the generating SEM.

### Prospective null definition

The replacement study will use an estimator-level null throughout. Let
$\eta_1^\star$ denote the saturated Gaussian-FIML pseudo-true mean/covariance
target under the joint outcome and missingness mechanism, and let
$\mathcal M_0=\{\eta(\theta):\theta\in\Theta_0\}$ denote the fitted SEM. A
condition is a null condition only when

$$
\eta_1^\star\in\mathcal M_0,
$$

equivalently when the population restricted--saturated Gaussian-FIML
discrepancy is zero. Satisfying the SEM in the generating complete-data
covariance is not enough. Proposed MAR nulls must be constructed or calibrated
to this pseudo-null and checked with very-large-sample restricted and saturated
fits, zero-under-H0 directions, and Monte Carlo uncertainty. Any mechanism
whose saturated pseudo-target lies outside $\mathcal M_0$ is a pseudo-model
alternative and belongs in a power or stress analysis, not a size table.

This rule is prospective. The tables below retain the retired pilot's original
classification so its saved results are described faithfully; its nonnormal-
MAR arms do not qualify as nulls for the replacement design.

A subsequent exact nonnormal-MAR pseudo-null diagnostic also changes the
planned score geometry. Under strong tail selection, expected-H0 score
pEBA(4) rejection fell to zero through $n=10{,}000$, whereas realized H0
sensitivity with the stable expected metric remained near nominal. Realized
saturated-H1 information is positive definite and shares the same population
sensitivity at the pseudo-null, but converged prohibitively slowly in this
design. The replacement nonnormal-MAR battery will therefore use observed-H0
sensitivity with the expected metric as its primary score construction and
retain expected-H0 and observed-H1 geometries as diagnostics. Full details are
in the [pseudo-null information note](notes/iteration-13-pseudonull-information.md).

## Preliminary story

### 1. Projected-score pEBA(4) is the strongest FIML result

```{r fiml-primary-table}
fiml_primary <- identified_scorecard[
  identified_scorecard$estimator == "FIML" &
    is.finite(identified_scorecard$mean_abs_size_error),
  c("method", "mean_abs_size_error", "min_cell_rejection",
    "max_cell_rejection")]
fiml_primary <- fiml_primary[order(fiml_primary$mean_abs_size_error), ]
names(fiml_primary) <- c("method", "mean |size-.05|", "minimum", "maximum")
knitr::kable(fiml_primary, digits = 3, row.names = FALSE)
```

Projected-score pEBA(4) has the smallest mean absolute size error (.013), and
its 55-cell range stops at .074. Projected-score SB is also competitive (.015,
range .034--.092). The pEBA advantage is not cellwise uniform; it comes from
containing the larger errors rather than winning most individual cells.

### 2. The correction and the base statistic cannot be separated

```{r score-lrt-crossing-table}
crossing <- data.frame(
  correction = c("SB", "SS", "pEBA(4)", "all"),
  score_method = c("Score FMG SB", "Score FMG SS",
                   "Score FMG pEBA(4)", "Score FMG all"),
  lrt_method = c("LR/D FMG SB", "LR/D FMG SS",
                 "LR/D FMG pEBA(4)", "LR/D FMG all"),
  stringsAsFactors = FALSE)
take_metric <- function(method, metric) identified_scorecard[
  identified_scorecard$estimator == "FIML" &
    identified_scorecard$method == method, metric]
crossing$score_mae <- vapply(
  crossing$score_method, take_metric, numeric(1L),
  metric = "mean_abs_size_error")
crossing$score_min <- vapply(
  crossing$score_method, take_metric, numeric(1L),
  metric = "min_cell_rejection")
crossing$score_max <- vapply(
  crossing$score_method, take_metric, numeric(1L),
  metric = "max_cell_rejection")
crossing$lrt_mae <- vapply(
  crossing$lrt_method, take_metric, numeric(1L),
  metric = "mean_abs_size_error")
crossing$lrt_min <- vapply(
  crossing$lrt_method, take_metric, numeric(1L),
  metric = "min_cell_rejection")
crossing$lrt_max <- vapply(
  crossing$lrt_method, take_metric, numeric(1L),
  metric = "max_cell_rejection")
crossing <- crossing[, c(
  "correction", "score_mae", "score_min", "score_max",
  "lrt_mae", "lrt_min", "lrt_max")]
names(crossing) <- c(
  "correction", "score MAE", "score min", "score max",
  "LR MAE", "LR min", "LR max")
knitr::kable(crossing, digits = 3, row.names = FALSE)
```

There is no generic ordering of scores above likelihood ratios. SB and pEBA(4)
work with the projected-score statistic, whereas score SS and score `all` are
strongly conservative. For the likelihood-ratio statistic the ordering
reverses: SS and `all` are the viable corrections, while SB and pEBA(4) are
liberal. In particular, LR SS does beat score SS, but the best projected-score
calibration beats the best likelihood-ratio calibration. The large reversal is
important enough to require a focused formula and implementation audit before
it becomes more than a preliminary finding.

### 3. Two-stage ML does not improve the useful methods

```{r estimator-comparison-table}
selected_methods <- c(
  "Score FMG pEBA(4)", "Score FMG SB", "LR/D FMG all", "LR/D FMG SS")
estimator_comparison <- do.call(rbind, lapply(selected_methods, function(method) {
  rows <- identified_scorecard[identified_scorecard$method == method, ]
  data.frame(
    method = method,
    FIML_MAE = rows$mean_abs_size_error[rows$estimator == "FIML"],
    FIML_min = rows$min_cell_rejection[rows$estimator == "FIML"],
    FIML_max = rows$max_cell_rejection[rows$estimator == "FIML"],
    ML2S_MAE = rows$mean_abs_size_error[rows$estimator == "ML2S"],
    ML2S_min = rows$min_cell_rejection[rows$estimator == "ML2S"],
    ML2S_max = rows$max_cell_rejection[rows$estimator == "ML2S"])
}))
names(estimator_comparison) <- c(
  "method", "FIML MAE", "FIML min", "FIML max",
  "ML2S MAE", "ML2S min", "ML2S max")
knitr::kable(estimator_comparison, digits = 3, row.names = FALSE)
```

ML2S is nearly indistinguishable from FIML under complete sampling, then
generally deteriorates under missingness. In the identified-null region,
projected-score pEBA(4) moves from .013 to .029 mean absolute error and its
maximum rejection rises from .074 to .162; projected-score SB deteriorates
more sharply. ML2S is therefore a secondary negative result rather than part
of the main replacement-method story.

## Retired matched-power snapshot

The exploratory FIML-only Modal continuation is nearly complete but formally
retired because the publication design will replace its models and MAR
mechanisms. The identified-null focus contains 164,275 of 165,000 planned
replications: 162 of 165 cells are complete, and the three partial cells retain
725, 650, and 900 replications. The separately labeled nonnormal-MAR stress
phase completed all 60 cells and 30,000 replications. It is archived but not
mixed into the power comparison below.

```{r retired-power-table}
retired_power <- power_overview[, c(
  "method", "size_mae", "size_min", "size_max",
  "sparse_matched", "diffuse_matched")]
names(retired_power) <- c(
  "method", "size MAE", "size min", "size max",
  "sparse calibrated power", "diffuse calibrated power")
retired_power$method[retired_power$method == "Score pEBA(4)"] <-
  "<strong>Score pEBA(4)</strong>"
knitr::kable(
  retired_power, digits = 3, row.names = FALSE, escape = FALSE)
```

Score pEBA(4) has the best null calibration in this larger run: mean absolute
size error is .010 and the cell range is .027--.067. Score SB has essentially
the same size-calibrated power, but a wider null range (.037--.088). After each
method is calibrated by its matching empirical-null fifth percentile, score
pEBA(4) rejects in .304 of sparse cells and .322 of diffuse cells on average.
LR SS and LR `all` are somewhat weaker, near .293 and .284. MLR appears much
more powerful at its nominal cutoff, but its null rejection reaches .893; after
matching its null cutoff, its power falls to .235 and .227. The apparent MLR
power advantage is therefore null inflation, not useful sensitivity.

The snapshot strengthens the preliminary recommendation of score pEBA(4), but
it does not freeze the paper design. Score SB remains a serious simple
competitor, and the alternatives were tuned for the retired pilot populations.

The retired run used two prespecified misspecifications for each model:

- a **sparse** alternative consisting of one omitted residual covariance; and
- a **diffuse** alternative consisting of an omitted method factor on a small,
  model-specific set of indicators or waves.

```{r power-design-table}
power_design <- data.frame(
  model = unname(model_labels[model_order]),
  sparse = c("x1 ~~ x2", "x1 ~~ x6", "x1 ~~ x5", "x1 ~~ x6",
             "y2 ~~ y3"),
  sparse_effect = c(.1068, .2391, .1562, .1881, .0684),
  diffuse = c("x1, x3, x5", "x1, x3, x6, x8", "x1, x5, x9",
              "x1, x6, x11", "y1, y3, y5"),
  diffuse_effect = c(.1433, .1239, .1264, .1586, .0688),
  stringsAsFactors = FALSE)
names(power_design) <- c(
  "model", "sparse residual pair", "effect", "diffuse method set", "effect")
knitr::kable(power_design, digits = 4, row.names = FALSE)
```

The sparse effect is the added residual covariance as a fraction of the two
variables' baseline standard deviations. The diffuse effect is the variance of
a method factor whose listed loadings equal the variables' baseline standard
deviations. Thus both perturbations are scale-free, but they represent visibly
different local failures: one concentrated residual edge versus shared
cross-block dependence.

Within each model and alternative, the perturbation is calibrated by population
normal-theory ML so that the ordinary likelihood-ratio test has 50% asymptotic
power at $n=200$. Equivalently, the population discrepancy satisfies
$(n-1)F_0=\lambda_{.50}$, where $\lambda_{.50}$ solves the noncentral
$\chi^2_{df}(\lambda)$ power equation at $\alpha=.05$. This common reference
does not tune the design in favor of any robust method. Null, sparse, and
diffuse cells share random-number seeds. The primary power comparison uses each
method's own empirical-null fifth-percentile cutoff in the matching model,
generator, and missingness cell; nominal-threshold power remains descriptive.

The focus was FIML-only over the 55 identified-null mechanisms. A 100-rep
normal-complete audit put ordinary-LR rejection near the intended .50 under
both alternatives in all five pilot models. The three incomplete cells are
shown below; their retained replications enter the exploratory averages rather
than being discarded.

```{r partial-power-cells}
partial_display <- partial_power_cells[, c(
  "cell_id", "model_label", "distribution", "missingness", "truth",
  "available_replications", "expected_replications")]
names(partial_display) <- c(
  "cell", "model", "generator", "missingness", "truth", "available", "planned")
knitr::kable(partial_display, row.names = FALSE)
```

No remaining Modal shards will be repaired. The next production power study
will be recalibrated for the source-based model panel and run through Slurm.

## Model specifications

These are the retired pilot populations used in both snapshots, not the final
paper panel. Their proposed source-based replacements and provenance gaps are
recorded in the [paper-design note](notes/iteration-08-paper-design.md).

```{r model-table}
model_table <- unique(design[, c(
  "model_id", "model_label", "p", "expected_df")])
model_table$free_parameters <- c(18L, 31L, 48L, 48L, 10L)[match(
  model_table$model_id, model_order)]
model_table$model_id <- factor(model_table$model_id, levels = model_order)
model_table <- model_table[order(model_table$model_id), ]
model_table$model <- unname(model_labels[as.character(model_table$model_id)])
model_table <- model_table[, c(
  "model", "p", "free_parameters", "expected_df")]
names(model_table) <- c("model", "observed variables", "free parameters", "df")
knitr::kable(model_table, row.names = FALSE)
```

1. **One-factor CFA, six indicators.** The population is
   $x_j=\lambda_j f+\varepsilon_j$, with
   $\lambda=(1,.8,.9,.7,1.1,.85)$, $\operatorname{Var}(f)=1$, and unique
   variances $(.5,.6,.55,.65,.5,.6)$. Indicator means are zero. The fitted CFA
   uses marker-variable identification and freely estimates means, unique
   variances, the factor variance, and the remaining loadings.

2. **Correlated two-factor CFA, ten indicators.** Indicators 1--5 load only on
   factor 1 with loadings $(.4,.5,.6,.8,.4)$; indicators 6--10 load only on
   factor 2 with loadings $(.4,.7,.6,.4,.8)$. Factor variances are one and the
   factor correlation is .5. Unique variances are $1-\lambda_j^2$, giving unit
   marginal variances, and all means are zero.

3. **Orthogonal bifactor model, twelve indicators.** Every indicator loads .6
   on a general factor. Three orthogonal specific factors each cover four
   indicators, with blockwise loadings .5, .4, and .45. All factor variances are
   one; unique variances are $1-.6^2-\lambda_{S,j}^2$ and means are zero. The
   fitted model uses `std.lv` identification.

4. **Five-wave linear growth model.** At $t=0,1,2,3,4$,
   $y_t=i+t s+\varepsilon_t$. The intercept/slope means are $(1,.2)$, their
   covariance matrix is

   $$
   \begin{pmatrix}
   1 & 0.1 \\
   0.1 & 0.1
   \end{pmatrix},
   $$

   and each residual variance is .5. The fitted model fixes the time scores
   and freely estimates the growth means, latent variance/covariance terms,
   and residual variances.

5. **Three-factor CFA, fifteen indicators.** Five indicators
   load on each of three correlated factors. Within-factor loadings range from
   .4 to .8; factor correlations are .3, .4, and .5; residual variances give
   unit indicator variances. The fitted model uses `std.lv` identification and
   freely estimates 15 loadings, 15 residual variances, 15 means, and three
   factor covariances: 48 parameters against 135 saturated moments, hence 87
   df.

## Methods

Each generated dataset in the completed null panel is fitted by both FIML and
ML2S. The retired power snapshot is FIML-only. The report compares these
calibrations:

- **Scalar robust:** Yuan--Bentler/Mplus MLR for FIML and the scaled two-stage
  statistic for ML2S.
- **LR/discrepancy FMG:** SB, SS, pEBA(4), and `all`, computed from the robust
  eigenvalue spectrum of the FIML likelihood-ratio or ML2S Stage-2 discrepancy.
- **Score FMG:** the same four spectrum transforms applied to the global
  projected score statistic and its own robust eigenvalues. The retired tables
  use expected-H0 sensitivity. The prospective nonnormal-MAR comparison uses
  observed-H0 sensitivity with the expected metric, with expected-H0 and
  observed-H1 variants reported as information diagnostics.

## Paired estimators and the larger model

The runner draws each dataset once, builds the saturated EM stage once, and fits
both FIML and normal-theory two-stage ML to that shared input. The 15-indicator
CFA is included in the same factorial panel as the four smaller models. Pairing
removes Monte Carlo differences in the generated samples when the two
estimators are compared.

## VM and IG generators

```{r generator-table}
generator_table <- data.frame(
  code = c("normal", "vm1", "ig1", "vm2", "ig2"),
  generator = c(
    "Multivariate normal", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)", "Vale--Maurelli / Fleishman (VM)",
    "Independent generator (IG)"),
  marginal_skewness = c(0, 2, 2, 3, 3),
  marginal_excess_kurtosis = c(0, 7, 7, 21, 21),
  severity = c("reference", "moderate", "moderate", "severe", "severe"),
  stringsAsFactors = FALSE)
knitr::kable(generator_table, row.names = FALSE)
```

**VM** is the Vale--Maurelli/Fleishman construction. Magmaan solves an
intermediate Gaussian correlation matrix and applies calibrated marginal
polynomials to recover the SEM population covariance and requested marginal
moments.

**IG** is the Foldnes--Olsson independent-generator construction. Independent
Pearson-family nonnormal generators are calibrated and mixed through a
symmetric covariance root. VM and IG therefore have the same target covariance
and marginal skewness/kurtosis but different multivariate higher-order
dependence.

Each generator is observed completely, with 30% MCAR on every variable except
the first indicator or wave, or with a calibrated 30% MAR mechanism. The MAR
arm keeps the first two variables observed and makes missingness elsewhere
depend on them. Under nonnormal MAR it is deliberately a pseudo-ML stress arm:
normal-theory FIML and saturated normal EM need not target the population
moments even when the SEM covariance structure is correct. In the replacement
study this arm is an alternative unless it is recalibrated so its saturated
Gaussian-FIML pseudo-target satisfies the SEM.

## Appendix: numerical and computational details {.appendix}

### Retired power at nominal and matched cutoffs

The matched columns use each method's empirical-null fifth percentile in the
same model, generator, and missingness cell. The nominal columns explain why
uncalibrated power is misleading when null behavior differs sharply.

```{r retired-power-raw-table}
power_raw_display <- power_overview[, c(
  "method", "sparse_raw", "sparse_matched", "diffuse_raw",
  "diffuse_matched")]
names(power_raw_display) <- c(
  "method", "sparse nominal", "sparse matched", "diffuse nominal",
  "diffuse matched")
knitr::kable(power_raw_display, digits = 3, row.names = FALSE)
```

### Complete 75-cell scorecard

This table reproduces the former headline across all 75 cells and both
estimators. It mixes the identified-null region with the 20 nonnormal-MAR
estimand-stress cells and is therefore descriptive rather than a pure Type-I
summary.

```{r headline-table}
headline <- scorecard[, c(
  "estimator", "method", "mean_cell_rejection", "mean_abs_size_error",
  "min_cell_rejection", "max_cell_rejection", "cells_in_band")]
headline <- headline[scorecard$finite > 0L, ]
names(headline) <- c(
  "estimator", "method", "mean size", "mean |size-.05|", "minimum", "maximum",
  "cells in [.025,.075]")
knitr::kable(headline, digits = 3, row.names = FALSE)
```

### Nonnormal-MAR estimand stress

```{r mar-stress-table}
mar_stress <- mar_stress_scorecard[
  is.finite(mar_stress_scorecard$mean_abs_size_error),
  c("estimator", "method", "mean_abs_size_error",
    "min_cell_rejection", "max_cell_rejection")]
names(mar_stress) <- c(
  "estimator", "method", "mean |rate-.05|", "minimum", "maximum")
knitr::kable(mar_stress, digits = 3, row.names = FALSE)
```

These rates describe rejection relative to the generating SEM moments, not
necessarily Type-I error for the Gaussian-FIML pseudo-true target.

### All 75 design cells

Each entry is the proportion of finite p-values below .05. The method closest
to .05 is bold in each row; exact ties are all bold. The table is intentionally
wide.

```{r all-cells-table}
all_cells <- cell_table[, c(
  "model", "generator", "missingness", wide_method_columns)]
cell_rates <- as.matrix(all_cells[, wide_method_columns])
cell_error <- abs(cell_rates - .05)
best_error <- apply(cell_error, 1L, min, na.rm = TRUE)
for (method in wide_method_columns) {
  value <- cell_rates[, method]
  rendered <- sprintf("%.3f", value)
  rendered[!is.finite(value)] <- "---"
  is_winner <- is.finite(cell_error[, method]) &
    abs(cell_error[, method] - best_error) < 1e-12
  rendered[is_winner] <- paste0("<strong>", rendered[is_winner], "</strong>")
  all_cells[[method]] <- rendered
}
knitr::kable(
  all_cells, row.names = FALSE, escape = FALSE,
  align = c("l", "l", "l", rep("r", length(wide_method_columns))))
```

### Method-specific denominators

Rejection rates use finite, method-specific denominators. Failures are not
counted as non-rejections.

```{r denominator-table}
denominators <- scorecard[, c("estimator", "method", "attempted", "finite")]
denominators <- denominators[denominators$finite > 0L, ]
denominators$non_finite <- denominators$attempted - denominators$finite
names(denominators) <- c(
  "estimator", "method", "attempted", "finite p-values",
  "non-finite or no fit")
knitr::kable(denominators, row.names = FALSE)
```

### Convergence

```{r convergence-table}
show_convergence <- convergence[, c(
  "estimator", "model_label", "attempted", "fit_ok")]
names(show_convergence) <- c(
  "estimator", "model", "attempted", "converged")
knitr::kable(show_convergence, row.names = FALSE)
```

The panel has four fit failures among 75,000 estimator rows: one FIML and three
ML2S. Method-specific non-finite values remain visible in the denominator table
rather than being counted as non-rejections.

### Runtime and next experiment

The checkpointed null panel took approximately 86 minutes on ten workers for
75,000 paired-estimator rows and the LR/discrepancy and score-FMG batteries.
The retired Modal continuation preserved 164,275 focus rows and all 30,000
nonnormal-MAR stress rows in the ignored local results cache.

### Reproduce

The archived null panel remains the report's default calibration input. The
retired power snapshot is read from `results/power-v1-20260821/`. To combine an
incomplete downloaded focus phase while retaining its coverage ledger:

```sh
Rscript experiments/77-fiml-global-gof-pilot/modal/combine.R \
  --out-dir experiments/77-fiml-global-gof-pilot/results/power-v1-20260821/focus \
  --profile focus --expected-cells 165 --allow-incomplete --skip-summaries
```

The Modal harness is retained only for provenance. The source-based replacement
design will use a Slurm harness and a new run identifier.
