---
title: "Heywood Box-Constraint Experiment"
format:
  html:
    toc: true
---

```{r}
#| echo: false
read_result <- function(name) {
  path <- file.path("results", name)
  if (!file.exists(path)) {
    stop("Missing result file: ", path, ". Run run_experiment.R first.")
  }
  read.csv(path, stringsAsFactors = FALSE)
}

lavaan_bounds <- read_result("lavaan_bounds.csv")
magmaan_unbounded <- read_result("magmaan_unbounded.csv")
summary <- read_result("summary.csv")
surface_gap <- read_result("surface_gap.csv")
```

## Question

The random-start cartography found stationary ML solutions for the hard Study 3
MSST cases, but those solutions were usually inadmissible: negative observed or
latent variance parameters even when the implied covariance matrix was positive
definite.

This experiment asks whether variance box constraints turn those cases into
admissible boundary optima.

## R Surface Status

```{r}
surface_gap
```

At the time of this experiment, lavaan bounded ML is the reference probe. The
magmaan C++ core accepts bounds for ML, but the R ML wrapper does not yet expose
that path.

## Lavaan Bounded ML

```{r}
summary[order(summary$case_key, summary$bounds, summary$rstarts), ]
```

## Magmaan Unbounded Baseline

```{r}
magmaan_unbounded[, c("case_key", "converged", "fmin", "grad_inf_norm",
                      "min_ov_var", "min_lv_var",
                      "n_neg_ov_var", "n_neg_lv_var", "error")]
```

## Boundary Interpretation

For bounded fits, a large raw gradient is not automatically a failure. If a
variance is active at its lower bound, the relevant diagnostic is a KKT
condition for the constrained problem, not unconstrained stationarity.

