---
title: "Heywood Box-Constraint Experiment"
format:
  html:
    toc: true
---

```{r}
#| echo: false
read_result <- function(name) {
  path <- file.path("results", name)
  if (!file.exists(path)) {
    stop("Missing result file: ", path, ". Run run_experiment.R first.")
  }
  read.csv(path, stringsAsFactors = FALSE)
}

lavaan_bounds <- read_result("lavaan_bounds.csv")
magmaan_bounds <- read_result("magmaan_bounds.csv")
lavaan_summary <- read_result("summary.csv")
magmaan_summary <- read_result("magmaan_summary.csv")
engine_comparison <- read_result("engine_comparison.csv")
surface_gap <- read_result("surface_gap.csv")

yes_no <- function(x) {
  ifelse(is.na(x), NA_character_, ifelse(x >= 1, "yes", "no"))
}

fmt <- function(x, digits = 4) {
  ifelse(is.na(x), NA_character_, formatC(x, digits = digits, format = "fg"))
}

comparison0 <- engine_comparison[engine_comparison$lavaan_rstarts == 0L, ]
headline <- comparison0[comparison0$bounds %in% c("none", "pos.var"), ]
headline <- headline[, c("case_key", "bounds",
                         "lavaan_converged", "lavaan_admissible",
                         "lavaan_active_variance_bounds",
                         "magmaan_converged", "magmaan_stationary",
                         "magmaan_admissible",
                         "magmaan_active_variance_bounds",
                         "fmin_gap_magmaan_minus_2_lavaan")]
headline[] <- Map(function(col, name) {
  if (name %in% c("lavaan_converged", "lavaan_admissible",
                  "magmaan_converged", "magmaan_stationary",
                  "magmaan_admissible")) {
    return(yes_no(col))
  }
  if (name == "fmin_gap_magmaan_minus_2_lavaan") return(fmt(col, 6))
  col
}, headline, names(headline))

posvar0 <- comparison0[comparison0$bounds == "pos.var", ]
posvar_admissible <- sum(posvar0$lavaan_admissible == 1 &
                           posvar0$magmaan_admissible == 1, na.rm = TRUE)
posvar_gap <- max(abs(posvar0$fmin_gap_magmaan_minus_2_lavaan), na.rm = TRUE)

posvar_r0 <- engine_comparison[engine_comparison$bounds == "pos.var" &
                                 engine_comparison$lavaan_rstarts == 0L,
                               c("case_key", "lavaan_fmin")]
posvar_r150 <- engine_comparison[engine_comparison$bounds == "pos.var" &
                                   engine_comparison$lavaan_rstarts == 150L,
                                 c("case_key", "lavaan_fmin")]
names(posvar_r0)[2L] <- "lavaan_fmin_r0"
names(posvar_r150)[2L] <- "lavaan_fmin_r150"
posvar_random <- merge(posvar_r0, posvar_r150, by = "case_key", all = TRUE)
posvar_random$delta <- posvar_random$lavaan_fmin_r150 -
  posvar_random$lavaan_fmin_r0
random_improved <- posvar_random$case_key[posvar_random$delta < -1e-6]
random_improved <- if (length(random_improved)) {
  paste(random_improved, collapse = ", ")
} else {
  "none"
}

wide0 <- comparison0[comparison0$bounds == "wide", ]
wide_admissible <- sum(wide0$lavaan_admissible == 1 &
                         wide0$magmaan_admissible == 1, na.rm = TRUE)

finding_summary <- data.frame(
  check = c(
    "pos.var admissibility",
    "pos.var objective agreement",
    "pos.var random-start sensitivity",
    "wide admissibility"
  ),
  result = c(
    paste0(posvar_admissible, "/", nrow(posvar0),
           " regular-start cases admissible in both engines"),
    paste0("max |magmaan_fmin - 2 * lavaan_fmin| = ",
           formatC(posvar_gap, digits = 4, format = "fg")),
    paste0("lavaan rstarts=150 found a lower pos.var basin for: ",
           random_improved),
    paste0(wide_admissible, "/", nrow(wide0),
           " regular-start cases admissible in both engines")
  ),
  takeaway = c(
    "The clean positive result is about nonnegative variance bounds.",
    "Regular-start magmaan matches regular-start lavaan for pos.var.",
    "Bounds do not remove the local-basin/start-value problem.",
    "Wide is a loose lavaan box, not a nonnegative-variance constraint."
  ),
  stringsAsFactors = FALSE
)
```

## Question

The random-start cartography found stationary ML solutions for the hard Study 3
MSST cases, but those solutions were usually inadmissible: negative observed or
latent variance parameters even when the implied covariance matrix was positive
definite.

This experiment asks whether variance box constraints turn those cases into
admissible boundary optima.

## Short Answer

Yes, the results matter, but narrowly. This is a mechanism probe, not a broad
success-rate benchmark. The useful question is whether the same sample
covariance that gives an unbounded Heywood solution can be fit as a constrained
boundary optimum when variance parameters are required to be nonnegative.

The clean pattern to look for is:

- `bounds = "none"` converges but has negative variance estimates;
- `bounds = "pos.var"` converges with no negative variance estimates;
- the constrained solution has active variance bounds;
- magmaan's `fmin` matches `2 * lavaan_fmin` to numerical tolerance;

then the experiment says something concrete: the failure mode is not just "the
optimizer got confused." It is the geometry of the unbounded SEM problem. The
bounded problem has a legitimate constrained optimum on the boundary, and our
convergence checks must be KKT/projected-gradient checks rather than ordinary
interior-gradient checks.

The `standard` and `wide` rows are supporting probes, not the central claim.
`standard` is another nonnegative-ish lavaan box in these cases, but `wide` is
a deliberately loose box and can still allow negative variances. Also, lavaan's
`rstarts = 150` condition checks basin sensitivity; if it finds a lower
objective than the regular start, bounds helped admissibility but did not solve
the start-value problem.

What it does *not* say: bounds should become a statistical default, random
starts are unnecessary, or ordinary interior standard errors are automatically
valid at a boundary solution.

## Headline Pattern

```{r}
knitr::kable(headline, row.names = FALSE)
```

Read this table as a small decision grid. The `none` rows show the unbounded ML
target. The `pos.var` rows show the smallest intervention: only free variance
diagonals get lower bound zero. `fmin_gap_magmaan_minus_2_lavaan` should be
near zero when magmaan and lavaan reached the same constrained objective value,
because magmaan reports the full ML discrepancy and lavaan reports half.

## Default Run Findings

```{r}
knitr::kable(finding_summary, row.names = FALSE)
```

The headline is therefore: `pos.var` is doing real work. It turns the default
hard cases into admissible constrained fits and magmaan agrees with lavaan's
regular-start objective. But there is still a basin problem in at least one
case: lavaan random starts find a lower `pos.var` objective for
`study3_msst:805`. So the result matters most as evidence for two separate
engineering requirements: bounded optimization and better start portfolios.

## R Surface Status

```{r}
surface_gap
```

The R surface now threads the named bound presets through complete-data ML, so
the same cases can be inspected with lavaan and magmaan. Remember that
magmaan's `fmin` is the full discrepancy, while lavaan reports half of it.

## Lavaan Bounded ML

```{r}
knitr::kable(lavaan_summary[order(lavaan_summary$case_key,
                                  lavaan_summary$bounds,
                                  lavaan_summary$rstarts), ],
             row.names = FALSE, digits = 4)
```

## Magmaan Bounded ML

```{r}
knitr::kable(magmaan_summary[, c("case_key", "bounds", "converged",
                                 "stationary", "admissible_variances", "fmin",
                                 "grad_inf_norm",
                                 "active_variance_bound_count",
                                 "audit_status", "error")],
             row.names = FALSE, digits = 4)
```

## Engine Agreement

```{r}
knitr::kable(engine_comparison[, c(
  "case_key", "bounds", "lavaan_rstarts",
  "lavaan_converged", "lavaan_admissible",
  "magmaan_converged", "magmaan_stationary", "magmaan_admissible",
  "fmin_gap_magmaan_minus_2_lavaan"
)], row.names = FALSE, digits = 6)
```

## Boundary Interpretation

For bounded fits, a large raw gradient is not automatically a failure. If a
variance is active at its lower bound, the relevant diagnostic is a KKT
condition for the constrained problem, not unconstrained stationarity.

That is why this experiment tracks both variance admissibility and active
bounds. A boundary optimum should generally have a nonzero raw gradient in some
bounded coordinates. The sign of the gradient, together with the active set, is
what matters. In magmaan this is surfaced through the terminal audit's
projected-gradient stationarity and the fit diagnostics' active-bound indices.

## Practical Takeaway

For these hard small-sample MSST cases, the result is worth keeping around as a
regression and design probe:

- It justifies exposing bounds in R, because the exact question cannot be
  studied honestly from the R surface without them.
- It gives a compact sanity check that nonnegative variance boxes can turn an
  inadmissible unbounded solution into an admissible constrained one.
- It forces convergence reporting to separate optimizer stationarity from SEM
  admissibility and from inference readiness.

It should not yet drive a default-policy change. Boundary solutions are
statistically delicate: they may be a reasonable constrained estimate, but they
can also be a symptom of small samples, misspecification, weak identification,
or a model that wants to leave the admissible parameter space. The next
experiment should ask when bounded starts, bounded optimization, and screened
start portfolios improve the probability of finding a useful local solution
across a larger portfolio, and how often the solution is truly boundary-active
versus merely protected from a bad basin.
