---
title: "Heywood Box-Constraint Experiment"
format:
  html:
    toc: true
---

```{r}
#| echo: false
read_result <- function(name) {
  path <- file.path("results", name)
  if (!file.exists(path)) {
    stop("Missing result file: ", path, ". Run run_experiment.R first.")
  }
  read.csv(path, stringsAsFactors = FALSE)
}

lavaan_bounds <- read_result("lavaan_bounds.csv")
magmaan_bounds <- read_result("magmaan_bounds.csv")
lavaan_summary <- read_result("summary.csv")
magmaan_summary <- read_result("magmaan_summary.csv")
engine_comparison <- read_result("engine_comparison.csv")
surface_gap <- read_result("surface_gap.csv")

yes_no <- function(x) {
  ifelse(is.na(x), NA_character_, ifelse(x >= 1, "yes", "no"))
}

fmt <- function(x, digits = 4) {
  ifelse(is.na(x), NA_character_, formatC(x, digits = digits, format = "fg"))
}

comparison0 <- engine_comparison[engine_comparison$lavaan_rstarts == 0L, ]
headline <- comparison0[comparison0$bounds %in% c("none", "pos.var"), ]
headline <- headline[, c("case_key", "identification", "bounds",
                         "lavaan_admissible",
                         "lavaan_n_neg_ov_var", "lavaan_n_neg_lv_var",
                         "magmaan_stationary", "magmaan_admissible",
                         "magmaan_n_neg_ov_var", "magmaan_n_neg_lv_var",
                         "magmaan_active_variance_bounds",
                         "fmin_gap_magmaan_minus_2_lavaan")]
headline[] <- Map(function(col, name) {
  if (name %in% c("lavaan_admissible", "magmaan_stationary",
                  "magmaan_admissible")) {
    return(yes_no(col))
  }
  if (name == "fmin_gap_magmaan_minus_2_lavaan") return(fmt(col, 6))
  col
}, headline, names(headline))

posvar0 <- comparison0[comparison0$bounds == "pos.var", ]
marker_posvar0 <- posvar0[posvar0$identification == "marker", ]
marker_posvar_good <- sum(marker_posvar0$lavaan_converged == 1 &
                            marker_posvar0$lavaan_admissible == 1 &
                            marker_posvar0$magmaan_converged == 1 &
                            marker_posvar0$magmaan_stationary == 1 &
                            marker_posvar0$magmaan_admissible == 1,
                          na.rm = TRUE)
marker_posvar_gap <- max(
  abs(marker_posvar0$fmin_gap_magmaan_minus_2_lavaan),
  na.rm = TRUE
)

stdlv_posvar0 <- posvar0[posvar0$identification == "std.lv", ]
stdlv_posvar_lavaan_converged <- sum(stdlv_posvar0$lavaan_converged == 1,
                                     na.rm = TRUE)
stdlv_posvar_gap <- max(abs(stdlv_posvar0$fmin_gap_magmaan_minus_2_lavaan),
                        na.rm = TRUE)

posvar_r0 <- engine_comparison[engine_comparison$bounds == "pos.var" &
                                 engine_comparison$lavaan_rstarts == 0L,
                               c("case_key", "identification", "lavaan_fmin")]
posvar_r150 <- engine_comparison[engine_comparison$bounds == "pos.var" &
                                   engine_comparison$lavaan_rstarts == 150L,
                                 c("case_key", "identification", "lavaan_fmin")]
names(posvar_r0)[3L] <- "lavaan_fmin_r0"
names(posvar_r150)[3L] <- "lavaan_fmin_r150"
posvar_random <- merge(posvar_r0, posvar_r150,
                       by = c("case_key", "identification"), all = TRUE)
posvar_random$delta <- posvar_random$lavaan_fmin_r150 -
  posvar_random$lavaan_fmin_r0
random_improved <- paste(posvar_random$case_key,
                         posvar_random$identification, sep = "/")
random_improved <- random_improved[posvar_random$delta < -1e-6]
random_improved <- if (length(random_improved)) {
  paste(random_improved, collapse = ", ")
} else {
  "none"
}

wide0 <- comparison0[comparison0$bounds == "wide", ]
wide_admissible <- sum(wide0$lavaan_admissible == 1 &
                         wide0$magmaan_admissible == 1, na.rm = TRUE)

stdlv_none0 <- comparison0[comparison0$bounds == "none" &
                             comparison0$identification == "std.lv", ]
stdlv_none_admissible <- sum(stdlv_none0$lavaan_admissible == 1 &
                               stdlv_none0$magmaan_admissible == 1,
                             na.rm = TRUE)

stdlv_probe <- comparison0[comparison0$bounds == "none",
                           c("case_key", "identification",
                             "lavaan_admissible",
                             "lavaan_min_ov_var", "lavaan_min_lv_var",
                             "magmaan_admissible",
                             "magmaan_min_ov_var", "magmaan_min_lv_var")]
stdlv_probe[] <- Map(function(col, name) {
  if (name %in% c("lavaan_admissible", "magmaan_admissible")) {
    return(yes_no(col))
  }
  if (name %in% c("lavaan_min_ov_var", "lavaan_min_lv_var",
                  "magmaan_min_ov_var", "magmaan_min_lv_var")) {
    return(fmt(col, 5))
  }
  col
}, stdlv_probe, names(stdlv_probe))

finding_summary <- data.frame(
  check = c(
    "std.lv without bounds",
    "marker pos.var",
    "marker pos.var objective agreement",
    "std.lv pos.var caveat",
    "pos.var random-start sensitivity",
    "wide admissibility"
  ),
  result = c(
    paste0(stdlv_none_admissible, "/", nrow(stdlv_none0),
           " std.lv regular-start cases admissible in both engines"),
    paste0(marker_posvar_good, "/", nrow(marker_posvar0),
           " regular-start marker cases converged, stationary, and admissible in both engines"),
    paste0("max |magmaan_fmin - 2 * lavaan_fmin| = ",
           formatC(marker_posvar_gap, digits = 4, format = "fg")),
    paste0("lavaan regular-start convergence ",
           stdlv_posvar_lavaan_converged, "/", nrow(stdlv_posvar0),
           "; max finite fmin gap = ",
           formatC(stdlv_posvar_gap, digits = 4, format = "fg")),
    paste0("lavaan rstarts=150 found a lower pos.var basin for: ",
           random_improved),
    paste0(wide_admissible, "/", nrow(wide0),
           " regular-start cases admissible in both engines")
  ),
  takeaway = c(
    "The latent metric alone does not remove the Heywood behavior.",
    "The clean positive result remains marker identification plus nonnegative variance bounds.",
    "Regular-start marker magmaan matches regular-start marker lavaan for pos.var.",
    "The bounded std.lv rows are diagnostic, not an engine-parity claim.",
    "Bounds do not remove the local-basin/start-value problem.",
    "Wide is a loose lavaan box, not a nonnegative-variance constraint."
  ),
  stringsAsFactors = FALSE
)
```

## Question

The random-start cartography found stationary ML solutions for the hard Study 3
MSST cases, but those solutions were usually inadmissible: negative observed or
latent variance parameters even when the implied covariance matrix was positive
definite.

This experiment asks whether variance box constraints turn those cases into
admissible boundary optima.

It now also asks a smaller identification question: without any bounds, does
`std.lv` merely remove latent-variance Heywoods by fixing factor variances, or
does the changed metric produce an admissible solution?

## Short Answer

Yes, the results matter, but narrowly. This is a mechanism probe, not a broad
success-rate benchmark. The useful question is whether the same sample
covariance that gives an unbounded Heywood solution can be fit as a constrained
boundary optimum when variance parameters are required to be nonnegative.

The clean pattern to look for is:

- `bounds = "none"` converges but has negative variance estimates;
- `bounds = "pos.var"` converges with no negative variance estimates;
- the constrained solution has active variance bounds;
- magmaan's `fmin` matches `2 * lavaan_fmin` to numerical tolerance;

then the experiment says something concrete: the failure mode is not just "the
optimizer got confused." It is the geometry of the unbounded SEM problem. The
bounded problem has a legitimate constrained optimum on the boundary, and our
convergence checks must be KKT/projected-gradient checks rather than ordinary
interior-gradient checks.

The `standard` and `wide` rows are supporting probes, not the central claim.
`standard` is another nonnegative-ish lavaan box in these cases, but `wide` is
a deliberately loose box and can still allow negative variances. Also, lavaan's
`rstarts = 150` condition checks basin sensitivity; if it finds a lower
objective than the regular start, bounds helped admissibility but did not solve
the start-value problem.

The `std.lv` rows answer a different question. They change the latent metric,
not the feasible set for residual variances. A successful `std.lv` row without
bounds would therefore be evidence that the marker metric was part of the
numerical problem. A failed `std.lv` row, especially one with negative observed
variances, says the Heywood behavior is not merely a latent-scale artifact.

What it does *not* say: bounds should become a statistical default, random
starts are unnecessary, or ordinary interior standard errors are automatically
valid at a boundary solution.

## Headline Pattern

```{r}
knitr::kable(headline, row.names = FALSE)
```

Read this table as a small decision grid. The `none` rows show the unbounded ML
target. The `pos.var` rows show the smallest intervention: only free variance
diagonals get lower bound zero. `fmin_gap_magmaan_minus_2_lavaan` should be
near zero when magmaan and lavaan reached the same constrained objective value,
because magmaan reports the full ML discrepancy and lavaan reports half.

## std.lv Without Bounds

```{r}
knitr::kable(stdlv_probe, row.names = FALSE)
```

Under `std.lv`, latent variances are fixed to one. The useful read is therefore
the observed-variance columns. If these are still negative without bounds, the
metric swap did not solve the admissibility problem; it only removed one place
where the problem can be expressed.

## Default Run Findings

```{r}
knitr::kable(finding_summary, row.names = FALSE)
```

The marker-identification headline is therefore: `pos.var` is doing real work.
It turns the default hard cases into admissible constrained fits and magmaan
agrees with lavaan's regular-start objective. But there is still a basin
problem in at least one case: lavaan random starts find a lower `pos.var`
objective for `study3_msst:805/marker`. The `std.lv` rows add a different
lesson: the metric swap alone does not remove the observed-variance Heywoods,
and bounded `std.lv` regular-start lavaan is not stable enough here to treat as
an objective-agreement check. So the result matters most as evidence for two
separate engineering requirements: bounded optimization and better start
portfolios.

## R Surface Status

```{r}
surface_gap
```

The R surface now threads the named bound presets through complete-data ML, so
the same cases can be inspected with lavaan and magmaan. Remember that
magmaan's `fmin` is the full discrepancy, while lavaan reports half of it.

## Lavaan Bounded ML

```{r}
knitr::kable(lavaan_summary[order(lavaan_summary$case_key,
                                  lavaan_summary$identification,
                                  lavaan_summary$bounds,
                                  lavaan_summary$rstarts), ],
             row.names = FALSE, digits = 4)
```

## Magmaan Bounded ML

```{r}
knitr::kable(magmaan_summary[, c("case_key", "identification", "bounds",
                                 "converged",
                                 "stationary", "admissible_variances", "fmin",
                                 "grad_inf_norm",
                                 "active_variance_bound_count",
                                 "audit_status", "error")],
             row.names = FALSE, digits = 4)
```

## Engine Agreement

```{r}
knitr::kable(engine_comparison[, c(
  "case_key", "identification", "bounds", "lavaan_rstarts",
  "lavaan_converged", "lavaan_admissible",
  "magmaan_converged", "magmaan_stationary", "magmaan_admissible",
  "fmin_gap_magmaan_minus_2_lavaan"
)], row.names = FALSE, digits = 6)
```

## Boundary Interpretation

For bounded fits, a large raw gradient is not automatically a failure. If a
variance is active at its lower bound, the relevant diagnostic is a KKT
condition for the constrained problem, not unconstrained stationarity.

That is why this experiment tracks both variance admissibility and active
bounds. A boundary optimum should generally have a nonzero raw gradient in some
bounded coordinates. The sign of the gradient, together with the active set, is
what matters. In magmaan this is surfaced through the terminal audit's
projected-gradient stationarity and the fit diagnostics' active-bound indices.

## Practical Takeaway

For these hard small-sample MSST cases, the result is worth keeping around as a
regression and design probe:

- It justifies exposing bounds in R, because the exact question cannot be
  studied honestly from the R surface without them.
- It gives a compact sanity check that nonnegative variance boxes can turn an
  inadmissible unbounded solution into an admissible constrained one.
- It forces convergence reporting to separate optimizer stationarity from SEM
  admissibility and from inference readiness.

It should not yet drive a default-policy change. Boundary solutions are
statistically delicate: they may be a reasonable constrained estimate, but they
can also be a symptom of small samples, misspecification, weak identification,
or a model that wants to leave the admissible parameter space. The next
experiment should ask when bounded starts, bounded optimization, and screened
start portfolios improve the probability of finding a useful local solution
across a larger portfolio, and how often the solution is truly boundary-active
versus merely protected from a bad basin.
