---
title: "Studentized permutation Wald tests for measurement invariance"
bibliography: references.bib
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
choose_run <- function(study) {
  candidates <- file.path("results", paste0(study, "-", c("full", "modal", "smoke")))
  found <- candidates[file.exists(file.path(candidates, "summary.csv"))]
  if (!length(found)) return(NULL)
  path <- found[[1L]]
  list(
    study = study,
    profile = sub(paste0("^", study, "-"), "", basename(path)),
    path = path,
    summary = read.csv(file.path(path, "summary.csv"), check.names = FALSE,
                       stringsAsFactors = FALSE),
    manifest = read.csv(file.path(path, "manifest.csv"), check.names = FALSE,
                        stringsAsFactors = FALSE),
    config = read.csv(file.path(path, "run_config.csv"), check.names = FALSE,
                      stringsAsFactors = FALSE))
}
runs <- Filter(Negate(is.null), lapply(c("reference", "imbalance"), choose_run))
if (!length(runs)) {
  stop("No new-format results. Run `Rscript run_experiment.R --smoke` first.")
}
all_summary <- do.call(rbind, lapply(runs, `[[`, "summary"))
profiles <- paste(vapply(runs, function(x)
  sprintf("%s=%s (%d rep, %d permutations)", x$study, x$profile,
          x$config$reps[[1L]], x$config$permutations[[1L]]), character(1)),
  collapse = "; ")
is_smoke <- any(vapply(runs, `[[`, character(1), "profile") == "smoke")
```

## Result status

This render uses `r profiles`.

```{r smoke-warning, results='asis'}
if (is_smoke) cat(
  "**Smoke results are wiring and timing evidence only.** With nine random ",
  "permutations the smallest attainable plus-one p-value is 0.10, so the ",
  "permutation rejection columns cannot estimate rejection rates at 0.05.\n")
```

## Scientific question

Does a fully refitted Wald statistic, studentized with a freshly estimated
empirical sandwich covariance after every group-label permutation, provide
better calibrated measurement-invariance inference when group observations are
not exchangeable? The design targets Chung and Romano's asymptotically pivotal
case: raw exchangeability is unnecessary if the recomputed statistic has a
common limiting null law.

The comparison includes ordinary and robust/scaled likelihood-ratio tests,
model-based multivariate Wald tests, sandwich Wald tests, the studentized
permutation Wald test, and a raw-contrast permutation negative control. In the
unbalanced study, the model-based Wald is Chen and Chao's MWT restriction test.

## Executed design

```{r design-table}
design_rows <- do.call(rbind, lapply(runs, function(run) {
  m <- run$manifest
  data.frame(
    Study = run$study, Profile = run$profile,
    Cells = nrow(m), Steps = paste(unique(m$step), collapse = ", "),
    Generators = paste(unique(m$generator), collapse = ", "),
    `N range` = paste(range(m$n_total), collapse = "--"),
    Allocations = paste(unique(m$allocation), collapse = ", "),
    check.names = FALSE)
}))
knitr::kable(design_rows, row.names = FALSE,
             caption = "Manifests used by this report render.")
```

The reference-law paper grid contains two correlated factors, `p = 8/16`, total
`N = 220/440/1760`, balanced and `1:3` allocation, normal and Pearson IG
nonnormal data, and ordinal-probit integer scores with explicit category and
threshold-shape factors. Its heterogeneous null retains equal measurement
parameters while factor means and factor/residual covariance differ. Separate
exchangeable controls make both group populations identical.

The unbalanced paper grid reproduces Chen and Chao's six-indicator population:
loading 0.6 and intercept 0 for each item, unit factor and residual variances,
group A `n = 200`, and group B from 400 through 3000. Metric/scalar violations
affect items `{2,4}` or `{2,3,4,5}` in group A using their small, large, mixed,
and non-uniform changes. Matched-balanced cells retain the same total N. Strict
invariance is the proportional residual-variance extension.

## Timing and availability

```{r timing-table}
timing <- aggregate(
  cbind(median_simulation_seconds, median_observed_seconds,
        median_permutation_seconds, median_total_seconds, failure_rate) ~
    study + generator + step,
  data = all_summary, FUN = function(x) median(x, na.rm = TRUE))
names(timing) <- c("Study", "Generator", "Step", "Simulation s", "Observed s",
                   "Permutation s", "Total s", "Failure rate")
num <- vapply(timing, is.numeric, logical(1))
timing[num] <- lapply(timing[num], function(x) round(x, 3))
knitr::kable(timing, row.names = FALSE,
             caption = "Median per-replication stage timings across selected cells.")
```

Invalid permutation fits are retained. `summary.csv` reports the median usable
permutation fraction and method-specific usable replication counts; any
observed-fit errors are also written to `failures.csv`.

## Rejection summaries

```{r rejection-table}
cols <- c("study", "cell_id", "p", "n1", "n2", "generator", "step", "truth",
          "null_kind", "pattern", "affected_fraction", "reps", "observed_ok",
          "reject_lrt", "reject_robust_lrt", "reject_model_wald",
          "reject_sandwich_wald", "reject_permutation_studentized",
          "reject_permutation_raw", "median_valid_permutation_fraction")
tab <- all_summary[, intersect(cols, names(all_summary)), drop = FALSE]
names(tab) <- gsub("_", " ", names(tab), fixed = TRUE)
numeric <- vapply(tab, is.numeric, logical(1))
tab[numeric] <- lapply(tab[numeric], function(x) round(x, 3))
knitr::kable(tab, row.names = FALSE,
             caption = "Cell-level rejection and availability summaries.")
```

Configural is fitted only as the metric test's internal H1. The reported
outcomes are metric versus configural, scalar versus the identified metric mean
chart, and strict versus scalar.

## Interpretation boundaries

- Reference alternatives use explicit step-specific deltas. The defaults are
  pilot starting values; paper claims require replacing them with departures
  calibrated to 50% studentized-permutation power in the normal balanced
  `N = 440` reference cell.
- Ordinal-probit responses are analyzed as ordinary numeric codes by continuous
  ML. This does not claim validity for a shuffled DWLS/polychoric statistic; that
  is a separate problem. Because discretization is nonlinear, latent-response
  invariance need not imply exact invariance of the continuous-ML pseudo-
  parameters, especially with asymmetric thresholds; that drift is part of the
  deliberately misspecified applied workflow and must be diagnosed separately
  from reference-law calibration.
- The asymptotic argument requires regular estimation under the pooled mixture.
  Exchangeable controls diagnose implementation errors, while the heterogeneous
  nulls test the scientifically relevant departure from raw exchangeability.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript calibrate_alternatives.R --smoke
Rscript run_experiment.R --study reference --modal --cores 10
Rscript run_experiment.R --study imbalance --modal --cores 10
quarto render report.qmd
```
