---
title: "Can a studentized permutation Wald test calibrate metric invariance?"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
summary_path <- file.path("results", "summary.csv")
if (!file.exists(summary_path)) {
  stop("No results. Run:\n  Rscript run_experiment.R --smoke\nthen:\n  Rscript run_experiment.R --full --reps 1000 --permutations 499")
}
results <- read.csv(summary_path, check.names = FALSE, stringsAsFactors = FALSE)
fmt <- function(x) formatC(x, format = "f", digits = 3)
```

## Question

Can a raw group-label permutation distribution of a fully recomputed,
studentized Wald statistic control Type-I error for continuous metric invariance
when group sample sizes, factor covariances, residual variances, and CFA size
vary? The comparison is against ordinary and robust likelihood-ratio and Wald
tests.

## Short answer

This feasibility experiment is intentionally a gate rather than a finished
method claim. The `invariant` rows show whether the studentized permutation
test stays near 0.05 under the heterogeneous null; the `loading_violation` rows
show whether that calibration retains useful power. The ordinary raw-contrast
permutation is included as a negative control: a gap between it and the
studentized column is the finite-sample behavior the Chung--Romano argument
predicts.

## Evidence

Rejection rates at the configured alpha. The two permutation methods refit the
configural CFA and recompute their statistic after every relabeling; they do
not reuse the observed covariance estimate.

```{r table}
cols <- c("model", "heterogeneity", "distribution", "scenario", "n1", "n2", "reps", "converged",
          "reject_lrt", "reject_robust_lrt", "reject_model_wald",
          "reject_sandwich_wald", "reject_permutation_studentized",
          "reject_permutation_raw", "median_valid_permutations")
tab <- results[, intersect(cols, names(results)), drop = FALSE]
names(tab) <- c("Model", "Heterogeneity", "Distribution", "Scenario", "N1", "N2", "Replications",
                "Converged", "LRT", "Robust LRT", "Model Wald",
                "Sandwich Wald", "Studentized permutation", "Raw permutation",
                "Median valid permutations")[seq_len(ncol(tab))]
numeric <- vapply(tab, is.numeric, logical(1))
tab[numeric] <- lapply(tab[numeric], function(x) round(x, 3))
knitr::kable(tab, row.names = FALSE,
             caption = "Metric-invariance rejection rates and permutation availability.")
```

## Caveats

- The Chung--Romano argument requires the estimator and its studentization to
  remain asymptotically regular under the pooled mixture, not merely under the
  two source groups. This experiment checks one- and three-factor CFA
  populations at standard and stronger group heterogeneity.
- The `exchangeable` scenario is a finite-sample control: both group
  distributions are identical, so a random label permutation is valid even for
  the raw contrast statistic. It distinguishes a resampling implementation bug
  from a failure of asymptotic robustness under heterogeneous populations.
- The ordinal arm is deliberately absent. A mixture of ordinal-probit groups is
  generally not ordinal-probit, so it needs a separate validity result or a
  different resampling scheme.
- Smoke output is a wiring and runtime check only. The full grid should use at
  least 1,000 replications before interpreting a 0.05 rejection rate.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --full --reps 1000 --permutations 499
quarto render report.qmd
```
