---
title: "Pivotal permutation tests for measurement invariance"
bibliography: references.bib
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
choose_run <- function(study) {
  candidates <- file.path(
    "results", paste0(study, "-", c("full", "sensitivity", "modal", "smoke")))
  found <- candidates[file.exists(file.path(candidates, "summary.csv"))]
  if (!length(found)) return(NULL)
  path <- found[[1L]]
  list(
    study = study,
    profile = sub(paste0("^", study, "-"), "", basename(path)),
    path = path,
    summary = read.csv(file.path(path, "summary.csv"), check.names = FALSE,
                       stringsAsFactors = FALSE),
    manifest = read.csv(file.path(path, "manifest.csv"), check.names = FALSE,
                        stringsAsFactors = FALSE),
    config = read.csv(file.path(path, "run_config.csv"), check.names = FALSE,
                      stringsAsFactors = FALSE))
}
runs <- Filter(Negate(is.null), lapply(c("reference", "imbalance"), choose_run))
if (!length(runs)) {
  stop("No new-format results. Run `Rscript run_experiment.R --smoke` first.")
}
all_summary <- do.call(rbind, lapply(runs, `[[`, "summary"))
profiles <- paste(vapply(runs, function(x)
  sprintf("%s=%s (%d rep, %d permutations)", x$study, x$profile,
          x$config$reps[[1L]], x$config$permutations[[1L]]), character(1)),
  collapse = "; ")
is_smoke <- any(vapply(runs, `[[`, character(1), "profile") == "smoke")

choose_score_run <- function() {
  candidates <- file.path(
    "results", paste0("score-pivot-", c("full", "probe", "smoke")))
  found <- candidates[file.exists(file.path(candidates, "summary.csv"))]
  if (!length(found)) return(NULL)
  path <- found[[1L]]
  list(
    profile = sub("^score-pivot-", "", basename(path)),
    path = path,
    summary = read.csv(file.path(path, "summary.csv"), check.names = FALSE,
                       stringsAsFactors = FALSE),
    replications = read.csv(
      file.path(path, "replications.csv"), check.names = FALSE,
      stringsAsFactors = FALSE),
    config = read.csv(file.path(path, "run_config.csv"), check.names = FALSE,
                      stringsAsFactors = FALSE))
}
score_run <- choose_score_run()
sensitivity_path <- file.path("results", "score-pivot-sensitivity")
score_sensitivity <- if (file.exists(
    file.path(sensitivity_path, "summary.csv"))) {
  list(
    summary = read.csv(
      file.path(sensitivity_path, "summary.csv"), check.names = FALSE,
      stringsAsFactors = FALSE),
    replications = read.csv(
      file.path(sensitivity_path, "replications.csv"), check.names = FALSE,
      stringsAsFactors = FALSE),
    config = read.csv(
      file.path(sensitivity_path, "run_config.csv"), check.names = FALSE,
      stringsAsFactors = FALSE))
} else NULL

read_score_arm <- function(stem) {
  path <- file.path("results", stem)
  if (!file.exists(file.path(path, "summary.csv"))) return(NULL)
  list(
    path = path,
    summary = read.csv(
      file.path(path, "summary.csv"), check.names = FALSE,
      stringsAsFactors = FALSE),
    replications = read.csv(
      file.path(path, "replications.csv"), check.names = FALSE,
      stringsAsFactors = FALSE),
    config = read.csv(
      file.path(path, "run_config.csv"), check.names = FALSE,
      stringsAsFactors = FALSE))
}
score_allocation <- read_score_arm("score-pivot-allocation")
score_allocation_reverse <- read_score_arm("score-pivot-allocation-reverse")
score_mixture <- read_score_arm("score-pivot-mixture")
score_power <- read_score_arm("score-pivot-power")

rejection_rate <- function(x, column, n_total = NULL,
                           regime = NULL) {
  if (!is.null(n_total)) x <- x[x$n_total == n_total, , drop = FALSE]
  if (!is.null(regime)) x <- x[x$regime == regime, , drop = FALSE]
  p <- x[[column]]
  mean(p[is.finite(p)] <= .05)
}
if (!is.null(score_sensitivity)) {
  sensitivity_rep <- score_sensitivity$replications
  sensitivity_size <- function(column, n) {
    rejection_rate(sensitivity_rep, column, n_total = n)
  }
}
if (!is.null(score_run)) {
  score_rep <- score_run$replications
  score_identity_error <- max(
    score_rep$permutation_rank_identity_error, na.rm = TRUE)
  score_control_size <- rejection_rate(
    score_rep, "p_permutation_score", regime = "exchangeable_normal")
  wald_control_size <- rejection_rate(
    score_rep, "p_permutation_wald", regime = "exchangeable_normal")
}
if (!is.null(score_allocation) && !is.null(score_allocation_reverse)) {
  allocation_arms <- list(
    `1:3` = score_allocation$replications,
    `3:1` = score_allocation_reverse$replications)
  allocation_size <- function(allocation, column, n) {
    rejection_rate(
      allocation_arms[[allocation]], column, n_total = n)
  }
  allocation_summary <- rbind(
    score_allocation$summary, score_allocation_reverse$summary)
  permutation_time_ratio <- with(
    allocation_summary,
    median_permutation_score_seconds / median_permutation_wald_seconds)
  allocation_identity_error <- max(
    allocation_summary$max_permutation_rank_identity_error, na.rm = TRUE)
}
if (!is.null(score_mixture)) {
  mixture_size <- function(column, n) {
    rejection_rate(score_mixture$replications, column, n_total = n)
  }
}
if (!is.null(score_power) && !is.null(score_allocation) &&
    !is.null(score_allocation_reverse)) {
  power_null <- rbind(
    score_allocation$replications,
    score_allocation_reverse$replications)
  power_rep <- score_power$replications
  power_methods <- c(
    p_permutation_score = "Permutation robust score",
    p_permutation_wald = "Permutation sandwich Wald",
    p_score_sb = "Effective score: SB",
    p_score_peba4 = "Effective score: pEBA4",
    p_score_all = "Effective score: numerical mixture")
  randomized_matched_power <- function(null_p, alternative_p, alpha = .05) {
    null_p <- null_p[is.finite(null_p)]
    alternative_p <- alternative_p[is.finite(alternative_p)]
    target <- alpha * length(null_p)
    critical <- sort(null_p)[ceiling(target)]
    n_below <- sum(null_p < critical)
    boundary_probability <-
      (target - n_below) / sum(null_p == critical)
    mean(alternative_p < critical) +
      boundary_probability * mean(alternative_p == critical)
  }
  matched_power_cell <- function(allocation, column, n, q) {
    keep_null <- power_null$allocation == allocation &
      power_null$n_total == n & power_null$q == q
    keep_power <- power_rep$allocation == allocation &
      power_rep$n_total == n & power_rep$q == q
    randomized_matched_power(
      power_null[[column]][keep_null],
      power_rep[[column]][keep_power])
  }
  matched_power <- function(allocation, column, n) {
    mean(vapply(
      c(4L, 8L), function(q) {
        matched_power_cell(allocation, column, n, q)
      }, numeric(1)))
  }
  overall_matched_power <- function(column) {
    mean(unlist(lapply(c("1:3", "3:1"), function(allocation) {
      vapply(c(120L, 240L, 480L, 960L), function(n) {
        matched_power(allocation, column, n)
      }, numeric(1))
    })))
  }
  power_columns <- names(power_methods)
  power_key <- interaction(
    power_rep$allocation, power_rep$n_total, power_rep$q, drop = TRUE)
  null_key <- interaction(
    power_null$allocation, power_null$n_total, power_null$q, drop = TRUE)
  power_cells <- split(power_rep[, power_columns], power_key)
  null_cells <- split(power_null[, power_columns], null_key)
  null_cells <- null_cells[names(power_cells)]
  bootstrap_matched_power <- function(null_cell, power_cell) {
    out <- vapply(seq_along(power_columns), function(j) {
      randomized_matched_power(null_cell[[j]], power_cell[[j]])
    }, numeric(1))
    names(out) <- power_columns
    out
  }
  one_power_bootstrap <- function() {
    cell_power <- mapply(function(null_cell, power_cell) {
      null_cell <- null_cell[
        sample.int(nrow(null_cell), replace = TRUE), , drop = FALSE]
      power_cell <- power_cell[
        sample.int(nrow(power_cell), replace = TRUE), , drop = FALSE]
      bootstrap_matched_power(null_cell, power_cell)
    }, null_cells, power_cells, SIMPLIFY = TRUE)
    rowMeans(cell_power)
  }
  set.seed(919)
  power_bootstrap <- t(replicate(2000L, one_power_bootstrap()))
  score_minus_comparator <- sweep(
    power_bootstrap[, -1L, drop = FALSE], 1L,
    power_bootstrap[, 1L], function(comparator, score) score - comparator)
  power_bootstrap_intervals <- apply(
    score_minus_comparator, 2L, stats::quantile, c(.025, .975))
}
```

## Question

Can the fully recomputed sandwich robust score serve as the pivotal statistic
for a group-label permutation test when the two groups are not exchangeable?
And does applying the ordinary-Hotelling reference to that score buy anything
inside the permutation test?

## Short answer

```{r score-short-answer, results='asis'}
if (is.null(score_run) || is.null(score_allocation) ||
    is.null(score_allocation_reverse) || is.null(score_mixture) ||
    is.null(score_power)) {
  cat("The score-pivot allocation, mixture, and power gates have not all run. ",
      "Execute their commands under **Reproduce**.\n")
} else {
  cat(sprintf(
    "The unequal-allocation gate reverses the balanced-grid verdict: the robust score is the better permutation pivot in this DGP. Pooled over $q=4,8$, score-permutation rejection was %.1f/%.1f/%.1f/%.1f%% for $N=120/240/480/960$ under $1{:}3$, and %.1f/%.1f/%.1f/%.1f%% after reversing the allocation to $3{:}1$. Wald permutation was orientation-sensitive: %.1f%% versus %.1f%% at $N=120$, converging only later. The score route cost %.1f%% more permutation time on average, so its win is calibration rather than speed. In the exchangeable-normal control, the score/Wald permutation sizes were %.1f%%/%.1f%%.",
    100 * allocation_size("1:3", "p_permutation_score", 120),
    100 * allocation_size("1:3", "p_permutation_score", 240),
    100 * allocation_size("1:3", "p_permutation_score", 480),
    100 * allocation_size("1:3", "p_permutation_score", 960),
    100 * allocation_size("3:1", "p_permutation_score", 120),
    100 * allocation_size("3:1", "p_permutation_score", 240),
    100 * allocation_size("3:1", "p_permutation_score", 480),
    100 * allocation_size("3:1", "p_permutation_score", 960),
    100 * allocation_size("1:3", "p_permutation_wald", 120),
    100 * allocation_size("3:1", "p_permutation_wald", 120),
    100 * mean(permutation_time_ratio - 1),
    100 * score_control_size, 100 * wald_control_size))
  cat(sprintf(
    " Under the common-sign local loading alternative $\\delta_N=5/\\sqrt{N}$, empirically 5%%-size-matched power averaged %.1f%% for score permutation, %.1f%% for SB, %.1f%% for pEBA4, %.1f%% for the numerical mixture, and %.1f%% for Wald permutation. The score's roughly one-point edge over the analytical score corrections was not resolved by a stratified bootstrap; its pooled %.1f-point edge over Wald was, but that average hides a strong allocation-direction reversal.",
    100 * overall_matched_power("p_permutation_score"),
    100 * overall_matched_power("p_score_sb"),
    100 * overall_matched_power("p_score_peba4"),
    100 * overall_matched_power("p_score_all"),
    100 * overall_matched_power("p_permutation_wald"),
    100 * (overall_matched_power("p_permutation_score") -
      overall_matched_power("p_permutation_wald"))))
  cat(sprintf(
    " The numerical limiting-mixture tail converged very slowly—%.1f%% at $N=480$ and %.1f%% even at $N=3840$—while SB reached %.1f%% and pEBA4 %.1f%% at the latter endpoint. MV and the numerical mixture made virtually identical decisions, so this is primarily slow finite-sample convergence to the mixture law, not MV tail-approximation error. “Permutation Hotelling” remained exactly the score permutation (maximum p-value difference %.1g).\n",
    100 * mixture_size("p_score_all", 480),
    100 * mixture_size("p_score_all", 3840),
    100 * mixture_size("p_score_sb", 3840),
    100 * mixture_size("p_score_peba4", 3840),
    allocation_identity_error))
}
```

## Evidence

```{r score-evidence}
if (!is.null(score_allocation) && !is.null(score_allocation_reverse)) {
  method_labels <- c(
    p_permutation_score = "Permutation robust score",
    p_permutation_wald = "Permutation sandwich Wald",
    p_score_sb = "Effective score: SB",
    p_score_peba4 = "Effective score: pEBA4",
    p_score_all = "Effective score: numerical mixture",
    p_wald_chisq = "Sandwich Wald: chi-square")
  score_table <- do.call(rbind, lapply(names(allocation_arms), function(a) {
    tab <- data.frame(
      Allocation = a, Method = unname(method_labels),
      stringsAsFactors = FALSE)
    for (n in c(120L, 240L, 480L, 960L)) {
      tab[[paste0("N=", n)]] <- round(
        100 * vapply(names(method_labels), function(method) {
          allocation_size(a, method, n)
        }, numeric(1)), 1)
    }
    tab
  }))
  knitr::kable(
    score_table, row.names = FALSE,
    caption = paste0(
      "Heterogeneous Vale–Maurelli null rejection at 5%, pooled equally over ",
      "q = 4, 8; 500 replications per q and 99 label permutations."))
}
```

Across the two allocation directions, the score-permutation size was
4.9%, 5.0%, 5.0%, and 5.5% over the four sample sizes. All 8,000
replications and all 792,000 score permutations were usable; 791,998 of
792,000 Wald permutations were usable.
The balanced sensitivity had looked less distinctive; unequal allocation is
the diagnostic setting because it exposes the Wald permutation's directional
finite-sample error.

### Local power after size matching

```{r power-evidence}
if (!is.null(score_power) && !is.null(score_allocation) &&
    !is.null(score_allocation_reverse)) {
  power_table <- do.call(rbind, lapply(c("1:3", "3:1"), function(allocation) {
    tab <- data.frame(
      Allocation = allocation, Method = unname(power_methods),
      stringsAsFactors = FALSE)
    for (n in c(120L, 240L, 480L, 960L)) {
      tab[[paste0("N=", n)]] <- round(
        100 * vapply(names(power_methods), function(method) {
          matched_power(allocation, method, n)
        }, numeric(1)), 1)
    }
    tab
  }))
  knitr::kable(
    power_table, row.names = FALSE,
    caption = paste0(
      "Empirically 5%-size-matched power under the local loading shift ",
      "$\\delta_N = 5/\\sqrt{N}$, pooled equally over $q = 4, 8$; 500 null and ",
      "500 alternative replications per q-cell."))
}
```

Size matching uses the empirical null distribution separately for every
method, allocation, $N$, and $q$. If the 5% empirical quantile is tied—as it
often is for the 99-permutation p-values—the boundary is randomized so the
calibration sample rejects exactly 5% in expectation. This prevents liberal or
conservative null behavior from masquerading as power.

The score permutation is practically tied with SB, pEBA4, and the numerical
mixture after that correction. A stratified two-stage bootstrap that resampled
both the 500 null and 500 alternative draws in every cell put the score-minus-
comparator intervals at $-0.6$ to $3.0$ percentage points for SB, $-0.7$ to
$3.0$ for pEBA4, and $-0.4$ to $3.3$ for the numerical mixture. Its pooled
advantage over Wald was 4.3 points (95% interval 2.6 to 6.3), but the table
shows why this is not a universal efficiency claim: Wald wins throughout the
$1{:}3$ arm and loses much more sharply throughout the $3{:}1$ mirror.

The bounded conclusion is therefore to retain the score permutation as a
calibration method, not to advertise a power gain. Its attractive numerator and
pivotal randomization reference avoid dependence on a slowly attained mixture
law; SB and pEBA4 remain simpler, competitive analytical defaults in this DGP.

### Slow convergence to the mixture law

```{r mixture-evidence}
if (!is.null(score_mixture)) {
  mixture_labels <- c(
    p_score_chisq = "Robust score: chi-square",
    p_score_hotelling = "Robust score: Hotelling",
    p_score_sb = "Effective score: SB",
    p_score_mv = "Effective score: MV",
    p_score_peba4 = "Effective score: pEBA4",
    p_score_all = "Effective score: numerical mixture")
  mixture_table <- data.frame(Method = unname(mixture_labels))
  for (n in c(480L, 960L, 1920L, 3840L)) {
    mixture_table[[paste0("N=", n)]] <- round(
      100 * vapply(names(mixture_labels), mixture_size,
                   numeric(1), n = n), 1)
  }
  knitr::kable(
    mixture_table, row.names = FALSE,
    caption = paste0(
      "Analytical null rejection at 5% under 1:3 heterogeneous VM, pooled ",
      "equally over q = 4, 8; 2,000 replications per q."))
}
```

MV and the numerical mixture disagreed on at most 0.05% of decisions at any
sample size. The numerical tail therefore does not rescue the finite-sample
statistic merely by avoiding a mixture approximation. Its rejection rate
increases from 2.9% to 4.3% as the limiting spectrum stabilizes, while the
simpler SB and pEBA4 corrections happen to compensate most of that slow
convergence.

### The statistic and the rank identity

Let $z_i(\widehat\theta_0)$ be the nuisance-orthogonal released-score row,
$u=\sum_i z_i$, and

$$
  B_r=\sum_{i=1}^n z_i z_i^\top,\qquad
  Q_r=u^\top B_r^{-1}u.
$$

The ordinary robust score compares the asymptotically pivotal $Q_r$ with
$\chi_q^2$. Writing $Q_c$ for the same numerator studentized by the
unbiased *centered* score covariance gives the exact rank-one identity

$$
  Q_r=\frac{nQ_c}{n-1+Q_c},\qquad
  F=\frac{n-q}{q(n-1)}Q_c
   =\frac{(n-q)Q_r}{q(n-Q_r)}.
$$

The map from $Q_r$ to its Hotelling tail is strictly decreasing for fixed
$(n,q)$. Consequently

$$
 \mathbf 1\{Q_r^\pi\ge Q_r\}
 =
 \mathbf 1\{p_H(Q_r^\pi)\le p_H(Q_r)\},
$$

so recomputing either representation after each label permutation yields
exactly the same Monte Carlo p-value. Hotelling remains a distinct
*analytical* finite-sample reference, but cannot improve a permutation test
based on ranks.

### Alternatives compared

- **Robust-score asymptotics:** raw-OPG sandwich $Q_r$ against
  $\chi_q^2$, or the algebraically centered version against ordinary
  Hotelling $F_{q,n-q}$.
- **Non-robust effective-score corrections:** mean-scaled
  Satorra--Bentler, mean-and-variance adjusted, pEBA4, and the numerical
  weighted-$\chi^2$ mixture tail. These are score tests, not robust LRT
  mixture limits.
- **Randomization:** fully recomputed robust-score label permutation.
  “Permutation Hotelling” is recorded only as an asserted rank-identity
  sentinel.
- **Wald comparators:** analytical sandwich Wald and the existing fully
  recomputed sandwich-Wald label permutation. The Wald permutation was
  direction-sensitive under unequal allocation; the score permutation was the
  stable method in this focused null probe.

## Original Wald-arm status

This render uses `r profiles`.

```{r smoke-warning, results='asis'}
if (is_smoke) cat(
  "**Smoke results are wiring and timing evidence only.** With nine random ",
  "permutations the smallest attainable plus-one p-value is 0.10, so the ",
  "permutation rejection columns cannot estimate rejection rates at 0.05.\n")
```

## Original Wald arm

Does a fully refitted Wald statistic, studentized with a freshly estimated
empirical sandwich covariance after every group-label permutation, provide
better calibrated measurement-invariance inference when group observations are
not exchangeable? The design targets Chung and Romano's asymptotically pivotal
case: raw exchangeability is unnecessary if the recomputed statistic has a
common limiting null law.

The comparison includes ordinary and robust/scaled likelihood-ratio tests,
model-based multivariate Wald tests, sandwich Wald tests, the semTools-style
ordinary permutation likelihood-ratio test, the studentized permutation Wald
test, and a raw-contrast permutation negative control. In the unbalanced study,
the model-based Wald is Chen and Chao's MWT restriction test. The ordinary
permutation LRT reproduces the established continuous-data comparison
[@jorgensen2018]; robust/scaled nested tests remain analytical comparators and
are not recalculated inside every permutation.

## Executed design

```{r design-table}
design_rows <- do.call(rbind, lapply(runs, function(run) {
  m <- run$manifest
  data.frame(
    Study = run$study, Profile = run$profile,
    Cells = nrow(m), Steps = paste(unique(m$step), collapse = ", "),
    Generators = paste(unique(m$generator), collapse = ", "),
    `N range` = paste(range(m$n_total), collapse = "--"),
    Allocations = paste(unique(m$allocation), collapse = ", "),
    check.names = FALSE)
}))
knitr::kable(design_rows, row.names = FALSE,
             caption = "Manifests used by this report render.")
```

The reference-law paper grid contains two correlated factors, `p = 8/16`, total
`N = 220/440/1760`, balanced and `1:3` allocation, normal and Pearson IG
nonnormal data, and ordinal-probit integer scores with explicit category and
threshold-shape factors. Its heterogeneous null retains equal measurement
parameters while factor means and factor/residual covariance differ. Separate
exchangeable controls make both group populations identical.

The unbalanced paper grid reproduces Chen and Chao's six-indicator population:
loading 0.6 and intercept 0 for each item, unit factor and residual variances,
group A `n = 200`, and group B from 400 through 3000. Metric/scalar violations
affect items `{2,4}` or `{2,3,4,5}` in group A using their small, large, mixed,
and non-uniform changes. Matched-balanced cells retain the same total N. Strict
invariance is the proportional residual-variance extension.

## Timing and availability

```{r timing-table}
timing <- aggregate(
  cbind(median_simulation_seconds, median_observed_seconds,
        median_permutation_seconds, median_total_seconds, failure_rate) ~
    study + generator + step,
  data = all_summary, FUN = function(x) median(x, na.rm = TRUE))
names(timing) <- c("Study", "Generator", "Step", "Simulation s", "Observed s",
                   "Permutation s", "Total s", "Failure rate")
num <- vapply(timing, is.numeric, logical(1))
timing[num] <- lapply(timing[num], function(x) round(x, 3))
knitr::kable(timing, row.names = FALSE,
             caption = "Median per-replication stage timings across selected cells.")
```

Invalid permutation fits are retained separately for the H1-based Wald and
raw-contrast statistics and the H0/H1 permutation LRT. `summary.csv` reports
both usable-permutation fractions and method-specific usable replication
counts; any observed-fit errors are also written to `failures.csv`.

## Rejection summaries

```{r rejection-table}
cols <- c("study", "cell_id", "p", "n1", "n2", "generator", "step", "truth",
          "null_kind", "pattern", "affected_fraction", "reps", "observed_ok",
          "reject_lrt", "reject_robust_lrt", "reject_model_wald",
          "reject_sandwich_wald", "reject_permutation_studentized",
          "reject_permutation_lrt", "reject_permutation_raw",
          "median_valid_permutation_fraction",
          "median_valid_permutation_lrt_fraction")
tab <- all_summary[, intersect(cols, names(all_summary)), drop = FALSE]
names(tab) <- gsub("_", " ", names(tab), fixed = TRUE)
numeric <- vapply(tab, is.numeric, logical(1))
tab[numeric] <- lapply(tab[numeric], function(x) round(x, 3))
knitr::kable(tab, row.names = FALSE,
             caption = "Cell-level rejection and availability summaries.")
```

Configural is fitted only as the metric test's internal H1. The reported
outcomes are metric versus configural, scalar versus the identified metric mean
chart, and strict versus scalar.

## Interpretation boundaries

- Reference alternatives use explicit step-specific deltas. The defaults are
  pilot starting values; paper claims require replacing them with departures
  calibrated to 50% studentized-permutation power in the normal balanced
  `N = 440` reference cell.
- Ordinal-probit responses are analyzed as ordinary numeric codes by continuous
  ML. This does not claim validity for a shuffled DWLS/polychoric statistic; that
  is a separate problem. Because discretization is nonlinear, latent-response
  invariance need not imply exact invariance of the continuous-ML pseudo-
  parameters, especially with asymmetric thresholds; that drift is part of the
  deliberately misspecified applied workflow and must be diagnosed separately
  from reference-law calibration.
- The asymptotic argument requires regular estimation under the pooled mixture.
  Exchangeable controls diagnose implementation errors, while the heterogeneous
  nulls test the scientifically relevant departure from raw exchangeability.
- The score-pivot extension covers one one-factor VM null and one aligned,
  common-sign local loading alternative. The size-matched result supports its
  calibration rationale but does not establish a power advantage or a broad
  method ranking. Sparse, mixed-sign, ordinal, and model-misspecified
  alternatives remain outside this bounded gate.
- The ordinary permutation LRT is intentionally not studentized. It is exact in
  the exchangeable controls and is the literature baseline, but it does not
  inherit the Chung--Romano guarantee under heterogeneous nuisance
  distributions. That contrast with the sandwich-studentized permutation Wald
  is a principal experiment outcome.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript scripts/run_score_pivot.R --smoke
Rscript scripts/run_score_pivot.R --probe --cores 10
Rscript scripts/run_score_pivot.R --sensitivity \
  --regimes heterogeneous_vm --n-total 60,240 --cores 10
Rscript scripts/run_score_pivot.R --allocation-gate --cores 10
Rscript scripts/run_score_pivot.R --allocation-reverse --cores 10
Rscript scripts/run_score_pivot.R --mixture-gate --cores 10
Rscript scripts/run_score_pivot.R --power-gate --local-strength 5 --cores 10
Rscript calibrate_alternatives.R --smoke
Rscript run_experiment.R --study reference --modal --cores 10
Rscript run_experiment.R --study imbalance --modal --cores 10
Rscript run_experiment.R --sensitivity --cores 10
quarto render report.qmd
```
