---
title: "Can a studentized permutation Wald test calibrate metric invariance?"
bibliography: references.bib
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
summary_path <- file.path("results", "summary.csv")
if (!file.exists(summary_path)) {
  stop("No results. Run:\n  Rscript run_experiment.R --smoke\nthen:\n  Rscript run_experiment.R --full --reps 1000 --permutations 499")
}
results <- read.csv(summary_path, check.names = FALSE, stringsAsFactors = FALSE)
fmt <- function(x) formatC(x, format = "f", digits = 3)
```

## Target Paper Design

The current numerical output below is the metric-invariance feasibility gate.
The paper study expands it into two complementary continuous-data simulations.
Study 1 follows the multi-group condition spine of Brace and Savalei
[@bracesavalei2017]. Study 2 directly compares the proposed test with the
unbalanced-data multivariate Wald design of Chen and Chao [@chenchao2026].

| Study | Population and allocation | Data conditions | Reported tests |
|---|---|---|---|
| 1. Reference-law study | Two groups, two correlated factors, `p = 8, 16`; total `N = 220, 440, 1760`; `1:1` and `1:3` allocation | Normal and Pearson IG `(2, 7)` / `(3, 21)`; ordinal-probit category scores analyzed by robust continuous ML | Metric, scalar, and strict LRT/Wald/permutation-Wald tests |
| 2. Unbalanced-data comparison | Chen--Chao one-factor, six-indicator model; group A `n = 200`, group B `n = 400, 600, 800, 1000, 1400, 2000, 3000`, plus matched balanced samples | Metric/scalar affected-item proportions `1/3` or `2/3`; small, large, mixed, and non-uniform effects; strict residual-variance extension | Chen--Chao multivariate Wald, ordinary/sandwich LRT and Wald, studentized permutation Wald |

The category-score arm deliberately treats ordered categories as continuous
numeric observations. This is not a polychoric or DWLS analysis. Its category
count and threshold-symmetry factors follow the conditions under which robust
continuous ML is, or is not, defensible in Rhemtulla, Brosseau-Liard, and
Savalei [@rhemtulla2012]. Robust continuous-ML inference follows
Maydeu-Olivares [@maydeuolivares2017].

All three reported invariance outcomes are adjacent comparisons: metric versus
configural, scalar versus metric, and strict versus scalar. The configural model
is fitted internally as the metric reference; this is an inference study rather
than a global-fit study.

## Question

Can a raw group-label permutation distribution of a fully recomputed,
studentized Wald statistic control Type-I error for continuous metric invariance
when group sample sizes, factor covariances, residual variances, and CFA size
vary? The comparison is against ordinary and robust likelihood-ratio and Wald
tests.

## Short answer

This feasibility experiment is intentionally a gate rather than a finished
method claim. The `invariant` rows show whether the studentized permutation
test stays near 0.05 under the heterogeneous null; the `loading_violation` rows
show whether that calibration retains useful power. The ordinary raw-contrast
permutation is included as a negative control: a gap between it and the
studentized column is the finite-sample behavior the Chung--Romano argument
predicts.

The loading-violation alternative is calibrated to comparable sandwich-Wald
noncentrality in the normal `n=(250,250)` reference cell, using raw loading
shifts of 0.18--0.21 according to CFA size and heterogeneity. The raw and
implied standardized shifts are both recorded in the results; equal
standardized shifts were rejected because they created radically unequal power
for the raw-loading invariance test.

## Evidence

Rejection rates at the configured alpha. The two permutation methods refit the
configural CFA and recompute their statistic after every relabeling; they do
not reuse the observed covariance estimate.

```{r table}
cols <- c("model", "heterogeneity", "distribution", "scenario", "n1", "n2",
          "loading_shift_raw", "loading_shift_standardized", "reps", "converged",
          "reject_lrt", "reject_robust_lrt", "reject_model_wald",
          "reject_sandwich_wald", "reject_permutation_studentized",
          "reject_permutation_raw", "median_valid_permutations")
tab <- results[, intersect(cols, names(results)), drop = FALSE]
labels <- c("Model", "Heterogeneity", "Distribution", "Scenario", "N1", "N2",
            "Raw loading shift", "Standardized loading shift", "Replications",
            "Converged", "LRT", "Robust LRT", "Model Wald",
            "Sandwich Wald", "Studentized permutation", "Raw permutation",
            "Median valid permutations")
names(labels) <- cols
names(tab) <- unname(labels[names(tab)])
numeric <- vapply(tab, is.numeric, logical(1))
tab[numeric] <- lapply(tab[numeric], function(x) round(x, 3))
knitr::kable(tab, row.names = FALSE,
             caption = "Metric-invariance rejection rates and permutation availability.")
```

## Caveats

- The Chung--Romano argument requires the estimator and its studentization to
  remain asymptotically regular under the pooled mixture, not merely under the
  two source groups. This experiment checks one- and three-factor CFA
  populations at standard and stronger group heterogeneity.
- The `exchangeable` scenario is a finite-sample control: both group
  distributions are identical, so a random label permutation is valid even for
  the raw contrast statistic. It distinguishes a resampling implementation bug
  from a failure of asymptotic robustness under heterogeneous populations.
- The ordinal arm is deliberately absent. A mixture of ordinal-probit groups is
  generally not ordinal-probit, so it needs a separate validity result or a
  different resampling scheme.
- Smoke output is a wiring and runtime check only. The full grid should use at
  least 1,000 replications before interpreting a 0.05 rejection rate.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --full --reps 1000 --permutations 499
quarto render report.qmd
```
