---
title: "Corrected corpus: optimizer settings and recovery"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
r <- function(name) {
  p <- file.path("results","full",paste0(name,".csv"))
  if(!file.exists(p)) stop("Missing results: run run_experiment.R and R/summarize.R first")
  read.csv(p,stringsAsFactors=FALSE)
}
s <- r("summary"); d <- r("comparison"); jobs <- r("jobs"); exclusions <- r("exclusions")
paired <- r("paired"); domain <- r("domain"); meta <- r("metadata")
retries <- r("retry_jobs")
primary <- c("lbfgs_default","port_default","slsqp_default")
labels <- c(lbfgs_default="L-BFGS",port_default="PORT",slsqp_default="SLSQP")
ps <- subset(s,arm %in% primary)
count <- function(est,arm,col) s[s$estimator==est & s$arm==arm,col][1]
```

## Question

On the corrected textbook corpus, which ordinary optimizers recover a good
solution from our current starts? Do the historical Newsom GLS failures still
justify changing the evaluator?

## Short answer

**Settings alone do not repair L-BFGS's early domain failures.** PORT recovers
cases that stop almost immediately under L-BFGS, even after increasing its
budget or tightening tolerances. In ML, default L-BFGS returns an accepted fit
matching the best observed objective on `r count("ML","lbfgs_default","accepted_match")`
of `r count("ML","lbfgs_default","attempted")` completed attempts; PORT does so on
`r count("ML","port_default","accepted_match")` of
`r count("ML","port_default","attempted")`. For GLS, tighter L-BFGS controls
raise accepted, matching fits from `r count("GLS","lbfgs_default","accepted_match")`
to `r count("GLS","lbfgs_tight","accepted_match")`;
PORT reaches `r count("GLS","port_default","accepted_match")`.
Missing attempts after job timeouts
are reported separately and prevent an unconditional ranking.

**The two historical Newsom GLS failures do not reproduce in these reruns.**
L-BFGS, PORT and SLSQP agree on their objective values, with small terminal
gradients. This run supplies no affirmative evidence for the old cancellation
explanation. It does not certify evaluator accuracy on every model.

**Some failures belong to the starting point, and success flags remain
insufficient.** The Little phantom model reaches a much worse stationary point from
ordinary starts under all three main optimizers. The curvature audit rejects
that point; a verified-solution restart recovers the lower objective and passes. A separate scalar probe shows that bounded SLSQP can report
success far from the known optimum; its independent terminal audit rejects it.
No default is changed on the basis of this study.

## Evidence

### Corpus recovery

“Accepted + matched” requires the library's authoritative acceptance and an
objective within `1e-6 * (1 + abs(best))` of the best objective among accepted fits and converged references.
That threshold is for this comparison; it is neither a production stopping rule
nor a global-optimum certificate.

```{r}
policies <- c(lbfgs_default="L-BFGS, default",lbfgs_budget="L-BFGS, larger budget",
  lbfgs_tight="L-BFGS, tighter",lbfgs_memory50="L-BFGS, memory 50",
  port_default="PORT, default",port_tight="PORT, tighter",
  slsqp_default="SLSQP, default",slsqp_tight="SLSQP, tighter",
  lbfgs_slsqp="L-BFGS → SLSQP fallback")
z <- data.frame(Policy=unname(policies))
for(est in c("ML","GLS")) z[[est]] <- vapply(names(policies),function(a)
  sprintf("%d / %d",count(est,a,"accepted_match"),count(est,a,"attempted")),character(1))
knitr::kable(z,col.names=c("Policy","ML Accepted + Matched / Attempts","GLS Accepted + Matched / Attempts"))
```

On paired cases, default PORT gains `r sum(paired$rescued[paired$estimator=="ML" & paired$arm=="port_default"])`
ML successes and loses `r sum(paired$lost[paired$estimator=="ML" & paired$arm=="port_default"])`
relative to default L-BFGS. For GLS it gains
`r sum(paired$rescued[paired$estimator=="GLS" & paired$arm=="port_default"])`
and loses `r sum(paired$lost[paired$estimator=="GLS" & paired$arm=="port_default"])`;
the loss is Kline's Worland structural-regression Step 2b. PORT therefore merits
an ML recovery-policy investigation, not an unconditional replacement across estimators.

The complete nine-policy comparison and verified-reference restarts are in
`results/full/comparison.csv`. Backend stopping status, acceptance, covariance
admissibility and objective quality remain separate outcomes. Settings do not
have the same meaning across algorithms; tighter numerical controls are not
in themselves a guarantee of a more accurate answer.

### Previously flagged GLS cases

```{r}
z <- subset(d, estimator=="GLS" & case %in% c("newsom_2015_ex5_4","newsom_2015_ex5_4c") & arm %in% primary)
z$case <- sub("newsom_2015_","",z$case); z$arm <- unname(labels[z$arm])
z$gradient <- formatC(z$gradient,format="e",digits=1)
knitr::kable(z[,c("case","arm","f","gradient","converged")],digits=9,
             col.names=c("Case","Optimizer","Objective","Terminal Gradient","Accepted"))
```

The original cancellation explanation came from older inputs and numerical
paths. A solver's noisy-gradient status can motivate investigation but cannot
establish where numerical error originates. A persistent discrepancy would
require an independent same-point objective and derivative comparison.

### Starting-point dependence

```{r}
z <- subset(d, estimator=="ML" & case=="little_2013_ch3_fig_3_11_longitudinal_cfa_phantom" &
              arm %in% c(primary,"lbfgs_reference_start"))
z$arm <- c(labels,lbfgs_reference_start="L-BFGS, verified start")[z$arm]
knitr::kable(z[,c("arm","f","converged","objective_gap")],digits=6,
             col.names=c("Policy","Objective","Accepted","Gap to Best Observed"))
```

The poor phantom endpoint is rejected for nonpositive curvature, despite its
small first-order gradient. It is not certified as a local minimum. The reference
restart is diagnostic, not an ordinary-start reliability result.
The old “12 failures” came from a smaller lower-level test with simple starts
and generic controls. This larger current-interface run is a new baseline,
not a direct rerun of that exact test protocol.

### Unit-scale domain recovery

A scalar variance likelihood has a known optimum. The problem is unchanged;
only units, optimizer coordinates and an explicit nonnegative bound vary.
These rows use a variance of one millionth and a start ten times the optimum.

```{r}
z <- subset(domain,variance==1e-6 & arm %in% c("lbfgs","port","slsqp"))
z$condition <- ifelse(z$normalized,"Normalized coordinates",
                       ifelse(z$bounded,"Raw coordinates, bounded","Raw coordinates, unbounded"))
z <- z[!duplicated(paste(z$condition,z$arm)),]
z$accurate <- is.finite(z$relative_error) & z$relative_error<1e-6
knitr::kable(z[,c("condition","arm","returned","stationary","accurate")],
             col.names=c("Condition","Optimizer","Returned","Audit Passed","Known Optimum Recovered"))
```

The larger-budget and tighter L-BFGS arms do not remove the small-scale failure.
Normalized coordinates recover the optimum in this probe. That is evidence for
work on scale-aware domain recovery, not a validated normalization policy for
arbitrary constrained SEMs.

## Caveats

The run scheduled `r nrow(jobs)` case/estimator jobs;
`r sum(jobs$exit != 0)` ended with a process failure or wall-clock timeout.
The initial cap was 120 seconds per case/estimator across all arms.
The `r nrow(retries)` omitted ordinary arms then received separate 120-second
attempts; `r sum(retries$exit != 0)` remained incomplete. These caps include
input preparation and reference fitting, so a timeout does not isolate optimizer
cost or establish numerical failure.
There were `r nrow(exclusions)` preparation exclusions. Categorical, FIML-only,
composite and multilevel models are outside this comparison; unsupported input
conventions and singular sample covariances are recorded rather than silently
changed. The included inputs comprise 304 cases for each estimator. This is an
ML/GLS experiment, not coverage of every supported estimator. `jobs.csv`, `exclusions.csv` and `input_contracts.csv` retain the details.
Completed-attempt denominators differ when a timeout interrupts the policy sequence.

Current start construction is held fixed across ordinary arms within each
case and estimator. ML retains its current coordinate-scaling policy, which
is not identical for every backend. The study compares available policies,
not isolated algorithms under identical preconditioning. Syntax hints are
honored; separate corpus solution-start files are reserved for references.

Only accepted fits and converged references define the comparison target;
failed or infeasible endpoints cannot lower it. The reference parameters are
evaluated under the same original objective.
Reference evaluation gradients live in full coordinates, whereas fitted
optimizer gradients may live in equality-reduced coordinates; they must not
be compared as though they were the same stationarity measure.

The engine is pinned to the installed build recorded in `metadata.csv` and
`library_hashes.csv`; the working repository may subsequently advance. Corpus
input hashes are checked before and after the run. Single-pass, concurrent
wall times are descriptive and order-dependent, not a controlled speed ranking.

The current default remains L-BFGS. A policy change needs recovery gains without
unexplained losses, preserved model/objective semantics and acceptable cost.
Reopen this experiment after a domain-recovery, scaling or start-policy change;
reopen the GLS accuracy diagnosis only with a current reproducible discrepancy.

## Reproduce

Install the intended `magmaanlab` build and provide the optional corrected
textbook corpus. Run from this experiment's directory:

```sh
export OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --results results/smoke
Rscript run_experiment.R --results results/full --workers 3 --timeout 120 \
  --engine-source COMMIT_USED_TO_BUILD_THE_PACKAGE
Rscript R/complete_timeouts.R ../../../external/textbook-corpus results/full
Rscript R/validate_inputs.R ../../../external/textbook-corpus results/full
Rscript R/summarize.R results/full
Rscript R/inspect_endpoints.R ../../../external/textbook-corpus results/full
cmake -S . -B results/build -DEIGEN_INCLUDE_DIR=/path/to/eigen
cmake --build results/build
results/build/domain_probe > results/full/domain.csv
quarto render report.qmd
```

`--cases` selects named corpus cases. Each case/estimator has its own process,
wall-clock cap, log and checkpointed attempts. The timeout follow-up preserves
completed arms and gives each omitted ordinary arm its own capped process.
Results and rendered output are
local, ignored artifacts. The report performs no fitting.
