---
title: "Latent Metric Identification Experiment"
format: html
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(
  getwd(),
  dirname(input_path),
  dirname(getwd())
))
result_candidates <- unique(c(
  file.path(candidate_dirs, "results"),
  candidate_dirs
))
hits <- result_candidates[file.exists(file.path(result_candidates, "fits.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
fits_path <- file.path(results_dir, "fits.csv")
pairs_path <- file.path(results_dir, "pairs.csv")
summary_path <- file.path(results_dir, "summary.csv")

if (!file.exists(fits_path) || !file.exists(pairs_path) || !file.exists(summary_path)) {
  stop("Missing experiment CSVs. Run: Rscript run_experiment.R")
}

fits <- utils::read.csv(fits_path, check.names = FALSE)
pairs <- utils::read.csv(pairs_path, check.names = FALSE)
summary <- utils::read.csv(summary_path, check.names = FALSE)

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), "", formatC(x, format = "g", digits = digits))
}
```

This experiment asks whether fixed latent-variance identification (`std_lv`)
changes numerical estimation behavior relative to marker-loading
identification, holding the implied model constant. Raw parameter estimates are
not compared directly because the two parameterizations use different latent
metrics.

```{r}
case_count <- length(unique(fits$case_id))
attempts <- nrow(fits)
ok <- sum(fits$status == "OK", na.rm = TRUE)
skips <- sum(fits$status == "SKIP", na.rm = TRUE)
errors <- sum(fits$status == "ERROR", na.rm = TRUE)
overview <- data.frame(
  cases = case_count,
  fit_rows = attempts,
  ok = ok,
  skipped = skips,
  errors = errors,
  pair_rows = nrow(pairs)
)
knitr::kable(overview)
```

## Summary

```{r}
summary_display <- summary
numeric_cols <- vapply(summary_display, is.numeric, logical(1))
summary_display[numeric_cols] <- lapply(summary_display[numeric_cols], fmt_num)
knitr::kable(summary_display)
```

## Pairwise Timing

The ratio below is `std_lv / marker`; values below 1 favor fixed latent
variance on elapsed time.

```{r}
complete_pairs <- pairs[pairs$status == "OK" &
                          is.finite(pairs$elapsed_ratio_std_lv_over_marker), ]
if (nrow(complete_pairs)) {
  boxplot(elapsed_ratio_std_lv_over_marker ~ backend,
          data = complete_pairs,
          ylab = "elapsed ratio: std_lv / marker",
          xlab = "backend",
          log = "y")
  abline(h = 1, lty = 2, col = "gray40")
} else {
  plot.new()
  text(0.5, 0.5, "No complete pair rows")
}
```

## Largest Differences

These rows highlight the largest implied-moment or objective differences. Large
values mean the two runs did not land at the same numerical solution and should
be inspected before interpreting speed or gradient differences.

```{r}
if (nrow(pairs)) {
  ranked <- pairs[order(-pairs$material_mismatch,
                        -pairs$implied_sigma_max_abs_diff,
                        -pairs$abs_fmin_diff), ]
  keep <- c("case_id", "backend", "replicate", "status",
            "marker_optimizer_status", "std_lv_optimizer_status",
            "abs_fmin_diff", "implied_sigma_max_abs_diff",
            "implied_mu_max_abs_diff", "elapsed_ratio_std_lv_over_marker",
            "grad_norm_ratio_std_lv_over_marker")
  top <- head(ranked[keep], 20L)
  numeric_cols <- vapply(top, is.numeric, logical(1))
  top[numeric_cols] <- lapply(top[numeric_cols], fmt_num)
  knitr::kable(top)
}
```

## Fit Errors And Skips

```{r}
problem_rows <- fits[fits$status != "OK", ]
if (nrow(problem_rows)) {
  keep <- c("case_id", "backend", "parameterization", "replicate", "status", "error")
  knitr::kable(head(problem_rows[keep], 40L))
} else {
  knitr::kable(data.frame(note = "No skipped or failed fit rows."))
}
```
