---
title: "Can score flipping rescue small-sample FIML nested tests?"
subtitle: "A focused probe in the published FIML--FMG measurement-invariance design"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages(library(ggplot2))
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
read_result <- function(name) read.csv(
  file.path(script_dir, "results", paste0(name, ".csv")),
  stringsAsFactors = FALSE, check.names = FALSE)
required <- c("design", "method_summary", "paired_summary", "matched_power",
              "timing_summary", "dgp_timing", "failures", "metadata")
paths <- file.path(script_dir, "results", paste0(required, ".csv"))
if (!all(file.exists(paths))) stop("run run_experiment.R before rendering")
objects <- lapply(required, read_result); names(objects) <- required
list2env(objects, environment())
meta <- setNames(metadata$value, metadata$key)
reps <- as.integer(meta[["reps"]]); flips <- as.integer(meta[["flips"]])

method_order <- c(
  "flip_basic", "flip_effective", "flip_standardized",
  "score_sb", "score_peba4", "score_sandwich", "score_all",
  "nested_sb_ml", "nested_peba4_ml", "nested_all_ml")
method_label <- c(
  flip_basic = "flip: basic",
  flip_effective = "flip: nuisance-effective",
  flip_standardized = "flip: standardized",
  score_sb = "score: SB",
  score_peba4 = "score: pEBA4",
  score_sandwich = "score: direct sandwich",
  score_all = "score: exact mixture",
  nested_sb_ml = "nested LR: SB",
  nested_peba4_ml = "nested LR: pEBA4",
  nested_all_ml = "nested LR: exact mixture")
label_method <- function(x) unname(method_label[x])

null <- subset(method_summary, truth == "null")
cell_mean <- function(method, distribution = NULL) {
  x <- null[null$method == method, , drop = FALSE]
  if (!is.null(distribution)) x <- x[x$distribution == distribution, , drop = FALSE]
  mean(x$rejection_rate, na.rm = TRUE)
}
weighted_cell_mean <- function(x, value) {
  ok <- is.finite(x[[value]]) & is.finite(x$n)
  stats::weighted.mean(x[[value]][ok], x$n[ok])
}

basic_rate <- cell_mean("flip_basic")
effective_rate <- cell_mean("flip_effective")
standardized_rate <- cell_mean("flip_standardized")
effective_normal <- cell_mean("flip_effective", "normal")
effective_pl <- cell_mean("flip_effective", "pl")
standardization_changes <- with(
  subset(paired_summary, truth == "null"),
  round(sum(n * disagreement_rate, na.rm = TRUE)))
fit_failures <- sum(!failures$fit_ok)
nested_only_failures <- sum(failures$fit_ok & failures$flip_ok & !failures$nested_ok)
flip_failures_after_fit <- sum(failures$fit_ok & !failures$flip_ok)
```

## Question

In the two-group, six-indicator FIML measurement-invariance design used by the
FIML--FMG work, can individual-score sign flipping repair the familiar
small-sample conservatism without making an ad hoc choice among chi-square
mixture approximations? Does the extra flip-specific standardization become
important once observations have different missingness patterns?

## Short answer

**Mostly yes for the nuisance correction; not yet uniformly under severe
nonnormality.** The basic flip rejected only
`r sprintf("%.1f%%", 100 * basic_rate)` of null samples. Projecting each
observed-data likelihood score away from the nuisance-score directions raised
that to `r sprintf("%.1f%%", 100 * effective_rate)`. The fully standardized
flip rejected `r sprintf("%.1f%%", 100 * standardized_rate)` and changed only
`r standardization_changes` null decisions out of about 1,800. In this
five-restriction problem, the expensive second correction is not earning its
keep.

The important qualification is distributional. Effective flipping rejected
`r sprintf("%.1f%%", 100 * effective_normal)` under normality but
`r sprintf("%.1f%%", 100 * effective_pl)` under the severe polynomial-logistic
(PL) distribution. Score pEBA4 was closer to nominal in PL, while the nested
FIML likelihood-ratio pEBA4/SB corrections remained conspicuously liberal.
This is a promising construction and a useful rescue of basic-flip
conservatism, not yet a universal finite-sample solution.

The run used **`r reps` replications per cell and `r flips` random signs per
replication**. Monte Carlo intervals are correspondingly wide; the cellwise
display below is the result, and pooled means are only orientation.

## What was flipped?

For case (i), let (s_i(\hat\theta_0)) be its score for the *observed-data*
FIML likelihood. Missing components are integrated out before this score is
formed. The effective contribution is

\[
z_i = G^\top s_i(\hat\theta_0),
\]

where (G) projects the tested loading-equality directions off the nuisance
score directions. Random Rademacher signs act on the rows (z_i), not on
labels, imputed data, residuals, or fitted parameters.

The standardized arm additionally evaluates each signed sum using its
flip-specific covariance. For FIML that covariance is assembled from the
conditional Fisher information for each observed-data pattern within group.
Thus the complete-data group correction extends to missingness without treating
all partially observed cases as equally informative.

## Design

```{r design-table}
design_show <- unique(design[c("n_group1", "n_group2", "n_total",
                               "missing_rate", "df")])
knitr::kable(design_show, digits = 2, row.names = FALSE,
             col.names = c("group 1 n", "group 2 n", "total n",
                           "MCAR target", "restrictions"))
```

The larger model is configural and the smaller model imposes equality of the
five non-marker loadings across groups. Both contain a mean structure and are
estimated by direct FIML. The power condition increases the second loading in
group 2 by 35%. Missingness is Savalei--Bentler-style MCAR with the first two
indicators kept intact. The distributions are multivariate normal and the
severe PL construction targeting marginal skewness 3 and excess kurtosis 21;
there is no (t_5) condition.

## Null calibration: retain the cells

```{r null-cells, fig.width=9, fig.height=12}
plot_methods <- c("flip_basic", "flip_effective", "flip_standardized",
                  "score_peba4", "score_sandwich",
                  "nested_peba4_ml", "nested_all_ml")
null_plot <- null[null$method %in% plot_methods, , drop = FALSE]
null_plot$method_label <- factor(label_method(null_plot$method),
  levels = rev(label_method(plot_methods)))
null_plot$missing_label <- factor(
  sprintf("%g%% MCAR", 100 * null_plot$missing_rate))
ggplot(null_plot,
       aes(n_group1, rejection_rate, colour = missing_label, group = missing_label)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = .025, ymax = .075,
           fill = "grey85", alpha = .55) +
  geom_hline(yintercept = .05, linetype = 2, colour = "grey30") +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 4,
                alpha = .35, position = position_dodge(width = 5)) +
  geom_line(position = position_dodge(width = 5), alpha = .6) +
  geom_point(position = position_dodge(width = 5), size = 1.7) +
  facet_grid(method_label ~ distribution) +
  scale_x_continuous(breaks = c(50, 100, 200)) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(x = "group 1 sample size (group 2 is 70% as large)",
       y = "null rejection probability", colour = NULL,
       caption = "Points are design cells; bars are 95% Wilson intervals. The shaded band is 2.5%--7.5%.") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.x = element_blank(), legend.position = "bottom",
        strip.text.y = element_text(angle = 0))
```

Three features matter. First, effective projection supplies the large rescue;
the basic sign statistic is too conservative throughout. Second, PL rather
than missingness is the main remaining calibration problem for the flips.
Third, the score transformations and their nested-LR names are not
interchangeable: pEBA4 applied to the score statistic is much better behaved
than pEBA4 applied to the nested FIML likelihood-ratio statistic in the severe
PL cells. That supports the earlier suspicion that a scalar SB success can be
accidental when the relevant null law has unequal eigenvalues.

## Does FIML standardization help?

```{r paired-standardization}
paired_null <- subset(paired_summary, truth == "null")
paired_by_rate <- do.call(rbind, lapply(split(paired_null,
                                               paired_null$missing_rate),
  function(x) data.frame(
    missing_rate = x$missing_rate[[1L]],
    n = sum(x$n),
    effective = weighted_cell_mean(x, "rejection_effective"),
    standardized = weighted_cell_mean(x, "rejection_standardized"),
    disagreement = weighted_cell_mean(x, "disagreement_rate"),
    mean_abs_delta_p = weighted_cell_mean(x, "mean_abs_delta_p"),
    median_relative_covariance_shift = median(x$median_variance_shift,
                                               na.rm = TRUE),
    median_patterns = median(x$median_pattern_count, na.rm = TRUE))))
knitr::kable(paired_by_rate, digits = 4, row.names = FALSE,
             col.names = c("MCAR target", "paired n", "effective reject",
                           "standardized reject", "decision disagreement",
                           "mean |delta p|", "median covariance shift",
                           "median patterns"))
```

Missingness does exactly what the construction predicts: it creates more
information strata and increases the difference between the identity-sign and
random-sign covariance matrices. But at rank five the numerical movement is
small. Every changed null decision moved from rejection to acceptance; none
moved the other way. Standardization may matter with more restrictions, more
unequal pattern information, or a much smaller effective sample, but this DGP
does not support paying for it by default.

## Power, calibrated to each method's null

```{r power-cells, fig.width=9, fig.height=7}
power_methods <- c("flip_effective", "flip_standardized", "score_peba4",
                   "score_sandwich", "nested_peba4_ml")
power_selected <- matched_power[matched_power$method %in% power_methods, ,
                                drop = FALSE]
power_selected$method_label <- factor(label_method(power_selected$method),
                                      levels = label_method(power_methods))
power_selected$missing_label <- factor(
  sprintf("%g%% MCAR", 100 * power_selected$missing_rate))
ggplot(power_selected,
       aes(n_group1, matched_power, colour = missing_label,
           group = missing_label)) +
  geom_line(alpha = .65) + geom_point(size = 1.8) +
  facet_grid(method_label ~ distribution) +
  scale_x_continuous(breaks = c(50, 100, 200)) +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::label_percent(accuracy = 1)) +
  labs(x = "group 1 sample size", y = "matched-null power", colour = NULL,
       caption = "Each threshold is estimated from only 100 null p-values in the matching cell; comparisons are illustrative.") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.x = element_blank(), legend.position = "bottom",
        strip.text.y = element_text(angle = 0))
```

Matched-null power prevents a liberal test from winning by definition. At this
probe budget, effective and standardized flips are essentially tied, and most
of the calibrated methods are also close. The severe PL cells have little
matched power for this loading perturbation; with only 100 null draws per cell,
the empirical 5% threshold itself is noisy. A larger run should therefore
spend first on null calibration, not add more test variants.

## Timing and numerical availability

```{r timing}
timing_focus <- subset(timing_summary,
  distribution == "normal" & truth == "null" & n_group1 == 100 &
    phase %in% c("fit", "flip_score", "flip_standardization", "nested_fmg",
                 "replication_total"))
timing_focus$median_ms <- 1000 * timing_focus$median_seconds
timing_wide <- reshape(timing_focus[c("missing_rate", "phase", "median_ms")],
                       idvar = "missing_rate", timevar = "phase",
                       direction = "wide")
names(timing_wide) <- sub("^median_ms\\.", "", names(timing_wide))
knitr::kable(timing_wide, digits = 2, row.names = FALSE,
             col.names = c("MCAR target", "fit", "flip score sums",
                           "flip standardization", "nested FMG battery",
                           "replication total"),
             caption = "Median milliseconds, normal-null n1=100 cell")
```

```{r failure-table}
availability_grid <- expand.grid(
  distribution = c("normal", "pl"), missing_rate = c(0, .15, .30),
  stringsAsFactors = FALSE)
availability_grid$fit_failures <- vapply(seq_len(nrow(availability_grid)),
  function(i) sum(!failures$fit_ok &
    failures$distribution == availability_grid$distribution[[i]] &
    failures$missing_rate == availability_grid$missing_rate[[i]]), integer(1L))
availability_grid$nested_only_failures <- vapply(seq_len(nrow(availability_grid)),
  function(i) sum(failures$fit_ok & failures$flip_ok & !failures$nested_ok &
    failures$distribution == availability_grid$distribution[[i]] &
    failures$missing_rate == availability_grid$missing_rate[[i]]), integer(1L))
availability_grid$flip_failures_after_fit <- vapply(seq_len(nrow(availability_grid)),
  function(i) sum(failures$fit_ok & !failures$flip_ok &
    failures$distribution == availability_grid$distribution[[i]] &
    failures$missing_rate == availability_grid$missing_rate[[i]]), integer(1L))
knitr::kable(availability_grid, row.names = FALSE,
             col.names = c("distribution", "MCAR target", "fit failures",
                           "nested-only failures", "flip failures after fit"))
```

At (n_1=100), pattern-specific standardization grows from under a millisecond
with complete data to roughly 5--7 ms with missingness and dominates the flip
calculation. The arithmetic is still modest, but effective flipping is cheaper
and gave the same decisions here. The full 3,600-replication run took
`r sprintf("%.1f", as.numeric(meta[["wall_seconds"]]))` wall seconds on
`r meta[["cores"]]` local workers; severe-PL generator calibration took about
`r sprintf("%.1f", sum(subset(dgp_timing, distribution == "pl")$generator_setup_seconds))`
seconds across its two truth conditions.

There were `r fit_failures` FIML fit failures and
`r nested_only_failures` additional nested-battery failures, concentrated in
small, 30%-missing PL cells. Conditional on a successful fit, the score-flip
path failed `r flip_failures_after_fit` times. The nested failures were
ill-conditioned or indefinite Satorra-2000 sandwich reductions, so method
availability is part of the comparison rather than a row to silently drop.

## Decision and next probe

Use the nuisance-effective flip as the primary randomization test in the next
FIML experiment. Keep the standardized arm as a diagnostic, not a default, and
retain score pEBA4 plus the direct-sandwich and exact-mixture score references.
The nested FIML corrections remain useful comparators but should not define the
answer in the severe PL cells.

The next useful spend is a larger null-only run in a reduced grid: normal and
PL, (n_1=50,100,200), 0% and 30% MCAR, with at least 500--1,000 successful
fits per cell. Only after that should the design vary restriction rank or add
MAR. That sequence tests whether the PL liberalism is persistent and whether
standardization starts paying when the tested subspace, rather than the menu of
methods, is enlarged.

## Reproduce

The compact summary CSVs are committed to freeze this probe's evidence; the
larger replication- and method-level files remain local and ignored.

```sh
Rscript experiments/66-fiml-score-flip-probe/run_experiment.R --smoke
Rscript experiments/66-fiml-score-flip-probe/run_experiment.R --probe --cores 4
quarto render experiments/66-fiml-score-flip-probe/report.qmd
```
