---
title: "88: Local convergence of ML and PSD-ML, ordinary vs sphere route (small version)"
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
res_dir <- file.path("results", "full")
fits_path <- file.path(res_dir, "fits.csv")
if (!file.exists(fits_path)) stop("run `Rscript run_experiment.R --full` first")
all_fits <- read.csv(fits_path, stringsAsFactors = FALSE)
x <- all_fits[all_fits$estimator == "ML", ]
psd <- all_fits[all_fits$estimator == "PSD-ML", ]
meta <- read.csv(file.path(res_dir, "metadata.csv"), stringsAsFactors = FALSE)
x$cell <- paste(x$ident, x$route)
cells <- c("marker ordinary", "marker sphere", "std_lv ordinary", "std_lv sphere")
pct <- function(v) round(100 * mean(v), 1)
wide <- function(expr_col, data = x) {
  a <- aggregate(data[[expr_col]] ~ design + n + cell, cbind(data, data.frame(tmp = 0)), pct)
  names(a)[4] <- "v"
  r <- reshape(a, idvar = c("design", "n"), timevar = "cell", direction = "wide")
  names(r) <- sub("^v\\.", "", names(r))
  r <- r[order(r$design, r$n), c("design", "n", intersect(cells, names(r)))]
  rownames(r) <- NULL
  r
}
x$runaway_stop <- grepl("runaway", x$status)
x$is_error <- x$status == "error"
x$is_flagged <- x$status == "flagged"
reps <- max(x$rep)
fl <- wide("is_flagged", x[x$route == "sphere", ])
```

## Question

With the canonical start and the scaled driven run, how often do ML and
PSD-ML reach a **local** optimum on the ordinary route and on the sphere
route, under the marker and std.lv identifications? When no route reaches
one, is that the optimizer's fault or is there no estimate to find? Global
optimality is not assessed.

## Short answer

```{r short}
all_fail <- aggregate(success ~ design + n + rep, x, function(v) all(!v))
af <- aggregate(success ~ design + n, all_fail, pct)
opt_fail <- table(x$cell[x$failure == "optimizer"])
```

```{r psd-short}
psd$cell <- paste(psd$ident, psd$route)
psd_rate <- tapply(psd$success, psd$cell, pct)
psd_cell_min <- aggregate(success ~ design + n + cell, psd, pct)
psd_min <- tapply(psd_cell_min$success, psd_cell_min$cell, min)
psd_flag <- pct(psd$status[psd$cell == "std_lv sphere"] == "flagged")
```

- **PSD-ML on the sphere converges essentially always.**
  - The sphere PSD-ML fit reaches a certified local optimum in
    `r psd_rate[["marker sphere"]]`% of all draws (lowest cell
    `r psd_min[["marker sphere"]]`%).
  - Reported in the std.lv identification it is always a success, though
    `r psd_flag`% of those optima lie outside std.lv.
  - The ordinary PSD-ML routes stall at their identification's pole. The
    marker route drops to `r psd_min[["marker ordinary"]]`% with a weak
    marker. The std.lv route drops to `r psd_min[["std_lv ordinary"]]`%
    with high R², where the PSD optimum puts Y's disturbance variance at
    zero in half the draws. 494 of its 500 failures are draws where the
    sphere reports the optimum as outside std.lv.
- **For ML, the identification no longer matters on the sphere route.** The marker
  and std.lv sphere columns agree within about one point in every cell.
- **Against the ordinary marker route, the sphere gains up to 10 points.**
  The gain is largest with a weak marker. Against the ordinary std.lv route
  the sphere is level from N = 20 and up to 5.5 points ahead at N = 10.
- **Most failures have no estimate to find.**
  - All four routes fail on the same draw in up to
    `r max(af$success[af$design == "ernst"])`% of Ernst draws.
  - With a weak marker, `r af$success[af$design == "weak_marker" & af$n == 100]`%
    still fail at N = 100.
  - These failures stop below the best admissible (PSD-ML) objective, and
    neither PORT nor SLSQP finds a certified local optimum. The likelihood
    keeps improving along a ridge.
- **Optimizer failures are rare.** Out of `r nrow(x) / 4` draws per route,
  another optimizer finds a local optimum the route missed in
  `r opt_fail[["marker ordinary"]]` (ordinary marker),
  `r opt_fail[["marker sphere"]]` (sphere marker),
  `r opt_fail[["std_lv ordinary"]]` (ordinary std.lv) and
  `r opt_fail[["std_lv sphere"]]` (sphere std.lv) draws.
- **Routes differ mainly in how they fail on ridges.**
  - The ordinary marker route mostly errors.
  - The ordinary std.lv route mostly reports convergence at a runaway
    point: a standardized loading, residual ratio or factor correlation
    beyond 10. That is a silent failure.
  - The sphere route sits in between.

## Evidence

### ML success rate

Share of draws (%) ending at a certified local optimum that is not a runaway
(for the sphere route: or flagged outside the identification). N = 100 is in
`results/full/fits.csv`; there Ernst and high-R² reach 100% on every route,
and the weak marker reaches 47.5% on the ordinary marker route and 54.5 to
55.5% on the other three.

```{r success}
sr <- wide("success")
sr <- sr[sr$n < 100, ]
names(sr) <- c("Design", "N", "Marker, ordinary", "Marker, sphere",
               "std.lv, ordinary", "std.lv, sphere")
knitr::kable(sr)
```

### PSD-ML success rate

Share of draws (%) ending at a certified local optimum, pooled over
N = 10 to 100, with the lowest single-N cell in parentheses. Every draw had
at least one PSD-ML route succeed, so an estimate always existed.

```{r psd-success}
pr <- aggregate(success ~ design + cell, psd, pct)
pm <- aggregate(success ~ design + cell, psd_cell_min, min)
pr$txt <- sprintf("%.1f (%.1f)", pr$success, pm$success[match(paste(pr$design, pr$cell), paste(pm$design, pm$cell))])
pw <- reshape(pr[, c("design", "cell", "txt")], idvar = "design", timevar = "cell", direction = "wide")
names(pw) <- sub("^txt\\.", "", names(pw))
pw <- pw[, c("design", cells)]
names(pw) <- c("Design", "Marker, ordinary", "Marker, sphere", "std.lv, ordinary", "std.lv, sphere")
knitr::kable(pw, row.names = FALSE)
```

### How the ML fits end

All designs pooled, N = 10 and 20 (%).

```{r ends}
small <- x[x$n <= 20, ]
ends <- do.call(rbind, lapply(cells, function(cl) {
  z <- small[small$cell == cl, ]
  data.frame(Route = sub("std_lv", "std.lv", sub(" ", ", ", cl)),
             `Local optimum` = pct(z$success),
             `Converged at a runaway` = pct(z$runaway_stop),
             `Error or not converged` = pct(z$status %in% c("error", "uncertified", "flagged_uncertified")),
             `Failure: no estimate` = pct(z$failure == "no_estimate"),
             `Failure: optimizer` = pct(z$failure == "optimizer"),
             check.names = FALSE)
}))
knitr::kable(ends)
```

- **Failures are mostly missing estimates.** The failure labels come from
  refitting with PORT and SLSQP and comparing with the best admissible
  (PSD-ML) fit. Where some local optimum exists, every route usually finds
  it.
- **The routes differ in how they fail.** The ordinary marker route errors
  out. The ordinary std.lv route mostly reports convergence at a runaway
  point, a silent failure. The sphere route has fewer runaway stops than
  std.lv and fewer errors than the marker route.
- **High R² and the std.lv identification.** In the high-R² design the
  sphere reports the optimum as lying outside the std.lv identification in
  `r round(min(fl$"std_lv sphere"[fl$design == "high_r2"]))` to
  `r round(max(fl$"std_lv sphere"[fl$design == "high_r2"]))`% of draws at
  every N: the fitted disturbance variance of Y is at or below zero. On
  352 of those 366 draws the ordinary std.lv route reports a certified local
  optimum with a worse objective (median gap 0.007 in the fit function).
  It counts as a success here, but it is the best point the std.lv
  identification can reach, not the model's optimum.

## Design

- **Model.** `X =~ x1 + x2 + x3`, `Y =~ y1 + y2 + y3`, `Y ~ X`. Normal-theory
  ML with the default optimizer, and PSD-ML (ML with every covariance
  matrix positive semidefinite) with SLSQP. Each is fitted to the same data
  four ways:
  - the ordinary fit, with the marker or the std.lv identification;
  - the sphere fit, with the marker or the std.lv identification.
- **Populations.** `var(X) = 1` and unit residual variances throughout.
  - `ernst`: loadings 1, .8, .6 and beta .25 (Ernst et al.).
  - `weak_marker`: loadings .1, .8, .6 on both factors.
  - `high_r2`: Y has R² = .98 (disturbance variance .02).
- **Grid.** N = `r paste(sort(unique(x$n)), collapse = ", ")`, with
  `r reps` replications per cell.
- **Runaway.** The largest of |standardized loading|, |residual variance /
  implied indicator variance| and |factor correlation| exceeds 10. The
  measure does not depend on the identification, and across all returned
  fits it is bimodal: most lie below 1.5, and a separate cluster lies
  between 30 and 1000.
- **Failure labels (ML).** Each ML failure is refitted with PORT and SLSQP,
  same route and start policy.
  - Optimizer failure: a refit stops cleanly at a certified, non-runaway
    local optimum.
  - No estimate: no refit does, and the failing point fits better than the
    PSD-ML solution.
  - Otherwise unclassified.

## Caveats

- **Scope.** One model family, three populations, 200 replications. Tomorrow's expansion adds designs, a least-squares estimator and N.
- **Labels are inferred.** "No estimate" relies on two alternative
  optimizers failing and on a single-start PSD-ML fit. The runaway bound is
  a heuristic, made robust by the bimodal measure.
- **Two magmaan behaviours were found along the way.** Both are in the
  backlog.
  - The sphere route can report "outside the identification" at a runaway
    stop. Here such stops count as failures.
  - Both routes can report convergence at runaway points. The runaway
    column counts them.

## Reproduce

```sh
cd experiments/engineering/88-sphere-local-convergence
Rscript run_experiment.R --full        # about 70 s on 6 cores
quarto render report.qmd
```

Run on magmaan `r meta$value[meta$key == "magmaan_git_head"]`.
