---
title: "FIML and two-stage measurement-invariance tests"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

## Question

How well do robust difference tests and direct FIML score tests calibrate
measurement-invariance restrictions under nonnormal incomplete data?

## Short answer

Spectrum corrections improve on naive difference tests, but do not settle
small-sample invariance testing. In the six-indicator, 5,000-replication null
study, FIML pEBA4 rejection ranged from 5.6–24.7% for weak and 5.6–35.6%
for strict restrictions across complete/MCAR and normal-MAR cells. Severe
nonnormality still exposes substantial inflation.

The separate direct-FIML score screen and independent hard-cell confirmation
favor score pEBA4 over effective or standardized flips. In the confirmation's
eight complete/MCAR or normal-MAR cells, score pEBA4 rejected 4.3–6.7%,
whereas effective flips reached 12.3%. Standardization supplied little gain.
These are distinct designs, not pooled replications. Scalar nesting,
larger-model coverage and size-matched power remain unfinished; multiplier
expansion is banked pending a concrete consumer.

```{r}
read_overview <- function(name) {
  path <- file.path("results", "overview", paste0(name, ".csv"))
  if (!file.exists(path)) stop("Missing retained overview: ", path,
    "; restore the tracked CSV or, with original runs present, run Rscript scripts/summarize.R.")
  read.csv(path, stringsAsFactors = FALSE)
}
percent <- function(x) sprintf("%.1f%%", 100*x)
range_percent <- function(lo, hi) paste(percent(lo), percent(hi), sep="–")
```

## Evidence

### Difference-test calibration

The executed study has a one-factor, six-indicator model at total N=400,
seven normal/VM/IG/PL generator families, balanced or 1:3 allocation, and
complete/MCAR/MAR patterns. Metadata records 70 completed generating cells
with 5,000 attempts, each evaluated with FIML and ML2S. The table includes
46 justified-null cells per estimator and step; nonnormal MAR is excluded.
Weak and strict restrictions are the implemented nested comparisons.

```{r}
c <- read_overview("calibration")
c <- c[c$method %in% c("SB","pEBA4"),]
knitr::kable(data.frame(Estimator=c$estimator, Restriction=c$step, Test=c$method,
  `Rejection range`=range_percent(c$min_rate,c$max_rate),
  `Mean absolute error`=percent(c$mean_cell_error),
  `Usable / cell`=paste0(c$min_valid,"–",c$max_valid," / 5000"), check.names=FALSE))
```

Naive rates reach 89.4% for FIML weak and 97.7% for FIML strict restrictions;
full summaries remain in `results/overview/calibration.csv`. Rejection is
conditional on a usable test; failures are not accepted as nonrejections.
A 5% rate with 5,000 usable replications has Monte Carlo standard error about
0.3 percentage points. The recorded run took 209.4 minutes.

The [population design](design/calibration_populations.md) preserves the
Brace–Savalei-inspired larger-model and power plan. Its proposed p=8/16/30
grid is **not** the executed six-indicator grid. The old `cells.csv` also
records that intended design rather than the actual run. Historical power
outputs remain under `results/calibration/_archive_pre_typeI/`; they are a
separate earlier run, not size-matched power for this null study.

### Direct-FIML score tests

The score-flips lane screens 240 cells with 500 attempts and 199 signs,
then independently confirms 11 predefined hard cells with 2,000 attempts
and 999 signs. Restriction ranks are 1/3/5. Its generators, sample sizes
and missingness laws differ from the difference-test lane. The table separates
nonnormal-MAR stress cells and reports only the eight justified-null hard
cells. Full region-separated summaries are in
`results/overview/flip_stress.csv`.

```{r}
f <- read_overview("flip_stress")
f <- f[f$profile=="confirm" & f$region=="complete/MCAR or normal MAR",]
labels <- c(flip_effective="Effective flip",flip_standardized="Standardized flip",
  score_sb="Score SB",score_peba4="Score pEBA4",nested_peba4_ml="Nested LR pEBA4")
knitr::kable(data.frame(Test=unname(labels[f$method]),
  `Mean cell rejection`=percent(f$mean_cell_rate),
  `Rejection range`=range_percent(f$min_rate,f$max_rate),
  `Usable / cell`=paste0(f$min_valid,"–",f$max_valid," / 2000"),check.names=FALSE))
```

These deliberately difficult cells are not a representative average. Flip
versus nested-test availability differs; the conditional rejection rates
cannot be read as a common denominator. The 12-worker screen and confirmation
took about 14.3 and 10.8 minutes. Median standardization cost increased to
94.8 ms at 999 signs, versus 12.1 ms for effective sign sums. There is no
independent evidence here for promoting standardized flips as a default.

### Oracle and pilot controls

The former showcase is now the oracle lane. Its three-replication local
normal-complete smoke checks numerical LR and UΓ agreement with lavaan:

```{r}
o <- read_overview("oracle_checks")
knitr::kable(data.frame(Quantity=o$metric, Comparisons=o$comparisons,
  `Maximum absolute difference`=format(o$max_abs_difference,digits=3),check.names=FALSE))
```

This checks implementation conventions; it cannot establish test size.
Maintained C++ golden/unit gates remain the correctness authority. The former
archived pilot retains its unique loading-power arm and paired controls in
`results/pilot/` and `results/overview/pilot_power.csv`. Its 100-replication
results are illustrative, and its seed base overlaps the later screen. It
is not an additional independent confirmation sample.

## Caveats

A generating-model invariance restriction under nonnormal MAR need not be a
restriction on the Gaussian-FIML pseudo-target. Those cells remain estimand
stress until independently justified. Information choices for parameter SEs,
global exact fit and nested constraints are different scientific questions.
This report joins overlapping nested-test evidence without claiming their
nulls or geometries are interchangeable.

The active work is valid metric-to-scalar nesting with released latent means,
a frozen larger-model null design, and size-matched power. Preserve completed
screen/confirmation evidence; reopen multiplier development only for a named
incomplete-data score consumer or a new rank/missingness failure regime.
No new estimator or inference default is selected by this reorganization.

## Reproduce

The report renders from the small tracked overview:

```sh
quarto render experiments/research/active/06-fiml-invariance-tests/report.qmd
```

From this study directory, inspect lane help and run fresh smoke outputs:

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --lane calibration --help
Rscript run_experiment.R --lane calibration --smoke \
  --results-dir /tmp/fiml-invariance-fresh
Rscript run_experiment.R --lane score-flips --help
Rscript run_experiment.R --lane oracle --help
Rscript run_experiment.R --lane pilot --help
```

`Rscript scripts/summarize.R` rebuilds the retained overview only with the
original runs present. Source paths and hashes are in
`results/overview/provenance.csv`. Every lane keeps its historical populations,
seeds, failure records and metadata. Source fingerprints changed when paths
moved; do not relax checkpoint guards or overwrite historical results.
