---
title: "Foldnes, Moss & Grønneberg (2026), Study 1: magmaan vs lavaan parity"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results <- file.path("results")
need <- c("lavaan_parity.csv", "summary_chi2.csv", "cells.csv", "metadata.csv")
missing <- need[!file.exists(file.path(results, need))]
if (length(missing)) {
  stop("missing results: ", paste(missing, collapse = ", "),
       "\nRun:  Rscript run_experiment.R --lavaan-parity")
}
parity  <- read.csv(file.path(results, "lavaan_parity.csv"))
summ    <- read.csv(file.path(results, "summary_chi2.csv"))
meta    <- read.csv(file.path(results, "metadata.csv"))
moments <- if (file.exists(file.path(results, "moments.csv")))
  read.csv(file.path(results, "moments.csv")) else NULL
mv <- function(k) meta$value[match(k, meta$key)]
fmt <- function(x, d = 0) formatC(x, format = "f", digits = d, big.mark = "")
```

## Question

The paper *"Penalized eigenvalue block averaging: Extension to nested model
comparison and Monte Carlo evaluations"* (Behavior Research Methods 2026,
58:107) evaluates goodness-of-fit test statistics for confirmatory factor
models under non-normality. Every statistic it studies is built on normal-theory
ML estimation and one of two base fit statistics, the standard $T_{\text{ML}}$
and Browne's reweighted least squares $T_{\text{RLS}}$, optionally rescaled by a
robust correction (Satorra–Bentler, scaled-and-shifted) computed from the
asymptotic covariance matrix $\Gamma$.

This experiment asks a parity question, not a methods question: **does magmaan
reproduce the Study-1 goodness-of-fit machinery, and does it match lavaan where
lavaan exposes the same statistic?** We deliberately keep the replication small
and correctly specified, in the spirit of the Maydeu-Olivares (2017) parity
experiment (`experiments/07-maydeu-olivares-2017`).

## Short answer

Yes. On the paper's correctly-specified five-factor design, with normal and
Vale–Maurelli non-normal data, magmaan matches lavaan to optimizer-level
tolerance on all four statistics lavaan exposes:

```{r headline}
anchor <- parity[parity$metric %in% c("T_ML", "T_RLS", "SB", "SS", "SB_scaling"), ]
maxd <- max(anchor$abs_diff, na.rm = TRUE)
cat(sprintf("Across %d cells and %d compared values, max |magmaan - lavaan| = %.2e.\n",
            length(unique(parity$cell_idx)), nrow(anchor), maxd))
```

$T_{\text{ML}}$ agrees to ~$10^{-9}$; $T_{\text{RLS}}$ and the robust statistics
agree to ~$10^{-4}$, the level set by independent optimizer convergence. The
Satorra–Bentler scaling factor matches lavaan including the cases where it falls
*below* one for these confirmatory models.

## Evidence

### lavaan parity

`magmaan` and `lavaan::cfa(..., estimator = "ML")` are compared on the first
replicate of each cell. The four anchors are lavaan's `standard`
($T_{\text{ML}}$), `browne.residual.nt.model` ($T_{\text{RLS}}$),
`satorra.bentler` (SB) and `scaled.shifted` (SS) tests, plus the SB scaling
factor.

```{r parity-table}
agg <- do.call(rbind, lapply(split(anchor, anchor$metric), function(g) {
  data.frame(Statistic = g$metric[1],
             `Max abs diff` = max(g$abs_diff, na.rm = TRUE),
             `Median abs diff` = median(g$abs_diff, na.rm = TRUE),
             check.names = FALSE)
}))
ord <- c("T_ML", "T_RLS", "SB", "SS", "SB_scaling")
agg <- agg[match(ord, agg$Statistic), ]
agg$Statistic <- c("T_ML (standard)", "T_RLS (Browne residual NT)",
                   "SB (Satorra-Bentler)", "SS (scaled-shifted)",
                   "SB scaling factor")[match(agg$Statistic, ord)]
knitr::kable(agg, digits = 9, row.names = FALSE,
             caption = "magmaan vs lavaan, aggregated over all cells (first replicate each).")
```

The largest discrepancies sit on $T_{\text{RLS}}$ and the robust statistics in
the largest model ($p = 30$), consistent with two optimizers stopping at
slightly different points rather than a formula difference; $T_{\text{ML}}$,
which depends only on $\hat\theta$ and the sample moments, agrees far more
tightly.

### The Satorra–Bentler scaling tracks lavaan, including below one

```{r scaling}
sc <- parity[parity$metric == "SB_scaling", c("dist", "p", "N", "magmaan", "lavaan")]
sc <- sc[order(sc$dist, sc$p, sc$N), ]
names(sc) <- c("Distribution", "p", "N", "magmaan c", "lavaan c")
knitr::kable(sc, digits = 4, row.names = FALSE,
             caption = "SB scaling factor c (mean of the UGamma eigenvalue spectrum).")
```

For these correctly-specified five-factor models the projected residual
non-normality does not always inflate the scaling: under the severe Vale–Maurelli
condition $c$ can drop below one. magmaan reproduces lavaan's value either way,
which is the point — the eigenvalue spectrum magmaan computes (and which the
paper's pEBA/pOLS tests consume) is the same object lavaan reduces to its
scaling factor.

### Non-normal margins

Non-normal data are generated with the Vale–Maurelli (1983) polynomial method,
self-contained in `R/population.R` (no `covsim` dependency). Achieved marginal
moments on a large draw:

```{r moments}
if (!is.null(moments)) {
  m <- moments[, c("dist", "skew_target", "kurt_target",
                   "skew_achieved", "kurt_achieved")]
  names(m) <- c("Distribution", "Target skew", "Target kurtosis",
                "Achieved skew", "Achieved kurtosis")
  knitr::kable(m, digits = 3, row.names = FALSE)
} else cat("No non-normal cells in this run.\n")
```

## Caveats

This experiment establishes **implementation parity on the lavaan-exposed
statistics**, nothing more. In particular:

- **Population is reconstructed, not the paper's.** We follow the verbal recipe
  (five factors, equal indicators, loadings $\sim U[0.5, 1.5]$, unit latent and
  error variances, interfactor correlations cycled through
  $\{-.3, -.2, 0, .1, .2, .3\}$). The paper's *realized* loading matrices and the
  exact assignment of interfactor correlations to factor pairs live only in the
  OSF supplement (<https://osf.io/h2y3n/>). Our populations are representative,
  not identical.
- **The eigenvalue tests are not parity-checked.** ALL, EBA, pEBA*x*, pOLS, pOLS2
  and the scaled-$F$ test are implemented in the `semTests` package, not lavaan,
  and `semTests` is not installed here. magmaan *does* produce the UGamma
  eigenvalue spectrum these tests consume (and its mean reproduces lavaan's SB
  scaling), but the block-averaging and penalization steps themselves are out of
  scope for a lavaan-parity check.
- **Small replication count.** This run uses `r mv("reps")` replicates per cell.
  That is enough to verify parity but cannot estimate a 5% rejection rate; the
  paper uses 2000. `results/summary_chi2.csv` reports rejection rates only as a
  shape check, not as a finding. We do not attach Monte Carlo standard errors.
- **Distributions limited to normal and Vale–Maurelli.** The paper also uses the
  independent-generator and piecewise-linear families (via `covsim`); those are
  not reproduced here. Parity is distribution-agnostic, so this does not weaken
  the conclusion.

### Is the simulation clearly described in the paper?

For reproduction purposes: the *design grid* is fully specified (factors, model
sizes $p \in \{15, 30, 60, 100\}$, $n \in \{400, 800, 2000\}$, seven
distributions with target skew/kurtosis $(2,7)$ and $(3,21)$, the RMSE/MAD/ARR
rejection-rate criteria, and 2000 replications). The *exact population matrices*
are not in the text and require the OSF supplement, as the paper states. A reader
can rebuild the structure from the article alone but cannot reproduce the precise
populations without the supplement.

## Reproduce

```bash
# from the repo root, with the magmaan R package installed (just r-install)
cd experiments/18-foldnes-moss-gronneberg-2026
Rscript run_experiment.R --lavaan-parity        # full 12-cell grid, 5 reps
Rscript run_experiment.R --smoke                 # one cell, one rep
Rscript run_experiment.R --cells "p=15,dist=vm2" --reps 20 --lavaan-parity
quarto render report.qmd
```

```{r footer}
cat(sprintf("magmaan %s, lavaan %s; reps=%s, cells=%s, seed-base=%s; run in %ss.\n",
            mv("magmaan_version"), mv("lavaan_version"), mv("reps"),
            mv("n_cells"), mv("seed_base"), mv("total_seconds")))
```
