---
title: "Foldnes, Moss & Grønneberg (2026), Study 1: magmaan parity"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results <- file.path("results")
need <- c("lavaan_parity.csv", "cells.csv", "metadata.csv")
missing <- need[!file.exists(file.path(results, need))]
if (length(missing)) {
  stop("missing results: ", paste(missing, collapse = ", "),
       "\nRun:  Rscript run_experiment.R --semtests-parity")
}
parity  <- read.csv(file.path(results, "lavaan_parity.csv"))
meta    <- read.csv(file.path(results, "metadata.csv"))
fmg     <- if (file.exists(file.path(results, "semtests_parity.csv")))
  read.csv(file.path(results, "semtests_parity.csv")) else NULL
mv <- function(k) meta$value[match(k, meta$key)]
anchor  <- parity[parity$metric %in% c("T_ML", "T_RLS", "SB", "SS"), ]
maxd_lav <- max(anchor$abs_diff, na.rm = TRUE)
maxd_st  <- if (!is.null(fmg)) max(fmg$abs_diff, na.rm = TRUE) else NA_real_
n_cells  <- length(unique(parity$cell_idx))
```

## Question

The paper evaluates goodness-of-fit tests for factor models under
non-normality, building everything on two base statistics (the standard
chi-square and Browne's reweighted least-squares statistic) and a family of
robust corrections, from the familiar Satorra–Bentler and scaled-shifted tests
up to the new penalized eigenvalue block-averaging tests (pEBA, pOLS) and their
relatives. Can magmaan reproduce that whole machinery, and does it agree with
the two reference implementations — lavaan for the base statistics, and the
paper's own `semTests` package for the eigenvalue tests?

## Short answer

Yes, to numerical tolerance. On the paper's correctly-specified five-factor
design, across `r n_cells` cells of normal and non-normal data, magmaan matches
**lavaan** on the standard chi-square, Browne's statistic, and the
Satorra–Bentler and scaled-shifted corrections to within
`r sprintf("%.0e", maxd_lav)` (the standard chi-square agrees to about
$10^{-9}$). And magmaan matches **`semTests`** on the full robust family — SB,
SS, the scaled-$F$ test, the block-averaging tests pEBA and pOLS, and their
penalized variants, in every combination of base statistic and covariance
estimator the paper crosses — with p-values agreeing to within
`r if (is.na(maxd_st)) "n/a" else sprintf("%.0e", maxd_st)`. The remaining
differences are the size of two estimators stopping at slightly different
points, not formula disagreements.

## Evidence

magmaan, lavaan, and `semTests` are compared on the same first-replicate data of
each cell. The base statistics are compared as statistic values (lavaan); the
robust family is compared as test p-values (`semTests`).

```{r parity-summary}
rows <- data.frame(
  Comparison = c("Standard chi-square (vs lavaan)",
                 "Browne reweighted statistic (vs lavaan)",
                 "Satorra-Bentler & scaled-shifted (vs lavaan)",
                 "Full robust family incl. pEBA / pOLS (vs semTests)"),
  Quantity = c("statistic", "statistic", "statistic", "p-value"),
  `Comparisons` = c(
    sum(parity$metric == "T_ML"),
    sum(parity$metric == "T_RLS"),
    sum(parity$metric %in% c("SB", "SS")),
    if (is.null(fmg)) 0L else nrow(fmg)),
  `Max abs diff` = c(
    max(parity$abs_diff[parity$metric == "T_ML"], na.rm = TRUE),
    max(parity$abs_diff[parity$metric == "T_RLS"], na.rm = TRUE),
    max(parity$abs_diff[parity$metric %in% c("SB", "SS")], na.rm = TRUE),
    if (is.null(fmg)) NA_real_ else max(fmg$abs_diff, na.rm = TRUE)),
  check.names = FALSE)
knitr::kable(rows, digits = 9, row.names = FALSE,
             caption = "magmaan vs the two reference implementations, worst case over all cells.")
```

One substantive aside worth noting: for these correctly-specified models the
robust correction does not always inflate the statistic. Under the more severe
non-normal condition the Satorra–Bentler scaling factor drops *below* one, and
magmaan reproduces lavaan's value in that regime just as well as in the usual
one — so the eigenvalue information the penalized tests depend on is being
computed correctly, not merely the average that the classical correction uses.

## Caveats

This experiment establishes that magmaan computes the same numbers as the
reference tools. It is not a methods study and says nothing new about which test
performs best.

- **The population is representative, not the paper's.** We follow the paper's
  recipe (five factors, equal indicators, loadings drawn uniformly, interfactor
  correlations spread over a small fixed set), but the exact population matrices
  live in the paper's online supplement and are not reproduced here. Parity does
  not depend on the specific population.
- **Few replicates.** This run uses `r mv("reps")` replicates per cell — enough
  to confirm parity, far too few to estimate a 5% rejection rate (the paper uses
  2000). The rejection-rate file in `results/` is a shape check only, and carries
  no Monte Carlo error bars.
- **Distribution families.** The run covers normal and the moderate/severe
  Vale-Maurelli, independent-generator, and piecewise-linear conditions through
  magmaan's VM, IG, and PLSIM generators. The IG cells use a symmetric root and
  Pearson generator marginals because that combination covers the whole
  experiment grid. Parity is distribution-agnostic; the generator cells make
  the smoke and timing runs closer to the paper's design.
- **Is the simulation reproducible from the paper alone?** The design grid is
  fully specified; the exact populations require the online supplement, as the
  paper itself states.

## Reproduce

```bash
# from the repo root, with magmaan, lavaan, and semTests installed
cd experiments/18-foldnes-moss-gronneberg-2026
Rscript run_experiment.R --semtests-parity      # full grid, both parity tracks
Rscript run_experiment.R --smoke                 # one IG cell, three reps
quarto render report.qmd
```

```{r footer}
cat(sprintf("magmaan %s, lavaan %s, semTests %s; reps=%s, cells=%s; run in %ss.\n",
            mv("magmaan_version"), mv("lavaan_version"),
            tryCatch(as.character(utils::packageVersion("semTests")),
                     error = function(e) "n/a"),
            mv("reps"), mv("n_cells"), mv("total_seconds")))
```
