---
title: "Complete-Data Estimator Timing"
format: html
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path), dirname(getwd())))
result_candidates <- unique(c(file.path(candidate_dirs, "results"), candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "fits.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

fits_path <- file.path(results_dir, "fits.csv")
pairs_path <- file.path(results_dir, "pairs_vs_nt.csv")
summary_path <- file.path(results_dir, "case_summary.csv")
if (!file.exists(fits_path) || !file.exists(pairs_path) || !file.exists(summary_path)) {
  stop("Missing experiment CSVs. Run: Rscript run_experiment.R")
}

fits <- utils::read.csv(fits_path, check.names = FALSE)
pairs <- utils::read.csv(pairs_path, check.names = FALSE)
case_summary <- utils::read.csv(summary_path, check.names = FALSE)

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), "", formatC(x, format = "g", digits = digits))
}
fmt_ms <- function(x) {
  ifelse(is.na(x), "", sprintf("%.3f", 1000 * x))
}
fmt_ratio <- function(x) {
  ifelse(is.na(x), "", sprintf("%.3f", x))
}
fmt_pct <- function(x) {
  ifelse(is.na(x), "", sprintf("%.1f%%", 100 * x))
}
median_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_real_)
  stats::median(x)
}
```

This is a curiosity-grade comparison of complete-data estimator paths in
magmaan. `NT` is normal-theory ML. `ULS` and `GLS` are the complete-data GMM
least-squares paths. Parameter estimates, implied moments, and a naive
chi-square-like value are compared back to NT.

The naive chi-square column is exactly
`magmaan_core$infer_chi2_stat(sample_stats, fit$fmin)`. For ULS/GLS this is a
rough common-scale diagnostic, not an official test statistic.

## Run Overview

```{r}
overview <- data.frame(
  cases = length(unique(fits$case_id)),
  labels = length(unique(fits$label)),
  fit_rows = nrow(fits),
  pair_rows = nrow(pairs),
  ok_fits = sum(fits$status == "OK", na.rm = TRUE),
  errors = sum(fits$status == "ERROR", na.rm = TRUE),
  skips = sum(fits$status == "SKIP", na.rm = TRUE),
  stringsAsFactors = FALSE
)
knitr::kable(overview)
```

## Estimator Summary

```{r}
fit_ok <- fits[fits$status == "OK", ]
pair_ok <- pairs[pairs$status == "OK", ]

fit_rows <- lapply(split(fit_ok, fit_ok$estimator), function(x) {
  data.frame(
    estimator = x$estimator[[1]],
    ok_fits = nrow(x),
    convergence_rate = mean(x$converged, na.rm = TRUE),
    median_elapsed_ms = median_na(x$elapsed_sec) * 1000,
    median_f_evals = median_na(x$f_evals),
    median_g_evals = median_na(x$g_evals),
    median_grad_norm = median_na(x$grad_norm),
    stringsAsFactors = FALSE
  )
})
fit_table <- do.call(rbind, fit_rows)
fit_table$convergence_rate <- fmt_pct(fit_table$convergence_rate)
fit_table$median_elapsed_ms <- fmt_num(fit_table$median_elapsed_ms)
fit_table$median_f_evals <- fmt_num(fit_table$median_f_evals)
fit_table$median_g_evals <- fmt_num(fit_table$median_g_evals)
fit_table$median_grad_norm <- fmt_num(fit_table$median_grad_norm)
knitr::kable(fit_table, row.names = FALSE)
```

## Drift From NT

```{r}
if (nrow(pair_ok)) {
  pair_rows <- lapply(split(pair_ok, pair_ok$estimator), function(x) {
    data.frame(
      estimator = x$estimator[[1]],
      ok_pairs = nrow(x),
      median_elapsed_ratio_vs_nt = median_na(x$elapsed_ratio_estimator_over_nt),
      median_param_max_abs_diff = median_na(x$param_max_abs_diff),
      median_param_rms_diff = median_na(x$param_rms_diff),
      median_sigma_max_abs_diff = median_na(x$implied_sigma_max_abs_diff),
      median_naive_chisq_diff = median_na(x$naive_chisq_diff_estimator_minus_nt),
      median_abs_naive_chisq_diff = median_na(x$abs_naive_chisq_diff),
      stringsAsFactors = FALSE
    )
  })
  drift_table <- do.call(rbind, pair_rows)
  num <- vapply(drift_table, is.numeric, logical(1))
  drift_table[num] <- lapply(drift_table[num], fmt_num)
  knitr::kable(drift_table, row.names = FALSE)
}
```

## Timing Ratios

Ratio below is estimator elapsed time divided by NT elapsed time for the same
case and replicate. Values below 1 mean the estimator was faster than NT in this
run.

```{r}
#| fig-width: 7
#| fig-height: 4
if (nrow(pair_ok) && any(is.finite(pair_ok$elapsed_ratio_estimator_over_nt))) {
  op <- par(mar = c(4, 5, 1, 1))
  boxplot(elapsed_ratio_estimator_over_nt ~ estimator,
          data = pair_ok,
          log = "y",
          ylab = "elapsed ratio vs NT",
          xlab = "estimator",
          outline = FALSE,
          col = "grey90")
  stripchart(elapsed_ratio_estimator_over_nt ~ estimator,
             data = pair_ok,
             add = TRUE,
             vertical = TRUE,
             method = "jitter",
             jitter = 0.15,
             pch = 16,
             col = grDevices::adjustcolor("steelblue", alpha.f = 0.35))
  abline(h = 1, lty = 2, col = "gray40")
  par(op)
} else {
  plot.new(); text(0.5, 0.5, "No complete pairs")
}
```

## Label-Sorted Case Table

```{r}
display <- case_summary[order(case_summary$label, case_summary$case_id), ]
keep <- c("label", "source", "family",
          "nt_time_sec", "uls_time_sec", "gls_time_sec",
          "uls_elapsed_ratio_over_nt", "gls_elapsed_ratio_over_nt",
          "uls_param_max_abs_diff", "gls_param_max_abs_diff",
          "uls_chisq_diff_vs_nt", "gls_chisq_diff_vs_nt")
display <- display[keep]
for (col in c("nt_time_sec", "uls_time_sec", "gls_time_sec")) {
  display[[col]] <- fmt_ms(display[[col]])
}
for (col in c("uls_elapsed_ratio_over_nt", "gls_elapsed_ratio_over_nt")) {
  display[[col]] <- fmt_ratio(display[[col]])
}
for (col in c("uls_param_max_abs_diff", "gls_param_max_abs_diff",
              "uls_chisq_diff_vs_nt", "gls_chisq_diff_vs_nt")) {
  display[[col]] <- fmt_num(display[[col]])
}
names(display) <- c("label", "source", "family",
                    "NT ms", "ULS ms", "GLS ms",
                    "ULS/NT", "GLS/NT",
                    "ULS max |dpar|", "GLS max |dpar|",
                    "ULS naive dchisq", "GLS naive dchisq")
knitr::kable(display, row.names = FALSE)
```

## Fit Errors And Skips

```{r}
problem_rows <- fits[fits$status != "OK", ]
if (nrow(problem_rows)) {
  keep <- c("case_id", "label", "estimator", "replicate", "status", "error")
  knitr::kable(head(problem_rows[keep], 60L), row.names = FALSE)
} else {
  knitr::kable(data.frame(note = "No skipped or failed fit rows."))
}
```
