---
title: "Complete-Data Estimator Timing"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path), dirname(getwd())))
result_candidates <- unique(c(file.path(candidate_dirs, "results"), candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "fits.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

fits_path <- file.path(results_dir, "fits.csv")
pairs_path <- file.path(results_dir, "pairs_vs_nt.csv")
case_path <- file.path(results_dir, "case_summary.csv")
group_path <- file.path(results_dir, "group_summary.csv")
coverage_path <- file.path(results_dir, "coverage.csv")
if (!file.exists(fits_path) || !file.exists(pairs_path) ||
    !file.exists(case_path) || !file.exists(group_path) ||
    !file.exists(coverage_path)) {
  stop("Missing experiment CSVs. Run: Rscript run_experiment.R")
}

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("The report requires ggplot2. Install it or inspect the CSVs directly.")
}

library(ggplot2)

fits <- utils::read.csv(fits_path, check.names = FALSE)
pairs <- utils::read.csv(pairs_path, check.names = FALSE)
case_summary <- utils::read.csv(case_path, check.names = FALSE)
group_summary <- utils::read.csv(group_path, check.names = FALSE)
coverage <- utils::read.csv(coverage_path, check.names = FALSE)

pair_ok <- pairs[pairs$status == "OK" &
                   is.finite(pairs$elapsed_ratio_estimator_over_nt) &
                   pairs$elapsed_ratio_estimator_over_nt > 0, , drop = FALSE]
fit_ok <- fits[fits$status == "OK" & is.finite(fits$elapsed_sec), , drop = FALSE]
problem_rows <- fits[fits$status != "OK", , drop = FALSE]

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), "", formatC(x, format = "g", digits = digits))
}
fmt_ms <- function(x) {
  ifelse(is.na(x), "", sprintf("%.3f", x))
}
fmt_ratio <- function(x) {
  ifelse(is.na(x), "", sprintf("%.3f", x))
}
fmt_pct <- function(x) {
  ifelse(is.na(x), "", sprintf("%.1f%%", 100 * x))
}
geo_mean <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (!length(x)) return(NA_real_)
  exp(mean(log(x)))
}
median_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_real_)
  stats::median(x)
}
ratio_breaks <- c(0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50)
theme_set(theme_minimal(base_size = 12))
```

This report compares wall time for three complete-data continuous estimator
paths in magmaan across the checked-out textbook corpus:

- **NT**: normal-theory ML via `fit_ml()`
- **ULS**: unweighted least squares via `fit_uls()`
- **GLS**: generalized least squares via `fit_gls()`

Timing ratios are computed within each `(case, replicate)` pair as
`estimator_time / NT_time`. Values below 1 mean the estimator was faster than
NT. Aggregates use geometric means, which are the natural average for ratios.

## Executive Summary

```{r}
total <- group_summary[group_summary$group_kind == "Total", , drop = FALSE]
all_tags <- sort(unique(unlist(strsplit(as.character(coverage$tags), ";", fixed = TRUE))))
all_tags <- all_tags[nzchar(all_tags)]
timed_coverage <- coverage[coverage$coverage_status == "timed_candidate", , drop = FALSE]
preload_skips <- coverage[coverage$coverage_status != "timed_candidate", , drop = FALSE]
complete_case_ids <- unique(pair_ok$case_id)
overview <- data.frame(
  Metric = c("Corpus Cases Considered", "Timed Candidates",
             "Cases With Complete Timing Pairs", "Pre-Timing Skips",
             "Textbook Subcorpora", "Tags", "Replicates",
             "Fit Rows", "Complete Timing Pairs", "Fit Errors/Skips"),
  Value = c(
    nrow(coverage),
    nrow(timed_coverage),
    length(complete_case_ids),
    nrow(preload_skips),
    length(unique(coverage$source)),
    length(all_tags),
    if (nrow(fits)) max(fits$replicate, na.rm = TRUE) else NA_integer_,
    nrow(fits),
    nrow(pair_ok),
    nrow(problem_rows)
  ),
  check.names = FALSE
)
knitr::kable(overview, row.names = FALSE)
```

```{r}
total_table <- total[, c("estimator", "n_cases", "n_pairs",
                         "geometric_mean_ratio", "median_ratio",
                         "p25_ratio", "p75_ratio",
                         "share_faster_than_nt",
                         "median_nt_ms", "median_estimator_ms")]
names(total_table) <- c("Estimator", "Cases", "Pairs", "Geomean Ratio",
                        "Median Ratio", "P25", "P75", "Faster Than NT",
                        "Median NT ms", "Median Estimator ms")
total_table$`Geomean Ratio` <- fmt_ratio(total_table$`Geomean Ratio`)
total_table$`Median Ratio` <- fmt_ratio(total_table$`Median Ratio`)
total_table$P25 <- fmt_ratio(total_table$P25)
total_table$P75 <- fmt_ratio(total_table$P75)
total_table$`Faster Than NT` <- fmt_pct(total_table$`Faster Than NT`)
total_table$`Median NT ms` <- fmt_ms(total_table$`Median NT ms`)
total_table$`Median Estimator ms` <- fmt_ms(total_table$`Median Estimator ms`)
knitr::kable(total_table, row.names = FALSE)
```

## Corpus Coverage

```{r}
coverage_status <- as.data.frame(table(coverage$coverage_status),
                                 stringsAsFactors = FALSE)
names(coverage_status) <- c("Status", "Cases")
knitr::kable(coverage_status, row.names = FALSE)
```

```{r}
if (nrow(preload_skips)) {
  skip_counts <- sort(table(preload_skips$skip_reason), decreasing = TRUE)
  skip_summary <- data.frame(
    Reason = names(skip_counts),
    Cases = as.integer(skip_counts),
    stringsAsFactors = FALSE
  )
  knitr::kable(skip_summary, row.names = FALSE)
} else {
  knitr::kable(data.frame(Note = "No cases were skipped before timing."))
}
```

## Overall Timing Ratios

```{r}
#| fig-width: 7
#| fig-height: 4.5
if (nrow(pair_ok)) {
  ggplot(pair_ok, aes(estimator, elapsed_ratio_estimator_over_nt,
                     color = estimator)) +
    geom_boxplot(outlier.shape = NA, width = 0.55, alpha = 0.15) +
    geom_jitter(width = 0.16, height = 0, alpha = 0.35, size = 1.2) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray35") +
    scale_y_log10(breaks = ratio_breaks) +
    labs(x = NULL, y = "Timing Ratio vs NT (log scale)",
         title = "Estimator Timing Ratios",
         subtitle = "Each point is one case/replicate pair") +
    theme(legend.position = "none")
}
```

## Subcorpus Breakdown

```{r}
subcorpus <- group_summary[group_summary$group_kind == "Subcorpus", , drop = FALSE]
subcorpus <- subcorpus[order(subcorpus$group, subcorpus$estimator), ]
sub_table <- subcorpus[, c("group", "estimator", "n_cases", "n_pairs",
                           "geometric_mean_ratio", "median_ratio",
                           "share_faster_than_nt")]
names(sub_table) <- c("Subcorpus", "Estimator", "Cases", "Pairs",
                      "Geomean Ratio", "Median Ratio", "Faster Than NT")
sub_table$`Geomean Ratio` <- fmt_ratio(sub_table$`Geomean Ratio`)
sub_table$`Median Ratio` <- fmt_ratio(sub_table$`Median Ratio`)
sub_table$`Faster Than NT` <- fmt_pct(sub_table$`Faster Than NT`)
knitr::kable(sub_table, row.names = FALSE)
```

```{r}
#| fig-width: 8
#| fig-height: 5
if (nrow(pair_ok)) {
  ggplot(pair_ok, aes(estimator, elapsed_ratio_estimator_over_nt,
                     color = estimator)) +
    geom_boxplot(outlier.shape = NA, width = 0.55, alpha = 0.12) +
    geom_jitter(width = 0.16, height = 0, alpha = 0.25, size = 0.9) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
    scale_y_log10(breaks = ratio_breaks) +
    facet_wrap(~ source, nrow = 1) +
    labs(x = NULL, y = "Timing Ratio vs NT (log scale)",
         title = "Timing Ratios By Subcorpus") +
    theme(legend.position = "none")
}
```

## Tag Breakdown

```{r}
tag_summary <- group_summary[group_summary$group_kind == "Tag", , drop = FALSE]
tag_summary <- tag_summary[order(tag_summary$estimator,
                                 tag_summary$geometric_mean_ratio), ]
tag_summary$tag_label <- factor(tag_summary$group,
                                levels = rev(unique(tag_summary$group)))

tag_table <- tag_summary[, c("group", "estimator", "n_cases", "n_pairs",
                             "geometric_mean_ratio", "p25_ratio", "p75_ratio")]
names(tag_table) <- c("Tag", "Estimator", "Cases", "Pairs",
                      "Geomean Ratio", "P25", "P75")
tag_table$`Geomean Ratio` <- fmt_ratio(tag_table$`Geomean Ratio`)
tag_table$P25 <- fmt_ratio(tag_table$P25)
tag_table$P75 <- fmt_ratio(tag_table$P75)
knitr::kable(tag_table, row.names = FALSE)
```

```{r}
#| fig-width: 8
#| fig-height: 6
if (nrow(tag_summary)) {
  ggplot(tag_summary,
         aes(geometric_mean_ratio, tag_label, color = estimator)) +
    geom_errorbarh(aes(xmin = p25_ratio, xmax = p75_ratio),
                   height = 0.18, alpha = 0.45,
                   position = position_dodge(width = 0.45)) +
    geom_point(size = 2.2, position = position_dodge(width = 0.45)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "gray40") +
    scale_x_log10(breaks = ratio_breaks) +
    labs(x = "Geometric Mean Timing Ratio vs NT (log scale)",
         y = NULL,
         color = "Estimator",
         title = "Geometric Mean Ratios By Tag",
         subtitle = "Horizontal bars show interquartile ranges") +
    theme(legend.position = "bottom")
}
```

## Wall Time Scale

```{r}
#| fig-width: 8
#| fig-height: 4.5
if (nrow(fit_ok)) {
  fit_ok$elapsed_ms <- 1000 * fit_ok$elapsed_sec
  ggplot(fit_ok, aes(estimator, elapsed_ms, color = estimator)) +
    geom_boxplot(outlier.shape = NA, width = 0.55, alpha = 0.12) +
    geom_jitter(width = 0.16, height = 0, alpha = 0.25, size = 0.9) +
    scale_y_log10() +
    facet_wrap(~ source, nrow = 1) +
    labs(x = NULL, y = "Wall Time (ms, log scale)",
         title = "Raw Wall Times By Subcorpus") +
    theme(legend.position = "none")
}
```

## Outliers To Inspect

These are not automatically bad results. They are the timing pairs that most
deviate from NT on a multiplicative scale and are worth eyeballing before
drawing conclusions.

```{r}
if (nrow(pair_ok)) {
  outliers <- do.call(rbind, lapply(split(pair_ok, paste(pair_ok$case_id, pair_ok$estimator)), function(x) {
    data.frame(
      Label = x$label[[1L]],
      Subcorpus = x$source[[1L]],
      Tag = x$primary_tag[[1L]],
      Estimator = x$estimator[[1L]],
      Reps = nrow(x),
      `Median NT ms` = 1000 * median_na(x$nt_elapsed_sec),
      `Median Estimator ms` = 1000 * median_na(x$estimator_elapsed_sec),
      `Median Ratio` = median_na(x$elapsed_ratio_estimator_over_nt),
      `Min Ratio` = min(x$elapsed_ratio_estimator_over_nt, na.rm = TRUE),
      `Max Ratio` = max(x$elapsed_ratio_estimator_over_nt, na.rm = TRUE),
      `Median |d Chi Sq|` = median_na(x$abs_naive_chisq_diff),
      `Median Max |d Par|` = median_na(x$param_max_abs_diff),
      check.names = FALSE
    )
  }))
  outliers$log_abs_ratio <- abs(log(outliers$`Median Ratio`))
  outliers <- head(outliers[order(-outliers$log_abs_ratio), ], 20L)
  out <- outliers[, c("Label", "Subcorpus", "Tag", "Estimator", "Reps",
                      "Median NT ms", "Median Estimator ms",
                      "Median Ratio", "Min Ratio", "Max Ratio",
                      "Median |d Chi Sq|", "Median Max |d Par|")]
  out$`Median NT ms` <- fmt_ms(out$`Median NT ms`)
  out$`Median Estimator ms` <- fmt_ms(out$`Median Estimator ms`)
  out$`Median Ratio` <- fmt_ratio(out$`Median Ratio`)
  out$`Min Ratio` <- fmt_ratio(out$`Min Ratio`)
  out$`Max Ratio` <- fmt_ratio(out$`Max Ratio`)
  out$`Median |d Chi Sq|` <- fmt_num(out$`Median |d Chi Sq|`)
  out$`Median Max |d Par|` <- fmt_num(out$`Median Max |d Par|`)
  knitr::kable(out, row.names = FALSE)
}
```

```{r}
#| fig-width: 8
#| fig-height: 5
if (nrow(pair_ok)) {
  plot_df <- pair_ok
  plot_df$nt_ms <- 1000 * plot_df$nt_elapsed_sec
  ggplot(plot_df, aes(nt_ms, elapsed_ratio_estimator_over_nt,
                      color = source, shape = estimator)) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
    scale_x_log10() +
    scale_y_log10(breaks = ratio_breaks) +
    labs(x = "NT Wall Time (ms, log scale)",
         y = "Timing Ratio vs NT (log scale)",
         color = "Subcorpus",
         shape = "Estimator",
         title = "Outliers Against NT Fit Cost") +
    theme(legend.position = "bottom")
}
```

## Fit Failures And Skips

```{r}
if (nrow(problem_rows)) {
  problem_summary <- do.call(rbind, lapply(split(problem_rows, problem_rows$error), function(x) {
    data.frame(
      Status = paste(sort(unique(x$status)), collapse = ";"),
      Cases = length(unique(x$case_id)),
      Rows = nrow(x),
      Error = x$error[[1L]],
      stringsAsFactors = FALSE
    )
  }))
  problem_summary <- problem_summary[order(-problem_summary$Rows, problem_summary$Error), ]
  knitr::kable(head(problem_summary, 20L), row.names = FALSE)

  examples <- unique(problem_rows[, c("case_id", "label", "source",
                                      "primary_tag", "status", "error")])
  examples <- examples[order(examples$source, examples$case_id), ]
  names(examples) <- c("Case", "Label", "Subcorpus", "Tag", "Status", "Error")
  knitr::kable(head(examples, 40L), row.names = FALSE)
} else {
  knitr::kable(data.frame(Note = "No skipped or failed fit rows."))
}
```
