---
title: "Deng & Chan (2017) alpha vs omega: a non-regular Wald test, fixed with Imhof"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results <- "results"
need <- c("nonregularity.csv", "validation_identity.csv", "validation_se.csv",
          "demo.csv", "simulation_summary.csv", "metadata.csv")
missing <- need[!file.exists(file.path(results, need))]
if (length(missing)) {
  stop("missing results: ", paste(missing, collapse = ", "),
       "\nRun:  Rscript run_experiment.R")
}
nonreg <- read.csv(file.path(results, "nonregularity.csv"))
ident  <- read.csv(file.path(results, "validation_identity.csv"))
seval  <- read.csv(file.path(results, "validation_se.csv"))
demo   <- read.csv(file.path(results, "demo.csv"))
sim    <- read.csv(file.path(results, "simulation_summary.csv"))
meta   <- read.csv(file.path(results, "metadata.csv"))
has_equal_wald <- "reject_wald_equal" %in% names(sim)
has_demo_equal_wald <- "p_wald_equal" %in% names(demo)
mv <- function(k) meta$value[match(k, meta$key)]
cellr <- function(pop, dist, N, col) {
  r <- sim[sim$population == pop & sim$dist == dist & sim$N == N, ]
  r[[col]][1]
}
pct <- function(x) sprintf("%.1f%%", 100 * x)
trinity_tau500 <- if (has_equal_wald) {
  paste0(pct(cellr("tau","normal",500,"reject_wald_equal")), ", ",
         pct(cellr("tau","normal",500,"reject_satorra_scaled")), ", and ",
         pct(cellr("tau","normal",500,"reject_score")))
} else {
  paste0(pct(cellr("tau","normal",500,"reject_satorra_scaled")), " and ",
         pct(cellr("tau","normal",500,"reject_score")),
         " for LR and score; rerun the experiment to populate the new regular Wald column")
}
```

## Question

Coefficient alpha equals coefficient omega exactly when a scale is
tau-equivalent (equal loadings), so testing $\alpha=\omega$ is testing
tau-equivalence within a one-factor model. Deng & Chan (2017) test it with a
Wald z on the reliability gap $\hat\omega-\hat\alpha$. Is that test valid; if
not, what is the right way to test the gap; and how do the regular tests of the
same null compare?

## Short answer

**The Wald z is not valid at the null it targets.** Because $\omega\ge\alpha$
with equality only at equal loadings, the contrast $\omega-\alpha$ sits at an
*interior minimum* under tau-equivalence — not a parameter boundary, but a point
where its gradient vanishes. The first-order delta method then collapses and
$\hat\omega-\hat\alpha$ becomes a **second-order** statistic: when $N$
quadruples its SD quarters (1/N), whereas each coefficient's SD halves
(1/$\sqrt N$); the two estimates become perfectly correlated
(`r sprintf("%.4f", nonreg$corr_omega_alpha[nrow(nonreg)])`) and the difference
stays right-skewed. So the z-test under-rejects badly — Type I error
`r pct(cellr("tau","normal",250,"reject_wald_sw"))` at $N=250$,
`r pct(cellr("tau","normal",500,"reject_wald_sw"))` at $N=500$, against a
nominal 5%. (Derivation: `notes/alpha_omega_second_order.qmd`.)

**An Imhof reference fixes it.** Keep Deng & Chan's interpretable statistic
$2n(\hat\omega-\hat\alpha)$, but reference it against its true law — a weighted
sum of $\chi^2_1$ — via Imhof instead of $N(0,1)$. Type I error returns toward
nominal (`r pct(cellr("tau","normal",250,"reject_imhof_sw"))` →
`r pct(cellr("tau","normal",500,"reject_imhof_sw"))` across $N$) and power rises
above the Wald test's.

**The regular trinity behaves.** The structural tests of the same null — the
normal-theory equal-loading Wald, Satorra-2000 scaled LR, and standard score
test, all proper $(p-1)$-df tests — are near nominal under tau-equivalence
(`r trinity_tau500` at $N=500$) and fully powered.
Under one-factor *misspecification* the reliability-gap tests and the structural
tests diverge: the structural tests increasingly reject (Satorra
`r pct(cellr("misspecified","normal",500,"reject_satorra_scaled"))` at $N=500$)
while the gap tests stay low — they answer different questions.

Everything is reachable with no new magmaan core: magmaan supplies the ML fits
and the moment-covariance matrices, the Imhof tail is a standard quadratic-form
computation, and Cronbach's alpha falls out exactly as the omega of a ULS
tau-fit (agreement `r sprintf("%.0e", ident$abs_diff[1])`).

## Evidence

### The contrast is second-order at the null

```{r diagnostic}
d <- nonreg[, c("N", "sd_diff", "sd_omega", "sd_diff_ratio", "sd_omega_ratio",
                "corr_omega_alpha")]
names(d) <- c("N", "SD(w-a)", "SD(w)", "SD(w-a) ratio", "SD(w) ratio",
              "corr(w,a)")
knitr::kable(d, digits = c(0, 5, 4, 2, 2, 4), row.names = FALSE,
  caption = "Tau-equivalent population. SD(omega-alpha) quarters when N quadruples (1/N, second-order) while SD(omega) halves (1/sqrt(N), regular). corr(omega,alpha) -> 1: the two estimates share one influence function at the null.")
```

The `SD(w-a) ratio` near 0.25 versus 0.50 for `SD(w)` is the fingerprint of a
second-order statistic — the reason a standard-normal Wald reference cannot be
right here.

### An Imhof reference restores calibration; the trinity behaves

```{r simulation}
sim_cols <- c("population", "N", "reject_wald_sw", "reject_imhof_sw")
if (has_equal_wald) sim_cols <- c(sim_cols, "reject_wald_equal")
sim_cols <- c(sim_cols, "reject_satorra_scaled", "reject_score")
tab <- sim[sim$dist == "normal", sim_cols]
tab <- tab[order(match(tab$population, c("tau","congeneric","misspecified")), tab$N), ]
names(tab) <- c("Population", "N", "Deng Wald z", "Imhof gap",
                if (has_equal_wald) "Equal-loading Wald",
                "Satorra LR", "Score")
knitr::kable(tab, digits = 3, row.names = FALSE,
  caption = "Rejection rate at 5%, normal data. Tau = Type I error (Deng Wald conservative; Imhof restored; structural tests nominal); congeneric = power; misspecified = divergence. Non-normal rates are in results/simulation_summary.csv.")
```

Reading the tau rows: the Wald z is stuck well below 5% and barely improves with
$N$ (it is converging to the wrong law), while the Imhof test climbs toward
nominal and the regular equal-loading Wald, Satorra LR, and score tests are
already there. The congeneric rows show all five are powerful against genuinely
unequal loadings, Imhof more so than the Deng Wald z. The misspecified rows show
the split: with a minor second factor the structural tests chase the model error
while the reliability-gap tests do not (the population gap is even *negative*
there — the extra factor lifts the model-free alpha above the one-factor omega).

### Implementation checks and real scales

```{r checks}
v <- data.frame(
  Check = c("alpha(S) == omega of ULS tau-fit",
            paste0("Deng-Chan sandwich SE vs bootstrap (", seval$dataset, ")")),
  Value = c(sprintf("|diff| = %.1e", ident$abs_diff[1]),
            sprintf("ratio = %.3f", seval$ratio_sw_boot)))
knitr::kable(v, caption = "Cronbach's alpha is exactly the ULS tau-equivalent omega, and the first-order sandwich SE matches a nonparametric bootstrap away from the null.")
```

```{r demo}
demo_cols <- c("scale", "p", "omega", "alpha", "diff", "p_imhof_sw")
if (has_demo_equal_wald) demo_cols <- c(demo_cols, "p_wald_equal")
demo_cols <- c(demo_cols, "p_satorra_scaled", "p_score")
dd <- demo[, demo_cols]
names(dd) <- c("Scale", "p", "Omega", "Alpha", "w - a", "Imhof p",
               if (has_demo_equal_wald) "Equal-loading Wald p",
               "Satorra p", "Score p")
demo_digits <- c(0,0,3,3,4,3, if (has_demo_equal_wald) 4, 4, 3)
knitr::kable(dd, digits = demo_digits, row.names = FALSE,
  caption = "Holzinger-Swineford subscales. On textual and speed the structural tests flag non-tau-equivalence while the reliability gap is a few thousandths; visual+textual is deliberately two-dimensional.")
```

## Proposed Expanded Study

The next simulation pass should separate three questions that are easy to
confound. First, exact tau-equivalent populations test calibration at the null,
crossing item count, sample size, and reliability level. Second, congeneric
one-factor populations test power against unequal loadings while the covariance
model remains correct. Third, minor-factor and correlated-residual populations
ask what happens when the fitted one-factor model is false; these cells are not
just stronger congeneric alternatives, because the structural tests and the
scalar reliability-gap tests answer different questions under misspecification.

The R support code now sketches this design in
`R/populations.R`. `proposed_study_grid()` returns a candidate grid over
$p\in\{4,6,12\}$, $N\in\{100,250,500,1000\}$, tau reliabilities
$.55/.70/.85$, several congeneric loading patterns, and distribution regimes
`normal`, `latent_chisq`, `error_chisq`, `all_chisq`, and
`paired_quadratic_residuals`. The paired-quadratic residual condition keeps the
SEM covariance model tau-equivalent or congeneric, but makes residuals
nonlinearly dependent within pairs while preserving zero residual covariance;
it is a fourth-moment stress test, not a correlated-residual misspecification.

For motivation, `alpha_avar_components()` computes the first-order
delta-method variance decomposition for coefficient alpha: normal-theory,
latent-kurtosis, residual-kurtosis, and paired-dependence increments. This lets
the paper report whether nonnormality matters through the common factor, the
residuals, or nonlinear residual dependence before relying on Monte Carlo
rejection rates.

## Caveats

- **What this establishes.** The non-regularity (diagnostic + derivation), an
  Imhof calibration of the reliability gap, and a finite-sample comparison of
  the gap tests against the regular trinity at $p=6$, $N\in\{250,500\}$,
  `r mv("reps")` replications per cell. It does not reproduce Deng & Chan's
  published tables: their data sets (a 26-item Holzinger-Swineford battery,
  FACES II, etc.) are not redistributable, so the real-data section uses
  lavaan's nine-item Holzinger-Swineford, whose subscales are three indicators
  each (a just-identified one-factor fit).
- **Not a boundary problem.** The tau-equivalent point is interior in the SEM
  parameterization; the singularity is in the *derived scalar contrast*, whose
  first derivative vanishes there. See `notes/alpha_omega_second_order.qmd`.
- **The score test is normal-theory only.** lavaan has no robust release-score
  test (it falls back to the ordinary one), and the equal-loading Wald column is
  the ordinary Wald test. Under non-normality only the sandwich Imhof and the
  empirical-$\Gamma$ Satorra LR are formally justified; the regular Wald and
  score columns are normal-theory references.
- **Estimation is traditional ML** throughout; non-normality is centred-scaled
  chi-square(5) components (closed-form covariance, no moment-matching).
- Rejection rates are point estimates; replicate counts are in the metadata.

## Reproduce

```sh
# full run (diagnostic + validation + real-data demo + simulation)
Rscript run_experiment.R

# fast smoke path
Rscript run_experiment.R --smoke

# a single slice
Rscript run_experiment.R --cells pop=tau,dist=normal --reps 2000
```
