---
title: "Deng & Chan (2017): a reliability-difference Wald test vs a Satorra-2000 LRT"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results <- "results"
need <- c("validation_identity.csv", "validation_se.csv", "demo.csv",
          "simulation_summary.csv", "metadata.csv")
missing <- need[!file.exists(file.path(results, need))]
if (length(missing)) {
  stop("missing results: ", paste(missing, collapse = ", "),
       "\nRun:  Rscript run_experiment.R")
}
ident <- read.csv(file.path(results, "validation_identity.csv"))
seval <- read.csv(file.path(results, "validation_se.csv"))
demo  <- read.csv(file.path(results, "demo.csv"))
sim   <- read.csv(file.path(results, "simulation_summary.csv"))
meta  <- read.csv(file.path(results, "metadata.csv"))
mv <- function(k) meta$value[match(k, meta$key)]

cellr <- function(pop, dist, N, col) {
  r <- sim[sim$population == pop & sim$dist == dist & sim$N == N, ]
  r[[col]][1]
}
pct <- function(x) sprintf("%.1f%%", 100 * x)
```

## Question

Coefficient alpha equals coefficient omega exactly when a scale's items are
tau-equivalent (equal loadings); otherwise alpha understates reliability. So
"is alpha different from omega?" and "are the loadings equal?" are the same null
hypothesis whenever a one-factor model holds. Deng & Chan (2017) test it one
way — a Wald z-test on the reliability difference $\hat\omega-\hat\alpha$ with a
delta-method standard error. magmaan already ships the other way — a
Satorra-2000 scaled likelihood-ratio test of the equal-loading restriction. Is
the Deng & Chan procedure reachable with today's magmaan, and how do the two
tests compare on calibration, power, and behaviour when the one-factor model is
wrong?

## Short answer

Reachable with no new magmaan core. The Satorra-2000 side is a single
`nestedTest()` call; the Deng & Chan side is a ~30-line delta method over
magmaan's one-factor fit and its $\Gamma$ helpers. Two checks say the
implementation is right: magmaan recovers Cronbach's alpha **exactly** as the
omega of a ULS-fitted tau-equivalent model (agreement
`r sprintf("%.0e", ident$abs_diff[1])`), and the analytic sandwich SE of
$\hat\omega-\hat\alpha$ matches a nonparametric bootstrap to within ~6%.

On the substance, the two tests agree only when a one-factor model holds, and
even then they are not interchangeable:

- **Tau-equivalent population (Type I error).** The Satorra scaled LRT sits near
  the nominal 5% (`r pct(cellr("tau","normal",500,"reject_satorra_scaled"))` at
  $N=500$); the Deng & Chan Wald is markedly **conservative**
  (`r pct(cellr("tau","normal",500,"reject_wald_sw"))`). Testing a difference
  that cannot be negative in the population pushes the two-sided z-test below
  nominal.
- **Congeneric population (power).** Both reject unequal loadings with high
  power even though the reliability gap is tiny ($\omega-\alpha\approx0.007$):
  Wald `r pct(cellr("congeneric","normal",250,"reject_wald_sw"))` and Satorra
  `r pct(cellr("congeneric","normal",250,"reject_satorra_scaled"))` at $N=250$.
  Deng & Chan's "the difference is minor in applications" is about its *size*,
  not its detectability.
- **Misspecified population (divergence).** With a minor second factor breaking
  unidimensionality the tests come apart: the Satorra LRT increasingly rejects
  (`r pct(cellr("misspecified","normal",250,"reject_satorra_scaled"))` at
  $N=250$ rising to
  `r pct(cellr("misspecified","normal",500,"reject_satorra_scaled"))` at
  $N=500$), while the reliability-difference Wald stays near nominal
  (`r pct(cellr("misspecified","normal",500,"reject_wald_sw"))`). They answer
  different questions: Satorra asks whether the loadings are equal *within* a
  one-factor model; Deng & Chan compare a model-based omega against a model-free
  alpha, two numbers that can track each other even when the model is wrong.

The same split appears on real Holzinger-Swineford subscales below.

## Evidence

### The implementation is correct

```{r validation}
v <- data.frame(
  Check = c("alpha(S) == omega of ULS tau-fit",
            paste0("sandwich SE vs bootstrap (", seval$dataset, ")")),
  Value = c(sprintf("|diff| = %.1e", ident$abs_diff[1]),
            sprintf("ratio = %.3f", seval$ratio_sw_boot)),
  stringsAsFactors = FALSE
)
knitr::kable(v, caption = "Validation: the model-free alpha is exactly the ULS tau-equivalent omega, and the Deng-Chan sandwich SE of the difference matches a nonparametric bootstrap.")
```

The first row is the reason alpha is not a magmaan primitive: it is the omega of
a *least-squares*-fitted tau model, not the ML one, and otherwise just a closed
form on the sample covariance.

### Calibration, power, and divergence

```{r simulation}
tab <- sim[sim$dist == "normal",
           c("population", "N", "gap", "reject_wald_sw", "reject_satorra_scaled")]
tab <- tab[order(match(tab$population, c("tau","congeneric","misspecified")), tab$N), ]
names(tab) <- c("Population", "N", "Pop. gap (w-a)",
                "Wald (sandwich)", "Satorra (scaled)")
rownames(tab) <- NULL
knitr::kable(tab, digits = 3, row.names = FALSE,
  caption = "Rejection rate at the 5% level, normal data. Tau row = Type I error; congeneric = power; misspecified = divergence. Non-normal data (in results/) is comparable, with the Satorra test slightly liberal under tau.")
```

The tau rows show the calibration split; the congeneric rows show both tests are
powerful against genuine inequality; the misspecified rows show the divergence
growing with $N$ — the structural LRT chases the model error while the
reliability contrast does not. Note the misspecified population gap is
*negative*: the minor factor inflates the model-free alpha above the one-factor
omega, so $\hat\omega-\hat\alpha$ has little to detect.

### The same pattern on real scales

```{r demo}
d <- demo
names(d) <- c("Scale", "p", "Omega", "Alpha", "w - a", "SE (sw)",
              "Wald p (sw)", "Satorra p")
knitr::kable(d, digits = c(0,0,3,3,4,4,3,4), row.names = FALSE,
  caption = "Holzinger-Swineford subscales. On textual and speed the Satorra LRT flags non-tau-equivalence while the Wald test does not; on the deliberately two-dimensional visual+textual block the split is extreme.")
```

For *textual* and *speed* the reliability gap is a few thousandths and the Wald
test is silent ($p>0.15$), yet the Satorra LRT rejects equal loadings
($p<0.03$). The `visual+textual` block is genuinely two factors: the Satorra LRT
rejects overwhelmingly while the Wald test sees a trivial reliability gap and
shrugs.

## Caveats

- **What this establishes.** A correctness check on the Deng & Chan SE (identity
  + bootstrap) and a finite-sample comparison of the two tests at $p=6$,
  $N\in\{250,500\}$, `r mv("reps")` replications per cell. It does not reproduce
  Deng & Chan's published tables: their data sets (a 26-item Holzinger-Swineford
  battery, FACES II, etc.) are not redistributable, so the real-data section
  uses lavaan's nine-item Holzinger-Swineford, whose subscales are only
  three indicators each (a just-identified one-factor fit).
- **Why the Wald test is conservative.** Under tau-equivalence the first-order
  influence functions of $\hat\omega$ and $\hat\alpha$ coincide, so
  $\hat\omega-\hat\alpha$ is a second-order statistic with a quadratic-form
  null law rather than a normal law. A standard two-sided z reference therefore
  under-rejects; see `notes/alpha_omega_second_order.qmd`.
- **Non-normality** here is centred-scaled chi-square(5) components (closed-form
  covariance, no moment-matching). The sandwich Wald and the empirical-$\Gamma$
  Satorra test both stay usable; rates are in `results/simulation_summary.csv`.
- Rejection rates are point estimates; replicate counts are in the metadata.

## Reproduce

```sh
# full run (validation + real-data demo + simulation)
Rscript run_experiment.R

# fast smoke path
Rscript run_experiment.R --smoke

# a single slice
Rscript run_experiment.R --cells pop=misspecified,dist=normal --reps 2000
```
