---
title: "Which information matrix produced the Savalei–Falk two-stage test?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(if (nzchar(override)) override else character(),
           file.path("results", c("replication", "pilot", "smoke")))
available <- roots[vapply(roots, function(path)
  file.exists(file.path(path, "metadata.csv")), logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
summary <- read_result("summary.csv")
design <- read_result("design.csv")
audit_frame <- read_result("audit.csv")
audit <- setNames(audit_frame$value, audit_frame$key)
meta_frame <- read_result("metadata.csv")
meta <- setNames(meta_frame$value, meta_frame$key)
reps <- as.integer(meta[["reps_per_cell"]])
profile <- meta[["profile"]]
labels <- c(
  saturated_expected = "Saturated plug-in (current ML2S)",
  saturated_observed = "Saturated observed",
  structured_expected = "Structured expected",
  structured_observed = "Structured observed (paper wording)")
k7 <- summary[summary$kurtosis == 7, , drop = FALSE]
k7$Method <- unname(labels[k7$information])
k7$Rejection <- 100 * k7$rejection_rate
k7$MC_SE <- 100 * k7$rejection_se
k7$Scale <- k7$mean_scaling_factor
k7 <- k7[match(names(labels), k7$information), ]
structured_observed_rate <- k7$rejection_rate[
  k7$information == "structured_observed"]
paper_se <- sqrt(0.10 * 0.90 / 1000)
combined_z <- abs(structured_observed_rate - 0.10) /
  sqrt(k7$rejection_se[k7$information == "structured_observed"]^2 +
         paper_se^2)
eligible <- 100 * design$theoretical_eligible_proportion[
  design$kurtosis == 7]
identity_error <- audit[["saturated_identity_max_abs"]]
default_error <- audit[["default_scale_max_abs"]]
indef <- aggregate(
  cbind(H = summary$information_indefinite_rate,
        U = summary$projector_indefinite_rate) ~ information,
  data = summary, mean)
indef$Method <- unname(labels[indef$information])
indef$H <- 100 * indef$H
indef$U <- 100 * indef$U
```

## Question

Savalei and Falk describe the two-stage correction with the complete-data
information matrix of the saturated moment model, while also saying that
observed information was used throughout. Did their reported statistic use the
saturated plug-in, structured expected information, or structured observed
information?

## Short answer

`r if (reps >= 1000) sprintf("The evidence identifies **structured observed information** as the paper's implementation. It gives %.1f%% rejection in the diagnostic cell versus the published 10.0%%; the %.1f-point difference is %.1f combined Monte Carlo standard errors. The saturated/current-ML2S and structured-expected alternatives are nowhere close.", 100 * structured_observed_rate, 100 * abs(structured_observed_rate - .10), combined_z) else sprintf("The %s run (%d samples per kurtosis cell) validates the crossed-information pipeline, but is too small to identify the paper's implementation from its 10.0%% rejection-rate target. A 1,000-replication run is required.", profile, reps)`

The equation-level checks passed: complete-data expected and observed
information agreed at the saturated estimate to
`r format(identity_error, digits = 3, scientific = TRUE)`, and the saturated
expected statistic reproduced magmaan's current ML2S scale to
`r format(default_error, digits = 3, scientific = TRUE)`.
The matching structured-observed matrix is indefinite in
`r sprintf("%.2f%%", 100 * indef$H[indef$information == "structured_observed"])`
of fitted samples, confirming that this historical choice has no
positive-semidefiniteness guarantee.

## Evidence

The focused cell is Savalei–Falk Model 2 with (N=200), skewness 2, kurtosis 7,
30% missingness, many patterns, and MAR-L deletion. The article reports a 10.0%
two-stage rejection rate there. Each row below is computed from the same fitted
sample; only the $H$ used in

\[
U=H-H\Delta(\Delta^\top H\Delta)^{-1}\Delta^\top H.
\]

changes.

```{r result-table}
knitr::kable(
  k7[, c("Method", "Rejection", "MC_SE", "Scale")],
  digits = c(NA, 1, 1, 3), row.names = FALSE,
  col.names = c("Information choice", "Reject (%)", "MC SE (points)",
                "Mean scale"),
  caption = "Model 2, N=200, MAR-L, MP, 30% missing, skewness 2, kurtosis 7. Published TS rejection: 10.0%.")
```

The saturated expected and saturated observed rows are an intentional identity
check: at the unrestricted complete-data moment estimate, observed and expected
information coincide. Structured observed information is evaluated at the
Stage-2 model-implied moments while retaining the saturated EM moments as the
pseudo-data. That point does not minimize the complete-data saturated
likelihood over unrestricted moments, so its observed Hessian need not be
positive semidefinite; the stability table shows that it generally is not.

```{r stability-table}
knitr::kable(
  indef[, c("Method", "H", "U")],
  digits = c(NA, 1, 1), row.names = FALSE,
  col.names = c("Information choice", "Indefinite H (%)",
                "Indefinite U (%)"),
  caption = "Finite-sample geometry diagnostics across both kurtosis cells.")
```

## Caveats

This is a targeted implementation-identification experiment, not a full
replication of the 192-cell study. The original random seeds and EQS random
number stream are unavailable, and the published 10.0% is itself a
1,000-replication Monte Carlo estimate. A difference smaller than roughly one
percentage point is not compelling identification evidence.

The generator uses the conventional monotone-increasing Fleishman root in
magmaan's Vale–Maurelli implementation and treats the paper's kurtosis values
as excess kurtosis. For the moderate condition this gives
`r sprintf("%.1f%%", eligible)` theoretically eligible observations under the
paper's positive-value MAR-L rule, matching the article's approximately 38%.
The missingness algorithm reaches exactly 30% missingness in each of the six
published deletion sets, conditional on the corresponding variable being
positive. Candidate samples in which a conditioning variable has fewer than
60 positive observations cannot satisfy that instruction and are regenerated;
this implicit edge case cost
`r sum(design$eligibility_rejections)` candidate samples across the two cells.
The compact replication summaries and equation-level audit are frozen with the
experiment so this report remains renderable; raw replication rows are local
generated output.

## Reproduce

```bash
cd experiments/80-savalei-falk-2014-ml2s-information

# Structural and identity check
Rscript run_experiment.R --smoke

# Identification run matching the paper's 1,000 replications per cell
Rscript run_experiment.R --replication --cores 8

quarto render report.qmd
```

## Mathematical definition of the four choices

Let $\widehat\beta=(\widehat\mu,\operatorname{vech}\widehat\Sigma)$ be the
Stage-1 saturated FIML estimate and let
$\widetilde\beta=\beta(\widetilde\theta)$ be the Stage-2 structured moments.
The four rows use

\[
\begin{aligned}
H_{SE}&=H_{\mathrm{exp}}(\widehat\beta), &
H_{SO}&=H_{\mathrm{obs}}(\widehat\beta;\widehat\beta),\\
H_{XE}&=H_{\mathrm{exp}}(\widetilde\beta), &
H_{XO}&=H_{\mathrm{obs}}(\widetilde\beta;\widehat\beta).
\end{aligned}
\]

The semicolon separates the point where the complete-data saturated likelihood
is differentiated from the pseudo-data moments entering that likelihood. In
every row, $\Omega=G_1^{-1}J_1G_1^{-1}$ is the same empirical Stage-1 FIML
sandwich, and $\Delta=\partial\beta(\theta)/\partial\theta^\top$ is the same
Stage-2 Jacobian. Thus only $H$ changes. At the saturated estimate,
$H_{SE}=H_{SO}$ exactly.
