---
title: "Which matrix choices produced the Savalei–Falk robust tests?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(if (nzchar(override)) override else character(),
           file.path("results", c("replication", "pilot", "smoke")))
available <- roots[vapply(roots, function(path)
  file.exists(file.path(path, "metadata.csv")), logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
summary <- read_result("summary.csv")
design <- read_result("design.csv")
audit_frame <- read_result("audit.csv")
audit <- setNames(audit_frame$value, audit_frame$key)
dgp_roots <- read_result("dgp_roots.csv")
meta_frame <- read_result("metadata.csv")
meta <- setNames(meta_frame$value, meta_frame$key)
reps <- as.integer(meta[["reps_per_cell"]])
profile <- meta[["profile"]]

fingerprint_override <- Sys.getenv(
  "MAGMAAN_EXPERIMENT_FINGERPRINT_RESULTS", unset = "")
fingerprint_root <- if (nzchar(fingerprint_override)) {
  fingerprint_override
} else file.path("results", "fingerprint")
if (!file.exists(file.path(fingerprint_root, "metadata.csv"))) {
  stop(paste(
    "missing fingerprint results; run",
    "`Rscript run_experiment.R --fingerprint --cores 8 --resume`"))
}
read_fingerprint <- function(name) read.csv(
  file.path(fingerprint_root, name), stringsAsFactors = FALSE,
  check.names = FALSE)
fingerprint_summary <- read_fingerprint("summary.csv")
fingerprint_design <- read_fingerprint("design.csv")
fingerprint_audit_frame <- read_fingerprint("audit.csv")
fingerprint_audit <- setNames(
  fingerprint_audit_frame$value, fingerprint_audit_frame$key)
k7 <- summary[summary$cell_id == "paper-mar-l-k7", , drop = FALSE]

one <- function(x, label) {
  if (nrow(x) != 1L) stop("expected one summary row for ", label)
  x
}
ml2s_default <- one(k7[
  k7$method == "ml2s" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "observed" &
    k7$omega_meat_point == "saturated" &
    k7$stage2_information == "saturated_expected", ], "ML2S default")
ml2s_source <- one(k7[
  k7$method == "ml2s" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "observed" &
    k7$omega_meat_point == "saturated" &
    k7$stage2_information == "structured_observed", ], "ML2S source row")
fiml_default <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "saturated_observed_h1" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "observed" &
    k7$omega_meat_point == "saturated", ], "FIML H1 default")
fiml_literal <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "structured_observed_h1" &
    k7$omega_bread_point == "structured" &
    k7$omega_bread_kind == "observed" &
    k7$omega_meat_point == "structured", ], "literal FIML equations row")

targets <- c(ml2s = .100, fiml = .633)
closest <- lapply(names(targets), function(method) {
  z <- k7[k7$method == method, , drop = FALSE]
  z[which.min(abs(z$rejection_rate - targets[[method]])), , drop = FALSE]
})
names(closest) <- names(targets)
ml2s_target_se <- sqrt(.1 * .9 / 1000)
fiml_target_se <- sqrt(.633 * .367 / 1000)
ml2s_z <- abs(ml2s_source$rejection_rate_n_minus_1 - .1) /
  sqrt(ml2s_source$rejection_se_n_minus_1^2 + ml2s_target_se^2)
fiml_closest <- closest$fiml
fiml_expected <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "saturated_expected_h1" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "expected" &
    k7$omega_meat_point == "saturated", ], "FIML closest row")
stopifnot(identical(fiml_closest$information, fiml_expected$information))
fiml_z <- abs(fiml_closest$rejection_rate - .633) /
  sqrt(fiml_closest$rejection_se^2 + fiml_target_se^2)
fiml_observed_residual <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "saturated_observed_h1" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "expected" &
    k7$omega_meat_point == "saturated", ], "FIML observed-residual alternative")
fiml_mlr <- one(k7[k7$method == "fiml_mlr", ], "named FIML MLR")
mlr_z <- abs(fiml_mlr$rejection_rate - .633) /
  sqrt(fiml_mlr$rejection_se^2 + fiml_target_se^2)

fingerprint_fiml <- function(residual, bread_point, bread_kind, meat_point,
                             label) {
  z <- fingerprint_summary[
    fingerprint_summary$method == "fiml" &
      fingerprint_summary$residual_information == residual &
      fingerprint_summary$omega_bread_point == bread_point &
      fingerprint_summary$omega_bread_kind == bread_kind &
      fingerprint_summary$omega_meat_point == meat_point, , drop = FALSE]
  if (nrow(z) != nrow(fingerprint_design)) {
    stop("incomplete FIML fingerprint route: ", label)
  }
  z
}
fingerprint_ml2s <- function(stage2_information, label) {
  z <- fingerprint_summary[
    fingerprint_summary$method == "ml2s" &
      fingerprint_summary$omega_bread_point == "saturated" &
      fingerprint_summary$omega_bread_kind == "observed" &
      fingerprint_summary$omega_meat_point == "saturated" &
      fingerprint_summary$stage2_information == stage2_information,
    , drop = FALSE]
  if (nrow(z) != nrow(fingerprint_design)) {
    stop("incomplete ML2S fingerprint route: ", label)
  }
  z
}
rate_by_cell <- function(z, column = "rejection_rate") {
  unname(setNames(z[[column]], z$cell_id)[fingerprint_design$cell_id])
}
rmse_points <- function(estimate, target) {
  100 * sqrt(mean((estimate - target)^2))
}
mae_points <- function(estimate, target) {
  100 * mean(abs(estimate - target))
}

yb_fiml <- fingerprint_fiml(
  "structured_expected_h1", "saturated", "expected", "saturated",
  "Yuan-Bentler expected route")
all_h1_expected_fiml <- fingerprint_fiml(
  "saturated_expected_h1", "saturated", "expected", "saturated",
  "all-H1 expected route")
literal_fiml_fingerprint <- fingerprint_fiml(
  "structured_observed_h1", "structured", "observed", "structured",
  "paper-literal route")
source_ml2s_fingerprint <- fingerprint_ml2s(
  "structured_observed", "source-implied route")

published_fiml <- fingerprint_design$paper_fiml_rejection
published_ml2s <- fingerprint_design$paper_ts_rejection
yb_fiml_rate <- rate_by_cell(yb_fiml)
all_h1_expected_rate <- rate_by_cell(all_h1_expected_fiml)
literal_fiml_rate <- rate_by_cell(literal_fiml_fingerprint)
ml2s_n_rate <- rate_by_cell(source_ml2s_fingerprint)
ml2s_n_minus_1_rate <- rate_by_cell(
  source_ml2s_fingerprint, "rejection_rate_n_minus_1")

fingerprint_table <- data.frame(
  N = fingerprint_design$n,
  Kurtosis = fingerprint_design$kurtosis,
  `Published FIML` = 100 * published_fiml,
  `YB expected FIML` = 100 * yb_fiml_rate,
  `All-H1 expected FIML` = 100 * all_h1_expected_rate,
  `Literal FIML` = 100 * literal_fiml_rate,
  `Published 2S` = 100 * published_ml2s,
  `2S (n-1)F/c` = 100 * ml2s_n_minus_1_rate,
  `2S nF/c` = 100 * ml2s_n_rate,
  check.names = FALSE)

fingerprint_metrics <- data.frame(
  Route = c("FIML: YB expected", "FIML: all-H1 expected",
            "FIML: literal structured observed",
            "2S: (n-1)F/c", "2S: nF/c"),
  RMSE = c(
    rmse_points(yb_fiml_rate, published_fiml),
    rmse_points(all_h1_expected_rate, published_fiml),
    rmse_points(literal_fiml_rate, published_fiml),
    rmse_points(ml2s_n_minus_1_rate, published_ml2s),
    rmse_points(ml2s_n_rate, published_ml2s)),
  MAE = c(
    mae_points(yb_fiml_rate, published_fiml),
    mae_points(all_h1_expected_rate, published_fiml),
    mae_points(literal_fiml_rate, published_fiml),
    mae_points(ml2s_n_minus_1_rate, published_ml2s),
    mae_points(ml2s_n_rate, published_ml2s)))

yb_mar <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "structured_expected_h1" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "expected" &
    k7$omega_meat_point == "saturated", ], "YB expected MAR row")
seven_fiml_targets <- c(published_fiml, .633)
seven_yb_rmse <- rmse_points(
  c(yb_fiml_rate, yb_mar$rejection_rate), seven_fiml_targets)
seven_all_h1_rmse <- rmse_points(
  c(all_h1_expected_rate, fiml_expected$rejection_rate),
  seven_fiml_targets)

evidence <- rbind(ml2s_source, ml2s_default, fiml_closest,
                  fiml_observed_residual, fiml_literal, fiml_default,
                  fiml_mlr)
evidence$Choice <- c("Source-implied candidate", "Current magmaan default",
                     "Closest: saturated expected throughout",
                     "Observed residual, expected sandwich",
                     "Literal all-structured observed equations",
                     "Current magmaan H1 metric",
                     "Named MLR: Yuan-Bentler/Mplus")
evidence$Method <- c("Two-stage", "Two-stage", rep("Direct FIML", 5L))
evidence$Published <- c(10.0, NA, 63.3, 63.3, 63.3, NA, 63.3)
evidence$Usable <- 100 * evidence$replications / reps
evidence$Rejection <- 100 * evidence$rejection_rate
evidence$MC_SE <- 100 * evidence$rejection_se
evidence$Rejection[[1L]] <- 100 * ml2s_source$rejection_rate_n_minus_1
evidence$MC_SE[[1L]] <- 100 * ml2s_source$rejection_se_n_minus_1
evidence$Scale <- evidence$mean_scaling_factor

eligible <- 100 * design$theoretical_eligible_proportion[
  design$cell_id == "paper-mar-l-k7"]
identity_error <- audit[["saturated_identity_max_abs"]]
default_error <- audit[["default_scale_max_abs"]]

point_labels <- c(saturated = "H₁", structured = "H₀")
kind_labels <- c(expected = "Expected (E)", observed = "Observed (O)")
residual_outer_point <- c(
  saturated_expected_h1 = "H₁",
  saturated_observed_h1 = "H₁",
  structured_expected_h1 = "H₀",
  structured_observed_h1 = "H₀",
  structured_observed_hessian = "H₀",
  mixed_saturated_observed_hessian = "H₁")
residual_outer_kind <- c(
  saturated_expected_h1 = "Expected (E)",
  saturated_observed_h1 = "Observed (O)",
  structured_expected_h1 = "Expected (E)",
  structured_observed_h1 = "Observed (O)",
  structured_observed_hessian = "Observed (O)",
  mixed_saturated_observed_hessian = "Observed (O)")
residual_theta_block <- c(
  saturated_expected_h1 = "Δ′AΔ",
  saturated_observed_h1 = "Δ′AΔ",
  structured_expected_h1 = "Δ′AΔ",
  structured_observed_h1 = "Δ′AΔ",
  structured_observed_hessian = "Full observed Hθθ(H₀)",
  mixed_saturated_observed_hessian = "Full observed Hθθ(H₀)")

fiml_options <- k7[k7$method == "fiml", , drop = FALSE]
fiml_options$Role <- ""
fiml_options$Role[
  fiml_options$residual_information == "structured_expected_h1" &
    fiml_options$omega_bread_point == "saturated" &
    fiml_options$omega_bread_kind == "expected" &
    fiml_options$omega_meat_point == "saturated"] <- "YB expected"
fiml_options$Role[
  fiml_options$residual_information == "saturated_expected_h1" &
    fiml_options$omega_bread_point == "saturated" &
    fiml_options$omega_bread_kind == "expected" &
    fiml_options$omega_meat_point == "saturated"] <- "Closest"
fiml_options$Role[
  fiml_options$residual_information == "structured_observed_h1" &
    fiml_options$omega_bread_point == "structured" &
    fiml_options$omega_bread_kind == "observed" &
    fiml_options$omega_meat_point == "structured"] <- "Paper-literal"
fiml_options$Role[
  fiml_options$residual_information == "saturated_observed_h1" &
    fiml_options$omega_bread_point == "saturated" &
    fiml_options$omega_bread_kind == "observed" &
    fiml_options$omega_meat_point == "saturated"] <- "magmaan default"
fiml_options$U_A_point <- unname(residual_outer_point[
  fiml_options$residual_information])
fiml_options$U_A_kind <- unname(residual_outer_kind[
  fiml_options$residual_information])
fiml_options$U_theta <- unname(residual_theta_block[
  fiml_options$residual_information])
fiml_options$Omega_A_point <- unname(point_labels[
  fiml_options$omega_bread_point])
fiml_options$Omega_A_kind <- unname(kind_labels[
  fiml_options$omega_bread_kind])
fiml_options$Omega_B_point <- unname(point_labels[
  fiml_options$omega_meat_point])
fiml_options$Invalid <- reps - fiml_options$replications
fiml_options$Reject <- 100 * fiml_options$rejection_rate
fiml_options$Gap <- fiml_options$Reject - 63.3
fiml_options$Scale <- fiml_options$mean_scaling_factor
fiml_options <- fiml_options[order(
  match(fiml_options$Role,
        c("Paper-literal", "YB expected", "Closest", "magmaan default", "")),
  match(fiml_options$U_A_point, c("H₁", "H₀")),
  match(fiml_options$U_A_kind, c("Observed (O)", "Expected (E)")),
  match(fiml_options$U_theta, c("Δ′AΔ", "Full observed Hθθ(H₀)")),
  match(fiml_options$omega_bread_point, c("saturated", "structured")),
  match(fiml_options$omega_bread_kind, c("observed", "expected")),
  match(fiml_options$omega_meat_point, c("saturated", "structured"))), ]

ml2s_options <- k7[k7$method == "ml2s", , drop = FALSE]
ml2s_options$Role <- ""
ml2s_options$Role[
  ml2s_options$omega_bread_point == "saturated" &
    ml2s_options$omega_bread_kind == "observed" &
    ml2s_options$omega_meat_point == "saturated" &
    ml2s_options$stage2_information == "structured_observed"] <-
  "Paper candidate"
ml2s_options$Role[
  ml2s_options$omega_bread_point == "saturated" &
    ml2s_options$omega_bread_kind == "observed" &
    ml2s_options$omega_meat_point == "saturated" &
    ml2s_options$stage2_information == "saturated_expected"] <-
  "magmaan default"
ml2s_options$Omega_A_point <- unname(point_labels[
  ml2s_options$omega_bread_point])
ml2s_options$Omega_A_kind <- unname(kind_labels[
  ml2s_options$omega_bread_kind])
ml2s_options$Omega_B_point <- unname(point_labels[
  ml2s_options$omega_meat_point])
ml2s_options$U_H_point <- unname(point_labels[
  sub("_.*", "", ml2s_options$stage2_information)])
ml2s_options$U_H_kind <- unname(kind_labels[
  sub(".*_", "", ml2s_options$stage2_information)])
ml2s_options$Invalid <- reps - ml2s_options$replications
ml2s_options$Reject_n_minus_1 <-
  100 * ml2s_options$rejection_rate_n_minus_1
ml2s_options$Gap_n_minus_1 <- ml2s_options$Reject_n_minus_1 - 10.0
ml2s_options$Reject_n <- 100 * ml2s_options$rejection_rate
ml2s_options$Gap_n <- ml2s_options$Reject_n - 10.0
ml2s_options$Scale <- ml2s_options$mean_scaling_factor
ml2s_options <- ml2s_options[order(
  match(ml2s_options$Role,
        c("Paper candidate", "magmaan default", "")),
  match(ml2s_options$omega_bread_point, c("saturated", "structured")),
  match(ml2s_options$omega_bread_kind, c("observed", "expected")),
  match(ml2s_options$omega_meat_point, c("saturated", "structured")),
  match(ml2s_options$stage2_information,
        c("saturated_observed", "structured_observed",
          "saturated_expected", "structured_expected"))), ]

normal <- summary[summary$cell_id == "normal-mar-l-k0", , drop = FALSE]
normal_rows <- rbind(
  one(normal[
    normal$method == "fiml" &
      normal$residual_information == "structured_observed_h1" &
      normal$omega_bread_point == "structured" &
      normal$omega_bread_kind == "observed" &
      normal$omega_meat_point == "structured", ], "normal paper-literal FIML"),
  one(normal[normal$method == "fiml_mlr", ], "normal MLR"),
  one(normal[
    normal$method == "fiml" &
      normal$residual_information == "saturated_expected_h1" &
      normal$omega_bread_point == "saturated" &
      normal$omega_bread_kind == "expected" &
      normal$omega_meat_point == "saturated", ], "normal closest FIML"),
  one(normal[
    normal$method == "ml2s" &
      normal$omega_bread_point == "saturated" &
      normal$omega_bread_kind == "observed" &
      normal$omega_meat_point == "saturated" &
      normal$stage2_information == "structured_observed", ],
      "normal paper-candidate ML2S"),
  one(normal[
    normal$method == "ml2s" &
      normal$omega_bread_point == "saturated" &
      normal$omega_bread_kind == "observed" &
      normal$omega_meat_point == "saturated" &
      normal$stage2_information == "saturated_expected", ],
      "normal default ML2S"))
normal_rows$Route <- c(
  "Direct FIML: paper-literal", "Direct FIML: named MLR",
  "Direct FIML: all-H₁ expected", "ML2S: paper candidate",
  "ML2S: magmaan default")
normal_rows$Usable <- normal_rows$replications
normal_rows$Reject <- 100 * normal_rows$rejection_rate
normal_rows$MC_SE <- 100 * normal_rows$rejection_se
normal_rows$Scale <- normal_rows$mean_scaling_factor

mcar <- summary[summary$cell_id == "diagnostic-mcar-k7", , drop = FALSE]
mcar_rows <- rbind(
  one(mcar[
    mcar$method == "fiml" &
      mcar$residual_information == "structured_expected_h1" &
      mcar$omega_bread_point == "saturated" &
      mcar$omega_bread_kind == "expected" &
      mcar$omega_meat_point == "saturated", ],
      "MCAR YB structured-U expected FIML"),
  one(mcar[
    mcar$method == "fiml" &
      mcar$residual_information == "saturated_expected_h1" &
      mcar$omega_bread_point == "saturated" &
      mcar$omega_bread_kind == "expected" &
      mcar$omega_meat_point == "saturated", ], "MCAR all-H1 expected FIML"),
  one(mcar[
    mcar$method == "fiml" &
      mcar$residual_information == "saturated_observed_h1" &
      mcar$omega_bread_point == "saturated" &
      mcar$omega_bread_kind == "expected" &
      mcar$omega_meat_point == "saturated", ],
      "MCAR observed-residual FIML"),
  one(mcar[
    mcar$method == "fiml" &
      mcar$residual_information == "structured_observed_h1" &
      mcar$omega_bread_point == "structured" &
      mcar$omega_bread_kind == "observed" &
      mcar$omega_meat_point == "structured", ], "MCAR literal FIML"),
  one(mcar[mcar$method == "fiml_mlr", ], "MCAR named MLR"),
  one(mcar[
    mcar$method == "ml2s" &
      mcar$omega_bread_point == "saturated" &
      mcar$omega_bread_kind == "observed" &
      mcar$omega_meat_point == "saturated" &
      mcar$stage2_information == "structured_observed", ],
      "MCAR paper-candidate ML2S"),
  one(mcar[
    mcar$method == "ml2s" &
      mcar$omega_bread_point == "saturated" &
      mcar$omega_bread_kind == "observed" &
      mcar$omega_meat_point == "saturated" &
      mcar$stage2_information == "saturated_expected", ],
      "MCAR default ML2S"))
mcar_rows$Route <- c(
  "Direct FIML: YB structured-U expected",
  "Direct FIML: all-H1 expected",
  "Direct FIML: observed residual, expected sandwich",
  "Direct FIML: paper-literal",
  "Direct FIML: named MLR",
  "ML2S: paper candidate",
  "ML2S: magmaan default")
mcar_rows$Published <- c(rep(45, 5L), rep(12, 2L))
mcar_rows$Usable <- mcar_rows$replications
mcar_rows$Reject <- 100 * mcar_rows$rejection_rate
mcar_rows$MC_SE <- 100 * mcar_rows$rejection_se
mcar_rows$Reject[[6L]] <- 100 * mcar_rows$rejection_rate_n_minus_1[[6L]]
mcar_rows$MC_SE[[6L]] <- 100 * mcar_rows$rejection_se_n_minus_1[[6L]]
mcar_rows$Scale <- mcar_rows$mean_scaling_factor

mcar_design <- design[design$cell_id == "diagnostic-mcar-k7", , drop = FALSE]
stopifnot(nrow(mcar_design) == 1L)
mcar_mean_patterns <- mcar_design$mean_realized_patterns
mcar_mean_missing <- mcar_design$mean_realized_missing_rate

dgp_roots$Branch <- ifelse(
  dgp_roots$same_joint_law_as_selected,
  "Selected law", "Alternative law")
dgp_roots$Monotonicity <- ifelse(
  dgp_roots$monotone_increasing, "Increasing",
  ifelse(dgp_roots$monotone_decreasing, "Decreasing", "Nonmonotone"))
dgp_roots$Solvable <- ifelse(
  dgp_roots$target_correlations_solvable, "Yes", "No")
dgp_roots$Selected <- ifelse(dgp_roots$selected, "Yes", "No")
```

## Question

Which choices of residual information, sandwich bread, and score meat reproduce
the robust FIML and robust two-stage tests reported by Savalei and Falk (2014)?
The answer determines the explicitly named paper-era routes that Experiment 77
should verify; it must not silently change magmaan's modern defaults.

**Goal.** Identify the finite-sample EQS/Yuan--Bentler route from a six-cell
MCAR fingerprint rather than one rejection rate; determine whether the
two-stage numerator was $nF$ or $(n-1)F$; enumerate all nearby matrix choices;
and turn the remaining ambiguity into a concrete Experiment 77 verification
gate without changing magmaan's defaults.

## Short answer

```{r short-answer, results='asis'}
if (reps >= 1000) {
  cat(sprintf(
    paste0(
      "Across the six graph-digitized **MCAR** cells, the best direct-FIML ",
      "fingerprint is the EQS/Yuan--Bentler expected-information hybrid: ",
      "$U$ uses expected information evaluated at the structured fit, while ",
      "$\\Omega$ uses the saturated estimator's expected bread and saturated ",
      "scores. Its rejection-rate RMSE is **%.1f points**, versus %.1f for ",
      "expected information everywhere at $H_1$ and %.1f for the literal ",
      "all-structured observed construction. The one exact MAR-L cell pulls ",
      "toward all-$H_1$ expected (%.1f%% versus %.1f%% for the hybrid, target ",
      "63.3%%), leaving the two expected routes essentially tied across all ",
      "seven cells (RMSE %.1f versus %.1f points).\n\n"),
    fingerprint_metrics$RMSE[[1L]], fingerprint_metrics$RMSE[[2L]],
    fingerprint_metrics$RMSE[[3L]],
    100 * fiml_expected$rejection_rate, 100 * yb_mar$rejection_rate,
    seven_yb_rmse, seven_all_h1_rmse))
  cat(sprintf(
    paste0(
      "For **two-stage FIML**, saturated observed Stage-1 bread, saturated ",
      "score meat, and structured observed Stage-2 information remains the ",
      "source-implied route. Using the paper's $(n-1)F/c$ numerator improves ",
      "its six-cell RMSE from %.1f to **%.1f points**. This is a modest but ",
      "consistent finite-sample improvement, not a new asymptotic method.\n\n"),
    fingerprint_metrics$RMSE[[5L]], fingerprint_metrics$RMSE[[4L]]))
  cat(paste0(
    "The plausible historical reading is therefore that Savalei and Falk ",
    "requested EQS's named Yuan--Bentler correction and reported what EQS 6.1 ",
    "returned. Their prose and syntax explicitly request analytic observed ",
    "information, yet the rejection fingerprint looks like an expected-",
    "information code path. That does **not** establish an execution mistake ",
    "by the authors: it may be a version-specific EQS convention or an ",
    "undocumented separation between the information used by the statistic ",
    "and by standard errors. Experiment 77 should verify both expected routes ",
    "and treat the structured-expected hybrid as the leading paper-era ",
    "candidate.\n"))
} else {
  cat(sprintf(
    paste0("The %s run (%d samples per cell) validates all 80 crossed matrix ",
           "constructions plus named MLR, but is too small to compare them reliably with ",
           "the paper's 10.0%% two-stage and 63.3%% FIML targets.\n"),
    profile, reps))
}
```

## Evidence

The primary fingerprint uses Savalei--Falk Model 2, MCAR, many missingness
patterns, 30% deletion within each missing-variable set, skewness 2, and both
paper kurtoses at $N=200,400,600$. The published values are approximate
readings of Figures 8--10; every simulation cell contains 1,000 samples.

```{r fingerprint-table}
knitr::kable(
  fingerprint_table,
  digits = c(0, 0, rep(1, 7)), row.names = FALSE,
  caption = paste(
    "Six-cell MCAR rejection-rate fingerprint. Published values are",
    "graph-digitized approximations; all entries after kurtosis are percentages."))
```

The six-cell RMSEs are 3.0 points for structured-$U$ expected FIML, 3.9
for all-$H_1$ expected FIML, and 8.5 for the literal structured-observed
route. For two-stage they are 1.8 points with $(n-1)F/c$ and 2.3 with $nF/c$.

The focused run had one numerically anomalous ML2S replication in the
$N=200$, kurtosis-15 cell. Its 32 ML2S matrix rows were excluded together; the
direct-FIML result from the same sample was retained.

```{r fingerprint-audit-text, results='asis'}
cat(sprintf(
  paste0(
    "The other 5,999 ML2S fits satisfy the saturated expected/observed ",
    "identity to %.3g and reproduce the fitted default scale to %.3g.\n"),
  fingerprint_audit[["saturated_identity_max_abs"]],
  fingerprint_audit[["default_scale_max_abs"]]))
```

### Exact MAR-L diagnostic

The earlier exact target cell uses Model 2, $N=200$, MAR-L, many missingness
patterns, 30% missingness, skewness 2, and kurtosis 7. Each method is fitted
once to a sample and every matrix choice is evaluated on that same fit. The
two-stage candidate below uses the paper's $(n-1)F/c$ convention.

```{r evidence-table}
knitr::kable(
  evidence[, c("Method", "Choice", "Published", "Usable", "Rejection",
                 "MC_SE", "Scale")],
  digits = c(NA, NA, 1, 1, 1, 1, 3), row.names = FALSE,
  col.names = c("Method", "Matrix choice", "Published (%)", "Usable (%)",
                "Reject (%)", "MC SE (points)", "Mean scale"),
  caption = "Published targets and reconstructed/default matrix choices in the diagnostic cell.")
```

```{r geometry-text, results='asis'}
cat(sprintf(
  paste0(
    "The literal direct-FIML residual matrix is indefinite in %.1f%% of its ",
    "usable samples, and %d of %d scales are nonpositive. The two-stage ",
    "structured observed residual matrix is indefinite in %.1f%%, yet its ",
    "trace remains positive in every sample. Indefiniteness is expected ",
    "finite-sample geometry because neither structured matrix is evaluated ",
    "at an unrestricted moment-model optimum.\n\n"),
  100 * fiml_literal$projection_indefinite_rate,
  reps - fiml_literal$replications, reps,
  100 * ml2s_source$projection_indefinite_rate))
cat(sprintf(
  paste0(
    "The complete-data saturated expected/observed identity holds to %.3g, ",
    "and the enumerated ML2S default reproduces magmaan's fitted scale to ",
    "%.3g.\n"),
  identity_error, default_error))
```

## Caveats

```{r caveat-text, results='asis'}
cat(sprintf(
  paste0(
    "This is a targeted implementation-identification experiment, not a ",
    "replication of all 192 published cells. It combines a six-cell MCAR ",
    "fingerprint with two exact nonnormal MAR-L cells, a normal control, and ",
    "the original single MCAR diagnostic. The original EQS random stream is ",
    "not available. The MCAR targets were read from plots and each published ",
    "percentage is based on 1,000 samples, so exact equality is neither ",
    "expected nor meaningful. The matrix-level EQS comparison remains the ",
    "decisive verification because the two expected-information FIML routes ",
    "are nearly tied across the seven target cells.\n\n")))
cat(sprintf(
  paste0(
    "The generator uses the monotone-increasing Fleishman root and treats the ",
    "stated kurtoses as excess kurtosis. In the moderate cell it yields %.1f%% ",
    "theoretically MAR-L-eligible observations, matching the paper's ",
    "approximately 38%%. The missingness routine enforces exactly 30%% ",
    "deletion within every published variable set and regenerates samples ",
    "that cannot meet that conditional instruction. The tracked CSV is a ",
    "compact summary; raw replication rows and per-cell checkpoints remain ",
    "generated local output.\n"),
  eligible))
```

## Reproduce

```{=html}
<pre><code class="literal-code">cd experiments/80-savalei-falk-2014-ml2s-information

Rscript run_experiment.R --smoke
Rscript run_experiment.R --replication --cores 8 --resume
Rscript run_experiment.R --fingerprint --cores 8 --resume
quarto render report.qmd</code></pre>
```

## What EQS and the papers actually specify

There are two different sample-size conventions, and they should not be
silently unified.

- Savalei and Falk's Equation (4) defines the two-stage statistic as
  $d(n-1)F_{ML}(\widetilde\theta)/\operatorname{tr}(\widehat\Omega_1U_2)$.
- Their direct-FIML Equation (6), and EQS Manual Equations (12.6)--(12.7),
  scale the raw likelihood ratio
  $T_{ML}=-2\{\ell(\widehat\theta)-\ell(\widehat\beta)\}$ directly. There is
  no extra $n-1$ multiplier to substitute in this route.

The public EQS 6 manual predates the study's two-stage implementation: it says
that this missing-data two-stage method was not then available, whereas the
article used EQS 6.1 with `MISSING=TS`. Consequently, the public manual cannot
independently confirm the 6.1 two-stage numerator. Equation (4) and the supplied
6.1 syntax are the relevant evidence; our $(n-1)$ result is also empirically
closer. For direct FIML, by contrast, the public manual documents the named
Yuan--Bentler route in detail.

For direct FIML, the [EQS 6 program manual](https://www.mvsoft.com/wp-content/uploads/2021/04/EQS_6_Prog_Manual_422pp.pdf)
calls the output a Yuan--Bentler correction and writes

$$
T_{YB}=\frac{T_{ML}}{c},\qquad
c=\frac{\operatorname{tr}(U_\beta V_\beta)}{d},
$$

with $V_\beta=A_\beta^{-1}B_\beta A_\beta^{-1}$ explicitly described as the
asymptotic covariance of the **saturated** estimator. Its Equation (12.11)
uses the simpler projected residual matrix

$$
U_\beta=A_\beta-A_\beta\Delta
  (\Delta^\top A_\beta\Delta)^{-1}\Delta^\top A_\beta,
$$

not the full structured-parameter observed Hessian. The manual says observed
information is the default finite-sample estimate of $A_\beta$, replaceable by
Fisher expected information on request. It does not say clearly whether the
$A_\beta$ inside $U_\beta$ and the breads inside saturated $V_\beta$ must be
evaluated at the same finite-sample point.

That omission is exactly where the leading empirical hybrid lives:

$$
U_\beta=U\{A^E_\beta(\widehat\beta_0)\},\qquad
V_\beta=
  A^E_\beta(\widehat\beta)^{-1}B_\beta(\widehat\beta)
  A^E_\beta(\widehat\beta)^{-1}.
$$

Savalei and Falk's Appendix B requests `SE=EXACT`, which they explain as an
analytic observed-information calculation, and the article says observed
information was used throughout. Thus the expected-information fingerprint is
a real implementation discrepancy, not something licensed by their written
method. The defensible conclusion is limited: the named EQS statistic appears
to have followed an expected-information-like path in this simulation, but
without EQS 6.1 output or source we cannot tell whether that path was selected
by the command, hard-coded for the statistic, or merely mimicked by another
finite-sample difference.

## Named MLR check

Magmaan's named Yuan--Bentler/Mplus MLR statistic is not a forty-ninth setting
of the crossed direct-FIML grid. It computes a scalar difference of traces,

$$
c_{\mathrm{MLR}}
=\frac{\frac{1}{2}\mathrm{tr}(H_1^{-1}J_1)
      -\frac{1}{2}\mathrm{tr}(H_0^{-1}J_0)}{d},
$$

using its separately fitted saturated $H_1$ blocks and the full observed
structured-parameter Hessian under $H_0$. In incomplete data this is not
identical to the superficially similar mixed moment-space row in the 48-choice
grid.

```{r mlr-text, results='asis'}
cat(sprintf(
  paste0(
    "In the 1,000-replication target cell, named MLR rejects **%.1f%%** ",
    "with mean scale %.3f. This is %.1f percentage points above the paper's ",
    "63.3%% and %.1f combined Monte Carlo SE away. It is therefore not the ",
    "missing Savalei--Falk implementation.\n"),
  100 * fiml_mlr$rejection_rate, fiml_mlr$mean_scaling_factor,
  100 * (fiml_mlr$rejection_rate - .633), mlr_z))
```

## Data-generating-process checks

The article used the Fleishman (1978) and Vale--Maurelli (1983) transformation
as implemented in EQS 6.1, with skewness 2 and kurtosis 7 or 15. It did **not**
include a normal population or a second nonnormal generator. The authors also
identified dependence on the nonnormal generator as a limitation. The normal
cell below is therefore a diagnostic control, not a published replication
cell.

```{r normal-control-table}
knitr::kable(
  normal_rows[, c("Route", "Usable", "Reject", "MC_SE", "Scale")],
  digits = c(NA, 0, 1, 1, 3), row.names = FALSE,
  col.names = c("Route", "Usable N", "Reject (%)", "MC SE (points)",
                "Mean c"),
  caption = "Normal-population diagnostic under the same N=200, MAR-L, many-pattern, 30%-missing design.")
```

Named MLR and the all-$H_1$ expected construction nearly coincide under
normality (26.2% and 26.3%, respectively), as they should when the information
identities become approximately valid. Their nonnominal rejection reflects the
very difficult $p=21$, $d=186$, $N=200$ incomplete-data setting; it does not
identify a VM branch. The paper-candidate ML2S route remains appreciably better
at 10.8%.

### MCAR mechanism check

To remove the thresholding and eligibility logic of MAR-L, the added MCAR cell
retains Model 2, $N=200$, MP, 30% missingness, and $s2k7$. For each of the six
missing-variable sets in Table 1, exactly 60 rows are sampled uniformly without
reference to any data value. Figure 8 supplies only plotted values; its matched
points are approximately 45% for direct FIML and 12% for TS, so they should be
treated as graph-digitized targets rather than exact tabulated percentages.

```{r mcar-control-table}
knitr::kable(
  mcar_rows[, c("Route", "Published", "Usable", "Reject", "MC_SE", "Scale")],
  digits = c(NA, 0, 0, 1, 1, 3), row.names = FALSE,
  col.names = c("Route", "Figure 8 (%)", "Usable N", "Reject (%)",
                "MC SE (points)", "Mean c"),
  caption = "Matched MCAR diagnostic; Figure 8 targets are approximate readings from the graph.")
```

The structured-$U$ expected direct-FIML construction gives 46.6%, closest to
the approximately 45% point. All-$H_1$ expected gives 41.8%. The competing
observed-residual, expected-sandwich construction that was near the MAR-L
target gives only 28.4%, so MCAR sharply discriminates against it. The
paper-literal route gives 33.5% among 947 usable tests, and named MLR gives
71.3%; neither explains the graph. For ML2S, the paper candidate with the
$n-1$ numerator gives 15.2% versus approximately 12%, whereas magmaan's
current default gives 43.3%.

```{r mcar-text, results='asis'}
cat(sprintf(
  paste0(
    "The simulation realizes an average of %.1f missingness patterns, ",
    "compared with the article's reported average of about 50 for ",
    "$N=200$, MP, and 30%% missingness. Its overall elementwise missing ",
    "rate is %.4f, exactly the expected $(12/21)\\times0.30$. These checks ",
    "substantially reduce the chance that the remaining result is an ",
    "artifact of our missingness implementation.\n"),
  mcar_mean_patterns, mcar_mean_missing))
```

The Fleishman moment equations themselves do have four real coefficient roots
for each nonnormal target. The table distinguishes algebraic roots from distinct
joint distributions and checks whether each branch can reproduce every target
correlation while retaining a positive-definite latent-normal correlation
matrix.

```{r dgp-root-table}
knitr::kable(
  dgp_roots[, c("excess_kurtosis", "Selected", "Branch", "Monotonicity",
                  "a", "b", "c", "d", "Solvable",
                  "intermediate_min_eigenvalue")],
  digits = c(0, NA, NA, NA, 6, 6, 6, 6, NA, 3), row.names = FALSE,
  col.names = c("Kurtosis", "Chosen", "Joint-law branch", "Map",
                "a", "b", "c", "d", "All correlations?",
                "Latent min eigenvalue"),
  caption = "All real Fleishman roots for skewness 2 and the two paper kurtoses.")
```

For kurtosis 7, the decreasing root in the selected branch is just the global
latent reflection $T(-Z)$ and generates the same joint distribution. The only
genuinely different, nonmonotone branch cannot reproduce all of Model 2's
negative correlations. Thus root nonuniqueness cannot explain the critical
63.3% discrepancy. For kurtosis 15, the alternative nonmonotone branch is
feasible and should be treated as a separate generator-sensitivity condition
if that cell is expanded, but it cannot explain the kurtosis-7 target.

## Complete option enumeration

These tables enumerate the full crossed grid, not merely the selected rows
above. `Invalid` counts samples with a nonpositive scaling factor, for which the
scaled chi-square is undefined. Rejection rates and mean scales use the
remaining usable samples. `Gap` is the rejection rate minus the corresponding
published rate, in percentage points.

The column notation separates decisions that the earlier labels accidentally
ran together:

- **Point** is where a matrix is evaluated: $H_1$ is the unrestricted saturated
  FIML estimate and $H_0$ is the fitted structured SEM. These labels do not
  change the null hypothesis being tested; every row tests the same SEM.
- **Type** applies to curvature matrices only: expected (E) means expected
  Hessian and observed (O) means the sample Hessian. “Expected” does not mean
  imputed or unobserved data. Under incomplete-data FIML, E and O need not agree
  in a finite sample.
- **B scores** is always an empirical score outer product, so it has an
  evaluation point but no expected/observed choice.
- **U: θ block** distinguishes the projected block $Delta^\top A\Delta$ from
  the full observed structured-parameter Hessian
  $H_{\theta\theta}^{O}(H_0)$, which additionally contains the second-derivative
  terms of the moment map.

### Direct FIML: all 48 choices

Each row independently chooses the $A$ used to construct $U$, the inner
$\theta$-curvature used inside $U$, the $A$ used as the sandwich bread, and the
point used for the empirical score meat $B$. The six $U$ constructions are
crossed with four sandwich breads and two score points.

The **paper-literal row is indeed observed**, as expected: $U$ uses $A^O(H_0)$
with inner block $\Delta^\top A\Delta$, while $\Omega$ uses $A^O(H_0)$ and
$B(H_0)$. The expected rows are diagnostic alternatives, not alternative
claims about the paper. They are retained because the paper-literal row misses
the rejection fingerprint badly. The structured-$U$ expected route is the best
six-cell MCAR match; all-$H_1$ expected is better in the exact MAR-L cell and
is nearly tied when all seven targets are combined.

```{r fiml-option-table}
knitr::kable(
  fiml_options[, c("Role", "U_A_point", "U_A_kind", "U_theta",
                     "Omega_A_point", "Omega_A_kind", "Omega_B_point",
                     "replications", "Invalid", "Reject", "Gap", "Scale")],
  digits = c(NA, NA, NA, NA, NA, NA, NA, 0, 0, 1, 1, 3),
  row.names = FALSE,
  col.names = c("Role", "U: A point", "U: A type", "U: θ block",
                "Ω: A point", "Ω: A type", "Ω: B scores", "Usable N",
                "Invalid", "Reject (%)", "Gap (points)", "Mean c"),
  caption = "All direct-FIML matrix choices, Model 2 moderate diagnostic cell.")
```

### Two-stage FIML: all 32 choices

The same split notation makes the two-stage table parallel. The four
incomplete-data Stage-1 breads and two score points are crossed with the four
complete-data Stage-2 information matrices. The paper candidate is $H_1$
observed Stage-1 bread, $H_1$ score meat, and $H_0$ observed Stage-2
information.

```{r ml2s-option-table}
knitr::kable(
  ml2s_options[, c("Role", "Omega_A_point", "Omega_A_kind",
                     "Omega_B_point", "U_H_point", "U_H_kind",
                     "replications", "Invalid", "Reject_n_minus_1",
                     "Gap_n_minus_1", "Reject_n", "Gap_n", "Scale")],
  digits = c(NA, NA, NA, NA, NA, NA, 0, 0, 1, 1, 1, 1, 3),
  row.names = FALSE,
  col.names = c("Role", "Ω₁: A point", "Ω₁: A type", "Ω₁: B scores",
                "U₂: H point", "U₂: H type", "Usable N", "Invalid",
                "Reject, (n−1)F (%)", "Gap, (n−1)F",
                "Reject, nF (%)", "Gap, nF", "Mean c"),
  caption = "All two-stage matrix choices, Model 2 moderate diagnostic cell.")
```

## Matrix grid

Let $\widehat\beta$ denote the unrestricted saturated FIML estimate of the
means and covariances, let
$\widehat\beta_0=\beta(\widehat\theta)$ denote the moments implied by the
direct-FIML fit, and let
$\widetilde\beta_0=\beta(\widetilde\theta)$ denote the moments implied by the
second-stage fit. Write

$
A_\beta^{r,k}=-E_k\!\left[\frac{1}{n}
  \frac{\partial^2\ell(\beta)}
       {\partial\beta\partial\beta^\top}\right]_{\beta=\beta_r},
\qquad
B_\beta^r=\frac{1}{n}\sum_{i=1}^n
  s_i(\beta_r)s_i(\beta_r)^\top ,
$

where $r\in\{H1,0\}$ selects saturated or structured moments and
$k\in\{E,O\}$ selects expected or observed information. Following the
paper's score-outer-product definition, $B$ is not sample-centered. The
sandwich is

$
\Omega(A,B)=A^{-1}BA^{-1}.
$

For direct FIML, four residual matrices use the H1-form projection

$
U(A)=A-A\Delta(\Delta^\top A\Delta)^{-1}\Delta^\top A,
\qquad
\Delta=\frac{\partial\beta(\theta)}{\partial\theta^\top},
$

with
$A\in\{A_\beta^{H1,E},A_\beta^{H1,O},
A_\beta^{0,E},A_\beta^{0,O}\}$. Two further choices replace
$\Delta^\top A\Delta$ by the full observed structured-parameter Hessian,
using either structured or saturated observed outer information. Each of those
six $U$ choices is crossed with four sandwich breads
$A_\beta^{H1,E},A_\beta^{H1,O},A_\beta^{0,E},A_\beta^{0,O}$ and two meats
$B_\beta^{H1},B_\beta^0$, giving $6\times4\times2=48$ tests. A literal
reading of Savalei and Falk's Equations (5)–(6) selects

$
U(A_\beta^{0,O}),\qquad
\Omega=\Omega(A_\beta^{0,O},B_\beta^0).
$

For two-stage FIML, the Stage-1 asymptotic covariance is independently crossed
over the same four incomplete-data breads and two meat points,

$
\Omega_1=\Omega(A_\beta^{r,k},B_\beta^m),
\qquad r,m\in\{H1,0\},\quad k\in\{E,O\}.
$

The second-stage residual matrix uses complete-data saturated-moment
information $H$, evaluated at either $\widehat\beta$ or
$\widetilde\beta_0$, with expected or observed curvature:

$
U(H)=H-H\widetilde\Delta
 (\widetilde\Delta^\top H\widetilde\Delta)^{-1}
 \widetilde\Delta^\top H.
$

This gives $4\times2\times4=32$ tests. Equation (2), Equation (4), and
SE=EXACT select saturated observed Stage-1 bread, saturated meat, and structured
observed $H$. Magmaan's current ML2S instead uses the same Stage-1 sandwich
with saturated expected $H$. Every row finally uses

$
c=\frac{\operatorname{tr}(U\Omega)}{d},
\qquad T_{\mathrm{scaled}}=\frac{T}{c},
\qquad d=186.
$

For direct FIML, $T$ is the raw saturated-versus-structured likelihood ratio
$T_{ML}$. For two-stage FIML the report records both $T=nF_{ML}$, magmaan's
generic convention, and $T=(n-1)F_{ML}$, the explicit Savalei--Falk
two-stage convention. Fingerprint comparisons use the latter.

The six direct-FIML residual choices follow the Equation-37 taxonomy in
Savalei and Rosseel (2022); the historical selections follow Savalei and Falk
(2014), especially Equations (2), (4)–(6), and Appendix B.
