---
title: "Which matrix choices produced the Savalei–Falk robust tests?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(if (nzchar(override)) override else character(),
           file.path("results", c("replication", "pilot", "smoke")))
available <- roots[vapply(roots, function(path)
  file.exists(file.path(path, "metadata.csv")), logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
summary <- read_result("summary.csv")
design <- read_result("design.csv")
audit_frame <- read_result("audit.csv")
audit <- setNames(audit_frame$value, audit_frame$key)
meta_frame <- read_result("metadata.csv")
meta <- setNames(meta_frame$value, meta_frame$key)
reps <- as.integer(meta[["reps_per_cell"]])
profile <- meta[["profile"]]
k7 <- summary[summary$kurtosis == 7, , drop = FALSE]

one <- function(x, label) {
  if (nrow(x) != 1L) stop("expected one summary row for ", label)
  x
}
ml2s_default <- one(k7[
  k7$method == "ml2s" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "observed" &
    k7$omega_meat_point == "saturated" &
    k7$stage2_information == "saturated_expected", ], "ML2S default")
ml2s_source <- one(k7[
  k7$method == "ml2s" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "observed" &
    k7$omega_meat_point == "saturated" &
    k7$stage2_information == "structured_observed", ], "ML2S source row")
fiml_default <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "saturated_observed_h1" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "observed" &
    k7$omega_meat_point == "saturated", ], "FIML H1 default")
fiml_literal <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "structured_observed_h1" &
    k7$omega_bread_point == "structured" &
    k7$omega_bread_kind == "observed" &
    k7$omega_meat_point == "structured", ], "literal FIML equations row")

targets <- c(ml2s = .100, fiml = .633)
closest <- lapply(names(targets), function(method) {
  z <- k7[k7$method == method, , drop = FALSE]
  z[which.min(abs(z$rejection_rate - targets[[method]])), , drop = FALSE]
})
names(closest) <- names(targets)
ml2s_target_se <- sqrt(.1 * .9 / 1000)
fiml_target_se <- sqrt(.633 * .367 / 1000)
ml2s_z <- abs(ml2s_source$rejection_rate - .1) /
  sqrt(ml2s_source$rejection_se^2 + ml2s_target_se^2)
fiml_closest <- closest$fiml
fiml_expected <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "saturated_expected_h1" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "expected" &
    k7$omega_meat_point == "saturated", ], "FIML closest row")
stopifnot(identical(fiml_closest$information, fiml_expected$information))
fiml_z <- abs(fiml_closest$rejection_rate - .633) /
  sqrt(fiml_closest$rejection_se^2 + fiml_target_se^2)
fiml_observed_residual <- one(k7[
  k7$method == "fiml" &
    k7$residual_information == "saturated_observed_h1" &
    k7$omega_bread_point == "saturated" &
    k7$omega_bread_kind == "expected" &
    k7$omega_meat_point == "saturated", ], "FIML observed-residual alternative")

evidence <- rbind(ml2s_source, ml2s_default, fiml_closest,
                  fiml_observed_residual, fiml_literal, fiml_default)
evidence$Choice <- c("Source-implied candidate", "Current magmaan default",
                     "Closest: saturated expected throughout",
                     "Observed residual, expected sandwich",
                     "Literal all-structured observed equations",
                     "Current magmaan H1 metric")
evidence$Method <- c("Two-stage", "Two-stage", rep("Direct FIML", 4L))
evidence$Published <- c(10.0, NA, 63.3, 63.3, 63.3, NA)
evidence$Usable <- 100 * evidence$replications / reps
evidence$Rejection <- 100 * evidence$rejection_rate
evidence$MC_SE <- 100 * evidence$rejection_se
evidence$Scale <- evidence$mean_scaling_factor

eligible <- 100 * design$theoretical_eligible_proportion[
  design$kurtosis == 7]
identity_error <- audit[["saturated_identity_max_abs"]]
default_error <- audit[["default_scale_max_abs"]]
```

## Question

Which choices of residual information, sandwich bread, and score meat reproduce
the robust FIML and robust two-stage tests reported by Savalei and Falk (2014)?
The answer determines the explicitly named paper-era routes that Experiment 77
should verify; it must not silently change magmaan's modern defaults.

## Short answer

`r if (reps >= 1000) sprintf("For **two-stage FIML**, the source-implied choice—saturated observed Stage-1 bread, saturated score meat, and structured observed Stage-2 information—is also the empirical match: %.1f%% rejection versus the published 10.0%% (%.1f combined Monte Carlo SE). For **direct FIML**, the grid exposes an unresolved conflict. The closest row uses saturated expected residual information and a saturated expected/saturated-score sandwich (%.1f%% versus 63.3%%; %.1f combined Monte Carlo SE), whereas the literal structured-observed equations produce only %.1f%% rejection among %.1f%% usable tests.", 100 * ml2s_source$rejection_rate, ml2s_z, 100 * fiml_expected$rejection_rate, fiml_z, 100 * fiml_literal$rejection_rate, 100 * fiml_literal$replications / reps) else sprintf("The %s run (%d samples per cell) validates all 80 matrix constructions, but is too small to compare them reliably with the paper's 10.0%% two-stage and 63.3%% FIML targets.", profile, reps)`

`r if (reps >= 1000) "Experiment 77 can proceed with the two-stage candidate, but it should **not yet add a paper-era direct-FIML route**. The next FIML gate must determine whether our structured observed-H1 construction matches EQS's analytic matrix, or whether EQS applied SE=EXACT only to standard errors while its robust test retained saturated expected information. The current magmaan defaults remain unchanged." else ""`

## Evidence

All rows use Savalei–Falk Model 2, (N=200), MAR-L, many missingness
patterns, 30% missingness, skewness 2, and kurtosis 7. Each method is fitted
once to a sample and every matrix choice is evaluated on that same fit.

```{r evidence-table}
knitr::kable(
  evidence[, c("Method", "Choice", "Published", "Usable", "Rejection",
                 "MC_SE", "Scale")],
  digits = c(NA, NA, 1, 1, 1, 1, 3), row.names = FALSE,
  col.names = c("Method", "Matrix choice", "Published (%)", "Usable (%)",
                "Reject (%)", "MC SE (points)", "Mean scale"),
  caption = "Published targets and reconstructed/default matrix choices in the diagnostic cell.")
```

The literal direct-FIML residual matrix is indefinite in
`r sprintf("%.1f%%", 100 * fiml_literal$projection_indefinite_rate)` of its
usable samples, and 65 of 1,000 scales are nonpositive. The two-stage structured
observed residual matrix is indefinite in
`r sprintf("%.1f%%", 100 * ml2s_source$projection_indefinite_rate)`, yet its
trace remains positive in every sample. Indefiniteness is expected finite-sample
geometry because neither structured matrix is evaluated at an unrestricted
moment-model optimum.

The complete-data saturated expected/observed identity holds to
`r format(identity_error, digits = 3, scientific = TRUE)`, and the enumerated
ML2S default reproduces magmaan's fitted scale to
`r format(default_error, digits = 3, scientific = TRUE)`.

## Caveats

This is a targeted two-cell implementation-identification experiment, not a
replication of all 192 published cells. The original EQS random stream is not
available, and each published percentage is itself based on 1,000 samples.
Agreement within roughly one to two percentage points is therefore the relevant
resolution, not exact equality. One published rejection rate cannot distinguish
every near-matching FIML row: the saturated-observed residual paired with the
same expected sandwich gives
`r sprintf("%.1f%%", 100 * fiml_observed_residual$rejection_rate)`. The
matrix-level EQS comparison is therefore essential rather than optional.

The generator uses the monotone-increasing Fleishman root and treats the stated
kurtoses as excess kurtosis. In the moderate cell it yields
`r sprintf("%.1f%%", eligible)` theoretically MAR-L-eligible observations,
matching the paper's approximately 38%. The missingness routine enforces exactly
30% deletion within every published variable set and regenerates samples that
cannot meet that conditional instruction. The tracked CSV is a compact summary;
raw replication rows and per-cell checkpoints remain generated local output.

## Reproduce

```bash
cd experiments/80-savalei-falk-2014-ml2s-information

Rscript run_experiment.R --smoke
Rscript run_experiment.R --replication --cores 8
quarto render report.qmd
```

## Matrix grid

Let \(\widehat\beta\) denote the unrestricted saturated FIML estimate of the
means and covariances, let
\(\widehat\beta_0=\beta(\widehat\theta)\) denote the moments implied by the
direct-FIML fit, and let
\(\widetilde\beta_0=\beta(\widetilde\theta)\) denote the moments implied by the
second-stage fit. Write

\[
A_\beta^{r,k}=-E_k\!\left[\frac{1}{n}
  \frac{\partial^2\ell(\beta)}
       {\partial\beta\partial\beta^\top}\right]_{\beta=\beta_r},
\qquad
B_\beta^r=\frac{1}{n}\sum_{i=1}^n
  s_i(\beta_r)s_i(\beta_r)^\top ,
\]

where \(r\in\{H1,0\}\) selects saturated or structured moments and
\(k\in\{E,O\}\) selects expected or observed information. Following the
paper's score-outer-product definition, \(B\) is not sample-centered. The
sandwich is

\[
\Omega(A,B)=A^{-1}BA^{-1}.
\]

For direct FIML, four residual matrices use the H1-form projection

\[
U(A)=A-A\Delta(\Delta^\top A\Delta)^{-1}\Delta^\top A,
\qquad
\Delta=\frac{\partial\beta(\theta)}{\partial\theta^\top},
\]

with
\(A\in\{A_\beta^{H1,E},A_\beta^{H1,O},
A_\beta^{0,E},A_\beta^{0,O}\}\). Two further choices replace
\(\Delta^\top A\Delta\) by the full observed structured-parameter Hessian,
using either structured or saturated observed outer information. Each of those
six \(U\) choices is crossed with four sandwich breads
\(A_\beta^{H1,E},A_\beta^{H1,O},A_\beta^{0,E},A_\beta^{0,O}\) and two meats
\(B_\beta^{H1},B_\beta^0\), giving \(6\times4\times2=48\) tests. A literal
reading of Savalei and Falk's Equations (5)–(6) selects

\[
U(A_\beta^{0,O}),\qquad
\Omega=\Omega(A_\beta^{0,O},B_\beta^0).
\]

For two-stage FIML, the Stage-1 asymptotic covariance is independently crossed
over the same four incomplete-data breads and two meat points,

\[
\Omega_1=\Omega(A_\beta^{r,k},B_\beta^m),
\qquad r,m\in\{H1,0\},\quad k\in\{E,O\}.
\]

The second-stage residual matrix uses complete-data saturated-moment
information \(H\), evaluated at either \(\widehat\beta\) or
\(\widetilde\beta_0\), with expected or observed curvature:

\[
U(H)=H-H\widetilde\Delta
 (\widetilde\Delta^\top H\widetilde\Delta)^{-1}
 \widetilde\Delta^\top H.
\]

This gives \(4\times2\times4=32\) tests. Equation (2), Equation (4), and
SE=EXACT select saturated observed Stage-1 bread, saturated meat, and structured
observed \(H\). Magmaan's current ML2S instead uses the same Stage-1 sandwich
with saturated expected \(H\). Every row finally uses

\[
c=\frac{\operatorname{tr}(U\Omega)}{d},
\qquad T_{\mathrm{scaled}}=\frac{T}{c},
\qquad d=186.
\]

The six direct-FIML residual choices follow the Equation-37 taxonomy in
Savalei and Rosseel (2022); the historical selections follow Savalei and Falk
(2014), especially Equations (2), (4)–(6), and Appendix B.
