---
title: "Can a projected Satterthwaite denominator calibrate pivotal SEM scores?"
subtitle: "A low-N goodness-of-fit probe with normal and Vale–Maurelli data"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
meta_frame <- read_result("metadata.csv")
meta <- setNames(meta_frame$value, meta_frame$key)
methods <- read_result("method_summary.csv")
diagnostics <- read_result("diagnostics_summary.csv")
reference <- read_result("reference_summary.csv")
failures <- read_result("failures.csv")
profile <- meta[["profile"]]
reps <- as.integer(meta[["reps_per_cell"]])
is_pilot <- identical(profile, "pilot") && reps >= 500L

labels <- c(
  score_chisq_centered = "Robust score chi-square (centered meat)",
  score_chisq_raw = "Robust score chi-square (raw OPG)",
  hotelling_gaussian = "Ordinary Hotelling (eta = n-1)",
  satterthwaite_normal_unprojected = "Normal-reference Satterthwaite, unprojected",
  satterthwaite_normal_projected = "Normal-reference Satterthwaite, projected",
  satterthwaite_oracle_projected = "DGP-oracle Satterthwaite, projected",
  satterthwaite_empirical_unprojected = "Empirical Satterthwaite, unprojected",
  satterthwaite_empirical_projected = "Empirical Satterthwaite, projected",
  fmg_sb_rls = "FMG SB-RLS",
  fmg_mv_rls = "FMG mean+variance RLS",
  fmg_peba4_rls = "FMG pEBA4-RLS")
methods$label <- unname(labels[methods$method])
score_names <- c(
  "score_chisq_centered",
  "score_chisq_raw",
  "hotelling_gaussian",
  "satterthwaite_normal_unprojected",
  "satterthwaite_normal_projected",
  "satterthwaite_oracle_projected",
  "satterthwaite_empirical_unprojected",
  "satterthwaite_empirical_projected")
context_names <- c("fmg_sb_rls", "fmg_mv_rls", "fmg_peba4_rls")
score_results <- methods[methods$method %in% score_names, , drop = FALSE]
context_results <- methods[methods$method %in% context_names, , drop = FALSE]
n_cells <- length(unique(methods$cell_id))
method_error <- do.call(rbind, lapply(
  split(score_results, score_results$method),
  function(x) data.frame(
    method = x$method[[1L]],
    valid_cells = sum(is.finite(x$rejection_rate)),
    mean_rejection = mean(x$rejection_rate, na.rm = TRUE),
    mean_absolute_error = mean(abs(x$rejection_rate - .05), na.rm = TRUE),
    stringsAsFactors = FALSE)))
eligible <- method_error$valid_cells == n_cells
best_method <- method_error[
  which(eligible)[which.min(method_error$mean_absolute_error[eligible])],
  , drop = FALSE]
rate <- function(method, distribution, n) {
  row <- methods[
    methods$method == method & methods$distribution == distribution &
      methods$n == n, , drop = FALSE]
  if (nrow(row) != 1L) return(NA_real_)
  100 * row$rejection_rate[[1L]]
}
mae <- function(method) {
  row <- method_error[method_error$method == method, , drop = FALSE]
  if (nrow(row) != 1L) return(NA_real_)
  100 * row$mean_absolute_error[[1L]]
}
wide_sizes <- function(x, method_order) {
  x$size_percent <- 100 * x$rejection_rate
  x$condition <- paste(x$distribution, paste0("N=", x$n))
  out <- reshape(
    x[c("method", "condition", "size_percent")],
    idvar = "method", timevar = "condition", direction = "wide")
  names(out) <- sub("size_percent.", "", names(out), fixed = TRUE)
  out <- out[match(method_order, out$method), , drop = FALSE]
  out <- out[!is.na(out$method), , drop = FALSE]
  out$method <- unname(labels[out$method])
  out
}
score_wide <- wide_sizes(score_results, score_names)
context_wide <- wide_sizes(context_results, context_names)
score_calibration_display <- method_error
score_calibration_display$method <-
  unname(labels[score_calibration_display$method])
score_calibration_display$mean_rejection <-
  100 * score_calibration_display$mean_rejection
score_calibration_display$mean_absolute_error <-
  100 * score_calibration_display$mean_absolute_error
names(score_calibration_display) <- c(
  "Method", "Valid cells", "Mean rejection (%)", "Mean absolute error (pp)")
score_calibration_display <- score_calibration_display[
  order(score_calibration_display[["Mean absolute error (pp)"]]), ,
  drop = FALSE]

reference_display <- reference[
  c("distribution", "shape_unprojected", "shape_projected",
    "shape_projected_p10", "shape_projected_p90",
    "projection_fraction")]
reference_display$projection_fraction <-
  100 * reference_display$projection_fraction
names(reference_display) <- c(
  "Distribution", "Unprojected shape", "Projected shape",
  "Projected p10", "Projected p90",
  "Variation removed (%)")

diagnostic_display <- diagnostics[
  c("distribution", "n", "median_q_raw_centered_ratio",
    "median_centered_meat_condition",
    "median_shape_empirical_projected",
    "eta_empirical_projected_availability",
    "fit_success", "score_success")]
diagnostic_display$eta_empirical_projected_availability <-
  100 * diagnostic_display$eta_empirical_projected_availability
diagnostic_display$fit_success <- 100 * diagnostic_display$fit_success
diagnostic_display$score_success <- 100 * diagnostic_display$score_success
names(diagnostic_display) <- c(
  "Distribution", "N", "median Qraw/Qcentered",
  "median meat condition", "median projected shape",
  "projected eta available (%)",
  "fit success (%)", "score success (%)")
```

## Question

For a goodness-of-fit score that is already sandwich-studentized and therefore
has a pivotal $\chi_q^2$ limit, can the randomness of its empirical meat be
summarized by a projected Satterthwaite degree of freedom? More concretely,
does that heavier finite-sample reference improve Type-I error at low $N$
without giving up the simple numerator?

$$
Q_c=n\bar\psi^\mathsf{T}\widehat B_c^{-1}\bar\psi.
$$

This is not a conventional SEM likelihood-ratio-mixture exercise. The target
statistic $Q_c$ is the robust score quadratic itself, and its asymptotic null
is pivotal. SB, mean-and-variance, and pEBA rows are included only as familiar
context.

## Short answer

`r if (is_pilot) {
  sprintf(
    "In the %d-replication pilot, the closest primary calibration to 5%% overall was %s (mean absolute size error %.2f percentage points across the eight cells).",
    reps, unname(labels[best_method$method[[1L]]]),
    100 * best_method$mean_absolute_error[[1L]])
} else {
  sprintf(
    "The %s run has only %d replications per cell. It validates the construction and exposes the denominator diagnostics, but it is not evidence for choosing a calibration. Run the pilot before reading rejection-rate differences.",
    profile, reps)
}`

`r if (is_pilot) {
  sprintf(
    "At N=30, centered-meat chi-square rejected %.1f%% (normal) and %.1f%% (VM); the fixed normal-reference projected correction rejected %.1f%% and %.1f%%, ordinary Hotelling rejected %.1f%% and %.1f%%, and the same-sample projected correction rejected %.1f%% and %.1f%%.",
    rate("score_chisq_centered", "normal", 30),
    rate("score_chisq_centered", "vm", 30),
    rate("satterthwaite_normal_projected", "normal", 30),
    rate("satterthwaite_normal_projected", "vm", 30),
    rate("hotelling_gaussian", "normal", 30),
    rate("hotelling_gaussian", "vm", 30),
    rate("satterthwaite_empirical_projected", "normal", 30),
    rate("satterthwaite_empirical_projected", "vm", 30))
} else {
  ""
}`

The decisive comparison is deliberately split in two. The fixed
normal-reference shape asks whether the proposed working-model denominator is
useful and cheap. The DGP-specific oracle asks whether the same approximation
could work with the correct score moments. The empirical row asks whether
estimating those third and fourth moments in the same low-$N$ sample destroys
the gain. If the oracle helps but the empirical version does not, the idea
generalizes mathematically but misses the simplicity/performance selling point.

## Evidence

### Common robust-score numerator

For each case, let $\psi_i\in\mathbb R^q$ be the model-centred saturated
covariance contribution after projection into magmaan's
expected-information residual basis. Here $p=5$, so the correctly specified
one-factor model has $q=5$ residual score directions. Define

$$
\bar\psi=\frac{1}{n}\sum_{i=1}^n\psi_i,\qquad
\widehat B_c=\frac{1}{n-1}\sum_{i=1}^n
  (\psi_i-\bar\psi)(\psi_i-\bar\psi)^\mathsf{T},
$$

and the raw outer-product meat

$$
\widehat B_r=\frac{1}{n}\sum_{i=1}^n\psi_i\psi_i^\mathsf{T}.
$$

The two sandwich quadratics actually computed are

$$
Q_c=n\bar\psi^\mathsf{T}\widehat B_c^{-1}\bar\psi,\qquad
Q_r=n\bar\psi^\mathsf{T}\widehat B_r^{-1}\bar\psi.
$$

Under the null and ordinary sandwich regularity conditions,
$Q_c\Rightarrow\chi_q^2$ and $Q_r\Rightarrow\chi_q^2$. Thus every score
calibration below retains a pivotal asymptotic null; the experiment asks only
whether a different finite-sample reference improves its low-$N$ behavior.

### The eight pivotal-score alternatives

For any effective meat degree of freedom $\eta>q-1$, define the common
Hotelling/Satterthwaite tail

$$
F_\eta(Q_c)=\frac{\eta-q+1}{\eta q}Q_c,\qquad
p_\eta=\Pr\!\left\{F_{q,\,\eta-q+1}\ge F_\eta(Q_c)\right\}.
$$

The report compares every score alternative generated by the experiment:

| Report label | Quadratic | Null reference or denominator rule | What it isolates |
|---|---:|---|---|
| Robust score chi-square (centered meat) | $Q_c$ | $\chi_q^2$ | The ordinary sandwich-robust score test and main baseline. |
| Robust score chi-square (raw OPG) | $Q_r$ | $\chi_q^2$ | The uncentered-OPG convention; it audits numerator leakage into the meat. |
| Ordinary Hotelling ($\eta=n-1$) | $Q_c$ | $p_\eta$ with $\eta=n-1$ | Treats the score vectors as Gaussian. This is a finite-sample benchmark, not an exact reference for quadratic SEM scores. |
| Normal-reference Satterthwaite, unprojected | $Q_c$ | $\eta=(n-1)s_{\mathrm{N},H}$ | Estimates meat variability from a large, independent normal-data reference without numerator projection. |
| Normal-reference Satterthwaite, projected | $Q_c$ | $\eta=(n-1)s_{\mathrm{N},H^\perp}$ | The cheap proposed rule: one fixed normal-working shape after projection. |
| DGP-oracle Satterthwaite, projected | $Q_c$ | $\eta=(n-1)s_{\mathrm{DGP},H^\perp}$ | Uses a large, independent draw from the true DGP; infeasible, but tests whether the approximation can work with the right moments. |
| Empirical Satterthwaite, unprojected | $Q_c$ | $\eta=(n-1)\widehat s_H$ | Estimates the unprojected third/fourth-score-moment shape in the test sample. |
| Empirical Satterthwaite, projected | $Q_c$ | $\eta=(n-1)\widehat s_{H^\perp}$ | Estimates the projected shape in the test sample; potentially adaptive, but no longer especially simple. |

The first two rows are the two chi-square calibrations and differ only in
whether the empirical meat is centered. The remaining six use the same
attractive numerator and centered sandwich denominator---hence the same
quadratic $Q_c$---and change only the finite-sample reference degree of freedom.

### Where the Satterthwaite denominator comes from

Whiten the centered score as
$z=B^{-1/2}\{\psi-E(\psi)\}$ and write the first-order meat fluctuation as

$$
H=zz^\mathsf{T}-I_q.
$$

The projected version removes the part linearly predictable from the
numerator:

$$
\operatorname{vec}(H^\perp)
=\operatorname{vec}(H)
-\operatorname{Cov}\{\operatorname{vec}(H),z\}z.
$$

Matching Frobenius variation to a $q$-variate Wishart meat gives the two shape
constants

$$
s_H=\frac{q(q+1)}{E\lVert H\rVert_F^2},\qquad
s_{H^\perp}=\frac{q(q+1)}{E\lVert H^\perp\rVert_F^2},
\qquad
\eta_n=(n-1)s.
$$

“Unprojected” versus “projected” chooses $H$ versus $H^\perp$.
“Normal-reference,” “DGP-oracle,” and “empirical” say where the expectation in
the matching formula is estimated: a large normal-data reference, a large
true-DGP reference, or the current test sample. Projection is the part intended
to preserve the clean numerator--denominator separation.

### The three FMG contextual comparators

The FMG rows deliberately do **not** belong to the pivotal-score family. They
start from the working-model residual quadratic

$$
T_{\mathrm{RLS}}=n\lVert\bar\psi\rVert^2,
\qquad
T_{\mathrm{RLS}}\ \Rightarrow\
\sum_{j=1}^q\lambda_j\chi_{1,j}^2,
$$

whose limiting weights $\lambda_1,\ldots,\lambda_q$ depend on the
$U\Gamma$ spectrum. They answer the conventional nonpivotal-mixture question
and are shown only to orient the low-$N$ results:

| Report label | Calibration of $T_{\mathrm{RLS}}$ |
|---|---|
| FMG SB-RLS | Mean scaling: $c=(\sum_j\lambda_j)/q$, then compare $T_{\mathrm{RLS}}/c$ with $\chi_q^2$. |
| FMG mean+variance RLS | Satterthwaite spectrum match: compare $T_{\mathrm{RLS}}(\sum_j\lambda_j)/(\sum_j\lambda_j^2)$ with $\chi_\nu^2$, where $\nu=(\sum_j\lambda_j)^2/\sum_j\lambda_j^2$. |
| FMG pEBA4-RLS | Sort the spectrum into four blocks, replace each block by its mean, shrink those block means halfway toward the global mean, and evaluate the resulting weighted-$\chi^2$ tail. |

These are useful comparators, but success here would not establish the
pivotal robust-score proposal that motivated the experiment.

### Design

The design crosses multivariate normal data with one moderate
Vale–Maurelli condition (marginal skewness 2, excess kurtosis 7), holding the
population covariance and the true one-factor model fixed, at
$N=30,50,100,200$. Large independent reference draws estimate the
normal-working and DGP-oracle shape constants; the same-sample empirical
version estimates $s$ in every replication.

```{r shape-table}
knitr::kable(
  reference_display, digits = 3, row.names = FALSE,
  caption = sprintf(
    "Reference score-shape constants: median and between-reference p10/p90 from %s independent draws of %s cases per distribution.",
    meta[["reference_reps"]], meta[["reference_n"]]))
```

The VM projected shape is only
`r sprintf("%.4f", reference$shape_projected[reference$distribution == "vm"])`.
Even at $n=200$, this implies $\eta_n=$
`r sprintf("%.2f", 199 * reference$shape_projected[reference$distribution == "vm"])`,
below the $q-1=4$ existence boundary for the inverse-Wishart/Hotelling
reference. Consequently the DGP-oracle p-value is unavailable in every VM cell,
not merely unstable at $n=30$. By contrast, the normal-reference projected
shape is
`r sprintf("%.3f", reference$shape_projected[reference$distribution == "normal"])`;
it exists at all four sample sizes but produces a much too heavy low-$n$ tail.

### Null rejection

```{r score-size-table}
ordered <- c(
  "method", "normal N=30", "normal N=50", "normal N=100", "normal N=200",
  "vm N=30", "vm N=50", "vm N=100", "vm N=200")
score_ordered <- intersect(ordered, names(score_wide))
knitr::kable(
  score_wide[score_ordered], digits = 1, row.names = FALSE,
  caption = sprintf(
    "All pivotal-score null rejection percentages at alpha=.05 (%d replications per cell). NA means that eta did not exceed q-1.",
    reps))
```

```{r score-calibration-table}
knitr::kable(
  score_calibration_display,
  digits = c(NA, 0, 2, 2), row.names = FALSE,
  caption = "Across-cell calibration summary for all pivotal-score alternatives. An eight-cell mean absolute error is comparable only when Valid cells = 8.")
```

Across the eight cells, ordinary Hotelling has mean absolute size error
`r sprintf("%.2f", mae("hotelling_gaussian"))` percentage points, versus
`r sprintf("%.2f", mae("satterthwaite_empirical_projected"))` for the
same-sample projected plug-in,
`r sprintf("%.2f", mae("satterthwaite_normal_projected"))` for the fixed
normal-reference projection, and
`r sprintf("%.2f", mae("score_chisq_centered"))` for centered-meat chi-square.
The empirical projected row is therefore interesting as a self-normalized
finite-sample rule, but not as an estimate of the population shape: at $n=30$
its median VM shape is
`r sprintf("%.3f", diagnostics$median_shape_empirical_projected[
  diagnostics$distribution == "vm" & diagnostics$n == 30])`,
roughly two orders of magnitude above the independent reference.

The following diagnostics are as important as size in this probe. The raw OPG
contains the rank-one $\bar\psi\bar\psi^\mathsf{T}$ numerator term; its
quadratic therefore obeys the exact identity

$$
Q_{\mathrm{raw}}=\frac{nQ_c}{n-1+Q_c}
$$

and shrinks relative to the centered-meat statistic. The
condition number and estimated shape show whether a nominal improvement costs
an unstable denominator.

```{r diagnostic-table}
knitr::kable(
  diagnostic_display, digits = 3, row.names = FALSE,
  caption = "Centering, conditioning, and same-sample score-shape diagnostics.")
```

For completeness, here are the three nonpivotal FMG contextual rows from the
same simulated samples:

```{r context-size-table}
context_ordered <- intersect(ordered, names(context_wide))
knitr::kable(
  context_wide[context_ordered], digits = 1, row.names = FALSE,
  caption = sprintf(
    "Context only: FMG calibrations of the nonpivotal RLS quadratic, as null rejection percentages at alpha=.05 (%d replications per cell).",
    reps))
```

## Caveats

- The projected $H$ used here is the fixed-geometry score contribution. It
  does not yet include the full influence of re-estimating the SEM nuisance
  parameters and residual basis inside the meat. A promising result would
  justify deriving that full fitted-nuisance influence; a null result does not
  rule it out.
- The DGP-oracle row is intentionally infeasible. It is an upper-bound
  diagnostic, not a proposed method. It also depends on eighth-order raw-data
  moments because the SEM score is quadratic and $H^2$ is quartic in that
  score; reference variability is therefore a result, not just Monte Carlo
  clutter.
- Same-sample third- and fourth-moment estimates can be noisy exactly where the
  correction is supposed to help. Shape failures and $\eta_n\le q-1$ appear
  as unavailable p-values rather than being silently truncated.
- This first probe has one model, one non-normal law, and Type-I error only.
  Power, nested restrictions, heterogeneous marginal shapes, missing data, and
  higher-dimensional GOF are follow-ups, not conclusions from this run.

## Reproduce

```sh
Rscript experiments/71-projected-score-satterthwaite/run_experiment.R --help
Rscript experiments/71-projected-score-satterthwaite/run_experiment.R --smoke
Rscript experiments/71-projected-score-satterthwaite/run_experiment.R --pilot
quarto render experiments/71-projected-score-satterthwaite/report.qmd
```
