---
title: "Can a projected Satterthwaite denominator calibrate pivotal SEM scores?"
subtitle: "A low-N goodness-of-fit probe with normal and Vale–Maurelli data"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
meta_frame <- read_result("metadata.csv")
meta <- setNames(meta_frame$value, meta_frame$key)
methods <- read_result("method_summary.csv")
diagnostics <- read_result("diagnostics_summary.csv")
reference <- read_result("reference_summary.csv")
failures <- read_result("failures.csv")
profile <- meta[["profile"]]
reps <- as.integer(meta[["reps_per_cell"]])
is_pilot <- identical(profile, "pilot") && reps >= 500L

labels <- c(
  score_chisq_centered = "Robust score chi-square (centered meat)",
  score_chisq_raw = "Robust score chi-square (raw OPG)",
  hotelling_gaussian = "Ordinary Hotelling (eta = n-1)",
  satterthwaite_normal_unprojected = "Normal-reference Satterthwaite, unprojected",
  satterthwaite_normal_projected = "Normal-reference Satterthwaite, projected",
  satterthwaite_oracle_projected = "DGP-oracle Satterthwaite, projected",
  satterthwaite_empirical_unprojected = "Empirical Satterthwaite, unprojected",
  satterthwaite_empirical_projected = "Empirical Satterthwaite, projected",
  fmg_sb_rls = "FMG SB-RLS",
  fmg_mv_rls = "FMG mean+variance RLS",
  fmg_peba4_rls = "FMG pEBA4-RLS")
methods$label <- unname(labels[methods$method])
primary_names <- c(
  "score_chisq_centered",
  "satterthwaite_normal_unprojected",
  "satterthwaite_normal_projected",
  "satterthwaite_oracle_projected",
  "satterthwaite_empirical_projected")
primary <- methods[methods$method %in% primary_names, , drop = FALSE]
method_error <- aggregate(
  abs(rejection_rate - .05) ~ method,
  data = primary, FUN = mean, na.rm = TRUE)
names(method_error)[[2L]] <- "mean_absolute_error"
best_method <- method_error[
  which.min(method_error$mean_absolute_error), , drop = FALSE]
rate <- function(method, distribution, n) {
  row <- methods[
    methods$method == method & methods$distribution == distribution &
      methods$n == n, , drop = FALSE]
  if (nrow(row) != 1L) return(NA_real_)
  100 * row$rejection_rate[[1L]]
}
mae <- function(method) {
  row <- method_error[method_error$method == method, , drop = FALSE]
  if (nrow(row) != 1L) return(NA_real_)
  100 * row$mean_absolute_error[[1L]]
}
primary$size_percent <- 100 * primary$rejection_rate
primary$condition <- paste(primary$distribution, paste0("N=", primary$n))
primary_wide <- reshape(
  primary[c("method", "condition", "size_percent")],
  idvar = "method", timevar = "condition", direction = "wide")
names(primary_wide) <- sub("size_percent.", "", names(primary_wide), fixed = TRUE)
primary_wide$method <- unname(labels[primary_wide$method])

reference_display <- reference[
  c("distribution", "shape_unprojected", "shape_projected",
    "shape_projected_p10", "shape_projected_p90",
    "projection_fraction")]
reference_display$projection_fraction <-
  100 * reference_display$projection_fraction
names(reference_display) <- c(
  "Distribution", "Unprojected shape", "Projected shape",
  "Projected p10", "Projected p90",
  "Variation removed (%)")

diagnostic_display <- diagnostics[
  c("distribution", "n", "median_q_raw_centered_ratio",
    "median_centered_meat_condition",
    "median_shape_empirical_projected",
    "eta_empirical_projected_availability",
    "fit_success", "score_success")]
diagnostic_display$eta_empirical_projected_availability <-
  100 * diagnostic_display$eta_empirical_projected_availability
diagnostic_display$fit_success <- 100 * diagnostic_display$fit_success
diagnostic_display$score_success <- 100 * diagnostic_display$score_success
names(diagnostic_display) <- c(
  "Distribution", "N", "median Qraw/Qcentered",
  "median meat condition", "median projected shape",
  "projected eta available (%)",
  "fit success (%)", "score success (%)")
```

## Question

For a goodness-of-fit score that is already sandwich-studentized and therefore
has a pivotal \(\chi_q^2\) limit, can the randomness of its empirical meat be
summarized by a projected Satterthwaite degree of freedom? More concretely:
does that heavier finite-sample reference improve Type-I error at low \(N\)
without giving up the simple numerator
\[
Q=n\bar\psi^\mathsf{T}\widehat B_c^{-1}\bar\psi?
\]

This is not a conventional SEM likelihood-ratio-mixture exercise. The target
statistic \(Q\) is the robust score quadratic itself; SB, mean-and-variance, and
pEBA rows are included only as familiar context.

## Short answer

`r if (is_pilot) {
  sprintf(
    "In the %d-replication pilot, the closest primary calibration to 5%% overall was %s (mean absolute size error %.2f percentage points across the eight cells).",
    reps, unname(labels[best_method$method[[1L]]]),
    100 * best_method$mean_absolute_error[[1L]])
} else {
  sprintf(
    "The %s run has only %d replications per cell. It validates the construction and exposes the denominator diagnostics, but it is not evidence for choosing a calibration. Run the pilot before reading rejection-rate differences.",
    profile, reps)
}`

`r if (is_pilot) {
  sprintf(
    "At N=30, centered-meat chi-square rejected %.1f%% (normal) and %.1f%% (VM); the fixed normal-reference projected correction rejected %.1f%% and %.1f%%, ordinary Hotelling rejected %.1f%% and %.1f%%, and the same-sample projected correction rejected %.1f%% and %.1f%%.",
    rate("score_chisq_centered", "normal", 30),
    rate("score_chisq_centered", "vm", 30),
    rate("satterthwaite_normal_projected", "normal", 30),
    rate("satterthwaite_normal_projected", "vm", 30),
    rate("hotelling_gaussian", "normal", 30),
    rate("hotelling_gaussian", "vm", 30),
    rate("satterthwaite_empirical_projected", "normal", 30),
    rate("satterthwaite_empirical_projected", "vm", 30))
} else {
  ""
}`

The decisive comparison is deliberately split in two. The fixed
normal-reference shape asks whether the proposed working-model denominator is
useful and cheap. The DGP-specific oracle asks whether the same approximation
could work with the correct score moments. The empirical row asks whether
estimating those third and fourth moments in the same low-\(N\) sample destroys
the gain. If the oracle helps but the empirical version does not, the idea
generalizes mathematically but misses the simplicity/performance selling point.

## Evidence

### Construction

For each case, let \(\psi_i\in\mathbb R^q\) be the model-centred saturated
covariance contribution after projection into magmaan's
expected-information residual basis. Here \(p=5\), so the correctly specified
one-factor model has \(q=5\) residual score directions. The centered empirical
meat is
\[
\widehat B_c={1\over n-1}\sum_{i=1}^n
  (\psi_i-\bar\psi)(\psi_i-\bar\psi)^\mathsf{T},
\qquad
Q=n\bar\psi^\mathsf{T}\widehat B_c^{-1}\bar\psi.
\]
Under the null and ordinary sandwich regularity conditions,
\(Q\Rightarrow\chi_q^2\): the limit is pivotal.

Whiten the centered score as \(z=B^{-1/2}(\psi-E\psi)\), and write the
first-order fluctuation of its meat as
\[
H=zz^\mathsf{T}-I_q.
\]
The part linearly predictable from the numerator is removed:
\[
\operatorname{vec}(H^\perp)
=\operatorname{vec}(H)
-\operatorname{Cov}\{\operatorname{vec}(H),z\}z.
\]
Matching its Frobenius variation to a \(q\)-variate Wishart meat gives
\[
s={q(q+1)\over E\lVert H^\perp\rVert_F^2},
\qquad
\eta_n=(n-1)s.
\]
The proposed tail is then
\[
{\,\eta_n-q+1\,\over \eta_n q}Q
\ \dot\sim\ F_{q,\,\eta_n-q+1}.
\]
Leaving \(H\) unprojected is the diagnostic that tells us whether the clean
numerator--denominator orthogonalization is doing real work.

The design crosses multivariate normal data with one moderate
Vale–Maurelli condition (marginal skewness 2, excess kurtosis 7), holding the
population covariance and the true one-factor model fixed, at
\(N=30,50,100,200\). Large independent reference draws estimate the
normal-working and DGP-oracle shape constants; the same-sample empirical
version estimates \(s\) in every replication.

```{r shape-table}
knitr::kable(
  reference_display, digits = 3, row.names = FALSE,
  caption = sprintf(
    "Reference score-shape constants: median and between-reference p10/p90 from %s independent draws of %s cases per distribution.",
    meta[["reference_reps"]], meta[["reference_n"]]))
```

The VM projected shape is only
`r sprintf("%.4f", reference$shape_projected[reference$distribution == "vm"])`.
Even at \(n=200\), this implies
\(\eta_n=\)
`r sprintf("%.2f", 199 * reference$shape_projected[reference$distribution == "vm"])`,
below the \(q-1=4\) existence boundary for the inverse-Wishart/Hotelling
reference. Consequently the DGP-oracle p-value is unavailable in every VM cell,
not merely unstable at \(n=30\). By contrast, the normal-reference projected
shape is
`r sprintf("%.3f", reference$shape_projected[reference$distribution == "normal"])`;
it exists at all four sample sizes but produces a much too heavy low-\(n\) tail.

### Null rejection

```{r size-table}
ordered <- c(
  "method", "normal N=30", "normal N=50", "normal N=100", "normal N=200",
  "vm N=30", "vm N=50", "vm N=100", "vm N=200")
ordered <- intersect(ordered, names(primary_wide))
knitr::kable(
  primary_wide[ordered], digits = 1, row.names = FALSE,
  caption = sprintf(
    "Null rejection percentages at alpha=.05 (%d replications per cell).",
    reps))
```

Across the eight cells, ordinary Hotelling has mean absolute size error
`r sprintf("%.2f", mae("hotelling_gaussian"))` percentage points, versus
`r sprintf("%.2f", mae("satterthwaite_empirical_projected"))` for the
same-sample projected plug-in,
`r sprintf("%.2f", mae("satterthwaite_normal_projected"))` for the fixed
normal-reference projection, and
`r sprintf("%.2f", mae("score_chisq_centered"))` for centered-meat chi-square.
The empirical projected row is therefore interesting as a self-normalized
finite-sample rule, but not as an estimate of the population shape: at \(n=30\)
its median VM shape is
`r sprintf("%.3f", diagnostics$median_shape_empirical_projected[
  diagnostics$distribution == "vm" & diagnostics$n == 30])`,
roughly two orders of magnitude above the independent reference.

The following diagnostics are as important as size in this probe. The raw OPG
contains the rank-one \(\bar\psi\bar\psi^\mathsf{T}\) numerator term; its
quadratic therefore obeys the exact identity
\[
Q_{\rm raw}={nQ_c\over n-1+Q_c}
\]
and shrinks relative to the centered-meat statistic. The
condition number and estimated shape show whether a nominal improvement costs
an unstable denominator.

```{r diagnostic-table}
knitr::kable(
  diagnostic_display, digits = 3, row.names = FALSE,
  caption = "Centering, conditioning, and same-sample score-shape diagnostics.")
```

The FMG SB-RLS, mean-and-variance RLS, and pEBA4-RLS rows are retained in
`method_summary.csv`. They calibrate the nonpivotal working-model quadratic and
answer a different asymptotic question, so they are not promoted into the
primary table.

## Caveats

- The projected \(H\) used here is the fixed-geometry score contribution. It
  does not yet include the full influence of re-estimating the SEM nuisance
  parameters and residual basis inside the meat. A promising result would
  justify deriving that full fitted-nuisance influence; a null result does not
  rule it out.
- The DGP-oracle row is intentionally infeasible. It is an upper-bound
  diagnostic, not a proposed method. It also depends on eighth-order raw-data
  moments because the SEM score is quadratic and \(H^2\) is quartic in that
  score; reference variability is therefore a result, not just Monte Carlo
  clutter.
- Same-sample third- and fourth-moment estimates can be noisy exactly where the
  correction is supposed to help. Shape failures and \(\eta_n\le q-1\) appear
  as unavailable p-values rather than being silently truncated.
- This first probe has one model, one non-normal law, and Type-I error only.
  Power, nested restrictions, heterogeneous marginal shapes, missing data, and
  higher-dimensional GOF are follow-ups, not conclusions from this run.

## Reproduce

```sh
Rscript experiments/71-projected-score-satterthwaite/run_experiment.R --help
Rscript experiments/71-projected-score-satterthwaite/run_experiment.R --smoke
Rscript experiments/71-projected-score-satterthwaite/run_experiment.R --pilot
quarto render experiments/71-projected-score-satterthwaite/report.qmd
```
