---
title: "FIML and two-stage global tests"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

## Question

Which global goodness-of-fit tests remain useful under nonnormal incomplete
data, and which information geometry should a FIML score test use?

## Short answer

The FIML score test uses **observed-H0 sensitivity with the expected
metric** (adopted for the ordinary policy on 2026-10-02). It is the consistent
choice. The pattern-conditional expected information equals the true Hessian
only for complete or MCAR data with a correct moment structure, so expected
sensitivity gives the wrong reference law under MAR. On an exact nonnormal-MAR
pseudo-null, observed-H0 rejected 4.0–6.1% at every positive selection slope and
every N from 200 to 10,000. Expected-H0 fell to 0.3–0.9% at slope 1.5 and to 0% at
slopes 3 and 5, without recovering with N.

The price is finite-sample conservatism that grows with df. On the same FIML
fits at N=200, observed-H0 pEBA4 rejected 0.0–0.7% at df 87 and 0.2–5.7% at
df 34, where expected-H0 rejected 3.3–5.7%. No further run is queued; a
finite-sample correction of the observed reference is a banked idea.

The representative 55-cell SEM panel ran **expected** sensitivity, since it
predates observed sensitivity. There, FIML score pEBA4 rejected 1.4–7.4%,
against 4.8–36.0% for LR pEBA4 and 4.8–89.2% for the scalar-robust comparator.
Each cell attempted 500 fits. The paired ML2S score pEBA4 arm was less stable
(3.0–16.2%). The older FIML-versus-MLR and two-stage comparisons have only
ten-replication local outputs and remain banked designs, not additional
calibration evidence.

```{r}
read_overview <- function(name) {
  path <- file.path("results", "overview", paste0(name, ".csv"))
  if (!file.exists(path)) stop("Missing retained overview: ", path,
    "; restore the tracked CSV or, with original runs present, run Rscript scripts/summarize.R.")
  read.csv(path, stringsAsFactors = FALSE)
}
percent <- function(x) sprintf("%.1f%%", 100*x)
range_percent <- function(lo, hi) paste(percent(lo), percent(hi), sep="–")
```

## Evidence

### Global SEM calibration

The retained panel crosses five latent models, five normal/VM/IG generators,
complete data, 30% MCAR and 30% MAR, and FIML/ML2S. It has 75,000 attempted
fits at N=200. Its directory name contains “n200-500”; metadata specifies
N=200 and **500 replications**, not two sample sizes. The table excludes
nonnormal MAR, leaving 55 cells per estimator with a justified estimator-level
null. Its score arms use expected-H0 sensitivity. Rates are conditional on usable results; unavailable calls are counted
separately. Score results also require the prespecified tangent rank.

```{r}
s <- read_overview("sem_overview")
knitr::kable(data.frame(
  Estimator=s$estimator, Test=s$method,
  `Rejection range`=range_percent(s$min_rate,s$max_rate),
  `Mean absolute error`=percent(s$mean_cell_error),
  `Minimum usable / cell`=paste0(s$min_usable," / ",s$attempts_per_cell),
  `Unavailable calls`=s$unavailable, check.names=FALSE))
```

Mean absolute error weights cells equally; it is descriptive, not a pooled
size estimate or decision criterion. The entire panel recorded four fit
failures, 447 FMG failures, four flip failures and five finite calls with
non-nominal geometry. Different tests have different usable denominators.
The approximately 86-minute wall time is historical, not a current benchmark.
At 500 usable replications, a 5% rejection rate has Monte Carlo standard
error about one percentage point. Full cell rates and denominators are in
`results/overview/sem_cells.csv`.

### Sensitivity and metric are separate choices

Four independent standardized exponentials supply an exact zero-covariance
pseudo-null. The first variable is always observed; logistic selection on it
jointly removes the other three. This isolates nonnormal MAR without assuming
that an arbitrary generating-model covariance is the FIML pseudo-target.
The retained slice uses N=2,000, selection slope 5 and 1,000 attempts per
construction.

```{r}
p <- read_overview("pseudo_null")
knitr::kable(data.frame(Geometry=p$geometry,
  `Rejection (pEBA4)`=percent(p$reject_peba4),
  `Usable / attempted`=paste0(p$usable," / ",p$attempted), check.names=FALSE))
```

Observed-H0 here changes sensitivity; “expected metric” specifies the
projection metric. Observed-H1 and expected-H0 are retained diagnostics.
The wider grid crosses the selection slope with N. Expected-H0 fails at every
N once selection is present, so its error is asymptotic, not finite-sample.

```{r}
g <- read_overview("pseudo_null_grid")
g$label <- ifelse(g$geometry == "expected-H0", "Expected-H0", "Observed-H0")
w <- reshape(g[, c("beta", "n", "label", "reject_peba4")],
  idvar = c("beta", "n"), timevar = "label", direction = "wide")
w <- w[order(w$beta, w$n), ]
knitr::kable(data.frame(`Selection slope`=w$beta, N=w$n,
  `Expected-H0 pEBA4`=percent(w$`reject_peba4.Expected-H0`),
  `Observed-H0 pEBA4`=percent(w$`reject_peba4.Observed-H0`),
  check.names=FALSE))
```

This targeted result does not prove every latent-model MAR null. The
[pseudo-null design](design/pseudo_null_information.md) and
[score geometry appendix](design/lrt_score_information_geometry.md)
preserve the derivation and wider historical checks.

### Where sensitivity enters

At the H0 fit the nuisance score is zero, so the sensitivity does not change
the observed score vector. It changes the weight of the score quadratic and,
decisively, the reference covariance from which the SB and pEBA4 eigenvalues
are computed. The realized score's limit covariance is set by the true
Hessian, whatever sensitivity is chosen, so the reference is correct only if
the sensitivity is consistent for that Hessian in its nuisance rows. The
normal log-likelihood Hessian is linear in centered first and second moments,
so the expected (pattern-conditional Fisher) information matches it only when
every missingness pattern's observed-variable moments equal the model's:
complete or MCAR data with a correct structure. MAR selection breaks this,
mildly for normal data (Kenward and Molenberghs, 1998) and strongly for
nonnormal data. The observed Hessian is consistent in every case.

### Observed sensitivity across latent models

A later paired run fitted FIML once per replication and computed both
sensitivities (expected metric) on the same fits: N=200, 1,000 replications
per cell, three latent models crossed with normal/VM2/IG2 data and complete or
30% MCAR data. It has no MAR cells, so it measures finite-sample behavior where
both sensitivities are consistent. Ranges are over each model's six cells.

```{r}
v <- read_overview("sensitivity_panel")
rng <- function(z, m) { r <- z$rejection_le_05[z$method == m]
  range_percent(min(r), max(r)) }
models <- unique(v[, c("model_id", "expected_df")])
models <- models[order(models$expected_df), ]
knitr::kable(do.call(rbind, lapply(seq_len(nrow(models)), function(i) {
  z <- v[v$model_id == models$model_id[i], ]
  data.frame(Model=models$model_id[i], df=models$expected_df[i],
    `Expected-H0 pEBA4`=rng(z, "expected_peba4"),
    `Observed-H0 pEBA4`=rng(z, "observed_peba4"),
    `Observed-H0 SB`=rng(z, "observed_sb"),
    `LR pEBA4`=rng(z, "lr_peba4"), check.names=FALSE)
})))
```

Observed sensitivity becomes conservative as df grows, on complete normal data
as well. A plausible mechanism, not separately verified: estimation noise in
the observed Hessian enters the nuisance projection and inflates the reference
covariance by a relative amount of order q/N (q free parameters). For the
three-factor model q/N is about .24, and an inflation of 16–24% gives
0.06–0.3% rejection at df 87. The nonnormal df-9 cells (up to 9.5%) show this
is not the whole story. The archived complete-data ML geometry study found the
same collapse (0% at p=40). Every finite-sample failure seen so far is
conservative except that df-9 VM2 cell.

### Two-stage statistic versus reference law

The retained ML2S diagnostic compares the ML discrepancy and residual
least-squares statistic while holding the estimated UΓ reference spectrum
fixed. Seven justified-null cells per N each have 10,000 usable replications;
nonnormal MAR remains a separate estimand-stress arm. The largest paired
rejection differences shrink with N, but changing the statistic alone does
not uniformly repair small-sample calibration. For IG2 complete data at
N=120, pEBA4 rejection is 8.72% versus 8.81%.

```{r}
b <- read_overview("ml2s_bases")
knitr::kable(data.frame(N=b$n, Cells=b$cells,
  `Usable / cell`=b$min_usable,
  `Largest SB difference`=percent(b$max_sb_size_difference),
  `Largest pEBA4 difference`=percent(b$max_peba4_size_difference),
  `Largest mixture difference`=percent(b$max_all_size_difference),check.names=FALSE))
```

Differences are percentage points between paired rejection rates, not errors
relative to 5%. The [two-stage design](design/two_stage_rls.md) retains the
smaller predecessor check; this table reads the later 10,000-replication run.

## Caveats

Nonnormal MAR can move the Gaussian-FIML pseudo-target away from the
unconditional generating covariance. Those panel cells are stress tests
until a model-specific pseudo-null is established, and are excluded above.
The affine covariance screen additionally includes infinite-moment, MNAR and
positivity-violation controls; it cannot establish latent-model validity.

The legacy-MLR lane retains a distinct two-group GOF/nested/power design;
the two-stage lane retains a one-factor estimator comparison. Their local
outputs are single ten-replication nonnormal-MAR cells. Neither supports a
method ranking. Reopen either only for a named paper question with a frozen
null and new substantive run. No older grid is queued by this consolidation.

Evidence limits of the adopted geometry: the sensitivity panel has no MAR
cells and a single N; latent-model nonnormal-MAR validity rests on the
four-variable pseudo-null; and observed-sensitivity nested evidence is the
df=1 normal family of the score-centering decision study (5.0–6.5%). The panel
ran magmaan 0.0.1 code of 2026-09-03 (`sensitivity_panel_metadata.csv`).
The prospective SEM design (`design/prospective_sem.md`) is not queued.

The study is banked. Reopen it for a finite-sample correction of the observed
reference (for example, removing the projection-noise term) or for a consumer
that needs FIML score power at high df/N. Implemented predeclared shrinkage
sensitivities exist for global tests only, and have never been run with SB or
pEBA4. Completed multiplier comparisons and full-observed-metric failures
remain diagnostics. Parameter SEs have a different target and remain in
research/39.

## Reproduce

Render the retained overview without the large ignored raw trees:

```sh
quarto render experiments/research/banked/44-fiml-global-tests/report.qmd
```

From this study directory, inspect the parent command reference or smoke the
paired SEM route into a fresh absolute directory:

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --lane sem --help
Rscript run_experiment.R --lane sem --smoke --models one_factor_6 \
  --distributions normal --missingness mcar_30 --max-cells 2 \
  --results-dir /tmp/fiml-global-fresh
Rscript run_experiment.R --lane two-stage --help
Rscript run_experiment.R --lane legacy-mlr --help
```

`--lane screen` selects the completed affine control. `--check` selects the
crossing, pseudo-null, pseudo-target, H1-quadratic, ML2S-RLS, MLR-parity or
power-calibration diagnostic; consult its source/help before a run.
`Rscript scripts/summarize.R` rebuilds the compact overview **only when the
original local runs are present**. `results/overview/provenance.csv` records
input paths and hashes. Original seeds, checkpoint metadata and cloud IDs
are preserved; moved source fingerprints cannot resume old checkpoints.
