---
title: "magmaan vs lavaan: speed bench"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
  gfm: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
results_dir <- file.path(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)), "results")
summary_path <- file.path(results_dir, "summary.csv")
if (!file.exists(summary_path)) stop("Missing results/summary.csv. Run: Rscript run_experiment.R")
summary_tbl <- utils::read.csv(summary_path, stringsAsFactors = FALSE)
geomean <- function(task) round(summary_tbl$speedup_geomean[summary_tbl$task == task])
```

How much faster is magmaan than lavaan on the same fitted model?

Fitting the identical one-factor CFA, magmaan is about `r geomean("continuous")`×
faster on plain ML, `r geomean("fiml")`× faster under 10% FIML missingness, and
`r geomean("ordinal")`× faster on ordinal DWLS — geometric means across three
population sizes (n=200-1000, p=6-12), with every fit checked to match
lavaan's estimates before it is timed.

## Evidence

```{r}
out <- data.frame(
  Fit = summary_tbl$label,
  Speedup = sprintf("%.0f×", summary_tbl$speedup_geomean)
)
knitr::kable(out, row.names = FALSE, align = c("l", "r"))
```

Per-size detail (all three population sizes, both engines' median times) is
in [`results/cases.csv`](results/cases.csv).

## Caveats

- Both engines are timed on the fit call only: parsing model syntax and
  building sample statistics happens once beforehand, outside the timer, for
  both sides. This is a repeated-fit speedup, not raw-syntax-to-answer speed.
- Estimate-only: no standard errors, test statistics, or fit indices are
  requested from either engine here.
- Small, synthetic, single-factor populations at three sizes. Not a general
  performance claim across model families or larger problems.
- Timed with a shared batched timer (seven batches, engine order rotated
  across batches, single BLAS/OpenMP thread), not a single `system.time()`
  call.
- A fuller, staged-attribution speed study (raw-to-report timing broken into
  parsing/fitting/inference) is tracked separately and does not yet have
  public numbers; see
  [project/validation/benchmark_plan.md](../../../project/validation/benchmark_plan.md).

## Reproduce

```sh
Rscript experiments/showcases/02-lavaan-speed-bench/run_experiment.R
(cd experiments/showcases/02-lavaan-speed-bench && quarto render report.qmd)
```

## See also

- [`project/validation/lavaan_tutorial_parity.md`](../../../project/validation/lavaan_tutorial_parity.md) — central lavaan-tutorial parity audit.
- [`benchmarks/README.md`](../../../benchmarks/README.md) — the shared benchmark harness this experiment builds on.
