---
title: "Intervals for bounded fitted functionals"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
read_evidence <- function(name) {
  path <- file.path("results", "overview", name)
  if (!file.exists(path)) stop("Missing retained summary: ", path,
                               "; recover original CSVs and run scripts/summarize.R")
  read.csv(path, stringsAsFactors = FALSE)
}
show_table <- function(x, digits = 3) knitr::kable(x, digits = digits, row.names = FALSE)
```

## Question

Which interval constructions remain worth pursuing for bounded fitted functionals,
and what prevents their small-sample correction from being operational?

## Short answer

This **banked** study consolidates bifactor maximal-reliability intervals and the
generic reliability profile engine. Oracle rescaling can rescue acceptance rates
in selected near-ceiling cases, but the feasible correction is weaker and actual
endpoint inversion needs separate validation. For bifactor H at $N=50$,
ordinary true-value acceptance is .526, oracle-rescaled acceptance .971 and
null-bootstrap acceptance .803 on 173 of 200 attempts. Reopen for a named bounded-functional
consumer that requires calibrated intervals; existing delta/Wald and uncorrected
profile primitives remain the cheaper choices.

The `bifactor` and `profile` lanes keep different models, coefficients, seeds,
fit exclusions and run metadata. They answer a shared construction question;
their results are never pooled.

## Evidence

### Delta intervals and bifactor constraints

The bifactor lane retains general/group maximal reliability, analytic gradients,
model-based and sandwich delta intervals, logit transformations, bias correction,
Cornish–Fisher, bootstrap and constrained-LR diagnostics. The retained main grid
has **40 attempts per cell at $N=100$**, not the larger historical design. Coverage
is conditional on admissible usable fits; the attempts and usable counts are
essential parts of the result. The displayed slice is normal general-factor
reliability; group-factor and nonnormal cells remain in
[the full summary](results/overview/bifactor_intervals.csv).

```{r bifactor}
x <- read_evidence("bifactor_intervals.csv")
x <- x[x$coef == "gen" & x$dist == "normal", ]
show_table(data.frame(Indicators=x$p, Method=x$method, Attempts=x$attempts,
  Usable=x$reps_ok, `Coverage (%)`=100*x$coverage, check.names=FALSE), digits=1)
```

The separate $N=50$, 18-indicator normal LR diagnostic has 200 attempts per cell,
with only about half usable. Its `cov_*` fields measure **acceptance at the true
functional**, not successful endpoint construction:

```{r bifactor-lr}
x <- read_evidence("bifactor_pointwise.csv")
show_table(data.frame(Coefficient=x$coef, Attempts=x$attempts, Usable=x$reps_ok,
  `Mean LR`=x$mean_LR, `Ordinary acceptance (%)`=100*x$cov_chi2,
  `Oracle acceptance (%)`=100*x$cov_bartlett, check.names=FALSE), digits=1)
```

The Bartlett factor is the cell mean of the same true-value LR statistics.
This oracle uses information unavailable to a single dataset and is evaluated
on the same Monte Carlo draws; it supplies a diagnostic, not a deployable recipe.

### Generic profile construction and oracle roots

The profile lane constrains a scalar functional of fitted loadings/residuals and
inverts the normal-theory likelihood-ratio test. Its examples cover one-factor
omega/H and bifactor total/hierarchical omega. H, general-factor maximal reliability
and total-score omega have different estimands and cannot be interchanged.

Eight dataset–functional pairs ($N=200$, two seeds) remain from the semlbci smoke
comparison. A bound is eligible for parity only if the constrained LR residual
is finite and within .02 for both engines:

```{r parity}
x <- read_evidence("parity.csv")
show_table(data.frame(Pairs=x$datasets_functionals, `Engine root failures`=x$engine_bad_roots,
  `Oracle root failures`=x$oracle_bad_roots, `Both valid`=x$both_valid,
  `Maximum bound difference`=format(x$max_bound_difference_both_valid, scientific=TRUE),
  check.names=FALSE))
```

One semlbci pair fails its own root check. Agreement on the other seven is
numerical construction evidence, not coverage calibration.

The separate $N=50$ diagnostics retain their own attempted/usable denominators:

```{r profile-calibration}
x <- read_evidence("profile_pointwise.csv")
show_table(data.frame(Functional=x$functional, Population=x$pop, Attempts=x$attempts,
  Usable=x$reps, `Ordinary (%)`=100*x$cov_nt, `Oracle (%)`=100*x$cov_bart_oracle,
  `Null bootstrap (%)`=100*x$cov_bart_null, check.names=FALSE), digits=1)
```

For bifactor H, ordinary true-value acceptance is about .526 on 173 usable fits
out of 200 attempts; oracle rescaling reaches .971, while null-bootstrap rescaling
reaches .803. The omega examples behave differently. These are pointwise tests;
end-to-end CI availability, connectedness, endpoint accuracy and cost remain
unproved. The older `coverage.csv`, double-bootstrap and Lawley outputs remain
separate historical diagnostics, with their original metadata.

## Caveats

Fit selection, boundary mass, biased estimates, correction noise and unstable
bootstrap fits can each change inference. Resampling is a comparator with its
own failures, not a universal gold standard. The retained secondary runs are
not one common grid; current CSVs do not establish all sample sizes described by
older drafts. A consumer-driven reopening should prespecify an operational
correction, held-out targets and full endpoint evaluation, including failures.
The generic parameter-profile study remains independent.

## Reproduce

The parent runner dispatches to lane-local modules. Inspect lane-specific help
before selecting a run; all commands accept fresh `--results-dir` paths.

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --lane bifactor --help
Rscript run_experiment.R --lane profile --help
Rscript run_experiment.R --lane bifactor --reps 2 --ns 100 \
  --results-dir /tmp/bounded_bifactor_check
Rscript run_experiment.R --lane profile --seeds 1 --ns 200 --funcs H_1f \
  --results-dir /tmp/bounded_profile_check
Rscript run_experiment.R --check lr --help
Rscript scripts/summarize.R
```

Other `--check` lanes retain bias, bootstrap, Cornish–Fisher, sampling, coverage,
Lawley and double-bootstrap diagnostics. Original results live under
`results/bifactor` and `results/profile`; [source hashes](results/overview/sources.csv)
make the frozen overview auditable. Rendering reads only these compact summaries.
