---
title: "Likelihood-score covariance centering"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
run_id <- Sys.getenv('SCORE_CENTERING_RUN', 'ml_confirmation')
root <- file.path('results', 'score-centering', run_id)
rd <- function(name) {
  path <- file.path(root, paste0(name, '.csv'))
  if (!file.exists(path)) stop('Missing ', path,
    '; run Rscript run_experiment.R --library PATH --run-id ', run_id)
  read.csv(path, stringsAsFactors = FALSE)
}
meta <- rd('metadata'); rates <- rd('rates'); cov <- rd('covariance')
pairs <- rd('paired'); checks <- rd('checks'); decisions <- rd('decisions')
value <- function(key) meta$value[match(key, meta$key)]
percent <- function(x) sprintf('%.1f%%', 100 * x)
fiml_root <- file.path('results', 'score-centering', 'fiml_confirmation')
fiml <- function(name) {
  path <- file.path(fiml_root, paste0(name, '.csv'))
  if (!file.exists(path)) stop('Missing ', path,
    '; run the FIML confirmation shown under Reproduce')
  read.csv(path, stringsAsFactors = FALSE)
}
fiml_rates <- fiml('rates'); fiml_cov <- fiml('covariance')
fiml_pairs <- fiml('paired'); fiml_checks <- fiml('checks')
fiml_decisions <- fiml('decisions'); fiml_meta <- fiml('metadata')
confirm_root <- file.path('results', 'score-centering', 'group_confirmation')
confirm <- function(name) {
  path <- file.path(confirm_root, paste0(name, '.csv'))
  if (!file.exists(path)) stop('Missing ', path,
    '; run the grouped confirmation and independent validator shown under Reproduce')
  read.csv(path, stringsAsFactors = FALSE)
}
group_cov <- confirm('covariance'); group_rates <- confirm('rates')
group_pairs <- confirm('paired'); group_checks <- confirm('checks')
group_decisions <- confirm('decisions'); oracle <- confirm('oracle_checks')
group <- group_cov[group_cov$role == 'null', ]
tab <- function(x, ...) knitr::kable(x, row.names = FALSE, ...)
```

## Default register

| Choice | Current value | Status |
|:--|:--|:--|
| ML parameter covariance | Uncentered exact likelihood-score meat, observed bread | Stationary equivalence confirmed; skewed-data Wald coverage remains open |
| Fixed-allocation score covariance | Within-group centered covariance rows | Formula settled; ordinary-policy extension banked |
| ML global/nested score reference | Uncentered efficient scores, expected geometry | Retain raw under the registered rule in tested regular ML families |
| Ordinary FIML inference | Unsupported | Prospective evidence only |
| Prospective FIML centering | Raw comparator retained within each geometry | No qualifying centering benefit; observed-sensitivity calibration remains open |
| Established moment-covariance and LR centering | Retained method-specific formulas | Redesign banked |

## Question

Can magmaan use uncentered likelihood-score second moments throughout its
parameter covariance and score-test calibration, or does the sampling law
require a distinction? All covariance arms use the same estimates, projection,
information matrices and observed score. The
[criteria](criteria/score_centering.md) were committed before the pilot and confirmation draws.

## Short answer

**Retain uncentered score references for the tested regular complete-data ML
families.** Independent confirmation used 32,000 datasets across global/nested
tests, normal/skewed innovations and N=80/300. Globally centering covariance rows
changed null rejection by at most 0.75 percentage points, with identical empirical
matched-null power in all SB/PEBA4 comparisons. It did not meet the preregistered
benefit or noninferiority conditions for a default change. Fits, information,
projection and observed statistics were held fixed; no default changes.

Parameter covariances were numerically identical across arms, but both Wald
intervals undercovered in skewed-data cells: nested null coverage at N=80 was
90.3% (95% Wilson interval 88.9–91.5%). This is a separate covariance/interval
validation task; score centering cannot correct it.

**The additional 32,000-dataset FIML confirmation supplies no qualifying
centering benefit**, separately under expected and observed sensitivity.
Matched-null power changed by at most 0.10 percentage points. Observed-sensitivity
global tests remain conservative, and centering worsens nested MCAR size at
N=80. This does not select an ordinary FIML default or settle sensitivity geometry.

The fixed-allocation covariance formula is settled separately: within-group
centering removes score-mean terms absent from conditional sampling variance.
Its ordinary-policy extension remains banked. Further centering comparisons are
now [banked with a reopening trigger](../../../project/backlog/speculative.md#likelihood-score-centering-alternatives);
FIML calibration and interval validation remain active. PSD boundary and penalty
inference are outside this grid.

## Evidence

### ML confirmation does not justify switching to global centering

Each null and power cell contains 2,000 usable pairs. The following slice shows
PEBA4 null rejection; all raw and centered SB/PEBA4 null Wilson intervals lie
inside the registered 3–7% band. The full rates, coverage, failures and paired
uncertainty are frozen in this run's CSVs.

```{r ml-confirmation}
z <- rates[rates$lane == 'ml' & rates$family != 'grouped_mean' &
  rates$role == 'null' & rates$method == 'peba4', ]
a <- z[z$arm == 'raw', c('family', 'n', 'rejection')]
b <- z[z$arm == 'global', c('family', 'n', 'rejection', 'lower', 'upper')]
w <- merge(a, b, by = c('family', 'n'), suffixes = c('_raw', '_centered'))
labels <- c(global_normal = 'Global, normal', global_skew = 'Global, skewed',
            nested_normal = 'Nested, normal', nested_skew = 'Nested, skewed')
tab(data.frame(Family = unname(labels[w$family]), N = w$n,
  'Raw rejection' = percent(w$rejection_raw),
  'Centered rejection' = percent(w$rejection_centered),
  'Centered 95% interval' = paste(percent(w$lower), percent(w$upper), sep = '–'),
  check.names = FALSE))
```

The registered benefit rule requires a paired 95% interval establishing at least
a one-percentage-point reduction in absolute size error in some cell. No cell
meets it. Absolute-size-error losses against raw were global normal N=300
(+0.10 points SB, +0.20 PEBA4), global skewed N=300 (+0.05 both), and global
skewed N=80 (+0.75 SB). The global N=80 SB paired upper bounds also exceed the
allowed one-point worsening margin (normal +1.001, skewed +1.05).

Matched-null power estimates are identical across arms in every comparison;
the lowest paired bootstrap bound for the difference is −0.15 percentage
points, within the registered two-point loss allowance. Both candidate components
are `not_eligible` in the frozen decision table. The implied outcome is to retain
raw provisionally in the tested scope, without claiming a blanket calibration
guarantee across models or sampling laws.

### Stationary parameter covariance agrees; Wald coverage needs investigation

The maximum raw/global parameter-covariance discrepancy is
`r format(max(cov$max_covariance_gap[cov$arm == 'global']), scientific=TRUE, digits=3)`
on the runner's comparison scale. Known-target Wald intervals are therefore the
same. Both arms fail the coverage rule in several cells, most visibly under
skewed innovations: nested null N=80 coverage is 90.3%, and nested power N=80
coverage is 90.25%. Finite-sample covariance bias, estimate bias and interval
calibration require separate checks against the declared population targets.
The nominal-size result for score tests does not establish Wald interval coverage.

### Disposition of the fixed-allocation case

Recorded 2026-10-01 after inspecting results: the fixed-allocation covariance
formula is settled by the sampling law, and its ordinary-policy extension is
[banked](../../../project/backlog/speculative.md#fixed-design-inference-under-mean-misspecification).
The [scope contract](../../../project/scope.md#group-allocation-and-likelihood-score-covariance)
distinguishes within-group covariance for fixed allocation from raw score second
moments for joint sampling of group labels. No further experiment is needed to
choose between those two population formulas. Finite-sample reference
calibration remains separate; the N=80 result cannot make raw covariance correct
for a fixed-allocation target. The original criteria and frozen decision outcomes
are retained. Absolute FIML calibration and excluded inference regimes remain open.

The grouped nested test targets the best-fitting shared mean, not equality of
the two true group means. Both nested models retain the wrong shared-mean
assumption; the additional restriction fixes that approximation mean at zero.
It therefore illustrates inference within misspecified models, not a correctly
specified full-model null.

### Grouped parameter covariance has an independent target

The groups have fixed sizes $N/4$ and $3N/4$, unit observation variance, and X
means $\mu+.6$ and $\mu-.2$. The fitted shared mean is the pooled average, whose
approximation target is $\mu$ and whose variance is exactly $1/N$. The between-
group squared mean displacement is $.25(.6)^2+.75(-.2)^2=.12$.

At the fitted mean, with divisor $N$, the expected reported variance ratios are

$$
E[N\widehat V_{\rm raw}]=E[N\widehat V_{\rm global}]=1.12-1/N,
\qquad E[N\widehat V_{\rm group}]=1-2/N.
$$

The second expression's small finite-sample bias is retained; this experiment
does not introduce an $N-1$ or group degrees-of-freedom correction.

```{r grouped-evidence}
tab(data.frame(
  N = group$n, Covariance = group$arm, Usable = group$usable,
  'Variance / true variance' = round(group$oracle_ratio, 3),
  'Expected ratio' = round(ifelse(group$arm == 'group', 1 - 2 / group$n, 1.12 - 1 / group$n), 3),
  '95% interval coverage' = percent(group$coverage), check.names = FALSE))
```

The raw-versus-group difference persists with sample size. Global score-mean
centering cannot remove it because the opposing group means cancel overall.
This fixed-allocation example establishes a scope limit for a universal raw-meat
policy, without generalizing group centering to random missingness patterns.

### Centered grouped scores fail the small-sample acceptance rule

The confirmation compares 2,000 paired null draws at each sample size. SB and
PEBA4 coincide in this one-dimensional case. The N=80 centered interval exceeds
the registered 7% upper limit, and its paired size-error comparison also fails
the allowed one-percentage-point worsening rule.

```{r grouped-score-evidence}
z <- group_rates[group_rates$role == 'null' & group_rates$method == 'peba4', ]
tab(data.frame(N = z$n, Reference = z$arm,
  'Null rejection' = percent(z$rejection),
  '95% Wilson interval' = paste(percent(z$lower), percent(z$upper), sep = '–'),
  check.names = FALSE))
```

Within-group minus raw matched-null power is +0.3 percentage points at N=80
and +0.2 at N=300; the lower bootstrap bounds are about −1.2 percentage points.
These comparisons meet the registered power-loss rule, but do not establish a
power gain. The group-centered parameter covariance is eligible in the tested
scope; neither centered score reference is eligible. Raw remains the provisional
score baseline in the primary joint-sampling scope, as assessed above.
This provisional baseline is not a fixed-allocation recommendation.

### FIML confirmation does not justify centering or validate the prospective geometry

The extension uses 2,000 fresh datasets per cell, separately reporting expected
and observed sensitivity with the same expected quadratic metric. Ordinary FIML
inference remains unsupported; raw is the native default comparator in each
stratum. This comparison holds geometry fixed and cannot choose between geometries.

The following slice shows PEBA4 null rejection at N=80. Both methods, N=300,
coverage and every loss against raw are retained in the frozen
[rates](results/score-centering/fiml_confirmation/rates.csv),
[covariance](results/score-centering/fiml_confirmation/covariance.csv) and
[paired comparisons](results/score-centering/fiml_confirmation/paired.csv).

```{r fiml-evidence}
z <- fiml_rates[fiml_rates$role == 'null' & fiml_rates$n == 80 &
  fiml_rates$method == 'peba4', ]
a <- z[z$arm == 'raw', c('family', 'geometry', 'rejection')]
b <- z[z$arm == 'global', c('family', 'geometry', 'rejection', 'lower', 'upper')]
w <- merge(a, b, by = c('family', 'geometry'), suffixes = c('_raw', '_centered'))
labels <- c(fiml_global_mcar = 'Global, MCAR', fiml_global_mar = 'Global, MAR',
            fiml_nested_mcar = 'Nested, MCAR', fiml_nested_mar = 'Nested, MAR')
tab(data.frame(Family = unname(labels[w$family]), Sensitivity = w$geometry,
  'Raw rejection' = percent(w$rejection_raw),
  'Centered rejection' = percent(w$rejection_centered),
  'Centered 95% interval' = paste(percent(w$lower), percent(w$upper), sep = '–'),
  check.names = FALSE))
```

Centering changes null rejection by at most one percentage point (expected
global MAR, N=80, SB: 3.9% to 4.9%). That cell's paired size-error interval is
−1.35 to +0.45 points, so it does not establish the required one-point benefit.
No other cell meets that benefit rule. The expected-sensitivity candidate passes
the size and power conditions but remains `not_eligible` without a qualifying
benefit. Its raw global and nested MAR N=80 PEBA4 Wilson lower bounds fall below
3%; retaining the raw comparator is not a claim that every baseline cell passes.

All absolute-size-error losses are small but retained: expected MCAR global
N=80 (+0.45 points SB), expected MCAR nested N=80/300 (+0.45/+0.20 points,
both methods), and observed nested MCAR N=80/300 (+0.75/+0.25 points) and MAR
N=80/300 (+0.30/+0.05 points), both methods. Observed nested MCAR N=80 centering
raises rejection from 6.5% to 7.25%, with a 6.19–8.47% Wilson interval and paired
worsening upper bounds of +1.05 points SB and +1.10 PEBA4. Observed global
MCAR/MAR tests reject only 1.5%/0.8% raw and 1.8%/1.0% centered at N=80;
centering does not solve their conservatism. The observed candidate fails the
calibration conditions. Both strata's score-calibration decisions are `not_eligible`.

Matched-null power differences range from −0.10 to +0.05 percentage points;
the lowest paired bootstrap bound is −0.15 points. Parameter covariances agree
to a maximum discrepancy of
`r format(max(fiml_cov$max_covariance_gap), scientific=TRUE, digits=3)`
on the runner's comparison scale. Nested MCAR N=80 null Wald coverage is 92.35%
(Wilson interval 91.10–93.44%), failing the registered coverage condition for
both arms. These results retain raw provisionally as a comparator; they leave
absolute calibration and interval accuracy open, not an ordinary FIML default.

### Bank and reopening conditions

Recorded 2026-10-01 after the ML, grouped and FIML confirmations: this decision
study is **banked in place**, with no queued centered/raw run. Its stable location
is indexed under [Decisions](../../README.md#decisions), and
[Likelihood-score centering alternatives](../../../project/backlog/speculative.md#likelihood-score-centering-alternatives)
owns the reopening trigger. Keep the current raw baseline and explicit lab
centering option. The fixed-allocation formula follows its separate scope contract.

Reopen a bounded comparison for a named supported inference component when a
reproducible case or independent derivation implicates raw score meat in a wrong
sampling covariance, or identifies a centering-specific calibration/numerical
benefit. Derive the covariance target first, preserve the observed test score,
and require fresh paired confirmation. A miscalibrated test or undercovered Wald
interval alone does not implicate centering. FIML policy integration/calibration,
Wald validation and excluded constraint/boundary contracts remain active under
their existing backlog owners; banking this comparison does not validate them.

### Construction and baseline checks

The largest relative score-Gram identity discrepancy is
`r format(checks$max_gram_identity_gap, scientific=TRUE, digits=3)`; the largest
ML policy score/p-value discrepancy is
`r format(checks$max_policy_gap, scientific=TRUE, digits=3)`; the parameter-
covariance discrepancy is
`r format(checks$max_covariance_policy_gap, scientific=TRUE, digits=3)`.
The grouped confirmation's independent replay checks estimates, parameter
covariance, the test statistic and both p-values against elementary Gaussian
mean formulas. All 24,000 arm rows pass; the largest estimate gap is
`r format(max(oracle$estimate_gap), scientific=TRUE, digits=3)` and the largest
variance gap is `r format(max(oracle$variance_gap), scientific=TRUE, digits=3)`.
None of the three confirmations has a failed fit or unavailable covariance/test.
The runner retains convergence failures and unavailable inference instead of
counting them as successful non-rejections. There are `r checks$fit_failures`
failed larger/unrestricted fits. Median task time is
`r sprintf('%.3f', checks$median_task_seconds)` seconds; the runner prints elapsed
wall time and a confirmation cost estimate. The broader pilot took about two
minutes, grouped confirmation about four minutes, and ML confirmation 342 seconds
(5.7 minutes), each with four workers. The native-binary and source hashes stayed
unchanged throughout the ML run; all 32,000 seeds were checked as unused and
every frozen rejection rate was reconstructed from retained raw rows.
FIML confirmation took 309 seconds (5.2 minutes), also with four workers and
32,000 previously unused seeds. Its maximum Gram/projection/numerator identity
gap is `r format(max(fiml_checks[, c('max_gram_identity_gap', 'max_projection_gap', 'max_numerator_gap')]), scientific=TRUE, digits=3)`.
Its source and native hashes stayed unchanged, every rejection rate was replayed
from raw rows, and observed statistics were identical across centering arms
within each sensitivity geometry. FIML has no ordinary-policy parity comparator;
the zero policy-gap fields in its checks are placeholders, not evidence of parity.

## Caveats

This grid covers regular ML and direct FIML, two sample sizes, one fixed-allocation
misspecified grouped mean family, normal/skewed complete-data innovations and
normal MCAR/MAR controls. It does not validate PSD boundary or penalized inference,
ordinal routes, arbitrary misspecified nested SEM, or growing dimension. It does
not reopen established moment Gamma/NACOV or LR-reference recipes. The grouped
algebraic witness does not establish which single-group spectral calibration has
better finite-sample behavior.

All three confirmations use the same frozen 0.1.0 prerelease native binary;
these results describe that implementation rather than later source edits.
The actual loaded package paths, versions and native-binary hashes are recorded.
Repository HEAD is separate provenance; it is not asserted to be the source of an
arbitrarily installed binary. Failed construction identities block interpretation.
The initial mechanics-only smoke started before the criteria commit (despite the original criteria header) and is
retained as development data; it supplies no decision evidence. No default is
changed by this report.

## Reproduce

Install matched magmaan/magmaanlab >=0.1.0 packages. An isolated installation can
be selected with `--library PATH`; otherwise the normal R library is used.

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --library PATH
Rscript run_experiment.R --kind pilot --reps 100 --run-id pilot --library PATH
Rscript run_experiment.R --kind confirmation --reps 2000 --families grouped_mean --run-id group_confirmation --library PATH
Rscript scripts/validate_grouped_results.R group_confirmation
Rscript run_experiment.R --kind confirmation --reps 2000 --families global_normal,global_skew,nested_normal,nested_skew --run-id ml_confirmation --library PATH
Rscript run_experiment.R --kind confirmation --reps 2000 --families fiml_global_mcar,fiml_global_mar,fiml_nested_mcar,fiml_nested_mar --run-id fiml_confirmation --library PATH
quarto render report.qmd
```

Each run directory is immutable. `--summarize --run-id ID` rescans its existing
local raw batches without fitting or replacing the design. Reports read only
this study's result CSVs and never run estimation or simulation.
