---
title: "Likelihood-score covariance centering"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
run_id <- Sys.getenv('SCORE_CENTERING_RUN', 'pilot')
root <- file.path('results', 'score-centering', run_id)
rd <- function(name) {
  path <- file.path(root, paste0(name, '.csv'))
  if (!file.exists(path)) stop('Missing ', path,
    '; run Rscript run_experiment.R --library PATH --run-id ', run_id)
  read.csv(path, stringsAsFactors = FALSE)
}
meta <- rd('metadata'); rates <- rd('rates'); cov <- rd('covariance')
pairs <- rd('paired'); checks <- rd('checks'); decisions <- rd('decisions')
value <- function(key) meta$value[match(key, meta$key)]
percent <- function(x) sprintf('%.1f%%', 100 * x)
confirm_root <- file.path('results', 'score-centering', 'group_confirmation')
confirm <- function(name) {
  path <- file.path(confirm_root, paste0(name, '.csv'))
  if (!file.exists(path)) stop('Missing ', path,
    '; run the grouped confirmation and independent validator shown under Reproduce')
  read.csv(path, stringsAsFactors = FALSE)
}
group_cov <- confirm('covariance'); group_rates <- confirm('rates')
group_pairs <- confirm('paired'); group_checks <- confirm('checks')
group_decisions <- confirm('decisions'); oracle <- confirm('oracle_checks')
group <- group_cov[group_cov$role == 'null', ]
tab <- function(x, ...) knitr::kable(x, row.names = FALSE, ...)
```

## Default register

| Choice | Current value | Status |
|:--|:--|:--|
| ML parameter covariance | Uncentered exact likelihood-score meat, observed bread | Current joint-sampling baseline; fixed allocation has a different target |
| Fixed-allocation score covariance | Within-group centered covariance rows | Formula settled; ordinary-policy extension banked |
| ML global/nested score reference | Uncentered efficient scores, expected geometry | Open centered/raw decision |
| Ordinary FIML inference | Unsupported | Prospective evidence only |
| Established moment-covariance and LR centering | Retained method-specific formulas | Redesign banked |

## Question

Can magmaan use uncentered likelihood-score second moments throughout its
parameter covariance and score-test calibration, or does the sampling law
require a distinction? All covariance arms use the same estimates, projection,
information matrices and observed score. The
[criteria](criteria/score_centering.md) were committed before the pilot and confirmation draws.

## Short answer

**A universal never-center rule fails for fixed groups with different population
score means.** In the independently specified grouped mean model, the true
parameter variance is exactly $1/N$. Raw and globally centered score meats have
an asymptotic variance ratio of 1.12; centering within the sampling groups removes
that extra term. This conclusion follows from the sampling covariance identity,
not from selecting an arm on simulation results.

The 8,000-draw independent grouped confirmation found a variance ratio of
**1.118 for raw/global centering versus 0.994 for within-group centering** at
N=300. Within-group parameter covariance meets the registered criteria in this
scope. The centered score test does not: at N=80 it rejects 6.2% under the null,
with a 95% interval of 5.2–7.3%, failing the calibration rule. Matched-null power
is similar. Fixing the sampling covariance alone does not settle small-sample
reference calibration.

The broader 3,600-draw pilot covers normal/skewed ML and MCAR/MAR FIML.
**It cannot choose a package default**; full independent confirmation remains
open. Stationary single-group parameter scores have approximately zero means,
whereas released test directions carry an observed signal. FIML evidence is
prospective, split by sensitivity geometry. No package default is changed.

## Evidence

### Disposition of the fixed-allocation case

Recorded 2026-10-01 after inspecting results: the fixed-allocation covariance
formula is settled by the sampling law, and its ordinary-policy extension is
[banked](../../../project/backlog/speculative.md#fixed-design-inference-under-mean-misspecification).
The [scope contract](../../../project/scope.md#group-allocation-and-likelihood-score-covariance)
distinguishes within-group covariance for fixed allocation from raw score second
moments for joint sampling of group labels. No further experiment is needed to
choose between those two population formulas. Finite-sample reference
calibration remains separate; the N=80 result cannot make raw covariance correct
for a fixed-allocation target. The original criteria and frozen decision outcomes
are retained. The other ML/FIML calibration lanes remain active.

The grouped nested test targets the best-fitting shared mean, not equality of
the two true group means. Both nested models retain the wrong shared-mean
assumption; the additional restriction fixes that approximation mean at zero.
It therefore illustrates inference within misspecified models, not a correctly
specified full-model null.

### Grouped parameter covariance has an independent target

The groups have fixed sizes $N/4$ and $3N/4$, unit observation variance, and X
means $\mu+.6$ and $\mu-.2$. The fitted shared mean is the pooled average, whose
approximation target is $\mu$ and whose variance is exactly $1/N$. The between-
group squared mean displacement is $.25(.6)^2+.75(-.2)^2=.12$.

At the fitted mean, with divisor $N$, the expected reported variance ratios are

$$
E[N\widehat V_{\rm raw}]=E[N\widehat V_{\rm global}]=1.12-1/N,
\qquad E[N\widehat V_{\rm group}]=1-2/N.
$$

The second expression's small finite-sample bias is retained; this experiment
does not introduce an $N-1$ or group degrees-of-freedom correction.

```{r grouped-evidence}
tab(data.frame(
  N = group$n, Covariance = group$arm, Usable = group$usable,
  'Variance / true variance' = round(group$oracle_ratio, 3),
  'Expected ratio' = round(ifelse(group$arm == 'group', 1 - 2 / group$n, 1.12 - 1 / group$n), 3),
  '95% interval coverage' = percent(group$coverage), check.names = FALSE))
```

The raw-versus-group difference persists with sample size. Global score-mean
centering cannot remove it because the opposing group means cancel overall.
This fixed-allocation example establishes a scope limit for a universal raw-meat
policy, without generalizing group centering to random missingness patterns.

### Centered grouped scores fail the small-sample acceptance rule

The confirmation compares 2,000 paired null draws at each sample size. SB and
PEBA4 coincide in this one-dimensional case. The N=80 centered interval exceeds
the registered 7% upper limit, and its paired size-error comparison also fails
the allowed one-percentage-point worsening rule.

```{r grouped-score-evidence}
z <- group_rates[group_rates$role == 'null' & group_rates$method == 'peba4', ]
tab(data.frame(N = z$n, Reference = z$arm,
  'Null rejection' = percent(z$rejection),
  '95% Wilson interval' = paste(percent(z$lower), percent(z$upper), sep = '–'),
  check.names = FALSE))
```

Within-group minus raw matched-null power is +0.3 percentage points at N=80
and +0.2 at N=300; the lower bootstrap bounds are about −1.2 percentage points.
These comparisons meet the registered power-loss rule, but do not establish a
power gain. The group-centered parameter covariance is eligible in the tested
scope; neither centered score reference is eligible. Raw remains the provisional
score baseline in the primary joint-sampling scope, with calibration still open.
This provisional baseline is not a fixed-allocation recommendation.

### Broader pilot: differences need confirming

Across the single-group pilot fits, globally centering parameter scores changes
covariance matrices by at most
`r format(max(cov$max_covariance_gap[cov$family != 'grouped_mean' & cov$arm == 'global']), scientific=TRUE, digits=3)`
relative to the comparison scale. This supports the expected stationary-fit
equivalence. Score-test reference changes require their own comparison.

The following pilot slice shows PEBA4 null rejection at N=80. FIML uses observed
sensitivity here; its expected-sensitivity comparator and both methods at N=300
remain in the per-cell CSVs. The paired tables also report SB and matched-null
power, with bootstrap uncertainty including re-estimation of null thresholds.

```{r score-evidence}
slice <- rates[rates$role == 'null' & rates$n == 80 & rates$method == 'peba4' &
  ((rates$lane == 'ml' & rates$geometry == 'expected') |
   (rates$lane == 'fiml' & rates$geometry == 'observed')), ]
wide <- reshape(slice[, c('family', 'arm', 'rejection')], idvar = 'family',
                timevar = 'arm', direction = 'wide')
names(wide) <- sub('rejection.', '', names(wide), fixed = TRUE)
for (name in setdiff(names(wide), 'family')) wide[[name]] <- percent(wide[[name]])
tab(wide)
```

The registered decision outcomes for the broader run are
`r paste(unique(decisions$status), collapse = ', ')`. Precision, availability,
coverage, per-cell calibration and matched-power conditions must all be assessed
before changing a tested score-reference default. Full rates, Wilson intervals,
paired comparisons and every failure are retained in `rates.csv`,
`covariance.csv`, `paired.csv`, `decisions.csv` and `failures.csv` for this run.

### Construction and baseline checks

The largest relative score-Gram identity discrepancy is
`r format(checks$max_gram_identity_gap, scientific=TRUE, digits=3)`; the largest
ML policy score/p-value discrepancy is
`r format(checks$max_policy_gap, scientific=TRUE, digits=3)`; the parameter-
covariance discrepancy is
`r format(checks$max_covariance_policy_gap, scientific=TRUE, digits=3)`.
The grouped confirmation's independent replay checks estimates, parameter
covariance, the test statistic and both p-values against elementary Gaussian
mean formulas. All 24,000 arm rows pass; the largest estimate gap is
`r format(max(oracle$estimate_gap), scientific=TRUE, digits=3)` and the largest
variance gap is `r format(max(oracle$variance_gap), scientific=TRUE, digits=3)`.
Neither run has a failed fit or unavailable covariance/test.
The runner retains convergence failures and unavailable inference instead of
counting them as successful non-rejections. There are `r checks$fit_failures`
failed larger/unrestricted fits. Median task time is
`r sprintf('%.3f', checks$median_task_seconds)` seconds; the runner prints elapsed
wall time and a confirmation cost estimate. The broader pilot took about two
minutes and the grouped confirmation about four minutes with four workers.
The full 36-cell, 2,000-draw confirmation is estimated at roughly 35–40 minutes
on this machine.

## Caveats

This grid covers regular ML and direct FIML, two sample sizes, one fixed-allocation
misspecified grouped mean family, normal/skewed complete-data innovations and
normal MCAR/MAR controls. It does not validate PSD boundary or penalized inference,
ordinal routes, arbitrary misspecified nested SEM, or growing dimension. It does
not reopen established moment Gamma/NACOV or LR-reference recipes. The grouped
algebraic witness does not establish which single-group spectral calibration has
better finite-sample behavior.

The actual loaded package paths, versions and native-binary hashes are recorded.
Repository HEAD is separate provenance; it is not asserted to be the source of an
arbitrarily installed binary. Failed construction identities block interpretation.
The initial mechanics-only smoke started before the criteria commit (despite the original criteria header) and is
retained as development data; it supplies no decision evidence. No default is
changed by this report.

## Reproduce

Install matched magmaan/magmaanlab >=0.1.0 packages. An isolated installation can
be selected with `--library PATH`; otherwise the normal R library is used.

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --library PATH
Rscript run_experiment.R --kind pilot --reps 100 --run-id pilot --library PATH
Rscript run_experiment.R --kind confirmation --reps 2000 --families grouped_mean --run-id group_confirmation --library PATH
Rscript scripts/validate_grouped_results.R group_confirmation
quarto render report.qmd
# Fresh confirming draws; estimate cost from the pilot before launching:
Rscript run_experiment.R --kind confirmation --reps 2000 --run-id confirmation --library PATH
SCORE_CENTERING_RUN=confirmation quarto render report.qmd
```

Each run directory is immutable. `--summarize --run-id ID` rescans its existing
local raw batches without fitting or replacing the design. Reports read only
this study's result CSVs and never run estimation or simulation.
