---
title: "Likelihood-score covariance centering"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
run_id <- Sys.getenv('SCORE_CENTERING_RUN', 'ml_confirmation')
root <- file.path('results', 'score-centering', run_id)
rd <- function(name) {
  path <- file.path(root, paste0(name, '.csv'))
  if (!file.exists(path)) stop('Missing ', path,
    '; run Rscript run_experiment.R --library PATH --run-id ', run_id)
  read.csv(path, stringsAsFactors = FALSE)
}
meta <- rd('metadata'); rates <- rd('rates'); cov <- rd('covariance')
pairs <- rd('paired'); checks <- rd('checks'); decisions <- rd('decisions')
value <- function(key) meta$value[match(key, meta$key)]
percent <- function(x) sprintf('%.1f%%', 100 * x)
pilot_root <- file.path('results', 'score-centering', 'pilot')
pilot_rates <- read.csv(file.path(pilot_root, 'rates.csv'), stringsAsFactors = FALSE)
confirm_root <- file.path('results', 'score-centering', 'group_confirmation')
confirm <- function(name) {
  path <- file.path(confirm_root, paste0(name, '.csv'))
  if (!file.exists(path)) stop('Missing ', path,
    '; run the grouped confirmation and independent validator shown under Reproduce')
  read.csv(path, stringsAsFactors = FALSE)
}
group_cov <- confirm('covariance'); group_rates <- confirm('rates')
group_pairs <- confirm('paired'); group_checks <- confirm('checks')
group_decisions <- confirm('decisions'); oracle <- confirm('oracle_checks')
group <- group_cov[group_cov$role == 'null', ]
tab <- function(x, ...) knitr::kable(x, row.names = FALSE, ...)
```

## Default register

| Choice | Current value | Status |
|:--|:--|:--|
| ML parameter covariance | Uncentered exact likelihood-score meat, observed bread | Stationary equivalence confirmed; skewed-data Wald coverage remains open |
| Fixed-allocation score covariance | Within-group centered covariance rows | Formula settled; ordinary-policy extension banked |
| ML global/nested score reference | Uncentered efficient scores, expected geometry | Retain raw under the registered rule in tested regular ML families |
| Ordinary FIML inference | Unsupported | Prospective evidence only |
| Established moment-covariance and LR centering | Retained method-specific formulas | Redesign banked |

## Question

Can magmaan use uncentered likelihood-score second moments throughout its
parameter covariance and score-test calibration, or does the sampling law
require a distinction? All covariance arms use the same estimates, projection,
information matrices and observed score. The
[criteria](criteria/score_centering.md) were committed before the pilot and confirmation draws.

## Short answer

**Retain uncentered score references for the tested regular complete-data ML
families.** Independent confirmation used 32,000 datasets across global/nested
tests, normal/skewed innovations and N=80/300. Globally centering covariance rows
changed null rejection by at most 0.75 percentage points, with identical empirical
matched-null power in all SB/PEBA4 comparisons. It did not meet the preregistered
benefit or noninferiority conditions for a default change. Fits, information,
projection and observed statistics were held fixed; no default changes.

Parameter covariances were numerically identical across arms, but both Wald
intervals undercovered in skewed-data cells: nested null coverage at N=80 was
90.3% (95% Wilson interval 88.9–91.5%). This is a separate covariance/interval
validation task; score centering cannot correct it.

The fixed-allocation covariance formula is settled separately: within-group
centering removes score-mean terms that are absent from conditional sampling
variance. Its ordinary-policy extension remains banked. The 3,600-draw pilot also
covers prospective MCAR/MAR FIML; that lane still needs independent confirmation.
PSD boundary and penalty inference are outside this grid.

## Evidence

### ML confirmation does not justify switching to global centering

Each null and power cell contains 2,000 usable pairs. The following slice shows
PEBA4 null rejection; all raw and centered SB/PEBA4 null Wilson intervals lie
inside the registered 3–7% band. The full rates, coverage, failures and paired
uncertainty are frozen in this run's CSVs.

```{r ml-confirmation}
z <- rates[rates$lane == 'ml' & rates$family != 'grouped_mean' &
  rates$role == 'null' & rates$method == 'peba4', ]
a <- z[z$arm == 'raw', c('family', 'n', 'rejection')]
b <- z[z$arm == 'global', c('family', 'n', 'rejection', 'lower', 'upper')]
w <- merge(a, b, by = c('family', 'n'), suffixes = c('_raw', '_centered'))
labels <- c(global_normal = 'Global, normal', global_skew = 'Global, skewed',
            nested_normal = 'Nested, normal', nested_skew = 'Nested, skewed')
tab(data.frame(Family = unname(labels[w$family]), N = w$n,
  'Raw rejection' = percent(w$rejection_raw),
  'Centered rejection' = percent(w$rejection_centered),
  'Centered 95% interval' = paste(percent(w$lower), percent(w$upper), sep = '–'),
  check.names = FALSE))
```

The registered benefit rule requires a paired 95% interval establishing at least
a one-percentage-point reduction in absolute size error in some cell. No cell
meets it. Absolute-size-error losses against raw were global normal N=300
(+0.10 points SB, +0.20 PEBA4), global skewed N=300 (+0.05 both), and global
skewed N=80 (+0.75 SB). The global N=80 SB paired upper bounds also exceed the
allowed one-point worsening margin (normal +1.001, skewed +1.05).

Matched-null power estimates are identical across arms in every comparison;
the lowest paired bootstrap bound for the difference is −0.15 percentage
points, within the registered two-point loss allowance. Both candidate components
are `not_eligible` in the frozen decision table. The implied outcome is to retain
raw provisionally in the tested scope, without claiming a blanket calibration
guarantee across models or sampling laws.

### Stationary parameter covariance agrees; Wald coverage needs investigation

The maximum raw/global parameter-covariance discrepancy is
`r format(max(cov$max_covariance_gap[cov$arm == 'global']), scientific=TRUE, digits=3)`
on the runner's comparison scale. Known-target Wald intervals are therefore the
same. Both arms fail the coverage rule in several cells, most visibly under
skewed innovations: nested null N=80 coverage is 90.3%, and nested power N=80
coverage is 90.25%. Finite-sample covariance bias, estimate bias and interval
calibration require separate checks against the declared population targets.
The nominal-size result for score tests does not establish Wald interval coverage.

### Disposition of the fixed-allocation case

Recorded 2026-10-01 after inspecting results: the fixed-allocation covariance
formula is settled by the sampling law, and its ordinary-policy extension is
[banked](../../../project/backlog/speculative.md#fixed-design-inference-under-mean-misspecification).
The [scope contract](../../../project/scope.md#group-allocation-and-likelihood-score-covariance)
distinguishes within-group covariance for fixed allocation from raw score second
moments for joint sampling of group labels. No further experiment is needed to
choose between those two population formulas. Finite-sample reference
calibration remains separate; the N=80 result cannot make raw covariance correct
for a fixed-allocation target. The original criteria and frozen decision outcomes
are retained. Prospective FIML calibration and excluded inference regimes remain open.

The grouped nested test targets the best-fitting shared mean, not equality of
the two true group means. Both nested models retain the wrong shared-mean
assumption; the additional restriction fixes that approximation mean at zero.
It therefore illustrates inference within misspecified models, not a correctly
specified full-model null.

### Grouped parameter covariance has an independent target

The groups have fixed sizes $N/4$ and $3N/4$, unit observation variance, and X
means $\mu+.6$ and $\mu-.2$. The fitted shared mean is the pooled average, whose
approximation target is $\mu$ and whose variance is exactly $1/N$. The between-
group squared mean displacement is $.25(.6)^2+.75(-.2)^2=.12$.

At the fitted mean, with divisor $N$, the expected reported variance ratios are

$$
E[N\widehat V_{\rm raw}]=E[N\widehat V_{\rm global}]=1.12-1/N,
\qquad E[N\widehat V_{\rm group}]=1-2/N.
$$

The second expression's small finite-sample bias is retained; this experiment
does not introduce an $N-1$ or group degrees-of-freedom correction.

```{r grouped-evidence}
tab(data.frame(
  N = group$n, Covariance = group$arm, Usable = group$usable,
  'Variance / true variance' = round(group$oracle_ratio, 3),
  'Expected ratio' = round(ifelse(group$arm == 'group', 1 - 2 / group$n, 1.12 - 1 / group$n), 3),
  '95% interval coverage' = percent(group$coverage), check.names = FALSE))
```

The raw-versus-group difference persists with sample size. Global score-mean
centering cannot remove it because the opposing group means cancel overall.
This fixed-allocation example establishes a scope limit for a universal raw-meat
policy, without generalizing group centering to random missingness patterns.

### Centered grouped scores fail the small-sample acceptance rule

The confirmation compares 2,000 paired null draws at each sample size. SB and
PEBA4 coincide in this one-dimensional case. The N=80 centered interval exceeds
the registered 7% upper limit, and its paired size-error comparison also fails
the allowed one-percentage-point worsening rule.

```{r grouped-score-evidence}
z <- group_rates[group_rates$role == 'null' & group_rates$method == 'peba4', ]
tab(data.frame(N = z$n, Reference = z$arm,
  'Null rejection' = percent(z$rejection),
  '95% Wilson interval' = paste(percent(z$lower), percent(z$upper), sep = '–'),
  check.names = FALSE))
```

Within-group minus raw matched-null power is +0.3 percentage points at N=80
and +0.2 at N=300; the lower bootstrap bounds are about −1.2 percentage points.
These comparisons meet the registered power-loss rule, but do not establish a
power gain. The group-centered parameter covariance is eligible in the tested
scope; neither centered score reference is eligible. Raw remains the provisional
score baseline in the primary joint-sampling scope, as assessed above.
This provisional baseline is not a fixed-allocation recommendation.

### FIML pilot still needs independent confirmation

The pilot uses 100 draws per cell. Its FIML lane remains prospective and cannot
choose an ordinary-package default.

The following pilot slice shows PEBA4 null rejection at N=80. FIML uses observed
sensitivity here; its expected-sensitivity comparator and both methods at N=300
remain in the per-cell CSVs. The paired tables also report SB and matched-null
power, with bootstrap uncertainty including re-estimation of null thresholds.

```{r score-evidence}
slice <- pilot_rates[pilot_rates$role == 'null' & pilot_rates$n == 80 &
  pilot_rates$method == 'peba4' & pilot_rates$lane == 'fiml' &
  pilot_rates$geometry == 'observed', ]
wide <- reshape(slice[, c('family', 'arm', 'rejection')], idvar = 'family',
                timevar = 'arm', direction = 'wide')
names(wide) <- sub('rejection.', '', names(wide), fixed = TRUE)
for (name in setdiff(names(wide), 'family')) wide[[name]] <- percent(wide[[name]])
tab(wide)
```

Pilot outcomes remain `open_pilot_only`; they do not choose FIML defaults.
Precision, availability, coverage, per-cell calibration and matched-power
conditions need independent confirmation before changing a prospective FIML
reference. Full summaries and failures are retained separately for each run.

### Construction and baseline checks

The largest relative score-Gram identity discrepancy is
`r format(checks$max_gram_identity_gap, scientific=TRUE, digits=3)`; the largest
ML policy score/p-value discrepancy is
`r format(checks$max_policy_gap, scientific=TRUE, digits=3)`; the parameter-
covariance discrepancy is
`r format(checks$max_covariance_policy_gap, scientific=TRUE, digits=3)`.
The grouped confirmation's independent replay checks estimates, parameter
covariance, the test statistic and both p-values against elementary Gaussian
mean formulas. All 24,000 arm rows pass; the largest estimate gap is
`r format(max(oracle$estimate_gap), scientific=TRUE, digits=3)` and the largest
variance gap is `r format(max(oracle$variance_gap), scientific=TRUE, digits=3)`.
Neither confirmation has a failed fit or unavailable covariance/test.
The runner retains convergence failures and unavailable inference instead of
counting them as successful non-rejections. There are `r checks$fit_failures`
failed larger/unrestricted fits. Median task time is
`r sprintf('%.3f', checks$median_task_seconds)` seconds; the runner prints elapsed
wall time and a confirmation cost estimate. The broader pilot took about two
minutes, grouped confirmation about four minutes, and ML confirmation 342 seconds
(5.7 minutes), each with four workers. The native-binary and source hashes stayed
unchanged throughout the ML run; all 32,000 seeds were checked as unused and
every frozen rejection rate was reconstructed from retained raw rows.

## Caveats

This grid covers regular ML and direct FIML, two sample sizes, one fixed-allocation
misspecified grouped mean family, normal/skewed complete-data innovations and
normal MCAR/MAR controls. It does not validate PSD boundary or penalized inference,
ordinal routes, arbitrary misspecified nested SEM, or growing dimension. It does
not reopen established moment Gamma/NACOV or LR-reference recipes. The grouped
algebraic witness does not establish which single-group spectral calibration has
better finite-sample behavior.

The actual loaded package paths, versions and native-binary hashes are recorded.
Repository HEAD is separate provenance; it is not asserted to be the source of an
arbitrarily installed binary. Failed construction identities block interpretation.
The initial mechanics-only smoke started before the criteria commit (despite the original criteria header) and is
retained as development data; it supplies no decision evidence. No default is
changed by this report.

## Reproduce

Install matched magmaan/magmaanlab >=0.1.0 packages. An isolated installation can
be selected with `--library PATH`; otherwise the normal R library is used.

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --library PATH
Rscript run_experiment.R --kind pilot --reps 100 --run-id pilot --library PATH
Rscript run_experiment.R --kind confirmation --reps 2000 --families grouped_mean --run-id group_confirmation --library PATH
Rscript scripts/validate_grouped_results.R group_confirmation
Rscript run_experiment.R --kind confirmation --reps 2000 --families global_normal,global_skew,nested_normal,nested_skew --run-id ml_confirmation --library PATH
quarto render report.qmd
# Fresh confirming draws; estimate cost from the pilot before launching:
Rscript run_experiment.R --kind confirmation --reps 2000 --run-id confirmation --library PATH
SCORE_CENTERING_RUN=confirmation quarto render report.qmd
```

Each run directory is immutable. `--summarize --run-id ID` rescans its existing
local raw batches without fitting or replacing the design. Reports read only
this study's result CSVs and never run estimation or simulation.
