---
title: "Ordinal functional inference: delta and fitted-profile lanes"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
read_evidence <- function(name) {
  path <- file.path("results", "overview", name)
  if (!file.exists(path)) stop("Missing retained summary: ", path,
                               "; recover original CSVs and run scripts/summarize.R")
  read.csv(path, stringsAsFactors = FALSE)
}
show_table <- function(x, digits = 3) knitr::kable(x, digits = digits, row.names = FALSE)
```

## Question

How do delta and profile intervals behave for ordinal functionals, and which
target and numerical failures explain their apparent calibration?

## Short answer

Keep one **active ordinal-functional-inference study** with two distinct targets.
The delta lane studies fit-free polychoric omega; the profile lane studies omega
of a fitted DWLS one-factor approximation. The substantive local runs support
continued work on sparse-category failures and profile endpoints. A promising
pointwise test does not establish a usable confidence interval.

The next check is a focused endpoint audit with failures, connectedness and
independent pseudo-target precision recorded. No pooled comparison or default
selection is justified by these different models and targets.

## Evidence

### Delta lane: Gaussian target and stress pseudo-targets

The four-indicator delta coverage run has **200 attempts per cell**, $N=50,100,250$
and balanced/extreme thresholds. The retained output is beyond smoke size:

```{r delta}
x <- read_evidence("delta_gaussian.csv")
show_table(x[, c("regime", "n", "reps", "failures", "coverage_95", "bias", "mean_width_95")])
```

The separate 200-attempt stress run uses skewed latent factors, $t$ innovations
and local dependence. Coverage below is around each fit-free pseudo-target,
estimated from an independent large sample of size 100,000, rather than an
assumed generating reliability:

```{r delta-stress}
x <- read_evidence("delta_stress.csv")
show_table(x[, c("dgp", "cuts", "n", "reps", "failures", "target_pseudo", "coverage_pseudo_95")])
```

### Profile lane: fitted target and failed inversion

The profile model has six indicators with a different loading schedule. Its
pseudo-target is the functional of the fitted DWLS approximation; the sample
polychoric target is retained as a distinct comparator. Under misspecification
these differ even within the same population.

The retained main stress run has **100 attempts per cell at $N=50$**, two
misspecified DGPs and extreme thresholds. Ordinary, robust-scaled and
misspecification-scaled pointwise tests share a cell but may have different
availability. Inversion was requested only for selected references:

```{r profile}
x <- read_evidence("profile_inversion.csv")
show_table(x[, c("dgp", "reference", "reps", "usable", "reference_coverage_95",
                 "ci_attempted", "ci_successful", "ci_failures", "ci_coverage_95")])
```

For strong local dependence, misspecification-scaled pointwise acceptance is
about .924 on 92 usable fits, whereas CI inversion fails on **91 of 92 attempts**.
The remaining single interval's conditional coverage of 1 says nothing about
calibration. Endpoint residuals and pointwise/interval disagreements are retained
in [the overview](results/overview/profile_inversion.csv), alongside targets.

A separate **300-attempt** pointwise-only grid covers $N=25,35,50,75$ and
extreme/sparse thresholds. At $N=25$ some cells lose nearly two thirds of fits.
The [full table](results/overview/profile_pointwise.csv) records each reference's
usable denominator. Those outputs are not endpoint-coverage evidence.

## Caveats

Pseudo-target Monte Carlo error needs independent confirmation and enters neither
lane as known truth. Sparse categories affect both the summary estimator and its
influence calculation. The $t_3$ stress law has finite variance but lacks the
fourth moments used by some normality-based arguments; an empirical success
there cannot establish those arguments. Conditional coverage among available
fits must stay separate from attempted-fit and endpoint availability.

The remaining work is to isolate numerical endpoint failures and the validity
of the local quadratic/reference law before enlarging a grid or choosing a
correction. Different indicator counts, weights, targets and seed schedules
remain lane-specific.

## Reproduce

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --lane delta-coverage --help
Rscript run_experiment.R --lane delta-stress --help
Rscript run_experiment.R --lane profile --help
Rscript run_experiment.R --lane delta-coverage --smoke --reps 2 --n-grid 100 \
  --results-dir /tmp/ordinal_delta_check
Rscript run_experiment.R --lane profile --smoke --reps 2 --n-grid 100 \
  --dgp one_factor --cut-regimes balanced --pseudo-n 2000 --ci \
  --results-dir /tmp/ordinal_profile_check
Rscript scripts/summarize.R
```

Historical metadata and result files are unchanged under `results/delta` and
`results/profile`. [Source hashes](results/overview/sources.csv) identify the
retained summaries. Rendering performs no fits or simulations.
