---
title: "Measurement-invariance difference tests under non-normal incomplete data: FMG/pEBA vs the MLR/SB default"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
res_dir <- "results"
sum_path <- file.path(res_dir, "summary_rejection.csv")
have_results <- file.exists(sum_path)
S <- if (have_results) read.csv(sum_path, stringsAsFactors = FALSE) else NULL
meta <- if (file.exists(file.path(res_dir, "metadata.csv")))
  read.csv(file.path(res_dir, "metadata.csv"), stringsAsFactors = FALSE) else NULL
reps <- if (!is.null(meta) && "reps" %in% meta$key) meta$value[meta$key == "reps"] else "?"
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

> **Status: DESIGN / pre-registration draft.** This document is the design we
> are pinning down before lighting up the compute. No calibration run has been
> committed yet; the Results section renders only once `run_experiment.R` has
> written `results/summary_rejection.csv`. It is meant to be read and torn apart
> (hi Njål) before we fix the grid.

## Question

Under **non-normality** and **missing data**, do the Foldnes-Moss-Gronneberg
eigenvalue p-values (SS / EBA / pEBA / pall / pOLS) calibrate the
**measurement-invariance difference test** better than the dominant applied
defaults, and do they keep that edge under the two estimators practitioners
actually use for incomplete data, **FIML** and **two-stage ML (ML2S)**?

The applied default for a nested invariance step (e.g. configural vs metric) is
the scaled chi-square difference test: lavaan's `estimator = "MLR"` plus the
Satorra-Bentler difference correction. Brace & Savalei (2017) is the canonical
simulation evaluation of that family on **complete non-normal** data; they
crowned the strictly-positive correction $D_{SB10}$ (Satorra & Bentler 2010).
The difference statistic is, like the omnibus statistic, asymptotically a
weighted sum of $\chi^2_1$ variables whose eigenvalues spread under
non-normality; the SB corrections match only the **mean** (or mean and
variance) of that spectrum, whereas pEBA approximates the whole spectrum. Our
own nested-pEBA result (experiment 18 / FMG 2026) showed the spectrum methods
win on **complete** data. **The open cell is missing data.** This experiment is
that cell.

## Where this sits in the literature

The robust-invariance-test literature splits cleanly, and our contribution is
the one combination nobody has run:

| | complete data | **incomplete (FIML / ML2S)** |
|---|---|---|
| scaled (SB / MLR) difference test | Brace & Savalei (2017) | ordinal-missing & FIML-vs-MI papers |
| **eigenvalue-spectrum (pEBA / FMG) difference test** | FMG (2026), exp 18 | **this experiment** |

The scaled-test missing-data work (e.g. Yang et al. on ordinal missing
invariance; FIML-vs-multiple-imputation comparisons) never uses an
eigenvalue-spectrum p-value; the nested-pEBA paper never touches missingness.
The estimator axis is deliberately two-wide because Savalei & Falk (2014)
showed **robust two-stage ML can beat robust FIML** under incomplete non-normal
data, so ML2S is not an afterthought; it is a predictable referee question we
answer head-on.

This is one focused story in the FMG/FIML program, a sibling of:

- **exp 21** -- FIML FMG feature-completeness audit for invariance (validation).
- **exp 23** -- FIML FMG vs MLR goodness-of-fit + nested, non-normality + MCAR.
- **exp 24** -- FIML and ML2S FMG goodness-of-fit under incomplete non-normal data.
- **exp 18** -- FMG 2026 nested-pEBA replication (complete data).

The scope fence (stated so reviewers do not ask): **continuous MG-CFA only**
(ordinal invariance is the companion `ordinal-fmg` paper); **measurement-model
invariance** (no structural-path invariance); single-occasion (no longitudinal
/ alignment / MNLFA). Each of those is named future work, not a gap.

## Models and the invariance ladder

Multi-group CFA **with mean structure** is the canonical and durable testbed for
measurement invariance (Meredith 1993; Vandenberg & Lance 2000; Millsap 2011).
Mean structure is non-negotiable here for two reasons: (1) the scalar rung *is*
about intercepts, and a loadings-only design earns an automatic "what about
scalar?"; (2) MAR bias concentrates in the means/intercepts and FIML's whole
selling point over listwise is unbiased means under MAR, so the mean structure
is where the missing-data contrast actually bites.

**Population.** Following the Brace & Savalei structural precedent: a two-group,
two-factor CFA, factor correlation $\phi = 0.5$, with model size

$$p \in \{8,\ 16,\ 30\}\quad(\text{2 factors} \times \{4, 8, 15\}\ \text{indicators}),$$

where $p = 30$ is the high-dimensional cell that is **pEBA's home turf** (FMG
2026 ran to 40 observed variables; the SB corrections degrade fastest exactly
there). A one-factor six-indicator reduction (identical to exp 21/23) is the
minimal smoke model.

**Ladder and difference tests.** configural $\to$ metric (loadings) $\to$
scalar (intercepts) $\to$ strict (residuals). Three difference tests:

| step | constrains | "is invariant at the …" |
|---|---|---|
| configural vs metric | loadings | weak / metric level |
| metric vs scalar | + intercepts | strong / scalar level |
| scalar vs strict | + residual variances | strict level |

**Type-I vs power, from one factor.** As in Brace & Savalei, the
*noninvariance-location* factor is the power machinery and the fully-invariant
population is the Type-I cell:

- **null** (fully invariant population) $\Rightarrow$ Type-I rate of each step.
- **targeted violation at a rung** $\Rightarrow$ power of that step (a planted
  loading / intercept / residual difference in group 2).

**What we add over Brace & Savalei.** They used a *single* effect size
(one $\Delta\text{RMSEA}$ per rung) and reported difference-test power without
size adjustment. We add:

1. an **effect-size gradient** (several $\Delta\text{RMSEA}$ levels including 0)
   so we get power *curves*, not points;
2. **partial invariance** as a factor (number of noninvariant indicators: 1 vs
   several), which both pre-empts the partial-invariance nag and is realistic;
3. **size-adjusted power** throughout (empirical-$\alpha$ calibration against the
   paired null cell), because tests with different Type-I are otherwise not
   comparable, exactly the conclusion exp 23 reached.

## Estimators and missingness

**Estimators (expensive axis, refit per cell):**

- **FIML** -- `magmaan::magmaan_core$fit_fiml` (lavaan `missing = "ml"` analogue).
- **ML2S** -- `magmaan::magmaan_core$fit_ml2s`, two-stage ML (saturated EM stage
  1, structured stage 2; lavaan `missing = "two.stage"`). Cheaper per ladder
  than FIML (stage-1 saturated fit + its ACOV are shared across all four rungs)
  and Savalei & Falk's winner under incomplete non-normal data.

Both estimators feed the **same** test battery: `fmg_tests()` auto-detects ML2S
and reads the two-stage spectrum, so each fit costs one battery call.

**Missingness (expensive axis):** complete, MCAR, MAR, each at rate
$\in \{.15, .30\}$, via the Savalei-Bentler (2005) generators in
`experiments/_support` (`sb2005_mcar`, `sb2005_mar`). MAR is driven by the
always-observed marker indicators (the MAR cause). Two regimes, made explicit:

- **MCAR** -- normal-theory FIML/ML2S is consistent for any distribution, so
  this is a **clean** test of the reference-law correction. The headline.
- **MAR + non-normality** -- the Gaussian likelihood imputes with linear
  conditional means, so the statistic is non-central even under a true model and
  **no** reference-law correction can fully restore level. We store the base
  statistic and the UGamma trace so the report can attribute MAR over-rejection
  to **estimator bias**, not the test (the diagnostic exp 23 introduced). MAR is
  the honest stress test, not the headline.

## Non-normality

Reuse the experiment-17 / FMG non-normal families rather than reinventing them:
`norm`, `vm1/2` (Vale-Maurelli), `ig1/2` (independent generator), `pl1/2`
(piecewise-linear, Foldnes & Gronneberg 2022), where `*1` is skew 2 / kurtosis 7
and `*2` is skew 3 / kurtosis 21. VM is a Gaussian copula, which is fine here
(continuous indicators); IG/PL bite harder than the copula when we want them to.

## The cost asymmetry, and the grid that exploits it

The key structural fact: **once the UGamma spectrum is formed, every p-value is
nearly free.** So the design splits into a free axis and an expensive factorial.

**Free axis (harvested per replication, NOT grid cells).** Every statistic is
computed from the one eigen-spectrum of each fit, for both `h1_information =`
`"saturated"` and `"structured"`:

```
GOF / per-rung:  naive, MLR (YB_mplus), YB_exact/SB, SS, SF,
                 EBA2/4/6, pEBA2/4/6, pall, pOLS, all
difference:      naive, SB, mean-var-adjusted, mixture, + pEBA-on-difference
```

"Lots of p-values" costs *columns*, not cells.

**Expensive factorial (each level needs a refit + a new spectrum):**

| axis | levels (starting point) | rationale |
|---|---|---|
| estimator | FIML, ML2S | Savalei-Falk; ML2S cheaper per ladder |
| missingness | complete, MCAR-.15/.30, MAR-.15/.30 | the new contribution |
| non-normality | norm, vm2, ig2, pl2 (+vm1 …) | reuse exp-17 families |
| total $N$ | 100, 200, 500, 1000 | push small; SB dies there |
| group ratio | 1:1, 1:3 | unevenness breaks SB-1 (B&S) |
| model size $p$ | 8, 16, 30 | $p=30$ is pEBA's home turf |
| invariance condition | null + {weak, strong, strict} $\times$ {small, med, large $\Delta\text{RMSEA}$} | Type-I (null) + power curves |

Full crossing $\approx 2 \times 5 \times 4 \times 4 \times 2 \times 3 \times 10
\approx 9{,}600$ cells. At the 5000-rep house standard (and **no Monte Carlo
error reported**, per house convention) compute is *not* the binding constraint
on a 100-CPU SLURM allocation; **presentability is**.

## Output schema (decide before the run)

One **long-format** row per `(cell, estimator, rung, outcome, method,
h1_information)`; per-replicate spectra to an `.rds` so any later p-value
transform is recomputable without refitting.

`results/fits.csv` columns:

```
# design keys
p, n_total, ratio, dist, mech, rate, realized_rate, invariance_cond,
delta_rmsea, n_noninvariant, estimator, rung, outcome (gof|nested),
# result
method, h1_information, p_value, base_stat, df, trace, scaling_factor, truth (h0|power)
```

Plus `spectra.rds` (per-rep eigenvalues; sufficient statistic),
`summary_rejection.csv` (rejection rate by design $\times$ method),
`power_adjusted.csv` (size-adjusted power vs paired null), `cells.csv`,
`metadata.csv`.

## Reporting plan (run big, report curated)

Run the superset once; present three stories, dump the rest to a supplement.

1. **Headline -- Type-I of the difference test.** pEBA vs $D_{SB10}$ vs MLR,
   sliced on non-normality $\times$ missingness, at $p=16$, MCAR (clean regime).
2. **Power curves.** size-adjusted power vs $\Delta\text{RMSEA}$, at the cells
   where Type-I is controlled.
3. **The flex.** $p=30$ + small $N$ + MAR: the corner where SB collapses and the
   spectrum methods earn their keep, with the trace-vs-mean(T) diagnostic
   attributing residual MAR over-rejection to estimator bias.

Everything else (full estimator $\times$ missingness crossing, all $j$-variants,
structured-vs-saturated H1) goes to the supplement. Do **not** cross the
effect-size gradient with every nuisance level: fix nuisances at a representative
level for the power sweeps, fully cross only for the Type-I headline.

## Open questions for Njål

The design decisions I could not settle from the methods side alone:

1. **Structural precedent.** Adopt the Brace & Savalei two-factor 4/8/15-indicator
   models, or anchor on a named real scale (more "applied" but less comparable)?
2. **Which rungs to feature.** Weak + strong (the applied core), or the full
   ladder including strict? Strict is where residual-variance non-invariance
   lives and SB overcorrects most -- a possible pEBA win, but rarely tested.
3. **Effect-size metric.** $\Delta\text{RMSEA}$ gradient vs number-of-noninvariant-items
   vs raw loading/intercept deltas, and how many levels.
4. **Partial invariance.** Include the 1-item-noninvariant condition? (I think yes.)
5. **Estimator framing.** FIML headline + ML2S secondary, or co-equal? (Savalei
   & Falk argues co-equal; co-equal is cheap given the shared battery.)
6. **Missingness.** Rates; add an MNAR robustness slice? MAR-cause design (marker
   indicators vs an external fully-observed covariate)?
7. **Non-normality set.** exp-17 families as above, or the exact FMG-2024 set?
8. **$N$ range.** Push to $N=100$ total (50/group)? That is where everything
   diverges and pEBA should shine, but convergence failures climb.
9. **High-dim cell.** $p=30$, or go to 40 to match FMG 2026 head-to-head?
10. **Reps.** 5000 (exp-23 house standard), or more for the Type-I headline?

## What the smoke established (and two real wiring items)

The scaffold runs end-to-end on the p=6 normal smoke model: both estimators
(FIML, ML2S), the full GOF battery (naive, MLR, SB, SS, SF, EBA2/4/6,
pEBA2/4/6, pall, pOLS, all) on each constrained model, and the Satorra-2000
difference test on the ladder steps, all written to the long-format schema. Two
findings fell out that turn into the paper's core wiring tasks:

1. **The mean-structure ("strong") difference test needs work.** The
   configural$\to$metric (loadings) and scalar$\to$strict (residuals) steps are
   clean parameter restrictions and the robust difference test runs. The
   **metric$\to$scalar step does not**: scalar ties intercepts *and* frees the
   group-2 latent mean, so it is not a strict parameter-subset of metric
   ($\text{npar}(\text{scalar}) = 37 > \text{npar}(\text{metric}) = 36$), and
   magmaan's exact Satorra-2000 route (`restriction_alpha_from_K`) requires
   $\text{npar}(H_0) < \text{npar}(H_1)$. lavaan sidesteps this via
   saturated-referenced chi-squares. **This is the same mean-structure issue
   reviewers of the pEBA-nested paper raise**: the scalar step is the one that
   matters for comparing means, and the robust nested test for it is a genuine
   methods/wiring task, not a smoke artifact. Decide with Njal: build a
   mean-structure-aware nested path, or reference both models to the saturated
   model the way lavaan does.

2. **pEBA-on-the-difference is not exposed yet -- but the wire is thin.**
   `nestedTest(method = "satorra.2000", A.method = "exact")` returns the
   difference-spectrum eigenvalues plus naive / SB / mean-var-adjusted / mixture
   p-values, but **not** the named pEBA-family difference p-values, which are the
   methodological crux of this paper. Concrete path (scoped 2026-06-16): every
   omnibus battery p-value is produced by one primitive,
   `infer_fmg_test(statistic, df, eigenvalues, method, param)` (see
   `.fmg_result_rows` in `r-package/R/fmg.R`). The same call applied to the
   *difference* statistic and `nestedTest`'s difference spectrum --
   `infer_fmg_test(T_diff, df_diff, nt$eigenvalues, method = "peba", param = 4)`
   -- gives pEBA-on-the-difference directly. Validate under normal complete data
   (difference eigenvalues ~1, so it should track the naive chi-square diff) then
   confirm calibration gains under non-normality. A short wire, but a methods
   decision (biased vs unbiased spectrum, param convention) for Njal to sign off.

## Results (Type-I run)

```{r results-guard, results="asis"}
if (!have_results) {
  cat("_No results yet._ Reproduce:\n\n```sh\n",
      "Rscript run_experiment.R --smoke                              # pipeline check\n",
      "Rscript run_experiment.R --conds null,weak,strict --reps 2500 # the Type-I run\n",
      "```\n", sep = "")
}
dist_order <- c("norm","vm1","vm2","ig1","ig2","pl1","pl2")
# wide dist x method table for one slice of the summary
slice_tab <- function(S, methods, oc, rng, est = "FIML", rat = "1:1",
                      mechs = c("complete","MCAR","MAR"), rate_hi = 0.30) {
  d <- S[S$outcome == oc & S$rung == rng & S$estimator == est & S$ratio == rat &
           S$truth == "h0" & S$method %in% methods &
           (S$mech == "complete" | S$rate == rate_hi) & S$mech %in% mechs, ]
  if (!nrow(d)) return(NULL)
  d$mech <- factor(d$mech, levels = c("complete","MCAR","MAR"))
  w <- reshape(d[, c("dist","mech","method","reject")],
               idvar = c("dist","mech"), timevar = "method", direction = "wide")
  names(w) <- sub("reject.", "", names(w))
  w <- w[order(w$mech, match(w$dist, dist_order)), c("dist","mech", methods)]
  w
}
```

```{r short-answer, results="asis"}
if (have_results) {
  cat(sprintf(
    paste0("**Short answer (Type-I, %s reps/cell, p=6 base model).** Three findings.\n\n",
      "1. **Nested difference test.** The configural-vs-metric Type-I reproduces ",
      "Brace-Savalei under non-normality and extends it to missing data: the naive ",
      "difference is destroyed (up to ~.92), Satorra-Bentler over-rejects for severe ",
      "skew/kurtosis, and **unequal groups (1:3) makes SB drift worse** (pl2 MAR.30: ",
      "SB ~.30 at 1:3 vs ~.19 at 1:1). Mean-variance-adjusted/mixture help but still ",
      "drift. The eigenvalue (pEBA) difference p-values are absent because ",
      "**pEBA-on-the-difference is not wired yet** -- the gap these numbers motivate.\n\n",
      "2. **Omnibus GOF.** The applied default **MLR fails hard under non-normality ",
      "(rejection .25-.65)**; the moment/eigenvalue methods recover. But at p=6 the ",
      "simple **mean-variance (SS) statistic already controls best (~.04-.16), and ",
      "pEBA4 does not beat it** -- expected, since pEBA's advantage is a ",
      "high-dimensional effect (FMG show it to p=40). This is the direct argument for ",
      "the p in {16,30} build-out, not evidence against pEBA.\n\n",
      "3. **Estimator.** Two-stage ML (ML2S) shows **better Type-I than FIML under ",
      "severe non-normality** (e.g. vm2 pEBA4 ~.16 vs ~.23), the Savalei-Falk ",
      "direction, though its naive statistic is wildly inflated (completes to N).\n\n"),
    reps))
}
```

```{r nested-table}
if (have_results) {
  nest <- slice_tab(S, c("naive","SB","adjusted","mixture"), "nested", "weak")
  if (!is.null(nest))
    knitr::kable(nest, digits = 3, row.names = FALSE,
                 col.names = c("Family","Missing","Naive","SB","Adj(mv)","Mixture"),
                 caption = "Nested weak-step (configural vs metric) Type-I, FIML, equal groups. complete / MCAR.30 / MAR.30. Nominal .05.")
}
```

```{r gof-table}
if (have_results) {
  gof <- slice_tab(S, c("naive","MLR","SB","SS","pEBA4","pall"), "gof", "metric")
  if (!is.null(gof))
    knitr::kable(gof, digits = 3, row.names = FALSE,
                 col.names = c("Family","Missing","Naive","MLR","SB","SS","pEBA4","pall"),
                 caption = "Omnibus GOF (metric model) Type-I, FIML, equal groups. The eigenvalue methods (SS/pEBA4/pall) vs the MLR/SB defaults under non-normality + missingness. Nominal .05.")
}
```

```{r caveats, results="asis"}
if (have_results) {
  cat(paste0(
    "_Caveats._ p=6 one-factor base model (B&S multi-factor p in {8,16,30} is the ",
    "build-out). FIML shown; ML2S and the unequal-groups (1:3) slice are in ",
    "`results/summary_rejection.csv` (1:3 is where SB drifts hardest, per B&S). ",
    "The weak (loadings) and strict (residuals) steps carry the difference test; ",
    "the strong (scalar) step is omitted here because its robust nested test is ",
    "not yet wired (mean-structure non-nesting). Power conditions were run at a ",
    "raw effect size, not yet calibrated to Delta-RMSEA, so treat power as ",
    "exploratory.\n"))
}
```

## References

- Brace, J. C., & Savalei, V. (2017). Type I error rates and power of several
  versions of scaled chi-square difference tests in investigations of
  measurement invariance. *Psychological Methods, 22*(3), 467-485.
- Foldnes, N., & Gronneberg, S. (2018). Approximating test statistics using
  eigenvalue block averaging. *Structural Equation Modeling, 25*(1), 101-114.
- Foldnes, N., Moss, J., & Gronneberg, S. (2026). Penalized eigenvalue block
  averaging: Extension to nested model comparison and Monte Carlo evaluations.
  *Behavior Research Methods.* (replicated in experiment 18)
- Savalei, V., & Falk, C. F. (2014). Robust two-stage approach outperforms
  robust FIML with incomplete non-normal data. *Structural Equation Modeling,
  21*(2), 280-302.
- Savalei, V., & Rosseel, Y. (2022). Computational options for standard errors
  and test statistics with incomplete normal and non-normal data in SEM.
  *Structural Equation Modeling, 29*(1), 163-181.
- Yuan, K.-H., & Bentler, P. M. (2000). Three likelihood-based methods for mean
  and covariance structure analysis with non-normal missing data. *Sociological
  Methodology, 30*, 165-200.
- Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
  analysis of moment structures. In *Innovations in Multivariate Statistical
  Analysis* (pp. 233-247).
- Satorra, A., & Bentler, P. M. (2001). A scaled difference chi-square test
  statistic for moment structure analysis. *Psychometrika, 66*(4), 507-514.
- Satorra, A., & Bentler, P. M. (2010). Ensuring positiveness of the scaled
  difference chi-square test statistic. *Psychometrika, 75*(2), 243-248.
- Asparouhov, T., & Muthen, B. (2013). Computing the strictly positive
  Satorra-Bentler chi-square test in Mplus. *Mplus Web Notes 12.*
- Meredith, W. (1993). Measurement invariance, factor analysis and factorial
  invariance. *Psychometrika, 58*(4), 525-543.
- Vandenberg, R. J., & Lance, C. E. (2000). A review and synthesis of the
  measurement invariance literature. *Organizational Research Methods, 3*(1),
  4-70.
- Millsap, R. E. (2011). *Statistical Approaches to Measurement Invariance.*
  Routledge.
- Putnick, D. L., & Bornstein, M. H. (2016). Measurement invariance conventions
  and reporting: The state of the art and future directions. *Developmental
  Review, 41*, 71-90.
- Foldnes, N., & Gronneberg, S. (2022). Non-normal data simulation using
  piecewise linear transforms. *Structural Equation Modeling, 29*(5), 783-790.
```
