---
title: "Ordinal Polychoric Omega: Coverage and Stress"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Question

Does the robust delta interval for fit-free polychoric omega cover the
appropriate target under Gaussian ordinal responses, nonnormality and local
dependence, especially with small samples or extreme thresholds?

## Short answer

```{r}
read_lane <- function(lane, filename) {
  path <- file.path("results", lane, filename)
  if (!file.exists(path)) stop("Missing ", path, "; run --lane ", lane)
  read.csv(path)
}
coverage <- read_lane("coverage", "simulation_summary.csv")
coverage_meta <- read_lane("coverage", "metadata.csv")
stress <- read_lane("stress", "simulation_summary.csv")
stress_meta <- read_lane("stress", "metadata.csv")
targets <- read_lane("stress", "pseudo_targets.csv")
```

```{r, results='asis'}
cat(sprintf(
  "The Gaussian arm has %d replications per cell: conditional 95%% coverage ranges from %.3f to %.3f, with %d construction failures. The stress arm has %d replications per cell: coverage of the polychoric pseudo-target ranges from %.3f to %.3f, with %d construction failures.\n\n",
  coverage_meta$reps, min(coverage$coverage_95), max(coverage$coverage_95),
  sum(coverage$failures), stress_meta$reps,
  min(stress$coverage_pseudo_95), max(stress$coverage_pseudo_95),
  sum(stress$failures)))
if (coverage_meta$smoke || stress_meta$smoke) {
  cat("At least one displayed arm is a smoke run. Its rates check the pipeline;
they are too noisy to rank interval methods or choose an inference policy.\n")
}
```

The two arms were formerly research/22 and research/23. They now share one
question and runner, while their result files and historical metadata remain
separate. Coverage is conditional on successful construction; failures are
shown explicitly and do not count as covered intervals.

## Gaussian ordinal responses

One-factor probit data with three-category indicators compare balanced and
extreme thresholds. Here the polychoric correlation-matrix coefficient targets
the generating latent-response coefficient.

```{r}
shown <- coverage[, c("regime", "n", "reps", "failures", "target",
  "mean_value", "bias", "sd_value", "mean_se", "coverage_95",
  "mean_width_95", "min_variable_category_count_p05")]
knitr::kable(shown, digits = 4, row.names = FALSE)
```

## Nonnormality and local dependence

Skewed factor scores, heavy-tailed factor/residual draws and residual local
dependence test the same fit-free coefficient. The inferential target is the
large-draw polychoric omega from the same ordinal-statistics pipeline.
Continuous latent-response Pearson omega is a separate bias diagnostic.

```{r}
knitr::kable(targets, digits = 4, row.names = FALSE)
shown <- stress[, c("dgp", "cuts", "n", "reps", "failures", "target_pseudo",
  "pseudo_minus_continuous", "mean_value", "bias_pseudo", "sd_value",
  "mean_se", "median_se", "coverage_pseudo_95", "coverage_continuous_95",
  "mean_width_95", "median_width_95", "min_variable_category_count_p05")]
knitr::kable(shown, digits = 4, row.names = FALSE)
```

## Caveats and next check

The pseudo-targets are Monte Carlo approximations, not analytic population
roots. A stress interval's coverage against the continuous target does not
measure calibration for its polychoric target. The heavy-tailed arm uses
Student t with three degrees of freedom, so fourth-moment assumptions require
particular care. The grids concern the ordinary Wald delta interval, not
fitted DWLS omega, observed-score reliability, profile-LR inversion or Bartlett
calibration.

Before a statistical recommendation, run both arms beyond smoke size and
report Monte Carlo uncertainty, sparse-category failures and pseudo-target
precision separately. Keep the generating and pseudo-true targets distinct.

## Reproduce

From this directory, or pass the runner's full path from the repository root:

```sh
Rscript run_experiment.R --lane coverage --smoke
Rscript run_experiment.R --lane stress --smoke
Rscript run_experiment.R --lane coverage --reps 500
Rscript run_experiment.R --lane stress --reps 500 --pseudo-n 100000
quarto render report.qmd
```

Use `--results-dir /absolute/path` to save a separate run. Rendering reads only
the retained `results/coverage/` and `results/stress/` tables.
