---
title: "DWLS spectral reference laws across realistic models"
format:
  html:
    toc: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Default register

| Choice | Current value | Status | Evidence |
|---|---|---|---|
| Global DWLS policy reference | All | Broader validation open | Earlier calibration; this study has development draws only |
| Nested DWLS policy reference | SB and PEBA4 | Provisional, TASK-81 | Textbook-model confirmation has not run |
| First-stage influence | Exact | Fixed baseline | Every comparator uses the same retained policy spectrum |

## Question

Which spectral reference law calibrates global and nested DWLS policy tests
across realistic textbook models?

## Short answer

The study is ready for production planning, but does not yet answer the
reference-law question. Eight frozen positive-definite populations cover CFA,
MTMM, higher-order, bifactor, longitudinal, structural and grouped invariance
models. Development draws check execution, availability and cost; two draws
per cell cannot establish calibration or select a default. The original grid projects 72.6 CPU-hours; the registered cost reduction
omits seven-category cells for the 30-item model, leaving 184 production cells
and a 44.5 CPU-hour planning estimate. Production has not been launched. Simbox remains unavailable.

## Development evidence

```{r}
root <- file.path("results", "pilot-full")
if (!file.exists(file.path(root, "summary.csv")))
  stop("Pilot summaries absent: run run_experiment.R --pilot --cell <pricing IDs> into fresh directories, then --combine <directories> --out-dir results/pilot-full")
s <- read.csv(file.path(root, "summary.csv"))
t <- read.csv(file.path(root, "timing.csv"))
x <- s[s$reference == "All", ]
x$step <- paste(x$model, x$test)
rows <- lapply(split(x, x$model), function(a) data.frame(
  Model = a$model[1], `Draws with tests available` = sum(a$available),
  `Attempted tests` = sum(a$attempted), `Unavailable tests` = sum(a$failed),
  check.names = FALSE))
knitr::kable(do.call(rbind, rows), caption = "Development availability across all cells. Global and nested tests are counted separately; typed binary threshold equivalence remains unavailable.")
```

Per-cell rejection rates, successful and attempted denominators, Wilson intervals,
joint reference availability, failures and ranking diagnostics are retained in
CSV. Those rankings are development diagnostics. A decision requires the
registered 1,000 fresh production draws per retained cell, with each model and
nested step reported separately. Per-model losses compare outside-band counts
and RMSE against All globally and against SB/PEBA4 for nested tests.

The per-draw raw artifact saves the statistic, df, full eigenvalue vector,
seed, convergence verdicts and failure reason for each test. All fourteen
reference labels are evaluated from that artifact. PEBA4 and pEBA4 both call
`peba` with four blocks and are retained as identical labels to match the
requested candidate list. The All tail truncates negative eigenvalues as in
the current policy: $T \sim \sum_j \max(\lambda_j,0) Z_j^2$, with independent standard-normal $Z_j$.
The same library truncation convention is used for the
comparator calls.

## Cost and scope

The permitted cost amendment removes eight seven-category long30 cells, while
preserving all other stable cell IDs and seeds. The reduced plan projects
44.54 CPU-hours: ideal 3.71 hours on 12 cores or 2.78 on 16. This estimate has
little margin against 45 CPU-hours, and actual elapsed time can exceed it.
Seven-category long30 evidence remains development-only and cannot support
adoption. The table below retains the original full-grid pricing.

```{r}
cost <- read.csv(file.path(root, "cost.csv"))
knitr::kable(data.frame(
  `Priced cells` = cost$observed_cells,
  `Unpriced cells` = cost$unpriced_cells,
  `Projected CPU hours` = round(cost$full_grid_cpu_hours, 1),
  `Ideal hours on 12 cores` = round(cost$full_grid_cpu_hours / 12, 1),
  `Ideal hours on 16 cores` = round(cost$full_grid_cpu_hours / 16, 1),
  check.names = FALSE), caption = "Matched-cell extrapolation to 1,000 draws per cell; ideal wall time excludes startup, serialization, reference evaluation and scheduling overhead.")
```

Timing includes generation, H1/H0 fitting and policy inference, with one math
thread per worker. Reference evaluation and summary writing happen after the
fit loop and are excluded from the recorded per-draw CPU time, so the estimate
is a lower planning estimate. Only two draws price each cell; difficult search
paths can vary substantially. The 30-item longitudinal model dominates cost.
Production requires a separate compute authorization and a working machine;
this task neither accesses simbox nor launches Modal. A Modal job can invoke
the same single-cell command and persist its attempt directory before marking
completion; the runner has no corpus or sibling-study dependency.

Small-N skewed cells can have empty categories or optimizer nonconvergence.
They retain failures rather than deleting categories, loosening tolerances or
changing the registered ordinary fitting defaults. Missing or numerically
unresolved cells leave the affected reference choice open. Binary MDD
threshold steps can be equivalent in moment space and retain that typed reason.

## Population construction

Only derived correlations, thresholds, syntax, restrictions and source hashes
are frozen. No corpus observations, paper code or source mirror is shipped.
Responses are standardized and null restrictions imposed as registered.
The MDD population replicates standardized group 1 in both groups; it does not
replicate the book's group differences.

Before simulation, the original Worland equal-Risk-path construction failed
the validity gate (standardized paths −1.59/−2.31; response-correlation minimum
eigenvalue −0.217). The approved, committed amendment instead equates the
second and third Cognitive loadings while preserving structural paths. Its
response-correlation minimum eigenvalue is 0.293. No PSD repair is used.
All eight residual covariance and response-correlation matrices pass the
positive-definiteness gate; the manifest records their eigenvalue minima.

## Reproduce

Use an installed current `magmaanlab` in a lane R library. From the repository root:

```sh
Rscript experiments/decisions/07-dwls-reference-law/run_experiment.R --help
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  R_LIBS="$HOME/.cache/magmaan-rlib/lane-d" nice -n 10 Rscript \
  experiments/decisions/07-dwls-reference-law/run_experiment.R \
  --smoke --cell 1,8 --workers 1 --out-dir /tmp/reference-smoke-new
```

Rebuilding populations is optional and local: use `scripts/build_populations.R
--corpus ROOT`. Production needs only the committed frozen inputs.

After compute authorization, the per-cell contract is
`run_experiment.R --production --cell ID --workers 1 --out-dir ATTEMPT`.
`scripts/run_cells.sh STUDY RLIB RUN_DIR WORKERS MODE IDS...` implements
`cells/cell_XXX`, fresh attempts and completion-last promotion. It skips finished
cells and refuses to overwrite an incomplete cell target. It can be used with
`smoke` to validate the driver. Keep the workstation's aggregate core budget.

The full-grid pilot source is commit `021d31bc`; use that version to
recombine the original attempts. Current sources register the reduced production
grid and require fresh attempts, as intended by the strict fingerprint check.

Combine with `run_experiment.R --combine DIR1,DIR2 --out-dir NEW_DIRECTORY`.
The combiner refuses incomplete attempts, differing source/binary/population/
design fingerprints, inconsistent modes/seeds/counts and duplicate cells.
Local two-cell fan-out reproduces serial summaries exactly; a deliberately
wrong fingerprint is rejected. Interrupted attempts remain diagnostic artifacts
and are rerun into fresh output directories.

The committed pilot summaries are small frozen development evidence. Raw RDS
spectra, p-values, generated reports and incomplete attempts remain ignored.
