---
title: "Do constrained non-iterative CFA fits depend on the marker chart?"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
raw_path <- file.path(results_dir, "chart_comparison_raw.csv")
sum_path <- file.path(results_dir, "chart_comparison_summary.csv")
lin_path <- file.path(results_dir, "linear_projection_summary.csv")
if (!file.exists(raw_path) || !file.exists(sum_path) || !file.exists(lin_path)) {
  stop("No results. Run:\n  Rscript run_experiment.R --smoke")
}
raw <- read.csv(raw_path, stringsAsFactors = FALSE, check.names = FALSE)
sumry <- read.csv(sum_path, stringsAsFactors = FALSE, check.names = FALSE)
lin <- read.csv(lin_path, stringsAsFactors = FALSE, check.names = FALSE)
meta <- tryCatch(read.csv(file.path(results_dir, "metadata.csv"),
                          stringsAsFactors = FALSE), error = function(e) NULL)
fmt <- function(x) formatC(x, format = "e", digits = 2)
fmt_fixed <- function(x) formatC(x, format = "f", digits = 3)
max_col <- function(col) max(raw[[col]], na.rm = TRUE)
```

## Question

The constrained estimator is an $\Omega$-metric projection of a configural
closed-form estimate onto the linear equalities carried by the lavaanified
partable. If those equalities are just a chart expression of the same
measurement-invariance surface, an arbitrary marker choice should change the
numeric coordinates but not the fitted covariance matrix or the Wald statistic.

This experiment compares two equivalent marker charts:

- `marker_x1x4`: default markers `x1` and `x4`;
- `marker_x2x5`: explicit markers `x2` and `x5`.

Both charts use the same observed covariance matrices, the same declared
indicator-factor blocks, the same non-iterative estimator, and the same
normal-theory delta covariance construction. We compare the chart-invariant
objects: the configural implied covariance, the constrained implied covariance,
and the metric-invariance Wald statistic. The raw parameter vectors are also
reported as a control; those should differ because they are chart coordinates.

## Short Answer

The pure projection algebra is fine, but the current CFA constraint target is
not fully marker-chart invariant away from the constraint surface.

The separate linear-algebra control passes: projecting in one coordinate system
and transforming the result agrees with projecting after a nonsingular linear
reparameterization, with maximum transformed-coordinate difference
`r fmt(lin$projected_theta_max_abs_diff_max)` and maximum Wald difference
`r fmt(lin$W_abs_diff_max)`.

For CFA, the configural map is chart invariant at the level of implied
covariances: the largest configural implied-covariance difference is
`r fmt(max_col("config_sigma_max_abs_diff"))`, while the raw coordinate vectors
differ by as much as `r fmt_fixed(max_col("raw_theta_max_abs_diff"))`. However,
the constrained projection depends on the marker chart when the unconstrained
estimate is off the metric-invariance surface: the largest constrained
implied-covariance difference is
`r fmt(max_col("constrained_sigma_max_abs_diff"))`, and the largest
Wald-statistic difference is `r fmt(max_col("W_abs_diff"))`. On the exact
invariant population these quantities collapse to roundoff, because the
projection is a no-op.

## Chart Comparison

The population rows use exact model covariances with very large `nobs`, so they
mostly expose algebraic or numerical defects. The finite-sample rows use the
same random sample covariance in both marker charts. `metric_viol` is
deliberately off the metric-invariance surface; this is the more important
case, because projection targets are where chart dependence would show up.

```{r population-table}
pop <- raw[raw$source == "population", ]
pop_out <- pop[c("scenario", "estimator", "k_x1x4", "k_x2x5",
                 "W_x1x4", "W_x2x5", "W_abs_diff",
                 "config_sigma_max_abs_diff",
                 "constrained_sigma_max_abs_diff",
                 "raw_theta_max_abs_diff")]
pop_out$W_x1x4 <- fmt_fixed(pop_out$W_x1x4)
pop_out$W_x2x5 <- fmt_fixed(pop_out$W_x2x5)
for (nm in c("W_abs_diff", "config_sigma_max_abs_diff",
             "constrained_sigma_max_abs_diff")) {
  pop_out[[nm]] <- fmt(pop_out[[nm]])
}
pop_out$raw_theta_max_abs_diff <- fmt_fixed(pop_out$raw_theta_max_abs_diff)
knitr::kable(pop_out, row.names = FALSE,
             caption = "Exact-population chart comparison.")
```

```{r finite-summary}
fin <- sumry[sumry$source == "finite_sample", ]
cols <- c("scenario", "estimator",
          "W_abs_diff_max", "W_abs_diff_median",
          "constrained_sigma_max_abs_diff_max",
          "constrained_sigma_max_abs_diff_median",
          "raw_theta_max_abs_diff_median")
fin_out <- fin[cols]
for (nm in setdiff(names(fin_out), c("scenario", "estimator"))) {
  fin_out[[nm]] <- if (grepl("raw_theta", nm)) fmt_fixed(fin_out[[nm]]) else fmt(fin_out[[nm]])
}
knitr::kable(fin_out, row.names = FALSE,
             caption = "Finite-sample chart comparison. Max/median are over replications.")
```

## Linear Projection Control

This control strips away CFA. For random $(\hat\theta,\Omega,R,c)$, it projects
onto $R\theta=c$. Then it applies a random nonsingular reparameterization
$\theta=B\eta$, transforms $\Omega$ and $R$, projects in $\eta$, and maps the
answer back. The two projections and Wald statistics should match.

```{r linear-table}
lin_out <- lin
for (nm in names(lin_out)) lin_out[[nm]] <- fmt(lin_out[[nm]])
knitr::kable(lin_out, row.names = FALSE,
             caption = "Pure projection equivariance control.")
```

## Interpretation

This is the important distinction:

- If the transformation is genuinely linear in the estimated coordinates, the
  minimum-distance projection is equivariant.
- The configural non-iterative CFA map itself is fine under a marker change:
  the implied covariance is unchanged up to roundoff.
- The marker-to-marker chart change is nonlinear in the loading coordinates.
  Therefore an $\Omega$-metric projection onto linear marker-chart equalities is
  only guaranteed to agree on the constraint surface. Under sampling noise or
  fixed misspecification, different marker charts induce different local
  distances to the same curved covariance-level surface.

So this experiment does not validate marker-invariant constrained targets for
the current implementation. It gives a concrete diagnostic: if we want an
off-model constrained Guttman estimator that is genuinely independent of marker
choice, the projection should be formulated in a chart-free object (for example
common fitted covariance / loading orbit) or with an explicit moving-chart
construction, not as a plain coordinate projection in whichever marker chart
the partable happens to use.

## Reproduce

```sh
Rscript run_experiment.R --smoke
quarto render report.qmd
```

For a less noisy finite-sample summary:

```sh
Rscript run_experiment.R --full --reps 300 --n 200,400,1000
quarto render report.qmd
```
