---
title: "Do constrained non-iterative CFA fits depend on the marker chart?"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
raw_path <- file.path(results_dir, "chart_comparison_raw.csv")
sum_path <- file.path(results_dir, "chart_comparison_summary.csv")
lin_path <- file.path(results_dir, "linear_projection_summary.csv")
if (!file.exists(raw_path) || !file.exists(sum_path) || !file.exists(lin_path)) {
  stop("No results. Run:\n  Rscript run_experiment.R --smoke")
}
raw <- read.csv(raw_path, stringsAsFactors = FALSE, check.names = FALSE)
sumry <- read.csv(sum_path, stringsAsFactors = FALSE, check.names = FALSE)
lin <- read.csv(lin_path, stringsAsFactors = FALSE, check.names = FALSE)
meta <- tryCatch(read.csv(file.path(results_dir, "metadata.csv"),
                          stringsAsFactors = FALSE), error = function(e) NULL)
fmt <- function(x) formatC(x, format = "e", digits = 2)
fmt_fixed <- function(x) formatC(x, format = "f", digits = 3)
max_col <- function(col) max(raw[[col]], na.rm = TRUE)
```

## Question

The estimator-side metric Guttman map estimates a common standardized loading
shape across groups and only then converts the result into the marker chart
carried by the lavaanified partable. If the construction is chart-correct, an
arbitrary marker choice should change the numeric coordinates but not the
Sigma/H-level constrained fitted covariance matrix.

This experiment compares two equivalent marker charts:

- `marker_x1x4`: default markers `x1` and `x4`;
- `marker_x2x5`: explicit markers `x2` and `x5`.

Both charts use the same observed covariance matrices, the same declared
indicator-factor blocks, and the same non-iterative estimator. We compare the
chart-invariant objects: the configural implied covariance and the estimator-
side metric-constrained implied covariance. The raw parameter vectors are also
reported as a control; those should differ because they are chart coordinates.

## Short Answer

The estimator-side metric map passes this marker-chart sanity check. The largest
chart difference in the configural implied covariance is
`r fmt(max_col("config_sigma_max_abs_diff"))`; the largest chart difference in
the metric-constrained implied covariance is
`r fmt(max_col("metric_sigma_max_abs_diff"))`. Both are roundoff. The coordinate
vectors still differ, with maximum configural coordinate difference
`r fmt_fixed(max_col("config_theta_max_abs_diff"))` and maximum metric-map
coordinate difference `r fmt_fixed(max_col("metric_theta_max_abs_diff"))`, which
is expected because the two partables use different marker charts.

The separate linear-algebra control is retained as a guardrail for the old
inference projection algebra: projecting in one coordinate system and
transforming the result agrees with projecting after a nonsingular linear
reparameterization, with maximum transformed-coordinate difference
`r fmt(lin$projected_theta_max_abs_diff_max)` and maximum Wald difference
`r fmt(lin$W_abs_diff_max)`.

## Chart Comparison

The population rows use exact model covariances with very large `nobs`, so they
mostly expose algebraic or numerical defects. The finite-sample rows use the
same random sample covariance in both marker charts. `metric_viol` is
deliberately off the metric-invariance surface; this is the more important
case, because off-surface constrained targets are where chart dependence would
show up.

```{r population-table}
pop <- raw[raw$source == "population", ]
pop_out <- pop[c("scenario", "estimator",
                 "config_sigma_max_abs_diff",
                 "metric_sigma_max_abs_diff",
                 "config_theta_max_abs_diff",
                 "metric_theta_max_abs_diff")]
for (nm in c("config_sigma_max_abs_diff", "metric_sigma_max_abs_diff")) {
  pop_out[[nm]] <- fmt(pop_out[[nm]])
}
pop_out$config_theta_max_abs_diff <- fmt_fixed(pop_out$config_theta_max_abs_diff)
pop_out$metric_theta_max_abs_diff <- fmt_fixed(pop_out$metric_theta_max_abs_diff)
knitr::kable(pop_out, row.names = FALSE,
             caption = "Exact-population chart comparison.")
```

```{r finite-summary}
fin <- sumry[sumry$source == "finite_sample", ]
cols <- c("scenario", "estimator",
          "config_sigma_max_abs_diff_max",
          "metric_sigma_max_abs_diff_max",
          "metric_sigma_max_abs_diff_median",
          "config_theta_max_abs_diff_median",
          "metric_theta_max_abs_diff_median")
fin_out <- fin[cols]
for (nm in setdiff(names(fin_out), c("scenario", "estimator"))) {
  fin_out[[nm]] <- if (grepl("theta", nm)) fmt_fixed(fin_out[[nm]]) else fmt(fin_out[[nm]])
}
knitr::kable(fin_out, row.names = FALSE,
             caption = "Finite-sample chart comparison. Max/median are over replications.")
```

## Linear Projection Control

This control strips away CFA. For random $(\hat\theta,\Omega,R,c)$, it projects
onto $R\theta=c$. Then it applies a random nonsingular reparameterization
$\theta=B\eta$, transforms $\Omega$ and $R$, projects in $\eta$, and maps the
answer back. The two projections and Wald statistics should match.

```{r linear-table}
lin_out <- lin
for (nm in names(lin_out)) lin_out[[nm]] <- fmt(lin_out[[nm]])
knitr::kable(lin_out, row.names = FALSE,
             caption = "Pure projection equivariance control.")
```

## Interpretation

This is the distinction the earlier Omega-projection version of this experiment
missed. The inference-side `noniterative_cfa_constrained()` function is an
Omega-metric projection of a configural estimator and is a valid Wald/minimum-
distance algebra object, but it is not the Sigma/H-level constrained estimator.
The estimator-side map tested here imposes the metric loading-shape constraint
inside the reconstruction, before marker output coordinates are chosen. That is
why the coordinate vectors differ while the implied metric-constrained
covariances agree at roundoff, even under the deliberately misspecified
`metric_viol` population.

## Reproduce

```sh
Rscript run_experiment.R --smoke
quarto render report.qmd
```

For a less noisy finite-sample summary:

```sh
Rscript run_experiment.R --full --reps 300 --n 200,400,1000
quarto render report.qmd
```
