---
title: "Latent Metric Parameterization"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
read_result <- function(name) {
  path <- file.path("results", name)
  if (!file.exists(path)) stop("Missing retained output: ",path,"; see Reproduce.")
  read.csv(path,stringsAsFactors=FALSE)
}
g <- read_result("geometry.csv")
i <- read_result("inference.csv")
c <- read_result("convergence.csv")
s <- read_result("structural.csv")
m <- read_result("metadata.csv")
```

## Question

How do marker, unit latent-variance and effect-coded charts change conditioning
and numerical behavior while describing the same identified covariance model?

## Short answer

The retained **six-replication smoke run** contains useful population
conditioning and chart-invariance diagnostics. It cannot establish that one
chart wins generally, quote earlier 2,000-draw failure rates, or prove that no
better chart exists. In particular, fixing an endogenous latent's residual
variance is not the same as standardizing its total variance.

This study is **banked**. Reopen for matched-start comparisons in a named
near-Heywood structural regime; existing marker/std.lv/effect conventions
remain available without a new default decision.

## Evidence

The current population CFA slice has six indicators and two marker strengths.
These are deterministic diagnostics for these populations, not simulation
estimates of optimality:

```{r geometry}
knitr::kable(data.frame(Indicators=g$p, `Marker loading`=g$lambda1, Chart=g$chart,
  `Information condition`=g$pcond_info, `Removable / intrinsic curvature`=g$pe_over_in,
  check.names=FALSE), digits=3,row.names=FALSE)
```

The inference lane checks invariant quantities on the same dataset across charts:

```{r invariance}
x <- do.call(rbind,lapply(split(i,paste(i$p,i$lambda1,i$replicate)),function(z)
  data.frame(chisq=diff(range(z$chisq_h1)),lrt=diff(range(z$lrt_p)),df=diff(range(z$df_h1)))))
knitr::kable(data.frame(Quantity=c("Chi-square","LRT p-value","Degrees of freedom"),
  `Maximum chart spread`=format(c(max(x$chisq),max(x$lrt),max(x$df)),scientific=TRUE),
  check.names=FALSE),row.names=FALSE)
```

For the retained structural populations, the chart's conditioning changes with
latent explained variance. A small selected population table is enough to show
why a CFA result cannot settle structural scaling:

```{r structural}
x <- unique(s[,c("r2","resid_var","chart","pcond_population")])
knitr::kable(data.frame(`Latent R-squared`=x$r2, `Residual variance`=x$resid_var,
  Chart=x$chart, `Population condition`=x$pcond_population, check.names=FALSE),
  digits=3,row.names=FALSE)
```

Raw matched-fit outcomes, timing and roundtrip diagnostics remain in
`results/convergence.csv`, `structural.csv`, `cost.csv` and `backconvert.csv`.
Their six-draw sampling frequencies are smoke diagnostics. Neither timing
ratios nor a zero-failure cell establishes a universal preference.

## Caveats

Conditioning depends on population, coordinate convention and numerical scale;
a lower condition number alone does not prove better fitting risk. Structural
runs use one latent regression and do not establish a universal crossover in
$R^2$. Growth models with user-fixed loadings are excluded: fixing an additional
latent variance there can add a restriction instead of merely changing charts.
The cheap available approach is to retain existing conventions and inspect the
requested chart's diagnostics. A new internal-chart policy needs independent
matched-start evidence and accurate back-transformation of estimates and SEs.

## Reproduce

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --results-dir /tmp/latent_chart_check
quarto render report.qmd
```

The larger `--full` design is retained but no run is queued. Use a fresh output
path so a smoke check cannot overwrite existing results.
