---
title: "Association-ML policy calibration"
format:
  html:
    toc: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r}
paths <- file.path('results','pilot',c('summary.csv','failures.csv','timing.csv','cost.csv','population_timing.csv','population_uncertainty.csv'))
if(!all(file.exists(paths))) stop('Pilot absent: Rscript run_experiment.R --pilot --workers 2')
s <- read.csv(paths[1]); failures <- read.csv(paths[2]); timing <- read.csv(paths[3]); cost <- read.csv(paths[4])
pt <- read.csv(paths[5]); uncertainty <- read.csv(paths[6])
```

## Default register

| Choice | Current value | Status | Evidence |
|---|---|---|---|
| Ordinary association-ML inference | Typed unavailable | Open | Production complete; user decision pending |
| Lab covariance | Exact Stage-1 IJ, observed sensitivity | Provisional for exposure | Component gates; frozen production; user decision pending |
| Lab global / nested references | All, SB, PEBA4 | Open | Production complete; user reference decision pending |
| Lab MI | Robust projected meat, observed Schur metric | Provisional for exposure | Component gates; frozen production; user decision pending |

## Question and short answer

Does the lab association-ML recipe calibrate well enough for ordinary exposure?
Production is complete: 192 cells with 2,000 draws each, frozen at `b37ebefb`.
Association ML has no coverage flags, seven global size flags and one nested
size flag; DWLS has one correlation-coverage flag. No MI size cell is flagged,
and the population approximation check passes the post-hoc 5.5% criterion below. Sparse ordinal samples
cause substantial unavailability. The production tables below
retain cell denominators and uncertainty. Ordinary exposure and the global and
nested reference choices remain user decisions.

## Pilot mechanics and development evidence

The pilot covers a difficult non-Gaussian misspecified population, single-group
coverage/global/MI and two-group metric nesting. It prices only
`r sum(cost$priced)` of `r nrow(cost)` registered cells. Gaussian global size,
other sample sizes and category imbalance panels remain unpriced. Per-cell
rates, attempted and successful denominators and Wilson intervals are in
`results/pilot/summary.csv`; the tiny rates are mechanics, not calibration.

```{r}
shown <- s[startsWith(s$arm,'association') & s$target %in% c('loading','threshold','correlation','mi','nested'),]
shown <- shown[!grepl('_nested_(SB|PEBA4)$',shown$arm),]
knitr::kable(data.frame(Component=shown$arm,Target=shown$target,
  Attempts=shown$attempted,Available=shown$successful,
  Rate=round(shown$rate,2),`Wilson lower`=round(shown$lower,2),`Wilson upper`=round(shown$upper,2)),check.names=FALSE)
```

The covariance recipe is $\mathrm{Cov}(\hat\alpha)=H^{-1}BH^{-\mathsf T}/N$,
where $H$ is observed sensitivity and $B$ transports exact Stage-1 influence.
The global All reference uses the residual correlation-ML metric and exact
Stage-1 sampling Gamma. The nested reference uses the larger model's observed
sensitivity and parameter-space IJ spectrum. Robust MI fixes a loading at its
unrestricted pseudo-true value for size, then 0.15 away for reported power.
DWLS uses its own weighted-discrepancy pseudo-true targets; its coverage is
reported separately and does not estimate a paired method effect.

Population setup uses two independent million-row draws, streamed into pair
counts and passed through library Stage 1. Association-ML and DWLS population
fits share the same first-stage summaries. Both sets of targets and their
differences are frozen in the population CSVs. The largest absolute target
difference is `r signif(max(abs(uncertainty$target_difference)),3)`; production
interpretation must assess these errors relative to interval widths and MI
null sensitivity.

## Pilot cost and caveats

```{r}
priced <- cost[cost$priced,]
knitr::kable(data.frame(`Total N`=priced$n,Groups=priced$groups,
  `Mean seconds per draw`=round(priced$mean_draw_seconds,3),
  `CPU hours for 2,000 draws`=round(priced$production_cpu_hours,3)),check.names=FALSE)
```

Population construction took `r round(sum(pt$elapsed_seconds),1)` seconds on
this workstation. The cost CSV lists every unpriced cell explicitly; the
per-cell estimates above exclude population setup and driver/package startup.
They do not price the full study. Worker parallelism is bounded at two and
math libraries use one thread. This pilot preceded the authorized simbox production.

Coverage targets are finite population-draw approximations. Gaussian correct
models supply global size gates; global rejection under non-normality or
misspecification is descriptive because the association model need not hold.
Identical groups preserve the pseudo-true metric null. Flags below 93% coverage
or outside 3–7% size at the registered sample sizes request finite-sample work,
never an inconsistent recipe. These pilot flags cannot settle defaults.

Initial development smoke/pilot attempts exposed an extraction error: DWLS
nested `p_all` is intentionally unavailable in that returned bundle. The final
runner computes All from its spectrum using the lab reference primitive.
The corrected smoke and pilot were rerun before freezing; no library change,
criterion change or inference-policy choice followed those attempts.


## Production confirmation

```{r}
production_dir <- 'results/production-2026-10-05'
read_production <- function(name) {
  path <- file.path(production_dir, paste0(name, '.csv'))
  if (!file.exists(path)) stop('Frozen production evidence absent: ', path)
  read.csv(path, stringsAsFactors=FALSE)
}
ps <- read_production('summary'); cells <- read_production('cells')
pf <- read_production('failure_counts'); runtime <- read_production('timing_summary')
pc <- read_production('cost'); pu <- read_production('population_uncertainty')
pp <- read_production('population_targets'); ppt <- read_production('population_timing')
pm <- read_production('metadata')
stopifnot(nrow(cells)==192, all(ps$attempted==2000),
          !anyDuplicated(ps[c('cell_id','arm','target')]))
z <- merge(ps,cells,by='cell_id',sort=FALSE)
z$family <- ifelse(startsWith(z$arm,'association'),'Association ML','DWLS')
z$coverage <- z$target %in% c('loading','threshold','correlation')
z$eligible <- (z$coverage & z$n>=300) |
  (z$target=='global' & z$n>=1000 & z$generator=='gaussian' & !z$misspecified) |
  (z$target %in% c('nested','mi') & z$n>=1000 & !grepl('power',z$arm))
z$flag <- z$eligible & is.finite(z$rate) &
  ifelse(z$coverage,z$rate<.93,z$rate<.03 | z$rate>.07)
z$clear_flag <- z$flag & ifelse(z$coverage,z$upper<.93,z$upper<.03 | z$lower>.07)
stopifnot(identical(z$flag, z$mechanical_flag))
z$mc_se <- sqrt(z$rate*(1-z$rate)/z$successful)
pct <- function(x) sprintf('%.2f',100*x)
interval <- function(lo,hi) paste0(pct(lo),'–',pct(hi))
cell_label <- function(d) paste0(d$cell_id, ': ',d$factors,'F/',d$categories,' categories/',
  d$imbalance,'/',d$generator,'/',ifelse(d$misspecified,'misspecified','correct'),
  '/N=',d$n,'/G=',d$groups)
rate_table <- function(d) data.frame(Cell=cell_label(d),Arm=d$arm,Target=d$target,
  Attempted=d$attempted,Successful=d$successful,`Rate (%)`=pct(d$rate),
  `Wilson 95% (%)`=interval(d$lower,d$upper),`MC SE (pp)`=pct(d$mc_se),
  Flag=d$flag,check.names=FALSE)
show <- function(d) knitr::kable(d,row.names=FALSE)
```

The registered grid crosses one/two factors, binary/five-category responses,
symmetric/skewed cuts, Gaussian/chi-square2 innovations, correct/misspecified
models, N=300/1,000/4,000 and one/two groups. Each cell uses fresh draws. Rates
are conditional on successful inference; attempted denominators also retain
failures. There is no pooled calibration verdict. Cell IDs below identify the
full design in `cells.csv`; exact rates and Wilson intervals are in `summary.csv`.

### Population approximation before MI interpretation

The first million-row population draw fixes each family's targets; the second
independent draw measures their uncertainty and never replaces the target.

```{r}
show(do.call(rbind,lapply(split(pu,list(pu$estimator,pu$target),drop=TRUE),function(d)
  data.frame(Family=ifelse(d$estimator[1]=='ML','Association ML','DWLS'),Target=d$target[1],
    Populations=nrow(d),`Median absolute target shift`=signif(median(abs(d$target_difference)),3),
    `Largest absolute target shift`=signif(max(abs(d$target_difference)),3),check.names=FALSE))))
```

### MI population approximation check (post-hoc amendment)

The user-approved criterion dated 2026-10-08 was written after seeing production
results. It requires maximum implied size at most 5.5%. For each single-group
MI size cell, let $\Delta=|\lambda_1-\lambda_2|$ be the absolute difference
between the two population loading targets, treated as a conservative bound
on draw-1 approximation error. Using the median loading SE among successful
association-IJ rows, compute $\mathrm{ncp}=(\Delta/\mathrm{SE})^2$ and
$P\{\chi^2_1(\mathrm{ncp})>\chi^2_{1,.95}\}$.

```{r}
mi_check <- read_production('mi_population_check')
stopifnot(nrow(mi_check)==sum(cells$groups==1),
          setequal(mi_check$cell_id,cells$cell_id[cells$groups==1]),
          all(is.finite(mi_check$implied_size)))
mi_worst <- mi_check[which.max(mi_check$implied_size),]
mi_pass <- max(mi_check$implied_size)<=.055
show(data.frame(Cells=nrow(mi_check),
  `Median implied size (%)`=sprintf('%.6f',100*median(mi_check$implied_size)),
  `Maximum implied size (%)`=sprintf('%.6f',100*mi_worst$implied_size),
  `Limit (%)`=5.5,`Maximum cell`=mi_worst$cell_id,
  Verdict=ifelse(mi_pass,'Pass','Fail'),check.names=FALSE))
```

The maximum is **`r sprintf('%.6f',100*mi_worst$implied_size)`%**, in
cell **`r mi_worst$cell_id`** (`r mi_worst$key`, N=`r mi_worst$n`).
The check **`r ifelse(mi_pass,'passes','fails')`** the 5.5% criterion.
The per-cell targets, shifts, median SEs, noncentralities and implied sizes are
frozen in `mi_population_check.csv`, derived from the production frozen at
`b37ebefb` and its local raw replicate SEs. This is a noncentral chi-square
approximation using a median SE, not a confidence bound on population error
or a finite-sample calibration guarantee. It resolves the specified population
approximation prerequisite for MI exposure. Group symmetry supplies the nested
pseudo-true null without fixing an approximated loading.

::: {.callout-note collapse="true" title="Population uncertainty and descriptive global rejection, per cell"}
```{r}
show(pu[c('key','estimator','target','estimate_first','estimate_second','target_difference')])
global_descriptive <- z[z$interpretation=='descriptive_global',]
global_descriptive$key <- with(global_descriptive,paste(factors,categories,imbalance,generator,misspecified,sep='_'))
global_descriptive$estimator <- ifelse(global_descriptive$family=='Association ML','ML','DWLS')
global_descriptive <- merge(global_descriptive,pp[pp$draw==1 & pp$target=='loading',
  c('key','estimator','population_fmin')],by=c('key','estimator'))
show(cbind(rate_table(global_descriptive),`Population discrepancy`=signif(global_descriptive$population_fmin,4)))
```
:::

Global rejection outside Gaussian correct models is descriptive, alongside the
family's population discrepancy; it is never a size gate.

### Registered flags and Monte Carlo uncertainty

Coverage below 93% is flagged at every registered N. Global size outside
[3%,7%] is flagged only at N≥1,000 for Gaussian correct models. Nested and MI
size use that band at N≥1,000 wherever their pseudo-true null holds. Power has
no gate. The point estimate triggers a flag; a Wilson interval wholly beyond
the relevant boundary identifies flags that remain beyond it at 95% precision.
At 2,000 successes, the MC standard error is about 0.49 percentage points for
5% size and 0.49 points for 95% coverage; losses of availability increase it.
Intervals are pointwise, with no simultaneous multiplicity claim.

```{r}
flag_counts <- do.call(rbind,lapply(split(z,list(z$family,z$arm,z$target),drop=TRUE),function(d)
 data.frame(Family=d$family[1],Arm=d$arm[1],Target=d$target[1],
   `Eligible cells`=sum(d$eligible),Flags=sum(d$flag),
   `Wilson wholly beyond boundary`=sum(d$clear_flag),check.names=FALSE)))
show(flag_counts[flag_counts$`Eligible cells`>0,])
```

::: {.callout-note collapse="true" title="Every flagged cell, with Wilson intervals and denominators"}
```{r}
show(rate_table(z[z$flag,][order(z$family[z$flag],z$arm[z$flag],z$cell_id[z$flag]),]))
```
:::

::: {.callout-note collapse="true" title="Covariance coverage per cell, separately by family"}
```{r}
coverage_rows <- z[z$coverage,]
show(rate_table(coverage_rows[order(coverage_rows$family,coverage_rows$cell_id,coverage_rows$target),]))
```
:::

::: {.callout-note collapse="true" title="All global/nested/MI rates, including reported power"}
```{r}
show(rate_table(z[!z$coverage,]))
```
:::

### Availability and failure classes

**NonPositiveDefiniteSample accounts for `r format(sum(pf$count[grepl('NonPositiveDefiniteSample',pf$error_class)]),big.mark=',')`
unavailable panel results. Association-ML cell/arm availability falls as low as
`r pct(min(z$successful[z$family=='Association ML']/z$attempted[z$family=='Association ML']))`%;
DWLS remains at or above
`r pct(min(z$successful[z$family=='DWLS']/z$attempted[z$family=='DWLS']))`%.
This is a panel-result count, not a count of distinct
draws:** a single failed sample is recorded across targets, references and both
families. Raw replicate keys are not in the frozen failure counts, so distinct
failed draws cannot be recovered by summing these rows.

```{r}
show(do.call(rbind,lapply(split(z,list(z$family,z$arm,z$target),drop=TRUE),function(d)
 data.frame(Family=d$family[1],Arm=d$arm[1],Target=d$target[1],
   Attempted=sum(d$attempted),Successful=sum(d$successful),Unavailable=sum(d$failures),
   `Lowest cell availability (%)`=pct(min(d$successful/d$attempted)),check.names=FALSE))))
pf$family <- ifelse(startsWith(pf$arm,'association'),'Association ML','DWLS')
show(aggregate(count~family+error_class,pf,sum))
stopifnot(sum(pf$count)==sum(z$failures))
npd <- pf[grepl('NonPositiveDefiniteSample',pf$error_class),]
npd_cells <- merge(aggregate(count~cell_id+family,npd,sum),cells,by='cell_id')
show(data.frame(Cell=cell_label(npd_cells),Family=npd_cells$family,
  `Unavailable panel results`=npd_cells$count,check.names=FALSE))
```

Availability totals above count target/reference results, and every arm's
individual cell denominator remains visible in the detailed rate tables.
Zero rejections among attempted draws are not substituted for unavailable tests.

### Comparator losses

All is the within-family reference comparator. A comparator loss means lower
availability, or a larger absolute deviation from nominal 5% size in eligible
null cells; descriptive global rejection and power do not select references.
Every loss is listed below, including differences smaller than MC uncertainty.
There are no paired uncertainty estimates in the frozen summaries.

```{r}
comparators <- z[grepl('_(SB|PEBA4)$',z$arm),]
comparators$baseline_arm <- sub('_(SB|PEBA4)$','_All',comparators$arm)
baseline <- z[c('cell_id','arm','target','rate','successful')]
names(baseline)[2:5] <- c('baseline_arm','target','all_rate','all_successful')
comparators <- merge(comparators,baseline,by=c('cell_id','baseline_arm','target'))
comparators$size_loss <- comparators$eligible &
  abs(comparators$rate-.05)>abs(comparators$all_rate-.05)
comparators$availability_loss <- comparators$successful<comparators$all_successful
losses <- comparators[comparators$size_loss | comparators$availability_loss,]
show(aggregate(cbind(size_loss,availability_loss)~family+arm,comparators,sum))
```

::: {.callout-note collapse="true" title="Every within-family comparator loss"}
```{r}
show(cbind(rate_table(losses),`All rate (%)`=pct(losses$all_rate),
  `All successful`=losses$all_successful,`Size distance loss`=losses$size_loss,
  `Availability loss`=losses$availability_loss))
```
:::

DWLS is a separate estimand comparator, with its own targets and policy. The
following inventory lists every cell where DWLS has lower coverage, lower
availability or greater eligible size deviation than association ML for the
matching target/reference. These are descriptive losses, not paired effects
or evidence that either family's target is preferable. DWLS has no MI arm.

::: {.callout-note collapse="true" title="Every same-draw DWLS comparator loss against association ML"}
```{r}
dw <- z[z$family=='DWLS',]; dw$matched_arm <- sub('^dwls','association',dw$arm)
ml <- z[z$family=='Association ML',c('cell_id','arm','target','rate','successful')]
names(ml)[2:5] <- c('matched_arm','target','ml_rate','ml_successful')
dw <- merge(dw,ml,by=c('cell_id','matched_arm','target'))
dw$coverage_loss <- dw$coverage & dw$rate<dw$ml_rate
dw$size_loss <- dw$eligible & !dw$coverage & abs(dw$rate-.05)>abs(dw$ml_rate-.05)
dw$availability_loss <- dw$successful<dw$ml_successful
dw_losses <- dw[dw$coverage_loss | dw$size_loss | dw$availability_loss,]
show(cbind(rate_table(dw_losses),`Association ML rate (%)`=pct(dw_losses$ml_rate),
  `Association ML successful`=dw_losses$ml_successful,Coverage_loss=dw_losses$coverage_loss,
  Size_loss=dw_losses$size_loss,Availability_loss=dw_losses$availability_loss))
```
:::

### Runtime

```{r}
stopifnot(all(runtime$draws==2000),all(pc$priced),
          abs(sum(runtime$total_seconds)/3600-sum(pc$production_cpu_hours))<1e-6)
rt <- merge(runtime,cells,by='cell_id')
show(do.call(rbind,lapply(split(rt,list(rt$groups,rt$n),drop=TRUE),function(d)
 data.frame(Groups=d$groups[1],N=d$n[1],Cells=nrow(d),
   `Summed elapsed hours`=round(sum(d$total_seconds)/3600,2),
   `Cell median seconds range`=paste(round(range(d$median_seconds),3),collapse='–'),
   `Largest cell p99 seconds`=round(max(d$p99_seconds),3),check.names=FALSE))))
```

The sum of draw elapsed times is `r round(sum(runtime$total_seconds)/3600,2)`
hours; `cost.csv` agrees and prices all 192 cells. These are measured serial
draw costs, not independently measured CPU utilization or parallel wall time.
Population timing rows include repeated reuse across cells; deduplicating
key/draw/N/time gives `r round(sum(unique(ppt)$elapsed_seconds)/3600,2)` hours
of recorded population setup. Startup, combine and transfer are excluded.
The metadata records source/package/criteria fingerprints and the simbox
combine command; its git-commit field is empty, so `b37ebefb` identifies the
repository freeze rather than a recorded executor commit.

## Reproduce

Criteria were committed before smoke in `e3333269`. Frozen CSVs are a narrow
exception for registered evidence; raw rows and provenance checkpoints remain
local and ignored. Completed local directories contain checksummed raw and
population files plus `COMPLETE`. Combine refuses source/binary/design,
checksum, mode, seed, replicate-count, duplicate-cell or target disagreements.

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --plan
Rscript run_experiment.R --smoke --out-dir results/smoke-new
Rscript run_experiment.R --pilot --workers 2 --out-dir results/pilot-new
bash scripts/run_cells.sh "$PWD" "$R_LIBS" results/fanout-new 2 smoke 1 97
Rscript run_experiment.R --combine results/fanout-new/cells/cell_001,results/fanout-new/cells/cell_097 --out-dir results/combined-new
Rscript scripts/mi_population_check.R results/production-2026-10-05
quarto render report.qmd
```

Run compute under `nice -n 10` with `OPENBLAS_NUM_THREADS=1`,
`OMP_NUM_THREADS=1` and `MKL_NUM_THREADS=1`. The local two-cell fan-out check
matches serial smoke statistical summaries and population targets; times are
allowed to vary. Use `--population-dir` only with a complete directory carrying
the same source/package fingerprint and mode. Production subsequently ran on simbox with the registered fresh seed base
2117000001; its combined evidence is frozen in `results/production-2026-10-05/`.


## Implication under the registered rules

Association ML has `r sum(z$flag[z$family=='Association ML' & z$coverage])`
coverage flags and `r sum(z$flag[z$family=='Association ML' & !z$coverage])`
size flags across target/reference cells; DWLS has
`r sum(z$flag[z$family=='DWLS' & z$coverage])` coverage flags and
`r sum(z$flag[z$family=='DWLS' & !z$coverage])` size flags. The flag table
separates global, nested and MI rules and identifies uncertainty at each boundary.
Sparse-sample non-positive-definiteness and the other listed failures limit
availability. The MI population approximation check passes the approved post-hoc criterion. These facts request finite-sample and availability review;
flags never justify OPG, expected bread or another inconsistent recipe.

Production reporting is complete. The user decision remains: should ordinary
association-ML inference be exposed, and with which global and nested reference
(All, SB or PEBA4), given the flags, MC uncertainty and availability? No adoption
or reference choice is made here.
