---
title: "Association-ML policy calibration"
format:
  html:
    toc: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r}
paths <- file.path('results','pilot',c('summary.csv','failures.csv','timing.csv','cost.csv','population_timing.csv','population_uncertainty.csv'))
if(!all(file.exists(paths))) stop('Pilot absent: Rscript run_experiment.R --pilot --workers 2')
s <- read.csv(paths[1]); failures <- read.csv(paths[2]); timing <- read.csv(paths[3]); cost <- read.csv(paths[4])
pt <- read.csv(paths[5]); uncertainty <- read.csv(paths[6])
```

## Default register

| Choice | Current value | Status | Evidence |
|---|---|---|---|
| Ordinary association-ML inference | Typed unavailable | Open | Production calibration pending |
| Lab covariance | Exact Stage-1 IJ, observed sensitivity | Provisional for exposure | Component gates; bounded pilot mechanics |
| Lab global / nested references | All, SB, PEBA4 | Open | Bounded pilot; no reference choice yet |
| Lab MI | Robust projected meat, observed Schur metric | Provisional for exposure | Component gates; bounded pilot mechanics |

## Question and short answer

Does the lab association-ML recipe calibrate well enough for ordinary exposure?
The smoke and bounded pilot exercise all four components and the DWLS comparator,
but cannot answer the calibration question. The pilot contains two draws each
in two cells (two-factor, five skewed categories, chi-square2 innovations,
omitted cross-loading, total N=300). It records `r nrow(failures)` unavailable
panel results. Ordinary exposure remains pending production.

## Evidence

The pilot covers a difficult non-Gaussian misspecified population, single-group
coverage/global/MI and two-group metric nesting. It prices only
`r sum(cost$priced)` of `r nrow(cost)` registered cells. Gaussian global size,
other sample sizes and category imbalance panels remain unpriced. Per-cell
rates, attempted and successful denominators and Wilson intervals are in
`results/pilot/summary.csv`; the tiny rates are mechanics, not calibration.

```{r}
shown <- s[startsWith(s$arm,'association') & s$target %in% c('loading','threshold','correlation','mi','nested'),]
shown <- shown[!grepl('_nested_(SB|PEBA4)$',shown$arm),]
knitr::kable(data.frame(Component=shown$arm,Target=shown$target,
  Attempts=shown$attempted,Available=shown$successful,
  Rate=round(shown$rate,2),`Wilson lower`=round(shown$lower,2),`Wilson upper`=round(shown$upper,2)),check.names=FALSE)
```

The covariance recipe is $\mathrm{Cov}(\hat\alpha)=H^{-1}BH^{-\mathsf T}/N$,
where $H$ is observed sensitivity and $B$ transports exact Stage-1 influence.
The global All reference uses the residual correlation-ML metric and exact
Stage-1 sampling Gamma. The nested reference uses the larger model's observed
sensitivity and parameter-space IJ spectrum. Robust MI fixes a loading at its
unrestricted pseudo-true value for size, then 0.15 away for reported power.
DWLS uses its own weighted-discrepancy pseudo-true targets; its coverage is
reported separately and does not estimate a paired method effect.

Population setup uses two independent million-row draws, streamed into pair
counts and passed through library Stage 1. Association-ML and DWLS population
fits share the same first-stage summaries. Both sets of targets and their
differences are frozen in the population CSVs. The largest absolute target
difference is `r signif(max(abs(uncertainty$target_difference)),3)`; production
interpretation must assess these errors relative to interval widths and MI
null sensitivity.

## Cost and caveats

```{r}
priced <- cost[cost$priced,]
knitr::kable(data.frame(`Total N`=priced$n,Groups=priced$groups,
  `Mean seconds per draw`=round(priced$mean_draw_seconds,3),
  `CPU hours for 2,000 draws`=round(priced$production_cpu_hours,3)),check.names=FALSE)
```

Population construction took `r round(sum(pt$elapsed_seconds),1)` seconds on
this workstation. The cost CSV lists every unpriced cell explicitly; the
per-cell estimates above exclude population setup and driver/package startup.
They do not price the full study. Worker parallelism is bounded at two and
math libraries use one thread. Simbox remains unavailable.

Coverage targets are finite population-draw approximations. Gaussian correct
models supply global size gates; global rejection under non-normality or
misspecification is descriptive because the association model need not hold.
Identical groups preserve the pseudo-true metric null. Flags below 93% coverage
or outside 3–7% size at the registered sample sizes request finite-sample work,
never an inconsistent recipe. These pilot flags cannot settle defaults.

Initial development smoke/pilot attempts exposed an extraction error: DWLS
nested `p_all` is intentionally unavailable in that returned bundle. The final
runner computes All from its spectrum using the lab reference primitive.
The corrected smoke and pilot were rerun before freezing; no library change,
criterion change or inference-policy choice followed those attempts.

## Reproduce

Criteria were committed before smoke in `e3333269`. Frozen CSVs are a narrow
exception for registered evidence; raw rows and provenance checkpoints remain
local and ignored. Completed local directories contain checksummed raw and
population files plus `COMPLETE`. Combine refuses source/binary/design,
checksum, mode, seed, replicate-count, duplicate-cell or target disagreements.

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --plan
Rscript run_experiment.R --smoke --out-dir results/smoke-new
Rscript run_experiment.R --pilot --workers 2 --out-dir results/pilot-new
bash scripts/run_cells.sh "$PWD" "$R_LIBS" results/fanout-new 2 smoke 1 97
Rscript run_experiment.R --combine results/fanout-new/cells/cell_001,results/fanout-new/cells/cell_097 --out-dir results/combined-new
quarto render report.qmd
```

Run compute under `nice -n 10` with `OPENBLAS_NUM_THREADS=1`,
`OMP_NUM_THREADS=1` and `MKL_NUM_THREADS=1`. The local two-cell fan-out check
matches serial smoke statistical summaries and population targets; times are
allowed to vary. Use `--population-dir` only with a complete directory carrying
the same source/package fingerprint and mode. Production commands are supported
for the later driver, but no production was run or authorized here.
