---
title: "What does covariance-honest NTML cost?"
subtitle: "Direct PSD fitting versus an audit-triggered fallback"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("full", "pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)
summary <- read_result("case_summary.csv")
validation <- read_result("validation.csv")
failures <- read_result("failures.csv")
baseline <- read.csv(
  file.path("results", "dense_lift_baseline.csv"),
  stringsAsFactors = FALSE, check.names = FALSE)

profile <- metadata[["profile"]]
repeats <- as.integer(metadata[["repeats"]])
is_evidence <- profile %in% c("pilot", "full") && repeats >= 20L
interior <- summary[
  summary$ordinary_admissible & summary$family == "interior scaling", ]
inadmissible <- summary[!summary$ordinary_admissible, ]
median_psd_ratio <- median(interior$psd_over_ordinary)
median_audit_interior <- median(interior$audit_over_ordinary)
median_audit_bad <- median(inadmissible$audit_over_ordinary)
has_ipopt <- "psd_ipopt_median_ms" %in% names(summary) &&
  any(is.finite(summary$psd_ipopt_median_ms))
median_ipopt_ratio <- if (has_ipopt) {
  median(interior$psd_ipopt_over_ordinary, na.rm = TRUE)
} else {
  NA_real_
}
median_ipopt_over_slsqp <- if (has_ipopt) {
  median(
    interior$psd_ipopt_median_ms / interior$psd_median_ms,
    na.rm = TRUE
  )
} else {
  NA_real_
}
median_audit_ipopt_bad <- if (has_ipopt) {
  median(inadmissible$audit_ipopt_over_ordinary, na.rm = TRUE)
} else {
  NA_real_
}
largest <- interior[which.max(interior$p), ]
fallback_count <- sum(!summary$ordinary_admissible)
all_valid <- nrow(failures) == 0L &&
  all(validation$pass)
comparison <- merge(
  baseline, summary,
  by = c("case_id", "p", "ordinary_admissible"),
  suffixes = c("_dense", "_structured"))
comparison$psd_speedup <-
  comparison$psd_median_ms_dense /
  comparison$psd_median_ms_structured
scaling_comparison <- comparison[
  grepl("^interior_cfa_", comparison$case_id), ]
p12_comparison <- scaling_comparison[scaling_comparison$p == 12L, ]

display_fields <- c(
  "case_id", "p", "ordinary_admissible",
  "ordinary_median_ms", "psd_median_ms", "audit_median_ms",
  "psd_over_ordinary", "audit_over_ordinary")
display_names <- c(
  "Case", "p", "Ordinary Admissible",
  "Ordinary (ms)", "PSD Direct (ms)", "Audit Policy (ms)",
  "PSD / Ordinary", "Audit / Ordinary")
if (has_ipopt) {
  display_fields <- c(
    display_fields,
    "psd_ipopt_median_ms", "audit_ipopt_median_ms",
    "psd_ipopt_over_ordinary", "audit_ipopt_over_ordinary"
  )
  display_names <- c(
    display_names,
    "PSD IPOPT (ms)", "Audit IPOPT (ms)",
    "IPOPT / Ordinary", "Audit IPOPT / Ordinary"
  )
}
display <- summary[, display_fields]
names(display) <- display_names
display$Case <- gsub("_", " ", display$Case, fixed = TRUE)
numeric_columns <- setdiff(
  names(display), c("Case", "p", "Ordinary Admissible"))
display[numeric_columns] <- lapply(
  display[numeric_columns], function(x) round(x, 2L))
```

## Question

How much slower is complete-data covariance-honest normal-theory ML than
ordinary NTML, and is the practical policy to fit ordinarily, inspect the
covariance audit, and refit only inadmissible cases?

## Short answer

`r if (!is_evidence) {
  sprintf("This is a %s timing run with %d repetitions per case. It validates the runner but is too small for performance conclusions.", profile, repeats)
} else {
  sprintf("After compiling covariance structure, direct PSD fitting took a median %.1fx as long as ordinary NTML across the interior CFA scaling cases. At p=12, PSD time fell %.1fx and its overhead dropped from %.1fx to %.1fx ordinary NTML. The audit-first policy cost %.2fx ordinary NTML when the first fit was admissible, while the deliberately inadmissible fixtures cost %.1fx because they required both fits.", median_psd_ratio, p12_comparison$psd_speedup, p12_comparison$psd_over_ordinary_dense, p12_comparison$psd_over_ordinary_structured, median_audit_interior, median_audit_bad)
}`
`r if (all_valid) sprintf(" All %d deliberately inadmissible cases were detected by the ordinary audit and repaired by PSD-ML; all final fits were admissible.", fallback_count) else " At least one validation or final-fit check failed; inspect the CSV outputs before interpreting timings."`
`r if (has_ipopt) sprintf(" IPOPT reached the same deterministic optima, but direct IPOPT was %.1fx ordinary NTML and %.1fx PSD-SLSQP across the interior cases; audit-triggered IPOPT was %.1fx ordinary NTML on the inadmissible cases.", median_ipopt_ratio, median_ipopt_over_slsqp, median_audit_ipopt_bad) else ""`

## Evidence

The model and sample-statistic construction is outside the timed region. The
direct policy runs the current PSD estimator from the standard magmaan start.
The audit policy first runs ordinary NTML and, only when its post-fit covariance
audit fails, initializes PSD-ML from that ordinary estimate.

```{r timing-table}
knitr::kable(display, row.names = FALSE)
```

The interior scaling rows use one-factor CFA models with 3, 6, 12, and, in the
full profile, 24 indicators. The remaining rows are exact-covariance fixtures:
a near-boundary uniqueness, a classical Heywood solution, jointly indefinite
latent and residual covariance matrices, a negative structural disturbance,
and an equality-constrained interior control.

The frozen baseline is the same 30-repetition pilot from commit `ee156162`,
before covariance-structure compilation. It was collected on the same host
with the same runner and fit policies.

```{r timing-plot, fig.width=7, fig.height=4.6}
plot_rows <- scaling_comparison[order(scaling_comparison$p), ]
ymax <- max(
  plot_rows$psd_over_ordinary_dense,
  plot_rows$psd_over_ordinary_structured,
  na.rm = TRUE)
old_par <- par(mar = c(4, 4, 1, 1))
on.exit(par(old_par), add = TRUE)
plot(
  plot_rows$p, plot_rows$psd_over_ordinary_dense,
  log = "xy", type = "b", pch = 16, col = "#B33A3A",
  xlab = "Indicators", ylab = "Direct PSD / ordinary NTML time",
  ylim = c(0.8, ymax * 1.15))
lines(
  plot_rows$p, plot_rows$psd_over_ordinary_structured,
  type = "b", pch = 17, col = "#245A9B")
abline(h = 1, lty = 2, col = "grey40")
axis(1, at = plot_rows$p)
legend(
  "topleft",
  legend = c("Dense lift baseline", "Structure-compiled lift"),
  pch = c(16, 17), col = c("#B33A3A", "#245A9B"),
  lty = 1, bty = "n")
```

## Caveats

This is a deterministic engineering benchmark, not an estimate of the
prevalence of improper SEM solutions. The current frontier estimator covers
complete-data normal-theory ML only; FIML and the continuous LS/GMM weight
families are not included.

`r if (has_ipopt) "The IPOPT rows use its limited-memory Hessian mode and create a fresh interior-point problem for every fit. For these small dense lifted SEMs, sparse KKT setup and factorization overhead dominate; the comparison does not claim that every IPOPT formulation or linear-solver build has the same cost." else "IPOPT is opt-in and was not included in this artifact."`

The timing panel deliberately exposes dimension scaling, but it is not a
textbook-corpus benchmark. The results show that the present implementation
still becomes relatively more expensive with indicator count, although the
previous timing cliff is gone. A separate ten-repetition `p=24` direction
probe put direct PSD at about 5.0 times ordinary NTML, down from roughly 711
times under the dense baseline. The ordinary `p=24` control ended with a soft
line-search-salvaged status, so that row is not part of the formal comparison.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript run_experiment.R --full
Rscript run_experiment.R --pilot --include-ipopt
quarto render report.qmd
```
