---
title: "Alpha–omega equality: retained nonregularity derivations"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
read_evidence <- function(name) {
  path <- file.path("results", "overview", name)
  if (!file.exists(path)) stop("Missing retained summary: ", path,
                               "; recover original CSVs and run scripts/summarize.R")
  read.csv(path, stringsAsFactors = FALSE)
}
show_table <- function(x, digits = 3) knitr::kable(x, digits = digits, row.names = FALSE)
```

## Question

What can be learned from the nonregular contrast between coefficient alpha and
fitted one-factor omega at tau-equivalence?

## Short answer

The first-order Wald contrast is singular at the interior equal-loading null.
Its second-order reference is a useful mathematical idea, but the retained main
run has only **eight replications per cell**. It cannot establish nominal size
or show that an Imhof reference fixes calibration. This study is **banked**.
Reopen for a named nonregular-functional testing consumer or a methods paper;
equal-loading Wald, score and LR tests are the cheaper existing alternative.

## Evidence

For a one-factor model with local independence, write $L=\sum_i\lambda_i$ and
$V=L^2+\sum_i\theta_i$. Then

$$
\omega-\alpha=
\frac{p\sum_i\lambda_i^2-L^2}{(p-1)V}\geq0.
$$

At equal loadings the numerator and its gradient vanish. A first-order delta
variance therefore collapses; under an interior regular parameter estimator,
$N(\widehat\omega-\widehat\alpha)$ has a quadratic-form limit rather than an
ordinary first-order normal limit. Its eigenvalues depend on the estimator's
sampling covariance. The derivations retain the assumptions and show why a
normal-theory covariance cannot simply be substituted for nonnormal inference:
[second-order reference](design/alpha_omega_second_order.md),
[nonnormal delta map](design/alpha_delta_nonnormality.md), and
[ULS identity](design/alpha_uls_identity.md).

A deterministic ULS tau-model check agrees with sample alpha:

```{r identity}
show_table(read_evidence("validation_identity.csv"), digits = 10)
```

The local main metadata records $N=250$, $p=6$, normal draws, eight replications
and bootstrap size 200. The table includes the alternative as a wiring check;
rejection proportions this coarse are not calibration evidence.

```{r smoke}
x <- read_evidence("simulation_summary.csv")
show_table(x[, c("population", "dist", "N", "p", "n_rel_ok", "n_wald_ok", "n_st_ok",
                 "reject_wald_nm", "reject_imhof_nm", "reject_score")])
```

The Holzinger–Swineford demonstration uses three-item subscales from lavaan's
public teaching data. It is not a reproduction of the original 26-item
Deng–Chan data. The bootstrap comparison is a diagnostic on one dataset.

## Caveats

The nonregular result concerns this contrast, estimator and interior null;
it does not supply a boundary law or a general reliability-testing policy.
The proposed larger indicator/sample-size and nonnormal design in `R/populations.R`
is preserved but has no queued run. The redundant corrective-note draft was
folded into this report and the mathematical derivations. Core functional and
structural-test APIs are unchanged.

## Reproduce

From this directory, inspect `Rscript run_experiment.R --help`. A cheap fresh
verification run is:

```bash
Rscript run_experiment.R --reps 2 --N 250 --boot-B 2 \
  --cells population=tau,dist=normal --results-dir /tmp/alpha_omega_check
Rscript scripts/summarize.R
```

The summary command reads the retained originals, not the temporary smoke run.
[Source hashes](results/overview/sources.csv) document the compact evidence.
