---
title: "Pairwise Composite Ordinal Nested LR Probe"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
rates_path <- file.path("results", "rejection_rates.csv")
p_path <- file.path("results", "p_values.csv")
if (!file.exists(rates_path) || !file.exists(p_path)) {
  stop("Run `Rscript run_experiment.R --smoke` first.", call. = FALSE)
}
rates <- read.csv(rates_path)
p_values <- read.csv(p_path)
meta <- tryCatch(read.csv(file.path("results", "metadata.csv")),
                 error = function(e) data.frame())

tab_long <- rates[rates$usable > 0 & is.finite(rates$rejection_rate),
                  c("truth", "method", "rejection_rate", "mc_se",
                    "mean_base", "mean_trace",
                    "mean_h1_condition_bread", "mean_h0_condition_bread")]
tab_long$condition <- ifelse(tab_long$truth == "fake", "Null", "Loading Shift")
tab_long$Rejection <- sprintf("%.3f", tab_long$rejection_rate)
tab_long$`MC SE` <- sprintf("%.3f", tab_long$mc_se)
tab_long$`Mean LR` <- sprintf("%.2f", tab_long$mean_base)
tab_long$`Mean Trace` <- sprintf("%.2f", tab_long$mean_trace)
tab_long$`H1 Bread Cond.` <- sprintf("%.1f", tab_long$mean_h1_condition_bread)
tab_long$`H0 Bread Cond.` <- sprintf("%.1f", tab_long$mean_h0_condition_bread)
tab <- tab_long[, c("condition", "method", "Rejection", "MC SE",
                    "Mean LR", "Mean Trace", "H1 Bread Cond.",
                    "H0 Bread Cond.")]
names(tab)[1:2] <- c("Condition", "Method")
tab <- tab[order(tab$Condition,
                 match(tab$Method, c("naive", "scaled", "adjusted", "mixture"))), ]

design <- if (nrow(meta)) meta[1L, ] else data.frame(
  reps = length(unique(p_values$rep)),
  n = paste(unique(p_values$n_per_group), collapse = ","),
  missing_rate = paste(unique(p_values$missing_rate), collapse = ","),
  truths = paste(unique(p_values$truth), collapse = ","))
```

## Simulated Model

This is a first smoke-testable probe for the frontier pairwise/composite ordinal
estimator's nested LR inference. It uses experiment 26's ordinal thresholds and
MCAR missingness pattern, narrowed to a single-group one-factor CFA so the
finite-difference Godambe calculation is cheap enough to run interactively:

```text
f =~ y1 + y2 + y3 + y4
```

The null constrains the `y2` and `y3` loadings equal. The `fake` condition is
the null, with both loadings set to 0.70. The `real` condition lowers the `y3`
loading to 0.45. MCAR missingness is applied independently to all four
indicators.

The test methods are:

- `naive`: unscaled \(2(\ell_1 - \ell_0)\) on the reference \(\chi^2\).
- `scaled`: mean-scaled Satorra-2000 statistic from the Godambe spectrum.
- `adjusted`: Satterthwaite-style mean/variance adjusted statistic.
- `mixture`: exact weighted-\(\chi^2\) mixture p-value.

## Rejection Rates

```{r}
knitr::kable(tab, row.names = FALSE)
```

## Run Metadata

```{r}
knitr::kable(design, row.names = FALSE)
```

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 50 --n 300 --missing-rate .20
quarto render report.qmd
```
