---
title: "Ordinal threshold constraints"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "threshold_constraints.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

constraints_path <- file.path(results_dir, "threshold_constraints.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(constraints_path)) {
  stop("Missing threshold_constraints.csv. Run: Rscript run_experiment.R --smoke")
}

constraints <- utils::read.csv(constraints_path, check.names = FALSE,
                               stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.2f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.3g", x))
fmt_diff <- function(x) ifelse(is.na(x), "", sprintf("%.2e", x))
max_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else max(x)
}
median_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else stats::median(x)
}
first_text <- function(x) {
  x <- x[nzchar(x)]
  if (!length(x)) "" else x[[1L]]
}

method_levels <- c("full_bounded", "threshold_profiled_bounded",
                   "full_threshold_snlls", "threshold_profiled_snlls")
method_labels <- c("Full bounded", "Threshold-profiled bounded",
                   "Full-threshold SNLLS", "Threshold-profiled SNLLS")
constraints$Method <- factor(constraints$path, levels = method_levels,
                             labels = method_labels)
constraints$Case <- factor(
  constraints$case_id,
  levels = c("free_thresholds", "shared_first_threshold",
             "one_item_symmetric_thresholds",
             "all_items_symmetric_thresholds"),
  labels = c("Free thresholds", "Shared first threshold",
             "One item symmetric thresholds",
             "All items symmetric thresholds")
)
ok <- constraints[constraints$status == "ok", ]
```

## Question

This experiment asks which ordinal fitting paths can handle equality
constraints on thresholds. The target use case is small but important:
threshold constraints should be available when the ordinal layer is part of the
model, not only when every threshold can be profiled away before optimization.

The design is deliberately single-group ordinal CFA. It avoids the R/lavaan
grouping machinery so that the C++ estimator behavior is isolated:

- free thresholds;
- one shared threshold label across items;
- one linear threshold symmetry constraint, `a + b == 0`;
- the same symmetry constraint repeated for all items.

## Short Answer

Shared threshold labels are still profileable: the threshold-profiled and
full-threshold paths agree. General linear threshold constraints are different.
The threshold-profiled bounded and threshold-profiled SNLLS paths reject them
because the ordinal threshold map cannot eliminate constrained thresholds. The
full bounded and full-threshold SNLLS paths keep thresholds in the parameter
vector and fit the constrained model successfully.

The full-threshold SNLLS row is the main research hook here: thresholds remain
in the full moment stack and enter the SNLLS decomposition as ordinary
Golub-Pereyra linear coordinates, so the estimator can support threshold
constraints without falling back to a fully nonlinear covariance block.

## Run

```{r}
data.frame(
  Item = metadata$key,
  Value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Support Matrix

```{r}
support <- aggregate(seed ~ Case + constraint_kind + estimator + Method + status,
                     data = constraints, FUN = length)
names(support) <- c("Case", "Constraint", "Estimator", "Method", "Status",
                    "Rows")
knitr::kable(support[order(support$Case, support$Estimator, support$Method), ],
             row.names = FALSE)
```

```{r}
errors <- constraints[constraints$status != "ok",
                      c("Case", "constraint_kind", "estimator", "Method",
                        "error")]
if (nrow(errors)) {
  errors <- aggregate(error ~ Case + constraint_kind + estimator + Method,
                      data = errors, FUN = first_text)
  names(errors) <- c("Case", "Constraint", "Estimator", "Method", "Error")
  knitr::kable(errors[order(errors$Case, errors$Estimator, errors$Method), ],
               row.names = FALSE)
}
```

## Agreement

```{r}
agreement <- aggregate(
  cbind(theta_diff_full_bounded, fmin_diff_full_bounded,
        theta_diff_profiled_bounded, fmin_diff_profiled_bounded) ~
    Case + constraint_kind + estimator + Method,
  data = ok,
  FUN = max_or_na,
  na.action = na.pass
)
names(agreement) <- c("Case", "Constraint", "Estimator", "Method",
                      "Max |theta - full|", "Max |fmin - full|",
                      "Max |theta - profiled|", "Max |fmin - profiled|")
for (j in 5:ncol(agreement)) agreement[[j]] <- fmt_diff(agreement[[j]])
knitr::kable(agreement[order(agreement$Case, agreement$Estimator,
                             agreement$Method), ],
             row.names = FALSE)
```

`Max |theta - profiled|` is blank for linear threshold-constraint cases because
the profiled paths intentionally reject those models.

## Timing Context

```{r}
wide <- reshape(
  ok[, c("case_id", "constraint_kind", "estimator", "path",
         "fit_median_ms")],
  idvar = c("case_id", "constraint_kind", "estimator"),
  timevar = "path",
  direction = "wide"
)
names(wide) <- sub("^fit_median_ms\\.", "", names(wide))
for (name in method_levels) {
  if (!name %in% names(wide)) wide[[name]] <- NA_real_
}
wide$profile_gain <- wide$full_bounded / wide$threshold_profiled_bounded
wide$full_snlls_gain <- wide$full_bounded / wide$full_threshold_snlls
wide$snlls_incremental <- wide$threshold_profiled_bounded /
  wide$threshold_profiled_snlls
wide$threshold_snlls_gain <- wide$full_threshold_snlls /
  wide$threshold_profiled_snlls
wide$total_gain <- wide$full_bounded / wide$threshold_profiled_snlls
wide$Case <- factor(
  wide$case_id,
  levels = c("free_thresholds", "shared_first_threshold",
             "one_item_symmetric_thresholds",
             "all_items_symmetric_thresholds"),
  labels = c("Free thresholds", "Shared first threshold",
             "One item symmetric thresholds",
             "All items symmetric thresholds")
)

timing <- wide[, c("Case", "constraint_kind", "estimator", "full_bounded",
                   "threshold_profiled_bounded", "full_threshold_snlls",
                   "threshold_profiled_snlls", "profile_gain",
                   "full_snlls_gain", "threshold_snlls_gain")]
names(timing) <- c("Case", "Constraint", "Estimator", "Full bounded ms",
                   "Profiled bounded ms", "Full-threshold SNLLS ms",
                   "Profiled SNLLS ms", "Full / profiled",
                   "Full / full-threshold SNLLS",
                   "Full-threshold / profiled SNLLS")
for (name in c("Full bounded ms", "Profiled bounded ms",
               "Full-threshold SNLLS ms", "Profiled SNLLS ms")) {
  timing[[name]] <- fmt_ms(timing[[name]])
}
for (name in c("Full / profiled", "Full / full-threshold SNLLS",
               "Full-threshold / profiled SNLLS")) {
  timing[[name]] <- fmt_num(timing[[name]])
}
knitr::kable(timing[order(timing$Case, timing$Estimator), ],
             row.names = FALSE)
```

Ratios greater than one favor the denominator. Timing is secondary here: these
are tiny models, fit-only timings exclude statistic construction, and the
linear-constraint rows have no threshold-profiled comparator by design.

## Cost Context

```{r}
cost <- aggregate(cbind(cache_setup_ms, weight_setup_ms) ~
                    Case + constraint_kind + estimator + Method,
                  data = constraints,
                  FUN = median_or_na,
                  na.action = na.pass)
names(cost) <- c("Case", "Constraint", "Estimator", "Method",
                 "Cache setup ms", "Weight setup ms")
for (j in 5:ncol(cost)) cost[[j]] <- fmt_ms(cost[[j]])
knitr::kable(cost[order(cost$Case, cost$Estimator, cost$Method), ],
             row.names = FALSE)
```

## Reproduce

```sh
Rscript experiments/12-ordinal-threshold-constraints/run_experiment.R --smoke --preset fast
(cd experiments/12-ordinal-threshold-constraints && quarto render report.qmd)
```

For less noisy timings, use `--preset opt` and increase `--reps`. This remains
a support experiment, not an end-to-end timing claim.
