---
title: "Ordinal SNLLS speed pilot"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "speed.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

speed_path <- file.path(results_dir, "speed.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(speed_path)) {
  stop("Missing speed.csv. Run: Rscript run_experiment.R --smoke")
}

speed <- utils::read.csv(speed_path, check.names = FALSE,
                         stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)
for (name in c("construction", "total_median_ms", "total_min_ms",
               "total_max_ms", "parameterization", "data_kind",
               "n_ordered", "n_continuous")) {
  if (!name %in% names(speed)) speed[[name]] <- NA
}
speed$parameterization[is.na(speed$parameterization)] <- "delta"
speed$data_kind[is.na(speed$data_kind)] <-
  ifelse(grepl("^mixed", speed$design), "mixed", "ordinal")
speed$n_ordered[is.na(speed$n_ordered)] <-
  ifelse(speed$data_kind == "ordinal", speed$p, NA_integer_)
speed$n_continuous[is.na(speed$n_continuous)] <-
  ifelse(speed$data_kind == "ordinal", 0L, NA_integer_)

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.2f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.3g", x))
fmt_diff <- function(x) ifelse(is.na(x), "", sprintf("%.2e", x))
max_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else max(x)
}
median_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else stats::median(x)
}

fit_method_levels <- c("full_bounded", "threshold_profiled_bounded",
                       "snlls_full_thresholds", "snlls")
e2e_levels <- c("snlls_e2e_legacy", "snlls_e2e_lazy",
                "full_bounded_e2e_legacy",
                "snlls_full_thresholds_e2e_legacy")
method_levels <- c(fit_method_levels, e2e_levels)
method_labels <- c("Full bounded", "Threshold-profiled bounded",
                   "Full-threshold SNLLS", "Threshold-profiled SNLLS",
                   "Legacy raw-to-SNLLS", "Lazy raw-to-SNLLS",
                   "Mixed raw-to-bounded",
                   "Mixed raw-to-full-threshold SNLLS")
speed$Method <- factor(speed$path, levels = method_levels,
                       labels = method_labels)
speed$Data <- factor(speed$data_kind,
                     levels = c("ordinal", "mixed"),
                     labels = c("All ordinal", "Mixed"))
speed$Parameterization <- factor(
  speed$parameterization,
  levels = c("delta", "theta"),
  labels = c("Delta", "Theta")
)
speed$timed_ms <- ifelse(is.finite(speed$total_median_ms),
                         speed$total_median_ms, speed$fit_median_ms)
design_levels <- c("kreiberg_scaling", "worked_example", "threshold_stress",
                   "mixed_scaling", "mixed_worked_example")
design_labels <- c("Kreiberg scaling", "Worked example", "Threshold stress",
                   "Mixed scaling", "Mixed worked example")
speed$Design <- factor(
  speed$design,
  levels = design_levels,
  labels = design_labels
)
ok <- speed[speed$status == "ok", ]

wide <- reshape(
  ok[, c("data_kind", "design", "case_id", "q", "p", "n_ordered",
         "n_continuous", "categories", "threshold_balance",
         "threshold_mode", "parameterization", "estimator", "path",
         "fit_median_ms")],
  idvar = c("data_kind", "design", "case_id", "q", "p", "n_ordered",
            "n_continuous", "categories",
            "threshold_balance", "threshold_mode", "parameterization",
            "estimator"),
  timevar = "path",
  direction = "wide"
)
names(wide) <- sub("^fit_median_ms\\.", "", names(wide))
wide$Parameterization <- factor(
  wide$parameterization,
  levels = c("delta", "theta"),
  labels = c("Delta", "Theta")
)
wide$Data <- factor(wide$data_kind,
                    levels = c("ordinal", "mixed"),
                    labels = c("All ordinal", "Mixed"))
for (name in fit_method_levels) {
  if (!name %in% names(wide)) wide[[name]] <- NA_real_
}
wide$profile_gain <- wide$full_bounded / wide$threshold_profiled_bounded
wide$full_snlls_gain <- wide$full_bounded / wide$snlls_full_thresholds
wide$snlls_incremental <- wide$threshold_profiled_bounded / wide$snlls
wide$threshold_snlls_gain <- wide$snlls_full_thresholds / wide$snlls
wide$total_gain <- wide$full_bounded / wide$snlls

e2e <- ok[ok$path %in% e2e_levels, ]
if (nrow(e2e)) {
  wide_e2e <- reshape(
    e2e[, c("data_kind", "design", "case_id", "q", "p", "n_ordered",
            "n_continuous", "categories", "threshold_balance",
            "threshold_mode", "parameterization", "estimator", "path",
            "timed_ms")],
    idvar = c("data_kind", "design", "case_id", "q", "p", "n_ordered",
              "n_continuous", "categories",
              "threshold_balance", "threshold_mode", "parameterization",
              "estimator"),
    timevar = "path",
    direction = "wide"
  )
  names(wide_e2e) <- sub("^timed_ms\\.", "", names(wide_e2e))
  for (name in e2e_levels) {
    if (!name %in% names(wide_e2e)) wide_e2e[[name]] <- NA_real_
  }
} else {
  wide_e2e <- data.frame(
    data_kind = character(),
    design = character(),
    case_id = character(),
    q = integer(),
    p = integer(),
    n_ordered = integer(),
    n_continuous = integer(),
    categories = integer(),
    threshold_balance = character(),
    threshold_mode = character(),
    parameterization = character(),
    estimator = character(),
    snlls_e2e_legacy = numeric(),
    snlls_e2e_lazy = numeric(),
    stringsAsFactors = FALSE
  )
}
wide_e2e$Data <- factor(wide_e2e$data_kind,
                        levels = c("ordinal", "mixed"),
                        labels = c("All ordinal", "Mixed"))
wide_e2e$Parameterization <- factor(
  wide_e2e$parameterization,
  levels = c("delta", "theta"),
  labels = c("Delta", "Theta")
)
for (name in e2e_levels) {
  if (!name %in% names(wide_e2e)) wide_e2e[[name]] <- NA_real_
}
wide_e2e$lazy_gain <- wide_e2e$snlls_e2e_legacy /
  wide_e2e$snlls_e2e_lazy
wide_e2e$mixed_e2e_gain <- wide_e2e$full_bounded_e2e_legacy /
  wide_e2e$snlls_full_thresholds_e2e_legacy
```

## Question

This pilot asks whether ordinal SNLLS shows the same basic speed pattern as
the continuous Kreiberg-style SNLLS examples once thresholds and ordinal
weights are in the objective. It now separates four costs where available:
legacy full bounded fitting, threshold-profiled bounded fitting,
full-threshold SNLLS, and threshold-profiled SNLLS, for both delta and theta
parameterizations. It also adds raw-data-to-fit SNLLS rows that compare the
legacy eager ordinal constructor against the lazy fit-only workspace constructor
for ULS/DWLS. Mixed continuous/ordinal delta rows extend the same full bounded
versus full-threshold SNLLS comparison, including raw-to-fit rows where mixed
statistics, starts, and the optimizer are all inside the timed operation. The
full direct bounded comparator is included for DWLS/WLS; ULS has the
cache-aware comparator paths.

The design deliberately includes three views:

- a repeated-measure four-factor family indexed by indicators per factor `q`;
- one worked repeated-measure example for table/prose use;
- a mixed continuous/ordinal repeated-measure family using paired ordinal and
  continuous indicators per factor;
- an ordinal-specific threshold stress slice varying categories, balance, and
  fixed/shared threshold constraints.

## Short Answer

In this pilot, the full direct comparator is a useful diagnostic rather than a
guaranteed slower baseline: fit-only timings can favor the legacy full bounded
path after statistic construction is excluded. For delta, the full-threshold
SNLLS row profiles thresholds as ordinary Golub-Pereyra linear parameters,
while the threshold-profiled SNLLS row applies the ordinal threshold map first
and then profiles the remaining conditionally-linear covariance block. For
theta, the standardized covariance moments make that covariance block
nonlinear, so SNLLS profiles threshold free parameters only. The fitted
solutions match the bounded objectives to numerical tolerance.

The end-to-end rows are the construction-aware check: legacy rows rebuild
`OrdinalStats`, project moments, build starts/cache, and fit profiled SNLLS
inside the timed operation; lazy rows rebuild `OrdinalWorkspace`, build starts,
and fit profiled SNLLS. That makes ULS the cleanest construction win. DWLS is
more nuanced because exact lazy construction still computes the score work
needed for the Gamma diagonal. The mixed rows are delta-only and compare the
materialized mixed full bounded path to materialized mixed full-threshold
SNLLS; mixed theta SNLLS and lazy mixed Gamma construction are deliberately not
claimed here. The result is still a pilot, not a paper-grade simulation: the
design is synthetic and there is no lavaan timing context here.

## Run

```{r}
data.frame(
  Item = metadata$key,
  Value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Status

```{r}
status <- aggregate(seed ~ Data + Design + Parameterization + estimator + Method +
                      status,
                    data = speed, FUN = length)
names(status) <- c("Data", "Design", "Parameterization", "Estimator",
                   "Method", "Status", "Rows")
knitr::kable(status, row.names = FALSE)
```

## Agreement

```{r}
agreement <- aggregate(
  cbind(theta_diff_profiled_bounded, fmin_diff_profiled_bounded,
        theta_diff_full_bounded, fmin_diff_full_bounded) ~
    Data + Design + Parameterization + estimator + Method,
  data = ok,
  FUN = max_or_na,
  na.action = na.pass
)
names(agreement) <- c("Data", "Design", "Parameterization", "Estimator",
                      "Method",
                      "Max |theta - profiled|", "Max |fmin - profiled|",
                      "Max |theta - full|", "Max |fmin - full|")
for (j in 6:ncol(agreement)) agreement[[j]] <- fmt_diff(agreement[[j]])
knitr::kable(agreement, row.names = FALSE)
```

## Speedup

```{r}
summary <- aggregate(cbind(profile_gain, full_snlls_gain, snlls_incremental,
                           threshold_snlls_gain, total_gain) ~
                       Data + design + Parameterization + estimator,
                     data = wide,
                     FUN = median_or_na,
                     na.action = na.pass)
summary$Design <- factor(
  summary$design,
  levels = design_levels,
  labels = design_labels
)
summary$`Full / profiled` <- fmt_num(summary$profile_gain)
summary$`Full / full-threshold SNLLS` <- fmt_num(summary$full_snlls_gain)
summary$`Profiled / profiled SNLLS` <- fmt_num(summary$snlls_incremental)
summary$`Full-threshold / profiled SNLLS` <-
  fmt_num(summary$threshold_snlls_gain)
summary$`Full / profiled SNLLS` <- fmt_num(summary$total_gain)
knitr::kable(summary[, c("Data", "Design", "Parameterization", "estimator",
                         "Full / profiled",
                         "Full / full-threshold SNLLS",
                         "Profiled / profiled SNLLS",
                         "Full-threshold / profiled SNLLS",
                         "Full / profiled SNLLS")],
             row.names = FALSE,
             col.names = c("Data", "Design", "Parameterization", "Estimator",
                           "Full / Profiled",
                           "Full / Full-Threshold SNLLS",
                           "Profiled / Profiled SNLLS",
                           "Full-Threshold / Profiled SNLLS",
                           "Full / Profiled SNLLS"))
```

Ratios greater than one favor the denominator. `Full / Profiled` can be below
one because the pilot excludes statistic construction and compares different
bounded objective implementations, not just the presence or absence of
threshold profiling.

```{r}
#| fig-width: 8
#| fig-height: 5
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  scaling <- wide[wide$design == "kreiberg_scaling", ]
  plot_data <- rbind(
    transform(scaling[is.finite(scaling$profile_gain), ],
              Ratio = profile_gain, Comparison = "Full / profiled"),
    transform(scaling[is.finite(scaling$full_snlls_gain), ],
              Ratio = full_snlls_gain,
              Comparison = "Full / full-threshold SNLLS"),
    transform(scaling[is.finite(scaling$snlls_incremental), ],
              Ratio = snlls_incremental,
              Comparison = "Profiled / profiled SNLLS"),
    transform(scaling[is.finite(scaling$threshold_snlls_gain), ],
              Ratio = threshold_snlls_gain,
              Comparison = "Full-threshold / profiled SNLLS"),
    transform(scaling[is.finite(scaling$total_gain), ],
              Ratio = total_gain, Comparison = "Full / profiled SNLLS")
  )
  ggplot(plot_data, aes(q, Ratio, color = estimator,
                        linetype = Parameterization)) +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_line() +
    geom_point(size = 2) +
    facet_wrap(~ Comparison, scales = "free_y") +
    labs(x = "Indicators per factor (q)",
         y = "Fit-time ratio",
         color = "Estimator",
         linetype = "Parameterization") +
    theme_minimal(base_size = 12)
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

## Worked Example

```{r}
worked <- wide[wide$design == "worked_example", ]
worked_table <- worked[, c("case_id", "q", "p", "categories",
                           "Parameterization", "estimator", "full_bounded",
                           "threshold_profiled_bounded",
                           "snlls_full_thresholds", "snlls", "profile_gain",
                           "full_snlls_gain", "snlls_incremental",
                           "threshold_snlls_gain", "total_gain")]
names(worked_table) <- c("Case", "q", "p", "Categories",
                         "Parameterization", "Estimator",
                         "Full bounded ms", "Profiled bounded ms",
                         "Full-threshold SNLLS ms",
                         "Profiled SNLLS ms", "Full / profiled",
                         "Full / full-threshold SNLLS",
                         "Profiled / profiled SNLLS",
                         "Full-threshold / profiled SNLLS",
                         "Full / profiled SNLLS")
worked_table$`Full bounded ms` <- fmt_ms(worked_table$`Full bounded ms`)
worked_table$`Profiled bounded ms` <-
  fmt_ms(worked_table$`Profiled bounded ms`)
worked_table$`Full-threshold SNLLS ms` <-
  fmt_ms(worked_table$`Full-threshold SNLLS ms`)
worked_table$`Profiled SNLLS ms` <- fmt_ms(worked_table$`Profiled SNLLS ms`)
worked_table$`Full / profiled` <- fmt_num(worked_table$`Full / profiled`)
worked_table$`Full / full-threshold SNLLS` <-
  fmt_num(worked_table$`Full / full-threshold SNLLS`)
worked_table$`Profiled / profiled SNLLS` <-
  fmt_num(worked_table$`Profiled / profiled SNLLS`)
worked_table$`Full-threshold / profiled SNLLS` <-
  fmt_num(worked_table$`Full-threshold / profiled SNLLS`)
worked_table$`Full / profiled SNLLS` <-
  fmt_num(worked_table$`Full / profiled SNLLS`)
knitr::kable(worked_table, row.names = FALSE)
```

## Mixed Continuous/Ordinal

```{r}
mixed <- wide[wide$data_kind == "mixed", ]
if (nrow(mixed)) {
  mixed_table <- mixed[, c("design", "case_id", "q", "p", "n_ordered",
                           "n_continuous", "categories", "estimator",
                           "full_bounded", "snlls_full_thresholds",
                           "full_snlls_gain")]
  mixed_table$Design <- factor(mixed_table$design,
                               levels = design_levels,
                               labels = design_labels)
  names(mixed_table) <- c("design", "Case", "q", "p", "Ordered",
                          "Continuous", "Categories", "Estimator",
                          "Full bounded ms",
                          "Full-threshold SNLLS ms",
                          "Full / full-threshold SNLLS", "Design")
  mixed_table$`Full bounded ms` <- fmt_ms(mixed_table$`Full bounded ms`)
  mixed_table$`Full-threshold SNLLS ms` <-
    fmt_ms(mixed_table$`Full-threshold SNLLS ms`)
  mixed_table$`Full / full-threshold SNLLS` <-
    fmt_num(mixed_table$`Full / full-threshold SNLLS`)
  knitr::kable(mixed_table[, c("Design", "Case", "q", "p", "Ordered",
                               "Continuous", "Categories", "Estimator",
                               "Full bounded ms",
                               "Full-threshold SNLLS ms",
                               "Full / full-threshold SNLLS")],
               row.names = FALSE)
} else {
  cat("No mixed continuous/ordinal rows in this result file.")
}
```

Mixed `q` counts ordinal/continuous pairs per factor: `q = 2` means four
factors with eight ordinal and eight continuous indicators in total. These rows
use the materialized mixed statistics path, so the fit-only comparison answers
whether the mixed Golub-Pereyra split helps after Gamma/weight construction has
already happened.

## Threshold Stress

```{r}
stress <- wide[wide$design == "threshold_stress", ]
stress_table <- stress[, c("categories", "threshold_balance",
                           "threshold_mode", "Parameterization", "estimator",
                           "profile_gain", "full_snlls_gain",
                           "snlls_incremental", "threshold_snlls_gain",
                           "total_gain")]
names(stress_table) <- c("Categories", "Threshold balance",
                         "Threshold mode", "Parameterization", "Estimator",
                         "Full / profiled",
                         "Full / full-threshold SNLLS",
                         "Profiled / profiled SNLLS",
                         "Full-threshold / profiled SNLLS",
                         "Full / profiled SNLLS")
stress_table$`Full / profiled` <- fmt_num(stress_table$`Full / profiled`)
stress_table$`Full / full-threshold SNLLS` <-
  fmt_num(stress_table$`Full / full-threshold SNLLS`)
stress_table$`Profiled / profiled SNLLS` <-
  fmt_num(stress_table$`Profiled / profiled SNLLS`)
stress_table$`Full-threshold / profiled SNLLS` <-
  fmt_num(stress_table$`Full-threshold / profiled SNLLS`)
stress_table$`Full / profiled SNLLS` <-
  fmt_num(stress_table$`Full / profiled SNLLS`)
knitr::kable(stress_table[order(stress_table$Categories,
                                stress_table$`Threshold balance`,
                                stress_table$`Threshold mode`,
                                stress_table$Parameterization,
                                stress_table$Estimator), ],
             row.names = FALSE)
```

Threshold stress keeps the same repeated-measure covariance model but changes
the ordinal layer. This is where fixed and shared threshold rows can erode or
preserve the profiling advantage.

## End-To-End Construction

```{r}
ordinal_e2e <- wide_e2e[wide_e2e$data_kind == "ordinal", ]
if (nrow(ordinal_e2e)) {
  e2e_summary <- aggregate(cbind(snlls_e2e_legacy, snlls_e2e_lazy,
                                 lazy_gain) ~ design + Parameterization +
                             estimator,
                           data = ordinal_e2e,
                           FUN = median_or_na,
                           na.action = na.pass)
  e2e_summary$Design <- factor(
    e2e_summary$design,
    levels = design_levels,
    labels = design_labels
  )
  e2e_summary$`Legacy raw-to-SNLLS ms` <-
    fmt_ms(e2e_summary$snlls_e2e_legacy)
  e2e_summary$`Lazy raw-to-SNLLS ms` <-
    fmt_ms(e2e_summary$snlls_e2e_lazy)
  e2e_summary$`Legacy / lazy` <- fmt_num(e2e_summary$lazy_gain)
  knitr::kable(e2e_summary[, c("Design", "Parameterization", "estimator",
                               "Legacy raw-to-SNLLS ms",
                               "Lazy raw-to-SNLLS ms", "Legacy / lazy")],
               row.names = FALSE,
               col.names = c("Design", "Parameterization", "Estimator",
                             "Legacy Raw-to-SNLLS ms",
                             "Lazy Raw-to-SNLLS ms", "Legacy / Lazy"))
} else {
  cat("No all-ordinal end-to-end rows in this result file.")
}

mixed_e2e <- wide_e2e[wide_e2e$data_kind == "mixed", ]
if (nrow(mixed_e2e)) {
  mixed_e2e_summary <- aggregate(
    cbind(full_bounded_e2e_legacy,
          snlls_full_thresholds_e2e_legacy,
          mixed_e2e_gain) ~ design + estimator,
    data = mixed_e2e,
    FUN = median_or_na,
    na.action = na.pass
  )
  mixed_e2e_summary$Design <- factor(mixed_e2e_summary$design,
                                     levels = design_levels,
                                     labels = design_labels)
  mixed_e2e_summary$`Mixed raw-to-bounded ms` <-
    fmt_ms(mixed_e2e_summary$full_bounded_e2e_legacy)
  mixed_e2e_summary$`Mixed raw-to-SNLLS ms` <-
    fmt_ms(mixed_e2e_summary$snlls_full_thresholds_e2e_legacy)
  mixed_e2e_summary$`Bounded / SNLLS` <-
    fmt_num(mixed_e2e_summary$mixed_e2e_gain)
  knitr::kable(mixed_e2e_summary[, c("Design", "estimator",
                                     "Mixed raw-to-bounded ms",
                                     "Mixed raw-to-SNLLS ms",
                                     "Bounded / SNLLS")],
               row.names = FALSE,
               col.names = c("Design", "Estimator",
                             "Mixed Raw-to-Bounded ms",
                             "Mixed Raw-to-SNLLS ms",
                             "Bounded / SNLLS"))
}
```

These rows include raw ordinal construction, start values, any needed
Gamma/cache work, and the threshold-profiled SNLLS fit. Ratios greater than
one favor lazy construction. WLS is legacy-only here because the fit-only WLS
plan still needs the full Gamma/inverse-weight materialization. The mixed rows
time raw-data mixed-stat construction, starts, and either full bounded or
full-threshold SNLLS fitting; ratios greater than one favor SNLLS.

```{r}
#| fig-width: 7
#| fig-height: 4
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  e2e_scaling <- wide_e2e[wide_e2e$design == "kreiberg_scaling" &
                            is.finite(wide_e2e$lazy_gain), ]
  ggplot(e2e_scaling, aes(q, lazy_gain, color = estimator,
                          linetype = Parameterization)) +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_line() +
    geom_point(size = 2) +
    labs(x = "Indicators per factor (q)",
         y = "Legacy / lazy raw-to-SNLLS time",
         color = "Estimator",
         linetype = "Parameterization") +
    theme_minimal(base_size = 12)
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

## Cost Context

```{r}
cost <- aggregate(cbind(stats_ms, cache_setup_ms, weight_setup_ms) ~
                    Data + Design + Parameterization + estimator + Method,
                  data = ok,
                  FUN = median_or_na,
                  na.action = na.pass)
names(cost) <- c("Data", "Design", "Parameterization", "Estimator", "Method",
                 "Stats ms", "Cache setup ms", "Weight setup ms")
for (j in 6:ncol(cost)) cost[[j]] <- fmt_ms(cost[[j]])
knitr::kable(cost, row.names = FALSE)
```

The original fit-only rows still exclude statistic construction; `stats_ms` is
shown because those rows start from legacy `OrdinalStats`, which estimates
polychorics, full NACOV, and legacy DWLS/WLS weights. The end-to-end rows above
are the construction-aware comparison. The `full_bounded` rows use the
materialized legacy path directly for DWLS/WLS; the cache-aware bounded and
SNLLS rows reuse the ordinal moment/Gamma cache. Mixed rows start from
materialized `MixedOrdinalStats`, which currently bundles NACOV and DWLS/WLS
weights rather than a lazy mixed Gamma cache.

## Caveats

This is a speed pilot, not a simulation study. It uses one synthetic
repeated-measure model family, all-ordinal complete data, delta/theta
parameterizations, a mixed continuous/ordinal delta slice, and NLopt L-BFGS. It
does not compare against lavaan timing, robust reporting, or alternate
optimizers. There is no full direct ULS comparator in this run, so ULS only
supports the cache-aware bounded/SNLLS splits. In theta rows, the
`threshold_profiled_bounded` label is the cache-aware bounded comparator and the
two SNLLS labels both use threshold-only Golub-Pereyra profiling. Mixed theta
SNLLS is intentionally absent. The end-to-end rows include raw category/stat
construction and starts, but still assume the parsed/lavaanified model has
already been built.

The default runner caps the Kreiberg-style scaling grid at `q <= 8` to keep
the preliminary run practical. Use `--max-q 12` for the fuller literature-like
range.

## Reproduce

```sh
Rscript experiments/11-ordinal-snlls-speed/run_experiment.R --smoke --preset fast
(cd experiments/11-ordinal-snlls-speed && quarto render report.qmd)
```

For less noisy timings, use `--preset opt`, increase `--reps`, and optionally
run `--max-q 12`.
