---
title: "Ordinal SNLLS speed pilot"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "speed.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

speed_path <- file.path(results_dir, "speed.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(speed_path)) {
  stop("Missing speed.csv. Run: Rscript run_experiment.R --smoke")
}

speed <- utils::read.csv(speed_path, check.names = FALSE,
                         stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.2f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.3g", x))
fmt_diff <- function(x) ifelse(is.na(x), "", sprintf("%.2e", x))
max_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else max(x)
}
median_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else stats::median(x)
}

speed$Method <- factor(speed$path, levels = c("bounded", "snlls"),
                       labels = c("Bounded", "SNLLS"))
speed$Design <- factor(
  speed$design,
  levels = c("kreiberg_scaling", "worked_example", "threshold_stress"),
  labels = c("Kreiberg scaling", "Worked example", "Threshold stress")
)
ok <- speed[speed$status == "ok", ]

wide <- reshape(
  ok[, c("design", "case_id", "q", "p", "categories", "threshold_balance",
         "threshold_mode", "estimator", "path", "fit_median_ms")],
  idvar = c("design", "case_id", "q", "p", "categories",
            "threshold_balance", "threshold_mode", "estimator"),
  timevar = "path",
  direction = "wide"
)
names(wide) <- sub("^fit_median_ms\\.", "", names(wide))
wide$speedup <- wide$bounded / wide$snlls
```

## Question

This pilot asks whether ordinal SNLLS shows the same basic speed pattern as
the continuous Kreiberg-style SNLLS examples once thresholds and ordinal
weights are in the objective.

The design deliberately includes three views:

- a repeated-measure four-factor family indexed by indicators per factor `q`;
- one worked repeated-measure example for table/prose use;
- an ordinal-specific threshold stress slice varying categories, balance, and
  fixed/shared threshold constraints.

## Short Answer

In this pilot, SNLLS is faster than bounded fitting in the repeated-measure
ordinal cells while matching the bounded objective to numerical tolerance.
The result is not yet a paper-grade simulation: statistic construction is still
the legacy full ordinal builder, and the design is synthetic. But it is the
right first sign that the Kreiberg-style scaling story survives ordinalization.

## Run

```{r}
data.frame(
  Item = metadata$key,
  Value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Status

```{r}
status <- aggregate(seed ~ Design + estimator + Method + status,
                    data = speed, FUN = length)
names(status) <- c("Design", "Estimator", "Method", "Status", "Rows")
knitr::kable(status, row.names = FALSE)
```

## Agreement

```{r}
agreement <- aggregate(
  cbind(theta_diff_bounded, fmin_diff_bounded) ~ Design + estimator + Method,
  data = ok,
  FUN = max_or_na,
  na.action = na.pass
)
names(agreement) <- c("Design", "Estimator", "Method",
                      "Max |theta - bounded|", "Max |fmin - bounded|")
for (j in 4:ncol(agreement)) agreement[[j]] <- fmt_diff(agreement[[j]])
knitr::kable(agreement, row.names = FALSE)
```

## Speedup

```{r}
summary <- aggregate(speedup ~ design + estimator, data = wide,
                     FUN = median_or_na)
summary$Design <- factor(
  summary$design,
  levels = c("kreiberg_scaling", "worked_example", "threshold_stress"),
  labels = c("Kreiberg scaling", "Worked example", "Threshold stress")
)
summary$`Median bounded / SNLLS` <- fmt_num(summary$speedup)
knitr::kable(summary[, c("Design", "estimator", "Median bounded / SNLLS")],
             row.names = FALSE,
             col.names = c("Design", "Estimator", "Median Bounded / SNLLS"))
```

```{r}
#| fig-width: 8
#| fig-height: 4
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  scaling <- wide[wide$design == "kreiberg_scaling", ]
  ggplot(scaling, aes(q, speedup, color = estimator)) +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_line() +
    geom_point(size = 2) +
    labs(x = "Indicators per factor (q)",
         y = "Bounded / SNLLS fit-time ratio",
         color = "Estimator") +
    theme_minimal(base_size = 12)
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

## Worked Example

```{r}
worked <- wide[wide$design == "worked_example", ]
worked_table <- worked[, c("case_id", "q", "p", "categories", "estimator",
                           "bounded", "snlls", "speedup")]
names(worked_table) <- c("Case", "q", "p", "Categories", "Estimator",
                         "Bounded ms", "SNLLS ms", "Bounded / SNLLS")
worked_table$`Bounded ms` <- fmt_ms(worked_table$`Bounded ms`)
worked_table$`SNLLS ms` <- fmt_ms(worked_table$`SNLLS ms`)
worked_table$`Bounded / SNLLS` <- fmt_num(worked_table$`Bounded / SNLLS`)
knitr::kable(worked_table, row.names = FALSE)
```

## Threshold Stress

```{r}
stress <- wide[wide$design == "threshold_stress", ]
stress_table <- stress[, c("categories", "threshold_balance",
                           "threshold_mode", "estimator", "speedup")]
names(stress_table) <- c("Categories", "Threshold balance",
                         "Threshold mode", "Estimator", "Bounded / SNLLS")
stress_table$`Bounded / SNLLS` <- fmt_num(stress_table$`Bounded / SNLLS`)
knitr::kable(stress_table[order(stress_table$Categories,
                                stress_table$`Threshold balance`,
                                stress_table$`Threshold mode`,
                                stress_table$Estimator), ],
             row.names = FALSE)
```

Threshold stress keeps the same repeated-measure covariance model but changes
the ordinal layer. This is where fixed and shared threshold rows can erode or
preserve the profiling advantage.

## Cost Context

```{r}
cost <- aggregate(cbind(stats_ms, cache_setup_ms, weight_setup_ms) ~
                    Design + estimator + Method,
                  data = ok,
                  FUN = median_or_na,
                  na.action = na.pass)
names(cost) <- c("Design", "Estimator", "Method", "Stats ms",
                 "Cache setup ms", "Weight setup ms")
for (j in 4:ncol(cost)) cost[[j]] <- fmt_ms(cost[[j]])
knitr::kable(cost, row.names = FALSE)
```

Fit timing excludes statistic construction, but `stats_ms` is shown because the
current pilot still starts from legacy `OrdinalStats`, which estimates
polychorics, full NACOV, and legacy DWLS/WLS weights. A raw/lazy ordinal
builder is the next experiment if we want end-to-end speed claims.

## Caveats

This is a speed pilot, not a simulation study. It uses one synthetic
repeated-measure model family, all-ordinal complete data, the delta
parameterization, and NLopt L-BFGS. It does not compare against lavaan timing,
mixed ordinal data, theta parameterization, robust reporting, or alternate
optimizers.

The default runner caps the Kreiberg-style scaling grid at `q <= 8` to keep
the preliminary run practical. Use `--max-q 12` for the fuller literature-like
range.

## Reproduce

```sh
Rscript experiments/11-ordinal-snlls-speed/run_experiment.R --smoke --preset fast
(cd experiments/11-ordinal-snlls-speed && quarto render report.qmd)
```

For less noisy timings, use `--preset opt`, increase `--reps`, and optionally
run `--max-q 12`.
