---
title: "Ordinal SNLLS speed pilot"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "speed.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

speed_path <- file.path(results_dir, "speed.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(speed_path)) {
  stop("Missing speed.csv. Run: Rscript run_experiment.R --smoke")
}

speed <- utils::read.csv(speed_path, check.names = FALSE,
                         stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.2f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.3g", x))
fmt_diff <- function(x) ifelse(is.na(x), "", sprintf("%.2e", x))
max_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else max(x)
}
median_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else stats::median(x)
}

method_levels <- c("full_bounded", "threshold_profiled_bounded",
                   "snlls_full_thresholds", "snlls")
method_labels <- c("Full bounded", "Threshold-profiled bounded",
                   "Full-threshold SNLLS", "Threshold-profiled SNLLS")
speed$Method <- factor(speed$path, levels = method_levels,
                       labels = method_labels)
speed$Design <- factor(
  speed$design,
  levels = c("kreiberg_scaling", "worked_example", "threshold_stress"),
  labels = c("Kreiberg scaling", "Worked example", "Threshold stress")
)
ok <- speed[speed$status == "ok", ]

wide <- reshape(
  ok[, c("design", "case_id", "q", "p", "categories", "threshold_balance",
         "threshold_mode", "estimator", "path", "fit_median_ms")],
  idvar = c("design", "case_id", "q", "p", "categories",
            "threshold_balance", "threshold_mode", "estimator"),
  timevar = "path",
  direction = "wide"
)
names(wide) <- sub("^fit_median_ms\\.", "", names(wide))
for (name in method_levels) {
  if (!name %in% names(wide)) wide[[name]] <- NA_real_
}
wide$profile_gain <- wide$full_bounded / wide$threshold_profiled_bounded
wide$full_snlls_gain <- wide$full_bounded / wide$snlls_full_thresholds
wide$snlls_incremental <- wide$threshold_profiled_bounded / wide$snlls
wide$threshold_snlls_gain <- wide$snlls_full_thresholds / wide$snlls
wide$total_gain <- wide$full_bounded / wide$snlls
```

## Question

This pilot asks whether ordinal SNLLS shows the same basic speed pattern as
the continuous Kreiberg-style SNLLS examples once thresholds and ordinal
weights are in the objective. It now separates four costs where available:
legacy full bounded fitting, threshold-profiled bounded fitting,
full-threshold SNLLS, and threshold-profiled SNLLS. The full direct bounded
comparator is included for DWLS/WLS; ULS has the three cache-aware paths.

The design deliberately includes three views:

- a repeated-measure four-factor family indexed by indicators per factor `q`;
- one worked repeated-measure example for table/prose use;
- an ordinal-specific threshold stress slice varying categories, balance, and
  fixed/shared threshold constraints.

## Short Answer

In this pilot, the full direct comparator is a useful diagnostic rather than a
guaranteed slower baseline: fit-only timings can favor the legacy full bounded
path after statistic construction is excluded. The full-threshold SNLLS row
profiles thresholds as ordinary Golub-Pereyra linear parameters, while the
threshold-profiled SNLLS row applies the ordinal threshold map first and then
profiles the remaining conditionally-linear covariance block. The fitted
solutions match the bounded objectives to numerical tolerance. The result is
not yet a paper-grade simulation: statistic construction is still the legacy
full ordinal builder, and the design is synthetic. But it is the right first
sign that the Kreiberg-style scaling story survives ordinalization.

## Run

```{r}
data.frame(
  Item = metadata$key,
  Value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Status

```{r}
status <- aggregate(seed ~ Design + estimator + Method + status,
                    data = speed, FUN = length)
names(status) <- c("Design", "Estimator", "Method", "Status", "Rows")
knitr::kable(status, row.names = FALSE)
```

## Agreement

```{r}
agreement <- aggregate(
  cbind(theta_diff_profiled_bounded, fmin_diff_profiled_bounded,
        theta_diff_full_bounded, fmin_diff_full_bounded) ~
    Design + estimator + Method,
  data = ok,
  FUN = max_or_na,
  na.action = na.pass
)
names(agreement) <- c("Design", "Estimator", "Method",
                      "Max |theta - profiled|", "Max |fmin - profiled|",
                      "Max |theta - full|", "Max |fmin - full|")
for (j in 4:ncol(agreement)) agreement[[j]] <- fmt_diff(agreement[[j]])
knitr::kable(agreement, row.names = FALSE)
```

## Speedup

```{r}
summary <- aggregate(cbind(profile_gain, full_snlls_gain, snlls_incremental,
                           threshold_snlls_gain, total_gain) ~
                       design + estimator,
                     data = wide,
                     FUN = median_or_na,
                     na.action = na.pass)
summary$Design <- factor(
  summary$design,
  levels = c("kreiberg_scaling", "worked_example", "threshold_stress"),
  labels = c("Kreiberg scaling", "Worked example", "Threshold stress")
)
summary$`Full / profiled` <- fmt_num(summary$profile_gain)
summary$`Full / full-threshold SNLLS` <- fmt_num(summary$full_snlls_gain)
summary$`Profiled / profiled SNLLS` <- fmt_num(summary$snlls_incremental)
summary$`Full-threshold / profiled SNLLS` <-
  fmt_num(summary$threshold_snlls_gain)
summary$`Full / profiled SNLLS` <- fmt_num(summary$total_gain)
knitr::kable(summary[, c("Design", "estimator", "Full / profiled",
                         "Full / full-threshold SNLLS",
                         "Profiled / profiled SNLLS",
                         "Full-threshold / profiled SNLLS",
                         "Full / profiled SNLLS")],
             row.names = FALSE,
             col.names = c("Design", "Estimator", "Full / Profiled",
                           "Full / Full-Threshold SNLLS",
                           "Profiled / Profiled SNLLS",
                           "Full-Threshold / Profiled SNLLS",
                           "Full / Profiled SNLLS"))
```

Ratios greater than one favor the denominator. `Full / Profiled` can be below
one because the pilot excludes statistic construction and compares different
bounded objective implementations, not just the presence or absence of
threshold profiling.

```{r}
#| fig-width: 8
#| fig-height: 5
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  scaling <- wide[wide$design == "kreiberg_scaling", ]
  plot_data <- rbind(
    transform(scaling[is.finite(scaling$profile_gain), ],
              Ratio = profile_gain, Comparison = "Full / profiled"),
    transform(scaling[is.finite(scaling$full_snlls_gain), ],
              Ratio = full_snlls_gain,
              Comparison = "Full / full-threshold SNLLS"),
    transform(scaling[is.finite(scaling$snlls_incremental), ],
              Ratio = snlls_incremental,
              Comparison = "Profiled / profiled SNLLS"),
    transform(scaling[is.finite(scaling$threshold_snlls_gain), ],
              Ratio = threshold_snlls_gain,
              Comparison = "Full-threshold / profiled SNLLS"),
    transform(scaling[is.finite(scaling$total_gain), ],
              Ratio = total_gain, Comparison = "Full / profiled SNLLS")
  )
  ggplot(plot_data, aes(q, Ratio, color = estimator)) +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_line() +
    geom_point(size = 2) +
    facet_wrap(~ Comparison, scales = "free_y") +
    labs(x = "Indicators per factor (q)",
         y = "Fit-time ratio",
         color = "Estimator") +
    theme_minimal(base_size = 12)
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

## Worked Example

```{r}
worked <- wide[wide$design == "worked_example", ]
worked_table <- worked[, c("case_id", "q", "p", "categories", "estimator",
                           "full_bounded", "threshold_profiled_bounded",
                           "snlls_full_thresholds", "snlls", "profile_gain",
                           "full_snlls_gain", "snlls_incremental",
                           "threshold_snlls_gain", "total_gain")]
names(worked_table) <- c("Case", "q", "p", "Categories", "Estimator",
                         "Full bounded ms", "Profiled bounded ms",
                         "Full-threshold SNLLS ms",
                         "Profiled SNLLS ms", "Full / profiled",
                         "Full / full-threshold SNLLS",
                         "Profiled / profiled SNLLS",
                         "Full-threshold / profiled SNLLS",
                         "Full / profiled SNLLS")
worked_table$`Full bounded ms` <- fmt_ms(worked_table$`Full bounded ms`)
worked_table$`Profiled bounded ms` <-
  fmt_ms(worked_table$`Profiled bounded ms`)
worked_table$`Full-threshold SNLLS ms` <-
  fmt_ms(worked_table$`Full-threshold SNLLS ms`)
worked_table$`Profiled SNLLS ms` <- fmt_ms(worked_table$`Profiled SNLLS ms`)
worked_table$`Full / profiled` <- fmt_num(worked_table$`Full / profiled`)
worked_table$`Full / full-threshold SNLLS` <-
  fmt_num(worked_table$`Full / full-threshold SNLLS`)
worked_table$`Profiled / profiled SNLLS` <-
  fmt_num(worked_table$`Profiled / profiled SNLLS`)
worked_table$`Full-threshold / profiled SNLLS` <-
  fmt_num(worked_table$`Full-threshold / profiled SNLLS`)
worked_table$`Full / profiled SNLLS` <-
  fmt_num(worked_table$`Full / profiled SNLLS`)
knitr::kable(worked_table, row.names = FALSE)
```

## Threshold Stress

```{r}
stress <- wide[wide$design == "threshold_stress", ]
stress_table <- stress[, c("categories", "threshold_balance",
                           "threshold_mode", "estimator", "profile_gain",
                           "full_snlls_gain", "snlls_incremental",
                           "threshold_snlls_gain", "total_gain")]
names(stress_table) <- c("Categories", "Threshold balance",
                         "Threshold mode", "Estimator", "Full / profiled",
                         "Full / full-threshold SNLLS",
                         "Profiled / profiled SNLLS",
                         "Full-threshold / profiled SNLLS",
                         "Full / profiled SNLLS")
stress_table$`Full / profiled` <- fmt_num(stress_table$`Full / profiled`)
stress_table$`Full / full-threshold SNLLS` <-
  fmt_num(stress_table$`Full / full-threshold SNLLS`)
stress_table$`Profiled / profiled SNLLS` <-
  fmt_num(stress_table$`Profiled / profiled SNLLS`)
stress_table$`Full-threshold / profiled SNLLS` <-
  fmt_num(stress_table$`Full-threshold / profiled SNLLS`)
stress_table$`Full / profiled SNLLS` <-
  fmt_num(stress_table$`Full / profiled SNLLS`)
knitr::kable(stress_table[order(stress_table$Categories,
                                stress_table$`Threshold balance`,
                                stress_table$`Threshold mode`,
                                stress_table$Estimator), ],
             row.names = FALSE)
```

Threshold stress keeps the same repeated-measure covariance model but changes
the ordinal layer. This is where fixed and shared threshold rows can erode or
preserve the profiling advantage.

## Cost Context

```{r}
cost <- aggregate(cbind(stats_ms, cache_setup_ms, weight_setup_ms) ~
                    Design + estimator + Method,
                  data = ok,
                  FUN = median_or_na,
                  na.action = na.pass)
names(cost) <- c("Design", "Estimator", "Method", "Stats ms",
                 "Cache setup ms", "Weight setup ms")
for (j in 4:ncol(cost)) cost[[j]] <- fmt_ms(cost[[j]])
knitr::kable(cost, row.names = FALSE)
```

Fit timing excludes statistic construction, but `stats_ms` is shown because the
current pilot still starts from legacy `OrdinalStats`, which estimates
polychorics, full NACOV, and legacy DWLS/WLS weights. A raw/lazy ordinal
builder is the next experiment if we want end-to-end speed claims. The
`full_bounded` rows use that materialized legacy path directly for DWLS/WLS;
the cache-aware bounded and SNLLS rows reuse the ordinal moment/Gamma cache.

## Caveats

This is a speed pilot, not a simulation study. It uses one synthetic
repeated-measure model family, all-ordinal complete data, the delta
parameterization, and NLopt L-BFGS. It does not compare against lavaan timing,
mixed ordinal data, theta parameterization, robust reporting, or alternate
optimizers. There is no full direct ULS comparator in this run, so ULS only
supports the cache-aware bounded/SNLLS splits.

The default runner caps the Kreiberg-style scaling grid at `q <= 8` to keep
the preliminary run practical. Use `--max-q 12` for the fuller literature-like
range.

## Reproduce

```sh
Rscript experiments/11-ordinal-snlls-speed/run_experiment.R --smoke --preset fast
(cd experiments/11-ordinal-snlls-speed && quarto render report.qmd)
```

For less noisy timings, use `--preset opt`, increase `--reps`, and optionally
run `--max-q 12`.
