---
title: "Latent Metric Identification — Pipeline Timing"
format: html
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(
  getwd(),
  dirname(input_path),
  dirname(getwd())
))
result_candidates <- unique(c(
  file.path(candidate_dirs, "results"),
  candidate_dirs
))
hits <- result_candidates[file.exists(file.path(result_candidates, "fits.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
fits_path <- file.path(results_dir, "fits.csv")
pairs_path <- file.path(results_dir, "pairs.csv")

if (!file.exists(fits_path) || !file.exists(pairs_path)) {
  stop("Missing experiment CSVs. Run: Rscript run_experiment.R")
}

fits <- utils::read.csv(fits_path, check.names = FALSE)
pairs <- utils::read.csv(pairs_path, check.names = FALSE)

ok <- pairs[pairs$status == "OK" &
              is.finite(pairs$elapsed_ratio_std_lv_over_marker), ]
valid <- ok[ok$same_implied_moments %in% TRUE, ]

skip_rows <- fits[fits$status == "SKIP" &
                    grepl("scale-equivalent", fits$error, fixed = TRUE), ]
n_skip_npar <- length(unique(skip_rows$case_id))
n_err <- sum(fits$status == "ERROR", na.rm = TRUE)
n_mismatch <- sum(pairs$material_mismatch, na.rm = TRUE)

n_cases <- length(unique(fits$case_id))
n_backends <- length(unique(pairs$backend))
n_reps <- if (nrow(pairs)) max(pairs$replicate, na.rm = TRUE) else NA_integer_

fmt_pct <- function(x) sprintf("%.1f%%", 100 * x)
fmt_ratio <- function(x) formatC(x, format = "f", digits = 3)
fmt_us <- function(x) sprintf("%.0f μs", x * 1e6)
```

The question this experiment answers: **if a user specifies a model in the
familiar marker form, would magmaan win by internally converting to `std_lv`,
fitting there, then back-converting to marker for the user?** That trades a
(usually faster) fit against the cost of a spec rebuild and a parameter
rescaling pass.

## Headline

Pairs are kept only if both fits converged AND landed at the same implied
moments (Δσ < 1e-5). Pairs where the two parameterizations disagree on the
model itself are not comparable on time and are excluded from these summaries.

```{r}
#| results: asis
if (nrow(valid)) {
  fr <- valid$elapsed_ratio_std_lv_over_marker
  pr <- valid$pipeline_ratio_std_lv_over_marker
  fqs <- stats::quantile(fr, c(0.25, 0.5, 0.75), na.rm = TRUE)
  pqs <- stats::quantile(pr, c(0.25, 0.5, 0.75), na.rm = TRUE)
  cat(sprintf(
    paste0(
      "Across **%d cases**, **%d backends**, %d reps — **%d valid ",
      "pairs** out of %d OK pairs.\n\n",
      "- **Fit only** (current question): `std_lv / marker` = **%s** ",
      "(IQR %s–%s); `std_lv` faster on **%s** of pairs.\n",
      "- **Full pipeline** (spec build + fit + backconvert vs spec build + ",
      "fit): ratio = **%s** (IQR %s–%s); `std_lv`-route faster on **%s** of pairs.\n"
    ),
    n_cases, n_backends, n_reps, nrow(valid), nrow(ok),
    fmt_ratio(fqs[[2L]]), fmt_ratio(fqs[[1L]]), fmt_ratio(fqs[[3L]]),
    fmt_pct(mean(fr < 1, na.rm = TRUE)),
    fmt_ratio(pqs[[2L]]), fmt_ratio(pqs[[1L]]), fmt_ratio(pqs[[3L]]),
    fmt_pct(mean(pr < 1, na.rm = TRUE))
  ))
} else {
  cat("No valid pairs to summarize.\n")
}
```

## Per-backend summary

```{r}
if (nrow(ok)) {
  by_be <- split(ok, ok$backend)
  rows <- lapply(by_be, function(x) {
    vx <- x[x$same_implied_moments %in% TRUE, ]
    fr <- vx$elapsed_ratio_std_lv_over_marker
    pr <- vx$pipeline_ratio_std_lv_over_marker
    data.frame(
      backend = x$backend[[1L]],
      ok_pairs = nrow(x),
      valid_pairs = nrow(vx),
      `fit ratio` = if (nrow(vx)) fmt_ratio(stats::median(fr, na.rm = TRUE)) else "—",
      `pipeline ratio` = if (nrow(vx)) fmt_ratio(stats::median(pr, na.rm = TRUE)) else "—",
      `pipeline faster` = if (nrow(vx)) fmt_pct(mean(pr < 1, na.rm = TRUE)) else "—",
      `same implied moments` = fmt_pct(mean(x$same_implied_moments, na.rm = TRUE)),
      check.names = FALSE,
      stringsAsFactors = FALSE
    )
  })
  knitr::kable(do.call(rbind, rows), row.names = FALSE)
}
```

## Per-step breakdown

Median wall time per pipeline step across valid pairs. Spec build is one call
to `model_spec()` (lavaanify + partable). Fit is the `fit_ml` call. Backconvert
is a vectorized R pass that rescales loadings, latent variances, latent
covariances, and latent regressions.

```{r}
if (nrow(valid)) {
  rows <- data.frame(
    step = c("spec build", "fit", "backconvert"),
    marker = c(fmt_us(stats::median(valid$marker_spec_sec, na.rm = TRUE)),
               fmt_us(stats::median(valid$marker_elapsed_sec, na.rm = TRUE)),
               "—"),
    std_lv = c(fmt_us(stats::median(valid$std_lv_spec_sec, na.rm = TRUE)),
               fmt_us(stats::median(valid$std_lv_elapsed_sec, na.rm = TRUE)),
               fmt_us(stats::median(valid$std_lv_backconvert_sec, na.rm = TRUE))),
    stringsAsFactors = FALSE
  )
  knitr::kable(rows, row.names = FALSE,
               col.names = c("step", "marker", "std_lv"))
}
```

## Pipeline ratio distribution

`log10(pipeline_std_lv_route / pipeline_marker_route)`; below zero means the
`std_lv`-with-backconvert route wins. Dashed line at zero.

```{r}
#| fig-width: 7
#| fig-height: 4
if (nrow(valid)) {
  valid$log_ratio <- log10(valid$pipeline_ratio_std_lv_over_marker)
  op <- par(mar = c(4, 6, 1, 1))
  boxplot(log_ratio ~ backend, data = valid, horizontal = TRUE,
          xlab = "log10(pipeline_ratio std_lv / marker)",
          ylab = "", las = 1, outline = FALSE, col = "grey90")
  stripchart(log_ratio ~ backend, data = valid, add = TRUE, vertical = FALSE,
             method = "jitter", jitter = 0.15, pch = 16,
             col = grDevices::adjustcolor("steelblue", alpha.f = 0.3))
  abline(v = 0, lty = 2, col = "gray40")
  par(op)
} else {
  plot.new(); text(0.5, 0.5, "no data")
}
```

## Pipeline ratio vs marker fit cost

Each point is a `(case, backend, rep)` pair. The horizontal axis is the
marker-side fit cost (where dominant cost lives); the vertical axis is the
pipeline ratio. As fit time grows, the fixed overhead (spec + backconvert)
becomes a smaller fraction and the ratio should approach the fit-only ratio.

```{r}
#| fig-width: 7
#| fig-height: 4
if (nrow(valid)) {
  op <- par(mar = c(4, 4, 1, 1))
  plot(valid$marker_elapsed_sec * 1e3,
       valid$pipeline_ratio_std_lv_over_marker,
       log = "xy",
       xlab = "marker fit time (ms)",
       ylab = "pipeline ratio std_lv / marker",
       pch = 16,
       col = grDevices::adjustcolor(
         ifelse(valid$backend == "port", "darkorange", "steelblue"),
         alpha.f = 0.4))
  abline(h = 1, lty = 2, col = "gray40")
  legend("topright", legend = sort(unique(valid$backend)),
         pch = 16, col = c("steelblue", "darkorange"), bty = "n")
  par(op)
}
```

## Caveats

```{r}
#| results: asis
cat(sprintf(
  paste0(
    "- **%d** case(s) skipped because `marker` and `std_lv` did not produce ",
    "the same number of free parameters (multi-factor / invariance models ",
    "whose loading-vs-variance fix isn't scale-equivalent).\n",
    "- **%d** pair(s) excluded from the timing summary because the two ",
    "parameterizations did not land at the same implied moments ",
    "(Δsigma > 1e-5 or Δfmin > 1e-5) — these are the cases where `std_lv` ",
    "is **not admissible** as an internal substitute; see `results/pairs.csv`.\n",
    "- **%d** fit error(s); see `results/fits.csv`.\n",
    "- The backconvert step is implemented in vectorized R for this study. ",
    "A C++ implementation would be substantially cheaper still, shifting the ",
    "pipeline ratio toward the fit-only ratio for small models.\n"
  ),
  n_skip_npar, n_mismatch, n_err
))
```
