\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}

\title{Analytic Saturated Information for FIML FMG}
\author{}
\date{\today}

\newcommand{\vech}{\operatorname{vech}}
\newcommand{\tr}{\operatorname{tr}}

\begin{document}
\maketitle

\section*{Purpose}

These notes derive the analytic observed-data saturated information used by
the FIML FMG machinery. The target is only the H1/saturated eta-space
information. The H0 model-parameter observed information can remain on its
existing finite-difference path.

For one independent block, the saturated parameter is
\[
  \eta = \begin{bmatrix}\mu \\ \vech(\Sigma)\end{bmatrix},
\]
where \(\vech\) stores the lower triangle once, in column-major order. A row
with observed index set \(o\) contributes the observed-subvector normal
deviance
\[
  f_o(\mu,\Sigma)
    = \log |\Sigma_{oo}|
      + (x_o-\mu_o)' \Sigma_{oo}^{-1} (x_o-\mu_o),
\]
up to constants independent of \(\eta\). Let
\[
  A = \Sigma_{oo}^{-1}, \qquad
  u = x_o-\mu_o, \qquad
  z = A u.
\]

\section*{Scores in the implementation convention}

The code stores row scores as gradients of the deviance \(f_o=-2\ell_o+c\).
The log-likelihood score is therefore \(-\frac12\) times this vector.

For the observed mean coordinates,
\[
  g_\mu = \frac{\partial f_o}{\partial \mu_o} = -2z.
\]

For a full covariance vech coordinate \(k\), let \(D_k\) denote the symmetric
matrix in observed-subvector coordinates induced by increasing that full
\(\vech(\Sigma)\) entry by one. Diagonal coordinates have one nonzero entry;
off-diagonal coordinates have the two mirrored nonzero entries. The deviance
score is
\[
  g_k
    = \frac{\partial f_o}{\partial \sigma_k}
    = \tr(A D_k) - z'D_kz.
\]
This is exactly the existing saturated-score code: off-diagonal covariance
scores receive the duplication factor because \(D_k\) contains both mirrored
entries.

\section*{Analytic row Hessian}

The row Hessian of the deviance score has three blocks.

Mean--mean:
\[
  H_{\mu\mu} = 2A.
\]

Mean--covariance:
\[
  H_{\mu k}
    = \frac{\partial g_\mu}{\partial \sigma_k}
    = 2 A D_k z.
\]
By symmetry, \(H_{k\mu}=H_{\mu k}'\).

Covariance--covariance:
\[
  H_{kl}
    = \frac{\partial g_k}{\partial \sigma_l}
    = -\tr(A D_l A D_k)
      + z'D_l A D_k z
      + z'D_k A D_l z.
\]
Equivalently, using \(D_k z\),
\[
  H_{kl}
    = -\tr(A D_l A D_k)
      + (D_lz)' A D_k z
      + (D_kz)' A D_l z.
\]
This form is convenient because \(D_k\) is sparse and \(A D_k z\) can be built
from one or two columns of \(A\).

\section*{Block aggregation and scaling}

For block \(b\) with \(n_b\) rows, the implementation averages the row
deviance Hessians first:
\[
  \bar H_{\mathrm{dev},b}
    = \frac{1}{n_b}\sum_{i=1}^{n_b}
      \frac{\partial^2 f_i}{\partial \eta_b\partial\eta_b'}.
\]
The saturated information reported by \texttt{saturated\_em\_moments()} is in
log-likelihood units, summed over rows:
\[
  H_b = -\frac{\partial^2 \ell_b}{\partial\eta_b\partial\eta_b'}
      = \frac{n_b}{2}\,\bar H_{\mathrm{dev},b}.
\]

The existing \(J\) construction remains unchanged. If \(G_{\mathrm{dev}}\) is
the matrix whose rows are deviance scores, then
\[
  J_b
    = \sum_i
      \frac{\partial\ell_i}{\partial\eta_b}
      \frac{\partial\ell_i}{\partial\eta_b'}
    = \frac14 G_{\mathrm{dev}}'G_{\mathrm{dev}}.
\]

The missing-data saturated moment covariance used by FIML FMG is then
\[
  \Gamma_{\mathrm{mis}} = H^{-1} J H^{-1},
\]
with \(H\) and \(J\) block-diagonal across independent groups/blocks.

\section*{Implementation implication}

The finite-difference step \(h\) is not part of the saturated H1 FMG
definition. It was only a numerical route to the same row Hessian. After this
analytic replacement, \texttt{h\_step} can remain on public calls for source
compatibility and for the still-numeric H0 parameter-information path, but it
does not tune saturated H1 information.

\end{document}
