\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}

\title{Asymptotics for Guttman's Multiple-Group CFA Estimator}
\author{}
\date{\today}

\newcommand{\T}{^{\mathsf T}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\vech}{\operatorname{vech}}
\newcommand{\vecop}{\operatorname{vec}}
\newcommand{\IF}{\operatorname{IF}}
\newcommand{\argmin}{\operatorname*{arg\,min}}

\theoremstyle{plain}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\begin{document}
\maketitle

\section*{Purpose}

Dhaene and Rosseel \cite{dhaene2024} report that Guttman's multiple-group
method \cite{guttman1952} is a surprisingly strong non-iterative CFA estimator
in their simulations. This note records the asymptotic objects needed to
evaluate that claim analytically against normal-theory ML and the
moment-quadratic estimators already implemented in \texttt{magmaan}.

The phrase ``multiple-group method'' below is Guttman's historical name for
forming factor score groups from indicators. It is not multi-group SEM. Actual
multi-group SEM enters only by stacking independent covariance blocks.

\section{Sample Moment Setup}

Let \(y_i\in\mathbb{R}^p\) have mean \(\mu_0\), covariance \(\Sigma_0\), and
finite fourth moments. Write
\[
  \hat\sigma=\vech(\hat\Sigma),\qquad
  \sigma_0=\vech(\Sigma_0),
\]
with the usual covariance influence rows
\[
  z_i=\vech\{(y_i-\mu_0)(y_i-\mu_0)\T-\Sigma_0\}.
\]
Then
\[
  \sqrt n(\hat\sigma-\sigma_0)
    = n^{-1/2}\sum_{i=1}^n z_i+o_p(1),
  \qquad
  \Gamma=\Var(z_i).
\]
Under multivariate normality, \(\Gamma\) reduces to the normal-theory
covariance of \(\vech(\hat\Sigma)\). Under misspecification or non-normality,
\(\Gamma\) is the actual fourth-moment covariance.

For independent sample groups \(b=1,\ldots,B\), stack
\(\hat\sigma=(\hat\sigma_1\T,\ldots,\hat\sigma_B\T)\T\). With
\(n_b/N\to\pi_b\),
\[
  \sqrt N(\hat\sigma-\sigma_0)
    \Rightarrow
  N\{0,\operatorname{blockdiag}(\Gamma_b/\pi_b)\}.
\]
All formulas below apply to this stacked vector after replacing \(\Gamma\) by
the block diagonal covariance.

\section{The Guttman Map}

Fix a simple CFA loading pattern. Let \(X\in\{0,1\}^{p\times k}\) be the
indicator-by-factor membership matrix, with one nonzero entry per row in the
clean simple-structure case. Let \(S(\sigma)\) be the symmetric matrix whose
lower triangle is \(\sigma\).

Guttman's method first replaces the diagonal of \(S\) by communalities. Write
\[
  h(\sigma)\in\mathbb{R}^p,\qquad
  C(\sigma)=S(\sigma)-\diag\{\diag S(\sigma)\}+\diag\{h(\sigma)\}.
\]
For the Spearman-Harman rule used by lavaan and magmaan starts, the unclamped
correlation-scale communality for indicator \(i\) in its factor block is
\[
  h_i^{\mathrm{cor}}(R)
    =
  {|P_i|}^{-1}\sum_{(j,\ell)\in P_i}
    \frac{R_{ij}R_{i\ell}}{R_{j\ell}},
  \qquad
  h_i(\Sigma)=\Sigma_{ii}h_i^{\mathrm{cor}}\{R(\Sigma)\},
\]
where \(P_i=\{(j,\ell):j<\ell,\ j\ne i,\ \ell\ne i\}\). The production start
routine clamps this value and falls back in singular cases; the smooth theory
below assumes an interior point where no clamp or fallback is active. A
different communality rule only changes \(Dh(\sigma_0)\).

Define
\[
  A=CX,\qquad
  B=X\T C X,\qquad
  D=\diag\{B\},\qquad
  Q=D^{-1/2}.
\]
Then
\[
  L=AQ,\qquad
  P=QBQ,\qquad
  K^\star=LP^{-1}.
\]
If \(m_f\) is the marker indicator for factor \(f\), set
\[
  M=\diag(K^\star_{m_1 1},\ldots,K^\star_{m_k k}).
\]
The marker-scaled Guttman estimates are
\[
  \Lambda_G=K^\star M^{-1},
  \qquad
  \Phi_G=MPM,
  \qquad
  \psi_G=\diag\{S-\Lambda_G\Phi_G\Lambda_G\T\}.
\]
The reported parameter map is a fixed selection
\[
  \tau_G(\sigma)
    =
  \bigl[
    \operatorname{sel}_{\Lambda}(\Lambda_G)\T,\ 
    \vech(\Phi_G)\T,\ 
    \psi_G\T
  \bigr]\T .
\]
This is the version closest to the current \texttt{guttman\_start\_values()}
code: the full \(K^\star\) is computed, but only the model's allowed loading
cells are read out. Dhaene and Rosseel mention a constrained quadratic program
for enforcing simple structure. If that projection is used instead, the same
delta method applies after differentiating the fixed-active-set KKT system.

\section{Guttman Influence Function}

Let \(dS\) be a symmetric covariance perturbation and write all quantities at
\(\sigma_0\). Let
\[
  dC=dS-\diag\{\diag(dS)\}+\diag\{Dh[\sigma_0]\,d\sigma\}.
\]
The differential chain is
\[
\begin{aligned}
  dA &= dC\,X,\\
  dB &= X\T dC\,X,\\
  dD &= \diag\{dB\},\\
  dQ &= -\frac12 D^{-3/2}dD,\\
  dL &= dA\,Q+A\,dQ,\\
  dP &= dQ\,B\,Q+Q\,dB\,Q+Q\,B\,dQ,\\
  dK^\star &= dL\,P^{-1}-L\,P^{-1}dP\,P^{-1}.
\end{aligned}
\]
Next
\[
  dM=\diag(dK^\star_{m_1 1},\ldots,dK^\star_{m_k k}),
\]
and
\[
\begin{aligned}
  d\Lambda_G
    &= dK^\star M^{-1}-K^\star M^{-1}dM M^{-1},\\
  d\Phi_G
    &= dM\,P\,M+M\,dP\,M+M\,P\,dM,\\
  d\psi_G
    &=
  \diag\{dS
    -d\Lambda_G\Phi_G\Lambda_G\T
    -\Lambda_G d\Phi_G\Lambda_G\T
    -\Lambda_G\Phi_G d\Lambda_G\T\}.
\end{aligned}
\]

\begin{proposition}[Guttman delta method]\label{prop:guttman}
Assume \(\Gamma\) is finite, \(h\) is differentiable at \(\sigma_0\), the
diagonal of \(B\) is positive, \(P\) is nonsingular, and all marker entries in
\(M\) are nonzero. Then
\[
  \sqrt n\{\tau_G(\hat\sigma)-\tau_G(\sigma_0)\}
    \Rightarrow
  N(0,\Omega_G),
  \qquad
  \Omega_G=J_G\Gamma J_G\T,
\]
where \(J_G=D\tau_G[\sigma_0]\) is the linear map defined by the differential
chain above. The observation-level influence function is
\[
  \IF_G(y_i)=J_G\,\vech\{(y_i-\mu_0)(y_i-\mu_0)\T-\Sigma_0\}.
\]
\end{proposition}

\begin{remark}
The current magmaan implementation already has most of the numerical map in
\texttt{src/estimate/start\_guttman.cpp}. Turning it into an estimator is not
hard. The open design choice is whether the estimator should be the clean
smooth map above, the lavaan-style clamped start map, or the constrained
projection variant used by Dhaene and Rosseel. For asymptotic comparisons, the
clean map is the right first target.
\end{remark}

\section{Guttman Omega and Alpha}

Let \(w\in\mathbb{R}^p\) be fixed score weights. For a diagonal-residual CFA
model, coefficient omega for the composite \(w\T y\) is
\[
  \omega(\Lambda,\Phi,\psi;w)
    =
  \frac{w\T\Lambda\Phi\Lambda\T w}{w\T S w},
\]
where \(S\) is the covariance matrix whose reliability is being summarized
\cite{mcdonald1999}. Applying this formula to the Guttman map gives
\[
  \omega_G(\sigma;w)
    =
  \frac{w\T\Lambda_G\Phi_G\Lambda_G\T w}{w\T S(\sigma) w}.
\]
This looks parameter-based, but the parameters cancel. Since
\[
  \Lambda_G\Phi_G\Lambda_G\T
    =
  K^\star P (K^\star)\T
    =
  L P^{-1}L\T
    =
  C X(X\T C X)^{-1}X\T C,
\]
we have the covariance-only form
\begin{equation}\label{eq:guttman-omega-general}
  \omega_G(\sigma;w)
    =
  \frac{
    w\T C X(X\T C X)^{-1}X\T C w
  }{
    w\T S(\sigma)w
  }.
\end{equation}
Thus, once the grouping matrix \(X\) and communality rule \(h(\sigma)\) are
fixed, Guttman omega is a population covariance functional. It does not require
iterative loadings, fitted residual variances, or marker choices. It is not
assumption-free: \(X\), \(h\), and the nonsingularity of \(X\T C X\) are part of
the definition.

For one factor, \(X=\mathbf{1}\). Put
\[
  a=C\mathbf{1},\qquad b=\mathbf{1}\T C\mathbf{1},\qquad
  T=\mathbf{1}\T S\mathbf{1}.
\]
Then
\[
  \omega_G(w)=\frac{(w\T a)^2}{b\,w\T S w}.
\]
For the unit-weight total score \(w=\mathbf{1}\), this collapses further:
\begin{equation}\label{eq:guttman-omega-one-factor}
  \omega_G
    =
  \frac{\mathbf{1}\T C\mathbf{1}}{\mathbf{1}\T S\mathbf{1}}
    =
  \frac{\sum_{i\ne j}S_{ij}+\sum_i h_i}{T}
    =
  1-\frac{\sum_i(S_{ii}-h_i)}{T}.
\end{equation}
Here \(h_i\) is the communality placed on item \(i\)'s diagonal in \(C\). For
the Spearman-Harman/Guttman rule above, in the one-factor case,
\[
  h_i
    =
  {\binom{p-1}{2}}^{-1}
  \left\{
    \sum_{1\le j<\ell\le p}
      \frac{S_{ij}S_{i\ell}}{S_{j\ell}}
    -(p-1)S_{ii}
  \right\}.
\]
This is the covariance-scale form of the correlation expression
\(S_{ii}R_{ij}R_{i\ell}/R_{j\ell}\): the item variance factors cancel, and the
unrestricted pair sum contains \(p-1\) pairs involving item \(i\), each reducing
to \(S_{ii}\). In a multi-factor simple-structure block, replace \(p\) by the
block size \(p_f\) and sum only within the block. Under an exact congeneric
one-factor population, this population rule recovers \(h_i=\lambda_i^2\phi\).
So the one-factor Guttman-derived omega is especially simple: replace each
item variance by its Guttman communality, sum all entries, and divide by the
observed total-score variance.

Cronbach's alpha for the same unit-weight score is
\cite{cronbach1951}
\begin{equation}\label{eq:alpha}
  \alpha
    =
  \frac{p}{p-1}
  \left(
    1-\frac{\tr(S)}{\mathbf{1}\T S\mathbf{1}}
  \right)
    =
  \frac{p}{p-1}\,
  \frac{\sum_{i\ne j}S_{ij}}{T}.
\end{equation}
Equations~\eqref{eq:guttman-omega-one-factor} and \eqref{eq:alpha} show the
contrast. Alpha extrapolates reliability from the off-diagonal covariance
average. Guttman omega uses the same off-diagonal covariance sum, but adds an
explicit diagonal common-variance estimate \(\sum_i h_i\).

Under an exact one-factor congeneric population,
\[
  \Sigma=\lambda\lambda\T\phi+\diag(\psi),
\]
and under a communality rule that recovers \(h_i=\lambda_i^2\phi\),
\[
  \omega_G
    =
  \frac{\phi(\mathbf{1}\T\lambda)^2}
       {\phi(\mathbf{1}\T\lambda)^2+\mathbf{1}\T\psi},
\]
which is the usual omega total for the unit-weight sum score. In the same
case,
\[
  \omega_G-\alpha
    =
  \frac{
    \phi\{p\sum_i\lambda_i^2-(\sum_i\lambda_i)^2\}
  }{
    (p-1)T
  }
  \ge 0,
\]
with equality exactly when the keyed loadings are equal. Thus alpha and
Guttman omega agree in the tau-equivalent case; with congeneric loadings,
alpha is lower because it has no direct diagonal common-variance estimate.

Away from the exact one-factor model, both quantities remain smooth covariance
functionals, but they target different summaries. Alpha needs no loading
pattern and is easiest to compute. Guttman omega is still non-iterative, but it
imports the CFA grouping pattern and the chosen communality estimator. Its
large-sample variance follows immediately by the delta method:
\[
  \sqrt n\{\hat\omega_G-\omega_G\}
    \Rightarrow
  N\{0,\nabla\omega_G(\sigma_0)\T\Gamma\nabla\omega_G(\sigma_0)\},
\]
and alpha has the same form with \(\nabla\alpha(\sigma_0)\).

\section{Moment-Quadratic Estimators}

Let \(m(\theta)\) be the model-implied covariance moment vector and
\[
  r(\theta,\sigma)=m(\theta)-\sigma.
\]
A fixed-weight ULS/GLS/WLS estimator minimizes
\[
  q_W(\theta,\sigma)=\frac12 r(\theta,\sigma)\T W r(\theta,\sigma),
\]
with pseudo-true value \(\theta_W\). Let
\[
  \Delta=\frac{\partial m(\theta_W)}{\partial\theta\T},
  \qquad
  H_W
    =
  \Delta\T W\Delta
    +\sum_{\ell}\{Wr(\theta_W,\sigma_0)\}_{\ell}
       \frac{\partial^2 m_\ell(\theta_W)}
            {\partial\theta\,\partial\theta\T}.
\]
The second term is the misspecification curvature term. It vanishes at exact
fit and is easy to miss if one silently substitutes Fisher or Gauss-Newton
information.

For fixed \(W\),
\[
  \IF_W(y_i)=
  H_W^{-1}\Delta\T W z_i,
  \qquad
  \Omega_W
    =
  H_W^{-1}\Delta\T W\Gamma W\Delta H_W^{-1}.
\]
If \(W=W(\hat\sigma)\), the weight influence is first-order under
misspecification:
\[
  \IF_{W(\hat\sigma)}(y_i)
    =
  H_W^{-1}\Delta\T W z_i
  -
  H_W^{-1}\Delta\T \{DW[\sigma_0]z_i\}\,r(\theta_W,\sigma_0).
\]
This second term drops out at exact fit. It is therefore most important in the
setting the user asked about: fixed model misspecification with GLS/WLS style
weights. This is also the term already handled by magmaan's continuous LS
influence machinery when the caller asks for data-dependent weight influence.

The efficient covariance-moment estimator for a fixed covariance target uses
\(W=\Gamma^{-1}\), giving the familiar
\[
  \Omega_{\mathrm{eff}}
    =
  (\Delta\T\Gamma^{-1}\Delta)^{-1}
\]
only at exact fit or when the curvature and weight-influence terms are absent.
Under misspecification the observed bread, not the Fisher simplification, is the
general object.

\section{Normal-Theory ML}

For covariance-only NT estimation,
\[
  F_{\mathrm{NT}}(\theta,\sigma)
    =
  \log|\Sigma_\theta|
  +\tr\{S(\sigma)\Sigma_\theta^{-1}\}
  -\log|S(\sigma)|-p.
\]
The term \(-\log|S|\) is irrelevant for the first-order condition. At the
pseudo-true point \(\theta_{\mathrm{NT}}\), define
\[
  U_a(\theta,\sigma)
    =
  \tr\left[
    \{\Sigma_\theta^{-1}
      -\Sigma_\theta^{-1}S(\sigma)\Sigma_\theta^{-1}\}
    \frac{\partial\Sigma_\theta}{\partial\theta_a}
  \right].
\]
Let \(H_{\mathrm{NT}}=\partial U/\partial\theta\T\) be the observed
pseudo-true bread. For a covariance perturbation \(dS\),
\[
  \{G_{\mathrm{NT}}d\sigma\}_a
    =
  -\tr\left[
    \Sigma_\theta^{-1}dS\Sigma_\theta^{-1}
    \frac{\partial\Sigma_\theta}{\partial\theta_a}
  \right].
\]
Thus
\[
  \IF_{\mathrm{NT}}(y_i)
    =
  -H_{\mathrm{NT}}^{-1}G_{\mathrm{NT}}z_i.
\]
When the normal model is correct this is the usual efficient likelihood
influence function. Under non-normality or fixed structural misspecification,
the sandwich \(H^{-1}G\Gamma G\T H^{-1}\) is the robust covariance around the
NT pseudo-true parameter.

\section{What Efficiency Means Under Misspecification}

Under fixed misspecification, Guttman, NT, ULS, GLS, and WLS generally estimate
different functionals of the same population covariance. A variance ratio is
therefore not automatically an MSE ratio.

For a scalar contrast \(a_e\{\tau_e(P)\}\), the variance-only ratio is
\[
  \rho_{G/e}
    =
  \frac{\nabla a_G\T \Omega_G \nabla a_G}
       {\nabla a_e\T \Omega_e \nabla a_e}.
\]
This is meaningful when the contrast names the same scientific quantity and
the pseudo-targets are close enough that first-order variance is the object of
interest.

For fixed misspecification, a finite-\(n\) target-risk comparison is more
honest:
\[
  R_e(n;A,\tau_0)
    =
  \{\tau_e(P)-\tau_0\}\T A\{\tau_e(P)-\tau_0\}
  +\frac1n\tr(A\Omega_e).
\]
Dhaene and Rosseel's MSE comparisons are of this type, with \(A\) selecting
groups of parameters and \(\tau_0\) the data-generating parameter vector. An
analytic experiment should report both \(R_e\) and the variance-only ratios.

A third option is local misspecification:
\[
  \sigma_n=\sigma\{\theta_0\}+n^{-1/2}\delta.
\]
Then bias and variance both contribute at first order, and the influence maps
above give closed-form local asymptotic risk. This is probably the cleanest
paper-level efficiency comparison.

\section{Proposed magmaan Experiment}

The experiment should be deterministic and population based first, with Monte
Carlo used only to validate the formulas.

\paragraph{1. Estimator maps.}
Implement a frontier Guttman CFA estimator that returns
\((\Lambda_G,\Phi_G,\psi_G)\) per covariance block using the clean map in this
note. Reuse \texttt{cfa\_block\_layouts()} and
\texttt{guttman\_start\_values()} as implementation scaffolding, but keep
finite-sample clamps out of the analytic target.

\paragraph{2. Jacobians.}
Start with central finite differences of \(\tau_G(\sigma)\) to verify the
chain above. Then add the analytic differential for speed and numerical
diagnostics. The Spearman-Harman communality derivative is a quotient-rule
calculation through the covariance-to-correlation map.

\paragraph{3. Comparator estimators.}
For each population covariance, fit the analysis model at the population
moments with NT, ULS, GLS, DWLS/WLS, and possibly empirical-ADF WLS. Use the
current observed-bread sandwich machinery for NT and continuous LS. For
data-dependent GLS/WLS weights, run both fixed-weight and weight-influence
versions to show how much the misspecification term matters.

\paragraph{4. Designs.}
Replicate the Dhaene-Rosseel three-factor, three-indicator design with
reliability \(0.5/0.8\), factor correlations \(0.1/0.4\), omitted
cross-loadings, and omitted residual covariances. Add one larger block design
with six to twelve indicators per factor, because Guttman's computational
appeal grows with \(p\).

\paragraph{5. Distributional covariance.}
Evaluate \(\Gamma\) under three choices: normal theory, an analytic elliptical
or independent-component heavy-tailed fourth-moment model, and empirical
fourth moments from a very large simulated DGP draw. This separates structural
misspecification from non-normal covariance noise.

\paragraph{6. Outputs.}
For each parameter group and selected scalar contrasts, report:
\[
  \diag(\Omega_G)/\diag(\Omega_e),\qquad
  \tr(A\Omega_G)/\tr(A\Omega_e),\qquad
  R_G(n;A,\tau_0)/R_e(n;A,\tau_0).
\]
Also report the pseudo-target displacement
\(\tau_e(P)-\tau_0\), because under fixed misspecification this often dominates
large-\(n\) MSE.

\paragraph{7. Validation.}
For a small subset, simulate many datasets, compute all estimators, and check
that \(n\,\widehat{\Cov}(\hat\tau_e)\) matches \(\Omega_e\). For Guttman this
is the main new validation. For NT and LS it should agree with existing
misspecification-robust sandwich code. This is implemented as the
\texttt{--validate} path in experiment 40: at the well-specified cell with
\(2000\) replications and \(N=1000\), the Guttman all-parameter asymptotic
SD matches simulation to within \(1\%\) (\(-0.8\%\)), bracketed by the
known-correct NT (\(-0.7\%\)) and oracle-WLS (\(-1.0\%\)) sandwiches, so the
finite-difference Jacobian of the Guttman map is as accurate as those
comparators.

\section{Takeaways}

Guttman's estimator has straightforward regular asymptotics because it is a
smooth covariance-summary map away from clamps, singular factor-score
correlations, and zero marker loadings. Its influence function is simply the
Jacobian of that map times the covariance influence row.

The comparison to NT, ULS, GLS, and WLS should not be framed as one universal
efficiency ranking. Under correct normal specification NT/GLS have the usual
optimality story. Under fixed misspecification, the relevant objects are
pseudo-targets, observed breads, empirical meats, and weight influence. That is
exactly the regime where an analytic magmaan experiment can be more informative
than another Monte Carlo table.

\begin{thebibliography}{9}

\bibitem{bollen1989}
Bollen, K. A. (1989).
\newblock \emph{Structural equations with latent variables}.
\newblock Wiley.

\bibitem{browne1974}
Browne, M. W. (1974).
\newblock Generalized least squares estimators in the analysis of covariance
structures.
\newblock \emph{South African Statistical Journal}, 8, 1--24.

\bibitem{cronbach1951}
Cronbach, L. J. (1951).
\newblock Coefficient alpha and the internal structure of tests.
\newblock \emph{Psychometrika}, 16, 297--334.
\newblock \href{https://doi.org/10.1007/BF02310555}
{doi:10.1007/BF02310555}.

\bibitem{dhaene2024}
Dhaene, S., \& Rosseel, Y. (2024).
\newblock An evaluation of non-iterative estimators in confirmatory factor
analysis.
\newblock \emph{Structural Equation Modeling: A Multidisciplinary Journal},
31(1), 1--13.
\newblock \href{https://doi.org/10.1080/10705511.2023.2187285}
{doi:10.1080/10705511.2023.2187285}.

\bibitem{guttman1952}
Guttman, L. (1952).
\newblock Multiple group methods for common-factor analysis: Their basis,
computation, and interpretation.
\newblock \emph{Psychometrika}, 17, 209--222.
\newblock \href{https://doi.org/10.1007/BF02288783}
{doi:10.1007/BF02288783}.

\bibitem{harman1976}
Harman, H. H. (1976).
\newblock \emph{Modern factor analysis} (3rd ed.).
\newblock University of Chicago Press.

\bibitem{mcdonald1999}
McDonald, R. P. (1999).
\newblock \emph{Test theory: A unified treatment}.
\newblock Lawrence Erlbaum.

\end{thebibliography}

\end{document}
