\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}

\title{Misspecification-Robust Standard Errors for the Moment-Quadratic
Estimators: Observed-Hessian Bread and the Infinitesimal Jackknife}
\author{}
\date{\today}

\newcommand{\T}{^{\mathsf T}}
\newcommand{\thh}{\widehat\theta}

\begin{document}
\maketitle

\section*{Context}

Implementation notes, not manuscript prose, for the \texttt{frontier} addition
that gives magmaan's moment-quadratic estimators (ordinal DWLS/WLSMV/ULSMV,
mixed/polyserial, and continuous ULS/GLS/WLS) a misspecification-robust standard
error, selectable as one regime keyword. Continuous ML and FIML already carried
this (their robust sandwich uses the observed Hessian by construction); the
moment-quadratic family did not.

\section*{The bread decomposition}

Every moment-quadratic estimator minimizes
\[
  F(\theta) = \tfrac12 \sum_b \frac{n_b}{N}\, d_b(\theta)\T W_b\, d_b(\theta),
  \qquad d_b(\theta) = \sigma_b(\theta) - s_b,
\]
with $s_b$ the sample moments (thresholds, polychorics, polyserials, or
$[\mu;\operatorname{vech}\Sigma]$), $\sigma_b(\theta)$ the model-implied moments,
$\Delta_b = \partial\sigma_b/\partial\theta$, and $W_b$ the estimator weight. The
exact Hessian is
\[
  H(\theta) = \sum_b \frac{n_b}{N}\Big[\,\Delta_b\T W_b \Delta_b
  \;+\; \textstyle\sum_k (W_b d_b)_k\, \partial^2\sigma_{b,k}/\partial\theta\,\partial\theta\T\,\Big].
\]
The first term is the Gauss--Newton (``expected'') bread $A=\Delta\T W\Delta$ that
the prevailing robust SE uses; the second is the residual-contracted curvature.
The sandwich $\operatorname{Var}(\thh)=A^{-1}BA^{-1}$ with meat
$B=\Delta\T W\,\Gamma\, W\Delta$ ($\Gamma$ the sampling ACOV of $s$, the NACOV)
is consistent for the variance of $\thh$ around its probability limit
\emph{only when $A$ is the true Hessian}.

\section*{Expected $=$ observed under the null}

Under correct specification the residual $d_b(\thh)\to 0$ in probability, the
curvature term vanishes, and $H\to A=\Delta\T W\Delta$. So the two breads
coincide at the null and the ``robust'' regime costs nothing there; the expected
bread is additionally the efficient (GLS-optimal) choice. Off the null
$W_b d_b\not\to 0$, the breads differ, and only the observed-Hessian sandwich is
consistent for $\operatorname{Var}(\thh)$. The error in the expected-bread SE is
first order in the residual, so even a model one would happily retain on fit
grounds can carry a non-trivial SE bias.

\section*{Implementation}

The observed bread is obtained by central-differencing the analytic per-unit
moment-LS gradient $g(\theta)=\sum_b (n_b/N)\,\Delta_b(\theta)\T W_b d_b(\theta)$
in full $\theta$-space (one model re-evaluation per $\theta\pm h$ coordinate),
then reducing to the constraint-space bread $H_\alpha=K\T H_\theta K$. This
mirrors the existing continuous-ML \texttt{information\_observed\_fd}; no analytic
moment-Hessian is needed (a future optimization). The result is injected as a
bread override into \texttt{robust\_weighted\_moments}; the meat ($\Gamma$/NACOV),
degrees of freedom, and reference $\chi^2$ are untouched. The keyword is
\texttt{bread = "expected" | "observed"}, the same vocabulary continuous ML/FIML
already use, surfaced in R as \texttt{vcov(fit, regime = "model" | "robust")},
which feeds \texttt{standardized(fit, vcov)} so the choice flows into
$\mathrm{SE}(\beta)$ for free.

\section*{Evidence}

Experiment 35 (\path{experiments/35-misspec-robust-se}). Two-factor
six-indicator model; item $y_4$ cross-loads on the wrong factor in the
data-generating process and the fitted model omits it. Focal: a free loading on
the distorted factor and its \texttt{std.all} counterpart, $N=600$. With the
model \emph{correct}, model and robust SE coincide and both match the empirical
sampling SD. With the cross-loading \emph{omitted}, the model SE underestimates
the sampling SD by about a third---the same for the raw and the standardized
loading---and the robust SE recovers it: essentially completely for continuous
ML (standardized loading $0.69\to1.01$ of truth) and substantially for ordinal
DWLS, the new bread ($0.78\to0.85$). The two regimes share the same empirical
meat, so the difference is the bread alone; the ordinal sandwich is asymptotic
and closes most, not all, of the SE ratio at this $N$ (DWLS std loading: $0.85$
at $N=600$, $0.93$ at $N=2400$).

A bootstrap arm (case resampling; percentile / BCa / studentized; coverage of
the pseudo-true value, nominal $0.95$) sharpens the verdict. Under
misspecification the \emph{expected}-bread interval (Muthen 1997 / lavaan
default) under-covers at $\sim0.87$, the \emph{observed}-bread interval recovers
nominal coverage at $\sim0.95$, and the bootstrap intervals slightly over-cover
($\sim0.97$). So for what actually matters---calibrated coverage---the analytic
observed bread is sufficient; the residual SE-ratio shortfall is heavy-tail mass
(the SD is tail-sensitive, central $95\%$ coverage is not), and the bootstrap is
a conservative safety net rather than a necessity here.

The one residue that is \emph{not} just tail mass: for a structural
($F\!\to\!F$) parameter under DWLS the observed-bread SE ratio plateaus below
$1$ and does \emph{not} improve with $n$ (DWLS $0.944$ at $N=2500$, $0.946$ at
$N=25000$ -- a $10\times$ bump moves it by $0.002$), whereas under ULS it is
$\approx 1$ at every $n$. This resolves Lai \& Simoes's ``slightly worse under
DWLS, unclear whether it vanishes'': it does not vanish. The new SE is the
asymptotic variance of the \emph{fixed}-weight estimator $\bar\theta=\phi(u)$,
but the practitioner uses the \emph{sample}-weight $\hat\theta$ with
$\hat W=\hat\Gamma_D^{-1}(u)$, whose influence carries an extra term
$\propto \hat W'(u)\,\varepsilon$ -- zero under the null (so the SE vanishes
there), but leading order under misspecification, and omitted by Lai's / our
observed-bread derivation. ULS ($W=I$, fixed) has no such term, so it is exact.
That analytic weight term is exactly the infinitesimal jackknife of the next
section; it recovers the missing variability with no resampling and no more $n$.

\section*{The infinitesimal jackknife: the estimated-weight influence}

The observed bread fixes $A$ but still treats $\hat W=\operatorname{diag}(\hat\Gamma)^{-1}$
as a constant. Under misspecification the estimation of $\hat W$ is leading order
(the Hall \& Inoue 2003 phenomenon for misspecified GMM): the categorical
estimator $\hat\theta$ depends on the data through both the stage-1 moments
$\hat\kappa=(\text{thresholds},\text{polychorics})$ and their empirical
covariance $\hat\Gamma$, and the standard sandwich meat propagates only the
former. The complete variance is the infinitesimal jackknife
\[
  \operatorname{Var}(\hat\theta)=\frac1{N^2}\sum_i \mathrm{IF}_i\,\mathrm{IF}_i\T,
  \qquad
  \mathrm{IF}_i = -H^{-1}\Delta\T\big[\hat W g_i + \mathrm{corr}_i\big],
\]
where $g_i$ is the per-case influence function of $\hat\kappa$ (magmaan already
forms it as $N\,\mathrm{SC}\,B^{-1\T}$, $B^{-1}$ the stage-1 estimating-equation
Jacobian inverse). The first bracket term reproduces the observed-bread sandwich
\emph{exactly}; in particular ULS, where $\hat W=I$ is constant and
$\mathrm{corr}_i=0$, is machine-identical to it. The second is the weight
influence, $\mathrm{corr}_{i,k} = d_k\,\mathrm{IF}_{i,k}(\hat\Gamma)/\hat\Gamma_{kk}^2$.

\paragraph{Three channels.} $\hat\Gamma=\hat A^{-1}\hat V\hat A^{-1}$ is itself a
sandwich ($\hat V$ the score cross-products, $\hat A$ the score Jacobian), so its
influence has three pieces, not two:
\begin{enumerate}\itemsep0pt
\item $\hat V$-direct (score-product wobble),
  $\hat A^{-1}(s_is_i\T-\hat V)\hat A^{-1}$, with diagonal $g_{i,k}^2-\hat\Gamma_{kk}$;
\item $\hat A$-direct (score-Jacobian wobble),
  $-\hat A^{-1}\delta\hat A_i\,\hat\Gamma-\hat\Gamma\,\delta\hat A_i\hat A^{-1}$,
  $\delta\hat A_i=b_i-\hat A$ with $b_i$ the per-case Jacobian contribution;
\item $\kappa$-movement ($\hat A,\hat V$ shift as $\hat\kappa$ shifts),
  $(\partial\hat\Gamma/\partial\kappa)\,g_i$.
\end{enumerate}
The naive ``meat $+$ FD-bread'' split keeps (1) and (3) and drops (2); since (2)
is negative here, that split overshoots (the SE ratio runs $1.03\!\to\!1.18$
against a target of $\approx1.08$). Channels (1)$+$(2), the per-case data-direct
sandwich influence, collapse to the closed diagonal
\[
  \mathrm{IF}_{i,k}(\hat\Gamma)^{\mathrm{data}}
  = g_{i,k}^2 + \hat\Gamma_{kk}
    - N\big[(B^{-1} b_i\hat\Gamma)_{kk} + (\hat\Gamma b_i B^{-1})_{kk}\big],
\]
which reduces to the $\hat V$-direct meat when $b_i\to\hat A=B/N$ (no per-case
Jacobian variation), a useful sanity check.

\paragraph{The structural catch.} $b_i$ is \emph{not} $\operatorname{mask}(s_is_i\T)$
and is \emph{not} recoverable from the stored $g_i$, because the blocks of the
two-stage Jacobian $B$ use different score functions: the threshold block
$A_{11}$ uses the \emph{marginal} univariate threshold scores, the
threshold-polychoric coupling $A_{21}$ uses the \emph{bivariate} pair scores, and
$A_{22}$ the polychoric scores. So the data-direct channel recomputes all
per-case scores from the integer data (reusing the polychoric pair-score routine)
rather than reconstructing them. No third derivatives appear: $\hat A$ is the BHHH
(outer-product) form, so $b_i$ is built from first-order scores, and
$\partial\hat\Gamma/\partial\kappa$ in channel (3) is finite-differenced (perturb
$\kappa$, recompute $\hat\Gamma$ from the integer data). The analytic
$\partial\hat\Gamma/\partial\kappa$ (third derivatives of the bivariate
polychoric likelihood) is a deferred optimization.

\paragraph{Implementation.} \texttt{ordinal\_gamma\_diag\_data\_influence}
(channels 1+2) and \texttt{ordinal\_gamma\_diag\_jacobian\_fd} (channel 3) live in
\path{src/data/ordinal.cpp}; \texttt{robust\_ordinal\_ij}
(\path{src/estimate/ordinal.cpp}) assembles $\mathrm{IF}_i$ and the sandwich.
\texttt{OrdinalStats} now carries $g_i$ (\texttt{moment\_influence}), the integer
data (\texttt{int\_data}), and $B^{-1}$ (\texttt{moment\_bread}), round-tripped
through R; the surface is \texttt{robust\_ordinal\_ij}. DWLS only (ULS needs no
correction; full WLS errors for now).

\paragraph{Evidence.} Coverage referee on the exp-35 structural focal ($N=2500$,
coverage of the pseudo-true value, nominal $0.95$): true sampling SD $0.0394$;
observed-bread SE $0.0366$ (ratio $0.93$, coverage $0.927$); the IJ SE $0.0388$
(ratio $0.985$, coverage $0.940$, best of the three); a delete-$d$ resampling
jackknife $0.0395$ serving only as the yardstick. The IJ recovers the true SD
analytically to $\sim1.5\%$, with no resampling. (150 replications, so the
coverage spread is within one Monte-Carlo standard error; the SE-vs-SD column is
the cleaner signal. One focal, one $N$, one misspecification.)

\section*{Where this sits relative to other software}

For continuous ML/FIML the observed-Hessian sandwich is ubiquitous
(\texttt{robust.huber.white}/MLR in lavaan, MLR in Mplus). For the
ordinal/WLS family it is, as far as we know, new surface: lavaan's DWLS/WLSMV
robust SE is the Gauss--Newton/expected (\texttt{robust.sem}-style) sandwich
only. So the correct-model cell of experiment 35 is the validation oracle (the
observed bread must coincide with the expected one there); there is no external
ordinal observed-bread SE to gate against. This is \texttt{frontier}, not
core lavaan parity.

The infinitesimal jackknife is further out still. Continuous MLR
(\texttt{robust.huber.white}/Mplus MLR) is misspecification-robust partly because
ML's weight $\Sigma(\theta)^{-1}$ is model-implied, not data-estimated, so it
carries no weight-influence term. The categorical analogue, a
misspecification-robust SE that accounts for the \emph{estimated} weight
$\hat W=\operatorname{diag}(\hat\Gamma)^{-1}$, has no counterpart in mainstream
software. Its only checks are internal: ULS is machine-identical to the observed
bread (the weight is fixed there), and a resampling jackknife serves as a
same-target yardstick.

\end{document}
