\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}

\title{Saturated FIML Information, MLR, and FMG under Non-normal
Incomplete Data}
\author{}
\date{\today}

\newcommand{\E}{\mathbb{E}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\avar}{\operatorname{avar}}
\newcommand{\vech}{\operatorname{vech}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\eig}{\operatorname{eig}}

\newtheorem*{claim}{Claim}
\newtheorem*{conclusion}{Conclusion}

\begin{document}
\maketitle

\section*{Purpose}

The useful question is not whether FIML under non-normality has a ``Gamma
matrix''. It does. The useful question is which curvature and which reference
law are being used.

For a FIML goodness-of-fit correction there are three separate objects:
\begin{enumerate}
  \item the saturated observed-data normal curvature \(V\);
  \item the saturated sandwich covariance
        \(\Gamma_{\mathrm{mis}} = H^{-1}JH^{-1}\);
  \item the reference-law reduction of \(U\Gamma_{\mathrm{mis}}\), either a
        scalar mean scaling or the full eigenvalue mixture.
\end{enumerate}
The MCAR simulations should be understood mainly through the third object.
Under a correct null and MCAR, the saturated and structured definitions of
\(V\) have the same asymptotic target. Non-normality changes
\(\Gamma_{\mathrm{mis}}\) and spreads the eigenvalues of
\(U\Gamma_{\mathrm{mis}}\); it does not by itself create a different
saturated-versus-structured \(V\) target.

\section*{Observed-data normal H1 model}

Let row \(i\) have observed index set \(o_i \subset \{1,\ldots,p\}\). The
working FIML H1 model is a normal model for the observed subvector,
\[
  \ell_i(\eta)
    =
  \log \phi_{|o_i|}
    \{x_{i,o_i}; \mu_{o_i}, \Sigma_{o_i o_i}\},
  \qquad
  \eta = (\mu, \vech(\Sigma)).
\]
The saturated H1 estimator \(\hat\eta\) maximizes
\(\sum_i \ell_i(\eta)\). This is a normal quasi-likelihood when the data are
non-normal; the general sandwich theory is the usual misspecified-likelihood
theory of White \cite{white1982}.

Define the saturated H1 score
\[
  h_i(\eta) = \frac{\partial \ell_i(\eta)}{\partial \eta}
\]
and let \(\eta^\dagger\) denote the population maximizer of
\(\E\{\ell_i(\eta)\}\). The population bread and meat are
\[
  H_\eta
    =
  -\E\left[
    \frac{\partial h_i(\eta^\dagger)}{\partial\eta'}
  \right],
  \qquad
  J_\eta
    =
  \E\{h_i(\eta^\dagger)h_i(\eta^\dagger)'\}.
\]
Then
\[
  \avar\{\sqrt N(\hat\eta-\eta^\dagger)\}
    =
  \Gamma_{\mathrm{mis}}
    =
  H_\eta^{-1}J_\eta H_\eta^{-1}.
\]
This is the missing-data analogue of the complete-data ADF covariance of
sample moments \cite{browne1984}. If the observed-data normal model is
correctly specified, information equality gives \(J_\eta = H_\eta\). Under
non-normality, \(J_\eta\) generally contains the relevant third- and
fourth-moment information and is not equal to \(H_\eta\).

\section*{The FIML likelihood-ratio statistic}

The restricted SEM is the null \(H_0:\eta=\eta(\theta)\) nested in the
saturated H1 model. Let \(\hat\eta\) be the saturated FIML maximizer and
\(\eta(\hat\theta)\) the restricted FIML maximizer of the same observed-data
normal log-likelihood. The goodness-of-fit statistic is the ordinary
likelihood-ratio,
\[
  T
    =
  2\left\{
    \textstyle\sum_i \ell_i(\hat\eta)
    -
    \textstyle\sum_i \ell_i(\eta(\hat\theta))
  \right\},
  \qquad
  \mathrm{df} = \dim(\eta) - \dim(\theta),
\]
i.e.\ lavaan's default \texttt{missing = "ml"} test (the same statistic the
nested configural-versus-metric comparison forms as a difference of two such
fits). Under a correctly specified normal model with ignorable missingness the
information equality \(J_\eta = H_\eta\) holds, so
\(U(V)\Gamma_{\mathrm{mis}}\) is a projector, all nonzero \(\lambda_j=1\), and
\[
  T \Rightarrow \chi^2_{\mathrm{df}}.
\]
This is the only regime in which the unscaled \(T\) is asymptotically pivotal.
Under non-normality the working normal likelihood is misspecified, information
equality fails, and \(T\) is no longer \(\chi^2_{\mathrm{df}}\); its limit is
the weighted mixture derived below, with mean \(\sum_j\lambda_j=\tr\{U(V)
\Gamma_{\mathrm{mis}}\}\) rather than \(\mathrm{df}\). Every robust correction
(MLR, SB, the FMG family) is a different way of referencing \(T\) to that
mixture instead of to \(\chi^2_{\mathrm{df}}\).

\section*{Definition of \(V\)}

The matrix \(V\) is the local curvature used to approximate the H1 likelihood
in saturated moment coordinates. In the saturated FIML FMG construction,
\[
  V = H_\eta(\eta^\dagger)
\]
in population notation. In the finite-sample magmaan implementation,
\[
  V_N
    =
  -\sum_{i=1}^N
    \frac{\partial^2 \ell_i(\hat\eta)}
         {\partial\eta\,\partial\eta'}
\]
and
\[
  \hat\Gamma_{\mathrm{mis}}
    =
  V_N^{-1}
  \left\{
    \sum_{i=1}^N h_i(\hat\eta)h_i(\hat\eta)'
  \right\}
  V_N^{-1}.
\]
This finite-sample \(\hat\Gamma_{\mathrm{mis}}\) carries \(1/N\) units, while
\(V_N\) carries \(N\) units; their product has the same limiting eigenvalues
as the population convention above.

Let the restricted SEM imply saturated coordinates \(\eta(\theta)\), and let
\[
  \Delta
    =
  \frac{\partial \eta(\theta)}{\partial\theta'}
\]
at the null value or at \(\hat\theta\). With equality constraints, \(\Delta\)
is first projected into the local free-parameter tangent space. The residual
operator for a chosen curvature \(V\) is
\[
  U(V)
    =
  V - V\Delta(\Delta'V\Delta)^{-1}\Delta'V.
\]
The robust reference law for the FIML LRT is the weighted mixture
\[
  T
    \Rightarrow
  \sum_{j=1}^{\mathrm{df}} \lambda_j \chi^2_{1j},
  \qquad
  \lambda_j
    =
  \eig_j\{U(V)\Gamma_{\mathrm{mis}}\},
\]
the same Satorra--Bentler \(U\Gamma\) structure used for robust moment
structure tests \cite{satorrabentler1994,satorra2000}. When the normal H1
model is correct, \(\Gamma_{\mathrm{mis}}=V^{-1}\), so
\(U(V)\Gamma_{\mathrm{mis}}\) is a rank-\(\mathrm{df}\) projector and the
nonzero eigenvalues are all one.

\section*{What MCAR buys}

Under MCAR, the missingness pattern is independent of the data values. Suppose
the full data have finite second moments \((\mu_0,\Sigma_0)\). For any
observed pattern \(o\),
\[
  \E(X_o-\mu_{0,o}) = 0,
  \qquad
  \E\{(X_o-\mu_{0,o})(X_o-\mu_{0,o})'\} = \Sigma_{0,oo}.
\]
Therefore the expected saturated normal score for that pattern is zero at
\(\eta_0=(\mu_0,\vech(\Sigma_0))\). Summing over the MCAR pattern
distribution preserves this zero-score property. Under ordinary uniqueness
conditions,
\[
  \eta^\dagger = \eta_0.
\]

If the restricted SEM is also correct, so that
\[
  \eta(\theta_0)=\eta_0,
\]
then the saturated curvature and the structured/model-implied curvature have
the same asymptotic target:
\[
  H_\eta(\eta^\dagger)
    =
  H_\eta\{\eta(\theta_0)\}.
\]
This statement does not require normality; it only uses the fact that, under
MCAR, the observed-pattern normal score is unbiased at the full-data first two
moments.

\begin{conclusion}
For a correct model under MCAR, a persistent asymptotic gap between saturated
FMG and lavaan-style MLR should not be explained by a different limiting
definition of \(V\). If the simulations show MCAR improvement for FMG over MLR,
the first explanation should be the reference-law reduction: full spectrum
versus scalar mean scaling.
\end{conclusion}

\section*{Where non-normality enters under MCAR}

Under non-normality and MCAR, \(\eta^\dagger\) can still equal the full-data
mean and covariance, but \(J_\eta\) is not the normal information matrix.
Consequently
\[
  \Gamma_{\mathrm{mis}}
    =
  H_\eta^{-1}J_\eta H_\eta^{-1}
    \ne
  H_\eta^{-1},
\]
and the nonzero eigenvalues of \(U(V)\Gamma_{\mathrm{mis}}\) need not be
equal. This is the direct missing-data analogue of the complete-data ADF
story \cite{browne1984,yuanbentler2000}.

The Yuan--Bentler / MLR mean-scaled statistic uses only
\[
  c_{\mathrm{SB}}
    =
  \frac{1}{\mathrm{df}}
  \tr\{U(V)\Gamma_{\mathrm{mis}}\}
    =
  \frac{1}{\mathrm{df}}\sum_{j=1}^{\mathrm{df}}\lambda_j,
\]
and compares \(T/c_{\mathrm{SB}}\) to \(\chi^2_{\mathrm{df}}\). That is exact
only in special cases, for example when the nonzero \(\lambda_j\)'s are all
equal. It can be a poor approximation when the spectrum is dispersed.

This \(c_{\mathrm{SB}}\) is precisely lavaan's default \texttt{estimator =
"MLR"} (Mplus \texttt{yuan.bentler.mplus}) scaling for incomplete data: the
``MLR'' branding hides that it is a \emph{first-moment} correction, matching
only the mean of the reference law. Its failure under a dispersed spectrum is
therefore not a new phenomenon but the classical Satorra--Bentler mean-scaling
limitation \cite{satorrabentler1994,yuanbentler2000} carried over to the FIML
setting; the Mplus variant differs from the asymptotic trace only in
finite-sample bookkeeping of \(\tr\{U(V)\Gamma_{\mathrm{mis}}\}\), not in being
mean-only. A dispersed spectrum is exactly what non-normality produces, so a
mean scaling must over-reject whenever \(\operatorname{var}_j(\lambda_j)\) is
large relative to \(\bar\lambda^2\).

The FMG family keeps the spectrum. The exact all-eigenvalue calculation uses
the distribution of \(\sum_j\lambda_j\chi^2_{1j}\), which can be evaluated by
standard quadratic-form methods such as Imhof's method \cite{imhof1961}. EBA
and penalized FMG variants approximate or shrink the same eigenvalue spectrum
\cite{foldnesgronneberg2018,foldnesmossgronneberg2025,foldnesgronnebergmoss2026}.
Thus the MCAR headline comparison is not
``different Gamma'' and not necessarily ``different \(V\)''. It is usually
\[
  \text{mean-only MLR scaling}
  \quad\text{versus}\quad
  \text{spectrum-aware FMG reference law}.
\]

\section*{Structured versus saturated \(V\)}

In words: the \emph{saturated} (``unstructured'') choice reads the H1 curvature
off the data alone, through the saturated EM moments \(\hat\eta\); the
\emph{structured} choice reads it off the fitted SEM, through the
model-implied moments \(\eta(\hat\theta)\). They share a probability limit
under a correct MCAR null, so the experiment's two \texttt{\_structured}
columns are a finite-sample / robustness check, not a different estimand. They
separate only under misspecification or non-normal MAR, as below.

The structured H1 analogue evaluates the same normal H1 curvature at the
restricted fitted moments,
\[
  V_{\mathrm{struct}}
    =
  H_\eta\{\eta(\hat\theta)\},
\]
whereas saturated FMG evaluates it at the saturated H1 fit,
\[
  V_{\mathrm{sat}}
    =
  H_\eta(\hat\eta).
\]
If both constructions are forced into the same eta-space comparison with the
same \(\Gamma_{\mathrm{mis}}\), then the only algebraic difference is
\[
  \eig\{U(V_{\mathrm{struct}})\Gamma_{\mathrm{mis}}\}
  \quad\text{versus}\quad
  \eig\{U(V_{\mathrm{sat}})\Gamma_{\mathrm{mis}}\}.
\]

Under the correct MCAR null these two \(V\)'s share the same probability
limit, but their finite-sample traces can differ because one uses
\(\hat\theta\)-implied moments and the other uses the saturated EM moments.
The advisory trace-convergence check in this repository is exactly about that
finite-sample distinction.

The \(V\)-evaluation distinction matters asymptotically in other regimes:
\begin{enumerate}
  \item under restricted-model misspecification, because
        \(\eta(\theta^\star)\ne\eta^\dagger\);
  \item under non-normal MAR working likelihoods, because the observed-data
        normal quasi-likelihood target can differ from the full-data moment
        target.
\end{enumerate}
The second case is not the clean MCAR calibration problem. Normal FIML has its
usual likelihood justification under MAR when the observed-data normal model
is correctly specified. Under non-normal MAR, Gaussian FIML is a working
likelihood; it may be biased for the full-data moment model, and the LRT can
be noncentral under the intended SEM null. No reference-law correction can
fully fix that estimator-target problem
\cite{yuanbentler2000,savalei2010}.

\section*{Two-variable \(V\) witness}

The following complete-data example is useful only as a witness for the
\(V_{\mathrm{struct}}\) versus \(V_{\mathrm{sat}}\) distinction. It is not the
MCAR null mechanism.

Let the saturated covariance of two observed variables be
\[
  \Sigma_Y =
  \begin{bmatrix}
    1 & \rho \\
    \rho & 1
  \end{bmatrix},
  \qquad \rho \ne 0,
\]
but fit the restricted independence model
\[
  \Sigma(\theta) =
  \begin{bmatrix}
    \theta_1 & 0 \\
    0 & \theta_2
  \end{bmatrix}.
\]
Then the restricted pseudo-true covariance is \(I_2\), while the saturated
covariance is \(\Sigma_Y\). The gap is misspecification; it exists even under
normal data.

Use covariance moments ordered as
\[
  \eta = (\sigma_{11},\sigma_{21},\sigma_{22})'.
\]
The restricted derivative is
\[
  \Delta =
  \begin{bmatrix}
    1 & 0 \\
    0 & 0 \\
    0 & 1
  \end{bmatrix}.
\]
At the structured point \(I_2\),
\[
  \Gamma_{\mathrm{NT}}(I_2)
    =
  \begin{bmatrix}
    2 & 0 & 0 \\
    0 & 1 & 0 \\
    0 & 0 & 2
  \end{bmatrix},
  \qquad
  V_{\mathrm{struct}} = \Gamma_{\mathrm{NT}}(I_2)^{-1},
\]
and
\[
  U_{\mathrm{struct}}
    =
  \begin{bmatrix}
    0 & 0 & 0 \\
    0 & 1 & 0 \\
    0 & 0 & 0
  \end{bmatrix}.
\]
At the saturated covariance,
\[
  \Gamma_{\mathrm{NT}}(\Sigma_Y)
    =
  \begin{bmatrix}
    2 & 2\rho & 2\rho^2 \\
    2\rho & 1+\rho^2 & 2\rho \\
    2\rho^2 & 2\rho & 2
  \end{bmatrix}.
\]
If the trace uses \(U_{\mathrm{struct}}\) with the saturated covariance-moment
matrix, then
\[
  c_{\mathrm{struct}}
    =
  \tr\{U_{\mathrm{struct}}\Gamma_{\mathrm{NT}}(\Sigma_Y)\}
    =
  1+\rho^2.
\]
By contrast, the unstructured/saturated choice uses
\[
  V_{\mathrm{sat}}=\Gamma_{\mathrm{NT}}(\Sigma_Y)^{-1},
\]
and \(U(V_{\mathrm{sat}})\Gamma_{\mathrm{NT}}(\Sigma_Y)\) is a rank-one
projector, so
\[
  c_{\mathrm{sat}} = 1.
\]
This proves that the evaluation point of \(V\) can matter. It does not prove
that \(V\) is the explanation for non-normal MCAR null calibration failures.

\section*{The MAR mechanism used (and why it is weak)}

MAR is not one mechanism, and the strength of the estimator-target problem
above depends entirely on which MAR is used. Experiment 23 uses the
Savalei--Bentler 2005 MAR generator \cite{savalei2010}: two indicators
\(x_1,x_2\) are always observed and act as the missingness cause, and the
remaining indicators \(x_3,\ldots,x_6\) are deleted by coarse \emph{sign}
rules on the always-observed pair (rows with \(x_1+x_2>0\), and rows with
\(x_1<0\)), with the deletion probabilities calibrated to the target overall
rate. This is a deliberately mild MAR for two reasons:
\begin{enumerate}
  \item the selection depends only on the \emph{sign} of always-observed
        indicators, a binary threshold rather than a steep continuous
        function, so the conditional missingness probability takes only a
        few levels; and
  \item because \(x_1,x_2\) enter the FIML likelihood, Gaussian FIML recovers
        the conditional mean of the missing entries given the observed ones
        \emph{exactly} whenever that conditional mean is linear, and the
        Vale--Maurelli / inverse-Gaussian / power-law generators keep
        conditional means close to linear. The non-normality lives in the
        marginals, not in a curved regression surface.
\end{enumerate}
Consequently the estimator-target bias that the structured-versus-saturated
section flags as the irreducible MAR limitation is \emph{small} in this design,
and the MAR cells behave almost like MCAR for the well-calibrated tests. This
is a property of \emph{this} MAR, not of MAR in general: a stronger mechanism
(steep continuous selection on the observed scores, or a generator with a
genuinely nonlinear missing-given-observed conditional mean) would make
\(\eta^\dagger \ne \eta_0\) bite, push the base statistic non-central under the
intended SEM null, and leave a residual no reference-law correction can remove.

\section*{Size-adjusted power}

Ordinary power rewards over-rejection: a test that rejects 27\% of \emph{true}
models also rejects more false ones, but that detection is borrowed from a
broken Type-I rate. Size-adjusted (empirical-size-corrected) power removes the
loan. For a fixed cell and method, let \(F_0\) be the simulated distribution of
the \(p\)-value under the true model, and set the critical threshold to its
\(\alpha\)-quantile,
\[
  c \;=\; F_0^{-1}(\alpha),
\]
the cutoff that \emph{would} give exactly size \(\alpha\) in that cell. The
size-adjusted power is then \(P(p<c)\) under the alternative. A well-calibrated
test has \(c\approx\alpha\) and ordinary \(\approx\) size-adjusted power; a
liberal test has \(c<\alpha\), which is exactly how its inflated raw power is
clawed back. This is a diagnostic, not an operational procedure: \(c\) is read
from the null simulation and is unavailable to a practitioner with one data
set. Its purpose is to compare the \emph{intrinsic} discriminating ability of
the statistics on equal honest footing.

\section*{Diagnostics for experiment 23}

For the MCAR cells, the clean diagnostic sequence is:
\begin{enumerate}
  \item Compare lavaan MLR with magmaan's FMG \(SB\) row. Both are mean-only
        trace scalings. If they agree and both miscalibrate, the culprit is the
        scalar reference-law approximation.
  \item Inspect the eigenvalue dispersion of \(U\Gamma_{\mathrm{mis}}\). Large
        spread predicts poor mean-scaling performance and gives FMG room to
        improve.
  \item Check that
        \(\sum_j\lambda_j\) from saturated FMG approaches the FIML MLR trace as
        \(N\) grows under MCAR. A persistent large gap points to a convention or
        implementation mismatch in \(V\), \(\Delta\), or scaling.
  \item Treat MAR cells separately. Mean base statistic exceeding mean trace is
        an estimator-target/noncentrality warning, not a failure of the
        eigenvalue reference law.
\end{enumerate}

The \texttt{noncentrality.csv} diagnostic reports, per \(h_0\) cell,
\[
  \widehat{\mathrm{ncp}}
    =
  \overline{T} - \overline{\tr\{U(V)\hat\Gamma_{\mathrm{mis}}\}}
    =
  \texttt{mean\_base} - \texttt{mean\_trace},
\]
the gap between the mean base LRT and the mean of its own reference law. With a
consistent estimator under a correct null \(\E[T]\to\tr\{U(V)
\Gamma_{\mathrm{mis}}\}\), so \(\widehat{\mathrm{ncp}}\to 0\); a positive value
is the estimated non-centrality the base statistic carries \emph{beyond} what
mean scaling can absorb. One caveat reading it: in finite samples the sandwich
trace \(\tr\{U(V)\hat\Gamma_{\mathrm{mis}}\}\) is itself downward biased under
heavy non-normality, so a moderate positive \(\widehat{\mathrm{ncp}}\) under
\emph{MCAR} is largely a finite-sample sandwich-bias artifact and shrinks with
\(N\); it does not contradict the calibration of the spectrum-aware tests,
which match the same finite-sample spectrum. Only a positive
\(\widehat{\mathrm{ncp}}\) that \emph{persists} as \(N\) grows -- the strong
non-normal MAR case -- signals the estimator-target bias that no reference-law
correction removes.

\begin{thebibliography}{9}

\bibitem{browne1984}
Browne, M. W. (1984).
Asymptotically distribution-free methods for the analysis of covariance
structures.
\emph{British Journal of Mathematical and Statistical Psychology}, 37(1),
62--83.
\href{https://doi.org/10.1111/j.2044-8317.1984.tb00789.x}{doi:10.1111/j.2044-8317.1984.tb00789.x}.

\bibitem{foldnesgronneberg2018}
Foldnes, N., \& Gronneberg, S. (2018).
Approximating test statistics using eigenvalue block averaging.
\emph{Structural Equation Modeling}, 25(1), 101--114.
\href{https://doi.org/10.1080/10705511.2017.1373021}{doi:10.1080/10705511.2017.1373021}.

\bibitem{foldnesmossgronneberg2025}
Foldnes, N., Moss, J., \& Gronneberg, S. (2025).
Improved goodness of fit procedures for structural equation models.
\emph{Structural Equation Modeling}, 32(1), 1--13.
\href{https://doi.org/10.1080/10705511.2024.2372028}{doi:10.1080/10705511.2024.2372028}.

\bibitem{foldnesgronnebergmoss2026}
Foldnes, N., Gronneberg, S., \& Moss, J. (2026).
Penalized eigenvalue block averaging: Extension to nested model comparison and
Monte Carlo evaluations.
\emph{Behavior Research Methods}, 58(4), Article 107.
\href{https://doi.org/10.3758/s13428-026-02968-4}{doi:10.3758/s13428-026-02968-4}.

\bibitem{imhof1961}
Imhof, J. P. (1961).
Computing the distribution of quadratic forms in normal variables.
\emph{Biometrika}, 48(3/4), 419--426.
\href{https://doi.org/10.1093/biomet/48.3-4.419}{doi:10.1093/biomet/48.3-4.419}.

\bibitem{satorra2000}
Satorra, A. (2000).
Scaled and adjusted restricted tests in multi-sample analysis of moment
structures.
In R. D. H. Heijmans, D. S. G. Pollock, \& A. Satorra (Eds.),
\emph{Innovations in Multivariate Statistical Analysis} (pp. 233--247).
Kluwer.
\href{https://doi.org/10.1007/978-1-4615-4603-0_17}{doi:10.1007/978-1-4615-4603-0\_17}.

\bibitem{satorrabentler1994}
Satorra, A., \& Bentler, P. M. (1994).
Corrections to test statistics and standard errors in covariance structure
analysis.
In A. von Eye \& C. C. Clogg (Eds.),
\emph{Latent Variables Analysis: Applications for Developmental Research}
(pp. 399--419). Sage.

\bibitem{savalei2010}
Savalei, V. (2010).
Expected versus observed information in SEM with incomplete normal and
nonnormal data.
\emph{Psychological Methods}, 15(4), 352--367.
\href{https://doi.org/10.1037/a0020143}{doi:10.1037/a0020143}.

\bibitem{white1982}
White, H. (1982).
Maximum likelihood estimation of misspecified models.
\emph{Econometrica}, 50(1), 1--25.
\href{https://doi.org/10.2307/1912526}{doi:10.2307/1912526}.

\bibitem{yuanbentler2000}
Yuan, K.-H., \& Bentler, P. M. (2000).
Three likelihood-based methods for mean and covariance structure analysis with
nonnormal missing data.
\emph{Sociological Methodology}, 30(1), 165--200.
\href{https://doi.org/10.1111/0081-1750.00078}{doi:10.1111/0081-1750.00078}.

\end{thebibliography}

\end{document}
