\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}

\title{Sum-Score Reliability Under Measurement Model Misspecification Is Target-Relative:\\
A Short Note on Omega Simulation Studies}
\author{}
\date{\today}

\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\T}{^{\mathsf T}}
\newcommand{\one}{\mathbf{1}}

\begin{document}
\maketitle

\begin{abstract}
Simulation studies of coefficient omega under measurement model
misspecification usually begin by choosing a population factor model, computing
the sum-score reliability implied by that model, fitting alternative models to
generated data, and reporting bias relative to the population value. This is a
coherent model-relative exercise. It is not, however, a model-free assessment of
bias for the composite's reliability. The reason is familiar from the
literature on equivalent structural equation models: the observed covariance
matrix does not, by itself, identify the substantive latent decomposition. For
omega as a composite reliability coefficient, the estimand is the proportion of
sum-score variance attributed to a named target factor. Equivalent
decompositions can therefore imply the same item covariance matrix while
assigning different reliabilities to the same sum score. We use this point to
reinterpret recent omega simulations as conditional sensitivity analyses and
argue that future simulations should separate estimator behavior from target
uncertainty.
\end{abstract}

\noindent\textbf{Keywords:} coefficient omega, sum-score reliability,
composite reliability, equivalent models, bifactor model

\section{The Problem}

Coefficient omega is attractive because it asks a substantively meaningful
question: how much of the observed sum score is attributable to the latent
variable the score is intended to measure? That question is also what makes
omega fragile. The answer depends on which latent variable is treated as the
target construct and which parts of the item covariance structure are treated as
nuisance, method, local dependence, specific-factor variance, or error.

Bell, Chalmers, and Flora \cite{bell2024} give a useful recent example. Their
simulation study is explicitly about reliability of the composite score
\(X=\one\T y\). They generate item covariances from four population families:
a simple one-factor model, a one-factor model with correlated errors, a
bifactor model, and a higher-order model. They then fit one-factor,
correlated-error one-factor, and bifactor models and compare the resulting
omega estimates with the omega value implied by the population model. In their
notation, \(\omega_u\) is the sum-score reliability with respect to the single
factor in a one-factor model; \(\omega_H\) is the sum-score reliability with
respect to the general factor in a bifactor model; and \(\omega_{ho}\) is the
sum-score reliability with respect to the higher-order factor through its
indirect effects.

The present note argues that there is a deeper target issue behind such
simulation results. Once a population model and target factor are stipulated,
the bias calculation is well-defined. But the word ``true'' supplies not only a
covariance matrix. It also supplies a substantive allocation of covariance to
the target factor rather than to other latent or residual sources. Under model
misspecification, that allocation is often exactly what is in question.

This is not a new problem in structural equation modeling. Equivalent models can
fit the same covariance matrix while representing different substantive
stories \cite{bollen1989,maccallum1993}. Fit alone cannot determine which story
is the data-generating one. The same difficulty becomes sharper for sum-score
reliability because the estimand itself can change across equivalent stories.
Reliability is not always a single property of a test; it can depend on what
the score is meant to measure \cite{borsboom2002,ellis2021,savalei2019}.

\section{Sum-Score Omega as a Target-Factor Quantity}

Let \(y\) be the vector of item scores and let \(X=\one\T y\) be the sum score.
Suppose a model says that a target factor \(f\), scaled to have variance one,
contributes loading vector \(\lambda\) to the items:
\[
  y=\lambda f+u,\qquad \Cov(f,u)=0,\qquad \Var(y)=\Sigma .
\]
The omega target for the sum score, with respect to \(f\), is
\begin{equation}\label{eq:omega}
  \omega_f
  =
  \frac{\Var(\one\T\lambda f)}{\Var(\one\T y)}
  =
  \frac{(\one\T\lambda)^2}{\one\T\Sigma\one}.
\end{equation}
This definition uses the full sum-score variance in the denominator. If the
model contains correlated residuals, specific factors, or lower-order factors,
their contribution remains in \(\one\T\Sigma\one\). What changes from model to
model is the numerator: which part of the item covariance is assigned to the
named target factor.

Thus, when Bell et al. label a population condition as having reliability .60 or
.85, that label is not a property of the covariance matrix alone. It is a
property of the covariance matrix together with a particular latent
interpretation: a particular \(\lambda\) is declared to be the contribution of
the target factor to the items.

\section{Equivalent Models Can Change the Sum-Score Target}

The issue can be seen with a simple construction. It is not meant as a proposed
measurement model. It is only a way to show what is, and is not, identified by
the covariance matrix. Let \(\Sigma\) be any positive definite item covariance
matrix. For any value \(r\) between zero and one, set
\[
  \lambda_r=\sqrt{r}\frac{\Sigma\one}{\sqrt{\one\T\Sigma\one}},
  \qquad
  \Psi_r=\Sigma-\lambda_r\lambda_r\T .
\]
Then
\[
  \Sigma=\lambda_r\lambda_r\T+\Psi_r
\]
is a one-factor model with correlated residuals. The residual covariance
\(\Psi_r\) is nonnegative definite by Cauchy's inequality, and
\[
  (\one\T\lambda_r)^2=r\,\one\T\Sigma\one .
\]
Equation~\eqref{eq:omega} therefore gives sum-score reliability \(r\) with
respect to the common factor in that decomposition.

Thus, if correlated residuals are allowed without further restriction, the same
observed covariance matrix can be paired with many different sum-score
reliabilities. An omega estimator that converges to a fixed function of
\(\Sigma\) can therefore be positively biased, unbiased, or negatively biased
depending only on which equivalent latent decomposition is called the population
model.

\section{A Worked Bell-Style Bifactor Example}

Consider six standardized items arranged in two three-item clusters, mimicking
the bifactor structure used by Bell et al. Let every item load .50 on a general
factor and .40 on one of two group factors, with all factors orthogonal and all
item residuals independent. The residual variance for each item is
\[
  1-.50^2-.40^2=.59 .
\]
The implied item covariance matrix has diagonal entries \(1\), within-cluster
off-diagonal entries
\[
  .50^2+.40^2=.41 ,
\]
and between-cluster off-diagonal entries
\[
  .50^2=.25 .
\]
For the sum score \(X=\one\T y\), the total variance is
\[
  \Var(X)
  =
  6+2\{6(.41)+9(.25)\}
  =
  15.42 .
\]
The general factor contributes
\[
  (\one\T\lambda_g)^2=(6\cdot .50)^2=9 ,
\]
so the Bell-style bifactor target is
\[
  \omega_H=\frac{9}{15.42}=.584 .
\]

Now hold the item covariance matrix fixed. The same covariance matrix can also
be written as a one-factor correlated-error model
\[
  \Sigma=\lambda_r\lambda_r\T+\Psi_r
\]
with a different target loading vector. If the desired sum-score reliability is
\(r=.75\), the construction above gives equal loadings
\[
  \lambda_r=.566789\,\one .
\]
The residual covariance matrix then has diagonal entries .678750,
within-cluster off-diagonal entries .088750, and between-cluster off-diagonal
entries \(-.071250\). These residual covariances reproduce exactly the same
item covariance matrix, but the sum-score reliability with respect to the
common factor is
\[
  \frac{(6\cdot .566789)^2}{15.42}=.75 .
\]
If instead \(r=.40\), the equal common-factor loading is .413924 and the
residual covariance entries are .828667 on the diagonal, .238667 within
clusters, and .078667 between clusters. The same observed covariance matrix now
has target-factor sum-score reliability .40.

The point is not that these alternative correlated-error decompositions are
substantively preferable to the bifactor model. The point is that they are
observationally equivalent at the covariance level and define different
sum-score reliability targets. In the bifactor interpretation the population
omega is .584; in the two one-factor local-dependence interpretations it is .75
or .40. The denominator, \(\Var(X)=15.42\), has not changed. Only the numerator,
the part of the sum-score variance assigned to the target factor, has changed.

\section{Re-reading Bell et al.'s Simulation Results}

The example clarifies what Bell et al.'s \cite{bell2024} simulations do and do
not show. Their population models define clear conditional targets. For a
one-factor correlated-error population, the target is the part of sum-score
variance due to the one common factor. For a bifactor population, the target is
the part due to the general factor. For a higher-order population, the target is
the part due to the higher-order factor through its indirect effects.

The limitation is that these are target-relative claims. In a bifactor
population,
\[
  \Sigma=\lambda_g\lambda_g\T+\Lambda_s\Lambda_s\T+\Theta ,
\]
the Bell target treats \(\lambda_g\lambda_g\T\) as construct-relevant
general-factor variance and treats \(\Lambda_s\Lambda_s\T+\Theta\) as
non-target variance for the total score. Marginally, however,
\(\Lambda_s\Lambda_s\T+\Theta\) is structured residual covariance around the
general factor. Another one-factor correlated-residual decomposition can imply
the same \(\Sigma\) but assign a different amount of sum-score variance to its
common factor. The observed item covariance matrix cannot decide which target
is substantively correct.

The higher-order case illustrates an important qualification. A higher-order
model and a restricted bifactor model can be related by the Schmid-Leiman
transformation \cite{schmid1957,yung1999}. When that transformation preserves
the vector of general-factor contributions to the items, omega is preserved too.
Equivalent models are not automatically a problem for sum-score reliability.
They become a problem when the equivalent representations change the target
loading vector in Equation~\eqref{eq:omega}.

Thus, simulations of misspecified omega estimates are best read as answering a
conditional question:
\[
  \text{If this latent target story is true, how does this fitted omega behave?}
\]
They do not answer the unconditional question:
\[
  \text{How biased is omega for the reliability encoded in the covariance matrix?}
\]
There is no such covariance-only sum-score reliability target without further
restrictions.

\section{Implications}

The practical implication is not that omega should be abandoned, or that
misspecification simulations are useless. The implication is that simulations
should separate two problems that are often blended together.

First, there is an estimator problem. Given a target model, a target factor, and
a population covariance matrix, what does a fitted omega estimate in finite
samples or asymptotically? Bell et al. address this problem.

Second, there is a target problem. Given the same observed covariance matrix,
which part of the sum-score variance should count as construct-relevant
reliability? This problem is not solved by Monte Carlo replication, larger
sample size, or better fit indices. It requires theory, design information,
external validation, or explicit sensitivity analysis over plausible equivalent
or near-equivalent decompositions.

Future simulation studies can make this distinction explicit by reporting their
results as conditional on a declared target factor, by noting whether the target
is invariant across relevant equivalent models, and by including sensitivity
conditions that move covariance between general factors, specific factors, and
correlated residuals while holding the item covariance matrix fixed or nearly
fixed. Such designs would not merely ask whether an estimator is biased under
misspecification. They would ask whether the proposed sum-score reliability
target itself is stable under the kinds of model uncertainty that applied
researchers face.

\section{Conclusion}

The established lesson from equivalent-model work in SEM is that the same
covariance matrix can support different substantive explanations. For
model-based composite reliability coefficients, the consequence is stronger:
different explanations can define different reliabilities for the same sum
score. Bias under misspecification is therefore not a model-free property of
coefficient omega. It is a property of an estimator relative to a declared
latent target. That framing preserves the value of recent omega simulations
while clarifying their scope. They are conditional sensitivity studies, not
definitive evidence about bias for a uniquely identified reliability of the
observed composite.

\begin{thebibliography}{9}

\bibitem{bell2024}
Bell, S. M., Chalmers, R. P., \& Flora, D. B. (2024).
The impact of measurement model misspecification on coefficient omega estimates
of composite reliability.
\emph{Educational and Psychological Measurement}, 84(1), 5--39.
\href{https://doi.org/10.1177/00131644231155804}{doi:10.1177/00131644231155804}.

\bibitem{bollen1989}
Bollen, K. A. (1989).
\emph{Structural equations with latent variables}.
Wiley.

\bibitem{borsboom2002}
Borsboom, D., \& Mellenbergh, G. J. (2002).
True scores, latent variables, and constructs: A comment on Schmidt and Hunter.
\emph{Intelligence}, 30(6), 505--514.
\href{https://doi.org/10.1016/S0160-2896(02)00082-X}{doi:10.1016/S0160-2896(02)00082-X}.

\bibitem{ellis2021}
Ellis, J. L. (2021).
A test can have multiple reliabilities.
\emph{Psychometrika}, 86(4), 869--876.
\href{https://doi.org/10.1007/s11336-021-09800-2}{doi:10.1007/s11336-021-09800-2}.

\bibitem{maccallum1993}
MacCallum, R. C., Wegener, D. T., Uchino, B. N., \& Fabrigar, L. R. (1993).
The problem of equivalent models in applications of covariance structure
analysis.
\emph{Psychological Bulletin}, 114(1), 185--199.
\href{https://doi.org/10.1037/0033-2909.114.1.185}{doi:10.1037/0033-2909.114.1.185}.

\bibitem{savalei2019}
Savalei, V., \& Reise, S. P. (2019).
Don't forget the model in your model-based reliability coefficients: A reply to
McNeish (2018).
\emph{Collabra: Psychology}, 5(1), Article 36.
\href{https://doi.org/10.1525/collabra.247}{doi:10.1525/collabra.247}.

\bibitem{schmid1957}
Schmid, J., \& Leiman, J. M. (1957).
The development of hierarchical factor solutions.
\emph{Psychometrika}, 22(1), 83--90.
\href{https://doi.org/10.1007/BF02289209}{doi:10.1007/BF02289209}.

\bibitem{yung1999}
Yung, Y. F., Thissen, D., \& McLeod, L. D. (1999).
On the relationship between the higher-order factor model and the hierarchical
factor model.
\emph{Psychometrika}, 64(2), 113--128.
\href{https://doi.org/10.1007/BF02294531}{doi:10.1007/BF02294531}.

\end{thebibliography}

% Development note: before submission, add a fuller review of equivalent-model
% results specific to CFA, local dependence, bifactor models, and reliability
% under multidimensional or method-factor interpretations.

\end{document}
