\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}

\title{Model-Relative Bias of Omega Under Equivalent Measurement Models}
\author{}
\date{\today}

\newcommand{\E}{\mathbb{E}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\vech}{\operatorname{vech}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\argmax}{\operatorname*{arg\,max}}
\newcommand{\plim}{\operatorname*{plim}}
\newcommand{\T}{^{\mathsf T}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\begin{document}
\maketitle

\section*{Purpose}

Bell, Chalmers, and Flora \cite{bell2024} define population reliability as the
omega parameter attached to the model that generated the covariance matrix.
They then call an estimated omega biased when its expectation differs from that
model-relative target. That is internally coherent.

The problem is that the target is not, in general, a functional of the observed
item distribution. Equivalent latent decompositions can imply the same
covariance matrix while assigning different amounts of composite variance to the
named target factor. In that case the same omega estimate has different
population bias under different, observationally indistinguishable, ``true''
models. This note gives a simple theorem and a numerical example where the bias
sign changes without changing the observable covariance matrix.

\section{Model-Relative Omega}

Let \(y\in\mathbb{R}^p\) be the item vector and let \(X=w\T y\) be the composite,
usually with \(w=\mathbf{1}\). Write \(\Sigma=\Var(y)\). If a target factor \(f\)
with \(\Var(f)=1\) contributes the item loading vector \(a\) in a decomposition
\[
  y=a f+u,\qquad \Cov(f,u)=0,\qquad \Var(u)=\Psi,
\]
then the model-relative omega target for the composite \(X\), with respect to
the named factor \(f\), is
\begin{equation}\label{eq:omega}
  \omega(a,\Sigma;w)
  =
  \frac{\Var(w\T a f)}{\Var(w\T y)}
  =
  \frac{(w\T a)^2}{w\T\Sigma w}
  =
  \frac{(w\T a)^2}{(w\T a)^2+w\T\Psi w}.
\end{equation}
The denominator is the full composite variance, including residual covariances.
It is not obtained by replacing \(\Psi\) with \(\operatorname{diag}(\Psi)\),
unless the residual covariance matrix is actually diagonal.

For a one-factor model, \(a\) is the loading vector. For a bifactor model, \(a\)
is the general-factor loading vector. For a higher-order model, \(a\) is the
vector of indirect effects from the higher-order factor to the items. The
Bell et al. bias calculation is therefore
\[
  \E(\hat\omega)-\omega(a_{\mathrm{true}},\Sigma;w),
\]
where \(a_{\mathrm{true}}\) is supplied by the chosen population model.

The word ``chosen'' matters. The covariance matrix alone need not determine
\(a_{\mathrm{true}}\).

\section{A Non-Identification Result}

\begin{theorem}[Equivalent models with arbitrary composite omega]
\label{thm:anyomega}
Let \(\Sigma\) be positive definite and let \(w\ne 0\). For every
\(r\in[0,1]\), define
\begin{equation}\label{eq:ar}
  a_r
  =
  \sqrt{r}\,
  \frac{\Sigma w}{\sqrt{w\T\Sigma w}},
  \qquad
  \Psi_r
  =
  \Sigma-a_r a_r\T .
\end{equation}
Then \(\Psi_r\succeq0\), the one-factor representation
\[
  y=a_r f_r+u_r,\qquad
  \Var(f_r)=1,\qquad
  \Cov(f_r,u_r)=0,\qquad
  \Var(u_r)=\Psi_r
\]
implies the same covariance matrix \(\Sigma\). In that representation, the omega
of the composite \(X=w\T y\) with respect to the common factor \(f_r\) is
\[
  \omega(a_r,\Sigma;w)=r .
\]
If all entries of \(\Sigma w\) are nonzero, the target factor loads on every
item for every \(r>0\).
\end{theorem}

\begin{proof}
The covariance equality is immediate:
\[
  \Var(a_r f_r+u_r)=a_r a_r\T+\Psi_r=\Sigma .
\]
For any vector \(x\), Cauchy's inequality in the inner product
\(\langle x,z\rangle_\Sigma=x\T\Sigma z\) gives
\[
  (x\T\Sigma w)^2
  \le
  (x\T\Sigma x)(w\T\Sigma w).
\]
Therefore
\[
  x\T\Psi_r x
  =
  x\T\Sigma x
  -
  r\frac{(x\T\Sigma w)^2}{w\T\Sigma w}
  \ge
  (1-r)x\T\Sigma x
  \ge 0 ,
\]
so \(\Psi_r\succeq0\). Finally,
\[
  (w\T a_r)^2
  =
  r\,w\T\Sigma w ,
\]
which proves \(\omega(a_r,\Sigma;w)=r\).
\end{proof}

\begin{corollary}[Bias sign is not identified]
\label{cor:bias}
Let an omega estimator converge under the observed covariance \(\Sigma\) to
some value \(T(\Sigma)\in(0,1)\). In the equivalent model family from
Theorem~\ref{thm:anyomega}, its asymptotic bias is
\[
  T(\Sigma)-r .
\]
Thus, without changing the observed covariance matrix, one can choose an
equivalent population model under which the estimator is positively biased,
unbiased, or negatively biased.
\end{corollary}

The corollary is the clean version of the objection. Bias is identifiable only
after fixing a target factor by theory or by a restrictive model class that makes
the target unique. It is not identified by the covariance matrix itself.

\section{Numerical Example}

Take \(p=8\), \(w=\mathbf{1}\), and the compound-symmetry correlation matrix
\[
  \Sigma_\rho=(1-\rho)I+\rho\mathbf{1}\mathbf{1}\T,
  \qquad
  \rho=.30 .
\]
The independent-error one-factor decomposition has
\[
  a_0=\sqrt{\rho}\,\mathbf{1},
  \qquad
  \Psi_0=(1-\rho)I .
\]
Its omega is
\[
  \omega_0
  =
  \frac{p^2\rho}{p+p(p-1)\rho}
  =
  0.774194 .
\]
A correctly fitted one-factor omega also converges to \(T(\Sigma_\rho)=0.774194\).

But the same \(\Sigma_\rho\) has the following equivalent one-factor
representations with correlated residuals:
\[
\begin{array}{c|c|c|c|c}
  r & a_r\text{ entry} & \Psi_r\text{ diagonal} &
  \Psi_r\text{ off diagonal} & T(\Sigma_\rho)-r\\
  \hline
  .600000 & .482183 & .767500 & .067500 & .174194\\
  .774194 & .547723 & .700000 & .000000 & .000000\\
  .850000 & .573912 & .670625 & -.029375 & -.075806
\end{array}
\]
All three rows generate exactly the same item covariance matrix. The first says
the same fitted omega is positively biased by about .17. The second says it is
unbiased. The third says it is negatively biased by about .08. The only thing
that changed is the latent story used to split \(\Sigma_\rho\) into target-factor
variance and residual covariance.

\section{Connection to Bell et al.'s Population Families}

Bell et al. use four population families. With \(d=w\T\Sigma w\), their target
vector \(a\) and target omega can be written as follows:
\[
\begin{aligned}
  \text{simple one factor:}\quad
  &\Sigma=\lambda\lambda\T+\Theta_D,
  &\omega_u&=(w\T\lambda)^2/d,\\
  \text{one factor with correlated errors:}\quad
  &\Sigma=\lambda\lambda\T+\Psi_G,
  &\omega_u&=(w\T\lambda)^2/d,\\
  \text{bifactor:}\quad
  &\Sigma=g g\T+SS\T+\Theta_D,
  &\omega_H&=(w\T g)^2/d,\\
  \text{higher order:}\quad
  &\Sigma=(\Lambda\gamma)(\Lambda\gamma)\T
    +\Lambda\Psi_z\Lambda\T+\Theta_D,
  &\omega_{ho}&=\{w\T(\Lambda\gamma)\}^2/d .
\end{aligned}
\]
Here \(\Theta_D\) is diagonal residual variance, \(\Psi_G\) is a residual
covariance matrix with Bell et al.'s selected local-dependence graph, \(S\) holds
the bifactor specific-factor loadings, and \(\Lambda\gamma\) is the vector of
indirect effects from the higher-order factor to the items.

This display makes the target-dependence visible.

\begin{corollary}[Bell's multidimensional covariances have one-factor
correlated-error alternatives]
\label{cor:bell}
Let \(\Sigma\) be any covariance matrix generated by Bell et al.'s correlated
error, bifactor, or higher-order population models. For every \(r\in[0,1]\),
there is a one-factor model with correlated residuals, a common factor loading
on every item whenever \(\Sigma w\) has no zero entry, and the same observed
\(\Sigma\), for which the omega of \(X=w\T y\) with respect to that model's
common factor equals \(r\):
\[
  \frac{\Var(w\T a_r f_r)}{\Var(w\T y)}
  =
  \frac{(w\T a_r)^2}{w\T(a_r a_r\T+\Psi_r)w}
  =
  r .
\]
\end{corollary}

\begin{proof}
Apply Theorem~\ref{thm:anyomega} to the covariance matrix generated by the Bell
population model.
\end{proof}

Corollary~\ref{cor:bell} is not a claim that Bell et al. secretly fit this
alternative model. It is a claim about what the observed covariance identifies.
A bifactor population covariance
\[
  \Sigma=g g\T+SS\T+\Theta_D
\]
is also a one-factor correlated-residual covariance
\[
  \Sigma=a_r a_r\T+\Psi_r
\]
for every target reliability \(r\). The same is true of a higher-order
population after writing its target contribution as \(a=\Lambda\gamma\). Thus
the Bell population labels \(.60\) and \(.85\) are not labels for the covariance
matrix alone. They are labels for a particular latent split of that covariance.

The four Bell families are therefore affected differently.

\begin{enumerate}
  \item In the simple one-factor model with independent errors, the target is
  comparatively well anchored. With at least three indicators and nonzero,
  sign-consistent off-diagonal covariances, \(\lambda\lambda\T\) is identified
  from the off-diagonal part of \(\Sigma\), up to a global sign. Since
  \((w\T\lambda)^2\) is invariant to that sign, \(\omega_u\) is identified
  inside this restricted model class. The equivalence objection does not attack
  Bell et al.'s correctly specified simple one-factor cells.

  \item In the one-factor correlated-error model, the target is conditional on
  the selected residual covariance graph and on the decision that those
  residual covariances, rather than the common factor, carry the extra
  covariance. If that graph is treated as fixed and the model is identified,
  Bell et al.'s target is again well-defined conditionally. If the graph is part
  of the uncertainty, or if the analyst is only committed to ``one common factor
  plus local dependence,'' Theorem~\ref{thm:anyomega} applies directly. The same
  covariance can be assigned a lower or higher target omega by moving covariance
  between \(\lambda\lambda\T\) and \(\Psi_G\).

  \item In the bifactor model, Bell et al.'s target is the contribution of the
  direct general factor \(g\). Marginally, however, the specific factors
  \(SS\T\) are just structured residual covariance relative to the general
  factor. A covariance generated by \(g g\T+SS\T+\Theta_D\) can be re-expressed
  as a one-factor correlated-error model with any target omega \(r\). Thus a
  statement such as ``\(\hat\omega_u\) is biased relative to the true bifactor
  omega'' is best read as ``biased relative to the declared bifactor target,''
  not as a covariance-only fact.

  \item In the higher-order model, the situation is less damaging for Bell et
  al.'s comparison between \(\omega_H\) and \(\omega_{ho}\). A Schmid-Leiman
  restricted bifactor representation can preserve the higher-order target vector
  \(a=\Lambda\gamma\). When it does, the numerator in \eqref{eq:omega} is the
  same and omega is target-preserving across those two representations. The
  equivalence objection returns only when the fitted or alternative bifactor
  representation changes the general-factor contribution vector.
\end{enumerate}

This also sharpens the role of model fit in Bell et al.'s RQ4. Exact equivalent
models have the same fit to \(\Sigma\). Fit indices can help reject visibly bad
approximations, but they cannot choose among target-changing equivalent
decompositions.

\section{What This Does and Does Not Refute}

The theorem does not show that Bell et al. made a computational mistake. Once
their population model and target factor are stipulated, their Monte Carlo bias
quantity is a legitimate model-relative quantity.

It does show that the broader language ``omega is biased for reliability under
misspecification'' needs a qualification. If reliability means the proportion of
composite variance attributable to a substantively named factor, then that
factor is part of the estimand. Equivalent measurement models that change the
factor contribution vector \(a\) change the estimand while leaving the data
distribution fixed. No sample size or fit index can recover which bias statement
is the correct one from \(\Sigma\) alone.

There is also an important non-objection. Some equivalent representations
preserve the target contribution vector. For example, the Schmid-Leiman
transformation relates restricted bifactor and higher-order models
\cite{schmid1957,yung1999}. When the bifactor general-factor loadings equal the
higher-order indirect effects, the numerator in \eqref{eq:omega} is the same,
so omega is invariant. Equivalence is damaging only when it is not
target-preserving.

\section{Implication for Simulation Design}

A simulation can still answer a conditional question:
\[
  \text{If this latent model is the true target story, what does this fitted
  omega estimate?}
\]
It cannot, by itself, answer an unconditional question:
\[
  \text{How biased is omega for the reliability encoded in the observed
  covariance matrix?}
\]
There is no such covariance-only reliability target unless additional
identifying restrictions are imposed.

For future reliability simulations, the defensible formulation is therefore:
\begin{quote}
  Bias is relative to a declared target model. Before interpreting it as a
  failure of an estimator, show that the target is invariant across the relevant
  equivalent models, or explicitly treat the result as a sensitivity analysis
  over plausible target decompositions.
\end{quote}

Under that standard, the user's objection is right in the strong sense: there
are equivalent ``true'' models under which the same fitted omega has smaller,
larger, zero, or opposite-signed bias. The objection does not make Bell et al.'s
conditional Monte Carlo invalid, but it blocks a covariance-only or
model-free reading of their bias claims.

\begin{thebibliography}{9}

\bibitem{bell2024}
Bell, S. M., Chalmers, R. P., \& Flora, D. B. (2024).
The impact of measurement model misspecification on coefficient omega estimates
of composite reliability.
\emph{Educational and Psychological Measurement}, 84(1), 5--39.
\href{https://doi.org/10.1177/00131644231155804}{doi:10.1177/00131644231155804}.

\bibitem{schmid1957}
Schmid, J., \& Leiman, J. M. (1957).
The development of hierarchical factor solutions.
\emph{Psychometrika}, 22(1), 83--90.
\href{https://doi.org/10.1007/BF02289209}{doi:10.1007/BF02289209}.

\bibitem{yung1999}
Yung, Y. F., Thissen, D., \& McLeod, L. D. (1999).
On the relationship between the higher-order factor model and the hierarchical
factor model.
\emph{Psychometrika}, 64(2), 113--128.
\href{https://doi.org/10.1007/BF02294531}{doi:10.1007/BF02294531}.

\end{thebibliography}

\end{document}
