\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,booktabs}
\usepackage{hyperref}

\title{A First Categorical DWLS Profile-Hessian Check}
\author{}
\date{\today}

\newcommand{\tr}{\operatorname{tr}}

\begin{document}
\maketitle

\section*{Purpose}

This note records a first diagnostic exploration of the regular profile-Hessian
roadmap in a genuinely categorical DWLS setting. The goal is not a finite-sample
simulation yet. The goal is to check whether the categorical DWLS weight channel
can visibly change the nested-test reference law once we leave exact fit.

The companion script is
\[
  \texttt{docs/research/notes/ordinal\_dwls\_profile\_check.R}.
\]

\section{Setup}

The population is a four-variable binary ordinal latent-response model. The
first-stage object is
\[
  x=(u,\gamma),
\]
where $u$ stacks the four thresholds and six tetrachoric correlations, and
\[
  \gamma=\operatorname{diag}(\Gamma_u)
\]
is the diagonal of the asymptotic covariance of $u$. DWLS uses
$W=\operatorname{diag}(\gamma)^{-1}$. The script computes $u$, $\gamma$, and the
joint covariance of $(u,\gamma)$ by finite-difference delta method from the
$2^4$ binary cell probabilities.

The nested pair is:
\[
  A:\ \text{congeneric one-factor latent-response correlations},
  \qquad
  B:\ \text{tau-equivalent one-factor latent-response correlations}.
\]
Thresholds are free in both models. Thus the nested contrast tests equality of
the loading structure, not threshold equality.

For the nested contrast $B$ versus $A$, the script compares three quadratic
reference laws:
\[
\begin{array}{ll}
U_{\mathrm{GN}}:     & \text{Gauss--Newton fixed-weight projector, using only }u,\\
Q_{\mathrm{fixed}}: & \text{fixed-weight profile Hessian in }u,\\
Q_{\mathrm{full}}:  & \text{full profile Hessian in }x=(u,\gamma).
\end{array}
\]
The last object is the categorical DWLS version of the general $Q_B-Q_A$
roadmap object.

\section{Results}

\begin{center}
\begin{tabular}{lrrrr}
\toprule
Scenario & $\sqrt{2f_A}$ & $\sqrt{2f_B}$ &
$\tr(Q_{\mathrm{fixed}}\Gamma_u)$ &
$\tr(Q_{\mathrm{full}}\Gamma_x)$ \\
\midrule
Exact nested null & 0.0000 & 0.0000 & 2.8925 & 2.8924 \\
Near-null misspec., $\epsilon=0.04$ & 0.0499 & 0.0499 & 2.9109 & 2.9391 \\
Near-null misspec., $\epsilon=0.08$ & 0.1004 & 0.1004 & 2.9708 & 3.1582 \\
Restriction false, congeneric & 0.0000 & 0.1260 & 2.9116 & 3.0915 \\
\bottomrule
\end{tabular}
\end{center}

The corresponding full/fixed trace ratios were
\[
  1.0000,\quad 1.0097,\quad 1.0631,\quad 1.0618 .
\]
The exact nested null behaves as expected: the residual is essentially zero, so
the DWLS weight channel is dormant and the full profile law collapses onto the
fixed-weight law. Under fixed misspecification, the full DWLS law separates from
the fixed-weight law. In the two near-null misspecification directions, the
trace increases by about $1.0\%$ and $6.3\%$.

The eigenvalue movement is more informative than the trace alone. For
$\epsilon=0.08$, the leading eigenvalues changed from
\[
  (1.2947,\ 0.9386,\ 0.7376)
\]
under the fixed-weight profile Hessian to
\[
  (1.5551,\ 0.9324,\ 0.6709)
\]
under the full DWLS profile Hessian. So the weight channel does not merely
rescale the law; it reshapes it.

\section{An Exact Symmetry Pseudo-Null}

The unequal-threshold rows above are useful diagnostics, but they are only
near-null examples. A cleaner categorical pseudo-null can be obtained by
symmetry. Set all binary thresholds to zero and start from the tau-equivalent
latent-response correlation matrix with loading $\lambda=0.55$, so every
off-diagonal correlation is $\lambda^2=0.3025$. Then add local dependence on the
four edges of a cycle:
\[
  R_{12}=R_{14}=R_{23}=R_{34}=0.3025+\epsilon,
  \qquad
  R_{13}=R_{24}=0.3025.
\]
All off-diagonal correlations remain nonzero, but the population has two pair
orbits: adjacent pairs and opposite pairs. The four-cycle is vertex-transitive.
With equal thresholds, this symmetry forces the congeneric pseudo-true loadings
and thresholds to be equal, so the larger model's pseudo-true point lies in the
tau-equivalent submodel even though both models are misspecified when
$\epsilon>0$.

Numerically, across the grid below the value difference satisfies
\(|f_B-f_A|<6\times10^{-14}\) and the profile-score difference is below
\(2.1\times10^{-7}\). The ordinary nested reference has three degrees of
freedom, with \(q_{0.95}(\chi^2_3)=7.8147\).

\begin{center}
\begin{tabular}{rrrrrr}
\toprule
$\epsilon$ & $\sqrt{2f_A}$ &
$\tr(Q_{\mathrm{fixed}}\Gamma_u)$ &
$\tr(Q_{\mathrm{full}}\Gamma_x)$ &
$q^{\mathrm{full}}_{0.95}$ &
$\Pr_{\mathrm{full}}\{T>q_{0.95}(\chi^2_3)\}$ \\
\midrule
0.02 & 0.0158 & 2.8544 & 2.8577 & 7.449 & 0.042\\
0.04 & 0.0317 & 2.8475 & 2.8606 & 7.471 & 0.043\\
0.06 & 0.0477 & 2.8408 & 2.8510 & 7.476 & 0.043\\
0.08 & 0.0638 & 2.8331 & 2.8842 & 7.585 & 0.045\\
0.10 & 0.0801 & 2.8234 & 2.8959 & 7.655 & 0.047\\
\bottomrule
\end{tabular}
\end{center}

The effect is smaller than in the unequal-threshold near-null direction, but it
is cleaner: the pseudo-null is protected by symmetry, the residual is nonzero,
and the $\gamma$ block still changes the full profile law. At
$\epsilon=0.10$, the full DWLS eigenvalues are approximately
\[
  (1.1151,\ 1.1151,\ 0.6656),
\]
whereas the fixed-weight trace is still below three. The ordinary
\(\chi^2_3\) cutoff is mildly conservative here rather than catastrophic. That
is a good sign for this as a realistic simulation design: the point is not to
manufacture a giant failure, but to quantify how much the estimated categorical
weight matters at plausible levels of limited-information misfit.

\section{Interpretation}

This is consistent with the general theory. The DWLS profile score has two
blocks:
\[
  s_u=-Wr,
  \qquad
  (s_\gamma)_k=-\frac12\,\frac{r_k^2}{\gamma_k^2}.
\]
The $\gamma$ block is residual-driven. It vanishes at exact fit and becomes a
leading component under fixed misspecification. This is exactly why categorical
DWLS is a strong applied example for the paper: the most common ordinal SEM
workflow uses an estimated diagonal weight, and that weight is asymptotically
irrelevant only in the exact-fit regime.

\section{Caveats}

The first near-null misspecification scenarios were constructed to keep the
congeneric and tau-equivalent fits nearly tied, but they are not exact
pseudo-null designs. The four-cycle construction fixes that problem for a
single-group binary CFA example. It is still only a population calculation:
finite-sample rejection rates have not yet been simulated.

The script is intentionally small and finite-difference based. It is useful for
orientation, not for final numerical claims. A production version should reuse
the analytic ordinal moment influence and DWLS weight machinery in
\texttt{magmaan}.

\section{Next Steps}

\begin{enumerate}
\item Run finite-sample simulations for the four-cycle pseudo-null, comparing
  ordinary $\chi^2$, fixed-weight
  $Q_{\mathrm{fixed}}$, and full DWLS $Q_{\mathrm{full}}$ references.
\item Search for larger but still plausible symmetry-protected categorical
  designs by varying the number of indicators, thresholds, categories, and local
  dependence graph.
\item Repeat for multigroup threshold/loading invariance, where categorical
  DWLS is most common in practice.
\item Port the profile-Hessian calculation to \texttt{magmaan} using the
  existing ordinal moment and weight caches.
\end{enumerate}

\end{document}
