\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}

\title{The pairwise-deletion weight matrix for WLSMV with ordinal
missing data}
\author{}
\date{\today}

\newcommand{\E}{\mathbb{E}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\plim}{\operatorname*{plim}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\pw}{\mathrm{pw}}
\newcommand{\Ghat}{\widehat{\Gamma}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\begin{document}
\maketitle

\section*{Purpose}

WLSMV is limited information: its sample statistics are the univariate
thresholds and bivariate polychoric correlations, so it can absorb missing
data by computing each statistic from its available cases (pairwise deletion,
PD). The point estimates are then standard; the open question is the weight
matrix. What is the asymptotic covariance $\Gamma^{\pw}$ of the PD statistic
vector, and how does it differ from the complete-data $\Gamma$ that the
conventional ``WLSMV with PD'' plugs in?

\textbf{Answer.} $\Gamma^{\pw}$ is the complete-data $\Gamma$ reweighted by
observation-overlap proportions. Writing $\pi_a$ for the probability that a
case observes every item needed for statistic $a$ and $\pi_{ab}$ for the
probability it observes the items needed for both $a$ and $b$,
\[
  \Gamma^{\pw}_{ab}
  \;=\;
  \frac{1}{\pi_a \pi_b}\,
  \E\!\left[\one\{S_a\cup S_b \subseteq R\}\,\psi^a\psi^b\right],
  \qquad\text{(Theorem~\ref{thm:gpw})}
\]
which under MCAR collapses to the Hadamard form
$\Gamma^{\pw}_{ab} = (\pi_{ab}/\pi_a\pi_b)\,\Gamma_{ab}$, with diagonal
$\Gamma^{\pw}_{aa} = \Gamma_{aa}/\pi_a$. This is the ordinal sibling of the
continuous covariance rule $\Gamma^{\mu,\pw}_{jk}=\sigma_{jk}\,\pi_{jk}/\pi_j\pi_k$
already used on the pairwise-GLS path. The conventional implementation omits
the reweighting (it uses one nominal $N$), so it understates the variance of
statistics estimated from fewer cases; that is the documented source of the
inflated $\Delta\chi^2$ \cite{chen2020}.

\section*{Setup and notation}

Let there be $p$ ordinal items, item $j$ with $C_j$ categories and thresholds
$\tau_j\in\mathbb{R}^{C_j-1}$. The WLSMV statistic vector stacks all thresholds
and all polychorics,
\[
  s = (\tau', \rho')',
  \qquad
  \rho = (\rho_{jk})_{j<k},
  \qquad
  p^\ast = \sum_j (C_j-1) + \binom{p}{2}.
\]
Index a generic component by $a\in\{1,\dots,p^\ast\}$ and let
$S_a\subseteq\{1,\dots,p\}$ be its \emph{support}: $|S_a|=1$ for a threshold
of item $j$ ($S_a=\{j\}$), $|S_a|=2$ for a polychoric ($S_a=\{j,k\}$).

Complete data are i.i.d.\ $z_i=(y_{i1},\dots,y_{ip})$. The staged estimator
(univariate thresholds, then bivariate polychorics with thresholds held fixed
\cite{muthen1984,joreskog1994}) is an M-estimator solving
$\sum_i g(z_i;\hat s)=0$, with bread $A=-\E[\partial g/\partial s']$ and casewise
influence
\[
  \psi_i = A^{-1} g(z_i;s_0),
  \qquad \E[\psi_i]=0,
  \qquad \Gamma=\E[\psi_i\psi_i'] .
\]
magmaan stores the influence rows $IF_i=-\psi_i A^{-\top}$ and forms
$\Ghat=\tfrac1N\sum_i IF_i IF_i'$ (cf.\ \path{robust_ordinal_gamma.tex}).

Missingness is encoded by the response set $R_i=\{j:y_{ij}\text{ observed}\}$.
Statistic $a$ is estimable on case $i$ iff $S_a\subseteq R_i$; write the
availability indicator and its proportions
\[
  d^a_i=\one\{S_a\subseteq R_i\},
  \quad
  \pi_a=\E[d^a_i],
  \quad
  \pi_{ab}=\E[d^a_i d^b_i]=\Pr(S_a\cup S_b\subseteq R),
  \quad
  N_a=\sum_i d^a_i .
\]
PD solves each block on its available cases; stacking with the diagonal mask
$D_i=\operatorname{diag}(d^a_i)$, the estimator solves
$\sum_i D_i\, g(z_i;\hat s^{\pw})=0$.

\section*{Influence sparsity}

\begin{lemma}\label{lem:sparse}
The influence $\psi^a_i$ depends on $z_i$ only through
$\{y_{ij}:j\in S_a\}$. Hence $\psi^{\tau_j}_i$ is a function of $y_{ij}$ alone,
and $\psi^{\rho_{jk}}_i$ a function of $(y_{ij},y_{ik})$ alone.
\end{lemma}

\begin{proof}
The stacked score is block lower triangular: threshold scores depend on their
own item; the $\rho_{jk}$ score depends on the pair $(j,k)$ and on
$\tau_j,\tau_k$. Thus $A$ is block lower triangular and so is $A^{-1}$, giving
\[
  \psi^{\rho_{jk}}_i
  = \frac{1}{\mathcal I_{\rho_{jk}}}
    \Big[\, s_{\rho_{jk}}(y_{ij},y_{ik})
      - \mathcal I_{\rho_{jk}\tau_j}\psi^{\tau_j}_i
      - \mathcal I_{\rho_{jk}\tau_k}\psi^{\tau_k}_i \Big],
\]
a combination of objects supported on $\{j,k\}$; the threshold rows are
supported on their single item.
\end{proof}

Lemma~\ref{lem:sparse} is what makes overlap proportions the right bookkeeping:
the only cases that carry information about the pair $(a,b)$ are those that
observe $S_a\cup S_b$.

\section*{The pairwise weight matrix}

\begin{theorem}\label{thm:gpw}
Let $s_0=\plim \hat s^{\pw}$ and let $\psi^a$ be the influence about $s_0$.
Then $\sqrt N(\hat s^{\pw}-s_0)\xrightarrow{d}N(0,\Gamma^{\pw})$ with
\[
  \Gamma^{\pw}_{ab}
  = \frac{1}{\pi_a\pi_b}\,
    \E\!\left[\one\{S_a\cup S_b\subseteq R\}\,\psi^a\psi^b\right].
\]
\end{theorem}

\begin{proof}
The available-case linearization of component $a$ is
$\hat s^{\pw}_a-s_{a,0}=N_a^{-1}\sum_i d^a_i\psi^a_i+o_p(N^{-1/2})$; the mask is
mean-zero against the influence at the pseudo-true value,
$\E[d^a_i\psi^a_i]=A^{-1}\E[d^a_i g_a(z_i;s_0)]=0$, by definition of $s_0$.
Independent cross terms vanish, and with $N_a=N\pi_a+o_p(N)$,
\[
  N\,\Cov(\hat s^{\pw}_a,\hat s^{\pw}_b)
  = \frac{N}{N_aN_b}\sum_i d^a_i d^b_i \psi^a_i\psi^b_i + o_p(1)
  \xrightarrow{p}
  \frac{1}{\pi_a\pi_b}\,\E[d^a d^b\psi^a\psi^b],
\]
and $d^a_id^b_i=\one\{S_a\cup S_b\subseteq R_i\}$.
\end{proof}

\begin{corollary}[MCAR closed form]\label{cor:mcar}
If $R\perp z$ then $s_0$ is the true value and the indicator factors out:
\[
  \Gamma^{\pw}_{ab}=\frac{\pi_{ab}}{\pi_a\pi_b}\,\Gamma_{ab},
  \qquad\text{equivalently}\qquad
  \Gamma^{\pw}=\Pi_{\!\div}\odot\Gamma,
  \quad
  (\Pi_{\!\div})_{ab}=\frac{\pi_{ab}}{\pi_a\pi_b},
\]
with diagonal $\Gamma^{\pw}_{aa}=\Gamma_{aa}/\pi_a$, i.e.\
$\Var(\hat s^{\pw}_a)=\Gamma_{aa}/N_a$.
\end{corollary}

\begin{corollary}[item-independent missingness]\label{cor:indep}
If items are observed independently with $\Pr(j\in R)=\pi_j$, then
$\pi_a=\prod_{j\in S_a}\pi_j$ and the overlap factor is the product of the
inverse observation probabilities of the \emph{shared} items,
\[
  \frac{\pi_{ab}}{\pi_a\pi_b}=\prod_{j\in S_a\cap S_b}\pi_j^{-1}.
\]
In particular the factor is $1$ for statistics with disjoint supports, and the
correction touches only diagonals and entries sharing an item.
\end{corollary}

The diagonal in Corollary~\ref{cor:mcar} is the whole story for Type I error:
a statistic seen by $N_a<N$ cases has variance $\Gamma_{aa}/N_a$, not
$\Gamma_{aa}/N$. Replacing $N_a$ by $N$ understates that variance, overstates
the corresponding precision in the weight, and inflates the quadratic-form
statistic. This is exactly the WLSMV{\_}PD behaviour reported in
\cite{chen2020} (Type I rate of $\Delta\chi^2$ up to $0.55$ at $50\%$ missing).

\begin{remark}[the correction preserves positive semidefiniteness]
$\Pi=[\pi_{ab}]$ is a Gram matrix ($\pi_{ab}=\E[d^a d^b]$), hence PSD; conjugating
by $\operatorname{diag}(\pi_a^{-1})$ keeps it PSD, so $\Pi_{\!\div}$ is PSD, and by
the Schur product theorem $\Gamma^{\pw}=\Pi_{\!\div}\odot\Gamma$ is PSD. The
overlap reweighting cannot by itself create an indefinite weight; any
non-PSD-ness is inherited from $\Gamma$ and handled by the existing repair
policy (\path{robust_ordinal_gamma.tex}).
\end{remark}

\begin{remark}[exact staged sandwich vs.\ the Hadamard form]
Theorem~\ref{thm:gpw} accounts each statistic at the level of its own influence,
which assigns the entire polychoric influence the pair availability $\pi_{jk}$.
The fully exact object is the masked sandwich
\[
  \Gamma^{\pw}=\bar A^{-1} M \bar A^{-\top},
  \quad
  \bar A=\E[D_i G_i],
  \quad
  M_{ab}=\E[\one\{S_a\cup S_b\subseteq R\}\,g_a g_b'],
  \quad G_i=\tfrac{\partial g}{\partial s'},
\]
which assigns the threshold-propagation terms of $\psi^{\rho_{jk}}$ (Lemma~\ref{lem:sparse})
the larger item availabilities $\pi_j,\pi_k\ge\pi_{jk}$. The Hadamard form is
therefore exact on the threshold block and a mild upper bound on the polychoric
block; the gap is of order
$\mathcal I_{\rho\tau}\,(\pi_{jk}^{-1}-\pi_j^{-1})$ and vanishes when thresholds
are well determined. Because it overstates the polychoric variances, the
Hadamard estimator errs \emph{conservative} (it under-rejects), the safe
direction for a Type I criterion, and it is the cheap option built directly on
the existing $IF$ rows.
\end{remark}

\begin{remark}[scope: this calibrates inference, not PD's MAR bias]
$\Gamma^{\pw}$ is the sampling covariance of the PD statistic about its own
probability limit $s_0$. Available-case polychorics are consistent for the true
$s$ under MCAR but not generally under MAR, where selection on an observed,
latent-correlated item shifts $s_0$ off the target. The corrected weight then
yields a correctly calibrated test of a slightly displaced pseudo-value; it does
not restore MAR consistency, which needs full information (FIML / two-stage) or
the composite-likelihood machinery with an MAR-valid sandwich
\cite{katsikatsou2012}. Empirically the displacement was small in
\cite{chen2020} (loading bias $|\mathrm{RB}|<0.1$), so the inference failure, not
the point estimate, dominated; the corrected $\Gamma^{\pw}$ targets exactly that
failure.
\end{remark}

\section*{What the conventional implementation omits}

Plugging the PD point estimates into the complete-data $\Gamma$ formula at a
single sample size is, in the language of Corollary~\ref{cor:mcar}, the choice
$\pi_a\equiv\pi_b\equiv\pi_{ab}$, i.e.\ $\Pi_{\!\div}\equiv\one\one'$ and
$\Gamma^{\pw}\equiv\Gamma$. The non-uniform $N_a$ across thresholds (item level)
and polychorics (pair level), and the missing-data covariance between them, are
discarded. Two consequences follow, both noted in \cite{chen2020}: the
per-statistic variances are wrong (the diagonal effect above), and the
correlations among statistics estimated on different overlapping subsamples are
wrong (the off-diagonal effect). Neither the mean nor the mean-and-variance
$\chi^2$ correction can compensate, because both read their scaling from this
same matrix.

\section*{Estimation}

The leading-order estimator is the existing influence machinery with overlap
masks. For statistics $a,b$ let $\widehat\psi^a_i$ be the centered PD influence
rows; then
\[
  \Ghat^{\pw}_{ab}
  = \frac{N}{N_a N_b}\sum_{i:\,S_a\cup S_b\subseteq R_i}
      \widehat\psi^a_i\,\widehat\psi^b_i .
\]
This is $\Ghat=IF'IF/N$ with the sum restricted to the overlap set and the
$1/N$ replaced by $N/N_aN_b$; under MCAR it converges to
Corollary~\ref{cor:mcar}, under MAR to Theorem~\ref{thm:gpw}. Implementation is
pattern-grouped, matching the continuous pairwise path: partition cases by
distinct $R_i$, accumulate $\widehat\psi^a\widehat\psi^b$ over patterns whose set
contains $S_a\cup S_b$, and divide once at the end. The two-level structure is
automatic from the supports: $\{j\}$ for thresholds feeds the item proportions
$\pi_j$, $\{j,k\}$ for polychorics feeds the pair proportions $\pi_{jk}$, and the
mixed block uses $\pi_{\{j,k,l\}}$. The exact sandwich (remark above) replaces the
own-influence rows by the masked score-and-bread accumulation
$\bar A^{-1}M\bar A^{-\top}$ when the conservative bias is judged too large.

\section*{Downstream: test statistic and FMG spectrum}

Everything past the weight matrix is the single substitution
$\Gamma\mapsto\Gamma^{\pw}$. Fit DWLS with weight $V=W_D^{-1}$,
$W_D=\operatorname{diag}(\Gamma^{\pw})$; the residual
$\hat e=\hat s-\sigma(\hat\theta)$ gives $T=N\,\hat e'V\hat e$, and
\[
  T \xrightarrow{d} \sum_{a=1}^{df}\lambda_a\,\chi^2_{1,a},
  \qquad
  \{\lambda_a\}=\operatorname{eig}(U\Gamma^{\pw}),
  \qquad
  U = V - V\Delta(\Delta'V\Delta)^{-1}\Delta'V,
\]
with $\Delta=\partial\sigma/\partial\theta$. The mean correction (WLSM) is
$\hat c=\tr(U\Gamma^{\pw})/df$; the mean-and-variance correction (WLSMV) matches
$\tr(U\Gamma^{\pw})$ and $\tr((U\Gamma^{\pw})^2)$ \cite{satorrabentler2001}; and
the FMG / pEBA test uses the full penalized spectrum of $U\Gamma^{\pw}$ rather
than its first two moments \cite{fmg2025}. The nested $\Delta\chi^2$ for
measurement invariance inherits the same substitution. The default's error is
precisely $U\Gamma$ in place of $U\Gamma^{\pw}$; with the corrected meat the
WLSMV missing-data statistic is the analogue of Savalei and Bentler's
statistically justified pairwise method for continuous nonnormal incomplete data
\cite{savaleibentler2005}, and an overlap-weighted $\Gamma^{\pw}$ is exactly what
an ordinal FMG difference test under missingness would consume.

\begin{thebibliography}{9}

\bibitem{muthen1984} Muth\'en, B. (1984). A general structural equation model
with dichotomous, ordered categorical, and continuous latent variable
indicators. \emph{Psychometrika}, 49(1), 115--132.
\href{https://doi.org/10.1007/BF02294210}{doi:10.1007/BF02294210}.

\bibitem{joreskog1994} J\"oreskog, K. G. (1994). On the estimation of polychoric
correlations and their asymptotic covariance matrix. \emph{Psychometrika},
59(3), 381--389. \href{https://doi.org/10.1007/BF02296131}{doi:10.1007/BF02296131}.

\bibitem{satorrabentler2001} Satorra, A., \& Bentler, P. M. (2001). A scaled
difference chi-square test statistic for moment structure analysis.
\emph{Psychometrika}, 66(4), 507--514.
\href{https://doi.org/10.1007/BF02296192}{doi:10.1007/BF02296192}.

\bibitem{savaleibentler2005} Savalei, V., \& Bentler, P. M. (2005). A
statistically justified pairwise ML method for incomplete nonnormal data: A
comparison with direct ML and pairwise ADF. \emph{Structural Equation
Modeling}, 12(2), 183--214.
\href{https://doi.org/10.1207/s15328007sem1202_1}{doi:10.1207/s15328007sem1202\_1}.

\bibitem{katsikatsou2012} Katsikatsou, M., Moustaki, I., Yang-Wallentin, F., \&
J\"oreskog, K. G. (2012). Pairwise likelihood estimation for factor analysis
models with ordinal data. \emph{Computational Statistics \& Data Analysis},
56(12), 4243--4258.
\href{https://doi.org/10.1016/j.csda.2012.04.010}{doi:10.1016/j.csda.2012.04.010}.

\bibitem{chen2020} Chen, P.-Y., Wu, W., Garnier-Villarreal, M., Kite, B. A., \&
Jia, F. (2020). Testing measurement invariance with ordinal missing data: A
comparison of estimators and missing data techniques. \emph{Multivariate
Behavioral Research}, 55(1), 87--101.
\href{https://doi.org/10.1080/00273171.2019.1608799}{doi:10.1080/00273171.2019.1608799}.

\bibitem{fmg2025} Foldnes, N., Moss, J., \& Gr\o nneberg, S. (2025). Improved
goodness of fit procedures for structural equation models (penalized eigenvalue
block averaging).

\end{thebibliography}

\end{document}
