\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}

\title{Relative-fit and nested difference tests for estimated-weight
  moment-quadratic estimators}
\author{}
\date{\today}

\newcommand{\E}{\mathbb{E}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\vech}{\operatorname{vech}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\argmax}{\operatorname*{arg\,max}}
\newcommand{\plim}{\operatorname*{plim}}
\newcommand{\T}{^{\mathsf{T}}}
\newcommand{\Wr}{W_{\!r}}
\newcommand{\rs}{r_{*}}
\newcommand{\ths}{\theta_{*}}

\theoremstyle{plain}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\begin{document}
\maketitle

\section*{Purpose}

The estimated-weight sandwich (companion note
\texttt{weighted\_moment\_ij\_grid}) corrects the parameter covariance by
adding a weight-derivative term. This note asks whether the same correction
extends to model-comparison tests, and whether the nested difference test
follows as a corollary. Three answers, stated up front.

\begin{enumerate}
\item The correction is proportional to the pseudo-true residual $\rs$
  (Proposition~\ref{prop:delin}). It is identically zero for the
  goodness-of-fit null, where the focal model is assumed correct, and present
  for every test whose null does not assume correctness.
\item For distinguishable models the relative-fit (Vuong) variance is a clean
  corollary: $\omega^{2}=c\T\Gamma c$ with the weight channel carried by an
  explicit term in $c$ (Proposition~\ref{prop:vuong}).
\item The nested, degenerate case is \emph{not} a clean corollary. The limiting
  quadratic form is the Hessian of the profiled discrepancy difference, not the
  covariance quadratic of the fitted residual fluctuation. The naive
  substitution $\Gamma\mapsto(W-\Wr)\Gamma(W-\Wr)\T$ into the $U$-matrix is
  wrong (Proposition~\ref{prop:nested}). It needs its own derivation.
\end{enumerate}

\section{Recap of the corrected influence}
\label{sec:recap}

The estimator minimises $F_{N}(\theta)=\tfrac12 r(\theta)\T\widehat{W}r(\theta)$
with residual $r(\theta)=\eta(\theta)-u$, Jacobian $\Delta=\partial\eta/\partial
\theta\T$, and weight $\widehat{W}=W(u)$ built from the sample moments $u$. Write
$g_{i}$ for the moment influence function, $\Gamma=\E[g_{i}g_{i}\T]$, and $\ths$,
$\rs=\eta(\ths)-u_{0}$ for the pseudo-true value and residual. The estimator is
asymptotically linear with influence function $\phi_{i}=H^{-1}\Delta\T
v_{i}$, where
\begin{equation}
  H = A + C,\quad A=\Delta\T W\Delta,\quad
  C=\sum_{k}(W\rs)_{k}\nabla^{2}\eta_{k}(\ths),\quad
  v_{i}=Wg_{i}-\dot W_{i}\,\rs,
  \label{eq:recap}
\end{equation}
where $\dot W_{i}$ is the casewise influence of the weight. When the weight is a
function of the modelled moments $u$, $\dot W_{i}\rs=\Wr g_{i}$ with
$\Wr=[\,\partial W/\partial u_{1}\,\rs,\ \dots,\ \partial W/\partial u_{m}\,
\rs\,]$ the residual-contracted weight derivative, so $v_{i}=(W-\Wr)g_{i}$ and
the corrected meat is $\Omega=(W-\Wr)\Gamma(W-\Wr)\T$. In general $\Omega=
\Var(v_{i})$ is formed from the casewise rows; the matrix form is the special
case used in Sections~\ref{sec:gof}--\ref{sec:nested}.

\section{The residual governs the correction}
\label{sec:delin}

\begin{proposition}
\label{prop:delin}
Every estimated-weight correction in \eqref{eq:recap}, namely the curvature
term $C$ and the weight term $\Wr$, is linear in the pseudo-true residual
$\rs$. Hence both vanish exactly when $\rs=0$. For a test whose null fixes the
focal model as correctly specified the correction is identically zero; for a
test whose null leaves the focal model misspecified the correction is present.
\end{proposition}

\begin{proof}
$C$ contracts $W\rs$ and $\Wr$ contracts $\rs$ columnwise, so both are $O(\lVert
\rs\rVert)$ and vanish at $\rs=0$. The goodness-of-fit null sets $\eta(\ths)=
u_{0}$, hence $\rs=0$. The relative-fit and the not-assuming-correct nested
nulls leave $\rs\neq0$.
\end{proof}

A second, sharper reason closes the goodness-of-fit case. At that null the
fitted residual is $r(\widehat\theta)=O_{p}(N^{-1/2})$, so the weight
fluctuation enters the residual only through $(\widehat W-W)r(\widehat\theta)=
O_{p}(N^{-1})$, which is negligible. The estimated weight is therefore
asymptotically irrelevant to the fit statistic, which is why the fixed-weight
projection theory of \cite{satorra1989} and the scaled statistic implemented in
practice are already correct under the null.

\section{Goodness of fit: the projection, unchanged}
\label{sec:gof}

For a fixed weight, expand the fitted residual at the goodness-of-fit null. With
$H=A$ and $\Wr=0$ there,
\[
  \sqrt{N}\,r(\widehat\theta)=-M\sqrt{N}(u-u_{0})+o_{p}(1),
  \qquad M=I-\Delta A^{-1}\Delta\T W,
\]
so $\Var(\sqrt N\,r(\widehat\theta))=M\Gamma M\T$ and the fit statistic
$N\,r\T W r$ converges to $\sum_{j}\lambda_{j}\chi^{2}_{1,j}$ with $\lambda_{j}=
\operatorname{eig}(U\Gamma)$ and $U=WM=W-W\Delta(\Delta\T W\Delta)^{-1}\Delta\T
W$ \cite{satorra1989}. When $W=\Gamma^{-1}$ the product $U\Gamma$ is idempotent
and the law collapses to $\chi^{2}_{\mathrm{df}}$; for $W\neq\Gamma^{-1}$ (DWLS,
ULS) it is a genuine mixture, the reason the scaled statistic
\cite{satorrabentler2001} exists. By Proposition~\ref{prop:delin} the estimated
weight contributes nothing here.

\section{Relative fit: a clean corollary}
\label{sec:vuong}

Compare two structural models $A$ and $B$ on the same indicators. The weight
$\widehat W=W(u)$ is a function of the sample moments alone, so it is shared,
$W_{A}=W_{B}=W$. Let $D=F_{A}(\widehat\theta_{A})-F_{B}(\widehat\theta_{B})$ be
the sample discrepancy difference and $\rs^{A},\rs^{B}$ the two pseudo-true
residuals.

\begin{proposition}
\label{prop:vuong}
The discrepancy difference is asymptotically linear, $\sqrt{N}\,D=N^{-1/2}\sum_{i}
c\T g_{i}+o_{p}(1)$, with
\begin{equation}
  c = -W(\rs^{A}-\rs^{B}) + \tfrac12(b^{A}-b^{B}),
  \qquad
  b^{A}_{l}=(\rs^{A})\T\frac{\partial W}{\partial u_{l}}\,\rs^{A},
  \label{eq:cvec}
\end{equation}
and $b^{B}_{l}$ defined likewise. For distinguishable models $\omega^{2}=c\T
\Gamma c>0$ and $\sqrt{N}\,D/\widehat\omega\to\mathcal N(0,1)$.
\end{proposition}

\begin{proof}
View $F_{A}$ as a functional of the sampling distribution through $u$, the
minimiser $\widehat\theta_{A}(u)$, and the weight $W(u)$. The minimiser is
interior, so the envelope theorem removes the $\widehat\theta_{A}$ channel and
\[
  \mathrm{IF}_{i}(F_{A})
  = \Bigl(\tfrac{\partial F_{A}}{\partial u}
    + \tfrac{\partial F_{A}}{\partial W}\,DW(u_{0})\Bigr)g_{i}
  = \bigl(-W\rs^{A}+\tfrac12 b^{A}\bigr)\T g_{i},
\]
using $\partial F_{A}/\partial u=-r\T W$ and $\partial(\tfrac12 r\T W r)/\partial
W$ contracted with $DW[g_{i}]=\sum_{l}(\partial W/\partial u_{l})g_{i,l}$.
Subtracting $\mathrm{IF}_{i}(F_{B})$ gives \eqref{eq:cvec}; the central limit
theorem gives the normal law.
\end{proof}

The term $-W(\rs^{A}-\rs^{B})$ is the difference of weighted residuals that a
fixed-weight account would give. The term $\tfrac12(b^{A}-b^{B})$ is the weight
channel, quadratic in each model's own residual and contracted with the weight
derivative. It is the testing analogue of $\Wr$ and is the new ingredient. Every
quantity is already available: $\rs^{A},\rs^{B}$ from the two fits, $\partial
W/\partial u$ from the same derivative the standard error uses, and $\Gamma$ from
the stage-one moments. The relative-fit test is therefore a corollary of the
standard-error machinery, with no projection involved because the envelope
theorem has removed the parameter channel.

\section{The nested case does not commute}
\label{sec:nested}

When $B$ is nested in $A$ the discrepancy difference degenerates under the null
that the restriction holds, $\rs^{A}=\rs^{B}$ in the relevant directions, so
$c=0$ and $\omega^{2}=0$. The first-order term vanishes and the limiting law is
the second-order quadratic form. The question is what second derivative that
quadratic form uses.

For one model $M$, let
\[
  f_{M}(u)=\min_{\theta}\tfrac12 r_{M}(\theta,u)\T W(u)r_{M}(\theta,u),
  \qquad r_{M}(\theta,u)=\eta_{M}(\theta)-u .
\]
The envelope gradient of this profiled scalar discrepancy is
\begin{equation}
  s_{M}(u)=\nabla_{u}f_{M}(u)
  =-W\rs^{M}+\tfrac12 b_{M},
  \qquad
  (b_{M})_{l}=(\rs^{M})\T\frac{\partial W}{\partial u_{l}}\rs^{M}.
  \label{eq:profile-score}
\end{equation}
This is the ``score'' used below: a moment-space / profile-value score, not the
parameter score $\partial F/\partial\theta$.

\begin{proposition}
\label{prop:nested}
Suppose the nested null is true at the pseudo-true level: the larger model's
pseudo-true point lies in the smaller model, so the two profiled values and
profile scores agree at $u_{0}$, $f_{A}(u_{0})=f_{B}(u_{0})$ and
$s_{A}(u_{0})=s_{B}(u_{0})$. Let
\[
  Q_{M}=D_{u}s_{M}(u_{0})
\]
be the Hessian of the profiled discrepancy for model $M$. Then
\[
  2N\{f_{B}(\widehat u)-f_{A}(\widehat u)\}
  \;\xrightarrow{d}\;
  z\T (Q_{B}-Q_{A})z,\qquad z\sim N(0,\Gamma),
\]
and the reference spectrum is
\[
  \lambda_{j}=\operatorname{eig}\{(Q_{B}-Q_{A})\Gamma\}.
\]
For a fixed weight,
\begin{equation}
  Q_{M}=W-W\Delta_{M}H_{M}^{-1}\Delta_{M}\T W,
  \qquad H_{M}=A_{M}+C_{M}.
  \label{eq:profile-hessian-fixed}
\end{equation}
For an estimated weight $W(u)$, $Q_{M}$ is the derivative of
\eqref{eq:profile-score}. It contains the fitted-residual derivative
\[
  D_{u}\rs^{M}[h]=
  \Delta_{M}H_{M}^{-1}\Delta_{M}\T\{Wh-DW[h]\rs^{M}\}-h
\]
and the first and second derivatives of the weight map. Consequently the nested
law is not obtained by replacing $\Gamma$ by
$(W-\Wr)\Gamma(W-\Wr)\T$ in Satorra's fixed-weight $U$ matrix.
\end{proposition}

\begin{proof}
Taylor expand the two profiled scalar functions around $u_{0}$:
\[
  f_{M}(u_{0}+N^{-1/2}z)
  =f_{M}(u_{0})+N^{-1/2}s_{M}(u_{0})\T z
   +\tfrac12N^{-1}z\T Q_{M}z+o_{p}(N^{-1}).
\]
The nested pseudo-null cancels the value and first derivative. Multiplying the
difference by $2N$ leaves $z\T(Q_{B}-Q_{A})z+o_{p}(1)$, giving the eigenvalue
law. For fixed $W$, differentiating the population first-order condition
$\Delta\T W\rs=0$ gives
$D\theta[h]=H^{-1}\Delta\T Wh$ and
$D\rs[h]=\Delta H^{-1}\Delta\T Wh-h$. Since $s=-W\rs$ when $W$ is fixed,
$Q[h]=-WD\rs[h]$, which is \eqref{eq:profile-hessian-fixed}. When $W=W(u)$, the
same calculation includes $DW[h]\rs$ in the first-order condition and the
derivative of the weight term in \eqref{eq:profile-score}; this introduces the
first and second weight derivatives.
\end{proof}

The fitted residual fluctuation is still useful for standard errors:
\[
  \sqrt{N}\,(\hat r_{M}-\rs^{M})
  =-\widetilde M_{M}z+o_{p}(1),\qquad
  \widetilde M_{M}=I-\Delta_{M}H_{M}^{-1}\Delta_{M}\T(W-\Wr^{M})
\]
when $W=W(u)$. But the covariance quadratic
$\widetilde M_{M}\T W\widetilde M_{M}$ is not the Hessian of the profiled
discrepancy unless $H_{M}=A_{M}$ and the weight terms satisfy special
cancellations. The nested test is a profile-value problem, not a residual-vcov
problem.

\section{Specialization: complete-data DWLS on a linear covariance model}
\label{sec:dwls}

The curvature term $C=\sum_{k}(W\rs)_{k}\nabla^{2}\eta_{k}(\ths)$ is the only
place the shape of the model surface enters. It vanishes whenever $\eta$ is
linear in $\theta$, which removes one of the two deformations and isolates the
weight channel.

\subsection{A model with $C=0$}
Take a linear covariance structure $\Sigma(\theta)=\sum_{k}\theta_{k}G_{k}$ with
fixed symmetric $G_{k}$, so $\eta(\theta)=\vech\Sigma(\theta)=\Delta\theta$ with
$\Delta=[\vech G_{1},\dots,\vech G_{q}]$ constant, $\nabla^{2}\eta_{k}=0$, hence
$C=0$ and $H=A=\Delta\T W\Delta$ constant. A concrete nested pair on $p$
continuous variables: $B$ is compound symmetry (one variance, one covariance),
nested in $A$ (one variance, free covariances), the difference testing equality
of the covariances. Both are linear, so the only surviving deformation is the
weight channel.

\begin{remark}[A test bed, not the headline model]
The parameter is restricted to the open cone $\{\theta:\Sigma(\theta)\succ0\}$;
for compound symmetry this is the wedge $a-b>0$ and $a+(p-1)b>0$ in
$\theta=(a,b)=(\text{variance},\text{covariance})$. This is an inequality
constraint, inactive at an interior pseudo-true value, so it does not enter the
first-order conditions or the asymptotics, and $C=0$ holds from the linearity of
$\eta$ regardless of it. The only design requirement is to keep the
data-generating $\theta_{*}$ interior, with no near-singular best fit. Linear
covariance structures are a classical class (variance components, exchangeable
and Toeplitz patterns, fixed-loading factor models), but pairing one with DWLS is
off the beaten path. Its role here is a curvature-free check: a deformation that
appears here cannot be curvature in disguise. The substantive case keeps a
genuine factor model and carries $C\neq0$ (Section~\ref{sec:cfa}).
\end{remark}

\subsection{DWLS ingredients}
The stage-one moments are $u=\vech(S)$ with casewise influence $g_{i}=\vech(z_{i}
z_{i}\T)-\sigma_{0}$, $z_{i}=y_{i}-\mu$, and $\Gamma=\E[g_{i}g_{i}\T]$ the Browne
fourth-moment matrix. DWLS uses $\widehat W=\diag(\widehat\Gamma)^{-1}$, so with
$\gamma=\diag(\Gamma)$ the population weight is $W=\diag(\gamma)^{-1}$. The weight
influence acts diagonally,
\[
  \dot w_{i,k}=-\gamma_{k}^{-2}\,\dot\gamma_{i,k},\qquad
  \dot\gamma_{i,k}=g_{i,k}^{2}-\gamma_{k}+(\text{centring corrections}),
\]
where $\dot\gamma_{i,k}$ is the casewise influence of the $k$-th diagonal of
$\widehat\Gamma$. The corrected casewise moment-space influence is
\begin{equation}
  v_{i,k}=\frac{g_{i,k}}{\gamma_{k}}
          +\frac{r_{*,k}\,\dot\gamma_{i,k}}{\gamma_{k}^{2}},
  \label{eq:dwls-v}
\end{equation}
which matches the implemented DWLS adapter.

\begin{remark}[The DWLS weight channel is not a matrix on $g_{i}$]
$\widehat W=\diag(\widehat\Gamma)^{-1}$ depends on fourth-order sample moments,
not on the modelled $u=\vech(S)$. Its influence $\dot\gamma_{i,k}$ is quadratic
in $g_{i}$ (through $g_{i,k}^{2}$), so the weight channel in \eqref{eq:dwls-v} is
a genuine casewise quantity, not $\Wr g_{i}$ for a matrix $\Wr$. The matrix
annihilator $\widetilde M$ of Proposition~\ref{prop:nested} therefore holds only
when the weight is a function of $u$; for DWLS the residual influence
$\rho_{i}=\Delta A^{-1}\Delta\T v_{i}-g_{i}$ stays casewise and carries a part
quadratic in $g_{i}$, so its covariance reaches eighth-order moments of the data.
\end{remark}

\subsection{The verification-friendly twin: normal-theory weight}
The normal-theory weight $W=W_{\mathrm{NT}}(S)$ is a function of $u$, so it is the
clean case. With $C=0$, the residual derivative uses
$\widetilde M=I-\Delta A^{-1}\Delta\T(W-\Wr)$, but the nested test uses
$Q=D_{u}s$, not $\widetilde M\T W\widetilde M$. A fixed weight would give
$Q=U=W-W\Delta A^{-1}\Delta\T W$, and with $W=\Gamma^{-1}$ the usual Satorra
spectrum collapses to $\chi^{2}_{df}$. A data-estimated normal-theory weight
adds the derivative of the envelope score \eqref{eq:profile-score}, including
second derivatives of $W_{\mathrm{NT}}(u)$. This isolates the weight channel from
curvature while keeping the actual test statistic in view.

\subsection{The geometry of the payoff}
The deformation of the profile Hessian is zero at $\rs=0$ and analytic in
$\rs$. The exploration is to sweep the direction of $\rs$ at fixed
$\lVert\rs\rVert$: generate from $\Sigma_{0}=\Sigma_{B}(\theta_{*})+\epsilon D$
and rotate $D$ from $W$-orthogonal to the constraint space toward aligned. The
conjecture, which the simulation tests rather than assumes, is that orthogonal
$\rs$ leaves the law near Satorra (the benign regime where robust intervals
already calibrate) and aligned $\rs$ deforms it most. The script
\texttt{relative-and-nested-fit-check.py} computes $Q_{B}-Q_{A}$ by finite
differences of the profile score and compares it with the ordinary Satorra
spectrum and with the residual-fluctuation quadratic.

\subsection{The residual deformation in closed form}
\label{sec:closedform}
With $C=0$ the fitted-residual fluctuation has an exact deformation. This is a
useful diagnostic for the standard-error machinery, but it is not the nested
test law once the weight is data-estimated.

\begin{proposition}[Closed-form deformation under $C=0$]
\label{prop:closed}
Let $H=A=\Delta\T W\Delta$ and $\widetilde M=I-\Delta A^{-1}\Delta\T(W-\Wr)$.
Then
\[
  \widetilde M\T W\widetilde M = U + R,\qquad
  U = M\T W M,\qquad
  R = (\Delta\T\Wr)\T A^{-1}(\Delta\T\Wr)\succeq 0,
\]
the cross terms vanishing by the annihilator orthogonality $\Delta\T W M=0$.
Hence the residual-vcov quadratic has spectrum
$\operatorname{eig}[(U_{B}-U_{A}+R_{B}-R_{A})\Gamma]$. This is not, in general,
the spectrum of $2N(f_{B}-f_{A})$; the latter is
$\operatorname{eig}[(Q_{B}-Q_{A})\Gamma]$ from Proposition~\ref{prop:nested}.
\end{proposition}

\begin{proof}
Write $\widetilde M=M+\delta$ with $\delta=\Delta A^{-1}\Delta\T\Wr\in\operatorname
{col}(\Delta)$. Since $M\T W\Delta=W\Delta-W\Delta A^{-1}(\Delta\T W\Delta)=0$,
both cross terms $M\T W\delta$ and $\delta\T WM$ vanish, and $\delta\T W\delta=
\Wr\T\Delta A^{-1}(\Delta\T W\Delta)A^{-1}\Delta\T\Wr=(\Delta\T\Wr)\T A^{-1}
(\Delta\T\Wr)=R$.
\end{proof}

The consequence is narrower than first hoped: $\Wr$ is linear in $\rs$, so this
residual-vcov deformation is quadratic in $\rs$ and vanishes when a model fits.
That statement validates the residual influence algebra, not the second-order
law of the profiled discrepancy. The numerical check confirms the identity to
machine precision and then separately computes the profile-Hessian law.

\begin{remark}[Why this is special to $C=0$]
The cancellation uses $A=\Delta\T W\Delta$. With curvature the bread is $H=A+C$,
$M\T W\Delta=W\Delta-W\Delta H^{-1}A\neq0$, and the cross terms survive. The clean
$U+R$ split is exactly what the entanglement of Proposition~\ref{prop:nested}
breaks; the linear test bed buys a closed form that the factor model does not have.
\end{remark}

\subsection{The substantive twin: a factor model with $C\neq0$}
\label{sec:cfa}
Dropping linearity costs the curvature term. For a confirmatory factor model
$\Sigma(\theta)=\Lambda\Psi\Lambda\T+\Theta$ the moments are quadratic in the
loadings, so $\nabla^{2}\eta_{k}$ is sparse and closed form (the $\lambda_{a}
\lambda_{b}$ blocks), and $H=A+C$ with $C=\sum_{k}(W\rs)_{k}\nabla^{2}\eta_{k}
(\ths)$ evaluated at the pseudo-true point. For fixed weights the profile Hessian
is $Q=W-W\Delta H^{-1}\Delta\T W$, so replacing $A$ by $H$ is already necessary.
For estimated weights, the derivative of \eqref{eq:profile-score} adds the
weight channel on top. A natural nested pair is configural versus metric
invariance across two groups, the difference testing equality of a block of
loadings. The price of realism is that the deformation now superposes two
channels, curvature and weight, so the clean attribution of the linear test bed
is lost. The CFA check script verifies that the old residual-vcov quadratic
breaks away from the finite-difference profile-Hessian law when $C\neq0$.

\section{Verdict}
\label{sec:verdict}

The testing extension partitions cleanly. The goodness-of-fit test needs no
correction (Proposition~\ref{prop:delin}); the prevailing scaled statistic is
already right. The relative-fit test for distinguishable models is a corollary
of the standard-error machinery, with $\omega^{2}=c\T\Gamma c$ and the weight
channel \eqref{eq:cvec} computable from quantities the standard error already
forms (Proposition~\ref{prop:vuong}). The nested difference test, in the regime
that does not assume the larger model correct, is its own project, because the
object is the Hessian of the profiled discrepancy difference. The fitted
residual influence and the parameter sandwich supply ingredients, but the
profile-Hessian law also needs curvature and, for estimated weights, second
derivatives of the weight map (Proposition~\ref{prop:nested}). For the
manuscript this argues to keep the estimated-weight standard error and the
distinguishable relative-fit test together, and to treat the nested
weighted-$\chi^{2}$ law as a sequel
\cite{vuong1989,merkleyou2016}.

\begin{thebibliography}{9}

\bibitem{satorra1989} Satorra, A. (1989). Alternative test criteria in
  covariance structure analysis: a unified approach. \emph{Psychometrika},
  54(1), 131--151.
  \href{https://doi.org/10.1007/BF02294453}{doi:10.1007/BF02294453}.

\bibitem{satorra2000} Satorra, A. (2000). Scaled and adjusted restricted tests
  in multi-sample analysis of moment structures. In R. D. H. Heijmans, D. S. G.
  Pollock, \& A. Satorra (Eds.), \emph{Innovations in Multivariate Statistical
  Analysis} (pp. 233--247). Springer.
  \href{https://doi.org/10.1007/978-1-4615-4603-0_17}{doi:10.1007/978-1-4615-4603-0\_17}.

\bibitem{satorrabentler2001} Satorra, A., \& Bentler, P. M. (2001). A scaled
  difference chi-square test statistic for moment structure analysis.
  \emph{Psychometrika}, 66(4), 507--514.
  \href{https://doi.org/10.1007/BF02296192}{doi:10.1007/BF02296192}.

\bibitem{vuong1989} Vuong, Q. H. (1989). Likelihood ratio tests for model
  selection and non-nested hypotheses. \emph{Econometrica}, 57(2), 307--333.
  \href{https://doi.org/10.2307/1912557}{doi:10.2307/1912557}.

\bibitem{merkleyou2016} Merkle, E. C., You, D., \& Preacher, K. J. (2016).
  Testing nonnested structural equation models. \emph{Psychological Methods},
  21(2), 151--163.
  \href{https://doi.org/10.1037/met0000038}{doi:10.1037/met0000038}.

\bibitem{laisimoes2023} Lai, K., \& Sim\~oes, F. (2023). Reflecting on the
  ``robust'' standard errors for two-stage structural equation modeling.
  \emph{Structural Equation Modeling}, 30(5).

\bibitem{hallinoue2003} Hall, A. R., \& Inoue, A. (2003). The large sample
  behaviour of the generalized method of moments estimator in misspecified
  models. \emph{Journal of Econometrics}, 114(2), 361--394.
  \href{https://doi.org/10.1016/S0304-4076(03)00075-0}{doi:10.1016/S0304-4076(03)00075-0}.

\end{thebibliography}

\end{document}
