\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}

\title{Notes on Smooth H Caps and Robust Ordinal Gamma}
\author{}
\date{\today}

\newcommand{\fhat}{\widehat f}
\newcommand{\Ghat}{\widehat\Gamma}

\begin{document}
\maketitle

\section*{Context}

These are implementation notes for the robust ordinal SEM paper track, not
manuscript prose. They summarize two observations from the first pilot run of
\path{docs/research/sims/r/robust_ordinal_sem_paper.R}.

\section*{Smooth cap tuning}

The current smooth h-score cap is parameterized by two transition points
$a < b$. It is exactly ML below $a$, transitions smoothly on $[a,b]$, and is
flat above $b$:
\[
  h(t)=t,\quad t \le a,
  \qquad
  h'(t)=0,\quad t \ge b.
\]
For the implemented quintic smootherstep transition, the plateau is
\[
  h_\infty = a + \frac{b-a}{2} = \frac{a+b}{2}.
\]
Therefore the pilot setting
\[
  a=1.6,\qquad b=2.2
\]
does \emph{not} smooth the Welz--Mair--Alfons hard cap with $k=1.6$; it
smooths to a less aggressive plateau,
\[
  h_\infty = 1.9.
\]
This explains why the default smooth cap was not an apples-to-apples
competitor to the WMA hard cap. To smooth a hard cap at $k=1.6$, use a
transition centered at $1.6$, for example
\[
  a=1.3,\qquad b=1.9,
\]
or a tighter transition such as
\[
  a=1.45,\qquad b=1.75.
\]
Early pilot checks show these centered smooth caps track the hard-cap estimates
much more closely than the original $(1.6,2.2)$ setting.

\section*{Which matrix failed?}

The pilot failures were not failures of the robust polychoric correlation
matrix $R$. In the pilot CSV, the minimum eigenvalue of $R$ remained positive
for the failing bivariate cases. The failure arose later, when building the
full WLS weight matrix from the robust moment covariance:
\[
  W_{\mathrm{WLS}} = \Ghat^{-1}.
\]
The implementation computes moment influence rows
\[
  IF_i = -\psi_i(\widehat\eta) A(\widehat\eta)^{-\top},
\]
then forms
\[
  \Ghat = \frac{1}{n}\sum_{i=1}^n IF_i IF_i^\top .
\]
This matrix is positive semidefinite by construction, but it is not guaranteed
to be positive definite in finite samples. It can be rank deficient whenever
the casewise influence matrix does not span every threshold/correlation moment
direction.

\section*{Why robust Gamma can be singular}

Clipping and h-score caps intentionally remove or attenuate information from
cells that look implausible under the fitted table. In finite samples, this can
erase enough directions that $\Ghat$ becomes singular or nearly singular.
This is especially likely when:

\begin{itemize}
\item the table is bivariate but the moment vector still contains many
  thresholds plus one correlation;
\item high contamination concentrates observations into a few cells;
\item hard caps or Huber residual clips set derivatives to zero in influential
  cells;
\item some categories are sparse, so several threshold directions are only
  weakly represented;
\item the bread matrix is itself ill-conditioned, magnifying numerical rank
  issues in the influence rows.
\end{itemize}

Thus a non-positive-definite $\Ghat$ is not necessarily evidence that the
robust correlation estimate is unusable. It mainly says that the full WLS
weight matrix is not invertible without a regularization decision. DWLS can
remain usable if the diagonal entries are finite and positive.

\section*{Repair policy to consider}

For the methods-developer surface, keep the raw objects visible and make any
repair explicit. A defensible contract is:

\begin{enumerate}
\item always return the robust moments and raw $\Ghat$ when they are finite;
\item build DWLS weights from the diagonal when the diagonal is positive;
\item build full WLS weights only after an explicit policy:
  \begin{itemize}
  \item error on non-positive-definite $\Ghat$;
  \item ridge/eigenvalue floor;
  \item Moore--Penrose or truncated eigenspace pseudoinverse;
  \item shrinkage toward the diagonal or identity.
  \end{itemize}
\item record diagnostics:
  \[
    \lambda_{\min}(\Ghat),\quad
    \kappa(\Ghat),\quad
    \mathrm{rank}(\Ghat),\quad
    \text{repair kind and intensity}.
  \]
\end{enumerate}

For the paper simulations, the first useful step is diagnostic rather than
automatic repair: add $\Ghat$ eigenvalue/rank/condition summaries to the CSV.
Then compare DWLS behavior before deciding whether full WLS repair belongs in
the main text or only in implementation notes.

\section*{Immediate simulation changes}

\begin{itemize}
\item Report the centered smooth cap $(a,b)=(1.3,1.9)$ as the primary smooth
  comparator to WMA $k=1.6$.
\item Keep the original $(1.6,2.2)$ only as a ``less aggressive smooth cap''
  sensitivity check if needed.
\item Add Gamma diagnostics to the runner before scaling replications.
\item Treat full-WLS failures as data about weight instability, not as missing
  robust polychoric estimates.
\end{itemize}

\end{document}
