---
title: "Misspecification-Robust Standard Errors: Model vs Robust Bread"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
sp <- file.path("results", "se_regime.csv")
mp <- file.path("results", "metadata.csv")
if (!file.exists(sp)) stop("Run `Rscript run_experiment.R` first.", call. = FALSE)
S <- read.csv(sp)
M <- read.csv(mp)
meta <- function(k) M$value[match(k, M$key)]
f2 <- function(x) sprintf("%.2f", x)
r <- function(est, cell, kind, col) {
  x <- S[[col]][S$estimator == est & S$cell == cell & S$param_kind == kind]
  if (length(x)) x[[1L]] else NA_real_
}
lab_est <- c(ML = "Continuous ML", FIML = "FIML (MCAR)",
             DWLS = "Ordinal DWLS")
rows <- expand.grid(estimator = names(lab_est), param_kind = c("raw", "std"),
                    stringsAsFactors = FALSE)
tab <- data.frame(
  Estimator = lab_est[rows$estimator],
  Parameter = ifelse(rows$param_kind == "raw", "loading", "std. loading"),
  `Null: model` = mapply(function(e, k) f2(r(e, "null", k, "ratio_model")), rows$estimator, rows$param_kind),
  `Null: robust` = mapply(function(e, k) f2(r(e, "null", k, "ratio_robust")), rows$estimator, rows$param_kind),
  `Misspec: model` = mapply(function(e, k) f2(r(e, "misspec", k, "ratio_model")), rows$estimator, rows$param_kind),
  `Misspec: robust` = mapply(function(e, k) f2(r(e, "misspec", k, "ratio_robust")), rows$estimator, rows$param_kind),
  check.names = FALSE)
```

## Question

The conventional ("model") standard error in SEM is built from a bread term that
is exact only when the structural model is correct. When you knowingly interpret
a misfitting model's parameters -- a loading, or its standardized form -- does a
misspecification-robust SE (the same sandwich but with the observed-Hessian
bread) recover the true sampling variability, and does it cost anything when the
model is right?

## Short Answer

No cost under the null, large repair under the alternative. With the model
correct, the model and robust SE coincide and both match the empirical sampling
SD. With a true cross-loading omitted, the model SE **underestimates** the
sampling SD of the focal loading by roughly a third -- the same for the raw and
the standardized loading -- and the robust SE recovers it. For continuous ML the
recovery is essentially complete (standardized loading: model
`r f2(r("ML","misspec","std","ratio_model"))` of truth, robust
`r f2(r("ML","misspec","std","ratio_robust"))`). FIML uses the same
observed-Hessian bread vocabulary with missing data (model
`r f2(r("FIML","misspec","std","ratio_model"))`, robust
`r f2(r("FIML","misspec","std","ratio_robust"))`). For ordinal DWLS, where the
robust bread is new, it moves the same direction and closes most of the gap
(`r f2(r("DWLS","misspec","std","ratio_model"))` to
`r f2(r("DWLS","misspec","std","ratio_robust"))`).

The numbers are ratios of mean SE to the empirical sampling SD: 1.0 is
calibrated, below 1.0 underestimates. The two regimes share the same empirical
meat, so the whole difference is the bread.

The practical reading is the wider lesson about settings made concrete: under a
believed-correct model the conventional SE is right and efficient; once you are
standing on a model you know is imperfect -- which is most of applied SEM -- you
want the robust regime, and the choice should be one explicit keyword. For
ordinal DWLS the robust regime is not the end of the story: the conventional
observed-Hessian SE (Lai & Simoes 2023) still under-covers, because it fixes the
estimated weight, and only the complete jackknife reaches nominal CI coverage as
`N` grows -- the sharper version of this finding (see Coverage across N).

## Evidence

A two-factor six-indicator model where item `y4` cross-loads on the wrong factor
in the data-generating process and the fitted model leaves that out; the
latent-normal measurement model is correct throughout, so the misfit is purely
structural. Focal parameter: a free loading on the distorted factor, and its
`std.all` counterpart. `N = r meta("n_total")`, `r meta("reps")` replications.

```{r}
knitr::kable(tab, row.names = FALSE)
```

## Coverage across N: the fixed-weight "robust" SE does not converge, the complete jackknife does

This is the step beyond Lai & Simoes (2023). Their corrected SE is the
observed-Hessian bread (the "robust" regime above); like the lavaan default, it
still treats the Stage-2 weight as fixed. magmaan's complete infinitesimal
jackknife adds the influence of that estimated weight. Scoring CI coverage of the
pseudo-true loading (nominal 0.95) across sample size tells the three apart.

```{r}
#| include: false
csp <- file.path("results", "coverage_sweep.csv")
CS <- if (file.exists(csp)) read.csv(csp) else NULL
csv <- function(cell, n, col) {
  if (is.null(CS)) return(NA_real_)
  x <- CS[[col]][CS$cell == cell & CS$n_total == n]
  if (length(x)) x[[1L]] else NA_real_
}
```

The headline: under misspecification only the complete jackknife holds nominal
coverage as `N` grows. The lavaan-default (expected) interval sits near
`r f2(csv("misspec", 6000, "cov_exp"))` and never reaches 0.95 -- the structural
misfit does not vanish with `N`. The observed bread (Lai & Simoes' corrected SE)
recovers part of the gap but settles around
`r f2(csv("misspec", 6000, "cov_obs"))`, short of nominal and not closing with
`N`, because the weight is held fixed. The complete jackknife covers at
`r f2(csv("misspec", 600, "cov_ij"))`--`r f2(csv("misspec", 6000, "cov_ij"))`
across `N = 600`--`6000`. Under the correct model all three coincide near nominal
at every `N` (no misfit, so the breads agree).

```{r}
if (!is.null(CS)) {
  mis <- CS[CS$cell == "misspec", ]
  ct <- data.frame(
    N = mis$n_total,
    `Expected (lavaan default)` = sprintf("%.2f", mis$cov_exp),
    `Observed (Lai & Simoes 2023)` = sprintf("%.2f", mis$cov_obs),
    `Complete jackknife (this work)` = sprintf("%.2f", mis$cov_ij),
    check.names = FALSE)
  knitr::kable(ct, row.names = FALSE, caption = paste(
    "Misspecified cell: CI coverage of the pseudo-true loading (nominal 0.95) by",
    "sample size; ordinal DWLS. The null cell is omitted -- all three coincide",
    "near nominal there."))
} else cat("Run `Rscript scripts/coverage_sweep.R` for the coverage-vs-N table.\n")
```

## Bootstrap: is it worth it here?

We also ran a case-resampling bootstrap (499 resamples per fit; BCa uses a
grouped jackknife) on the ordinal DWLS focal loading, scoring CI coverage of the
pseudo-true value (nominal 0.95). The plot twist: the analytic observed-bread
interval already covers, and the bootstrap is, if anything, mildly conservative.

```{r}
bp <- file.path("results", "bootstrap_coverage.csv")
if (file.exists(bp)) {
  B <- read.csv(bp)
  bf <- function(x) sprintf("%.2f", x)
  bt <- data.frame(
    Cell = ifelse(B$cell == "null", "Correct model", "Misspecified"),
    `Expected normal` = bf(B$cov_norm_exp),
    `Observed normal` = bf(B$cov_norm_obs),
    Percentile = bf(B$cov_pct), BCa = bf(B$cov_bca),
    Studentized = bf(B$cov_stud), check.names = FALSE)
  knitr::kable(bt, row.names = FALSE)
} else cat("Run `Rscript scripts/bootstrap.R` for the bootstrap coverage table.\n")
```

At this single sample size the expected-bread (lavaan-default) interval
under-covers (~0.87), the observed bread is near nominal, and the three bootstrap
intervals slightly over-cover (~0.97). The bootstrap re-estimates the weight on
every resample, so -- unlike the fixed-weight analytic breads -- it carries the
weight's own sampling variability, the same term the complete jackknife adds
analytically (coverage-vs-N section above). That the resampling intervals land at
or above nominal while the fixed-weight breads fall short is the cross-check that
the weight term is the missing piece. The practical reading: the analytic
jackknife reaches calibrated coverage without 499 refits per fit, so the
bootstrap is a confirmation here, not a necessity. (The bare observed-bread *mean
SE* underestimates the empirical SD by exactly that weight term; the jackknife SE
matches it -- see the `se` columns of the sweep.)

## Caveats

- One focal parameter, one design point. The observed-Hessian ("robust") SE is
  an asymptotic sandwich: continuous ML closes the SE gap fully, but for ordinal
  DWLS the observed bread leaves a residual that does **not** vanish with `N` --
  the influence of the estimated Stage-2 weight, which under-covers by ~2--3
  points (Coverage across N). The complete jackknife adds that term and recovers
  nominal coverage at every `N`. The SE-ratio table above reports the observed
  bread for all three estimators, so its DWLS column carries this residual; read
  it alongside the jackknife coverage, not as the final word for DWLS.
- The coverage-vs-N sweep is ordinal DWLS, one focal loading, 400 replications
  per cell. The ordering (expected < observed < jackknife, only the jackknife at
  nominal) is stable across `N = 600`--`6000`; absolute rates carry Monte Carlo
  noise (~1.1 points at `p = 0.95`).
- The null-cell raw ratios carry Monte Carlo noise in the empirical SD at this
  replication count; the model-equals-robust coincidence (not the absolute
  level) is the point there.
- magmaan-only study. For continuous ML the observed-Hessian sandwich is
  standard everywhere (robust.huber.white / MLR); for the ordinal/WLS family it
  is new surface, so the correct-model cell -- where the robust bread must
  coincide with the conventional one -- is the validation oracle.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 250 --n-total 600 --cross 0.4
Rscript scripts/coverage_sweep.R --reps 400 --ns 600,1500,3000,6000
Rscript scripts/bootstrap.R --reps 150 --boot 499 --n-total 600 --cross 0.4
quarto render report.qmd
```
