---
title: "Misspecification-Robust Standard Errors for Ordinal DWLS"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
sp <- file.path("results", "se_regime.csv")
mp <- file.path("results", "metadata.csv")
if (!file.exists(sp)) stop("Run `Rscript run_experiment.R` first.", call. = FALSE)
S <- read.csv(sp)
M <- read.csv(mp)
meta <- function(k) M$value[match(k, M$key)]
fmt <- function(x, d = 3) sprintf(paste0("%.", d, "f"), x)
tab <- S[S$param_kind == "raw", ]
tab <- data.frame(
  Cell = ifelse(tab$cell == "null", "Correct model", "Misspecified"),
  `Empirical SD` = fmt(tab$emp_sd, 4),
  `Model SE` = fmt(tab$se_model, 4),
  `Robust SE` = fmt(tab$se_robust, 4),
  `Model / truth` = fmt(tab$ratio_model, 3),
  `Robust / truth` = fmt(tab$ratio_robust, 3),
  check.names = FALSE)
```

## Question

When an ordinal CFA is fit by DWLS (the WLSMV family) and the structural model
is *wrong*, the conventional ("model") standard error is built from a bread term
that is only exact when the model is correct. Does a misspecification-robust SE,
which swaps in the observed-Hessian bread, recover the true sampling variability
of the estimates -- and does it cost anything when the model is right?

## Short Answer

It does, and it costs nothing under the null. When the fitted model is correct,
the conventional SE and the robust SE are indistinguishable and both match the
empirical sampling SD (ratios `r fmt(S$ratio_model[S$cell=="null"&S$param_kind=="raw"],2)`
vs `r fmt(S$ratio_robust[S$cell=="null"&S$param_kind=="raw"],2)`). When the model
omits a true cross-loading, the conventional SE **underestimates** the sampling
SD by about 14% (ratio `r fmt(S$ratio_model[S$cell=="misspec"&S$param_kind=="raw"],2)`),
while the robust SE recovers it to within about 4%
(ratio `r fmt(S$ratio_robust[S$cell=="misspec"&S$param_kind=="raw"],2)`). The two
regimes use the *same* sampling-based meat and differ only in the bread, so this
is purely the bread doing its job: it is the right thing under the alternative
and reduces to the conventional bread under the null.

The practical reading mirrors the wider lesson about settings: under a believed-
correct model the conventional SE is fine and efficient; once you are knowingly
interpreting a misfitting model's parameters, you want the robust regime. The
choice should be explicit, which is why it is one keyword.

## Evidence

Focal parameter: the factor covariance, in a two-factor six-indicator ordinal
CFA at `N = r meta("n_total")`, `r meta("reps")` replications. The "misspecified"
cell generates a cross-loading the fitted model leaves out; the latent-normal
measurement model is correct throughout, so this is purely structural misfit.

```{r}
knitr::kable(tab, row.names = FALSE)
```

The empirical SD is the truth (the sampling SD of the estimate across
replications). A ratio of 1 means the SE is calibrated; below 1 means it
underestimates.

## Caveats

- One focal parameter, one design point. The standardized counterpart of this
  parameter (a correlation) is barely affected by the regime here; the
  underestimation and its repair are clearest on the unstandardized focal. The
  standardized-loading amplification is a separate parameter to probe.
- The robust SE is an asymptotic sandwich: at finite `N` and strong
  misspecification it closes most, not all, of the gap. Larger `N` or milder
  misspecification tightens it.
- This is a magmaan-only study (no Mplus/lavaan oracle): no mainstream library
  exposes an observed-Hessian bread for the ordinal/WLS family, so there is
  nothing to gate against. The internal check is the correct-model cell, where
  the robust bread must coincide with the conventional one.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 300 --n-total 1000 --cross 0.3
quarto render report.qmd
```
