---
title: "Misspecification-Robust Standard Errors: Model vs Robust Bread"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
sp <- file.path("results", "se_regime.csv")
mp <- file.path("results", "metadata.csv")
if (!file.exists(sp)) stop("Run `Rscript run_experiment.R` first.", call. = FALSE)
S <- read.csv(sp)
M <- read.csv(mp)
meta <- function(k) M$value[match(k, M$key)]
f2 <- function(x) sprintf("%.2f", x)
r <- function(est, cell, kind, col) {
  S[[col]][S$estimator == est & S$cell == cell & S$param_kind == kind]
}
lab_est <- c(ML = "Continuous ML", DWLS = "Ordinal DWLS")
rows <- expand.grid(estimator = c("ML", "DWLS"), param_kind = c("raw", "std"),
                    stringsAsFactors = FALSE)
tab <- data.frame(
  Estimator = lab_est[rows$estimator],
  Parameter = ifelse(rows$param_kind == "raw", "loading", "std. loading"),
  `Null: model` = mapply(function(e, k) f2(r(e, "null", k, "ratio_model")), rows$estimator, rows$param_kind),
  `Null: robust` = mapply(function(e, k) f2(r(e, "null", k, "ratio_robust")), rows$estimator, rows$param_kind),
  `Misspec: model` = mapply(function(e, k) f2(r(e, "misspec", k, "ratio_model")), rows$estimator, rows$param_kind),
  `Misspec: robust` = mapply(function(e, k) f2(r(e, "misspec", k, "ratio_robust")), rows$estimator, rows$param_kind),
  check.names = FALSE)
```

## Question

The conventional ("model") standard error in SEM is built from a bread term that
is exact only when the structural model is correct. When you knowingly interpret
a misfitting model's parameters -- a loading, or its standardized form -- does a
misspecification-robust SE (the same sandwich but with the observed-Hessian
bread) recover the true sampling variability, and does it cost anything when the
model is right?

## Short Answer

No cost under the null, large repair under the alternative. With the model
correct, the model and robust SE coincide and both match the empirical sampling
SD. With a true cross-loading omitted, the model SE **underestimates** the
sampling SD of the focal loading by roughly a third -- the same for the raw and
the standardized loading -- and the robust SE recovers it. For continuous ML the
recovery is essentially complete (standardized loading: model
`r f2(r("ML","misspec","std","ratio_model"))` of truth, robust
`r f2(r("ML","misspec","std","ratio_robust"))`). For ordinal DWLS, where the
robust bread is new, it moves the same direction and closes most of the gap
(`r f2(r("DWLS","misspec","std","ratio_model"))` to
`r f2(r("DWLS","misspec","std","ratio_robust"))`).

The numbers are ratios of mean SE to the empirical sampling SD: 1.0 is
calibrated, below 1.0 underestimates. The two regimes share the same empirical
meat, so the whole difference is the bread.

The practical reading is the wider lesson about settings made concrete: under a
believed-correct model the conventional SE is right and efficient; once you are
standing on a model you know is imperfect -- which is most of applied SEM -- you
want the robust regime, and the choice should be one explicit keyword.

## Evidence

A two-factor six-indicator model where item `y4` cross-loads on the wrong factor
in the data-generating process and the fitted model leaves that out; the
latent-normal measurement model is correct throughout, so the misfit is purely
structural. Focal parameter: a free loading on the distorted factor, and its
`std.all` counterpart. `N = r meta("n_total")`, `r meta("reps")` replications.

```{r}
knitr::kable(tab, row.names = FALSE)
```

## Caveats

- One focal parameter, one design point, `N = r meta("n_total")`. The robust SE
  is an asymptotic sandwich: continuous ML closes the gap fully here, while
  ordinal DWLS at this `N` and a strong misspecification closes most of it.
  Larger `N` or milder misspecification tightens the ordinal case.
- The null-cell raw ratios carry Monte Carlo noise in the empirical SD at this
  replication count; the model-equals-robust coincidence (not the absolute
  level) is the point there.
- magmaan-only study. For continuous ML the observed-Hessian sandwich is
  standard everywhere (robust.huber.white / MLR); for the ordinal/WLS family it
  is new surface, so the correct-model cell -- where the robust bread must
  coincide with the conventional one -- is the validation oracle.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 250 --n-total 600 --cross 0.4
quarto render report.qmd
```
