---
title: "A Misspecification-Robust Standard Error for Ordinal DWLS"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
csp <- file.path("results", "coverage_sweep.csv")
if (!file.exists(csp)) {
  stop("Run `Rscript scripts/coverage_sweep.R` first.", call. = FALSE)
}
CS <- read.csv(csp)
f2 <- function(x) sprintf("%.2f", x)
csv <- function(cell, n, col) {
  x <- CS[[col]][CS$cell == cell & CS$n_total == n]
  if (length(x)) x[[1L]] else NA_real_
}
nmin <- min(CS$n_total); nmax <- max(CS$n_total); reps <- max(CS$reps)
```

## Question

When you interpret the parameters of a model you know is misspecified -- the
common case in applied SEM -- does the standard error give an interval that
actually covers the (pseudo-)true value? For ordinal DWLS specifically: the
conventional "robust" SE corrects the bread but still treats the estimated weight
matrix as fixed. Does accounting for the weight's own sampling variability matter
for coverage, and across what sample sizes?

## Short Answer

Yes, and only the SE that accounts for the estimated weight is calibrated. Under
a misspecified model the lavaan-default ("expected") SE covers the focal loading
about `r f2(csv("misspec", nmax, "cov_exp"))` of the time at a nominal 0.95, and
the conventional observed-Hessian ("robust") SE -- the corrected SE of Lai &
Simoes (2023) -- recovers part of the gap but settles around
`r f2(csv("misspec", nmax, "cov_obs"))`, still short of nominal. Both shortfalls
are **independent of `N`**: the misspecification gap does not shrink as the
sample grows. The complete infinitesimal jackknife, which adds the influence of
the data-estimated weight, covers near nominal --
`r f2(csv("misspec", nmin, "cov_ij"))` to `r f2(csv("misspec", nmax, "cov_ij"))`
across `N` from `r nmin` to `r nmax` -- and its mean SE tracks the empirical
sampling SD throughout. Under a correctly specified model all three coincide
near nominal at every `N`.

The practical reading: standing on a model you know is imperfect, the ordinary
"robust" SE for ordinal DWLS is not enough; it under-covers by a margin that does
not vanish with `N`, and the fix is to account for the estimated weight.

## Evidence

A two-factor six-indicator ordinal model (three categories each) in which item
`y4` cross-loads on the wrong factor in the data-generating process and the
fitted model omits it; the latent-normal measurement model is correct throughout,
so the misfit is purely structural. Focal parameter: a free loading on the
distorted factor. We score CI coverage of its pseudo-true value (the probability
limit of the estimate, from one huge-`N` fit) at a nominal 0.95, over `r reps`
replications per cell, for three standard errors: the expected bread (lavaan
default), the observed bread (Lai & Simoes 2023), and the complete jackknife.

```{r}
#| label: fig-coverage-n
#| fig-width: 7
#| fig-height: 4.2
#| fig-cap: "Misspecified cell: CI coverage of the pseudo-true ordinal-DWLS loading versus sample size (nominal 0.95, dashed). Bands are Wilson 95% binomial intervals over the replications. Only the complete jackknife straddles nominal across N; the expected (lavaan) and observed (Lai & Simoes 2023) breads sit below by an N-independent margin. This is the well-conditioned design -- a harder one could break DWLS convergence at small N and drag every method down."
if (requireNamespace("ggplot2", quietly = TRUE)) {
  suppressPackageStartupMessages(library(ggplot2))
  mis <- CS[CS$cell == "misspec", ]
  series <- list(c("cov_exp", "Expected (lavaan default)"),
                 c("cov_obs", "Observed (Lai & Simoes 2023)"),
                 c("cov_ij",  "Complete jackknife (this work)"))
  long <- do.call(rbind, lapply(series, function(m) data.frame(
    N = mis$n_total, reps = mis$reps, method = m[2], p = mis[[m[1]]],
    stringsAsFactors = FALSE)))
  z <- qnorm(0.975)                                  # Wilson binomial interval
  den <- 1 + z^2 / long$reps
  ctr <- (long$p + z^2 / (2 * long$reps)) / den
  half <- (z / den) * sqrt(long$p * (1 - long$p) / long$reps +
                           z^2 / (4 * long$reps^2))
  long$lo <- ctr - half; long$hi <- ctr + half
  long$method <- factor(long$method, levels = vapply(series, `[`, "", 2))
  pal <- c("Expected (lavaan default)" = "#D55E00",
           "Observed (Lai & Simoes 2023)" = "#E69F00",
           "Complete jackknife (this work)" = "#0072B2")
  ggplot(long, aes(N, p, color = method, fill = method)) +
    geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey40") +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, color = NA) +
    geom_line(linewidth = 0.6) + geom_point(size = 1.7) +
    scale_x_log10(breaks = c(100, 200, 400, 800, 1500)) +
    scale_color_manual(values = pal) + scale_fill_manual(values = pal) +
    coord_cartesian(ylim = c(0.80, 1.00)) +
    labs(x = "N (log scale)", y = "CI coverage (nominal 0.95)",
         color = NULL, fill = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
} else cat("Install ggplot2 to render the coverage figure.\n")
```

## Caveats

- One focal parameter, one design point. This is the well-conditioned design
  (moderate loadings, symmetric three-category thresholds): DWLS converges on all
  replications even at `N = 100`, so the estimator recovers the pseudo-true value
  and coverage is meaningful at every `N`. A harder design (extreme or asymmetric
  thresholds, weak loadings, sparse cells) could break convergence or bias the
  estimate at small `N`, dragging coverage down for every method -- a
  point-estimate problem no SE correction can fix.
- The jackknife is the only SE here whose interval is calibrated; the expected and
  observed breads under-cover by an `N`-independent margin, the signature of a
  misspecification gap rather than a finite-sample one. The bands in the figure
  are Wilson binomial intervals, so the residual wiggle is Monte Carlo noise.
- magmaan-only study. For the ordinal/WLS family the observed-Hessian bread and
  the estimated-weight jackknife are new surface (no mainstream software exposes
  them), so the correct-model cell -- where all three SEs must coincide -- is the
  validation oracle.

## Reproduce

```sh
Rscript scripts/coverage_sweep.R --smoke
Rscript scripts/coverage_sweep.R --reps 2000
quarto render report.qmd
```
