---
title: "Robust parameter intervals: Wald, score and profile LR"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r}
path <- "results/frozen"
if (!file.exists(file.path(path, "summary.csv")))
  stop("Run the experiment and scripts/freeze_results.R as described in Reproduce.")
s <- read.csv(file.path(path, "summary.csv"))
pairs <- read.csv(file.path(path, "paired.csv"))
common <- read.csv(file.path(path, "common_valid.csv"))
failures <- read.csv(file.path(path, "failures.csv"))
integrity <- read.csv(file.path(path, "integrity.csv"))
primary <- c("wald_O", "score_E", "lr_E")
labels <- c(wald_O="Wald-O", score_E="Score-E", lr_E="LR-E",
            wald_E="Wald-E", score_O="Score-O", lr_O="LR-O",
            wald_naive="Uncorrected Wald", score_naive="Uncorrected score",
            lr_naive="Uncorrected LR")
colours <- c(wald_O="#333333", score_E="#0072B2", lr_E="#D55E00")
cell <- function(n, distribution, target, method) {
  z <- s[s$n==n & s$distribution==distribution & s$target==target & s$method==method,]
  stopifnot(nrow(z)==1L);z
}
pc <- function(x) sprintf("%.1f%%",100*x)
interval_range <- function(x) paste(sprintf("%.1f",100*range(x)),collapse="–")
```

## Question

For interior SEM parameters estimated by complete-data ML, do robust score
or profile-LR intervals improve on robust Wald intervals under non-normal
sampling, and at what cost?

## Short answer

**The benefit depended on the parameter.** In the difficult N = 100 skewed
correlation cell, Score-E covered 94.9% of valid intervals, versus 89.6% for
Wald-O; counting failures against all attempts reduced Score-E to 92.6%.
LR-E reached 93.8% conditional coverage and 90.8% successful-and-covering.
For the skewed loading, all three remained below nominal coverage (88.6–90.8%).

Observed-information inversion failed in up to 24.2% of one cell under the
prespecified search rule. Expected-information inversion was more reliable,
but cost roughly 26–33 times as much as Wald here. These 3,000 datasets support
further study of score inversion, with no ordinary-user default change.

## Evidence

The comparison uses the same datasets and unrestricted estimates for all
methods. Wald-O uses observed information; Score-E and LR-E use expected
information at each candidate restriction. The secondary methods change only
that information choice. All six robust methods use uncentered empirical score
second moments. The three uncorrected methods are diagnostic controls.

```{r}
#| fig-width: 10
#| fig-height: 6.2
#| fig-cap: "Coverage among valid intervals. Bars show approximate 95% Monte Carlo intervals; the dashed line is 95% coverage. Failures are reported separately below."
layout(rbind(c(7,7,7),matrix(1:6,nrow=2,byrow=TRUE)),heights=c(.35,3,3))
par(mar=c(3,3,2.2,.6),oma=c(.5,2,0,0))
for(target in c("loading","correlation")) for(d in c("normal","t10","heterogeneous_gamma")) {
  z <- s[s$target==target & s$distribution==d & s$method%in%primary,]
  yr <- c(85,99)
  plot(NA,xlim=c(.65,2.35),ylim=yr,xaxt="n",xlab="",ylab="",
       main=paste(if(target=="loading") "Loading" else "Correlation",
                  c(normal="Normal",t10="t(10)",heterogeneous_gamma="Gamma")[[d]],sep=" · "))
  axis(1,1:2,c("N = 100","N = 300"));abline(h=95,lty=2,col="grey60")
  for(j in seq_along(primary)) {
    v <- z[z$method==primary[j],];v <- v[order(v$n),]
    xx <- 1:2+(j-2)*.12;yy <- 100*v$coverage;err <- 196*v$coverage_mcse
    arrows(xx,yy-err,xx,yy+err,angle=90,code=3,length=.04,col=colours[j])
    points(xx,yy,pch=16,col=colours[j]);lines(xx,yy,col=colours[j],lty=3)
  }
}
mtext("Coverage (%)",side=2,outer=TRUE,line=.5)
par(mar=rep(0,4));plot.new()
legend("center",legend=labels[primary],col=colours,pch=16,horiz=TRUE,bty="n")
```

For N = 100 skewed correlations, the common-valid paired improvement over
Wald-O was 5.53 percentage points for Score-E (Monte Carlo SE 1.04 points)
and 4.55 points for LR-E (SE 0.99). On the 484 datasets valid for all three,
coverage was 89.3%, 94.8% and 93.8%, respectively. This supports the direction
of the correlation finding without ignoring its search failures. In the same
population, the loading difference was only +0.4 points for Score-E (SE 0.63)
and −1.8 points for LR-E (SE 0.91).

Loading errors were asymmetric: at N = 100 under skewness, the true loading
lay above the upper endpoint in 7.2% of Wald-O intervals, 8.0% of Score-E
intervals and 10.2% of LR-E intervals. Thus wider intervals alone did not
repair the problematic tail. Across cells, mean Score-E widths were 0.9–8.8%
larger than Wald-O and LR-E widths 0.6–6.5% larger, using each method's valid
intervals; paired width differences are also retained.

```{r}
#| fig-width: 8
#| fig-height: 3.2
#| fig-cap: "Skewed loading, N = 100: which side misses? All 500 intervals per primary method were valid. The dashed line marks 5% total noncoverage."
z <- s[s$n==100 & s$distribution=="heterogeneous_gamma" & s$target=="loading",]
z <- z[match(primary,z$method),]
par(mar=c(4,5,3,.5))
barplot(100*t(as.matrix(z[c("lower_miss","upper_miss")])),horiz=TRUE,
        names.arg=labels[primary],las=1,col=c("#56B4E9","#E69F00"),border=NA,
        xlim=c(0,14),xlab="Noncoverage (%)")
abline(v=5,lty=2,col="grey40")
legend("top",inset=c(0,-.25),legend=c("Truth below interval","Truth above interval"),
       fill=c("#56B4E9","#E69F00"),horiz=TRUE,bty="n",cex=.9,xpd=NA)
```

The uncorrected controls show why non-normal inference matters: for the
t(10) correlation at N = 300, uncorrected coverage was 90.0–90.4%, compared
with 94.2–95.0% for the primary robust methods. Robust calibration was not
uniformly better in the normal control: the N = 100 loading had 92.6% LR-E
coverage versus 94.6% for uncorrected LR. These are descriptive controls,
not additional candidates for a robust default.

The full per-cell coverage, tail misses, widths and failure counts are in
[summary.csv](results/frozen/summary.csv). Paired coverage differences and
Monte Carlo standard errors use datasets valid for both compared methods in
[paired.csv](results/frozen/paired.csv); the stricter subset valid for all three
primary methods is in [common_valid.csv](results/frozen/common_valid.csv).
The successful-and-covering rate counts failures against all 500 attempts.

```{r}
show <- c(primary,"wald_E","score_O","lr_O")
cost <- do.call(rbind,lapply(show,function(m) {
  z <- s[s$method==m,]
  data.frame(Method=labels[[m]],`Valid / attempted`=paste(sum(z$valid),"/",sum(z$attempts)),
             `Mean time (ms)`=round(1000*mean(z$mean_seconds),1),
             `Mean refits`=round(mean(z$mean_refits),1),
             `Correlation endpoints outside domain`=sum(z$out_of_domain),check.names=FALSE)
}))
knitr::kable(cost,align=c("l","r","r","r","r"))
```

Wald-O returned all 6,000 intervals; Score-E failed 15 times and LR-E 18.
Score-O failed 148 times and LR-O 211. Failures were concentrated: for
N = 100 under skewness, LR-O failed on 121/500 loading intervals (24.2%)
and Score-O on 96/500 correlation intervals (19.2%). Across all cells,
293 observed-information failures involved nonpositive sensitivity and 66
involved an unresolved acceptance set. Of their 359 failures, 293 occurred
during the wider audit, beyond the candidate endpoint searches.

All 18 LR-E failures and 14 of 15 Score-E failures ended at a candidate
correlation outside [-1,1]. The remaining Score-E failure was an endpoint
check. These domain/search events warrant follow-up before policy exposure;
they are not replaced by Wald. The failure types and search stages are retained
in [failures.csv](results/frozen/failures.csv) and
[failure_stages.csv](results/frozen/failure_stages.csv).

Times are incremental per attempted interval, including failed searches, and
exclude the shared unrestricted fit (about 9 ms per dataset). These local measurements include four
concurrent workers. The mean shared fit cost is recorded separately in
[timing_summary.csv](results/frozen/timing_summary.csv). Widths are reported
in the original parameter units; no interval is clipped to the correlation
domain. Primitive covariance inadmissibility during profile searches is
flagged in the summary rather than silently repaired. Sixteen of the 3,000
unrestricted fits were primitive-inadmissible; they remain in all comparisons.
Wald-O and Wald-E each had one correlation interval outside the domain.

The final audit checked 922,207 candidate evaluations. All valid inversions
agreed with the pointwise test at truth, every recorded equality residual was
zero, and the largest valid endpoint-statistic gap was 0.00000243 (tolerance
0.0001). All 66 numerical validation checks passed, as did method-order
invariance and checkpoint reproduction. The compact audit is in
[integrity.csv](results/frozen/integrity.csv).

## Caveats

These are two interior targets in one correctly specified covariance model.
The study does not establish performance under covariance misspecification,
at boundaries, with missing or ordinal data, or with penalized fits. A method
that fails more often cannot be judged from its successful-interval coverage
alone. Monte Carlo standard errors near 95% coverage are about one percentage
point per cell, and the two targets from each dataset are paired.

The profile search checks a finite wider grid for additional crossings. That
is a numerical audit, not a proof that the acceptance set has no further
components. Unresolved searches remain failures; their convex hull is never
reported as a confidence interval. Fits require a positive implied covariance,
but the latent correlation can leave its admissible range during the search.
Some failures occur only during the wider audit, beyond the proposed interval;
the counts characterize this search procedure, not mathematical nonexistence
of a confidence set. Robust scalar scaling is an asymptotic
correction and does not eliminate finite-sample uncertainty in the estimated
scale. Centered and uncentered empirical score moments can differ at a
restricted fit, so these results should not be transferred silently between
those conventions. Bartlett and bootstrap corrections are outside this study.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --smoke --output results/smoke-confirmation
Rscript run_experiment.R --main --reps 50 --seed-base 2026132751 --output results/confirmation
Rscript run_experiment.R --main --reps 500 --seed-base 2026132751 --output results/confirmation
Rscript scripts/freeze_results.R results/confirmation
quarto render report.qmd --to html
```

The 50-per-cell confirmation pilot is a prefix of the full run, using unchanged
computation. The original-seed development pilot is retained locally under
`results/main/`; a logging refinement preserves constrained-fit convergence
when a later information calculation fails. The confirmation run uses a fresh
seed after that refinement. No population or method was selected or changed
using pilot coverage. All numerical validation checks must pass before a run.
Source and package-binary hashes protect resumption against changed computation.

The small frozen CSV summaries and provenance under `results/frozen/` are
committed so the findings remain reviewable without rerunning the simulation.
Raw intervals, complete candidate traces and per-replicate checkpoints stay
local under `results/confirmation/`. Rendering reads only the frozen results.

## Prespecified design

Use eight centered continuous indicators, four per factor. Fix both factor
variances to one and set their correlation to 0.4. Loadings on each factor are
(0.8, 0.7, 0.9, 0.6); residual variances equal one minus the squared loading, with zero
residual covariances. Fit all eight loadings, eight residual variances and the factor
correlation, retaining unit factor variances and positive factor orientation.
Targets are the second loading (0.7) and factor correlation (0.4), tested
separately with all other free parameters treated as nuisance parameters.

Define mean-zero, unit-variance innovations z1 through z10.
Construct f1 = z1, f2 = 0.4 z1 + sqrt(1 - 0.4^2) z2, and indicator residuals
from z3 through z10 times their population residual SDs. Fix this construction;
changing the covariance square root changes the non-normal population.

Use three distributions:

- **Normal control:** all innovations standard normal.
- **Symmetric heavy tails:** begin with ten independent standard normals,
  then multiply the entire ten-dimensional vector for each observation by
  sqrt(8/U), where U is an independent chi-square variate with 10 degrees of
  freedom. The resulting indicators have standardized multivariate t(10)
  sampling and the same population covariance as the normal control.
- **Heterogeneous skewness:** independent standardized gamma innovations
  (G-k)/sqrt(k), G ~ Gamma(shape=k, scale=1), with shapes
  (2, 0.5, 2, 0.5, 2, 0.5, 2, 0.5, 2, 0.5) in z1,...,z10 order.

These constructions preserve the target covariance analytically. Do not
force simulated sample covariances to equal that target. Center each sample
and use the ML covariance divisor N. Use N = 100 and 300 with 500 datasets per
distribution/sample-size cell: 3,000 datasets and two targets per dataset.
The prespecified seed bases were 2026092751 for the main run and
2026102751 for smoke runs. The fresh confirmation base is 2026132751.

Compare 95% robust Wald, robust efficient-score inversion and robust profile-LR
inversion on the same datasets and unrestricted fits. The primary comparison
uses the current ordinary-user Wald policy and expected-information geometry
for the two inverted tests. A prespecified secondary comparison changes the
information choice within each method. Exact definitions are fixed in the
implementation requirements before simulation. Include uncorrected intervals
as diagnostics of the effect of non-normality, never as robust candidates.

Use the same unpenalized ML objective, optimizer, bounds, starts policy and
covariance domain across methods. Record them before running the pilot.
Boundary-active fits are flagged and reported separately; do not replace them
with repaired or penalized estimates. Avoid post-hoc changes to the population
or method definitions in response to observed coverage.

## Outcomes and run stages

First validate the numerical constructions. Then run two datasets per cell
with both targets to measure cost and detect failures. After passing these
checks, run 50 datasets per cell, followed by the full 500 using deterministic
replicate IDs and resumable output. The first 50 can be a prefix of the final
run only if code and method definitions remain unchanged. After a statistical
or numerical fix, retain old results as development runs and use a fresh seed
base for the confirming run.

Record coverage and each tail's noncoverage, mean and median width, endpoint
asymmetry, inversion failures, boundary events, refit counts and elapsed time.
Report method-specific successful-interval coverage, the successful-and-covering
fraction of all attempts, and paired method differences on common-valid
datasets. Never replace a failed score or LR interval with Wald. Save enough
candidate-level diagnostics to explain failed or disconnected confidence sets.

Timing includes candidate corrections and all constrained refits. Report
shared unrestricted-fit cost separately from incremental interval cost.
Present Monte Carlo uncertainty and paired differences; do not declare a
winner from a small apparent coverage improvement or pool away a poor cell.
