---
title: "Confidence Intervals for Maximal Reliability in Bifactor Models"
format:
  html:
    toc: true
    number-sections: false
    fig-width: 9
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
---

```{r setup}
res_dir <- params$results_dir
need <- file.path(res_dir, c("coverage.csv", "metadata.csv"))
if (!all(file.exists(need))) {
  stop("Missing result files. Run: Rscript experiments/43-li-savalei-2026-maximal-reliability-ci/run_experiment.R",
       call. = FALSE)
}
cov <- read.csv(file.path(res_dir, "coverage.csv"))
meta <- read.csv(file.path(res_dir, "metadata.csv"))
meta <- setNames(meta$value, meta$key)
boot_path <- file.path(res_dir, "bootstrap.csv")
boot <- if (file.exists(boot_path)) read.csv(boot_path) else NULL

has_gg <- requireNamespace("ggplot2", quietly = TRUE)
method_lab <- c(wald_model = "normal SE", wald_robust = "robust SE",
                logit_model = "normal SE, logit", logit_robust = "robust SE, logit",
                boot_perc = "bootstrap, percentile", boot_basic = "bootstrap, basic")
coef_lab <- c(gen = "general factor", grp = "group factor")
cov$method <- factor(cov$method,
  levels = c("wald_model", "wald_robust", "logit_model", "logit_robust"))
g <- function(coef, p, n, dist, method, col = "coverage")
  cov[[col]][cov$coef==coef & cov$p==p & cov$n==n & cov$dist==dist & cov$method==method]
fp <- function(x) sprintf("%.2f", x)
```

## Question

Li & Savalei (2026, *MBR* 61:3, 469-490) derive the correct point formulas for
maximal reliability in a bifactor model (an optimally weighted composite for the
general factor, and the corrected coefficient H for a group factor) but leave
standard errors and confidence intervals to future work. At the sample sizes these
models are actually fit at, can you put a trustworthy interval around either
coefficient, and what does it take?

## Short answer

The obvious interval does not work and the reason is not the standard error. A
textbook 95% interval (ordinary standard error, no transform) covers the
general-factor coefficient as little as
**`r fp(g("gen",18,50,"normal","wald_model"))`** of the time at 18 indicators and
N=50, and is either uselessly wide or under-covering for the weaker group factor. A
robust standard error fixes part of it, a logit-scale interval fixes another part,
and only the two together give a usable interval: near nominal by N around 200 at 9
indicators and N around 500 at 18. But even that best interval falls short in the
hardest cells, because the estimate itself is biased upward, by about
**`r fp(g("grp",18,50,"normal","wald_model","bias"))`** (more than half the true
value) for the group-factor coefficient at 18 indicators and N=50. A nonparametric
bootstrap confirms it: the percentile interval inherits the same bias and
under-covers, while an interval that reflects the bias out does markedly better. So
the missing piece Li & Savalei point to is a *bias correction*, not just a variance
estimate. The coefficients sit in the two regimes they describe: the general factor
is reliable (population value
`r fp(unique(cov$rho_pop[cov$coef=="gen" & cov$p==9]))`-`r fp(unique(cov$rho_pop[cov$coef=="gen" & cov$p==18]))`),
the group factor is not
(`r fp(unique(cov$rho_pop[cov$coef=="grp" & cov$p==9]))`-`r fp(unique(cov$rho_pop[cov$coef=="grp" & cov$p==18]))`).

## Evidence

### A usable interval needs both fixes

```{r fig-coverage}
#| fig-cap: "Coverage of nominal-95% intervals vs sample size (1000 replications). Rows: coefficient; columns: number of indicators x distribution. Dashed line is 0.95."
if (has_gg) {
  library(ggplot2)
  d <- cov; d$panel <- paste0(d$p, " indicators, ", d$dist); d$coeff <- coef_lab[d$coef]
  ggplot(d, aes(n, coverage, colour = method, group = method)) +
    geom_hline(yintercept = 0.95, linetype = 2, colour = "grey50") +
    geom_line() + geom_point(size = 1.3) +
    facet_grid(coeff ~ panel) +
    scale_x_continuous(trans = "log10", breaks = sort(unique(d$n))) +
    scale_colour_discrete(labels = method_lab[levels(d$method)]) +
    labs(x = "N (log scale)", y = "coverage", colour = NULL) +
    coord_cartesian(ylim = c(0.35, 1)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "top", panel.grid.minor = element_blank())
} else {
  knitr::kable(cov[cov$n %in% c(50,200), c("coef","p","dist","n","method","coverage")], digits = 2)
}
```

The ordinary interval under-covers the general factor, worse as the model grows and
under non-normal data. The robust standard error repairs the variance
under-estimation; the logit scale repairs the boundary distortion and the absurd
width at low reliability (the group-factor interval at 9 indicators, N=50 shrinks
from a mean width of `r fp(g("grp",9,50,"normal","wald_model","mean_width"))` to
`r fp(g("grp",9,50,"normal","logit_model","mean_width"))` at the same coverage).
Together they are the best simple interval, near nominal once N is moderate.

### What it cannot fix is the bias

```{r fig-bias}
#| fig-cap: "Upward bias of the maximal-reliability estimate (mean estimate minus population value), by coefficient and model size. Independent of the interval method."
b <- cov[cov$method == "wald_model", ]
if (has_gg) {
  b$coeff <- coef_lab[b$coef]; b$panel <- paste0(b$p, " indicators")
  ggplot(b, aes(n, bias, colour = dist, group = dist)) +
    geom_hline(yintercept = 0, colour = "grey60") +
    geom_line() + geom_point(size = 1.4) +
    facet_grid(coeff ~ panel) +
    scale_x_continuous(trans = "log10", breaks = sort(unique(b$n))) +
    labs(x = "N (log scale)", y = "bias of the estimate", colour = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "top", panel.grid.minor = element_blank())
} else {
  knitr::kable(b[, c("coef","p","dist","n","bias","mean_est","rho_pop")], digits = 3)
}
```

The estimate is biased upward, more so with more free loadings and less so with N.
For the group factor at 18 indicators, N=50, the bias is about
`r fp(g("grp",18,50,"normal","wald_model","bias"))` on a true value of
`r fp(unique(cov$rho_pop[cov$coef=="grp" & cov$p==18]))`: roughly half the estimate.
No interval centered on the estimate can recover the true value when it is displaced
that far, which is why every method still falls short there. A nonparametric
bootstrap makes the point directly:

```{r boot, eval = !is.null(boot)}
boot$method <- factor(boot$method, levels = c("boot_perc", "boot_basic"))
bt <- boot[order(boot$coef, boot$dist, boot$method),
           c("coef","dist","n","method","coverage","mean_width")]
bt$coef <- coef_lab[bt$coef]; bt$method <- method_lab[as.character(bt$method)]
names(bt) <- c("Coefficient","Distribution","N","Interval","Coverage","Mean width")
knitr::kable(bt, row.names = FALSE, digits = 2,
  caption = "Bootstrap intervals at 18 indicators, N=100. The percentile interval, built around the biased estimate, under-covers; the basic interval reflects the bias out and recovers much of the coverage.")
```

```{r boot-missing, eval = is.null(boot), results = "asis"}
cat("*Bootstrap reference not yet generated. Run `Rscript",
    "experiments/43-li-savalei-2026-maximal-reliability-ci/scripts/bootstrap.R --pkeys p18 --ns 100`.*\n")
```

## Caveats

- One model family: an orthogonal bifactor with three subscales, equal loadings,
  correct specification, complete data, continuous ML. The two distributions are
  normal and a strongly skewed one with the same covariance.
- The bias is the headline obstacle but this run only quantifies it; it does not
  build or test a bias-corrected estimate or a BCa-type interval, which is the
  natural next step.
- The bootstrap arm is a focused reference at 18 indicators, N=100. A full bootstrap
  sweep is impractical at the smallest cells: at 9 indicators and N=50 the resampled
  bifactor models frequently fail to converge, which is itself a reason to prefer an
  analytic bias-corrected interval over bootstrapping here.
- Coverage rates are from 1000 replications (the bootstrap arm from 150); read them
  as approximate.

## Reproduce

```sh
# main coverage sweep (~3 min)
Rscript experiments/43-li-savalei-2026-maximal-reliability-ci/run_experiment.R
# focused bootstrap reference (~6 min)
Rscript experiments/43-li-savalei-2026-maximal-reliability-ci/scripts/bootstrap.R --pkeys p18 --ns 100
quarto render experiments/43-li-savalei-2026-maximal-reliability-ci/report.qmd
```

Smoke paths: add `--smoke` to either runner. This experiment is the finite-sample
evidence behind the maximal-reliability entry in the project backlog; see the paper
eval (`docs/research/paper-evals/2026-li-bifactor-maximal-reliability.md`) for
context.

```{r meta}
show <- meta[c("model","loadings","coefficients","estimator","reps","ns","distributions")]
knitr::kable(data.frame(Setting = names(show), Value = unname(show)),
             row.names = FALSE, caption = "Run settings.")
```
