---
title: "Small-sample intervals for covariance functionals"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
read_evidence <- function(name) {
  path <- file.path("results", "overview", name)
  if (!file.exists(path)) stop("Missing retained summary: ", path,
                               "; recover original CSVs and run scripts/summarize.R")
  read.csv(path, stringsAsFactors = FALSE)
}
show_table <- function(x, digits = 3) knitr::kable(x, digits = digits, row.names = FALSE)
```

## Question

Can an effective-df reference and a variance-stabilizing transform address
different small-sample errors in intervals for covariance functionals?

## Short answer

Keep this as a **banked covariance-functional interval study**. Correlation and
alpha are demonstrations of a broader idea: an influence-kurtosis effective-df
$t$ reference changes tail thickness, while Fisher $z$ or logit changes interval
asymmetry. The retained 300-replication grid is exploratory evidence, not a
distribution-free guarantee or an adopted interval recipe.

## Evidence

The runner estimates casewise influence for a covariance functional, its robust
delta SE and effective degrees of freedom from influence kurtosis. It compares
ordinary Wald (`wald`), effective-df $t$ (`kc`), transformed Wald (`twald`), and
transformed effective-df $t$ (`tkc`). Transform choice is functional-specific:
Fisher $z$ for correlation and logit for alpha. Their effects need not compose
well for every population.

The local run has 300 attempts per cell, $N=20,100$, correlation $0,.9$, and
normal/contaminated laws; the alpha population is about .852. Failure/nonfinite
interval handling remains in the original runner. The displayed $N=20$ contaminated slice keeps
both missed tails and interval width alongside coverage. The
[full summary](results/overview/coverage.csv) retains the other sample size and normal controls:

```{r intervals}
x <- read_evidence("coverage.csv")
x <- x[x$n == 20 & x$law == "contaminated", ]
x$functional <- ifelse(x$quantity == "alpha", "Alpha", paste0("Correlation ",x$param))
show_table(data.frame(Functional=x$functional, Method=x$method, Attempts=x$reps,
  `Coverage (%)`=100*x$coverage, `Miss lower (%)`=100*x$miss_lower,
  `Miss upper (%)`=100*x$miss_upper, `Mean width`=x$mean_width, check.names=FALSE), digits=1)
```

At 95% coverage, 300 independent usable draws give a binomial MCSE of roughly
1.3 percentage points. Differences of a few points here need fresh confirmation.

## Caveats

Influence-based robust intervals still need the relevant finite moments and a
stable covariance estimator. Heavy-tail and near-boundary behavior may defeat
both corrections. The run's historical `smoke` label and its limited design
remain in the metadata. No general default follows from these cells.

Reopen for a named covariance-functional consumer with demonstrably inadequate
ordinary coverage, then evaluate one prespecified recipe on independent
heavy-tail and boundary cases. Existing delta/Wald inference is the cheaper
available approach until that trigger is met.

## Reproduce

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --reps 10 --ns 20 --rhos 0,0.9 \
  --results-dir /tmp/covariance_intervals_check
Rscript scripts/summarize.R
```

The runner now honors `--results-dir`; historical results stay untouched.
[Source hashes](results/overview/sources.csv) identify the retained evidence.
