---
title: "Does PSD-ML reach the same maximum from different starts?"
subtitle: "A multistart basin audit for the small-N SEM"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)
screen <- read_result("screen.csv")
datasets <- read_result("dataset_summary.csv")
classification <- read_result("classification_summary.csv")
starts <- read_result("start_summary.csv")
candidates <- read_result("confirmation_candidates.csv")
has_confirmation <- file.exists(file.path(root, "confirmation_summary.csv"))
confirmation <- if (has_confirmation) {
  read_result("confirmation_summary.csv")
} else {
  data.frame()
}
has_profiles <- file.exists(file.path(root, "profile_summary.csv"))
profiles <- if (has_profiles) {
  read_result("profile_summary.csv")
} else {
  data.frame()
}

profile <- metadata[["profile"]]
core <- classification[classification$scope == "random_core", ]
is_evidence <- profile == "pilot" &&
  as.integer(metadata[["core_reps"]]) >= 100L
fmt_pct <- function(x) sprintf("%.1f%%", 100 * x)

screen_rows <- lapply(split(screen, screen$n), function(x) {
  data.frame(
    N = x$n[[1L]],
    Screened = nrow(x),
    `Default KKT + admissible` = fmt_pct(mean(x$eligible)),
    `Default covariance boundary` = fmt_pct(mean(x$covariance_boundary)),
    check.names = FALSE
  )
})
screen_display <- do.call(rbind, screen_rows)

core_display <- core[, c(
  "n", "datasets", "default_best_rate", "ordinary_warm_best_rate",
  "competing_basin_rate", "different_parameter_rate",
  "different_moment_rate", "start_sensitive_rate",
  "best_boundary_rate", "max_default_gap"
)]
names(core_display) <- c(
  "N", "Datasets", "Default hits best", "Ordinary warm hits best",
  "Competing basins", "Same fit/different parameters",
  "Equal objective/different moments", "Start-sensitive eligibility",
  "Best solution on boundary", "Maximum default objective gap"
)
for (field in c(
  "Default hits best", "Ordinary warm hits best", "Competing basins",
  "Same fit/different parameters", "Equal objective/different moments",
  "Start-sensitive eligibility", "Best solution on boundary"
)) {
  core_display[[field]] <- vapply(
    core_display[[field]], fmt_pct, character(1)
  )
}
core_display$`Maximum default objective gap` <- format(
  core_display$`Maximum default objective gap`,
  scientific = TRUE, digits = 3
)

start_core <- starts[starts$scope == "random_core", c(
  "n", "start_family", "attempts", "eligible_rate", "best_hit_rate",
  "median_elapsed_ms", "median_f_evals"
)]
start_core$eligible_rate <- vapply(
  start_core$eligible_rate, fmt_pct, character(1))
start_core$best_hit_rate <- vapply(
  start_core$best_hit_rate, fmt_pct, character(1))
names(start_core) <- c(
  "N", "Start family", "Attempts", "Eligible", "Hits best attained",
  "Median time (ms)", "Median objective evaluations"
)

default_miss <- sum(
  datasets$random_core & !datasets$default_best
)
core_n <- sum(datasets$random_core)
upper <- if (core_n) {
  stats::binom.test(default_miss, core_n, alternative = "less")$conf.int[[2L]]
} else {
  NA_real_
}
competing <- sum(
  datasets$random_core &
    datasets$objective_basin_class == "competing"
)
different_moments <- sum(
  datasets$random_core &
    datasets$best_equivalence_class ==
      "equal-objective-different-moments"
)
different_parameters <- sum(
  datasets$random_core &
    datasets$best_equivalence_class ==
      "same-fit-different-parameters"
)
candidate_n <- nrow(candidates)
confirmed_display <- if (has_confirmation) {
  do.call(rbind, lapply(sort(unique(datasets$n)), function(n_value) {
    core_n_value <- sum(datasets$random_core & datasets$n == n_value)
    initial_n <- sum(
      datasets$random_core & datasets$n == n_value &
        datasets$objective_basin_class == "competing"
    )
    confirmed_n <- sum(
      confirmation$random_core & confirmation$n == n_value &
        confirmation$persistent_competing
    )
    data.frame(
      N = n_value,
      `Random-core datasets` = core_n_value,
      `Initial competing basins` = initial_n,
      `Persistent after tight restart` = confirmed_n,
      `Persistent rate` = fmt_pct(confirmed_n / core_n_value),
      check.names = FALSE
    )
  }))
} else {
  data.frame()
}
profile_display <- if (has_profiles) {
  do.call(rbind, lapply(sort(unique(profiles$n)), function(n_value) {
    subset <- profiles[profiles$n == n_value, ]
    multiple <- sum(subset$local_envelope_minima >= 2L)
    data.frame(
      N = n_value,
      `Profiled cases` = nrow(subset),
      `At least two envelope minima` = multiple,
      `Rate among targeted cases` = fmt_pct(multiple / nrow(subset)),
      check.names = FALSE
    )
  }))
} else {
  data.frame()
}
short_answer <- if (!is_evidence) {
  sprintf(
    paste(
      "This is a %s run. It validates the multistart machinery but is too",
      "small for a basin-frequency conclusion."
    ),
    profile
  )
} else {
  sprintf(
    paste(
      "Across the %d-dataset random core, the default start missed the",
      "best attained admissible KKT objective %d times (one-sided 95%%",
      "upper bound %.1f%%). We found %d datasets with competing stationary",
      "objective basins, %d with equal best objectives but different",
      "implied moments, and %d with the same fit but distinct parameter",
      "representatives. %d enriched or random-core datasets require the",
      "separate confirmation stage."
    ),
    core_n, default_miss, 100 * upper, competing, different_moments,
    different_parameters, candidate_n
  )
}
if (is_evidence && has_confirmation) {
  confirmed_total <- sum(
    confirmation$random_core & confirmation$persistent_competing
  )
  short_answer <- paste0(
    short_answer,
    sprintf(
      " Tight 20,000-iteration restarts retained competing basins in %d ",
      confirmed_total
    ),
    "random-core datasets."
  )
}
if (is_evidence && has_profiles) {
  multiple_profiles <- sum(profiles$local_envelope_minima >= 2L)
  short_answer <- paste0(
    short_answer,
    sprintf(
      " Of %d targeted beta profiles, %d had at least two local minima ",
      nrow(profiles), multiple_profiles
    ),
    "in the feasible lower envelope."
  )
}
```

## Question

Does covariance-honest complete-data ML converge to the same likelihood
maximum from a reasonable portfolio of standard, warm, restart, and perturbed
starts?

## Short answer

`r short_answer`

“Best attained” means the smallest objective among the admissible,
KKT-stationary solutions found by the prespecified portfolio. It is not a
claim that a global maximum has been proved.

## Screen

The screen independently generates the \(N=10,20,50\) datasets and runs the
standard PSD-SLSQP fit once. A random core supports frequency statements.
Every default failure and the largest-\(|\hat\beta|\) estimates are added only
as diagnostic enrichment.

```{r screen-table}
knitr::kable(screen_display, row.names = FALSE)
```

## Basin audit

Objective values, original partable parameters, and implied moments are
clustered separately. This distinguishes an inferior stationary basin from an
observationally equivalent parameter representative. Start-sensitive
eligibility is another axis: it is possible for some starts to fail even when
all successful starts reach one objective basin.

```{r core-table}
knitr::kable(core_display, row.names = FALSE)
```

The objective-equivalence threshold is
\(10^{-6}\max(1,|F_\star|)\), with raw gaps retained for sensitivity checks.
Parameter and implied-moment equivalence use scaled distances of \(10^{-4}\)
and \(10^{-6}\), respectively.

## Start families

```{r start-table}
knitr::kable(start_core, row.names = FALSE)
```

The ordinary-ML fit is used only as a source of starting values. An
inadmissible ordinary estimate is not treated as a candidate solution in the
PSD model class.

## Confirmation boundary

Only competing objective basins or equal-objective/different-moment cases are
written to `confirmation_candidates.csv`. They receive high-budget restarts
and a bidirectional profile over the structural coefficient before being
described as evidence of multiple maxima. IPOPT remains an optional backend
cross-check from the near-solution representatives; the pilot did not require
it after experiment research/42 established its non-finite-callback failure mode. The
base runner deliberately does not launch adaptive confirmation automatically.

```{r confirmation-results}
if (has_confirmation) {
  cat(
    "One representative from each initial cluster was restarted with a ",
    "20,000-iteration budget and tighter tolerances. The table reports ",
    "persistence against the original 100-dataset random core at each N.\n\n",
    sep = ""
  )
  print(knitr::kable(confirmed_display, row.names = FALSE))
}
```

All initially flagged equal-objective/different-moment cases collapsed under
the tight restart. They are therefore treated as numerical convergence
artifacts, not distinct fitted distributions.

```{r profile-results}
if (has_profiles) {
  cat(
    "\nThe beta-profile tier targeted persistent random-core cases, ",
    "prioritizing default misses with the largest objective gaps. ",
    "It is diagnostic and intentionally not a frequency sample.\n\n",
    sep = ""
  )
  print(knitr::kable(profile_display, row.names = FALSE))
}
```

## Caveats

This experiment concerns one correctly specified two-factor Gaussian SEM and
one explicit start portfolio. Failure to find competing basins is evidence
about that design, not a general proof of globality for SEM likelihoods.
Boundary maxima are retained without regularization because they belong to
the specified PSD model class.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript run_confirmation.R
Rscript run_profiles.R
quarto render report.qmd
```

## Study design

### Model and data

Use a two-factor, six-indicator Gaussian SEM:

```text
Y =~ y1 + y2 + y3
X =~ x1 + x2 + x3
Y ~ X
```

The population loadings are \(1.0, 0.8, 0.6\), the structural coefficient is
\(0.25\), the exogenous \(X\) variance and \(Y\) disturbance variance are one,
and all six indicator residual variances are one. The experiment is
self-contained: it reimplements this short DGP and does not read another
experiment's source or results.

The pilot uses \(N\in\{10,20,50\}\). These cells cover the severe-boundary
regime, the transition regime, and a mostly stable control. For each \(N\),
screen 1,000 datasets under the standard PSD-SLSQP start using the documented
seed rule

```text
seed = 20260729 + 1000003 * N + replication
```

The screen supplies the default result and defines the replay sample:

1. a deterministic simple random sample of 100 datasets per \(N\);
2. every dataset where the default fit fails to return or fails the terminal
   KKT audit; and
3. the 25 audit-converged datasets per \(N\) with the largest
   \(|\hat\beta|\).

Deduplicate datasets while retaining all selection tags. Primary frequency
statements use only the random core. The failure and extreme-estimate
enrichments are diagnostic and are reported separately, never pooled into a
prevalence estimate.

### Start portfolio

All primary fits use PSD-ML with NLopt SLSQP. Starts are expressed in the
ordinary free partable coordinates; the production PSD routine constructs its
usual covariance-factor start from each vector. Shared free indices are
perturbed once, so equality-label semantics are preserved.

For every replay dataset, run:

1. the standard FABIN start;
2. the ordinary L-BFGS ML estimate as a warm start, when that fit returns;
3. the default PSD estimate as a restart, when it returns;
4. three moderate perturbations of the FABIN start;
5. three broad perturbations of the FABIN start;
6. two moderate perturbations of the ordinary-ML start; and
7. two broad perturbations of the ordinary-ML start.

This gives at most 13 starts per dataset. Perturbation directions are
deterministic standard-normal draws indexed by dataset seed, start family, and
start number. For free coordinate \(k\), use

\[
  \theta^{(j)}_k =
  \theta^{(0)}_k + a_j\max(|\theta^{(0)}_k|,0.25)z_{jk},
\]

with \(a_j=0.10\) for moderate and \(a_j=0.50\) for broad starts. Do not redraw
or silently repair a start that fails; record the failure. Primitive covariance
starts still pass through the estimator's documented eigenvalue projection.

### Eligibility and clustering

A returned solution is eligible only if:

- the equality-constrained terminal audit reports convergence;
- the primitive covariance audit reports admissibility; and
- the objective and implied moments are finite.

For each dataset, let \(F_\star\) be the smallest objective among eligible
solutions. “Best attained” means

\[
  F-F_\star \leq 10^{-6}\max(1,|F_\star|).
\]

Always retain the raw objective gaps and repeat summaries at \(10^{-7}\) and
\(10^{-5}\). This makes the result insensitive to a single arbitrary numerical
cutoff.

Within the best-attained objective cluster, compare:

- free partable estimates using
  \(\max_k |\Delta\theta_k|/(1+|\theta_{\star k}|)\);
- implied covariance and mean vectors using relative Frobenius distance; and
- the fitted primitive-covariance ranks and minimum eigenvalues.

Use \(10^{-4}\) as the parameter-equivalence threshold and \(10^{-6}\) as the
implied-moment-equivalence threshold, while retaining the continuous
distances. Classify each dataset on three nonexclusive axes rather than forcing
distinct phenomena into one label:

1. **objective basins:** unresolved (no eligible solution), single eligible
   objective cluster, or competing eligible objective clusters;
2. **best-cluster equivalence:** unique parameters and moments; same implied
   moments but different parameters; or equal objective but different implied
   moments; and
3. **start eligibility:** all available starts eligible or start-sensitive
   eligibility.

“Stable unique” is the conjunction of one objective basin, unique
best-cluster parameters/moments, and eligibility from every available start.
Parameter multiplicity is not called multiple local maxima merely because
lift coordinates or observationally equivalent representatives differ.

### Confirmation of suspicious datasets

Escalate only datasets classified as competing stationary basins or equal
objective/different moments:

1. restart one representative from each cluster with a 20,000-iteration
   budget and tighter tolerances;
2. run IPOPT from those near-solution starts as an optional backend
   cross-check, recording failure rather than requiring it; and
3. profile the structural coefficient \(\beta\) over an adaptive grid spanning
   the attained estimates and \(0.25\), fixing \(\beta\) by a model equality
   and fitting nuisance parameters by warm continuation from both grid
   directions.

A competing maximum is reported only when distinct feasible KKT-stationary
clusters survive the high-budget restart. The profile is diagnostic evidence
about the principal structural direction, not a proof of globality over the
full nuisance space.

### Primary outputs

Report by \(N\), and separately for the random core and each enrichment:

- probability that the default start hits the best attained objective;
- probability that the ordinary-ML warm start hits it;
- number and gap of eligible objective clusters;
- rates and cross-tabulations of the three classification axes;
- return, KKT, and best-hit rates by start family;
- parameter and implied-moment distances within the best cluster; and
- covariance-boundary/rank summaries by basin classification.

The raw table has one row per dataset and start. A dataset summary table holds
the cluster assignments and best-attained comparisons. A cluster-membership
table retains the representative starts used by the confirmation stage.

### Profiles and stopping rule

- **Smoke:** 30 screened datasets per \(N\), 10 random-core datasets per
  \(N\), all screen failures, the three largest-\(|\hat\beta|\) datasets, and
  six starts (default, ordinary warm, and two perturbations around each).
- **Pilot:** the 1,000/100/25 design above and the full 13-start portfolio.
- **Full:** not predefined. Expand the random core or start portfolio only if
  the pilot finds a material default-to-best gap, competing stationary
  basins, or insufficient precision for their frequency.

If the pilot finds no competing basins and the upper confidence bound for the
random-core default-miss rate is below 1% after pooling only where the
\(N\)-specific results are compatible, stop. The conclusion is then limited
to this model, DGP, and start portfolio.


## Activity after the engineering review

Complete; retained as covariance-honest paper evidence. Keep the prespecified
random core, diagnostic enrichment and confirmation tiers distinct from later
sphere development cases. No result CSVs were present locally at the
30 September 2026 review: recover the original artifacts or reproduce the
prespecified run before rendering. Do not infer absence of evidence from the
local cache, or delete the paper-supporting pipeline.
