---
title: "Barrier penalties against improper solutions"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
profile <- Sys.getenv("E47_PROFILE", "full")
res <- file.path("results", profile)
need <- file.path(res, c("fits.csv", "params.csv", "design.csv", "truth.csv"))
if (!all(file.exists(need))) {
  stop("Missing results. Run: Rscript run_experiment.R --full", call. = FALSE)
}
fits <- read.csv(need[[1]], stringsAsFactors = FALSE)
params <- read.csv(need[[2]], stringsAsFactors = FALSE)
design <- read.csv(need[[3]], stringsAsFactors = FALSE)
truth <- read.csv(need[[4]], stringsAsFactors = FALSE)
key_truth <- setNames(design$key_truth, design$design)
key_proper <- setNames(design$key_truth_proper, design$design)
fits$err <- fits$key_est - key_truth[fits$design]
fits$adm <- fits$rls - fits$browne
params$cover <- abs(params$est - params$truth) <= 1.96 * params$se

design_label <- c(
  f1_p3 = "One factor, 3 indicators (.9, .3, .9)",
  f1_p5 = "One factor, 5 indicators (.9, .3, .9, .3, .9)",
  f2_r90 = "Two factors, r = .90",
  f2_r97 = "Two factors, r = .97",
  f2_r99 = "Two factors, r = .99",
  path_r2_90 = "Three-latent path, R²(f3) = .90",
  f1_p3_face = "One factor, 3 indicators (1, .3, .9)",
  f1_p5_face = "One factor, 5 indicators (1, .3, .9, .3, .9)",
  f2_r100 = "Two factors, r = 1",
  path_r2_100 = "Three-latent path, R²(f3) = 1",
  imp_f1_p5 = "One factor, 5 indicators, first loading 1.03",
  imp_f2_r103 = "Two factors, r = 1.03",
  mis_f2_resid = "Two factors, r = .95, omitted residual covariances")
key_label <- c(theta_x1 = "residual variance of x1",
               factor_correlation = "factor correlation", r2_f3 = "R² of f3")
key_of <- setNames(key_label[design$key_quantity], design$design)
interior <- c("f1_p3", "f1_p5", "f2_r90", "f2_r97", "f2_r99", "path_r2_90")
face <- c("f1_p3_face", "f1_p5_face", "f2_r100", "path_r2_100")
imp <- c("imp_f1_p5", "imp_f2_r103", "mis_f2_resid")
proper_d <- c(interior, face)
lambda_of <- c(pen_l010 = 0.1, pen_l025 = 0.25, pen_l050 = 0.5, pen_l100 = 1,
               pen_l200 = 2, det_l010 = 0.1, det_l025 = 0.25, det_l050 = 0.5,
               det_l100 = 1, det_l200 = 2)
pens <- c("pen_l010", "pen_l025", "pen_l050", "pen_l100", "pen_l200")
dets <- sub("pen", "det", pens)

pct <- function(x) sprintf("%.1f", 100 * x)
num <- function(x) sprintf("%.3f", x)
cell <- function(d, n, m) fits[fits$design == d & fits$n == n & fits$method == m, ]
rate <- function(d, n, m, f) mean(f(cell(d, n, m)), na.rm = TRUE)
key_rmse <- function(d, n, m) sqrt(mean(cell(d, n, m)$err^2, na.rm = TRUE))
key_bias <- function(d, n, m) mean(cell(d, n, m)$err, na.rm = TRUE)
key_mean <- function(d, n, m) mean(cell(d, n, m)$key_est, na.rm = TRUE)
improper <- function(d, n, m) rate(d, n, m, function(s) s$classification == "improper")
boundary <- function(d, n, m) rate(d, n, m, function(s) s$classification == "boundary")
failed <- function(d, n, m) rate(d, n, m, function(s) s$status != "ok" | !s$converged)
nonint <- function(s) s$classification != "interior" | s$status != "ok" | !s$converged
nonint_rate <- function(d, n, ms) {
  s <- fits[fits$design == d & fits$n == n & fits$method %in% ms, ]
  mean(nonint(s), na.rm = TRUE)
}
# Rejection at 5% against chi-square(df); `stat` is chi2 (LR), rls, or browne.
reject <- function(d, n, m, stat = "chi2") {
  s <- cell(d, n, m)
  s <- s[!is.na(s$df) & s$df > 0, ]
  if (!nrow(s)) return(NA_real_)
  mean(pchisq(s[[stat]], s$df, lower.tail = FALSE) < 0.05, na.rm = TRUE)
}
# The admissibility statistic rls - browne is the score statistic for the
# estimate's displacement from the ML optimum. At a single face with the truth
# on it, its limit law is U^2 1{U < 0} for PSD-ML and (t(U) - U)^2 for a
# barrier, U ~ N(0, 1), t(u) = (u + sqrt(u^2 + 4 lambda)) / 2.
t_lambda <- function(u, l) (u + sqrt(u^2 + 4 * l)) / 2
adm_crit <- function(m, alpha = 0.05) {
  z <- qnorm(alpha)
  if (m == "psd") return(z^2)
  (t_lambda(z, lambda_of[[m]]) - z)^2
}
adm_reject <- function(d, n, m) mean(cell(d, n, m)$adm > adm_crit(m), na.rm = TRUE)
min_cover <- function(d, n, m, reps = NULL) {
  s <- params[params$design == d & params$n %in% n & params$method == m, ]
  if (!is.null(reps)) s <- s[paste(s$n, s$rep) %in% reps, ]
  min(tapply(s$cover, s$param, mean, na.rm = TRUE))
}
tbl <- function(rows, cols, f, first = "Design", labels = design_label) {
  out <- data.frame(labels[rows], check.names = FALSE, row.names = NULL)
  names(out) <- first
  for (k in names(cols)) out[[cols[[k]]]] <- sapply(rows, function(d) f(d, k))
  out
}
kable_r <- function(tab) knitr::kable(tab, align = paste0("l", strrep("r", ncol(tab) - 1L)))
span <- function(x, f = pct) {
  x <- x[is.finite(x)]
  paste0(f(min(x)), " to ", f(max(x)))
}
bar_fits <- fits[fits$method %in% c(pens, dets), ]
n_bar <- nrow(bar_fits)
n_bar_err <- sum(bar_fits$status != "ok" | !bar_fits$converged, na.rm = TRUE)
n_bar_bad <- sum(bar_fits$status == "ok" & bar_fits$classification != "interior", na.rm = TRUE)
err_imp <- sum((bar_fits$status != "ok" | !bar_fits$converged) &
                 bar_fits$design == "imp_f1_p5", na.rm = TRUE)
tdes <- setdiff(proper_d, c("f1_p3", "f1_p3_face"))
face_ratio <- function(d, n = 400, m = "det_l025") key_rmse(d, n, m) / key_rmse(d, n, "psd")
fmt2 <- function(x) sprintf("%.2f", x)
```

## Question

Improper SEM solutions (negative residual variances, latent correlations
beyond one, latent R² of one) all say that some equation has no error. Two
log-barrier penalties add λ times a log-determinant to the log-likelihood. The
joint barrier uses the correlation matrix of the complete latent-plus-observed
vector, R. The latent-determinacy barrier, or Q-barrier, uses Q, the covariance
of the latents given the observed variables standardized by the latent
variances. They differ by one term, log det R = log det Q + log det Corr(Σ), so
the Q-barrier drops the shrinkage of the model-implied observed correlations,
which the likelihood already keeps positive definite, and it is zero for a
model without latents.

Three questions. Do the barriers remove improper solutions without costing the
accuracy of PSD-constrained ML, including when the truth sits exactly on a
face, where PSD-ML is at its best? Which fit test belongs with a barrier
estimate? And does the local limit derived for the barrier describe it at
these sample sizes?

## Short answer

Both barriers keep every estimate admissible.
`r if (n_bar_bad == 0) "None" else n_bar_bad` of the
`r format(n_bar, big.mark = ",")` barrier fits
`r if (n_bar_bad == 1) "was" else "were"` improper or on the boundary.
`r n_bar_err` (`r sprintf("%.2f", 100 * n_bar_err / n_bar)`%)
failed in the optimizer, `r err_imp` of them in the design whose population is
itself improper, where the barrier optimum hugs the face.

Against PSD-ML the barrier trades accuracy on the face for accuracy inside.
With the truth inside the admissible set, the Q-barrier at λ = 0.25 is more
accurate when ML is often improper but the truth lies some way inside (RMSE
`r num(key_rmse("f1_p3", 50, "det_l025"))` against
`r num(key_rmse("f1_p3", 50, "psd"))` for the Heywood-prone residual variance,
`r num(key_rmse("path_r2_90", 50, "det_l025"))` against
`r num(key_rmse("path_r2_90", 50, "psd"))` for the path R², N = 50), and less
accurate when the truth is within a standard error or so of the face
(`r num(key_rmse("f2_r99", 50, "det_l025"))` against
`r num(key_rmse("f2_r99", 50, "psd"))` at r = .99). With the truth exactly on a
face, PSD-ML is exactly right about half the time and the barrier never is. The
local limit puts the barrier's RMSE there at √(1 + 2λ) times PSD-ML's, 1.22 at
λ = 0.25, and at N = 400 the observed ratio is
`r fmt2(face_ratio("f2_r100"))` for r = 1 and
`r fmt2(face_ratio("path_r2_100"))` for R² = 1. The same limit, the barrier
estimate t(u) = (u + √(u² + 4λ)) / 2 standard errors from the face when ML sits
at u, matches the simulated estimates closely at λ = 0.25.

For testing, Browne's residual statistic is the model test to use with any of
these estimators. It gives the same rejection rate at ML, PSD-ML, and either
barrier, and at N = 50 it is better calibrated than the likelihood ratio test
(`r span(sapply(tdes, function(d) reject(d, 50, "det_l025", "browne")))`% at the
Q-barrier estimate, against
`r span(sapply(tdes, function(d) reject(d, 50, "ml")))`% for the LR test at ML).
The likelihood ratio at an admissible estimate adds a second part, the score for
the estimator's displacement from the ML optimum, which tests admissibility.
Referred to its reference law at a face, it rejects
`r span(c(sapply(face, function(d) adm_reject(d, 50, "det_l025")), sapply(face, function(d) adm_reject(d, 400, "det_l025"))))`%
with the truth on a face, less inside, and
`r pct(adm_reject("imp_f1_p5", 50, "det_l025"))`% at N = 50 when the population
has a residual variance of −0.061, which no test of the unconstrained model can
detect (Browne rejects `r pct(reject("imp_f1_p5", 50, "det_l025", "browne"))`%).
Against a factor correlation of 1.03 it has little power
(`r pct(adm_reject("imp_f2_r103", 400, "det_l025"))`% at N = 400).

The Q-barrier is at least as accurate as the joint barrier in every design at
every weight, and at λ = 1 its worst coverage is higher by up to
`r sprintf("%.1f", 100 * max(sapply(proper_d, function(d) min_cover(d, 50, "det_l100") - min_cover(d, 50, "pen_l100"))))`
points.

## Evidence

### Every converged barrier fit is admissible

Share of samples at N = 50 (%). "Joint" and "Q" pool five weights each
(λ = 0.1 to 2).

```{r admissibility}
all_d <- c(interior, face, imp)
tab <- data.frame(
  Design = design_label[all_d],
  `ML improper` = sapply(all_d, function(d) pct(improper(d, 50, "ml"))),
  `ML failed` = sapply(all_d, function(d) pct(failed(d, 50, "ml"))),
  `Bounded ML improper` = sapply(all_d, function(d) pct(improper(d, 50, "ml_bounded"))),
  `PSD-ML on the boundary` = sapply(all_d, function(d) pct(boundary(d, 50, "psd"))),
  `Joint not interior` = sapply(all_d, function(d) pct(nonint_rate(d, 50, pens))),
  `Q not interior` = sapply(all_d, function(d) pct(nonint_rate(d, 50, dets))),
  check.names = FALSE, row.names = NULL)
kable_r(tab)
```

### Accuracy with the truth inside

RMSE of each design's key quantity at N = 50. The key quantity of the
one-factor designs is the residual variance of x1 (truth .19), a single named
parameter.

```{r accuracy}
cols <- c(ml = "ML", psd = "PSD-ML", det_l010 = "Q, λ = 0.1",
          det_l025 = "Q, λ = 0.25", det_l050 = "Q, λ = 0.5",
          det_l100 = "Q, λ = 1", det_l200 = "Q, λ = 2")
tab <- tbl(interior, cols, function(d, m) num(key_rmse(d, 50, m)))
tab <- cbind(tab[1], `Key quantity` = unname(key_of[interior]), tab[-1])
kable_r(tab)
```

### Accuracy with the truth on the face

When the truth is exactly on a face, PSD-ML is exactly right whenever it lands
on the boundary, and a barrier estimate is never exactly right. RMSE and bias
of the key quantity, and the share of PSD-ML fits on the boundary, at N = 50.

```{r face}
cols <- c(psd = "RMSE, PSD-ML", det_l025 = "RMSE, Q λ = 0.25",
          det_l100 = "RMSE, Q λ = 1", psd_b = "Bias, PSD-ML",
          det_l025_b = "Bias, Q λ = 0.25", det_l100_b = "Bias, Q λ = 1",
          psd_on = "PSD-ML on the boundary (%)")
tab <- tbl(face, cols, function(d, k) {
  if (k == "psd_on") return(pct(boundary(d, 50, "psd")))
  if (grepl("_b$", k)) return(num(key_bias(d, 50, sub("_b$", "", k))))
  num(key_rmse(d, 50, k))
})
tab <- cbind(tab[1], `Key quantity` = unname(key_of[face]), tab[-1])
kable_r(tab)
```

PSD-ML wins in three of the four designs, and its margin does not shrink with
N. In the local limit, PSD-ML's RMSE on the face is √½ standard error and a
barrier's is √(½ + λ). The exception is the three-indicator design, where the
standard error of the residual variance is about
`r fmt2(truth$se_unit[truth$design == "f1_p3_face" & truth$param == "x1 ~~ x1"] / sqrt(50))`
at N = 50, so the likelihood is far from quadratic over the admissible range
and the local limit does not apply yet.

### Where the gains and costs come from

RMSE of the key quantity and the worst coverage of a nominal 95% Wald interval
over the free parameters, N ≤ 100, split by whether ordinary ML was improper on
the same sample. On proper-ML samples PSD-ML equals ML, so any difference there
is the barrier's cost.

```{r strata}
ml_state <- fits[fits$method == "ml", c("design", "n", "rep", "classification")]
ml_state$state <- ifelse(!is.na(ml_state$classification) &
                           ml_state$classification == "improper", "ML improper", "ML proper")
g <- merge(fits[fits$n <= 100, ], ml_state[, c("design", "n", "rep", "state")],
           by = c("design", "n", "rep"))
strata_d <- c("f1_p3", "f2_r90", "f2_r97", "path_r2_90", "f2_r100", "path_r2_100")
rows <- expand.grid(state = c("ML improper", "ML proper"), design = strata_d,
                    stringsAsFactors = FALSE)
cnt <- mapply(function(d, st) sum(g$design == d & g$state == st & g$method == "ml"),
              rows$design, rows$state)
rows <- rows[cnt >= 20, ]
ms <- c(psd = "PSD-ML", det_l025 = "Q, λ = 0.25")
tab <- data.frame(Design = design_label[rows$design], Samples = rows$state,
                  check.names = FALSE, row.names = NULL)
tab$Share <- mapply(function(d, st) {
  pct(mean(g$state[g$design == d & g$method == "ml"] == st))
}, rows$design, rows$state)
for (m in names(ms)) {
  tab[[paste("RMSE,", ms[[m]])]] <- mapply(function(d, st) {
    s <- g[g$design == d & g$state == st & g$method == m, ]
    num(sqrt(mean(s$err^2, na.rm = TRUE)))
  }, rows$design, rows$state)
}
for (m in names(ms)) {
  tab[[paste("Worst coverage,", ms[[m]])]] <- mapply(function(d, st) {
    reps <- with(ml_state[ml_state$design == d & ml_state$n <= 100 &
                            ml_state$state == st, ], paste(n, rep))
    pct(min_cover(d, c(50, 100), m, reps))
  }, rows$design, rows$state)
}
knitr::kable(tab, align = paste0("ll", strrep("r", ncol(tab) - 2L)))
```

On ML-improper samples the barrier beats PSD-ML's boundary estimate when the
truth is inside and loses to it when the truth is on the face, where the
boundary is the truth. On ML-proper samples it pays a small shrinkage cost near
a face. Worst coverage at λ = 0.25 stays within two points of PSD-ML in every
stratum.

### The local limit describes the barrier

Near a single face the barrier estimate is a fixed function of the ML estimate.
In standard errors from the face, ML at u gives the barrier estimate
t(u) = (u + √(u² + 4λ)) / 2 and PSD-ML gives max(u, 0). The figure pools the
residual variance of x1 over the five one-factor designs and all sample sizes,
standardized by its population standard error, and shows the median estimate
in bins of a quarter standard error with at least 40 samples.

```{r local-limit, fig.width = 9, fig.height = 3.6}
x1 <- params[params$param == "x1 ~~ x1" &
               params$design %in% c("f1_p3", "f1_p5", "f1_p3_face", "f1_p5_face", "imp_f1_p5"),
             c("design", "n", "rep", "method", "est")]
tr <- truth[truth$param == "x1 ~~ x1", ]
x1$s <- setNames(tr$se_unit, tr$design)[x1$design] / sqrt(x1$n)
w <- reshape(x1, idvar = c("design", "n", "rep", "s"), timevar = "method",
             direction = "wide")
w$u <- w$est.ml / w$s
lm_methods <- c(psd = "PSD-ML", pen_l025 = "Joint", det_l025 = "Q",
                pen_l100 = "Joint", det_l100 = "Q")
lm_panel <- c(psd = "PSD-ML", pen_l025 = "λ = 0.25", det_l025 = "λ = 0.25",
              pen_l100 = "λ = 1", det_l100 = "λ = 1")
long <- do.call(rbind, lapply(names(lm_methods), function(m) {
  data.frame(method = m, barrier = lm_methods[[m]], panel = lm_panel[[m]],
             u = w$u, a = w[[paste0("est.", m)]] / w$s)
}))
long <- long[is.finite(long$u) & is.finite(long$a) & abs(long$u) <= 4, ]
long$bin <- round(long$u * 4) / 4
agg <- aggregate(a ~ method + barrier + panel + bin, long, median)
cnt <- aggregate(a ~ method + bin, long, length)
names(cnt)[3] <- "count"
agg <- merge(agg, cnt, by = c("method", "bin"))
agg <- agg[agg$count >= 40, ]
ugrid <- seq(-4, 4, by = 0.02)
theory <- rbind(
  data.frame(panel = "PSD-ML", u = ugrid, a = pmax(ugrid, 0)),
  data.frame(panel = "λ = 0.25", u = ugrid, a = t_lambda(ugrid, 0.25)),
  data.frame(panel = "λ = 1", u = ugrid, a = t_lambda(ugrid, 1)))
panels <- c("PSD-ML", "λ = 0.25", "λ = 1")
agg$panel <- factor(agg$panel, panels)
theory$panel <- factor(theory$panel, panels)
cols <- c(Q = "#2a78d6", Joint = "#eb6834", `PSD-ML` = "#1baf7a")
shapes <- c(Q = 16, Joint = 17, `PSD-ML` = 15)
library(ggplot2)
ggplot() +
  geom_line(data = theory, aes(u, a), colour = "#52514e", linetype = "22",
            linewidth = 0.5) +
  geom_point(data = agg, aes(bin, a, colour = barrier, shape = barrier), size = 2.2) +
  facet_wrap(~panel, nrow = 1) +
  scale_colour_manual(values = cols, name = NULL) +
  scale_shape_manual(values = shapes, name = NULL) +
  labs(x = "ML estimate (standard errors from the face)",
       y = "Estimate (SE from the face)") +
  coord_cartesian(xlim = c(-4, 4), ylim = c(-0.2, 4)) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

```{r local-gap}
gap <- function(m, lo, hi) {
  s <- agg[agg$method == m & agg$bin >= lo & agg$bin <= hi, ]
  max(abs(s$a - t_lambda(s$bin, lambda_of[[m]])))
}
```

Dashed lines: the local limit. At λ = 0.25 the medians stay within
`r fmt2(gap("det_l025", -4, 2))` standard errors of the curve for u up to 2. At
λ = 1 they stay within `r fmt2(gap("det_l100", -4, 0))` at and beyond the face
(u ≤ 0) but fall below the curve inside it. There the penalty's pull on one
residual variance is weaker than a pure log barrier, because Q responds to the
information of all indicators together, so the limit overstates the shrinkage
of estimates well inside the admissible set.

### Which fit test goes with a barrier estimate

Three statistics were computed at every estimate: the likelihood ratio
statistic n F(θ̂), the reweighted least squares quadratic
(n/2) tr{[Σ̂⁻¹(S − Σ̂)]²}, and Browne's residual statistic, which is the same
quadratic with the model tangent space projected out (lavaan's
`browne.residual.nt.model`). At the ML optimum the last two are equal. At any
other estimate the projection removes the estimate's own displacement, so
Browne's statistic tests the unconstrained model whichever estimator produced
θ̂. The difference, RLS minus Browne, is the score statistic for that
displacement, the part of the fit statistic that measures how far the
estimator had to move to stay admissible.

Rejection of the model at 5% against χ²(df), N = 50 (%). Every model is
correct and every truth is proper. The three-indicator designs have df = 0.

```{r model-test}
tdes <- setdiff(proper_d, c("f1_p3", "f1_p3_face"))
cols <- c(ml_lr = "LR at ML", ml_rls = "RLS at ML",
          psd_br = "Browne at PSD-ML", det_l025_br = "Browne at Q λ = 0.25",
          det_l100_br = "Browne at Q λ = 1", psd_lr = "LR at PSD-ML",
          det_l025_lr = "LR at Q λ = 0.25", det_l100_lr = "LR at Q λ = 1")
stat_of <- c(lr = "chi2", rls = "rls", br = "browne")
tab <- tbl(tdes, cols, function(d, k) {
  m <- sub("_(lr|rls|br)$", "", k)
  pct(reject(d, 50, m, stat_of[[sub("^.*_", "", k)]]))
})
kable_r(tab)
```

Browne's statistic gives the same answer whichever estimator produced θ̂, so
the model test depends neither on λ nor on whether ML was improper. It is also
the better-calibrated statistic at N = 50. The LR test at ML over-rejects in
the path designs, and RLS at ML, which equals Browne's statistic there, does
not. The LR evaluated at PSD-ML or a barrier estimate over-rejects further near
a face, and that excess is the admissibility part below.

The admissibility part has its own reference law. With the truth on a single
face it is U² 1{U < 0} for PSD-ML and (t(U) − U)² for a barrier, U standard
normal, which gives 5% critical values of
`r sprintf("%.2f", adm_crit("psd"))` for PSD-ML,
`r sprintf("%.2f", adm_crit("det_l025"))` for λ = 0.25, and
`r sprintf("%.2f", adm_crit("det_l100"))` for λ = 1. The face is the least
favorable proper truth, so the test is conservative inside. Rejection at those
critical values (%):

```{r adm-test}
cols <- c(psd_50 = "PSD-ML, N = 50", det_l025_50 = "Q λ = 0.25, N = 50",
          det_l100_50 = "Q λ = 1, N = 50", psd_400 = "PSD-ML, N = 400",
          det_l025_400 = "Q λ = 0.25, N = 400", det_l100_400 = "Q λ = 1, N = 400")
tab <- tbl(proper_d, cols, function(d, k) {
  n <- as.integer(sub("^.*_", "", k))
  pct(adm_reject(d, n, sub("_[0-9]+$", "", k)))
})
kable_r(tab)
```

The test holds its level on the face and is conservative inside. At λ = 1 it
is also somewhat conservative on the face.

### Improper truths

In the last three designs no proper parameter value reproduces the population.
In two the population equals Σ(θ) for an improper θ (a residual variance of
−0.061, a factor correlation of 1.03), so the unconstrained model fits exactly
and only a test of the proper model can reject. In the third an omitted
residual covariance makes the pseudo-true factor correlation
`r num(key_truth[["mis_f2_resid"]])` and the model also misfits. Rejection at
5% (%):

```{r power}
rows <- expand.grid(n = c(50, 400), design = imp, stringsAsFactors = FALSE)
tab <- data.frame(Design = design_label[rows$design], N = rows$n,
                  check.names = FALSE, row.names = NULL)
tab$`LR at ML` <- mapply(function(d, n) pct(reject(d, n, "ml")), rows$design, rows$n)
tab$`Browne at Q λ = 0.25` <- mapply(function(d, n) pct(reject(d, n, "det_l025", "browne")),
                                     rows$design, rows$n)
tab$`Admissibility, PSD-ML` <- mapply(function(d, n) pct(adm_reject(d, n, "psd")),
                                      rows$design, rows$n)
tab$`Admissibility, Q λ = 0.25` <- mapply(function(d, n) pct(adm_reject(d, n, "det_l025")),
                                          rows$design, rows$n)
tab$`LR at PSD-ML` <- mapply(function(d, n) pct(reject(d, n, "psd")), rows$design, rows$n)
tab$`LR at Q λ = 0.25` <- mapply(function(d, n) pct(reject(d, n, "det_l025")),
                                 rows$design, rows$n)
knitr::kable(tab, align = paste0("lr", strrep("r", ncol(tab) - 2L)))
```

The unconstrained tests stay at their nominal level in the two exact-fit
designs, because the model with an improper parameter fits. Only the
admissibility test sees these populations. It detects a residual variance of
−0.061 almost surely from N = 100
(`r pct(adm_reject("imp_f1_p5", 100, "det_l025"))`%), but a factor correlation
of 1.03 is only about a standard error beyond the face even at N = 400. Under
misfit the model test carries the power and the admissibility test adds
little. PSD-ML and the barrier converge to the face as N grows, while ML
converges to the improper value. In the exact r = 1.03 design the mean
Q-barrier estimate at λ = 0.25 is `r num(key_mean("imp_f2_r103", 50, "det_l025"))`
at N = 50 and `r num(key_mean("imp_f2_r103", 400, "det_l025"))` at N = 400, PSD-ML's
`r num(key_mean("imp_f2_r103", 50, "psd"))` and
`r num(key_mean("imp_f2_r103", 400, "psd"))`, and ML's
`r num(key_mean("imp_f2_r103", 400, "ml"))` at N = 400.

### The joint barrier and the Q-barrier

RMSE of the key quantity at N = 50, joint barrier against Q-barrier at the same
weight.

```{r q-accuracy}
lams <- c(`0.25` = "l025", `1` = "l100", `2` = "l200")
jd <- c(interior, face)
tab <- data.frame(Design = design_label[jd], check.names = FALSE, row.names = NULL)
for (l in names(lams)) {
  tab[[paste0("λ = ", l, ", joint")]] <- sapply(jd, function(d) num(key_rmse(d, 50, paste0("pen_", lams[[l]]))))
  tab[[paste0("λ = ", l, ", Q")]] <- sapply(jd, function(d) num(key_rmse(d, 50, paste0("det_", lams[[l]]))))
}
tab$`Worst coverage, λ = 1, joint` <- sapply(jd, function(d) pct(min_cover(d, 50, "pen_l100")))
tab$`Worst coverage, λ = 1, Q` <- sapply(jd, function(d) pct(min_cover(d, 50, "det_l100")))
kable_r(tab)
```

```{r joint-check}
worse <- sum(sapply(proper_d, function(d) sapply(c("l025", "l100", "l200"), function(l) {
  key_rmse(d, 50, paste0("det_", l)) > key_rmse(d, 50, paste0("pen_", l)) + 5e-4
})))
```

In `r if (worse == 0) "none" else worse` of these thirty cells is the
Q-barrier less accurate than the joint barrier by more than 0.0005, and its
advantage grows with λ. The joint
barrier's extra term shrinks the observed correlations, which costs accuracy
and coverage without protecting any face.

## Caveats

- Data are multivariate normal. The three test statistics are normal-theory
  statistics.
- The barrier fits that failed did so in the optimizer's line search, mostly
  at λ = 0.1 and N = 400 in the design whose population is improper
  (`r pct(mean(cell("imp_f1_p5", 400, "det_l010")$status != "ok"))`% of those
  fits). They are excluded from the accuracy summaries.
- The admissibility reference law is derived for a single face. Several
  nearby faces at once (for example two residual variances near zero) need
  the joint law, which this run does not test.
- Standard errors are unpenalized expected-information errors at each
  method's estimate. Wald intervals at a face are a poor reference for PSD-ML
  and a barrier alike, and no profile or penalized-likelihood interval was
  tried.
- λ = 0.25 is the current default of the joint barrier. The Q-barrier has no
  default yet.
- Coverage and the local limit are computed for λ = 0.25 and 1 only.

## Reproduce

```sh
just r-dev
Rscript experiments/research/active/47-multiinfo-penalty-improper/run_experiment.R --full --cores 4
quarto render experiments/research/active/47-multiinfo-penalty-improper/report.qmd
```
