---
title: "Maydeu-Olivares (2017) replication"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "metadata.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
read_csv_safely <- function(name) {
  p <- file.path(results_dir, name)
  if (!file.exists(p)) return(NULL)
  utils::read.csv(p, check.names = FALSE, stringsAsFactors = FALSE)
}
meta       <- read_csv_safely("metadata.csv")
cells      <- read_csv_safely("cells.csv")
thresholds <- read_csv_safely("thresholds.csv")
se_long    <- read_csv_safely("fits_se.csv")
chi2_long  <- read_csv_safely("fits_chi2.csv")
sum_se     <- read_csv_safely("summary_se.csv")
sum_chi2   <- read_csv_safely("summary_chi2.csv")
parity     <- read_csv_safely("lavaan_parity.csv")
if (is.null(meta) || is.null(sum_se) || is.null(sum_chi2)) {
  stop("Missing result CSVs. Run: Rscript run_experiment.R")
}
fmt_pct   <- function(x) ifelse(is.na(x), "", sprintf("%.2f", x))
fmt_pct3  <- function(x) ifelse(is.na(x), "", sprintf("%.3f", x))
fmt_num   <- function(x) ifelse(is.na(x), "", sprintf("%.4g", x))
dist_pretty <- function(d) {
  c(norm = "skew=0, kurt=0", kurt2 = "skew=0, kurt=2",
    skew_kurt = "skew=-2, kurt=3.18")[d]
}
```

This experiment reproduces the simulation in
**Maydeu-Olivares (2017), "Maximum Likelihood Estimation of Structural
Equation Models for Continuous Data: Standard Errors and Goodness of Fit"**
against magmaan's own ML inference. The paper compares seven standard-error
methods and five χ² adjustments on a two-factor CFA fit unidimensionally
under three nonnormality regimes, three sample sizes, two model sizes, and
three degrees of misspecification.

## Run

```{r}
knitr::kable(meta, row.names = FALSE)
```

## Threshold sets

Item-level nonnormality is induced by 5-category discretization of MVN data
at thresholds chosen so the marginal moments hit the (skew, excess-kurt)
targets.

```{r}
knitr::kable(thresholds, row.names = FALSE,
             col.names = c("dist", "skew target", "kurt target",
                           "skew achieved", "kurt achieved",
                           "τ₁", "τ₂", "τ₃", "τ₄", "loss"),
             digits = 4)
```

## Cell grid

```{r}
knitr::kable(cells, row.names = FALSE)
```

## Standard-error relative bias

Mean relative bias of factor-loading SEs across replicates, by method.
Negative values indicate the SEs are too small (variability of `θ̂`
underestimated). Maydeu-Olivares' Table 2 cutoff for "trivial bias" is
|·| < 0.10.

```{r}
if (!is.null(sum_se)) {
  sum_se$dist <- dist_pretty(sum_se$dist)
  sum_se$mean_rel_bias_se <- fmt_pct(sum_se$mean_rel_bias_se)
  wide <- stats::reshape(sum_se[, c("n_items", "rho", "dist", "N",
                                    "method", "mean_rel_bias_se")],
                          idvar = c("n_items", "rho", "dist", "N"),
                          timevar = "method", direction = "wide")
  names(wide) <- sub("^mean_rel_bias_se\\.", "", names(wide))
  wide <- wide[order(wide$n_items, wide$rho, wide$dist, wide$N), ]
  knitr::kable(wide, row.names = FALSE)
}
```

## χ² Type I error / power

Empirical rejection rate at the 5% nominal level. With ρ = 1.0 the
unidimensional model is correctly specified — these cells give the Type I
error. ρ = 0.8 and ρ = 0.4 give the power for "minor" and "large"
misspecification.

Bradley's accuracy band for Type I error is `[0.025, 0.075]` (paper
convention). Cells outside that band are highlighted in the prose below.

```{r}
if (!is.null(sum_chi2)) {
  sum_chi2$dist <- dist_pretty(sum_chi2$dist)
  sum_chi2$reject_05 <- fmt_pct3(sum_chi2$reject_05)
  wide <- stats::reshape(sum_chi2[, c("n_items", "rho", "dist", "N",
                                      "stat_name", "reject_05")],
                          idvar = c("n_items", "rho", "dist", "N"),
                          timevar = "stat_name", direction = "wide")
  names(wide) <- sub("^reject_05\\.", "", names(wide))
  wide <- wide[order(wide$n_items, wide$rho, wide$dist, wide$N), ]
  knitr::kable(wide, row.names = FALSE)
}
```

## RMSEA: sample vs population

Mean sample RMSEA across replicates, by χ² adjustment.

```{r}
if (!is.null(sum_chi2)) {
  sum_chi2$mean_rmsea <- fmt_pct3(sum_chi2$mean_rmsea)
  wide <- stats::reshape(sum_chi2[, c("n_items", "rho", "dist", "N",
                                      "stat_name", "mean_rmsea")],
                          idvar = c("n_items", "rho", "dist", "N"),
                          timevar = "stat_name", direction = "wide")
  names(wide) <- sub("^mean_rmsea\\.", "", names(wide))
  wide <- wide[order(wide$n_items, wide$rho, wide$dist, wide$N), ]
  knitr::kable(wide, row.names = FALSE)
}
```

## SRMR

Mean SRMR across replicates (SRMR is identical across χ² rows by
construction, so we show it once per cell).

```{r}
if (!is.null(sum_chi2)) {
  srmr_table <- unique(sum_chi2[, c("n_items", "rho", "dist", "N", "mean_srmr")])
  srmr_table$mean_srmr <- fmt_pct3(srmr_table$mean_srmr)
  srmr_table <- srmr_table[order(srmr_table$n_items, srmr_table$rho,
                                  srmr_table$dist, srmr_table$N), ]
  knitr::kable(srmr_table, row.names = FALSE)
}
```

## Lavaan parity check

Only present when the runner was invoked with `--lavaan-parity`. Compares
magmaan to lavaan on rep 1 of each cell. ML and MLM are compared on the
statistic; MLMV and MLR are compared via the upper-tail p-value because
magmaan uses Satterthwaite-df parameterization and lavaan uses the
scaled-shifted linear transform — different parameterizations of the same
asymptotic adjustment.

```{r}
if (!is.null(parity) && nrow(parity) > 0L) {
  parity$dist <- dist_pretty(parity$dist)
  parity$magmaan <- fmt_num(parity$magmaan)
  parity$lavaan  <- fmt_num(parity$lavaan)
  parity$abs_diff <- fmt_num(parity$abs_diff)
  knitr::kable(parity, row.names = FALSE)
} else {
  cat("(no lavaan parity table — run with `--lavaan-parity`)\n")
}
```

## Notes and caveats

- Default `--reps 10` is for fast iteration; the rejection rates and bias
  estimates are noisy at this size. The paper uses `--reps 1000`. Per
  internal style, no Monte-Carlo error bars are reported.
- Threshold sets are produced by numerical moment-matching with
  deterministic restarts — the table is reproducible across runs.
- Categories may be empty for extreme distributions at small `N` — in the
  `skew_kurt` set the upper two categories are rare, so a few rows may have
  fewer than 5 observed levels. ML treats the integer codes as continuous
  regardless.
- MLM, MLMV, and MLR-exp share the same SE (`bread=expected,
  meat=empirical` sandwich); they differ only in the χ² adjustment.
- MLR uses `bread=observed, meat=empirical`.
- MLF uses the new `inference_information_cross_products` primitive
  introduced for this experiment.
