---
title: "Normal-theory parameter intervals: Wald, score and profile LR"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
---

## Question

For interior SEM parameters under normal sampling, how do Wald, score and
profile likelihood-ratio intervals compare, and does a bootstrap Bartlett
correction improve profile-LR inversion enough to justify its cost?

## Short answer

**Score and LR improve the small-sample correlation interval in this pilot;
there is no clear advantage for the loading.** At N = 100, correlation coverage
is 88.5% for Wald, 93.0% for score and 92.0% for LR, against nominal 95%.
The paired gains over Wald are 4.5 percentage points (Monte Carlo SE 1.6) and
3.5 points (SE 1.5), respectively. At N = 300 and 1,000 their correlation
coverages are within half a percentage point of one another. All 4,800 ordinary
intervals completed, with no truth-test/inversion disagreements.

```{r setup}
#| include: false
paths <- c("results/frozen/main_summary.csv", "results/frozen/bartlett-pilot_summary.csv",
           "results/frozen/stability.csv", "results/frozen/bootstrap_noise.csv",
           "results/frozen/endpoint_corrections.csv")
if (!all(file.exists(paths))) stop("Complete the runs and run Rscript scripts/summarize.R before rendering.")
main <- read.csv(paths[1])
bs <- read.csv(paths[2])
stability <- read.csv(paths[3])
noise <- read.csv(paths[4])
ec <- read.csv(paths[5])
br <- bs[bs$method == "lr_bartlett", ]
stable_valid <- stability[stability$valid, ]
fmt <- function(x, digits=1) formatC(x, format="f", digits=digits)
safe_max <- function(x) if(any(is.finite(x))) max(x[is.finite(x)]) else NA_real_
```

The candidate-specific Bartlett pilot returned **`r br$valid` of 50 intervals**,
with `r fmt(100*br$coverage)`% coverage among successful intervals. Both ordinary
LR and corrected LR produced 47 successful, covering intervals out of 50
attempts. On the 49 datasets where both succeeded, Bartlett covered one
additional case; this is too little evidence to establish a benefit. In the
prespecified endpoint checks, `r sum(stable_valid$unstable)` of
`r nrow(stable_valid)` valid comparisons moved an endpoint by more than 5% of
the ordinary LR interval's width. This is a feasibility and noise check;
50 datasets cannot establish a small coverage improvement.

## Evidence

### Ordinary interval coverage

```{r}
methods <- c("wald_expected", "score_expected", "lr")
key <- unique(main[,c("n","target")]); key <- key[order(key$target,key$n),]
values <- sapply(methods,function(method) vapply(seq_len(nrow(key)),function(i)
  100*main$coverage[main$n==key$n[i] & main$target==key$target[i] & main$method==method],numeric(1)))
tab <- data.frame(Target=tools::toTitleCase(key$target),N=key$n,values,check.names=FALSE)
names(tab)[3:5] <- c("Wald (%)", "Score (%)", "LR (%)")
knitr::kable(tab,digits=1,row.names=FALSE,
  caption="Coverage of nominal 95% intervals; 200 paired datasets per sample size.")
```

The N = 100 correlation Wald interval misses below its lower endpoint in
8.0% of samples and above its upper endpoint in 3.5%. Score gives 3.0% and
4.0%; LR gives 4.0% and 4.0%. The improvement is mainly in the placement of the
interval: mean widths are 0.374 for Wald, 0.380 for score and 0.374 for LR.
Observed-information Wald has the same correlation coverage as the primary
Wald method in all three cells. Loading coverage across methods and sample
sizes ranges from 94.0% to 97.5%; the paired differences are small and do not
support a general method ranking.

### Bartlett inversion: correction noise and cost

```{r}
brief <- bs[bs$method %in% c("lr","lr_bartlett"),]
knitr::kable(data.frame(
  Method=ifelse(brief$method=="lr","Ordinary LR","Bootstrap Bartlett LR"),
  Attempted=brief$attempted,Successful=brief$valid,
  `Coverage (%)`=100*brief$coverage,
  `Successful and Covering (%)`=100*brief$successful_covering,
  `Mean Width`=brief$mean_width,
  `Mean Seconds`=brief$mean_seconds,check.names=FALSE),digits=c(0,0,0,1,1,3,2),
  caption="Paired N = 100 loading pilot; 399 bootstrap draws per candidate.")
```

Increasing the bootstrap size from 399 to 1,999 and changing the random stream
are checked on the first ten prespecified datasets. The larger run returned
nine intervals and failed once because a bootstrap fit did not converge; the
independent stream returned all ten. The independent-stream
comparison at 1,999 draws has `r sum(noise$valid)` valid pairs; its largest
endpoint displacement is `r fmt(100*safe_max(noise$endpoint_shift_fraction))`%
of the ordinary LR width. The median endpoint correction is
`r fmt(median(ec$correction[ec$run=="bartlett-pilot"]),3)` with median Monte Carlo
SE `r fmt(median(ec$mcse[ec$run=="bartlett-pilot"]),3)` at 399 draws, versus
`r fmt(median(ec$mcse[ec$run=="bartlett-large"]),3)` at 1,999 draws. These checks
separate finite-bootstrap noise from error in the underlying LR approximation.

A compact evidence snapshot in `results/frozen/` is deliberately tracked so
these findings and the rendered report remain reviewable without repeating the
nested bootstrap. It contains summaries, numerical checks, stability checks
run metadata and interval-failure summaries only. Raw intervals, candidate
diagnostics and individual bootstrap diagnostics remain local and ignored.

## Caveats

The population has a correct covariance model and interior parameters. This
does not settle inference for non-normal data, misspecified covariance models,
boundary parameters, missing data, or penalized estimates. Two targets in one
dataset are paired outcomes, not independent replications. At 200 replications,
the Monte Carlo SE of 95% coverage is about 1.5 percentage points.

The run used ordinary unpenalized likelihood fits with a positive-definite
implied covariance, without bounds on primitive variances. One of the 200
N = 100 unrestricted fits was covariance-inadmissible; none at larger N were.
Candidate searches also visit inadmissible primitive covariance parameters.
These events are retained and separately tabulated; this is not a test of
intervals constrained to the physically admissible parameter space. A finite
wider-grid check found no additional acceptance components, but is not a proof
of connectedness over the entire parameter domain. The failed Bartlett interval had one nonconvergent bootstrap dataset even
after its predetermined fallback. Timing of Wald covariance
setup was shared and not measured separately; no Wald speed ratio is claimed. Dataset workers run concurrently, so interval
elapsed times include scheduling contention and are not isolated microbenchmarks.

Bartlett correction is compatible with test inversion: invert the corrected
test rather than the uncorrected one. A simulated null mean is an estimate of
the correction, however; neither higher-order accuracy nor better coverage is
guaranteed for this SEM implementation. Bootstrap mean correction is an
established construction, for example in
[Loose, Bayer and Pereira, *Bootstrap Bartlett correction in inflated beta regression*](https://arxiv.org/abs/1506.03993).
That application is methodological background, not evidence about SEM coverage.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --plan
```

The runner is implemented and resumable. The completed commands are:

```sh
Rscript run_experiment.R --validate
Rscript run_experiment.R --smoke --workers 2 --output smoke-v2
Rscript run_experiment.R --main --workers 4
Rscript run_experiment.R --bartlett-pilot --workers 4
Rscript run_experiment.R --bartlett-pilot --reps 10 --bootstrap-reps 1999 --workers 2 --output bartlett-large
Rscript run_experiment.R --bartlett-pilot --reps 10 --bootstrap-reps 1999 --stream 1 --workers 2 --output bartlett-independent
Rscript scripts/summarize.R
quarto render report.qmd --to html
```

Use a new output name for a changed code version or statistical configuration.
An identical command resumes completed datasets. Source fingerprints, package
versions, seeds, numerical checks and run metadata accompany the results.
All 14 numerical checks passed: agreement with lavaan for the LR, an independent
score derivative check, information scaling, endpoint equations, truth-test
agreement and reproducibility of the bootstrap correction.

## Prespecified design

Use eight centered continuous indicators, four per factor. Both factor
variances are fixed to one, their correlation is 0.4, and the loadings on each
factor are (0.8, 0.7, 0.9, 0.6). Residuals are independent with variances
one minus the squared loading. Factor and residual innovations are independent
standard normals. Thus each indicator has population variance one. Fit the
same two-factor model with all eight loadings, eight residual variances and
the factor correlation free, retaining unit factor variances. Orient each
factor positively. Center each sample and use the ML covariance divisor N;
do not mix N and N-1 likelihood conventions.

The targets are the second loading (0.7) and factor correlation (0.4), tested
separately while all other free parameters are nuisance parameters. Estimate
the same unpenalized complete-data ML model for every interval method. Record
the exact optimizer, bounds and covariance-domain policy before the smoke run;
hold these fixed across methods and bootstrap fits. Boundary-active fits are
reported separately rather than repaired, clipped or silently removed.

Compare 95% model-based Wald intervals using expected information, efficient
score inversion using expected information at the restricted fit, and ordinary
profile-LR inversion. Observed-information model-based Wald is a secondary
diagnostic. Use N = 100, 300 and 1,000, with 200 datasets per cell. Compute all
methods on every dataset. The main run uses seed base 2026092750; smoke runs
use 2026102750. Select replicate IDs before examining fits or results.

For every method record coverage, each tail's noncoverage, median and mean
width, endpoint asymmetry, failure and boundary rates, constrained-fit counts,
and elapsed time. Report coverage among successful intervals alongside the
successful-and-covering fraction of all attempts, and show paired comparisons
on common-valid datasets separately. Retain failures by method and reason.

## Bartlett pilot

Write the profile statistic as

$$
T(b)=2\{\ell(\widehat\theta)-\ell(\widetilde\theta_b)\}.
$$

For **each candidate** b, simulate B datasets from the normal model fitted
under b, refit both unrestricted and b-restricted models in every bootstrap
dataset, and estimate the rank-one null mean

$$
\widehat c(b)=B^{-1}\sum_{j=1}^{B}T_j^*(b).
$$

The corrected confidence set accepts b when
$T(b)/\widehat c(b)\leq\chi^2_{1,.95}$. This is a null-fitted parametric
bootstrap mean correction, not a bootstrap quantile test. Re-estimate nuisance
parameters inside every bootstrap refit. Do not reuse a factor calculated at
the true parameter or at the unrestricted estimate as the primary correction.

Start with the first 50 prespecified N = 100 datasets, for the loading only,
using B = 399. Compute the ordinary methods on the same subset. Repeat the
first ten datasets with B = 1,999, retaining the original bootstrap draws as a
prefix, and with an independent bootstrap seed stream. Log the Monte Carlo SE
of the estimated correction. This is a feasibility pilot: at 50 datasets the
Monte Carlo SE of 95% coverage is about 3.1 percentage points.

Keep base normal draws fixed across candidate values within a dataset and
target, transforming them by each restricted covariance. This makes the
objective reproducible and reduces artificial roughness. Check candidate
endpoints by direct evaluation, not solely interpolation. If correction noise
moves an endpoint by more than 5% of the uncorrected interval width, mark that
dataset unstable; report its frequency without discarding it. Do not claim
that 5% is a statistical optimality threshold. A larger coverage run is a
later design amendment, conditional on measured computational cost and endpoint
stability, not on favorable pilot coverage.

An optional independently simulated population-mean factor is a diagnostic
benchmark only. It must use separate draws and be labeled as unavailable to
users. No truth-based factor is allowed in the feasible method.
