---
title: "Ordinal observed-score omega: target and uncertainty"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
required <- c("results/target/target_summary.csv", "results/target/metadata.csv",
              "results/sampling/simulation_summary.csv", "results/sampling/metadata.csv")
if (!all(file.exists(required))) {
  stop("Missing results. From this directory run: Rscript run_experiment.R --smoke")
}
target <- read.csv(required[[1]])
target_meta <- read.csv(required[[2]])
sampling <- read.csv(required[[3]])
sampling_meta <- read.csv(required[[4]])
worst <- target[which.max(target$abs_difference), ]
```

## Question

What does ordinal observed-score covariance omega estimate, and do its delta
intervals cover that target in correctly specified ordinal data?

## Short answer

Covariance omega and direct one-factor true-score reliability are different
targets. In the retained `r target_meta$n_cells`-cell population audit, their
largest absolute difference is `r sprintf("%.4f", worst$abs_difference)`:
covariance omega `r sprintf("%.4f", worst$current_covariance_omega)` against
direct reliability `r sprintf("%.4f", worst$direct_true_score_omega)`.

The separate sampling check has `r sampling_meta$reps` replicates per threshold
regime at N = `r sampling_meta$n`. Its covariance-target coverage is
`r paste(sprintf("%s: %.1f%%", sampling$regime, 100 * sampling$coverage_95), collapse = "; ")`.
That check does not establish coverage for direct true-score reliability.

## Evidence

### Population targets

```{r}
# Keep the largest discrepancies; the full grid remains in results/target/.
x <- head(target[order(-target$abs_difference), ], 4)
x <- x[c("loading_pattern", "threshold_pattern", "current_covariance_omega",
         "direct_true_score_omega", "difference")]
names(x) <- c("Loadings", "Thresholds", "Covariance Omega", "Direct Reliability", "Difference")
x[] <- lapply(x, function(v) if (is.numeric(v)) round(v, 4) else v)
knitr::kable(x, row.names = FALSE)
```

The gap comes from item true-score variances: the covariance coefficient
reconstructs them from between-item covariances, while direct reliability
integrates conditional expected category scores.

### Sampling behavior for covariance omega

```{r}
x <- sampling[c("regime", "reps", "failures", "bias", "mean_se", "sd_value", "coverage_95")]
names(x) <- c("Thresholds", "Replicates", "Failures", "Bias", "Mean SE", "Sampling SD", "95% Coverage")
x[] <- lapply(x, function(v) if (is.numeric(v)) round(v, 4) else v)
knitr::kable(x, row.names = FALSE)
```

## Caveats

The population audit assumes one-factor probit responses, local independence
and integer category scores. The sampling target is a large Monte Carlo
category-score covariance. Both studies use correct-model data; neither covers
misspecification, nonnormal latent responses, multi-group pooling or direct-target
intervals. Small runs establish execution, not calibrated coverage.

The former engineering/07 sampling study and engineering/08 target audit were
merged on 30 September 2026. Their historical CSVs and metadata are unchanged
under `results/sampling/` and `results/target/`; they are not pooled.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --lane target
Rscript run_experiment.R --lane sampling --reps 200 --n 500 --n-pop 500000
quarto render report.qmd
```

## Activity

Banked; no run is queued. Reopen for a named consumer of direct true-score
reliability or a specified coverage study. Keep the direct and covariance targets
explicit before extending inference. The polychoric-omega studies address a
separate latent-response target.
