---
title: "When does closed-form Guttman CFA work?"
subtitle: "A simulation sketch of accuracy, inference, admissibility, and restrictions"
format:
  html:
    toc: true
    number-sections: false
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: results/paper-full
---

```{r setup}
results_dir <- params$results_dir
need <- c("metadata", "design", "estimate_summary", "whole_joint_summary",
          "diagnostic_summary", "progress")
paths <- setNames(file.path(results_dir, paste0(need, ".csv")), need)
if (!all(file.exists(paths))) {
  stop("Missing results. Run the paper grid shown under Reproduce, then render again.")
}

metadata <- read.csv(paths[["metadata"]], stringsAsFactors = FALSE)
design <- read.csv(paths[["design"]], stringsAsFactors = FALSE)
est <- read.csv(paths[["estimate_summary"]], stringsAsFactors = FALSE)
whole <- read.csv(paths[["whole_joint_summary"]], stringsAsFactors = FALSE)
diag <- read.csv(paths[["diagnostic_summary"]], stringsAsFactors = FALSE)
progress <- read.csv(paths[["progress"]], stringsAsFactors = FALSE)

meta <- function(key) {
  x <- metadata$value[metadata$key == key]
  if (length(x)) x[[1]] else ""
}
fmt <- function(x, digits = 2) {
  ifelse(is.finite(x), formatC(x, format = "f", digits = digits), "NA")
}
pct <- function(x, digits = 1) {
  ifelse(is.finite(x), paste0(formatC(100 * x, format = "f", digits = digits), "%"), "NA")
}
med <- function(x) median(x[is.finite(x)], na.rm = TRUE)
collapse_values <- function(x) paste(sort(unique(x)), collapse = ", ")

# The completed 2026-07-11 configural run predates the cleaner population-model
# column. Normalize it here so the report remains an honest reader of that run.
normalize_population <- function(x) {
  if (!"population_model" %in% names(x)) {
    x$population_model <- if ("constraint" %in% names(x) &
                                any(x$constraint == "tau_resid")) {
      ifelse(x$constraint == "tau_resid", "tau_resid", "congeneric")
    } else {
      "congeneric"
    }
  }
  x
}
design <- normalize_population(design)
est <- normalize_population(est)
whole <- normalize_population(whole)
diag <- normalize_population(diag)
progress <- normalize_population(progress)

label_map <- c(
  guttman_std = "Guttman",
  guttman_std_raw = "Guttman",
  guttman_legacy = "Guttman, legacy",
  ml_emp = "ML",
  uls_emp = "ULS",
  guttman_std_raw_restricted = "Restricted Guttman",
  ml_emp_restricted = "Restricted ML",
  uls_emp_restricted = "Restricted ULS")
label <- function(x) {
  out <- unname(label_map[x])
  out[is.na(out)] <- x[is.na(out)]
  out
}
first_available <- function(ids) {
  found <- ids[ids %in% unique(whole$estimator)]
  if (length(found)) found[[1]] else NA_character_
}
raw_std <- first_available(c("guttman_std_raw", "guttman_std"))
if (is.na(raw_std)) {
  stop("Results must contain the standardized Guttman arm (guttman_std_raw).")
}

restricted_present <- any(design$population_model == "tau_resid") &&
  any(grepl("_restricted$", whole$estimator))
config <- whole[whole$population_model == "congeneric", , drop = FALSE]
config_diag <- diag[diag$population_model == "congeneric", , drop = FALSE]
config_est <- est[est$population_model == "congeneric", , drop = FALSE]

rho_value <- function(estimator, rho) {
  x <- config$common_rmse_ratio[config$estimator == estimator & config$rho == rho]
  med(x)
}
raw_moderate_rho <- setNames(vapply(c(0, .35, .6, .8), function(r) {
  med(config$common_rmse_ratio[config$estimator == raw_std &
                                 config$rho == r & config$strength == "moderate"])
}, numeric(1)), c("0", ".35", ".6", ".8"))

design_table <- data.frame(
  Dimension = c("Population model", "Observed variables", "Factors",
                "Indicators per factor", "Standardized loadings",
                "Factor correlation", "Sample size", "Data law",
                "Observed-variable scales", "Replications"),
  `Levels in this result set` = c(
    if (restricted_present) "Congeneric; tau-equivalent with equal residual variances" else
      "Congeneric only; restricted regime not yet run",
    {
      p <- vapply(seq_len(nrow(design)), function(i) {
        q <- design$factors[[i]]
        m <- design$indicators[[i]]
        if (m == "u") sum(rep(c(3L, 6L), length.out = q)) else q * as.integer(m)
      }, integer(1))
      paste0(min(p), " to ", max(p))
    },
    collapse_values(design$factors),
    "3, 5, or alternating 3/6 (unbalanced)",
    "Moderate or weak; heterogeneous within factor in the congeneric regime",
    collapse_values(design$rho),
    collapse_values(design$n),
    "Normal; mildly non-normal independent generator; five-category ordinal scores",
    collapse_values(design$scale),
    meta("reps")),
  check.names = FALSE)

status_table <- data.frame(
  Population = c("Congeneric", "Tau-equivalent + equal residual variances"),
  Fits = c("Configural Guttman, legacy Guttman, ML, ULS",
           "The same configural fits plus correctly restricted Guttman, ML, and ULS"),
  Status = c(
    if (any(design$population_model == "congeneric")) paste0(
      sum(design$population_model == "congeneric"), " cells complete") else "Not run",
    if (restricted_present) paste0(sum(design$population_model == "tau_resid"),
                                  " cells complete") else "Not in this result set"),
  check.names = FALSE)

rho_estimators <- c(raw_std, "guttman_legacy", "ml_emp", "uls_emp")
rho_estimators <- rho_estimators[rho_estimators %in% unique(config$estimator)]
rho_long <- aggregate(common_rmse_ratio ~ rho + estimator,
                      config[config$estimator %in% rho_estimators, ], med)
rho_long$Estimator <- factor(label(rho_long$estimator), levels = label(rho_estimators))
rho_wide <- reshape(rho_long[c("rho", "Estimator", "common_rmse_ratio")],
                    idvar = "rho", timevar = "Estimator", direction = "wide")
names(rho_wide) <- sub("common_rmse_ratio\\.", "", names(rho_wide))
names(rho_wide)[names(rho_wide) == "rho"] <- "Factor Correlation"
for (j in setdiff(names(rho_wide), "Factor Correlation")) rho_wide[[j]] <- fmt(rho_wide[[j]])

strength_long <- aggregate(common_rmse_ratio ~ rho + strength + estimator,
                           config[config$estimator %in% c(raw_std, "guttman_legacy"), ], med)
strength_long$Estimator <- label(strength_long$estimator)
strength_wide <- reshape(strength_long[c("rho", "strength", "Estimator", "common_rmse_ratio")],
                         idvar = c("rho", "strength"), timevar = "Estimator",
                         direction = "wide")
names(strength_wide) <- sub("common_rmse_ratio\\.", "", names(strength_wide))
names(strength_wide)[1:2] <- c("Factor Correlation", "Loading Strength")
strength_wide$`Loading Strength` <- tools::toTitleCase(strength_wide$`Loading Strength`)
for (j in 3:ncol(strength_wide)) strength_wide[[j]] <- fmt(strength_wide[[j]])

diag_estimators <- rho_estimators
admissibility <- aggregate(cbind(fit_ok_rate, improper_rate) ~ estimator,
                           config_diag[config_diag$estimator %in% diag_estimators, ],
                           mean, na.rm = TRUE)
admissibility$Estimator <- factor(label(admissibility$estimator),
                                  levels = label(diag_estimators))
admissibility <- admissibility[order(admissibility$Estimator), ]
admissibility <- data.frame(
  Estimator = admissibility$Estimator,
  `Fit Returned` = pct(admissibility$fit_ok_rate),
  `Improper Among Returned Fits` = pct(admissibility$improper_rate),
  check.names = FALSE)

coverage_estimators <- c(raw_std, "guttman_legacy", "ml_emp", "uls_emp")
coverage_estimators <- coverage_estimators[coverage_estimators %in% unique(config_est$estimator)]
coverage_long <- aggregate(coverage ~ estimator + group,
                           config_est[config_est$estimator %in% coverage_estimators, ], med)
coverage_long$Estimator <- factor(label(coverage_long$estimator),
                                  levels = label(coverage_estimators))
coverage_wide <- reshape(coverage_long[c("Estimator", "group", "coverage")],
                         idvar = "Estimator", timevar = "group", direction = "wide")
names(coverage_wide) <- sub("coverage\\.", "", names(coverage_wide))
names(coverage_wide) <- sub("factor_cov", "Factor Covariances", names(coverage_wide))
names(coverage_wide) <- sub("factor_var", "Factor Variances", names(coverage_wide))
names(coverage_wide) <- sub("loading", "Loadings", names(coverage_wide))
names(coverage_wide) <- sub("residual_var", "Residual Variances", names(coverage_wide))
for (j in setdiff(names(coverage_wide), "Estimator")) coverage_wide[[j]] <- pct(coverage_wide[[j]])

raw_improper <- mean(config_diag$improper_rate[config_diag$estimator == raw_std], na.rm = TRUE)
restricted_sentence <- if (restricted_present) {
  "The correctly restricted regime is present and reported separately below."
} else {
  "This result set contains no paper-grade restricted cells, so it cannot yet answer whether imposing correct tau/equal-residual restrictions improves the estimator."
}
```

## Question

Guttman's multiple-group construction gives a closed-form CFA estimator
[@guttman1952], and recent work has renewed interest in non-iterative CFA as a
practical estimator class [@dhaene2024]. We study the chosen recipe, the aligned
map with extended triad least-squares communalities and a standardized composite.
We ask where it delivers useful point estimates and empirical-sandwich intervals,
how it compares to the legacy Guttman construction and to ML and ULS, and how
much is gained when correct loading and residual restrictions are imposed.

## Short answer

**For the completed configural regime, the method works well only away from the
combination of weak loadings and strongly correlated factors.** With moderate
loadings, Guttman's median common-covariance RMSE relative to ML rises from
`r fmt(raw_moderate_rho[["0"]])` at factor correlation 0 to
`r fmt(raw_moderate_rho[[".8"]])` at .80. With weak loadings, the corresponding
high-correlation deterioration is orders of magnitude larger. Improper returned
solutions are reported alongside RMSE (`r pct(raw_improper)` of returned Guttman
fits in the configural regime), since accuracy conditional on a returned fit
cannot be read without them. Conditional empirical-SE coverage remains near
nominal for the Guttman arm, but cannot be interpreted without the separate fit
and admissibility rates.

`r restricted_sentence`

## Study design

The simulation treats the fitted common covariance as the primary point target.
For each design cell, RMSE is divided by the RMSE of the relevant ML fit on the
same successful replications. Values near 1 indicate ML-level accuracy; values
above 1 indicate efficiency loss. Parameter-specific bias, RMSE, empirical-SE
coverage, fit availability, and improper solutions are secondary outcomes.
Reporting failures separately follows the longstanding warning that convergence
and improper solutions are themselves simulation outcomes in CFA [@boomsma1985].
The design and performance measures are fixed before inspecting individual
cells, in the spirit of simulation-study guidance from @morris2019.

```{r design-table}
knitr::kable(design_table, row.names = FALSE, caption = "Conditions represented in the loaded results.")
```

The two population regimes have different jobs. The congeneric regime asks how
the unrestricted estimator behaves under heterogeneous loadings. The restricted
regime generates tau-equivalent loadings and equal residual variances within
each factor, then fits configural and correctly restricted estimators to the
same draws. This pairing identifies the gain from using valid restrictions;
misspecified restrictions are a separate robustness question.

```{r design-status}
knitr::kable(status_table, row.names = FALSE, caption = "Which parts of the paper design are complete in this result set.")
```

Five-category ordinal variables are analyzed through their ordinary score
covariance rather than polychoric correlations. That arm therefore studies an
ordinal-as-continuous use case, whose adequacy is known to depend on the number
of categories and marginal shape [@rhemtulla2012]. The mildly non-normal arm
uses empirical fourth moments for sandwich SEs, consistent with the
distribution-free covariance-structure framework [@browne1984].

## Configural evidence

### Aggregate accuracy

The first aggregation keeps factor correlation visible because averaging over
it hides the main failure mode. The entries are median cell-wise RMSE ratios
relative to ML.

```{r rho-evidence}
knitr::kable(rho_wide, row.names = FALSE,
             caption = "Common-covariance RMSE relative to ML, aggregated over the other configural conditions.")
```

The chosen aligned map and the legacy Guttman construction separate most
clearly as factors become correlated: the aligned map carries the cross-factor
information the legacy incidence map discards.

### Where the aggregate comes from

Separating loading strength reveals the interaction concealed by the preceding
margin. Moderate-loading cells deteriorate gradually as factors become more
correlated. Weak-loading cells become unstable much earlier. This is the main
design slice to carry into the paper.

```{r strength-evidence}
knitr::kable(strength_wide, row.names = FALSE,
             caption = "RMSE ratio by loading strength and factor correlation.")
```

Sample size moderates but does not remove this interaction. The following plot
shows the median over the remaining conditions; the logarithmic vertical scale
is necessary because the stressed cells span several orders of magnitude.

```{r n-rho-plot, fig.width=8, fig.height=5.5, fig.cap="RMSE ratio by sample size and factor correlation. Lines summarize the chosen Guttman map and the legacy construction over the remaining configural conditions."}
plot_arms <- c(raw_std, "guttman_legacy")
n_slice <- aggregate(common_rmse_ratio ~ estimator + n + rho,
                     config[config$estimator %in% plot_arms, ], med)
old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1), oma = c(0, 0, 1, 0))
for (r in sort(unique(n_slice$rho))) {
  z <- n_slice[n_slice$rho == r, ]
  yr <- range(z$common_rmse_ratio[z$common_rmse_ratio > 0], finite = TRUE)
  plot(NA, xlim = range(z$n), ylim = yr, log = "xy", xlab = "Sample size",
       ylab = "RMSE ratio vs ML", main = paste0("Factor correlation = ", r),
       xaxt = "n")
  axis(1, at = sort(unique(z$n)))
  abline(h = 1, col = "grey70", lty = 2)
  for (i in seq_along(plot_arms)) {
    zz <- z[z$estimator == plot_arms[i], ]
    zz <- zz[order(zz$n), ]
    lines(zz$n, zz$common_rmse_ratio, type = "b", pch = c(16, 1)[i],
          lty = i, col = c("#1B4F72", "#B03A2E")[i])
  }
}
legend("topleft", legend = c("Guttman (chosen)", "Guttman (legacy)"), lty = 1:2,
       pch = c(16, 1), col = c("#1B4F72", "#B03A2E"), bty = "n")
par(old_par)
```

### Inference and admissibility

Coverage below is conditional on a finite estimate and SE. It is summarized by
parameter family because a single overall coverage number overweights models
with more indicators. The Guttman intervals are generally close to 95%; ML and
ULS show more small-sample undercoverage in this grid.

```{r coverage-evidence}
knitr::kable(coverage_wide, row.names = FALSE,
             caption = "Median 95% empirical-sandwich coverage across configural cells.")
```

That conditional result is only one part of operating performance. A returned
fit is called improper when it has a negative residual variance or a non-positive
fitted covariance eigenvalue; the closed-form map returns these on the stressed
cells, so improper rates belong beside the coverage numbers.

```{r admissibility-evidence}
knitr::kable(admissibility, row.names = FALSE,
             caption = "Fit availability and admissibility in the configural regime.")
```

## Correctly restricted evidence

```{r restricted-evidence, results='asis'}
if (!restricted_present) {
  cat(paste0(
    "This section is intentionally empty: the loaded run used `include_restricted=FALSE`. ",
    "The small development smoke under `results/check-restricted` is not evidence. ",
    "Run the paper grid below. It fits configural and correctly restricted models ",
    "to each tau/equal-residual draw, so the comparison is paired.\n"))
} else {
  tau <- whole[whole$population_model == "tau_resid", , drop = FALSE]
  pairs <- data.frame(
    Family = c("Guttman", "ML", "ULS"),
    configural = c(raw_std, "ml_emp", "uls_emp"),
    restricted = c("guttman_std_raw_restricted",
                   "ml_emp_restricted", "uls_emp_restricted"),
    stringsAsFactors = FALSE)
  rows <- lapply(seq_len(nrow(pairs)), function(i) {
    a <- aggregate(common_rmse_ratio ~ rho,
                   tau[tau$estimator == pairs$configural[[i]], ], med)
    b <- aggregate(common_rmse_ratio ~ rho,
                   tau[tau$estimator == pairs$restricted[[i]], ], med)
    names(a)[[2]] <- "Configural Fit"
    names(b)[[2]] <- "Correctly Restricted Fit"
    x <- merge(a, b, by = "rho", all = TRUE)
    x$`Estimator Family` <- pairs$Family[[i]]
    x
  })
  tab <- do.call(rbind, rows)
  names(tab)[names(tab) == "rho"] <- "Factor Correlation"
  tab <- tab[c("Factor Correlation", "Estimator Family", "Configural Fit",
               "Correctly Restricted Fit")]
  tab$`Configural Fit` <- fmt(tab$`Configural Fit`)
  tab$`Correctly Restricted Fit` <- fmt(tab$`Correctly Restricted Fit`)
  cat("\n\n")
  print(knitr::kable(
    tab, row.names = FALSE,
    caption = "Paired RMSE in the tau/equal-residual population, relative to restricted ML."))

  tau_diag <- diag[diag$population_model == "tau_resid", , drop = FALSE]
  tau_est <- est[est$population_model == "tau_resid", , drop = FALSE]
  keep <- unique(c(pairs$configural, pairs$restricted))
  op_rows <- lapply(keep, function(estimator) {
    d <- tau_diag[tau_diag$estimator == estimator, ]
    e <- tau_est[tau_est$estimator == estimator, ]
    cov_group <- function(group) med(e$coverage[e$group == group])
    data.frame(
      Estimator = label(estimator),
      Loadings = pct(cov_group("loading")),
      `Factor Covariances` = pct(cov_group("factor_cov")),
      `Residual Variances` = pct(cov_group("residual_var")),
      `Fit Returned` = pct(mean(d$fit_ok_rate, na.rm = TRUE)),
      `Improper Among Returned Fits` = pct(mean(d$improper_rate, na.rm = TRUE)),
      check.names = FALSE)
  })
  op <- do.call(rbind, op_rows)
  cat("\n\n")
  print(knitr::kable(
    op, row.names = FALSE,
    caption = "Coverage and admissibility in the correctly restricted population."))
  cat("\n\n")
}
```

## Interpretation

The current evidence supports a narrow claim. Closed-form Guttman CFA is
competitive when factors are weakly correlated and the measurement blocks are
well determined. It is not a drop-in replacement for ML under weak measurement
and strong factor correlation. The estimator is reported raw: no admissibility
clamp or score/spectral conditioning is applied, because separate calibration of
those repairs produced no acceptable accuracy gain, and improper solutions are
reported as an outcome rather than suppressed.

The restriction question is co-primary for the paper. The paired restricted
regime is designed to show whether correct tau/equal-residual information
recovers enough efficiency and stability to change that conclusion. Until that
regime is complete, the report is a configural result plus a preregistered
restricted design, not the finished paper simulation.

## Caveats

- RMSE ratios are conditional on the same finite replications as the ML
  denominator. Fit availability and improper solutions must be reported beside
  them.
- Cell medians describe the typical design condition. The disaggregated tables
  and plot are needed because tail behavior is scientifically important here.
- The ordinal arm concerns Pearson covariances of category scores. It does not
  establish performance with polychoric input.
- Correct restrictions and misspecified restrictions answer different
  questions. This design covers the former only.

## Reproduce

```sh
# Fast validation of both population regimes and every estimator family.
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R \
  --smoke --generators normal --factors 2 --indicators 3 --rho 0 \
  --scales equal --strength moderate --n 120 --reps 2 --cores 1 \
  --results-dir experiments/58-guttman-rmse-coverage/results/probe

# Paper grid. Restricted cells are included by default and use paired
# configural/restricted fits on the same draws.
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R \
  --full --reps 150 --cores 8 \
  --results-dir experiments/58-guttman-rmse-coverage/results/paper-full

quarto render experiments/58-guttman-rmse-coverage/report.qmd \
  -P results_dir:results/paper-full
```

## References

::: {#refs}
:::
