---
title: "Raw and soft-clamped Guttman CFA: RMSE, coverage, and admissibility"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: results
---

```{r setup}
results_dir <- params$results_dir
need <- c("metadata", "design", "estimate_summary", "whole_joint_summary",
          "diagnostic_summary", "timing_summary", "progress")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run the smoke check or full grid shown under Reproduce.")
}
metadata <- read.csv(paths[1]); design <- read.csv(paths[2])
est <- read.csv(paths[3]); whole_joint <- read.csv(paths[4])
diag <- read.csv(paths[5]); timing <- read.csv(paths[6]); progress <- read.csv(paths[7])

meta <- function(key) {
  x <- metadata$value[metadata$key == key]
  if (length(x)) x[[1]] else ""
}
fmt <- function(x, digits = 2) {
  ifelse(is.finite(x), formatC(x, format = "f", digits = digits), "NA")
}
pct <- function(x, digits = 1) {
  ifelse(is.finite(x), paste0(formatC(100 * x, format = "f", digits = digits), "%"), "NA")
}
first_available <- function(ids) {
  found <- ids[ids %in% unique(whole_joint$estimator)]
  if (length(found)) found[[1]] else NA_character_
}

# `guttman_std` names the completed historical raw screen. New runs label the
# policy explicitly, while retaining this reader for the existing evidence.
raw_std <- first_available(c("guttman_std_raw", "guttman_std"))
soft_std <- first_available(c("guttman_std_soft"))
legacy <- first_available(c("guttman_legacy"))
ml <- first_available(c("ml_emp"))
uls <- first_available(c("uls_emp"))
if (is.na(raw_std) || is.na(ml)) stop("Results lack the standardized raw map or ML reference.")

label_map <- c(
  guttman_std = "Standardized Guttman (raw)",
  guttman_std_raw = "Standardized Guttman (raw)",
  guttman_std_soft = "Standardized Guttman (soft clamp)",
  guttman_legacy = "Legacy Guttman",
  ml_emp = "ML (reference)",
  uls_emp = "ULS")
label <- function(x) unname(label_map[x])
primary <- c(raw_std, soft_std, legacy, ml, uls)
primary <- unique(primary[!is.na(primary)])

rmse_by_rho <- function(estimator) {
  w <- whole_joint[whole_joint$estimator == estimator &
                     is.finite(whole_joint$common_rmse_ratio), ]
  d <- diag[diag$estimator == estimator, ]
  rmse <- aggregate(common_rmse_ratio ~ rho, w, median, na.rm = TRUE)
  fit <- aggregate(fit_ok_rate ~ rho, d, mean, na.rm = TRUE)
  improper <- aggregate(improper_rate ~ rho, d, mean, na.rm = TRUE)
  out <- Reduce(function(x, y) merge(x, y, by = "rho", all = TRUE),
                list(rmse, fit, improper))
  out$Estimator <- label(estimator)
  out
}
rho_summary <- do.call(rbind, lapply(primary, rmse_by_rho))
rho_summary <- rho_summary[, c("Estimator", "rho", "common_rmse_ratio",
                               "fit_ok_rate", "improper_rate")]
names(rho_summary) <- c("Estimator", "Factor correlation", "Median RMSE vs ML",
                        "Fit success", "Improper solution")
rho_summary$`Median RMSE vs ML` <- fmt(rho_summary$`Median RMSE vs ML`)
rho_summary$`Fit success` <- pct(rho_summary$`Fit success`)
rho_summary$`Improper solution` <- pct(rho_summary$`Improper solution`)

coverage_table <- function(estimators) {
  rows <- lapply(estimators, function(estimator) {
    x <- est[est$estimator == estimator & is.finite(est$coverage), ]
    by_generator <- tapply(x$coverage, x$generator, median, na.rm = TRUE)
    at <- function(name) {
      if (name %in% names(by_generator)) by_generator[[name]] else NA_real_
    }
    data.frame(
      Estimator = label(estimator),
      Normal = pct(at("normal")),
      `Non-normal` = pct(at("ig")),
      Ordinal = pct(at("ordinal")),
      check.names = FALSE)
  })
  do.call(rbind, rows)
}
coverage_summary <- coverage_table(primary)

ratio_at <- function(rho) {
  x <- whole_joint[whole_joint$estimator == raw_std & whole_joint$rho == rho &
                     is.finite(whole_joint$common_rmse_ratio), "common_rmse_ratio"]
  median(x)
}
raw_ratios <- setNames(vapply(c(0, .35, .6, .8), ratio_at, numeric(1)),
                        c("0", ".35", ".6", ".8"))
raw_coverage <- median(est$coverage[est$estimator == raw_std & is.finite(est$coverage)])
soft_note <- if (is.na(soft_std)) {
  "The completed screen predates the soft-clamped arm. The runner now fixes that arm before the next confirmation run; this report does not impute an outcome for it."
} else {
  paste0("The soft-clamped arm is present in these results. Its row is shown alongside the raw control in both evidence tables.")
}
```

## Question

When does a closed-form Guttman CFA map remain a useful full estimator—point
estimate plus empirical standard error—relative to ML and ULS? In particular,
does smoothly enforcing the physically admissible communality range make the
standardized aligned map a better practical estimator than its raw counterpart?

## Short answer

**The completed raw screen identifies factor correlation, not distributional
shape, as the decisive stressor.** The raw standardized map is at ML parity at
rho = 0 (median RMSE ratio `r fmt(raw_ratios[["0"]])`), rises to
`r fmt(raw_ratios[[".35"]])` at rho = .35, `r fmt(raw_ratios[[".6"]])` at .60,
and `r fmt(raw_ratios[[".8"]])` at .80. Its median conditional empirical-SE
coverage is `r pct(raw_coverage)`, but that statement is conditional on a
returned fit and must be read with the improper-solution rates below.

`r soft_note`

The paper-primary admissible variant is therefore the standardized aligned map
with a symmetric smooth communality clamp: margin .001, base sharpness 4, and
sample-size rate 1. Score-covariance and full-H spectral conditioning remain
raw; the calibration evidence does not support either repair as a default.

## Evidence

The completed raw screen has `r nrow(progress)` configural cells and
`r meta("reps")` replications per cell. RMSE is calculated on the same draws as
ML and only when both fits are finite. “Improper solution” is reported over
returned fits, separately from fit failure.

```{r rho-evidence}
knitr::kable(rho_summary, row.names = FALSE)
```

The raw map’s conditional coverage remains close to 95% across all three data
generators, while ML and ULS are included to show the same small-sample and
non-normal context. Coverage does not erase a point-estimation or admissibility
failure.

```{r coverage-evidence}
knitr::kable(coverage_summary, row.names = FALSE)
```

## Conditions

The completed screen is configural only: 2, 3, or 5 factors; 3, 5, or an
unbalanced 3/6 indicator pattern; sample sizes 50, 100, 300, and 800; factor
correlations 0, .35, .60, and .80; equal or unequal scales; moderate or weak
loadings; and normal, mildly non-normal, or ordinal-score data. The ordinal arm
fits the ordinary covariance of integer category scores, not polychorics.

New runs compare five arms: raw standardized Guttman, soft-clamped standardized
Guttman, legacy Guttman, ML, and ULS. The soft clamp is deliberately an
estimator choice rather than a hidden cleanup. The retired adaptive composite is
not a primary competitor here; its compatibility behavior remains available in
the library and in the earlier raw screen.

## Caveats

- This completed evidence is a broad raw baseline, not the soft-clamp
  confirmation. The next run must include the configured soft arm before any
  claim that it improves the final estimator.
- The smooth communality clamp prevents impossible communality values, but does
  not by itself guarantee a positive-definite factor covariance. The raw
  high-rho failure remains a substantive finding, not a result to conceal.
- RMSE and coverage are conditional on finite fits; success and improper rates
  are part of the result, not implementation footnotes.
- The constrained tau-equivalent regime is intentionally not reported here:
  it was not included in the completed full run and needs its own completed
  confirmation before it is added to this report.

## Reproduce

```sh
# Configural smoke check; writes a separate result directory.
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R \
  --smoke --max-cells 1 --reps 2 \
  --results-dir experiments/58-guttman-rmse-coverage/results/probe

# Configural full grid, including the raw and soft-clamped standardized maps.
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R \
  --full --reps 150 --cores 8 \
  --results-dir experiments/58-guttman-rmse-coverage/results

# Add the constrained regime only for a dedicated follow-up run.
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R \
  --full --restricted --reps 150 --cores 8 \
  --results-dir /tmp/exp58-restricted

quarto render experiments/58-guttman-rmse-coverage/report.qmd
```
