---
title: "Guttman RMSE and Coverage Grid"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: results
---

```{r setup}
results_dir <- params$results_dir
need <- c("metadata", "design", "population", "estimate_summary",
          "estimate_joint_summary", "whole_summary", "whole_joint_summary",
          "diagnostic_summary", "timing_summary", "progress")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --probe\n",
       "or:\n  Rscript run_experiment.R --full --cores 3")
}
metadata <- read.csv(paths[1])
design <- read.csv(paths[2])
population <- read.csv(paths[3])
est <- read.csv(paths[4])
est_joint <- read.csv(paths[5])
whole <- read.csv(paths[6])
whole_joint <- read.csv(paths[7])
diag <- read.csv(paths[8])
timing <- read.csv(paths[9])
progress <- read.csv(paths[10])

meta <- function(key) {
  z <- metadata$value[metadata$key == key]
  if (length(z)) z[[1]] else ""
}
fmt <- function(x, d = 3) ifelse(is.finite(x), formatC(x, format = "f", digits = d), "NA")
pct <- function(x) ifelse(is.finite(x), sprintf("%.1f%%", 100 * x), "NA")

est_label <- c(
  guttman_std = "Std Guttman ILS",
  guttman_gls = "GLS-aligned Guttman ILS",
  guttman_unit = "Unit-composite Guttman ILS",
  guttman_legacy = "Legacy Guttman",
  ml_emp = "NTML (emp SE)",
  uls_emp = "ULS (emp SE)",
  guttman_std_restricted = "Std restricted Guttman ILS",
  guttman_gls_restricted = "GLS-aligned restricted Guttman ILS",
  ml_emp_restricted = "Restricted NTML (emp SE)",
  uls_emp_restricted = "Restricted ULS (emp SE)"
)
label <- function(x) ifelse(x %in% names(est_label), est_label[x], x)
largest_n <- max(design$n)
```

## Question

How do the implemented Guttman CFA maps behave as full estimators once we vary
sample size, number of factors, factor correlations, item scales, nonnormal
continuous data, and category-score data?

The ordinal cells are deliberately **not** polychoric CFA cells. They use
`magmaan::sim_ordcorr_calibrate(metric = "pearson_codes")` to generate ordinal
category scores whose observed Pearson-code correlation matrix is the declared
CFA target. Guttman, NTML, and ULS then all see the same sample covariance of
those numeric category scores.

## Design

```{r design-table}
knitr::kable(data.frame(
  Quantity = c("Mode", "Replications per cell", "Cells requested", "Cells complete",
               "Generators", "Factors", "Indicators", "n", "Categorical target"),
  Value = c(meta("mode"), meta("reps"), nrow(design), sum(progress$draw_ok),
            meta("generators"), meta("factors"), meta("indicators"), meta("n"),
            meta("categorical_metric"))
), row.names = FALSE)
```

Constrained cells use true within-factor tau-equivalent loadings and equal
residual variances, then fit the estimator-side residual/loading restricted
Guttman map. Configural cells use the same data-generation family without
imposing those equalities during estimation.

## Whole-Estimator RMSE

The table reports whole common-covariance RMSE ratios against the matching NTML
arm on the same successful replications. Values above 1 mean higher RMSE than
NTML for the same condition.

```{r whole-table}
show <- subset(whole_joint, n == largest_n &
                 estimator %in% c("guttman_std", "guttman_gls",
                                  "guttman_unit", "guttman_legacy",
                                  "guttman_std_restricted",
                                  "guttman_gls_restricted",
                                  "ml_emp", "ml_emp_restricted",
                                  "uls_emp", "uls_emp_restricted"))
show$Estimator <- label(show$estimator)
show$`Common / NTML` <- fmt(show$common_rmse_ratio, 2)
show$`Sigma / NTML` <- fmt(show$sigma_rmse_ratio, 2)
show$`Common diag / NTML` <- fmt(show$common_diag_rmse_ratio, 2)
show$`Residual diag / NTML` <- fmt(show$residual_diag_rmse_ratio, 2)
show$`Joint reps` <- show$n_joint_reps
show <- show[order(show$constraint, show$generator, show$factors,
                   show$indicators, show$rho, show$scale, show$Estimator), ]
knitr::kable(show[, c("constraint", "generator", "factors", "indicators",
                      "rho", "scale", "Estimator", "Common / NTML",
                      "Sigma / NTML", "Common diag / NTML",
                      "Residual diag / NTML", "Joint reps")],
             row.names = FALSE)
```

## Parameter Coverage

Wald coverage uses empirical fourth-moment covariance for all reported arms.
The NTML rows are still normal-theory ML point estimates; only the SE is the
empirical sandwich version.

```{r coverage-table}
cov <- subset(est, n == largest_n &
                estimator %in% c("guttman_std", "guttman_gls",
                                 "guttman_unit", "guttman_legacy",
                                 "ml_emp", "uls_emp",
                                 "guttman_std_restricted",
                                 "guttman_gls_restricted",
                                 "ml_emp_restricted", "uls_emp_restricted"))
cov$Estimator <- label(cov$estimator)
cov$Coverage <- pct(cov$coverage)
cov$RMSE <- fmt(cov$rmse)
cov$`Avg SE` <- fmt(cov$avg_se)
cov$`Finite reps` <- cov$n_finite_reps
cov <- cov[order(cov$constraint, cov$generator, cov$factors, cov$indicators,
                 cov$rho, cov$scale, cov$group, cov$Estimator), ]
knitr::kable(cov[, c("constraint", "generator", "factors", "indicators",
                     "rho", "scale", "group", "Estimator", "Coverage",
                     "RMSE", "Avg SE", "Finite reps")],
             row.names = FALSE)
```

## Failures And Heywood-Like Diagnostics

`improper_rate` flags a negative residual variance or non-positive fitted
Sigma eigenvalue. These rows are important for the small-n cells: RMSE ratios
are same-replication comparisons, but failure rates tell us which estimators
are reaching that comparison set.

```{r diag-table}
d <- subset(diag, n == largest_n)
d$Estimator <- label(d$estimator)
d$`Fit OK` <- pct(d$fit_ok_rate)
d$Converged <- pct(d$converged_rate)
d$`SE OK` <- pct(d$se_ok_rate)
d$Improper <- pct(d$improper_rate)
d$`Residual p05` <- fmt(d$min_resid_p05)
d$`Sigma eig p05` <- fmt(d$sigma_min_eig_p05)
d <- d[order(d$constraint, d$generator, d$factors, d$indicators, d$rho,
             d$scale, d$Estimator), ]
knitr::kable(d[, c("constraint", "generator", "factors", "indicators",
                   "rho", "scale", "Estimator", "Fit OK", "Converged",
                   "SE OK", "Improper", "Residual p05", "Sigma eig p05",
                   "first_error")],
             row.names = FALSE)
```

## Timing

The timing includes fitting plus the SE path. It is therefore closer to the
workload of the RMSE/coverage tables than a communality-only benchmark.

```{r timing-table}
t <- timing
t$Estimator <- label(t$estimator)
t$`Fit median ms` <- fmt(t$fit_median_ms, 1)
t$`SE median ms` <- fmt(t$se_median_ms, 1)
t$`Total median ms` <- fmt(t$total_median_ms, 1)
t$`Total p90 ms` <- fmt(t$total_p90_ms, 1)
t <- t[order(t$constraint, t$generator, t$factors, t$indicators, t$rho,
             t$scale, t$Estimator), ]
knitr::kable(t[, c("constraint", "generator", "factors", "indicators",
                   "rho", "scale", "Estimator", "Fit median ms",
                   "SE median ms", "Total median ms", "Total p90 ms")],
             row.names = FALSE)
```

## Progress

```{r progress-table}
p <- progress
p$`Draw ms` <- fmt(p$draw_ms, 1)
p$`Cell elapsed sec` <- fmt(p$elapsed_sec, 1)
knitr::kable(p[, c("cell", "generator", "factors", "indicators", "rho",
                   "scale", "loading", "constraint", "n", "reps",
                   "draw_ok", "Draw ms", "Cell elapsed sec", "error")],
             row.names = FALSE)
```
