---
title: "Guttman CFA maps as full estimators: efficiency, coverage, cost"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: results
---

```{r setup}
results_dir <- params$results_dir
need <- c("metadata", "design", "population", "estimate_summary",
          "estimate_joint_summary", "whole_summary", "whole_joint_summary",
          "diagnostic_summary", "timing_summary", "progress")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --probe\n",
       "or the full grid (see Reproduce).")
}
metadata <- read.csv(paths[1]); design <- read.csv(paths[2])
population <- read.csv(paths[3]); est <- read.csv(paths[4])
est_joint <- read.csv(paths[5]); whole <- read.csv(paths[6])
whole_joint <- read.csv(paths[7]); diag <- read.csv(paths[8])
timing <- read.csv(paths[9]); progress <- read.csv(paths[10])

meta <- function(key) {
  z <- metadata$value[metadata$key == key]; if (length(z)) z[[1]] else ""
}
fmt <- function(x, d = 2) ifelse(is.finite(x), formatC(x, format = "f", digits = d), "NA")
pct <- function(x, d = 0) ifelse(is.finite(x), paste0(formatC(100 * x, format = "f", digits = d), "%"), "NA")

est_label <- c(
  guttman_gls = "GLS-aligned Guttman", guttman_std = "Standardized Guttman",
  guttman_unit = "Unit-weight Guttman", guttman_legacy = "Legacy Guttman",
  ml_emp = "ML (reference)", uls_emp = "ULS",
  guttman_gls_restricted = "GLS-aligned Guttman", guttman_std_restricted = "Standardized Guttman",
  ml_emp_restricted = "ML (reference)", uls_emp_restricted = "ULS")
label <- function(x) ifelse(x %in% names(est_label), est_label[x], x)

# estimator sets per constraint
cfg_ratio <- c("guttman_gls", "guttman_std", "guttman_unit", "guttman_legacy", "uls_emp")
res_ratio <- c("guttman_gls_restricted", "guttman_std_restricted", "uls_emp_restricted")
cfg_all <- c("guttman_gls", "guttman_std", "guttman_unit", "guttman_legacy", "ml_emp", "uls_emp")
res_all <- c("guttman_gls_restricted", "guttman_std_restricted", "ml_emp_restricted", "uls_emp_restricted")
aligned <- c("guttman_gls", "guttman_std", "guttman_unit")

# --- table builders (constraint-specific, apples-to-apples) ---
eff_table <- function(constraint, ests) {
  do.call(rbind, lapply(ests, function(e) {
    x <- whole_joint[whole_joint$estimator == e & whole_joint$constraint == constraint &
                       is.finite(whole_joint$common_rmse_ratio), ]
    data.frame(
      Estimator = label(e),
      `Median RMSE vs ML` = fmt(median(x$common_rmse_ratio)),
      `Beats ML` = pct(mean(x$common_rmse_ratio < 1)),
      `Median at n=100` = fmt(median(x$common_rmse_ratio[x$n == 100])),
      `Median at n=800` = fmt(median(x$common_rmse_ratio[x$n == 800])),
      check.names = FALSE)
  }))
}
cov_table <- function(constraint, ests) {
  ce <- est[est$estimator %in% ests & est$constraint == constraint, ]
  m <- tapply(ce$coverage, list(ce$estimator, ce$generator), median, na.rm = TRUE)
  ord <- ests[ests %in% rownames(m)]
  data.frame(
    Estimator = label(ord),
    Normal = pct(m[ord, "normal"], 1),
    `Non-normal` = pct(m[ord, "ig"], 1),
    Ordinal = pct(m[ord, "ordinal"], 1),
    check.names = FALSE)
}
cost_table <- function(constraint, ests) {
  tt <- timing[timing$estimator %in% ests & timing$constraint == constraint, ]
  big <- tt[tt$factors == 5 & as.character(tt$indicators) == "5", ]
  do.call(rbind, lapply(ests, function(e) {
    data.frame(
      Estimator = label(e),
      `Fit-only (ms)` = fmt(median(tt$fit_mean_ms[tt$estimator == e]), 2),
      `Fit+SE (ms)` = fmt(median(tt$total_median_ms[tt$estimator == e]), 0),
      `25-var fit-only (ms)` = fmt(median(big$fit_mean_ms[big$estimator == e]), 2),
      `25-var fit+SE (ms)` = fmt(median(big$total_median_ms[big$estimator == e]), 0),
      check.names = FALSE)
  }))
}

# --- headline numbers reused in prose ---
gutt <- c("guttman_gls", "guttman_std", "guttman_unit", "guttman_legacy")
wj <- whole_joint[is.finite(whole_joint$common_rmse_ratio), ]
wjg <- wj[wj$estimator %in% gutt, ]
med_ratio <- median(wjg$common_rmse_ratio)
wjg$cell <- with(wjg, paste(generator, factors, indicators, rho, scale, constraint, n))
best_ratio <- tapply(wjg$common_rmse_ratio, wjg$cell, min)
best_n <- tapply(wjg$n, wjg$cell, `[`, 1)
win_best <- mean(best_ratio < 1)
win_100 <- mean(best_ratio[best_n == 100] < 1)
win_800 <- mean(best_ratio[best_n == 800] < 1)
cov_g <- est$coverage[est$estimator %in% gutt & is.finite(est$coverage)]
cov_lo <- quantile(cov_g, 0.1, names = FALSE); cov_hi <- quantile(cov_g, 0.9, names = FALSE)
# aligned vs legacy, configural only
wj_cfg <- whole_joint[whole_joint$constraint == "configural" & is.finite(whole_joint$common_rmse_ratio), ]
med_aligned <- median(wj_cfg$common_rmse_ratio[wj_cfg$estimator %in% aligned])
med_legacy <- median(wj_cfg$common_rmse_ratio[wj_cfg$estimator == "guttman_legacy"])
```

## Question

We have four closed-form (non-iterative) Guttman CFA maps. Used as full
estimators (point estimate plus standard error), how close do they get to
iterative maximum likelihood on accuracy and coverage across sample size, model
size, factor correlation, item scaling, non-normal data, and ordinal
category-score data? And a second question aimed at our own machinery: do the
three *aligned* maps, which spend the full within-factor covariance, actually
beat the plain *legacy* incidence map that does not?

## Short answer

**The closed-form maps are competitive full estimators, and the aligned ones do
not earn their extra machinery here.** Across the grid the maps carry a median of
only about **`r fmt((med_ratio - 1) * 100, 0)`% more error** than ML on the
common covariance. ML is modestly more efficient on average, which is what theory
predicts under a correct, near-normal model, so that small gap is the expected
price of a non-iterative fit. ML does not win everywhere, though: the best map
beats it in **`r pct(win_best)` of conditions**, rising to **`r pct(win_100)` at
n = 100** (down to `r pct(win_800)` at n = 800). Coverage of the empirical
standard error stays near the nominal 95% (about
`r fmt(cov_lo*100,0)`–`r fmt(cov_hi*100,0)`%) for every map, uniformly on normal,
non-normal, and ordinal-score data.

On the aligned-versus-legacy question the answer so far is **no**: the three
aligned maps (median RMSE `r fmt(med_aligned)` vs ML) track the legacy incidence
map (`r fmt(med_legacy)`) to within about a percent. Spending the full covariance
buys essentially nothing in this grid. The likely reason is that factor
correlations are low here (rho ≤ 0.35), so the cross-factor information the
aligned reconstruction exploits has little to bite on; whether it pays off at
higher rho is left open.

## What the conditions mean

Every cell is one combination of the following. The full grid is
`r nrow(design)` cells, `r meta("reps")` replications each.

- **Generator — how the data are drawn.** *Normal*: multivariate normal.
  *Non-normal* (`ig`): mildly skewed, heavier-tailed marginals (skewness 0.2,
  excess kurtosis 0.5). *Ordinal*: ordered-category scores calibrated so the
  ordinary correlation of the integer codes equals the CFA target; the ordinal
  arm fits the plain covariance of those codes and is **not** a polychoric
  analysis.
- **Model size.** `r meta("factors")` factors, `r meta("indicators")` indicators
  per factor, so 6 to 25 observed variables.
- **Factor correlation (rho).** `r meta("rho")` — orthogonal versus mildly
  correlated factors.
- **Item scaling.** *Equal* versus *unequal* indicator scales (or ordinal
  marginals).
- **Configural vs restricted.** *Configural*: loadings and residual variances all
  free. *Restricted*: the population is genuinely tau-equivalent (equal loadings
  within a factor, equal residual variances) and the estimator imposes that, so
  the restricted maps are judged on a correctly specified target, never under
  misspecification. The two regimes are reported separately below.
- **Estimators.** Four closed-form Guttman maps, in two families. The three
  *aligned* maps take each item's communality from the full within-factor
  covariance (a pooled, block-GLS estimate) and reconstruct loadings from an
  aligned composite, differing only in the composite weight — *standardized*,
  *GLS-aligned*, or *unit*. The *legacy* map is the classic incidence
  construction: communalities from the Spearman/Guttman (1952)
  **average-of-ratios** formula (the same one lavaan uses,
  `h²ₖ = mean of Rₐₖ·R_bₖ / Rₐb` over item pairs, with wide bounds), and a unit
  incidence composite. Average-of-ratios has wider sampling variance than the
  pooled communality the aligned maps use, so legacy is the natural "does the
  extra machinery help?" baseline. Iterative **ML** and **ULS** are the
  references. Every arm reports an empirical standard error. Only the aligned
  maps are run in the restricted regime.

## Efficiency: error relative to ML

Same-replication RMSE on the common covariance, as a ratio to ML (below 1 = lower
error than ML), with the median split by sample size.

**Configural.**

```{r eff-configural}
knitr::kable(eff_table("configural", cfg_ratio), row.names = FALSE)
```

**Restricted (tau-equivalent).** The constraint removes most of the room for the
estimators to differ, so the aligned maps sit even closer to ML.

```{r eff-restricted}
knitr::kable(eff_table("tau_resid", res_ratio), row.names = FALSE)
```

In neither regime do the aligned maps open a clear gap on legacy; per-cell detail
is in `results/whole_joint_summary.csv`.

## Coverage: are the intervals honest?

Median Wald coverage of the empirical standard error, nominal 95%, split by how
the data were generated. This holds across data types, not just under normality.

**Configural.**

```{r cov-configural}
knitr::kable(cov_table("configural", cfg_all), row.names = FALSE)
```

**Restricted (tau-equivalent).**

```{r cov-restricted}
knitr::kable(cov_table("tau_resid", res_all), row.names = FALSE)
```

Every map and ML land within about a point of nominal; ULS runs consistently a
little low.

## Cost and reliability

Fit on its own versus fit **plus** standard error. The closed-form maps fit in a
fraction of a millisecond, several times faster than ML's fit, but the empirical
standard error re-evaluates the map many times and dominates the total, so on the
25-variable models the aligned maps are slower than ML end to end.

**Configural.**

```{r cost-configural}
knitr::kable(cost_table("configural", cfg_all), row.names = FALSE)
```

**Restricted (tau-equivalent).**

```{r cost-restricted}
knitr::kable(cost_table("tau_resid", res_all), row.names = FALSE)
```

The clock is millisecond-resolution, so the sub-millisecond fits round to a zero
median; the fit-only column is therefore the mean, which recovers the sub-ms
average. Convergence and improper-solution (Heywood) rates are low and
essentially the same across estimators.

## Caveats

- **The aligned-vs-legacy verdict is conditional on low rho.** Factor
  correlations here are 0 and 0.35. The aligned reconstruction's advantage should
  grow with rho, so "aligned buys nothing" is a statement about this grid, not a
  general one; a higher-rho arm is the obvious next test.
- **Correct specification only.** Every population is a true CFA and, for the
  restricted cells, truly tau-equivalent. This measures efficiency, not
  robustness to a wrong model; ML's small average edge is expected precisely
  because the model is right.
- **Mild non-normality.** The non-normal arm is a gentle stress (skewness 0.2,
  excess kurtosis 0.5); stronger targets failed calibration on the 25-variable
  cells.
- **Ordinal means category codes, not polychorics.**
- **RMSE ratios are same-replication.** A few small-n cells produce extreme
  ratios in either direction; read the medians and win-rates, not the tails.

## Reproduce

```sh
# full grid (about 75 min on 8 cores)
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R \
  --full --reps 150 --cores 8 \
  --results-dir experiments/58-guttman-rmse-coverage/results

# fast smoke check
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R --probe \
  --results-dir experiments/58-guttman-rmse-coverage/results/probe

# render this report
quarto render experiments/58-guttman-rmse-coverage/report.qmd
```
