---
title: "Guttman CFA maps as full estimators: efficiency, coverage, cost"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: results
---

```{r setup}
results_dir <- params$results_dir
need <- c("metadata", "design", "population", "estimate_summary",
          "estimate_joint_summary", "whole_summary", "whole_joint_summary",
          "diagnostic_summary", "timing_summary", "progress")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --probe\n",
       "or the full grid (see Reproduce).")
}
metadata <- read.csv(paths[1]); design <- read.csv(paths[2])
population <- read.csv(paths[3]); est <- read.csv(paths[4])
est_joint <- read.csv(paths[5]); whole <- read.csv(paths[6])
whole_joint <- read.csv(paths[7]); diag <- read.csv(paths[8])
timing <- read.csv(paths[9]); progress <- read.csv(paths[10])

meta <- function(key) {
  z <- metadata$value[metadata$key == key]; if (length(z)) z[[1]] else ""
}
fmt <- function(x, d = 2) ifelse(is.finite(x), formatC(x, format = "f", digits = d), "NA")
pct <- function(x, d = 0) ifelse(is.finite(x), paste0(formatC(100 * x, format = "f", digits = d), "%"), "NA")

est_label <- c(
  guttman_gls = "GLS-aligned Guttman", guttman_std = "Standardized Guttman",
  guttman_unit = "Unit-weight Guttman", guttman_legacy = "Legacy Guttman",
  ml_emp = "ML (reference)", uls_emp = "ULS",
  guttman_gls_restricted = "GLS-aligned Guttman", guttman_std_restricted = "Standardized Guttman",
  ml_emp_restricted = "ML (reference)", uls_emp_restricted = "ULS")
label <- function(x) ifelse(x %in% names(est_label), est_label[x], x)

# --- headline numbers computed once, reused in prose ---
gutt <- c("guttman_gls", "guttman_std", "guttman_unit", "guttman_legacy")
wj <- whole_joint[is.finite(whole_joint$common_rmse_ratio), ]
wjg <- wj[wj$estimator %in% gutt, ]
med_ratio <- median(wjg$common_rmse_ratio)
# best closed-form map per condition cell vs ML on the same replications
wjg$cell <- with(wjg, paste(generator, factors, indicators, rho, scale, constraint, n))
best_ratio <- tapply(wjg$common_rmse_ratio, wjg$cell, min)
best_n <- tapply(wjg$n, wjg$cell, `[`, 1)
win_best <- mean(best_ratio < 1)
win_100 <- mean(best_ratio[best_n == 100] < 1)
win_800 <- mean(best_ratio[best_n == 800] < 1)
cov_g <- est$coverage[est$estimator %in% gutt & is.finite(est$coverage)]
cov_lo <- quantile(cov_g, 0.1, names = FALSE); cov_hi <- quantile(cov_g, 0.9, names = FALSE)
```

## Question

We have four closed-form (non-iterative) Guttman CFA maps. Used as full
estimators (point estimate plus standard error), how close do they get to
iterative maximum likelihood on accuracy and confidence-interval coverage, and
does that hold up as we vary sample size, model size, factor correlation, item
scaling, non-normal data, and ordinal category-score data?

## Short answer

**They are competitive full estimators, not a poor relation of ML.** Across the
grid the closed-form maps carry a median of only about
**`r fmt((med_ratio - 1) * 100, 0)`% more error** than ML on the common part of
the covariance. ML is modestly more efficient on average, which is exactly what
theory predicts when the model is correct and the data are near-normal, so that
small gap is the expected price of a non-iterative fit.

But ML does **not** win everywhere. On the same replications, the best
closed-form map beats ML in **`r pct(win_best)` of conditions**, and that share
climbs to **`r pct(win_100)` at n = 100** (falling to `r pct(win_800)` at
n = 800): ML's efficiency edge is real but concentrates at large samples and
erodes when n is small. Meanwhile the empirical standard error keeps
**coverage near the nominal 95%** (roughly `r fmt(cov_lo*100,0)`–`r fmt(cov_hi*100,0)`%)
for every closed-form map, and it does so uniformly on normal, non-normal, and
ordinal-score data. The one thing the closed form does *not* buy here is speed:
the fit is cheap, but its empirical standard error costs more than the whole ML
fit.

## What the conditions mean

Every cell of the grid is one combination of the following. The full grid is
`r nrow(design)` cells, `r meta("reps")` replications each.

- **Generator — how the data are drawn.** *Normal*: multivariate normal.
  *Non-normal* (`ig`): mildly skewed, heavier-tailed marginals (skewness 0.2,
  excess kurtosis 0.5). *Ordinal*: ordered-category scores calibrated so the
  ordinary correlation of the integer category codes equals the CFA target. The
  ordinal arm fits the plain covariance of those codes and is deliberately
  **not** a polychoric analysis.
- **Model size.** `r meta("factors")` factors, `r meta("indicators")` indicators
  per factor, so 6 to 25 observed variables.
- **Factor correlation (rho).** `r meta("rho")` — orthogonal versus mildly
  correlated factors.
- **Item scaling.** *Equal*: every indicator on the same scale. *Unequal*:
  heterogeneous indicator scales (or ordinal marginals).
- **Configural vs restricted.** *Configural*: loadings and residual variances
  all free. *Restricted*: the population is genuinely tau-equivalent (equal
  loadings within a factor, equal residual variances) and the estimator imposes
  that, so the restricted maps are judged on a correctly specified target, never
  under misspecification.
- **Estimators.** Four closed-form Guttman maps (GLS-aligned, standardized,
  unit-weight, legacy) plus iterative **ML** and **ULS** as references. Every
  arm reports an empirical (data-driven) standard error, so coverage is
  comparable across all of them.

## Efficiency: error relative to ML

Same-replication RMSE on the common covariance, as a ratio to ML. Below 1 means
lower error than ML. The last two columns show the median split by sample size.

```{r eff-table}
ests <- c("guttman_gls", "guttman_std", "guttman_unit", "guttman_legacy", "uls_emp")
eff <- do.call(rbind, lapply(ests, function(e) {
  x <- whole_joint[whole_joint$estimator == e & whole_joint$constraint == "configural" &
                     is.finite(whole_joint$common_rmse_ratio), ]
  data.frame(
    Estimator = label(e),
    `Median RMSE vs ML` = fmt(median(x$common_rmse_ratio)),
    `Beats ML` = pct(mean(x$common_rmse_ratio < 1)),
    `Median at n=100` = fmt(median(x$common_rmse_ratio[x$n == 100])),
    `Median at n=800` = fmt(median(x$common_rmse_ratio[x$n == 800])),
    check.names = FALSE)
}))
knitr::kable(eff, row.names = FALSE)
```

The restricted (tau-equivalent) maps sit even closer to ML, essentially tied
(median ratio about
`r fmt(median(whole_joint$common_rmse_ratio[whole_joint$estimator %in% c("guttman_gls_restricted","guttman_std_restricted") & is.finite(whole_joint$common_rmse_ratio)]))`),
because the constraint removes most of the room for the estimators to differ.
Per-cell detail lives in `results/whole_joint_summary.csv`.

## Coverage: are the intervals honest?

Median Wald coverage of the empirical standard error, nominal 95%, split by how
the data were generated. This is the robustness result: it holds across data
types, not just under normality.

```{r cov-table}
ests <- c("guttman_gls", "guttman_std", "guttman_unit", "guttman_legacy", "ml_emp", "uls_emp")
ce <- est[est$estimator %in% ests, ]
m <- tapply(ce$coverage, list(ce$estimator, ce$generator), median, na.rm = TRUE)
ord <- ests[ests %in% rownames(m)]
covtab <- data.frame(
  Estimator = label(ord),
  Normal = pct(m[ord, "normal"], 1),
  `Non-normal` = pct(m[ord, "ig"], 1),
  Ordinal = pct(m[ord, "ordinal"], 1),
  check.names = FALSE)
knitr::kable(covtab, row.names = FALSE)
```

Every closed-form map and ML land within about a point of nominal on all three
data types. ULS is the only arm that runs consistently a little low.

## Cost and reliability

Median total time for one fit **plus** its standard error, and how often the
fit is well behaved. The closed-form maps fit almost instantly, but their
empirical standard error re-evaluates the map many times, and on the 25-variable
models that dominates and makes them slower than ML end to end.

```{r cost-table}
ests <- c("guttman_gls", "guttman_std", "guttman_unit", "guttman_legacy", "ml_emp", "uls_emp")
tt <- timing[timing$estimator %in% ests, ]
big <- tt[tt$factors == 5 & tt$indicators == 5, ]
cost <- do.call(rbind, lapply(ests, function(e) {
  data.frame(
    Estimator = label(e),
    `Fit+SE median (ms)` = fmt(median(tt$total_median_ms[tt$estimator == e]), 0),
    `25-var median (ms)` = fmt(median(big$total_median_ms[big$estimator == e]), 0),
    `Min converged` = pct(min(diag$converged_rate[diag$estimator == e])),
    `Improper at n=100` = pct(median(diag$improper_rate[diag$estimator == e & diag$n == 100]), 1),
    check.names = FALSE)
}))
knitr::kable(cost, row.names = FALSE)
```

Convergence and improper-solution (Heywood) rates are low and essentially the
same across estimators, so no arm is reaching the accuracy comparison on a
different footing.

## Caveats

- **Correct specification only.** Every population here is a true CFA and, for
  the restricted cells, truly tau-equivalent. This measures efficiency, not
  robustness to a wrong model. ML's small average edge is expected precisely
  because the model is right; the interesting open question is what happens
  under misspecification.
- **Mild non-normality.** The non-normal arm is a gentle stress (skewness 0.2,
  excess kurtosis 0.5); stronger targets failed calibration on the 25-variable
  cells. Heavier tails could widen the coverage story.
- **Ordinal means category codes, not polychorics.** The ordinal result speaks
  to fitting the covariance of integer codes, not to a polychoric CFA.
- **RMSE ratios are same-replication.** A few small-n cells produce extreme
  ratios in either direction; read the medians and win-rates, not the tails.

## Reproduce

```sh
# full grid (about 75 min on 8 cores)
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R \
  --full --reps 150 --cores 8 \
  --results-dir experiments/58-guttman-rmse-coverage/results

# fast smoke check
Rscript experiments/58-guttman-rmse-coverage/run_experiment.R --probe \
  --results-dir experiments/58-guttman-rmse-coverage/results/probe

# render this report
quarto render experiments/58-guttman-rmse-coverage/report.qmd
```
```{r reproduce-note}
```
