---
title: "Chen 2020 WLSMV_PD Replication Probe"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
rates_path <- file.path("results", "rejection_rates.csv")
p_path <- file.path("results", "p_values.csv")
if (!file.exists(rates_path) || !file.exists(p_path)) {
  stop("Run `Rscript run_experiment.R --smoke` first.", call. = FALSE)
}
rates <- read.csv(rates_path)
p_values <- read.csv(p_path)

fmt_rate <- function(x) ifelse(is.finite(x), sprintf("%.3f", x), "")
main_method <- "lavaan_wlsmv_pairwise_satorra2000"
focus <- rates[rates$method == main_method, , drop = FALSE]
focus <- focus[order(focus$n_total, focus$thresholds, focus$missing_rate), ]
if (!nrow(focus)) {
  stop("No Satorra 2000 lavaan WLSMV results in results/rejection_rates.csv.",
       call. = FALSE)
}
target <- focus[order(abs(focus$n_total - 1000L),
                      focus$thresholds != "symmetric",
                      abs(focus$missing_rate - .50)), ][1L, ]
short_rate <- fmt_rate(target$rejection_rate)
short_chen <- fmt_rate(target$chen_type1)
short_df <- if (is.finite(target$df_diff_min) &&
                target$df_diff_min == target$df_diff_max) {
  sprintf("%.0f", target$df_diff_min)
} else {
  sprintf("%.0f-%.0f", target$df_diff_min, target$df_diff_max)
}

pretty_method <- function(x) {
  out <- x
  out[out == "lavaan_wlsmv_pairwise_satorra2000"] <- "lavaan WLSMV pairwise, Satorra 2000"
  out[out == "lavaan_wlsmv_pairwise_naive_diff"] <- "lavaan WLSMV pairwise, naive diff"
  out
}

tab <- rates[, c("thresholds", "n_total", "missing_rate", "method", "usable",
                 "rejection_rate", "mc_se", "median_p", "mean_df_diff",
                 "chen_type1")]
tab <- tab[order(tab$n_total, tab$thresholds, tab$missing_rate, tab$method), ]
tab$thresholds <- tools::toTitleCase(tab$thresholds)
tab$method <- pretty_method(tab$method)
tab$missing_rate <- sprintf("%.0f%%", 100 * tab$missing_rate)
tab$rejection_rate <- fmt_rate(tab$rejection_rate)
tab$mc_se <- fmt_rate(tab$mc_se)
tab$median_p <- fmt_rate(tab$median_p)
tab$mean_df_diff <- sprintf("%.1f", tab$mean_df_diff)
tab$chen_type1 <- fmt_rate(tab$chen_type1)
names(tab) <- c("Thresholds", "Total N", "Missing", "Method", "Usable",
                "Type-I", "MC SE", "Median p", "Mean df diff",
                "Chen Table 2")
```

## Question

Can we cheaply reproduce the poor WLSMV_PD Type-I behavior reported by
Chen et al. (2020) for the invariant one-factor ordinal model with missing data?
The target cell is the symmetric-threshold, total N = 1000, 50% missing condition,
where Table 2 reports a WLSMV_PD rejection rate of 0.550.

## Short Answer

This lavaan replication does not reproduce the published bad cell. In the closest
current lavaan WLSMV pairwise run here, with the auxiliary variable included as a
saturated observed covariate, the Type-I rate is `r short_rate` versus Chen's
`r short_chen`, and the lavaan nested test uses `r short_df` df rather than the
38 df reported for Mplus WLSMV_PD. The naive chi-square difference diagnostic is
inflated, but that is not the robust WLSMV difference test used for the main
comparison.

## Evidence

```{r}
knitr::kable(tab, row.names = FALSE)
```

The simulated p-values are in `results/p_values.csv`; the table above is the
condition-level summary in `results/rejection_rates.csv`.

## Caveats

This is a targeted probe, not a full reproduction of the paper. It uses current
lavaan WLSMV with `missing = "pairwise"` and `lavTestLRT(..., method =
"satorra.2000")`, not Mplus 8.0 DIFFTEST. The auxiliary variable is included as
saturated observed covariances, which is a lavaan approximation to the paper's
Mplus auxiliary-variable setup. The missingness generator is an exact weighted
rank sample on group B items 8-10, matching the rank-MAR direction and target
rate but not necessarily Mplus's exact random-until-target implementation. Small
smoke runs are only syntax checks; use at least a few hundred replications before
treating a rate as stable.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 500 --n-total 1000 --missing-rate .50 --thresholds symmetric
quarto render report.qmd
```
