---
title: "Ordinal Polychoric Omega Coverage Probe"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- "results/simulation_summary.csv"
metadata_path <- "results/metadata.csv"
if (!file.exists(summary_path) || !file.exists(metadata_path)) {
  stop("Missing results. Run: Rscript run_experiment.R --smoke")
}
summary <- read.csv(summary_path)
metadata <- read.csv(metadata_path)
worst <- summary[which.min(summary$coverage_95), ]
```

## Question

Does the ordinal polychoric omega delta-method interval cover its population
latent-response target in the small-N cells we would use before profile-LR work?

## Short Answer

```{r, results='asis'}
cat(sprintf(
  "At %s reps per cell and target omega %.3f, coverage among successful draws ranges from %.3f to %.3f. The weakest cell is %s at N = %d. The grid recorded %d construction failures.",
  metadata$reps, metadata$target, min(summary$coverage_95),
  max(summary$coverage_95), worst$regime, worst$n, sum(summary$failures)))
```

## Evidence

```{r}
shown <- summary[, c(
  "regime", "n", "failures", "mean_value", "bias", "sd_value", "mean_se",
  "coverage_95", "mean_width_95", "min_variable_category_count_p05"
)]
names(shown) <- c(
  "Regime", "N", "Failures", "Mean Omega", "Bias", "SD", "Mean SE",
  "Coverage", "Mean Width", "P05 Min Variable Category Count"
)
shown[] <- lapply(shown, function(x) if (is.numeric(x)) round(x, 4) else x)
knitr::kable(shown)
```

## Caveats

This is a correctly specified ordinal probit simulation with one factor and
three-category indicators. It only evaluates the ordinary Wald delta interval
for the polychoric correlation-matrix omega; it does not test observed-score
Green-Yang omega, nonnormal latent responses, profile-LR inversion, or Bartlett
calibration.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R
quarto render report.qmd
```
