---
title: "Ordinal LS / SNLLS probe"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "fits.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

fits_path <- file.path(results_dir, "fits.csv")
construction_path <- file.path(results_dir, "construction.csv")
attempts_path <- file.path(results_dir, "attempts.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(fits_path)) {
  stop("Missing fits.csv. Run: Rscript run_experiment.R")
}

fits <- utils::read.csv(fits_path, check.names = FALSE,
                        stringsAsFactors = FALSE)
construction <- utils::read.csv(construction_path, check.names = FALSE,
                                stringsAsFactors = FALSE)
attempts <- utils::read.csv(attempts_path, check.names = FALSE,
                            stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.2f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.4g", x))
fmt_diff <- function(x) ifelse(is.na(x), "", sprintf("%.2e", x))
meta_value <- function(key) metadata$value[match(key, metadata$key)]
```

This experiment asks a narrow question: how much of the current ordinal
least-squares cost is moment construction, how much is the fit, and what
happens if we try to route ordinal data through ULS or SNLLS today?

The important caveat is that **native ordinal SNLLS is not implemented in this
codebase yet**. The SNLLS condition below is a proxy: it runs continuous ULS
SNLLS on the polychoric correlation matrix and ignores thresholds. The ULS
condition is also a proxy: identity weights are injected through the current
DWLS weight slot. DWLS and WLS are the native ordinal estimators.

## Run

```{r}
data.frame(
  item = metadata$key,
  value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Cases

The cases are checked-in repo fixtures: two synthetic ordinal CFA fixtures and
the BFI ordinal parity fixture.

```{r}
case_table <- construction[, c("case_label", "source", "n_obs", "n_vars",
                               "levels", "n_moments",
                               "n_threshold_moments", "n_assoc_moments",
                               "construct_ms_median")]
names(case_table) <- c("Case", "Source", "n", "Observed vars", "Levels",
                       "Moment dim", "Thresholds", "Associations",
                       "Construction ms")
case_table$`Construction ms` <- fmt_ms(case_table$`Construction ms`)
knitr::kable(case_table, row.names = FALSE)
```

Construction here is the full current ordinal data build:
`data_ordinal_stats_from_df()` estimates thresholds and polychorics, builds
`NACOV`, and materializes both `W_dwls` and `W_wls`. This experiment therefore
separates construction from fitting, but it does not isolate DWLS-only weight
construction from WLS-only weight construction.

## Fit Timings

```{r}
fit_table <- fits[, c("case_label", "condition", "status", "fit_ms_median",
                      "whole_ms_median", "iterations", "optimizer_status",
                      "grad_norm", "fmin", "max_abs_theta_diff")]
names(fit_table) <- c("Case", "Condition", "Status", "Fit ms",
                      "Whole ms", "Iterations", "Optimizer", "Grad norm",
                      "fmin", "Max |theta diff|")
fit_table$`Fit ms` <- fmt_ms(fit_table$`Fit ms`)
fit_table$`Whole ms` <- fmt_ms(fit_table$`Whole ms`)
fit_table$`Grad norm` <- fmt_num(fit_table$`Grad norm`)
fit_table$fmin <- fmt_num(fit_table$fmin)
fit_table$`Max |theta diff|` <- fmt_diff(fit_table$`Max |theta diff|`)
knitr::kable(fit_table, row.names = FALSE)
```

```{r}
#| fig-width: 8
#| fig-height: 4
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ok <- fits[fits$status == "ok", ]
  ok$condition <- factor(
    ok$condition,
    levels = c("DWLS", "WLS", "ULS_identity_proxy",
               "ULS_SNLLS_polychoric_proxy")
  )
  ggplot(ok, aes(condition, fit_ms_median, fill = condition)) +
    geom_col(width = 0.62, show.legend = FALSE) +
    facet_wrap(~ case_label, scales = "free_y") +
    labs(x = NULL, y = "Median fit-only milliseconds",
         title = "Fit-only cost after ordinal moments are already built") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

## Unsupported Attempts

```{r}
attempt_table <- attempts[, c("case_label", "condition", "status", "message")]
names(attempt_table) <- c("Case", "Attempt", "Status", "Message")
knitr::kable(attempt_table, row.names = FALSE)
```

These failures are expected with the current surface. `magmaan(...,
estimator = "ULS", ordered = ...)` does not dispatch to an ordinal ULS
objective, and `fit_uls_snlls()` accepts continuous sample statistics rather
than ordinal threshold-plus-polychoric moment objects.

## Reading

The current ordinal path is a bounded least-squares problem over thresholds
and polychoric correlations. DWLS and WLS differ only in the weight selected
from the already-built ordinal statistics. The identity-weight ULS proxy uses
the same residual stack as DWLS/WLS, so it is a useful fit-loop comparison, but
it is not a public estimator contract.

The polychoric SNLLS proxy is even narrower: it answers "what if we fit the
polychoric correlation matrix as if it were continuous covariance input?"
That can be a useful performance smoke test for the separable continuous path,
but it is not ordinal SNLLS because it drops threshold residuals and does not
profile the ordinal parameter blocks.

## What Must Change

Native ordinal ULS would need one small API/enum extension: add an identity
weight option to the ordinal weight kind, expose `fit_uls_ordinal()` on the
R side, and document chi-square semantics for the identity-weight
threshold-plus-polychoric objective.

Native ordinal SNLLS is a larger design item. The implementation needs an
ordinal residual/Jacobian path that includes thresholds, classifies threshold
rows outside the current matrix-cell parameter-location scheme, proves which
latent-response covariance pieces remain conditionally linear, and exposes a
profiled bounded ordinal fit entry guarded against the existing DWLS/WLS
fixtures.

## Reproduce

```sh
Rscript experiments/06-ordinal-snlls-probe/run_experiment.R
(cd experiments/06-ordinal-snlls-probe && quarto render report.qmd)
```
