---
title: "Ordinal workspace and SNLLS probe"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "workspace.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

workspace_path <- file.path(results_dir, "workspace.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(workspace_path)) {
  stop("Missing workspace.csv. Run: Rscript run_experiment.R --smoke")
}

workspace <- utils::read.csv(workspace_path, check.names = FALSE,
                             stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)

meta_value <- function(key) metadata$value[match(key, metadata$key)]
fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.2f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.4g", x))
fmt_diff <- function(x) ifelse(is.na(x), "", sprintf("%.2e", x))
max_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else max(x)
}
median_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else stats::median(x)
}

workspace$Path <- factor(
  workspace$path,
  levels = c("materialized_bounded", "cache_bounded", "cache_snlls"),
  labels = c("Materialized bounded", "Cache bounded", "Cache SNLLS")
)
workspace$Thresholds <- factor(
  workspace$threshold_mode,
  levels = c("free", "fixed_or_shared", "linear_map"),
  labels = c("Free", "Fixed", "Shared")
)
ok <- workspace[workspace$fit_status == "ok", ]
```

## Question

Now that all-ordinal delta fitting has a cache-aware path and a native
Golub-Pereyra SNLLS path, this experiment asks what those paths buy us:

- Do cache-aware bounded fits reproduce the existing materialized DWLS/WLS
  results?
- Does ordinal SNLLS land on the same objective as bounded fitting?
- Which Gamma/weight pieces are actually materialized for ULS, DWLS, and WLS?

## Short Answer

The smoke design is a correctness-and-shape probe, not a final performance
claim. In the generated cells, cache-aware bounded fits agree with the
materialized bounded path for DWLS/WLS, and cache-aware SNLLS agrees with
cache-aware bounded fits to numerical tolerance.

The materialization flags show the intended workspace split: ULS carries no
Gamma cache, fit-only DWLS keeps only the diagonal Gamma, and WLS uses the full
Gamma plus the WLS weight. That is the main engineering result this experiment
is meant to keep visible.

## Run

```{r}
data.frame(
  Item = metadata$key,
  Value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Status

```{r}
status <- aggregate(
  seed ~ Thresholds + estimator + Path + fit_status,
  data = workspace,
  FUN = length
)
names(status) <- c("Thresholds", "Estimator", "Path", "Status", "Rows")
knitr::kable(status, row.names = FALSE)
```

## Equivalence

```{r}
equiv <- aggregate(
  cbind(theta_diff_bounded, fmin_diff_bounded,
        theta_diff_legacy, fmin_diff_legacy) ~ Thresholds + estimator + Path,
  data = ok,
  FUN = max_or_na,
  na.action = na.pass
)
names(equiv) <- c("Thresholds", "Estimator", "Path",
                  "Max |theta - bounded|", "Max |fmin - bounded|",
                  "Max |theta - materialized|", "Max |fmin - materialized|")
for (j in 4:ncol(equiv)) equiv[[j]] <- fmt_diff(equiv[[j]])
knitr::kable(equiv, row.names = FALSE)
```

`Materialized bounded` is available only for DWLS and WLS because the old
ordinal API has no identity-weight ULS estimator. `Cache SNLLS` is compared to
the cache-aware bounded fit for every estimator and, where available, to the
materialized bounded fit.

## Cost Shape

```{r}
cost <- aggregate(
  cbind(fit_median_ms, cache_setup_ms, weight_setup_ms,
        stats_ms, starts_ms) ~ estimator + Path,
  data = ok,
  FUN = median_or_na,
  na.action = na.pass
)
names(cost) <- c("Estimator", "Path", "Fit ms", "Cache setup ms",
                 "Weight setup ms", "Stats ms", "Starts ms")
for (j in 3:ncol(cost)) cost[[j]] <- fmt_ms(cost[[j]])
knitr::kable(cost, row.names = FALSE)
```

```{r}
#| fig-width: 8
#| fig-height: 4
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ggplot(ok, aes(Path, fit_median_ms, fill = estimator)) +
    geom_col(position = "dodge", width = 0.7) +
    facet_grid(Thresholds ~ p, labeller = label_both) +
    labs(x = NULL, y = "Median fit milliseconds") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

The `stats_ms` column is intentionally separate from fit timing. It is the
current full ordinal statistic construction from integer data, including NACOV
and legacy DWLS/WLS weights. The cache setup columns measure only the adapter
work after those statistics already exist.

## Workspace Flags

```{r}
flags <- unique(ok[, c("estimator", "Path", "cache_blocks",
                      "cache_has_diagonal", "cache_has_full",
                      "cache_has_dwls_weight", "cache_has_wls_weight")])
names(flags) <- c("Estimator", "Path", "Cache blocks", "Diagonal Gamma",
                  "Full Gamma", "DWLS weight", "WLS weight")
knitr::kable(flags[order(flags$Estimator, flags$Path), ], row.names = FALSE)
```

For fit-only DWLS, the cache-aware path does not materialize a DWLS weight
matrix; it uses the diagonal Gamma directly. WLS necessarily materializes the
full inverse weight. ULS uses identity weights and therefore has no cache.

## Caveats

This is an advisory benchmark harness. It uses synthetic single-group
all-ordinal CFA cells, the delta parameterization, and the NLopt L-BFGS
backend. Mixed ordinal/continuous data, theta parameterization, robust
fit-plus-inference reuse, and R-facing convenience wrappers are separate
follow-up experiments.

The runner still begins by constructing legacy `OrdinalStats` so we can compare
against the existing path in one executable. A future raw-data lazy-workspace
experiment should split threshold/polychoric estimation, diagonal Gamma, full
Gamma, and weight inversion at their true construction boundaries.

## Reproduce

```sh
Rscript experiments/06-ordinal-snlls-probe/run_experiment.R --smoke --preset fast
(cd experiments/06-ordinal-snlls-probe && quarto render report.qmd)
```

For less noisy timings, use `--preset opt` and increase `--reps`.
