---
title: "Which information matrix should FIML use?"
subtitle: "A focused computational validation, not a new general simulation result"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("overnight", "pilot", "smoke")))
available <- roots[vapply(roots, function(path)
  file.exists(file.path(path, "metadata.csv")), logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke --lavaan-parity`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
meta_frame <- read_result("metadata.csv")
meta <- setNames(meta_frame$value, meta_frame$key)
cells <- read_result("cell_method_summary.csv")
parity <- read_result("lavaan_parity.csv")
projection <- read_result("runtime_projection.csv")
profile <- meta[["profile"]]
reps <- as.integer(meta[["reps_per_cell"]])
is_overnight <- identical(profile, "overnight") &&
  as.integer(meta[["cells"]]) == 48L && reps >= 200000L

correct <- subset(cells, specification == "correct")
ranked <- aggregate(
  cbind(
    coverage = correct$coverage_true,
    coverage_error = abs(correct$coverage_true - 0.95)
  ),
  list(method = correct$method), mean, na.rm = TRUE)
ranked <- ranked[order(ranked$coverage_error), , drop = FALSE]

cells$coverage_primary <- ifelse(
  cells$specification == "correct",
  cells$coverage_true,
  cells$coverage_centered)
cells$condition <- paste(cells$specification, cells$distribution, sep = "_")
coverage_summary <- aggregate(
  coverage_primary ~ method + condition,
  data = cells, mean, na.rm = TRUE)
coverage_wide <- reshape(
  coverage_summary, idvar = "method", timevar = "condition",
  direction = "wide")
names(coverage_wide) <- sub(
  "coverage_primary.", "", names(coverage_wide), fixed = TRUE)
method_labels <- c(
  expected_model = "Expected, model-based",
  expected_sandwich = "Expected + sandwich",
  observed_h1_model = "Observed H1, model-based",
  observed_h1_sandwich = "Observed H1 + sandwich",
  observed_hessian_model = "Full observed Hessian, model-based",
  observed_hessian_sandwich = "Full observed Hessian + sandwich")
coverage_wide <- coverage_wide[
  match(names(method_labels), coverage_wide$method), , drop = FALSE]
coverage_wide$method <- unname(method_labels[coverage_wide$method])
coverage_wide[-1L] <- 100 * coverage_wide[-1L]

failure_rows <- unique(cells[
  c("cell_id", "information", "reps_fit", "reps_method")])
failure_rows$inversion_failure_rate <- with(
  failure_rows,
  ifelse(reps_fit > 0, (reps_fit - reps_method) / reps_fit, NA_real_))
failure_display <- aggregate(
  inversion_failure_rate ~ information,
  data = failure_rows, FUN = max)
failure_display$inversion_failure_rate <-
  100 * failure_display$inversion_failure_rate
```

## Question

For continuous FIML, should standard errors use the structured expected Fisher
information, the analytic observed-H1 approximation, or the full observed
Hessian? Does the answer change when the covariance is model based versus an
empirical-score sandwich?

This is a focused computational validation and a small extension, not an
attempt to rediscover the main result of Savalei (2010). That paper already
showed that expected information is inconsistent under general MAR and
recommended observed information. The local questions are whether the three
computational conventions agree with an established implementation,
whether observed-H1 is a stable analytic substitute for the full Hessian, and
whether the same-meat sandwich behaves as expected under nonnormality and
structural misspecification.

## Short answer

`r if (is_overnight) {
  sprintf(
    "Across the completed 48-cell, %d-replication design, %s had the smallest mean absolute 95%%-coverage error in the correctly specified cells (%.3f percentage points), with mean coverage of %.3f%%. These are local computational results, interpreted alongside—not instead of—the existing theory and simulations.",
    reps, unname(method_labels[ranked$method[[1L]]]),
    100 * ranked$coverage_error[[1L]],
    100 * ranked$coverage[[1L]])
} else {
  sprintf(
    "The %s run (%d replications per cell) validates the pipeline and estimates runtime, but it is not large enough to choose a default. The overnight result is still required before using finite-sample differences as evidence.",
    profile, reps)
}`

`r if (nrow(parity)) {
  sprintf(
    "The optional lavaan sentinel passed %d/%d parameter-level SE comparisons; max |difference| = %.4g.",
    sum(parity$pass), nrow(parity), max(parity$abs_difference))
} else {
  "The lavaan sentinel was not requested for this run."
}`

## Evidence

The design has 48 cells:

- multivariate normal versus elliptical multivariate \(t_5\), both with the
  same population covariance;
- correct two-factor CFA versus an omitted 0.25 cross-loading;
- complete data, 30% MCAR, and 30% MAR driven by always-observed factor
  markers;
- \(N = 150, 300, 600, 1200\).

Two parameters are followed: a free loading and the factor covariance. Every
replication is fit once. The three information matrices are then combined with
either their inverse (model covariance) or the identical empirical
outer-product-of-scores meat (sandwich covariance). Thus the three sandwich
columns differ only in their bread.

```{r coverage-table}
knitr::kable(
  coverage_wide[
    c(
      "method", "correct_normal", "correct_t5",
      "omitted_crossloading_normal", "omitted_crossloading_t5")
  ],
  digits = 2, row.names = FALSE,
  col.names = c(
    "Method", "Correct, normal", "Correct, t5",
    "Misspecified, normal", "Misspecified, t5"),
  caption = paste(
    "Mean 95% coverage across sample sizes, missingness mechanisms,",
    "and focal parameters. Misspecified cells use centered coverage.")
)
```

For misspecified cells, the population pseudo-true parameter is not assumed
known. Coverage is therefore centered on the leave-one-out cell mean, which
tests whether an SE estimates sampling variation around the pseudo-limit
without conflating that question with pseudo-target calibration.

Rare converged but inadmissible (Heywood) or nearly singular solutions create
extreme estimate and standard-error tails. They are retained in coverage when
the standard error is finite, but unstable arithmetic mean SE/SD ratios are
omitted. Admissibility-conditioned results require a prospectively recorded
fit-level flag rather than an arbitrary post-hoc cutoff.

```{r failure-table}
knitr::kable(
  failure_display, digits = 4, row.names = FALSE,
  col.names = c("Information", "Worst cell inversion-failure rate (%)"),
  caption = paste(
    "Information-matrix stability among converged fits;",
    "failures are retained as outcomes.")
)
```

The measured worker time was
`r sprintf("%.3f", projection$mean_worker_seconds_per_rep)` seconds per
replication. On the configured
`r projection$assumed_cores` cores, the 48-by-200,000 profile projects to
`r sprintf("%.2f", projection$projected_hours_with_15pct_overhead)` hours,
including a 15% scheduling and I/O allowance.

## Sanity checks

The optional sentinel regenerates a spread of simulation cells and compares
all six parameter SEs against explicit settings in an established
implementation:

| Experiment label | Information | Observed-information form | SE |
|---|---|---|---|
| expected model | `expected` | `hessian` | `standard` |
| expected sandwich | `expected` | `hessian` | `robust.huber.white` |
| observed-H1 model | `observed` | `h1` | `standard` |
| observed-H1 sandwich | `observed` | `h1` | `robust.huber.white` |
| observed-Hessian model | `observed` | `hessian` | `standard` |
| observed-Hessian sandwich | `observed` | `hessian` | `robust.huber.white` |

## Caveats

The elliptical \(t_5\) condition stresses fourth moments but not skewness.
The MAR mechanism is deliberately legible rather than exhaustive. There are
only six indicators, one kind of structural misspecification, and two focal
parameters. Accordingly, this run can validate the computational choice and
expose finite-sample stability problems; it cannot supersede the broader
literature.

Expected-information standard errors are theoretically restricted to complete
data or MCAR under the relevant model assumptions. The observed-H1 versus full
observed-Hessian comparison is the genuinely useful finite-sample part of this
run. Under misspecification, a model-based inverse is not made robust merely by
choosing an observed bread; the empirical sandwich is the relevant comparison.

Ordinal/polychoric estimation is outside this experiment. Its sparse-table,
threshold, correlation, and estimated-weight issues require a separate design.

## Reproduce

```bash
# Fast structural check plus the independent lavaan oracle
Rscript run_experiment.R --smoke --lavaan-parity --cores 4

# Timing and failure-rate pilot
Rscript run_experiment.R --pilot --lavaan-parity --cores 8

# Checkpointed machine-sized run (9.6 million fits)
Rscript run_experiment.R --overnight --lavaan-parity --cores 8
```

The runner uses deterministic replication seeds, atomic per-chunk checkpoints,
strict manifest/code-hash validation, progress/ETA messages, and automatic
resume. Use a new `--results-dir` after changing code or design settings.

## References

Savalei, V. (2010). Expected versus observed information in SEM with incomplete
normal and nonnormal data. *Psychological Methods, 15*(4), 352--367.
[doi:10.1037/a0020143](https://doi.org/10.1037/a0020143).

Savalei, V., & Rosseel, Y. (2022). Computational options for standard errors
and test statistics with incomplete normal and nonnormal data in SEM.
*Structural Equation Modeling, 29*(2), 163--181.
[doi:10.1080/10705511.2021.1877548](https://doi.org/10.1080/10705511.2021.1877548).

Yuan, K.-H., & Bentler, P. M. (2000). Three likelihood-based methods for mean
and covariance structure analysis with nonnormal missing data.
*Sociological Methodology, 30*, 165--200.
[doi:10.1111/0081-1750.00078](https://doi.org/10.1111/0081-1750.00078).
