---
title: "Near-Singular ML Continuation"
format: html
---

## What Was Fit

For a sample covariance matrix `S`, the continuation path fits a sequence of
ordinary ML problems to blended covariance matrices,

```text
S_lambda = (1 - lambda) S + lambda T(S),
```

where `T(S)` is the chosen shrinkage target. The wrapper calls the tuning
parameter `alpha`, but it is the same object as `lambda` here.

The default sequence is:

```text
0.50 -> 0.20 -> 0.10 -> 0.05 -> 0.01 -> 0.00
```

The final `0.00` stage is appended automatically when `include_endpoint =
TRUE`. That final stage is the original ML problem. Each stage is a full ML fit,
and its parameter vector is used as the start value for the next stage, so this
is a hot-started chain of multiple fits.

The comparison grid now tries three targets:

```text
diagonal         T(S) = diag(S)
scaled_identity  T(S) = mean(diag(S)) I
identity         T(S) = I
```

and four fixed lambda profiles:

```text
legacy    0.50, 0.20, 0.10, 0.05, 0.01, 0
light     0.10, 0.05, 0.02, 0.01, 0.005, 0
endpoint  0.20, 0.10, 0.05, 0.025, 0.0125, 0.006, 0.003, 0.001, 0
dense     0.50, 0.35, 0.25, 0.18, 0.13, 0.09, 0.06, 0.04, 0.025, 0.015, 0.008, 0
```

## How To Read The Tables

`baseline_ml` is one direct ML fit on the original sample covariance.
`ridge_continuation` is the hot-started sequence above.

`attempts` is the number of attempted fits. `ok` means the wrapper returned a
fit object instead of an error. `converged` means the terminal audit accepted
the final point as stationary, using magmaan's lavaan-style default
`||projected gradient||_inf <= 0.001`. `optimizer_converged` is narrower: it
means the backend itself reported a clean stop. A line-search salvage can fail
that optimizer flag while still passing the audit. `line_search_failures`
counts trapped hard L-BFGS line-search errors. `line_search_salvaged` counts
non-clean backend stops whose returned point passed the audit.
`median_total_f_evals` is the median number of objective evaluations, counting
all stages for continuation.

The paired table is often the most interpretable one. `baseline_only` means the
direct fit converged and continuation did not. `continuation_only` means
continuation rescued a replicate that direct ML did not converge on. `neither`
means both failed.

## Toy Smoke Run

This is the artificial near-singular one-factor generator from
`run_experiment.R`. It is useful as a quick smoke test, not as the main evidence.

```{r}
fits_path <- file.path("results", "fits.csv")
summary_path <- file.path("results", "summary.csv")
if (!file.exists(fits_path) || !file.exists(summary_path)) {
  stop("Run `Rscript run_experiment.R` first.")
}
fits <- read.csv(fits_path)
summary <- read.csv(summary_path)
summary
```

```{r}
fits[order(fits$replicate, fits$method),
     c("replicate", "method", "sample_min_eigen", "sample_condition",
       "ok", "converged", "optimizer_converged", "optimizer_status",
       "fmin", "audit_grad_inf_norm", "audit_stationarity_rhs",
       "total_f_evals", "elapsed_sec", "error")]
```

## Convergence-Note Designs

This section uses the same convergence-simulation factories and seed rule as
`papers/convergence-note/pilot/scripts/run_convergence_sims.R`, with
`optimizer = "nlopt-lbfgs"` and control
`max_iter = 5000, ftol = 1e-10, gtol = 1e-7`.

```{r}
cn_summary_path <- file.path("results", "convergence_note_summary.csv")
cn_paired_path <- file.path("results", "convergence_note_paired.csv")
if (!file.exists(cn_summary_path) || !file.exists(cn_paired_path)) {
  cat("Run `Rscript run_convergence_note_comparison.R` to populate this section.")
} else {
  cn_summary <- read.csv(cn_summary_path)
  cn_paired <- read.csv(cn_paired_path)
  cn_summary[, c("design", "method", "lambda_profile", "target",
                 "converged_rate", "optimizer_converged_rate",
                 "line_search_failures", "line_search_salvaged",
                 "median_total_f_evals", "median_audit_grad_inf_norm")]
}
```

```{r}
if (exists("cn_paired")) {
  cn_paired[, c("design", "lambda_profile", "target", "both_converged",
                "baseline_only", "continuation_only", "neither",
                "median_fmin_diff_cont_minus_base")]
}
```

```{r}
if (exists("cn_summary")) {
  cont <- cn_summary[cn_summary$method == "ridge_continuation", ]
  top <- do.call(
    rbind,
    lapply(split(cont, cont$design), function(x) {
      x[order(-x$converged_rate, x$median_total_f_evals), ][1L, ]
    })
  )
  top[, c("design", "lambda_profile", "target", "converged_rate",
          "line_search_failures", "median_total_f_evals",
          "median_audit_grad_inf_norm")]
}
```

```{r, results='asis'}
if (exists("cn_summary") && exists("cn_paired")) {
  baseline <- cn_summary[cn_summary$method == "baseline_ml", ]
  continuation <- cn_summary[cn_summary$method == "ridge_continuation", ]
  best <- do.call(
    rbind,
    lapply(split(continuation, continuation$design), function(x) {
      x[order(-x$converged_rate, x$median_total_f_evals), ][1L, ]
    })
  )
  pct <- function(x) paste0(round(100 * x), "%")
  cat(
    "In this run, direct ML converged in ",
    pct(min(baseline$converged_rate)), " to ",
    pct(max(baseline$converged_rate)),
    " of replicates across designs. The best continuation profile per design ",
    "converged in ",
    pct(min(best$converged_rate)), " to ",
    pct(max(best$converged_rate)),
    " by the terminal-audit criterion. Across all continuation scenarios, ",
    "paired rescues (`continuation_only`) were ",
    pct(min(cn_paired$continuation_only)), " to ",
    pct(max(cn_paired$continuation_only)),
    ", while paired losses (`baseline_only`) were ",
    pct(min(cn_paired$baseline_only)), " to ",
    pct(max(cn_paired$baseline_only)),
    ".\n",
    sep = ""
  )
}
```

## Conclusion

This experiment does not support covariance continuation as a default
convergence aid for the current ML path. The idea is simple and attractive:
solve a slightly regularized covariance problem, then hot-start ordinary ML
from there. In these simulations, however, the extra stages mostly add more
places for L-BFGS to fail.

The most useful outcome is diagnostic. Using magmaan's terminal audit changes
the story relative to raw optimizer return codes: some fits that look like
clean optimizer stops are not stationary by the lavaan-style gradient check.
So future convergence studies should continue reporting audit convergence, not
only backend status.

Among the continuation variants, lighter lambda paths were generally better
than the original coarse path, and `scaled_identity` was the most consistent
target. Raw identity was sometimes competitive, but not enough to change the
overall conclusion. The best continuation variant still trailed direct ML on
the convergence-note designs and required more objective evaluations.

The practical decision is to close this branch of the experiment for now. Keep
the code as frontier research tooling, but do not promote it as a recommended
estimation strategy. More promising next directions are different starts,
bounded or screened multi-start portfolios, and cost-normalized optimizer
comparisons using the same terminal-audit criterion.
