---
title: "Improved goodness-of-fit tests (Foldnes, Moss & Grønneberg 2024): magmaan replication"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results <- "results"
need <- file.path(results, c("rejection_rates.csv", "metadata.csv"))
if (!all(file.exists(need))) {
  stop("Missing results. Run:\n",
       "  Rscript run_experiment.R --reps 100 --semtests-parity\n",
       "from the experiment directory first.")
}
rr   <- utils::read.csv(file.path(results, "rejection_rates.csv"))
meta <- utils::read.csv(file.path(results, "metadata.csv"))
par_file <- file.path(results, "semtests_parity.csv")
parity <- if (file.exists(par_file)) utils::read.csv(par_file) else NULL
reps <- meta$value[match("reps", meta$key)]
fmt_pct <- function(x) sprintf("%.1f", 100 * x)
```

## Question

The paper proposes penalized eigenvalue-block-averaging goodness-of-fit tests
(pEBA, pOLS) that are meant to hold the nominal Type I error rate under
non-normality where the classical ML chi-square fails. Two questions:

1. **Does magmaan compute these tests correctly** — do its p-values match the
   paper's own `semTests` implementation, fit for fit?
2. **Do the recommended tests behave** — under correct specification, does
   `pEBA4_RLS` / `pOLS_RLS` hold ~5% rejection where the ML chi-square inflates
   as the model grows and the data turn non-normal?

## Short answer

**Yes on both.** magmaan's FMG family, computed through `fmg_pvalues()`,
reproduces `semTests` p-value for p-value: across every test and cell the largest
disagreement is **~1e-5** (the biased-Gamma tests agree to ~5e-7; the
unbiased-Gamma variants to ~1e-5). And the qualitative Type I picture replicates:
the classical **ML chi-square over-rejects** — worse as `p` grows and under severe
non-normality — while **`pEBA4_RLS` and `pOLS_RLS` stay near 5%**.

```{r parity-headline, results='asis'}
if (!is.null(parity) && nrow(parity)) {
  comp <- parity[is.finite(parity$abs_diff), ]
  cat(sprintf("\n> Procedure parity: **max |Δp| = %.2g** over %d magmaan-vs-semTests comparisons.\n",
              max(comp$abs_diff), nrow(comp)))
}
```

## Evidence

### Type I error: ML inflates, penalized tests hold

Empirical rejection rates at α = 5% (correctly-specified model, so every rejection
is a Type I error). Averaged over sample sizes; one row per model size ×
distribution, a few representative tests. Nominal is 5.0.

```{r rr-table}
key <- c(std_ml = "ML chi-square", sb_ug_rls = "SB (UG, RLS)",
         peba4_rls = "pEBA4 (RLS)", pols2_rls = "pOLS (RLS)")
sub <- rr[rr$test %in% names(key), ]
agg <- aggregate(reject_05 ~ p + dist + test, data = sub, FUN = mean)
wide <- reshape(agg, idvar = c("p", "dist"), timevar = "test",
                direction = "wide")
names(wide) <- sub("reject_05\\.", "", names(wide))
ord <- c("std_ml", "sb_ug_rls", "peba4_rls", "pols2_rls")
ord <- ord[ord %in% names(wide)]
wide <- wide[order(wide$dist, wide$p), c("p", "dist", ord)]
disp <- wide
disp[ord] <- lapply(disp[ord], fmt_pct)
colnames(disp) <- c("p", "Distribution", unname(key[ord]))
knitr::kable(disp, align = "llrrrr",
             caption = "Rejection rate (%) at alpha = 5%. Full grid in results/rejection_rates.csv.")
```

The ML column drifts above 5% as `p` rises and under the severe (`*2`)
distributions; the penalized RLS columns track 5% across the grid. (Per-cell
detail, all tests and sample sizes, lives in `results/rejection_rates.csv`.)

### Procedure parity: magmaan vs semTests

```{r parity-table}
if (is.null(parity) || !nrow(parity)) {
  cat("Parity not run. Re-run with --semtests-parity to populate this section.\n")
} else {
  comp <- parity[is.finite(parity$abs_diff), ]
  by_test <- aggregate(abs_diff ~ test, data = comp, FUN = max)
  by_test <- by_test[order(-by_test$abs_diff), ]
  by_test$abs_diff <- signif(by_test$abs_diff, 2)
  colnames(by_test) <- c("Test", "Max |Δp| vs semTests")
  knitr::kable(by_test, align = "lr",
               caption = "Largest magmaan-vs-semTests p-value gap per test (rep 1 of each cell).")
}
```

The split is exactly as expected: tests on the biased Gamma estimator match to
near machine precision; the unbiased-Gamma (`_ug`) tests are looser at ~1e-5,
since magmaan and semTests apply the bias correction at different stages. Both
are far tighter than any Monte Carlo rejection-rate cell can resolve.

## Caveats

- The design now matches the paper: its **exact** two-factor populations, the
  full 84-condition grid (`p ∈ {10, 20, 40}` × `N ∈ {400, 800, 1500, 3000}` ×
  seven distributions), and its full battery of 42 test statistics. The one
  omission is the **Bollen-Stine bootstrap** (the paper's 43rd test), which
  magmaan does not implement. The only remaining knob is replicate count:
  `r reps` here versus the paper's 3000.
- At `r reps` replicates the rejection rates carry visible Monte Carlo noise; the
  table establishes the *direction and ordering* of the effect, not precise rates.
  The procedure-parity finding, by contrast, is exact and replicate-independent.
- Non-normal data uses magmaan's own simulators for the IG and PL families and a
  self-contained Fleishman power method for VM; the paper generates IG/PL with
  covsim and VM with lavaan. The target distributions (covariance plus marginal
  skewness and kurtosis) are identical, but individual datasets are not.

## Reproduce

```sh
# full 84-condition grid (p {10,20,40} x N {400,800,1500,3000} x 7 dists), parity
Rscript run_experiment.R --reps 100 --semtests-parity

# the large/heavy cells are p=40 at N=3000; scope with --cells when iterating
Rscript run_experiment.R --reps 1000 --cells p=10,N=400
Rscript run_experiment.R --reps 1000 --cells dist=pl2

Rscript run_experiment.R --help

# paper-scale rejection rates: parallel over cells (forks the 84 cells, chunked
# so memory stays light). ~30-40 min on 10 cores; writes rejection_rates.csv.
Rscript scripts/run_parallel.R --reps 3000 --cores 10
```
