---
title: "Improved goodness-of-fit tests (Foldnes, Moss & Grønneberg 2024): magmaan replication"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results <- "results"
need <- file.path(results, c("rejection_rates.csv", "metadata.csv"))
if (!all(file.exists(need))) {
  stop("Missing results. Run:\n",
       "  Rscript run_experiment.R --reps 100 --semtests-parity\n",
       "from the experiment directory first.")
}
rr   <- utils::read.csv(file.path(results, "rejection_rates.csv"))
meta <- utils::read.csv(file.path(results, "metadata.csv"))
par_file <- file.path(results, "semtests_parity.csv")
parity <- if (file.exists(par_file)) utils::read.csv(par_file) else NULL
reps <- meta$value[match("reps", meta$key)]
fmt_pct <- function(x) sprintf("%.1f", 100 * x)
```

## Question

The paper proposes penalized eigenvalue-block-averaging goodness-of-fit tests
(pEBA, pOLS) that are meant to hold the nominal Type I error rate under
non-normality where the classical ML chi-square fails. Two questions:

1. **Does magmaan compute these tests correctly** — do its p-values match the
   paper's own `semTests` implementation, fit for fit?
2. **Do the recommended tests behave** — under correct specification, does
   `pEBA4_RLS` / `pOLS_RLS` hold ~5% rejection where the ML chi-square inflates
   as the model grows and the data turn non-normal?

## Short answer

**Yes on both.** magmaan's FMG family, computed through `fmg_pvalues()`,
reproduces `semTests` p-value for p-value: the biased-Gamma tests — including the
recommended `pEBA4_RLS` and `pOLS_RLS` — agree to **~3e-6** across the whole grid.
The unbiased-Gamma (`_ug`) variants agree to ~1e-5 in small models and widen to
**~9e-3** only at `p = 40, N = 400`, where the two packages stage the bias
correction differently and Γ is near-singular — still far below what 3000
replicates resolve. And the qualitative Type I picture replicates: the classical
**ML chi-square over-rejects** — worse as `p` grows and under severe
non-normality — while **`pEBA4_RLS` and `pOLS_RLS` stay near 5%**.

```{r parity-headline, results='asis'}
if (!is.null(parity) && nrow(parity)) {
  comp <- parity[is.finite(parity$abs_diff), ]
  biased <- comp[!grepl("_ug_", comp$test), ]
  cat(sprintf(paste0("\n> Procedure parity: biased-Γ tests match `semTests` to ",
              "**max |Δp| = %.1g**; the unbiased-Γ (`_ug`) variants to **%.1g** ",
              "(worst at p=40/N=400). %d comparisons total.\n"),
              max(biased$abs_diff), max(comp$abs_diff), nrow(comp)))
}
```

## Evidence

### Type I error: ML inflates, penalized tests hold

Empirical rejection rates at α = 5% from `r reps` replicates (correctly-specified
model, so every rejection is a Type I error). Averaged over sample sizes; one row
per model size × distribution, a few representative tests. Nominal is 5.0.

```{r rr-table}
key <- c(std_ml = "ML chi-square", sb_ug_rls = "SB (UG, RLS)",
         peba4_rls = "pEBA4 (RLS)", pols2_rls = "pOLS (RLS)")
sub <- rr[rr$test %in% names(key), ]
agg <- aggregate(reject_05 ~ p + dist + test, data = sub, FUN = mean)
wide <- reshape(agg, idvar = c("p", "dist"), timevar = "test",
                direction = "wide")
names(wide) <- sub("reject_05\\.", "", names(wide))
ord <- c("std_ml", "sb_ug_rls", "peba4_rls", "pols2_rls")
ord <- ord[ord %in% names(wide)]
wide <- wide[order(wide$dist, wide$p), c("p", "dist", ord)]
disp <- wide
disp[ord] <- lapply(disp[ord], fmt_pct)
colnames(disp) <- c("p", "Distribution", unname(key[ord]))
knitr::kable(disp, align = "llrrrr",
             caption = "Rejection rate (%) at alpha = 5%. Full grid in results/rejection_rates.csv.")
```

The ML column drifts above 5% as `p` rises and under the severe (`*2`)
distributions; the penalized RLS columns track 5% across the grid. (Per-cell
detail, all tests and sample sizes, lives in `results/rejection_rates.csv`.)

### Procedure parity: magmaan vs semTests

```{r parity-table}
if (is.null(parity) || !nrow(parity)) {
  cat("Parity not run. Re-run with --semtests-parity to populate this section.\n")
} else {
  comp <- parity[is.finite(parity$abs_diff), ]
  by_test <- aggregate(abs_diff ~ test, data = comp, FUN = max)
  by_test <- by_test[order(-by_test$abs_diff), ]
  by_test$abs_diff <- signif(by_test$abs_diff, 2)
  colnames(by_test) <- c("Test", "Max |Δp| vs semTests")
  knitr::kable(by_test, align = "lr",
               caption = "Largest magmaan-vs-semTests p-value gap per test (rep 1 of each cell).")
}
```

The split is exactly as expected: tests on the biased Gamma estimator match to
near machine precision (~3e-6) across all model sizes. The unbiased-Gamma (`_ug`)
tests are looser — ~1e-5 at p=10/20, widening to ~9e-3 at p=40/N=400 — because
the two packages stage the bias correction differently and that gap amplifies
where N is small relative to the number of covariances and Γ is near-singular.
Even the widest gap is far tighter than any Monte Carlo rejection-rate cell can
resolve at 3000 replicates.

## Caveats

- This is the paper's full design at full scale: its **exact** two-factor
  populations, the complete 84-condition grid (`p ∈ {10, 20, 40}` ×
  `N ∈ {400, 800, 1500, 3000}` × seven distributions), the full battery of
  42 test statistics, and `r reps` replicates per cell. The sole omission is the
  **Bollen-Stine bootstrap** (the paper's 43rd test), which magmaan does not
  implement.
- Non-normal data uses magmaan's own simulators for the IG and PL families and a
  self-contained Fleishman power method for VM; the paper generates IG/PL with
  covsim and VM with lavaan. The target distributions (covariance plus marginal
  skewness and kurtosis) are identical, but individual datasets are not.

## Cost

The full `r reps`-replicate grid (84 cells × 42 tests) ran in **2 h 17 m** on a
13th-gen Intel Core i7-1355U (10 cores / 12 threads; 2 performance + 8
efficiency) with 10 parallel workers. Throughput at `p = 40` is
memory-bandwidth-bound and the sustained all-core load thermally throttles, so
wall time scales sublinearly with cores.

## Reproduce

```sh
# full 84-condition grid (p {10,20,40} x N {400,800,1500,3000} x 7 dists), parity
Rscript run_experiment.R --reps 100 --semtests-parity

# the large/heavy cells are p=40 at N=3000; scope with --cells when iterating
Rscript run_experiment.R --reps 1000 --cells p=10,N=400
Rscript run_experiment.R --reps 1000 --cells dist=pl2

Rscript run_experiment.R --help

# paper-scale rejection rates (these results): parallel over cells, forks the 84
# cells and chunks the replicate loop so memory stays light. Writes
# rejection_rates.csv. Pair with the parity command above for the parity table.
Rscript scripts/run_parallel.R --reps 3000 --cores 10
```
