---
title: "Corrected corpus: optimizer settings and recovery"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
r <- function(name) {
  p <- file.path("results","full",paste0(name,".csv"))
  if(!file.exists(p)) stop("Missing results: run run_experiment.R and R/summarize.R first")
  read.csv(p,stringsAsFactors=FALSE)
}
s <- r("summary"); d <- r("comparison"); jobs <- r("jobs"); exclusions <- r("exclusions")
paired <- r("paired"); domain <- r("domain"); meta <- r("metadata")
retries <- r("retry_jobs")
primary <- c("lbfgs_default","port_default","slsqp_default")
labels <- c(lbfgs_default="L-BFGS",port_default="PORT",slsqp_default="SLSQP")
ps <- subset(s,arm %in% primary)
count <- function(est,arm,col) s[s$estimator==est & s$arm==arm,col][1]
```

## Question

On the corrected textbook corpus, which ordinary optimizers recover a good
solution from our current starts? Do the historical Newsom GLS failures still
justify changing the evaluator?
Where the starts are at fault, does a structure-aware start fix it without
losing cases that already work? Which optimizer coordinates make PORT and
L-BFGS indifferent to the data's units?

## Short answer

```{r layered-setup}
lay <- r("layered/summary"); lp <- r("layered/paired"); eng <- r("engines_summary")
lc <- function(est,arm,start) lay$good_now[lay$estimator==est & lay$arm==arm & lay$start==start]
lr <- function(col) sum(lp[[col]])
```

```{r scaling-setup}
sc <- r("scaling_summary"); fiml <- r("fiml_scaling"); lane <- r("constrained_lane")
g <- function(mode,est,start,opt,scal,col="best_known") sc[[col]][sc$mode==mode & sc$estimator==est & sc$start==start & sc$optimizer==opt & sc$scaling==scal]
gc <- function(mode,est,start,opt,scal) g(mode,est,start,opt,scal,"certified")
lv <- function(mode,est,start,col) g(mode,est,start,"nlminb","lavaan",col)
np <- function(mode,est) sc$pairs[sc$mode==mode & sc$estimator==est & sc$engine=="magmaan"][1]
rng <- function(...) {x <- c(...); if (min(x)==max(x)) x[1] else paste0(min(x),"–",max(x))}
fg <- function(mode,opt,scal,col="accepted") sum(fiml[[col]][fiml$mode==mode & fiml$optimizer==opt & fiml$scaling==scal])
fn <- length(unique(fiml$case))
```

**How fits are scored.** A fit is solved when magmaan's verdict, the Newton
check, certifies its endpoint as a local minimum. Reaching the best known
objective is secondary; a certified endpoint above it is another local
minimum, since no start constructor can be required to find the best basin.
Every engine gets the same judge: lavaan's estimates are checked by the same
verdict, and lavaan's own convergence flag, which accepts saddles, is reported
only alongside. The `r np("native","ML")` standard pairs per estimator exclude
Mplus ex6.17, whose nonlinear equality constraints make it a different
optimization problem; SLSQP solves it (`constrained_lane.csv`).

**Unit-equivariant optimizer coordinates make PORT and L-BFGS indifferent to
the data's units; they are now the default.** Both optimizers search in each
parameter's sample unit, refined by the expected information at the start, with
means centered at their starts. From the layered start every standard ML pair
ends at a certified local minimum with both optimizers
(`r gc("native","ML","layered","nlopt-lbfgs","information")` and
`r gc("native","ML","layered","port","information")` of `r np("native","ML")`; raw
coordinates: `r gc("native","ML","layered","nlopt-lbfgs","none")` and
`r gc("native","ML","layered","port","none")`), and
`r g("native","ML","layered","port","information")` reach the best known objective.
GLS certifies `r gc("native","GLS","layered","port","information")`. With the current
default start, ML L-BFGS certifies `r gc("native","ML","current","nlopt-lbfgs","information")`
(raw: `r gc("native","ML","current","nlopt-lbfgs","none")`). After multiplying each
observed variable by a power of ten between 0.01 and 100, on the
`r np("rescaled","ML")` ML and `r np("rescaled","GLS")` GLS pairs whose optimum survives
the change, scaled runs from the layered start still certify
`r rng(gc("rescaled","ML","layered","nlopt-lbfgs","information"),gc("rescaled","ML","layered","port","information"))` and
`r rng(gc("rescaled","GLS","layered","nlopt-lbfgs","information"),gc("rescaled","GLS","layered","port","information"))`;
raw coordinates certify
`r rng(gc("rescaled","ML","layered","nlopt-lbfgs","none"),gc("rescaled","ML","layered","port","none"))` and
`r rng(gc("rescaled","GLS","layered","nlopt-lbfgs","none"),gc("rescaled","GLS","layered","port","none"))`, and lavaan's
defaults `r lv("rescaled","ML","default","certified")` and `r lv("rescaled","GLS","default","certified")`.
The old ML scaling gave every observed variable in a regression the block's
mean standard deviation; lavaan's start-based scale is not unit-equivariant.

**A layered moment start removes most start failures and loses few cases.**
It builds the start from the data in layers (measurement, latent covariance,
identification scale, structure) instead of cell by cell. With the same engine
and default controls, accepted fits matching the best observed objective rise
for ML from `r lc("ML","port_default","current")` to `r lc("ML","port_default","layered")`
of 304 with PORT and from `r lc("ML","lbfgs_default","current")` to
`r lc("ML","lbfgs_default","layered")` with L-BFGS. For GLS they rise from
`r lc("GLS","port_default","current")` to `r lc("GLS","port_default","layered")` with PORT
and from `r lc("GLS","lbfgs_default","current")` to `r lc("GLS","lbfgs_default","layered")`
with L-BFGS. Across the four estimator/optimizer pairs, `r lr("rescued")` case fits
are gained and `r lr("lost")` lost. Every loss starts from a lower objective than
the current start and fails in the optimizer: an early L-BFGS line-search abort,
or a PORT or L-BFGS stop short of the optimum on raw-unit data. Information
coordinates recover all of them. Most original
failures share one mechanism: a latent scaled by a fixed variance
(a std.lv or phantom factor) has a sign-reflection symmetry, and a start with all
its sign-odd paths at zero sits on it with exactly zero gradient in those paths.
No gradient-based optimizer leaves that subspace, lavaan included. This development comparison predates the layered start's promotion on
26 September 2026. It is now the ordinary ML/GLS default.

**Under the same judge, magmaan from the layered start certifies more fits
than lavaan's defaults and exactly as many as lavaan from the same start.** On the `r np("native","ML")` standard ML pairs, lavaan's
defaults end at a certified local minimum in `r lv("native","ML","default","certified")`
(`r lv("native","ML","default","best_known")` at the best known objective); lavaan's own
flag claims convergence at `r lv("native","ML","default","flag_only")` further endpoints
that the Newton check rejects. For GLS: `r lv("native","GLS","default","certified")`
(`r lv("native","GLS","default","best_known")`), with `r lv("native","GLS","default","flag_only")`
rejected claims. From the layered start lavaan certifies `r lv("native","ML","layered","certified")`
ML and `r lv("native","GLS","layered","certified")` GLS, and magmaan PORT in information
coordinates `r gc("native","ML","layered","port","information")` and
`r gc("native","GLS","layered","port","information")`. Both engines from the layered start
settle in the same three worse ML minima (Little's two Chapter 8 ALT models and
Kline's Worland step 2a), reached better only from the book's or lavaan's own
starts. After rescaling the units, lavaan's flag claims convergence at
`r lv("rescaled","ML","default","flag_only")` ML and `r lv("rescaled","GLS","default","flag_only")`
GLS endpoints the Newton check rejects.

**Settings alone do not repair L-BFGS's early domain failures.** PORT recovers
cases that stop almost immediately under L-BFGS, even after increasing its
budget or tightening tolerances. In ML, default L-BFGS returns an accepted fit
matching the best observed objective on `r count("ML","lbfgs_default","accepted_match")`
of `r count("ML","lbfgs_default","attempted")` completed attempts; PORT does so on
`r count("ML","port_default","accepted_match")` of
`r count("ML","port_default","attempted")`. For GLS, tighter L-BFGS controls
raise accepted, matching fits from `r count("GLS","lbfgs_default","accepted_match")`
to `r count("GLS","lbfgs_tight","accepted_match")`;
PORT reaches `r count("GLS","port_default","accepted_match")`.
Missing attempts after job timeouts
are reported separately and prevent an unconditional ranking.

**The two historical Newsom GLS failures do not reproduce in these reruns.**
L-BFGS, PORT and SLSQP agree on their objective values, with small terminal
gradients. This run supplies no affirmative evidence for the old cancellation
explanation. It does not certify evaluator accuracy on every model.

**Some failures belong to the starting point, and success flags remain
insufficient.** The Little phantom model reaches a much worse stationary point from
ordinary starts under all three main optimizers. The curvature audit rejects
that point; a verified-solution restart recovers the lower objective and passes. A separate scalar probe shows that bounded SLSQP can report
success far from the known optimum; its independent terminal audit rejects it.
A focused start experiment recovers 10 of the 12 rejected ML cases in both
PORT and lavaan by changing free, zero-valued latent paths to 0.5. Lavaan's own
defaults also fail to recover many of these reference solutions. This implicates
start construction; it does not establish that the diagnostic 0.5 rule is a
suitable general policy. A narrower sample-based initializer recovers three ML
cases and one GLS case with each optimizer, but fails unit-invariance checks
when composed with the existing loading starts. It is not ready for promotion.
The layered start below supersedes this prototype. No default is changed on the basis of this study.

## Evidence

### Corpus recovery

“Accepted + matched” requires the library's authoritative acceptance and an
objective within `1e-6 * (1 + abs(best))` of the best objective among accepted fits and converged references.
That threshold is for this comparison; it is neither a production stopping rule
nor a global-optimum certificate.

```{r}
policies <- c(lbfgs_default="L-BFGS, default",lbfgs_budget="L-BFGS, larger budget",
  lbfgs_tight="L-BFGS, tighter",lbfgs_memory50="L-BFGS, memory 50",
  port_default="PORT, default",port_tight="PORT, tighter",
  slsqp_default="SLSQP, default",slsqp_tight="SLSQP, tighter",
  lbfgs_slsqp="L-BFGS → SLSQP fallback")
z <- data.frame(Policy=unname(policies))
for(est in c("ML","GLS")) z[[est]] <- vapply(names(policies),function(a)
  sprintf("%d / %d",count(est,a,"accepted_match"),count(est,a,"attempted")),character(1))
knitr::kable(z,col.names=c("Policy","ML Accepted + Matched / Attempts","GLS Accepted + Matched / Attempts"))
```

On paired cases, default PORT gains `r sum(paired$rescued[paired$estimator=="ML" & paired$arm=="port_default"])`
ML successes and loses `r sum(paired$lost[paired$estimator=="ML" & paired$arm=="port_default"])`
relative to default L-BFGS. For GLS it gains
`r sum(paired$rescued[paired$estimator=="GLS" & paired$arm=="port_default"])`
and loses `r sum(paired$lost[paired$estimator=="GLS" & paired$arm=="port_default"])`;
the loss is Kline's Worland structural-regression Step 2b. PORT therefore merits
an ML recovery-policy investigation, not an unconditional replacement across estimators.

The complete nine-policy comparison and verified-reference restarts are in
`results/full/comparison.csv`. Backend stopping status, acceptance, covariance
admissibility and objective quality remain separate outcomes. Settings do not
have the same meaning across algorithms; tighter numerical controls are not
in themselves a guarantee of a more accurate answer.

### What PORT missed, and which identification it used

PORT's 17 misses per estimator are not 17 solver breakdowns. A targeted rerun
reproduces all 34 outcomes and separates them as follows:

| Outcome | ML | GLS |
|---|---:|---:|
| Non-finite objective error | 2 | 0 |
| Unsupported nonlinear equalities | 1 | 1 |
| Returned but rejected by the fit audit | 12 | 7 |
| Accepted but worse than the comparison objective | 2 | 9 |

The two ML errors are Little Chapter 7 Table 2's `homogc_gender` and
`vars_gender`: both supplied starts already have infinite ML objectives.
Mplus Chapter 6 Example 6.17 requires nonlinear equality constraints that the
PORT route does not support. Neither category establishes a PORT line-search
failure from a valid supported problem.

All twelve ML audit rejections report nonpositive curvature: Little Chapter 10
Figure 12; Chapter 3 Figure 3.11; Chapter 7 Figure 7.10, Table 7.6
(weak/strong/partial), and Table 7.8 (pruned initial/partial/variance test);
and Chapter 8 Figure 7, Figure 9, and Table 8 bullying models. The two accepted
but worse ML fits are Kline's Worland Step 2a and Little Chapter 8 Figure 3b.

The seven GLS audit rejections are Geiser's Chapter 4 latent autoregressive
cross-lagged model; Kline's Sabatelli Heywood model and Worland Step 2b;
the three Chapter 8 Little bullying models above; and Newsom Example 5.5b.
The nine accepted but worse GLS fits are the remaining nine Little models in
the twelve-model ML curvature list. Full case IDs, objective gaps, messages,
and reproduced curvature statuses are in `results/full/port_failures.csv`.

**This was not an identification sweep.** We retained each corpus model's
syntax. The builder defaults were `auto_fix_first=TRUE`, `std_lv=FALSE`, and
`effect_coding=FALSE`, but explicit syntax overrides automatic markers and can
impose effect coding directly. Consequently:

- Kline's Worland factors use `verbal`, `read`, `motive`, and `parent` as
  unit-loading markers. Sabatelli uses `problems` and `father`.
- Geiser's depression/competence factors use `d11/d12/d13` and `c11/c12/c13`.
  Newsom's simplex uses unit-loading single indicators `srh1` through `srh6`.
- Little Figure 7.10 uses indicator 2 for both positive and negative affect
  at each wave. Table 7.8 uses indicator 2 for positive affect and indicator 3
  for negative affect. Both also contain unit-variance phantom factors.
- Little Table 7.6 has no observed marker loading: unit-variance phantoms and
  paths fixed to one in the reference group anchor the scales; shared loadings
  link groups. This is not the same as requesting a blanket `std.lv=TRUE` fit.
- Little Chapter 3 Figure 3.11, Chapter 10 Figure 12, and the Chapter 8 bullying
  models combine explicitly effect-coded measurement blocks with unit-variance
  higher-order/phantom factors. The two Chapter 7 Table 2 error cases also use
  effect coding, plus unit-loading single-indicator covariate factors.
- Little Chapter 8 Figure 3b fixes all level loadings to one and the slope's
  `NegT2` loading to one; its `NegT3/NegT4` loadings are free. Mplus Example 6.17
  uses fixed intercept loadings and slope times 0, 1, 2, 3.

`port_factor_identification.csv` lists every factor and group in the affected
models, including fixed zero loadings; those zeros are not scale markers.
`port_identification_constraints.csv` retains their equality constraints.
The clustering among phantom models motivates checking starts and curvature;
it does not establish that a particular identification causes the failures.

### Previously flagged GLS cases

```{r}
z <- subset(d, estimator=="GLS" & case %in% c("newsom_2015_ex5_4","newsom_2015_ex5_4c") & arm %in% primary)
z$case <- sub("newsom_2015_","",z$case); z$arm <- unname(labels[z$arm])
z$gradient <- formatC(z$gradient,format="e",digits=1)
knitr::kable(z[,c("case","arm","f","gradient","converged")],digits=9,
             col.names=c("Case","Optimizer","Objective","Terminal Gradient","Accepted"))
```

The original cancellation explanation came from older inputs and numerical
paths. A solver's noisy-gradient status can motivate investigation but cannot
establish where numerical error originates. A persistent discrepancy would
require an independent same-point objective and derivative comparison.

### Starting-point dependence

```{r}
z <- subset(d, estimator=="ML" & case=="little_2013_ch3_fig_3_11_longitudinal_cfa_phantom" &
              arm %in% c(primary,"lbfgs_reference_start"))
z$arm <- c(labels,lbfgs_reference_start="L-BFGS, verified start")[z$arm]
knitr::kable(z[,c("arm","f","converged","objective_gap")],digits=6,
             col.names=c("Policy","Objective","Accepted","Gap to Best Observed"))
```

The poor phantom endpoint is rejected for nonpositive curvature, despite its
small first-order gradient. It is not certified as a local minimum. The reference
restart is diagnostic, not an ordinary-start reliability result.
The old “12 failures” came from a smaller lower-level test with simple starts
and generic controls. This larger current-interface run is a new baseline,
not a direct rerun of that exact test protocol.

### Crossing starting values and engines

The follow-up targets the 19 audit-rejected PORT fits plus its two non-finite
ML errors: 14 ML and seven GLS cases. All 21 capped jobs completed. Counts below
require each engine's convergence flag and an objective at least as good as the
**earlier** reference, within the same comparison tolerance. Lavaan's flag is
not equivalent to magmaan's curvature/feasibility audit.

```{r}
sx <- r("starts/summary")
sx$policy <- paste(sx$engine,sx$start)
policies <- c("PORT current","PORT simple","PORT lavaan_default",
              "PORT nonzero_latent_paths","PORT reference_solution",
              "lavaan lavaan_default","lavaan current","lavaan nonzero_latent_paths")
labels <- c("PORT: current starts","PORT: simple starts","PORT: lavaan starts",
            "PORT: nonzero latent paths","PORT: reference solution",
            "lavaan: own defaults","lavaan: our current starts","lavaan: nonzero latent paths")
z <- data.frame(Policy=labels)
for(est in c("ML","GLS")) {
  x <- sx[sx$estimator==est,]; x<-x[match(policies,x$policy),]
  z[[est]] <- paste(x$at_least_reference,x$attempts,sep=" / ")
}
knitr::kable(z,col.names=c("Engine and Start","ML Recovered / Attempts","GLS Recovered / Attempts"))
```

The diagnostic changes only free latent-to-latent regression coefficients
whose supplied start is zero to 0.5. It preserves model specification and fixed
parameters; it is not a new production default. Both engines recover ten of the
twelve ML curvature cases. The two bullying ALT models instead reach the same
slightly worse ML objective (0.547575 versus 0.543219) in both engines. The two
invalid-start models remain failures in both. PORT succeeds on all 21 cases
when initialized at the converged reference solution.

The phantom CFA is particularly informative: current, simple, and lavaan starts
all set the two phantom paths to zero. Both engines then reach objective
0.893810, which lavaan accepts but magmaan rejects for curvature. Setting the
paths to 0.5 lets both reach 0.011198. A zero scale path can eliminate first-order
sensitivity to its variance contribution, since that contribution is quadratic
in the path. This is a concrete start-induced degeneracy to investigate.

The reference's provenance matters. Little's original LISREL input explicitly
starts these paths at 0.7. The corpus records verification of a lavaan refit
from the original input's ST/VA/MA values against the supplied LISREL output,
and stores the verified solution in `start_values.csv`. Our reference run then
uses that solution as a start for a fresh lavaan fit. This is evidence for a
known attainable solution, not evidence that LISREL has a superior optimizer
or a certificate of global optimality.

GLS separates from the ML story. Lavaan reaches or improves on the old target in
all seven cases from its own defaults, and from our supplied current starts.
PORT still exhausts its budget on Geiser's cross-lagged model, Kline's Sabatelli
model and Worland Step 2b under the nonzero-path start. These retain an
optimizer/scaling/control question. The two bullying ALT GLS cases reach
0.501215, below the earlier 0.501320 reference, in lavaan and in PORT with the
nonzero-path start. The earlier reference is therefore explicitly not a global
optimum; historical sweep counts are retained, rather than silently redefined.

`results/full/starts/comparison.csv` retains all endpoints, and the per-case
`__vectors.csv` files retain the supplied parameter starts. Corpus solution
files are used only for the reference-solution arm; ordinary lavaan runs retain
syntax hints but do not receive those files.

### General start-policy gap: structure versus spelling

A controlled equivalent-model probe confirms a syntax-dependent start choice.
The phantom paths written as latent regressions receive simple starts of zero;
written as free higher-order loadings they receive 0.7. The free/fixed row pattern
matches after reversing the loading notation, and objectives agree at shared
zero and nonzero parameter vectors. The models are equivalent; their current
start heuristics are not. `results/full/start_semantics.csv` records the check.

The current loading constructors leave structural paths at the simple baseline.
The missing capability is therefore an explicit structural initialization pass:
recognize variance-carrying latent paths and initialize them together with the
measurement block's scale and residual variance. Ordinary regressions need not
all receive nonzero starts. A narrowly supported single-parent block can use an
estimated target variance for its child to initialize the magnitude of a scale
path; multiple-parent, cyclic, constrained, and sign-ambiguous blocks need their
own handling or an explicit fallback. Fixed innovation variances must not be
confused with total latent variances.

A production candidate should preserve explicit hints and fixed values, respect
cross-group constraints, and check the resulting implied covariance and the
sensitivity of affected blocks. It should compose with the existing constructors
and identification transport as a separately callable operation. Its validation
must include equivalent loading/regression spellings, marker and unit-variance
identification, changes of observed units, sign conventions, explicit zero hints,
and the full corpus including cases that already work. The observed 0.5 rescue
is motivation for this general work, not its implementation.

### Testing a sample-based structural initializer

The first prototype targets measured latent variables with zero innovation
variance and exactly one latent parent. It estimates the child's total variance
from off-diagonal indicator moments, conditional on the supplied loading starts,
and initializes the path magnitude relative to the parent's implied variance.
It excludes cross-loaded blocks and free correlated-residual pairs, preserves
explicit hints, and projects proposals onto the applicable affine constraints.
Unsupported, degenerate or non-finite proposals leave the original start intact.
This is experiment-local R code, not a package implementation or default.

The scan covered all 608 prepared case/estimator pairs. Six models changed under
each estimator; both optimizers were refitted, giving 24 new fits. Unchanged
vectors reuse their frozen baseline outcomes. Every job completed.

```{r}
t <- r("structural/comparison")
z <- aggregate(cbind(baseline_good,good,rescued,lost)~estimator+arm,t,sum)
z$arm <- ifelse(z$arm=="port_default","PORT","L-BFGS")
knitr::kable(z,col.names=c("Estimator","Optimizer","Baseline Recovered","Prototype Recovered","Gains","Losses"))
```

The recovered ML models are the phantom CFA and Little Table 7.8's pruned
initial and partial models. The phantom CFA is also recovered under GLS.
Previously successful cases retain their old starts. The three Table 7.6 models
improve in objective but still miss the target; Figure 7.10 and Table 7.8's
variance-test model fall back after degenerate constrained proposals. The
broader 0.5 diagnostic recovered ten of the twelve ML curvature cases, whereas
this restricted moment rule recovers three. It is therefore a limited proof of
useful sample-based initialization, not yet a general replacement.

**Property tests prevent promotion despite those gains.** Twenty-five of 28
constructor/property combinations pass. Equivalent syntax, explicit hints
(including zero and negative hints), parent-variance rescaling, effect-coding
equalities and ordinary-regression preservation pass. End-to-end observed-unit
rescaling fails with both the simple and FABIN3 baselines; a reverse-keyed
indicator example also falls back under simple starts. The moment rule itself
passes the unit and sign checks when measurement starts are transformed
consistently. The inherited loading constructors treat a fixed zero innovation
variance as though it fixed the factor's total scale. That changes loading
starts with observed units even when a marker loading remains fixed, and the
simple branch also bypasses marker-based signs in this situation.

The next change must coordinate measurement and structural scale initialization,
separating innovation variance from total latent variance. It should then revisit
constraint-coupled blocks and unsupported higher-order/growth structures. Keep
the failed property checks as explicit promotion blockers. The full policy
comparison and fallback reasons are in `results/full/structural/comparison.csv`;
`structural_properties.csv` records both passing and failing checks.

### The layered moment start

At the current start, the phantom CFA, the second-order model and Table 7.6 have
gradient exactly zero in every phantom or higher-order path, with second
derivatives between −1.9 and −17. This is the reflection symmetry at work, not
slow progress: the implied covariance depends on each such path only through its
square or through products with the latent covariances, which also start at zero.

The layered start estimates loading shapes per latent on the standardized
covariance, fits the covariance of the measured latents to the indicator
covariances, rescales each latent to the model's own identification (markers,
effect coding, fixed variances, loading equalities across groups or time), and
then fits the paths and latent covariances to that latent covariance, starting
from moment-based, nonzero magnitudes for variance-carrying paths. Means are fitted
last and the result is projected onto the model's equality constraints. The same
change corrects a defect in the simple and FABIN starts: a disturbance fixed at
zero was read as a std.lv scale, which made loading starts depend on observed units.

```{r}
z <- data.frame(Estimator=c("ML","ML","GLS","GLS"),
  Optimizer=c("PORT","L-BFGS","PORT","L-BFGS"),
  arm=c("port_default","lbfgs_default","port_default","lbfgs_default"))
z$Current <- mapply(function(e,a) lc(e,a,"current"),z$Estimator,z$arm)
z$Layered <- mapply(function(e,a) lc(e,a,"layered"),z$Estimator,z$arm)
z$Gained <- mapply(function(e,a) lp$rescued[lp$estimator==e & lp$arm==a],z$Estimator,z$arm)
z$Lost <- mapply(function(e,a) lp$lost[lp$estimator==e & lp$arm==a],z$Estimator,z$arm)
knitr::kable(z[,c("Estimator","Optimizer","Current","Layered","Gained","Lost")],
  col.names=c("Estimator","Optimizer","Current Start","Layered Start","Gained","Lost"))
```

Counts are accepted fits within the comparison tolerance of the best objective
seen in any run, out of 304 per estimator. Both starts use the same installed
engine. Of the 21 targeted pairs from the crossing study, all seven GLS pairs
recover with both optimizers, and 12 of the 14 ML pairs; the exceptions are
Little's two Chapter 8 ALT models. From the layered start they reach 0.547575, as they do
from the 0.5-path diagnostic and in lavaan, against 0.543219 from the verified
LISREL start. This looks like a second local optimum, not a start defect.
The two cases with an infinite objective at the current start (Table 2 gender
models, where fixed covariances meet constant 0.05 variance starts) now start
finite and recover.

The remaining losses are optimizer events from better starts. L-BFGS aborts its
first line search on Geiser's quadratic growth model, whose solution lies near
the edge of the admissible region. PORT stops with variances 1 to 4% short on
Kline's raw-unit Roth models, and L-BFGS stops early on Lynam's GLS model.
Trait-state models, whose latent covariance only the structural part identifies,
keep the FABIN3 start. Full endpoints, start objectives and start notes are in
`results/full/layered/`.

```{r}
z <- eng[,c("estimator","label","good","false_success","failed","median_seconds")]
z$median_seconds <- round(1000*z$median_seconds,1)
knitr::kable(z,col.names=c("Estimator","Engine and Start","Recovered","Converged Elsewhere","Failed","Median Fit (ms)"))
```

This earlier comparison scores lavaan by its own flag; the short answer gives
the comparison under the shared judge. "Converged elsewhere" counts runs that
report success at an objective above the shared target. For lavaan these are mostly the reflection-trap saddles (the
phantom, second-order and Table 7.6 models), which lavaan reports as converged.
Where lavaan's defaults beat magmaan's layered L-BFGS fit, three cases are
L-BFGS's early line-search abort (PORT recovers them), one needs nonlinear
equality constraints these routes do not support (Mplus Example 6.17), and one is
a genuine second optimum: in Kline's Worland Step 2a, lavaan's cold start reaches
a lower, extreme suppression solution (exogenous latent correlation −0.94), while
the layered start, in both engines and in lavaan, reaches the moderate one. Fit
times exclude input preparation and start construction (about 1 ms for the
layered start); they are single concurrent executions, not a controlled speed ranking.

### Why lavaan's PORT differs from magmaan's

Both engines call the same PORT routine, and magmaan's adapter reproduces R's
`nlminb` exactly: from the layered start on Kline's Roth path model, R's
`nlminb` on magmaan's objective stops after 11 iterations at the same point as
magmaan's PORT. The objective and gradient also agree with lavaan's at the start.
The difference is what lavaan hands to `nlminb`. It sets PORT's diagonal scale to
1/|start| for every parameter above 1 in magnitude. Unscaled, variances in the
thousands dominate PORT's relative step test, only the paths move, and the run
stops early by X-convergence. With lavaan's scale vector, the same objective reaches
the optimum in 25 iterations, as lavaan does.

```{r}
ps <- r("port_scaling")
ps$case <- sub("^kline_2023_ch9_roth_illness_","Kline Roth ",sub("^kline_2023_ch1[45]_","Kline ",sub("^geiser_2013_ch4_","Geiser ",ps$case)))
ps$message <- sub(" \\(.*","",ps$message)
knitr::kable(ps[,c("case","estimator","run","f","iterations","message")],digits=6,
  col.names=c("Case","Estimator","Run","Objective","Iterations","Stop"))
```

The three GLS cases where lavaan succeeded from magmaan's earlier start are
different. Neither the scale vector nor lavaan's GLS objective (magmaan's times
(N−1)/N, the same minimizer) reproduces lavaan's endpoint systematically. With
both, a replication tracks lavaan's iterates to six digits for 40 iterations,
then drifts and ends at a singular point. These are path-sensitive, ill-conditioned
problems, not a systematic lavaan advantage; the layered start recovers all three
in both magmaan optimizers. The two engines also assess convergence differently.
lavaan accepts PORT's return code; magmaan requires its terminal audit. In this
comparison the stricter audit never rejected a magmaan fit at the target
objective, while lavaan's flag accepted the reflection-trap saddles.

### Optimizer coordinates

Quasi-Newton searches depend on the coordinates they run in. L-BFGS starts
from an identity metric and steps along the raw gradient. PORT starts from a
Hessian guess set by its scale vector, bounds its trust region in scaled steps,
and stops by X-convergence when the largest scaled step is small relative to the
largest scaled coordinate; in raw units one large variance or mean dominates that
test. Three coordinate systems were compared on every pair, with PORT and
L-BFGS, from the current and the layered start: raw coordinates; each
parameter's sample unit (observed standard deviations and each latent's
identification: marker, fixed variance, higher-order marker, cross-group
equality, or effect coding); and those units refined downward by the expected
information at the start. Both scaled systems transform like the parameters
when the observed variables change units.

```{r}
rows <- expand.grid(opt=c("nlopt-lbfgs","port"),est=c("ML","GLS"),start=c("layered","current"),stringsAsFactors=FALSE)
cb <- function(mode,e,st,o,sc) paste0(gc(mode,e,st,o,sc)," / ",g(mode,e,st,o,sc))
tab <- data.frame(Start=rows$start,Estimator=rows$est,Optimizer=ifelse(rows$opt=="port","PORT","L-BFGS"),
  None=mapply(cb,"native",rows$est,rows$start,rows$opt,"none"),
  Units=mapply(cb,"native",rows$est,rows$start,rows$opt,"sample_units"),
  Information=mapply(cb,"native",rows$est,rows$start,rows$opt,"information"))
knitr::kable(tab,col.names=c("Start","Estimator","Optimizer","Raw","Sample Units","Information"),
  caption=paste0("Certified local minima / of those, at the best known objective; ",np("native","ML")," standard pairs, native units."))
```

From the layered start the information coordinates lose no pair to raw
coordinates or to sample units and need the fewest evaluations
(`scaling_summary.csv`). From the current start they add many L-BFGS
recoveries; PORT from those far starts gains and loses a few pairs to path
sensitivity (`scaling_paired.csv`, `scaling_changes.csv`). Sample units alone
leave a well-determined parameter's first L-BFGS step too long; the information
refinement shrinks it. The refinement never enlarges a unit: at a start with a
collapsed variance the loadings carry almost no information, and the first run,
which let information enlarge units up to a thousandfold, lost the Geiser and
Newsom trait-state models that way. It also takes the start's own implied
covariance into account, so a start close to a singular covariance does not
receive a step its own curvature forbids.

```{r}
rs <- expand.grid(opt=c("nlopt-lbfgs","port"),est=c("ML","GLS"),stringsAsFactors=FALSE)
lcb <- function(e,st) paste0(lv("rescaled",e,st,"certified")," / ",lv("rescaled",e,st,"best_known"))
rt <- data.frame(Estimator=rs$est,Optimizer=ifelse(rs$opt=="port","PORT","L-BFGS"),
  Pairs=sapply(rs$est,function(e)np("rescaled",e)),
  None=mapply(cb,"rescaled",rs$est,"layered",rs$opt,"none"),
  Information=mapply(cb,"rescaled",rs$est,"layered",rs$opt,"information"),
  Lavaan=sapply(rs$est,function(e)lcb(e,"default")),
  LavaanLayered=sapply(rs$est,function(e)lcb(e,"layered")))
knitr::kable(rt,col.names=c("Estimator","Optimizer","Pairs","Raw","Information","lavaan","lavaan, Layered Start"),
  caption="Rescaled units (each observed variable times a power of ten between 0.01 and 100), unit-invariant pairs: certified / best known. magmaan from the layered start; lavaan judged by the same verdict.")
```

Rescaling keeps the optimum only when no fixed value or equality ties
variables whose units change differently; `R/check_invariance.R` keeps
`r np("rescaled","ML")` ML and `r np("rescaled","GLS")` GLS pairs whose native optimum,
carried to the new units and refined, attains the native objective.
The current start is not unit-equivariant (its start objective moves on most
invariant pairs), so its rescaled counts are lower; the layered start moves on
32 pairs, which the start backlog records.

The one nonlinear-constraints model, Mplus ex6.17, is outside these tables:
PORT and L-BFGS reject its constraints, and SLSQP in information coordinates
reaches the best known objective from both starts under ML and GLS
(`r sum(lane$best)` of `r nrow(lane)` fits, `constrained_lane.csv`).

For FIML, `R/test_fiml_scaling.R` fits the `r fn` single-group continuous FIML
cases with L-BFGS. Information coordinates certify `r fg("native","nlopt-lbfgs","information")`
in native units (raw: `r fg("native","nlopt-lbfgs","none")`) and
`r fg("rescaled","nlopt-lbfgs","information")` after rescaling (raw:
`r fg("rescaled","nlopt-lbfgs","none")`), every certified fit at the best known
objective. The native case it misses reaches the best objective but stops
short of the Newton check's accuracy under FIML's generic stopping controls.

### Unit-scale domain recovery

A scalar variance likelihood has a known optimum. The problem is unchanged;
only units, optimizer coordinates and an explicit nonnegative bound vary.
These rows use a variance of one millionth and a start ten times the optimum.

```{r}
z <- subset(domain,variance==1e-6 & arm %in% c("lbfgs","port","slsqp"))
z$condition <- ifelse(z$normalized,"Normalized coordinates",
                       ifelse(z$bounded,"Raw coordinates, bounded","Raw coordinates, unbounded"))
z <- z[!duplicated(paste(z$condition,z$arm)),]
z$accurate <- is.finite(z$relative_error) & z$relative_error<1e-6
knitr::kable(z[,c("condition","arm","returned","stationary","accurate")],
             col.names=c("Condition","Optimizer","Returned","Audit Passed","Known Optimum Recovered"))
```

The larger-budget and tighter L-BFGS arms do not remove the small-scale failure.
Normalized coordinates recover the optimum in this probe. That is evidence for
work on scale-aware domain recovery, not a validated normalization policy for
arbitrary constrained SEMs.

## Caveats

The scoring rule changed on 2026-09-26. The short answer and the optimizer
coordinates section count certified local minima first and the best known
objective second, with lavaan judged by the same verdict. The older sections
(corpus recovery, starts, the layered start) keep their original rule,
accepted fits at the best objective, and lavaan's own convergence flag. The
shared judge also exposed that `evaluate_at` computed the Newton check only
for ML; it now does so for every moment estimator, so lavaan's GLS endpoints
are judged as magmaan's own are.

The information refinement was revised twice on this corpus: the first run let
information enlarge units (losing the trait-state models) and weighted by the
sample covariance only (a runaway on a latent change score FIML fit in the unit
tests). The final rule is a principled restriction of both, but the counts are
in-sample; a fresh panel should confirm them. Rescaling by variable name breaks
invariance for longitudinal models whose waves share loadings, so the rescaled
panel keeps fewer than half of the pairs. The FIML check covers 13 cases.

The layered comparison reruns both starts on one engine build that includes the
start change, so its current-start counts differ slightly from the frozen main
study. Its target is the frozen best observed objective, lowered when a new fit
is better; it is not a global-optimum certificate. Start objectives and counts
are descriptive for this corpus. Fresh synthetic panels, other estimators and
the sanitized build remain to be checked before a default change.

The run scheduled `r nrow(jobs)` case/estimator jobs;
`r sum(jobs$exit != 0)` ended with a process failure or wall-clock timeout.
The initial cap was 120 seconds per case/estimator across all arms.
The `r nrow(retries)` omitted ordinary arms then received separate 120-second
attempts; `r sum(retries$exit != 0)` remained incomplete. These caps include
input preparation and reference fitting, so a timeout does not isolate optimizer
cost or establish numerical failure.
There were `r nrow(exclusions)` preparation exclusions. Categorical, FIML-only,
composite and multilevel models are outside this comparison; unsupported input
conventions and singular sample covariances are recorded rather than silently
changed. The included inputs comprise 304 cases for each estimator. This is an
ML/GLS experiment, not coverage of every supported estimator. `jobs.csv`, `exclusions.csv` and `input_contracts.csv` retain the details.
Completed-attempt denominators differ when a timeout interrupts the policy sequence.

Current start construction is held fixed across ordinary arms within each
case and estimator. ML retains its current coordinate-scaling policy, which
is not identical for every backend. The study compares available policies,
not isolated algorithms under identical preconditioning. Syntax hints are
honored; separate corpus solution-start files are reserved for references.

Only accepted fits and converged references define the comparison target;
failed or infeasible endpoints cannot lower it. The reference parameters are
evaluated under the same original objective.
Reference evaluation gradients live in full coordinates, whereas fitted
optimizer gradients may live in equality-reduced coordinates; they must not
be compared as though they were the same stationarity measure.

The engine is pinned to the installed build recorded in `metadata.csv` and
`library_hashes.csv`; the working repository may subsequently advance. Corpus
input hashes are checked before and after the run. Single-pass, concurrent
wall times are descriptive and order-dependent, not a controlled speed ranking.

The current default remains L-BFGS. A policy change needs recovery gains without
unexplained losses, preserved model/objective semantics and acceptable cost.
Reopen this experiment after a domain-recovery, scaling or start-policy change;
reopen the GLS accuracy diagnosis only with a current reproducible discrepancy.

## Reproduce

Install the intended `magmaanlab` build and provide the optional corrected
textbook corpus. Run from this experiment's directory:

```sh
export OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --results results/smoke
Rscript run_experiment.R --results results/full --workers 3 --timeout 120 \
  --engine-source COMMIT_USED_TO_BUILD_THE_PACKAGE
Rscript R/complete_timeouts.R ../../../../external/textbook-corpus results/full
Rscript R/validate_inputs.R ../../../../external/textbook-corpus results/full
Rscript R/summarize.R results/full
Rscript R/inspect_endpoints.R ../../../../external/textbook-corpus results/full
Rscript R/inspect_port.R ../../../../external/textbook-corpus results/full
Rscript R/compare_starts.R ../../../../external/textbook-corpus results/full
Rscript R/summarize_starts.R results/full/starts
Rscript R/probe_start_semantics.R ../../../../external/textbook-corpus results/full
Rscript R/check_structural_starts.R results/full
Rscript R/test_structural_starts.R ../../../../external/textbook-corpus results/full
Rscript R/test_layered_starts.R ../../../../external/textbook-corpus results/full --workers 3
Rscript R/summarize_layered.R results/full/layered
Rscript R/test_lavaan_defaults.R ../../../../external/textbook-corpus results/full --workers 3
Rscript R/summarize_engines.R results/full
Rscript R/probe_port_scaling.R ../../../../external/textbook-corpus results/full
Rscript R/test_coordinate_scaling.R ../../../../external/textbook-corpus results/full --workers 3
Rscript R/test_coordinate_scaling.R ../../../../external/textbook-corpus results/full --rescaled --workers 3
Rscript R/check_invariance.R ../../../../external/textbook-corpus results/full
Rscript R/classify_problems.R ../../../../external/textbook-corpus results/full
Rscript R/test_constrained_lane.R ../../../../external/textbook-corpus results/full
Rscript R/test_fiml_scaling.R ../../../../external/textbook-corpus results/full
Rscript R/summarize_scaling.R results/full
cmake -S . -B results/build -DEIGEN_INCLUDE_DIR=/path/to/eigen
cmake --build results/build
results/build/domain_probe > results/full/domain.csv
quarto render report.qmd
```

`--cases` selects named corpus cases. Each case/estimator has its own process,
wall-clock cap, log and checkpointed attempts. The timeout follow-up preserves
completed arms and gives each omitted ordinary arm its own capped process.
Results and rendered output are
local, ignored artifacts. The report performs no fitting.


## Activity after the engineering review

Active for remaining recovery failures. The layered start and information-refined
coordinates have since landed; their historical development comparisons below
are not current default-selection runs. Isolate domain/line-search failures under
the shipped policy and verify each proposed recovery on fresh problems without
losing existing successes. Certified worse local minima remain a separate outcome.
