---
title: "SAM Efficiency And Stability"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- file.path("results", "summary.csv")
rep_path <- file.path("results", "replicates.csv")
if (!file.exists(summary_path) || !file.exists(rep_path)) {
  stop("Missing results. Run: Rscript experiments/51-sam-efficiency-stability/run_experiment.R --budget-min 30")
}
tab <- read.csv(summary_path, stringsAsFactors = FALSE)
reps <- read.csv(rep_path, stringsAsFactors = FALSE)
fmt <- function(x, digits = 3) {
  ifelse(is.na(x), "", format(round(x, digits), nsmall = digits, trim = TRUE))
}
```

## Question

How do local SAM and joint ML compare on a simple two-factor structural path when the data are normal, nonnormal from magmaan's independent-generator stack, or five-category ordinal scores treated as continuous?

The point is not a definitive simulation result; it is a bounded engineering probe for SE calibration, optimizer failures, and runtime before broader SAM work.

## Short Answer

```{r}
if (!nrow(tab)) {
  cat("No completed cells are available yet.")
} else {
  ok <- aggregate(ok ~ estimator, reps, mean)
  names(ok)[2] <- "ok_rate"
  med <- aggregate(elapsed_ms ~ estimator, reps, median)
  merged <- merge(ok, med, by = "estimator", all = TRUE)
  best <- merged[order(-merged$ok_rate, merged$elapsed_ms), , drop = FALSE][1L, ]
  cat(sprintf(
    "This run completed %d fits. The highest observed convergence rate was %s for %s; median fit+SE time for that estimator was %.1f ms.",
    nrow(reps), fmt(best$ok_rate, 3), best$estimator, best$elapsed_ms
  ))
}
```

## Evidence

```{r}
show <- tab
show$ok_rate <- fmt(show$ok_rate)
show$failure_rate <- fmt(show$failure_rate)
show$bias <- fmt(show$bias)
show$emp_sd <- fmt(show$emp_sd)
show$mean_se <- fmt(show$mean_se)
show$se_to_sd <- fmt(show$se_to_sd)
show$coverage95 <- fmt(show$coverage95)
show$median_ms <- fmt(show$median_ms, 1)
show$p90_ms <- fmt(show$p90_ms, 1)
names(show) <- c("Generator", "N", "Estimator", "Reps", "OK Rate",
                 "Failure Rate", "Bias", "Emp SD", "Mean SE", "SE / SD",
                 "Coverage 95", "Median Time (ms)", "P90 Time (ms)")
knitr::kable(show)
```

The raw replicate-level output is in `results/replicates.csv`; the table above is `results/summary.csv`.

## Caveats

This is a small-N engineering probe. The ordinal condition is deliberately pseudo-continuous: it feeds integer category scores to the current continuous SAM and ML paths, so its target is estimator-specific and estimated by a large reference draw.

The default runner enforces the budget at cell boundaries, so a 30-minute run may finish only part of the requested grid.

## Reproduce

```sh
Rscript experiments/51-sam-efficiency-stability/run_experiment.R --budget-min 30
Rscript experiments/51-sam-efficiency-stability/run_experiment.R --smoke
```
