---
title: "Sphere reference fits: a small exploratory pilot"
params:
  run: reference-pilot
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
read_result <- function(name) {
  path <- file.path("results", params$run, paste0(name, ".csv"))
  if (!file.exists(path)) stop("Missing results: run Rscript run_experiment.R --pilot")
  read.csv(path, stringsAsFactors = FALSE)
}
s <- read_result("summary"); refs <- read_result("references"); metadata <- read_result("metadata")
probe_path <- file.path("results", params$run, "witness_probe.csv")
probe <- if (file.exists(probe_path)) read.csv(probe_path) else NULL
count <- function(domain, label) sum(refs$domain == domain & refs$reference_label == label)
tab <- function(x) knitr::kable(x, row.names = FALSE, digits = 3)
```

## Question

Can a small collection of sphere fits provide useful reference solutions for
ML and PSD-ML, while keeping local numerical accuracy, chart problems and
uncertainty about the best solution separate? Can those references explain
ordinary-route failures without changing the estimator used in a paper?

## Short answer

Across `r sum(refs$domain == "ML")` datasets, the best screened sphere objective
is reached from at least two start recipes on
`r count("ML", "repeated_best_observed")` ML problems and
`r count("PSD", "repeated_best_observed")` PSD problems. There is no screened
sphere reference on `r count("ML", "no_screened_reference")` ML problems and
`r count("PSD", "no_screened_reference")` PSD problems. Those remain unresolved;
we do not label them nonexistent estimates or proven poles.

This pilot supplies reference **candidates**, not certified optima. It uses an
existing local-accuracy check after re-expressing each endpoint with its
strongest standardized indicator as marker. A sphere-native curvature check
remains missing. Agreement across starts strengthens the numerical evidence but
does not establish global optimality. No library default changes here.

```{r results='asis'}
if (!is.null(probe)) {
  p <- probe[1, ]
  cat(sprintf(paste0("The targeted follow-up supplies a concrete counterexample to calling failed multistarts ",
    "\"no estimate\": on %s draw %d at N=%d, lavaan finds a screened objective of %.6f. ",
    "%d sphere backend(s) recover a screened candidate when started there. ",
    "The original finite portfolio missed a point the sphere route can represent.\n"),
    p$design, p$rep, p$n, p$witness_objective, sum(probe$label == "screened_candidate")))
}
```

## Evidence

### How often a reference is found

All datasets remain in the denominator. ML and PSD references are built
separately, using sphere attempts only. Repetition means different start
recipes reached the best observed objective; two optimizers from the same
start do not suffice.

```{r}
a <- aggregate(list(Datasets = refs$rep), refs[c("design", "domain", "reference_label")], length)
names(a)[1:3] <- c("Design", "Domain", "Reference evidence")
tab(a)
```

### Which sphere configurations reach the reference

The same random starting points are supplied to each configuration on a draw.
Each configuration has canonical and layered starts, plus
`r metadata$value[metadata$key == "random_starts"]` random starts.
The random starts have positive disturbances but varying loading signs and
latent paths. Their limited distribution does not exhaust possible basins.
The table pools starts and the two sample sizes within each design, so its
attempts are repeated fits, not independent datasets. Screening time is included.

```{r}
z <- s[s$route == "sphere", ]
z$screened <- z$fits * (z$label == "screened_candidate")
z$matched <- z$fits * (z$comparison == "matches_sphere_reference")
a <- aggregate(cbind(fits, screened, matched) ~ design + domain + backend + preconditioning, z, sum)
names(a) <- c("Design", "Domain", "Optimizer", "PSD scaling", "Attempts", "Screened", "At reference")
tab(a)
```

### Ordinary fits compared with the sphere reference

These attempts never define the sphere reference. A better ordinary endpoint is
retained as evidence that the sphere search missed something; the reference is
not silently replaced. Every finite returned ordinary endpoint is checked in
the same strongest-indicator chart. Its original convergence flag is retained
separately in the raw results.

```{r}
z <- s[s$route != "sphere", ]
a <- aggregate(fits ~ domain + route + backend + comparison, z, sum)
names(a) <- c("Domain", "Route", "Optimizer", "Comparison", "Attempts")
tab(a)
```

The ordinary PSD route remains ordinary PSD-ML. Sphere is an auxiliary
engineering reference: this study does not introduce sphere fitting into the
ordinary estimator, change paper protocols, or count a sphere rescue as an
ordinary-route success. A paper can report ordinary performance against a
best-observed numerical reference while describing that reference transparently.

## Caveats

- This is exploratory development data, not a held-out default-selection run.
  The initial pilot has three two-factor models, N = 20 and 100, ten draws per
  cell, complete Gaussian data, one group and native units. Both latents must
  be covered by the sphere transformation. No conclusions extend to uncovered
  structures, FIML, DWLS or barrier fitting.
- Fits use unit-free sphere coordinates with user-chart polishing disabled.
  The current route still finalizes in the original chart when possible; a
  finalization error can therefore hide an otherwise usable sphere endpoint.
  Such attempts remain errors, never evidence that no solution exists.
- The strongest-indicator assessment is a local Newton cross-check, not an
  intrinsic sphere certificate or an identification proof. PSD checks use the
  PSD domain. Original-chart verdicts do not decide the reference screen.
- A first-order sphere check, norm residual at most 1e-6, finite consistent
  objectives, and the local Newton budget .01 must pass. PSD also requires
  covariance admissibility. These are an explicit exploratory screen, not a
  replacement library verdict. Nonpositive curvature, unavailable accuracy and
  excessive Newton distance are recorded separately in the raw evidence.
- Standardized extent above 10 is an **extreme-parameter screen**. Such points
  remain recorded but are excluded from the main reference pool. They are not
  called runaways. A finite improper optimum can be ML's best answer; being
  improper alone never excludes an ML candidate. A validated escape diagnostic
  and threshold-sensitivity study remain open.
- Objective agreement uses 1e-6 times one plus the absolute reference objective.
  Equal objectives need not imply equal parameters or the same basin. The
  number of observed objective levels is not a count of local minima.
- The marker diagnostic is evaluated at each endpoint, not at a known optimum.
  A level below 1e-6 means numerical proximity to the requested chart's boundary,
  not a proven zero loading. Levels and classifications at 1e-4, 1e-6 and 1e-8
  are retained. No chart condition is counted as convergence on its own.
- Sphere may remove identification poles while other parameter escape paths
  persist. Neither unsuccessful multistarts nor an ML objective below the PSD
  objective proves nonattainment.

## Reproduce

```sh
cd experiments/engineering/15-sphere-reference-fits
Rscript scripts/check_labels.R
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript scripts/probe_reference_gap.R reference-pilot
quarto render report.qmd
# Optional small paired unit check; a distinct run, not default-selection evidence:
Rscript run_experiment.R --smoke --transforms native,x0.01,x100 --no-lavaan --run-id reference-units
```

The runner refuses to overwrite a completed run; choose a fresh `--run-id` for
reruns and render it with `quarto render report.qmd -P run:NAME`. The seed base
is separate for smoke and pilot profiles. Raw fits, endpoint parameter tables,
random starts and comparison rows stay local under `results/`. Small pilot
summaries, per-problem references and metadata are frozen with the source so
the report can be rendered without rerunning fits. Metadata identifies the
installed package and source checkout; a dirty checkout is recorded explicitly.

## Label contract

Labels describe the evidence collected, never a mathematical existence claim.
The raw columns retain the ingredients so a later screen can be compared without
rewriting the original run.

| Label | Meaning | Does not imply |
| --- | --- | --- |
| `screened_candidate` | Endpoint passes the stated numerical and extent screen | A certified local minimum, identification, or global optimality |
| `screened_extreme` | Local checks pass, but standardized extent exceeds 10 or is undefined | A proven runaway; a finite optimum is impossible |
| `accuracy_failed` | Available Newton distance exceeds its budget | No better start or backend can solve the problem |
| `accuracy_unresolved` | Curvature/conditioning does not support the requested accuracy assessment; consult `audit_status` | Every unresolved case has the same mechanism |
| `sphere_first_order_failed` / `sphere_norm_failed` | The driven first-order or unit-norm check fails | The endpoint is a local optimum at a pole |
| `partial_sphere` | At least one relevant latent did not use the sphere | This attempt is a pole-free reference |
| `nonfinite_objective` / `objective_disagreement` / `psd_infeasible` | A finite, consistently evaluated point in the requested domain was not established | A valid reference candidate |
| `assessment_unavailable` / `call_error` | Endpoint assessment or fitting could not return the required evidence | Nonexistence |
| `single_start_best_observed` | Best screened sphere objective has one supporting start recipe | An unreliable point by definition, or a unique minimum |
| `repeated_best_observed` | At least two start recipes reach that objective within tolerance | Independent replications or exhaustive search |
| `no_screened_reference` | The finite portfolio supplies no qualifying reference | An unattainable problem |
| `near_chart_boundary` | Endpoint's original marker-chart level is below the declared tolerance | A mathematically exact pole or failed sphere optimization |
| `better_than_sphere_reference` | Screened ordinary endpoint improves on the sphere reference | The ordinary endpoint is globally optimal |
| `above_sphere_reference` | Screened endpoint has a worse objective | It lies in a distinct basin; convergence failed |
| `matches_sphere_reference` | Screened endpoint agrees in objective within tolerance | Equality of parameter estimates |
| `not_screened` / `no_sphere_reference` | Comparison lacks a qualifying endpoint or a sphere reference | Inferiority to a known optimum |

## Current choice and next step

The current engineering setup is a small sphere portfolio, strongest-indicator
cross-check, and separate original-chart diagnostics. Alternatives compared are
L-BFGS versus PORT for ML, and diagonal versus no PSD scaling with SLSQP.
A configuration deserves a larger follow-up only if it repeatedly finds the
best screened candidates without concentrating unresolved outcomes in a design;
this pilot does not select a library default.

The targeted witness follow-up is selected after inspecting the pilot; it is
not part of the original reference pool. Both recoveries are recorded in
`witness_probe.csv`. The unit smoke run separately retains paired rescalings;
it is too small to establish general unit invariance.

Next inspect disagreements and single-start references, then decide whether a
sphere-native assessment is necessary for the next claim. Preserve the existing
ordinary-parameterization comparisons for PSD work. Any production promotion
requires a fresh decision study with criteria written before its run.

This study was renamed from `15-sphere-local-convergence` on 2026-09-27.
Historical `results/full` and `results/smoke` are preserved but are not read by
this report. The former `no_estimate` classification combined failed alternative
fits with an objective below PSD; it was heuristic and must not be used as
proof of nonexistence. Historical sources remain available in Git history.
