---
title: "Sphere reference fits: a small exploratory pilot"
params:
  run: reference-pilot
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r}
read_result <- function(name) {
  path <- file.path("results", params$run, paste0(name, ".csv"))
  if (!file.exists(path)) stop("Missing results: run Rscript run_experiment.R --pilot")
  read.csv(path, stringsAsFactors = FALSE)
}
s <- read_result("summary"); refs <- read_result("references"); metadata <- read_result("metadata")
probe_path <- file.path("results", params$run, "witness_probe.csv")
probe <- if (file.exists(probe_path)) read.csv(probe_path) else NULL
count <- function(domain, label) sum(refs$domain == domain & refs$reference_label == label)
tab <- function(x) knitr::kable(x, row.names = FALSE, digits = 3)
```

## Question

Can a small collection of sphere fits provide useful reference solutions for
ML and PSD-ML, while keeping local numerical accuracy, chart problems and
uncertainty about the best solution separate? Can those references explain
ordinary-route failures without changing the estimator used in a paper?

## Short answer

Of the nine inspected spherical endpoints, two are near-pole failures in the
requested marker chart. The other seven now have numerical evidence of finite
local minima in the original marker identification. Five refine directly to
stable points; one requires the better starting point found earlier; the last
has an exceptionally flat but finite minimum, confirmed at 60 and 90 decimal
digits. Six refined points pass the library's ordinary-precision accuracy check.
The very flat case still fails that check and should remain a numerical failure,
not be called nonattainment. None of this proves global optimality.

All seven finite ML witnesses have negative primitive variance components.
That is separate from chart availability and numerical accuracy, and excludes
them as PSD-ML solutions. In the separate retained fresh PSD batch, FABIN3-auto
plus positive spectral starts matches 51/60 sphere references under either
scaling choice. The remaining search gap and the requested-chart rejection gate
still block promotion. Marker changes in this study are diagnostic only;
automatic substitution of the user's markers is not the adopted policy.

```{r results='asis'}
if (!is.null(probe)) {
  p <- probe[1, ]
  cat(sprintf(paste0("The targeted follow-up supplies a concrete counterexample to calling failed multistarts ",
    "\"no estimate\": on %s draw %d at N=%d, lavaan finds a screened objective of %.6f. ",
    "%d sphere backend(s) recover a screened candidate when started there. ",
    "The original finite portfolio missed a point the sphere route can represent.\n"),
    p$design, p$rep, p$n, p$witness_objective, sum(probe$label == "screened_candidate")))
}
```

## Evidence

### What became of the other seven?

We checked the seven endpoints that pass the study's marker-proximity test
using an independent high-precision calculation of the same unrestricted
Gaussian ML objective. The original markers x1 and y1 stay fixed at 1. Both
the saved sphere endpoint and the best locally checked point found earlier
were used as starting points. The calculation retains the starting sign of
X's variance while refining, so a trajectory is not silently carried through
a singular regression representation. It is a local diagnostic, not a new
production fitter or an exhaustive search.

The derivative formulas agree with independent numerical differentiation at
high precision. Every retained finite witness has a positive definite observed
covariance, a vanishing gradient to the stated numerical precision, and positive
curvature in every free parameter direction. These are strong numerical local
minimum checks, not exact-arithmetic or global-optimality proofs. The resulting
points reproduce the library's objective in the original marker chart to
less than 1e-14 and all pass the explicit 1e-4 chart-proximity check.

```{r}
finite <- read.csv(file.path("results", "open-case-conclusions", "finite_witnesses.csv"))
accuracy <- read.csv(file.path("results", "open-case-conclusions", "library_crosscheck.csv"))
finite <- merge(finite, accuracy[c("case_id","accuracy_status","accuracy_passed")], by="case_id", sort=FALSE)
tab(data.frame(Batch=finite$batch, Design=finite$design, N=finite$n, Draw=finite$rep,
  Objective=finite$objective, `Largest loading`=finite$max_loading,
  `Ordinary-precision accuracy`=ifelse(finite$accuracy_passed,"Pass","Ill-conditioned"),check.names=FALSE))
```

Five saved spherical endpoints refine directly to the same finite minimum as
the earlier checked point. Their large or negative variance components do not
by themselves indicate a runaway path. Two cases need a more specific explanation:

**Fresh weak-marker N=20 draw 1: the sphere had followed a losing path.**
Sign-preserving continuation from that endpoint approaches X variance zero,
while the regression coefficient and compensating disturbance grow without
bound numerically. The continuation reaches X variance about 2e-32 and a
regression magnitude above 5e30 before stopping; its objective is about
0.2920422 and its gradient has not vanished. The previously found negative-X-
variance point instead refines to a finite local minimum at 0.2917589, with
regression about 6.33 in the requested marker chart. Both sphere backends
recover it when started from that witness. This identifies a missed basin,
not nonattainment of the entire problem. It does not establish what other
paths or minima exist.

**Development weak-marker N=100 draw 2: a very flat finite minimum.**
Simply increasing ordinary refinement effort made little progress. Holding
the y3 loading fixed and optimizing the remaining parameters shows the objective
first decreasing and then increasing. Solving for the bottom of that local
profile gives loading 740.4555345 and objective 0.0246899172027636. A subsequent
unrestricted gradient/curvature check confirms a finite numerical local minimum;
60- and 90-digit calculations agree. This is evidence against interpreting the
investigated branch as an endlessly improving large-loading sequence.

```{r}
profile <- read.csv(file.path("results", "delicate-loading-profile", "summary.csv"))
minimum <- read.csv(file.path("results", "delicate-finite-minimum", "summary.csv"))
plot(profile$fixed_y3_loading, (profile$objective-minimum$objective)*1e9,
  log="xy", type="b", pch=19, xlab="Fixed y3 loading (y1 marker = 1)",
  ylab="Objective above local minimum (× 10⁹)")
abline(v=minimum$max_loading, lty=2, col="steelblue")
```

The ordinary-precision library still reports this point as ill-conditioned,
with an accuracy-condition measure around 5e12, even when supplied the refined
point. It should remain rejected or unresolved on numerical accuracy grounds.
Finding the minimum at high precision does not make routine parameter estimates
or inference reliable. The other six refined witnesses pass that accuracy check.
All seven are improper ML points; none may be substituted for a PSD fit.

### Requested-chart failure: the 184,000-loading example

**Decision:** reject the fresh weak-marker N=100 draw 4 endpoint in the requested
identification. Do not automatically switch its marker. Report that the fitted
point is too close to a pole of the requested marker chart to return as an
accepted fit in that chart. The alternative-marker point and implied covariance
may be retained for diagnosis, and the user can explicitly choose a different
identification.

The saved normalized chart level is 5.44e-6. The existing translation tolerance
1e-6 admits it; the already-reported study threshold 1e-4 rejects it. The new
required-failure check confirms this using the saved point without optimization
or marker changes. The 1e-4 value is an explicit study check, not a newly adopted
library-wide tolerance. A general near-pole acceptance criterion still needs
validation before promotion.

Of the other eight saved endpoints, development weak-marker N=20 draw 4 also
fails that proximity check (level 1.15e-6) and already fails sphere stationarity.
The other seven pass it; their negative variance components and, in two cases,
failed/unresolved accuracy are separate issues. Passing this proximity check
does not establish that a solution is regular, proper or globally optimal.
We should therefore not call all nine “poles,” nor claim that the seven are
proved free of every identification problem.

Suggested failure explanation for the named case: “The requested marker is
extremely weak relative to the other indicators at the fitted point. Fixing its
loading to 1 produces extremely large parameter estimates. The fit is rejected
in this identification; no marker was changed automatically.”

### Looking at the same spherical point in familiar coordinates

This inspection does not fit anything again. It takes each case's lowest-objective
evaluable saved sphere endpoint and translates it to the original marker, a
better-loaded marker, and ordinary std.lv. Successful translations preserve the
implied covariance and objective to numerical precision. “No” for std.lv means
that this same point cannot be expressed there; it is not a new failed fit.

```{r}
translations <- read.csv(file.path("results", "sphere-translations-inspected", "summary.csv"))
first <- translations[translations$chart == "first_marker", ]
strong <- translations[translations$chart == "strongest_marker", ]
std <- translations[translations$chart == "std_lv", ]
stopifnot(identical(first$case_id, strong$case_id), identical(first$case_id, std$case_id))
tab(data.frame(Batch=first$batch, Design=first$design, N=first$n, Draw=first$rep,
  `Largest loading, first marker`=first$max_abs_loading,
  `Largest loading, better marker`=strong$max_abs_loading,
  `Negative variance rows`=strong$negative_variances,
  `std.lv possible`=ifelse(std$translated,"Yes","No"),
  `Local checks passed`=ifelse(first$source_local_checked,"Yes","No"),check.names=FALSE))
```

Here `X` denotes the exogenous latent variance, `Y` the disturbance variance in
`Y ~ X`, and lowercase indicator names denote their residual variances. In this
regression model, std.lv fixes X's variance and Y's disturbance variance to +1.
A real rescaling cannot change a negative variance into a positive one. Negative
indicator residuals, by contrast, do not themselves prevent std.lv translation.

Three concrete examples explain most of what to look for:

- **High-R2, development N=20 draw 4:** with better markers, loadings are at
  most 1 and the regression is about 0.53. Y's disturbance variance is about
  -0.377, almost cancelling the explained variance and leaving total variance
  about 0.0021. The point is locally checked but improper; enormous standardized
  ratios do not mean enormous unstandardized parameters here.
- **Weak marker, fresh N=100 draw 4:** the largest loading falls from about
  184,000 to 1 after changing markers. That dramatic loading was a coordinate
  issue. The negative Y disturbance (about -27.7 under the better marker) and
  large residual compensation remain. Translation clarifies the point but does
  not make it an admissible covariance model.
- **Weak marker, development N=100 draw 2:** std.lv translation works, with a
  largest loading about 12.6, but x3 and y3 have negative residual variances.
  The largest absolute indicator residual variance is about 168 times that
  indicator's sample variance. This endpoint also has unresolved local accuracy,
  so it warrants further numerical inspection.

All nine saved endpoints have negative primitive variance components. This is
allowed by the unrestricted observed-covariance ML objective, but excluded by
PSD-ML. Six pass the local accuracy checks; the other three are retained as
endpoints to inspect, not accepted estimates. Translation alone does not tell
us whether a path is runaway or whether the endpoint is globally optimal.

```{r results='asis'}
pt <- read.csv(file.path("results", "sphere-translations-inspected", "parameters.csv"))
for (i in first$case_id) {
  x <- first[first$case_id == i, ]
  cat(sprintf("\n<details><summary>%s, %s, N=%d, draw %d: parameter tables</summary>\n\n",
    x$batch, x$design, x$n, x$rep))
  z <- pt[pt$case_id == i, ]
  for (chart in c("first_marker", "strongest_marker", "std_lv")) {
    q <- z[z$chart == chart, c("lhs", "op", "rhs", "est")]
    cat(sprintf("\n**%s**\n\n", gsub("_", " ", chart)))
    if (nrow(q)) cat(as.character(knitr::kable(q, digits=6, row.names=FALSE)), sep="\n") else
      cat("This point cannot be translated to positive std.lv.\n")
    cat("\n\n")
  }
  cat("</details>\n\n")
}
```

### Inspecting the unresolved draws

The inspection selects all four development and five retained fresh ML problems
with no screened sphere reference. It fits nine combinations of marker
indicators, effect coding, positive std.lv and the sphere, with L-BFGS and PORT.
The four spectral starting points are transported across charts; their implied
covariances are checked for agreement before fitting. Native constructors are
separate arms. Positive std.lv cannot represent negative latent disturbance
variances, so unavailable starts are recorded rather than silently changed.
Failure in that sign-restricted chart is not failure of unrestricted ML.

For each optimizer, the lowest-objective evaluable endpoint in the marker,
effect-coded, std.lv and sphere families is continued at tighter tolerances
with budgets of 1,000, 10,000 and 50,000. These are allowed budgets, not evidence
that the optimizer used every evaluation. The inspection retains covariance
drift, objective, local accuracy and both the old extent and a new diagnostic
of scale-invariant component magnitudes. Neither magnitude statistic is a
proof of runaway or nonattainment.

```{r}
escape <- read.csv(file.path("results", "unresolved-identifications", "candidate_evidence.csv"))
tab(data.frame(Batch=escape$batch, Design=escape$design, N=escape$n, Draw=escape$rep,
  Objective=escape$objective, `Old extent`=escape$std_extent,
  `Component extent`=escape$component_extent, check.names=FALSE))
```

These are the lowest-objective locally checked candidates in this inspection,
not certified global optima. The magnitude-only screen still labels seven of
these `screened_extreme`; the original records are preserved.

- Development weak-marker N=20 draw 4 and fresh weak-marker N=20 draw 1 now
  pass the original screen. Marker charts find lower finite candidates than
  the sphere attempts in this inspection. Those sphere attempts can show
  large components on losing paths without establishing nonattainment of the
  problem itself.
- Both high-R2 development cases have moderate component magnitudes, around
  1.1–1.3, despite old standardized extents of roughly 18 and 177. A small
  total latent variance makes standardized ratios large; the fitted primitive
  components need not diverge. Multiple charts recover the same objective.
- Fresh Ernst N=20 draw 1 and weak-marker N=20 draw 4, N=100 draw 1, and N=100
  draw 4 have stable locally checked candidates with component magnitudes
  around 13–28. Their large signed variance components remain finite in the
  continuations. These are improper ML candidates, not PSD candidates.
- Development weak-marker N=100 draw 2 remains delicate. Several charts give
  almost the same objective, around 0.024689918, while representative component
  magnitudes range roughly from 125 to 169 and local-check outcomes differ.
  The best evaluable endpoint is not the best locally checked endpoint.
  Stagnation at tighter tolerances does not resolve finite attainment here.

The component diagnostic uses loading energies (squared loadings divided by
sample indicator variances) to scale latent total/disturbance covariances and
structural paths; residual contributions are scaled by indicator variances.
Unlike the old standardized extent, it never divides by a fitted latent total
variance. Both statistics are reported, and neither replaces the accuracy check.
The retained candidate parameters and implied covariances permit inspection
without rerunning the optimizers.

The practical conclusion is to separate magnitude flags from local accuracy
and from evidence about attainment. We do not need every unrestricted ML
problem to have a solution, and making PSD solve a different constrained
problem cannot answer whether ML attains its infimum. However, the original
five missing fresh references cannot be described collectively as five
nonconvergent or apparently nonattained fits.

### Known geometry before promotion

Five deterministic covariance witnesses isolate behavior from sampling noise.
The requested identification is the first marker except for the std.lv case.
The alternative chart always fixes the second indicator, selected before any
fit. ML uses L-BFGS and PORT with native layered and feasible-to-construct signed
spectral starts; PSD uses SLSQP with FABIN3 automatic transport and positive
spectral starts under both no and diagonal preconditioning. The sphere uses its
canonical start with polishing disabled. Its internal scaling is recorded
separately from ordinary ML's information scaling.

```{r}
g <- read.csv(file.path("results", "geometry-witnesses", "fits.csv"))
geom_names <- c("interior", "marker_pole", "residual_face", "stdlv_pole", "nonattainment")
recovery <- function(name, route, domain) {
  z <- g[g$design == name & g$route == route & g$domain == domain, ]
  if (any(z$screened & z$matches_covariance)) "Yes" else "No"
}
tab(data.frame(Case=c("Interior", "Marker pole", "Zero residual variance", "std.lv pole", "ML nonattainment"),
  `Requested ML`=vapply(geom_names, recovery, "", route="requested", domain="ML"),
  `Requested PSD`=vapply(geom_names, recovery, "", route="requested", domain="PSD"),
  `Alternative marker ML / PSD`=vapply(geom_names, function(n) paste(recovery(n,"alternate_marker","ML"), recovery(n,"alternate_marker","PSD"),sep=" / "), ""),
  `Sphere ML / PSD`=vapply(geom_names, function(n) paste(recovery(n,"sphere","ML"), recovery(n,"sphere","PSD"),sep=" / "), ""),
  check.names=FALSE))
```

“Yes” requires both the local endpoint screen and a maximum covariance error,
standardized by the target indicator variances, below 1e-6. It means exact-target
recovery to this tolerance, not simply a convergence flag. The table deliberately
shows no recovery for PSD on the last row: that target is outside the PSD model,
so PSD must minimize a different, strictly positive discrepancy.

At the marker pole the first loading is zero while the other indicators share
nonzero covariance. The original marker cannot represent this point. Its
spectral construction explicitly returns `start_unavailable`; choosing a
representable chart permits fitting. At the std.lv pole the endogenous latent's
disturbance variance is zero, whereas the requested identification fixes it at
one. Ordinary fits report convergence and pass the local screen at finite
approximations, but their standardized covariance gaps range from about 3e-6
(PSD) to 2e-4 (ML). Alternative-marker and sphere gaps are below 3e-8. Thus
convergence and a small discrepancy do not establish finite attainment in the
requested chart. Successful alternative fits retain a translation-failure or
near-boundary flag for that original chart.

The zero-residual witness attains zero discrepancy with an observed covariance
that is still positive definite. Every tested PSD configuration recovers it;
a primitive covariance boundary is therefore not a failed fit. The numerical
face nomination uses a 1e-7 threshold on scaled primitive eigenvalues, with
PSD feasibility and accuracy assessed separately. The scale is invariant to
indicator units and latent rescaling for these simple-structure models; it is
not a general boundary certificate.

#### A nonattainment witness, independently of optimization

For a one-factor model with diagonal residual covariance, consider

$$
S=\begin{pmatrix}1&0.3&0.4\\0.3&1&0\\0.4&0&1\end{pmatrix}.
$$

The nonzero covariances $s_{12}=\lambda_1\lambda_2\phi$ and
$s_{13}=\lambda_1\lambda_3\phi$ force all three loadings and $\phi$ to be
nonzero at a finite solution. Then $s_{23}=\lambda_2\lambda_3\phi$ cannot
be zero. No finite factor parameters reproduce $S$, even with signed variances.
Yet the sequence

$$
\lambda(t)=(t,0.3/t,0.4/t)^\top,\qquad
\phi=1,\qquad \theta_i(t)=1-\lambda_i(t)^2
$$

keeps the diagonal and first two covariances fixed while
$s_{23}(t)=0.12/t^2\to0$. Its observed covariance is positive definite for
$t\ge1$. Since the Gaussian covariance discrepancy is nonnegative and vanishes
only at equality, its infimum is zero and is not attained. This is a proved
closure/nonattainment example; no claim about a particular differential-geometric
fold type is needed. The stored sequence and checks verify the covariance map
and the library's half-discrepancy convention independently of optimization.

For $t>1$ this sequence has a negative first residual variance. PSD excludes it.
All ten tested PSD attempts across the requested, alternative and sphere routes
instead agree at a boundary candidate with reported discrepancy 0.009509003 and
$s_{23}\simeq0.12$. Their agreement and local checks are not a global-optimality
proof. No unrestricted ML attempt passes the local screen on this witness.

### Fresh PSD reference coverage

The retained fresh batch now has a separate PSD reference pool: canonical,
positive spectral and three reproducible random sphere starts under each of
the two preconditioning choices. The ordinary arms run independently on every
draw, with no reference-selected retries. All 60 draws have screened sphere
references: 58 reached by multiple start recipes and two by only one. All are
marker-representable at 1e-6; 42 draws have a matching reference candidate near
a PSD face under the numerical nomination above.

```{r}
psd <- read.csv(file.path("results", "psd-fresh", "summary.csv"))
tab(data.frame(Preconditioning=psd$preconditioning,
  `FABIN3-auto matches`=psd$fabin3_hit, `Positive spectral matches`=psd$spectral_hit,
  `Combined matches`=psd$portfolio_hit, `Any screened candidate`=psd$any_screened,
  check.names=FALSE))
```

Counts are out of 60. Both settings leave eight draws with screened but inferior
ordinary candidates and one with no screened ordinary candidate. No ordinary
candidate improves on the sphere reference. This is evidence for a remaining
search problem, not evidence that these nine datasets lack a PSD solution.
The references remain best observed candidates with a marker-chart accuracy
cross-check; they do not replace the missing sphere-native assessment. The
batch was already used in the direct-start comparison, so it is not a new
held-out decision study.

### What a spectral start does

A start is an initial guess for every parameter. Each indicator block supplies
a correlation pattern. The positive recipe chooses a loading direction that
captures positive shared variation. The negative recipe chooses an alternative
direction suited to a negative latent variance. Changing direction changes the
relative loadings. The construction adjusts residual variances so the implied
observed covariance is positive definite at the start.

Each factor receives one of those two guesses. Regressions and between-factor
covariances start at zero. The optimizer then estimates all free parameters
normally: the initial signs are not constraints, and every recipe targets the
same ML objective. Negative latent variances can be allowed by ML while the
observed covariance remains positive definite; such endpoints may be
substantively inadmissible and are flagged separately. These recipes are not
being proposed for PSD estimation.

### Direct ordinary-chart starts

The direct probe constructs every start from the sample covariance and runs
all recipes on every draw. It reads saved sphere references only after fitting;
no sphere endpoint, reference-selected marker, or reference-dependent retry
enters this portfolio. The spectral directions are converted algebraically to
the requested marker units: divide each loading by its marker loading and
multiply the latent variance by the squared marker loading. A zero marker
component would prevent that conversion; none occurred in these runs.

ML uses native layered starts plus four spectral alternatives, under information
scaling, separately with L-BFGS and PORT. PSD uses FABIN3 with automatic std.lv
transport plus the positive spectral alternative, under SLSQP with either no
preconditioning or diagonal preconditioning. Negative-variance recipes are
excluded from PSD. Each batch contains 60 draws from the original three
populations; the fresh batch is the previously retained start-study batch,
not a newly independent confirmation after this probe.

```{r}
direct <- read.csv(file.path("results", "direct-starts-information", "summary.csv"))
d <- direct[direct$reference_available > 0, ]
tab(data.frame(Batch=d$batch, Domain=d$domain, Backend=d$backend,
  Scaling=d$preconditioning, `Available references`=d$reference_available,
  `Constructor matches`=d$constructor_hit, `Portfolio matches`=d$portfolio_hit,
  check.names=FALSE))
```

Counts are out of 60 draws per batch. A match requires the existing endpoint
screen and objective agreement with the saved domain-specific reference.
No screened ordinary point improved on an available reference in this run.
The constructor is layered for ML and FABIN3 with automatic transport for PSD.
At the time of this direct run, the fresh PSD arms produced screened candidates
on 59/60 draws under either scaling choice, but had no sphere reference pool.
Those frozen match counts remain unavailable; the later PSD-specific run above
adds an independent reference comparison on the same draws. Raw attempts and starting vectors stay
local; the paired rows, summaries and reference classifications are retained.

The earlier reference-informed diagnostic is separate: translating a known
sphere endpoint recovered all 16 ML misses per backend and all six PSD misses
with references. That establishes that those points can be reached from a
known solution; it does not establish a sample-only recipe. Historical files
labelled `default` describe the then-used constructor and settings, not a
stable or recommended configuration. New ordinary probes name these settings
explicitly.

### What happened to fold / pole / fine?

There is no inferred fold/pole/fine partition for the random-draw batches.
The later deterministic geometry witnesses above supply known cases instead. The retained chart flag is
proximity to the requested marker-chart boundary, at tolerances 1e-4, 1e-6 and
1e-8. Numerical accuracy, parameter extent and PSD admissibility are separate
flags. At 1e-6, all 56 development ML references, all 55 fresh ML references,
and all 60 development PSD references are representable. The remaining four
and five ML problems have no screened sphere reference and remain unresolved.
A representable candidate is not a proof of a global optimum or of good
behavior in every chart. The ordinary probe does not measure std.lv proximity.

Thus the recovered misses with finite representable references provide evidence
of start/optimization trouble. They do not support labelling those targets
marker poles. A near-boundary flag alone would only nominate a possible pole;
a fold would require separate evidence that the limiting covariance is not
attained in any relevant chart. Failed multistarts and an extreme endpoint
cannot supply that conclusion.

### Expansion to new model families

The expansion uses Gaussian data, weak first indicators, unequal residual
variances, and N=30/100 with 20 draws per cell. New families are one factor with
five indicators (including a negative loading), two correlated factors with
four indicators each (mixed loading signs), and a three-factor regression chain
with three indicators per factor. Families and recipes were fixed before this
run. These models expand structural scope but still have simple measurement
structure, no equality constraints, and no missing data.

```{r}
expanded <- read.csv(file.path("results", "expanded-families", "summary.csv"))
x <- expanded[expanded$backend == "both", c("design", "portfolio", "matches_best_observed")]
x <- reshape(x, idvar="design", timevar="portfolio", direction="wide")
x <- x[c("design", paste0("matches_best_observed.", c("canonical", "single_negative", "all_signed", "positive_control")))]
names(x) <- c("Family (40 problems)", "Canonical", "+ Single negatives", "+ All negative subsets", "+ Positive control")
tab(x)
```

All entries pool the same two backends and compare against the full portfolio's
best screened objective. Canonical plus every nonempty negative subset uses
2^k starts for k factors; retaining the positive spectral control uses 2^k+1.
That is 6, 10 and 18 fits per problem for these one-, two- and three-factor
families with both backends. The full expansion uses 1,360 fits; recorded fitting
and endpoint-assessment time totals about 54 seconds on this machine. The three-factor full portfolio takes about 21 times
the canonical fitting time. Every draw's coverage and timing are retained in
`expanded-families/coverage.csv`.

The positive control's unique improvement is one-factor N=30 draw 1: PORT
finds a screened objective of 0.190643 versus canonical's 0.199426. The lower
point is inadmissible by the substantive variance screen, but remains a valid
candidate for unconstrained ML. This also shows why starts are search choices,
not declarations of endpoint admissibility. PORT alone reaches all 118 expanded
references, versus 117 for L-BFGS; this small result does not select a backend.

The two unresolved problems differ. One-factor N=30 draw 10 has locally screened
endpoints excluded only by the extent rule. Three-factor N=100 draw 6 has no
endpoint passing the full local screen. Neither is labelled nonexistent. Across
the expansion, 97 references have support from multiple start recipes and 21
from just one; the latter merit inspection before stronger reference claims.

### How often a reference is found

All datasets remain in the denominator. ML and PSD references are built
separately, using sphere attempts only. Repetition means different start
recipes reached the best observed objective; two optimizers from the same
start do not suffice.

```{r}
a <- aggregate(list(Datasets = refs$rep), refs[c("design", "domain", "reference_label")], length)
names(a)[1:3] <- c("Design", "Domain", "Reference evidence")
tab(a)
```

### Which sphere configurations reach the reference

The same random starting points are supplied to each configuration on a draw.
Each configuration has canonical and layered starts, plus
`r metadata$value[metadata$key == "random_starts"]` random starts.
The random starts have positive disturbances but varying loading signs and
latent paths. Their limited distribution does not exhaust possible basins.
The table pools starts and the two sample sizes within each design, so its
attempts are repeated fits, not independent datasets. Screening time is included.

```{r}
z <- s[s$route == "sphere", ]
z$screened <- z$fits * (z$label == "screened_candidate")
z$matched <- z$fits * (z$comparison == "matches_sphere_reference")
a <- aggregate(cbind(fits, screened, matched) ~ design + domain + backend + preconditioning, z, sum)
names(a) <- c("Design", "Domain", "Optimizer", "PSD scaling", "Attempts", "Screened", "At reference")
tab(a)
```

### Ordinary fits compared with the sphere reference

These attempts never define the sphere reference. A better ordinary endpoint is
retained as evidence that the sphere search missed something; the reference is
not silently replaced. Every finite returned ordinary endpoint is checked in
the same strongest-indicator chart. Its original convergence flag is retained
separately in the raw results.

```{r}
z <- s[s$route != "sphere", ]
a <- aggregate(fits ~ domain + route + backend + comparison, z, sum)
names(a) <- c("Domain", "Route", "Optimizer", "Comparison", "Attempts")
tab(a)
```

The ordinary PSD route remains ordinary PSD-ML. Sphere is an auxiliary
engineering reference: this study does not introduce sphere fitting into the
ordinary estimator, change paper protocols, or count a sphere rescue as an
ordinary-route success. A paper can report ordinary performance against a
best-observed numerical reference while describing that reference transparently.

### Purposeful starts: loading direction matters

The missed lavaan witness has a negative X-factor variance. The pilot's random
starts used positive latent variances. For this witness, merely making X's
starting variance negative fails; changing its loading direction to the
smallest-eigenvalue direction succeeds with either initial variance sign.
Thus the sign alone does not explain the miss. These are ML starts: negative
latent variances are outside the PSD estimator's domain.

The follow-up uses each factor's within-block correlation matrix with its
diagonal removed. A positive factor uses its largest-eigenvalue direction;
a negative factor uses its smallest. Both start with standardized variance
magnitude 0.5, zero structural regression, and residual variances chosen to
preserve observed diagonal variances. The implied observed covariance is
positive definite for either sign. The first follow-up used only the original
two-factor design. The expansion applies the same construction to other simple
structures; extension to general SEM remains open.

Four spectral recipes (both positive, negative X, negative Y, both negative)
were tested with canonical starts under both ML backends. The recipes were
fixed before the fresh run; the recommendation to drop the all-positive recipe
is a retrospective simplification, still needing confirmation. Fresh means new
random draws of the same three populations, not held-out model families.

```{r}
start_tables <- lapply(c("development", "fresh"), function(phase) {
  path <- file.path("results", paste0("start-", phase), "coverage_summary.csv")
  if (!file.exists(path)) stop("Missing start evidence: run scripts/tune_starts.R and scripts/summarize_starts.R")
  x <- read.csv(path)
  x <- x[x$backend == "both", ]
  aggregate(x[c("matches_best_observed", "attempts")], x[c("phase", "portfolio")], sum)
})
x <- do.call(rbind, start_tables)
names(x) <- c("Batch (60 problems)", "Start portfolio", "Best observed reached", "Fits")
tab(x)
```

Coverage means at least one endpoint matches the best screened objective in
that batch's reference pool. Development references additionally retain the
original pilot candidates; fresh references use only the new portfolio. The
canonical figure is therefore 45/60 on development against the expanded pool,
versus 46 against the older pilot pool. Adding both single-negative recipes
reaches 54 development and 52 fresh references; the double-negative recipe
adds two and three, respectively. Neither backend dominates: each alone reaches
54/60 fresh references with the full portfolio; together they reach 55/60.
The retained `coverage.csv` gives every paired case, backend and fit cost.

Three of the original nine missing references already had endpoints passing
the local checks but excluded by the extent screen. All three remain excluded.
Of the other six, five now have a screened candidate, leaving weak-marker
N=20 draw 4 unresolved. The development reference pool now has 50 repeated,
six single-start and four missing references. The fresh pool has 38 repeated,
17 single-start and five missing references. Supporting starts are recipes,
not independent replications; agreement by two backends alone is not repetition.

A canonical fit can pass the screen and still miss a better observed solution.
Therefore a retry triggered only by failed convergence would miss some of this
benefit. Keep the small portfolio as an exploratory reference search before
trying to invent an adaptive retry rule. No additional optimizer or termination
machinery was needed for these recoveries.

## Caveats

- This is exploratory development data, not a held-out default-selection run.
  The initial pilot has three two-factor models, N = 20 and 100, ten draws per
  cell, complete Gaussian data, one group and native units. Both latents must
  be covered by the sphere transformation. No conclusions extend to uncovered
  structures, FIML, DWLS or barrier fitting.
- Fits use unit-free sphere coordinates with user-chart polishing disabled.
  The current route still finalizes in the original chart when possible; a
  finalization error can therefore hide an otherwise usable sphere endpoint.
  Such attempts remain errors, never evidence that no solution exists.
- The strongest-indicator assessment is a local Newton cross-check, not an
  intrinsic sphere certificate or an identification proof. PSD checks use the
  PSD domain. Original-chart verdicts do not decide the reference screen.
- A first-order sphere check, norm residual at most 1e-6, finite consistent
  objectives, and the local Newton budget .01 must pass. PSD also requires
  covariance admissibility. These are an explicit exploratory screen, not a
  replacement library verdict. Nonpositive curvature, unavailable accuracy and
  excessive Newton distance are recorded separately in the raw evidence.
- Standardized extent above 10 is an **extreme-parameter screen**. Such points
  remain recorded but are excluded from the main reference pool. They are not
  called runaways. A finite improper optimum can be ML's best answer; being
  improper alone never excludes an ML candidate. A validated escape diagnostic
  and threshold-sensitivity study remain open.
- Objective agreement uses 1e-6 times one plus the absolute reference objective.
  Equal objectives need not imply equal parameters or the same basin. The
  number of observed objective levels is not a count of local minima.
- The marker diagnostic is evaluated at each endpoint, not at a known optimum.
  A level below 1e-6 means numerical proximity to the requested chart's boundary,
  not a proven zero loading. Levels and classifications at 1e-4, 1e-6 and 1e-8
  are retained. No chart condition is counted as convergence on its own.
- Sphere may remove identification poles while other parameter escape paths
  persist. Neither unsuccessful multistarts nor an ML objective below the PSD
  objective proves nonattainment.

## Reproduce

The high-precision diagnostic scripts require Python with `mpmath`; they are
study-local checks of the six-indicator, two-factor covariance model only.

```sh
cd experiments/engineering/15-sphere-reference-fits
Rscript scripts/check_labels.R
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript scripts/probe_reference_gap.R reference-pilot
Rscript scripts/tune_starts.R --phase witness
Rscript scripts/tune_starts.R --phase development
Rscript scripts/tune_starts.R --phase fresh
Rscript scripts/direct_starts.R
Rscript scripts/probe_geometry.R
Rscript scripts/check_geometry.R
Rscript scripts/psd_fresh.R
Rscript scripts/inspect_unresolved.R
Rscript scripts/check_escape.R
Rscript scripts/translate_sphere.R
Rscript scripts/check_requested_chart.R
Rscript scripts/export_open_cases.R
python3 scripts/refine_open_cases.py
python3 scripts/profile_delicate.py
python3 scripts/resolve_profile_minimum.py
python3 scripts/resolve_profile_minimum.py --digits 90 --run-id delicate-finite-minimum-90
python3 scripts/check_open_cases.py
Rscript scripts/check_open_case_translation.R
# Optional reference-informed diagnostic; requires retained raw pilot/start fits:
Rscript scripts/probe_ordinary.R --run-id ordinary-explicit
Rscript scripts/summarize_starts.R
Rscript scripts/check_expansion.R
Rscript scripts/expand_starts.R --smoke
Rscript scripts/expand_starts.R
quarto render report.qmd
# Optional small paired unit check; a distinct run, not default-selection evidence:
Rscript run_experiment.R --smoke --transforms native,x0.01,x100 --no-lavaan --run-id reference-units
```

The runner refuses to overwrite a completed run; choose a fresh `--run-id` for
reruns and render it with `quarto render report.qmd -P run:NAME`. The seed base
is separate for smoke and pilot profiles. Raw fits, endpoint parameter tables,
random starts and comparison rows stay local under `results/`. Small pilot
summaries, per-problem references, paired start coverage and metadata are frozen with the source so
the report can be rendered without rerunning fits. The start-follow-up sections
always show the frozen development/fresh and expanded-family batches. Witness and development
runners require the original pilot raw fits, regenerated by the pilot command;
the fresh runner does not. Metadata identifies the
installed package and source checkout; a dirty checkout is recorded explicitly.

## Label contract

Labels describe the evidence collected, never a mathematical existence claim.
The raw columns retain the ingredients so a later screen can be compared without
rewriting the original run.

| Label | Meaning | Does not imply |
| --- | --- | --- |
| `screened_candidate` | Endpoint passes the stated numerical and extent screen | A certified local minimum, identification, or global optimality |
| `screened_extreme` | Local checks pass, but standardized extent exceeds 10 or is undefined | A proven runaway; a finite optimum is impossible |
| `accuracy_failed` | Available Newton distance exceeds its budget | No better start or backend can solve the problem |
| `accuracy_unresolved` | Curvature/conditioning does not support the requested accuracy assessment; consult `audit_status` | Every unresolved case has the same mechanism |
| `sphere_first_order_failed` / `sphere_norm_failed` | The driven first-order or unit-norm check fails | The endpoint is a local optimum at a pole |
| `partial_sphere` | At least one relevant latent did not use the sphere | This attempt is a pole-free reference |
| `nonfinite_objective` / `objective_disagreement` / `psd_infeasible` | A finite, consistently evaluated point in the requested domain was not established | A valid reference candidate |
| `assessment_unavailable` / `call_error` | Endpoint assessment or fitting could not return the required evidence | Nonexistence |
| `single_start_best_observed` | Best screened sphere objective has one supporting start recipe | An unreliable point by definition, or a unique minimum |
| `repeated_best_observed` | At least two start recipes reach that objective within tolerance | Independent replications or exhaustive search |
| `no_screened_reference` | The finite portfolio supplies no qualifying reference | An unattainable problem |
| `near_chart_boundary` | Endpoint's original marker-chart level is below the declared tolerance | A mathematically exact pole or failed sphere optimization |
| `better_than_sphere_reference` | Screened ordinary endpoint improves on the sphere reference | The ordinary endpoint is globally optimal |
| `above_sphere_reference` | Screened endpoint has a worse objective | It lies in a distinct basin; convergence failed |
| `matches_sphere_reference` | Screened endpoint agrees in objective within tolerance | Equality of parameter estimates |
| `not_screened` / `no_sphere_reference` | Comparison lacks a qualifying endpoint or a sphere reference | Inferiority to a known optimum |

## Current choice and next step

The current engineering setup is a small sphere portfolio, strongest-indicator
cross-check, and separate original-chart diagnostics. Alternatives compared are
L-BFGS versus PORT for ML, and diagonal versus no PSD scaling with SLSQP.
A configuration deserves a larger follow-up only if it repeatedly finds the
best screened candidates without concentrating unresolved outcomes in a design;
this pilot does not select a library default.

The targeted witness follow-up is selected after inspecting the pilot; it is
not part of the original reference pool. Both recoveries are recorded in
`witness_probe.csv`. The unit smoke run separately retains paired rescalings;
it is too small to establish general unit invariance.

Retain the positive spectral control alongside canonical and signed alternatives:
the new-family run refutes dropping it universally. The exhaustive sign portfolio
remains an exploratory reference tool for at most three factors. Before adding
more starts, inspect the 21 single-start references and the two unresolved cases.
Developing a cheaper selection rule is open; changing only one factor at a time
already loses ten references here. A sphere-native assessment remains necessary for stronger local
accuracy claims. Preserve the existing
ordinary-parameterization comparisons for PSD work. Any production promotion
requires a fresh decision study with criteria written before its run.

This study was renamed from `15-sphere-local-convergence` on 2026-09-27.
Historical `results/full` and `results/smoke` are preserved but are not read by
this report. The former `no_estimate` classification combined failed alternative
fits with an objective below PSD; it was heuristic and must not be used as
proof of nonexistence. Historical sources remain available in Git history.
