---
title: "Sphere reference fits: a small exploratory pilot"
params:
  run: reference-pilot
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
read_result <- function(name) {
  path <- file.path("results", params$run, paste0(name, ".csv"))
  if (!file.exists(path)) stop("Missing results: run Rscript run_experiment.R --pilot")
  read.csv(path, stringsAsFactors = FALSE)
}
s <- read_result("summary"); refs <- read_result("references"); metadata <- read_result("metadata")
probe_path <- file.path("results", params$run, "witness_probe.csv")
probe <- if (file.exists(probe_path)) read.csv(probe_path) else NULL
count <- function(domain, label) sum(refs$domain == domain & refs$reference_label == label)
tab <- function(x) knitr::kable(x, row.names = FALSE, digits = 3)
```

## Question

Can a small collection of sphere fits provide useful reference solutions for
ML and PSD-ML, while keeping local numerical accuracy, chart problems and
uncertainty about the best solution separate? Can those references explain
ordinary-route failures without changing the estimator used in a paper?

## Short answer

Purposeful starts help beyond the original two-factor examples. In 120 fresh
datasets from three new model families, canonical starts reach the best
screened objective on 93 problems; the expanded portfolio reaches it on 118.
The largest gain is in the three-factor chain: 24/40 to 39/40. These are best
observed candidates within the tested portfolio, not certified optima.

The expansion also reverses one proposed simplification: the all-positive
spectral start contributes a better candidate on one problem. Keep it for now.
Single-factor negative alternatives miss ten references found by combined sign
patterns. Exhaustive combinations are useful for this small reference study,
but their cost grows exponentially with the factor count; this is not yet a
production retry policy.

The earlier pilot and start-development results remain below as separate
batches. All screens still use a strongest-indicator marker chart for the local
Newton cross-check; a sphere-native curvature assessment remains missing.
Two expanded problems have no screened reference. That does not establish
nonexistence or a pole. No library default changes here.

```{r results='asis'}
if (!is.null(probe)) {
  p <- probe[1, ]
  cat(sprintf(paste0("The targeted follow-up supplies a concrete counterexample to calling failed multistarts ",
    "\"no estimate\": on %s draw %d at N=%d, lavaan finds a screened objective of %.6f. ",
    "%d sphere backend(s) recover a screened candidate when started there. ",
    "The original finite portfolio missed a point the sphere route can represent.\n"),
    p$design, p$rep, p$n, p$witness_objective, sum(probe$label == "screened_candidate")))
}
```

## Evidence

### What a spectral start does

A start is an initial guess for every parameter. Each indicator block supplies
a correlation pattern. The positive recipe chooses a loading direction that
captures positive shared variation. The negative recipe chooses an alternative
direction suited to a negative latent variance. Changing direction changes the
relative loadings. The construction adjusts residual variances so the implied
observed covariance is positive definite at the start.

Each factor receives one of those two guesses. Regressions and between-factor
covariances start at zero. The optimizer then estimates all free parameters
normally: the initial signs are not constraints, and every recipe targets the
same ML objective. Negative latent variances can be allowed by ML while the
observed covariance remains positive definite; such endpoints may be
substantively inadmissible and are flagged separately. These recipes are not
being proposed for PSD estimation.

### Expansion to new model families

The expansion uses Gaussian data, weak first indicators, unequal residual
variances, and N=30/100 with 20 draws per cell. New families are one factor with
five indicators (including a negative loading), two correlated factors with
four indicators each (mixed loading signs), and a three-factor regression chain
with three indicators per factor. Families and recipes were fixed before this
run. These models expand structural scope but still have simple measurement
structure, no equality constraints, and no missing data.

```{r}
expanded <- read.csv(file.path("results", "expanded-families", "summary.csv"))
x <- expanded[expanded$backend == "both", c("design", "portfolio", "matches_best_observed")]
x <- reshape(x, idvar="design", timevar="portfolio", direction="wide")
x <- x[c("design", paste0("matches_best_observed.", c("canonical", "single_negative", "all_signed", "positive_control")))]
names(x) <- c("Family (40 problems)", "Canonical", "+ Single negatives", "+ All negative subsets", "+ Positive control")
tab(x)
```

All entries pool the same two backends and compare against the full portfolio's
best screened objective. Canonical plus every nonempty negative subset uses
2^k starts for k factors; retaining the positive spectral control uses 2^k+1.
That is 6, 10 and 18 fits per problem for these one-, two- and three-factor
families with both backends. The full expansion uses 1,360 fits; recorded fitting
and endpoint-assessment time totals about 54 seconds on this machine. The three-factor full portfolio takes about 21 times
the canonical fitting time. Every draw's coverage and timing are retained in
`expanded-families/coverage.csv`.

The positive control's unique improvement is one-factor N=30 draw 1: PORT
finds a screened objective of 0.190643 versus canonical's 0.199426. The lower
point is inadmissible by the substantive variance screen, but remains a valid
candidate for unconstrained ML. This also shows why starts are search choices,
not declarations of endpoint admissibility. PORT alone reaches all 118 expanded
references, versus 117 for L-BFGS; this small result does not select a backend.

The two unresolved problems differ. One-factor N=30 draw 10 has locally screened
endpoints excluded only by the extent rule. Three-factor N=100 draw 6 has no
endpoint passing the full local screen. Neither is labelled nonexistent. Across
the expansion, 97 references have support from multiple start recipes and 21
from just one; the latter merit inspection before stronger reference claims.

### How often a reference is found

All datasets remain in the denominator. ML and PSD references are built
separately, using sphere attempts only. Repetition means different start
recipes reached the best observed objective; two optimizers from the same
start do not suffice.

```{r}
a <- aggregate(list(Datasets = refs$rep), refs[c("design", "domain", "reference_label")], length)
names(a)[1:3] <- c("Design", "Domain", "Reference evidence")
tab(a)
```

### Which sphere configurations reach the reference

The same random starting points are supplied to each configuration on a draw.
Each configuration has canonical and layered starts, plus
`r metadata$value[metadata$key == "random_starts"]` random starts.
The random starts have positive disturbances but varying loading signs and
latent paths. Their limited distribution does not exhaust possible basins.
The table pools starts and the two sample sizes within each design, so its
attempts are repeated fits, not independent datasets. Screening time is included.

```{r}
z <- s[s$route == "sphere", ]
z$screened <- z$fits * (z$label == "screened_candidate")
z$matched <- z$fits * (z$comparison == "matches_sphere_reference")
a <- aggregate(cbind(fits, screened, matched) ~ design + domain + backend + preconditioning, z, sum)
names(a) <- c("Design", "Domain", "Optimizer", "PSD scaling", "Attempts", "Screened", "At reference")
tab(a)
```

### Ordinary fits compared with the sphere reference

These attempts never define the sphere reference. A better ordinary endpoint is
retained as evidence that the sphere search missed something; the reference is
not silently replaced. Every finite returned ordinary endpoint is checked in
the same strongest-indicator chart. Its original convergence flag is retained
separately in the raw results.

```{r}
z <- s[s$route != "sphere", ]
a <- aggregate(fits ~ domain + route + backend + comparison, z, sum)
names(a) <- c("Domain", "Route", "Optimizer", "Comparison", "Attempts")
tab(a)
```

The ordinary PSD route remains ordinary PSD-ML. Sphere is an auxiliary
engineering reference: this study does not introduce sphere fitting into the
ordinary estimator, change paper protocols, or count a sphere rescue as an
ordinary-route success. A paper can report ordinary performance against a
best-observed numerical reference while describing that reference transparently.

### Purposeful starts: loading direction matters

The missed lavaan witness has a negative X-factor variance. The pilot's random
starts used positive latent variances. For this witness, merely making X's
starting variance negative fails; changing its loading direction to the
smallest-eigenvalue direction succeeds with either initial variance sign.
Thus the sign alone does not explain the miss. These are ML starts: negative
latent variances are outside the PSD estimator's domain.

The follow-up uses each factor's within-block correlation matrix with its
diagonal removed. A positive factor uses its largest-eigenvalue direction;
a negative factor uses its smallest. Both start with standardized variance
magnitude 0.5, zero structural regression, and residual variances chosen to
preserve observed diagonal variances. The implied observed covariance is
positive definite for either sign. The first follow-up used only the original
two-factor design. The expansion applies the same construction to other simple
structures; extension to general SEM remains open.

Four spectral recipes (both positive, negative X, negative Y, both negative)
were tested with canonical starts under both ML backends. The recipes were
fixed before the fresh run; the recommendation to drop the all-positive recipe
is a retrospective simplification, still needing confirmation. Fresh means new
random draws of the same three populations, not held-out model families.

```{r}
start_tables <- lapply(c("development", "fresh"), function(phase) {
  path <- file.path("results", paste0("start-", phase), "coverage_summary.csv")
  if (!file.exists(path)) stop("Missing start evidence: run scripts/tune_starts.R and scripts/summarize_starts.R")
  x <- read.csv(path)
  x <- x[x$backend == "both", ]
  aggregate(x[c("matches_best_observed", "attempts")], x[c("phase", "portfolio")], sum)
})
x <- do.call(rbind, start_tables)
names(x) <- c("Batch (60 problems)", "Start portfolio", "Best observed reached", "Fits")
tab(x)
```

Coverage means at least one endpoint matches the best screened objective in
that batch's reference pool. Development references additionally retain the
original pilot candidates; fresh references use only the new portfolio. The
canonical figure is therefore 45/60 on development against the expanded pool,
versus 46 against the older pilot pool. Adding both single-negative recipes
reaches 54 development and 52 fresh references; the double-negative recipe
adds two and three, respectively. Neither backend dominates: each alone reaches
54/60 fresh references with the full portfolio; together they reach 55/60.
The retained `coverage.csv` gives every paired case, backend and fit cost.

Three of the original nine missing references already had endpoints passing
the local checks but excluded by the extent screen. All three remain excluded.
Of the other six, five now have a screened candidate, leaving weak-marker
N=20 draw 4 unresolved. The development reference pool now has 50 repeated,
six single-start and four missing references. The fresh pool has 38 repeated,
17 single-start and five missing references. Supporting starts are recipes,
not independent replications; agreement by two backends alone is not repetition.

A canonical fit can pass the screen and still miss a better observed solution.
Therefore a retry triggered only by failed convergence would miss some of this
benefit. Keep the small portfolio as an exploratory reference search before
trying to invent an adaptive retry rule. No additional optimizer or termination
machinery was needed for these recoveries.

## Caveats

- This is exploratory development data, not a held-out default-selection run.
  The initial pilot has three two-factor models, N = 20 and 100, ten draws per
  cell, complete Gaussian data, one group and native units. Both latents must
  be covered by the sphere transformation. No conclusions extend to uncovered
  structures, FIML, DWLS or barrier fitting.
- Fits use unit-free sphere coordinates with user-chart polishing disabled.
  The current route still finalizes in the original chart when possible; a
  finalization error can therefore hide an otherwise usable sphere endpoint.
  Such attempts remain errors, never evidence that no solution exists.
- The strongest-indicator assessment is a local Newton cross-check, not an
  intrinsic sphere certificate or an identification proof. PSD checks use the
  PSD domain. Original-chart verdicts do not decide the reference screen.
- A first-order sphere check, norm residual at most 1e-6, finite consistent
  objectives, and the local Newton budget .01 must pass. PSD also requires
  covariance admissibility. These are an explicit exploratory screen, not a
  replacement library verdict. Nonpositive curvature, unavailable accuracy and
  excessive Newton distance are recorded separately in the raw evidence.
- Standardized extent above 10 is an **extreme-parameter screen**. Such points
  remain recorded but are excluded from the main reference pool. They are not
  called runaways. A finite improper optimum can be ML's best answer; being
  improper alone never excludes an ML candidate. A validated escape diagnostic
  and threshold-sensitivity study remain open.
- Objective agreement uses 1e-6 times one plus the absolute reference objective.
  Equal objectives need not imply equal parameters or the same basin. The
  number of observed objective levels is not a count of local minima.
- The marker diagnostic is evaluated at each endpoint, not at a known optimum.
  A level below 1e-6 means numerical proximity to the requested chart's boundary,
  not a proven zero loading. Levels and classifications at 1e-4, 1e-6 and 1e-8
  are retained. No chart condition is counted as convergence on its own.
- Sphere may remove identification poles while other parameter escape paths
  persist. Neither unsuccessful multistarts nor an ML objective below the PSD
  objective proves nonattainment.

## Reproduce

```sh
cd experiments/engineering/15-sphere-reference-fits
Rscript scripts/check_labels.R
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript scripts/probe_reference_gap.R reference-pilot
Rscript scripts/tune_starts.R --phase witness
Rscript scripts/tune_starts.R --phase development
Rscript scripts/tune_starts.R --phase fresh
Rscript scripts/summarize_starts.R
Rscript scripts/check_expansion.R
Rscript scripts/expand_starts.R --smoke
Rscript scripts/expand_starts.R
quarto render report.qmd
# Optional small paired unit check; a distinct run, not default-selection evidence:
Rscript run_experiment.R --smoke --transforms native,x0.01,x100 --no-lavaan --run-id reference-units
```

The runner refuses to overwrite a completed run; choose a fresh `--run-id` for
reruns and render it with `quarto render report.qmd -P run:NAME`. The seed base
is separate for smoke and pilot profiles. Raw fits, endpoint parameter tables,
random starts and comparison rows stay local under `results/`. Small pilot
summaries, per-problem references, paired start coverage and metadata are frozen with the source so
the report can be rendered without rerunning fits. The start-follow-up sections
always show the frozen development/fresh and expanded-family batches. Witness and development
runners require the original pilot raw fits, regenerated by the pilot command;
the fresh runner does not. Metadata identifies the
installed package and source checkout; a dirty checkout is recorded explicitly.

## Label contract

Labels describe the evidence collected, never a mathematical existence claim.
The raw columns retain the ingredients so a later screen can be compared without
rewriting the original run.

| Label | Meaning | Does not imply |
| --- | --- | --- |
| `screened_candidate` | Endpoint passes the stated numerical and extent screen | A certified local minimum, identification, or global optimality |
| `screened_extreme` | Local checks pass, but standardized extent exceeds 10 or is undefined | A proven runaway; a finite optimum is impossible |
| `accuracy_failed` | Available Newton distance exceeds its budget | No better start or backend can solve the problem |
| `accuracy_unresolved` | Curvature/conditioning does not support the requested accuracy assessment; consult `audit_status` | Every unresolved case has the same mechanism |
| `sphere_first_order_failed` / `sphere_norm_failed` | The driven first-order or unit-norm check fails | The endpoint is a local optimum at a pole |
| `partial_sphere` | At least one relevant latent did not use the sphere | This attempt is a pole-free reference |
| `nonfinite_objective` / `objective_disagreement` / `psd_infeasible` | A finite, consistently evaluated point in the requested domain was not established | A valid reference candidate |
| `assessment_unavailable` / `call_error` | Endpoint assessment or fitting could not return the required evidence | Nonexistence |
| `single_start_best_observed` | Best screened sphere objective has one supporting start recipe | An unreliable point by definition, or a unique minimum |
| `repeated_best_observed` | At least two start recipes reach that objective within tolerance | Independent replications or exhaustive search |
| `no_screened_reference` | The finite portfolio supplies no qualifying reference | An unattainable problem |
| `near_chart_boundary` | Endpoint's original marker-chart level is below the declared tolerance | A mathematically exact pole or failed sphere optimization |
| `better_than_sphere_reference` | Screened ordinary endpoint improves on the sphere reference | The ordinary endpoint is globally optimal |
| `above_sphere_reference` | Screened endpoint has a worse objective | It lies in a distinct basin; convergence failed |
| `matches_sphere_reference` | Screened endpoint agrees in objective within tolerance | Equality of parameter estimates |
| `not_screened` / `no_sphere_reference` | Comparison lacks a qualifying endpoint or a sphere reference | Inferiority to a known optimum |

## Current choice and next step

The current engineering setup is a small sphere portfolio, strongest-indicator
cross-check, and separate original-chart diagnostics. Alternatives compared are
L-BFGS versus PORT for ML, and diagonal versus no PSD scaling with SLSQP.
A configuration deserves a larger follow-up only if it repeatedly finds the
best screened candidates without concentrating unresolved outcomes in a design;
this pilot does not select a library default.

The targeted witness follow-up is selected after inspecting the pilot; it is
not part of the original reference pool. Both recoveries are recorded in
`witness_probe.csv`. The unit smoke run separately retains paired rescalings;
it is too small to establish general unit invariance.

Retain the positive spectral control alongside canonical and signed alternatives:
the new-family run refutes dropping it universally. The exhaustive sign portfolio
remains an exploratory reference tool for at most three factors. Before adding
more starts, inspect the 21 single-start references and the two unresolved cases.
Developing a cheaper selection rule is open; changing only one factor at a time
already loses ten references here. A sphere-native assessment remains necessary for stronger local
accuracy claims. Preserve the existing
ordinary-parameterization comparisons for PSD work. Any production promotion
requires a fresh decision study with criteria written before its run.

This study was renamed from `15-sphere-local-convergence` on 2026-09-27.
Historical `results/full` and `results/smoke` are preserved but are not read by
this report. The former `no_estimate` classification combined failed alternative
fits with an objective below PSD; it was heuristic and must not be used as
proof of nonexistence. Historical sources remain available in Git history.
