---
title: "How fast are ML and PSD?"
subtitle: "Normalization off versus on · 27 September 2026"
format:
  html:
    toc: false
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r}
p <- 'results/normalization-timing/'
x <- read.csv(paste0(p,'overall.csv'))
r <- read.csv(paste0(p,'round_totals.csv'))
methods <- c('ml','direct','cold','warm')
labels <- c('Ordinary ML','Direct PSD','Independent two-stage','Warm-start two-stage')
a <- x[!x$normalized,];b <- x[x$normalized,]
a <- a[match(methods,a$method),];b <- b[match(methods,b$method),]
direct <- b[b$method=='direct',]
ms <- function(v) sprintf('%.2f',v)
```

**These are millisecond-scale fits.** With normalization enabled, a typical
direct PSD fit takes **`r ms(direct$median_ms)` ms**; its mean is
**`r ms(direct$sum_ms/direct$cases)` ms** because a few difficult fits cost much
more. The earlier half-second figure was for **all 120 datasets**, not one fit.

## Normalization off → on

All entries below are **milliseconds per fit**. The median describes a typical
case; the mean includes the cost of difficult fits and failures.

```{r}
t <- data.frame(Method=labels,
 `Median off → on`=paste0(ms(a$median_ms),' → ',ms(b$median_ms)),
 `Mean off → on`=paste0(ms(a$sum_ms/a$cases),' → ',ms(b$sum_ms/b$cases)),
 `Total-time on/off`=paste0(sprintf('%.2f',b$sum_ms/a$sum_ms),'×'),check.names=FALSE)
knitr::kable(t,row.names=FALSE)
```

**Normalization barely changes typical fit times.** Direct PSD's total cost
rises about 14%, but almost all of that increase is one previously certified
case that now exhausts its budget: 1.46 → 68.76 ms. It missed the saved target
in both settings. Combined time on cases certified in both settings is
essentially unchanged. Warm two-stage's larger slowdown is also dominated by
additional failures; ordinary ML's total cost falls about 5%.

For normalized direct PSD, 95% of these per-case times are below
**`r ms(direct$p95_ms)` ms**; the slowest is **`r ms(direct$max_ms)` ms**.
The six failed direct-PSD fits consume about **69% of its total time**.
The six-indicator, two-factor models are small. These numbers do not predict
runtime for a large SEM, multiple groups, missing data or inference.

Ordinary ML optimizes a different domain and can accept negative variances;
it is a speed reference, not a substitute for PSD. Its standalone start is
layered/native. Both two-stage routes use FABIN3-auto for ordinary ML, matching
the earlier policy comparison; independent fallback constructs PSD's own start.

**What this measures:** complete fits with normalization explicitly off/on,
including automatic starts, optimizer work, library diagnostics and R return
objects. Any timing difference can include a changed optimization path—not just
the arithmetic cost of scaling. Faster does not necessarily mean a better
solution; saved outcome counts accompany the timing data.

The same 120 challenge datasets were measured serially in four rounds, with
rotating route order and alternating normalization order. Three calls per block
give **11,520 timed policy calls**, after warmup. Each case's reported time is
the median of its four block averages. Model/data preparation, manual garbage
collection and outcome verification are outside the timer. Failures count.
Every measured block's final outcome matches the saved verdict and objective.

Hardware: Intel Core i7-1355U, single-threaded math. These are the current
R-accessible workflows; independent fallback uses up to two library calls,
while warm fallback uses native orchestration. No production behavior changed.

[Per-family results](results/normalization-timing/family.csv) ·
[Cost changes by verdict](results/normalization-timing/cost_changes.csv) ·
[Repeat totals](results/normalization-timing/round_totals.csv) ·
[Accuracy comparison](two-stage-summary.html) ·
[Reproduction](report.html#normalization-timing)
