---
title: "Does ordinary-first PSD save time?"
subtitle: "Repeated serial timings · 27 September 2026"
format:
  html:
    toc: false
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r}
p <- 'results/two-stage-timing/'
o <- read.csv(paste0(p,'overall.csv'))
f <- read.csv(paste0(p,'family.csv'))
s <- read.csv(paste0(p,'path.csv'))
r <- read.csv(paste0(p,'round_totals.csv'))
cost <- function(m) o$sum_ms[o$method==m]
ratio <- function(m) sprintf('%.2f',cost(m)/cost('direct'))
seconds <- function(m) sprintf('%.3f',cost(m)/1000)
```

**Direct PSD is fastest on this challenge set.** Independent two-stage takes
**`r ratio('cold')` times** its total fitting time; warm-start two-stage takes
**`r ratio('warm')` times**. Estimated time for one pass over the 120 datasets is
**`r seconds('direct')` s direct**, **`r seconds('cold')` s independent two-stage**,
and **`r seconds('warm')` s warm-start two-stage**. Failures are included.

This is a cost comparison, not a new accuracy test. Independent two-stage and
direct PSD reach the same 101 saved targets; warm-start two-stage reaches 100,
with different gains and losses. The times summarize repeated measurements,
not one unusually fast or slow pass.

## Where the time goes

**Mean milliseconds per dataset**, including failures. Each row contains 20
datasets; each dataset's cost is the median of six timed block averages.

```{r}
k <- expand.grid(design=c('ernst','weak_marker','high_r2'),batch=c('development','fresh'),stringsAsFactors=FALSE)
t <- data.frame(Batch=ifelse(k$batch=='development','Earlier','Later'),
 Problem=unname(c(ernst='Ernst',weak_marker='Weak markers',high_r2='High R²')[k$design]))
for(m in c('direct','cold','warm')) {
 z <- f[f$method==m,];z <- z[match(paste(k$batch,k$design),paste(z$batch,z$design)),]
 t[[switch(m,direct='Direct PSD',cold='Independent two-stage',warm='Warm two-stage')]] <- z$sum_ms/z$cases
}
knitr::kable(t,row.names=FALSE,digits=2)
```

The ordinary-first shortcut works when ordinary ML is accepted. But only
**34 of 120** datasets take that shortcut; the other **86** pay for both attempts.

```{r}
t <- data.frame(Path=c('Ordinary accepted (34)','PSD needed (86)'))
for(m in c('direct','cold','warm')) {
 z <- s[s$method==m,];z <- z[match(c(TRUE,FALSE),z$ordinary_accepted),]
 t[[switch(m,direct='Direct PSD',cold='Independent two-stage',warm='Warm two-stage')]] <- z$sum_ms/z$cases
}
knitr::kable(t,row.names=FALSE,digits=2,caption='Mean milliseconds per dataset, grouped by whether ordinary ML is accepted.')
```

**Practical conclusion:** independent fallback is a reasonable policy, but
it is not a speed optimization for this difficult problem mix. Use direct PSD
when this evidence is the guide; an ordinary-first speed claim needs a workload
where ordinary ML succeeds much more often or costs much less.

## How much to trust the timing

One serial R process, single-threaded math, on an Intel Core i7-1355U. Every
case/method is warmed up before measurement. Six rounds use all six method
orders, with shuffled case order; each timed block contains five calls.
That is **10,800 timed policy calls**. Samples/models are prepared beforehand;
garbage collection and outcome verification are outside timed blocks. Failed
fits count toward cost. Every block's final result is checked against the saved
verdict and objective.

The measurements include automatic starts, library diagnostics, returned R
objects and R orchestration. Independent fallback currently uses up to two
library calls; warm fallback uses one native orchestration call. These are
end-to-end timings of the available workflows, not an isolated C++ comparison.
They do not establish timings for a future integrated independent-fallback API,
larger models, or a representative user workload.

[Per-case timings](results/two-stage-timing/per_case.csv) ·
[Repeat totals](results/two-stage-timing/round_totals.csv) ·
[Accuracy comparison](two-stage-summary.html) ·
[Reproduction](report.html#two-stage-timing)
