---
title: "Testing Alpha-Omega Equality Is a Non-Regular Reliability-Gap Problem"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results <- "../results"
need <- c("nonregularity.csv", "simulation_summary.csv", "demo.csv",
          "validation_identity.csv", "validation_se.csv", "metadata.csv")
missing <- need[!file.exists(file.path(results, need))]
if (length(missing)) {
  stop("missing results: ", paste(missing, collapse = ", "),
       "\nRun from the experiment root: Rscript run_experiment.R")
}
nonreg <- read.csv(file.path(results, "nonregularity.csv"))
sim <- read.csv(file.path(results, "simulation_summary.csv"))
demo <- read.csv(file.path(results, "demo.csv"))
ident <- read.csv(file.path(results, "validation_identity.csv"))
seval <- read.csv(file.path(results, "validation_se.csv"))
meta <- read.csv(file.path(results, "metadata.csv"))
has_equal_wald <- "reject_wald_equal" %in% names(sim)
has_demo_equal_wald <- "p_wald_equal" %in% names(demo)
pct <- function(x) sprintf("%.1f%%", 100 * x)
mv <- function(k) meta$value[match(k, meta$key)]
```

## Abstract

Deng and Chan (2017) proposed a Wald test for the difference between
coefficient omega and coefficient alpha. We show that this test is non-regular
at the null hypothesis it is meant to test. In a one-factor model,
$\omega \ge \alpha$, with equality exactly at tau-equivalence, so the scalar
gap $\omega-\alpha$ has an interior minimum at the null and its first derivative
vanishes. The resulting statistic has a second-order quadratic-form limit, not a
normal limit. A short simulation shows the Deng-Chan Wald test is conservative
under tau-equivalence, while an Imhof calibration of the reliability gap and the
regular equal-loading Wald, likelihood-ratio, and score tests behave as expected.

## Problem

Coefficient alpha is the covariance-based reliability coefficient that is exact
for a tau-equivalent item sum. Coefficient omega is the one-factor/congeneric
reliability coefficient for the same total score.

For a unidimensional congeneric model, alpha equals omega only when the item
loadings are equal. Thus the hypothesis $\alpha=\omega$ is not merely a
numerical comparison between two reliability summaries; it is the
tau-equivalence restriction in disguise.

Deng and Chan's statistic tests the derived scalar gap
$\hat\omega-\hat\alpha$ with a first-order Wald reference. The regular
structural tests instead test the equal-loading restriction directly, with
$p-1$ degrees of freedom. These are different Wald problems: the scalar
reliability-gap Wald is singular at the null, while the equal-loading Wald test
is regular under the usual SEM conditions.

## Why the Scalar Wald Test Fails

On the correctly specified one-factor model,

$$
\omega-\alpha =
\frac{p\sum_{j=1}^p(\lambda_j-\bar\lambda)^2}
{(p-1)\{(\mathbf 1^\top\lambda)^2+\mathbf 1^\top\psi\}} .
$$

The numerator is loading dispersion. It is zero at tau-equivalence and
nonnegative nearby, so the null is an interior minimum of the derived
reliability gap. Consequently the first-order derivative of
$d(s)=\omega(s)-\alpha(s)$ is zero at the null.

The usual first-order delta method is therefore the wrong approximation. If
$\sqrt n(\hat s-s_0)\Rightarrow Z$ and $Z\sim N(0,\Gamma)$, then

$$
n\{\hat\omega-\hat\alpha\}\Rightarrow \frac{1}{2}Z^\top HZ,
$$

where $H$ is the Hessian of the fitted reliability gap with respect to the
sample covariance vector. This is a quadratic form in normals, usually written
as a weighted sum of independent $\chi^2_1$ variables.

## Numerical Check

```{r scaling-table}
sc <- nonreg[, c("N", "sd_diff", "sd_omega", "sd_diff_ratio",
                 "sd_omega_ratio", "corr_omega_alpha")]
names(sc) <- c("N", "SD(omega-alpha)", "SD(omega)",
               "SD gap ratio", "SD omega ratio", "corr(omega, alpha)")
knitr::kable(sc, digits = c(0, 5, 4, 2, 2, 4), row.names = FALSE,
             caption = "Tau-equivalent population. The reliability gap scales at the second-order 1/N rate, while omega itself scales at the ordinary 1/sqrt(N) rate.")
```

The diagnostic is the non-regularity in one table: when $N$ quadruples, the
standard deviation of the gap quarters, while the standard deviation of omega
roughly halves.

```{r test-table}
sim_cols <- c("population", "N", "reject_wald_sw", "reject_imhof_sw")
if (has_equal_wald) sim_cols <- c(sim_cols, "reject_wald_equal")
sim_cols <- c(sim_cols, "reject_satorra_scaled", "reject_score")
tab <- sim[sim$dist == "normal", sim_cols]
tab <- tab[order(match(tab$population, c("tau", "congeneric", "misspecified")),
                 tab$N), ]
names(tab) <- c("Population", "N", "Deng Wald", "Imhof gap",
                if (has_equal_wald) "Equal-loading Wald",
                "Satorra LR", "Score")
knitr::kable(tab, digits = 3, row.names = FALSE,
             caption = "Rejection rate at 5%, normal data. The tau rows are Type I error; the congeneric rows are power; the misspecified rows show that reliability-gap tests and structural tau-equivalence tests answer different questions.")
```

The Deng-Chan Wald test is conservative at the tau-equivalent null. The Imhof
test keeps the scalar reliability gap but uses the correct second-order
reference law. The regular Wald/LR/score tests target the structural
equal-loading restriction directly.

## Real-Data Illustration

```{r demo-table}
demo_cols <- c("scale", "p", "omega", "alpha", "diff", "p_imhof_sw")
if (has_demo_equal_wald) demo_cols <- c(demo_cols, "p_wald_equal")
demo_cols <- c(demo_cols, "p_satorra_scaled", "p_score")
dd <- demo[, demo_cols]
names(dd) <- c("Scale", "p", "Omega", "Alpha", "Omega - alpha", "Imhof p",
               if (has_demo_equal_wald) "Equal-loading Wald p",
               "Satorra p", "Score p")
knitr::kable(dd, digits = c(0, 0, 3, 3, 4, 3,
                            if (has_demo_equal_wald) 4, 4, 3),
             row.names = FALSE,
             caption = "Holzinger-Swineford subscales. Structural tests can reject tau-equivalence even when the reliability gap is only a few thousandths.")
```

## Recommendation

For a test of tau-equivalence, use the regular equal-loading Wald, LR, or score
test. These are the tests of the model restriction that makes alpha and omega
equal.

If the estimand of interest is specifically the scalar reliability gap, do not
use a normal Wald reference at the null. Calibrate the gap by its second-order
quadratic-form law, as in the Imhof test here, or by a null bootstrap.

## Appendix Sketch

The full appendix should contain four pieces.

1. Algebra showing that $\omega-\alpha$ is proportional to loading dispersion
   on the one-factor model.
2. A second-order delta-method derivation of the quadratic-form limit.
3. A derivation showing that Deng and Chan's estimated first-order SE is also
   second-order at the null, producing a non-pivotal Wald ratio.
4. Implementation checks: alpha equals the omega from a ULS tau-equivalent fit
   (`r sprintf("%.1e", ident$abs_diff[1])` absolute difference), and the
   sandwich SE agrees with a nonparametric bootstrap away from the null
   (`r sprintf("%.3f", seval$ratio_sw_boot[1])` ratio for
   `r seval$dataset[1]`).

## Citation Audit Before Submission

The first web pass did not find an obvious published correction using searches
for Deng-Chan plus non-regularity, Imhof, singularity, and second-order delta
method. This is not enough for a novelty claim. Before submission, screen the
citing-paper set from Google Scholar and one reproducible source such as
OpenAlex, Semantic Scholar, or Scopus for prior corrections or bootstrap-based
workarounds.
