---
title: "Ordinal construction boundary"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "construction.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

construction_path <- file.path(results_dir, "construction.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(construction_path)) {
  stop("Missing construction.csv. Run: Rscript run_experiment.R --smoke")
}

construction <- utils::read.csv(construction_path, check.names = FALSE,
                                stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.3f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.3g", x))
fmt_ratio <- function(x) ifelse(is.na(x), "", sprintf("%.3g", x))
median_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else stats::median(x)
}

operation_levels <- c(
  "lazy_uls_workspace_from_raw",
  "lazy_dwls_workspace_from_raw",
  "legacy_stats_full",
  "moments_projection_from_stats",
  "cache_diagonal_from_legacy_gamma",
  "cache_full_gamma_copy",
  "cache_from_stats_copy_all",
  "extract_diagonal_from_full_gamma",
  "dwls_weight_from_diagonal",
  "wls_weight_reinvert_full_gamma"
)
operation_labels <- c(
  "Lazy ULS workspace",
  "Lazy DWLS workspace",
  "Legacy full stats",
  "Project moments",
  "Copy diagonal cache",
  "Copy full Gamma",
  "Copy all from stats",
  "Extract diagonal from full",
  "DWLS weight from diagonal",
  "Reinvert WLS weight"
)
construction$Operation <- factor(construction$operation,
                                 levels = operation_levels,
                                 labels = operation_labels)
ok <- construction[construction$status == "ok", ]

wide <- reshape(
  ok[, c("design", "n", "p", "categories", "moment_dim", "gamma_mb",
         "diagonal_mb", "operation", "median_ms")],
  idvar = c("design", "n", "p", "categories", "moment_dim", "gamma_mb",
            "diagonal_mb"),
  timevar = "operation",
  direction = "wide"
)
names(wide) <- sub("^median_ms\\.", "", names(wide))
for (name in operation_levels) {
  if (!name %in% names(wide)) wide[[name]] <- NA_real_
}
wide$wls_reinvert_share <- wide$wls_weight_reinvert_full_gamma /
  wide$legacy_stats_full
wide$lazy_uls_share <- wide$lazy_uls_workspace_from_raw /
  wide$legacy_stats_full
wide$lazy_dwls_share <- wide$lazy_dwls_workspace_from_raw /
  wide$legacy_stats_full
wide$diag_cache_share <- wide$cache_diagonal_from_legacy_gamma /
  wide$legacy_stats_full
wide$moments_share <- wide$moments_projection_from_stats /
  wide$legacy_stats_full
wide$full_gamma_mb_per_diag_mb <- wide$gamma_mb / wide$diagonal_mb
```

## Question

The ordinal SNLLS speed pilot times fitting after the legacy ordinal statistic
builder has already run. This experiment asks what that hides, and now times
the lazy fit-only boundary alongside the legacy all-ordinal constructor:

- lazy raw construction for fit-only ULS, returning `OrdinalMoments` only;
- lazy raw construction for fit-only DWLS, returning `OrdinalMoments` plus the
  Gamma diagonal;
- `OrdinalStats` from raw integer categories, which currently estimates
  thresholds/polychorics, builds full `NACOV`, and materializes DWLS/WLS
  weights;
- projection from that full object to `OrdinalMoments`;
- diagonal/full Gamma cache copies and weight construction from already
  materialized legacy pieces.

## Short Answer

The lazy raw boundary is now real for fit-only ULS/DWLS. ULS stops after
thresholds and polychoric `R`; DWLS additionally computes the Gamma diagonal
without materializing the full square Gamma or WLS inverse. WLS and
fit-plus-inference still fall back to the eager legacy path.

This turns the earlier caveat into a measurable implementation split. The
remaining question is whether the ULS/DWLS construction savings are large
enough, relative to pairwise polychoric estimation itself, to matter in the
paper-facing end-to-end speed story.

## Run

```{r}
data.frame(
  Item = metadata$key,
  Value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Status

```{r}
status <- aggregate(seed ~ p + categories + Operation + status,
                    data = construction, FUN = length)
names(status) <- c("p", "Categories", "Operation", "Status", "Rows")
knitr::kable(status[order(status$p, status$Categories, status$Operation), ],
             row.names = FALSE)
```

## Timing

```{r}
timing <- ok[, c("p", "categories", "moment_dim", "Operation", "median_ms")]
names(timing) <- c("p", "Categories", "Moment dim", "Operation",
                   "Median ms")
timing$`Median ms` <- fmt_ms(timing$`Median ms`)
knitr::kable(timing[order(timing$p, timing$Categories, timing$Operation), ],
             row.names = FALSE)
```

These rows are not additive. They are boundary probes. For example, "Reinvert
WLS weight" starts from an already available full Gamma and isolates the cost
of inverting it again for a cache.

## Relative Costs

```{r}
summary <- wide[, c("p", "categories", "moment_dim", "legacy_stats_full",
                    "lazy_uls_share", "lazy_dwls_share", "moments_share",
                    "diag_cache_share",
                    "wls_reinvert_share")]
names(summary) <- c("p", "Categories", "Moment dim", "Legacy full stats ms",
                    "Lazy ULS / legacy", "Lazy DWLS / legacy",
                    "Moments / legacy", "Diagonal cache / legacy",
                    "WLS reinvert / legacy")
summary$`Legacy full stats ms` <- fmt_ms(summary$`Legacy full stats ms`)
summary$`Lazy ULS / legacy` <- fmt_ratio(summary$`Lazy ULS / legacy`)
summary$`Lazy DWLS / legacy` <- fmt_ratio(summary$`Lazy DWLS / legacy`)
summary$`Moments / legacy` <- fmt_ratio(summary$`Moments / legacy`)
summary$`Diagonal cache / legacy` <- fmt_ratio(summary$`Diagonal cache / legacy`)
summary$`WLS reinvert / legacy` <- fmt_ratio(summary$`WLS reinvert / legacy`)
knitr::kable(summary[order(summary$p, summary$Categories), ],
             row.names = FALSE)
```

Ratios are relative to the eager legacy `OrdinalStats` construction for the
same synthetic data block. The lazy rows start from raw categories and are the
actual construction alternatives for fit-only ULS/DWLS. The cache-copy rows
start after legacy stats already exist, so they remain boundary probes rather
than alternatives to raw construction.

```{r}
#| fig-width: 8
#| fig-height: 5
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  plot_data <- ok
  ggplot(plot_data, aes(moment_dim, median_ms, color = Operation)) +
    geom_line(aes(group = interaction(Operation, categories))) +
    geom_point(size = 2) +
    facet_wrap(~ categories, labeller = label_both, scales = "free_y") +
    scale_y_log10() +
    labs(x = "Moment dimension", y = "Median ms, log scale",
         color = "Operation") +
    theme_minimal(base_size = 12)
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

## Payload Size

```{r}
payload <- unique(wide[, c("p", "categories", "moment_dim", "gamma_mb",
                           "diagonal_mb", "full_gamma_mb_per_diag_mb")])
names(payload) <- c("p", "Categories", "Moment dim", "Full Gamma MB",
                    "Diagonal MB", "Full / diagonal bytes")
payload$`Full Gamma MB` <- fmt_num(payload$`Full Gamma MB`)
payload$`Diagonal MB` <- fmt_num(payload$`Diagonal MB`)
payload$`Full / diagonal bytes` <- fmt_num(payload$`Full / diagonal bytes`)
knitr::kable(payload[order(payload$p, payload$Categories), ],
             row.names = FALSE)
```

This is the memory side of the same boundary: fit-only ULS needs no Gamma,
fit-only DWLS needs only the Gamma diagonal, while the legacy constructor
always creates the full square matrix and its WLS inverse.

## Reproduce

```sh
Rscript experiments/13-ordinal-construction-boundary/run_experiment.R --smoke --preset fast
(cd experiments/13-ordinal-construction-boundary && quarto render report.qmd)
```

For the preliminary sweep used here, run with `--preset opt --reps 3 --max-p 16`.
