---
title: "Ordinal construction boundary"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "construction.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

construction_path <- file.path(results_dir, "construction.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(construction_path)) {
  stop("Missing construction.csv. Run: Rscript run_experiment.R --smoke")
}

construction <- utils::read.csv(construction_path, check.names = FALSE,
                                stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.3f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.3g", x))
fmt_ratio <- function(x) ifelse(is.na(x), "", sprintf("%.3g", x))
median_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else stats::median(x)
}

operation_levels <- c(
  "legacy_stats_full",
  "moments_projection_from_stats",
  "cache_diagonal_from_legacy_gamma",
  "cache_full_gamma_copy",
  "cache_from_stats_copy_all",
  "extract_diagonal_from_full_gamma",
  "dwls_weight_from_diagonal",
  "wls_weight_reinvert_full_gamma"
)
operation_labels <- c(
  "Legacy full stats",
  "Project moments",
  "Copy diagonal cache",
  "Copy full Gamma",
  "Copy all from stats",
  "Extract diagonal from full",
  "DWLS weight from diagonal",
  "Reinvert WLS weight"
)
construction$Operation <- factor(construction$operation,
                                 levels = operation_levels,
                                 labels = operation_labels)
ok <- construction[construction$status == "ok", ]

wide <- reshape(
  ok[, c("design", "n", "p", "categories", "moment_dim", "gamma_mb",
         "diagonal_mb", "operation", "median_ms")],
  idvar = c("design", "n", "p", "categories", "moment_dim", "gamma_mb",
            "diagonal_mb"),
  timevar = "operation",
  direction = "wide"
)
names(wide) <- sub("^median_ms\\.", "", names(wide))
for (name in operation_levels) {
  if (!name %in% names(wide)) wide[[name]] <- NA_real_
}
wide$wls_reinvert_share <- wide$wls_weight_reinvert_full_gamma /
  wide$legacy_stats_full
wide$diag_cache_share <- wide$cache_diagonal_from_legacy_gamma /
  wide$legacy_stats_full
wide$moments_share <- wide$moments_projection_from_stats /
  wide$legacy_stats_full
wide$full_gamma_mb_per_diag_mb <- wide$gamma_mb / wide$diagonal_mb
```

## Question

The ordinal SNLLS speed pilot times fitting after the legacy ordinal statistic
builder has already run. This experiment asks what that hides. It times the
current all-ordinal construction boundary:

- `OrdinalStats` from raw integer categories, which currently estimates
  thresholds/polychorics, builds full `NACOV`, and materializes DWLS/WLS
  weights;
- projection from that full object to `OrdinalMoments`;
- diagonal/full Gamma cache copies and weight construction from already
  materialized legacy pieces.

## Short Answer

The current public raw boundary is still eager. `ordinal_stats_from_integer_data`
materializes the full Gamma and WLS inverse before any fit path gets to choose
ULS, DWLS, WLS, profiled thresholds, or SNLLS. The cache-aware fitting code is
therefore proving that downstream consumers can avoid *using* full Gamma, but
not yet that raw construction can avoid *building* it.

That makes the next C++ target crisp: add a raw/lazy all-ordinal builder that
can return `OrdinalMoments` plus only the requested Gamma materialization. The
first useful version should support fit-only ULS/DWLS without full Gamma or WLS
inverse construction; WLS and robust inference can continue to request full
Gamma.

## Run

```{r}
data.frame(
  Item = metadata$key,
  Value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Status

```{r}
status <- aggregate(seed ~ p + categories + Operation + status,
                    data = construction, FUN = length)
names(status) <- c("p", "Categories", "Operation", "Status", "Rows")
knitr::kable(status[order(status$p, status$Categories, status$Operation), ],
             row.names = FALSE)
```

## Timing

```{r}
timing <- ok[, c("p", "categories", "moment_dim", "Operation", "median_ms")]
names(timing) <- c("p", "Categories", "Moment dim", "Operation",
                   "Median ms")
timing$`Median ms` <- fmt_ms(timing$`Median ms`)
knitr::kable(timing[order(timing$p, timing$Categories, timing$Operation), ],
             row.names = FALSE)
```

These rows are not additive. They are boundary probes. For example, "Reinvert
WLS weight" starts from an already available full Gamma and isolates the cost
of inverting it again for a cache.

## Relative Costs

```{r}
summary <- wide[, c("p", "categories", "moment_dim", "legacy_stats_full",
                    "moments_share", "diag_cache_share",
                    "wls_reinvert_share")]
names(summary) <- c("p", "Categories", "Moment dim", "Legacy full stats ms",
                    "Moments / legacy", "Diagonal cache / legacy",
                    "WLS reinvert / legacy")
summary$`Legacy full stats ms` <- fmt_ms(summary$`Legacy full stats ms`)
summary$`Moments / legacy` <- fmt_ratio(summary$`Moments / legacy`)
summary$`Diagonal cache / legacy` <- fmt_ratio(summary$`Diagonal cache / legacy`)
summary$`WLS reinvert / legacy` <- fmt_ratio(summary$`WLS reinvert / legacy`)
knitr::kable(summary[order(summary$p, summary$Categories), ],
             row.names = FALSE)
```

Ratios are relative to the eager legacy `OrdinalStats` construction for the
same synthetic data block. Values near zero do not mean those operations are
unimportant generally; they mean the current public raw boundary has already
paid the larger estimation and full-Gamma costs.

```{r}
#| fig-width: 8
#| fig-height: 5
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  plot_data <- ok
  ggplot(plot_data, aes(moment_dim, median_ms, color = Operation)) +
    geom_line(aes(group = interaction(Operation, categories))) +
    geom_point(size = 2) +
    facet_wrap(~ categories, labeller = label_both, scales = "free_y") +
    scale_y_log10() +
    labs(x = "Moment dimension", y = "Median ms, log scale",
         color = "Operation") +
    theme_minimal(base_size = 12)
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

## Payload Size

```{r}
payload <- unique(wide[, c("p", "categories", "moment_dim", "gamma_mb",
                           "diagonal_mb", "full_gamma_mb_per_diag_mb")])
names(payload) <- c("p", "Categories", "Moment dim", "Full Gamma MB",
                    "Diagonal MB", "Full / diagonal bytes")
payload$`Full Gamma MB` <- fmt_num(payload$`Full Gamma MB`)
payload$`Diagonal MB` <- fmt_num(payload$`Diagonal MB`)
payload$`Full / diagonal bytes` <- fmt_num(payload$`Full / diagonal bytes`)
knitr::kable(payload[order(payload$p, payload$Categories), ],
             row.names = FALSE)
```

This is the memory side of the same boundary: a fit-only DWLS path only needs
the Gamma diagonal, while the legacy constructor always creates the full square
matrix and its WLS inverse.

## Reproduce

```sh
Rscript experiments/13-ordinal-construction-boundary/run_experiment.R --smoke --preset fast
(cd experiments/13-ordinal-construction-boundary && quarto render report.qmd)
```

For the preliminary sweep used here, run with `--preset opt --reps 3 --max-p 16`.
