---
title: "A generic profile-LR confidence interval for the reliability family"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
res <- file.path("results", "semlbci_parity.csv")
if (!file.exists(res)) stop("Run: Rscript run_experiment.R  (then re-render)")
d <- read.csv(res, stringsAsFactors = FALSE)
fmt <- function(x, k = 4) formatC(x, format = "f", digits = k)
```

## Question

We want small-sample, range-respecting, transformation-invariant confidence
intervals for the whole reliability family (omega_total, omega_h, H, and their
kin) by inverting the likelihood-ratio test of `g(theta) = g0`. The first thing
to establish is that a *single generic engine*, which knows nothing about which
coefficient it is bounding, reproduces the established construction. The
reference is `semlbci` (the R implementation of the Wu-Neale / Pek-Wu
likelihood-based CI). Does one profile-LR engine, fed only a functional closure,
match `semlbci` across the family under normal-theory ML on complete data?

## Short answer

Yes. Across omega_total and H on a one-factor congeneric model and omega_total
and omega_h on an orthogonal three-group bifactor, wherever `semlbci` converges
the engine's interval reproduces its likelihood-based interval to `1e-7`. The
engine is the *same* code for all four coefficients: only the scalar functional
passed in changes. Two things went beyond a plain match. First, every engine
bound satisfies its own defining equation, `T(bound) = qchisq(.95, 1)`, to
machine tolerance. Second, on some bifactor upper bounds `semlbci`'s default
(Wu-Neale) search returns a bound where the true likelihood ratio is well below
the threshold (its constrained refit is stuck at a suboptimal optimum, and its
own post-check re-uses that stuck fit and so certifies the wrong value); the
engine's constrained fit is strictly better there and its bound is the one that
solves the equation. So this is the normal-theory foundation the small-sample
(Bartlett) and robust tiers build on, and the engine is at least as reliable as
the reference at the boundary.

## Evidence

```{r summary}
vtol <- 0.02
summ <- do.call(rbind, lapply(split(d, d$func), function(g) {
  ok_lo <- abs(g$tdev_sl_lo) <= vtol; ok_hi <- abs(g$tdev_sl_hi) <= vtol
  agree <- max(c(abs(g$d_lo[ok_lo]), abs(g$d_hi[ok_hi])), na.rm = TRUE)
  data.frame(
    Coefficient = g$label[1],
    Cells = nrow(g),
    `Engine bounds off eqn` = sum(abs(c(g$tdev_eng_lo, g$tdev_eng_hi)) > vtol, na.rm = TRUE),
    `semlbci bounds off eqn` = sum(abs(c(g$tdev_sl_lo, g$tdev_sl_hi)) > vtol, na.rm = TRUE),
    `Max diff (semlbci valid)` = formatC(agree, format = "e", digits = 2),
    check.names = FALSE)
}))
knitr::kable(summ[order(summ$Coefficient), ], row.names = FALSE,
             caption = "Per coefficient: number of bounds failing T=q (|T(bound)-qchisq(.95,1)|>0.02) for each method, and the worst engine-vs-semlbci bound difference over cells where semlbci's bound is valid. All seeds and sample sizes.")
```

Each cell is one simulated normal dataset; both methods invert the same
normal-theory ML likelihood ratio, so the comparison isolates the engine's
constrained-fit and bisection machinery from any modelling choice. The interval
is `{g0 : N (F_con(g0) - F_unc) <= q}` with `q = qchisq(0.95, 1)`; the constrained
fit imposes `g(theta) = g0` through a nonlinear equality constraint, so the bound
always stays inside the admissible reliability range. The "off eqn" columns count
bounds where the true likelihood ratio at the reported bound misses the threshold
by more than `0.02`, an independent, solver-agnostic check of correctness: the
engine misses on none, `semlbci` on a few bifactor upper bounds.

## Caveats

- Normal-theory only, complete data, correctly specified model. This run
  establishes construction parity, not coverage. The point of the project, the
  small-sample (Bartlett) rescaling that makes the bounded interval actually
  cover at the sample sizes reliability is reported at, is a separate step and is
  not exercised here.
- The oracle coefficients are written in `semlbci`'s `:=` algebra, which is why
  the functionals chosen here avoid a genuine matrix inverse (H on one factor
  uses the diagonal-residual reduction). The engine itself has no such
  restriction; it takes an opaque closure and can bound functionals `semlbci`
  cannot express.
- Small grid (a few seeds per cell). The agreement is a numerical-identity check,
  not a statistical estimate, so a large grid adds nothing.
- The `semlbci` boundary failures are a note about the reference implementation's
  default search on this harder model, not a coverage claim; they do not affect
  the one-factor cases, where both methods agree to `1e-7` on every bound.
- A few bifactor cells are dropped because the ML fit is improper (a badly
  negative residual variance, or non-convergence) at these sample sizes;
  reliability is undefined there and both methods skip them. Every retained cell
  is a proper solution.

## Reproduce

```sh
cd experiments/45-profile-lr-reliability-ci
Rscript run_experiment.R --smoke          # quick check
Rscript run_experiment.R --seeds 1,2,3,4,5,6,7,8 --ns 100,200
quarto render report.qmd
```
