---
title: "A small-sample-corrected profile-LR confidence interval for the reliability family"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
fmt <- function(x, k = 3) formatC(x, format = "f", digits = k)
rd <- function(f) if (file.exists(file.path("results", f))) read.csv(file.path("results", f)) else NULL
par_d <- rd("semlbci_parity.csv")
cov_h <- rd("coverage_H.csv")
cov_o <- rd("coverage_omega.csv")
lw <- rd("lawley_bf.csv")
```

## Question

We are building small-sample, range-respecting confidence intervals for the
reliability family (omega_total, omega_h, coefficient H, maximal reliability) by
inverting the likelihood-ratio test of `g(theta) = g0`. Two things must hold for
the engine to be trustworthy: (1) it must reproduce the established construction,
and (2) its intervals must actually cover at the sample sizes reliability is
reported at. Does a single generic engine, fed only a functional closure,
reproduce `semlbci`, and where the plain profile-LR miscalibrates at small N,
does a Bartlett correction fix it?

## Short answer

**Construction is exact.** Wherever `semlbci` (the Wu-Neale reference) converges,
the engine reproduces its likelihood-based bounds to `1e-7`, across omega_total,
H, and omega_h on one- and three-factor models. The engine also carries a
first-principles gate (`T(bound) = qchisq(.95,1)`) that every engine bound passes
and `semlbci`'s default search fails on some bifactor upper bounds.

**Calibration depends sharply on the coefficient.** For omega the plain
profile-LR is already near-exact (coverage ~0.93-0.97). For **maximal reliability
it collapses**: on a bifactor general factor at N=50 the nominal-95% interval
covers only **0.61**, because the LR statistic is inflated to `E[T] ~ 4.2` instead
of 1. Rescaling the chi-square threshold by the Bartlett factor `c = E[T]` is
necessary and, at the oracle factor, sufficient (0.61 -> 0.97). The feasible
factor from a *null-model* parametric bootstrap recovers most of the gap
(0.61 -> 0.87 at N=50, 0.92 at N=100) and fully fixes the milder one-factor H
case (0.92 -> 0.94, beating Wald's 0.88); closing the last points at the hardest
near-ceiling functionals is
the open estimator problem. Everything shown here is normal-theory; the robust
and misspec tiers are the same engine with a swappable reference scaling that, as
expected, reduces to 1 under normal data.

## Evidence

### Construction parity vs semlbci

```{r parity}
if (is.null(par_d)) cat("Run: Rscript run_experiment.R\n") else {
  vtol <- 0.02
  s <- do.call(rbind, lapply(split(par_d, par_d$func), function(g) {
    ok_lo <- abs(g$tdev_sl_lo) <= vtol; ok_hi <- abs(g$tdev_sl_hi) <= vtol
    data.frame(Coefficient = g$label[1], Cells = nrow(g),
      `Engine off eqn` = sum(abs(c(g$tdev_eng_lo, g$tdev_eng_hi)) > vtol, na.rm = TRUE),
      `semlbci off eqn` = sum(abs(c(g$tdev_sl_lo, g$tdev_sl_hi)) > vtol, na.rm = TRUE),
      `Max diff (semlbci valid)` = formatC(max(c(abs(g$d_lo[ok_lo]), abs(g$d_hi[ok_hi])), na.rm = TRUE), format = "e", digits = 1),
      check.names = FALSE)
  }))
  knitr::kable(s[order(s$Coefficient), ], row.names = FALSE,
    caption = "Engine profile-LR vs semlbci (Wu-Neale). 'off eqn' counts bounds failing T=qchisq(.95,1).")
}
```

### Small-sample coverage of maximal reliability, and the Bartlett fix

```{r coverage}
if (is.null(cov_h)) cat("Run: Rscript scripts/coverage.R --func H\n") else {
  lab <- c(high = "H, 1-factor (omega~.87)", mod = "H, 1-factor (omega~.67)",
           bf = "max reliability, bifactor")
  tab <- data.frame(
    Coefficient = lab[cov_h$pop], N = cov_h$n, `E[T]` = fmt(cov_h$mean_T, 2),
    `Uncorrected` = fmt(cov_h$cov_nt), `Bartlett oracle` = fmt(cov_h$cov_bart_oracle),
    `Bartlett null-boot` = fmt(cov_h$cov_bart_null),
    `Wald` = ifelse(is.na(cov_h$cov_wald), "-", fmt(cov_h$cov_wald)),
    check.names = FALSE)
  knitr::kable(tab, row.names = FALSE,
    caption = "Coverage of the population value (nominal 0.95), normal data. E[T]=1 would mean the chi-square-1 reference is exact.")
}
```

The uncorrected profile-LR is exact only when `E[T] ~ 1`. It holds for omega and
one-factor H, and breaks for maximal reliability, where the near-ceiling,
strongly nonlinear functional inflates the LR. The oracle Bartlett factor
(cell-mean `E[T]`, infeasible) shows the correction is sufficient; the feasible
null-model bootstrap (simulate from the constrained MLE that imposes `g = g0`,
average the LR) recovers most of it. omega coverage (separate run) stays
0.93-0.97 across N with `E[T] ~ 1`, so the correction is a near-no-op there.

### Why the correction must be a calibrated constant

Two natural shortcuts both fail for the case that needs correcting, and the way
they fail says what the right factor looks like.

*Not an analytic formula.* The maximal-reliability inflation is not a smooth
property of the coefficient: it lives almost entirely in samples that land near a
boundary (a near-zero fitted uniqueness, where the coefficient becomes unstable).
Splitting `E[T]` by how close the fitted model is to that boundary shows the
overall value is set by the near-boundary quarter, not by the typical (interior)
sample, and a smooth factor sees only the interior.

```{r lawley}
if (is.null(lw)) cat("Run: Rscript scripts/lawley.R --pop bf\n") else {
  r <- lw[lw$func == "rho", ]
  tab <- data.frame(N = r$n,
    `E[T], all samples` = fmt(r$oracle_ET, 2),
    `E[T], near-boundary quarter` = fmt(r$et_near_bd, 2),
    `E[T], interior quarter` = fmt(r$et_far_bd, 2),
    check.names = FALSE)
  knitr::kable(tab, row.names = FALSE,
    caption = "Maximal reliability: the LR inflation concentrates in near-boundary samples (bifactor, normal data). E[T]=1 is the calibrated target.")
}
```

A smooth factor built from the curvature of the coefficient tracks the interior
quarter but cannot see the near-boundary mass, so it recovers only about a quarter
of the inflation and does not improve with N. (The members where such a factor is
accurate, omega and one-factor H, are already near-nominal and need no correction.)

*Not re-estimated per dataset.* The opposite shortcut, a bootstrap that estimates
the factor separately for each sample, also fails, and instructively. The factor a
sample needs is largest exactly when its statistic is largest (near a boundary),
but the bootstrap estimate is *smallest* there; the two are anti-correlated. So
although the average bootstrap factor is roughly right, applied one dataset at a
time its variance and this anti-correlation cost about ten points of coverage,
and neither a larger bootstrap nor an iterated (double) bootstrap closes the gap.
The identical average factor applied as a single constant recovers those points.

```{r perssample}
db <- rd("double_boot_bf_n50.csv")
if (is.null(db)) cat("Run: Rscript scripts/double_boot.R --pop bf --samples 200\n") else {
  q1 <- qchisq(0.95, 1)
  oc <- if (!is.null(lw)) lw$oracle_ET[lw$func == "rho" & lw$n == 50][1] else NA
  covp <- function(cs) mean(db$T_real <= pmax(1, cs) * q1)   # per-sample factor
  covc <- function(k)  mean(db$T_real <= k * q1)             # same mean, as a constant
  tab <- data.frame(
    `Factor` = c("uncorrected", "one-level bootstrap", "double bootstrap", "oracle constant"),
    `Applied per sample` = c(fmt(covc(1)), fmt(covp(db$c_plugin)), fmt(covp(db$c_mult)), "-"),
    `Same mean, as a constant` = c("-", fmt(covc(mean(db$c_plugin))), fmt(covc(mean(db$c_mult))), fmt(covc(oc))),
    check.names = FALSE)
  knitr::kable(tab, row.names = FALSE,
    caption = "Maximal reliability coverage (nominal 0.95, bifactor, N=50). A per-sample factor caps near 0.85; the same factor as a constant reaches 0.93-0.98.")
}
```

So the small-sample factor for the hard members must be a stable, model-level
constant (calibrated by simulation), not a smooth formula and not a per-dataset
resampling estimate.

## Caveats

- **Normal data throughout.** This establishes the NT tier. The robust and
  misspec tiers are the same engine with a reference scaling equal to a variance
  ratio (sandwich over expected, or over observed, information); under normal
  data both reduce to ~1 (verified), so they coincide with NT here. Non-normal
  stress is the next run.
- **The feasible small-sample factor is not fully solved for the hardest cases.**
  A per-dataset factor (one-level or double bootstrap) caps coverage near 0.85 at
  the bifactor ceiling; an analytic factor is ruled out (boundary-driven, not
  smooth). The same factor as a model-level constant reaches 0.93-0.98, so the
  open task is a low-variance, calibrated constant for the model class rather than
  a per-dataset estimate; the point-estimate near-Heywood bias is the deeper
  remaining source.
- A distribution-free Bollen-Stine bootstrap and an unconstrained-DGP bootstrap
  are available as options for later non-normal comparisons; they are not the
  headline.
- Modest reps per cell and a few improper (Heywood) drops at the smallest N;
  the qualitative pattern (collapse for maximal reliability, near-exact for
  omega) is stable across runs.

## Reproduce

```sh
cd experiments/45-profile-lr-reliability-ci
Rscript run_experiment.R                          # construction parity vs semlbci
Rscript scripts/coverage.R --func H  --scalings   # maximal-reliability collapse + Bartlett
Rscript scripts/coverage.R --func omega           # omega is near-nominal (control)
Rscript scripts/lawley.R --pop bf                 # analytic factor cannot see the boundary mass
Rscript scripts/double_boot.R --pop bf --samples 200  # per-sample factor fails; a constant works
quarto render report.qmd
```
