---
title: "Bell-Style Omega Bias at Population Moments"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
---

```{r}
res_dir <- params$results_dir
needed <- file.path(res_dir, c("omega_bias.csv", "population_targets.csv",
                               "cfa_reliability.csv",
                               "cfa_sensitivity_summary.csv",
                               "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/42-bell-omega-bias/run_experiment.R",
       call. = FALSE)
}
bias <- read.csv(file.path(res_dir, "omega_bias.csv"))
targets <- read.csv(file.path(res_dir, "population_targets.csv"))
cfa <- read.csv(file.path(res_dir, "cfa_reliability.csv"))
cfa_summary <- read.csv(file.path(res_dir, "cfa_sensitivity_summary.csv"))
metadata <- read.csv(file.path(res_dir, "metadata.csv"))
meta <- setNames(metadata$value, metadata$key)
fmt <- function(x, digits = 3) {
  ifelse(is.na(x), "", format(round(x, digits), nsmall = digits))
}
pct <- function(x) {
  ifelse(is.na(x), "", sprintf("%+.1f%%", 100 * x))
}
label_est <- c(ML = "ML", ULS = "ULS", GLS = "GLS",
               closed_form = "Spearman-Guttman")
label_denom <- c(model_implied = "Model-implied",
                 observed_total = "Observed total")
focus <- bias[bias$omega_variant %in%
                c("omega_u_model_implied",
                  "omega_u_observed_total",
                  "spearman_guttman_covariance_omega"), ]
focus$Estimator <- unname(label_est[focus$estimator])
focus$Estimator[is.na(focus$Estimator)] <- focus$estimator[is.na(focus$Estimator)]
focus$Denominator <- unname(label_denom[focus$denominator])
focus$Denominator[is.na(focus$Denominator)] <- focus$denominator[is.na(focus$Denominator)]
focus$Reliability <- tools::toTitleCase(focus$reliability_level)
focus$Population <- focus$population_label

display <- focus[focus$reliability_level == "low", ]
display <- display[order(display$population, display$Estimator,
                         display$Denominator), ]
tab <- data.frame(
  Population = display$Population,
  p = display$p,
  Estimator = display$Estimator,
  Denominator = display$Denominator,
  Omega = fmt(display$estimate),
  Target = fmt(display$true_reliability),
  Bias = fmt(display$bias),
  `Relative Bias` = pct(display$relative_bias),
  `F` = fmt(display$population_discrepancy),
  `RMSEA-like` = fmt(display$rmsea_like),
  check.names = FALSE
)

ml_simple <- bias[bias$estimator == "ML" &
                    bias$denominator == "model_implied" &
                    bias$omega_variant == "omega_u_model_implied", ]
sg <- bias[bias$estimator == "closed_form", ]
fit_only <- focus[is.finite(focus$rmsea_like) &
                    focus$denominator == "model_implied", ]
worst_ml <- ml_simple[which.max(abs(ml_simple$bias)), ]
worst_sg <- sg[which.max(abs(sg$bias)), ]
worst_fit <- fit_only[which.max(fit_only$rmsea_like), ]
one_factor_ml <- ml_simple[ml_simple$population == "one_factor", ]
max_well_spec <- max(abs(one_factor_ml$bias), na.rm = TRUE)
cfa_summary$Reliability <- tools::toTitleCase(cfa_summary$reliability_level)
cfa_summary$Population <- cfa_summary$population_label
cfa_low <- cfa_summary[cfa_summary$reliability_level == "low", ]
cfa_low <- cfa_low[order(cfa_low$population), ]
cfa_tab <- data.frame(
  Population = cfa_low$Population,
  p = cfa_low$p,
  Target = fmt(cfa_low$true_reliability),
  `One-factor bias` = fmt(cfa_low$one_factor_bias),
  `One-factor RMSEA-like` = fmt(cfa_low$one_factor_rmsea_like),
  `Omega range` = fmt(cfa_low$omega_range),
  `Best CFA model` = cfa_low$best_cfa_model,
  `Best abs. bias` = fmt(cfa_low$min_abs_bias),
  check.names = FALSE
)
cfa_correct <- cfa_summary[cfa_summary$population == "one_factor", ]
cfa_misspec <- cfa_summary[cfa_summary$population != "one_factor", ]
max_cfa_range_correct <- max(cfa_correct$omega_range, na.rm = TRUE)
min_cfa_range_misspec <- min(cfa_misspec$omega_range, na.rm = TRUE)
max_cfa_range_misspec <- max(cfa_misspec$omega_range, na.rm = TRUE)
best_cfa <- cfa_summary[which.min(cfa_summary$min_abs_bias), ]
```

## Question

Bell et al. define bias as the omega estimate minus the population reliability:
the total-score variance due to the data-generating factor common to all items,
divided by total-score variance. What happens if we evaluate ordinary one-factor
omega directly at those population covariance matrices, and swap ML for ULS/GLS
or the Spearman-Guttman closed form?

## Short Answer

At population moments, the ordinary one-factor omega functional is essentially
unbiased in the correctly specified one-factor cells: the largest absolute ML
model-implied bias there is `r fmt(max_well_spec, 4)`. Under Bell-style
misspecification, the same functional can be badly biased against Bell's target:
the largest displayed ML/model-implied bias is `r fmt(worst_ml$bias)` in the
`r worst_ml$population_label` / `r worst_ml$reliability_level` cell.

Changing the denominator from model-implied to observed total variance is a small
correction, not a cure. The Spearman-Guttman covariance omega follows the same
logic: exact on the one-factor manifold, but its largest absolute bias here is
`r fmt(worst_sg$bias)` in the `r worst_sg$population_label` /
`r worst_sg$reliability_level` cell.

The fit diagnostics tell the same broad story but should not be read as a clean
ML-versus-ULS-versus-GLS ranking: the largest model-implied RMSEA-like value is
`r fmt(worst_fit$rmsea_like)` for `r worst_fit$Estimator` in the
`r worst_fit$population_label` / `r worst_fit$reliability_level` cell, using that
estimator's own discrepancy scale.

The CFA model-set check is more directly about reliability than ordinary
fit. The omega range across candidate CFA specifications is at most
`r fmt(max_cfa_range_correct, 4)` in the correctly specified one-factor cells,
but ranges from `r fmt(min_cfa_range_misspec)` to
`r fmt(max_cfa_range_misspec)` in the misspecified cells. When the candidate set
contains a model that isolates the generating main factor, the best CFA bias is
numerically zero; the low-reliability correlated-errors cell is the exception in
this first implementation, with best absolute bias
`r fmt(cfa_summary$min_abs_bias[cfa_summary$population == "correlated_errors" &
                                cfa_summary$reliability_level == "low"])`.

## CFA Reliability Functional

Let $S$ be the population covariance matrix used as sample covariance in this
experiment and let $w = \mathbf 1_p$ form the unit-weighted total score. For a
candidate CFA model $M$, fit the model at $S$ and let
$\hat\theta_M = \arg\min_{\theta \in M} F_{\text{ML}}(S, \Sigma(\theta))$.
Write $C_g(\hat\theta_M)$ for the covariance contribution of the declared
main factor $g$. The CFA-specific reliability coefficient used here is

$$
\omega_g^{\text{obs}}(M)
  =
  \frac{w^\top C_g(\hat\theta_M) w}{w^\top S w}.
$$

For the one-factor and bifactor candidates,
$C_g(\hat\theta_M)=\lambda_g\lambda_g^\top$, because the main factor is
standardized. For the higher-order candidate, the item coefficients on the
higher-order factor are indirect: if $\Lambda$ contains item loadings on the
first-order factors and $\gamma$ contains first-order loadings on the
higher-order factor, then
$C_g(\hat\theta_M)=(\Lambda\gamma)(\Lambda\gamma)^\top$.

Against Bell et al.'s target reliability
$\omega_{\text{Bell}} = w^\top C_{\text{gen}} w / w^\top S w$, the
population bias for candidate $M$ is

$$
b_g(M) = \omega_g^{\text{obs}}(M) - \omega_{\text{Bell}}.
$$

The model-set sensitivity diagnostic is just the range of that coefficient over
a small CFA set $\mathcal M$:

$$
R_\omega(\mathcal M)
  =
  \max_{M \in \mathcal M}\omega_g^{\text{obs}}(M)
  -
  \min_{M \in \mathcal M}\omega_g^{\text{obs}}(M).
$$

This is not a new fit index. It asks whether plausible CFA ways of protecting a
declared main factor produce materially different main-factor reliability
claims.

## Evidence

The table shows the low-reliability cells, where Bell et al. also report the
largest misspecification effects. Bias is estimate minus Bell's target. `F` is
the fitted discrepancy at population moments (`2 * fmin`), and `RMSEA-like` is
`sqrt(F / df)`. Those two fit columns are within-discrepancy diagnostics, not
cross-discrepancy evidence.

```{r}
knitr::kable(tab, row.names = FALSE)
```

The next table uses the CFA functional above. `Omega range` is
$R_\omega(\mathcal M)$ over the candidate set: one factor, one factor with
within-block residual covariances, bifactor blocks, and for p=12 also
higher-order blocks.

```{r}
knitr::kable(cfa_tab, row.names = FALSE)
```

The full population-moment tables, including high-reliability cells, are in
`r file.path(res_dir, "omega_bias.csv")` and
`r file.path(res_dir, "cfa_reliability.csv")`.

## Caveats

This is not a finite-sample replication. There is no $N$, no convergence
replacement loop, no standard errors, and no coverage. The experiment asks only
what Bell's bias definition becomes after sample noise is removed.

The population values are Bell-inspired p=8 cells plus the p=12 higher-order
cell from their loading table.

The CFA sensitivity section is not an identified bias estimator. It depends on
the candidate model set and on which factor is declared to be the main factor.
The residual-block model is deliberately flexible: in these toy populations it
can reproduce some bifactor or higher-order nuisance covariance while preserving
the total-score contribution attributed to the main factor. That is useful as a
diagnostic, but it is not proof that the fitted model has recovered the
substantive factor.

## Reproduce

```sh
Rscript experiments/42-bell-omega-bias/run_experiment.R
quarto render experiments/42-bell-omega-bias/report.qmd
```
