---
title: "Bell-Style Omega Bias at Population Moments"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
---

```{r}
res_dir <- params$results_dir
needed <- file.path(res_dir, c("omega_bias.csv", "population_targets.csv",
                               "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/42-bell-omega-bias/run_experiment.R",
       call. = FALSE)
}
bias <- read.csv(file.path(res_dir, "omega_bias.csv"))
targets <- read.csv(file.path(res_dir, "population_targets.csv"))
metadata <- read.csv(file.path(res_dir, "metadata.csv"))
meta <- setNames(metadata$value, metadata$key)
fmt <- function(x, digits = 3) {
  ifelse(is.na(x), "", format(round(x, digits), nsmall = digits))
}
pct <- function(x) {
  ifelse(is.na(x), "", sprintf("%+.1f%%", 100 * x))
}
label_est <- c(ML = "ML", ULS = "ULS", GLS = "GLS",
               closed_form = "Spearman-Guttman")
label_denom <- c(model_implied = "Model-implied",
                 observed_total = "Observed total")
focus <- bias[bias$omega_variant %in%
                c("omega_u_model_implied",
                  "omega_u_observed_total",
                  "spearman_guttman_covariance_omega"), ]
focus$Estimator <- unname(label_est[focus$estimator])
focus$Estimator[is.na(focus$Estimator)] <- focus$estimator[is.na(focus$Estimator)]
focus$Denominator <- unname(label_denom[focus$denominator])
focus$Denominator[is.na(focus$Denominator)] <- focus$denominator[is.na(focus$Denominator)]
focus$Reliability <- tools::toTitleCase(focus$reliability_level)
focus$Population <- focus$population_label

display <- focus[focus$reliability_level == "low", ]
display <- display[order(display$population, display$Estimator,
                         display$Denominator), ]
tab <- data.frame(
  Population = display$Population,
  p = display$p,
  Estimator = display$Estimator,
  Denominator = display$Denominator,
  Omega = fmt(display$estimate),
  Target = fmt(display$true_reliability),
  Bias = fmt(display$bias),
  `Relative Bias` = pct(display$relative_bias),
  `F` = fmt(display$population_discrepancy),
  `RMSEA-like` = fmt(display$rmsea_like),
  check.names = FALSE
)

ml_simple <- bias[bias$estimator == "ML" &
                    bias$denominator == "model_implied" &
                    bias$omega_variant == "omega_u_model_implied", ]
sg <- bias[bias$estimator == "closed_form", ]
fit_only <- focus[is.finite(focus$rmsea_like) &
                    focus$denominator == "model_implied", ]
worst_ml <- ml_simple[which.max(abs(ml_simple$bias)), ]
worst_sg <- sg[which.max(abs(sg$bias)), ]
worst_fit <- fit_only[which.max(fit_only$rmsea_like), ]
one_factor_ml <- ml_simple[ml_simple$population == "one_factor", ]
max_well_spec <- max(abs(one_factor_ml$bias), na.rm = TRUE)
```

## Question

Bell et al. define bias as the omega estimate minus the population reliability:
the total-score variance due to the data-generating factor common to all items,
divided by total-score variance. What happens if we evaluate ordinary one-factor
omega directly at those population covariance matrices, and swap ML for ULS/GLS
or the Spearman-Guttman closed form?

## Short Answer

At population moments, the ordinary one-factor omega functional is essentially
unbiased in the correctly specified one-factor cells: the largest absolute ML
model-implied bias there is `r fmt(max_well_spec, 4)`. Under Bell-style
misspecification, the same functional can be badly biased against Bell's target:
the largest displayed ML/model-implied bias is `r fmt(worst_ml$bias)` in the
`r worst_ml$population_label` / `r worst_ml$reliability_level` cell.

Changing the denominator from model-implied to observed total variance is a small
correction, not a cure. The Spearman-Guttman covariance omega follows the same
logic: exact on the one-factor manifold, but its largest absolute bias here is
`r fmt(worst_sg$bias)` in the `r worst_sg$population_label` /
`r worst_sg$reliability_level` cell.

The fit diagnostics tell the same broad story but should not be read as a clean
ML-versus-ULS-versus-GLS ranking: the largest model-implied RMSEA-like value is
`r fmt(worst_fit$rmsea_like)` for `r worst_fit$Estimator` in the
`r worst_fit$population_label` / `r worst_fit$reliability_level` cell, using that
estimator's own discrepancy scale.

## Evidence

The table shows the low-reliability cells, where Bell et al. also report the
largest misspecification effects. Bias is estimate minus Bell's target. `F` is
the fitted discrepancy at population moments (`2 * fmin`), and `RMSEA-like` is
`sqrt(F / df)`. Those two fit columns are within-discrepancy diagnostics, not
cross-discrepancy evidence.

```{r}
knitr::kable(tab, row.names = FALSE)
```

The full population-moment table, including high-reliability cells, is in
`r file.path(res_dir, "omega_bias.csv")`.

## Caveats

This is not a finite-sample replication. There is no \(N\), no convergence
replacement loop, no standard errors, and no coverage. The experiment asks only
what Bell's bias definition becomes after sample noise is removed.

The population values are Bell-inspired p=8 cells plus the p=12 higher-order
cell from their loading table. The fitted model is deliberately the ordinary
one-factor model with independent residuals; bifactor omega-hierarchical and
one-factor models with freed residual covariances are outside this first slice.

## Reproduce

```sh
Rscript experiments/42-bell-omega-bias/run_experiment.R
quarto render experiments/42-bell-omega-bias/report.qmd
```
