---
title: "Robust test calibration"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Question

How do robust score and LR tests calibrate across SEMs and sampling
distributions, and which spectral or computational controls explain the differences?

## Short answer

**Score SB and pEBA4 performed better than their LR counterparts in the frozen
null-size battery, especially at small N and high residual rank.** Across
48 nested-test cells, score SB rejected 3.6–5.8% and score pEBA4 2.8–5.5% at
nominal 5%; LR reached 6.6–17.7% at N=100 with those calibrations. These rates
are close to nominal in the tested designs, not a general exactness result.

Global tests remain sensitive to heterogeneous innovations: score pEBA4
reached 12.3% and score SB 14.1% in the worst IG cells. Homogeneous-kurtosis
controls explain why changing rank or clean information geometry alone did
not recreate that failure. Correctness and timing diagnostics accompany the
statistical evidence. No new inference default follows from this consolidation.

## Evidence

```{r setup}
read_result <- function(lane, name) {
  p <- file.path("results", lane, name)
  if (!file.exists(p)) stop("Missing result: ", p, "; see Reproduce.")
  read.csv(p, stringsAsFactors = FALSE)
}
battery <- read_result("battery", "frozen/summary.csv")
failures <- read_result("battery", "frozen/failures.csv")
dfs <- read_result("drivers", "df_sweep.csv")
structure <- read_result("drivers", "structure.csv")
checks <- read_result("diagnostics", "trace-sb-nested/checks.csv")
```

### Broad frozen battery

Seven textbook-derived population models, N=100/300/1000 and four generators
(normal, homogeneous Vale–Maurelli, independent heterogeneous innovations,
and five-point discretization) give 84 global and 48 nested cells, with 5,000
replications per cell. Score and LR calibrations use their respective own
null-fit spectra; controls include lavaan RLS and MV-family corrections.

```{r}
z <- battery[battery$subset == "converged" &
  battery$arm %in% c("score_own_sb", "score_own_peba4", "lr_own_sb", "lr_own_peba4"), ]
tab <- do.call(rbind, lapply(split(z, interaction(z$test,z$arm)), function(x)
  data.frame(Test=ifelse(x$test[1]=="gof","Global fit","Nested"),
    Arm=unname(c(score_own_sb="Score SB",score_own_peba4="Score pEBA4",
      lr_own_sb="LR SB",lr_own_peba4="LR pEBA4")[x$arm[1]]), Cells=nrow(x),
    `Minimum rejection (%)`=100*min(x$reject_05),
    `Maximum rejection (%)`=100*max(x$reject_05),
    `Mean absolute error (points)`=100*mean(abs(x$reject_05-.05)),check.names=FALSE)))
knitr::kable(tab, digits=1,row.names=FALSE,
  caption="Ranges over all generators, models and N; averages describe cell errors, not pooled rejection.")
```

For global normal/discretized data, score SB rejected 4.3–6.2% and score
pEBA4 3.2–6.0%. LR-SB and LR-pEBA4 reached 20.8% and 15.8% at N=100 in the
18-indicator longitudinal model (df=102). MV-family score corrections tended
to conservatism; corrected MV did not outperform pEBA4 under heterogeneous
innovations. Five-point Gaussian-copula discretization behaved similarly to
normal data here, which is not a result about arbitrary ordinal DGPs.

With 5,000 valid draws, a 5% rejection estimate has Monte Carlo SE about
0.31 percentage points (roughly ±0.6 points for a marginal 95% interval).
Cell ranges span many comparisons and are not simultaneous confidence bounds.
Rates condition on convergence: about 9.4% of N=100 second-order CFA fits and
2% of Worland SEM fits did not converge. Frozen failure and admissible-only
summaries allow a separate audit of that selection.

### Mechanism controls

Controls use homogeneous nonnormality and vary tested rank (1–28), information
geometry and residual conditioning. They have 300 rank-sweep and 240 structure
replications per cell, not extra replications for the battery. Eigenvalue
dispersion governs the information a mean-only SB correction omits: for a
weighted chi-square limit its variance relative to the equal-eigenvalue
reference involves $1+\mathrm{CV}(\lambda)^2$. Rank one removes spectral-shape
mismatch in that limit; it does not make the fitted finite-sample test exact.

```{r}
z <- dfs[dfs$distribution == "vm", ]
methods <- c(score_sb="Score SB",score_peba4="Score pEBA4",flip_effective="Effective flip")
cols <- c("#0072B2","#D55E00","#555555")
plot(range(z$df_obs),c(0,.12),type="n",xlab="Restriction rank",ylab="Rejection at nominal 5%")
abline(h=.05,lty=2,col="grey50")
for (i in seq_along(methods)) {
  o <- order(z$df_obs)
  lines(z$df_obs[o],z[[names(methods)[i]]][o],type="b",col=cols[i],pch=i)
}
legend("topright",legend=unname(methods),col=cols,pch=seq_along(methods),lty=1,bty="n")
```

Clean rank and structure changes did not break SB in these controls. The
near-Heywood case exposed boundary conditioning and also degraded flips.
These homogeneous designs do not validate concentrated directional kurtosis;
heterogeneous spectral calibration remains an independent research question.

### Correctness and cost diagnostics

Extracted predecessor diagnostics compare trace-only SB with spectral SB,
match nested statistics/pEBA4 against lavaan, and separate setup, fitting,
projection and calibration cost. Frozen checks are deterministic agreement
evidence, rather than another null-size or power grid.

```{r}
cols <- c("sb_trace_vs_spectral", "sb_spectral_vs_lavaan",
          "nested_Tdiff_vs_lavaan", "nested_peba4_delta_vs_lavaan")
knitr::kable(data.frame(Check=cols,
  `Largest absolute difference`=formatC(vapply(checks[cols], max, numeric(1), na.rm=TRUE),format="e",digits=2),
  check.names=FALSE),row.names=FALSE)
```

Phase timings, wrapper checks and matched correctness gates remain under
`results/diagnostics/`. Historical comparison outputs are retained for
provenance; the overlapping score-versus-LR simulation runner is retired.
An optional FIML diagnostic is a convention investigation, not parity before
its discrepancy is resolved.

## Caveats

This is exploratory null-size evidence. It gives no power ranking, does not
settle larger-model misspecification and does not repair heterogeneous global
calibration. Driver designs differ from the battery. Timing is machine/package
dependent. Keep statistical and failure conventions intact when extending a lane.

Source populations retain their model definitions and optional corpus builder.
Frozen metadata retain original study numbers (research/52, /38 and former
/45), which describe the original runs.

```{r units}
units <- read_result("battery", "frozen/units_check.csv")
```

The frozen run used a standardized two-group path population (`mg_path5`),
a workaround for a unit-dependent rank check fixed on 2026-10-01; the
population now keeps the book's units, whose variances span a ratio near 300.
A check with the main seeds (`r units$cells` cells, `r units$reps_per_cell`
replications each) reproduces the standardized run in original units:
`r units$failures_original` failures in both, `r units$decisions_differing_at_05`
of `r format(units$rejection_decisions, big.mark = ",")` rejection decisions at
the 5% level differ, and statistics agree within optimizer tolerance (largest
absolute difference `r units$max_abs_difference`).

## Reproduce

```sh
Rscript run_experiment.R --lane battery --plan
Rscript run_experiment.R --lane battery --smoke --output /tmp/calibration-battery
Rscript run_experiment.R --lane drivers --reps 1 --flips 9 --max-cells 1 --results-dir /tmp/calibration-drivers
Rscript run_experiment.R --lane diagnostics --check trace --smoke --no-time
quarto render report.qmd
```

Use `--lane NAME --help` for commands. Trace smoke writes the diagnostics
lane's `results/trace-sb-nested/`; copy the self-contained diagnostics lane
to a temporary directory for verification without replacing frozen evidence.
Additional checks are `runtime`, `setup`, `calibration` and `validation`.
Rendering performs no fits or simulation.
