---
title: "Lambda6 Reliability Experiment"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
---

```{r}
res_dir <- params$results_dir
needed <- file.path(res_dir, c("simulation_summary.csv",
                               "population_targets.csv",
                               "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/41-reliability-lambda6/run_experiment.R --smoke",
       call. = FALSE)
}
summary <- read.csv(file.path(res_dir, "simulation_summary.csv"))
targets <- read.csv(file.path(res_dir, "population_targets.csv"))
metadata <- read.csv(file.path(res_dir, "metadata.csv"))
meta <- setNames(metadata$value, metadata$key)
fmt <- function(x, digits = 3) {
  ifelse(is.na(x), "", format(round(x, digits), nsmall = digits))
}
pct <- function(x) {
  ifelse(is.na(x), "", sprintf("%.1f%%", 100 * x))
}
label_method <- c(
  alpha = "Alpha",
  lambda6 = "Lambda6",
  spearman_guttman_omega = "Spearman-Guttman omega",
  nt_omega_model_se = "NT omega, model SE",
  nt_omega_robust_se = "NT omega, robust SE"
)
summary$Method <- unname(label_method[summary$method])
summary$Method[is.na(summary$Method)] <- summary$method[is.na(summary$Method)]
summary$Population <- summary$population_label
summary$N <- summary$n

show <- summary[summary$n == min(summary$n), ]
show <- show[order(show$population, match(show$method, names(label_method))), ]
tab <- data.frame(
  Population = show$Population,
  N = show$N,
  Method = show$Method,
  Target = fmt(show$target_functional),
  `True reliability` = fmt(show$true_reliability),
  `RMSE vs target` = fmt(show$rmse_functional),
  `RMSE vs true` = fmt(show$rmse_true),
  `Coverage vs target` = pct(show$coverage_functional),
  `Coverage vs true` = pct(show$coverage_true),
  `Mean SE` = fmt(show$mean_se),
  check.names = FALSE
)

best_target <- do.call(rbind, lapply(split(show, show$population), function(d) {
  d[which.min(d$rmse_functional), , drop = FALSE]
}))
best_true <- do.call(rbind, lapply(split(show, show$population), function(d) {
  d[which.min(d$rmse_true), , drop = FALSE]
}))
best_sentence <- paste(sprintf("%s target-RMSE: %s; true-RMSE: %s",
                               best_target$population_label,
                               best_target$Method,
                               best_true$Method),
                       collapse = ". ")
target_slice <- targets[targets$population == "one_factor", ]
lambda6_target <- target_slice$target_functional[target_slice$method == "lambda6"]
alpha_target <- target_slice$target_functional[target_slice$method == "alpha"]
```

## Question

If Guttman's lambda6 is just a sample-covariance functional, how does it behave
as a reliability estimator beside alpha, the Spearman-Guttman covariance omega,
and fitted normal-theory omega?

## Short Answer

For the correctly specified one-factor population, the Spearman-Guttman
covariance omega and fitted omega target the same total-score reliability,
whereas lambda6 targets `r fmt(lambda6_target)` and alpha targets
`r fmt(alpha_target)`. In the displayed `r meta[["reps"]]`-rep smoke slice at
the smallest \(N\), `r best_sentence`.

Under the two-factor population fit as one factor, every number should be read
against its own population functional unless the column explicitly says "true":
the robust fitted-omega interval is the misspecification-aware one, while the
model-SE interval is the ordinary one-factor NT interval.

## Evidence

```{r}
knitr::kable(tab, row.names = FALSE)
```

The raw replicate file also records squared error and coverage against the
generating reliability, so the persistent bias of alpha and lambda6 is visible
separately from interval calibration for their own covariance functionals.

## Caveats

This is a small normal-data run. It checks the first implementation slice:
sample covariance, empirical-Gamma delta intervals for covariance functionals,
and one fitted one-factor omega. The misspecified population is deliberately
simple: two correlated factors summarized by one fitted factor. Non-normal data,
multi-group reliability, and analytic gradients for the Spearman-Guttman
covariance omega are left for later runs.

Smoke results are noisy. For coverage, rerun with at least a few hundred
replications before making a statistical call.

## Reproduce

```sh
Rscript experiments/41-reliability-lambda6/run_experiment.R --smoke
Rscript experiments/41-reliability-lambda6/run_experiment.R --reps 500 --n 250,500
quarto render experiments/41-reliability-lambda6/report.qmd
```
