---
title: "Schlechter 2024 ASCI CFA parity"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "metadata.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
read_csv_safely <- function(name) {
  p <- file.path(results_dir, name)
  if (!file.exists(p)) return(NULL)
  utils::read.csv(p, check.names = FALSE, stringsAsFactors = FALSE)
}
meta <- read_csv_safely("metadata.csv")
fits <- read_csv_safely("model_fits.csv")
if (is.null(meta) || is.null(fits)) {
  stop("Missing result CSVs. Run: Rscript run_experiment.R")
}
fmt <- function(x, digits = 3) ifelse(is.na(x), "", formatC(x, digits = digits, format = "f"))
```

## Question

Are the final Schlechter et al. (2024) Attitudes Towards Social Comparison
Inventory measurement models inside magmaan's SEM scope? The OSF scripts include
exploratory factor analysis, reliability, correlations, and t-tests, but the
final ASCI measurement models are ordinary two-factor ordinal CFAs fit with
lavaan WLSMV.

## Short answer

Yes, for the final single-sample ASCI CFA models. magmaan's ordinal DWLS path
matches lavaan's WLSMV/DWLS free estimates across Study 1's held-out sample,
Study 2 English, and Study 3 German with maximum free-parameter difference
`r fmt(max(fits$max_abs_free_param_diff), 6)`. On lavaan's ordinal DWLS
chi-square scale, the maximum chi-square gap is
`r fmt(max(fits$abs_chisq_diff), 6)`. The robust scaled WLSMV statistics are
reported as lavaan context, not as the parity target.

## Evidence

```{r}
show <- fits[, c("study", "label", "nobs_lavaan", "lavaan_df",
                 "magmaan_chisq_lavaan_scale", "lavaan_chisq",
                 "abs_chisq_diff", "max_abs_free_param_diff",
                 "lavaan_chisq_scaled", "lavaan_cfi_scaled",
                 "lavaan_rmsea_scaled", "lavaan_srmr")]
names(show) <- c("Study", "Sample", "N", "df", "magmaan DWLS T",
                 "lavaan DWLS T", "|T diff|", "Max |param diff|",
                 "lavaan WLSMV T", "lavaan scaled CFI",
                 "lavaan scaled RMSEA", "lavaan SRMR")
num_cols <- sapply(show, is.numeric)
show[num_cols] <- lapply(show[num_cols], function(x) round(x, 6))
knitr::kable(show, row.names = FALSE)
```

## Caveats

This is a parity probe for the final CFA measurement models, not a reproduction
of every analysis in the OSF scripts. The exploratory EFA models use lavaan's
`efa("efa")` rotation machinery, and the paper also reports omega coefficients,
scale-score correlations, and t-tests; those are outside magmaan's SEM target.

The gender measurement-invariance ladders are partly in scope, but they are not
included here. Configural/metric/scalar estimates are close in quick checks, but
the ordinal fit-statistic conventions and residual-invariance step need their
own focused probe before being presented as reproduced.

## Reproduce

```sh
Rscript experiments/32-schlechter-2024-asci-cfa/run_experiment.R
quarto render experiments/32-schlechter-2024-asci-cfa/report.qmd
```
