---
title: "Recentered bootstrap calibration of the small-N nested ML score"
format:
  html:
    toc: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Question

Can case resampling with recentered likelihood scores and refitted observed
geometry reduce the small-N conservatism of the nested ML score, while allowing
the larger working SEM to be misspecified?

```{r}
pilot <- file.path('results','pilot')
if (!file.exists(file.path(pilot,'rates.csv')))
  stop('Pilot absent. Run Rscript run_experiment.R --reps 40 --bootstrap 199 --run-id pilot.')
rates <- read.csv(file.path(pilot,'rates.csv'))
cells <- read.csv(file.path(pilot,'cells.csv'))
meta <- read.csv(file.path(pilot,'metadata.csv'))
value <- function(k) meta$value[match(k,meta$key)]
rates <- merge(cells,rates,by='cell_id',sort=FALSE)
failure <- read.csv(file.path(pilot,'failures.csv'))
```

## Short answer

This development run establishes bootstrap feasibility and compute cost, not a
correction to adopt. It completed `r value('completed_reps')` fresh draws per
cell, with `r value('bootstrap')` resamples per draw, in
`r round(as.numeric(value('elapsed_s'))/60,1)` minutes on two workers.
Current SB/pEBA4 and fixed-geometry bootstrap rejected 0–2.5% in each cell;
refitted geometry rejected 0–5.1% among available tests. Eight bootstrap
optimization failures made three of 160 refitted tests unavailable (all in
skewed cells); the other arms were available on every draw. Per-cell Wilson
intervals below are too imprecise to distinguish 2% from 5% size.
The fixed-geometry arm controls resampling cost; the refitted arm addresses the
joint nuisance-projection/metric mechanism. Neither the library nor its
ordinary inference defaults changes.

## Evidence

```{r}
labels <- c(current_sb='Current SB',current_peba4='Current pEBA4',
            fixed='Fixed geometry bootstrap',refit='Refitted geometry bootstrap')
x <- rates[order(rates$cell_id,match(rates$arm,names(labels))),]
knitr::kable(data.frame(Population=paste(x$distribution,x$larger),
  Test=unname(labels[x$arm]),Available=paste0(x$available,'/',x$attempted),
  `Rejection (%)`=round(100*x$rejection,1),
  `95% Wilson interval (%)`=sprintf('%.1f–%.1f',100*x$lower,100*x$upper),check.names=FALSE),
  caption='Nominal 5% null tests; fresh development draws, no pooled estimate.')
```

These four cells use decisions/04's frozen constants: two factors with four
indicators each, loadings .7/.8/.75/.65, factor correlation .4, two groups
with 100 rows each, and six loading-equality restrictions. Mild
misspecification omits a residual correlation .3 between x1 and x5. Both
groups have the same population, so equality holds at the symmetric
pseudo-true solution. Skewed indicators use Vale–Maurelli skewness 2 and excess
kurtosis 7. Only design constants were copied; this leaf reads no sibling code
or results. Complete cell summaries are deliberately retained here because
four development cells are the full scope, with uncertainty essential to their
interpretation. Raw resample statistics and per-draw rows remain ignored.

## Caveats and methods

Let $S$ be the total score at the restricted fit. The adopted test projects
with observed sensitivity $H$ and measures the quadratic with expected
information $I$:

$$
G=D-K(K^\top H K)^{-1}K^\top H D,
\qquad Q=(G^\top S)^\top(G^\top I G)^{-1}(G^\top S).
$$

The current SB/pEBA4 references use the fitted raw case-score spectrum. The
fixed bootstrap resamples those case rows within each group, subtracts their
original total and keeps $G,I$ fixed. The refitted bootstrap resamples cases,
refits restricted and larger models using ordinary ML, obtains $S_b,H_b,I_b$,
and replaces $S_b$ by $S_b-S$ in common parameter coordinates. It then
recomputes the quadratic geometry. Bootstrap p-values are
$(1+\#\{Q_b\ge Q\})/(B+1)$; ties count in the upper tail.
All arms use the strict rejection rule $p<.05$. With B=199 the largest
rejected bootstrap p-value is .045; this finite-grid conservatism must be
considered when comparing rates. Production should preregister its tie and
threshold convention, with larger B.

The [design note](../../../../project/design/nested-score-finite-sample.md)
derives first-order validity of this post-fit recentering and describes an
exactly recentered objective, which has not been implemented or run. It also
derives a bootstrap-assisted moment adjustment and explains why an analytic
Bartlett coefficient is not yet available. No bootstrap from a fitted Gaussian
SEM is used: that would discard the misspecification being protected against.
Fixed B leaves Monte Carlo calibration error, and B=199 has a coarse tail.
These identical-group cells do not validate heterogeneous-group score-mean
handling. Strong misspecification, ten-restriction models, power and larger N
are absent. Bootstrap failure makes the refitted arm unavailable for that
draw; resamples are never dropped or replaced. Wilson intervals condition on
available arms; failures remain separate against all attempted draws.

The maximum independently reconstructed policy statistic discrepancy was
`r format(as.numeric(value('max_policy_gap')),digits=3)`. The pilot deliberately exits nonzero after writing all summaries when any
bootstrap fails. Its eight failures were line-search/convergence failures,
not a statistic reconstruction discrepancy. The smoke completed without
failures. No convergence rule was altered.

Summary CSVs are frozen
as tiny development/pricing evidence with source/native fingerprints and
package versions, not as confirmation or an optimizer gate.

## Next registration and price

Proposed follow-up: six- and ten-restriction families; N=100/300/1000 per group;
normal/skewed; correct/mild/strong shared misspecification; null and matched
power. This yields 72 cells, 2000 null and 1000 power draws per cell (108000
outer draws), B=999. Current observed SB/pEBA4, fixed-geometry and post-fit
refitted-geometry bootstraps are required arms; add exact adjusted-objective
recentring only after gradient/stationarity gates and an independent
implementation audit. Population pseudo-null/curvature and actual power
restrictions must be checked and committed before registration. Report size,
paired size-error differences, size-adjusted power, every loss and failures
per cell. No default follows from this pilot.

```{r}
wall <- as.numeric(value('elapsed_s'))
draws <- 4*as.numeric(value('completed_reps'))
b <- as.numeric(value('bootstrap'))
cpu_per_draw <- 2*wall/draws
hours <- cpu_per_draw*(999/b)*108000/3600
knitr::kable(data.frame(
  Quantity=c('Pilot CPU seconds per outer draw (two-worker wall bound)',
             'Proposed 108000-draw B=999 run (CPU hours)',
             'With 20% contingency (CPU hours)'),
  Estimate=round(c(cpu_per_draw,hours,1.2*hours),1)),
  caption='Rough pricing extrapolates N=100, six-restriction fits; larger models/N need a new timing pilot.')
```

This CPU bound charges both workers for all elapsed wall time. It is not a
measurement of user+system CPU and excludes an additional exact-objective arm.
The extrapolation assumes linear B cost and unchanged fit cost, which larger
models and larger N may violate. A stratified timing pilot is required before
booking compute. Proposed fresh draw base 1226100101 and summary base
1426100101 are reserved and unused. Production must assign nonoverlapping
resample streams. Pilot draw seeds are base 1126100091 + 10000*cell ID + draw;
bootstrap index seeds add 1000000 + resample ID (overlap across outer draws is
recorded and should be removed for production). Simbox is unavailable; this
report proposes a later registration and authorizes no remote run.

## Reproduce

From the repository root, with installed lane packages:

```sh
Rscript experiments/research/active/54-nested-score-small-n/run_experiment.R --help
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1 \
R_LIBS="$HOME/.cache/magmaan-rlib/lane-e" nice -n 10 Rscript \
  experiments/research/active/54-nested-score-small-n/run_experiment.R \
  --smoke --run-id fresh-smoke
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1 \
R_LIBS="$HOME/.cache/magmaan-rlib/lane-e" nice -n 10 Rscript \
  experiments/research/active/54-nested-score-small-n/run_experiment.R \
  --reps 40 --bootstrap 199 --workers 2 --run-id fresh-pilot
quarto render experiments/research/active/54-nested-score-small-n/report.qmd
```

Fresh directories are mandatory; the runner checkpoints every balanced batch
and stops starting batches before its 1100-second budget. Rendering reads only
this leaf's frozen pilot summaries and performs no fits or simulations.
