---
title: "Nested ML geometry calibration"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Default register

| Choice | Current value | Status | Evidence |
|---|---|---|---|
| Nested ML LR spectrum | Observed information at the larger model | Adopted by principle; confirmed | Production 2026-10-03: expected drifts under strong misspecification (11–16% at nominal 5%), observed approaches 5% as N grows |
| Nested ML score projection | Observed sensitivity, expected metric | Adopted by principle; conservative at N = 100 per group | Production 2026-10-03: 2–3% at N = 100, 3.6–5.7% at N = 300; finite-sample follow-up open |
| Nested ML restricted-mean meat | Exact likelihood rows at the larger fit | Adopted; calibration open | Structured-mean lane pilot checks mechanics and prices held-out production |
| Global ML geometry | Expected | Outside this decision | No change proposed |

## Question

How well does the adopted observed nested ML geometry calibrate, in particular
when the larger model is misspecified, and how does the expected geometry
compare? Since 2026-10-03 the observed geometry is fixed by the
misspecification-robust requirement ([scope](../../../project/scope.md));
this study is calibration evidence, not a selection.

## Short answer

**Under strong misspecification of the larger model, the expected geometry's
LR test is inconsistent and the observed geometry fixes it.** With an omitted
.4 cross-loading (population RMSEA .127), expected-geometry LR-PEBA4 rejects
13.5% and 11.6% of true nulls at N = 100 and 300 per group (normal data),
against 7.8% and 6.4% for the observed geometry; skewed data show the same
pattern (14.9%/10.4% against 10.2%/6.9%). The expected excess barely shrinks
with N; the observed one does.

**The observed score test is conservative in small samples.** At N = 100 per
group it rejects 2.0–3.3% where the expected-sensitivity score rejects
4.2–6.3%; at N = 300 the gap is at most one point. This is what tripped the
registered rule (`reconsider_observed`, mild and correct cells, score arms).
Under the 2026-10-03 amendment it calls for finite-sample work on the observed
score, not a return to expected sensitivity. Under strong misspecification both
score geometries stay near 5%.

The LR tests over-reject with skewed data at N = 100 in every geometry
(10–16%), consistent with the score-first reporting decision. Seven of 36,000
draws (strong cells) produced no LR test: the larger model converged to a worse
local optimum than the restricted model (lavaan reaches the same endpoints), so
the difference statistic is negative and the policy reports a typed failure.

```{r}
prod <- file.path('results', 'nested-geometry', 'production-2026-10-03')
if (!file.exists(file.path(prod, 'rates.csv')))
  stop('Production results absent. Run Rscript run_experiment.R --production --workers 4 --run-id production-2026-10-03.')
rates <- read.csv(file.path(prod, 'rates.csv'))
null <- rates[rates$role == 'null', ]
w <- reshape(null[, c('n', 'distribution', 'larger', 'test', 'method', 'geometry', 'rejection')],
  idvar = c('n', 'distribution', 'larger', 'test', 'method'), timevar = 'geometry',
  direction = 'wide')
w <- w[order(match(w$larger, c('correct', 'mild', 'strong')), w$test, w$method,
             w$distribution, w$n), ]
knitr::kable(data.frame(`Larger model` = w$larger, Test = w$test, Method = w$method,
  Data = w$distribution, `N per group` = w$n,
  `Expected (%)` = round(100 * w$rejection.expected, 2),
  `Observed (%)` = round(100 * w$rejection.observed, 2), check.names = FALSE),
  caption = 'Null rejection at nominal 5%, 2,000 replicates per cell. Wilson intervals and paired differences are in rates.csv and paired.csv.')
checks <- read.csv(file.path(prod, 'checks.csv'))
knitr::kable(data.frame(Draws = checks$draws, `Failed draws` = checks$failed_draws,
  `Failed arms` = checks$failed_arms, `Maximum policy gap` = checks$max_policy_gap,
  `Registered rule` = read.csv(file.path(prod, 'decisions.csv'))$status,
  check.names = FALSE))
```

## Evidence and decision rule

The registered [criteria](criteria/nested_geometry.md) compare expected and
observed score and LR quadratics, each with SB and PEBA4. A two-group CFA tests
six loading restrictions. Equal population loadings make the null hold both
under the correct larger model and under mild and strong misspecification shared by both groups. Mild omits a
residual correlation of .3; strong additionally omits a standardized .4
cross-loading of x5 on f1, with residual variance adjusted to keep variance 1. Normal and Vale–Maurelli skewed indicators are studied at 100
and 300 observations per group; power lowers group b's standardized x3 loading
by .15.

Retain the observed geometry unless, in some misspecified-larger-model null cell, its PEBA4 or SB size error exceeds the expected geometry's by more than 1 percentage point with a paired 95% interval excluding zero. Report correct-model cells as controls; there both geometries are consistent. Size-adjusted power is reported, not gating.

Per-cell Wilson intervals, paired bootstrap intervals, failures, matching-null
p-value critical values and size-adjusted power are in the run's summary CSVs.
Production has 2000 null and 1000 power replicates per cell (36000 draws).
Both fits use ordinary default ML estimation and library convergence verdicts;
the quadratics reuse one shared prepared inference dataset.

## Population misfit and geometry gaps (pilot)

```{r}
out <- file.path('results', 'nested-geometry', 'pilot-strong')
```

The exact-covariance configural fits quantify misspecification independently of
sampling noise. Population RMSEA uses the limiting ML discrepancy; residuals
are standardized by population marginal standard deviations.

```{r}
population <- read.csv(file.path(out, 'population.csv'))
knitr::kable(data.frame(Level = population$larger,
  `Population RMSEA` = round(population$population_rmsea, 4),
  `Largest Standardized Residual` = round(population$largest_standardized_residual, 4)),
  check.names = FALSE)
```

Paired diagnostics are stored per cell in `geometry_gaps.csv`. The table shows
ranges of cell medians across the four null cells per level; power cells remain
in the CSV. Relative score differences divide the absolute observed–expected
statistic difference by the absolute expected statistic. These gaps describe
how much the geometries differ, rather than which calibration is better.

```{r}
gaps <- read.csv(file.path(out, 'geometry_gaps.csv'))
null_gaps <- gaps[gaps$role == 'null', ]
range_text <- function(x) paste(sprintf('%.4f', range(x)), collapse = '–')
gap_table <- do.call(rbind, lapply(population$larger, function(level) {
  x <- null_gaps[null_gaps$larger == level, ]
  data.frame(Level = level,
    `Score PEBA4 Gap` = range_text(x$median_absolute_peba4_gap[x$test == 'score']),
    `LR PEBA4 Gap` = range_text(x$median_absolute_peba4_gap[x$test == 'lr']),
    `Relative Score Difference` = range_text(x$median_relative_score_gap[x$test == 'score']),
    check.names = FALSE)
}))
knitr::kable(gap_table)
```

The original 16-cell pilot remains frozen under `results/nested-geometry/pilot`.
The amended pilot preserves those cells' seeds and adds strong cells 17–24;
therefore its existing-cell draws overlap the original pilot. Neither pilot is
confirmatory evidence. Frozen population and geometry summaries document the
amendment and its compute pricing; raw replicate batches remain ignored.

## Caveats

The pilots' twenty draws per cell assess mechanics and timing only; the
production run (2,000 null and 1,000 power replicates per cell) is the
evidence. One model family (two-group, two-factor CFA, six loading
restrictions) is covered; scalar invariance, larger df and incomplete data
are not. Four-core pricing assumes ideal scaling; worker startup,
summary computation adds time. No Modal price model
was available in this session, so the estimate is CPU-hours only. Generator
calibration is included in the measured per-draw cost. Complete-pair inference
conditions on available tests; all failures remain separately visible.

## Reproduce

Use the installed lane-c library; one math thread per worker:

```bash
export R_LIBS="$HOME/.cache/magmaan-rlib/lane-c"
export OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1
nice -n 10 Rscript run_experiment.R --help
nice -n 10 Rscript run_experiment.R --smoke --run-id new_smoke
nice -n 10 Rscript run_experiment.R --pilot --run-id new_pilot --workers 4
quarto render report.qmd
```

The production run of 2026-10-03 used
`nice -n 10 Rscript run_experiment.R --production --workers 4 --run-id production-2026-10-03`
(about 18 minutes on four local cores).
Each run uses a fresh output directory and separate mode seeds. Raw batches
stay local; frozen pilot summaries and source/native fingerprints provide
pricing provenance. Earlier failed smoke attempts remain local diagnostics.

## Small-sample score mechanism replay

The [task-73 diagnostic](diagnostics/note.md) replays 200 production draws in
each N=100 correct/mild normal/skewed null cell. The observed statistic's
variance is 34–58% below the expected-sensitivity comparator's, while raw meat
traces closely match Monte Carlo projected-score covariance traces. Freezing
population sensitivity recovers part of the variance; fixing the population
metric recovers more. This implicates fitted projection and studentization,
with additional reference-shape error under skewness. It motivates a registered
nuisance-refitting bootstrap or higher-order moment study, not a scalar meat
correction or a default change. Frozen summaries and provenance are under
`results/score-diagnostic/frozen-2026-10-05/`; the note describes normalization,
uncertainty and reproduction.


## Restricted-mean misspecification lane (TASK-75)

The [structured-mean registration draft](criteria/structured_mean.md) extends
this study beyond saturated means. Equal observed intercepts and fixed zero
latent means in the larger model leave group mean differences unmodeled.
Symmetric shifts of .2 SD on one item or .4 SD on two items preserve the
loading-equality null at the symmetric pseudo-true solution. Exact-moment
population fits verify this local null and positive observed curvature; global
uniqueness is not established. Power changes one group loading by .15 and has
a verified nonzero pseudo-true loading departure.

The full-grid twenty-draw pilot completed 1,440 draws, but one draw
(cell 2, replicate 16, seed 1526120047) returned an unexplained R/native
conversion error (`type=character; target=double`) after both fits converged.
All three arms for that draw are retained as unavailable. A fresh serial replay,
a fresh two-worker cell replay, and a replay including the population-check
setup all succeeded; the original failure remains frozen and unresolved.
The runner exited with its failure guard. Registration acceptance and production
are blocked pending the maintainer decision on this failure.

The full-grid twenty-draw pilot checks construction and prices production;
it cannot establish size or choose a default. Policy LR and score SB/PEBA4
are compared with an explicit pre-66 LR meat, using centered moment influence
without the fitted-mean likelihood correction. An independent exact-row
reconstruction checks that the observed bread and restriction projection agree
with the policy spectrum. The flag rule remains the parent's paired size-error
rule; a flag calls for finite-sample work on the adopted policy. The merger must
review this draft before any held-out production on Modal.

```{r}
structured <- file.path('results','structured-mean','pilot-2026-10-05')
if (!file.exists(file.path(structured,'checks.csv')))
  stop('Structured-mean pilot absent: run Rscript run_experiment.R --lane structured-mean --pilot --workers 2 --run-id pilot-2026-10-05')
sc <- read.csv(file.path(structured,'checks.csv'))
st <- read.csv(file.path(structured,'timing.csv'))
knitr::kable(data.frame(Draws=sc$draws, `Failed arms`=sc$failed_arms,
  `Maximum spectrum discrepancy`=signif(sc$max_reference_gap,3),
  `Projected production CPU hours`=round(sum(st$production_cpu_hours),2),
  `Ideal two-worker hours`=round(sum(st$production_cpu_hours)/2,2),
  check.names=FALSE))
```

The frozen summaries cover all 72 cells: one-factor/six-indicator and
two-factor/twelve-indicator models, normal/skewed data, N=100/300/1000 per group,
correct/mild/strong mean restrictions, and matched null/power. Production is
108,000 draws (2,000 null and 1,000 power per cell). CPU pricing includes the
explicit numerical-Jacobian comparator and generator; ideal scaling excludes
startup, summaries, cloud overhead and monetary rates. Per-cell rates, Wilson
intervals, paired flags, failures and population checks remain in the frozen
CSVs. Pilot flags are development diagnostics and the decision stays open.

```bash
export R_LIBS="$HOME/.cache/magmaan-rlib/lane-e"
export OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1
nice -n 10 Rscript run_experiment.R --lane structured-mean --smoke --workers 2 --run-id new_structured_smoke
nice -n 10 Rscript run_experiment.R --lane structured-mean --pilot --workers 2 --run-id new_structured_pilot
```

No production run was launched by TASK-75. Prior saturated-mean evidence and
its conclusions remain separate from this new calibration question.
