---
title: "Nested ML geometry calibration"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Default register

| Choice | Current value | Status | Evidence |
|---|---|---|---|
| Nested ML score projection and LR spectrum | Observed geometry | Adopted; confirmatory calibration open | Amended 24-cell pilot checks construction, population misfit and geometry gaps; production pending |
| Global ML geometry | Expected | Outside this decision | No change proposed |

## Question

Does observed nested ML geometry calibrate at least as well as expected geometry,
particularly when the larger model is misspecified?

## Short answer

Confirmation has not run. The pilot prices the amended 24-cell study and checks
that the observed arm reproduces the ordinary policy. Strong misspecification raises population RMSEA to .1274 (mild: .0587) and
its null-cell median LR PEBA4 gaps span .0331–.0863, compared with
.0054–.0177 for mild. Score gaps still overlap across levels. Its rejection
rates cannot justify a geometry decision.

```{r}
out <- file.path('results', 'nested-geometry', 'pilot-strong')
if (!file.exists(file.path(out, 'timing.csv')))
  stop('Pilot results absent. Run Rscript run_experiment.R --pilot --run-id pilot-strong --workers 4 with the lane library.')
timing <- read.csv(file.path(out, 'timing.csv'))
checks <- read.csv(file.path(out, 'checks.csv'))
metadata <- read.csv(file.path(out, 'metadata.csv'))
hours <- sum(timing$production_cpu_hours)
knitr::kable(data.frame(`Production CPU Hours` = round(hours, 2),
  `Ideal Four Core Hours` = round(hours / 4, 2), `Pilot Draws` = checks$draws,
  `Failed Arms` = checks$failed_arms, `Maximum Policy Gap` = checks$max_policy_gap),
  check.names = FALSE)
```

## Evidence and decision rule

The registered [criteria](criteria/nested_geometry.md) compare expected and
observed score and LR quadratics, each with SB and PEBA4. A two-group CFA tests
six loading restrictions. Equal population loadings make the null hold both
under the correct larger model and under mild and strong misspecification shared by both groups. Mild omits a
residual correlation of .3; strong additionally omits a standardized .4
cross-loading of x5 on f1, with residual variance adjusted to keep variance 1. Normal and Vale–Maurelli skewed indicators are studied at 100
and 300 observations per group; power lowers group b's standardized x3 loading
by .15.

Retain the observed geometry unless, in some misspecified-larger-model null cell, its PEBA4 or SB size error exceeds the expected geometry's by more than 1 percentage point with a paired 95% interval excluding zero. Report correct-model cells as controls; there both geometries are consistent. Size-adjusted power is reported, not gating.

Per-cell Wilson intervals, paired bootstrap intervals, failures, matching-null
p-value critical values and size-adjusted power are in the run's summary CSVs.
Production has 2000 null and 1000 power replicates per cell (36000 draws).
Both fits use ordinary default ML estimation and library convergence verdicts;
the quadratics reuse one shared prepared inference dataset.

## Population misfit and geometry gaps

The exact-covariance configural fits quantify misspecification independently of
sampling noise. Population RMSEA uses the limiting ML discrepancy; residuals
are standardized by population marginal standard deviations.

```{r}
population <- read.csv(file.path(out, 'population.csv'))
knitr::kable(data.frame(Level = population$larger,
  `Population RMSEA` = round(population$population_rmsea, 4),
  `Largest Standardized Residual` = round(population$largest_standardized_residual, 4)),
  check.names = FALSE)
```

Paired diagnostics are stored per cell in `geometry_gaps.csv`. The table shows
ranges of cell medians across the four null cells per level; power cells remain
in the CSV. Relative score differences divide the absolute observed–expected
statistic difference by the absolute expected statistic. These gaps describe
how much the geometries differ, rather than which calibration is better.

```{r}
gaps <- read.csv(file.path(out, 'geometry_gaps.csv'))
null_gaps <- gaps[gaps$role == 'null', ]
range_text <- function(x) paste(sprintf('%.4f', range(x)), collapse = '–')
gap_table <- do.call(rbind, lapply(population$larger, function(level) {
  x <- null_gaps[null_gaps$larger == level, ]
  data.frame(Level = level,
    `Score PEBA4 Gap` = range_text(x$median_absolute_peba4_gap[x$test == 'score']),
    `LR PEBA4 Gap` = range_text(x$median_absolute_peba4_gap[x$test == 'lr']),
    `Relative Score Difference` = range_text(x$median_relative_score_gap[x$test == 'score']),
    check.names = FALSE)
}))
knitr::kable(gap_table)
```

The original 16-cell pilot remains frozen under `results/nested-geometry/pilot`.
The amended pilot preserves those cells' seeds and adds strong cells 17–24;
therefore its existing-cell draws overlap the original pilot. Neither pilot is
confirmatory evidence. Frozen population and geometry summaries document the
amendment and its compute pricing; raw replicate batches remain ignored.

## Caveats

Twenty draws per cell assess mechanics and timing, not one-percentage-point
size differences. Four-core pricing assumes ideal scaling; worker startup,
summary computation adds time. No Modal price model
was available in this session, so the estimate is CPU-hours only. Generator
calibration is included in the measured per-draw cost. Complete-pair inference
conditions on available tests; all failures remain separately visible.

## Reproduce

Use the installed lane-c library; one math thread per worker:

```bash
export R_LIBS="$HOME/.cache/magmaan-rlib/lane-c"
export OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1
nice -n 10 Rscript run_experiment.R --help
nice -n 10 Rscript run_experiment.R --smoke --run-id new_smoke
nice -n 10 Rscript run_experiment.R --pilot --run-id new_pilot --workers 4
quarto render report.qmd
```

After the task-43 compute decision, the production command is
`nice -n 10 Rscript run_experiment.R --production --workers 4`.
Each run uses a fresh output directory and separate mode seeds. Raw batches
stay local; frozen pilot summaries and source/native fingerprints provide
pricing provenance. Earlier failed smoke attempts remain local diagnostics.
