---
title: "Ordinal missingness: mechanism and test-reference diagnostics"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
read_result <- function(name) {
  p <- file.path("results",name)
  if (!file.exists(p)) stop("Missing retained output: ",p,"; see Reproduce.")
  read.csv(p, stringsAsFactors=FALSE)
}
s <- read_result("summary.csv")
p <- read_result("p_values.csv")
m <- read_result("metadata.csv")
value <- function(key) m$value[match(key,m$key)]
```

## Question

Can matched MCAR/MAR cells distinguish a biased ordinal summary target from
failure of the nested-test reference law?

## Short answer

The retained output is **only a three-attempt smoke run per cell**, with total
$N=300$. It cannot establish nominal MCAR size, quantify MAR inflation or rank
p-value corrections. The earlier report's larger-run numerical claims are not
supported by the artifacts now present. Keep this study **banked**: reopen for
a named missing-data consumer after recovering the original substantive run
or independently validating estimator targets and bias.

## Evidence

The design uses a ten-indicator one-factor scalar-invariance comparison.
Group-B items v8–v10 have either uniformly random missingness or deletion
through a factor-correlated auxiliary; complete data is a shared baseline.
The runner applies alternative p-value laws to the same estimated summaries.
It preserves failed p-value attempts and missingness diagnostics.

The scaled-shifted method supplies a small routing/availability witness:

```{r witness}
x <- s[s$pd_gamma == "overlap" & s$method == "scaled_shifted", ]
knitr::kable(data.frame(Mechanism=toupper(x$missing_mechanism),
  `Missing fraction`=x$missing_rate, Attempts=x$reps, Usable=x$usable,
  Failures=x$failures, `Mean difference statistic`=x$mean_T_diff,
  check.names=FALSE), digits=3, row.names=FALSE)
```

The full method battery stays in `results/summary.csv`. Changes in the estimated
summary's limit and changes in the reference law are distinct possible causes;
this pilot does not identify either cause. A follow-up should first estimate
and validate each mechanism-specific pseudo-target, then calibrate around a
justified null instead of assuming the generating invariant model is the null
for every missing-data estimator.

## Caveats

A rejection rate from three draws moves in steps of one third. At 5% its
binomial MCSE would be about 12.6 percentage points; a reported zero does not
show exact size. This local run has no power arm and cannot establish a
practical missing-data recommendation. Mplus-parity evidence lives in a
separate retained study; it is not read by this runner or report.

## Reproduce

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --results-dir /tmp/ordinal_missingness_check
quarto render report.qmd
```

No larger run is queued. Use a fresh output directory for verification and
recover original substantive artifacts before treating old numerical claims
as evidence.
