---
title: "Two Small-Sample Defects of Distribution-Free Intervals, and Two Corrections That Compose"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
---

```{r}
res_dir <- params$results_dir
needed <- file.path(res_dir, c("coverage.csv", "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/44-alpha-kc-coverage/run_experiment.R",
       call. = FALSE)
}
d <- read.csv(file.path(res_dir, "coverage.csv"))
reps <- max(d$reps)
fmt <- function(x, k = 3) formatC(x, format = "f", digits = k)
rhos <- sort(unique(d$param[d$quantity == "correlation"]))
ns <- sort(unique(d$n)); nlo <- min(ns)
mlab <- c(wald = "raw + z (Wald)", kc = "raw + KC t",
          twald = "transform + z", tkc = "transform + KC t")
mcol <- c(wald = "#b2182b", kc = "#ef8a62", twald = "#67a9cf", tkc = "#1b7837")
cor_pick <- function(rho, law, n, method, col)
  d[[col]][d$quantity == "correlation" & d$param == rho & d$law == law &
           d$n == n & d$method == method][1]
```

## The question and the answer

A distribution-free (ADF) Wald interval for a covariance functional $g(S)$ -- a
correlation, Cronbach's alpha -- undercovers at small $N$ for **two separate
reasons that need two separate corrections**:

1. **Variance of the variance.** The influence-function SE $se^2=\operatorname{var}(v)/n$
   is itself a noisy estimate, so the $z$ interval is too short. Kauermann &
   Carroll (2001): the deficit is governed by the kurtosis of the influence
   values; the fix is a $t$ on effective df $f=2N/(\hat\kappa-1)$. This hits the
   **two-sided** coverage and is roughly **constant in the correlation $\rho$**.
2. **Skewness / boundedness.** The raw estimate is skewed and range-bounded, so a
   symmetric interval misses asymmetrically. The fix is a variance-stabilizing
   transform (Fisher's $z$ for a correlation, the logit for a reliability). This
   hits the **left/right balance** of the misses, and the imbalance **grows with
   $\rho$**.

Sweeping $\rho$ pulls the two apart. Raw two-sided coverage is flat in $\rho$,
but its miss turns almost entirely one-sided: at $N=`r nlo`$ (normal) the raw
Wald interval misses below/above the true $\rho$ about
`r fmt(1000*cor_pick(0.9,"normal",nlo,"wald","miss_lower"),0)`/`r fmt(1000*cor_pick(0.9,"normal",nlo,"wald","miss_upper"),0)`
per 1000 at $\rho=0.9$ versus a balanced
`r fmt(1000*cor_pick(0,"normal",nlo,"wald","miss_lower"),0)`/`r fmt(1000*cor_pick(0,"normal",nlo,"wald","miss_upper"),0)`
at $\rho=0$. The transform keeps the two sides balanced at every $\rho$ but does
not lift the total; the KC $t$ lifts the total but stays imbalanced. **Only the
combination (transform + KC $t$) is both near-nominal and balanced**, across
$\rho$ and $N$, for the correlation and for alpha alike. The two corrections are
orthogonal and they compose.

```{r, fig.width=9, fig.height=4.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.4, 1), cex = 0.9)
meths <- names(mlab)
# Panel 1: two-sided coverage vs rho
cov_at <- function(m) sapply(rhos, function(r) cor_pick(r, "normal", nlo, m, "coverage"))
plot(NA, xlim = range(rhos), ylim = c(0.80, 0.97), xlab = expression(rho),
     ylab = "two-sided coverage", main = sprintf("Coverage (normal, N=%d)", nlo))
abline(h = 0.95, lty = 3, col = "grey50")
for (m in meths) lines(rhos, cov_at(m), type = "b", pch = 19, col = mcol[[m]])
legend("bottomleft", bty = "n", lwd = 2, col = mcol[meths], legend = mlab[meths], cex = 0.85)
# Panel 2: signed one-sidedness (miss-below minus miss-above) vs rho
imb_at <- function(m) sapply(rhos, function(r)
  cor_pick(r, "normal", nlo, m, "miss_lower") - cor_pick(r, "normal", nlo, m, "miss_upper"))
yl <- range(0, sapply(meths, imb_at))
plot(NA, xlim = range(rhos), ylim = yl, xlab = expression(rho),
     ylab = "P(true below) - P(true above)", main = sprintf("Imbalance (normal, N=%d)", nlo))
abline(h = 0, lty = 3, col = "grey50")
for (m in meths) lines(rhos, imb_at(m), type = "b", pch = 19, col = mcol[[m]])
par(op)
```

Left: two-sided coverage is flat in $\rho$; the raw + KC $t$ (orange) and
transform + KC $t$ (green) lift it, the latter most. Right: the signed imbalance
of the misses grows with $\rho$ for the **raw** intervals (red, orange) and stays
near zero for the **transform** intervals (blue, green). The KC $t$ moves points
up in the left panel; the transform moves them toward zero in the right panel.
The two corrections act on different panels.

## Two-sided coverage across $N$

```{r}
laws <- c("normal", "contaminated")
rho_hi <- max(rhos[rhos <= 0.8])
rowblock <- function(q, param, plab) {
  do.call(rbind, lapply(laws, function(law) do.call(rbind, lapply(ns, function(n) {
    g <- function(m) d$coverage[d$quantity == q &
      (if (is.na(param)) is.na(d$param) else d$param == param) &
      d$law == law & d$n == n & d$method == m][1]
    data.frame(Quantity = plab, Law = law, N = n,
      `raw+z` = fmt(g("wald")), `raw+KC` = fmt(g("kc")),
      `tr+z` = fmt(g("twald")), `tr+KC` = fmt(g("tkc")),
      check.names = FALSE, stringsAsFactors = FALSE)
  }))))
}
tab <- rbind(rowblock("correlation", rho_hi, sprintf("correlation (rho=%.1f)", rho_hi)),
             rowblock("alpha", NA, "alpha"))
knitr::kable(tab, align = "lllrrrr")
```

Two-sided 95% coverage (`r reps` reps). Reading across the row, `raw+z` is lowest,
the KC $t$ and the transform each help, and `tr+KC` is the best in nearly every
cell; the gains are largest under contaminated data (heavier influence kurtosis)
and at small $N$. Alpha (logit transform) shows the same ordering, with a smaller
transform effect since $\alpha\approx0.85$ is less boundary-pressed than a high
correlation.

## How much to trust it

- **The two axes are real, not an artifact.** The $\rho$ sweep separates them
  cleanly: the KC gain is ~flat in $\rho$ (it is a kurtosis effect), the
  imbalance is ~zero at $\rho=0$ and grows with $\rho$ (a skew effect). Neither
  correction touches the other's panel.
- **The corrections compose for a free reason.** The transform rescales every
  influence value by one constant, which leaves the kurtosis -- hence the KC
  effective df -- unchanged, so "transform then correct" and "correct then
  transform" are the same interval. The df lives in the quantile; the transform
  lives in the geometry.
- **Honest residual.** Even `tr+KC` undercovers somewhat at $N\le 30$ under heavy
  tails: $\hat\kappa$ is a fourth moment estimated from few cases, and the ADF SE
  is biased low there (a meat-debiasing job, Bell-McCaffrey, not a df or transform
  job). No distribution-free interval recovers fourth-moment information the
  sample does not contain.

## Reproduce

```text
Rscript experiments/44-alpha-kc-coverage/run_experiment.R          # full grid
Rscript experiments/44-alpha-kc-coverage/run_experiment.R --smoke  # quick check
quarto render experiments/44-alpha-kc-coverage/report.qmd
```

Background and the reference set (Kauermann-Carroll, Bell-McCaffrey,
Imbens-Kolesár, Yuan-Bentler, Satterthwaite) live in
`external/refs/corrections/`.
