---
title: "A Variance-of-Variance Correction for Distribution-Free Reliability and Correlation Intervals"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
---

```{r}
res_dir <- params$results_dir
needed <- file.path(res_dir, c("coverage.csv", "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/44-alpha-kc-coverage/run_experiment.R",
       call. = FALSE)
}
d <- read.csv(file.path(res_dir, "coverage.csv"))
reps <- max(d$reps)
fmt <- function(x, k = 3) formatC(x, format = "f", digits = k)
pick <- function(q, law, n, method, col)
  d[[col]][d$quantity == q & d$law == law & d$n == n & d$method == method][1]
ns <- sort(unique(d$n)); nlo <- min(ns); nhi <- max(ns)
quants <- c("alpha", "correlation")
qlab <- c(alpha = "Cronbach's alpha", correlation = "Pearson correlation")
laws <- c("normal", "contaminated")
```

## The question and the answer

Cronbach's alpha and the Pearson correlation are both smooth covariance
functionals $g(S)$, so each has a distribution-free (ADF) Wald interval built on
the influence-function SE, $se^2 = \operatorname{var}(v)/n$ with
$v_k = d_k' G\, d_k$ the per-case influence values ($d_k = x_k-\bar x$,
$G = \partial g/\partial\Sigma$). Both **undercover at small $N$**, for the reason
**Kauermann & Carroll (2001)** identified: the variance estimate is itself noisy,
with relative variability set by the *kurtosis of the influence values*, which is
large because the $v_k$ are quadratic forms (heavy-tailed **even when the data are
normal**).

Replacing the normal quantile with a $t$ on Kauermann-Carroll **effective degrees
of freedom**, $f = 2N/(\hat\kappa-1)$ computed from the same $v_k$, restores most
of the coverage, for **both** functionals and both data laws. At $N=`r nlo`$:

| | Wald (z) | KC (t) |
|---|--:|--:|
| alpha, normal | `r fmt(pick("alpha","normal",nlo,"wald","coverage"))` | **`r fmt(pick("alpha","normal",nlo,"kc","coverage"))`** |
| alpha, contaminated | `r fmt(pick("alpha","contaminated",nlo,"wald","coverage"))` | **`r fmt(pick("alpha","contaminated",nlo,"kc","coverage"))`** |
| correlation, normal | `r fmt(pick("correlation","normal",nlo,"wald","coverage"))` | **`r fmt(pick("correlation","normal",nlo,"kc","coverage"))`** |
| correlation, contaminated | `r fmt(pick("correlation","contaminated",nlo,"wald","coverage"))` | **`r fmt(pick("correlation","contaminated",nlo,"kc","coverage"))`** |

By $N=50$-$100$ the corrected interval is essentially nominal. It needs **no fit
and no bootstrap**, only the influence values already used for the SE, and it is
**design-free** (the same two lines serve any covariance functional). One honest
asymmetry: the correlation residual at the smallest $N$ is larger than alpha's,
because raw-scale $r$ is skewed and bounded, and the symmetric $t$-widening does
not repair skewness. That part is the job of a variance-stabilizing transform
(Fisher's $z$ for $r$, the logit for a reliability coefficient), complementary to
the KC correction, not a substitute for it.

```{r, fig.width=8, fig.height=7}
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2.4, 1), cex = 0.9)
for (q in quants) for (law in laws) {
  cw <- sapply(ns, function(n) pick(q, law, n, "wald", "coverage"))
  ck <- sapply(ns, function(n) pick(q, law, n, "kc", "coverage"))
  yl <- range(0.74, 0.97, cw, ck)
  plot(ns, ck, type = "b", pch = 19, col = "#1b7837", ylim = yl, log = "x",
       xlab = "N (log scale)", ylab = "95% CI coverage",
       main = sprintf("%s, %s data", qlab[[q]], law), xaxt = "n")
  axis(1, at = ns, labels = ns)
  abline(h = 0.95, lty = 3, col = "grey50")
  lines(ns, cw, type = "b", pch = 17, col = "#b2182b")
  legend("bottomright", bty = "n", pch = c(17, 19),
         col = c("#b2182b", "#1b7837"), legend = c("Wald (z)", "KC (t, eff. df)"))
}
par(op)
```

Coverage of the nominal-95% interval (`r reps` replications per cell). The
Kauermann-Carroll interval (green) tracks the 0.95 line far more closely than the
plain Wald interval (red), for both quantities, both data laws, at every $N$.

## The numbers

```{r}
w <- d[d$method == "wald", c("quantity", "law", "n", "coverage", "mean_kappa", "mean_se", "mc_sd")]
kc <- d[d$method == "kc", c("quantity", "law", "n", "coverage", "mean_df")]
m <- merge(w, kc, by = c("quantity", "law", "n"), suffixes = c("_wald", "_kc"))
m <- m[order(match(m$quantity, quants), match(m$law, laws), m$n), ]
tab <- data.frame(
  Quantity = qlab[m$quantity], Law = m$law, N = m$n,
  `Wald cover` = fmt(m$coverage_wald),
  `KC cover`   = fmt(m$coverage_kc),
  `KC eff. df` = fmt(m$mean_df, 1),
  `infl. kurtosis` = fmt(m$mean_kappa, 1),
  `SE / MC-SD` = fmt(m$mean_se / m$mc_sd, 2),
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE)
knitr::kable(tab, align = "lllrrrrr")
```

The **influence kurtosis** is well above the normal value of 3 even under normal
data (the $v_k$ are quadratic forms) and grows under contamination, so the
**effective df** $f=2N/(\hat\kappa-1)$ sits far below $N$. The **SE / MC-SD**
column is the downward bias of the ADF SE: near 1 at large $N$ (the SE is right),
but well below 1 at small $N$, which the df-widening compensates for but does not
formally repair.

## How much to trust it

- **Mechanism is verified, not assumed.** The SE/MC-SD ratio reaches ~1 at
  $N=`r nhi`$ for both quantities, confirming the influence-function SE is the
  correct ADF SE (for the correlation it reproduces the textbook
  $(1-r^2)^2/N$). The small-$N$ undercoverage is the Kauermann-Carroll
  variance-of-the-variance effect, plus an SE downward bias, plus point bias.
- **The correction is the cited one, not ad hoc.** $f=2N/(\hat\kappa-1)$ is the
  Satterthwaite realization of Kauermann & Carroll's quantile adjustment
  (undercoverage $=c_p\,\operatorname{var}(\hat\sigma^2)/\sigma^4$ with the
  relative variance equal to the influence kurtosis), and it is design-free, so
  alpha and the correlation use identical code.
- **Honest residual.** Under heavy tails at $N\le 30$ the corrected interval still
  undercovers, more so for the correlation: $\hat\kappa$ is itself a fourth moment
  estimated from few cases, the ADF SE is biased low there (the HC2/Bell-McCaffrey
  meat-debiasing's job, not the df's), and the raw-scale skew of $r$ is a separate
  defect a symmetric interval cannot fix. No distribution-free interval recovers
  fourth-moment information the sample does not contain.

## Reproduce

```text
Rscript experiments/44-alpha-kc-coverage/run_experiment.R          # full grid (~4s)
Rscript experiments/44-alpha-kc-coverage/run_experiment.R --smoke  # quick check
quarto render experiments/44-alpha-kc-coverage/report.qmd
```

Background and the reference set (Kauermann-Carroll, Bell-McCaffrey,
Imbens-Kolesár, Yuan-Bentler, Satterthwaite) live in
`external/refs/corrections/`.
