---
title: "A Variance-of-Variance Correction for the Distribution-Free Alpha Interval"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
---

```{r}
res_dir <- params$results_dir
needed <- file.path(res_dir, c("coverage.csv", "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/44-alpha-kc-coverage/run_experiment.R",
       call. = FALSE)
}
d <- read.csv(file.path(res_dir, "coverage.csv"))
meta <- read.csv(file.path(res_dir, "metadata.csv"))
true_alpha <- d$true_alpha[1]
reps <- max(d$reps)
fmt <- function(x, k = 3) formatC(x, format = "f", digits = k)
pick <- function(law, n, method, col)
  d[[col]][d$law == law & d$n == n & d$method == method][1]
ns <- sort(unique(d$n)); nlo <- min(ns); nhi <- max(ns)
```

## The question and the answer

Cronbach's alpha is a covariance functional, $\alpha = \tfrac{p}{p-1}\bigl(1-\operatorname{tr}(S)/\mathbf 1'S\mathbf 1\bigr)$,
so its distribution-free (ADF) Wald interval **undercovers at small $N$**, for the
reason **Kauermann & Carroll (2001)** identified: the variance estimate is itself
noisy, with relative variability set by the *kurtosis of alpha's per-case
influence values* $v_k=d_k'\,G\,d_k$ (with $d_k=x_k-\bar x$, $G=\partial\alpha/\partial\Sigma$).
That kurtosis is large because the influence values are quadratic forms, so they
are heavy-tailed **even when the data are normal**.

Replacing the normal quantile with a $t$ on Kauermann-Carroll **effective degrees
of freedom**, $f = 2N/(\hat\kappa-1)$ with $\hat\kappa$ the influence kurtosis,
computed from the same $v_k$, recovers most of the gap. At $N=`r nlo`$ it lifts
coverage from `r fmt(pick("normal",nlo,"wald","coverage"))` to
`r fmt(pick("normal",nlo,"kc","coverage"))` under normal data and from
`r fmt(pick("contaminated",nlo,"wald","coverage"))` to
`r fmt(pick("contaminated",nlo,"kc","coverage"))` under contaminated-normal data;
by $N=50$-$100$ the corrected interval is essentially nominal in both. It needs
**no fit and no bootstrap**, only the influence values already used for the SE.

It does **not** fully close the gap at the very smallest $N$ under heavy tails.
Two things the df-widening cannot fix live there: alpha's own downward point bias,
and a downward-biased ADF SE (the mean SE is only `r fmt(100*pick("contaminated",nlo,"mean_se","mean_se")/pick("contaminated",nlo,"mean_se","mc_sd"),0)`%
of the Monte-Carlo SD at $N=`r nlo`$ under contamination). The Kauermann-Carroll
correction targets the variance-of-the-variance; closing the rest would need the
HC2-style meat de-biasing (Bell-McCaffrey) and is bounded below by the point bias.

```{r, fig.width=8, fig.height=3.8}
laws <- c("normal", "contaminated")
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.4, 1), cex = 0.9)
for (law in laws) {
  cw <- sapply(ns, function(n) pick(law, n, "wald", "coverage"))
  ck <- sapply(ns, function(n) pick(law, n, "kc", "coverage"))
  plot(ns, ck, type = "b", pch = 19, col = "#1b7837", ylim = c(0.80, 0.97),
       log = "x", xlab = "N (log scale)", ylab = "95% CI coverage",
       main = sprintf("%s data", law), xaxt = "n")
  axis(1, at = ns, labels = ns)
  abline(h = 0.95, lty = 3, col = "grey50")
  lines(ns, cw, type = "b", pch = 17, col = "#b2182b")
  legend("bottomright", bty = "n", pch = c(17, 19),
         col = c("#b2182b", "#1b7837"), legend = c("Wald (z)", "KC (t, eff. df)"))
}
par(op)
```

Coverage of the nominal-95% interval for true $\alpha=`r fmt(true_alpha)`$ (`r reps`
replications per cell, $p=`r meta$p[1]`$ congeneric items). The Kauermann-Carroll
interval (green) tracks the 0.95 line far more closely than the plain Wald interval
(red), in both data laws, at every $N$.

## The numbers

```{r}
tab <- do.call(rbind, lapply(laws, function(law) do.call(rbind, lapply(ns, function(n) {
  data.frame(Law = law, N = n,
    `Wald cover` = fmt(pick(law, n, "wald", "coverage")),
    `KC cover`   = fmt(pick(law, n, "kc", "coverage")),
    `KC eff. df` = fmt(pick(law, n, "kc", "mean_df"), 1),
    `infl. kurtosis` = fmt(pick(law, n, "wald", "mean_kappa"), 1),
    `SE / MC-SD` = fmt(pick(law, n, "wald", "mean_se") / pick(law, n, "wald", "mc_sd"), 2),
    check.names = FALSE, stringsAsFactors = FALSE)
}))))
knitr::kable(tab, align = "llrrrrr")
```

Reading the table: the **influence kurtosis** is well above the normal value of 3
even under normal data (the influence values are quadratic forms), and it grows
under contamination, so the **effective df** $f=2N/(\hat\kappa-1)$ sits far below
$N$ (at $N=`r nhi`$ contaminated it is only ~`r fmt(pick("contaminated",nhi,"kc","mean_df"),0)`).
The **SE / MC-SD** column is the downward bias of the ADF SE: near 1 at large $N$
(the SE is right), but well below 1 at small $N$, which is the part the
df-widening compensates for but does not formally repair.

## How much to trust it

- **Mechanism is verified, not assumed.** The SE/MC-SD ratio reaching ~1 at
  $N=`r nhi`$ confirms the influence-function SE is the correct ADF SE; the
  undercoverage at small $N$ is the Kauermann-Carroll variance-of-the-variance
  effect, plus an SE downward bias, plus alpha's point bias.
- **The correction is the cited one, not ad hoc.** $f=2N/(\hat\kappa-1)$ is the
  Satterthwaite realization of Kauermann & Carroll's quantile adjustment (their
  undercoverage $=c_p\,\operatorname{var}(\hat\sigma^2)/\sigma^4$, with the
  relative variance equal to the influence kurtosis).
- **Honest residual.** Under heavy tails at $N\le 30$ the corrected interval still
  undercovers, because $\hat\kappa$ is itself a fourth moment estimated from few
  cases and because no quantile widening removes a biased point estimate. This is
  the expected floor: no distribution-free interval recovers fourth-moment
  information that the sample does not contain.

## Reproduce

```text
Rscript experiments/44-alpha-kc-coverage/run_experiment.R          # full grid (~2s)
Rscript experiments/44-alpha-kc-coverage/run_experiment.R --smoke  # quick check
quarto render experiments/44-alpha-kc-coverage/report.qmd
```

Background and the reference set (Kauermann-Carroll, Bell-McCaffrey,
Imbens-Kolesár, Yuan-Bentler, Satterthwaite) live in
`external/refs/corrections/`.
