---
title: "Ordinal inference cache probe"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "inference_workspace.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

workspace_path <- file.path(results_dir, "inference_workspace.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(workspace_path)) {
  stop("Missing inference_workspace.csv. Run: Rscript run_experiment.R --smoke")
}

workspace <- utils::read.csv(workspace_path, check.names = FALSE,
                             stringsAsFactors = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE,
                            stringsAsFactors = FALSE)

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.2f", x))
fmt_num <- function(x) ifelse(is.na(x), "", sprintf("%.4g", x))
fmt_diff <- function(x) ifelse(is.na(x), "", sprintf("%.2e", x))
max_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else max(x)
}
median_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else stats::median(x)
}

workspace$Path <- factor(
  workspace$path,
  levels = c("materialized_bounded_robust",
             "cache_fit_only_then_inference",
             "cache_fit_plus_inference_reuse"),
  labels = c("Materialized", "Fit-only + separate inference",
             "Fit-plus-inference reuse")
)
workspace$Thresholds <- factor(
  workspace$threshold_mode,
  levels = c("free", "fixed", "shared"),
  labels = c("Free", "Fixed", "Shared")
)
ok <- workspace[workspace$status == "ok", ]
```

## Question

This experiment is the inference companion to the ordinal fit-only workspace
probe. It asks whether carrying a full cache through fitting helps when robust
ordinal reporting is requested immediately after a bounded DWLS/WLS fit.

The compared paths are:

- `Materialized`: legacy `OrdinalStats` fit and robust reporting.
- `Fit-only + separate inference`: minimal fit cache, then a separate full
  inference cache for robust reporting.
- `Fit-plus-inference reuse`: one full cache prepared for the fit and reused
  by robust reporting.

## Short Answer

The smoke design should be read as a shape and equivalence check. Cache-aware
robust reporting matches the legacy materialized path, and the reuse path makes
the duplicated inference setup explicit: the separate path pays for an
additional full Gamma/weight cache, while the fit-plus-inference path reuses
the cache already attached to the fit.

## Run

```{r}
data.frame(
  Item = metadata$key,
  Value = metadata$value,
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

## Status

```{r}
status <- aggregate(
  seed ~ Thresholds + estimator + Path + status,
  data = workspace,
  FUN = length
)
names(status) <- c("Thresholds", "Estimator", "Path", "Status", "Rows")
knitr::kable(status, row.names = FALSE)
```

## Equivalence

```{r}
equiv <- aggregate(
  cbind(theta_diff_legacy, fmin_diff_legacy, se_diff_legacy,
        vcov_diff_legacy, chisq_diff_legacy) ~ Thresholds + estimator + Path,
  data = ok,
  FUN = max_or_na,
  na.action = na.pass
)
names(equiv) <- c("Thresholds", "Estimator", "Path",
                  "Max |theta - legacy|", "Max |fmin - legacy|",
                  "Max |SE - legacy|", "Max |vcov - legacy|",
                  "Max |chisq - legacy|")
for (j in 4:ncol(equiv)) equiv[[j]] <- fmt_diff(equiv[[j]])
knitr::kable(equiv, row.names = FALSE)
```

## Cost Shape

```{r}
cost <- aggregate(
  cbind(fit_cache_setup_ms, fit_weight_setup_ms,
        inference_cache_setup_ms, inference_weight_setup_ms,
        fit_median_ms, robust_median_ms) ~ estimator + Path,
  data = ok,
  FUN = median_or_na,
  na.action = na.pass
)
names(cost) <- c("Estimator", "Path", "Fit cache setup ms",
                 "Fit weight setup ms", "Inference cache setup ms",
                 "Inference weight setup ms", "Fit ms", "Robust ms")
for (j in 3:ncol(cost)) cost[[j]] <- fmt_ms(cost[[j]])
knitr::kable(cost, row.names = FALSE)
```

```{r}
#| fig-width: 8
#| fig-height: 4
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  plot_data <- ok
  plot_data$total_fit_robust_ms <-
    plot_data$fit_median_ms + plot_data$robust_median_ms
  ggplot(plot_data, aes(Path, total_fit_robust_ms, fill = estimator)) +
    geom_col(position = "dodge", width = 0.7) +
    facet_grid(Thresholds ~ p, labeller = label_both) +
    labs(x = NULL, y = "Median fit + robust milliseconds") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
} else {
  cat("ggplot2 not installed; skipping figure.")
}
```

The figure excludes cache setup columns so fit and robust reporting are not
mixed with one-time materialization. The setup table above is the better place
to see where duplicated full-Gamma or weight work appears.

## Cache Flags

```{r}
flags <- unique(ok[, c("estimator", "Path",
                      "fit_cache_has_diagonal", "fit_cache_has_full",
                      "fit_cache_has_dwls_weight", "fit_cache_has_wls_weight",
                      "inference_cache_has_diagonal",
                      "inference_cache_has_full",
                      "inference_cache_has_dwls_weight",
                      "inference_cache_has_wls_weight")])
names(flags) <- c("Estimator", "Path", "Fit diagonal", "Fit full",
                  "Fit DWLS weight", "Fit WLS weight",
                  "Inference diagonal", "Inference full",
                  "Inference DWLS weight", "Inference WLS weight")
knitr::kable(flags[order(flags$Estimator, flags$Path), ], row.names = FALSE)
```

For DWLS, fit-only fitting needs only the diagonal Gamma; robust reporting
needs full Gamma and a DWLS diagonal weight matrix. For WLS, both fitting and
robust reporting need the full WLS weight, so separate fit/inference caches
duplicate the inversion unless the fit-plus-inference cache is reused.

## Caveats

This is still an advisory synthetic benchmark. It covers bounded all-ordinal
delta DWLS/WLS with robust reporting. It does not cover ULS robust inference,
SNLLS followed by robust reporting, theta parameterization, mixed ordinal data,
or a raw-data lazy Gamma builder that avoids constructing legacy `OrdinalStats`
up front.

## Reproduce

```sh
Rscript experiments/_archive/10-ordinal-inference-cache-probe/run_experiment.R --smoke --preset fast
(cd experiments/_archive/10-ordinal-inference-cache-probe && quarto render report.qmd)
```

For less noisy timings, use `--preset opt` and increase `--reps`.
