---
title: "Li (2016) replication — ordinal SEM, ML vs DWLS"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "metadata.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
rd <- function(name) {
  p <- file.path(results_dir, name)
  if (!file.exists(p)) return(NULL)
  utils::read.csv(p, check.names = FALSE, stringsAsFactors = FALSE)
}
meta    <- rd("metadata.csv")
thr     <- rd("thresholds.csv")
sparam  <- rd("summary_param.csv")
schi2   <- rd("summary_chi2.csv")
sconv   <- rd("summary_conv.csv")
parity  <- rd("lavaan_parity.csv")
if (is.null(meta) || is.null(sparam) || is.null(schi2)) {
  stop("Missing result CSVs. Run: Rscript run_experiment.R")
}
pct  <- function(x, d = 1) ifelse(is.na(x), "", sprintf(paste0("%.", d, "f"), 100 * x))
num3 <- function(x) ifelse(is.na(x), "", sprintf("%.3f", x))
fam_lab <- c(loading = "λ loadings", factor_cor = "φ inter-factor",
             gamma = "γ paths", beta = "β paths")
shape_lab <- c(sym = "symmetric", slight = "slightly asym.",
               moderate = "moderately asym.")
```

This experiment replicates **Cheng-Hsien Li (2016), "The Performance of ML,
DWLS, and ULS Estimation With Robust Corrections in Structural Equation Models
With Ordinal Variables", *Psychological Methods* 21(3):369-387** against
magmaan. Li's population model is the same five-factor structural model used in
Li (2021) (experiment 16), but here *all 20 indicators are ordinal*. Two
estimators are fit on identical categorized data per replicate: **DWLS** (lavaan
WLSMV; polychoric correlations) and **ML** (robust ML treating the integer codes
as continuous). The fitted model is correctly specified, so χ² rejection rates
are Type I error rates.

## Question

When a five-factor structural model is measured entirely by ordinal indicators,
does treating them as categorical (DWLS on polychorics) or as continuous (robust
ML) recover loadings and structural paths better, and which controls χ² Type I
error, as the number of categories, distribution asymmetry, and sample size
vary?

## Short answer

magmaan reproduces Li's central contrasts. **DWLS recovers loadings and
structural paths with smaller bias than ML, and the gap widens with
asymmetry**; ML's downward bias on loadings is its signature failure. ML's
robust standard errors are at least as well calibrated as DWLS's. But on the
fit statistic the order reverses: **DWLS's mean-and-variance adjusted χ²
controls Type I error close to nominal, while robust ML over-rejects at small N
and few categories**, only catching up by N ≈ 1000 — Li's headline finding.

Both estimators match lavaan on the same data (`standardizedSolution`; WLSMV for
DWLS, MLM for ML) to the tolerances in the parity table.

## Design

```{r}
knitr::kable(meta, row.names = FALSE)
```

One population model (Li Figure 1): a five-factor SEM with two exogenous and
three endogenous factors, structural paths `Γ`/`B` and inter-factor correlation
φ₁₂ = .3, all latent variances standardized to 1. Each factor has four ordinal
indicators with standardized loadings .8/.7/.6/.5 (20 indicators total). The
design crosses three distribution shapes (Li Figure 2 response-probability
profiles) × number of categories {4, 5, 6, 7} × N {200, 300, 400, 500, 750,
1000, 1500} — 84 cells. Parameters and SEs are reported in the standardized
metric, split by family.

```{r}
if (!is.null(thr)) {
  t2 <- thr; t2$shape <- shape_lab[t2$shape]
  t2$skew_achieved <- num3(t2$skew_achieved)
  t2$kurt_achieved <- num3(t2$kurt_achieved)
  knitr::kable(t2[, c("shape", "categories", "proportions",
                      "skew_achieved", "kurt_achieved")],
               row.names = FALSE,
               col.names = c("Shape", "K", "Proportions", "Skew", "Ex. kurt."),
               caption = "Achieved marginal moments match Li's reported values (sym ~0/-.48, slight ~-.91/.82, moderate ~-1.39/1.17). Thresholds in results/thresholds.csv.")
}
```

## Evidence

### Parameter relative bias (%), by family

Averaged over replicates and same-family parameters; negative = underestimate.
Li's "trivial bias" band is |RB| < 5%, "substantial" is > 10%.

```{r}
if (!is.null(sparam)) {
  d <- sparam[sparam$N == 500 & sparam$param_type %in%
                c("loading", "gamma", "beta"), ]
  d$family <- fam_lab[d$param_type]
  w <- stats::reshape(
    d[, c("shape", "categories", "estimator", "param_type", "rel_bias")],
    idvar = c("shape", "categories", "param_type"),
    timevar = "estimator", direction = "wide")
  names(w) <- sub("^rel_bias\\.", "", names(w))
  w$DWLS <- pct(w$DWLS); w$ML <- pct(w$ML)
  w$family <- fam_lab[w$param_type]; w$shape <- shape_lab[w$shape]
  w <- w[order(match(w$shape, shape_lab), w$param_type, w$categories),
         c("shape", "family", "categories", "DWLS", "ML")]
  knitr::kable(w, row.names = FALSE,
               col.names = c("Shape", "Family", "K", "DWLS RB%", "ML RB%"),
               caption = "N = 500. Full grid in results/summary_param.csv.")
}
```

The signature contrast: **ML's loading and path bias grows with asymmetry while
DWLS stays near zero.** Bias attenuates with more categories and larger N for
both.

### Standard-error relative bias and coverage

```{r}
if (!is.null(sparam)) {
  d <- sparam[sparam$shape == "moderate" & sparam$N == 500, ]
  d$family <- fam_lab[d$param_type]
  d$se_rel_bias <- pct(d$se_rel_bias); d$coverage <- num3(d$coverage)
  d$rel_bias <- pct(d$rel_bias)
  d <- d[order(d$param_type, d$categories, d$estimator),
         c("categories", "estimator", "family", "rel_bias",
           "se_rel_bias", "coverage")]
  knitr::kable(d, row.names = FALSE,
               col.names = c("K", "Est.", "Family", "Param RB%", "SE RB%",
                             "Coverage"),
               caption = "Moderately asymmetric, N = 500.")
}
```

### χ² rejection rate (Type I error at α = .05)

DWLS uses the mean-and-variance adjusted statistic (lavaan WLSMV); ML the
Satorra-Bentler and mean-variance adjusted statistics. Bradley's band is
[.025, .075].

```{r}
if (!is.null(schi2)) {
  d <- schi2[schi2$stat_name == "MV" & schi2$shape == "moderate" &
               schi2$estimator %in% c("DWLS", "ML"), ]
  d$tag <- paste0(d$estimator, "-MV")
  w <- stats::reshape(d[, c("categories", "N", "tag", "reject_05")],
                      idvar = c("categories", "N"),
                      timevar = "tag", direction = "wide")
  names(w) <- sub("^reject_05\\.", "", names(w))
  for (nm in c("DWLS-MV", "ML-MV")) if (nm %in% names(w)) w[[nm]] <- num3(w[[nm]])
  w <- w[order(w$N, w$categories), ]
  knitr::kable(w, row.names = FALSE,
               caption = "Moderately asymmetric. Empirical rejection at α = .05; nominal .05.")
}
```

DWLS-MV sits near .05 across the grid; ML-MV runs high at small N and pulls
toward nominal as N grows.

### Convergence and improper solutions

```{r}
if (!is.null(sconv)) {
  d <- sconv
  d$conv_rate <- num3(d$conv_rate); d$improper_rate <- num3(d$improper_rate)
  d$shape <- shape_lab[d$shape]
  bad <- d[as.numeric(d$conv_rate) < 1 | as.numeric(d$improper_rate) > 0, ]
  if (nrow(bad)) {
    knitr::kable(bad[order(bad$N, bad$categories),
                     c("categories", "shape", "N", "estimator",
                       "conv_rate", "improper_rate")],
                 row.names = FALSE,
                 col.names = c("K", "Shape", "N", "Est.", "Conv.", "Improper"),
                 caption = "Cells with any nonconvergence or improper solutions.")
  } else {
    cat("All cells converged with no improper solutions.\n")
  }
}
```

### lavaan parity (rep 1 per cell)

Present only when run with `--lavaan-parity`. Max absolute difference between
magmaan and lavaan (`standardizedSolution`; WLSMV for DWLS, MLM for ML) on the
same data.

```{r}
if (!is.null(parity) && nrow(parity)) {
  a <- aggregate(abs_diff ~ estimator + metric, parity, max)
  a$abs_diff <- sprintf("%.2e", a$abs_diff)
  knitr::kable(a[order(a$estimator, a$metric), ], row.names = FALSE,
               col.names = c("Est.", "Metric", "Max |diff|"))
} else {
  cat("(no parity table — run with `--lavaan-parity`)\n")
}
```

## Caveats

- **Two of Li's three estimators.** Li compares robust ML, DWLS *and* ULS;
  this run covers ML and DWLS. ULS (ULSMV) on polychorics is a planned
  follow-up; it is central to Li's three-way comparison, so the ML-vs-DWLS
  contrasts here are the replication's first half, not the whole story.
- Default `--reps` is small for iteration; rejection rates and biases are noisy
  until `--reps` is large (Li uses 500). Per repo style, no Monte-Carlo error
  bars are reported.
- Estimates are standardized to lavaan's `std.all` metric; for DWLS the
  loadings use the unit-variance latent-response scale and structural paths are
  rescaled by the model-implied latent SDs.
- DWLS's χ² parity uses the mean-and-variance adjusted statistic; magmaan's MV
  uses a Satterthwaite df, lavaan's WLSMV default is scaled-shifted, so the
  per-rep p-values differ by a parameterization (compared, not the statistics).
- At small N the 20-indicator polychoric weight matrix is often singular; DWLS
  (diagonal weight) is unaffected and matches lavaan WLSMV.

## Reproduce

```sh
# fast smoke (a few cells across shapes)
Rscript run_experiment.R --reps 30 --smoke

# a single cell with the lavaan cross-check
Rscript run_experiment.R --reps 100 --cells "categories=5,N=500" --lavaan-parity

# the full paper-scale grid (slow)
Rscript run_experiment.R --reps 500
```

`Rscript run_experiment.R --help` lists all options. Outputs land in
`results/`; this report reads only from there.
