---
title: "Does covariance-honest ML improve small-sample convergence?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("full", "pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
raw <- read_result("raw.csv")
summary <- read_result("summary.csv")
paired <- read_result("paired.csv")
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)
profile <- metadata[["profile"]]
reps <- as.integer(metadata[["reps"]])

method_labels <- c(
  `ml-lbfgs` = "Ordinary ML, L-BFGS",
  `ml-slsqp` = "Ordinary ML, SLSQP",
  `psd-ml-slsqp` = "PSD-ML, SLSQP",
  `psd-ml-ipopt` = "PSD-ML, IPOPT"
)

rate_at <- function(n, method, field) {
  x <- summary[summary$n == n & summary$method == method, field]
  if (length(x) == 1L) x else NA_real_
}
fmt_pct <- function(x) sprintf("%.1f%%", 100 * x)
fmt_num <- function(x, digits = 3L) formatC(
  x, format = "f", digits = digits)

low <- raw[raw$n < 50L, ]
pair_rates <- function(baseline, alternative = "psd-ml-slsqp") {
  left <- low[low$method == baseline, ]
  right <- low[low$method == alternative, ]
  x <- merge(
    left, right,
    by = c("n", "rep", "seed"),
    suffixes = c("_baseline", "_alternative")
  )
  c(
    convergence_rescue = mean(!x$audit_converged_baseline &
                                x$audit_converged_alternative),
    convergence_harm = mean(x$audit_converged_baseline &
                              !x$audit_converged_alternative),
    usability_rescue = mean(!x$usable_baseline & x$usable_alternative),
    usability_harm = mean(x$usable_baseline & !x$usable_alternative)
  )
}
lbfgs_pair <- pair_rates("ml-lbfgs")
slsqp_pair <- pair_rates("ml-slsqp")
has_ipopt <- "psd-ml-ipopt" %in% raw$method

display_n <- intersect(c(10L, 20L, 50L, 100L), summary$n)
beta_rows <- lapply(
  split(raw, interaction(raw$n, raw$method, drop = TRUE)),
  function(x) {
    beta <- x$beta_hat[x$audit_converged & is.finite(x$beta_hat)]
    data.frame(
      n = x$n[[1L]],
      method = x$method[[1L]],
      beta_median = if (length(beta)) median(beta) else NA_real_,
      beta_abs_q99 = if (length(beta)) {
        as.numeric(quantile(abs(beta), 0.99, names = FALSE))
      } else {
        NA_real_
      },
      stringsAsFactors = FALSE
    )
  }
)
beta_summary <- do.call(rbind, beta_rows)
display <- summary[summary$n %in% display_n, c(
  "n", "method", "audit_convergence_rate", "admissible_rate",
  "usable_rate", "covariance_boundary_rate",
  "median_elapsed_ms")]
display <- merge(
  display, beta_summary,
  by = c("n", "method"), all.x = TRUE, sort = FALSE)
display <- display[order(display$n, display$method), ]
display$method <- unname(method_labels[display$method])
names(display) <- c(
  "N", "Method", "Converged", "Admissible",
  "Converged + Admissible", "Non-interior Covariance",
  "Median Time (ms)", "Median Beta", "99th Percentile |Beta|")
for (field in c("Converged", "Admissible", "Converged + Admissible",
                "Non-interior Covariance")) {
  display[[field]] <- vapply(display[[field]], fmt_pct, character(1))
}
display$`Median Beta` <- round(display$`Median Beta`, 3L)
display$`99th Percentile |Beta|` <-
  round(display$`99th Percentile |Beta|`, 3L)
display$`Median Time (ms)` <- round(display$`Median Time (ms)`, 1L)

is_evidence <- profile %in% c("pilot", "full") && reps >= 200L
psd_all_admissible <- all(
  raw$admissible[
    grepl("^psd-ml-", raw$method) & raw$audit_converged])
psd_returned <- raw[
  raw$method == "psd-ml-slsqp" & raw$returned, ]
psd_solver_rejected <- raw[
  raw$method == "psd-ml-slsqp" & raw$solver_converged &
    !raw$audit_converged, ]
psd_max_constraint_violation <- max(
  psd_returned$constraint_violation_inf, na.rm = TRUE)
psd_max_constraint_violation_label <- format(
  psd_max_constraint_violation, scientific = TRUE, digits = 3)
psd_rejected_kkt <- range(
  psd_solver_rejected$stationarity_inf, na.rm = TRUE)
n10_psd_convergence <- rate_at(10L, "psd-ml-slsqp",
                               "audit_convergence_rate")
n10_ml_convergence <- rate_at(10L, "ml-lbfgs",
                              "audit_convergence_rate")
n10_psd_boundary <- rate_at(10L, "psd-ml-slsqp",
                            "covariance_boundary_rate")
n10_psd <- raw[
  raw$n == 10L & raw$method == "psd-ml-slsqp" &
    raw$audit_converged, ]
n10_beta_median <- median(n10_psd$beta_hat, na.rm = TRUE)
n10_beta_q99 <- as.numeric(
  quantile(abs(n10_psd$beta_hat), 0.99, na.rm = TRUE))
n10_beta_max <- max(abs(n10_psd$beta_hat), na.rm = TRUE)
short_answer <- if (!is_evidence) {
  sprintf(
    paste(
      "This is a %s run with %d replications per sample size.",
      "It validates the comparison but is too small for a convergence",
      "conclusion."
    ),
    profile, reps
  )
} else {
  sprintf(
    paste(
      "PSD-ML substantially improves numerical convergence: at N = 10,",
      "%.1f%% passed the common convergence audit versus %.1f%% for",
      "ordinary ML. Across N < 50 it rescued %.1f%% of ordinary-ML",
      "failures and lost %.1f%% of ordinary-ML successes. Holding SLSQP",
      "fixed, the rescue and harm rates were %.1f%% and %.1f%%, so the",
      "gain is not merely an optimizer switch. It is not a small-sample",
      "cure, however: %.1f%% of N = 10 PSD fits were on a covariance",
      "boundary; beta had median %.3f but a 99th absolute percentile of",
      "%.2f and maximum of %.1f (population value 0.25)."
    ),
    100 * n10_psd_convergence,
    100 * n10_ml_convergence,
    100 * lbfgs_pair[["convergence_rescue"]],
    100 * lbfgs_pair[["convergence_harm"]],
    100 * slsqp_pair[["convergence_rescue"]],
    100 * slsqp_pair[["convergence_harm"]],
    100 * n10_psd_boundary,
    n10_beta_median,
    n10_beta_q99,
    n10_beta_max
  )
}
ipopt_sentence <- if (has_ipopt) {
  ipopt_n10 <- summary[
    summary$n == 10L & summary$method == "psd-ml-ipopt", ]
  slsqp_n10 <- summary[
    summary$n == 10L & summary$method == "psd-ml-slsqp", ]
  ipopt_n10_pair <- paired[
    paired$n == 10L &
      paired$baseline == "psd-ml-slsqp" &
      paired$alternative == "psd-ml-ipopt", ]
  sprintf(
    paste(
      "In the IPOPT cross-check at N = 10, PSD-SLSQP was usable in %.1f%%",
      "of replications versus %.1f%% for IPOPT. Relative to PSD-SLSQP,",
      "IPOPT rescued %.1f%% but harmed %.1f%% and cost a median %.1fx as",
      "much."
    ),
    100 * slsqp_n10$usable_rate,
    100 * ipopt_n10$usable_rate,
    100 * ipopt_n10_pair$usability_rescue_rate,
    100 * ipopt_n10_pair$usability_harm_rate,
    ipopt_n10_pair$median_time_ratio
  )
} else {
  ""
}
admissibility_sentence <- if (psd_all_admissible) {
  "Every audit-converged PSD-ML fit was covariance-admissible."
} else {
  paste(
    "At least one audit-converged PSD-ML fit failed the primitive covariance",
    "audit; inspect the raw results."
  )
}
```

## Question

On the six-indicator small-sample SEM used by
[De Jonckere and Rosseel (2022)](https://doi.org/10.1080/10705511.2021.1982716)
and [Ernst and colleagues (2023)](https://doi.org/10.1007/s11336-022-09891-5),
does covariance-honest normal-theory ML converge more reliably than ordinary
normal-theory ML?

## Short answer

`r short_answer` `r admissibility_sentence` `r ipopt_sentence`

## Evidence

Every method saw the same generated dataset and the same model start. A fit is
called converged only when its returned solution passes the optimizer terminal
audit: the box-projected objective gradient for ordinary ML and the
equality-KKT Lagrangian residual plus primal feasibility for lifted PSD-ML.
The stricter combined column additionally requires
the fitted primitive covariance matrices to be positive semidefinite. The
non-interior column means that the smallest primitive-covariance eigenvalue is
at or below \(10^{-6}\); for ordinary ML this includes estimates beyond the
covariance cone, whereas PSD-ML can only reach its boundary. The beta summaries
are conditional on convergence and expose the heavy tails that a binary
convergence rate hides.

The full artifact was refreshed after the equality-KKT audit was added on
2026-07-30. All returned PSD fits have finite KKT, raw-gradient,
constraint-violation, and constraint-Jacobian-rank telemetry. Their largest
equality violation was `r psd_max_constraint_violation_label`, so equality
feasibility was numerically tight. The refresh reproduced the 96.8% versus
54.1% \(N=10\) headline exactly.

```{r evidence-table}
knitr::kable(display, row.names = FALSE)
```

The ordinary-SLSQP row is an attribution control. Comparing PSD-ML only with
ordinary L-BFGS would combine a changed model class with a changed optimizer.
Comparing the two SLSQP rows holds the optimizer fixed.

`r if (has_ipopt) "The IPOPT row is a backend robustness cross-check, not another model class. Most IPOPT failures in the N = 10 panel had status Invalid_Number_Detected, consistent with the interior-point path reaching a singular observed-covariance or structural-transform domain where the ML callback is non-finite. The current adapter does not identify which callback rejected the point. NLopt SLSQP can reject such a point with an infinite objective and shorten its step; IPOPT requires a finite smooth callback over its bounded search domain. Returned IPOPT fits nevertheless satisfied the lifted equalities to numerical tolerance." else ""`

## Caveats

This experiment studies one correctly specified, six-indicator Gaussian SEM.
It uses magmaan's current FABIN starting values rather than the all-ones starts
of the original bounded-estimation study. The current `r profile` run has
`r reps` rather than the published 10,000 replications per sample size, and
uses the common optimizer audit rather than requiring an additional successful
standard-error calculation. It can establish whether PSD-ML helps this known
convergence stress design; it cannot establish a general prevalence or
small-sample inference result.

Rank-boundary solutions are valid members of the PSD model class, but ordinary
interior information-matrix inference does not apply there. In the full run,
`r nrow(psd_solver_rejected)` clean PSD solver returns were rejected by the
terminal audit. Both were covariance-admissible and equality-feasible, but
their KKT residuals were `r fmt_num(psd_rejected_kkt[[1L]])` and
`r fmt_num(psd_rejected_kkt[[2L]])`, rather than at the \(10^{-3}\) audit
scale. Solver return codes are therefore not used as the convergence verdict.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript run_experiment.R --full
Rscript run_experiment.R --pilot --reps 200 --n-values 10 \
  --methods ml-lbfgs,ml-slsqp,psd-ml-slsqp,psd-ml-ipopt
quarto render report.qmd
```
