---
title: "Latent Metric Parameterization: Geometry, Cost, and Inference"
format: html
execute:
  echo: false
  warning: false
  message: false
---

```{r}
cand <- unique(c(getwd(), dirname(getwd())))
rd <- cand[file.exists(file.path(cand, "results", "geometry.csv"))]
rd <- if (length(rd)) file.path(rd[[1L]], "results") else file.path(getwd(), "results")
need <- c("geometry.csv", "cost.csv", "inference.csv", "metadata.csv")
miss <- need[!file.exists(file.path(rd, need))]
if (length(miss)) stop("missing: ", paste(miss, collapse = ", "),
                       ". Run: Rscript run_experiment.R")
geom <- utils::read.csv(file.path(rd, "geometry.csv"))
cost <- utils::read.csv(file.path(rd, "cost.csv"))
infr <- utils::read.csv(file.path(rd, "inference.csv"))
meta <- utils::read.csv(file.path(rd, "metadata.csv"))
mv <- function(k, d = "?") { r <- meta$value[meta$key == k]; if (length(r)) r[[1L]] else d }
wide <- function(d, val, id = "lambda1") {
  w <- stats::reshape(d[, c(id, "chart", val)], idvar = id, timevar = "chart",
                      direction = "wide")
  names(w) <- sub(paste0("^", val, "\\."), "", names(w))
  w[order(w[[id]]), ]
}
# Must print explicitly: expressions inside a braced `if` block are not
# auto-printed the way top-level chunk expressions are, so a bare kable() inside
# one silently emits nothing except when it happens to be the block's last value.
kb <- function(x, ...) print(knitr::kable(x, row.names = FALSE, ...))
skew <- function(x) { x <- x[is.finite(x)]; if (length(x) < 3L) return(NA_real_)
  m <- mean(x); mean((x - m)^3) / mean((x - m)^2)^1.5 }
PMAX <- max(geom$p)
```

Mode **`r mv("mode")`**, reps **`r mv("reps")`**, n **`r mv("n")`**, p **`r mv("p")`**,
λ₁ **`r mv("lambda1")`**, backends **`r mv("backends")`**.

This supersedes `experiments/_archive/02-latent-metric-identification`, which timed
marker against std_lv on the golden-fixture models and found a fit-only ratio of
0.839 that decayed to 0.996 once the back-conversion was counted. That result was
inconclusive for three reasons, each addressed here: there was no marker-strength
axis even though marker strength controls the entire effect, wall time was the only
outcome at model sizes where timing is noise, and no statistical quality measure
was recorded at all.

## What a chart can and cannot move

A latent scaling convention is a gauge choice. The scaling indeterminacy is a group
orbit `c -> (c*lambda, phi/c^2)`, and each convention is a hypersurface cutting
those orbits. Nothing gauge-invariant can move.

```{r}
inv <- do.call(rbind, lapply(split(infr, paste(infr$p, infr$lambda1, infr$replicate)),
  function(d) data.frame(chisq = diff(range(d$chisq_h1)),
                         lrt = diff(range(d$lrt_p)),
                         df = diff(range(d$df_h1)))))
kb(data.frame(quantity = c("chi-square (H1)", "LRT p-value", "df"),
              `max across-chart spread` = c(
                format(max(inv$chisq, na.rm = TRUE), digits = 3),
                format(max(inv$lrt, na.rm = TRUE), digits = 3),
                format(max(inv$df, na.rm = TRUE), digits = 3)),
              check.names = FALSE),
   caption = "Invariance check. All three charts fit the same manifold, so these must be numerical noise. This is a correctness gate on the whole run, not a finding.")
```

## Geometry

Bates and Watts (1980) split the second-derivative array of the moment map into a
component tangent to the model manifold, the parameter-effects curvature, and a
component normal to it, the intrinsic curvature. The tangent part depends on the
chart and is in principle removable. The normal part is a property of the manifold
and is invariant, which is why its agreement across charts below is a second
correctness gate. The ratio PE/IN is the headroom: a chart whose PE has been driven
down to the order of IN has essentially nothing left to gain from any further
reparameterization.

```{r}
g <- geom[geom$p == PMAX, ]
kb(wide(g, "pe_over_in"), digits = 3,
   caption = sprintf("PE/IN by chart (p = %d). Lower is better, and IN is the floor.", PMAX))
kb(wide(g, "pcond_info"), digits = 4,
   caption = sprintf("Pseudo-condition number of the expected information (p = %d).", PMAX))
kb(data.frame(check = c("closed-form chart map vs population Sigma",
                        "magmaan's implied Sigma vs population Sigma",
                        "IN agreement across charts (max relative spread)"),
              `max abs` = c(format(max(geom$validate_own_map, na.rm = TRUE), digits = 3),
                            format(max(geom$validate_vs_magmaan, na.rm = TRUE), digits = 3),
                            format(max(unlist(lapply(
                              split(geom$curv_in, paste(geom$p, geom$lambda1)),
                              function(x) diff(range(x)) / mean(x))), na.rm = TRUE), digits = 3)),
              check.names = FALSE),
   caption = "Validation. The geometry arm uses explicit closed-form chart maps because effect coding's raw parameter vector lives on a constraint surface and cannot be differenced in place. Those maps are pinned to magmaan's own moment map here.")
```

Two readings. First, **marker degrades with marker strength and the other two do
not**: std_lv and effect coding are flat in λ₁ to three digits, while marker's PE/IN
and conditioning both blow up as λ₁ falls. That is the transversality mechanism. The
marker chart anchors the gauge with λ₁, so its slice becomes tangent to the gauge
orbit as λ₁ goes to zero, while std_lv anchors on a factor variance and effect coding
on a sum of loadings, neither of which can approach zero in an identified model.

Second, **std_lv and effect coding trade places depending on the axis**. std_lv has
the lower parameter-effects curvature, which is the part a second-order optimizer
cannot absorb. Effect coding has the better linear conditioning at moderate to strong
λ₁, which is the part full Newton and Fisher scoring are affine-invariant to and
therefore get for free. Neither dominates, and the practical implication is that the
choice between them is close to immaterial.

## Cost

Wall time here is measured properly, in batches, and it is not noise. The earlier
claim in this report that microsecond fits are untimeable was wrong. What actually
breaks is `system.time()`, which quantises to about 1 ms on Linux: ten timings of the
same 225 µs fit return a minimum of 0.000, a median of 0.001 and a maximum of 0.007.
That is quantisation rather than variance, and it is the real reason exp 02 could not
resolve what it was looking for, since it timed each fit once. Batching removes the
problem outright. `time_per_call()` calibrates a batch size so each batch runs at
least 50 ms, then reports the median over five batches.

```{r}
prec <- stats::median(cost$fit_sec_iqr / cost$fit_sec, na.rm = TRUE)
kb(data.frame(`median batch size` = stats::median(cost$fit_batch_k, na.rm = TRUE),
              `median IQR / median` = sprintf("%.1f%%", 100 * prec),
              check.names = FALSE),
   caption = "Timing precision actually achieved. Roughly two percent relative, against a quantisation floor that made a 20 percent effect unmeasurable.")
ft <- stats::aggregate(fit_sec ~ chart + p + backend, cost, median)
fw <- stats::reshape(ft, idvar = c("p", "backend"), timevar = "chart", direction = "wide")
names(fw) <- sub("^fit_sec\\.", "", names(fw))
if (all(c("marker", "std_lv") %in% names(fw))) fw$`std_lv/marker` <- fw$std_lv / fw$marker
kb(fw[order(fw$backend, fw$p), ], digits = 4,
   caption = "Median per-call fit time by chart, and the ratio exp 02 was trying to measure.")
```

`f_evals` is reported alongside as the mechanism, since it counts optimizer work
directly and is immune to machine and BLAS variation.

```{r}
be <- if ("nlopt-lbfgs" %in% cost$backend) "nlopt-lbfgs" else cost$backend[[1L]]
a <- stats::aggregate(f_evals ~ chart + lambda1 + p, cost[cost$backend == be, ], median)
w <- stats::reshape(a, idvar = c("p", "lambda1"), timevar = "chart", direction = "wide")
names(w) <- sub("^f_evals\\.", "", names(w))
kb(w[order(w$p, w$lambda1), ], caption = sprintf("Median f_evals (%s).", be))

r <- stats::aggregate(f_evals ~ chart + backend, cost, median)
rw <- stats::reshape(r, idvar = "backend", timevar = "chart", direction = "wide")
names(rw) <- sub("^f_evals\\.", "", names(rw))
if (all(c("marker", "std_lv") %in% names(rw))) rw$`marker/std_lv` <- rw$marker / rw$std_lv
kb(rw, digits = 2, caption = "Median f_evals by backend. A more metric-aware optimizer should be less exposed to a badly conditioned chart.")

bc <- cost[cost$chart != "marker" & is.finite(cost$backconvert_sec), ]
if (nrow(bc)) {
  b <- stats::aggregate(cbind(fit_sec, backconvert_sec) ~ p + chart, bc, median)
  b$`backconvert % of fit` <- 100 * b$backconvert_sec / b$fit_sec
  kb(b[order(b$chart, b$p), c("p", "chart", "backconvert % of fit")], digits = 3,
     caption = "Back-convert cost relative to the fit. This is the exp 02 question: a library wanting a better internal chart must still report the user's chart, and the back-convert is O(p) while the fit is superlinear, so the overhead should shrink as p grows.")
}
```

Three readings. The ratio exp 02 reported as 0.839 is **0.43 at p = 24 under
nlopt-lbfgs**, and it improves with p rather than washing out. The backend split
confirms the mechanism: nlopt-lbfgs gains far more than port, because a quasi-Newton
method starting from `H_0 = I` is exposed to the chart's conditioning while a more
metric-aware method is not.

And std_lv costs roughly a third of marker's function evaluations, an advantage that
does not disappear at strong markers. That is more than the geometry tables predict,
since at λ₁ = 0.9 marker is competitive on both curvature and conditioning yet still
needs several times more evaluations. **Start values are the obvious remaining
suspect and are not controlled here.** That is the clearest follow-up this experiment
leaves open, because it is the one part of the cost gap better geometry does not
explain.

## Back-conversion

Using a better chart internally is viable only if converting back to the user's chart
is both exact and cheap. exp 02 measured only the cost, and only for point estimates.

```{r}
if (file.exists(file.path(rd, "backconvert.csv"))) {
  bcv <- utils::read.csv(file.path(rd, "backconvert.csv"))
  rt <- bcv$roundtrip_max_abs_sigma[is.finite(bcv$roundtrip_max_abs_sigma)]
  kb(data.frame(check = "round-trip: back-converted theta reproduces the native marker implied Sigma",
                `max abs` = format(max(rt), digits = 3), check.names = FALSE),
     caption = "Exactness. This is the disqualifying check, and it comes first: no amount of speed rescues a lossy substitution.")
  s <- bcv[bcv$chart == "std_lv", ]
  s <- stats::aggregate(cbind(fit_sec, bc_point_sec, bc_vcov_sec,
                              bc_point_pct_of_fit, bc_vcov_pct_of_fit) ~ p, s, median)
  kb(s[order(s$p), ], digits = 4,
     caption = "Cost against the fit it is trying to save. Point estimates are an O(p) loading rescale. The vcov needs the delta-method sandwich J V J', which is dense O(p^3) -- and a user who asks for the marker chart wants marker standard errors, not just marker point estimates. That second term is the one exp 02 never counted.")
} else cat("_backconvert arm not run._\n")
```

Point-estimate conversion is flat in absolute terms near 1.8 µs and falls to well
under a tenth of a percent of the fit by p = 48, so it is free. The vcov transform
grows as O(p³), but the fit grows superlinearly too, so its share **plateaus near
three percent rather than shrinking**. That corrects the earlier claim in this report
that the overhead simply falls with p: it falls for point estimates and levels off for
the vcov.

Either way the verdict is clear, and it reverses exp 02's. Roughly three percent total
overhead against a thirty to fifty-seven percent saving on the fit means the
internal-chart substitution pays, and exp 02's roughly eighteen percent figure was an
artifact of measuring a quantised timer on small models.

## Convergence at small n

Conditioning should bite hardest where the problem is hard, and the geometry arm is a
population-level statement that says nothing about actual failure. Every chart sees
the same dataset per cell, so the charts can be compared per draw.

```{r}
if (file.exists(file.path(rd, "convergence.csv"))) {
  cvv <- utils::read.csv(file.path(rd, "convergence.csv"))
  cvv$ok <- cvv$converged & !cvv$errored
  be2 <- if ("nlopt-lbfgs" %in% cvv$backend) "nlopt-lbfgs" else cvv$backend[[1L]]
  pp <- max(cvv$p)
  s <- cvv[cvv$p == pp & cvv$backend == be2, ]
  a <- stats::aggregate(list(fail = !s$ok), by = list(chart = s$chart, n = s$n), FUN = mean)
  w <- stats::reshape(a, idvar = "n", timevar = "chart", direction = "wide")
  names(w) <- sub("^fail\\.", "", names(w))
  kb(w[order(w$n), ], digits = 3,
     caption = sprintf("Non-convergence rate (p = %d, %s).", pp, be2))
  k <- paste(cvv$p, cvv$lambda1, cvv$n, cvv$replicate, cvv$backend)
  mk <- k[cvv$chart == "marker"]; sk <- k[cvv$chart == "std_lv"]
  mo <- cvv$ok[cvv$chart == "marker"]; so <- cvv$ok[cvv$chart == "std_lv"]
  ix <- match(mk, sk); keep <- !is.na(ix)
  kb(data.frame(outcome = c("marker fails, std_lv succeeds",
                            "std_lv fails, marker succeeds", "both fail",
                            "matched draws"),
                draws = c(sum(!mo[keep] & so[ix[keep]]), sum(mo[keep] & !so[ix[keep]]),
                          sum(!mo[keep] & !so[ix[keep]]), sum(keep))),
     caption = "Per-draw discordance. A dominance relation is much stronger evidence than a rate comparison, because it rules out the two charts simply failing on different draws at similar rates.")
  nimp <- sum(as.logical(cvv$improper), na.rm = TRUE)
  imr <- mean(as.logical(cvv$improper), na.rm = TRUE)
  cat(sprintf("\nImproper solutions across the whole arm: %d of %d draws (%.4f).%s\n",
    nimp, nrow(cvv), imr,
    if (imr < 0.01) " That is far too few to compare charts on, so **this design does not stress admissibility** and says nothing about the Heywood question. Loadings of 0.7 with psi = 0.51 are not extreme enough, and the archived exp 03 needed deliberately hard cases to produce them. Treat the improper column as uninformative here rather than as evidence that no chart has a problem." else ""))
} else cat("_convergence arm not run._\n")
```

## Inference

The restriction under test ties the loadings of the last two indicators, which is
true in the population, so rejection is Type-I error.

```{r}
i2 <- infr[infr$p == PMAX, ]
s <- stats::aggregate(loading_a ~ chart + lambda1, i2, skew)
kb(wide(s, "loading_a"), digits = 3,
   caption = sprintf("Skewness of the fitted loading in each chart's own coordinates (p = %d). These are deliberately different functionals, since reshaping them is exactly what a chart is allowed to do.", PMAX))
im <- stats::aggregate(improper ~ chart + lambda1, i2, function(x) mean(as.logical(x), na.rm = TRUE))
kb(wide(im, "improper"), digits = 3,
   caption = "Improper-solution rate, read off every estimated variance in the chart. exp 03 found std_lv relocates a Heywood case from the latent variance to the observed ones rather than removing it, so the detector cannot look at the latent variance alone.")
mn <- stats::aggregate(min_est_var ~ chart + lambda1, i2, median)
kb(wide(mn, "min_est_var"), digits = 4,
   caption = "Median smallest estimated variance. Distance to the admissibility boundary.")
```

```{r}
if (as.integer(mv("reps", "0")) < 200L) {
  cat("> **The skewness and improper-rate tables above are not evidence at this replication count.** Rejection rates and third moments need several hundred replications before they separate. Run `--full` before reading them. The geometry arm needs no replication, because those are population quantities, and the invariance gates are exact, so both are trustworthy from a smoke run.\n")
}
```

## Status

Trustworthy from any run: the invariance gates, the validation table, the geometry arm
(deterministic), the round-trip exactness check, and the timing tables, since the
batched timer supplies its own precision and replications only add data variation.

Needs `--full`: skewness, and the convergence rates, where the interesting cells sit
at a couple of percent and 25 replications put one or two draws in each.

Open, and the reason the convergence arm is not the last word: this population does
not produce improper solutions at all, so it cannot speak to the archived exp 03
finding that std_lv relocates Heywood cases rather than removing them. A harder
population is needed for that. The other open item is the start-value confound in the
cost arm.
