---
title: "Latent Metric Parameterization: Geometry, Cost, and Inference"
format: html
execute:
  echo: false
  warning: false
  message: false
---

```{r}
cand <- unique(c(getwd(), dirname(getwd())))
rd <- cand[file.exists(file.path(cand, "results", "geometry.csv"))]
rd <- if (length(rd)) file.path(rd[[1L]], "results") else file.path(getwd(), "results")
need <- c("geometry.csv", "cost.csv", "inference.csv", "metadata.csv")
miss <- need[!file.exists(file.path(rd, need))]
if (length(miss)) stop("missing: ", paste(miss, collapse = ", "),
                       ". Run: Rscript run_experiment.R")
geom <- utils::read.csv(file.path(rd, "geometry.csv"))
cost <- utils::read.csv(file.path(rd, "cost.csv"))
infr <- utils::read.csv(file.path(rd, "inference.csv"))
meta <- utils::read.csv(file.path(rd, "metadata.csv"))
mv <- function(k, d = "?") { r <- meta$value[meta$key == k]; if (length(r)) r[[1L]] else d }
wide <- function(d, val, id = "lambda1") {
  w <- stats::reshape(d[, c(id, "chart", val)], idvar = id, timevar = "chart",
                      direction = "wide")
  names(w) <- sub(paste0("^", val, "\\."), "", names(w))
  w[order(w[[id]]), ]
}
kb <- function(x, ...) knitr::kable(x, row.names = FALSE, ...)
skew <- function(x) { x <- x[is.finite(x)]; if (length(x) < 3L) return(NA_real_)
  m <- mean(x); mean((x - m)^3) / mean((x - m)^2)^1.5 }
PMAX <- max(geom$p)
```

Mode **`r mv("mode")`**, reps **`r mv("reps")`**, n **`r mv("n")`**, p **`r mv("p")`**,
λ₁ **`r mv("lambda1")`**, backends **`r mv("backends")`**.

This supersedes `experiments/_archive/02-latent-metric-identification`, which timed
marker against std_lv on the golden-fixture models and found a fit-only ratio of
0.839 that decayed to 0.996 once the back-conversion was counted. That result was
inconclusive for three reasons, each addressed here: there was no marker-strength
axis even though marker strength controls the entire effect, wall time was the only
outcome at model sizes where timing is noise, and no statistical quality measure
was recorded at all.

## What a chart can and cannot move

A latent scaling convention is a gauge choice. The scaling indeterminacy is a group
orbit `c -> (c*lambda, phi/c^2)`, and each convention is a hypersurface cutting
those orbits. Nothing gauge-invariant can move.

```{r}
inv <- do.call(rbind, lapply(split(infr, paste(infr$p, infr$lambda1, infr$replicate)),
  function(d) data.frame(chisq = diff(range(d$chisq_h1)),
                         lrt = diff(range(d$lrt_p)),
                         df = diff(range(d$df_h1)))))
kb(data.frame(quantity = c("chi-square (H1)", "LRT p-value", "df"),
              `max across-chart spread` = c(
                format(max(inv$chisq, na.rm = TRUE), digits = 3),
                format(max(inv$lrt, na.rm = TRUE), digits = 3),
                format(max(inv$df, na.rm = TRUE), digits = 3)),
              check.names = FALSE),
   caption = "Invariance check. All three charts fit the same manifold, so these must be numerical noise. This is a correctness gate on the whole run, not a finding.")
```

## Geometry

Bates and Watts (1980) split the second-derivative array of the moment map into a
component tangent to the model manifold, the parameter-effects curvature, and a
component normal to it, the intrinsic curvature. The tangent part depends on the
chart and is in principle removable. The normal part is a property of the manifold
and is invariant, which is why its agreement across charts below is a second
correctness gate. The ratio PE/IN is the headroom: a chart whose PE has been driven
down to the order of IN has essentially nothing left to gain from any further
reparameterization.

```{r}
g <- geom[geom$p == PMAX, ]
kb(wide(g, "pe_over_in"), digits = 3,
   caption = sprintf("PE/IN by chart (p = %d). Lower is better, and IN is the floor.", PMAX))
kb(wide(g, "pcond_info"), digits = 4,
   caption = sprintf("Pseudo-condition number of the expected information (p = %d).", PMAX))
kb(data.frame(check = c("closed-form chart map vs population Sigma",
                        "magmaan's implied Sigma vs population Sigma",
                        "IN agreement across charts (max relative spread)"),
              `max abs` = c(format(max(geom$validate_own_map, na.rm = TRUE), digits = 3),
                            format(max(geom$validate_vs_magmaan, na.rm = TRUE), digits = 3),
                            format(max(unlist(lapply(
                              split(geom$curv_in, paste(geom$p, geom$lambda1)),
                              function(x) diff(range(x)) / mean(x))), na.rm = TRUE), digits = 3)),
              check.names = FALSE),
   caption = "Validation. The geometry arm uses explicit closed-form chart maps because effect coding's raw parameter vector lives on a constraint surface and cannot be differenced in place. Those maps are pinned to magmaan's own moment map here.")
```

Two readings. First, **marker degrades with marker strength and the other two do
not**: std_lv and effect coding are flat in λ₁ to three digits, while marker's PE/IN
and conditioning both blow up as λ₁ falls. That is the transversality mechanism. The
marker chart anchors the gauge with λ₁, so its slice becomes tangent to the gauge
orbit as λ₁ goes to zero, while std_lv anchors on a factor variance and effect coding
on a sum of loadings, neither of which can approach zero in an identified model.

Second, **std_lv and effect coding trade places depending on the axis**. std_lv has
the lower parameter-effects curvature, which is the part a second-order optimizer
cannot absorb. Effect coding has the better linear conditioning at moderate to strong
λ₁, which is the part full Newton and Fisher scoring are affine-invariant to and
therefore get for free. Neither dominates, and the practical implication is that the
choice between them is close to immaterial.

## Cost

`f_evals` is the primary outcome rather than wall time. The models here fit in a few
hundred microseconds, which is inside timing noise, and that is what made exp 02
inconclusive.

```{r}
be <- if ("nlopt-lbfgs" %in% cost$backend) "nlopt-lbfgs" else cost$backend[[1L]]
a <- stats::aggregate(f_evals ~ chart + lambda1 + p, cost[cost$backend == be, ], median)
w <- stats::reshape(a, idvar = c("p", "lambda1"), timevar = "chart", direction = "wide")
names(w) <- sub("^f_evals\\.", "", names(w))
kb(w[order(w$p, w$lambda1), ], caption = sprintf("Median f_evals (%s).", be))

r <- stats::aggregate(f_evals ~ chart + backend, cost, median)
rw <- stats::reshape(r, idvar = "backend", timevar = "chart", direction = "wide")
names(rw) <- sub("^f_evals\\.", "", names(rw))
if (all(c("marker", "std_lv") %in% names(rw))) rw$`marker/std_lv` <- rw$marker / rw$std_lv
kb(rw, digits = 2, caption = "Median f_evals by backend. A more metric-aware optimizer should be less exposed to a badly conditioned chart.")

bc <- cost[cost$chart != "marker" & is.finite(cost$backconvert_sec), ]
if (nrow(bc)) {
  b <- stats::aggregate(cbind(fit_sec, backconvert_sec) ~ p + chart, bc, median)
  b$`backconvert % of fit` <- 100 * b$backconvert_sec / b$fit_sec
  kb(b[order(b$chart, b$p), c("p", "chart", "backconvert % of fit")], digits = 3,
     caption = "Back-convert cost relative to the fit. This is the exp 02 question: a library wanting a better internal chart must still report the user's chart, and the back-convert is O(p) while the fit is superlinear, so the overhead should shrink as p grows.")
}
```

std_lv costs roughly a third of marker's function evaluations, and the advantage does
not disappear at strong markers. That is more than the geometry tables predict on
their own, since at λ₁ = 0.9 marker is competitive on both curvature and conditioning
yet still needs several times more evaluations. **Start values are the obvious
remaining suspect and are not controlled here.** That is the clearest follow-up this
experiment leaves open, and it matters because it is the one part of the cost gap
that better geometry does not explain.

The back-convert overhead falls as p grows, which supports the internal-chart idea on
the cost side. exp 02 measured it at roughly 18 percent of the fit and concluded the
substitution was a wash. Here it is smaller and shrinking.

## Inference

The restriction under test ties the loadings of the last two indicators, which is
true in the population, so rejection is Type-I error.

```{r}
i2 <- infr[infr$p == PMAX, ]
s <- stats::aggregate(loading_a ~ chart + lambda1, i2, skew)
kb(wide(s, "loading_a"), digits = 3,
   caption = sprintf("Skewness of the fitted loading in each chart's own coordinates (p = %d). These are deliberately different functionals, since reshaping them is exactly what a chart is allowed to do.", PMAX))
im <- stats::aggregate(improper ~ chart + lambda1, i2, function(x) mean(as.logical(x), na.rm = TRUE))
kb(wide(im, "improper"), digits = 3,
   caption = "Improper-solution rate, read off every estimated variance in the chart. exp 03 found std_lv relocates a Heywood case from the latent variance to the observed ones rather than removing it, so the detector cannot look at the latent variance alone.")
mn <- stats::aggregate(min_est_var ~ chart + lambda1, i2, median)
kb(wide(mn, "min_est_var"), digits = 4,
   caption = "Median smallest estimated variance. Distance to the admissibility boundary.")
```

```{r}
if (as.integer(mv("reps", "0")) < 200L) {
  cat("> **The skewness and improper-rate tables above are not evidence at this replication count.** Rejection rates and third moments need several hundred replications before they separate. Run `--full` before reading them. The geometry arm needs no replication, because those are population quantities, and the invariance gates are exact, so both are trustworthy from a smoke run.\n")
}
```

## Status

Trustworthy from any run: the invariance gates, the validation table, and the whole
geometry arm, which is deterministic. Needs `--full`: skewness, improper rates, and
anything else averaged over replications. Open: the start-value confound in the cost
arm, which is the one result here that the geometry does not account for.
