---
title: "What does covariance-honest NTML cost?"
subtitle: "Direct PSD fitting versus an audit-triggered fallback"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("full", "pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)
summary <- read_result("case_summary.csv")
validation <- read_result("validation.csv")
failures <- read_result("failures.csv")

profile <- metadata[["profile"]]
repeats <- as.integer(metadata[["repeats"]])
is_evidence <- profile %in% c("pilot", "full") && repeats >= 20L
interior <- summary[
  summary$ordinary_admissible & summary$family == "interior scaling", ]
inadmissible <- summary[!summary$ordinary_admissible, ]
median_psd_ratio <- median(interior$psd_over_ordinary)
median_audit_interior <- median(interior$audit_over_ordinary)
median_audit_bad <- median(inadmissible$audit_over_ordinary)
largest <- interior[which.max(interior$p), ]
fallback_count <- sum(!summary$ordinary_admissible)
all_valid <- nrow(failures) == 0L &&
  all(validation$pass)

display <- summary[, c(
  "case_id", "p", "ordinary_admissible",
  "ordinary_median_ms", "psd_median_ms", "audit_median_ms",
  "psd_over_ordinary", "audit_over_ordinary")]
names(display) <- c(
  "Case", "p", "Ordinary Admissible",
  "Ordinary (ms)", "PSD Direct (ms)", "Audit Policy (ms)",
  "PSD / Ordinary", "Audit / Ordinary")
display$Case <- gsub("_", " ", display$Case, fixed = TRUE)
numeric_columns <- c(
  "Ordinary (ms)", "PSD Direct (ms)", "Audit Policy (ms)",
  "PSD / Ordinary", "Audit / Ordinary")
display[numeric_columns] <- lapply(
  display[numeric_columns], function(x) round(x, 2L))
```

## Question

How much slower is complete-data covariance-honest normal-theory ML than
ordinary NTML, and is the practical policy to fit ordinarily, inspect the
covariance audit, and refit only inadmissible cases?

## Short answer

`r if (!is_evidence) {
  sprintf("This is a %s timing run with %d repetitions per case. It validates the runner but is too small for performance conclusions.", profile, repeats)
} else {
  sprintf("Across the interior CFA scaling cases, direct PSD fitting took a median %.1fx as long as ordinary NTML. The audit-first policy cost %.2fx ordinary NTML when the first fit was admissible, while the deliberately inadmissible fixtures cost %.1fx because they required both fits. At p=%d, the direct PSD/ordinary ratio was %.1fx.", median_psd_ratio, median_audit_interior, median_audit_bad, largest$p, largest$psd_over_ordinary)
}`
`r if (all_valid) sprintf(" All %d deliberately inadmissible cases were detected by the ordinary audit and repaired by PSD-ML; all final fits were admissible.", fallback_count) else " At least one validation or final-fit check failed; inspect the CSV outputs before interpreting timings."`

## Evidence

The model and sample-statistic construction is outside the timed region. The
direct policy runs the current PSD estimator from the standard magmaan start.
The audit policy first runs ordinary NTML and, only when its post-fit covariance
audit fails, initializes PSD-ML from that ordinary estimate.

```{r timing-table}
knitr::kable(display, row.names = FALSE)
```

The interior scaling rows use one-factor CFA models with 3, 6, 12, and, in the
full profile, 24 indicators. The remaining rows are exact-covariance fixtures:
a near-boundary uniqueness, a classical Heywood solution, jointly indefinite
latent and residual covariance matrices, a negative structural disturbance,
and an equality-constrained interior control.

```{r timing-plot, fig.width=8, fig.height=4.8}
plot_rows <- summary[order(summary$p, summary$case_id), ]
x <- seq_len(nrow(plot_rows))
ymax <- max(
  plot_rows$psd_over_ordinary,
  plot_rows$audit_over_ordinary,
  na.rm = TRUE)
old_par <- par(mar = c(8, 4, 1, 1))
on.exit(par(old_par), add = TRUE)
plot(
  x, plot_rows$psd_over_ordinary,
  log = "y", pch = 16, col = "#B33A3A", xaxt = "n",
  xlab = "", ylab = "Time ratio versus ordinary NTML",
  ylim = c(0.8, ymax * 1.15))
points(
  x, plot_rows$audit_over_ordinary,
  pch = 17, col = "#245A9B")
abline(h = 1, lty = 2, col = "grey40")
axis(
  1, at = x,
  labels = gsub("_", " ", plot_rows$case_id, fixed = TRUE),
  las = 2, cex.axis = .75)
legend(
  "topleft",
  legend = c("PSD direct", "Audit then PSD if needed"),
  pch = c(16, 17), col = c("#B33A3A", "#245A9B"),
  bty = "n")
```

## Caveats

This is a deterministic engineering benchmark, not an estimate of the
prevalence of improper SEM solutions. It measures the current PSD estimator
with the one constrained backend available on the benchmark host. The current
frontier estimator covers complete-data normal-theory ML only; FIML and the
continuous LS/GMM weight families are not included.

The timing panel deliberately exposes dimension scaling, but it is not a
textbook-corpus benchmark. The results show that the present implementation
scales poorly with indicator count and motivate a more structure-aware
formulation, but cannot establish the numerical reliability of an alternative.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript run_experiment.R --full
quarto render report.qmd
```
