---
title: "Newton check in the FIML and least-squares verdicts"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
res <- file.path("results", Sys.getenv("E19_PROFILE", "current"))
need <- file.path(res, "fits.csv")
if (!file.exists(need)) stop("Missing results. Run: Rscript run_experiment.R", call. = FALSE)
f <- read.csv(need, stringsAsFactors = FALSE)
cases <- unique(f$case)
excluded <- f[f$estimator == "none", ]
fits <- f[f$estimator != "none", ]
errors <- fits[is.na(fits$converged), ]
ok <- fits[!is.na(fits$converged), ]
ok$old <- ok$first_order %in% TRUE
ok$new <- ok$converged %in% TRUE
ok$change <- ifelse(ok$old & !ok$new, "rejected now",
             ifelse(!ok$old & ok$new, "accepted now", "unchanged"))
ok$cause <- with(ok, ifelse(change != "rejected now", "",
  ifelse(newton_status == "nonpositive_curvature", "not a minimum (nonpositive curvature)",
  ifelse(newton_status == "available" & distance > 0.01, "Newton step above budget",
  ifelse(newton_status %in% c("ill_conditioned", "solve_unreliable"), "ill-conditioned",
  ifelse(newton_status == "unavailable", "curvature unavailable", "other"))))))
n_est <- function(e) sum(ok$estimator == e)
```

## Question

Complete-data ML decides convergence with a Newton check that rejects stationary
points that are not minima and measures the remaining Newton step in standard
errors. FIML, the least-squares fitters (GLS, ULS) and the categorical DWLS
fitter decided with a first-order check only. What does moving them to the
Newton check change on the textbook corpus, and are the changed verdicts right?

## Short answer

```{r short}
tab <- table(ok$estimator, ok$change)
```

The corpus gave `r length(cases)` cases, `r nrow(excluded)` excluded before
fitting (categorical, composite, multilevel, a free-parameter pattern that
differs from lavaan's, or a non-positive-definite sample covariance). Of the
`r nrow(ok)` fits that returned, `r sum(ok$change == "rejected now")` change
from accepted to rejected and `r sum(ok$change == "accepted now")` from
rejected to accepted. Every new rejection is a bad endpoint or a near miss of
the budget. Most are stationary points that are not minima, where the
first-order check had accepted a saddle, and a few are fits that stopped far
from the optimum with a small gradient. Only near misses (Newton step between
0.01 and 0.06 standard errors) are rescued by tighter stopping controls alone,
which is the case for adopting complete-data ML's stopping controls for these
families next. The two new acceptances are fits whose remaining step is under
the budget although their absolute gradient exceeded the first-order
tolerance.

## Evidence

### Verdict changes by estimator

```{r changes}
knitr::kable(as.data.frame.matrix(table(ok$estimator, ok$change)))
```

### Causes of the new rejections

```{r causes}
r <- ok[ok$change == "rejected now", ]
if (nrow(r)) knitr::kable(as.data.frame.matrix(table(r$cause, r$estimator))) else cat("None.")
```

### Every rejected fit, checked

For each newly rejected fit: the objective lavaan reaches with the same
estimator from its own starts (comparable for GLS and ULS; not for FIML, whose
constant differs, and only roughly for DWLS against lavaan's WLSMV), and a refit by magmaan under complete-data ML's tighter
stopping controls (relative objective tolerance $10^{-12}$, relative step
tolerance $10^{-10}$, 5000 evaluations), with its verdict and Newton distance.

```{r rejected}
v <- read.csv(file.path(res, "rejections.csv"), stringsAsFactors = FALSE)
v <- v[order(v$status, v$estimator, v$case), ]
lower <- with(v, estimator %in% c("GLS", "ULS") & !is.na(lavaan_f) & lavaan_f < f - 1e-6 * (1 + abs(f)))
rescued <- v$tight_conv %in% TRUE
knitr::kable(data.frame(Case = v$case, Estimator = v$estimator, Status = v$status,
                        d = v$d, Objective = v$f, `lavaan objective` = v$lavaan_f,
                        `Tight refit objective` = v$tight_f, `Tight refit converged` = v$tight_conv,
                        check.names = FALSE), row.names = FALSE)
```

```{r gaps}
gap <- with(v, (f - lavaan_f) / (1 + abs(f)))
big <- lower & gap > 1e-3
```

Of the `r nrow(v)` rejections, lavaan reaches a lower objective on
`r sum(lower)` of the `r sum(v$estimator %in% c("GLS", "ULS") & !is.na(v$lavaan_f))` GLS
and ULS fits it could fit, by more than a thousandth of the objective on
`r sum(big)`. The small gaps remain where magmaan's own tight refit converges,
so they reflect a difference between the two programs' objectives or basins,
not a convergence failure. The tight refit converges on `r sum(rescued)` of
the rejections (`r sum(rescued & v$estimator == "FIML")` of the
`r sum(v$estimator == "FIML")` FIML fits, whose default fits had stopped far
from the optimum). The remaining curvature rejections are the Little models on
which ordinary starts lead every optimizer to the same non-minimum, a start
problem, and the far-off budget rejections (Brown, Sabatelli, Geiser, Robbins)
where lavaan reaches a much lower objective.

```{r dwls}
w <- v[v$estimator == "DWLS", ]
```

The `r nrow(w)` DWLS rejections are the largest gaps of all: magmaan's
categorical fits from raw data end at objectives of
`r signif(min(w$f), 2)` to `r signif(max(w$f), 2)` where lavaan's WLSMV reaches
at most `r signif(max(w$lavaan_f, na.rm = TRUE), 2)` (a just-identified probit
regression, Mplus example 3.4, sits at `r signif(w$f[w$case == "mplus_users_guide_v8_ch3_ex3_4"], 2)`),
the tight refit stays there, and the exact Hessian says the point is not a
minimum of magmaan's own objective. The first-order check accepted all of
them. The size of the gaps points at the categorical fit path's model setup or
starts (probit regressions on covariates, growth models with released scales),
not at the verdict.

### Newton distance of the fits accepted under both checks

```{r distance}
a <- ok[ok$old & ok$new & !is.na(ok$distance), ]
knitr::kable(do.call(rbind, lapply(split(a$distance, a$estimator), function(d)
  data.frame(Fits = length(d), Median = signif(median(d), 2),
             `90th percentile` = signif(unname(quantile(d, .9)), 2), Max = signif(max(d), 2),
             check.names = FALSE))))
```

### Fits that returned an error

`r nrow(errors)` fits returned an error before any verdict; they are unaffected
by the change.

## Reproduce

```sh
just r-dev
Rscript experiments/engineering/active/19-newton-verdict-migration/run_experiment.R --workers 4
Rscript experiments/engineering/active/19-newton-verdict-migration/R/verify_rejections.R
quarto render experiments/engineering/active/19-newton-verdict-migration/report.qmd
```

## Decision record

- Current default: the Newton check decides the verdict of complete-data ML,
  FIML (ordinary and PSD), every moment-quadratic fit (GLS, ULS, WLS, DWLS,
  GMM, SNLLS), the ordinal and mixed least-squares fits, and the barrier
  fitters, with an analytic Hessian of the fitted objective. FIML and the
  barriers use their own Hessian as metric, the continuous least-squares fits
  the normal-theory sandwich, and the ordinal fits the Gauss-Newton sandwich.
- Multi-group categorical cases are excluded here: the R interface's data
  conversion rejects them (group count mismatch), which is unrelated to the
  verdict.
- Acceptance: every fit that changes verdict is explained, and no fit changes
  from accepted to rejected at a point that is a genuine local minimum with a
  step inside the budget.
- Reopening trigger: a rejected fit at a verified local minimum, or a
  family whose default stopping controls routinely leave the step above the
  budget.


## Activity after the engineering review

Active for the failures exposed by the completed verdict rollout. Resolve
categorical bad endpoints and stopping-control near misses against independently
checked objectives. Keep every changed verdict explained; changing the acceptance
rule to hide an optimization or model-setup failure is not a resolution.
