---
title: "RMSEA-like CATML-at-DWLS Probe"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- file.path("results", "summary.csv")
target_path <- file.path("results", "targets.csv")
if (!file.exists(summary_path) || !file.exists(target_path)) {
  stop("Run `Rscript run_experiment.R --smoke` first.", call. = FALSE)
}
summary <- read.csv(summary_path)
targets <- read.csv(target_path)

fmt <- function(x, digits = 5) ifelse(is.finite(x), formatC(x, digits = digits, format = "f"), "")
tab <- summary
tab$missing_rate <- sprintf("%.0f%%", 100 * tab$missing_rate)
tab$target_F <- fmt(tab$target_F)
tab$mean_F <- fmt(tab$mean_F)
tab$bias_F <- fmt(tab$bias_F)
tab$mae_F <- fmt(tab$mae_F)
tab$target_rmsea <- fmt(tab$target_rmsea)
tab$mean_rmsea <- fmt(tab$mean_rmsea)
tab$bias_rmsea <- fmt(tab$bias_rmsea)
tab$mae_rmsea <- fmt(tab$mae_rmsea)
tab$mean_elapsed_ms <- fmt(tab$mean_elapsed_ms, digits = 1)
names(tab) <- c("N", "Missing", "Usable", "Target F", "Mean F", "Bias F",
                "MAE F", "Target RMSEA", "Mean RMSEA", "Bias RMSEA",
                "MAE RMSEA", "Mean ms")
target_short <- targets[1L, ]
```

## Question

For ordinal DWLS/WLSMV, lavaan's robust categorical RMSEA is not the DWLS
least-squares discrepancy. It is an RMSEA-like statistic built from a CATML
normal-theory correlation criterion evaluated at the already-estimated DWLS
solution, plus a robust scale correction. Under misspecification this is a
criterion-at-estimator pair:

`D_CATML(theta_DWLS)`, not `min_theta D_CATML(theta)`.

This probe checks whether the statistic behaves consistently for that hybrid
estimand in a deliberately misspecified setting: six ordinal indicators generated
from two correlated Gaussian factors, fit by a one-factor ordinal DWLS model with
pairwise MCAR missingness.

## Short Answer

In the smoke/current run, `XX3 / N` moves toward the large-`N` reference
`D_CATML(theta_DWLS)` target. The RMSEA transform moves with the same target
because, under fixed misspecification, the finite robust subtraction is
asymptotically negligible. This supports the interpretation that lavaan's
categorical robust RMSEA is a consistent RMSEA-like statistic for the hybrid
CATML-at-DWLS estimand, not evidence that DWLS and CATML estimate the same
pseudo-true parameters.

The first target row used `N_ref = `r target_short$n`` and missing rate
`r sprintf("%.0f%%", 100 * target_short$missing_rate)`, giving
`F_CATML@DWLS = `r fmt(target_short$F_catml_at_dwls)`` and robust RMSEA
`r fmt(target_short$rmsea_robust)``.

## Evidence

```{r}
knitr::kable(tab, row.names = FALSE)
```

Raw replicate rows are in `results/replicates.csv`; large-`N` references are in
`results/targets.csv`.

## Interpretation

This is the same conceptual shape as GIC-like diagnostics: the statistic is
defined by both the criterion and the estimator whose pseudo-true point is being
plugged into that criterion. If the model is misspecified, changing either side
changes the population target.

For lavaan's categorical robust RMSEA, the population object is naturally read as
`RMSEA_CATML(theta_DWLS)`. A "DWLS RMSEA" based on the DWLS quadratic criterion
would be a different statistic; a refitted CATML RMSEA would be another one.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 50 --n 500,2000,8000 --missing-rate .30 --n-ref 200000
quarto render report.qmd
```

