---
title: "Ernst IRLS ML Convergence"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path), dirname(getwd())))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[
  file.exists(file.path(result_candidates,
                        "irls_ernst_convergence_summary.csv"))
]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

summary_path <- file.path(results_dir, "irls_ernst_convergence_summary.csv")
kappa_path <- file.path(results_dir, "irls_ernst_convergence_kappa.csv")
improper_path <- file.path(results_dir, "irls_ernst_improper_by_n.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(summary_path) || !file.exists(kappa_path) ||
    !file.exists(improper_path) || !file.exists(metadata_path)) {
  stop("Missing experiment CSVs. Run: Rscript run_experiment.R")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("The report requires ggplot2. Install it or inspect the CSVs directly.")
}

library(ggplot2)
summary_n <- utils::read.csv(summary_path, check.names = FALSE)
summary_k <- utils::read.csv(kappa_path, check.names = FALSE)
improper_n <- utils::read.csv(improper_path, check.names = FALSE)
metadata <- utils::read.csv(metadata_path, check.names = FALSE)

meta_value <- function(key) {
  hit <- metadata$value[metadata$key == key]
  if (length(hit)) hit[[1L]] else ""
}
fmt_pct <- function(x) ifelse(is.na(x), "", sprintf("%.1f%%", 100 * x))
fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), "", formatC(x, format = "g", digits = digits))
}
fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.3f", x / 1000))
# method_levels <- c("ml-lbfgs", "ml-port", "ml-fisher", "ml-fisher-snlls",
#                    "ml-irls", "ml-irls-snlls")
# method_labels <- c(
#   `ml-lbfgs` = "ML L-BFGS",
#   `ml-port` = "ML PORT",
#   `ml-fisher` = "ML Fisher",
#   `ml-fisher-snlls` = "ML Fisher-SNLLS",
#   `ml-irls` = "IRLS-ML",
#   `ml-irls-snlls` = "SNLLS-IRLS-ML"
# )
method_levels <- c("ml-lbfgs", "ml-fisher", "ml-fisher-snlls")
method_labels <- c(
  `ml-lbfgs` = "ML L-BFGS",
  `ml-fisher` = "ML Fisher",
  `ml-fisher-snlls` = "ML Fisher-SNLLS"
)
summary_n$method <- factor(summary_n$method, levels = method_levels)
summary_k$method <- factor(summary_k$method, levels = method_levels)
theme_set(theme_minimal(base_size = 12))

method_totals <- do.call(rbind, lapply(split(summary_n, summary_n$method),
                                       function(x) {
  data.frame(
    Method = method_labels[as.character(x$method[[1L]])],
    Reps = sum(x$reps),
    `Convergence Rate` = sum(x$n_converged) / sum(x$reps),
    `Solver Error Rate` = sum(x$n_error) / sum(x$reps),
    `Median Time ms` = stats::median(x$median_wall_usec, na.rm = TRUE) / 1000,
    `Median F Evals` = stats::median(x$median_fevals, na.rm = TRUE),
    check.names = FALSE
  )
}))
```

## Question

Does Fisher-scoring IRLS for complete-data ML improve convergence on the Ernst
small-sample two-factor SEM design, and does the Golub-Pereyra SNLLS inner step
help or hurt that story?

The comparison is against direct ML with NLopt L-BFGS and PORT. Convergence is
the magmaan terminal-audit verdict, not merely the backend return code.

## Short Answer

```{r}
best <- method_totals[which.max(method_totals$`Convergence Rate`), ]
worst <- method_totals[which.min(method_totals$`Convergence Rate`), ]
cat(
  "Across ", meta_value("n_cells"), " sample-size cells and ",
  meta_value("reps"), " replicates per cell, the best method was ",
  best$Method, " with ", fmt_pct(best$`Convergence Rate`),
  " audit convergence. The weakest was ", worst$Method, " with ",
  fmt_pct(worst$`Convergence Rate`), ".\n",
  sep = ""
)
```

```{r}
total_table <- method_totals
total_table$`Convergence Rate` <- fmt_pct(total_table$`Convergence Rate`)
total_table$`Solver Error Rate` <- fmt_pct(total_table$`Solver Error Rate`)
total_table$`Median Time ms` <- sprintf("%.3f", total_table$`Median Time ms`)
total_table$`Median F Evals` <- fmt_num(total_table$`Median F Evals`)
knitr::kable(total_table, row.names = FALSE)
```

## Evidence

```{r}
#| fig-width: 8
#| fig-height: 4.8
ggplot(summary_n,
       aes(n, convergence_rate, color = method,
           group = method)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  scale_color_discrete(labels = method_labels) +
  labs(x = "Sample Size", y = "Audit Convergence Rate",
       color = "Method",
       title = "Convergence By Sample Size") +
  theme(legend.position = "bottom")
```

```{r}
keep_n <- c(10L, 15L, 25L, 40L, 70L, 100L)
slice <- summary_n[summary_n$n %in% keep_n, ]
slice <- slice[order(slice$n, slice$method), ]
table_n <- data.frame(
  N = slice$n,
  Method = method_labels[as.character(slice$method)],
  `Convergence Rate` = fmt_pct(slice$convergence_rate),
  `Solver Error Rate` = fmt_pct(slice$solver_error_rate),
  `Median F Evals` = fmt_num(slice$median_fevals),
  `Median Time ms` = fmt_ms(slice$median_wall_usec),
  check.names = FALSE
)
knitr::kable(table_n, row.names = FALSE)
```

```{r}
#| fig-width: 8
#| fig-height: 4.8
ggplot(summary_n,
       aes(n, median_wall_usec / 1000, color = method,
           group = method)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  scale_y_log10() +
  scale_color_discrete(labels = method_labels) +
  labs(x = "Sample Size", y = "Median Wall Time (ms, log scale)",
       color = "Method",
       title = "Per-Fit Cost By Sample Size") +
  theme(legend.position = "bottom")
```

```{r}
#| fig-width: 8
#| fig-height: 4.8
summary_k$kbin_label <- factor(summary_k$kbin,
                               levels = unique(summary_k$kbin[order(summary_k$kappa_med)]))
ggplot(summary_k,
       aes(kappa_med, convergence_rate, color = method,
           group = method)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  scale_x_log10() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  scale_color_discrete(labels = method_labels) +
  labs(x = "Median kappa(S) In Bin (log scale)",
       y = "Audit Convergence Rate",
       color = "Method",
       title = "Convergence By Sample Covariance Conditioning") +
  theme(legend.position = "bottom")
```

```{r}
imp <- improper_n[improper_n$n %in% keep_n, ]
imp_table <- data.frame(
  N = imp$n,
  Datasets = imp$datasets,
  `Improper Rate` = fmt_pct(imp$improper_rate),
  check.names = FALSE
)
knitr::kable(imp_table, row.names = FALSE)
```

## Caveats

This is an engineering convergence check, not a paper-grade Monte Carlo table.
The default run uses `r meta_value("reps")` replicates per sample-size cell.
Wall times are single-fit timings inside the simulation loop, so they are good
for coarse cost ordering but not microbenchmark claims.

The SNLLS-IRLS path is only meaningful for separable models; this Ernst design
is separable, so the experiment exercises its intended domain rather than its
applicability boundary.

## Reproduce

```{r}
repro <- data.frame(
  Item = c("Command", "Seed Base", "Replicates", "N Values", "Methods"),
  Value = c(
    "Rscript run_experiment.R",
    meta_value("seed_base"),
    meta_value("reps"),
    meta_value("n_values"),
    meta_value("methods")
  ),
  check.names = FALSE
)
knitr::kable(repro, row.names = FALSE)
```
