---
title: "Mixed first-stage statistics: does continuous FIML buy efficiency over pairwise deletion?"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path), dirname(getwd())))
result_candidates <- unique(c(file.path(candidate_dirs, "results"), candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "efficiency.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
need <- file.path(results_dir, c("efficiency.csv", "per_param.csv", "convergence.csv"))
if (!all(file.exists(need))) {
  stop("Missing results. Run: Rscript run_experiment.R")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) stop("The report requires ggplot2.")
library(ggplot2)

eff  <- utils::read.csv(file.path(results_dir, "efficiency.csv"), check.names = FALSE)
pp   <- utils::read.csv(file.path(results_dir, "per_param.csv"), check.names = FALSE)
conv <- utils::read.csv(file.path(results_dir, "convergence.csv"), check.names = FALSE)
meta_path <- file.path(results_dir, "metadata.csv")
meta <- if (file.exists(meta_path)) utils::read.csv(meta_path, check.names = FALSE) else NULL
meta_lookup <- function(key, default = "?") {
  if (is.null(meta)) return(default)
  row <- meta[meta$key == key, , drop = FALSE]
  if (!nrow(row)) default else as.character(row$value[[1L]])
}
n_reps <- meta_lookup("reps")

block_label <- c(continuous = "Continuous block", ordinal = "Ordinal block",
                 cross = "Cross (factor cor.)")
mech_label  <- c(mcar = "MCAR", mar = "MAR")
eff$Block      <- factor(block_label[eff$block], levels = block_label)
eff$Mechanism  <- mech_label[eff$mechanism]

# Headline numbers used in prose.
pct <- function(x) sprintf("%.0f%%", 100 * (1 - x))
mcar_cont <- eff[eff$mechanism == "mcar" & eff$block == "continuous", ]
miss_lo <- min(eff$missing_rate); miss_hi <- max(eff$missing_rate)
n_hi <- max(eff$n)
val <- function(d, mech, block, n, mr) {
  r <- d[d$mechanism == mech & d$block == block & d$n == n & d$missing_rate == mr, ]
  if (!nrow(r)) NA_real_ else r$median_rel_rmse[[1L]]
}
valv <- function(d, mech, block, n, mr) {
  r <- d[d$mechanism == mech & d$block == block & d$n == n & d$missing_rate == mr, ]
  if (!nrow(r)) NA_real_ else r$median_rel_var[[1L]]
}
cvar_lo <- valv(eff, "mcar", "continuous", n_hi, miss_lo)
cvar_hi <- valv(eff, "mcar", "continuous", n_hi, miss_hi)
ord_rrmse <- eff$median_rel_rmse[eff$block == "ordinal"]

# MAR bias on the continuous block at the highest missing rate.
mar_bias <- eff[eff$mechanism == "mar" & eff$block == "continuous" &
                eff$n == n_hi & eff$missing_rate == miss_hi, ]
```

## Question

When a mixed continuous/ordinal SEM is fit from summary statistics with
item-level missing data, the continuous means and covariances can be formed two
ways: by **pairwise deletion** (each moment from the cases that observe its
pair) or by **continuous FIML** (the saturated normal likelihood over all
observed continuous values). The ordinal thresholds, polychorics, and
ordinal/continuous polyserials are pairwise either way. Does swapping pairwise
continuous moments for FIML continuous moments improve the resulting
estimates, and which parameters feel it?

## Short answer

Yes, modestly, and exactly where the theory says it should. The hybrid changes
only the continuous moment block, so the ordinal parameters are **bit-for-bit
identical** (relative RMSE = `r sprintf("%.2f", stats::median(ord_rrmse))`),
which is the clean control. On the continuous block under MCAR the FIML hybrid
is more efficient, and the gain grows with the missing rate: at n = `r n_hi` the
sampling variance of the continuous parameters drops by about `r pct(cvar_lo)`
at `r sprintf("%.0f%%", 100*miss_lo)` missing and about `r pct(cvar_hi)` at
`r sprintf("%.0f%%", 100*miss_hi)` missing, both estimators being unbiased here.
Under MAR the decisive effect is **bias**: pairwise continuous moments are
selected on the observed anchors and become inconsistent, while continuous FIML
stays consistent. At `r sprintf("%.0f%%", 100*miss_hi)` missing the worst
continuous-parameter bias is `r sprintf("%.3f", mar_bias$max_abs_bias_pp)`
(pairwise) versus `r sprintf("%.3f", mar_bias$max_abs_bias_pf)` (hybrid), a
roughly `r sprintf("%.0f-fold", mar_bias$max_abs_bias_pp / mar_bias$max_abs_bias_pf)`
reduction. The cross-block factor correlation, governed by the unchanged
polyserial moments, barely moves.

## Evidence

### Efficiency: relative RMSE by block

Each point is the median over the free parameters in a block of
RMSE(hybrid) / RMSE(pairwise); below 1 means the FIML hybrid wins. The ordinal
line sits on 1 by construction (the hybrid never touches it); the continuous
line drops further below 1 as missingness rises.

```{r efficiency-plot, fig.width = 8, fig.height = 4.5}
d <- eff[eff$block %in% c("continuous", "ordinal"), ]
ggplot(d, aes(missing_rate, median_rel_rmse, colour = Block,
              group = interaction(Block, n), linetype = factor(n))) +
  geom_hline(yintercept = 1, colour = "grey60") +
  geom_line() + geom_point(size = 2) +
  facet_wrap(~ Mechanism) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = c("Continuous block" = "#1f78b4",
                                 "Ordinal block" = "#b2182b",
                                 "Cross (factor cor.)" = "#7f7f7f")) +
  labs(x = "Missing rate", y = "Relative RMSE (hybrid / pairwise)",
       colour = NULL, linetype = "n",
       caption = "Below 1: FIML-continuous hybrid more efficient.") +
  theme_minimal(base_size = 12)
```

### Bias under MAR (continuous block)

Pairwise deletion under MAR selects the continuous covariances on the observed
anchors; FIML does not. The table reports the largest absolute bias among the
continuous-block parameters (loadings, residual variances, factor variance,
intercepts) at each design cell.

```{r mar-bias-table}
b <- eff[eff$mechanism == "mar" & eff$block == "continuous",
         c("n", "missing_rate", "max_abs_bias_pp", "max_abs_bias_pf")]
b <- b[order(b$n, b$missing_rate), ]
b$missing_rate <- sprintf("%.0f%%", 100 * b$missing_rate)
names(b) <- c("n", "Missing", "Max |bias| pairwise", "Max |bias| hybrid")
knitr::kable(b, digits = 3, row.names = FALSE)
```

### SE calibration and convergence

Both first stages feed the same misspecification-robust IJ standard errors. The
table reports, per estimator, the convergence rate and the median fit time
(statistics + DWLS fit + IJ standard errors).

```{r conv-table}
cs <- aggregate(cbind(conv_rate, median_ms) ~ estimator, data = conv, FUN = mean)
cs$estimator <- c(pp = "Pairwise x pairwise", pf = "Pairwise x FIML")[cs$estimator]
names(cs) <- c("Estimator", "Mean convergence rate", "Median time (ms)")
knitr::kable(cs, digits = c(0, 3, 0), row.names = FALSE)
```

## Caveats

- This is an efficiency probe, not a parity gate. The population is a single
  two-factor mixed CFA (four continuous indicators on one factor, four
  three-category ordinal indicators on the other); the gains will differ for
  other ratios of continuous to ordinal information and other correlation
  strengths.
- Reference ("true") values are a large-sample fit, not closed-form, so biases
  carry a small reference-noise floor.
- Only DWLS stage-two weights are exercised. WLS and ULS are deferred.
- `r n_reps` replications per cell; small relative-RMSE gaps near 1 are within
  Monte Carlo noise.

## Reproduce

```sh
cd experiments/37-mixed-fiml-pairwise-efficiency
Rscript run_experiment.R            # full grid
Rscript run_experiment.R --smoke    # fast sanity path
quarto render report.qmd
```
