---
title: "Near-Singular ML Continuation"
format: html
---

## What Was Fit

For a sample covariance matrix `S`, the continuation path fits a sequence of
ordinary ML problems to blended covariance matrices,

```text
S_lambda = (1 - lambda) S + lambda T(S),
```

where `T(S)` is the diagonal target in these runs. The wrapper calls the
tuning parameter `alpha`, but it is the same object as `lambda` here.

The default sequence is:

```text
0.50 -> 0.20 -> 0.10 -> 0.05 -> 0.01 -> 0.00
```

The final `0.00` stage is appended automatically when `include_endpoint =
TRUE`. That final stage is the original ML problem. Each stage is a full ML fit,
and its parameter vector is used as the start value for the next stage, so this
is a hot-started chain of multiple fits.

## How To Read The Tables

`baseline_ml` is one direct ML fit on the original sample covariance.
`ridge_continuation` is the hot-started sequence above.

`attempts` is the number of attempted fits. `ok` means the wrapper returned a
fit object instead of an error. `converged` means the terminal audit accepted
the final point as stationary, using magmaan's lavaan-style default
`||projected gradient||_inf <= 0.001`. `optimizer_converged` is narrower: it
means the backend itself reported a clean stop. A line-search salvage can fail
that optimizer flag while still passing the audit. `line_search_failures`
counts trapped hard L-BFGS line-search errors. `line_search_salvaged` counts
non-clean backend stops whose returned point passed the audit.
`median_total_f_evals` is the median number of objective evaluations, counting
all stages for continuation.

The paired table is often the most interpretable one. `baseline_only` means the
direct fit converged and continuation did not. `continuation_only` means
continuation rescued a replicate that direct ML did not converge on. `neither`
means both failed.

## Toy Smoke Run

This is the artificial near-singular one-factor generator from
`run_experiment.R`. It is useful as a quick smoke test, not as the main evidence.

```{r}
fits_path <- file.path("results", "fits.csv")
summary_path <- file.path("results", "summary.csv")
if (!file.exists(fits_path) || !file.exists(summary_path)) {
  stop("Run `Rscript run_experiment.R` first.")
}
fits <- read.csv(fits_path)
summary <- read.csv(summary_path)
summary
```

```{r}
fits[order(fits$replicate, fits$method),
     c("replicate", "method", "sample_min_eigen", "sample_condition",
       "ok", "converged", "optimizer_converged", "optimizer_status",
       "fmin", "audit_grad_inf_norm", "audit_stationarity_rhs",
       "total_f_evals", "elapsed_sec", "error")]
```

## Convergence-Note Designs

This section uses the same convergence-simulation factories and seed rule as
`papers/convergence-note/pilot/scripts/run_convergence_sims.R`, with
`optimizer = "nlopt-lbfgs"` and control
`max_iter = 5000, ftol = 1e-10, gtol = 1e-7`.

```{r}
cn_summary_path <- file.path("results", "convergence_note_summary.csv")
cn_paired_path <- file.path("results", "convergence_note_paired.csv")
if (!file.exists(cn_summary_path) || !file.exists(cn_paired_path)) {
  cat("Run `Rscript run_convergence_note_comparison.R` to populate this section.")
} else {
  cn_summary <- read.csv(cn_summary_path)
  cn_paired <- read.csv(cn_paired_path)
  cn_summary
}
```

```{r}
if (exists("cn_paired")) {
  cn_paired
}
```

```{r, results='asis'}
if (exists("cn_summary") && exists("cn_paired")) {
  baseline <- cn_summary[cn_summary$method == "baseline_ml", ]
  continuation <- cn_summary[cn_summary$method == "ridge_continuation", ]
  pct <- function(x) paste0(round(100 * x), "%")
  cat(
    "In this run, direct ML converged in ",
    pct(min(baseline$converged_rate)), " to ",
    pct(max(baseline$converged_rate)),
    " of replicates across designs. The continuation path converged in ",
    pct(min(continuation$converged_rate)), " to ",
    pct(max(continuation$converged_rate)),
    " by the terminal-audit criterion. Paired rescues (`continuation_only`) were ",
    pct(min(cn_paired$continuation_only)), " to ",
    pct(max(cn_paired$continuation_only)),
    ", while paired losses (`baseline_only`) were ",
    pct(min(cn_paired$baseline_only)), " to ",
    pct(max(cn_paired$baseline_only)),
    ".\n",
    sep = ""
  )
}
```

## Current Takeaway

On the convergence-note designs, this first continuation path is not a
recommendation yet. The audit-based table is stricter than the raw optimizer
table: several direct L-BFGS fits that report a clean backend stop still fail
the lavaan-style gradient check. Continuation rescues a few of those audited
failures, but it still loses more paired replicates than it rescues and spends
more objective evaluations. The final objective values are essentially the same
when both methods pass the audit, so the current issue is path stability and
terminal stationarity, not a different endpoint.
