---
title: "Near-Singular ML Continuation"
format: html
---

## What Was Fit

For a sample covariance matrix `S`, the continuation path fits a sequence of
ordinary ML problems to blended covariance matrices,

```text
S_lambda = (1 - lambda) S + lambda T(S),
```

where `T(S)` is the diagonal target in these runs. The wrapper calls the
tuning parameter `alpha`, but it is the same object as `lambda` here.

The default sequence is:

```text
0.50 -> 0.20 -> 0.10 -> 0.05 -> 0.01 -> 0.00
```

The final `0.00` stage is appended automatically when `include_endpoint =
TRUE`. That final stage is the original ML problem. Each stage is a full ML fit,
and its parameter vector is used as the start value for the next stage, so this
is a hot-started chain of multiple fits.

## How To Read The Tables

`baseline_ml` is one direct ML fit on the original sample covariance.
`ridge_continuation` is the hot-started sequence above.

`attempts` is the number of attempted fits. `ok` means the wrapper returned a
fit object instead of an error. `converged` means the optimizer reported
convergence. `*_rate` is the corresponding proportion. `line_search_failures`
counts trapped L-BFGS line-search errors. `median_total_f_evals` is the median
number of objective evaluations, counting all stages for continuation.

The paired table is often the most interpretable one. `baseline_only` means the
direct fit converged and continuation did not. `continuation_only` means
continuation rescued a replicate that direct ML did not converge on. `neither`
means both failed.

## Toy Smoke Run

This is the artificial near-singular one-factor generator from
`run_experiment.R`. It is useful as a quick smoke test, not as the main evidence.

```{r}
fits_path <- file.path("results", "fits.csv")
summary_path <- file.path("results", "summary.csv")
if (!file.exists(fits_path) || !file.exists(summary_path)) {
  stop("Run `Rscript run_experiment.R` first.")
}
fits <- read.csv(fits_path)
summary <- read.csv(summary_path)
summary
```

```{r}
fits[order(fits$replicate, fits$method),
     c("replicate", "method", "sample_min_eigen", "sample_condition",
       "ok", "converged", "optimizer_status", "fmin", "grad_norm",
       "total_f_evals", "elapsed_sec", "error")]
```

## Convergence-Note Designs

This section uses the same convergence-simulation factories and seed rule as
`papers/convergence-note/pilot/scripts/run_convergence_sims.R`, with
`optimizer = "nlopt-lbfgs"` and control
`max_iter = 5000, ftol = 1e-10, gtol = 1e-7`.

```{r}
cn_summary_path <- file.path("results", "convergence_note_summary.csv")
cn_paired_path <- file.path("results", "convergence_note_paired.csv")
if (!file.exists(cn_summary_path) || !file.exists(cn_paired_path)) {
  cat("Run `Rscript run_convergence_note_comparison.R` to populate this section.")
} else {
  cn_summary <- read.csv(cn_summary_path)
  cn_paired <- read.csv(cn_paired_path)
  cn_summary
}
```

```{r}
if (exists("cn_paired")) {
  cn_paired
}
```

```{r, results='asis'}
if (exists("cn_summary") && exists("cn_paired")) {
  baseline <- cn_summary[cn_summary$method == "baseline_ml", ]
  continuation <- cn_summary[cn_summary$method == "ridge_continuation", ]
  pct <- function(x) paste0(round(100 * x), "%")
  cat(
    "In this run, direct ML converged in ",
    pct(min(baseline$converged_rate)), " to ",
    pct(max(baseline$converged_rate)),
    " of replicates across designs. The continuation path converged in ",
    pct(min(continuation$converged_rate)), " to ",
    pct(max(continuation$converged_rate)),
    ". Paired rescues (`continuation_only`) were ",
    pct(min(cn_paired$continuation_only)), " to ",
    pct(max(cn_paired$continuation_only)),
    ", while paired losses (`baseline_only`) were ",
    pct(min(cn_paired$baseline_only)), " to ",
    pct(max(cn_paired$baseline_only)),
    ".\n",
    sep = ""
  )
}
```

## Current Takeaway

On the convergence-note designs, this first continuation path is not a
recommendation yet. It can rescue a small number of direct L-BFGS failures, but
it more often introduces a line-search failure at one of the intermediate
stages. The final objective values are essentially the same when both methods
converge, so the current issue is the stability of the path, not a different
endpoint.
