---
title: "Barrier defaults"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
run <- "2026-09-27"
rd <- function(name) {
  p <- file.path("results", "barrier-ml", run, paste0(name, ".csv"))
  if (!file.exists(p)) stop("Missing ", p, ": run Rscript run_experiment.R")
  read.csv(p, stringsAsFactors = FALSE)
}
tab <- function(x, ...) knitr::kable(x, row.names = FALSE, format.args = list(big.mark = ","), ...)
fam <- c(research47 = "research/47", engineering15 = "engineering/15", boomsma = "Boomsma 1985",
         wolf = "Wolf et al. 2013", chen = "Chen et al. 2001", constrained = "equality-constrained",
         ernst = "Ernst (control)")
arms <- c(default = "Default (FABIN3, PORT)", layered_port = "Layered, PORT",
          default_lbfgs = "FABIN3, L-BFGS", layered_lbfgs = "Layered, L-BFGS")
rates <- rd("rates_by_family"); rt <- rd("rates_by_family_transform")
choice <- rd("choice"); dec <- rd("decision"); loss <- rd("losses"); runaways <- rd("runaways")
f0 <- function(x) format(round(x), big.mark = ",")
tot <- function(r, arm, col = "success", fams = NULL) sum(r[[col]][r$role == "test" & r$arm == arm &
  (is.null(fams) | r$family %in% fams)])
other <- setdiff(names(fam), c("constrained", "ernst"))
x001 <- function(arm) sum(rt$attainable[rt$role == "test" & rt$family == "constrained" &
  rt$transform == "x0.01" & rt$arm == arm] - rt$success[rt$role == "test" &
  rt$family == "constrained" & rt$transform == "x0.01" & rt$arm == arm])
lp <- loss[loss$role == "test" & loss$candidate == "layered_port", ]
lp_worse <- sum(is.finite(lp$fmin) & lp$fmin > lp$best_known + 1e-6 * (1 + abs(lp$best_known)))
lp_npc <- sum(lp$newton_status == "nonpositive_curvature")
ll <- loss[loss$role == "test" & loss$candidate == "layered_lbfgs", ]
dl <- loss[loss$role == "test" & loss$candidate == "default_lbfgs", ]
ls_fail <- function(d) sum(grepl("LineSearch", d$message))
pole <- function(arm) sum(runaways$chart_beyond_1000[runaways$role == "test" & runaways$arm == arm])
```

## Question

Which start and optimizer should the complete-data ML barrier fitter
(`frontier_fit_ml_multiinfo`, here with the latent-determinacy penalty
$\lambda \log\det Q$) use by default? Its current defaults were set before the
optimizer coordinates and the penalized Newton verdict existed, and no study
compared its start with the layered start that complete-data ML now uses. The
criteria were committed before the run, in
[criteria/barrier-ml.md](criteria/barrier-ml.md) (commit `677e78bc`).

## Answer

**No candidate is eligible, so the default stays: transported FABIN3 with
PORT.** Every candidate certifies more fits in total than the default
(`r f0(choice$success[choice$candidate == "default_lbfgs"])` for FABIN3 with
L-BFGS, against `r f0(choice$default_success[1])`). The whole gain comes from
one family, the equality-constrained models, and there it comes from a units
bug in the default. In the other test families the default wins or is within
a few fits.

- **The layered start is wrong for the barrier.** Layered with PORT loses
  `r f0(nrow(lp))` test fits against the default, mostly at $N = 25$. In
  `r f0(lp_worse)` of them it stops at a worse penalized objective, and
  `r f0(lp_npc)` fail the Newton check for nonpositive curvature. The inspected
  cases sit on the factor-disappearance set. There a latent variance is zero and
  one loading has the wrong sign. The layered start begins next to that set
  (for example a factor variance of 0.01 when two indicators are uncorrelated
  in the sample). Ordinary ML walks away from it, but the barrier does not,
  because $\log\det Q$ is largest (zero) where a factor disappears. The penalty
  itself pulls toward the set, which is the non-disappearance condition of the
  sem-barrier paper's attainment result showing up in practice.
- **L-BFGS stalls, as _archive/barrier-optimizer found.** With the default start it loses
  `r f0(nrow(dl))` test fits against PORT: `r f0(ls_fail(dl))` line-search
  failures, and the rest uncertified stops. With the layered start both
  problems add up (`r f0(nrow(ll))` losses).
- **The default has a units bug in constrained models.** At $\times 0.01$ it
  fails `r f0(x001("default"))` attainable equality-constrained fits,
  against `r f0(x001("default_lbfgs"))` for FABIN3 with L-BFGS and
  `r f0(x001("layered_port"))` for layered with PORT. In the inspected cases
  the start transport falls back to a native FABIN3 start that is not in the
  data's units. PORT then reports convergence after about a dozen evaluations
  without moving, and the verdict correctly rejects the point. The default
  wins the other units in this family.
- **PORT certifies marker poles, L-BFGS never does.** Certified fits with a
  chart extent above 1000 number `r f0(pole("default"))` for the default and
  `r f0(pole("layered_port"))` for layered with PORT, and zero for both L-BFGS
  arms. The pole rule scores them as failures, and the library verdict still
  calls them converged. This is the barrier's version of the ML runaway
  problem.

Per the criteria, the author decides what follows. The report's reading: fix
the $\times 0.01$ start fallback and look at a pole check in the verdict
before any rerun. The layered start should not be retried for the barrier
unless it is made to stay away from vanishing latent variances.

## The register

| Default | Current value | Status after this run | Evidence |
|---|---|---|---|
| barrier start | transported FABIN3 (pinned apart from the ML default) | settled: keep. The layered start fails rule 1 or 2 in five of six test families | lane barrier-ml, run 2026-09-27 |
| barrier optimizer | PORT | settled for now: L-BFGS fails rules 1 and 2 in three test families (line-search stalls) | lane barrier-ml; _archive/barrier-optimizer |
| barrier start at $\times 0.01$ in constrained models | transport falls back to native FABIN3, PORT stops at the start | open: a units bug, to fix before the next run | lane barrier-ml failures |
| verdict at marker poles | the Newton check certifies PORT points far toward a marker pole | open: scoring has a pole rule (chart extent > 1000), the verdict does not | lane barrier-ml runaways |
| barrier weight $\lambda$ | 0.25 | not decided here; the sem-barrier paper owns it | none |
| FIML barrier start and optimizer | transported FABIN3, PORT | not covered | none |

## Evidence

Certified fits among attainable test problems, both weights pooled.

```{r by-family}
r <- rates[rates$arm %in% names(arms), ]
a <- aggregate(cbind(attainable, success) ~ role + family + arm, r, sum)
w <- reshape(a[c("role", "family", "arm", "success")], idvar = c("role", "family"),
             timevar = "arm", direction = "wide")
att <- a[a$arm == "default", c("role", "family", "attainable")]
w <- merge(att, w)
w <- w[order(w$role == "control", w$family), ]
out <- data.frame(Family = fam[w$family], Attainable = w$attainable,
                  lapply(names(arms), function(k) w[[paste0("success.", k)]]))
names(out)[-(1:2)] <- arms
tab(out)
```

- The default leads in Boomsma, engineering/active/15-sphere-reference-fits and Wolf, and ties Chen. In
  research/47 FABIN3 with L-BFGS is five fits ahead, within its rule. In the
  equality-constrained family the default's losses are the $\times 0.01$
  fallback above. Outside that family
  the default certifies `r f0(tot(rates, "default", fams = other))` against
  `r f0(tot(rates, "layered_port", fams = other))` for layered with PORT and
  `r f0(tot(rates, "default_lbfgs", fams = other))` for FABIN3 with L-BFGS.
- The rules fail in every family where a candidate loses; `decision.csv` has
  each rule per family and weight, `losses.csv` every lost fit with the arms
  that solved it.
- The two weights agree: every rule that fails at one weight fails at the
  other, except Chen's best-known rule for the layered arms (four fits, at
  $\lambda = 1$ only).
- Runs are cheap. The median fit takes about 20 ms in every arm.

## Caveats

- These are complete-data ML fits of correctly specified or mildly
  misspecified models. FIML, multi-group models and the joint penalty are not
  covered.
- The loss mechanisms were read from inspected cases and from the verdict and
  optimizer statuses of all losses. The factor-disappearance reading covers
  the layered start's typical loss, not every one.
- The pole rule's bound (1000) was chosen before the run. The runaway table
  (`runaways.csv`) reports 10 and 100 as well. The population-start witness
  also reaches poles under PORT, so some weak-marker draws may have no
  attainable finite optimum in the marker chart.

## Reproduce

```sh
cd experiments/decisions/02-barrier-defaults
Rscript run_experiment.R --smoke          # seed base + 1, one draw per family
nice -n 10 Rscript run_experiment.R --run-id 2026-09-27   # about 6 minutes on 4 cores
Rscript run_experiment.R --run-id 2026-09-27 --summarize  # rescore raw batches
```
