---
title: "Parameter Profile-LR Small-N Probe"
format:
  html:
    toc: true
---

```{r}
lrt_summary <- read.csv("results/profile_lrt_summary.csv")
lrt_failures <- read.csv("results/profile_lrt_failures.csv")
ci_summary <- read.csv("results/profile_ci_summary.csv")
ci_failures <- read.csv("results/profile_ci_failures.csv")
metadata <- read.csv("results/metadata.csv")
knitr::kable(metadata)
```

## ML LRT Calibration

```{r}
knitr::kable(lrt_summary, digits = 4)
```

The target is the true loading of `x2` in a correctly specified one-factor normal-theory ML model. `mean_T` should be near 1 under the ordinary `chi^2_1` reference; `type1_05` should be near 0.05. This probe is only the interior-parameter baseline for the new C++ scalar/profile-LR surface.

## Robust CI Coverage

```{r}
knitr::kable(ci_summary, digits = 4)
```

The CI branch uses multivariate `t(5)` data scaled to the same population covariance and inverts the continuous GMM profile-LR surface for the `x2` loading. ULS keeps the original fixed/fitted profile-weight comparison. GLS and WLS add fixed-profile ordinary, fixed robust, and estimated-weight robust references, where `robust_estimated_weight` routes the scale through the complete IJ weight influence. This is a non-normal stress check for the continuous robust profile machinery; it is not a Bartlett or small-sample correction.

## Failures

```{r}
if (nrow(lrt_failures)) {
  cat("ML LRT failures\n\n")
  knitr::kable(head(lrt_failures, 20))
} else {
  cat("No ML LRT failures recorded.\n")
}
cat("\n\n")
if (nrow(ci_failures)) {
  cat("Robust CI failures\n\n")
  knitr::kable(head(ci_failures, 20))
} else {
  cat("No robust CI failures recorded.\n")
}
```
