---
title: "Reduced-bias SEM: estimation risk and correction tails"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results/tough-n30-ml-gls"
---

```{r}
res_dir <- params$results_dir
needed <- file.path(res_dir, c("method_summary.csv", "parameter_summary.csv",
                               "fit_status.csv", "metadata.csv"))
if (!all(file.exists(needed))) {
  stop("Missing result files. Run: Rscript experiments/research/banked/53-rbm-estimation-risk/run_experiment.R --reps 5000 --n 30 --ref-n 100000 --estimators ml,gls --results-dir results/tough-n30-ml-gls",
       call. = FALSE)
}
method_summary <- read.csv(file.path(res_dir, "method_summary.csv"))
parameter_summary <- read.csv(file.path(res_dir, "parameter_summary.csv"))
fit_status <- read.csv(file.path(res_dir, "fit_status.csv"))
metadata <- read.csv(file.path(res_dir, "metadata.csv"))
meta <- setNames(metadata$value, metadata$key)
meta_value <- function(key, default = NA_character_) {
  if (key %in% names(meta)) meta[[key]] else default
}
fmt <- function(x, digits = 4) format(round(x, digits), nsmall = digits)
avg_rel <- as.numeric(meta_value("average_item_reliability"))
if (!length(avg_rel)) avg_rel <- NA_real_
avg_rel_text <- if (is.na(avg_rel)) "not recorded" else fmt(avg_rel)
estimators_text <- paste(toupper(sort(unique(method_summary$estimator))),
                         collapse = ", ")

if (!"mean_abs_error_primary" %in% names(method_summary) &&
    "mean_abs_bias_primary" %in% names(method_summary)) {
  method_summary$mean_abs_error_primary <- method_summary$mean_abs_bias_primary
}
if (!all(c("mean_bias_primary", "mean_abs_mean_bias_primary") %in%
         names(method_summary))) {
  derived <- lapply(seq_len(nrow(method_summary)), function(i) {
    key <- method_summary[i, ]
    d <- parameter_summary[parameter_summary$estimator == key$estimator &
                             parameter_summary$method == key$method &
                             parameter_summary$param_class %in%
                               c("loading", "latent_variance"), ]
    data.frame(
      mean_bias_primary = mean(d$bias, na.rm = TRUE),
      mean_abs_mean_bias_primary = mean(abs(d$bias), na.rm = TRUE)
    )
  })
  derived <- do.call(rbind, derived)
  if (!"mean_bias_primary" %in% names(method_summary)) {
    method_summary$mean_bias_primary <- derived$mean_bias_primary
  }
  if (!"mean_abs_mean_bias_primary" %in% names(method_summary)) {
    method_summary$mean_abs_mean_bias_primary <-
      derived$mean_abs_mean_bias_primary
  }
}
for (nm in c("primary_p99_abs_error", "primary_extreme_rate_abs10",
             "primary_extreme_rate_abs100")) {
  if (!nm %in% names(method_summary)) method_summary[[nm]] <- NA_real_
}

method_summary$Method <- factor(method_summary$method,
                                levels = c("standard", "posthoc", "integrated"))
primary <- method_summary[order(method_summary$estimator,
                                method_summary$Method), ]
best <- do.call(rbind, lapply(split(primary, primary$estimator), function(d) {
  d[which.min(d$mise_primary), , drop = FALSE]
}))
```

## Question

When do reduced-bias corrections improve point-estimation error, and when do
extreme corrections or failed fits erase that gain? Compare ordinary estimates,
post-hoc corrections and integrated reduced-bias fits, keeping failure rates
alongside errors among accepted fits.

## Short Answer

The retained 5,000-replication N=30 ML/GLS run exposes extreme post-hoc correction
tails. Its primary mean squared errors are about 5.3e13 for ML and 2.3e10 for
GLS, versus 0.65 and 1.05 for ordinary estimates. Integrated corrections have
lower error among accepted fits (0.15 and 0.79), but acceptance falls to 96.0%
and 88.8%, versus 99.96% and 100% for ordinary fits. This is historical evidence
for a stability investigation, not a current-library guarantee or an
unconditional risk ranking.

For the selected run (`r meta[["reps"]]` replicates, N = `r meta[["n"]]`), the
smallest error among accepted fits is:
`r paste(sprintf("%s: %s (%s)", toupper(best$estimator), best$method,
                 fmt(best$mise_primary)), collapse = "; ")`.
The default report selects `results/tough-n30-ml-gls`; the small four-estimator
smoke run remains available through `-P results_dir:results`.

## Evidence

```{r}
tab <- primary[c("estimator", "method", "mean_bias_primary",
                 "mean_abs_mean_bias_primary", "mean_abs_error_primary",
                 "mean_rmse_primary", "mise_primary",
                 "primary_extreme_rate_abs100",
                 "relative_mise_primary", "mean_abs_pu_error_primary",
                 "acceptance_rate", "median_elapsed_ms")]
names(tab) <- c("Estimator", "Method", "Mean Bias", "Mean |Bias|", "MAE",
                "Mean RMSE", "MISE", "|Err| > 100", "MISE / Standard",
                "Mean |PU - .5|", "Acceptance", "Median Time (ms)")
tab$Estimator <- toupper(tab$Estimator)
tab$Method <- c(standard = "Standard", posthoc = "Post-hoc",
                integrated = "Integrated")[tab$Method]
tab$`Mean Bias` <- fmt(tab$`Mean Bias`)
tab$`Mean |Bias|` <- fmt(tab$`Mean |Bias|`)
tab$MAE <- fmt(tab$MAE)
tab$`Mean RMSE` <- fmt(tab$`Mean RMSE`)
tab$MISE <- fmt(tab$MISE)
tab$`|Err| > 100` <- ifelse(is.na(tab$`|Err| > 100`), "",
                            sprintf("%.2f%%", 100 * tab$`|Err| > 100`))
tab$`MISE / Standard` <- fmt(tab$`MISE / Standard`, 3)
tab$`Mean |PU - .5|` <- fmt(tab$`Mean |PU - .5|`, 3)
tab$Acceptance <- sprintf("%.2f%%", 100 * tab$Acceptance)
tab$`Median Time (ms)` <- round(tab$`Median Time (ms)`)
knitr::kable(tab, row.names = FALSE)
```

The primary summary uses only the free factor loadings and latent variance, so
the requested estimator rows are comparable. MISE is the average, over
accepted replications, of the mean squared error across that primary parameter
vector. Failed fits are excluded from these error summaries and reported through
acceptance; an unconditional risk comparison needs an explicit failure policy.
`MAE` is mean absolute estimation error, not absolute mean bias. The full
parameter summaries, including residual variances, thresholds, and FIML
intercepts where present, are in `r file.path(res_dir, "parameter_summary.csv")`.

## Caveats

This is still not a replication of the paper's full design: it uses one
correctly specified CFA population, one sample size, normal latent responses,
MCAR missingness for FIML, and two thresholds for DWLS. The ML branch is the
paper-relevant likelihood path, but the paper's two-factor and growth-curve
grids remain larger and harder. This probe can catch gross direction and
stability problems in the standard/post-hoc/integrated comparison, but it cannot
establish uniform bias reduction, coverage, or optimizer reliability.

## Reproduce

```sh
Rscript experiments/research/banked/53-rbm-estimation-risk/run_experiment.R --smoke
Rscript experiments/research/banked/53-rbm-estimation-risk/run_experiment.R --reps 250 --n 160 --ref-n 50000 --cores auto --chunk-size 120
Rscript experiments/research/banked/53-rbm-estimation-risk/run_experiment.R --reps 5000 --n 30 --ref-n 100000 --estimators ml,gls --cores 1 --chunk-size 100 --results-dir results/tough-n30-ml-gls
quarto render experiments/research/banked/53-rbm-estimation-risk/report.qmd
quarto render experiments/research/banked/53-rbm-estimation-risk/report.qmd -P results_dir:results
```


## Next question

Recheck correction tails with the current library and explicit conditioning
diagnostics, preserving extreme draws and reporting failure rates. Compare
post-hoc and integrated corrections on error, tail quantiles and acceptance
before broadening the model grid. The original engineering/06 result trees moved here unchanged on 30 September 2026; no simulation was rerun
as part of the reorganization.
