---
title: "The closed-form Guttman recipe versus NTML"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: results
---

```{r setup}
results_dir <- params$results_dir
need <- c("metadata", "estimate_summary", "estimate_joint_summary",
          "whole_joint_summary", "diagnostic_summary", "timing_summary",
          "population", "progress")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run the grid first (see Reproduce), or point\n",
       "  -P results_dir:<dir> at a completed run.")
}
metadata   <- read.csv(paths[1]); est      <- read.csv(paths[2])
est_joint  <- read.csv(paths[3]); wjoint    <- read.csv(paths[4])
diag       <- read.csv(paths[5]); timing    <- read.csv(paths[6])
population <- read.csv(paths[7]); progress <- read.csv(paths[8])

fmt <- function(x, d = 2) ifelse(is.finite(x), formatC(x, format = "f", digits = d), "NA")
pct <- function(x, d = 0) ifelse(is.finite(x), paste0(formatC(100 * x, format = "f", digits = d), "%"), "NA")
med <- function(x) median(x[is.finite(x)])
# p = number of observed variables, joined by (factors, indicators)
pkey <- function(df) population$p[match(paste(df$factors, df$indicators),
                                        paste(population$factors, population$indicators))]
est$p <- pkey(est); est_joint$p <- pkey(est_joint); timing$p <- pkey(timing)

G <- "guttman"; M <- "ml_emp"
# headline numbers
rr    <- est_joint$rmse_ratio[est_joint$estimator == G & is.finite(est_joint$rmse_ratio)]
covG  <- med(est$coverage[est$estimator == G]); covM <- med(est$coverage[est$estimator == M])
sfit  <- med(wjoint$sigma_rmse_ratio[wjoint$estimator == G])
tg    <- tapply(timing$total_median_ms[timing$estimator == G], timing$p[timing$estimator == G], med)
tm    <- tapply(timing$total_median_ms[timing$estimator == M], timing$p[timing$estimator == M], med)
speed_big <- { ps <- as.integer(names(tg)); p <- max(ps); tm[[as.character(p)]] / tg[[as.character(p)]] }
failG <- 1 - med(diag$fit_ok_rate[diag$estimator == G] * diag$converged_rate[diag$estimator == G])
failM <- 1 - med(diag$fit_ok_rate[diag$estimator == M] * diag$converged_rate[diag$estimator == M])
ncells <- nrow(progress); ncells_ok <- sum(progress$draw_ok)
```

## Question

Does the recommended closed-form recipe -- `extended_triad_ls` communality with
the `standardized` composite on the aligned Guttman map -- match normal-theory ML
(NTML) on parameter RMSE and empirical-SE coverage, including the high-rho /
weak-loading regime that broke the retired adaptive composite, while being faster
and never failing to fit?

## Short answer

Across `r ncells_ok` grid cells (n, generator, factors, indicators, latent
correlation, scales, loading strength), the closed-form recipe recovers the free
parameters at a **median RMSE of `r fmt(med(rr))`x NTML** -- effectively a
dead heat -- with **empirical-SE coverage of `r pct(covG)`** (NTML `r pct(covM)`)
at a nominal 95%. It pays a modest **`r fmt(sfit)`x on implied-covariance fit**
(it is a covariance functional, not a discrepancy minimizer) and runs
**`r fmt(speed_big)`x faster than NTML+SE** at the largest models here, while
never failing to fit (Guttman fit failures `r pct(failG)`, NTML `r pct(failM)`).

```{r partial}
total_cells <- 1152
if (ncells_ok < total_cells)
  cat(sprintf("_Partial run: %d / %d cells complete._\n", ncells_ok, total_cells))
```

## Evidence

### Parameter RMSE vs NTML (same-replication ratio)

```{r rmse}
sub <- est_joint[est_joint$estimator == G & is.finite(est_joint$rmse_ratio), ]
m <- tapply(sub$rmse_ratio, list(sub$strength, sub$rho), med)
knitr::kable(round(m, 3), caption = "Median RMSE ratio (Guttman / NTML) by loading strength x latent correlation. 1.00 = tie.")
```

Values near 1 across the grid -- including `rho = 0.8` with weak loadings -- are
the headline: the closed form tracks ML on the parameters even in the regime that
used to blow up under the adaptive composite.

### Empirical-SE coverage by data type

```{r coverage}
cc <- est[est$estimator %in% c(G, M), ]
tab <- tapply(cc$coverage, list(cc$estimator, cc$generator), med)
rownames(tab) <- c(guttman = "Guttman (extended+std)", ml_emp = "NTML")[rownames(tab)]
knitr::kable(round(tab, 3), caption = "Median 95% CI coverage by generator (normal, inverse-Gaussian, ordinal codes).")
```

### The one cost: implied-covariance fit

```{r sigmafit}
sj <- wjoint[wjoint$estimator == G & is.finite(wjoint$sigma_rmse_ratio), ]
knitr::kable(
  data.frame(
    Statistic = c("Common-part RMSE ratio", "Sigma RMSE ratio", "Residual-diag RMSE ratio"),
    `Median vs NTML` = c(fmt(med(sj$common_rmse_ratio)), fmt(med(sj$sigma_rmse_ratio)),
                         fmt(med(sj$residual_diag_rmse_ratio))),
    check.names = FALSE),
  caption = "Whole-matrix recovery vs NTML. >1 = closed form fits the covariance less tightly.")
```

### Speed and reliability

```{r speed}
ps <- sort(unique(timing$p))
sp <- data.frame(
  p = ps,
  `Guttman total ms` = fmt(sapply(ps, function(pp) med(timing$total_median_ms[timing$estimator==G & timing$p==pp]))),
  `NTML total ms`    = fmt(sapply(ps, function(pp) med(timing$total_median_ms[timing$estimator==M & timing$p==pp]))),
  check.names = FALSE)
knitr::kable(sp, caption = "Median fit+SE time (ms) by number of observed variables p.")
```

Fit/convergence failures: Guttman `r pct(failG)`, NTML `r pct(failM)`. The
closed form is a deterministic map, so it always returns an estimate.

## Caveats

- Per-fit timings use a millisecond clock; sub-millisecond fits at small `p` read
  as ~1 ms, so the speed gap is clearest at larger `p`.
- The recipe runs through the restricted-map entry with an unconstrained partable
  (a bit-exact proxy) because the configural entry does not yet expose communality
  selection; the numbers are the configural aligned map's.
- Categorical cells fit the observed Pearson-code covariance directly; no
  polychoric statistics enter either arm.

## Reproduce

```sh
Rscript experiments/59-guttman-vs-ntml/run_experiment.R --full \
  --reps 2000 --cores 6 \
  --results-dir experiments/59-guttman-vs-ntml/results

quarto render experiments/59-guttman-vs-ntml/report.qmd
```
```
