---
title: "Pairwise Fit Speed: ML, Σ-only GLS, Γ_NT^pw GLS"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path), dirname(getwd())))
hits <- file.path(candidate_dirs, "results")
hits <- hits[file.exists(file.path(hits, "summary.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
summary_df <- utils::read.csv(file.path(results_dir, "summary.csv"),
                              check.names = FALSE)
if (!requireNamespace("ggplot2", quietly = TRUE))
  stop("The report requires ggplot2.")
library(ggplot2)
metadata <- utils::read.csv(file.path(results_dir, "metadata.csv"),
                            check.names = FALSE)
meta <- function(k, d = "") {
  v <- metadata$value[metadata$key == k]
  if (!length(v)) d else as.character(v[[1L]])
}
```

## Question

Three pairwise estimators of an SEM on incomplete continuous data —
which one is fastest? Trivial follow-up to experiment 08, which
established a Γ_NT^pw GLS ≥ pairwise ML > Σ-only GLS ranking on
statistical efficiency. Inference cost likely dominates the total
wall-time downstream, so the absolute differences here are mostly a
sanity check.

## Short Answer

**Pairwise ML is uniformly the fastest** — 1.2× faster than the GLS
variants at the 5-indicator model and 4×–5× faster at the
9-indicator model. The two GLS variants are roughly comparable to
each other (within ~50% in either direction across cells), both
several times slower than ML at the medium model size.

Absolute wall-times stay sub-25 ms per fit at p = 9, so the gap is
unlikely to matter for one-off model fitting. It does matter for
Monte-Carlo simulations (~1000 reps × ~10 cells: 4 minutes of GLS vs
50 seconds of ML), which is exactly the place pairwise estimators get
benchmarked.

## Evidence

```{r evidence-table}
estimator_pretty <- c(sigma = "Σ-only GLS", ml = "Pairwise ML",
                      gamma = "Γ_NT^pw GLS")
model_pretty <- c(brown_1f5i = "Brown-style 1F·5I",
                  hs1939_3f9i = "HS-1939-style 3F·9I")
tbl <- summary_df
tbl$Estimator <- estimator_pretty[tbl$estimator]
tbl$Model     <- model_pretty[tbl$model]
tbl$Mechanism <- toupper(tbl$mechanism)
tbl$median_ms <- round(tbl$median_ms, 2)
tbl$mad_ms    <- round(tbl$mad_ms, 2)
tbl$ratio_to_fastest <- round(tbl$ratio_to_fastest, 2)
out <- tbl[, c("Model", "Mechanism", "Estimator",
               "median_ms", "mad_ms", "ratio_to_fastest")]
names(out) <- c("Model", "Mechanism", "Estimator",
                "Median (ms)", "MAD (ms)", "× Slower than Fastest")
knitr::kable(out, caption = paste(
  "Median per-fit wall-time across microbenchmark replications.",
  "MAD is the median absolute deviation; ratio gives the relative slowdown",
  "vs the fastest estimator in the (Model, Mechanism) cell."
))
```

```{r evidence-plot, fig.width=7.5, fig.height=3.6}
plot_df <- summary_df
plot_df$estimator_pretty <- factor(
  estimator_pretty[plot_df$estimator],
  levels = c("Σ-only GLS", "Pairwise ML", "Γ_NT^pw GLS"))
plot_df$model_pretty <- model_pretty[plot_df$model]
plot_df$mechanism <- toupper(plot_df$mechanism)
ggplot(plot_df,
       aes(x = estimator_pretty, y = median_ms, fill = mechanism)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_errorbar(aes(ymin = pmax(median_ms - mad_ms, 0),
                    ymax = median_ms + mad_ms),
                position = position_dodge(width = 0.75), width = 0.2,
                linewidth = 0.4) +
  facet_wrap(~ model_pretty, scales = "free_y") +
  scale_fill_manual(values = c(MCAR = "#1f78b4", MAR = "#e31a1c")) +
  labs(x = NULL, y = "Median fit time (ms)", fill = "Mechanism") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 15, hjust = 1),
        legend.position = "top")
```

ML's per-iteration cost in magmaan is dominated by a Cholesky on Σ(θ)
and a log-det/trace pair; the GLS path additionally evaluates a
p* × p* weighted quadratic at each iteration, which grows as p⁴. At
p = 9 (p* = 45) that's already a meaningful constant. At p = 5
(p* = 15) it's not.

The MCAR vs MAR comparison shows essentially no effect on speed —
the missingness only changes which `apply_missingness` function ran
once at setup; the fitting cost is the same.

## Caveats

- Two models only. The qualitative story (ML ≪ GLS at meaningful p)
  is expected to hold up across larger CFA / SEM designs, but the
  multiplier at, say, p = 30 / 3-factor / cross-loadings isn't tested.
- One sample size (n = `r meta("n", "?")`), one missing rate
  (`r meta("rate", "?")`). Speed at p, p* sizes isn't strongly
  sensitive to n once the data fits in cache; missing rate affects
  the upstream `pairwise_sample_stats` call but that's outside the
  microbenchmark window.
- `microbenchmark` reports wall-time only; we don't separate optimizer
  inner iterations from per-iteration cost. A spike in iter count for
  a poorly-conditioned weight would show up here as a slow median.
- Inference (SE, chi-square, etc.) is not benchmarked. The user's prior
  expectation was that inference cost dominates the total runtime of a
  pairwise-SEM analysis; this experiment doesn't contradict that.

## Reproduce

```bash
cd experiments/09-pairwise-fit-speed
Rscript run_experiment.R --smoke    # quick check
Rscript run_experiment.R            # full bench (≈ 3 s)
quarto render report.qmd
```
