---
title: "lavaan audit parity"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
options(width = 120)

candidate_dirs <- c(getwd(), dirname(getwd()),
                    file.path("experiments", "00-lavaan-parity"))
result_candidates <- unique(file.path(candidate_dirs, "results"))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "audit_summary.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
need_file <- function(name) {
  path <- file.path(results_dir, name)
  if (!file.exists(path)) {
    stop("Missing results/", name,
         ". Run `Rscript experiments/00-lavaan-parity/run_experiment.R` first.",
         call. = FALSE)
  }
  path
}
read_result <- function(name) {
  utils::read.csv(need_file(name), stringsAsFactors = FALSE,
                  check.names = FALSE, na.strings = c("", "NA"))
}
read_optional <- function(name) {
  path <- file.path(results_dir, name)
  if (!file.exists(path)) return(data.frame())
  utils::read.csv(path, stringsAsFactors = FALSE, check.names = FALSE,
                  na.strings = c("", "NA"))
}

audit_summary <- read_result("audit_summary.csv")
current_failures <- read_result("current_failures.csv")
v7_disagree <- read_result("audit_disagreements_enriched.csv")
by_book <- read_result("audit_by_book.csv")
by_weight <- read_result("audit_by_weight.csv")
refit_summary <- read_result("refit_summary.csv")
lcs_summary <- read_result(file.path("lcs-objective-gap", "summary.csv"))
drivers <- read_result("driver_inventory.csv")
metadata <- read_result("metadata.csv")
v7_agreement_cells <- as.integer(audit_summary$n_with_both_fits[[1L]] -
                                   nrow(v7_disagree))
refit <- read_optional(file.path("refit-disagreements",
                                 "refit_disagreements_v8.csv"))

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), "", formatC(x, format = "g", digits = digits))
}
fmt_pct <- function(x) {
  ifelse(is.na(x), "", sprintf("%.1f%%", 100 * x))
}
html_escape <- function(x) {
  x <- as.character(x)
  x[is.na(x)] <- ""
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x <- gsub('"', "&quot;", x, fixed = TRUE)
  x
}
html_table <- function(x, status_col = NULL, limit = 30L) {
  if (is.null(x) || !nrow(x)) {
    cat("<p class=\"empty-table\">No rows.</p>\n")
    return(invisible(NULL))
  }
  if (!is.null(limit) && nrow(x) > limit) x <- x[seq_len(limit), , drop = FALSE]
  classes <- rep("", nrow(x))
  if (!is.null(status_col) && status_col %in% names(x)) {
    status <- tolower(as.character(x[[status_col]]))
    classes[grepl("fail|disagree|conditioning|gap", status)] <- " class=\"status-fail\""
    classes[grepl("error", status)] <- " class=\"status-error\""
    classes[grepl("agree|pass", status)] <- " class=\"status-pass\""
  }
  cat("<table class=\"audit-table\"><thead><tr>")
  cat(paste0("<th>", html_escape(names(x)), "</th>", collapse = ""))
  cat("</tr></thead><tbody>\n")
  for (i in seq_len(nrow(x))) {
    cat("<tr", classes[[i]], ">")
    cat(paste0("<td>", html_escape(x[i, , drop = TRUE]), "</td>",
               collapse = ""))
    cat("</tr>\n")
  }
  cat("</tbody></table>\n")
}
meta_value <- function(key) {
  hit <- metadata$value[metadata$key == key]
  if (length(hit)) hit[[1L]] else ""
}
```

<style>
.scope-box {
  border-left: 4px solid #475569;
  padding: 0.7rem 1rem;
  background: #f8fafc;
  margin: 1rem 0;
}
.failure-box {
  border-left: 4px solid #b91c1c;
  padding: 0.7rem 1rem;
  background: #fff1f2;
  margin: 1rem 0;
}
.pass-box {
  border-left: 4px solid #15803d;
  padding: 0.7rem 1rem;
  background: #f0fdf4;
  margin: 1rem 0;
}
.audit-table {
  border-collapse: collapse;
  width: 100%;
  margin: 0.5rem 0 1.4rem 0;
  font-size: 0.92rem;
}
.audit-table th,
.audit-table td {
  border: 1px solid #d7dde5;
  padding: 0.35rem 0.45rem;
  vertical-align: top;
}
.audit-table th {
  background: #e2e8f0;
}
.audit-table tr.status-fail td {
  background: #fff1f2;
}
.audit-table tr.status-error td {
  background: #fffbeb;
}
.audit-table tr.status-pass td {
  background: #f8fafc;
}
.empty-table {
  color: #64748b;
  font-style: italic;
}
</style>

## Scope

<div class="scope-box">
This experiment presents the paper-side lavaan audit-parity validation in a
single inspectable place. The full sweep fits each SNLLS textbook-corpus
<code>(case, weight)</code> cell in <code>lavaan</code>, aligns lavaan's
returned free-parameter vector to magmaan's model, evaluates the same point
with <code>magmaan_core$evaluate_at()</code>, and compares lavaan's
<code>lavInspect(fit, "converged")</code> self-report with magmaan's terminal
audit stationarity verdict. This is a convergence-verdict validation, not a
new estimator, not a fixture generator, and not the benchmark speed surface.
</div>

The tables below read the latest normalized outputs written by
`run_experiment.R`. Plain `run_experiment.R` copies the legacy paper-side
combined sweep, `r meta_value("audit_run")`; `--from-oracle` instead reuses a
lavaan-only cache and reruns just magmaan's audit. The targeted refit is not a
full sweep: it reruns only the 24 v7 disagreement cells and projects the
current failure surface under the assumption that the `r v7_agreement_cells`
v7 agreement cells did not regress.

## Oracle Cache

```{r}
#| results: asis
cache_tbl <- data.frame(
  Field = c("Audit Mode", "Audit Source", "Oracle Run", "lavaan Version",
            "Refit Source"),
  Value = c(meta_value("audit_mode"), meta_value("audit_source"),
            meta_value("oracle_run"), meta_value("lavaan_version"),
            meta_value("refit_source")),
  stringsAsFactors = FALSE
)
cache_tbl <- cache_tbl[!is.na(cache_tbl$Value) & nzchar(cache_tbl$Value),
                       , drop = FALSE]
html_table(cache_tbl, limit = NULL)
```

## Current Failures

```{r}
#| results: asis
if (nrow(current_failures)) {
  cat("<div class=\"failure-box\"><strong>Current/projected failures are shown first.</strong> ",
      "When the v8 refit table is present, this block is based on the 24-cell targeted refit; ",
      "otherwise it falls back to the v7 disagreement table.</div>\n", sep = "")
  if ("v8_agrees" %in% names(current_failures)) {
    tbl <- current_failures[, c("case", "v8_lav_conv", "v8_audit_stat",
                                "v8_grad_inf", "v8_magmaan_fmin",
                                "v8_lav_fx_bare", "v8_ratio", "v8_err"),
                            drop = FALSE]
    tbl$v8_grad_inf <- fmt_num(tbl$v8_grad_inf)
    tbl$v8_magmaan_fmin <- fmt_num(tbl$v8_magmaan_fmin)
    tbl$v8_lav_fx_bare <- fmt_num(tbl$v8_lav_fx_bare)
    tbl$v8_ratio <- fmt_num(tbl$v8_ratio)
    tbl$Status <- "current failure"
    names(tbl) <- c("Case", "lavaan Converged", "Audit Stationary",
                    "Audit Grad Inf", "magmaan fmin", "lavaan fmin bare",
                    "fmin Ratio", "Error", "Status")
    html_table(tbl, status_col = "Status", limit = NULL)
  } else {
    adf_cols <- c("adf_gamma_rank", "adf_gamma_dim",
                  "adf_conditioning_warning",
                  "adf_conditioned_fmin",
                  "adf_conditioned_grad_inf",
                  "adf_conditioned_stationary")
    for (nm in setdiff(adf_cols, names(current_failures))) {
      current_failures[[nm]] <- NA
    }
    tbl <- current_failures[, c("case_id", "classification", "weight",
                                "audit_grad_inf", "magmaan_fmin",
                                "lavaan_fx_bare", "fmin_ratio",
                                "adf_gamma_rank", "adf_gamma_dim",
                                "adf_conditioning_warning",
                                "adf_conditioned_fmin",
                                "adf_conditioned_grad_inf",
                                "adf_conditioned_stationary"),
                            drop = FALSE]
    tbl$adf_gamma_rank <- ifelse(
      is.na(tbl$adf_gamma_rank), "",
      paste0(tbl$adf_gamma_rank, "/", tbl$adf_gamma_dim)
    )
    tbl$audit_grad_inf <- fmt_num(tbl$audit_grad_inf)
    tbl$magmaan_fmin <- fmt_num(tbl$magmaan_fmin)
    tbl$lavaan_fx_bare <- fmt_num(tbl$lavaan_fx_bare)
    tbl$fmin_ratio <- fmt_num(tbl$fmin_ratio)
    tbl$adf_gamma_dim <- NULL
    tbl$adf_conditioned_fmin <- fmt_num(tbl$adf_conditioned_fmin)
    tbl$adf_conditioned_grad_inf <- fmt_num(tbl$adf_conditioned_grad_inf)
    names(tbl) <- c("Case", "Classification", "Weight", "Audit Grad Inf",
                    "magmaan fmin", "lavaan fmin bare", "fmin Ratio",
                    "ADF Γ Rank", "ADF Warning", "Conditioned fmin",
                    "Conditioned Grad Inf", "Conditioned Stationary")
    html_table(tbl, status_col = "Classification", limit = NULL)
  }
} else {
  cat("<div class=\"pass-box\"><strong>No current/projected failures.</strong></div>\n")
}
```

## Block A: Full-Sweep Headline

```{r}
#| results: asis
headline <- audit_summary
headline$agreement_rate <- fmt_pct(headline$agreement_rate)
names(headline) <- c("Audit Run", "Cells", "Cells With Both Fits", "Errors",
                     "Agree", "lavaan Yes / Audit No",
                     "lavaan No / Audit Yes", "Agreement Rate")
html_table(headline, limit = NULL)
```

```{r}
#| results: asis
if (nrow(refit_summary)) {
  html_table(refit_summary, limit = NULL)
}
```

## Block B: Disagreements From The Last Full Sweep

These are the v7 off-diagonal cells before the targeted v8 refit. The
classification column separates the LCS / phantom-latent objective gap, the
single-indicator `auto.fix.single` parity gap, the Muthen ADF conditioning side
finding, and ordinary gradient-verdict splits.

```{r}
#| results: asis
tbl <- v7_disagree[, c("case_id", "classification", "book", "family", "weight",
                       "lavaan_converged", "audit_stationary",
                       "audit_grad_inf", "magmaan_fmin", "lavaan_fx_bare",
                       "fmin_ratio"),
                   drop = FALSE]
for (nm in c("audit_grad_inf", "magmaan_fmin", "lavaan_fx_bare",
             "fmin_ratio")) {
  tbl[[nm]] <- fmt_num(tbl[[nm]])
}
names(tbl) <- c("Case", "Classification", "Book", "Family", "Weight",
                "lavaan Converged", "Audit Stationary", "Audit Grad Inf",
                "magmaan fmin", "lavaan fmin bare", "fmin Ratio")
html_table(tbl, status_col = "Classification", limit = NULL)
```

## Block C: Breakdown By Book And Weight

```{r}
#| results: asis
book_tbl <- by_book
book_tbl$agreement_rate <- fmt_pct(book_tbl$agreement_rate)
names(book_tbl) <- c("Block", "Book", "Cells", "Agree",
                     "lavaan Yes / Audit No", "lavaan No / Audit Yes",
                     "Agreement Rate")
html_table(book_tbl, limit = NULL)
```

```{r}
#| results: asis
weight_tbl <- by_weight
weight_tbl$agreement_rate <- fmt_pct(weight_tbl$agreement_rate)
names(weight_tbl) <- c("Block", "Weight", "Cells", "Agree",
                       "lavaan Yes / Audit No", "lavaan No / Audit Yes",
                       "Agreement Rate")
html_table(weight_tbl, limit = NULL)
```

## Block D: LCS Objective-Gap Probe

The LCS probe compares `Σhat(theta_hat)`, `muhat(theta_hat)`, sample
covariances, and sample means element by element on three target cells. It
diagnoses the old large fmin-ratio outliers as an implied-moment assembly
problem in LCS / phantom-latent patterns.

```{r}
#| results: asis
lcs_tbl <- lcs_summary
for (nm in c("magmaan_fmin", "lavaan_fx", "ratio", "audit_grad_inf",
             "worst_magmaan", "worst_lavaan", "worst_diff",
             "worst_rel_diff")) {
  if (nm %in% names(lcs_tbl)) lcs_tbl[[nm]] <- fmt_num(lcs_tbl[[nm]])
}
html_table(lcs_tbl, limit = NULL)
```

## Block E: Driver Inventory

```{r}
#| results: asis
html_table(drivers, limit = NULL)
```

## Reproduce

Normalize existing outputs and render this report:

```sh
Rscript experiments/00-lavaan-parity/run_experiment.R
(cd experiments/00-lavaan-parity && quarto render report.qmd)
```

Collect a new lavaan-only oracle cache using the newest CRAN lavaan, then audit
current magmaan against that cached oracle:

```sh
Rscript experiments/00-lavaan-parity/run_experiment.R --collect-lavaan --install-latest-lavaan
Rscript experiments/00-lavaan-parity/run_experiment.R --from-oracle
```

Run the legacy combined paper sweep only when needed:

```sh
Rscript experiments/00-lavaan-parity/run_experiment.R --run-full
```

Run the cheap targeted probes:

```sh
Rscript experiments/00-lavaan-parity/scripts/refit_disagreements.R
Rscript experiments/00-lavaan-parity/scripts/diagnose_adf_conditioning.R
Rscript experiments/00-lavaan-parity/run_experiment.R --run-lcs
```
