---
title: "Guttman CFA Asymptotic Efficiency"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- "results/asymptotic_summary.csv"
points_path <- "results/point_estimates.csv"
diag_path <- "results/diagnostics.csv"
if (!file.exists(summary_path) || !file.exists(points_path) || !file.exists(diag_path)) {
  stop("Missing results. Run: Rscript run_experiment.R --smoke")
}
summary <- read.csv(summary_path)
points <- read.csv(points_path)
diagnostics <- read.csv(diag_path)

focus <- subset(summary, group == "all" & n_ref == 100)
best <- do.call(rbind, lapply(split(focus, paste(focus$reliability, focus$factor_cor, focus$misspec)), function(x) {
  x[which.min(x$risk), , drop = FALSE]
}))
best_counts <- sort(table(best$estimator), decreasing = TRUE)
best_sentence <- paste(paste(names(best_counts), as.integer(best_counts), sep = "="), collapse = ", ")
guttman_rows <- subset(focus, estimator == "Guttman")
guttman_nt_median <- median(guttman_rows$risk_ratio_to_nt, na.rm = TRUE)
guttman_nt_min <- min(guttman_rows$risk_ratio_to_nt, na.rm = TRUE)
guttman_nt_max <- max(guttman_rows$risk_ratio_to_nt, na.rm = TRUE)

show <- subset(focus, reliability == min(focus$reliability) & factor_cor == min(focus$factor_cor))
show$Misspecification <- c(
  none = "None",
  cross_loading = "Omitted loading",
  residual_covariance = "Omitted residual covariance"
)[show$misspec]
show$Estimator <- show$estimator
show$`Bias Norm` <- round(show$bias_norm, 3)
show$`Asymptotic SD Norm` <- round(show$asymp_sd_norm, 3)
show$`Risk / NT` <- round(show$risk_ratio_to_nt, 2)
show$`Risk / Guttman` <- round(show$risk_ratio_to_guttman, 2)
show <- show[order(show$misspec, show$estimator),
             c("Misspecification", "Estimator", "Bias Norm",
               "Asymptotic SD Norm", "Risk / NT", "Risk / Guttman")]
```

## Question

Does Guttman's closed-form CFA estimator have competitive first-order variance
and fixed-misspecification risk relative to NT, ULS, GLS, and oracle WLS in the
small three-factor CFA design used by Dhaene and Rosseel?

## Short Answer

In this population-level run, the lowest all-parameter risk at \(N=100\) is split
as follows across design cells: `r best_sentence`. Guttman's median risk ratio
to NT is `r round(guttman_nt_median, 2)` across the same cells, ranging from
`r round(guttman_nt_min, 2)` to `r round(guttman_nt_max, 2)`.

Interpret this as an asymptotic target check, not a finite-sample convergence
study. It compares pseudo-target bias plus first-order sampling variance; it
does not simulate Heywood cases or optimizer failures.

## Evidence

The table shows the hardest displayed slice: reliability 0.5, factor correlation
0.1, and \(N=100\). Ratios below 1 beat the denominator on the displayed
all-parameter risk.

```{r}
knitr::kable(show)
```

## Caveats

The runner uses finite differences for the Guttman covariance map and for the
implicit estimator maps. The diagnostics file records convergence and covariance
matrix eigenvalues; inspect it before trusting a new grid.

The WLS column is an oracle fixed-weight moment estimator using the population
normal fourth-moment covariance. GLS recomputes the normal-theory sample weight
as the covariance summary changes. Under non-normal data these two should be
extended with non-normal fourth moments, which is the next useful slice.

The experiment is covariance-only and uses one small simple-structure CFA
family. Larger indicator blocks are where Guttman's computational appeal should
be strongest.

## Reproduce

```sh
cd experiments/40-guttman-cfa-asymptotics
Rscript run_experiment.R --smoke
Rscript run_experiment.R --full
quarto render report.qmd
```
