---
title: "What does covariance-honest NTML repair?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("full", "pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
corpus <- read_result("corpus_anatomy.csv")
changes <- read_result("corpus_parameter_changes.csv")
raw <- read_result("risk_raw.csv")
pairs <- read_result("risk_pairs.csv")
method_summary <- read_result("risk_method_summary.csv")
pair_summary <- read_result("risk_pair_summary.csv")
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)
profile <- metadata[["profile"]]
reps <- as.integer(metadata[["reps"]])
is_evidence <- profile %in% c("pilot", "full") && reps >= 200L

fmt_pct <- function(x) sprintf("%.1f%%", 100 * x)
fmt_num <- function(x, digits = 3L) formatC(
  x, format = "f", digits = digits)

ordinary_bad <- sum(!pairs$ordinary_admissible)
all_psd_usable <- all(pairs$psd_usable)
repaired <- pairs[pairs$both_analysis_ok & !pairs$ordinary_admissible, ]
mean_repaired_psi <- mean(repaired$psi_error_improvement)
mean_repaired_sigma <- mean(repaired$sigma_error_improvement)
mean_repaired_kl <- mean(repaired$kl_improvement)
rate_repaired_psi <- mean(repaired$psi_error_improvement > 1e-10)
rate_repaired_sigma <- mean(repaired$sigma_error_improvement > 1e-10)
rate_repaired_kl <- mean(repaired$kl_improvement > 1e-10)
median_time_ratio <- median(pairs$time_ratio, na.rm = TRUE)

short_answer <- if (!is_evidence) {
  sprintf(
    paste(
      "This is a %s validation run with %d replications per cell.",
      "It verifies the design and report, but it is not the prespecified",
      "risk pilot."
    ), profile, reps)
} else {
  sprintf(
    paste(
      "Across %d generated datasets, ordinary NTML left the primitive",
      "covariance domain %d times; every warm PSD refit was KKT-converged",
      "and admissible. Conditional on an ordinary inadmissible estimate,",
      "PSD-ML reduced mean relative error by %.4f for Psi and %.4f for",
      "Sigma, and reduced mean population KL loss by %.4f. It improved",
      "all three losses in %.1f%%, %.1f%%, and %.1f%% of repaired datasets,",
      "respectively. The median PSD/ordinary fit-time ratio was %.2f."
    ),
    nrow(pairs), ordinary_bad,
    mean_repaired_psi, mean_repaired_sigma, mean_repaired_kl,
    100 * rate_repaired_psi, 100 * rate_repaired_sigma,
    100 * rate_repaired_kl, median_time_ratio)
}

corpus_display <- corpus[, c(
  "set", "id", "geometry", "ordinary_psi_min", "psd_psi_min",
  "lr_scale_increase", "max_abs_parameter_change",
  "implied_sigma_relative_distance")]
names(corpus_display) <- c(
  "Set", "Case", "Geometry", "Ordinary min eig(Psi)",
  "PSD min eig(Psi)", "LR-scale cost", "Max parameter change",
  "Relative implied-Sigma change")
for (field in names(corpus_display)[4:8]) {
  corpus_display[[field]] <- format(
    corpus_display[[field]], digits = 4L, scientific = TRUE)
}

display_n <- sort(unique(pair_summary$n))
display_lambda <- sort(unique(pair_summary$lambda_min))
positive_lambda <- display_lambda[display_lambda > 0]
lambda_offset <- if (length(positive_lambda)) min(positive_lambda) / 2 else 1
lambda_coordinate <- function(x) log10(x + lambda_offset)
risk_display <- pair_summary[, c(
  "n", "lambda_min", "ordinary_inadmissible_rate",
  "psd_boundary_rate", "repaired_mean_psi_error_improvement",
  "repaired_mean_sigma_error_improvement",
  "repaired_mean_kl_improvement", "median_time_ratio")]
names(risk_display) <- c(
  "N", "lambda", "Ordinary inadmissible", "PSD boundary",
  "Psi error improvement", "Sigma error improvement",
  "KL improvement", "Time ratio")
risk_display$`Ordinary inadmissible` <- vapply(
  risk_display$`Ordinary inadmissible`, fmt_pct, character(1))
risk_display$`PSD boundary` <- vapply(
  risk_display$`PSD boundary`, fmt_pct, character(1))
for (field in c("Psi error improvement", "Sigma error improvement",
                "KL improvement", "Time ratio")) {
  risk_display[[field]] <- round(risk_display[[field]], 4L)
}
```

## Question

When ordinary complete-data normal-theory ML estimates an indefinite primitive
LISREL covariance, what does an explicit covariance-honest refit change? In a
controlled Gaussian model generated from a genuinely PSD population, does the
repair improve estimation risk near the covariance boundary?

## Short answer

`r short_answer`

This is an estimation experiment. The objective difference is descriptive and
is not assigned a p-value or likelihood-ratio reference distribution.

## Repair anatomy

```{r corpus-table}
knitr::kable(corpus_display, align = "lllrrrrr")
```

The four deterministic cases show that “an improper solution” is not a single
phenomenon. The first Little model permits a nontrivial reallocation between
residual and latent level variance with effectively unchanged likelihood and
implied observed covariance. Its linear-growth companion pays a substantial
fit cost and moves the fitted observed covariance materially. The Geiser
second-order disturbance is repaired at negligible likelihood cost, whereas
the jointly indefinite quadratic-growth covariance produces an intermediate
change despite all individual variances being positive.

The LR-scale cost is $2N(F_{PSD}-F_{ordinary})$. It is shown only to make the
size of the likelihood sacrifice legible; it is not a test statistic in this
report.

## Controlled covariance path

The risk design fixes indicator residual variance at $0.2$ and varies the
smallest eigenvalue of a three-dimensional latent covariance:

\[
  \operatorname{eig}(\Psi)=(1,0.4,\lambda_{min}),\qquad
  \Sigma=\Psi+0.2I.
\]

Consequently, the observed Gaussian likelihood remains regular even when the
primitive latent covariance is exactly singular. Ordinary NTML is saturated in
the observed covariance; PSD-ML changes it only when the implied primitive
estimate would leave the covariance domain.

```{r inadmissibility-plot, fig.width=7.5, fig.height=4.8}
colors <- grDevices::hcl.colors(length(display_n), "Dark 3")
plot(
  range(lambda_coordinate(display_lambda)), c(0, 1), type = "n",
  xlab = expression(lambda[min]),
  ylab = "Proportion",
  main = "When does ordinary NTML leave the PSD domain?",
  xaxt = "n")
axis(1, at = lambda_coordinate(display_lambda),
     labels = format(display_lambda, scientific = FALSE, trim = TRUE))
for (i in seq_along(display_n)) {
  x <- pair_summary[pair_summary$n == display_n[[i]], ]
  lines(lambda_coordinate(x$lambda_min), x$ordinary_inadmissible_rate,
        type = "b", pch = 14L + i, col = colors[[i]], lwd = 2)
}
legend("topright", legend = paste0("N = ", display_n),
       col = colors, pch = 15L + seq_along(display_n), lwd = 2,
       bty = "n")
```

The horizontal coordinate is $\log_{10}(\lambda+c)$, with $c$ equal to half
the smallest positive grid value, so the exact zero boundary remains visible.
At the exact population boundary, the constrained estimate may itself land either on the
boundary or in the interior. The fitted rank is therefore a description of the
sample solution, not a rank-selection claim.

```{r risk-table}
knitr::kable(risk_display, align = "rrrrrrrr")
```

Positive error improvements favor PSD-ML and are calculated conditional on the
ordinary estimate being inadmissible. `NA` cells indicate that no repair was
triggered in that cell. When ordinary NTML is already admissible, the paired
solutions agree to numerical precision.

## Population KL loss

```{r kl-plot, fig.width=8, fig.height=7}
old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1))
on.exit(par(old_par), add = TRUE)
method_colors <- c(ordinary = "#B23A48", psd = "#246A73")
for (n_value in display_n) {
  x <- method_summary[method_summary$n == n_value, ]
  ylim <- range(x$mean_population_kl, finite = TRUE)
  plot(
    range(lambda_coordinate(display_lambda)), ylim, type = "n", xaxt = "n",
    xlab = expression(lambda[min]), ylab = "Mean population KL",
    main = paste0("N = ", n_value))
  axis(1, at = lambda_coordinate(display_lambda),
       labels = format(display_lambda, scientific = FALSE, trim = TRUE))
  for (method in c("ordinary", "psd")) {
    z <- x[x$method == method, ]
    lines(lambda_coordinate(z$lambda_min), z$mean_population_kl, type = "b",
          pch = if (method == "ordinary") 16 else 17,
          col = method_colors[[method]], lwd = 2)
  }
  if (n_value == display_n[[1L]]) {
    legend("topright", legend = c("Ordinary", "PSD"),
           col = method_colors, pch = c(16, 17), lwd = 2, bty = "n")
  }
}
```

The constraint is beneficial where the ordinary estimate violates a true
population restriction. As sample size grows or the population moves inward,
ordinary inadmissibility becomes rare and the two estimators coincide. The
experiment therefore supports an audit-first policy, not unconditional silent
replacement of ordinary NTML.

## Numerical checks and limits

- The run contains `r nrow(pairs)` paired datasets and
  `r nrow(raw)` fitted-result rows.
- `r if (all_psd_usable) "Every PSD refit passed the KKT and covariance-admissibility audits." else "At least one PSD refit failed; inspect the raw results."`
- Ordinary and PSD estimates both returned for
  `r fmt_pct(mean(pairs$both_analysis_ok))` of datasets.
- The median warm-PSD/ordinary time ratio was
  `r fmt_num(median_time_ratio, 2L)`.

This deliberately transparent model does not represent the full difficulties
of a latent SEM with estimated loadings, structural regressions, equality
links, or competing likelihood basins. Those geometries are represented by the
four repair cases and prior validation experiments, but the risk conclusion is
limited to this controlled DGP. No inferential coverage or hypothesis-testing
claim is made.

## Reproduce

From this directory:

- **Smoke:** `N = 30,100`, `lambda = 0,.01,.20`, five replications.
- **Pilot:** `N = 25,50,100,250`,
  `lambda = 0,.001,.01,.05,.20`, 200 replications.
- **Full:** the pilot grid with 1,000 replications. Run only if the pilot is too
  imprecise for the descriptive risk comparisons.

Seeds are

```text
20260731 + 100003 * N + 1009 * lambda_index + replication.
```

Run with:

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
quarto render report.qmd
```

All result artifacts are ignored. The report reads only this experiment's
`results/` directory.

## Study design

### Part 1: repair anatomy

Four self-contained model/data summaries reproduce distinct cases from the
Little and Geiser continuous corpus:

1. a same-fit level/residual covariance reallocation;
2. a materially different linear-growth optimum;
3. a negative second-order disturbance repaired at negligible likelihood cost;
   and
4. a jointly indefinite quadratic-growth covariance with positive individual
   variances.

The local input file records the model syntax, sample covariance and mean,
sample size, and ordinary start. It is a private copy owned by this experiment;
the runner does not read another experiment, a test fixture, or a paper tree.

For each case, fit ordinary NTML by L-BFGS from the recorded ordinary solution,
then warm-start PSD-ML with SLSQP. Record:

- primitive covariance minimum eigenvalues and admissibility;
- objective difference and the descriptive scale `2 * N * delta_F`;
- maximum free-parameter displacement and its partable label;
- relative Frobenius distance between the two implied observed covariances;
- KKT-audited convergence; and
- fit time.

The scaled objective difference is called an “LR-scale cost” only because it is
on the familiar likelihood-ratio scale. It is not calibrated and must never be
reported as a likelihood-ratio test.

### Part 2: controlled risk path

Use three single-indicator latent variables with fixed unit loadings and fixed
indicator residual variances of `0.2`:

```text
f1 =~ 1*x1
f2 =~ 1*x2
f3 =~ 1*x3
x1 ~~ 0.2*x1
x2 ~~ 0.2*x2
x3 ~~ 0.2*x3
f1 ~~ f1 + f2 + f3
f2 ~~ f2 + f3
f3 ~~ f3
```

Let the population latent covariance be

\[
  \Psi(\lambda)=Q\operatorname{diag}(1,0.4,\lambda)Q',
\]

where the columns of $Q$ are the fixed orthonormal vectors

\[
  (1,1,1)'/\sqrt3,\quad
  (1,-1,0)'/\sqrt2,\quad
  (1,1,-2)'/\sqrt6.
\]

The observed covariance is

\[
  \Sigma(\lambda)=\Psi(\lambda)+0.2I.
\]

Thus $\Sigma$ remains safely positive definite even at the exact primitive
boundary $\lambda=0$. The design isolates covariance-component honesty from
singular observed likelihoods, loading indeterminacy, and structural weak
identification. Unconstrained NTML is saturated in the observed covariance and
may estimate an indefinite $\Psi$; PSD-ML fits the same likelihood over
$\Psi\succeq0$.

For every generated dataset, fit ordinary NTML and then an ordinary-warm
PSD-ML refit. Compare:

- convergence and covariance admissibility;
- fitted primitive rank and minimum eigenvalue;
- relative Frobenius error in $\Psi$ and $\Sigma$;
- population Gaussian KL loss;
- descriptive likelihood cost; and
- computation time.

The paired risk improvement is defined as ordinary error minus PSD error, so a
positive value favors PSD-ML. Summaries are reported both unconditionally and,
more importantly, conditional on the ordinary estimate being inadmissible.
The latter describes what the explicit audit/refit policy changes.

Fitted rank is descriptive. At a true covariance boundary, a constrained MLE
can lie on the boundary or in the interior with nonvanishing probability; this
experiment does not treat fitted rank as a consistent rank selector.
