---
title: "Does PSD-ML reach the same maximum from different starts?"
subtitle: "A multistart basin audit for the small-N SEM"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("pilot", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)
screen <- read_result("screen.csv")
datasets <- read_result("dataset_summary.csv")
classification <- read_result("classification_summary.csv")
starts <- read_result("start_summary.csv")
candidates <- read_result("confirmation_candidates.csv")
has_confirmation <- file.exists(file.path(root, "confirmation_summary.csv"))
confirmation <- if (has_confirmation) {
  read_result("confirmation_summary.csv")
} else {
  data.frame()
}
has_profiles <- file.exists(file.path(root, "profile_summary.csv"))
profiles <- if (has_profiles) {
  read_result("profile_summary.csv")
} else {
  data.frame()
}

profile <- metadata[["profile"]]
core <- classification[classification$scope == "random_core", ]
is_evidence <- profile == "pilot" &&
  as.integer(metadata[["core_reps"]]) >= 100L
fmt_pct <- function(x) sprintf("%.1f%%", 100 * x)

screen_rows <- lapply(split(screen, screen$n), function(x) {
  data.frame(
    N = x$n[[1L]],
    Screened = nrow(x),
    `Default KKT + admissible` = fmt_pct(mean(x$eligible)),
    `Default covariance boundary` = fmt_pct(mean(x$covariance_boundary)),
    check.names = FALSE
  )
})
screen_display <- do.call(rbind, screen_rows)

core_display <- core[, c(
  "n", "datasets", "default_best_rate", "ordinary_warm_best_rate",
  "competing_basin_rate", "different_parameter_rate",
  "different_moment_rate", "start_sensitive_rate",
  "best_boundary_rate", "max_default_gap"
)]
names(core_display) <- c(
  "N", "Datasets", "Default hits best", "Ordinary warm hits best",
  "Competing basins", "Same fit/different parameters",
  "Equal objective/different moments", "Start-sensitive eligibility",
  "Best solution on boundary", "Maximum default objective gap"
)
for (field in c(
  "Default hits best", "Ordinary warm hits best", "Competing basins",
  "Same fit/different parameters", "Equal objective/different moments",
  "Start-sensitive eligibility", "Best solution on boundary"
)) {
  core_display[[field]] <- vapply(
    core_display[[field]], fmt_pct, character(1)
  )
}
core_display$`Maximum default objective gap` <- format(
  core_display$`Maximum default objective gap`,
  scientific = TRUE, digits = 3
)

start_core <- starts[starts$scope == "random_core", c(
  "n", "start_family", "attempts", "eligible_rate", "best_hit_rate",
  "median_elapsed_ms", "median_f_evals"
)]
start_core$eligible_rate <- vapply(
  start_core$eligible_rate, fmt_pct, character(1))
start_core$best_hit_rate <- vapply(
  start_core$best_hit_rate, fmt_pct, character(1))
names(start_core) <- c(
  "N", "Start family", "Attempts", "Eligible", "Hits best attained",
  "Median time (ms)", "Median objective evaluations"
)

default_miss <- sum(
  datasets$random_core & !datasets$default_best
)
core_n <- sum(datasets$random_core)
upper <- if (core_n) {
  stats::binom.test(default_miss, core_n, alternative = "less")$conf.int[[2L]]
} else {
  NA_real_
}
competing <- sum(
  datasets$random_core &
    datasets$objective_basin_class == "competing"
)
different_moments <- sum(
  datasets$random_core &
    datasets$best_equivalence_class ==
      "equal-objective-different-moments"
)
different_parameters <- sum(
  datasets$random_core &
    datasets$best_equivalence_class ==
      "same-fit-different-parameters"
)
candidate_n <- nrow(candidates)
confirmed_display <- if (has_confirmation) {
  do.call(rbind, lapply(sort(unique(datasets$n)), function(n_value) {
    core_n_value <- sum(datasets$random_core & datasets$n == n_value)
    initial_n <- sum(
      datasets$random_core & datasets$n == n_value &
        datasets$objective_basin_class == "competing"
    )
    confirmed_n <- sum(
      confirmation$random_core & confirmation$n == n_value &
        confirmation$persistent_competing
    )
    data.frame(
      N = n_value,
      `Random-core datasets` = core_n_value,
      `Initial competing basins` = initial_n,
      `Persistent after tight restart` = confirmed_n,
      `Persistent rate` = fmt_pct(confirmed_n / core_n_value),
      check.names = FALSE
    )
  }))
} else {
  data.frame()
}
profile_display <- if (has_profiles) {
  do.call(rbind, lapply(sort(unique(profiles$n)), function(n_value) {
    subset <- profiles[profiles$n == n_value, ]
    multiple <- sum(subset$local_envelope_minima >= 2L)
    data.frame(
      N = n_value,
      `Profiled cases` = nrow(subset),
      `At least two envelope minima` = multiple,
      `Rate among targeted cases` = fmt_pct(multiple / nrow(subset)),
      check.names = FALSE
    )
  }))
} else {
  data.frame()
}
short_answer <- if (!is_evidence) {
  sprintf(
    paste(
      "This is a %s run. It validates the multistart machinery but is too",
      "small for a basin-frequency conclusion."
    ),
    profile
  )
} else {
  sprintf(
    paste(
      "Across the %d-dataset random core, the default start missed the",
      "best attained admissible KKT objective %d times (one-sided 95%%",
      "upper bound %.1f%%). We found %d datasets with competing stationary",
      "objective basins, %d with equal best objectives but different",
      "implied moments, and %d with the same fit but distinct parameter",
      "representatives. %d enriched or random-core datasets require the",
      "separate confirmation stage."
    ),
    core_n, default_miss, 100 * upper, competing, different_moments,
    different_parameters, candidate_n
  )
}
if (is_evidence && has_confirmation) {
  confirmed_total <- sum(
    confirmation$random_core & confirmation$persistent_competing
  )
  short_answer <- paste0(
    short_answer,
    sprintf(
      " Tight 20,000-iteration restarts retained competing basins in %d ",
      confirmed_total
    ),
    "random-core datasets."
  )
}
if (is_evidence && has_profiles) {
  multiple_profiles <- sum(profiles$local_envelope_minima >= 2L)
  short_answer <- paste0(
    short_answer,
    sprintf(
      " Of %d targeted beta profiles, %d had at least two local minima ",
      nrow(profiles), multiple_profiles
    ),
    "in the feasible lower envelope."
  )
}
```

## Question

Does covariance-honest complete-data ML converge to the same likelihood
maximum from a reasonable portfolio of standard, warm, restart, and perturbed
starts?

## Short answer

`r short_answer`

“Best attained” means the smallest objective among the admissible,
KKT-stationary solutions found by the prespecified portfolio. It is not a
claim that a global maximum has been proved.

## Screen

The screen independently generates the \(N=10,20,50\) datasets and runs the
standard PSD-SLSQP fit once. A random core supports frequency statements.
Every default failure and the largest-\(|\hat\beta|\) estimates are added only
as diagnostic enrichment.

```{r screen-table}
knitr::kable(screen_display, row.names = FALSE)
```

## Basin audit

Objective values, original partable parameters, and implied moments are
clustered separately. This distinguishes an inferior stationary basin from an
observationally equivalent parameter representative. Start-sensitive
eligibility is another axis: it is possible for some starts to fail even when
all successful starts reach one objective basin.

```{r core-table}
knitr::kable(core_display, row.names = FALSE)
```

The objective-equivalence threshold is
\(10^{-6}\max(1,|F_\star|)\), with raw gaps retained for sensitivity checks.
Parameter and implied-moment equivalence use scaled distances of \(10^{-4}\)
and \(10^{-6}\), respectively.

## Start families

```{r start-table}
knitr::kable(start_core, row.names = FALSE)
```

The ordinary-ML fit is used only as a source of starting values. An
inadmissible ordinary estimate is not treated as a candidate solution in the
PSD model class.

## Confirmation boundary

Only competing objective basins or equal-objective/different-moment cases are
written to `confirmation_candidates.csv`. They receive high-budget restarts
and a bidirectional profile over the structural coefficient before being
described as evidence of multiple maxima. IPOPT remains an optional backend
cross-check from the near-solution representatives; the pilot did not require
it after experiment 74 established its non-finite-callback failure mode. The
base runner deliberately does not launch adaptive confirmation automatically.

```{r confirmation-results}
if (has_confirmation) {
  cat(
    "One representative from each initial cluster was restarted with a ",
    "20,000-iteration budget and tighter tolerances. The table reports ",
    "persistence against the original 100-dataset random core at each N.\n\n",
    sep = ""
  )
  print(knitr::kable(confirmed_display, row.names = FALSE))
}
```

All initially flagged equal-objective/different-moment cases collapsed under
the tight restart. They are therefore treated as numerical convergence
artifacts, not distinct fitted distributions.

```{r profile-results}
if (has_profiles) {
  cat(
    "\nThe beta-profile tier targeted persistent random-core cases, ",
    "prioritizing default misses with the largest objective gaps. ",
    "It is diagnostic and intentionally not a frequency sample.\n\n",
    sep = ""
  )
  print(knitr::kable(profile_display, row.names = FALSE))
}
```

## Caveats

This experiment concerns one correctly specified two-factor Gaussian SEM and
one explicit start portfolio. Failure to find competing basins is evidence
about that design, not a general proof of globality for SEM likelihoods.
Boundary maxima are retained without regularization because they belong to
the specified PSD model class.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript run_confirmation.R
Rscript run_profiles.R
quarto render report.qmd
```
