---
title: "Savalei–Falk (2014): test conventions and numerical fingerprints"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
read_evidence <- function(name) {
  path <- file.path("results", "overview", name)
  if (!file.exists(path)) stop("Missing retained summary: ", path,
                               "; recover original CSVs and run scripts/summarize.R")
  read.csv(path, stringsAsFactors = FALSE)
}
show_table <- function(x, digits = 3) knitr::kable(x, digits = digits, row.names = FALSE)
```

## Question

Which published conventions and numerical matrix choices explain the
Savalei–Falk robust FIML and two-stage tests, and what still blocks exact replication?

## Short answer

This **active reference study** combines the definitions audit and numerical
matrix investigation. The written observed-information convention and the
closest empirical rejection-rate fingerprint differ. The expected-information
hybrid fits six MCAR targets best among the displayed candidates; all-saturated
expected information fits the harsh MAR target better. Neither identifies the
paper-era EQS 6.1 statistic exactly.

The next gate is an independent same-data matrix/trace reference plus original
EQS output or source. The full 192-cell published design is preserved as a
reference, not a queued simulation.

## Evidence

### Definitions lane

The exact-fit null is a structured CFA mean/covariance model, with saturated
observed-data moments as $H_1$. Direct FIML scales the raw likelihood ratio;
two-stage uses $(n-1)F_{ML}$ after Stage-1 saturated EM. Means are freely estimated
and cancel from the degrees of freedom. `SE=EXACT` is described by the article
as analytic observed information; that statement does not identify every
finite-sample matrix used internally by the robust statistic.

The [published conventions](design/published_conventions.md) retain the formulas,
loading and deletion maps, df conventions and proposed matrix checks. The
[matrix conventions](design/matrix_conventions.md) distinguish the public EQS
manual, the article and the expected-information-like numerical fingerprint.
The publications and manual are evidence for their written formulas; current
lavaan reconstructions are secondary sentinels, not EQS oracles.

A deterministic six-indicator MAR routing witness distinguishes implementations
on one sample. Its [retained table](results/overview/routing_witness.csv) is a
wiring check, not Monte Carlo evidence or a paper replication cell.

### Matrices lane

The runner crosses 48 FIML and 32 ML2S matrix settings, with named MLR kept
separate. Six MCAR cells retain 1,000 attempts each, matching $N=200,400,600$
and excess kurtosis 7/15. The paper targets are approximate graph readings from
1,000 replications. RMSE is in percentage points:

```{r fingerprint}
show_table(read_evidence("fingerprint.csv"))
```

The table reports usable counts because indefinite information/projectors and
other failures differ by route. Selecting a close fingerprint after seeing
these targets is exploratory; rejection-rate similarity cannot prove an
information-matrix identity. Both Monte Carlo and graph-reading uncertainty
limit interpretation of small differences.

The exact paper MAR-L/kurtosis-7 diagnostic tells a different part of the story:

```{r mar}
show_table(read_evidence("mar_comparison.csv"))
```

Its FIML paper target is .633. The saturated-expected candidate reaches .619,
while the six-cell-winning hybrid reaches .572. The written observed proxy is
farther away and loses fits. The source-style two-stage $(n-1)$ numerator is
closer than the $n$ alternative. Four diagnostic cells are retained separately;
none constitutes the full published grid.

## Caveats

The selected Fleishman root, excess-kurtosis convention and joint distribution
are substantive DGP choices. Alternative roots remain recorded. Conditional
MAR deletion proceeds set by set; infeasible candidate samples are regenerated
as documented by the runner. A generating exact-fit model under that selection
law is not automatically an estimator-level pseudo-null.

Original EQS 6.1 source/output is unavailable. The public manual predates its
two-stage implementation and cannot independently certify its numerator.
The literal observed route remains a proxy for the written formula. The retained
metadata's short resume time covers reused cells and is not a full-run benchmark.
No current library route or inference default changes here.

## Reproduce

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --lane definitions --audit \
  --results-dir /tmp/savalei_falk_definitions_check
Rscript run_experiment.R --lane matrices --smoke --reps 1 --cores 1 \
  --results-dir /tmp/savalei_falk_matrices_check
Rscript scripts/summarize.R
```

Inspect each lane's `--help` for its focused designs. Historical sources,
raw/checkpoint trees, published maps and frozen matrix summaries retain their
seed schedules and identities. [Source hashes](results/overview/sources.csv)
identify the inputs to this compact report. New runs use fresh output directories;
source-fingerprint resume guards stay strict after the move.
