---
title: "Where does SEM runtime go?"
subtitle: "Experiment showcases/08 · continuous ML pilot"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
params:
  run: pilot
  audit: tail-audit
---

```{r}
path <- file.path('results',params$run)
if(!file.exists(file.path(path,'summary.csv')))
  stop('Run Rscript experiments/showcases/08-speed-attribution/run_experiment.R first, or set -P run:<run-id>.')
s <- read.csv(file.path(path,'summary.csv'))
c <- read.csv(file.path(path,'checks.csv'))
v <- read.csv(file.path(path,'values.csv'))
d <- read.csv(file.path(path,'diagnostics.csv'))
t <- read.csv(file.path(path,'samples.csv'))
meta <- read.csv(file.path(path,'metadata.csv'))
case_labels <- c(hs='HS CFA',democracy='Political democracy')
main <- s[s$workload=='sb_wald' & s$boundary=='raw',]
ratios <- main$paired_ratio[main$accepted]
rng <- function(x,digits=1) paste(formatC(range(x),digits=digits,format='f'),collapse='–')
same <- v[v$boundary=='same_spectrum',]
same <- aggregate(value ~ case + engine, same, median)
wide <- reshape(same,idvar='case',timevar='engine',direction='wide')
wide$relative_gap <- abs(wide$value.magmaan/wide$value.lavaan-1)
worst <- wide[which.max(wide$relative_gap),]
failed <- unique(c[c$comparison=='cross_engine' & !c$passed,c('case','workload','metric')])
audit_path <- file.path('results',params$audit)
has_audit <- file.exists(file.path(audit_path,'independent.csv'))
if(has_audit) {
  audit <- read.csv(file.path(audit_path,'tolerance-sweep.csv'))
  independent <- read.csv(file.path(audit_path,'independent.csv'))
}
```

## Question

How much time does a complete SEM analysis take in magmaan and lavaan, and
how does the comparison change when model/data preparation is reused or only
inference remains?

## Short answer

On these two datasets, the accepted raw-data analyses producing SB goodness
of fit and robust 95% Wald intervals were **`r if(length(ratios)) rng(ratios) else 'not comparably'` times
faster in magmaan**. These are local measurements from
`r length(unique(t$session))` fresh R sessions, not a general speed claim.

The numerical screen also found a pEBA-4 discrepancy. For
`r unname(case_labels[worst$case])`, the p-values differ by
`r sprintf('%.1f%%',100*worst$relative_gap)` even with the same fitted statistic
and eigenvalues. Ratios for workloads failing the numerical screen are withheld.
`r if(has_audit) 'A follow-up calculation identifies the cause as numerical integration tolerance in the tiny upper tail, not a different fit or information-matrix convention. Tightening lavaan’s tolerance agrees with magmaan and an independent reference.' else 'The experiment establishes a usable timing protocol and a calibration discrepancy requiring a follow-up check.'`

## Evidence

### Reusing preparation changes the comparison

```{r}
z <- s[s$workload=='sb_wald' & s$boundary %in% c('raw','prepared','postfit'),]
z <- z[order(z$case,match(z$boundary,c('raw','prepared','postfit'))),]
knitr::kable(data.frame(Model=unname(case_labels[z$case]),
  Boundary=unname(c(raw='Raw data to report',prepared='Prepared inputs to report',postfit='Fit to inference')[z$boundary]),
  `Magmaan (ms)`=round(z$magmaan_ms,2),`Lavaan (ms)`=round(z$lavaan_ms,2),
  `Ratio (IQR)`=ifelse(z$accepted,sprintf('%.1f (%.1f–%.1f)',z$paired_ratio,z$ratio_q25,z$ratio_q75),'withheld'),
  check.names=FALSE))
```

Every row is a directly timed call returning the requested outputs. Prepared
inputs exclude model/data preparation for both packages, but still include
fitting, finalization, and inference. Fit-to-inference begins with an
estimate-only fit and includes the data work needed for robust inference.
These boundaries use different entry points; their differences are not an
additive estimate of object-construction overhead.

### Numerical agreement limits the pEBA comparison

```{r}
knitr::kable(data.frame(Model=unname(case_labels[wide$case]),
  `Magmaan p`=formatC(wide$value.magmaan,digits=6,format='g'),
  `Lavaan p`=formatC(wide$value.lavaan,digits=6,format='g'),
  `Relative Difference`=sprintf('%.2f%%',100*wide$relative_gap),check.names=FALSE))
```

This table holds the statistic and spectrum identical, so the remaining
difference arises after their construction.

```{r results='asis'}
if(has_audit) {
  cat('The follow-up audit holds those inputs fixed and tightens only the tail integration tolerances. An independent high-precision calculation gives **',
      sprintf('%.10g',independent$p_value),
      '**. Magmaan gives ',sprintf('%.10g',audit$magmaan_p[1]),
      '; lavaan at tolerance 1e-13 gives ',sprintf('%.10g',tail(audit$lavaan_p,1)),
      '. The default lavaan value is ',sprintf('%.10g',audit$lavaan_p[1]),
      ', an absolute difference of about 1.13e-8. Its default absolute integration tolerance is 1e-6.\n\n',sep='')
  cat('This is a numerical precision limitation in an extreme tail. The independent reference sums twelve exponential waiting times, which is exactly the same distribution as this four-block pEBA law; calculations at 50 and 70 decimal digits agree. The original timing ratios remain withheld because a tighter-accuracy reference route must be timed before an accuracy-matched speed claim.\n\n')
} else cat('The same-input comparison alone does not establish which implementation is correct.\n\n')
```

 The complete checks include parameter estimates,
objective values, covariance matrices, interval endpoints, SB scaling and
statistic, GOF degrees of freedom, spectra, and log p-values. The log-p check
prevents an absolute tolerance from concealing a disagreement in a small tail.
Prepared and post-fit adapters are also checked against full calls separately
within each package.

### Optimization and evaluation costs remain separate questions

```{r}
a <- d[!duplicated(d[c('case','engine')]),]
options(knitr.kable.NA='—')
knitr::kable(data.frame(Model=unname(case_labels[a$case]),Engine=a$engine,
  Iterations=ifelse(a$iterations>0,a$iterations,NA),
  `Function Calls`=a$f_evals,`Gradient Calls`=a$g_evals,check.names=FALSE))
```

A reported iteration can contain several evaluations. Missing counts remain
missing; they are not replaced with iterations. The run also retains isolated
setup timings and lavaan objective/gradient probes at the fitted point. Those
unpaired probes and lavaan's coarse internal timing fields are diagnostics,
not evidence of a pure R-versus-C++ effect. The controlled cross-engine replay
and common-optimizer comparison remain the next experimental arm.

## Caveats

This pilot covers two complete-data, single-group ML models, without a mean
structure. It requests no baseline model or extra fit indices. It uses the
same expected-information convention for both packages' intervals and SB
comparison. All timings include the requested output extraction but exclude
printing and package startup. Natural garbage collection remains inside timed
batches. Comparable arms alternate order, and processes run serially with one
numerical-library thread.

These data and model sizes do not establish scaling behavior. Timing repeats
measure runtime variability on fixed data, not statistical sampling uncertainty.
The numerical screen combines magmaan's fit verdict with lavaan convergence,
a gradient check, covariance admissibility, and output agreement; it is not a
common cross-engine stationarity metric or a global-optimum certificate.
The full covariance and tail checks can reject a workload even when its fitted
parameters agree.

Package and runner fingerprints, raw batches, options, inputs, checks, errors
(if any), and session information are retained under this report's result
folder. No historical benchmark timings enter these results. General public
claims still require broader cases, timings at matched tail accuracy, a
controlled evaluator/backend comparison, and independent-machine reproduction.

## Reproduce

From the repository root, with an optimized magmaan installation and lavaan
0.7-2:

```sh
Rscript experiments/showcases/08-speed-attribution/run_experiment.R --help
Rscript experiments/showcases/08-speed-attribution/run_experiment.R --smoke --out smoke
Rscript experiments/showcases/08-speed-attribution/run_experiment.R --out pilot
quarto render experiments/showcases/08-speed-attribution/report.qmd -P run:pilot
```

Output directories must be new. Use `R_LIBS` to select a private package
installation. The runner starts three fresh sessions by default, with seven
paired batches per workload and a 100 ms target per batch. The smoke path uses
one session and two shorter batches. Rendering reads only the selected results;
it performs no fitting or timing. `summary.csv` contains all five workloads;
`samples.csv` retains every batch and arm order, and `checks.csv` retains every
numerical check, including failures. The setup and fitted-point probes are
also in `summary.csv`, with no cross-engine ratio.

The optional tail audit uses the pilot's saved inputs and requires Python's
`mpmath` package. It neither changes installed lavaan nor rewrites timing results:

```sh
Rscript experiments/showcases/08-speed-attribution/scripts/audit_tail.R pilot tail-audit
quarto render experiments/showcases/08-speed-attribution/report.qmd -P run:pilot -P audit:tail-audit
```
