---
title: "Can efficient residual flipping calibrate SEM goodness of fit?"
subtitle: "A small Foldnes--Moss--Gronneberg-style DGP probe"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages(library(ggplot2))
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
read_result <- function(name) read.csv(
  file.path(script_dir, "results", paste0(name, ".csv")),
  stringsAsFactors = FALSE, check.names = FALSE)
required <- c("design", "replications", "method_summary", "paired_summary",
              "matched_power", "timing_summary", "dgp_timing", "failures",
              "methods", "metadata")
paths <- file.path(script_dir, "results", paste0(required, ".csv"))
if (!all(file.exists(paths))) stop("run run_experiment.R before rendering")
objects <- lapply(required, read_result); names(objects) <- required
list2env(objects, environment())
meta <- setNames(metadata$value, metadata$key)
reps <- as.integer(meta[["reps"]]); flips <- as.integer(meta[["flips"]])
is_small_probe <- reps >= 100L
primary <- c("flip_effective", "flip_standardized", "gof_std_ml", "gof_sb_ml",
             "gof_mv_ml", "gof_peba4_rls", "gof_all_rls")
pretty <- function(x) {
  x <- sub("^gof_", "", x)
  x <- gsub("_", " ", x, fixed = TRUE)
  sub("flip standardized chisq", "standardized: chi-square", x, fixed = TRUE)
}
cell_rate <- function(method, p, n, distribution) {
  x <- method_summary[method_summary$truth == "null" &
                        method_summary$method == method &
                        method_summary$p == p & method_summary$n == n &
                        method_summary$distribution == distribution,
                      "rejection_rate"]
  if (length(x) == 1L) x else NA_real_
}
```

## Result first

This probe implements a coherent GOF analogue of efficient score flipping.
For every case, the saturated covariance contribution is centred at the fitted
model and projected onto the expected-information complement of the model
tangent. The all-plus quadratic is the structured RLS GOF statistic; random
signs supply its reference distribution.

The optional standardized arm whitens those same projected contributions by
their empirical covariance. It is therefore an operation on the efficient
residual scores themselves. It is not the nested test's flip-specific nuisance
correction, because the GOF projection has already removed the nuisance tangent.

This report contains **`r reps` replications per cell and `r flips` random
flips**. `r if (is_small_probe) "The run is an exploratory probe, not a method-ranking simulation." else "This is a smoke run: rejection and power estimates are not evidence."`

## Identity and availability checks

```{r checks}
checks <- data.frame(
  check = c("replications", "failed fit/test batteries",
            "largest |flip statistic - RLS statistic|",
            "standardization available"),
  value = c(nrow(replications), nrow(failures),
            format(max(replications$flip_rls_abs_difference, na.rm = TRUE),
                   scientific = TRUE, digits = 3),
            sprintf("%d/%d", sum(replications$standardization_available,
                                 na.rm = TRUE),
                    sum(!is.na(replications$standardization_available)))))
knitr::kable(checks, row.names = FALSE)
```

```{r availability}
availability <- aggregate(
  standardization_available ~ p + n + df + n_over_df,
  replications, mean, na.rm = TRUE)
knitr::kable(availability, digits = 3, row.names = FALSE,
             col.names = c("p", "n", "residual df", "n/df",
                           "standardization availability"))
```

The unavailable corner is structural, not a software failure: at `p=20`,
residual df is 170, so an unregularized empirical covariance cannot have full
rank at `n=100`. No ridge or eigenvalue truncation is inserted. At `n=400` the
matrix is invertible, but conditioning remains part of the result rather than
being declared harmless.

## Design

```{r design}
knitr::kable(unique(design[c("p", "n", "df", "n_over_df", "rho")]),
             digits = 2, row.names = FALSE,
             col.names = c("indicators", "n", "GOF df", "n/df",
                           "omitted residual covariance under power"))
```

The null is a correctly specified one-factor covariance model. The power DGP
adds one residual covariance of .25 between the first two indicators but fits
the same CFA. Normal data are paired with the severe polynomial-logistic
construction targeting marginal skewness 3 and excess kurtosis 21. There is no
`t5` cell and no unbiased-Gamma variant.

## Null calibration retains the cells

```{r cellwise, fig.width=9, fig.height=6}
null_cells <- subset(method_summary, truth == "null" & method %in% primary)
null_cells$label <- factor(pretty(null_cells$method),
                           levels = rev(pretty(primary)))
ggplot(null_cells, aes(rejection_rate, label, colour = distribution)) +
  annotate("rect", xmin = .025, xmax = .075, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = .55) +
  geom_vline(xintercept = .05, linetype = 2, colour = "grey30") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = .16,
                 position = position_dodge(width = .45), alpha = .5) +
  geom_point(aes(shape = factor(p), size = factor(n)),
             position = position_dodge(width = .45)) +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1)) +
  scale_size_manual(values = c("100" = 2, "400" = 3), name = "n") +
  labs(x = "Cellwise null rejection probability", y = NULL,
       colour = "distribution", shape = "p",
       caption = "Every point is a design cell; Wilson intervals show Monte Carlo uncertainty. No pooled rate is used as a substitute for cell variation.") +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid.major.y = element_blank(), legend.position = "bottom")
```

```{r p-hist, fig.width=9, fig.height=5.5}
hist_rows <- do.call(rbind, lapply(names(c(
  p_flip_effective = "effective flip",
  p_flip_standardized = "standardized flip",
  p_gof_peba4_rls = "pEBA4 RLS",
  p_gof_mv_ml = "MV ML")), function(column) {
    data.frame(replications[c("p", "n", "distribution", "truth")],
               method = c(
                 p_flip_effective = "effective flip",
                 p_flip_standardized = "standardized flip",
                 p_gof_peba4_rls = "pEBA4 RLS",
                 p_gof_mv_ml = "MV ML")[[column]],
               p_value = replications[[column]])
  }))
hist_rows <- subset(hist_rows, truth == "null" & is.finite(p_value))
ggplot(hist_rows, aes(p_value, fill = distribution)) +
  geom_histogram(binwidth = .05, boundary = 0, position = "identity",
                 alpha = .45) +
  facet_grid(method ~ p + n, scales = "free_y") +
  coord_cartesian(xlim = c(0, 1)) +
  labs(x = "Null p-value", y = "count",
       caption = "Histograms expose discreteness and shape; missing standardized panels correspond to singular empirical covariance.") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom")
```

## When might standardization earn its keep?

```{r conditioning}
condition_summary <- aggregate(
  cbind(empirical_eigen_min, empirical_eigen_mean, empirical_eigen_cv,
        empirical_eigen_ratio, empirical_identity_distance) ~
    p + n + distribution + truth,
  replications, median, na.rm = TRUE)
knitr::kable(condition_summary, digits = 3, row.names = FALSE)
```

```{r paired-scatter, fig.width=8, fig.height=5}
paired_rows <- subset(replications,
  is.finite(p_flip_effective) & is.finite(p_flip_standardized))
paired_rows$delta <- abs(paired_rows$p_flip_standardized -
                           paired_rows$p_flip_effective)
ggplot(paired_rows, aes(empirical_eigen_ratio, delta,
                        colour = distribution, shape = truth)) +
  geom_point(alpha = .7) +
  scale_x_log10() +
  facet_wrap(~ p + n, scales = "free_x") +
  labs(x = "Empirical projected-meat condition number (log scale)",
       y = "|standardized p - effective p|",
       caption = "The useful question is paired movement versus a visible covariance-geometry change, not dimension alone.") +
  theme_minimal(base_size = 10.5) +
  theme(legend.position = "bottom")
```

The a priori prediction is non-monotone. Standardization should have little to
do under well-sampled normal theory, may help under nonnormal anisotropy, and
becomes unstable or undefined as residual df approaches sample rank. The probe
therefore treats availability and condition number as part of the method, not
as implementation trivia.

## What this probe says

At `p=5`, effective-flip null rejection was
`r sprintf("%.02f/%.02f", cell_rate("flip_effective", 5, 100, "normal"), cell_rate("flip_effective", 5, 400, "normal"))`
for normal `n=100/400` and
`r sprintf("%.02f/%.02f", cell_rate("flip_effective", 5, 100, "pl"), cell_rate("flip_effective", 5, 400, "pl"))`
for PL. With only 100 replications the intervals are wide, but nothing here
rules out useful low-rank calibration.

At `p=20` (df 170), the story changes. Effective rejection was
`r sprintf("%.02f/%.02f", cell_rate("flip_effective", 20, 100, "normal"), cell_rate("flip_effective", 20, 400, "normal"))`
under normality and
`r sprintf("%.02f/%.02f", cell_rate("flip_effective", 20, 100, "pl"), cell_rate("flip_effective", 20, 400, "pl"))`
under PL. Standardization was unavailable at `n=100`; at `n=400` it rejected
`r sprintf("%.02f/%.02f", cell_rate("flip_standardized", 20, 400, "normal"), cell_rate("flip_standardized", 20, 400, "pl"))`
under normal/PL, versus the effective arm's
`r sprintf("%.02f/%.02f", cell_rate("flip_effective", 20, 400, "normal"), cell_rate("flip_effective", 20, 400, "pl"))`.
The corresponding median empirical-meat condition numbers were about 37 and
42,000. That is evidence against treating raw empirical whitening as the
automatic GOF analogue of nested-test standardization.

The high-rank failure is not unique to flipping. At `p=20,n=100`, ordinary ML
and SB were badly liberal, whereas MV/SS/ALL were often extremely conservative
under PL; pEBA4-RLS sat between those extremes but still rejected .20. The
cellwise plot is the result: no pooled average should hide this geometry.

## Power

```{r power}
power_show <- subset(matched_power, method %in% primary)
power_show$method <- pretty(power_show$method)
knitr::kable(power_show[c("p", "n", "df", "distribution", "method",
                          "n_null", "n_power", "nominal_power",
                          "matched_power", "ci_lower", "ci_upper")],
             digits = 3, row.names = FALSE)
```

Matched power calibrates each method against its empirical null p-values in the
same `(p,n,distribution)` corner. With a small probe it is noisy, but it keeps
size distortion from masquerading as power. The single omitted covariance is
also deliberately sparse: dilution in the 170-dimensional residual space is a
feature of the question.

## Timing

```{r timing}
time_tab <- aggregate(median_seconds ~ p + n + phase,
                      timing_summary, median, na.rm = TRUE)
time_wide <- reshape(time_tab, idvar = c("p", "n"), timevar = "phase",
                     direction = "wide")
knitr::kable(time_wide, digits = 4, row.names = FALSE)
```

```{r generator-timing}
knitr::kable(dgp_timing, digits = 3, row.names = FALSE,
             caption = "One-time generator setup, shared across n=100 and n=400")
```

The relevant incremental cost is
`standardization_setup + flip_standardization`, not the whole model fit. Unlike
the nested flip-specific correction, this empirical whitening is computed once
and reused for all signs; its statistical conditioning is expected to become a
problem before its arithmetic cost does. Polynomial-logistic calibration is
the dominant one-time setup at `p=20` and is kept outside replication timing.
For this 100-replication run, the four-worker wall time was
`r sprintf("%.1f", as.numeric(meta[["elapsed_seconds"]]))` seconds. At
`p=20,n=400`, the median replication took about 55 ms: 5 ms for effective flip
resampling, 7 ms to factor the empirical standardizer, 2 ms to apply it over
the flips, and 33 ms for the 12-test FMG battery. Arithmetic cost is modest;
statistical instability is the binding issue.

## Decision after the probe

The algebraic bridge is sound and the low-rank cells justify another focused
look. The present gate does **not** justify a general core default: effective
flipping deteriorated when residual rank was large relative to sample size,
and raw empirical standardization was unavailable or worse. The next probe
should stay at low/moderate df, increase the null budget, and vary `n/df`
directly. If rescuing high-df cells requires a shrinkage parameter, that is a
separate method-selection problem—not a free extension of sign flipping.
