---
title: "Ordinary fitting: starts, optimizer controls and convergence audits"
params:
  lane: uls_guard
  guard_run: uls-guard-calibration-v3
  factor_run: uls-factor-square-root-v2
  factor_baseline: uls-factor-baseline-v3
  factor_replay: uls-factor-native-replay
  study_run: uls-reliability-final-v2
  study_confirm: uls-reliability-reference-90
  uls_run: uls-unit-final-60
  uls_confirm: uls-unit-final-90
  run: ordinary-layered-pilot-v2
  units_run: ordinary-layered-units-v2
  witness_run: ordinary-layered-witnesses-v2
  audit_run: ordinary-layered-exact-v3
  scaling_units: ordinary-scaling-units-v4
  scaling_fresh: ordinary-scaling-fresh-units-v2
  scaling_pilot: ordinary-scaling-pilot-v2
  scaling_witnesses: ordinary-scaling-witnesses-v2
  scaling_audit: ordinary-scaling-exact-v2
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| results: asis
if (identical(params$lane, "uls_guard")) {
  read_guard <- function(name) {
    path <- file.path("results",params$guard_run,paste0(name,".csv"))
    if (!file.exists(path)) stop("Missing guard calibration: ",path,"; run --ordinary --uls-audit-guard")
    read.csv(path,stringsAsFactors=FALSE)
  }
  v <- read_guard("verification"); sweep <- read_guard("guard_summary")
  points <- read_guard("comparisons"); timing <- read_guard("reference_metadata")
  refs <- points[points$target_distance==0,]
  flat <- refs[refs$case_id %in% c(21,22),]
  cat("## Question\n\nHow should a numerical acceptance guard account for uncertainty in the ULS Newton distance, and do the two flat Hessians require regularization?\n\n")
  cat(sprintf("## Short answer\n\nAn uncertainty interval is needed near the acceptance budget. Among %d controls on 25 development datasets, %d are close to 0.01. The distance error remains below %.2g, but it reverses %d raw budget comparisons at points nominated to lie exactly on the threshold. Conditional intervals based on independently measured errors cover every reference distance: %d controls are decisively within budget, %d above it, and %d unresolved, with no wrong decisions.\n\n",v$points,v$near_budget_controls,v$max_distance_error,v$raw_budget_classification_errors,v$measured_within,v$measured_above,v$measured_unresolved))
  cat(sprintf("The flat curvature has a promising coordinate remedy. In 90-digit objective-Jacobian QR coordinates, the two Hessian conditions become %.2f and %.2f, preserving the full observed correction and the ULS target. The largest condition across all 25 minima is %.1f. This nominates a stable curvature implementation before regularization; it does not establish a production error bound. The library's acceptance guard remains unchanged and passes zero controls.\n\n",flat$qr_curvature_condition[1],flat$qr_curvature_condition[2],v$max_reference_qr_curvature_condition))
  cat("## Evidence\n\n### Distance uncertainty\n\nThe controls comprise a rounded refined minimum and six positive x1-variance perturbations nominated for distances 0.009, 0.00999, 0.01, 0.01001, 0.011 and 0.02. The exact calculation judges the actual floating-point inputs. All 16 raw classification reversals occur among the 25 nominal 0.01 controls; none occurs in the controls on either side.\n\n")
  cat(sprintf("The largest bound using measured factor/residual construction errors and measured projection arithmetic error is %.2g. These measured errors require the independent calculation, so they cannot themselves be used by the production checker. They test the perturbation mechanism and expose the margin that a usable error model must cover.\n\n",v$max_measured_bound))
  table <- data.frame(`Assumed relative allowance`=c("8 × machine epsilon","64 × machine epsilon","8 × rows × columns × machine epsilon"),
    `Within budget`=sweep$within,`Above budget`=sweep$above,Unresolved=sweep$unresolved,
    `Input allowance misses`=sweep$input_allowance_failures,`Minima within / 25`=sweep$reference_within,check.names=FALSE)
  cat(knitr::kable(table,row.names=FALSE),sep="\n")
  cat("\n\nEvery interval in this sensitivity sweep covers its reference distance and makes no wrong decisive classification. The 8-epsilon input-error allowance nevertheless misses the construction error for all seven points of one dataset; interval coverage alone does not validate that model. The 64-epsilon allowance covers the observed construction errors but leaves 83 controls unresolved, including one reference minimum. The larger dimensional allowance is much more conservative. These are declared numerical assumptions, not candidates selected for a new default.\n\n")
  cat("The separation between backward error, conditioning and output error follows the [LAPACK error-analysis guidance](https://www.netlib.org/lapack/lug/node78.html). A small solve residual alone is insufficient when the factor is sensitive or when cancellation occurs in constructing the residual.\n\n### Curvature coordinates\n\n")
  table <- data.frame(`Weak-marker case`=flat$case_id,
    `Diagonal Hessian condition`=format(flat$reference_hessian_condition,digits=3,scientific=TRUE),
    `QR-coordinate Hessian condition`=round(flat$qr_curvature_condition,2),check.names=FALSE)
  cat(knitr::kable(table,row.names=FALSE),sep="\n")
  cat(sprintf("\n\nAcross the minima, the smallest eigenvalue in these reference coordinates is %.3f. The reparameterization normalizes the objective Jacobian; it does not make statistical weak identification disappear. A synthetic saddle retains negative observed curvature despite a positive Gauss-Newton term. A zero singular-value control yields unresolved rank. Native unidentified and saddle rejection remains covered by the existing owning C++ tests.\n\n",v$min_reference_qr_curvature_eigenvalue))
  cat("### Conditional bound and proposed implementation\n\nWrite $\\Omega=A^\\top A$, $G=A^\\top b$, and $d=\\|P_A b\\|$, where $P_A$ projects onto the factor's column space. After column scaling, suppose factor construction and QR contribute a perturbation of norm at most $\\delta_A$, residual construction contributes at most $\\delta_b$, and $\\delta_A<\\sigma_{\\min}(A)$. Then\n\n$$|d_{\\mathrm{true}}-d_{\\mathrm{computed}}|\\;\\leq\\;\\|b\\|\\frac{\\delta_A}{\\sigma_{\\min}(A)-\\delta_A}+\\delta_b,$$\n\nwith any separately measured projection arithmetic error added in the measured-error arm. The bound follows by applying the complementary projector to the perturbed full-rank factor, bounding its pseudoinverse norm by the reciprocal smallest singular value, and using the norm triangle inequality. The implementation uses Frobenius norms as upper bounds on spectral perturbations. Rank uncertainty and intervals crossing 0.01 remain unresolved.\n\n")
  cat("For curvature, retain the total objective Jacobian $J$ and the analytic observed correction $C$, with $H=J^\\top J+C$. A full-rank QR $J=QR$ yields the congruent Hessian $I+R^{-\\top}CR^{-1}$. This preserves the observed curvature and provides a coordinate change for the Newton solve. The next core change must assemble the correction separately and solve through these coordinates; applying the transform to an already rounded cross-product Hessian cannot recover lost digits. No ridge or substitution of Gauss-Newton curvature is warranted by these results.\n\n")
  cat(sprintf("## Caveats\n\nThis is a fixed-point calibration on the same 25 development datasets, with no optimizer calls or new statistical target. The actual parameter and sample inputs are reconstructed as binary64 values for the 90-digit reference. The sample's lower triangle is mirrored, matching the objective packing and factor Cholesky convention. Some model-implied covariances are indefinite under the explicitly unrestricted target. The reference calculations took about %.1f seconds on one math thread under nice 10. Small summaries and fingerprints are frozen; matrices and point inputs stay local and are regenerated from this leaf's frozen bank.\n\n",timing$elapsed_s))
  cat("The measured-error arm validates a conditional numerical mechanism, not a deployable forward-error guarantee. Construction allowances, reliable curvature computation, the flat finite NTML witness and broader fresh confirmation remain open. A numerical distance interval describes uncertainty in the local audit calculation; it is not a bound on distance to a global optimum or a guarantee of inferential accuracy. Earlier square-root and search evidence remain available with `-P lane:uls_audit` and `-P lane:uls_study`.\n\n## Reproduce\n\nUse the square-root magmaanlab build, Python with mpmath 1.3.0, single-threaded math and fresh output directories.\n\n```sh\nRscript run_experiment.R --ordinary --uls-audit-guard --help\nOMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 MKL_NUM_THREADS=1 nice -n 10 Rscript run_experiment.R --ordinary --uls-audit-guard --smoke --run-id guard-smoke-new\nOMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 MKL_NUM_THREADS=1 nice -n 10 Rscript run_experiment.R --ordinary --uls-audit-guard --run-id guard-new\npython3 scripts/verify_uls_guard.py --run-dir results/guard-new\nquarto render report.qmd -P guard_run:guard-new\n```\n\n")
  knitr::knit_exit()
} else if (identical(params$lane, "uls_audit")) {
  read_factor <- function(run,name) {
    path <- file.path("results",run,paste0(name,".csv"))
    if (!file.exists(path)) stop("Missing factor-audit evidence: ",path,"; see reproduction commands")
    read.csv(path,stringsAsFactors=FALSE)
  }
  v <- read_factor(params$factor_run,"verification")
  points <- read_factor(params$factor_run,"comparisons")
  meta <- read_factor(params$factor_run,"metadata")
  cat("## Question\n\nCan a square-root representation compute the unrestricted ULS Newton accuracy distance reliably at the mixed-unit reference minima and at deliberately inaccurate points, without regularizing the estimator?\n\n")
  cat(sprintf("## Short answer\n\nThe square-root audit computes distances for all %d points, including all %d reference minima. The former cross-product calculation supplies %d distances and only %d at the minima. Against independent 90-digit calculations, the new largest absolute distance error is %.2g, and it makes no wrong decisions about the 0.01 distance budget across %d accurate and %d inaccurate controls.\n\n",v$factor_finite_distances,v$factor_reference_distances,v$baseline_finite_distances,v$baseline_reference_distances,v$max_absolute_distance_error,v$accurate_controls,v$inaccurate_controls))
  cat("This repairs distance computation. The unchanged production condition guard still rejects all points, including every minimum. No statistical regularization, new optimizer, acceptance override or default change follows. The numerical guard follow-up is available in the `uls_guard` report lane.\n\n## Evidence\n\n")
  table <- data.frame(Method=c("Cross-product / LLT","Square-root / QR"),
    `Distances / 100`=c(v$baseline_finite_distances,v$factor_finite_distances),
    `Distances at minima / 25`=c(v$baseline_reference_distances,v$factor_reference_distances),check.names=FALSE)
  cat(knitr::kable(table,row.names=FALSE),sep="\n")
  cat(sprintf("\n\nThe largest new distance at a reference minimum is %.2g. The existing squared metric condition reaches %.2g; its cutoff remains 1e12. Two weak-marker minima also exceed the Hessian cutoff; the larger condition is near %.2g. A factor calculation cannot repair missing or unresolved positive curvature.\n\n",v$max_reference_distance,v$max_reference_factor_condition,v$max_reference_curvature_condition))
  cat(sprintf("An exploratory factor-aware rule would compare the numerical cap with the factor condition, while retaining the Hessian cap, full rank, residual guard and 0.01 distance budget. It accepts %d/25 minima and %d/75 accurate controls, with %d inaccurate acceptances here. This rule is diagnostic only: the library still applies its original guard. It needs error-bound calibration and broader confirmation before becoming an acceptance policy.\n\n",v$candidate_reference_passes,v$candidate_accurate_passes,v$candidate_false_acceptances))
  replay <- read_factor(params$factor_replay,"verification")
  paired <- read_factor(params$factor_replay,"replay_comparison")
  cat(sprintf("The repeated native grid retains all %d cold search outcomes and the signed-start sphere PORT-NLS portfolio still matches 25/25 references. The production audit passes remain zero. The new grid took about 50 seconds; these runs do not provide a balanced speed comparison. Fixed-point audit times were below one second per 100 points for both builds.\n\n",paired$fits))
  cat("### What changed numerically\n\nFor the total gradient variance, retain $\\Omega=A^\\top A$ directly from the sample covariance and the moment Jacobian. Reduce the factor through the same equality or sphere tangent coordinates. Column equilibration and pivoted QR avoid forming the ill-conditioned cross-product. With the whitened residual $b$ satisfying $G=A^\\top b$, the unconstrained distance is $\\|Q^\\top b\\|$; this also avoids amplifying cancellation error in a nearly zero gradient. Box corrections retain a triangular solve for their actual correction score. The analytic observed Hessian still determines positive curvature. Rank loss fails explicitly; no pseudoinverse or ridge is substituted.\n\n")
  cat("Owning C++ regression tests cover a rounded singular cross-product with a full-rank factor, a truly rank-deficient factor, accurate and inaccurate scores, agreement with the old calculation in well-conditioned systems, sampling-factor identities across weights/groups/means, and existing sphere saddle/unidentified controls. The study compares fixed parameter vectors; no optimizer is invoked.\n\n## Caveats\n\nThese are 25 development datasets from the preceding search study, with four points per dataset: the refined minimum and three positive x1-variance perturbations. Every point is evaluated under the original mixed-unit, unrestricted ULS target with explicit infinite bounds. Some implied covariances are indefinite; the sampling metric is built from the positive-definite sample covariance. The independent calculations evaluate the saved floating-point parameters, rather than assuming that a rounded high-precision minimum has exactly zero score.\n\n")
  cat("The distance check is a local standard-error accuracy calculation, not a global-optimum proof. Agreement on these controls does not establish a general condition policy or a statistical inference default. The unchanged curvature and metric guards remain enforced; calibration must retain inaccurate and unidentified controls and include the flat NTML witness. The earlier sample-only signed-start search results remain available in the `uls_study` report lane.\n\n## Reproduce\n\nRun with single-threaded math under `nice -n 10`. Preserve the baseline installation and use a separate matching square-root installation. Requires Python mpmath. Choose fresh run IDs.\n\n```sh\nRscript run_experiment.R --ordinary --uls-audit-factor --smoke --run-id factor-smoke\nRscript run_experiment.R --ordinary --uls-audit-factor --run-id uls-factor-baseline-v3\n# Switch the installed magmaanlab to the square-root build.\nRscript run_experiment.R --ordinary --uls-audit-factor --run-id uls-factor-square-root-v2\npython3 scripts/verify_uls_factor.py --run-dir results/uls-factor-square-root-v2 --baseline-dir results/uls-factor-baseline-v3\nquarto render report.qmd\n```\n\n")
  knitr::knit_exit()
} else if (identical(params$lane, "uls_study")) {
  read_study <- function(run, name) {
    file <- file.path("results", run, paste0(name, ".csv"))
    if (!file.exists(file)) stop("Missing ULS study evidence: ", file, "; see reproduction commands in report.qmd")
    read.csv(file, stringsAsFactors = FALSE)
  }
  portfolios <- read_study(params$study_run, "portfolio_summary")
  refs <- read_study(params$study_run, "reference_summary")
  confirm <- read_study(params$study_confirm, "reference_summary")
  core <- read_study(params$study_run, "reference_core_checks")
  transport <- read_study(params$study_run, "verification")
  failures <- read_study(params$study_run, "inconsistency_witnesses")
  meta <- read_study(params$study_run, "metadata")
  ref_meta <- read_study(params$study_run, "reference_metadata")
  cat("## Question\n\nCan sample-only signed starts and existing least-squares solvers recover the known mixed-unit ULS minima, and does the improvement survive fresh regular and weak-marker samples? Does profiling or a change of coordinates help under the same objective?\n\n")
  cat("## Short answer\n\nSphere PORT-NLS with four sample-only signed-moment starts recovers all five retained references, all ten fresh regular references and all ten fresh weak-marker references. With one shared FABIN3 start, the same solver recovers 2/5, 9/10 and 7/10 respectively. The bounded portfolio gains seven targets and loses none in these comparisons. No fitted reference is used to construct a start.\n\n")
  cat("This improves search, but does not yet give certified ordinary fits: every exact reference fails the unrestricted point audit, and none of the 2,400 fitting calls yields a native audit pass. High-precision calculations find sandwich-metric conditions above the unchanged 1e12 guard for all 25 minima. Profiling helps L-BFGS, especially in standardized coordinates, but does not outperform the signed-start sphere PORT-NLS portfolio here. A checker-domain mismatch was also exposed: point evaluation needs explicit unrestricted bounds to keep its Newton audit from being replaced by the default positive-variance preset's first-order fallback. No default or threshold changes.\n\n")
  cat("## Evidence\n\n### Search: starts and optimizer representation\n\n")
  choices <- data.frame(route = c("regression_marker", "regression_sphere", "covariance_full", "covariance_profiled", "covariance_profiled", "covariance_profiled"),
    coordinates = c(rep("mixed", 5), "standardized"), optimizer = c(rep("port-nls", 4), rep("nlopt-lbfgs", 2)),
    Method = c("Regression marker / PORT-NLS", "Regression sphere / PORT-NLS", "Covariance full / PORT-NLS", "Covariance profiled / PORT-NLS", "Covariance profiled / L-BFGS", "Covariance profiled / L-BFGS, standardized"), stringsAsFactors = FALSE)
  table <- choices["Method"]
  for (family in c("retained", "regular", "weak")) {
    role <- if (family == "retained") "retained" else "fresh"
    design <- if (family == "weak") "weak_marker" else "ernst"
    table[[switch(family, retained = "Retained / 5", regular = "Fresh regular / 10", weak = "Fresh weak / 10")]] <- vapply(seq_len(nrow(choices)), function(i) {
      p <- choices[i, ]; d <- portfolios[portfolios$role == role & portfolios$design == design & portfolios$route == p$route &
        portfolios$coordinates == p$coordinates & portfolios$optimizer == p$optimizer, ]
      paste(d$matches[d$portfolio == "fabin3"], "→", d$matches[d$portfolio == "all"])
    }, "")
  }
  cat(knitr::kable(table, row.names = FALSE), sep = "\n")
  cat("\n\nCells show reference matches from one shared FABIN3 start → the six-start portfolio (FABIN3, layered and four signed-moment candidates). The four signed candidates alone attain the same 5/5, 10/10 and 10/10 for sphere PORT-NLS. The full unprofiled L-BFGS routes match no reference under either coordinate setting, even with all six starts. Standardizing coordinates without changing the weights back to the mixed-unit target would fit a different estimator; this study restores those weights exactly. Full PORT-NLS generally gains little from that coordinate change. Profiling has a clearer effect on L-BFGS.\n\n")
  cat("The covariance parameterization expresses the same two-factor covariance model through three free latent covariance entries instead of a regression coefficient and disturbance variance. Its existing SNLLS implementation profiles those three entries and six indicator variances, leaving four nonlinear loading parameters. It runs the same ULS target; there is no new optimizer implementation. At the retained second minimum, profiled PORT-NLS can meet objective tolerance while missing the stricter covariance tolerance. The joint match requirement retains that loss.\n\n")
  cat(sprintf("Across all transported starts, the largest relative objective discrepancy is %.2g and the largest covariance discrepancy standardized by sample SDs is %.2g. Fitted endpoints are restored to the original mixed-unit regression chart and re-evaluated before comparison. %d returned profiled PORT-NLS calls fail objective consistency and are excluded from diagnostic selection; their case/start/route records remain frozen. No loss or failed call is removed from the denominators.\n\n", transport$max_start_objective_gap, transport$max_start_sigma_gap, nrow(failures)))
  cat("### References, audit domain and certification\n\n")
  table <- data.frame(Family = c("Retained regular", "Fresh regular", "Fresh weak marker"), Datasets = c(5, 10, 10), check.names = FALSE)
  ids <- list(1:5, 6:15, 16:25)
  table$`Finite local references` <- vapply(ids, function(id) sum(tolower(refs$finite_local_minimum[refs$case_id %in% id]) == "true"), 0L)
  table$`Exact-point audit passes` <- vapply(ids, function(id) sum(core$audit[core$case_id %in% id] == "passed"), 0L)
  table$`Indefinite implied covariances` <- vapply(ids, function(id) sum(refs$sigma_min_eigenvalue[refs$case_id %in% id] < 0), 0L)
  cat(knitr::kable(table, row.names = FALSE), sep = "\n")
  stopifnot(all(tolower(refs$finite_local_minimum) == "true"), all(tolower(confirm$finite_local_minimum) == "true"),
    max(abs(refs$objective - confirm$objective)) <= 1e-14, all(core$audit == "failed"))
  cat(sprintf("\n\nAll 25 references have positive exact profile and full Hessians at 90 digits. Their sandwich-metric conditions range from %.2g to %.2g; two fresh weak-marker references also have Hessian condition above 1e12. The 60-digit pilot left two fresh refinements unresolved; the higher-precision calculation resolves them, and a separate 90-digit refinement reproduces every reference. The full cold grid also reproduces its search outcomes across runs. These are finite strict local minima, with no global-optimality claim.\n\n", min(refs$metric_condition), max(refs$metric_condition)))
  cat(sprintf("Point evaluation's omitted bounds argument supplies a positive-variance preset. At %d endpoints it produces a first-order pass even though the Newton evidence fails; that domain is inappropriate for this unrestricted study. Supplying explicitly infinite bounds restores the intended Newton assessment. Native fit audits remain unchanged and none pass. The correction declares the study domain; it neither fits with positive-variance constraints nor relaxes the accuracy guard. The evaluator-domain behavior is retained as a separate witness for future contract review.\n\n", transport$implicit_bound_passes))
  cat("## What remains\n\nThe search result makes signed-start sphere PORT-NLS a candidate for a broader study, not an adopted default. The next priority is stable computation and calibration of the sandwich accuracy metric: the current search can recover these reference covariances while certification remains unavailable. Then test the candidate on new populations, ordinary measurement units and both ULS/GLS targets, recording family-specific losses and the cost of the extra starts. Investigate the profiled PORT-NLS objective inconsistencies separately.\n\n")
  cat("## Caveats and cost\n\nThis exploratory study contains five retained regular N=100 problems plus ten fresh regular and ten fresh weak-marker draws. Both fresh families were chosen before running the full comparison; promising methods were compared together on them, so this is not a prespecified default-selection gate. All observations share the fixed multiplier (.01, 100, 2, .3, 10, .1). No means, groups, equality constraints, FIML, ordinal fitting, PSD constraints or statistical barriers are added. Signed primitive and residual variances are permitted; indefinite reference covariances remain visible.\n\n")
  cat("All six starts are constructed from sample moments before fitting. The single-start baseline uses the shared FABIN3 vector; canonical sphere LS's auxiliary ML-start path is outside this comparison. The four moment candidates use the (+,+), (+,−), (−,+) and (−,−) latent-variance sign patterns, covariance-ratio loading formulas and a weighted cross-factor covariance estimate; their six indicator variances match the sample diagonal. This recipe is specific to this simple two-factor model. Diagnostic portfolio selection chooses the lowest consistent original objective even if it fails certification. A match requires objective agreement within 1e-6(1+|f|) and implied covariance agreement within 1e-5 after standardizing by sample SDs. Fresh references refine the best consistent candidate from the full cold grid. This checks local accuracy rather than recovery of globally or separately selected optima. Reference refinement happens after all cold fits and uses exact derivatives with the indicator variances profiled; every reference objective agrees with the independent C++ target within 1e-8(1+|f|).\n\n")
  fit_time <- as.numeric(meta$value[meta$key == "fit_elapsed_s"])
  cost <- portfolios[portfolios$route == "regression_sphere" & portfolios$coordinates == "mixed" & portfolios$optimizer == "port-nls" & portfolios$portfolio %in% c("fabin3", "signed"), ]
  cat(sprintf("The 2,400 cold fits took %.1f seconds locally, including domain checks and result handling; 90-digit reference refinement took %.1f seconds. The sphere four-start portfolio uses 100 optimizer calls across 25 datasets versus 25 calls for FABIN3 alone; measured fitting time was %.2f versus %.2f seconds in this run. These short serial timings are approximate, with no balanced speed claim. All work used one math thread under nice 10 on the workstation. Compact per-family summaries, portfolio checks, reference inputs/parameters/traces, failed-call witnesses and fingerprints are frozen; raw cold fits and RDSs stay local. Earlier diagnosis remains available with `-P lane:uls`.\n\n", fit_time, ref_meta$elapsed_s, sum(cost$seconds[cost$portfolio == "signed"]), sum(cost$seconds[cost$portfolio == "fabin3"])))
  cat("## Reproduce\n\nWith the recorded magmaanlab numerical baseline installed, and Python with mpmath 1.3.0 available, from this study directory:\n\n```sh\nRscript run_experiment.R --ordinary --uls-study --help\nOMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 nice -n 10 Rscript run_experiment.R --ordinary --uls-study --smoke --reference-digits 90 --run-id uls-study-smoke-new\nOMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 nice -n 10 Rscript run_experiment.R --ordinary --uls-study --reference-digits 90 --run-id uls-study-new\npython scripts/verify_uls_study.py --run-dir results/uls-study-new\nquarto render report.qmd -P study_run:uls-study-new\n```\n\n`--fresh-reps`, `--designs`, `--seed-base` and `--python` control the grid and environment. Fresh run IDs preserve attempts. The retained sample inputs are frozen in this leaf and need no local RDS files. For a separate high-precision repeat, copy the new run's `covariances.csv` and `reference_starts.csv` into a fresh result directory and run `python scripts/uls_unit_reference.py --run-dir DIRECTORY --digits 90`, then render with `-P study_confirm:NAME`. The report only reads saved results.\n")
  knitr::knit_exit()
}
if (identical(params$lane, "uls")) {
  read_uls <- function(run, name) {
    file <- file.path("results", run, paste0(name, ".csv"))
    if (!file.exists(file)) stop("Missing ULS evidence: ", file, "; see reproduction commands in report.qmd")
    read.csv(file, stringsAsFactors = FALSE)
  }
  refs <- read_uls(params$uls_run, "reference_summary")
  confirm <- read_uls(params$uls_confirm, "reference_summary")
  hits <- read_uls(params$uls_run, "reference_hits")
  summary <- read_uls(params$uls_run, "summary")
  warm <- read_uls(params$uls_run, "warm_checks")
  meta <- read_uls(params$uls_run, "metadata")
  ref_meta <- read_uls(params$uls_run, "reference_metadata")
  cat("## Question\n\nWhy does mixed-unit ULS fail in both marker and sphere charts after the unit repair that cleared NTML and GLS? Do the failures reflect search, convergence auditing, or a lack of finite minima?\n\n")
  cat("## Short answer\n\nBoth search and auditing need work. Independent 60- and 90-digit calculations establish finite strict local ULS minima for all five datasets. Warm diagnostic fits reproduce all five in both charts, but none of those ten fits passes the unchanged production audit: the normal-theory gradient-variance metric has condition about 1e16–1e18, above its 1e12 guard, even though the objective Hessian passes that guard. Failed certification therefore does not imply nonexistence.\n\n")
  cat("The existing PORT nonlinear least-squares backend reaches three of the five references from most cold configurations, while L-BFGS and scalar PORT do not match any reference to the joint objective/covariance tolerance. Two lower finite minima require diagnostic signed-variance starts; cold fits miss them. Three reference implied covariances are indefinite. These are minima of unrestricted ULS, not PSD fits or evidence of statistical admissibility. No optimizer default or audit threshold changes.\n\n")
  cat("## Evidence\n\n### Finite local references and the conditioning guard\n\n")
  table <- data.frame(Dataset = refs$case_id, Objective = signif(refs$objective, 7),
    `Hessian condition` = format(refs$hessian_condition, digits = 3, scientific = TRUE),
    `Gradient-variance condition` = format(refs$metric_condition, digits = 3, scientific = TRUE),
    `Implied covariance` = ifelse(refs$sigma_min_eigenvalue > 0, "Positive definite", "Indefinite"), check.names = FALSE)
  cat(knitr::kable(table, row.names = FALSE), sep = "\n")
  stopifnot(all(abs(refs$objective - confirm$objective) <= 1e-14 * (1 + abs(refs$objective))))
  cat("\n\nReference objectives agree in ordinary floating-point precision across the 60- and 90-digit calculations. Each reference has a vanishing score and positive exact profile and full Hessians. The Hessian and gradient-variance conditions use the same diagonal equilibration as the production audit, in the original marker regression chart. Its LS accuracy distance uses the sandwich metric, not Hessian curvature alone. The independent calculation resolves a positive but excessively ill-conditioned metric; ordinary floating-point factorization may instead report numerical singularity. Both prevent certification under the existing guard.\n\n")
  cat(sprintf("Warm PORT-NLS fits reproduce %d/%d references (objective tolerance 1e-6 times 1+|f|; covariance tolerance 1e-5 after standardizing by sample SDs), with %d production passes. Their maximum covariance gap is %.2g. These imported fitted values are diagnostic warm starts constructed after the cold grid; they never enter sample-only start selection.\n\n", sum(warm$reference_match), nrow(warm), sum(warm$accepted), max(warm$standardized_sigma_gap)))
  cat("### Cold search under a fixed ULS target\n\n")
  selected <- summary[summary$objective_scale == 1, ]
  base_hits <- hits[hits$objective_scale == 1, ]
  selected$matches <- base_hits$reference_matches[match(paste(selected$chart, selected$profile), paste(base_hits$chart, base_hits$profile))]
  profiles <- c("lbfgs", "lbfgs_more", "port", "port_more", "nls", "nls_more")
  table <- data.frame(Profile = c("L-BFGS", "L-BFGS: longer/tighter", "PORT", "PORT: longer/tighter", "PORT-NLS", "PORT-NLS: longer/tighter"), check.names = FALSE)
  for (chart in c("marker", "sphere")) {
    d <- selected[selected$chart == chart, ]; d <- d[match(profiles, d$profile), ]
    table[[paste(chart, "returned / 10")]] <- d$returned
    table[[paste(chart, "reference matches / 10")]] <- d$matches
  }
  cat(knitr::kable(table, row.names = FALSE), sep = "\n")
  cat("\n\nEach chart/profile fits five samples from two shared sample-only starts: FABIN3(auto) and layered(native). ‘Returned’ includes endpoints carried by sphere failure conditions; it is not acceptance. All 360 cold fits fail the original-target audit. PORT-NLS is the existing NL2SOL residual solver, used here as a diagnostic comparison. Longer budgets and tighter objective/step stops do not recover the two missed lower minima. In unscaled sphere ULS, shared FABIN3 also misses dataset 1; layered recovers it.\n\n")
  cat("Multiplying the identity weight by 1e-4 or 1e-8 preserves the ULS minimizers. At 1e-8 L-BFGS returns in every configuration, but still matches none of the reference covariances jointly with objective precision. PORT-NLS matches 6/10 per chart at either smaller scale. These rescalings change absolute backend stops and the sphere pin/objective ratio; every endpoint is therefore re-evaluated under the original unscaled, unpinned ULS objective. Rescaling success cannot substitute for an original-target audit.\n\n")
  cat("### Audit classification repair\n\nThe sphere audit previously labeled a singular or numerically unresolved gradient-variance metric as nonpositive curvature even after its Hessian had passed. It now reports ill conditioning, matching the ordinary LS audit. Separate curvature-system and accuracy-metric evidence, plus the reduced metric, are exposed in R. A rank-one sample regression demonstrates positive objective curvature with a singular sandwich metric. This repairs the failure label; genuine Hessian curvature failures, the .01 accuracy budget and the 1e12 condition cap remain unchanged.\n\n")
  cat("## What remains\n\n1. **Optimizer options:** test objective-aware LS preconditioning or damped Gauss–Newton steps on this fixed target, against the existing PORT-NLS comparison. Retain covariance accuracy and every failed endpoint; scalar rescaling alone is insufficient.\n2. **Starts:** investigate a bounded sample-only signed-basin alternative on the two missed minima, then fresh draws. Diagnostic reference starts are not default-start evidence.\n3. **Convergence audits:** investigate stable representation and calibration of the sandwich conditioning guard, together with the retained flat NTML witness. Do not relax a threshold merely to improve pass counts.\n\n")
  cat("## Caveats and cost\n\nThis is an exploratory five-dataset diagnosis, not a global-minimum proof or a default-selection study. The retained unit witness and four fresh regular draws use the same two-factor regression population at N=100. Indicators are multiplied by (.01, 100, 2, .3, 10, .1). ULS weights raw covariance discrepancies equally, so changing measurement units changes its statistical target; this lane fixes that mixed-unit target throughout. Means, groups, FIML, ordinal data, PSD restrictions and statistical barriers are outside this study. Signed primitive and indicator variances remain unrestricted.\n\n")
  cat("For the independent reference, six free indicator variances match the sample diagonal exactly. Four loading ratios and three latent covariance entries minimize the fifteen remaining weighted discrepancies. Exact profile curvature and the six diagonal residual directions establish strict local curvature; the full regression Hessian independently confirms it. Refinement begins from the best finite unscaled PORT-NLS candidate. If that does not qualify, three signed-moment alternatives are retained with their failed traces. This limited diagnostic search makes no exhaustive basin or global claim. The saved decimal sample covariances are symmetrized, and the resulting reference objectives agree with the C++ original objective within 1e-8(1+|f|).\n\n")
  elapsed <- as.numeric(meta$value[meta$key == "elapsed_s"])
  cat(sprintf("The 360-fit cold grid took %.1f seconds locally; the 60-digit reference calculation took %.1f seconds. A separate 90-digit run repeats the grid and reference calculation. These include result handling and are not balanced speed comparisons. Compact summaries, warm checks, reference inputs/parameters, failed refinement traces and fingerprints are frozen; full cold fits and RDS objects stay local. Earlier coordinate-repair evidence remains available with `-P lane:scaling`, and the ordinary start/control pilot with `-P lane:ordinary`.\n\n", elapsed, ref_meta$elapsed_s))
  cat("## Reproduce\n\nUse an installed magmaanlab build from this branch and Python with mpmath 1.3.0 available (for example through an isolated environment). From this study directory, first regenerate the local sample inputs with fresh run IDs:\n\n```sh\nRscript run_experiment.R --ordinary --smoke --transforms native,mixed --run-id units-new\nRscript run_experiment.R --ordinary --smoke --reps 4 --ns 100 --designs ernst --seed-base 620260004 --transforms native,mixed --run-id fresh-units-new\nRscript run_experiment.R --ordinary --uls-unit-probe --help\nRscript run_experiment.R --ordinary --uls-unit-probe --source-run units-new --fresh-run fresh-units-new --reference-digits 60 --run-id uls-new-60\nRscript run_experiment.R --ordinary --uls-unit-probe --source-run units-new --fresh-run fresh-units-new --reference-digits 90 --run-id uls-new-90\nquarto render report.qmd -P uls_run:uls-new-60 -P uls_confirm:uls-new-90\n```\n\nAdd `--smoke` to restrict the probe to the retained unit witness; `--python PATH` selects the Python environment. Frozen `covariances.csv` and `reference_starts.csv` also permit independent refinement without reconstructing R objects: copy them into a fresh local directory and run `python scripts/uls_unit_reference.py --run-dir DIRECTORY --digits 90`. Reports only read saved results.\n")
  knitr::knit_exit()
}
if (identical(params$lane, "scaling")) {
  read_scaling <- function(run, name) {
    file <- file.path("results", run, paste0(name, ".csv"))
    if (!file.exists(file)) stop("Missing scaling evidence: ", file, "; see reproduction commands in report.qmd")
    read.csv(file, stringsAsFactors = FALSE)
  }
  units <- read_scaling(params$scaling_units, "repair_comparison")
  fresh <- read_scaling(params$scaling_fresh, "unit_summary")
  fresh_checks <- read_scaling(params$scaling_fresh, "unit_checks")
  pilot <- read_scaling(params$scaling_pilot, "repair_comparison")
  witnesses <- read_scaling(params$scaling_witnesses, "repair_comparison")
  changed <- read_scaling(params$scaling_witnesses, "repair_changes")
  known <- read_scaling(params$scaling_audit, "known_target_summary")
  warm <- read_scaling(params$scaling_audit, "warm_summary")
  none <- read_scaling("ordinary-scaling-after-none", "summary")
  explicit <- read_scaling("ordinary-scaling-after-sample_units", "summary")
  cat("## Question\n\nWhy do sphere NTML and GLS fits fail after changing indicator units, even with the same layered starts? Can a small coordinate repair clear that failure while retaining the ordinary-fit audit?\n\n")
  cat("## Short answer\n\nThe retained measurement-unit failures come from optimizer coordinates. Correcting the internal latent units and applying the requested scaling to LS and PORT raises sphere NTML passes from 2/12 to 12/12 and GLS from 1/12 to 12/12. All 48 configurations per estimator also pass on four fresh regular draws. Their objectives and implied covariances agree after undoing the unit change. The starts, stopping thresholds, evaluation budgets and .01 Newton audit are unchanged.\n\n")
  cat("The repair is deliberately narrower than a reliability guarantee. Mixed-unit ULS still fails in both charts, and the hard-case bank retains basin and conditioning problems. Two shared-FABIN3 PORT ML fits gain acceptance and two lose it; layered-start target recovery stays at 3/6 qualified ML minima versus marker's 1/6. Those losses are retained, rather than hidden by pooled pass counts.\n\n")
  cat("## Evidence\n\n### Measurement units\n\n")
  table <- units[units$transform == "mixed" & units$chart == "sphere", c("estimator", "fits", "before", "after")]
  table$fresh_pairs <- vapply(table$estimator, function(est) {
    row <- fresh[fresh$estimator == est & fresh$chart == "sphere", ]
    if (nrow(row)) sprintf("%d/%d", row$matched, row$pairs) else "different target"
  }, "")
  names(table) <- c("Estimator", "Witness configurations", "Before: passes", "After: passes", "Fresh unit pairs matched")
  cat(knitr::kable(table, row.names = FALSE), sep = "\n")
  cat(sprintf("\n\nThe witness and four fresh draws are regular two-factor regression samples at N=100. Each draw is fitted under three starts and four optimizer profiles, in both charts. Indicators are multiplied by (.01, 100, 2, .3, 10, .1), with paired fits sharing the underlying observations. Maximum fresh sphere objective gap is %.2g; maximum covariance gap after undoing units and standardizing by sample SDs is %.2g. These meet the existing comparison tolerances (objective 1e-6 and standardized covariance 1e-5). Marker NTML/GLS also match all 48 fresh pairs.\n\n",
    max(fresh_checks$objective_gap[fresh_checks$chart == "sphere"]), max(fresh_checks$standardized_sigma_gap[fresh_checks$chart == "sphere"])))
  cat("Explicit `coordinate_scaling = \"sample_units\"` gives the same 12/12 witness passes. With scaling disabled, sphere falls back to 2/12 NTML and 1/12 GLS. Before the repair, requesting sample-unit scaling left those sphere failures unchanged. Thus tighter stopping and another start were insufficient to clear this particular failure. ULS weights raw covariance discrepancies equally, so heterogeneous units change its target; its 0/12 witness and 0/48 fresh mixed-unit passes in both charts remain a separate numerical problem.\n\n")
  cat("### What changed\n\nUnit-free sphere loadings already carry each indicator's sample SD through the loading map. The gauge latents therefore use dimensionless internal coordinates. Inferring their units from the gauge-free partable instead assigned the mean indicator SD to those latents, producing inappropriate scales for latent variances and regressions. The rest-coordinate scales now use the sphere chart's units; passthrough latents keep their ordinary units. LS and PORT now apply the existing requested scaling, which those paths previously skipped. Beta directions and the artificial radial pin retain their existing scales. The unscaled endpoint is still audited with the pin excluded.\n\n")
  cat("This changes search paths, including where PORT stops, but introduces no start selector, retry, new backend or statistical-domain restriction. Both nonzero scaling settings use the existing sample-unit rule here; this repair does not introduce an information-based tangent preconditioner. Nonlinear constraints and unsupported bound configurations retain their previous scaling limitations. PSD and FIML routes are outside this repair.\n\n")
  cat("### Pilot and hard-case tradeoffs\n\n")
  cat(sprintf("The 864-fit ordinary pilot gains %d accepted sphere ULS fit and loses %d accepted fits; ML and GLS acceptance counts stay unchanged. In the 648-fit hard-case replay, sphere ULS gains %d fit. Sphere ML gains %d and loses %d configurations, all shared-FABIN3 PORT or tight-PORT arms. Every ordinary marker result reproduces.\n\n", sum(pilot$gains), sum(pilot$losses), sum(witnesses$gains[witnesses$estimator == "ULS"]), sum(witnesses$gains[witnesses$estimator == "ML"]), sum(witnesses$losses[witnesses$estimator == "ML"])))
  losses <- changed[changed$loss, c("design", "n", "rep", "profile", "before_objective", "after_objective", "after_newton_status")]
  names(losses) <- c("Population", "N", "Draw", "Profile", "Before objective", "After objective", "After Newton status")
  cat(knitr::kable(losses, row.names = FALSE, digits = 6), sep = "\n")
  cat("\n\nThose two losses reach a lower objective but fail the curvature audit. They are numerical failures; a lower objective cannot substitute for a qualified endpoint. The two ML gains occur near the flat weak-marker finite witness: nearby endpoints now pass, while its independently refined minimum still exceeds the fixed conditioning guard. This sensitivity remains unresolved. Changes are recorded in `repair_changes.csv`; the original hard-case conclusions are not replaced by a universal success claim.\n\n")
  cat("Exact-point checks preserve the same six passing refined ML minima and one ill-conditioned finite minimum. Warm diagnostic fits still recover all six passing targets under both charts and both backends. Under default L-BFGS, shared-layered sphere cold fits still recover 3/6 qualified targets versus marker's 1/6; canonical sphere recovers 2/6. Warm LS comparisons retain their estimator-specific targets. Earlier start/control evidence remains available by rendering with `-P lane:ordinary`.\n\n")
  cat("## What remains\n\n1. **Starts:** explain the retained finite-basin misses, including the shared-FABIN3 worse GLS local minimum; confirm any replacement on fresh problems.\n2. **Optimizer options:** investigate mixed-unit ULS under its own fixed target and the changed PORT failures; keep search and stopping evidence separate.\n3. **Audits:** calibrate the conditioning guard on the independently established flat finite witness; validate requested-chart reporting separately. A failed guard does not establish nonexistence.\n\n")
  cat("## Limits and cost\n\nFive regular unit-change datasets and a small development bank establish a reproducible defect and repair, not broad reliability or global optimality. This covariance-only, single-group scope excludes means, cross-group/equality constraints, FIML, ordinal data, PSD and statistical barriers. The repair has an independent C++ regression for ML/GLS under L-BFGS/PORT, comparing audited endpoints, objectives and covariances. The focused C++ and R sphere suites pass. Runs took seconds locally; concurrent execution and result handling prevent a balanced speed claim. Compact summaries, paired unit checks, changed outcomes and package/source fingerprints are frozen; raw fit CSVs and RDSs stay local.\n\n")
  cat("## Reproduce\n\nThe before rows use `bf150875`. Rebuild that revision and regenerate raw baseline grids with fresh run IDs before rebuilding this repair to use `--compare-run`; frozen summaries alone do not contain the raw endpoints. With an installed repaired magmaanlab build, from this study directory:\n\n```sh\nRscript run_experiment.R --ordinary --help\nRscript run_experiment.R --ordinary --smoke --transforms native,mixed --run-id units-new\nRscript run_experiment.R --ordinary --smoke --reps 4 --ns 100 --designs ernst --seed-base 620260004 --transforms native,mixed --run-id fresh-units-new\nRscript run_experiment.R --ordinary --smoke --transforms native,mixed --coordinate-scaling none --run-id units-none-new\nRscript run_experiment.R --ordinary --pilot --run-id pilot-new\nRscript run_experiment.R --ordinary --witnesses --run-id witnesses-new\nRscript run_experiment.R --ordinary --witness-audit --cold-run witnesses-new --run-id exact-new\nquarto render report.qmd\n```\n\nUse the identical grid and `--compare-run BEFORE_RAW_RUN` to regenerate paired gain/loss CSVs. Render the retained before-repair start study with `-P lane:ordinary`; render the older reference investigation with `-P lane:reference -P run:reference-pilot`. Rendering only reads this leaf's results.\n")
  knitr::knit_exit()
}
if (identical(params$lane, "ordinary")) {
  read_ordinary <- function(name, run = params$run) {
    path <- file.path("results", run, paste0(name, ".csv"))
    if (!file.exists(path)) stop("Missing ordinary results: Rscript run_experiment.R --ordinary --pilot --run-id ", run)
    read.csv(path, stringsAsFactors = FALSE)
  }
  s <- read_ordinary("summary"); a <- read_ordinary("audit_summary")
  refs <- read_ordinary("references"); changes <- read_ordinary("changes")
  units <- read_ordinary("summary", params$units_run)
  witness <- read_ordinary("summary", params$witness_run)
  points <- read_ordinary("exact_points", params$audit_run)
  targets <- read_ordinary("known_target_summary", params$audit_run)
  warm <- read_ordinary("warm_summary", params$audit_run)
  metadata <- read_ordinary("metadata")
  meta <- function(key) metadata$value[metadata$key == key][1]
  cat("## Question\n\nCan a small complete-data NTML, ULS and GLS programme separate failures caused by starts, optimizer stopping, and endpoint accuracy, while comparing marker and sphere coordinates?\n\n")
  cat(sprintf("## Short answer\n\nShared layered starts improve sphere ULS and GLS by one accepted fit out of twelve each, with NTML unchanged, in the %d-fit pilot. The ordinary NTML/GLS layered defaults are reproduced by the shared vectors to rounding error. The original 576 fits reproduce after adding the layered arms; the earlier result was not a FABIN3-only comparison.\n\n", sum(s$fits)))
  cat("On the nine retained hard datasets, sphere with shared layered starts recovers three of the six qualified historical ML minima under default L-BFGS, versus one for marker with the identical starts. Canonical sphere recovers two. Starting directly from the refined points recovers all six under both charts and both backends. Thus the earlier sphere advantage remains visible, together with a search/accuracy gap; this pilot does not reverse it.\n\n")
  cat("Six of the seven independently refined finite minima pass today's same-point audit. The very flat seventh remains rejected by the conditioning guard. Shared layered starts do not repair the separate heterogeneous-unit sphere failures. No default or audit threshold changes follow from these development runs.\n\n")
  cat("## Evidence\n\n### Starts and optimizer controls\n\n")
  d <- s[s$profile == "default" & s$start_id %in% c("native", "shared_layered"), ]
  compact <- aggregate(list(Passes = d$audited), d[c("estimator", "chart", "start_id")], sum)
  compact$Configuration <- paste(compact$chart, compact$start_id, sep = ":")
  compact <- reshape(compact[c("estimator", "Configuration", "Passes")], idvar = "estimator", timevar = "Configuration", direction = "wide")
  compact <- compact[c("estimator", "Passes.marker:native", "Passes.marker:shared_layered", "Passes.sphere:native", "Passes.sphere:shared_layered")]
  names(compact) <- c("Estimator", "Marker current default", "Marker shared layered", "Sphere canonical", "Sphere shared layered")
  cat(knitr::kable(compact, row.names = FALSE), sep = "\n")
  cat("\n\nCounts are out of twelve per estimator, with default L-BFGS. The two sphere gains are GLS and ULS on Ernst N=20 draw 2; no passing canonical sphere fit is lost by layered starts in this pilot. Family-specific counts are retained in `summary.csv`; each family has only four draws (two at each N). Every gain and loss against the no-options fit in the same chart is retained in `changes.csv`.\n\n")
  cat("The initial optimizer result reproduces: with native starts, tight L-BFGS changes marker/sphere passes from 22/23 to 23/22 out of 36; PORT gives 25/21. Supplying shared layered starts gives 25 sphere passes under either backend, though one PORT endpoint is above the best observed objective. These pooled counts describe the development bank and do not choose a default.\n\n")
  cat("On Ernst N=20 draw 1, the shared FABIN3 start reaches a GLS endpoint about 0.0176 above the best observed half discrepancy, across both charts and both optimizers. Its Newton distance passes comfortably. Thus the local audit and basin recovery must remain separate. Another tightened marker GLS endpoint passes the .01 local budget while remaining about 2.44e-6 above the best observed objective; an audit budget is an accuracy convention, not an objective equality test.\n\n")
  cat("### What the audit catches\n\n")
  audits <- aggregate(list(Fits = a$fits), a[c("chart", "newton_status", "accepted")], sum)
  names(audits)[1:3] <- c("Chart", "Newton status", "Accepted")
  cat(knitr::kable(audits, row.names = FALSE), sep = "\n")
  cat("\n\nNonpositive curvature and ill conditioning stay visible; a solver's convergence flag cannot substitute for the audit. No acceptance target was relaxed. Near-chart failures with failed native audits are not called chart-singular optima. Backend stops, including soft exits whose iterates are discarded before the R audit, are counted in `stop_summary.csv` and preserved in local raw records. Alternative-chart verdicts are recorded at the same endpoints without refitting. In the original default sphere arm they agree on all 23 passes and 13 failures.\n\n")
  cat("### Retained witnesses and known local minima\n\n")
  known <- targets[targets$profile == "default", c("chart", "start_id", "target_available", "known_local_hit")]
  names(known) <- c("Chart", "Start", "Qualified ML targets", "Targets recovered")
  cat(knitr::kable(known, row.names = FALSE), sep = "\n")
  cat("\n\nThe cold replay fits all nine retained datasets with NTML, ULS and GLS: 648 fits over the same start/control grid. The table judges ML against independently refined finite local minima, rather than the lowest result of the cold portfolio. Shared layered sphere recovers the fresh Ernst N=20 witness, while canonical sphere remains just above the accuracy budget. All cold starts still miss several qualified targets; warm diagnostic fits from the six passing refined witnesses recover their ML objectives in both charts under both backends. These imported estimates never enter sample-only start selection.\n\n")
  cat("The direct no-optimization check reproduces all nine saved objectives. Six saved points pass; the other three fail the accuracy budget, conditioning guard, or curvature check. Six of seven refined finite minima pass in both strongest and original marker charts. The seventh is a known finite but exceptionally flat minimum with condition above 1e12. A nearby cold endpoint can pass the guard and .01 budget while the refined minimum fails the guard; this is a conditioning-policy calibration issue to investigate, not evidence of nonexistence. The saved 184,000-loading case still passes endpoint accuracy but fails the explicit study chart tolerance 1e-4; the library tolerance 1e-6 admits it. Chart reporting and endpoint accuracy remain separate.\n\n")
  cat("Warm LS fits use the ML witnesses only as starting points, then minimize their own objective. Across seven such starts, ULS passes in 3/7 marker fits versus 4/7 sphere fits under either backend; GLS passes in 4/7 versus 3/7 with L-BFGS and 3/7 versus 4/7 with PORT. Every passing warm LS endpoint matches the best locally accepted cold/warm objective for that dataset and estimator. This supplies useful local references but does not establish existence or global optimality for the remaining LS cases. `ls_warm_checks.csv` preserves those outcomes.\n\n")
  cat("### Measurement-unit smoke\n\n")
  u <- aggregate(units[c("fits", "audited")], units[c("transform", "estimator", "chart")], sum)
  names(u) <- c("Units", "Estimator", "Chart", "Fits", "Local audit passes")
  cat(knitr::kable(u, row.names = FALSE), sep = "\n")
  cat("\n\nThe two unit settings reuse one regular N=100 draw, multiplied indicator-wise by (.01, 100, 2, .3, 10, .1). NTML and GLS retain their estimator under this change. ULS weights raw covariance discrepancies equally, so its target changes; its counts are a numerical stress check, not a same-target invariance claim. All twelve marker NTML/GLS configurations pass in mixed units; sphere passes only 2/12 NTML and 1/12 GLS. Every shared-layered sphere configuration fails in mixed units. The failures include line-search errors and inaccurate stops; changing the start alone does not fix this witness. This does not by itself identify a derivative defect.\n\n")
  cat("## Minimal programme\n\n")
  cat("1. **Starts:** retain the matched native/FABIN3/layered comparisons across NTML, ULS and GLS. Explain the demonstrated sphere gains and retained basin losses before confirming any replacement on fresh draws. Do not choose a start by fitting the reference first.\n")
  cat("2. **Optimizer options:** compare stock L-BFGS and PORT with tighter relative objective/step stops. Evaluation budgets, L-BFGS memory and internal gradient tolerance remain unchanged. Next isolate sphere coordinate/preconditioning effects on the retained unit witness before adding more backends or random starts.\n")
  cat("3. **Convergence audits:** keep objective consistency, feasibility, positive curvature, conditioning and the .01 Newton budget fixed during search comparisons. Investigate conditioning-guard sensitivity on the very flat finite witness and validate chart reporting separately. A failed guard is a limitation of the numerical accuracy statement, not a nonexistence result.\n\n")
  cat("A demonstrated deterministic defect becomes a regression test and a small repair. A candidate default goes to the existing optimizer decision study with fresh held-out draws, family-specific results, every baseline loss and a cost comparison. Completing this programme does not require joint sphere/PSD boundary geometry or barrier inference.\n\n")
  cat("## Caveats\n\nThis is a 12-dataset exploratory development pilot, not confirming evidence. It covers one covariance-only, single-group two-factor regression model under three populations, at N=20 and 100. No means, cross-group/equality constraints, FIML, ordinal data, PSD or statistical barriers are fitted. Signed primitive variances remain permitted. The sphere's artificial radial pin is excluded from its endpoint audit and reference objective.\n\n")
  cat("All arms consume exactly the same sample covariance with divisor N. ULS uses identity weight; GLS freezes its normal-theory weight from that sample. Comparisons never pool objectives across estimators. Ordinary route-native ML/GLS starts are layered and ULS uses native FABIN3; sphere canonical LS currently includes an auxiliary ML fit. Shared layered(native) and FABIN3(auto-transported) vectors are constructed once per sample and supplied explicitly to each chart and estimator. The shared layered vector agrees with actual ordinary ML/GLS inputs on all twelve pilot draws to less than 1e-13 absolute difference. Sphere polishing is disabled to expose the endpoint being studied.\n\n")
  cat("Newton distances use observed information for NTML and the normal-theory sandwich metric for LS. First-order residuals are telemetry: marker and sphere residuals use different declared metrics and are not compared by a common raw-gradient cutoff. A repeated reference requires passing library audits on both backends, original-objective agreement within 1e-6(1+|f|), and standardized implied-covariance agreement within 1e-5. These exploratory matching tolerances do not alter the library's convergence criterion. A missing accepted reference is not evidence of nonattainment.\n\n")
  cat(sprintf("The pilot took %.1f seconds in this local install, including fitting and result handling. This is not a balanced speed benchmark. Raw fit/start/pair CSVs and endpoint/sample RDSs remain local; compact summaries, references, changed outcomes and source/package fingerprints are frozen for review. Sphere evaluation counters are not exposed and stay missing rather than being inferred.\n\n", as.numeric(meta("elapsed_s"))))
  cat("## Reproduce\n\nUse an installed magmaanlab build containing the sphere-native audit; metadata records its path, build stamp and binary/R-code hashes. From this study directory:\n\n```sh\nRscript run_experiment.R --ordinary --help\nRscript run_experiment.R --ordinary --smoke --run-id ordinary-smoke-new\nRscript run_experiment.R --ordinary --pilot --run-id ordinary-pilot-new\nRscript run_experiment.R --ordinary --witnesses --run-id ordinary-witnesses-new\nRscript run_experiment.R --ordinary --witness-audit --cold-run ordinary-witnesses-new --run-id ordinary-exact-new\nRscript run_experiment.R --ordinary --smoke --transforms native,mixed --run-id ordinary-units-new\nquarto render report.qmd -P run:ordinary-pilot-new -P units_run:ordinary-units-new -P witness_run:ordinary-witnesses-new -P audit_run:ordinary-exact-new\n```\n\nFresh run IDs preserve failed and completed attempts. Seeds are stable under subset/reordering and paired unit transforms; smoke and pilot use different seed bases. The runner writes checkpoints per draw and prints progress every twelve fits. The retained witness seeds and fitted-point inputs stay frozen. Use `--shared-starts fabin3` to reproduce the first 576-fit grid. For the retained earlier ML/PSD reference investigation, render with `-P lane:reference -P run:reference-pilot`. Its targets and historical screens are separate.\n")
  knitr::knit_exit()
}
read_result <- function(name) {
  path <- file.path("results", params$run, paste0(name, ".csv"))
  if (!file.exists(path)) stop("Missing results: run Rscript run_experiment.R --pilot")
  read.csv(path, stringsAsFactors = FALSE)
}
s <- read_result("summary"); refs <- read_result("references"); metadata <- read_result("metadata")
probe_path <- file.path("results", params$run, "witness_probe.csv")
probe <- if (file.exists(probe_path)) read.csv(probe_path) else NULL
count <- function(domain, label) sum(refs$domain == domain & refs$reference_label == label)
tab <- function(x) knitr::kable(x, row.names = FALSE, digits = 3)
```

## Question

Can a small collection of sphere fits provide useful reference solutions for
ML and PSD-ML, while keeping local numerical accuracy, chart problems and
uncertainty about the best solution separate? Can those references explain
ordinary-route failures without changing the estimator used in a paper?

## Short answer

**Latest rerun:** shared normalization does not improve target recovery on the
120 retained challenge draws. ML and direct PSD recover the same targets;
two-stage PSD loses three in the later batch. The [executive summary](executive-summary.html)
contains the compact comparison; see [normalization revisit](#normalization-revisit)
for controls and reproduction. Earlier reference-building results follow.

Of the nine inspected spherical endpoints, two are near-pole failures in the
requested marker chart. The other seven now have numerical evidence of finite
local minima in the original marker identification. Five refine directly to
stable points; one requires the better starting point found earlier; the last
has an exceptionally flat but finite minimum, confirmed at 60 and 90 decimal
digits. Six refined points pass the library's ordinary-precision accuracy check.
The very flat case still fails that check and should remain a numerical failure,
not be called nonattainment. None of this proves global optimality.

All seven finite ML witnesses have negative primitive variance components.
That is separate from chart availability and numerical accuracy, and excludes
them as PSD-ML solutions. In the separate retained fresh PSD batch, FABIN3-auto
plus positive spectral starts matches 51/60 sphere references under either
scaling choice. The remaining search gap and the requested-chart rejection gate
still block promotion. Marker changes in this study are diagnostic only;
automatic substitution of the user's markers is not the adopted policy.

A retrospective check of three sample-only, single-fit start selectors finds no
clear replacement for the current starts. The strongest signed spectral rule
adds two reference matches on the later 60 draws in ordinary ML with either
backend, but loses two previously successful fits and brings no net gain on the
earlier 60. Selecting by the lowest initial discrepancy performs worse. All
three rules miss the finite negative-X witness above. Keep the current starts;
this small check does not establish that they are optimal.

```{r results='asis'}
if (!is.null(probe)) {
  p <- probe[1, ]
  cat(sprintf(paste0("The targeted follow-up supplies a concrete counterexample to calling failed multistarts ",
    "\"no estimate\": on %s draw %d at N=%d, lavaan finds a screened objective of %.6f. ",
    "%d sphere backend(s) recover a screened candidate when started there. ",
    "The original finite portfolio missed a point the sphere route can represent.\n"),
    p$design, p$rep, p$n, p$witness_objective, sum(probe$label == "screened_candidate")))
}
```

## Evidence

### Can a single sample-based start do better?

The layered measurement construction initially keeps the latent covariance
positive. That suggests a general possible improvement for unrestricted ML:
let the observed correlations choose a negative variance direction when they
support one. This is inapplicable to PSD-ML, whose domain excludes those points.

We checked three deterministic rules on all 120 retained draws. The first uses,
for each factor, the eigenvector of its standardized off-diagonal correlation
matrix with the largest absolute eigenvalue, retaining that eigenvalue's sign.
It constructs one spectral start with strength 0.5. The second keeps the route's
constructor start whenever both selected signs are positive, and otherwise uses
the same signed spectral start. The third evaluates the initial ML discrepancy
of the four existing spectral vectors and chooses the smallest. It evaluates
starting points, not fitted endpoints. Each rule requires only one optimization;
none changes the requested markers or uses a sphere solution to initialize an
ordinary fit.

Selections were computed from regenerated sample moments before reading any
fit outcomes. We then looked up the already saved fit for the chosen start,
so no optimization was rerun. Both batches have already been inspected: these
are retrospective engineering checks, not fresh confirmation data. The ordinary
comparison is layered/native with information scaling; the sphere comparison
is canonical with diagonal scaling and polishing disabled.

```{r}
selection <- read.csv(file.path("results", "single-start-selection", "summary.csv"))
a <- selection[selection$batch == "fresh", ]
a <- a[order(a$route, a$backend, a$policy), ]
tab(data.frame(Route=a$route, Backend=a$backend, Rule=a$policy,
  `Constructor matches`=a$baseline_hit, `Selected matches`=a$selected_hit,
  Gains=a$gain, Losses=a$loss, check.names=FALSE))
```

These counts use the frozen reference-hit screen, including its historical
magnitude cutoff; they measure continuity with the earlier experiment, not a
new convergence verdict. Of the later 60 draws, 55 have a reference under that
screen. Library convergence verdicts and every gain/loss are retained separately
in the paired results, along with summaries by population.

The strongest signed rule gives a small ordinary-route gain on the later batch,
but exchanges four gains for two losses with each backend. On the earlier batch
it exchanges one gain for one loss. For sphere it loses three net matches on the
earlier batch and gains two on the later one. Keeping the constructor when both
signs are positive does not remove those tradeoffs. Lowest initial discrepancy
loses matches on both batches and both routes: a better initial objective does
not reliably identify a better destination.

Most tellingly, all three rules choose a negative-Y start for the inspected
fresh weak-marker N=20 draw 1. None reaches its better finite negative-X minimum.
The local correlation signal misses the sign needed by the jointly fitted model.
Thus negative directions matter, but this check does not supply a reliable
sample-only rule for choosing them. A model-wide moment construction retaining
signed latent covariances is a possible future investigation; it would need to
handle structural paths and observed-covariance positivity together and is not
a small clipping change. We do not promote a new start rule on this evidence.
The pole rejection policy and the very flat finite minimum remain separate
issues that a starting-value change need not cure.

### What became of the other seven?

We checked the seven endpoints that pass the study's marker-proximity test
using an independent high-precision calculation of the same unrestricted
Gaussian ML objective. The original markers x1 and y1 stay fixed at 1. Both
the saved sphere endpoint and the best locally checked point found earlier
were used as starting points. The calculation retains the starting sign of
X's variance while refining, so a trajectory is not silently carried through
a singular regression representation. It is a local diagnostic, not a new
production fitter or an exhaustive search.

The derivative formulas agree with independent numerical differentiation at
high precision. Every retained finite witness has a positive definite observed
covariance, a vanishing gradient to the stated numerical precision, and positive
curvature in every free parameter direction. These are strong numerical local
minimum checks, not exact-arithmetic or global-optimality proofs. The resulting
points reproduce the library's objective in the original marker chart to
less than 1e-14 and all pass the explicit 1e-4 chart-proximity check.

```{r}
finite <- read.csv(file.path("results", "open-case-conclusions", "finite_witnesses.csv"))
accuracy <- read.csv(file.path("results", "open-case-conclusions", "library_crosscheck.csv"))
finite <- merge(finite, accuracy[c("case_id","accuracy_status","accuracy_passed")], by="case_id", sort=FALSE)
tab(data.frame(Batch=finite$batch, Design=finite$design, N=finite$n, Draw=finite$rep,
  Objective=finite$objective, `Largest loading`=finite$max_loading,
  `Ordinary-precision accuracy`=ifelse(finite$accuracy_passed,"Pass","Ill-conditioned"),check.names=FALSE))
```

Five saved spherical endpoints refine directly to the same finite minimum as
the earlier checked point. Their large or negative variance components do not
by themselves indicate a runaway path. Two cases need a more specific explanation:

**Fresh weak-marker N=20 draw 1: the sphere had followed a losing path.**
Sign-preserving continuation from that endpoint approaches X variance zero,
while the regression coefficient and compensating disturbance grow without
bound numerically. The continuation reaches X variance about 2e-32 and a
regression magnitude above 5e30 before stopping; its objective is about
0.2920422 and its gradient has not vanished. The previously found negative-X-
variance point instead refines to a finite local minimum at 0.2917589, with
regression about 6.33 in the requested marker chart. Both sphere backends
recover it when started from that witness. This identifies a missed basin,
not nonattainment of the entire problem. It does not establish what other
paths or minima exist.

**Development weak-marker N=100 draw 2: a very flat finite minimum.**
Simply increasing ordinary refinement effort made little progress. Holding
the y3 loading fixed and optimizing the remaining parameters shows the objective
first decreasing and then increasing. Solving for the bottom of that local
profile gives loading 740.4555345 and objective 0.0246899172027636. A subsequent
unrestricted gradient/curvature check confirms a finite numerical local minimum;
60- and 90-digit calculations agree. This is evidence against interpreting the
investigated branch as an endlessly improving large-loading sequence.

```{r}
profile <- read.csv(file.path("results", "delicate-loading-profile", "summary.csv"))
minimum <- read.csv(file.path("results", "delicate-finite-minimum", "summary.csv"))
plot(profile$fixed_y3_loading, (profile$objective-minimum$objective)*1e9,
  log="xy", type="b", pch=19, xlab="Fixed y3 loading (y1 marker = 1)",
  ylab="Objective above local minimum (× 10⁹)")
abline(v=minimum$max_loading, lty=2, col="steelblue")
```

The ordinary-precision library still reports this point as ill-conditioned,
with an accuracy-condition measure around 5e12, even when supplied the refined
point. It should remain rejected or unresolved on numerical accuracy grounds.
Finding the minimum at high precision does not make routine parameter estimates
or inference reliable. The other six refined witnesses pass that accuracy check.
All seven are improper ML points; none may be substituted for a PSD fit.

### Requested-chart failure: the 184,000-loading example

**Decision:** reject the fresh weak-marker N=100 draw 4 endpoint in the requested
identification. Do not automatically switch its marker. Report that the fitted
point is too close to a pole of the requested marker chart to return as an
accepted fit in that chart. The alternative-marker point and implied covariance
may be retained for diagnosis, and the user can explicitly choose a different
identification.

The saved normalized chart level is 5.44e-6. The existing translation tolerance
1e-6 admits it; the already-reported study threshold 1e-4 rejects it. The new
required-failure check confirms this using the saved point without optimization
or marker changes. The 1e-4 value is an explicit study check, not a newly adopted
library-wide tolerance. A general near-pole acceptance criterion still needs
validation before promotion.

Of the other eight saved endpoints, development weak-marker N=20 draw 4 also
fails that proximity check (level 1.15e-6) and already fails sphere stationarity.
The other seven pass it; their negative variance components and, in two cases,
failed/unresolved accuracy are separate issues. Passing this proximity check
does not establish that a solution is regular, proper or globally optimal.
We should therefore not call all nine “poles,” nor claim that the seven are
proved free of every identification problem.

Suggested failure explanation for the named case: “The requested marker is
extremely weak relative to the other indicators at the fitted point. Fixing its
loading to 1 produces extremely large parameter estimates. The fit is rejected
in this identification; no marker was changed automatically.”

### Looking at the same spherical point in familiar coordinates

This inspection does not fit anything again. It takes each case's lowest-objective
evaluable saved sphere endpoint and translates it to the original marker, a
better-loaded marker, and ordinary std.lv. Successful translations preserve the
implied covariance and objective to numerical precision. “No” for std.lv means
that this same point cannot be expressed there; it is not a new failed fit.

```{r}
translations <- read.csv(file.path("results", "sphere-translations-inspected", "summary.csv"))
first <- translations[translations$chart == "first_marker", ]
strong <- translations[translations$chart == "strongest_marker", ]
std <- translations[translations$chart == "std_lv", ]
stopifnot(identical(first$case_id, strong$case_id), identical(first$case_id, std$case_id))
tab(data.frame(Batch=first$batch, Design=first$design, N=first$n, Draw=first$rep,
  `Largest loading, first marker`=first$max_abs_loading,
  `Largest loading, better marker`=strong$max_abs_loading,
  `Negative variance rows`=strong$negative_variances,
  `std.lv possible`=ifelse(std$translated,"Yes","No"),
  `Local checks passed`=ifelse(first$source_local_checked,"Yes","No"),check.names=FALSE))
```

Here `X` denotes the exogenous latent variance, `Y` the disturbance variance in
`Y ~ X`, and lowercase indicator names denote their residual variances. In this
regression model, std.lv fixes X's variance and Y's disturbance variance to +1.
A real rescaling cannot change a negative variance into a positive one. Negative
indicator residuals, by contrast, do not themselves prevent std.lv translation.

Three concrete examples explain most of what to look for:

- **High-R2, development N=20 draw 4:** with better markers, loadings are at
  most 1 and the regression is about 0.53. Y's disturbance variance is about
  -0.377, almost cancelling the explained variance and leaving total variance
  about 0.0021. The point is locally checked but improper; enormous standardized
  ratios do not mean enormous unstandardized parameters here.
- **Weak marker, fresh N=100 draw 4:** the largest loading falls from about
  184,000 to 1 after changing markers. That dramatic loading was a coordinate
  issue. The negative Y disturbance (about -27.7 under the better marker) and
  large residual compensation remain. Translation clarifies the point but does
  not make it an admissible covariance model.
- **Weak marker, development N=100 draw 2:** std.lv translation works, with a
  largest loading about 12.6, but x3 and y3 have negative residual variances.
  The largest absolute indicator residual variance is about 168 times that
  indicator's sample variance. This endpoint also has unresolved local accuracy,
  so it warrants further numerical inspection.

All nine saved endpoints have negative primitive variance components. This is
allowed by the unrestricted observed-covariance ML objective, but excluded by
PSD-ML. Six pass the local accuracy checks; the other three are retained as
endpoints to inspect, not accepted estimates. Translation alone does not tell
us whether a path is runaway or whether the endpoint is globally optimal.

```{r results='asis'}
pt <- read.csv(file.path("results", "sphere-translations-inspected", "parameters.csv"))
for (i in first$case_id) {
  x <- first[first$case_id == i, ]
  cat(sprintf("\n<details><summary>%s, %s, N=%d, draw %d: parameter tables</summary>\n\n",
    x$batch, x$design, x$n, x$rep))
  z <- pt[pt$case_id == i, ]
  for (chart in c("first_marker", "strongest_marker", "std_lv")) {
    q <- z[z$chart == chart, c("lhs", "op", "rhs", "est")]
    cat(sprintf("\n**%s**\n\n", gsub("_", " ", chart)))
    if (nrow(q)) cat(as.character(knitr::kable(q, digits=6, row.names=FALSE)), sep="\n") else
      cat("This point cannot be translated to positive std.lv.\n")
    cat("\n\n")
  }
  cat("</details>\n\n")
}
```

### Inspecting the unresolved draws

The inspection selects all four development and five retained fresh ML problems
with no screened sphere reference. It fits nine combinations of marker
indicators, effect coding, positive std.lv and the sphere, with L-BFGS and PORT.
The four spectral starting points are transported across charts; their implied
covariances are checked for agreement before fitting. Native constructors are
separate arms. Positive std.lv cannot represent negative latent disturbance
variances, so unavailable starts are recorded rather than silently changed.
Failure in that sign-restricted chart is not failure of unrestricted ML.

For each optimizer, the lowest-objective evaluable endpoint in the marker,
effect-coded, std.lv and sphere families is continued at tighter tolerances
with budgets of 1,000, 10,000 and 50,000. These are allowed budgets, not evidence
that the optimizer used every evaluation. The inspection retains covariance
drift, objective, local accuracy and both the old extent and a new diagnostic
of scale-invariant component magnitudes. Neither magnitude statistic is a
proof of runaway or nonattainment.

```{r}
escape <- read.csv(file.path("results", "unresolved-identifications", "candidate_evidence.csv"))
tab(data.frame(Batch=escape$batch, Design=escape$design, N=escape$n, Draw=escape$rep,
  Objective=escape$objective, `Old extent`=escape$std_extent,
  `Component extent`=escape$component_extent, check.names=FALSE))
```

These are the lowest-objective locally checked candidates in this inspection,
not certified global optima. The magnitude-only screen still labels seven of
these `screened_extreme`; the original records are preserved.

- Development weak-marker N=20 draw 4 and fresh weak-marker N=20 draw 1 now
  pass the original screen. Marker charts find lower finite candidates than
  the sphere attempts in this inspection. Those sphere attempts can show
  large components on losing paths without establishing nonattainment of the
  problem itself.
- Both high-R2 development cases have moderate component magnitudes, around
  1.1–1.3, despite old standardized extents of roughly 18 and 177. A small
  total latent variance makes standardized ratios large; the fitted primitive
  components need not diverge. Multiple charts recover the same objective.
- Fresh Ernst N=20 draw 1 and weak-marker N=20 draw 4, N=100 draw 1, and N=100
  draw 4 have stable locally checked candidates with component magnitudes
  around 13–28. Their large signed variance components remain finite in the
  continuations. These are improper ML candidates, not PSD candidates.
- Development weak-marker N=100 draw 2 remains delicate. Several charts give
  almost the same objective, around 0.024689918, while representative component
  magnitudes range roughly from 125 to 169 and local-check outcomes differ.
  The best evaluable endpoint is not the best locally checked endpoint.
  Stagnation at tighter tolerances does not resolve finite attainment here.

The component diagnostic uses loading energies (squared loadings divided by
sample indicator variances) to scale latent total/disturbance covariances and
structural paths; residual contributions are scaled by indicator variances.
Unlike the old standardized extent, it never divides by a fitted latent total
variance. Both statistics are reported, and neither replaces the accuracy check.
The retained candidate parameters and implied covariances permit inspection
without rerunning the optimizers.

The practical conclusion is to separate magnitude flags from local accuracy
and from evidence about attainment. We do not need every unrestricted ML
problem to have a solution, and making PSD solve a different constrained
problem cannot answer whether ML attains its infimum. However, the original
five missing fresh references cannot be described collectively as five
nonconvergent or apparently nonattained fits.

### Known geometry before promotion

Five deterministic covariance witnesses isolate behavior from sampling noise.
The requested identification is the first marker except for the std.lv case.
The alternative chart always fixes the second indicator, selected before any
fit. ML uses L-BFGS and PORT with native layered and feasible-to-construct signed
spectral starts; PSD uses SLSQP with FABIN3 automatic transport and positive
spectral starts under both no and diagonal preconditioning. The sphere uses its
canonical start with polishing disabled. Its internal scaling is recorded
separately from ordinary ML's information scaling.

```{r}
g <- read.csv(file.path("results", "geometry-witnesses", "fits.csv"))
geom_names <- c("interior", "marker_pole", "residual_face", "stdlv_pole", "nonattainment")
recovery <- function(name, route, domain) {
  z <- g[g$design == name & g$route == route & g$domain == domain, ]
  if (any(z$screened & z$matches_covariance)) "Yes" else "No"
}
tab(data.frame(Case=c("Interior", "Marker pole", "Zero residual variance", "std.lv pole", "ML nonattainment"),
  `Requested ML`=vapply(geom_names, recovery, "", route="requested", domain="ML"),
  `Requested PSD`=vapply(geom_names, recovery, "", route="requested", domain="PSD"),
  `Alternative marker ML / PSD`=vapply(geom_names, function(n) paste(recovery(n,"alternate_marker","ML"), recovery(n,"alternate_marker","PSD"),sep=" / "), ""),
  `Sphere ML / PSD`=vapply(geom_names, function(n) paste(recovery(n,"sphere","ML"), recovery(n,"sphere","PSD"),sep=" / "), ""),
  check.names=FALSE))
```

“Yes” requires both the local endpoint screen and a maximum covariance error,
standardized by the target indicator variances, below 1e-6. It means exact-target
recovery to this tolerance, not simply a convergence flag. The table deliberately
shows no recovery for PSD on the last row: that target is outside the PSD model,
so PSD must minimize a different, strictly positive discrepancy.

At the marker pole the first loading is zero while the other indicators share
nonzero covariance. The original marker cannot represent this point. Its
spectral construction explicitly returns `start_unavailable`; choosing a
representable chart permits fitting. At the std.lv pole the endogenous latent's
disturbance variance is zero, whereas the requested identification fixes it at
one. Ordinary fits report convergence and pass the local screen at finite
approximations, but their standardized covariance gaps range from about 3e-6
(PSD) to 2e-4 (ML). Alternative-marker and sphere gaps are below 3e-8. Thus
convergence and a small discrepancy do not establish finite attainment in the
requested chart. Successful alternative fits retain a translation-failure or
near-boundary flag for that original chart.

The zero-residual witness attains zero discrepancy with an observed covariance
that is still positive definite. Every tested PSD configuration recovers it;
a primitive covariance boundary is therefore not a failed fit. The numerical
face nomination uses a 1e-7 threshold on scaled primitive eigenvalues, with
PSD feasibility and accuracy assessed separately. The scale is invariant to
indicator units and latent rescaling for these simple-structure models; it is
not a general boundary certificate.

#### A nonattainment witness, independently of optimization

For a one-factor model with diagonal residual covariance, consider

$$
S=\begin{pmatrix}1&0.3&0.4\\0.3&1&0\\0.4&0&1\end{pmatrix}.
$$

The nonzero covariances $s_{12}=\lambda_1\lambda_2\phi$ and
$s_{13}=\lambda_1\lambda_3\phi$ force all three loadings and $\phi$ to be
nonzero at a finite solution. Then $s_{23}=\lambda_2\lambda_3\phi$ cannot
be zero. No finite factor parameters reproduce $S$, even with signed variances.
Yet the sequence

$$
\lambda(t)=(t,0.3/t,0.4/t)^\top,\qquad
\phi=1,\qquad \theta_i(t)=1-\lambda_i(t)^2
$$

keeps the diagonal and first two covariances fixed while
$s_{23}(t)=0.12/t^2\to0$. Its observed covariance is positive definite for
$t\ge1$. Since the Gaussian covariance discrepancy is nonnegative and vanishes
only at equality, its infimum is zero and is not attained. This is a proved
closure/nonattainment example; no claim about a particular differential-geometric
fold type is needed. The stored sequence and checks verify the covariance map
and the library's half-discrepancy convention independently of optimization.

For $t>1$ this sequence has a negative first residual variance. PSD excludes it.
All ten tested PSD attempts across the requested, alternative and sphere routes
instead agree at a boundary candidate with reported discrepancy 0.009509003 and
$s_{23}\simeq0.12$. Their agreement and local checks are not a global-optimality
proof. No unrestricted ML attempt passes the local screen on this witness.

### Fresh PSD reference coverage

The retained fresh batch now has a separate PSD reference pool: canonical,
positive spectral and three reproducible random sphere starts under each of
the two preconditioning choices. The ordinary arms run independently on every
draw, with no reference-selected retries. All 60 draws have screened sphere
references: 58 reached by multiple start recipes and two by only one. All are
marker-representable at 1e-6; 42 draws have a matching reference candidate near
a PSD face under the numerical nomination above.

```{r}
psd <- read.csv(file.path("results", "psd-fresh", "summary.csv"))
tab(data.frame(Preconditioning=psd$preconditioning,
  `FABIN3-auto matches`=psd$fabin3_hit, `Positive spectral matches`=psd$spectral_hit,
  `Combined matches`=psd$portfolio_hit, `Any screened candidate`=psd$any_screened,
  check.names=FALSE))
```

Counts are out of 60. Both settings leave eight draws with screened but inferior
ordinary candidates and one with no screened ordinary candidate. No ordinary
candidate improves on the sphere reference. This is evidence for a remaining
search problem, not evidence that these nine datasets lack a PSD solution.
The references remain best observed candidates with a marker-chart accuracy
cross-check; they do not replace the missing sphere-native assessment. The
batch was already used in the direct-start comparison, so it is not a new
held-out decision study.

### What a spectral start does

A start is an initial guess for every parameter. Each indicator block supplies
a correlation pattern. The positive recipe chooses a loading direction that
captures positive shared variation. The negative recipe chooses an alternative
direction suited to a negative latent variance. Changing direction changes the
relative loadings. The construction adjusts residual variances so the implied
observed covariance is positive definite at the start.

Each factor receives one of those two guesses. Regressions and between-factor
covariances start at zero. The optimizer then estimates all free parameters
normally: the initial signs are not constraints, and every recipe targets the
same ML objective. Negative latent variances can be allowed by ML while the
observed covariance remains positive definite; such endpoints may be
substantively inadmissible and are flagged separately. These recipes are not
being proposed for PSD estimation.

### Direct ordinary-chart starts

The direct probe constructs every start from the sample covariance and runs
all recipes on every draw. It reads saved sphere references only after fitting;
no sphere endpoint, reference-selected marker, or reference-dependent retry
enters this portfolio. The spectral directions are converted algebraically to
the requested marker units: divide each loading by its marker loading and
multiply the latent variance by the squared marker loading. A zero marker
component would prevent that conversion; none occurred in these runs.

ML uses native layered starts plus four spectral alternatives, under information
scaling, separately with L-BFGS and PORT. PSD uses FABIN3 with automatic std.lv
transport plus the positive spectral alternative, under SLSQP with either no
preconditioning or diagonal preconditioning. Negative-variance recipes are
excluded from PSD. Each batch contains 60 draws from the original three
populations; the fresh batch is the previously retained start-study batch,
not a newly independent confirmation after this probe.

```{r}
direct <- read.csv(file.path("results", "direct-starts-information", "summary.csv"))
d <- direct[direct$reference_available > 0, ]
tab(data.frame(Batch=d$batch, Domain=d$domain, Backend=d$backend,
  Scaling=d$preconditioning, `Available references`=d$reference_available,
  `Constructor matches`=d$constructor_hit, `Portfolio matches`=d$portfolio_hit,
  check.names=FALSE))
```

Counts are out of 60 draws per batch. A match requires the existing endpoint
screen and objective agreement with the saved domain-specific reference.
No screened ordinary point improved on an available reference in this run.
The constructor is layered for ML and FABIN3 with automatic transport for PSD.
At the time of this direct run, the fresh PSD arms produced screened candidates
on 59/60 draws under either scaling choice, but had no sphere reference pool.
Those frozen match counts remain unavailable; the later PSD-specific run above
adds an independent reference comparison on the same draws. Raw attempts and starting vectors stay
local; the paired rows, summaries and reference classifications are retained.

The earlier reference-informed diagnostic is separate: translating a known
sphere endpoint recovered all 16 ML misses per backend and all six PSD misses
with references. That establishes that those points can be reached from a
known solution; it does not establish a sample-only recipe. Historical files
labelled `default` describe the then-used constructor and settings, not a
stable or recommended configuration. New ordinary probes name these settings
explicitly.

### What happened to fold / pole / fine?

There is no inferred fold/pole/fine partition for the random-draw batches.
The later deterministic geometry witnesses above supply known cases instead. The retained chart flag is
proximity to the requested marker-chart boundary, at tolerances 1e-4, 1e-6 and
1e-8. Numerical accuracy, parameter extent and PSD admissibility are separate
flags. At 1e-6, all 56 development ML references, all 55 fresh ML references,
and all 60 development PSD references are representable. The remaining four
and five ML problems have no screened sphere reference and remain unresolved.
A representable candidate is not a proof of a global optimum or of good
behavior in every chart. The ordinary probe does not measure std.lv proximity.

Thus the recovered misses with finite representable references provide evidence
of start/optimization trouble. They do not support labelling those targets
marker poles. A near-boundary flag alone would only nominate a possible pole;
a fold would require separate evidence that the limiting covariance is not
attained in any relevant chart. Failed multistarts and an extreme endpoint
cannot supply that conclusion.

### Expansion to new model families

The expansion uses Gaussian data, weak first indicators, unequal residual
variances, and N=30/100 with 20 draws per cell. New families are one factor with
five indicators (including a negative loading), two correlated factors with
four indicators each (mixed loading signs), and a three-factor regression chain
with three indicators per factor. Families and recipes were fixed before this
run. These models expand structural scope but still have simple measurement
structure, no equality constraints, and no missing data.

```{r}
expanded <- read.csv(file.path("results", "expanded-families", "summary.csv"))
x <- expanded[expanded$backend == "both", c("design", "portfolio", "matches_best_observed")]
x <- reshape(x, idvar="design", timevar="portfolio", direction="wide")
x <- x[c("design", paste0("matches_best_observed.", c("canonical", "single_negative", "all_signed", "positive_control")))]
names(x) <- c("Family (40 problems)", "Canonical", "+ Single negatives", "+ All negative subsets", "+ Positive control")
tab(x)
```

All entries pool the same two backends and compare against the full portfolio's
best screened objective. Canonical plus every nonempty negative subset uses
2^k starts for k factors; retaining the positive spectral control uses 2^k+1.
That is 6, 10 and 18 fits per problem for these one-, two- and three-factor
families with both backends. The full expansion uses 1,360 fits; recorded fitting
and endpoint-assessment time totals about 54 seconds on this machine. The three-factor full portfolio takes about 21 times
the canonical fitting time. Every draw's coverage and timing are retained in
`expanded-families/coverage.csv`.

The positive control's unique improvement is one-factor N=30 draw 1: PORT
finds a screened objective of 0.190643 versus canonical's 0.199426. The lower
point is inadmissible by the substantive variance screen, but remains a valid
candidate for unconstrained ML. This also shows why starts are search choices,
not declarations of endpoint admissibility. PORT alone reaches all 118 expanded
references, versus 117 for L-BFGS; this small result does not select a backend.

The two unresolved problems differ. One-factor N=30 draw 10 has locally screened
endpoints excluded only by the extent rule. Three-factor N=100 draw 6 has no
endpoint passing the full local screen. Neither is labelled nonexistent. Across
the expansion, 97 references have support from multiple start recipes and 21
from just one; the latter merit inspection before stronger reference claims.

### How often a reference is found

All datasets remain in the denominator. ML and PSD references are built
separately, using sphere attempts only. Repetition means different start
recipes reached the best observed objective; two optimizers from the same
start do not suffice.

```{r}
a <- aggregate(list(Datasets = refs$rep), refs[c("design", "domain", "reference_label")], length)
names(a)[1:3] <- c("Design", "Domain", "Reference evidence")
tab(a)
```

### Which sphere configurations reach the reference

The same random starting points are supplied to each configuration on a draw.
Each configuration has canonical and layered starts, plus
`r metadata$value[metadata$key == "random_starts"]` random starts.
The random starts have positive disturbances but varying loading signs and
latent paths. Their limited distribution does not exhaust possible basins.
The table pools starts and the two sample sizes within each design, so its
attempts are repeated fits, not independent datasets. Screening time is included.

```{r}
z <- s[s$route == "sphere", ]
z$screened <- z$fits * (z$label == "screened_candidate")
z$matched <- z$fits * (z$comparison == "matches_sphere_reference")
a <- aggregate(cbind(fits, screened, matched) ~ design + domain + backend + preconditioning, z, sum)
names(a) <- c("Design", "Domain", "Optimizer", "PSD scaling", "Attempts", "Screened", "At reference")
tab(a)
```

### Ordinary fits compared with the sphere reference

These attempts never define the sphere reference. A better ordinary endpoint is
retained as evidence that the sphere search missed something; the reference is
not silently replaced. Every finite returned ordinary endpoint is checked in
the same strongest-indicator chart. Its original convergence flag is retained
separately in the raw results.

```{r}
z <- s[s$route != "sphere", ]
a <- aggregate(fits ~ domain + route + backend + comparison, z, sum)
names(a) <- c("Domain", "Route", "Optimizer", "Comparison", "Attempts")
tab(a)
```

The ordinary PSD route remains ordinary PSD-ML. Sphere is an auxiliary
engineering reference: this study does not introduce sphere fitting into the
ordinary estimator, change paper protocols, or count a sphere rescue as an
ordinary-route success. A paper can report ordinary performance against a
best-observed numerical reference while describing that reference transparently.

### Purposeful starts: loading direction matters

The missed lavaan witness has a negative X-factor variance. The pilot's random
starts used positive latent variances. For this witness, merely making X's
starting variance negative fails; changing its loading direction to the
smallest-eigenvalue direction succeeds with either initial variance sign.
Thus the sign alone does not explain the miss. These are ML starts: negative
latent variances are outside the PSD estimator's domain.

The follow-up uses each factor's within-block correlation matrix with its
diagonal removed. A positive factor uses its largest-eigenvalue direction;
a negative factor uses its smallest. Both start with standardized variance
magnitude 0.5, zero structural regression, and residual variances chosen to
preserve observed diagonal variances. The implied observed covariance is
positive definite for either sign. The first follow-up used only the original
two-factor design. The expansion applies the same construction to other simple
structures; extension to general SEM remains open.

Four spectral recipes (both positive, negative X, negative Y, both negative)
were tested with canonical starts under both ML backends. The recipes were
fixed before the fresh run; the recommendation to drop the all-positive recipe
is a retrospective simplification, still needing confirmation. Fresh means new
random draws of the same three populations, not held-out model families.

```{r}
start_tables <- lapply(c("development", "fresh"), function(phase) {
  path <- file.path("results", paste0("start-", phase), "coverage_summary.csv")
  if (!file.exists(path)) stop("Missing start evidence: run scripts/tune_starts.R and scripts/summarize_starts.R")
  x <- read.csv(path)
  x <- x[x$backend == "both", ]
  aggregate(x[c("matches_best_observed", "attempts")], x[c("phase", "portfolio")], sum)
})
x <- do.call(rbind, start_tables)
names(x) <- c("Batch (60 problems)", "Start portfolio", "Best observed reached", "Fits")
tab(x)
```

Coverage means at least one endpoint matches the best screened objective in
that batch's reference pool. Development references additionally retain the
original pilot candidates; fresh references use only the new portfolio. The
canonical figure is therefore 45/60 on development against the expanded pool,
versus 46 against the older pilot pool. Adding both single-negative recipes
reaches 54 development and 52 fresh references; the double-negative recipe
adds two and three, respectively. Neither backend dominates: each alone reaches
54/60 fresh references with the full portfolio; together they reach 55/60.
The retained `coverage.csv` gives every paired case, backend and fit cost.

Three of the original nine missing references already had endpoints passing
the local checks but excluded by the extent screen. All three remain excluded.
Of the other six, five now have a screened candidate, leaving weak-marker
N=20 draw 4 unresolved. The development reference pool now has 50 repeated,
six single-start and four missing references. The fresh pool has 38 repeated,
17 single-start and five missing references. Supporting starts are recipes,
not independent replications; agreement by two backends alone is not repetition.

A canonical fit can pass the screen and still miss a better observed solution.
Therefore a retry triggered only by failed convergence would miss some of this
benefit. Keep the small portfolio as an exploratory reference search before
trying to invent an adaptive retry rule. No additional optimizer or termination
machinery was needed for these recoveries.

## Caveats

- This is exploratory development data, not a held-out default-selection run.
  The initial pilot has three two-factor models, N = 20 and 100, ten draws per
  cell, complete Gaussian data, one group and native units. Both latents must
  be covered by the sphere transformation. No conclusions extend to uncovered
  structures, FIML, DWLS or barrier fitting.
- Fits use unit-free sphere coordinates with user-chart polishing disabled.
  The current route still finalizes in the original chart when possible; a
  finalization error can therefore hide an otherwise usable sphere endpoint.
  Such attempts remain errors, never evidence that no solution exists.
- The strongest-indicator assessment is a local Newton cross-check, not an
  intrinsic sphere certificate or an identification proof. PSD checks use the
  PSD domain. Original-chart verdicts do not decide the reference screen.
- A first-order sphere check, norm residual at most 1e-6, finite consistent
  objectives, and the local Newton budget .01 must pass. PSD also requires
  covariance admissibility. These are an explicit exploratory screen, not a
  replacement library verdict. Nonpositive curvature, unavailable accuracy and
  excessive Newton distance are recorded separately in the raw evidence.
- Standardized extent above 10 is an **extreme-parameter screen**. Such points
  remain recorded but are excluded from the main reference pool. They are not
  called runaways. A finite improper optimum can be ML's best answer; being
  improper alone never excludes an ML candidate. A validated escape diagnostic
  and threshold-sensitivity study remain open.
- Objective agreement uses 1e-6 times one plus the absolute reference objective.
  Equal objectives need not imply equal parameters or the same basin. The
  number of observed objective levels is not a count of local minima.
- The marker diagnostic is evaluated at each endpoint, not at a known optimum.
  A level below 1e-6 means numerical proximity to the requested chart's boundary,
  not a proven zero loading. Levels and classifications at 1e-4, 1e-6 and 1e-8
  are retained. No chart condition is counted as convergence on its own.
- Sphere may remove identification poles while other parameter escape paths
  persist. Neither unsuccessful multistarts nor an ML objective below the PSD
  objective proves nonattainment.

## Reproduce

The single-start selection check reuses raw fits from `direct_starts.R` and
`tune_starts.R` (development and fresh phases). Its small selections, paired
outcomes, summaries, and metadata are frozen for reading the report:

```sh
Rscript scripts/select_single_start.R --help
Rscript scripts/select_single_start.R --smoke
Rscript scripts/select_single_start.R
```


The high-precision diagnostic scripts require Python with `mpmath`; they are
study-local checks of the six-indicator, two-factor covariance model only.

```sh
cd experiments/engineering/active/15-sphere-reference-fits
Rscript scripts/check_labels.R
Rscript run_experiment.R --smoke
Rscript run_experiment.R --pilot
Rscript scripts/probe_reference_gap.R reference-pilot
Rscript scripts/tune_starts.R --phase witness
Rscript scripts/tune_starts.R --phase development
Rscript scripts/tune_starts.R --phase fresh
Rscript scripts/direct_starts.R
Rscript scripts/probe_geometry.R
Rscript scripts/check_geometry.R
Rscript scripts/psd_fresh.R
Rscript scripts/inspect_unresolved.R
Rscript scripts/check_escape.R
Rscript scripts/translate_sphere.R
Rscript scripts/check_requested_chart.R
Rscript scripts/export_open_cases.R
python3 scripts/refine_open_cases.py
python3 scripts/profile_delicate.py
python3 scripts/resolve_profile_minimum.py
python3 scripts/resolve_profile_minimum.py --digits 90 --run-id delicate-finite-minimum-90
python3 scripts/check_open_cases.py
Rscript scripts/check_open_case_translation.R
# Optional reference-informed diagnostic; requires retained raw pilot/start fits:
Rscript scripts/probe_ordinary.R --run-id ordinary-explicit
Rscript scripts/summarize_starts.R
Rscript scripts/check_expansion.R
Rscript scripts/expand_starts.R --smoke
Rscript scripts/expand_starts.R
quarto render report.qmd
# Optional small paired unit check; a distinct run, not default-selection evidence:
Rscript run_experiment.R --smoke --transforms native,x0.01,x100 --no-lavaan --run-id reference-units
```

The runner refuses to overwrite a completed run; choose a fresh `--run-id` for
reruns and render it with `quarto render report.qmd -P run:NAME`. The seed base
is separate for smoke and pilot profiles. Raw fits, endpoint parameter tables,
random starts and comparison rows stay local under `results/`. Small pilot
summaries, per-problem references, paired start coverage and metadata are frozen with the source so
the report can be rendered without rerunning fits. The start-follow-up sections
always show the frozen development/fresh and expanded-family batches. Witness and development
runners require the original pilot raw fits, regenerated by the pilot command;
the fresh runner does not. Metadata identifies the
installed package and source checkout; a dirty checkout is recorded explicitly.

## Label contract

Labels describe the evidence collected, never a mathematical existence claim.
The raw columns retain the ingredients so a later screen can be compared without
rewriting the original run.

| Label | Meaning | Does not imply |
| --- | --- | --- |
| `screened_candidate` | Endpoint passes the stated numerical and extent screen | A certified local minimum, identification, or global optimality |
| `screened_extreme` | Local checks pass, but standardized extent exceeds 10 or is undefined | A proven runaway; a finite optimum is impossible |
| `accuracy_failed` | Available Newton distance exceeds its budget | No better start or backend can solve the problem |
| `accuracy_unresolved` | Curvature/conditioning does not support the requested accuracy assessment; consult `audit_status` | Every unresolved case has the same mechanism |
| `sphere_first_order_failed` / `sphere_norm_failed` | The driven first-order or unit-norm check fails | The endpoint is a local optimum at a pole |
| `partial_sphere` | At least one relevant latent did not use the sphere | This attempt is a pole-free reference |
| `nonfinite_objective` / `objective_disagreement` / `psd_infeasible` | A finite, consistently evaluated point in the requested domain was not established | A valid reference candidate |
| `assessment_unavailable` / `call_error` | Endpoint assessment or fitting could not return the required evidence | Nonexistence |
| `single_start_best_observed` | Best screened sphere objective has one supporting start recipe | An unreliable point by definition, or a unique minimum |
| `repeated_best_observed` | At least two start recipes reach that objective within tolerance | Independent replications or exhaustive search |
| `no_screened_reference` | The finite portfolio supplies no qualifying reference | An unattainable problem |
| `near_chart_boundary` | Endpoint's original marker-chart level is below the declared tolerance | A mathematically exact pole or failed sphere optimization |
| `better_than_sphere_reference` | Screened ordinary endpoint improves on the sphere reference | The ordinary endpoint is globally optimal |
| `above_sphere_reference` | Screened endpoint has a worse objective | It lies in a distinct basin; convergence failed |
| `matches_sphere_reference` | Screened endpoint agrees in objective within tolerance | Equality of parameter estimates |
| `not_screened` / `no_sphere_reference` | Comparison lacks a qualifying endpoint or a sphere reference | Inferiority to a known optimum |

## Current choice and next step

The current engineering setup is a small sphere portfolio, strongest-indicator
cross-check, and separate original-chart diagnostics. Alternatives compared are
L-BFGS versus PORT for ML, and diagonal versus no PSD scaling with SLSQP.
A configuration deserves a larger follow-up only if it repeatedly finds the
best screened candidates without concentrating unresolved outcomes in a design;
this pilot does not select a library default.

The targeted witness follow-up is selected after inspecting the pilot; it is
not part of the original reference pool. Both recoveries are recorded in
`witness_probe.csv`. The unit smoke run separately retains paired rescalings;
it is too small to establish general unit invariance.

Retain the positive spectral control alongside canonical and signed alternatives:
the new-family run refutes dropping it universally. The exhaustive sign portfolio
remains an exploratory reference tool for at most three factors. Before adding
more starts, inspect the 21 single-start references and the two unresolved cases.
Developing a cheaper selection rule is open; changing only one factor at a time
already loses ten references here. A sphere-native assessment remains necessary for stronger local
accuracy claims. Preserve the existing
ordinary-parameterization comparisons for PSD work. Any production promotion
requires a fresh decision study with criteria written before its run.

This study was renamed from `15-sphere-local-convergence` on 2026-09-27.
Historical `results/full` and `results/smoke` are preserved but are not read by
this report. The former `no_estimate` classification combined failed alternative
fits with an objective below PSD; it was heuristic and must not be used as
proof of nonexistence. Historical sources remain available in Git history.

## Normalization revisit

A paired single-start rerun on the 120 retained draws compares normalization
explicitly off/on, with the same current library judge in both arms. No sphere
optimization, retries or endpoint-derived starts occur. ML uses layered/native
starts with L-BFGS and PORT, information scaling; PSD uses FABIN3/auto, SLSQP and
diagonal preconditioning. Two-stage PSD first uses L-BFGS/information. All have
5,000 evaluations, relative objective tolerance 1e-12 and parameter tolerance
1e-10; PSD floor and feasibility tolerance are 1e-6. The sample seed rules and
library fingerprint are in the frozen metadata.

ML targets use the saved reference plus the seven refined finite witnesses;
the two known near-pole cases are excluded from recovery denominators and
reported separately. PSD targets use the earlier pilot and retained fresh PSD
references. A target hit requires the current fit verdict, requested-marker
accuracy, chart translation at the explicit study tolerance 1e-4, and an
objective no higher than the target by more than 1e-6 times one plus its absolute
value. PSD additionally requires admissibility. Better objectives are flagged
separately; neither of the two lower-objective PORT endpoints passes accuracy.
The historical extent/strong-marker screen is retained as a separate column,
not the acceptance rule. Thus the new totals need not reproduce old screening
counts (such as 35 versus 36 ML hits). Compare the two current arms to isolate
normalization; do not interpret differences from older runs as its effect.

There are no target-hit gains or losses for either ML backend or direct PSD.
Two-stage loses three later-batch targets: Ernst N=20 draws 1 and 3, high-R²
N=20 draw 7. Two exhaust the PSD budget; Ernst draw 3 reaches a worse local
solution. Direct PSD also loses one certified fit that was already above its
reference, so unchanged target counts do not mean every outcome is unchanged.
PORT gains three certified fits in the later batch, all above their targets.
Every changed verdict, acceptance or material objective is retained in
`results/normalization-revisit/changes.csv`; per-family counts are in
`by_family.csv`. The compact results table is in the executive summary.

All seven high-precision witnesses reproduce their saved objectives when
re-evaluated with the current library. Six pass requested-marker accuracy and
one remains ill-conditioned. The two saved near-pole endpoints are still
rejected at the study tolerance 1e-4 and admitted at 1e-6; neither sample yields
a certified ML fit in this rerun. This is not evidence that a general production
chart gate is implemented, or that these samples cannot have other solutions.

These saved batches are development evidence, not fresh held-out validation.
No algorithms were tuned during this comparison. Reproduce from this folder
with the current ML/PSD normalization implementation installed:

```sh
Rscript scripts/normalization_revisit.R --smoke
Rscript scripts/normalization_revisit.R
Rscript scripts/check_normalization_witnesses.R
```

The runner refuses to overwrite results; use a fresh `--run-id NAME` and pass
that name to the witness checker when rerunning. Full fits and parameters remain
local, while compact summaries and all changes are tracked. The two-stage
regressions and the remaining finite-reference misses are the next diagnostic
targets; this run does not promote spectral starts or multistart.

## Two-stage handoff investigation

The [short two-stage report](two-stage-summary.html) summarizes the follow-up.
The three later-batch regressions were replayed with their original controls.
Both ordinary endpoints were then supplied explicitly to PSD with normalization
off/on, alongside independent FABIN3-auto controls, at budgets of 5,000 and
25,000. Every combination is retained; these diagnostic fits are not a new
multistart policy. No sphere reference enters a start.

All six ordinary endpoints were returned and used as warm starts. Ernst draw 1
and high-R² draw 7 fail ordinary accuracy under both normalizations; Ernst draw 3
passes accuracy but is inadmissible. The standalone PSD fits reproduce the
wrapper outcomes at the same endpoint, normalization and budget. This supports
handoff/path sensitivity rather than an unexplained wrapper discrepancy.

For Ernst draw 1, either endpoint reaches the PSD target 0.13842345 with
normalization off; with it on, both exhaust 5,000 evaluations and eventually
return objective 0.22897031 at the larger budget. A cold FABIN3-auto start reaches
the target under either normalization. For Ernst draw 3, the old ordinary
endpoint leads to PSD objective 0.39193273 under both normalizations. The new
ordinary endpoint has a better unrestricted objective (0.33911821 versus
0.37150774), yet leads to the worse PSD objective 0.40385641 under both
normalizations, as does the cold start. This loss is endpoint selection, not a
failure of PSD accuracy at the returned local solution.

For high-R² draw 7, the old endpoint reaches 0.35601702 only with PSD
normalization off; turning it on reaches 0.39103251. The new endpoint exhausts
5,000 under either normalization, then returns 0.44965490 at the larger budget.
The cold start reaches 0.36575495 under either normalization. The extended-budget
Ernst/high-R² paths have loadings in the millions and are not recovered targets;
a finite return or convergence flag alone does not justify accepting those
requested-chart endpoints. The study's target comparison remains separate.

A retrospective rule permits warm starts only from certified ordinary fits;
rejected ordinary estimates instead use the original FABIN3-auto start. This
was checked on all 120 normalized cases, using the saved direct-PSD fit as that
counterfactual. The library fingerprint is verified, and every current two-stage
verdict and returned objective is replay-checked against the saved run. Ten
handoffs change. Earlier-batch target recovery stays 54/60 (one gain and one
loss); later-batch recovery rises 46→47/60 (one gain). Gains: earlier Ernst
N=20 draw 9 and later Ernst N=20 draw 1. Loss: earlier weak-marker N=100 draw 9.
Per-family outcomes and every pair are saved. This is not held-out evidence and
does not justify promoting the rule. No production code is changed.

Reproduce using the same library as the saved normalization revisit:

```sh
Rscript scripts/probe_two_stage.R
Rscript scripts/check_two_stage_handoff.R
```

Both refuse to overwrite an existing directory; use a fresh `--run-id NAME`.
Frozen evidence is under `results/two-stage-probe/` and
`results/two-stage-handoff/`. Keep the current handoff while two-stage remains
provisional; broader start/PSD-boundary work is separate from normalization.

## Independent PSD fallback

The user's intended policy is ordinary-first acceptance followed, when needed,
by PSD with its own start. It does not reuse ordinary estimates, even when
ordinary ML converges but is inadmissible. This differs from the earlier
certified-only handoff probe, which still warmed PSD from converged inadmissible
ordinary solutions. The [current short report](two-stage-summary.html) presents
this exact policy, superseding its earlier handoff recommendation.

All 120 saved challenge datasets were fitted explicitly under normalization.
Ordinary ML uses FABIN3-auto, L-BFGS/information; its convergence and admissibility
verdicts decide whether PSD is needed. PSD independently constructs FABIN3-auto
starts in an unchanged model, with SLSQP/diagonal, without receiving ordinary
parameters. Controls remain 5,000 evaluations, relative objective tolerance
1e-12, parameter tolerance 1e-10, and PSD floor/feasibility tolerance 1e-6. These
starts match the previous comparison; choosing a different ordinary start is
outside this isolated handoff test. Final target judging retains PSD accuracy
and the requested-chart check at 1e-4, without a magnitude cutoff.

Ordinary ML is accepted on 34 datasets (18 earlier, 16 later); 86 require PSD.
Every cold fallback reproduces the saved direct PSD return status, verdict,
acceptance and objective. Accepted ordinary fits also reproduce their saved
ordinary-first results. All 120 final acceptance and target-hit statuses agree
with direct PSD; the maximum objective difference across returned pairs is
9.41e-14 (rounded upward). Earlier target recovery is 54/60 and later 47/60,
versus 54/60 and 46/60 for the warm-start route. The gains are earlier Ernst
N=20 draw 9 and later Ernst N=20 draw 1; the loss is earlier weak-marker N=100
draw 9. Per-family counts and every stage are preserved. The fact that the
previous certified-only candidate had the same hit counts is an observed result,
not an equivalence between those policies.

Independent fallback is a reasonable candidate and warm starts are not required
by the mathematics. These development results do not establish a universal
advantage or a speed benefit: 34 PSD calls are avoided, but the other 86 cases
incur an additional ordinary attempt. Production warm-start behavior is unchanged;
a future API/policy change should expose this distinction explicitly.

With the library and raw comparisons from the normalization revisit available:

```sh
Rscript scripts/check_two_stage_cold.R --smoke
Rscript scripts/check_two_stage_cold.R
```

Use a fresh `--run-id NAME` for reruns. The runner checks the library fingerprint
and reproducibility against the saved direct and warm fits. Frozen evidence is
in `results/two-stage-cold/`; full fitted parameters remain local. No reference
solution enters a starting point, and no retries or multistart are used.

## Two-stage timing

The [short timing report](timing-summary.html) compares direct PSD,
independent two-stage, and the existing warm-start two-stage on all 120 retained
cases. A single R process with single-threaded math warms every case/method,
then measures all six method orders per case across six rounds. Case order is
shuffled deterministically each round; five calls form each timed block,
for 10,800 timed policy calls. Model/sample preparation, manual garbage
collection, warmup and result verification are outside measured blocks.
Failures remain in the timing. Every block's final result reproduces the saved
verdict and objective. The original fixed starts, normalization, tolerances and
5,000-evaluation budgets are unchanged.

Per-case costs are medians of six block-average wall times. Their sums give
estimated costs per 120-case pass: direct 0.553 s, independent two-stage 0.703 s,
and warm two-stage 0.839 s. Independent is about 27% slower than direct and 16%
faster than warm; warm is about 52% slower than direct. Independent and direct
retain the same target recovery, while warm has different gains and losses.
On the 34 ordinary-accepted cases, mean independent time is 0.35 ms versus
0.96 ms direct. On the other 86, independent costs 8.03 ms versus 6.06 ms direct.
That mix explains why the shortcut does not pay overall. Weak-marker cases
dominate total cost. Per-family, per-case, path and round summaries are frozen
under `results/two-stage-timing/`; raw block measurements stay local.

Absolute round totals drift upward during the run, but direct PSD is fastest
in every round. Use the per-round ratios and repeated costs rather than one
pass as the evidence. Wall and process-CPU totals closely agree. These are
end-to-end R-accessible workflows on an Intel Core i7-1355U, including automatic
starts, library diagnostics and returned R objects. Independent fallback uses
up to two library calls, while warm fallback is orchestrated natively. The
result is not an isolated C++ timing or a prediction for a future integrated
independent-fallback API, large models or a representative user workload.

Independent fallback remains a reasonable policy, but it is not a speed
optimization for this saved difficult-case mix. No production policy changes.
Reproduce with the same library fingerprint and saved two-stage-cold results:

```sh
Rscript scripts/time_two_stage.R --smoke
Rscript scripts/time_two_stage.R
```

Use a fresh `--run-id NAME` to rerun. The runner prints elapsed time and an ETA,
checks all saved outcomes outside timed blocks, and records timing scope and
platform metadata. Rendering the report performs no fitting.

## Normalization timing

The [updated timing summary](timing-summary.html) answers the absolute-speed and
normalization-off/on questions. It adds standalone ordinary ML with the saved
layered/native start and repeats direct, independent two-stage and warm two-stage
under both normalization settings. Two-stage starts remain FABIN3-auto; all
optimizers and budgets are unchanged. Four serial rounds rotate the four routes
through every position once per case and alternate normalization order. Three
calls per timed block give 11,520 policy calls, excluding warmup. Preparation,
manual garbage collection and verification remain outside timing. Every block's
last outcome matches its saved return status, objective and convergence flag.

Typical (median) milliseconds per fit, off→on: ordinary ML 0.47→0.48, direct
PSD 1.07→1.09, independent two-stage 1.38→1.42, warm two-stage 1.37→1.37.
Means including difficult/failed fits are 2.06→1.96, 4.22→4.81, 5.33→6.06 and
5.40→7.22 ms respectively. Thus normalization's aggregate cost ratio is 0.95
for ML, 1.14 for direct PSD, 1.14 for independent two-stage and 1.34 for warm
two-stage. Per-round ratios retain the same direction: ML 0.941–0.953,
direct 1.125–1.148, independent 1.127–1.146 and warm 1.332–1.383.

These are full-fit timing changes, not isolated normalization overhead.
Direct PSD's 70.52 ms extra bank cost is dominated by 67.30 ms on one newly
failing case: later weak-marker N=20 draw 2 takes 1.46→68.76 ms. It missed the
reference under both settings. The total on cases certified in both settings
falls by 0.27 ms; old failures add 3.50 ms. Normalized direct PSD's six failures
consume 400.0 of its 577.0 ms total, about 69%. Its median is 1.09 ms, 95th
percentile 14.86 ms and slowest 68.76 ms. This explains the gap between typical
and mean times. The warm route's three newly failed cases similarly contribute
195.75 ms of its 217.88 ms increase. All outcome-dependent costs are retained;
no failed fit is removed to improve a timing comparison.

The results concern small six-indicator, two-factor covariance models with
prepared samples/models, no inference, and the current R-accessible workflows
on an Intel Core i7-1355U with single-threaded math. They do not predict large-
model, missing-data, multigroup or full user-call runtimes. Independent fallback
still incurs up to two library calls versus native warm orchestration.
No production policy or controls changed. Prior timing figures remain frozen
as a separate run; modest absolute differences across runs are expected.

Reproduce with the saved normalization-revisit comparisons and same library:

```sh
Rscript scripts/time_normalization.R --smoke
Rscript scripts/time_normalization.R
```

Use a fresh `--run-id NAME` when rerunning. Frozen per-case, per-family,
convergence-status, aggregate and repeat summaries are under
`results/normalization-timing/`; raw measured blocks stay local. Model/sample
preparation, extra audits and report rendering are not timed.


## Activity after the engineering review

Active. Normalized complete-data ML and direct PSD remain the current routes;
independent PSD fallback is an exploratory alternative. The next checks are a
requested-chart rejection criterion that distinguishes poles from finite improper
minima, and a fresh comparison of fallback handoffs. The retained cases are
development witnesses, not held-out default-selection evidence.
