---
title: "Optimizer defaults"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
run <- "2026-09-26"
rd <- function(lane, name) {
  p <- file.path("results", lane, run, paste0(name, ".csv"))
  if (!file.exists(p)) stop("Missing ", p, ": run run_experiment.R --lane ", lane)
  read.csv(p, stringsAsFactors = FALSE)
}
tab <- function(x, ...) knitr::kable(x, row.names = FALSE, format.args = list(big.mark = ","), ...)
fam <- c(research47 = "research/47", engineering15 = "engineering/15", boomsma = "Boomsma 1985",
         wolf = "Wolf et al. 2013", chen = "Chen et al. 2001", constrained = "equality-constrained",
         ernst = "Ernst (control)")
ml_rates <- rd("ml-gls", "rates_by_family"); ml_pair <- rd("ml-gls", "paired")
ml_dec <- rd("ml-gls", "decision"); ml_loss <- rd("ml-gls", "losses")
ml_breaks <- rd("ml-gls", "invariance_breaks")
ps_rates <- rd("psd-ml", "rates_by_family"); ps_pair <- rd("psd-ml", "paired")
ps_dec <- rd("psd-ml", "decision"); ps_fail <- rd("psd-ml", "failures")
ps_stage <- rd("psd-ml", "twostage_stages"); ps_tol <- rd("psd-ml", "tolerance")
rdc <- function(name) { run <<- "2026-09-26-confirm"; on.exit(run <<- "2026-09-26"); rd("ml-gls", name) }
cf_dec <- rdc("decision"); cf_rates <- rdc("rates_by_family"); cf_pair <- rdc("paired"); cf_esc <- rdc("escapes")
ml_esc <- rd("ml-gls", "escapes"); ps_rt <- rd("psd-ml", "rates_by_family_transform")
tot <- function(r, arm, col, est = NULL) sum(r[[col]][r$role == "test" & r$arm == arm &
  (is.null(est) | r$estimator %in% est)])
pr <- function(p, cand, base, col) sum(p[[col]][p$role == "test" & p$candidate == cand & p$baseline == base])
f0 <- function(x) format(round(x), big.mark = ",")
```

## Question

Which start, optimizer and PSD route should each estimation route use by
default? Every fit is judged by the library verdict, on simulated problems
the candidates were not developed on. The criteria were committed before the
runs, in [criteria/ml-gls.md](criteria/ml-gls.md) and
[criteria/psd-ml.md](criteria/psd-ml.md) (commit `ec5b0c96`).

## Answer

**ML and GLS: nothing is promoted yet. The pre-registered confirmation run
failed, and it exposed a flaw in how fits are scored.**

- The first run favoured the layered start with PORT (ML
  `r f0(tot(ml_rates, "layered_port", "success", "ML"))` against
  `r f0(tot(ml_rates, "default", "success", "ML"))` certified,
  GLS `r f0(tot(ml_rates, "layered_port", "success", "GLS"))` against
  `r f0(tot(ml_rates, "default", "success", "GLS"))`). Two small rule failures
  pointed at PORT's generic budget.
- The confirmation run, on fresh draws, gave PORT ML's budget. Its candidate
  fails rule 1 (Chen ML) and rule 2 in five ML cells. The extra budget made
  PORT *worse* on ML than generic PORT
  (`r sum(cf_pair$wins[cf_pair$role == "test" & cf_pair$candidate == "layered_port_ml" & cf_pair$baseline == "layered_port" & cf_pair$estimator == "ML"])`
  wins,
  `r sum(cf_pair$losses[cf_pair$role == "test" & cf_pair$candidate == "layered_port_ml" & cf_pair$baseline == "layered_port" & cf_pair$estimator == "ML"])`
  losses) and better on GLS.
- **The reason is runaway certifications.** On draws where ordinary ML has no
  proper minimum, PORT keeps walking down the divergent (Heywood) path. Where
  its budget stops it, the Newton check often accepts the point. In native
  units every PORT arm has about
  `r f0(sum(ml_esc$escapes[ml_esc$role == "test" & ml_esc$arm == "layered_port" & ml_esc$estimator == "ML"]))`
  such ML certifications with a largest parameter above 100 (median about
  9,600). L-BFGS has none.
- **Without them, the ML gain almost vanishes.** It stays large for GLS (see
  "Confirmation run" below). The layered start with L-BFGS fails only the Chen
  ML rule-1 check (4 losses, no wins), where today's default is already
  perfect.
- **Next:** add a runaway rule to the judge or the scoring, and pre-register
  a new run with the layered start with L-BFGS and the layered start with
  PORT as candidates.

**PSD ML: keep today's direct configuration. The two-stage route is better
only in rescaled units.**

- **Rule A fails.** The layered cold start loses in four of the six test
  families.
- **Rule B fails.** Dropping the diagonal preconditioning loses
  `r f0(pr(ps_pair, "psd_default_none", "psd_default", "losses"))` fits
  against `r f0(pr(ps_pair, "psd_default_none", "psd_default", "wins"))` and
  takes 2.4 times as long.
- **The two-stage route beats the direct fit.** It fits ordinary ML first
  with today's default and refits under the PSD constraint only when needed.
  It wins `r pr(ps_pair, "twostage_default", "psd_default", "wins")` and loses
  `r pr(ps_pair, "twostage_default", "psd_default", "losses")` certified fits
  against the direct fit, reaches the best known objective
  `r pr(ps_pair, "twostage_default", "psd_default", "best_wins")` against
  `r pr(ps_pair, "twostage_default", "psd_default", "best_losses")` times, and
  takes `r round(100 * (1 - tot(ps_rates, "twostage_default", "total_seconds") / tot(ps_rates, "psd_default", "total_seconds")))`%
  less time. The advantage is mostly a units effect, though. In native units
  the direct fit certifies slightly more
  (`r f0(sum(ps_rt$success[ps_rt$role == "test" & ps_rt$transform == "native" & ps_rt$arm == "psd_default"]))`
  against
  `r f0(sum(ps_rt$success[ps_rt$role == "test" & ps_rt$transform == "native" & ps_rt$arm == "twostage_default"]))`).
  Two-stage wins at $\times 100$ and $\times 0.01$, where the direct fit's
  absolute preconditioning clamp bites; with that fixed, the two routes may
  end up equal. The author decides the route.
- **Its ordinary stage must not inherit the layered start with PORT.** With
  that ordinary stage the two-stage route is worse than the direct fit: the
  PSD refit succeeds less often when it is warm-started from PORT's rejected
  or improper endpoints than from L-BFGS's.
- **Rule D passes.** The absolute feasibility tolerances do not bite.

Two issues found along the way:

- Direct PSD fits exhaust their budget more often after rescaling, in the
  pattern the preconditioning's absolute clamp predicts.
- A lift round-trip error hits the std.lv equal-loadings CFA (about 4% of
  fits at $N = 25$).

## The register

| Default | Current value | Status after this run | Evidence |
|---|---|---|---|
| ML and GLS start | transported FABIN3 (ML), native FABIN3 (GLS) | open: the layered start gains on GLS, only slightly on ML; rerun with a runaway rule | lane ml-gls, both runs; engineering/17 corpus |
| ML and GLS optimizer | NLopt L-BFGS | open: PORT's ML gain is mostly runaway certifications; its GLS gain is real | lane ml-gls confirmation run |
| judge on runaway paths | the Newton check accepts some divergent-path points | open: add a runaway rule before the next ML/GLS run | lane ml-gls confirmation run |
| optimizer coordinates | information-refined sample units | settled (2026-09-26); every arm here used them, and the candidate's success and objective agree across units | engineering/17; lane ml-gls invariance |
| PSD cold start | transported FABIN3 | settled: keep (rule A fails in 4 of 6 families) | lane psd-ml |
| PSD preconditioning | lifted information diagonal | settled: keep (rule B fails in every family) | lane psd-ml |
| PSD route | direct PSD fit | open, for the author: two-stage from today's ordinary default is better and faster | lane psd-ml |
| PSD feasibility tolerances | absolute | settled: keep (rule D passes, 0 fits) | lane psd-ml |
| PSD preconditioning clamp | $[10^{-4}, 10^{4}]$, absolute | open: costs direct PSD fits at $\times 100$ | lane psd-ml failures |

## How much to trust it

- **The problems are held out.**
  - 51 test populations and one control: research/47,
    engineering/15's weak marker and high $R^2$, Boomsma (1985), Wolf et al.
    (2013), Chen et al. (2001) with four fitted models, and ten
    equality-constrained models. The control is Ernst, used in the 2026-09-22
    start work.
  - Each population is drawn at a small $N$ and at $N = 100$: 100
    replications per $N$ for ml-gls, 50 for psd-ml.
  - Every draw is fitted in native units, $\times 100$, $\times 0.01$ and
    separate units.
  - None of these populations shaped the layered start, the coordinates or
    the PSD scaling.
- **One judge.** Success is the library verdict (the Newton check), plus
  admissibility for PSD fits. A problem counts only if some arm, or a PORT
  or PSD fit from the population values, certifies it.
- **The engine.** The `magmaanlab` build from `30514b09`, built 2026-09-26
  04:58 UTC. The engine source is unchanged through the pre-registration
  commit.
- **Limits.**
  - The data are multivariate normal and complete, and all groups are single.
  - FIML, least squares, ordinal routes and lavaan are not covered; the
    textbook corpus is engineering/17's evidence.
  - Counts of fits are descriptive; paired wins and losses on the same draws
    carry the comparisons.

## Lane ml-gls

### Rules

```{r ml-rules}
tab(ml_dec, col.names = c("Rule", "Test family and estimator", "Value", "Pass"))
```

### Certified fits by family

Attainable problems and certified fits, pooled over $N$, units and
replications.

```{r ml-rates}
w <- reshape(ml_rates[c("role", "family", "estimator", "arm", "attainable", "success")],
             idvar = c("role", "family", "estimator"), timevar = "arm", direction = "wide")
w <- w[order(w$role == "control", w$estimator, w$family), ]
tab(data.frame(Family = fam[w$family], Estimator = w$estimator, Attainable = w$attainable.default,
  Default = w$success.default, `Default + PORT` = w$success.default_port,
  `Layered + L-BFGS` = w$success.layered_lbfgs, `Layered + PORT` = w$success.layered_port,
  check.names = FALSE))
```

### The candidate's losses

Every loss is an optimizer termination; none is an error. The columns show
which single-change arm solves the same problem.

```{r ml-losses}
l <- ml_loss[ml_loss$role == "test", ]
l$attribution <- ifelse(grepl("layered_lbfgs:ok", l$arms), "the layered start with L-BFGS solves it",
                 ifelse(grepl("default_port:ok", l$arms), "today's start with PORT solves it",
                        "only today's default solves it"))
tab(as.data.frame.matrix(table(l$attribution, l$estimator)) |>
      (\(m) cbind(Attribution = rownames(m), m))())
tab(as.data.frame.matrix(table(l$optimizer_status, l$estimator)) |>
      (\(m) cbind(`PORT status` = rownames(m), m))())
```

Most losses are draws where the layered start leads both optimizers to a
worse place, and PORT then stops at its 1000-iteration budget. The single
Chen loss is of this kind: model m0 at $N = 25$ in separate units, where PORT
exhausted its budget at an objective of 0.763 against the default's certified
0.665.

### Invariance

Breaks among the draws of the candidate and the baseline, by pattern:

```{r ml-breaks}
b <- ml_breaks[ml_breaks$role == "test" & ml_breaks$arm %in% c("default", "layered_port"), ]
b <- stats::aggregate(groups ~ arm + estimator + units + pattern, b, sum)
tab(b[order(b$estimator, b$units, -b$groups), ], col.names = c("Arm", "Estimator", "Units", "Pattern", "Draws"))
```

The candidate flips between certified and not across units less often than
the default. Its excess on common units in Boomsma and Wolf ML is certified
fits whose objectives differ by $10^{-6}$ to $10^{-5}$ of the objective. That
is consistent with PORT stopping on its generic tolerances rather than ML's
tighter profile; the confirmation run with ML's controls will show it.

### Confirmation run

Seed base 2026092602, candidate `layered_port_ml`, with criteria in the dated
section of [criteria/ml-gls.md](criteria/ml-gls.md). Failing rules:

```{r cf-rules}
tab(cf_dec[!cf_dec$pass, ], col.names = c("Rule", "Test family and estimator", "Value", "Pass"))
```

Native units, test families, fresh draws: certified fits, runaway
certifications among them (largest $\lvert\theta\rvert > 100$; every
population has unit-order parameters), and certified fits without them.

```{r cf-escapes}
e <- stats::aggregate(cbind(fits, escapes, success, success_no_escape) ~ estimator + arm,
                      cf_esc[cf_esc$role == "test", ], sum)
tab(e[order(e$estimator, e$arm), ], col.names = c("Estimator", "Arm", "Fits", "Runaways", "Certified", "Certified, no runaways"))
```

- L-BFGS never certifies a runaway. PORT does, from either start.
- The PORT runaways that the 1000-iteration budget certified sit at a median
  largest parameter of about 9,600 (10th percentile 140). With 5000
  iterations PORT runs further, and the verdict rejects most of them. That is
  why the larger budget "lost" ML fits.
- Without runaways, on ML the layered start with L-BFGS and with PORT gain
  little over today's default. On GLS both gain several hundred fits.
- Both the pre-registered rules and the attainability witnesses count
  runaway certifications, so the next ML/GLS run needs a runaway rule
  written into its criteria.

## Lane psd-ml

### Rules

```{r psd-rules}
tab(ps_dec, col.names = c("Rule", "Test family", "Value", "Pass"))
```

### Routes

Attainable problems, successes (certified and admissible), best-known
objective reached, and total seconds, on the test families.

```{r psd-routes}
a <- stats::aggregate(cbind(attainable, success, best, total_seconds) ~ arm, ps_rates[ps_rates$role == "test", ], sum)
a$total_seconds <- round(a$total_seconds)
tab(a, col.names = c("Arm", "Attainable", "Success", "Best known", "Seconds"))
```

Which stage returned the two-stage fits, and how often it succeeded:

```{r psd-stages}
st <- stats::aggregate(cbind(fits, success) ~ arm + stage, ps_stage[ps_stage$role == "test", ], sum)
st$rate <- sprintf("%.1f%%", 100 * st$success / st$fits)
tab(st, col.names = c("Arm", "Stage", "Fits", "Success", "Rate"))
```

When the ordinary fit is rejected or improper, the PSD refit warm-starts from
the ordinary endpoint; when the ordinary fit errors, it starts from the
original start. From L-BFGS endpoints that refit succeeds far more often than
from PORT endpoints.

A probe on 30 draws where the layered-start PORT two-stage fails and the
direct fit succeeds (native units, mostly $N = 25$ with weak loadings) shows
why. On these draws the ordinary likelihood keeps improving along a Heywood
path: a residual variance goes negative while loadings grow.

- **L-BFGS** usually aborts early: 18 of 30 fits end in a line-search error
  after about 100 evaluations. The PSD refit then starts cold from FABIN3 and
  succeeds.
- **PORT** follows the path for its whole budget: 21 of 30 fits exhaust it.
  It ends far out, with a median most-negative variance of $-33$ and a median
  largest parameter of 211; the worst case reaches $-2543$. No PSD refit
  from these endpoints succeeds.

So PORT's persistence, an advantage when a minimum exists, turns into a poor
warm start when none does.

### Units and failure causes

The pre-registered tolerance check (rule D) finds no fit that fails only in
rescaled units because of link feasibility or admissibility.

```{r psd-tolerance}
tab(ps_tol, col.names = c("Units", "Arm", "Cause", "Fits"))
```

Two other failures depend on the model or the units:

```{r psd-failures}
f <- ps_fail[ps_fail$arm == "psd_default" & ps_fail$cause %in% c("budget exhausted", "link round-trip"), ]
f <- stats::aggregate(fits ~ family + cause + transform, f, sum)
f <- reshape(f, idvar = c("family", "cause"), timevar = "transform", direction = "wide")
names(f) <- sub("fits.", "", names(f), fixed = TRUE); f[is.na(f)] <- 0
tab(f[c("family", "cause", intersect(c("native", "x100", "x0.01", "mixed"), names(f)))])
```

- **Budget exhaustion rises after rescaling, at both ends of the clamp.**
  The lifted information scale $s_k = 1/\sqrt{I_{kk}}$ is clamped to
  $[10^{-4}, 10^{4}]$, and a variance of size $v$ wants $s \approx v$.
  - The chain SEM with equal paths goes from 0 exhausted fits in native units
    to 53 at $\times 100$, where its variances are near $10^{4}$.
  - With equal disturbances too, it goes to 54 at $\times 0.01$, where the
    disturbance variances are near $4 \cdot 10^{-5}$.
  - The pattern fits the clamp but is not yet proven.
- **The lift round-trip error ("could not round-trip the terminal covariance
  links")** occurs only in the two std.lv equal-loadings CFAs, mostly with
  loadings .4. It hits about 4% of their fits at $N = 25$ and 0.5% at
  $N = 100$. It occurs in native and $\times 0.01$ units about equally and
  never at $\times 100$. The mechanism is not yet known.

## Reproduce

```sh
export OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1
Rscript run_experiment.R --help
Rscript run_experiment.R --lane ml-gls --smoke
nice -n 10 Rscript run_experiment.R --lane ml-gls --workers 4   # about 10 minutes
nice -n 10 Rscript run_experiment.R --lane psd-ml --workers 4   # about 8 minutes
quarto render report.qmd
```

`--summarize` rescores existing raw rows (about 30 seconds per lane). Raw
rows stay local; the summaries in `results/<lane>/2026-09-26/` are tracked.
