---
title: "Optimizer defaults"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
run <- "2026-09-26"
rd <- function(lane, name) {
  p <- file.path("results", lane, run, paste0(name, ".csv"))
  if (!file.exists(p)) stop("Missing ", p, ": run run_experiment.R --lane ", lane)
  read.csv(p, stringsAsFactors = FALSE)
}
tab <- function(x, ...) knitr::kable(x, row.names = FALSE, format.args = list(big.mark = ","), ...)
fam <- c(research47 = "research/47", engineering15 = "engineering/15", boomsma = "Boomsma 1985",
         wolf = "Wolf et al. 2013", chen = "Chen et al. 2001", constrained = "equality-constrained",
         ernst = "Ernst (control)")
ml_rates <- rd("ml-gls", "rates_by_family"); ml_pair <- rd("ml-gls", "paired")
ml_dec <- rd("ml-gls", "decision"); ml_loss <- rd("ml-gls", "losses")
ml_breaks <- rd("ml-gls", "invariance_breaks")
ps_rates <- rd("psd-ml", "rates_by_family"); ps_pair <- rd("psd-ml", "paired")
ps_dec <- rd("psd-ml", "decision"); ps_fail <- rd("psd-ml", "failures")
ps_stage <- rd("psd-ml", "twostage_stages"); ps_tol <- rd("psd-ml", "tolerance")
tot <- function(r, arm, col, est = NULL) sum(r[[col]][r$role == "test" & r$arm == arm &
  (is.null(est) | r$estimator %in% est)])
pr <- function(p, cand, base, col) sum(p[[col]][p$role == "test" & p$candidate == cand & p$baseline == base])
f0 <- function(x) format(round(x), big.mark = ",")
```

## Question

Which start, optimizer and PSD route should each estimation route use by
default? Every fit is judged by the library verdict, on simulated problems
the candidates were not developed on. The criteria were committed before the
runs, in [criteria/ml-gls.md](criteria/ml-gls.md) and
[criteria/psd-ml.md](criteria/psd-ml.md) (commit `ec5b0c96`).

## Answer

**ML and GLS: the layered start with PORT is much better than today's
default, but PORT needs ML's budget before it becomes the default.**

- On the test families, today's default (FABIN3 with L-BFGS) certifies
  `r f0(tot(ml_rates, "default", "success", "ML"))` of
  `r f0(tot(ml_rates, "default", "attainable", "ML"))` attainable ML problems,
  and the candidate certifies
  `r f0(tot(ml_rates, "layered_port", "success", "ML"))`.
- For GLS the counts are `r f0(tot(ml_rates, "default", "success", "GLS"))`
  and `r f0(tot(ml_rates, "layered_port", "success", "GLS"))` of
  `r f0(tot(ml_rates, "default", "attainable", "GLS"))`.
- Most of the gain is the start: the layered start with L-BFGS already
  certifies `r f0(tot(ml_rates, "layered_lbfgs", "success", "ML"))` ML and
  `r f0(tot(ml_rates, "layered_lbfgs", "success", "GLS"))` GLS problems.
  PORT from today's start is slightly worse than L-BFGS.

Two pre-registered checks fail, both small:

- **One loss in the Chen family.** A single ML draw, where the default
  certifies all of that family's problems.
- **Invariance under common units** for ML on the Boomsma and Wolf families.
  - The candidate's success flips across units about as often as the
    default's there (91 against 94 draws), and far less often over all test
    families (ML 138 against 537, GLS 19 against 4,321).
  - Its excess there is 41 draws certified in every unit system at objectives
    that differ by $10^{-6}$ to $10^{-5}$ of the objective; the default has 2.
    At $N = 25$ a Newton distance of .01 allows differences of about this
    size, so these are the same minimum reached to slightly different
    precision.

Both trace to one configuration gap: the ML option profile sets NLopt's budget
and tolerances (5000 evaluations, $10^{-12}$ and $10^{-10}$), while PORT runs
with its generic 1000 iterations and its own tolerances. Most of the
candidate's losses end with PORT's budget exhausted. **Recommendation:**
promote the layered start with PORT after PORT gets ML's budget and
tolerances, and confirm that on fresh draws (a new seed base, about ten
minutes).

**PSD ML: keep today's direct configuration; the two-stage route is better.**

- **Rule A fails.** The layered cold start loses in four of the six test
  families.
- **Rule B fails.** Dropping the diagonal preconditioning loses
  `r f0(pr(ps_pair, "psd_default_none", "psd_default", "losses"))` fits
  against `r f0(pr(ps_pair, "psd_default_none", "psd_default", "wins"))` and
  takes 2.4 times as long.
- **The two-stage route beats the direct fit.** It fits ordinary ML first
  with today's default and refits under the PSD constraint only when needed.
  It wins `r pr(ps_pair, "twostage_default", "psd_default", "wins")` and loses
  `r pr(ps_pair, "twostage_default", "psd_default", "losses")` certified fits
  against the direct fit, reaches the best known objective
  `r pr(ps_pair, "twostage_default", "psd_default", "best_wins")` against
  `r pr(ps_pair, "twostage_default", "psd_default", "best_losses")` times, and
  takes `r round(100 * (1 - tot(ps_rates, "twostage_default", "total_seconds") / tot(ps_rates, "psd_default", "total_seconds")))`%
  less time. The author decides the route.
- **Its ordinary stage must not inherit the layered start with PORT.** With
  that ordinary stage the two-stage route is worse than the direct fit: the
  PSD refit succeeds less often when it is warm-started from PORT's rejected
  or improper endpoints than from L-BFGS's.
- **Rule D passes.** The absolute feasibility tolerances do not bite.

Two issues found along the way:

- The preconditioning's absolute clamp does cost fits in large units.
- A lift round-trip error in equality-constrained models occurs in every unit
  system.

## The register

| Default | Current value | Status after this run | Evidence |
|---|---|---|---|
| ML and GLS start | transported FABIN3 (ML), native FABIN3 (GLS) | open: promote the layered start together with the optimizer below | lane ml-gls; engineering/17 corpus (303 + 303 pairs, no loss) |
| ML and GLS optimizer | NLopt L-BFGS | open: promote PORT once it gets ML's budget and tolerances, after a confirmation run on fresh draws | lane ml-gls |
| optimizer coordinates | information-refined sample units | settled (2026-09-26); every arm here used them, and the candidate's success and objective agree across units | engineering/17; lane ml-gls invariance |
| PSD cold start | transported FABIN3 | settled: keep (rule A fails in 4 of 6 families) | lane psd-ml |
| PSD preconditioning | lifted information diagonal | settled: keep (rule B fails in every family) | lane psd-ml |
| PSD route | direct PSD fit | open, for the author: two-stage from today's ordinary default is better and faster | lane psd-ml |
| PSD feasibility tolerances | absolute | settled: keep (rule D passes, 0 fits) | lane psd-ml |
| PSD preconditioning clamp | $[10^{-4}, 10^{4}]$, absolute | open: costs direct PSD fits at $\times 100$ | lane psd-ml failures |

## How much to trust it

- **The problems are held out.**
  - 51 test populations and one control: research/47,
    engineering/15's weak marker and high $R^2$, Boomsma (1985), Wolf et al.
    (2013), Chen et al. (2001) with four fitted models, and ten
    equality-constrained models. The control is Ernst, used in the 2026-09-22
    start work.
  - Each population is drawn at a small $N$ and at $N = 100$: 100
    replications per $N$ for ml-gls, 50 for psd-ml.
  - Every draw is fitted in native units, $\times 100$, $\times 0.01$ and
    separate units.
  - None of these populations shaped the layered start, the coordinates or
    the PSD scaling.
- **One judge.** Success is the library verdict (the Newton check), plus
  admissibility for PSD fits. A problem counts only if some arm, or a PORT
  or PSD fit from the population values, certifies it.
- **The engine.** The `magmaanlab` build from `30514b09`, built 2026-09-26
  04:58 UTC. The engine source is unchanged through the pre-registration
  commit.
- **Limits.**
  - The data are multivariate normal and complete, and all groups are single.
  - FIML, least squares, ordinal routes and lavaan are not covered; the
    textbook corpus is engineering/17's evidence.
  - Counts of fits are descriptive; paired wins and losses on the same draws
    carry the comparisons.

## Lane ml-gls

### Rules

```{r ml-rules}
tab(ml_dec, col.names = c("Rule", "Test family and estimator", "Value", "Pass"))
```

### Certified fits by family

Attainable problems and certified fits, pooled over $N$, units and
replications.

```{r ml-rates}
w <- reshape(ml_rates[c("role", "family", "estimator", "arm", "attainable", "success")],
             idvar = c("role", "family", "estimator"), timevar = "arm", direction = "wide")
w <- w[order(w$role == "control", w$estimator, w$family), ]
tab(data.frame(Family = fam[w$family], Estimator = w$estimator, Attainable = w$attainable.default,
  Default = w$success.default, `Default + PORT` = w$success.default_port,
  `Layered + L-BFGS` = w$success.layered_lbfgs, `Layered + PORT` = w$success.layered_port,
  check.names = FALSE))
```

### The candidate's losses

Every loss is an optimizer termination; none is an error. The columns show
which single-change arm solves the same problem.

```{r ml-losses}
l <- ml_loss[ml_loss$role == "test", ]
l$attribution <- ifelse(grepl("layered_lbfgs:ok", l$arms), "the layered start with L-BFGS solves it",
                 ifelse(grepl("default_port:ok", l$arms), "today's start with PORT solves it",
                        "only today's default solves it"))
tab(as.data.frame.matrix(table(l$attribution, l$estimator)) |>
      (\(m) cbind(Attribution = rownames(m), m))())
tab(as.data.frame.matrix(table(l$optimizer_status, l$estimator)) |>
      (\(m) cbind(`PORT status` = rownames(m), m))())
```

Most losses are draws where the layered start leads both optimizers to a
worse place, and PORT then stops at its 1000-iteration budget. The single
Chen loss is of this kind: model m0 at $N = 25$ in separate units, where PORT
exhausted its budget at an objective of 0.763 against the default's certified
0.665.

### Invariance

Breaks among the draws of the candidate and the baseline, by pattern:

```{r ml-breaks}
b <- ml_breaks[ml_breaks$role == "test" & ml_breaks$arm %in% c("default", "layered_port"), ]
b <- stats::aggregate(groups ~ arm + estimator + units + pattern, b, sum)
tab(b[order(b$estimator, b$units, -b$groups), ], col.names = c("Arm", "Estimator", "Units", "Pattern", "Draws"))
```

The candidate flips between certified and not across units less often than
the default. Its excess on common units in Boomsma and Wolf ML is certified
fits whose objectives differ by $10^{-6}$ to $10^{-5}$ of the objective. That
is consistent with PORT stopping on its generic tolerances rather than ML's
tighter profile; the confirmation run with ML's controls will show it.

## Lane psd-ml

### Rules

```{r psd-rules}
tab(ps_dec, col.names = c("Rule", "Test family", "Value", "Pass"))
```

### Routes

Attainable problems, successes (certified and admissible), best-known
objective reached, and total seconds, on the test families.

```{r psd-routes}
a <- stats::aggregate(cbind(attainable, success, best, total_seconds) ~ arm, ps_rates[ps_rates$role == "test", ], sum)
a$total_seconds <- round(a$total_seconds)
tab(a, col.names = c("Arm", "Attainable", "Success", "Best known", "Seconds"))
```

Which stage returned the two-stage fits, and how often it succeeded:

```{r psd-stages}
st <- stats::aggregate(cbind(fits, success) ~ arm + stage, ps_stage[ps_stage$role == "test", ], sum)
st$rate <- sprintf("%.1f%%", 100 * st$success / st$fits)
tab(st, col.names = c("Arm", "Stage", "Fits", "Success", "Rate"))
```

When the ordinary fit is rejected or improper, the PSD refit warm-starts from
the ordinary endpoint. From L-BFGS endpoints that refit succeeds far more
often than from PORT endpoints. Why is not established; a likely reason is
that PORT, stopping at its budget, ends further along divergent paths.

### Units and failure causes

The pre-registered tolerance check (rule D) finds no fit that fails only in
rescaled units because of link feasibility or admissibility.

```{r psd-tolerance}
tab(ps_tol, col.names = c("Units", "Arm", "Cause", "Fits"))
```

Two other failures depend on the model or the units:

```{r psd-failures}
f <- ps_fail[ps_fail$arm == "psd_default" & ps_fail$cause %in% c("budget exhausted", "link round-trip"), ]
f <- stats::aggregate(fits ~ family + cause + transform, f, sum)
f <- reshape(f, idvar = c("family", "cause"), timevar = "transform", direction = "wide")
names(f) <- sub("fits.", "", names(f), fixed = TRUE); f[is.na(f)] <- 0
tab(f[c("family", "cause", intersect(c("native", "x100", "x0.01", "mixed"), names(f)))])
```

- **Budget exhaustion rises at $\times 100$.** This is most marked in the
  equality-constrained family. The lifted information scale is clamped to
  $[10^{-4}, 10^{4}]$, which binds on variances of order $10^{4}$.
- **The lift round-trip error ("could not round-trip the terminal covariance
  links")** occurs only in equality-constrained models, in native units as
  often as rescaled. It is a defect of the PSD lift with cross-variable
  equalities, not a units effect.

## Reproduce

```sh
export OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1
Rscript run_experiment.R --help
Rscript run_experiment.R --lane ml-gls --smoke
nice -n 10 Rscript run_experiment.R --lane ml-gls --workers 4   # about 10 minutes
nice -n 10 Rscript run_experiment.R --lane psd-ml --workers 4   # about 8 minutes
quarto render report.qmd
```

`--summarize` rescores existing raw rows (about 30 seconds per lane). Raw
rows stay local; the summaries in `results/<lane>/2026-09-26/` are tracked.
