---
title: "Optimizer defaults"
format:
  html:
    toc: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
run <- "2026-09-26"
rd <- function(lane, name) {
  p <- file.path("results", lane, run, paste0(name, ".csv"))
  if (!file.exists(p)) stop("Missing ", p, ": run run_experiment.R --lane ", lane)
  read.csv(p, stringsAsFactors = FALSE)
}
tab <- function(x, ...) knitr::kable(x, row.names = FALSE, format.args = list(big.mark = ","), ...)
fam <- c(research47 = "research/47", engineering15 = "engineering/15", boomsma = "Boomsma 1985",
         wolf = "Wolf et al. 2013", chen = "Chen et al. 2001", constrained = "equality-constrained",
         ernst = "Ernst (control)")
ml_rates <- rd("ml-gls", "rates_by_family"); ml_pair <- rd("ml-gls", "paired")
ml_dec <- rd("ml-gls", "decision"); ml_loss <- rd("ml-gls", "losses")
ml_breaks <- rd("ml-gls", "invariance_breaks")
ps_rates <- rd("psd-ml", "rates_by_family"); ps_pair <- rd("psd-ml", "paired")
ps_dec <- rd("psd-ml", "decision"); ps_fail <- rd("psd-ml", "failures")
ps_stage <- rd("psd-ml", "twostage_stages"); ps_tol <- rd("psd-ml", "tolerance")
rdc <- function(name) { run <<- "2026-09-26-confirm"; on.exit(run <<- "2026-09-26"); rd("ml-gls", name) }
cf_dec <- rdc("decision"); cf_rates <- rdc("rates_by_family"); cf_pair <- rdc("paired"); cf_esc <- rdc("escapes")
ml_esc <- rd("ml-gls", "escapes"); ps_rt <- rd("psd-ml", "rates_by_family_transform")
rdt <- function(name) { run <<- "2026-09-26-third"; on.exit(run <<- "2026-09-26"); rd("ml-gls", name) }
t3_choice <- rdt("choice"); t3_dec <- rdt("decision"); t3_rates <- rdt("rates_by_family")
t3_run <- rdt("runaways"); t3_loss <- rdt("losses"); t3_pair <- rdt("paired")
rd2 <- function(name) { run <<- "2026-09-26-second"; on.exit(run <<- "2026-09-26"); rd("psd-ml", name) }
p2_dec <- rd2("decision"); p2_rates <- rd2("rates_by_family"); p2_pair <- rd2("paired")
p2_rt <- rd2("rates_by_family_transform"); p2_loss <- rd2("losses")
t3 <- function(arm, est, col = "success") tot(t3_rates, arm, col, est)
t3r <- function(arm, est) sum(t3_run$beyond_10[t3_run$role == "test" & t3_run$arm == arm & t3_run$estimator == est])
tot <- function(r, arm, col, est = NULL) sum(r[[col]][r$role == "test" & r$arm == arm &
  (is.null(est) | r$estimator %in% est)])
pr <- function(p, cand, base, col) sum(p[[col]][p$role == "test" & p$candidate == cand & p$baseline == base])
f0 <- function(x) format(round(x), big.mark = ",")
```

## Question

Which start, optimizer and PSD route should each estimation route use by
default? Every fit is judged by the library verdict, on simulated problems
the candidates were not developed on. The criteria were committed before the
runs, in [criteria/ml-gls.md](criteria/ml-gls.md) and
[criteria/psd-ml.md](criteria/psd-ml.md) (commit `ec5b0c96`).

## Answer

**ML and GLS: the layered start is the default since 2026-09-26, and the
optimizer stays L-BFGS (the author's decision after the third run).** With
runaways scored as failures, the layered start wins for both estimators. PORT
adds certified fits, but also many runaway fits that the library still calls
converged. PORT therefore waits for a runaway check in the verdict, the next
optimization project.

- A certified fit whose standardized extent exceeds 10 now counts as a
  runaway, a failure. The extent is the largest absolute standardized
  loading, latent path, latent correlation or residual/disturbance ratio. It
  is unit-free, and without this rule PORT's persistence was being rewarded.
- **GLS: the pre-registered choice is the layered start with PORT.** Both
  candidates pass every rule. Certified fits: today
  `r f0(t3("default", "GLS"))`, layered with L-BFGS
  `r f0(t3("layered_lbfgs", "GLS"))`, layered with PORT
  `r f0(t3("layered_port", "GLS"))`, of `r f0(t3("default", "GLS", "attainable"))`.
- **ML: no candidate is eligible, so the author decides.** Both candidates
  pass every rule except in the Chen family.
  - Certified fits: today `r f0(t3("default", "ML"))`, layered with L-BFGS
    `r f0(t3("layered_lbfgs", "ML"))`, layered with PORT
    `r f0(t3("layered_port", "ML"))`, of `r f0(t3("default", "ML", "attainable"))`.
  - Chen fails rule 1 because today's default certifies every attainable Chen
    problem, so any loss fails. The losses are 5 of about 800 draws (model m0
    at $N = 25$, 4 draws; model m1, 1 draw), each counted in all four unit
    systems. On those draws the layered start sends both optimizers to a
    worse basin (objective 0.09–0.47 higher), ending at saddles or
    ill-conditioned points.
- **What users would see.** The runaway rule is in the scoring, not in the
  library's verdict. Among certified test-family fits, runaways number:

  | Estimator | Today | Layered + L-BFGS | Layered + PORT |
  |---|---:|---:|---:|
  | ML | `r f0(t3r("default", "ML"))` | `r f0(t3r("layered_lbfgs", "ML"))` | `r f0(t3r("layered_port", "ML"))` |
  | GLS | `r f0(t3r("default", "GLS"))` | `r f0(t3r("layered_lbfgs", "GLS"))` | `r f0(t3r("layered_port", "GLS"))` |

  PORT would hand users several times as many runaway fits marked as
  converged, unless the verdict gets a runaway check.
- Runs 1 and 2 kept their pre-registered scoring and failed their rules. Run 2
  exposed the runaway problem.
- **The shipped default is the evaluated candidate.** After the promotion, a
  run of the library's `default` arm on the third run's draws reproduces
  `layered_lbfgs` exactly on all 92,000 fits: the same certification and the
  same objective (`results/ml-gls/2026-09-26-library/reproduction.csv`).

**PSD ML: keep FABIN3 in both PSD routes, confirmed after the ML/GLS
promotion (second run).** The layered start does not carry over to PSD. In
the second run, on fresh draws with the runaway rule, it fails rule A again
(direct PSD fits, five of six test families) and rule E (the fallback's
ordinary step, four of six). Most of its losses exhaust SLSQP's budget. So
`frontier_fit_ml_psd_fallback` keeps its ordinary step on FABIN3 with L-BFGS,
although complete-data ML's own default is now the layered start. The route
result replicates: two-stage is better overall, and its whole advantage comes from rescaled units.

The first run's findings:

- **Rule A fails.** The layered cold start loses in four of the six test
  families.
- **Rule B fails.** Dropping the diagonal preconditioning loses
  `r f0(pr(ps_pair, "psd_default_none", "psd_default", "losses"))` fits
  against `r f0(pr(ps_pair, "psd_default_none", "psd_default", "wins"))` and
  takes 2.4 times as long.
- **The two-stage route beats the direct fit.** It fits ordinary ML first
  with today's default and refits under the PSD constraint only when needed.
  It wins `r pr(ps_pair, "twostage_default", "psd_default", "wins")` and loses
  `r pr(ps_pair, "twostage_default", "psd_default", "losses")` certified fits
  against the direct fit, reaches the best known objective
  `r pr(ps_pair, "twostage_default", "psd_default", "best_wins")` against
  `r pr(ps_pair, "twostage_default", "psd_default", "best_losses")` times, and
  takes `r round(100 * (1 - tot(ps_rates, "twostage_default", "total_seconds") / tot(ps_rates, "psd_default", "total_seconds")))`%
  less time. The advantage is mostly a units effect, though. In native units
  the direct fit certifies slightly more
  (`r f0(sum(ps_rt$success[ps_rt$role == "test" & ps_rt$transform == "native" & ps_rt$arm == "psd_default"]))`
  against
  `r f0(sum(ps_rt$success[ps_rt$role == "test" & ps_rt$transform == "native" & ps_rt$arm == "twostage_default"]))`).
  Two-stage wins at $\times 100$ and $\times 0.01$, where the absolute preconditioning clamp was a suspected cause. The
  follow-up below tests that hypothesis; the earlier comparison did not isolate it. The author kept the direct fit for now (2026-09-27); the
  clamp fix and a route rerun are the priority follow-up.
- **Its ordinary stage must not inherit the layered start with PORT.** With
  that ordinary stage the two-stage route is worse than the direct fit: the
  PSD refit succeeds less often when it is warm-started from PORT's rejected
  or improper endpoints than from L-BFGS's.
- **Rule D passes.** The absolute feasibility tolerances do not bite.

Two issues found along the way:

- Direct PSD fits exhaust their budget more often after rescaling, in the
  pattern the preconditioning's absolute clamp predicts.
- A lift round-trip error hits the std.lv equal-loadings CFA (about 4% of
  fits at $N = 25$).

The PSD coordinate-scale clamp has since been made relative to measurement
units. A small paired follow-up tests that correction and repeats the route
comparison with explicit, unchanged starts. The coordinate tests pass, but direct PSD still certifies 1,328/1,356 fits:
eight gains offset eight losses. With the corrected scales, two-stage certifies
1,332 and matches more best observed objectives (1,326 versus 1,298), while
direct remains slightly ahead in native-unit certification (351 versus 350).
The [PSD scaling follow-up](#psd-scaling-follow-up) retains every paired loss. This diagnostic does not promote
a route or introduce multistart.

A later unconstrained pilot tries full normalization before starts and fitting.
It removes all unit-dependent verdict and objective differences for direct
marker PSD in this sample, with exact backtransformation to numerical precision.
It does not uniformly improve the solutions found. A subsequent accuracy-only
fix resolves five unit-induced failures without refitting; all 4,742 saved
points now agree across original and normalized representations. See the
[unconstrained normalization pilot](#unconstrained-normalization-pilot);
full fitting normalization is not promoted by this experiment.

After this pilot, the author requested enabling shared normalization throughout
complete-data ML and PSD fitting (2026-09-27), including multiple groups and
linear equality constraints. That implementation is checked by unit/fixture and
R integration tests. This is an explicit adoption decision, not a new held-out
run establishing better basin selection; the mixed pilot results remain relevant.

## The register

| Default | Current value | Current status | Evidence |
|---|---|---|---|
| ML and GLS start | the layered moment start (formerly transported FABIN3 for ML, native FABIN3 for GLS) | settled 2026-09-26 (author, after run 3). Named losses: 5 Chen draws at $N = 25$, and one real-data model found by the R examples (Geiser's latent AR cross-lagged model, marker identification; backlog) | lane ml-gls, third run |
| ML and GLS optimizer | NLopt L-BFGS | provisional: PORT certifies more but returns about eight times as many runaways; reopen once the verdict flags runaways | lane ml-gls, runs 2 and 3 |
| verdict on runaway paths | the Newton check accepts divergent-path points | open: scoring has a runaway rule (extent > 10); the library verdict does not | lane ml-gls, runs 2 and 3 |
| optimizer coordinates | information-refined sample units | settled (2026-09-26); every arm here used them, and the candidate's success and objective agree across units | engineering/17; lane ml-gls invariance |
| PSD cold start | transported FABIN3 | settled: keep (rule A fails in 4 of 6 families, and in 5 of 6 in the second run) | lane psd-ml, runs 1 and 2 |
| PSD fallback ordinary step | transported FABIN3 with L-BFGS, pinned apart from the ML default | settled: keep (rule E fails in 4 of 6 families) | lane psd-ml, second run |
| PSD preconditioning | lifted information diagonal | settled: keep (rule B fails in every family) | lane psd-ml |
| PSD route | direct PSD fit (`psd = TRUE`); two-stage is `frontier_fit_ml_psd_fallback` | settled for now (author, 2026-09-27): direct. Two-stage is better and faster overall only in rescaled units (replicated), where the clamp below is the suspect; in native units direct is slightly ahead. The clamp correction and a small paired route rerun are now recorded below; remaining unit-sensitive failures are still open | lane psd-ml, runs 1 and 2 |
| ML and PSD fitting normalization | sample-normalized model and starts; estimates returned in requested units | enabled by author request, 2026-09-27; complete-data, single/multigroup, linear equalities | exploratory pilot below plus library regression checks; no claim of uniformly better local solutions |
| PSD feasibility tolerances | absolute in the internal normalized units | numerical threshold unchanged; its units now follow the fitting normalization | lane psd-ml and subsequent author-requested normalization |
| PSD preconditioning clamp | $[10^{-4}u, 10^{4}u]$, relative to sample-derived units | implemented; further normalization now precedes fitting | PSD scaling follow-up below |

## How much to trust it

- **The problems are held out.**
  - 51 test populations and one control: research/47,
    engineering/15's weak marker and high $R^2$, Boomsma (1985), Wolf et al.
    (2013), Chen et al. (2001) with four fitted models, and ten
    equality-constrained models. The control is Ernst, used in the 2026-09-22
    start work.
  - Each population is drawn at a small $N$ and at $N = 100$: 100
    replications per $N$ for ml-gls, 50 for psd-ml.
  - Every draw is fitted in native units, $\times 100$, $\times 0.01$ and
    separate units.
  - None of these populations shaped the layered start, the coordinates or
    the PSD scaling.
- **One judge.** Success is the library verdict (the Newton check), plus
  admissibility for PSD fits. A problem counts only if some arm, or a PORT
  or PSD fit from the population values, certifies it.
- **The engine.** The `magmaanlab` build from `30514b09`, built 2026-09-26
  04:58 UTC. The engine source is unchanged through the pre-registration
  commit.
- **Limits.**
  - The data are multivariate normal and complete, and all groups are single.
  - FIML, least squares, ordinal routes and lavaan are not covered; the
    textbook corpus is engineering/17's evidence.
  - Counts of fits are descriptive; paired wins and losses on the same draws
    carry the comparisons.

## Lane ml-gls

### Rules

```{r ml-rules}
tab(ml_dec, col.names = c("Rule", "Test family and estimator", "Value", "Pass"))
```

### Certified fits by family

Attainable problems and certified fits, pooled over $N$, units and
replications.

```{r ml-rates}
w <- reshape(ml_rates[c("role", "family", "estimator", "arm", "attainable", "success")],
             idvar = c("role", "family", "estimator"), timevar = "arm", direction = "wide")
w <- w[order(w$role == "control", w$estimator, w$family), ]
tab(data.frame(Family = fam[w$family], Estimator = w$estimator, Attainable = w$attainable.default,
  Default = w$success.default, `Default + PORT` = w$success.default_port,
  `Layered + L-BFGS` = w$success.layered_lbfgs, `Layered + PORT` = w$success.layered_port,
  check.names = FALSE))
```

### The candidate's losses

Every loss is an optimizer termination; none is an error. The columns show
which single-change arm solves the same problem.

```{r ml-losses}
l <- ml_loss[ml_loss$role == "test", ]
l$attribution <- ifelse(grepl("layered_lbfgs:ok", l$arms), "the layered start with L-BFGS solves it",
                 ifelse(grepl("default_port:ok", l$arms), "today's start with PORT solves it",
                        "only today's default solves it"))
tab(as.data.frame.matrix(table(l$attribution, l$estimator)) |>
      (\(m) cbind(Attribution = rownames(m), m))())
tab(as.data.frame.matrix(table(l$optimizer_status, l$estimator)) |>
      (\(m) cbind(`PORT status` = rownames(m), m))())
```

Most losses are draws where the layered start leads both optimizers to a
worse place, and PORT then stops at its 1000-iteration budget. The single
Chen loss is of this kind: model m0 at $N = 25$ in separate units, where PORT
exhausted its budget at an objective of 0.763 against the default's certified
0.665.

### Invariance

Breaks among the draws of the candidate and the baseline, by pattern:

```{r ml-breaks}
b <- ml_breaks[ml_breaks$role == "test" & ml_breaks$arm %in% c("default", "layered_port"), ]
b <- stats::aggregate(groups ~ arm + estimator + units + pattern, b, sum)
tab(b[order(b$estimator, b$units, -b$groups), ], col.names = c("Arm", "Estimator", "Units", "Pattern", "Draws"))
```

The candidate flips between certified and not across units less often than
the default. Its excess on common units in Boomsma and Wolf ML is certified
fits whose objectives differ by $10^{-6}$ to $10^{-5}$ of the objective. That
is consistent with PORT stopping on its generic tolerances rather than ML's
tighter profile; the confirmation run with ML's controls will show it.

### Confirmation run

Seed base 2026092602, candidate `layered_port_ml`, with criteria in the dated
section of [criteria/ml-gls.md](criteria/ml-gls.md). Failing rules:

```{r cf-rules}
tab(cf_dec[!cf_dec$pass, ], col.names = c("Rule", "Test family and estimator", "Value", "Pass"))
```

Native units, test families, fresh draws: certified fits, runaway
certifications among them (largest $\lvert\theta\rvert > 100$; every
population has unit-order parameters), and certified fits without them.

```{r cf-escapes}
e <- stats::aggregate(cbind(fits, escapes, success, success_no_escape) ~ estimator + arm,
                      cf_esc[cf_esc$role == "test", ], sum)
tab(e[order(e$estimator, e$arm), ], col.names = c("Estimator", "Arm", "Fits", "Runaways", "Certified", "Certified, no runaways"))
```

- L-BFGS never certifies a runaway. PORT does, from either start.
- The PORT runaways that the 1000-iteration budget certified sit at a median
  largest parameter of about 9,600 (10th percentile 140). With 5000
  iterations PORT runs further, and the verdict rejects most of them. That is
  why the larger budget "lost" ML fits.
- Without runaways, on ML the layered start with L-BFGS and with PORT gain
  little over today's default. On GLS both gain several hundred fits.
- Both the pre-registered rules and the attainability witnesses count
  runaway certifications, so the next ML/GLS run needs a runaway rule
  written into its criteria.

### Third run: the runaway rule

Seed base 2026092603, candidates `layered_lbfgs` and `layered_port`, and the
runaway rule at 10. Criteria are in the third dated section of
[criteria/ml-gls.md](criteria/ml-gls.md).

```{r t3-choice}
tab(t3_choice, col.names = c("Estimator", "Eligible", "Certified", "PORT no net loss vs L-BFGS", "Choice"))
tab(t3_dec[!t3_dec$pass, ], col.names = c("Candidate", "Rule", "Test family and estimator", "Value", "Pass"))
```

Certified fits beyond each runaway bound (all units; the extent is unit-free):

```{r t3-runaways}
w <- stats::aggregate(cbind(certified, beyond_5, beyond_10, beyond_100) ~ estimator + arm,
                      t3_run[t3_run$role == "test", ], sum)
tab(w[order(w$estimator, w$arm), ], col.names = c("Estimator", "Arm", "Certified", "> 5", "> 10", "> 100"))
```

## Lane psd-ml

### Rules

```{r psd-rules}
tab(ps_dec, col.names = c("Rule", "Test family", "Value", "Pass"))
```

### Routes

Attainable problems, successes (certified and admissible), best-known
objective reached, and total seconds, on the test families.

```{r psd-routes}
a <- stats::aggregate(cbind(attainable, success, best, total_seconds) ~ arm, ps_rates[ps_rates$role == "test", ], sum)
a$total_seconds <- round(a$total_seconds)
tab(a, col.names = c("Arm", "Attainable", "Success", "Best known", "Seconds"))
```

Which stage returned the two-stage fits, and how often it succeeded:

```{r psd-stages}
st <- stats::aggregate(cbind(fits, success) ~ arm + stage, ps_stage[ps_stage$role == "test", ], sum)
st$rate <- sprintf("%.1f%%", 100 * st$success / st$fits)
tab(st, col.names = c("Arm", "Stage", "Fits", "Success", "Rate"))
```

When the ordinary fit is rejected or improper, the PSD refit warm-starts from
the ordinary endpoint; when the ordinary fit errors, it starts from the
original start. From L-BFGS endpoints that refit succeeds far more often than
from PORT endpoints.

A probe on 30 draws where the layered-start PORT two-stage fails and the
direct fit succeeds (native units, mostly $N = 25$ with weak loadings) shows
why. On these draws the ordinary likelihood keeps improving along a Heywood
path: a residual variance goes negative while loadings grow.

- **L-BFGS** usually aborts early: 18 of 30 fits end in a line-search error
  after about 100 evaluations. The PSD refit then starts cold from FABIN3 and
  succeeds.
- **PORT** follows the path for its whole budget: 21 of 30 fits exhaust it.
  It ends far out, with a median most-negative variance of $-33$ and a median
  largest parameter of 211; the worst case reaches $-2543$. No PSD refit
  from these endpoints succeeds.

So PORT's persistence, an advantage when a minimum exists, turns into a poor
warm start when none does.

### Units and failure causes

The pre-registered tolerance check (rule D) finds no fit that fails only in
rescaled units because of link feasibility or admissibility.

```{r psd-tolerance}
tab(ps_tol, col.names = c("Units", "Arm", "Cause", "Fits"))
```

Two other failures depend on the model or the units:

```{r psd-failures}
f <- ps_fail[ps_fail$arm == "psd_default" & ps_fail$cause %in% c("budget exhausted", "link round-trip"), ]
f <- stats::aggregate(fits ~ family + cause + transform, f, sum)
f <- reshape(f, idvar = c("family", "cause"), timevar = "transform", direction = "wide")
names(f) <- sub("fits.", "", names(f), fixed = TRUE); f[is.na(f)] <- 0
tab(f[c("family", "cause", intersect(c("native", "x100", "x0.01", "mixed"), names(f)))])
```

- **Budget exhaustion rises after rescaling, at both ends of the clamp.**
  The lifted information scale $s_k = 1/\sqrt{I_{kk}}$ is clamped to
  $[10^{-4}, 10^{4}]$, and a variance of size $v$ wants $s \approx v$.
  - The chain SEM with equal paths goes from 0 exhausted fits in native units
    to 53 at $\times 100$, where its variances are near $10^{4}$.
  - With equal disturbances too, it goes to 54 at $\times 0.01$, where the
    disturbance variances are near $4 \cdot 10^{-5}$.
  - The pattern fits the clamp but is not yet proven.
- **The lift round-trip error ("could not round-trip the terminal covariance
  links")** occurs only in the two std.lv equal-loadings CFAs, mostly with
  loadings .4. It hits about 4% of their fits at $N = 25$ and 0.5% at
  $N = 100$. It occurs in native and $\times 0.01$ units about equally and
  never at $\times 100$. The mechanism is not yet known.

### Second run: after the ML/GLS promotion

Pre-registered in the dated section of
[criteria/psd-ml.md](criteria/psd-ml.md) (commit `d0413f2c`), on seed base
2026092604 with the runaway rule. The question: now that complete-data ML
starts from the layered start, should the PSD routes follow?

```{r psd2-rules}
tab(p2_dec[c("rule", "scope", "value", "pass")],
    col.names = c("Rule", "Test family", "Value", "Pass"))
```

```{r psd2-routes}
a <- stats::aggregate(cbind(attainable, success, best, total_seconds) ~ arm, p2_rates[p2_rates$role == "test", ], sum)
a$total_seconds <- round(a$total_seconds)
tab(a, col.names = c("Arm", "Attainable", "Success", "Best known", "Seconds"))
```

- **Neither rule passes, so nothing changes.** Rule A fails in five of six
  test families and rule E in four. The equality-constrained family is the
  exception, and a large one: there the layered start avoids most of
  FABIN3's exhausted budgets (direct PSD fits: 203 exhausted with FABIN3, 23
  with the layered start). That family alone gives `psd_layered` the higher
  total.
- **The losses are mostly exhausted budgets.** Of the named losses in
  `losses.csv`:
  - rule A: 231 exhaust SLSQP's budget and 46 stop at a saddle (nonpositive
    curvature), out of 301;
  - rule E: 266 of 344 exhaust the budget.
- **Where rule E loses.** 153 of its losses come from draws where the ordinary
  step errors. The PSD refit then starts cold from the ordinary step's own
  start, now the layered start, which rule A rejects for PSD. The other
  losses are warm starts from rejected (115) or improper (76) layered-start
  endpoints.
- **No PSD fit is a runaway.** Every certified PSD fit, in every arm, has a
  standardized extent of at most 10. So the runaway rule changes nothing in
  this lane.
- **The route result replicates.** Two-stage (with the FABIN3 ordinary step)
  certifies more than the direct fit and takes about 40% less time. In native
  units the direct fit is again slightly ahead
  (`r f0(sum(p2_rt$success[p2_rt$role == "test" & p2_rt$transform == "native" & p2_rt$arm == "psd_default"]))`
  against
  `r f0(sum(p2_rt$success[p2_rt$role == "test" & p2_rt$transform == "native" & p2_rt$arm == "twostage_default"]))`);
  two-stage wins after rescaling.

A probe before the pre-registration, on the covariance-honest paper's own
designs, pointed the same way (disclosed in the criteria): with marker
identification the layered ordinary step lost 33 fits and won 2.

## PSD scaling follow-up

The absolute clamp is a real coordinate-scaling defect, but correcting it alone
does not make the fitting routes interchangeable. The change keeps the same
clamp range relative to each parameter's measurement units. It also makes the
zero-information fallback follow those units. Existing information scales are
left numerically unchanged when the clamp and fallback are inactive. Estimated
variances and PSD boundary solutions are not clamped by this change.

The criteria were committed before the runs in
[criteria/psd-scale-followup.md](criteria/psd-scale-followup.md). All test
populations contribute three replications at each sample size, with a new seed
base, giving 1,356 problems per route after model and unit transformations.
Both versions use FABIN3 with automatic transport, SLSQP and diagonal
preconditioning for direct PSD, and FABIN3 with automatic transport and L-BFGS
for the ordinary first step of the two-stage route. A rejected or inadmissible
ordinary fit triggers the existing SLSQP PSD fallback. Budgets and tolerances
are fixed. The same samples are used before and after the scale change.

```{r}
pscale <- read.csv(file.path("results", "psd-scale", "comparison", "summary.csv"))
psum <- function(scale, arm, field) sum(pscale[[field]][pscale$scale == scale & pscale$arm == arm])
units <- c("native", "x0.01", "x100", "mixed")
pcol <- function(scale, arm, field) {
  z <- pscale[pscale$scale == scale & pscale$arm == arm, ]
  z[[field]][match(units, z$transform)]
}
tab(data.frame(Units=units, Problems=pcol("relative", "direct", "fit_count"),
  `Direct before`=pcol("absolute", "direct", "success"),
  `Direct after`=pcol("relative", "direct", "success"),
  `Two-stage before`=pcol("absolute", "two_stage", "success"),
  `Two-stage after`=pcol("relative", "two_stage", "success"),check.names=FALSE))
```

After the change, direct PSD certifies `r psum("relative", "direct", "success")`
fits and two-stage certifies `r psum("relative", "two_stage", "success")`.
Their counts matching the best certified objective among the four paired
attempts are `r psum("relative", "direct", "hit")` and
`r psum("relative", "two_stage", "hit")`. These are observed comparison targets,
not independently established global optima. Paired gains, losses, objective
changes and population breakdowns are retained in `results/psd-scale/comparison/`.
The failure records retain budget exhaustion, accuracy rejection and covariance
round-trip failures rather than collapsing them into one cause.

The final implementation applies units directly to the clamp bounds. An initial,
mathematically equivalent divide-and-multiply implementation was also inspected
locally, but introduced unnecessary rounding changes even for unclamped scales;
its results are not the final comparison above. Both implementations implement
the criterion's same relative range. This numerical refinement makes the final
comparison exploratory, not an untouched confirmation run.

Tests isolate the coordinate correction by transporting the same starting point
and keeping its covariance projection inactive. They cover marker and std.lv
identification, correlated residuals, equality constraints, zero-information
loadings, and uniform and mixed changes of observed units. A fitted PSD boundary
case also preserves its objective and covariance under ×0.01 and ×100 units.
All 535 estimation tests pass.

One retained example makes the remaining problem concrete. For the equal-path,
equal-disturbance chain with coefficient 0.2, N=25 draw 3, direct PSD certifies
objective 1.555299 in native units and 2.198990 at ×100. Two-stage reaches
1.555299 at ×100. The issue therefore includes the destination reached, not just
whether the optimizer reports success.

Remaining unit sensitivity has other possible sources. The start covariance
projection still uses an absolute eigenvalue floor, and covariance-link residuals
retain physical units. Those facts do not establish the cause of each remaining
failure. The next diagnostic should separate them on the retained losses before
a larger route decision. Spectral starts remain a separate banked candidate;
this comparison does not change the start recipe or add multistart.

The run is small (three draws per population/N), and transformations share the
same draws. Timing is advisory: compilations overlapped portions of the runs,
so the before/after elapsed times cannot establish a clamp speedup. Both routes'
times are retained for inspection. The direct route remains the ordinary-user
policy and two-stage remains an explicit frontier call.

## Unconstrained normalization pilot

Full sample normalization improves consistency across measurement units, most
clearly for direct marker-identified PSD fitting. It is not yet an across-the-board
improvement in which local solution is reached, and post-fit accuracy checks
were subsequently made consistent in the scoped accuracy check described below.

The pilot divides each observed variable by its sample standard deviation,
constructs FABIN3-auto starts and fits in those normalized units, then transforms
the parameters back. It retains the existing marker indicator; std.lv models
are tested separately. There are no equality constraints or arbitrary nonzero
fixed values. Correlated residuals and cross-loadings are included. A separate
check verifies mean transport. This is a study-local prototype, not an R
implementation of a new estimator or a change to the production fitter.

The criteria were committed in
[criteria/psd-normalization-pilot.md](criteria/psd-normalization-pilot.md) before
the runs. Three fresh draws per population/N produce 300 fitted model/sample
combinations. Each is fitted under four observed-unit transformations, two
identifications, two routes and two normalization choices: 9,600 attempts.
Both choices use the same explicit FABIN3-auto, L-BFGS ordinary and SLSQP PSD
configurations as the scale follow-up, with their starts constructed in their
respective fitting units. Every transformed sample is normalized and fitted
independently; no endpoint is reused to manufacture agreement. Because starts
are constructed after normalization, this tests the proposed fitting workflow
as a whole; it does not isolate changed starts from changed optimizer behavior.
Unit comparisons stay within each identification.

```{r}
ni <- read.csv(file.path("results", "psd-normalization", "pilot", "invariance.csv"))
a <- ni[ni$arm == "current", ]; b <- ni[ni$arm == "normalized", ]
w <- merge(a, b, by=c("chart", "route"), suffixes=c("_current", "_normalized"))
tab(data.frame(Identification=w$chart, Route=w$route,
  `Verdict differences before`=w$verdict_changed_current,
  `Verdict differences normalized`=w$verdict_changed_normalized,
  `Objective differences before`=w$objective_mismatch_current,
  `Objective differences normalized`=w$objective_mismatch_normalized,
  check.names=FALSE))
```

Each row represents 900 native-versus-rescaled comparisons. A verdict difference
means only one member succeeds. Objective differences are counted only when
both members succeed, using tolerance 1e-6 times one plus the native objective's
absolute value. Direct marker PSD also has no fitted-covariance differences
above the specified 1e-5 threshold after normalization. The normalized sample
matrices themselves differ by at most 4.45e-16 across unit transformations;
remaining discrepancies therefore expose sensitivity to rounding as well as
route selection, rather than a large discrepancy in the normalized input.

Backtransformation preserves every returned normalized point. The largest
covariance discrepancy, measured relative to sample standard deviations, is
5.94e-16; the largest objective discrepancy is 2.71e-13. The guards reject
repeated parameter labels and unsupported fixed values instead of fitting a
changed model silently.

The practical gains are mixed. Direct marker PSD's internal success count rises
from 1,179 to 1,184 out of 1,200 fits. Its best-observed objective matches change
from 1,148 to 1,146. Two-stage marker matches fall from 1,178 to 1,149 despite
unchanged internal success count (1,182). These references are the best certified
objectives observed among the four attempts within each identification and unit
condition; they are not global optima. Normalization can make a fitter more
consistent while consistently sending it to a less favorable local solution.
Every paired gain/loss and unit exception is retained in the pilot results.

With the previous accuracy check, five normalized marker fits pass internally but fail the PSD accuracy check
when evaluated in original units. We rechecked those exact points without any
optimization: all pass in normalized and native units, while the rescaled
checks report ill-conditioning. Condition measures increase from about
1e3–1e6 to 1e13–1e15. In two mixed-unit cases the detected covariance nullity
also changes. Covariance and objective transport still agree. This identifies
unit dependence in the diagnostics, not a changed fitted point. The table
above uses the internal verdict plus successful transport and original-unit
PSD admissibility; `output_success` in the summaries additionally requires
the original-unit PSD accuracy check, reducing normalized marker counts to
1,182 direct and 1,179 two-stage. These verdicts remain explicitly separate.

Remaining two-stage sensitivity and worse local solutions still need
inspection. Equality-constrained models, general fixed-value transport,
multi-group models and inference are outside this pilot. The small number of
draws and their repeated unit transformations limit any policy claim; no
fitting normalization or route change is adopted here.

### Accuracy checks in normalized units

The scoped PSD accuracy check now normalizes each supplied point and its sample
before assessing boundary geometry and accuracy. It preserves the requested
marker and uses the same accuracy thresholds. Originally single-group PSD,
this now extends to ordinary and PSD complete-data ML with multiple groups,
equal labels and general linear equalities.
Estimates remain in requested units. The equality extension is covered by
regression tests of equivalent models in mixed units, including boundary and
inaccurate points; the saved-point pilot below contains no equality constraints.

We re-audited all 4,742 returned normalized-pilot points, without fitting.
All five earlier false failures now pass. Original and normalized
representations agree on every verdict and on the numbers of null and held
boundary directions. Two original-unit checks newly fail with nonpositive
curvature; these are two unit representations of the Boomsma 4US, N=25,
replicate 3, marker two-stage case. Both fits already failed their internal
accuracy check, so no previously successful fit is lost.

```{r}
a <- read.csv(file.path("results", "psd-accuracy-normalization", "points", "summary.csv"))
tab(data.frame(
  Identification=a$chart, Route=a$route, `Saved points`=a$points,
  `New passes`=a$gain, `New rejections`=a$loss,
  `Unit verdict disagreements`=a$verdict_disagreement,
  `Unit geometry disagreements`=a$geometry_disagreement,
  check.names=FALSE))
```

These are development regression checks of the diagnostic, not fresh draws or
a new comparison of fitting routes. The original pilot summaries remain frozen;
the separate accuracy results retain every changed verdict. Other fit checks
and the choice of local solution can still depend on units.

## Reproduce

For the unconstrained normalization pilot, use the library with the relative
PSD scale fix (`9c6146a1`). Raw fits and parameter/covariance rows stay local;
small summaries, exceptions and accuracy rechecks are retained for rendering.

```sh
Rscript scripts/psd_normalization.R --help
Rscript scripts/psd_normalization.R --smoke
Rscript scripts/psd_normalization.R
Rscript scripts/summarize_psd_normalization.R
Rscript scripts/inspect_psd_normalization.R
Rscript scripts/check_psd_normalization.R
```


For the scale follow-up, run the same script against the pre-fix and fixed
library builds, retaining separate result IDs; do not generate both IDs with
the same library. Use an isolated R library for the changed build. The final
paired summaries are frozen, so rendering does not require raw fit files.

```sh
Rscript scripts/psd_scale_followup.R --help
Rscript scripts/psd_scale_followup.R --run-id absolute-controls  # pre-fix library
Rscript scripts/psd_scale_followup.R --run-id relative-final     # fixed library
Rscript scripts/summarize_psd_scale.R
```


```sh
export OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1
Rscript run_experiment.R --help
Rscript run_experiment.R --lane ml-gls --smoke
nice -n 10 Rscript run_experiment.R --lane ml-gls --workers 4   # about 10 minutes
nice -n 10 Rscript run_experiment.R --lane psd-ml --workers 4   # about 13 minutes
nice -n 10 Rscript run_experiment.R --lane psd-ml --workers 4 --seed-base 2026092604 \
  --run-id 2026-09-26-second --candidate psd_layered --runaway-bound 10 \
  --arms psd_default,psd_layered,twostage_default,twostage_layered_lbfgs,witness   # about 7 minutes
quarto render report.qmd
```

`--summarize` rescores existing raw rows (about 30 seconds per lane). Raw
rows stay local; the summaries in `results/<lane>/2026-09-26/` are tracked.

For the saved-point accuracy regression, install the normalized PSD accuracy
build and run `Rscript scripts/recheck_psd_accuracy.R --smoke`, then
`Rscript scripts/recheck_psd_accuracy.R`. It reads the local pilot fit and
parameter files and regenerates samples from their recorded seeds; it never
runs an optimizer. Summary, metadata and changed verdicts are retained under
`results/psd-accuracy-normalization/points/`; full point rows remain local.
