---
title: "DWLS threshold-invariance calibration"
format:
  html:
    toc: true
execute:
  echo: false
---

## Default register

| Choice | Current value | Status | Evidence |
|---|---|---|---|
| Threshold-step reference | Moment-nested policy All; SB/PEBA4 comparators | Adopted 2026-10-06 (board TASK-81); confirmed on fresh draws | Production and confirmation 2026-10-05 (576,000 + 288,000 draws) |

**Decision (2026-10-06):** Adopt All for every DWLS nested spectral test; retain SB/PEBA4 as comparators (TASK-81).

## Question

Does the DWLS configural-to-threshold-invariant test keep nominal size when the
larger model is correct or misspecified?

## Short answer

**Not with SB or PEBA4 once items have many categories.** Their null rejection
rises with the number of categories, that is with the number of threshold
restrictions: at N ≥ 500 per group the SB medians are about 5.0% with four
categories, 5.5% with five and 7% with seven. The registered rule flags 22
policy arm-cells, all with seven categories (SB 7.05–7.95%, PEBA4 7.05–7.5%).
The pattern is the same with a correct or a misspecified larger model and with
symmetric or skewed cuts.

**The All reference stays calibrated everywhere.** The exact weighted
chi-square tail on every positive eigenvalue of the same spectrum rejects
3.5–6.0% in every null cell, at every N, category count and misspecification.
Size-adjusted power is the same for all three arms (median 91%). This mirrors
the global DWLS test, where the user chose All on 2026-10-03.

**All passes its fresh-draw confirmation.** As registered, flags call for
investigation and fresh confirmation, never a silent recipe switch. A dated
post-production amendment ([criteria](criteria/threshold_invariance.md))
reran the 144 null cells with four or more categories on fresh draws, with
All's criterion fixed in advance: rejection within [3%, 7%] at N ≥ 500 per
group. All rejected 3.45–6.0% in every cell, at every N; SB again exceeded 7%
in 13 cells and PEBA4 in 3, all with five or seven categories. The user adopted
All for nested DWLS on 2026-10-06.

```{r}
prod <- file.path("results", "production-2026-10-05")
if (!file.exists(file.path(prod, "summary.csv")))
  stop("Production summaries absent: modal run modal/app.py --mode production --run-id production-2026-10-05")
s <- merge(read.csv(file.path(prod, "summary.csv")), read.csv(file.path(prod, "cells.csv")), by = "cell_id")
x <- s[s$role == "null" & s$categories > 3 & s$arm %in% c("policy_sb", "policy_peba4", "all"), ]
f <- function(v) sprintf("%.1f (%.1f–%.1f)", 100 * median(v), 100 * min(v), 100 * max(v))
w <- aggregate(rate ~ categories + n + arm, x, f)
w <- reshape(w, idvar = c("categories", "n"), timevar = "arm", direction = "wide")
w <- w[order(w$categories, w$n), ]
knitr::kable(data.frame(Categories = w$categories, `N per group` = w$n,
  `Policy SB (%)` = w$rate.policy_sb, `Policy PEBA4 (%)` = w$rate.policy_peba4,
  `All (%)` = w$rate.all, check.names = FALSE),
  caption = "Null rejection at nominal 5%: median (range) over the 16 cells per row (one or two factors, symmetric or skewed cuts, correct or misspecified larger model, delta or theta), 2,000 draws per cell. Per-cell Wilson intervals are in summary.csv.")
```

## Production evidence

Production ran on Modal under the registered criteria (seed base 830150001):
384 cells, 2,000 null and 1,000 power draws per cell, 576,000 draws and about
52 CPU-hours. 34 draws produced no test (25 non-converged fits, 9 with an empty
category in one group); they stay in the denominators. Three-category
controls return the registered zero-df `equivalent_models` reason, and the
delta WLSMV comparator returns `not_nested` for this moment-nested step, as
registered. `failure_counts.csv` lists every typed reason by cell and arm.

## Fresh-draw confirmation

The confirmation used seed base 840150001 and 2,000 draws in each of the 144
null cells with four, five or seven categories (288,000 draws). Ten draws
produced no test; the WLSMV comparator again returns `not_nested`. The cells
ran on Modal; the Modal workspace was disabled before its combine step, so the
same `modal/combine.R` combined the downloaded cells locally.

```{r}
conf <- file.path("results", "confirm-2026-10-05")
if (!file.exists(file.path(conf, "summary.csv")))
  stop("Confirmation summaries absent: modal run app.py --mode confirm --run-id confirm-2026-10-05 --cells <null cells, four or more categories>")
cs <- merge(read.csv(file.path(conf, "summary.csv")), read.csv(file.path(conf, "cells.csv")), by = "cell_id")
cx <- cs[cs$role == "null" & cs$arm %in% c("policy_sb", "policy_peba4", "all"), ]
cw <- aggregate(rate ~ categories + n + arm, cx, f)
cw <- reshape(cw, idvar = c("categories", "n"), timevar = "arm", direction = "wide")
cw <- cw[order(cw$categories, cw$n), ]
knitr::kable(data.frame(Categories = cw$categories, `N per group` = cw$n,
  `Policy SB (%)` = cw$rate.policy_sb, `Policy PEBA4 (%)` = cw$rate.policy_peba4,
  `All (%)` = cw$rate.all, check.names = FALSE),
  caption = "Fresh-draw confirmation: null rejection at nominal 5%, median (range) over 16 cells per row, 2,000 draws per cell.")
```

## Pilot (development)

```{r}
base <- "results/pilot"
if (!file.exists(file.path(base, "summary.csv")))
  stop("Run run_experiment.R --pilot --cell 17,18,65,66,209,210,257,258 --workers 2 --out-dir results/pilot")
s <- read.csv(file.path(base, "summary.csv"))
cost <- read.csv(file.path(base, "cost.csv"))
tab <- cost[c("observed_cells", "full_cells", "production_cpu_hours", "unpriced_cells")]
names(tab) <- c("Pilot Cells", "Full Grid Cells", "Estimated CPU Hours", "Unpriced Cells")
knitr::kable(tab, digits=2)
```

Full cell rates, Wilson intervals, availability and failures are in the frozen
CSV files. Pilot flags are development diagnostics. The cost extrapolation uses
factor/parameterization family means and excludes cloud startup/build time.

## Caveats

Six policy cell/arm rates fall outside the registered band in this tiny pilot;
these are diagnostic flags, not confirming evidence. Twenty draws have
substantial sampling uncertainty. The bounded pilot covers matched null/power cells for both factor counts and
parameterizations at N = 500 with four symmetric categories and correct models.
It does not cover other categories, sample sizes or misspecification. The full
20-draw grid would exceed the ten-minute pilot budget based on smoke timing. Identical
group distributions preserve the pseudo-null; the one-factor misspecification
uses a residual correlation because a cross-loading requires two factors.
The requested phase seed bases overlap across the 384-cell grid; registration
must amend them to disjoint ranges before held-out production.
The registration amended the production seed base to 830150001 before
production.

## Reproduce

From this directory, using a lane R library and one BLAS/OpenMP thread:

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --workers 2 --out-dir results/smoke
Rscript run_experiment.R --pilot --cell 17,18,65,66,209,210,257,258 --workers 2 --out-dir results/pilot
```

Production and the confirmation run on Modal from `modal/`:
`modal run app.py --mode production --run-id production-2026-10-05` and
`modal run app.py --mode confirm --run-id confirm-2026-10-05 --cells <the 144 null cells with four or more categories>`.
Modal fans out stable cell IDs; `modal/combine.R` applies the same summary
function to the cell outputs and validates seeds, provenance and completeness.
