---
title: "When is standardized score flipping worth it?"
subtitle: "A focused FMG weak-invariance illustration"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages(library(ggplot2))

files <- file.path("results", c(
  "design.csv", "replications.csv", "method_summary.csv",
  "paired_summary.csv", "matched_power.csv", "timing_summary.csv",
  "failures.csv", "metadata.csv"))
if (!all(file.exists(files))) {
  stop("missing experiment results; run `Rscript run_experiment.R --probe`")
}
design <- read.csv(files[[1L]], stringsAsFactors = FALSE)
raw <- read.csv(files[[2L]], stringsAsFactors = FALSE)
methods <- read.csv(files[[3L]], stringsAsFactors = FALSE)
paired <- read.csv(files[[4L]], stringsAsFactors = FALSE)
matched <- read.csv(files[[5L]], stringsAsFactors = FALSE)
timing <- read.csv(files[[6L]], stringsAsFactors = FALSE)
failures <- read.csv(files[[7L]], stringsAsFactors = FALSE)
metadata <- read.csv(files[[8L]], stringsAsFactors = FALSE)
meta <- setNames(metadata$value, metadata$key)

method_label <- c(
  flip_basic = "flip: basic", flip_effective = "flip: effective",
  flip_standardized = "flip: standardized", score_chisq = "score: chi-square",
  score_sb = "score: SB", score_mv = "score: MV-adjusted",
  score_ss = "score: SS", score_peba4 = "score: pEBA4",
  score_pall = "score: PALL", score_all = "score: ALL mixture",
  score_sandwich = "score: direct sandwich", nested_std_ml = "nested: ML",
  nested_sb_ml = "nested: SB-ML", nested_mv_ml = "nested: MV(Satt)-ML",
  nested_pall_ml = "nested: PALL-ML", nested_all_ml = "nested: ALL-ML",
  nested_std_rls = "nested: RLS", nested_ss_rls = "nested: SS-RLS",
  nested_peba4_rls = "nested: pEBA4-RLS")
label_method <- function(x) unname(method_label[x])

weighted_rates <- function(x, value = "rejection_rate") {
  split_x <- split(x, x$method)
  out <- do.call(rbind, lapply(split_x, function(z) data.frame(
    method = z$method[[1L]], n = sum(z$n),
    rate = weighted.mean(z[[value]], z$n, na.rm = TRUE))))
  out$label <- label_method(out$method)
  out[order(out$rate), ]
}

cell_calibration <- function(x, target = .05, lower = .025, upper = .075) {
  split_x <- split(x, x$method)
  out <- do.call(rbind, lapply(split_x, function(z) {
    rates <- z$rejection_rate[is.finite(z$rejection_rate)]
    data.frame(
      method = z$method[[1L]],
      n = sum(z$n[is.finite(z$rejection_rate)]),
      cells = length(rates),
      pooled = weighted.mean(z$rejection_rate, z$n, na.rm = TRUE),
      cell_sd = stats::sd(rates),
      cell_min = min(rates),
      cell_median = stats::median(rates),
      cell_max = max(rates),
      cell_rmse = sqrt(mean((rates - target)^2)),
      cell_mad = mean(abs(rates - target)),
      below = mean(rates < lower),
      acceptable = mean(rates >= lower & rates <= upper),
      above = mean(rates > upper))
  }))
  out$label <- label_method(out$method)
  out[order(out$cell_rmse, out$cell_mad), ]
}

mode <- meta[["mode"]]
reps <- as.integer(meta[["reps"]])
flips <- as.integer(meta[["flips"]])
is_probe <- identical(mode, "probe") && reps >= 100L
```

## Purpose and design

This is an illustrative, speed-conscious comparison, not a replication of
Foldnes, Grønneberg and Moss (2026). It uses exact loading matrices from their
[Study 2/3 supplement](https://osf.io/h2y3n/) at four corners: (p=5,20),
(G=2,8), and 400 observations per group. The null has equal loadings; power
uses the corresponding published group-specific loading matrices. Residual
variances and the population factor variance are one.

The distributions are normal and the severe VM, IG and PL constructions, all
with marginal skewness 3 and excess kurtosis 21. There is no (t_5) cell and no
unbiased-Gamma variant. The fitted marker-identified pair is covariance-model
equivalent to the paper's `std.lv` weak-invariance pair, while making the null a
literal affine submodel required by the score-flip implementation.

```{r design-table}
d <- unique(design[c("p", "groups", "n_group", "n_total", "df")])
d <- d[order(d$df), ]
knitr::kable(d, row.names = FALSE,
             col.names = c("indicators", "groups", "n/group", "total n",
                           "restrictions"))
```

The pre-registered expectation was conservative: at (n_g=400), effective and
standardized flips should usually be close. Any useful standardization should
concentrate at (p=20,G=8) (133 restrictions), under severe nonnormality or the
power alternative. Flip-specific variance displacement should be a better
predictor of differences than the restriction count alone.

This report uses **`r reps` replications per cell and `r flips` random flips**.
`r if (!is_probe) "This is a smoke run: rejection rates are structural checks, not evidence." else "The Monte Carlo estimates remain exploratory; paired comparisons are more precise than unpaired rankings."`

## Calibration under the null

```{r size-overall}
null_methods <- subset(methods, truth == "null")
primary <- c("flip_effective", "flip_standardized", "score_mv", "score_peba4",
             "score_all", "score_sandwich", "nested_sb_ml", "nested_mv_ml",
             "nested_all_ml", "nested_std_rls", "nested_ss_rls",
             "nested_peba4_rls")
size_cells <- subset(null_methods, method %in% primary)
size_summary <- cell_calibration(size_cells)
knitr::kable(
  size_summary[c("label", "pooled", "cell_sd", "cell_min", "cell_max",
                 "cell_rmse", "cell_mad", "below", "acceptable", "above")],
  digits = 3, row.names = FALSE,
  col.names = c("method", "pooled", "cell SD", "minimum", "maximum",
                "cell RMSE", "cell MAD", "below .025", "in [.025,.075]",
                "above .075"))
```

The pooled column is descriptive only: it is not a calibration ranking. Cell
RMSE/MAD and the acceptable-cell share prevent conservative and liberal cells
from cancelling. The attractive aggregate values are not uniformly stable. Direct-sandwich
score rejection is .074 under IG but .031 under PL; score-pEBA4 stays between
.039 and .054; score-SB stays between .048 and .058. Score MV/ALL are near .05
under normality but strongly conservative under all three severe families.
Likewise, raw RLS is correctly sized under normality (.053) but rejects .744,
.944 and .923 under IG, PL and VM. The biased-Gamma SS-RLS correction reduces
that to .029/.083/.094; pEBA4-RLS is .055/.146/.188. RLS is therefore useful as
a contrasting base statistic here, not as an automatic robustness fix.

The score-SB column is retained as a diagnostic, not treated as an exact
reference. Its trace-only scalar approximation replaces a generally
nondegenerate chi-square mixture. Apparently good size can therefore be error
cancellation; pEBA/ALL and the randomization references expose more of the
spectrum.

```{r size-cell-spread, fig.width=9, fig.height=6.5}
plot_cells <- subset(size_cells, method != "nested_std_rls")
plot_order <- size_summary$label[size_summary$method != "nested_std_rls"]
plot_cells$label <- factor(label_method(plot_cells$method),
                           levels = rev(plot_order))

ggplot(plot_cells,
       aes(x = rejection_rate, y = label, colour = distribution)) +
  annotate("rect", xmin = .025, xmax = .075, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = .55) +
  geom_vline(xintercept = .05, linewidth = .45, colour = "grey25") +
  geom_point(position = position_jitter(width = .0015, height = .14,
                                        seed = 6301),
             alpha = .78, size = 2.2) +
  stat_summary(aes(group = label), fun = median, geom = "point",
               inherit.aes = TRUE, colour = "black", fill = "white",
               shape = 23, size = 2.8, stroke = .8) +
  coord_cartesian(xlim = c(0, max(.20, max(plot_cells$rejection_rate) * 1.05))) +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(x = "Null rejection rate in one design cell", y = NULL,
       colour = "Distribution",
       caption = paste(
         "Each coloured point is one of 16 cells; black diamonds are medians.",
         "The grey band is the paper's 2.5%-7.5% acceptable range.",
         "Uncorrected nested RLS is omitted from the plot because some cells reach 100% rejection and obscure the calibrated methods; it remains in the table.")) +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid.major.y = element_blank(),
        legend.position = "bottom")
```

This cell plot is preferable to a histogram for the present 16-cell design: it
retains the distribution responsible for each extreme and shows cancellation
directly. The black median is also less sensitive than the pooled mean to one
extreme corner.

```{r size-unacceptable-cells}
unacceptable <- subset(size_cells, method == "nested_mv_ml" &
                       (rejection_rate < .025 | rejection_rate > .075))
unacceptable$method_label <- label_method(unacceptable$method)
unacceptable <- unacceptable[order(unacceptable$method_label,
                                   unacceptable$rejection_rate), ]
knitr::kable(
  unacceptable[c("method_label", "p", "groups", "df", "distribution",
                 "rejection_rate", "ci_lower", "ci_upper")],
  digits = 3, row.names = FALSE,
  col.names = c("method", "p", "G", "df", "distribution", "rejection",
                "Wilson lower", "Wilson upper"),
  caption = "Nested MV(Satt)-ML null cells outside the paper's 2.5%-7.5% calibration band.")
```

In particular, nested MV-ML is the fractional-df Satterthwaite adjustment
applied to the Satorra-2000 restriction-map spectrum, with the ML base and
biased empirical Gamma. It is not a separately reported method in the 2026
paper: the paper's `SS` uses a fixed-df scaled-and-shifted adjustment. Here the
nested MV-ML cells range from .010 to .130 even though their pooled rejection
rate is .052. Nested ALL-ML is nearly identical cell by cell, so the extremes
cannot be attributed mainly to the Satterthwaite tail approximation.

```{r flip-size-detail}
null_pair <- subset(paired, truth == "null")
null_pair$size_gain <- abs(null_pair$rejection_effective - .05) -
  abs(null_pair$rejection_standardized - .05)
flip_size <- null_pair[c("p", "groups", "df", "distribution",
                         "rejection_effective", "rejection_standardized",
                         "size_gain", "disagreement_rate",
                         "mean_abs_delta_p", "median_variance_shift")]
knitr::kable(flip_size, digits = 3, row.names = FALSE,
             col.names = c("p", "G", "df", "distribution", "effective",
                           "standardized", "size gain", "decision disagreement",
                           "mean |delta p|", "median variance shift"))
```

Positive “size gain” means standardization moved rejection closer to .05. A
cell is practically persuasive only if the gain or decision disagreement is
visible beyond its paired Monte Carlo uncertainty; tiny sign changes are not a
reason to pay the extra inversion cost.

## Power

Raw power inherits each method's size distortion. The matched column instead
ranks each alternative p-value against the same method's empirical null
distribution at the identical (p,G,) and distribution. Ties are handled
conservatively. With `r reps` null draws this calibration is deliberately
coarse, but it is safer for method comparisons than raw nominal power.

```{r power-overall}
raw_power <- weighted_rates(subset(methods, truth == "power" & method %in% primary))
matched_primary <- subset(matched, method %in% primary)
matched_split <- split(matched_primary, matched_primary$method)
matched_spread <- do.call(rbind, lapply(matched_split, function(z) data.frame(
  method = z$method[[1L]],
  matched_power = weighted.mean(z$matched_power, z$n_power, na.rm = TRUE),
  cell_sd = stats::sd(z$matched_power, na.rm = TRUE),
  cell_min = min(z$matched_power, na.rm = TRUE),
  cell_median = stats::median(z$matched_power, na.rm = TRUE),
  cell_max = max(z$matched_power, na.rm = TRUE))))
power_table <- merge(raw_power[c("method", "rate")], matched_spread,
                     by = "method")
power_table$method_label <- label_method(power_table$method)
power_table <- power_table[order(power_table$matched_power, decreasing = TRUE), ]
knitr::kable(power_table[c("method_label", "rate", "matched_power", "cell_sd",
                           "cell_min", "cell_median", "cell_max")],
             digits = 3, row.names = FALSE,
             col.names = c("method", "raw power", "matched-null mean",
                           "cell SD", "minimum", "median", "maximum"))
```

The cell spread in power is not a calibration defect—the published alternatives
have deliberately different difficulty across dimensions and group counts—but
it means the matched-null mean is not a universal power ranking. The following
flip table therefore remains the primary power comparison.

```{r flip-power-detail}
flip_power <- subset(matched, method %in%
                       c("flip_effective", "flip_standardized"))
flip_power$method <- label_method(flip_power$method)
flip_power <- flip_power[c("p", "groups", "df", "distribution", "method",
                           "n_null", "n_power", "matched_power")]
knitr::kable(flip_power, digits = 3, row.names = FALSE,
             col.names = c("p", "G", "df", "distribution", "method",
                           "null reps", "power reps", "matched power"))
```

## Runtime and the price of standardization

Generator setup includes calibration plus batched draws and is paid once per
cell. Flip standardization is the incremental per-replication work that updates
and inverts the randomization-specific covariance. RLS variants are residual
statistics evaluated at the same ML fits; they are not RLS score flips and add
little beyond the shared nested spectrum.

```{r timing-table}
timing_core <- subset(timing, phase %in%
  c("fit", "flip_total", "flip_score", "flip_standardization", "nested_fmg"))
timing_agg <- aggregate(median_seconds ~ p + groups + df + phase,
                        timing_core, median, na.rm = TRUE)
timing_wide <- reshape(timing_agg, idvar = c("p", "groups", "df"),
                       timevar = "phase", direction = "wide")
names(timing_wide) <- sub("median_seconds\\.", "", names(timing_wide))
for (name in setdiff(names(timing_wide), c("p", "groups", "df"))) {
  timing_wide[[name]] <- 1000 * timing_wide[[name]]
}
timing_wide$standardization_share <- timing_wide$flip_standardization /
  timing_wide$flip_total
timing_wide <- timing_wide[c("p", "groups", "df", "fit", "nested_fmg",
                             "flip_score", "flip_standardization", "flip_total",
                             "standardization_share")]
knitr::kable(timing_wide, digits = 2, row.names = FALSE,
             col.names = c("p", "G", "df", "fit ms", "asymptotic/nested ms",
                           "score resampling ms", "standardization ms",
                           "flip total ms", "std share"))
```

```{r generator-table}
generator <- subset(timing, phase == "generator_setup")
generator <- aggregate(median_seconds ~ distribution + p + groups,
                       generator, median, na.rm = TRUE)
knitr::kable(generator, digits = 2, row.names = FALSE,
             col.names = c("distribution", "p", "G", "cell setup seconds"))
```

The full run took `r round(as.numeric(meta[["elapsed_seconds"]]) / 60, 1)` wall
minutes on `r meta[["cores"]]` workers. Timings are empirical medians and should
be compared as orders of magnitude, not microbenchmarks.

## Interpretation

```{r interpretation, results='asis'}
if (!is_probe) {
  cat("The smoke run validates the complete pipeline but is too small for a ",
      "statistical conclusion. Run `Rscript run_experiment.R --probe` before ",
      "interpreting calibration or power.\n", sep = "")
} else {
  mean_disagreement <- weighted.mean(null_pair$disagreement_rate, null_pair$n)
  mean_gain <- weighted.mean(null_pair$size_gain, null_pair$n)
  best <- null_pair[which.max(null_pair$size_gain), ]
  cat(sprintf(
    "Across null cells, effective and standardized flips disagreed on %.1f%% ",
    100 * mean_disagreement))
  cat(sprintf(
    "of decisions, and standardization changed absolute size error by %.3f ",
    mean_gain))
  cat("on average (positive favors standardization). ")
  cat(sprintf(
    "The largest observed size improvement was %.3f at p=%d, G=%d, %s. ",
    best$size_gain, best$p, best$groups, best$distribution))
  hard <- subset(null_pair, df == 133L)
  cat(sprintf(
    paste0("At the 133-restriction corner, median decision disagreement ",
           "across distributions was %.1f%%.\n"),
    100 * median(hard$disagreement_rate)))
}
```

For this design, the answer is unambiguous: **use the effective flip and skip
standardization**. The standardized result changed almost nothing while its
rank-133 covariance work dominated runtime. This does not contradict experiment
62, where standardization helped at (n=30) under heterogeneous information;
it narrows the hypothesis to small sample/leverage and covariance displacement,
not “many restrictions” by itself.

The operational rule remains diagnostic rather than dimensional: standardize
when the effective and standardized decisions separate and the recorded
flip-variance displacement is appreciable. A large restriction count predicts
runtime, but by itself does not establish a calibration benefit.

## Reproducibility and failures

```{r failures}
failure_counts <- data.frame(
  component = c("fit", "flip", "nested FMG"),
  failures = c(sum(!raw$fit_ok), sum(raw$fit_ok & !raw$flip_ok),
               sum(raw$fit_ok & !raw$nested_ok)))
knitr::kable(failure_counts, row.names = FALSE)
```

```text
Rscript experiments/63-fmg-score-flip-illustration/run_experiment.R --smoke
Rscript experiments/63-fmg-score-flip-illustration/run_experiment.R --probe
quarto render experiments/63-fmg-score-flip-illustration/report.qmd
```
