---
title: "Robust invariance testing: the base statistic beats the transform"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
files <- file.path("results", c(
  "summary.csv", "expansion_summary.csv", "expansion_replications.csv",
  "expansion_metadata.csv", "timing_summary.csv", "power_summary.csv",
  "power_replications.csv", "power_metadata.csv", "stress_summary.csv",
  "dimension_summary.csv"))
if (any(!file.exists(files))) {
  stop(paste(
    "missing results; run run_experiment.R, run_expansion.R --probe,",
    "run_expansion.R --power, run_expansion.R --stress, run_dimension.R --probe,",
    "and run_timing.R"))
}
s <- read.csv(files[[2L]])
r <- read.csv(files[[3L]]); r_all_ok <- r$ok; r <- r[r$ok, , drop = FALSE]
m <- read.csv(files[[4L]])
timing <- read.csv(files[[5L]])
power <- read.csv(files[[7L]]); power <- power[power$ok, , drop = FALSE]
power_metadata <- read.csv(files[[8L]])
stress <- read.csv(files[[9L]])
dim_summary <- read.csv(files[[10L]])

# Method naming: {flip | score | LR} x transform. Every method is built on the
# same efficient score; what differs is how the reference is obtained (flip =
# sign-flip resampling; score/LR = FMG spectrum transforms on the score
# statistic's own eigenvalues, or on the Satorra-2000 LR construction).
method_labels <- c(
  p_effective    = "flip (effective)",
  p_standardized = "flip (standardized)",
  p_chisq        = "score: chi-square",
  p_mean_scaled  = "score: SB",
  p_score_ss     = "score: SS",
  p_score_mv     = "score: MV",
  p_score_sf     = "score: SF",
  p_score_peba4  = "score: pEBA4",
  p_mixture      = "score: ALL",
  p_fmg_sb       = "LR: SB",
  p_fmg_ss       = "LR: SS",
  p_fmg_mv       = "LR: MV",
  p_fmg_sf       = "LR: SF",
  p_fmg_peba4    = "LR: pEBA4",
  p_fmg_all      = "LR: ALL")

# Same FMG transform on two base statistics: score vs LR.
fmg_transforms <- c("SB", "SS", "MV", "SF", "pEBA4", "ALL")
score_cols <- c(SB = "p_mean_scaled", SS = "p_score_ss", MV = "p_score_mv",
                SF = "p_score_sf", pEBA4 = "p_score_peba4", ALL = "p_mixture")
lr_cols <- c(SB = "p_fmg_sb", SS = "p_fmg_ss", MV = "p_fmg_mv",
             SF = "p_fmg_sf", pEBA4 = "p_fmg_peba4", ALL = "p_fmg_all")
base_compare_reps <- function(data) data.frame(
  Transform = fmg_transforms,
  `Score base` = vapply(fmg_transforms,
    function(t) mean(data[[score_cols[[t]]]] < .05, na.rm = TRUE), numeric(1L)),
  `LR base` = vapply(fmg_transforms,
    function(t) mean(data[[lr_cols[[t]]]] < .05, na.rm = TRUE), numeric(1L)),
  check.names = FALSE, row.names = NULL)
base_compare_summary <- function(data) data.frame(
  Transform = fmg_transforms,
  `Score base` = vapply(fmg_transforms, function(t)
    mean(data[[sub("^p_", "rejection_", score_cols[[t]])]], na.rm = TRUE), numeric(1L)),
  `LR base` = vapply(fmg_transforms, function(t)
    mean(data[[sub("^p_", "rejection_", lr_cols[[t]])]], na.rm = TRUE), numeric(1L)),
  check.names = FALSE, row.names = NULL)
base_null <- base_compare_reps(r)
base_stress <- base_compare_summary(stress[stress$df == 8L, , drop = FALSE])
flip_std_overall <- mean(r$p_standardized < .05, na.rm = TRUE)

# Normal-data N-convergence: with no non-normality, any gap is small-sample.
norm_conv <- do.call(rbind, lapply(c(60L, 100L, 200L), function(nn) {
  sub <- r[r$distribution == "normal" & r$n_total == nn, , drop = FALSE]
  data.frame(`Total N` = nn,
    `score: SB` = mean(sub$p_mean_scaled < .05, na.rm = TRUE),
    `LR: SB` = mean(sub$p_fmg_sb < .05, na.rm = TRUE),
    `flip (standardized)` = mean(sub$p_standardized < .05, na.rm = TRUE),
    check.names = FALSE)
}))
t5_lr_sb_n200 <- mean(r$p_fmg_sb[r$distribution == "t5" & r$n_total == 200L] < .05,
                      na.rm = TRUE)

# Severe-copula stress helpers.
stress_cols <- c(
  `flip (standardized)` = "rejection_standardized",
  `score: SB`           = "rejection_mean_scaled",
  `score: ALL`          = "rejection_mixture",
  `LR: SB`              = "rejection_fmg_sb",
  `LR: pEBA4`           = "rejection_fmg_peba4")
stress_row <- function(d, dff, het = NULL, nn = NULL) {
  sub <- stress[stress$distribution == d & stress$df == dff, , drop = FALSE]
  if (!is.null(het)) sub <- sub[sub$heterogeneity == het, , drop = FALSE]
  if (!is.null(nn)) sub <- sub[sub$n_total == nn, , drop = FALSE]
  vapply(stress_cols, function(cc) mean(sub[[cc]], na.rm = TRUE), numeric(1L))
}

# Dimension-scaling helpers.
dim_cols <- c(`score: SB` = "rejection_score_sb",
              `score: ALL` = "rejection_score_all",
              `LR: SB` = "rejection_lr_sb",
              `flip (standardized)` = "rejection_flip_standardized")
dim_table <- function(adeq) {
  sub <- dim_summary[dim_summary$adequacy == adeq, , drop = FALSE]
  agg <- aggregate(sub[unname(dim_cols)], list(p = sub$p), mean)
  names(agg) <- c("p", names(dim_cols)); agg
}
dim_adequate <- dim_table("adequate")
dim_pick <- function(adeq, pp, col)
  mean(dim_summary[dim_summary$adequacy == adeq & dim_summary$p == pp, col], na.rm = TRUE)

# Standardization sub-study: decision changes by N and restriction count.
by_ndf <- aggregate(
  r[c("reject_effective", "reject_standardized", "changed_to_accept")],
  r[c("n_total", "df")], mean)

# Matched-empirical-null (size-adjusted) power.
power_methods <- intersect(names(method_labels), names(power))
empirical_critical <- function(x, alpha = .05) {
  x <- x[is.finite(x)]
  candidates <- sort(unique(x))
  rates <- vapply(candidates, function(cut) mean(x < cut), numeric(1L))
  valid <- candidates[rates <= alpha]
  if (!length(valid)) -Inf else max(valid)
}
keys <- c("df", "n_total", "balance", "heterogeneity", "distribution")
power_cells <- unique(power[c(keys, "alternative", "effect")])
calibrated <- do.call(rbind, lapply(seq_len(nrow(power_cells)), function(i) {
  cell <- power_cells[i, , drop = FALSE]
  null_match <- rep(TRUE, nrow(r)); power_match <- rep(TRUE, nrow(power))
  for (key in keys) {
    null_match <- null_match & r[[key]] == cell[[key]]
    power_match <- power_match & power[[key]] == cell[[key]]
  }
  power_match <- power_match & power$alternative == cell$alternative &
    power$effect == cell$effect
  do.call(rbind, lapply(power_methods, function(v) {
    critical <- empirical_critical(r[[v]][null_match])
    data.frame(cell, method = unname(method_labels[[v]]),
      calibrated_power = mean(power[[v]][power_match] < critical, na.rm = TRUE))
  }))
}))
calibrated_power <- aggregate(calibrated["calibrated_power"],
  calibrated[c("alternative", "effect", "method")], mean)
calibrated_power <- reshape(calibrated_power,
  idvar = c("alternative", "effect"), timevar = "method", direction = "wide")
names(calibrated_power) <- sub("^calibrated_power\\.", "", names(calibrated_power))
power_show_cols <- function(tab) intersect(c(
  "alternative", "effect", "flip (standardized)", "score: SB", "score: MV",
  "score: pEBA4", "LR: SB", "LR: MV", "LR: pEBA4"), names(tab))

# Timing.
timing_ms <- timing
for (name in grep("seconds$", names(timing), value = TRUE))
  timing_ms[[sub("seconds$", "ms", name)]] <- 1000 * timing[[name]]
timing_499 <- subset(timing_ms, scenario == "easy" & n_flips == 499,
  select = c(df, median_setup_ms, median_resampling_score_ms,
             median_standardization_ms, median_asymptotic_ms, median_core_total_ms))
null_elapsed <- as.numeric(m$value[m$key == "elapsed_seconds"])
```

## Question

When you test measurement invariance under non-normality, which robust test
controls Type I error? We compare the sign-flip resampling test against the
Foldnes-Moss-Grønneberg spectrum transforms (Satorra-Bentler `SB`,
scaled-shifted `SS`, mean-variance `MV`, scaled-$F$ `SF`, `pEBA`, and the exact
mixture `ALL`), each applied to two base statistics: the score/Lagrange-multiplier
statistic, and the Satorra-2000 likelihood-ratio statistic that lavaan and Mplus
use by default.

## Short answer

The base statistic dominates the transform. On the null grid, the FMG transforms
on the score statistic's own eigenvalues are near nominal for every transform
(about `r sprintf("%.2f", min(base_null[["Score base"]]))` to
`r sprintf("%.2f", max(base_null[["Score base"]]))`), while the identical
transforms on the LR base are uniformly liberal (about
`r sprintf("%.2f", min(base_null[["LR base"]]))` to
`r sprintf("%.2f", max(base_null[["LR base"]]))`). Which spectrum transform you
pick barely matters next to which base statistic you read it from.

The score construction is not exotic. It is Neyman's C(α) test, the same object
an SEM modification index computes, and it inherits the Lagrange-multiplier
test's well-known small-sample size advantage. Much of the headline gap is the
classic over-rejection of likelihood-ratio tests at small $N$, which adequate
sample size removes. A residual robustness margin remains under genuine
non-normality.

That margin holds up. From ten to thirty variables the score base stays near
nominal while the LR base stays liberal, at matched sample adequacy. Sample
adequacy, not dimension, is what governs. And under a severe non-Gaussian copula,
where the LR base rejects up to `r sprintf("%.2f", max(base_stress[["LR base"]]))`
and the sign-flip test reaches
`r sprintf("%.2f", stress_row("pl", 8L)[["flip (standardized)"]])` at 8 df, the
score-base transforms are the only members of the battery that stay near nominal.

Two practical notes. After matched-null calibration all tests reach similar
power, so the LR base's higher nominal power was size distortion rather than
detection. And standardizing the sign-flip test is a cheap, minor improvement,
useful mainly at small $N$ with several restrictions, not a robustness fix.

## The base statistic dominates the transform

The Foldnes, Moss, and Grønneberg (2026) transforms exist because the two
established robust nested tests fail in opposite directions under non-normality:
Satorra-Bentler (`SB`) is chronically liberal and scaled-shifted (`SS`) is
chronically conservative. `MV`, `SF`, and the eigenvalue-averaging families
(`EBA`, `pEBA`) are the proposed repairs. All of them transform the eigenvalue
spectrum of a weighted-chi-square limit.

The score test carries its own spectrum. Its statistic
$T = u^\top V^{-1} u$ over the efficient-score subspace has an asymptotic
weighted-chi-square limit whose eigenvalues come straight from the model, so the
identical transforms can be applied to those eigenvalues instead of to the
Satorra-2000 construction. The table applies each transform to both bases, on the
full null grid.

```{r base-null}
knitr::kable(base_null, digits = 3)
```

On the LR base every transform is liberal; on the score base every transform is
near nominal. The more aggressive score-base transforms (`SS`, `MV`, `SF`,
`ALL`) are mildly conservative rather than liberal, trading a little power for
size control. The gap holds even at a single restriction, where the transforms
collapse to nearly one scalar test: there the LR base still rejects about
`r sprintf("%.2f", mean(r$p_fmg_sb[r$df == 1L] < .05, na.rm = TRUE))` for `SB`
while the score base is already near
`r sprintf("%.2f", mean(r$p_mean_scaled[r$df == 1L] < .05, na.rm = TRUE))`.

### Why the score base wins, and how much is small-sample

The efficient-score construction is Neyman's C(α) test: the score of the tested
parameters, orthogonalized against the nuisance parameters and standardized by
the efficient information. In SEM that is exactly what a modification index
computes, only univariate. Assembled into a joint test with a spectrum reference
it is a robust Lagrange-multiplier test, and Lagrange-multiplier tests are the
least liberal of the classical trio in small samples (the Berndt-Savin ordering
places them below the likelihood-ratio and Wald statistics). So part of the
score base's advantage is the documented small-sample over-rejection of the LR
statistic, not robustness.

The normal-data cells separate the two effects. With no non-normality to correct,
any gap must be small-sample.

```{r norm-convergence}
knitr::kable(norm_conv, digits = 3)
```

Under normality the LR base falls from about 0.12 at $N=60$ toward nominal as $N$
grows, while the score base stays flat near 0.05. Adequate sample size removes
that part of the gap. What remains under non-normality is the genuine robustness
margin: under $t_5$ the LR base is still about
`r sprintf("%.2f", t5_lr_sb_n200)` at $N=200$, where the score base is near 0.05.
That margin is smaller than the raw gap but does not wash out.

As a validity check, the LR-base `SB`, `MV`, and `SS` values reproduce the
independent nested-test scaled, mean-variance, and scaled-shifted routes to a
maximum p-value gap of
`r format(max(r$fmg_duplicate_max_gap, na.rm = TRUE), scientific = TRUE, digits = 2)`.

## The copula, not the marginals

Mild non-normality does not stress these tests: an elliptical $t_5$ or a
coordinatewise skew stays close to asymptotic robustness. The lever that breaks
them is the joint dependence structure, the copula, not the marginal skewness and
kurtosis (Foldnes et al., 2026). Two data sets can share a covariance matrix and
identical marginal moments yet differ in their copula, and normal-theory ML
inference is not asymptotically robust to that difference.

The stress block fixes the marginals at a severe level (skewness 3, excess
kurtosis 21) and switches only the copula: Vale-Maurelli is a Gaussian copula,
piecewise-linear is not. The table applies each transform to both bases at the
8-df joint test, averaged over $N$ and the homogeneous and geometry populations.

```{r base-stress}
knitr::kable(base_stress, digits = 3)
```

The score base survives and the LR base does not. On the LR base every transform
rejects four to eight times nominal. On the score base the aggressive transforms
stay near or below 0.05, with `SB` and `pEBA4` only mildly liberal. The sign-flip
test lands between them, about
`r sprintf("%.2f", stress_row("vm", 8L)[["flip (standardized)"]])` to
`r sprintf("%.2f", stress_row("pl", 8L)[["flip (standardized)"]])`, and
standardization does not repair it: the score sign-symmetry it relies on is
itself strained by a severe copula.

The copula, not group heterogeneity, is the cause. In homogeneous populations,
where the two groups share covariance and means exactly, the 8-df LR-`SB` test
still rejects about
`r sprintf("%.2f", stress_row("vm", 8L, "homogeneous")[["LR: SB"]])` under the
Gaussian copula and
`r sprintf("%.2f", stress_row("pl", 8L, "homogeneous")[["LR: SB"]])` under the
non-Gaussian one. Piecewise-linear is uniformly harder than Vale-Maurelli, as the
copula explanation predicts. The distortion shrinks with sample size but does not
calibrate: for the hardest cell the LR-`SB` test goes
`r sprintf("%.2f", stress_row("pl", 8L, "geometry", 60L)[["LR: SB"]])`,
`r sprintf("%.2f", stress_row("pl", 8L, "geometry", 100L)[["LR: SB"]])`,
`r sprintf("%.2f", stress_row("pl", 8L, "geometry", 200L)[["LR: SB"]])` at
$N=60, 100, 200$. This is an asymptotic-robustness failure, not a small-sample
one.

## Dimension scaling

Foldnes et al. (2026) show robust-test over-rejection grows with model dimension,
so the score base's edge might be a ten-variable artifact. This probe holds the
tested restriction count fixed at 8 loadings and grows the ambient model from two
to six factors ($p = 10, 20, 30$) under the severe copula, at a matched per-group
sample-to-parameter ratio of about 6. The table averages over the two copulas.

```{r dim-adequate}
knitr::kable(dim_adequate, digits = 3)
```

The edge is not a ten-variable artifact: the score base stays near nominal at
every dimension while the LR base is liberal at every dimension. But sample
adequacy, not dimension, governs. Because a fixed ratio grows $N$ with $p$, the
LR base's excess shrinks toward its residual robustness margin as the model grows
(LR-`SB` falls from about
`r sprintf("%.2f", dim_pick("adequate", 10, "rejection_lr_sb"))` at $p=10$ to
`r sprintf("%.2f", dim_pick("adequate", 30, "rejection_lr_sb"))` at $p=30$). At
an inadequate ratio near 2.5 the LR base over-rejects worst where $N$ is smallest,
about `r sprintf("%.2f", dim_pick("tight", 10, "rejection_lr_sb"))` at $p=10$,
$N=150$, falling to `r sprintf("%.2f", dim_pick("tight", 30, "rejection_lr_sb"))`
at $p=30$ where the same ratio buys a larger $N$. The score base holds near
nominal throughout, and `score: ALL` is its most consistently calibrated member.

## Power

Nominal power ranks the LR base above the score base, but only because the LR
base rejects more of everything. The honest comparison holds each method to its
own realized null size: each method and matched null cell gets its own critical
p-value (the largest cutoff whose observed null rejection does not exceed 0.05),
and power is read off independent alternative replications.

```{r calibrated-power}
knitr::kable(calibrated_power[power_show_cols(calibrated_power)], digits = 3)
```

Once size is matched, the flip, score-base, and LR-base tests are within a point
or two across all four alternative families. The score base's mild conservatism
therefore costs almost no power: it buys size control nearly for free. As
expected, a sparse violation concentrated in one loading is easier to detect than
the same norm spread across all tested loadings, and no single spectrum transform
dominates once size is aligned.

## Standardizing the sign-flip test

Standardization is the original subject of this probe and it is a minor effect.
Across `r format(nrow(r), big.mark = ",")` null replications it moved
`r sum(r$changed_to_accept)` decisions from rejection to acceptance and
`r sum(r$changed_to_reject)` the other way, always in the conservative direction.
Its value concentrates at small $N$ with several restrictions.

```{r n-df-table}
tab <- by_ndf
names(tab) <- c("total N", "df", "effective", "standardized",
                "effective-only rejection")
knitr::kable(tab, digits = 3)
```

At total $N=60$ standardization removes little for one restriction but a few
tenths of a percentage point for four and eight, and it does not fully repair the
liberal 8-df test. At $N=200$ the changes are negligible. Standardize by default
because it is nearly free, but do not read it as a small-sample correction.

## Timing

Cost is a few milliseconds and linear in the number of flips. At 499 flips,
median core latency was about 2.3 ms at 1 df, 5.4 ms at 4 df, and 8.7 ms at 8 df,
of which flip-specific standardization is roughly 0.7, 2.7, and 4.2 ms. The FMG
transforms on the score eigenvalues are essentially free once the flip test has
run, since the eigenvalues are already computed.

```{r timing-df}
names(timing_499) <- c("df", "setup ms", "resampling ms", "standardization ms",
                       "asymptotic ms", "total core ms")
knitr::kable(timing_499, digits = 3)
```

## Recommendation

- **Put the FMG spectrum transforms on the score base, not the LR base.** The
  same transform is near nominal on the score/LM statistic's eigenvalues and
  uniformly liberal on the Satorra-2000 LR construction that lavaan and Mplus
  default to. This is the largest calibration lever in the study, and it is free:
  the score eigenvalues are already computed by the flip test.
- Among the score-base transforms, the aggressive members (`SS`, `MV`, `ALL`)
  give the tightest size control, including under the severe copula, at a mild
  cost in conservatism that matched-null power shows is nearly free. `SB` and
  `pEBA4` sit closer to nominal.
- Read the advantage correctly: much of the raw gap is the classic small-sample
  over-rejection of likelihood-ratio tests, which adequate $N$ removes. The
  genuine robustness margin under non-normality is smaller but real.
- Keep `p_standardized` as a cheap resampling option, but it is not better
  calibrated than the score-base asymptotics and is more strained by a severe
  copula.
- Do not compare nominal power without comparing null size, and do not extend
  "robust" to a severe non-Gaussian copula: there only the score-base transforms
  hold near nominal, and even that is a probe result, not a guarantee.
- Next: vary nuisance leverage and conditioning at fixed dimension and adequacy,
  and push $p$ further with a proportionally larger $N$ to confirm the score
  base's margin does not close.

## Design and caveats

The null grid crosses 1/4/8 loading-equality restrictions, total $N=60/100/200$,
1:1 and 1:3 allocation, three group-information geometries, and normal, $t_5$,
and skewed innovations, at 200 replications and 499 flips per cell. Maximum cell
fit-failure rate was `r sprintf("%.3f", max(s$failure_rate))`; there were
`r sum(!r_all_ok)` failed fits overall. The power grid adds sparse and dense
alternatives at two effect sizes.

Two boundary probes sit outside the main grid so they do not unbalance it. The
severe-copula stress block crosses two copulas at fixed severe marginals with
1/8 restrictions, $N=60/100/200$, and two geometries. The dimension-scaling probe
crosses two copulas, $p=10/20/30$ at fixed 8-df restrictions, and two
sample-to-parameter ratios (about 2.5 and 6), with $N$ set from the ratio so it
grows with $p$; the ratios are approximate because the parameter count is
estimated. The independent-generator copula was excluded because its marginal
moment solve is infeasible at this severity.

Monte Carlo and binomial uncertainty remain visible in cell-level 0.05 decisions.
Score flipping here is a frontier method for affine, complete-data ML models. The
results do not automatically transfer to FIML, ordinal, estimated-weight,
clustered, boundary, or nonlinear-constraint tests.

## Reproduce

```sh
Rscript experiments/62-standardized-flip-score/run_experiment.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --power
Rscript experiments/62-standardized-flip-score/run_expansion.R --stress
Rscript experiments/62-standardized-flip-score/run_dimension.R --probe
Rscript experiments/62-standardized-flip-score/run_timing.R
quarto render experiments/62-standardized-flip-score/report.qmd
```
