---
title: "Score-flip tests, FMG transforms on the score vs LR base, and power"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
files <- file.path("results", c(
  "summary.csv", "expansion_summary.csv", "expansion_replications.csv",
  "expansion_metadata.csv", "timing_summary.csv", "power_summary.csv",
  "power_replications.csv", "power_metadata.csv", "stress_summary.csv",
  "dimension_summary.csv"))
if (any(!file.exists(files))) {
  stop(paste(
    "missing results; run run_experiment.R, run_expansion.R --probe,",
    "run_expansion.R --power, run_expansion.R --stress, run_dimension.R --probe,",
    "and run_timing.R"))
}
legacy <- read.csv(files[[1L]])
s <- read.csv(files[[2L]])
r <- read.csv(files[[3L]])
m <- read.csv(files[[4L]])
timing <- read.csv(files[[5L]])
power_summary <- read.csv(files[[6L]])
power <- read.csv(files[[7L]])
power_metadata <- read.csv(files[[8L]])
stress <- read.csv(files[[9L]])
dim_summary <- read.csv(files[[10L]])
r <- r[r$ok, , drop = FALSE]
power <- power[power$ok, , drop = FALSE]

# Dimension-scaling probe: tested df = 8 fixed, ambient p in {10,20,30}, severe
# copula, two sample-adequacy regimes. Aggregate over the two copulas.
dim_cols <- c(`score: SB` = "rejection_score_sb",
              `score: ALL` = "rejection_score_all",
              `LR: SB` = "rejection_lr_sb",
              `flip (standardized)` = "rejection_flip_standardized")
dim_table <- function(adeq) {
  sub <- dim_summary[dim_summary$adequacy == adeq, , drop = FALSE]
  agg <- aggregate(sub[unname(dim_cols)], list(p = sub$p), mean)
  names(agg) <- c("p", names(dim_cols))
  agg
}
dim_adequate <- dim_table("adequate")
dim_pick <- function(adeq, pp, col)
  mean(dim_summary[dim_summary$adequacy == adeq & dim_summary$p == pp, col], na.rm = TRUE)

# Severe non-normal copula stress block (vm = Gaussian copula, pl = non-Gaussian,
# matched severe marginals). Aggregated for the report.
stress_cols <- c(
  `flip (standardized)` = "rejection_standardized",
  `score: SB`           = "rejection_mean_scaled",
  `score: ALL`          = "rejection_mixture",
  `LR: SB`              = "rejection_fmg_sb",
  `LR: pEBA4`           = "rejection_fmg_peba4")
stress_row <- function(d, dff, het = NULL, nn = NULL) {
  sub <- stress[stress$distribution == d & stress$df == dff, , drop = FALSE]
  if (!is.null(het)) sub <- sub[sub$heterogeneity == het, , drop = FALSE]
  if (!is.null(nn)) sub <- sub[sub$n_total == nn, , drop = FALSE]
  vapply(stress_cols, function(cc) mean(sub[[cc]], na.rm = TRUE), numeric(1L))
}
s$calibration_effective <- abs(s$rejection_effective - .05)
s$calibration_standardized <- abs(s$rejection_standardized - .05)
s$calibration_gain <- s$calibration_effective - s$calibration_standardized

means <- function(vars, by) {
  aggregate(s[vars], s[by], mean)
}
by_df <- means(c("rejection_effective", "rejection_standardized",
                 "changed_to_accept", "mean_abs_delta_p",
                 "mean_variance_relative_shift"), "df")
by_n <- means(c("rejection_effective", "rejection_standardized",
                "changed_to_accept", "mean_abs_delta_p",
                "mean_variance_relative_shift"), "n_total")
by_dist <- means(c("rejection_effective", "rejection_standardized"),
                 "distribution")
by_balance <- means(c("rejection_effective", "rejection_standardized",
                      "changed_to_accept"), "balance")
by_ndf <- aggregate(
  r[c("reject_effective", "reject_standardized", "changed_to_accept")],
  r[c("n_total", "df")], mean)

timing_ms <- timing
for (name in grep("seconds$", names(timing), value = TRUE)) {
  timing_ms[[sub("seconds$", "ms", name)]] <- 1000 * timing[[name]]
}
timing_499 <- subset(
  timing_ms, scenario == "easy" & n_flips == 499,
  select = c(df, median_setup_ms, median_resampling_score_ms,
             median_standardization_ms, median_asymptotic_ms,
             median_core_total_ms))
timing_scale <- subset(
  timing_ms, scenario == "easy" & df == 8,
  select = c(n_flips, median_resampling_score_ms,
             median_standardization_ms, median_core_total_ms))

# Every method here is built on the same efficient score. What distinguishes
# them is (i) how the reference distribution is obtained and (ii) which base
# statistic and eigenvalues feed it. Three groups:
#   flip:  sign-flip resampling of the score (basic/effective/standardized).
#   score: FMG spectrum transforms on the SCORE statistic's own eigenvalues.
#   LR:    the same FMG transforms on the Satorra-2000 LR construction.
method_labels <- c(
  p_effective    = "flip (effective)",
  p_standardized = "flip (standardized)",
  p_chisq        = "score: chi-square",
  p_mean_scaled  = "score: SB",
  p_score_ss     = "score: SS",
  p_score_mv     = "score: MV",
  p_score_sf     = "score: SF",
  p_score_peba4  = "score: pEBA4",
  p_mixture      = "score: ALL",
  p_fmg_sb       = "LR: SB",
  p_fmg_ss       = "LR: SS",
  p_fmg_mv       = "LR: MV",
  p_fmg_sf       = "LR: SF",
  p_fmg_peba4    = "LR: pEBA4",
  p_fmg_all      = "LR: ALL")

# The same FMG transform on two base statistics: score vs LR.
fmg_transforms <- c("SB", "SS", "MV", "SF", "pEBA4", "ALL")
score_cols <- c(SB = "p_mean_scaled", SS = "p_score_ss", MV = "p_score_mv",
                SF = "p_score_sf", pEBA4 = "p_score_peba4", ALL = "p_mixture")
lr_cols <- c(SB = "p_fmg_sb", SS = "p_fmg_ss", MV = "p_fmg_mv",
             SF = "p_fmg_sf", pEBA4 = "p_fmg_peba4", ALL = "p_fmg_all")
base_compare_reps <- function(data) data.frame(
  Transform = fmg_transforms,
  `Score base` = vapply(fmg_transforms,
    function(t) mean(data[[score_cols[[t]]]] < .05, na.rm = TRUE), numeric(1L)),
  `LR base` = vapply(fmg_transforms,
    function(t) mean(data[[lr_cols[[t]]]] < .05, na.rm = TRUE), numeric(1L)),
  check.names = FALSE, row.names = NULL)
base_compare_summary <- function(data) data.frame(
  Transform = fmg_transforms,
  `Score base` = vapply(fmg_transforms, function(t)
    mean(data[[sub("^p_", "rejection_", score_cols[[t]])]], na.rm = TRUE), numeric(1L)),
  `LR base` = vapply(fmg_transforms, function(t)
    mean(data[[sub("^p_", "rejection_", lr_cols[[t]])]], na.rm = TRUE), numeric(1L)),
  check.names = FALSE, row.names = NULL)
base_null <- base_compare_reps(r)
base_stress <- base_compare_summary(stress[stress$df == 8L, , drop = FALSE])

method_size <- do.call(rbind, lapply(names(method_labels), function(v) {
  data.frame(
    method = unname(method_labels[[v]]),
    overall = mean(r[[v]] < .05, na.rm = TRUE),
    df1 = mean(r[[v]][r$df == 1L] < .05, na.rm = TRUE),
    df4 = mean(r[[v]][r$df == 4L] < .05, na.rm = TRUE),
    df8 = mean(r[[v]][r$df == 8L] < .05, na.rm = TRUE))
}))

power_methods <- intersect(names(method_labels), names(power))

raw_power <- do.call(rbind, lapply(power_methods, function(v) {
  out <- aggregate(power[[v]] < .05,
                   power[c("alternative", "effect")], mean, na.rm = TRUE)
  names(out)[[3L]] <- "power"
  out$method <- unname(method_labels[[v]])
  out
}))
raw_power <- reshape(raw_power, idvar = c("alternative", "effect"),
                     timevar = "method", direction = "wide")
names(raw_power) <- sub("^power\\.", "", names(raw_power))

# Use the largest strict critical value whose observed matched-cell null size
# does not exceed alpha. This respects the discreteness of the 499-flip tests.
empirical_critical <- function(x, alpha = .05) {
  x <- x[is.finite(x)]
  candidates <- sort(unique(x))
  rates <- vapply(candidates, function(cut) mean(x < cut), numeric(1L))
  valid <- candidates[rates <= alpha]
  if (!length(valid)) -Inf else max(valid)
}

keys <- c("df", "n_total", "balance", "heterogeneity", "distribution")
power_cells <- unique(power[c(keys, "alternative", "effect")])
calibrated <- do.call(rbind, lapply(seq_len(nrow(power_cells)), function(i) {
  cell <- power_cells[i, , drop = FALSE]
  null_match <- rep(TRUE, nrow(r))
  power_match <- rep(TRUE, nrow(power))
  for (key in keys) {
    null_match <- null_match & r[[key]] == cell[[key]]
    power_match <- power_match & power[[key]] == cell[[key]]
  }
  power_match <- power_match & power$alternative == cell$alternative &
    power$effect == cell$effect
  do.call(rbind, lapply(power_methods, function(v) {
    null_p <- r[[v]][null_match]
    alt_p <- power[[v]][power_match]
    critical <- empirical_critical(null_p)
    data.frame(
      cell, method = unname(method_labels[[v]]), critical = critical,
      achieved_size = mean(null_p < critical, na.rm = TRUE),
      calibrated_power = mean(alt_p < critical, na.rm = TRUE))
  }))
}))

calibrated_power <- aggregate(
  calibrated[c("achieved_size", "calibrated_power")],
  calibrated[c("alternative", "effect", "method")], mean)
calibrated_power <- reshape(
  calibrated_power, idvar = c("alternative", "effect"),
  timevar = "method", direction = "wide")
names(calibrated_power) <- sub("^calibrated_power\\.", "", names(calibrated_power))

overall_calibrated <- aggregate(
  calibrated[c("achieved_size", "calibrated_power")],
  calibrated[c("method")], mean)

power_by_df <- aggregate(
  calibrated$calibrated_power,
  calibrated[c("df", "alternative", "effect", "method")], mean)
names(power_by_df)[[5L]] <- "power"
power_by_df <- subset(
  power_by_df,
  method %in% c("flip (standardized)", "score: MV", "LR: MV", "LR: pEBA4"))

null_elapsed <- as.numeric(m$value[m$key == "elapsed_seconds"])
power_elapsed <- as.numeric(power_metadata$value[power_metadata$key == "elapsed_seconds"])
parallel_timing <- data.frame(
  stage = c("two fits", "499-flip score battery", "nested LR battery",
            "FMG battery"),
  median_ms = 1000 * c(
    median(r$fit_seconds, na.rm = TRUE),
    median(r$flip_elapsed_seconds, na.rm = TRUE),
    median(r$nested_seconds, na.rm = TRUE),
    median(r$fmg_seconds, na.rm = TRUE)))
```

## Bottom line

Standardization is cheap and directionally useful, but its practical value is
concentrated. Across `r format(nrow(r), big.mark = ",")` successful null
replications, it changed `r sum(r$changed_to_accept)` decisions from rejection
to acceptance and `r sum(r$changed_to_reject)` in the opposite direction. The
mean p-value increased from the effective reference by
`r sprintf("%.4f", mean(r$p_standardized - r$p_effective))`; no replication
had a smaller standardized p-value.

The larger comparator result reframes the question. Every robust test here
carries the same FMG spectrum transform (SB, SS, MV, SF, pEBA, ALL) on one of
two base statistics: the score/Lagrange-multiplier statistic's own eigenvalues,
or the Satorra-2000 likelihood-ratio construction that lavaan and Mplus use by
default. **The base statistic dominates the transform.** Across the null grid
the score base is near nominal for every transform (about
`r sprintf("%.2f", min(base_null[["Score base"]]))`--`r sprintf("%.2f", max(base_null[["Score base"]]))`),
while the identical transforms on the LR base are uniformly liberal (about
`r sprintf("%.2f", min(base_null[["LR base"]]))`--`r sprintf("%.2f", max(base_null[["LR base"]]))`).
The standardized flip test sits at
`r sprintf("%.3f", method_size$overall[method_size$method == "flip (standardized)"])`.
Matched-empirical-null calibration equalizes the flip and LR tests on power, so
the LR family's larger nominal power was mostly size distortion, not a real
advantage; but the simpler fix is to run the FMG transforms on the score base,
where recalibration is barely needed.

The strongest interaction was sample size by restriction dimension:

```{r n-df-table}
tab <- by_ndf
names(tab) <- c("total N", "df", "effective", "standardized",
                "effective-only rejection")
knitr::kable(tab, digits = 3)
```

At total $N=60$, standardization removes little for one restriction, but about
0.45 and 0.62 percentage points for 4 and 8 restrictions. It does not fully
repair the 8-df test, which remains liberal at 0.077. At $N=200$ the absolute
changes are only 0.06--0.17 percentage points. The answer is therefore not
"always standardize because the unstandardized test is broken". It is closer
to: **standardize by default because it is inexpensive, but expect a material
calibration difference mainly for small samples and moderate-to-large joint
tests.**

This is where the base statistic matters most. Under a severe non-Gaussian
copula (marginals fixed at skewness 3 and excess kurtosis 21), the standardized
flip rejects about 0.11--0.13 at 8 df and the LR base is far worse, both
persisting as $N$ grows. The score base is the exception: its SS, MV, and ALL
transforms stay near or below nominal even here. So standardization is a cheap
default but not a robustness guarantee, whereas the score-base FMG transforms
are the only members of the battery that hold up when the copula turns severe.

## Hypotheses and what happened

### H1: small samples should matter most — supported

The flip-specific covariance correction depends on group sign sums, whose
relative size shrinks with group sample size. Both the measured covariance
shift and the p-value movement decline rapidly with $N$.

```{r n-table}
names(by_n) <- c("total N", "effective", "standardized",
                 "effective-only rejection", "mean |delta p|",
                 "mean relative covariance shift")
knitr::kable(by_n, digits = 4)
```

### H2: more restrictions should increase the payoff — supported

The mean covariance shift itself is only slightly larger at 8 df, but the
multivariate quadratic aggregates those changes across more directions. The
mean absolute p-value movement rises from 0.0014 at 1 df to 0.0046 at 8 df.

```{r df-table}
names(by_df) <- c("df", "effective", "standardized",
                  "effective-only rejection", "mean |delta p|",
                  "mean relative covariance shift")
knitr::kable(by_df, digits = 4)
```

This makes "many restrictions" a better predictor than "large raw covariance
shift" alone. The replication-level correlation between mean relative
covariance shift and absolute p-value movement was
`r sprintf("%.3f", cor(r$mean_variance_relative_shift, r$abs_delta_p))`.

### H3: imbalance should increase the payoff — weakly supported

The 1:3 design produced more effective-only rejections, but did not have a
larger average covariance shift. Imbalance appears to make a given correction
more consequential near the rejection boundary rather than mechanically
making every flip variance more different.

```{r balance-table}
names(by_balance) <- c("balance", "effective", "standardized",
                       "effective-only rejection")
knitr::kable(by_balance, digits = 4)
```

### H4: heterogeneous populations should matter most — not as stated

Homogeneous, common-scale, and strongly geometry-heterogeneous populations had
similar effective-to-standardized rejection changes. The homogeneous cells
actually had the largest mean relative covariance shift. Raw variance
heterogeneity is therefore not a sufficient proxy: the relevant quantity is
the group-specific tested--nuisance information cross-geometry after the H0
restriction map is applied.

### H5: nonnormality should increase the need — consequential, not causal

The covariance-shift diagnostic was almost unchanged across normal, $t_5$, and
skewed generators. Nonnormality did make both flip tests more liberal, so the
same conservative standardization adjustment bought more calibration under
$t_5$ and skew. It did not eliminate the remaining misspecification effect.
The flat diagnostic is a property of these generators, not of non-normality in
general: all three are close to asymptotically robust. The severe-copula stress
section below shows what happens when the generator is not.

```{r distribution-table}
names(by_dist) <- c("distribution", "effective", "standardized")
knitr::kable(by_dist, digits = 4)
```

## FMG transforms and the choice of base statistic

These comparators are not arbitrary. Foldnes, Moss, and Grønneberg (2026)
document that the two established robust nested tests fail in opposite
directions under non-normality: the Satorra-Bentler scaled test (`SB`) is
chronically liberal, and the scaled-and-shifted test (`SS`) is chronically
conservative. The mean/variance-adjusted (`MV`), scaled-$F$ (`SF`), and
eigenvalue-averaging (EBA/pEBA) families are the proposed repairs. All of them
are transforms of the eigenvalue spectrum of a weighted-chi-square limit.

The key observation is that the score test carries its own spectrum. Its
statistic $T = u^\top V^{-1} u$ over the efficient-score subspace has an
asymptotic weighted-chi-square limit whose eigenvalues come straight from the
model, so the identical FMG transforms can be applied to those eigenvalues
instead of to the Satorra-2000 likelihood-ratio construction that lavaan and
Mplus use by default. The table applies each transform to both base statistics,
on the full null grid.

```{r base-compare-null}
knitr::kable(base_null, digits = 3)
```

The base statistic dominates the transform. On the LR base every transform is
uniformly liberal, around 0.10--0.14; on the score base the same transforms are
near nominal, around 0.03--0.05. The spread across transforms within a base is
small next to the gap between the two bases. It holds even at a single
restriction, where the transforms collapse to nearly one scalar test: the LR
base still rejects about
`r sprintf("%.2f", mean(r$p_fmg_sb[r$df == 1L] < .05, na.rm = TRUE))`
for SB while the score base is already near
`r sprintf("%.2f", mean(r$p_mean_scaled[r$df == 1L] < .05, na.rm = TRUE))`.
On the score base the more aggressive transforms (SS, MV, SF, ALL) are mildly
conservative rather than liberal, which trades a little power for size control.

The LR-base SB/MV/SS values are also a wiring check against the nested-test
scaled, mean/variance-adjusted, and scaled-shifted routes: they agree to a
maximum absolute p-value gap of
`r format(max(r$fmg_duplicate_max_gap, na.rm = TRUE), scientific = TRUE,
digits = 2)`.

## Severe non-normality: the copula, not the marginals

The three generators above (normal, $t_5$, exponential skew) are mild. $t_5$ is
elliptical and the skew cell mixes independent coordinates, so neither breaks
the asymptotic robustness the robust tests rely on. This is why H5 saw so little
movement, not evidence that non-normality is harmless. The lever that actually
breaks these tests is the joint dependence structure, the copula, not the
marginal skewness and kurtosis (Foldnes et al., 2026). Two data sets can share a
covariance matrix and identical marginal moments yet differ in their copula, and
normal-theory ML inference is not asymptotically robust to that difference.

The stress block holds the marginals fixed at a severe level (skewness 3, excess
kurtosis 21) and switches only the copula: Vale-Maurelli is a Gaussian copula,
piecewise-linear is not. Both use magmaan's native generators. The base-statistic
comparison carries straight through: the table applies each FMG transform to the
score and LR bases at the 8-df joint test, averaged over $N$ and the homogeneous
and geometry populations.

```{r base-compare-stress}
knitr::kable(base_stress, digits = 3)
```

The split that was visible in the mild grid becomes a chasm:

- **The score base survives the severe copula; the LR base does not.** On the
  LR base every transform rejects around 0.28--0.39, four to eight times
  nominal. On the score base the more aggressive transforms (SS, MV, SF, ALL)
  stay near or below 0.05, with SB and pEBA4 only mildly liberal. The same
  spectrum transform, read off a different base statistic, is the difference
  between a broken test and a working one.
- **The sign-flip test lands between them and does not survive.** The
  standardized flip rejects about
  `r sprintf("%.2f", stress_row("vm", 8L)[["flip (standardized)"]])`--`r sprintf("%.2f", stress_row("pl", 8L)[["flip (standardized)"]])`,
  better than the LR base but far worse than the score base, and standardization
  does not repair it. The score sign-symmetry the flip relies on is itself
  strained by a severe copula.
- **The copula, not the heterogeneity, is the cause.** In the homogeneous
  populations, where the two groups share covariance and mean structure exactly,
  the 8-df LR-SB test still rejects about
  `r sprintf("%.2f", stress_row("vm", 8L, "homogeneous")[["LR: SB"]])`
  under the Gaussian copula and
  `r sprintf("%.2f", stress_row("pl", 8L, "homogeneous")[["LR: SB"]])`
  under the non-Gaussian one. Same marginals, different copula, very different
  size. Piecewise-linear is uniformly harder than Vale-Maurelli, as the copula
  explanation predicts.

The distortion shrinks with sample size but does not calibrate. For the hardest
cell (piecewise-linear, 8 df, geometry heterogeneity), the LR-SB test rejects
about
`r sprintf("%.2f", stress_row("pl", 8L, "geometry", 60L)[["LR: SB"]])`,
`r sprintf("%.2f", stress_row("pl", 8L, "geometry", 100L)[["LR: SB"]])`,
and `r sprintf("%.2f", stress_row("pl", 8L, "geometry", 200L)[["LR: SB"]])`
at $N=60, 100, 200$. This is an asymptotic-robustness failure, not a small-sample
one: it does not vanish as $N$ grows. The score-base transforms are the only
members of the battery that hold near nominal across the whole stress block.

## Dimension scaling: does the score base keep its edge?

Everything so far is at ten observed variables. Foldnes et al. (2026) show that
robust-test over-rejection grows with model dimension, so the score base's
advantage could be a low-dimensional artifact. This probe holds the tested
restriction count fixed at 8 loadings and grows the ambient model from two to
six factors ($p = 10, 20, 30$) under the severe copula, at a matched per-group
sample-to-parameter ratio of about 6. The table averages over the two copulas.

```{r dim-adequate}
knitr::kable(dim_adequate, digits = 3)
```

The advantage is not a ten-variable artifact: the score base stays near nominal
at every dimension while the LR base is liberal at every dimension. But the
governing variable is sample adequacy, not dimension. Because a fixed ratio
means $N$ grows with $p$, the LR base's excess shrinks toward its residual
robustness margin as the model grows (LR-SB falls from about
`r sprintf("%.2f", dim_pick("adequate", 10, "rejection_lr_sb"))` at $p=10$ to
`r sprintf("%.2f", dim_pick("adequate", 30, "rejection_lr_sb"))` at $p=30$),
while the score base stays flat. The exact-mixture transform on the score base
(`score: ALL`) is the most consistently calibrated member across the whole
probe.

Cutting the ratio to about 2.5, an inadequate sample, tells the complementary
story. There the LR base over-rejects worst where $N$ is smallest: at $p=10$,
$N=150$ it reaches about
`r sprintf("%.2f", dim_pick("tight", 10, "rejection_lr_sb"))`, falling to about
`r sprintf("%.2f", dim_pick("tight", 30, "rejection_lr_sb"))` at $p=30$, where
the same ratio buys a larger $N$. The score base holds near nominal throughout.
So the largest score-vs-LR gaps were driven mostly by small-sample
likelihood-ratio over-rejection, not by dimension; the score base's robustness
margin over the LR base is smaller than those gaps suggest but persists as the
model grows.

## Power

The independent power run contains
`r format(nrow(power), big.mark = ",")` successful fits from 81,000 attempts.
For `sparse`, the first tested loading departs by 0.25 or 0.50. For `dense`, the
same Euclidean violation norm is divided equally over all tested restrictions.
The df=1 dense cells are omitted because they duplicate sparse exactly.

### Nominal power is confounded by size

```{r raw-power}
power_show_cols <- function(tab) intersect(c(
  "alternative", "effect", "flip (standardized)", "score: SB", "score: MV",
  "score: pEBA4", "LR: SB", "LR: MV", "LR: pEBA4"), names(tab))
knitr::kable(raw_power[power_show_cols(raw_power)], digits = 3)
```

At face value the LR-base SB and pEBA4 look substantially more powerful. Their
null sizes show why that ranking is misleading. For example, sparse 0.50 nominal
power is about 0.50 for LR-base SB, 0.49 for LR-base pEBA4, and 0.38 for
standardized flips, but their overall null rejection rates are about 0.138,
0.130, and 0.063. The score-base transforms show lower nominal power because
they are near or below nominal size to begin with, not because they detect less.

### Matched empirical-null calibration

For a fairer diagnostic, each method and each of the 162 matched null cells
gets its own critical p-value: the largest strict cutoff whose observed null
rejection rate does not exceed 0.05. Power is then evaluated on independent
alternative replications using that cutoff. This respects the discrete
499-flip distribution. It also consumes only 200 null replications per cell,
so differences of a few points should not be over-interpreted.

```{r calibrated-power}
knitr::kable(calibrated_power[power_show_cols(calibrated_power)], digits = 3)
```

The dramatic nominal separation largely vanishes. Across all four alternative
families, standardized flips, the LR-base MV/pEBA4/ALL, and the score-base
transforms are generally within one or two percentage points once each is held
to its own realized size. The score base's mild conservatism therefore costs
little power after calibration: it buys size control almost for free. Averaged
over all power cells, standardized/effective flips reach about 0.186 calibrated
power versus about 0.184 for the LR-base MV/pEBA4/ALL; that gap is too small to
call a general advantage.

The all-method average makes the comparison among the EBA orders and the basic
FMG variants explicit. `achieved_size` is the average rejection rate obtained
by the matched empirical critical values, not the original nominal size.

```{r all-method-calibrated}
knitr::kable(overall_calibrated, digits = 3)
```

### H6: sparse violations should be easier than dense equal-norm violations — supported

Concentrating the same Euclidean violation in one loading is much easier to
detect. At effect 0.50, calibrated standardized-flip power is about 0.350 for
sparse alternatives and 0.134 for dense alternatives. The comparison is not
available at df=1 because those constructions coincide, but the result persists
within both df=4 and df=8.

### H7: FMG's spectrum transforms should help more with many restrictions — not supported here

Before size calibration, pEBA4's apparent advantage grows with df, exactly
where its null over-rejection also grows. After calibration, pEBA4, MV, and ALL
are nearly indistinguishable. For the sparse 0.50 case, their calibrated power
is around 0.37 at df=4 and 0.28 at df=8. Standardized flips are about 0.39 and
0.30. This probe gives no evidence that a particular FMG transform dominates
once size is aligned.

### H8: more restrictions should always improve power — rejected

Restriction dimension and alternative geometry interact. Dense equal-norm
signals are diluted across coordinates, and the df=8 tests pay a larger
critical-value penalty. Sparse power also falls at df=8 because the joint test
adds seven null directions around one signal. "Many restrictions" predicts
when flip standardization changes calibration, but does not by itself predict
higher power.

## Timing

The controlled benchmark is single-process, warm-cache, and excludes model
fitting. At 499 flips, median core latency was about 2.3 ms at 1 df, 5.4 ms at
4 df, and 8.7 ms at 8 df. Flip-specific standardization accounts for roughly
0.7, 2.7, and 4.2 ms of those totals.

```{r timing-df}
names(timing_499) <- c("df", "setup ms", "basic/effective resampling ms",
                       "standardization ms", "asymptotic comparator ms",
                       "total core ms")
knitr::kable(timing_499, digits = 3)
```

Cost is essentially linear in the number of flips. For the 8-df easy case:

```{r timing-flips}
names(timing_scale) <- c("flips", "basic/effective resampling ms",
                         "standardization ms", "total core ms")
knitr::kable(timing_scale, digits = 3)
```

The two model fits took 3--5 ms in the warmed serial benchmark. In the full
parallel experiment, fit and score timings include worker contention. The
following medians average across restriction dimensions; the complete FMG
battery means all 13 requested transformations, not one p-value.

```{r parallel-timing}
knitr::kable(parallel_timing, digits = 1)
```

The 162-cell null run (200 replications per cell, 499 flips, 16 workers) took
`r sprintf("%.1f", null_elapsed)` seconds, or about
`r sprintf("%.0f", nrow(r) / null_elapsed)` successful replications per
wall-clock second. The 540-cell power run (150 replications per cell) took
`r sprintf("%.1f", power_elapsed)` seconds, or about
`r sprintf("%.0f", nrow(power) / power_elapsed)` per second. In a single call,
the FMG battery's median 8 ms is less than half the median score-flip battery;
one FMG transform would be cheaper still.

## Recommendation

- **Put the FMG spectrum transforms on the score base, not the LR base.** The
  identical transform (SB, SS, MV, SF, pEBA, ALL) is near nominal on the
  score/LM statistic's own eigenvalues and uniformly liberal on the
  Satorra-2000 LR construction that lavaan and Mplus default to. This is the
  largest calibration lever in the study and it costs nothing extra: the score
  eigenvalues are already computed by the flip test.
- Among the score-base transforms, the more aggressive members (SS, MV, ALL)
  are mildly conservative and give the tightest size control, including under
  the severe copula; SB and pEBA4 sit a little closer to nominal. Their
  calibrated power is within a point or two of the flip test, so the
  conservatism costs little.
- Keep `p_standardized` if you want a resampling test: it is cheap and every
  decision change went in the conservative direction. But it is not better
  calibrated than the score-base asymptotics, and it is more strained by severe
  non-normality. Expect the standardization-vs-effective difference to matter
  mainly at total $N\leq100$ with 4--8 restrictions and group imbalance; it is a
  small adjustment, not a small-sample correction.
- Do not compare raw power for any transform without also comparing null size.
  The apparent LR-base pEBA/SB power advantage disappears under matched
  empirical-null calibration.
- Do not extend "robust" to a severe non-Gaussian copula uncritically. The LR
  base and the flip test over-reject badly and do not calibrate as $N$ grows;
  only the score-base transforms hold near nominal there. That edge persists
  from ten to thirty variables at matched sample adequacy, but sample adequacy,
  not dimension, is the governing variable: below about $n/q = 3$ everything
  degrades, the LR base most.
- Next stress tests: hold df fixed while varying nuisance leverage and
  conditioning at fixed dimension and adequacy; and push $p$ further with a
  proportionally larger $N$ to confirm the score base's margin does not close.

## Design and caveats

The null expansion crosses 1/4/8 loading-equality restrictions; total
$N=60/100/200$; 1:1/1:3 group allocation; homogeneous, common latent-scale,
and strong factor/residual-geometry heterogeneity; and normal, $t_5$, and
standardized skewed innovations. Each of 162 null cells has 200 attempted
replications and 499 random flips. There were
`r sum(!read.csv(files[[3L]])$ok)` failed fits (maximum cell failure rate
`r sprintf("%.3f", max(s$failure_rate))`). Monte Carlo and binomial uncertainty
remain visible, especially in cell-by-cell 0.05 decisions.

The power expansion adds sparse/dense alternatives and two effect sizes, for
540 cells and 150 attempts per cell. It had
`r 81000 - nrow(power)` fit failures. Comparator-specific failures are retained
as missing p-values rather than discarding otherwise valid flip results; they
occurred in about `r sprintf("%.3f", mean(nzchar(power$nested_error)))` of
successful power fits.

The severe-copula stress block is a separate 24-cell null grid: two copulas
(Vale-Maurelli and piecewise-linear) at fixed severe marginals, crossed with
1/8 restrictions, $N=60/100/200$, and the homogeneous and geometry populations,
at 200 replications per cell. It is kept out of the main aggregations so it does
not unbalance them. Its purpose is a boundary probe, not a calibrated rejection
estimate; the independent-generator copula was excluded because its marginal
moment solve is infeasible at this severity.

The dimension-scaling probe is a separate 12-cell null grid: two copulas, three
ambient sizes ($p=10/20/30$, two to six factors) at fixed 8-df tested
restrictions, and two per-group sample-to-parameter ratios (about 2.5 and 6),
homogeneous populations, 200 replications per cell. Total $N$ is set from the
target ratio, so it grows with $p$; the ratios are approximate because the free
parameter count is estimated, not read from the fit.

Score flipping here remains a frontier method for affine, complete-data ML
models. These results do not automatically transfer to FIML, ordinal,
estimated-weight, clustered, boundary, or nonlinear-constraint tests.

## Reproduce

```sh
Rscript experiments/62-standardized-flip-score/run_experiment.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --power
Rscript experiments/62-standardized-flip-score/run_expansion.R --stress
Rscript experiments/62-standardized-flip-score/run_dimension.R --probe
Rscript experiments/62-standardized-flip-score/run_timing.R
quarto render experiments/62-standardized-flip-score/report.qmd
```
