---
title: "Standardized score flips for small-sample metric invariance"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
need <- file.path("results", c("summary.csv", "failures.csv", "metadata.csv"))
if (any(!file.exists(need))) {
  stop("missing results; run: Rscript run_experiment.R --probe")
}
s <- read.csv(need[[1L]])
f <- read.csv(need[[2L]])
m <- read.csv(need[[3L]])
focus <- s[!s$alternative & s$distribution == "t5" & s$phi_ratio == 3, ]
wide <- reshape(focus[, c("n", "method", "rejection")], idvar = "n",
                timevar = "method", direction = "wide")
names(wide) <- sub("rejection\\.", "", names(wide))
at50 <- focus[focus$n == 50, ]
rate <- function(method) at50$rejection[match(method, at50$method)]
```

## Question

Can nuisance-effective score sign flips improve the small-sample metric-
invariance test, and does flip-specific variance standardization repair the
finite-sample mismatch between the observed and transformed statistics?

## Short answer

At the focal heavy-tailed, heterogeneous (n=50) cell, rejection was
`r sprintf("%.3f", rate("flip_basic"))` for basic flips,
`r sprintf("%.3f", rate("flip_effective"))` for effective flips, and
`r sprintf("%.3f", rate("flip_standardized"))` after flip-specific
standardization. The corresponding robust score-mixture and Satorra-2000 rates
were `r sprintf("%.3f", rate("score_mixture"))` and
`r sprintf("%.3f", rate("satorra2000_scaled"))`. This is a calibration probe,
not a support claim; the table below shows whether the pattern persists with
sample size.

## Evidence

Heavy-tailed groups with a threefold factor-variance difference; target level
is 0.05.

```{r table}
keep <- c("n", "flip_basic", "flip_effective", "flip_standardized",
          "score_mixture", "satorra2000_scaled")
tab <- wide[, intersect(keep, names(wide)), drop = FALSE]
names(tab) <- c("n per group", "basic flip", "effective flip",
                "standardized flip", "robust score mixture", "Satorra 2000")
knitr::kable(tab, digits = 3)
```

## Caveats

- The probe uses continuous complete-data ML, affine metric restrictions, and
  normal or symmetric heavy-tailed data. Skewed, missing, ordinal, fixed-X, and
  boundary cases remain outside the implemented contract.
- Each cell has `r m$value[m$key == "reps"]` replications and
  `r m$value[m$key == "flips"]` random flips. A 0.05 rejection rate therefore
  has visible binomial and Monte Carlo noise.
- Score flipping is asymptotic for covariance-parameter scores; it is not an
  exact finite-sample randomization test.

## Reproduce

```sh
Rscript experiments/62-standardized-flip-score/run_experiment.R --smoke
Rscript experiments/62-standardized-flip-score/run_experiment.R --probe
quarto render experiments/62-standardized-flip-score/report.qmd
```
