---
title: "Robust invariance testing: score bases, spectral shrinkage, and sandwich inversion"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
files <- file.path("results", c(
  "summary.csv", "expansion_summary.csv", "expansion_replications.csv",
  "expansion_metadata.csv", "timing_summary.csv", "power_summary.csv",
  "power_replications.csv", "power_metadata.csv", "stress_summary.csv",
  "dimension_summary.csv", "sandwich_power_replications.csv",
  "dimension_sandwich_summary.csv", "multiplier_probe_summary.csv",
  "multiplier_probe_replications.csv", "multiplier_probe_moments.csv",
  "multiplier_probe_metadata.csv", "multiplier_refine_summary.csv",
  "multiplier_refine_replications.csv", "multiplier_refine_moments.csv",
  "multiplier_refine_metadata.csv", "multiplier_confirm_summary.csv",
  "multiplier_confirm_replications.csv", "multiplier_confirm_moments.csv",
  "multiplier_confirm_metadata.csv", "multiplier_power_summary.csv",
  "multiplier_power_replications.csv", "multiplier_power_moments.csv",
  "multiplier_power_metadata.csv"))
if (any(!file.exists(files))) {
  stop(paste(
    "missing results; run run_experiment.R, run_expansion.R --probe,",
    "run_expansion.R --power, --power-sandwich, and --stress,",
    "run_dimension.R --big and --sandwich,",
    "run_multiplier_probe.R --probe, --refine, --confirm, and",
    "--confirm-power, and run_timing.R"))
}
s <- read.csv(files[[2L]])
r <- read.csv(files[[3L]]); r_all_ok <- r$ok; r <- r[r$ok, , drop = FALSE]
m <- read.csv(files[[4L]])
timing <- read.csv(files[[5L]])
power <- read.csv(files[[7L]]); power <- power[power$ok, , drop = FALSE]
power_metadata <- read.csv(files[[8L]])
stress <- read.csv(files[[9L]])
dim_summary <- read.csv(files[[10L]])
sandwich_power <- read.csv(files[[11L]])
dim_sandwich <- read.csv(files[[12L]])
multiplier_summary <- read.csv(files[[13L]])
multiplier_reps <- read.csv(files[[14L]])
multiplier_moments <- read.csv(files[[15L]])
multiplier_metadata <- read.csv(files[[16L]])
refine_summary <- read.csv(files[[17L]])
refine_reps <- read.csv(files[[18L]])
refine_moments <- read.csv(files[[19L]])
refine_metadata <- read.csv(files[[20L]])
confirm_summary <- read.csv(files[[21L]])
confirm_reps <- read.csv(files[[22L]])
confirm_moments <- read.csv(files[[23L]])
confirm_metadata <- read.csv(files[[24L]])
multiplier_power_summary <- read.csv(files[[25L]])
multiplier_power_reps <- read.csv(files[[26L]])
multiplier_power_moments <- read.csv(files[[27L]])
multiplier_power_metadata <- read.csv(files[[28L]])

# The one-flip replay uses the same deterministic cells and replications as the
# original power grid; merge only the new asymptotic sandwich result.
merge_keys <- c("rep", "df", "n_total", "balance", "heterogeneity",
                "distribution", "alternative", "effect")
sandwich_power <- sandwich_power[sandwich_power$ok,
  c(merge_keys, "p_sandwich", "sandwich_available", "sandwich_condition")]
power <- merge(power, sandwich_power, by = merge_keys, all.x = TRUE,
               sort = FALSE)

# Method naming: {flip | score | LR} x transform. Every method is built on the
# same efficient score; what differs is how the reference is obtained (flip =
# sign-flip resampling; score/LR = FMG spectrum transforms on the score
# statistic's own eigenvalues, or on the Satorra-2000 LR construction).
method_labels <- c(
  p_effective    = "flip (effective)",
  p_standardized = "flip (standardized)",
  p_chisq        = "score: chi-square",
  p_mean_scaled  = "score: SB",
  p_sandwich     = "score: sandwich",
  p_score_ss     = "score: SS",
  p_score_mv     = "score: MV",
  p_score_sf     = "score: SF",
  p_score_peba4  = "score: pEBA4",
  p_mixture      = "score: ALL",
  p_fmg_sb       = "LR: SB",
  p_fmg_ss       = "LR: SS",
  p_fmg_mv       = "LR: MV",
  p_fmg_sf       = "LR: SF",
  p_fmg_peba4    = "LR: pEBA4",
  p_fmg_all      = "LR: ALL")

# Same FMG transform on two base statistics: score vs LR.
fmg_transforms <- c("SB", "SS", "MV", "SF", "pEBA4", "ALL")
score_cols <- c(SB = "p_mean_scaled", SS = "p_score_ss", MV = "p_score_mv",
                SF = "p_score_sf", pEBA4 = "p_score_peba4", ALL = "p_mixture")
lr_cols <- c(SB = "p_fmg_sb", SS = "p_fmg_ss", MV = "p_fmg_mv",
             SF = "p_fmg_sf", pEBA4 = "p_fmg_peba4", ALL = "p_fmg_all")
base_compare_reps <- function(data) data.frame(
  Transform = fmg_transforms,
  `Score base` = vapply(fmg_transforms,
    function(t) mean(data[[score_cols[[t]]]] < .05, na.rm = TRUE), numeric(1L)),
  `LR base` = vapply(fmg_transforms,
    function(t) mean(data[[lr_cols[[t]]]] < .05, na.rm = TRUE), numeric(1L)),
  check.names = FALSE, row.names = NULL)
base_compare_summary <- function(data) data.frame(
  Transform = fmg_transforms,
  `Score base` = vapply(fmg_transforms, function(t)
    mean(data[[sub("^p_", "rejection_", score_cols[[t]])]], na.rm = TRUE), numeric(1L)),
  `LR base` = vapply(fmg_transforms, function(t)
    mean(data[[sub("^p_", "rejection_", lr_cols[[t]])]], na.rm = TRUE), numeric(1L)),
  check.names = FALSE, row.names = NULL)
base_null <- base_compare_reps(r)
base_stress <- base_compare_summary(stress[stress$df == 8L, , drop = FALSE])
flip_std_overall <- mean(r$p_standardized < .05, na.rm = TRUE)
score_null_by_df <- aggregate(
  r[c("p_sandwich", "p_mean_scaled", "p_score_peba4", "p_mixture",
      "p_standardized")] < .05,
  r["df"], mean, na.rm = TRUE)
names(score_null_by_df) <- c("df", "sandwich", "SB", "pEBA4", "ALL", "flip")
sandwich_diag <- aggregate(
  r[c("sandwich_condition", "score_eigen_cv")],
  r[c("df", "n_total")], median, na.rm = TRUE)

# Normal-data N-convergence: with no non-normality, any gap is small-sample.
norm_conv <- do.call(rbind, lapply(c(60L, 100L, 200L), function(nn) {
  sub <- r[r$distribution == "normal" & r$n_total == nn, , drop = FALSE]
  data.frame(`Total N` = nn,
    `score: SB` = mean(sub$p_mean_scaled < .05, na.rm = TRUE),
    `score: sandwich` = mean(sub$p_sandwich < .05, na.rm = TRUE),
    `LR: SB` = mean(sub$p_fmg_sb < .05, na.rm = TRUE),
    `flip (standardized)` = mean(sub$p_standardized < .05, na.rm = TRUE),
    check.names = FALSE)
}))
t5_lr_sb_n200 <- mean(r$p_fmg_sb[r$distribution == "t5" & r$n_total == 200L] < .05,
                      na.rm = TRUE)

# Severe-copula stress helpers.
stress_cols <- c(
  `flip (standardized)` = "rejection_standardized",
  `score: SB`           = "rejection_mean_scaled",
  `score: sandwich`     = "rejection_sandwich",
  `score: ALL`          = "rejection_mixture",
  `LR: SB`              = "rejection_fmg_sb",
  `LR: pEBA4`           = "rejection_fmg_peba4")
stress_row <- function(d, dff, het = NULL, nn = NULL) {
  sub <- stress[stress$distribution == d & stress$df == dff, , drop = FALSE]
  if (!is.null(het)) sub <- sub[sub$heterogeneity == het, , drop = FALSE]
  if (!is.null(nn)) sub <- sub[sub$n_total == nn, , drop = FALSE]
  vapply(stress_cols, function(cc) mean(sub[[cc]], na.rm = TRUE), numeric(1L))
}

# Dimension-scaling helpers.
dim_cols <- c(`score: SB` = "rejection_score_sb",
              `score: ALL` = "rejection_score_all",
              `LR: SB` = "rejection_lr_sb",
              `flip (standardized)` = "rejection_flip_standardized")
dim_table <- function(adeq) {
  sub <- dim_summary[dim_summary$adequacy == adeq, , drop = FALSE]
  agg <- aggregate(sub[unname(dim_cols)], list(p = sub$p), mean)
  names(agg) <- c("p", names(dim_cols)); agg
}
dim_adequate <- dim_table("adequate")
dim_pick <- function(adeq, pp, col)
  mean(dim_summary[dim_summary$adequacy == adeq & dim_summary$p == pp, col], na.rm = TRUE)
dim_sandwich_table <- aggregate(
  dim_sandwich[c("rejection_score_sandwich", "sandwich_available_rate",
                 "median_sandwich_condition", "median_score_eigen_cv")],
  dim_sandwich[c("adequacy", "p")], mean, na.rm = TRUE)

# Standardization sub-study: decision changes by N and restriction count.
by_ndf <- aggregate(
  r[c("reject_effective", "reject_standardized", "changed_to_accept")],
  r[c("n_total", "df")], mean)

# Multiplier-law sanity check on the q=8 severe-copula cells where Rademacher
# effective flips were most liberal. All rows use the same data and seeds as
# the original stress block.
multiplier_cols <- c(
  Rademacher = "rejection_rademacher",
  Mammen = "rejection_mammen",
  Gaussian = "rejection_gaussian",
  `centered exponential` = "rejection_centered_exponential",
  sandwich = "rejection_sandwich",
  pEBA4 = "rejection_peba4",
  ALL = "rejection_all")
multiplier_table <- multiplier_summary[
  c("n_total", "distribution", "heterogeneity", unname(multiplier_cols))]
names(multiplier_table) <- c(
  "N", "copula", "geometry", names(multiplier_cols))
multiplier_overall <- vapply(
  unname(multiplier_cols),
  function(v) mean(multiplier_summary[[v]], na.rm = TRUE),
  numeric(1L))
names(multiplier_overall) <- names(multiplier_cols)
names(multiplier_moments) <- c(
  "multiplier", "mean", "variance", "third moment", "fourth moment")
refine_cols <- c(
  `two-point gamma=0.50` = "rejection_two_point_0_50",
  `two-point gamma=0.75` = "rejection_two_point_0_75",
  `two-point gamma=0.90` = "rejection_two_point_0_90",
  Mammen = "rejection_two_point_1_00",
  `centered Mammen` = "rejection_mammen_centered",
  `weighted-meat Mammen` = "rejection_mammen_weighted_meat",
  `centered weighted-meat` =
    "rejection_mammen_centered_weighted_meat",
  sandwich = "rejection_sandwich",
  pEBA4 = "rejection_peba4")
refine_table <- refine_summary[
  c("n_total", "distribution", "heterogeneity", unname(refine_cols))]
names(refine_table) <- c("N", "copula", "geometry", names(refine_cols))
refine_overall <- vapply(
  unname(refine_cols),
  function(v) mean(refine_summary[[v]], na.rm = TRUE),
  numeric(1L))
names(refine_overall) <- names(refine_cols)
refine_rank <- data.frame(
  method = names(refine_cols),
  rejection = unname(refine_overall),
  minimum = vapply(
    unname(refine_cols), function(v) min(refine_summary[[v]]), numeric(1L)),
  maximum = vapply(
    unname(refine_cols), function(v) max(refine_summary[[v]]), numeric(1L)),
  mean_absolute_error = vapply(
    unname(refine_cols),
    function(v) mean(abs(refine_summary[[v]] - .05)), numeric(1L)))
refine_rank <- refine_rank[order(refine_rank$mean_absolute_error), ]

# Independent centered-Mammen confirmation.
confirm_cols <- c(
  `raw Mammen` = "rejection_mammen",
  `centered Mammen` = "rejection_mammen_centered",
  sandwich = "rejection_sandwich",
  pEBA4 = "rejection_peba4")
confirm_rank <- data.frame(
  method = names(confirm_cols),
  rejection = vapply(
    unname(confirm_cols),
    function(v) mean(confirm_summary[[v]], na.rm = TRUE), numeric(1L)),
  N60 = vapply(
    unname(confirm_cols),
    function(v) mean(confirm_summary[[v]][confirm_summary$n_total == 60L]),
    numeric(1L)),
  N200 = vapply(
    unname(confirm_cols),
    function(v) mean(confirm_summary[[v]][confirm_summary$n_total == 200L]),
    numeric(1L)),
  minimum = vapply(
    unname(confirm_cols), function(v) min(confirm_summary[[v]]), numeric(1L)),
  maximum = vapply(
    unname(confirm_cols), function(v) max(confirm_summary[[v]]), numeric(1L)),
  mean_absolute_error = vapply(
    unname(confirm_cols),
    function(v) mean(abs(confirm_summary[[v]] - .05)), numeric(1L)))
confirm_rank <- confirm_rank[order(confirm_rank$mean_absolute_error), ]

# Matched-empirical-null (size-adjusted) power.
power_methods <- intersect(names(method_labels), names(power))
empirical_critical <- function(x, alpha = .05) {
  x <- x[is.finite(x)]
  candidates <- sort(unique(x))
  rates <- vapply(candidates, function(cut) mean(x < cut), numeric(1L))
  valid <- candidates[rates <= alpha]
  if (!length(valid)) -Inf else max(valid)
}
keys <- c("df", "n_total", "balance", "heterogeneity", "distribution")
power_cells <- unique(power[c(keys, "alternative", "effect")])
calibrated <- do.call(rbind, lapply(seq_len(nrow(power_cells)), function(i) {
  cell <- power_cells[i, , drop = FALSE]
  null_match <- rep(TRUE, nrow(r)); power_match <- rep(TRUE, nrow(power))
  for (key in keys) {
    null_match <- null_match & r[[key]] == cell[[key]]
    power_match <- power_match & power[[key]] == cell[[key]]
  }
  power_match <- power_match & power$alternative == cell$alternative &
    power$effect == cell$effect
  do.call(rbind, lapply(power_methods, function(v) {
    critical <- empirical_critical(r[[v]][null_match])
    data.frame(cell, method = unname(method_labels[[v]]),
      calibrated_power = mean(power[[v]][power_match] < critical, na.rm = TRUE))
  }))
}))
calibrated_power <- aggregate(calibrated["calibrated_power"],
  calibrated[c("alternative", "effect", "method")], mean)
calibrated_power <- reshape(calibrated_power,
  idvar = c("alternative", "effect"), timevar = "method", direction = "wide")
names(calibrated_power) <- sub("^calibrated_power\\.", "", names(calibrated_power))
power_show_cols <- function(tab) intersect(c(
  "alternative", "effect", "flip (standardized)", "score: sandwich",
  "score: SB", "score: MV",
  "score: pEBA4", "LR: SB", "LR: MV", "LR: pEBA4"), names(tab))

# The focused multiplier power gate uses the independent confirmation null,
# cell-specific critical p-values, and the same four surviving methods.
confirm_reps <- confirm_reps[confirm_reps$ok, , drop = FALSE]
multiplier_power_reps <-
  multiplier_power_reps[multiplier_power_reps$ok, , drop = FALSE]
confirm_power_cols <- c(
  `raw Mammen` = "p_mammen",
  `centered Mammen` = "p_mammen_centered",
  sandwich = "p_sandwich",
  pEBA4 = "p_peba4")
confirm_keys <- c(
  "df", "n_total", "balance", "heterogeneity", "distribution")
confirm_power_cells <- unique(multiplier_power_reps[
  c(confirm_keys, "alternative", "effect")])
matched_multiplier_power <- do.call(
  rbind, lapply(seq_len(nrow(confirm_power_cells)), function(i) {
    cell <- confirm_power_cells[i, , drop = FALSE]
    null_match <- rep(TRUE, nrow(confirm_reps))
    power_match <- rep(TRUE, nrow(multiplier_power_reps))
    for (key in confirm_keys) {
      null_match <- null_match & confirm_reps[[key]] == cell[[key]]
      power_match <-
        power_match & multiplier_power_reps[[key]] == cell[[key]]
    }
    power_match <-
      power_match &
      multiplier_power_reps$alternative == cell$alternative &
      multiplier_power_reps$effect == cell$effect
    do.call(rbind, lapply(names(confirm_power_cols), function(method) {
      value <- confirm_power_cols[[method]]
      critical <- empirical_critical(confirm_reps[[value]][null_match])
      data.frame(
        cell, method = method, critical = critical,
        null_size = mean(confirm_reps[[value]][null_match] < critical),
        nominal_power =
          mean(multiplier_power_reps[[value]][power_match] < .05),
        calibrated_power =
          mean(multiplier_power_reps[[value]][power_match] < critical))
    }))
  }))
matched_power_overall <- aggregate(
  matched_multiplier_power[
    c("null_size", "nominal_power", "calibrated_power")],
  matched_multiplier_power["method"], mean)
matched_power_overall <- matched_power_overall[
  match(names(confirm_power_cols), matched_power_overall$method), ]
names(matched_power_overall) <- c(
  "method", "matched null size", "nominal power", "matched-null power")
matched_power_cells <- reshape(
  matched_multiplier_power[
    c(confirm_keys, "method", "calibrated_power")],
  idvar = confirm_keys, timevar = "method", direction = "wide")
centered_raw_power_difference <-
  matched_power_cells$`calibrated_power.centered Mammen` -
  matched_power_cells$`calibrated_power.raw Mammen`

# Timing.
timing_ms <- timing
for (name in grep("seconds$", names(timing), value = TRUE))
  timing_ms[[sub("seconds$", "ms", name)]] <- 1000 * timing[[name]]
timing_499 <- subset(timing_ms, scenario == "easy" & n_flips == 499,
  select = c(df, median_setup_ms, median_resampling_score_ms,
             median_standardization_ms, median_asymptotic_ms, median_core_total_ms))
null_elapsed <- as.numeric(m$value[m$key == "elapsed_seconds"])
```

## Question

When you test measurement invariance under non-normality, which robust test
controls Type I error? We compare the sign-flip resampling test against the
Foldnes-Moss-Grønneberg spectrum transforms (Satorra-Bentler `SB`,
scaled-shifted `SS`, mean-variance `MV`, scaled-$F$ `SF`, `pEBA`, and the exact
mixture `ALL`), each applied to two base statistics: the score/Lagrange-multiplier
statistic, and the Satorra-2000 likelihood-ratio statistic that lavaan and Mplus
use by default. We also include the direct sandwich-studentized score statistic
$u^\top(G^\top\hat B G)^{-1}u$, whose asymptotic reference is $\chi_q^2$.
The focused follow-up asks whether Mammen multiplier weights, with score rows
recentered to impose the resampling null, improve the effective-score
resampling reference without losing matched-null power.

## Short answer

The score base is substantially better behaved than the Satorra-2000 LR base,
but the transform still matters. Score-SB's overall null rejection of
`r sprintf("%.3f", mean(r$p_mean_scaled < .05, na.rm = TRUE))` looks almost
perfect because it averages an exact one-df case with relatively spherical
population spectra and because trace-only scaling regularizes noisy sample
eigenvalues. At eight restrictions it rejects
`r sprintf("%.3f", mean(r$p_mean_scaled[r$df == 8L] < .05, na.rm = TRUE))`;
under the severe copulas it reaches about 0.08. It is not generally exact.

The direct sandwich score test supplies the missing asymptotically pivotal
comparison, but it moves too far in the other direction at small $N$: overall
null rejection is `r sprintf("%.3f", mean(r$p_sandwich < .05, na.rm = TRUE))`
and only `r sprintf("%.3f", mean(r$p_sandwich[r$df == 8L] < .05, na.rm = TRUE))`
at eight restrictions. Full meat inversion amplifies estimation noise and can
be poorly conditioned. In this probe pEBA4 gives the best compromise: it shrinks
the spectrum without discarding all shape information or inverting the raw meat.

Much of the score-versus-LR gap remains the familiar small-sample difference
between LM and LR bases. The revised conclusion is about regularization, not a
universally winning base or transform: score-SB is trace shrinkage, ALL is a
noisy plug-in mixture, and direct sandwich inversion is asymptotically clean but
finite-sample fragile.

Centered Mammen is the promising resampling repair. In a fresh 12-cell
normal/$t_5$/skew confirmation it rejects 5.2% overall, versus 4.5% for raw
Mammen, 3.8% for direct sandwich, and 5.3% for pEBA4. After cellwise
empirical-null calibration, its sparse-alternative power is 35.6%, essentially
the same as raw Mammen's 35.3% and pEBA4's 36.0%. This is evidence for
recentering rather than tuned multiplier moments, but only for complete-data
affine ML at eight restrictions; it is not yet a general default.

## Score base helps, but the transform still matters

The Foldnes, Moss, and Grønneberg (2026) transforms exist because the two
established robust nested tests fail in opposite directions under non-normality:
Satorra-Bentler (`SB`) is chronically liberal and scaled-shifted (`SS`) is
chronically conservative. `MV`, `SF`, and the eigenvalue-averaging families
(`EBA`, `pEBA`) are the proposed repairs. All of them transform the eigenvalue
spectrum of a weighted-chi-square limit.

The score test carries its own spectrum. Its statistic
$T = u^\top V^{-1} u$ over the efficient-score subspace has an asymptotic
weighted-chi-square limit whose eigenvalues come straight from the model, so the
identical transforms can be applied to those eigenvalues instead of to the
Satorra-2000 construction. The table applies each transform to both bases, on the
full null grid.

```{r base-null}
knitr::kable(base_null, digits = 3)
```

On the LR base every transform is liberal. On the score base the methods span a
meaningful range: SB and pEBA4 sit near nominal overall, while SS, MV, SF, and
ALL are conservative. The gap holds even at a single restriction, where the transforms
collapse to nearly one scalar test: there the LR base still rejects about
`r sprintf("%.2f", mean(r$p_fmg_sb[r$df == 1L] < .05, na.rm = TRUE))` for `SB`
while the score base is already near
`r sprintf("%.2f", mean(r$p_mean_scaled[r$df == 1L] < .05, na.rm = TRUE))`.

### Why score-SB looked suspiciously good

The normal-theory efficient-score statistic has limit
$\sum_j\lambda_j\chi^2_{1,j}$. SB replaces every eigenvalue by their mean.
This is exact at one df and for a spherical population spectrum, but its
reference variance $2q\bar\lambda^2$ is smaller than the true
$2\sum_j\lambda_j^2$ whenever the eigenvalues differ.

Finite-sample eigenvalue noise reverses the practical comparison with ALL. At
$N=200,q=8$, median estimated spectral CV remains about
`r sprintf("%.2f", median(r$score_eigen_cv[r$n_total == 200L & r$df == 8L]))`.
The plug-in ALL tail treats this noise as population heterogeneity and becomes
conservative. SB retains only the more stable trace, acting as maximal spectral
shrinkage. Disaggregating by df prevents the exact scalar case from hiding the
joint-test behavior.

```{r score-null-by-df}
knitr::kable(score_null_by_df, digits = 3)
```

The direct sandwich statistic avoids the mixture approximation, but it must
invert the full efficient meat. It is available in every successful main-grid
fit, yet conditioning worsens rapidly with restriction dimension and shrinking
sample size.

```{r sandwich-diagnostics}
knitr::kable(sandwich_diag, digits = 2)
```

The derivation, diagnostic values, and iteration checklist are preserved in
[`notes/score-sb-audit.tex`](notes/score-sb-audit.tex).

### How much of the base gap is small-sample

The efficient-score construction is Neyman's C(α) test: the score of the tested
parameters, orthogonalized against the nuisance parameters and standardized by
the efficient information. In SEM that is exactly what a modification index
computes, only univariate. Assembled into a joint test with a spectrum reference
it is a robust Lagrange-multiplier test, and Lagrange-multiplier tests are the
least liberal of the classical trio in small samples (the Berndt-Savin ordering
places them below the likelihood-ratio and Wald statistics). So part of the
score base's advantage is the documented small-sample over-rejection of the LR
statistic, not robustness.

The normal-data cells separate the two effects. With no non-normality to correct,
any gap must be small-sample.

```{r norm-convergence}
knitr::kable(norm_conv, digits = 3)
```

Under normality the LR base falls from about 0.12 at $N=60$ toward nominal as $N$
grows, while the score base stays flat near 0.05. Adequate sample size removes
that part of the gap. What remains under non-normality is the genuine robustness
margin: under $t_5$ the LR base is still about
`r sprintf("%.2f", t5_lr_sb_n200)` at $N=200$, where the score base is near 0.05.
That margin is smaller than the raw gap but does not wash out.

As a validity check, the LR-base `SB`, `MV`, and `SS` values reproduce the
independent nested-test scaled, mean-variance, and scaled-shifted routes to a
maximum p-value gap of
`r format(max(r$fmg_duplicate_max_gap, na.rm = TRUE), scientific = TRUE, digits = 2)`.

## The copula, not the marginals

Mild non-normality does not stress these tests: an elliptical $t_5$ or a
coordinatewise skew stays close to asymptotic robustness. The lever that breaks
them is the joint dependence structure, the copula, not the marginal skewness and
kurtosis (Foldnes et al., 2026). Two data sets can share a covariance matrix and
identical marginal moments yet differ in their copula, and normal-theory ML
inference is not asymptotically robust to that difference.

The stress block fixes the marginals at a severe level (skewness 3, excess
kurtosis 21) and switches only the copula: Vale-Maurelli is a Gaussian copula,
piecewise-linear is not. The table applies each transform to both bases at the
8-df joint test, averaged over $N$ and the homogeneous and geometry populations.

```{r base-stress}
knitr::kable(base_stress, digits = 3)
```

The LR base fails most dramatically, but the score methods do not share one
calibration. Score-SB rejects about 0.08, ALL about 0.03, and pEBA4 about 0.06.
Direct sandwich ranges from about
`r sprintf("%.2f", stress_row("pl", 8L)[["score: sandwich"]])` under PL to
`r sprintf("%.2f", stress_row("vm", 8L)[["score: sandwich"]])` under VM; its
mean cellwise meat condition numbers are roughly 251 and 96. The sign-flip test
is more liberal still, about
`r sprintf("%.2f", stress_row("vm", 8L)[["flip (standardized)"]])` to
`r sprintf("%.2f", stress_row("pl", 8L)[["flip (standardized)"]])`, and
standardization does not repair it: the score sign-symmetry it relies on is
itself strained by a severe copula. pEBA4 is the only score method that remains
consistently close to 0.05 across both severe copulas.

The copula, not group heterogeneity, is the cause. In homogeneous populations,
where the two groups share covariance and means exactly, the 8-df LR-`SB` test
still rejects about
`r sprintf("%.2f", stress_row("vm", 8L, "homogeneous")[["LR: SB"]])` under the
Gaussian copula and
`r sprintf("%.2f", stress_row("pl", 8L, "homogeneous")[["LR: SB"]])` under the
non-Gaussian one. Piecewise-linear is uniformly harder than Vale-Maurelli, as the
copula explanation predicts. The distortion shrinks with sample size but does not
calibrate: for the hardest cell the LR-`SB` test goes
`r sprintf("%.2f", stress_row("pl", 8L, "geometry", 60L)[["LR: SB"]])`,
`r sprintf("%.2f", stress_row("pl", 8L, "geometry", 100L)[["LR: SB"]])`,
`r sprintf("%.2f", stress_row("pl", 8L, "geometry", 200L)[["LR: SB"]])` at
$N=60, 100, 200$. This is an asymptotic-robustness failure, not a small-sample
one.

## Dimension scaling

Foldnes et al. (2026) show robust-test over-rejection grows with model dimension,
so the score base's edge might be a low-dimensional artifact. This probe holds the
tested restriction count fixed at 8 loadings and grows the ambient model from two
to eight factors ($p = 10, 20, 30, 40$) under the severe copula, over both
population geometries and both copulas, at a matched per-group
sample-to-parameter ratio of about 6 and 1700 replications per cell. The table
averages over geometry and copula.

```{r dim-adequate}
knitr::kable(dim_adequate, digits = 3)
```

The score base remains less liberal than the LR base, but its transforms bracket
0.05 rather than agreeing: SB runs high at tight adequacy, while ALL runs low.
Sample adequacy, not dimension alone, governs. Because a fixed ratio
grows $N$ with $p$, the LR base's excess shrinks toward its residual robustness
margin as the model grows, from about
`r sprintf("%.2f", dim_pick("adequate", 10, "rejection_lr_sb"))` at $p=10$ to
`r sprintf("%.2f", dim_pick("adequate", 40, "rejection_lr_sb"))` at $p=40$. The
largest gaps sit at small $p$ and small $N$, where the small-sample
likelihood-ratio effect is strongest: at an inadequate ratio near 2.5 the LR base
reaches about `r sprintf("%.2f", dim_pick("tight", 10, "rejection_lr_sb"))` at
$p=10$, falling to `r sprintf("%.2f", dim_pick("tight", 40, "rejection_lr_sb"))`
at $p=40$ where the same ratio buys a larger $N$. Score-SB moves toward nominal
as adequacy increases; pEBA4 is more stable around 0.05, whereas ALL remains
conservative between about
`r sprintf("%.2f", min(dim_summary$rejection_score_all, na.rm = TRUE))` and
`r sprintf("%.2f", max(dim_summary$rejection_score_all, na.rm = TRUE))`.

The direct-sandwich replay uses the same 32 design cells with 200 replications
and one nominal flip per replication. It is intentionally separate from the
1,700-replication spectrum study so the original precision is retained.

```{r dim-sandwich}
knitr::kable(dim_sandwich_table, digits = 3)
```

This tests asymptotic recovery directly: sandwich rejection should approach
0.05 and meat conditioning should improve as adequacy and total information
grow. At the tight end it is a diagnostic, not a recommended reference.

## Power

Nominal power ranks the LR base above the score base, but only because the LR
base rejects more of everything. The honest comparison holds each method to its
own realized null size: each method and matched null cell gets its own critical
p-value (the largest cutoff whose observed null rejection does not exceed 0.05),
and power is read off independent alternative replications.

```{r calibrated-power}
knitr::kable(calibrated_power[power_show_cols(calibrated_power)], digits = 3)
```

Once size is matched, the flip, spectrum-score, LR-base, and direct-sandwich
tests are broadly similar for weak dense effects. Direct sandwich remains lower
for the stronger alternatives even after size alignment, suggesting a power
cost from noisy directional meat inversion rather than nominal conservatism
alone. As
expected, a sparse violation concentrated in one loading is easier to detect than
the same norm spread across all tested loadings, and no single spectrum transform
dominates once size is aligned.

## Standardizing the sign-flip test

Standardization is the original subject of this probe and it is a minor effect.
Across `r format(nrow(r), big.mark = ",")` null replications it moved
`r sum(r$changed_to_accept)` decisions from rejection to acceptance and
`r sum(r$changed_to_reject)` the other way, always in the conservative direction.
Its value concentrates at small $N$ with several restrictions.

```{r n-df-table}
tab <- by_ndf
names(tab) <- c("total N", "df", "effective", "standardized",
                "effective-only rejection")
knitr::kable(tab, digits = 3)
```

At total $N=60$ standardization removes little for one restriction but a few
tenths of a percentage point for four and eight, and it does not fully repair the
liberal 8-df test. At $N=200$ the changes are negligible. Standardize by default
because it is nearly free, but do not read it as a small-sample correction.

## Changing the multiplier law

The effective score reference does not mathematically require the multiplier to
be a sign. For fixed fitted score rows, draw independent weights with
$E(w_i)=0$ and $E(w_i^2)=1$ and recompute

$$
T^\ast =
\left(n^{-1/2}\sum_i w_i\hat u_i\right)^\top
\hat V^{-1}
\left(n^{-1/2}\sum_i w_i\hat u_i\right).
$$

The first two moments preserve the conditional covariance, while the third and
fourth moments control finite-sample skewness and tail thickness. We replayed
only the eight 8-df severe-copula cells where Rademacher effective flips had
performed poorly, using the original observations and random seeds, 200
replications per cell, and 499 draws per law. The alternatives were:

- Rademacher signs, the existing symmetric two-point law and the only option
  with a sign-randomization interpretation;
- Mammen's asymmetric two-point law, which matches a nonzero third moment while
  keeping a modest fourth moment;
- standard Gaussian weights, whose conditional quadratic-form law converges
  exactly to score-ALL for the fitted score geometry;
- centered unit-rate exponential weights, an intentionally highly skewed and
  heavy-tailed stress case.

```{r multiplier-moments}
knitr::kable(multiplier_moments, digits = 0)
```

```{r multiplier-null}
knitr::kable(multiplier_table, digits = 3)
```

Rademacher replayed the old effective-flip p-values exactly and remained badly
liberal: its rejection rate averaged
`r sprintf("%.3f", multiplier_overall[["Rademacher"]])`. Gaussian weights
averaged `r sprintf("%.3f", multiplier_overall[["Gaussian"]])`, close to ALL's
`r sprintf("%.3f", multiplier_overall[["ALL"]])`; their mean absolute p-value
gap was
`r sprintf("%.3f", mean(multiplier_reps$gaussian_all_abs_gap, na.rm = TRUE))`.
That is the intended numerical checksum and shows that changing weights is
changing the finite multiplier reference, not the observed numerator or fitted
score geometry.

Mammen is the only promising alternative in this small gate. It reduced average
rejection to `r sprintf("%.3f", multiplier_overall[["Mammen"]])`, with cell
rates between
`r sprintf("%.3f", min(multiplier_summary$rejection_mammen))` and
`r sprintf("%.3f", max(multiplier_summary$rejection_mammen))`. Centered
exponential weights went much too far, averaging
`r sprintf("%.3f", multiplier_overall[["centered exponential"]])`; their large
fourth moment makes the conditional reference far too heavy-tailed here.

This does not validate Mammen as a default. Unlike Rademacher it is a multiplier
bootstrap, not an exact sign randomization, and the current flip-specific
nuisance standardization relies on $w_i^2=1$. The implementation therefore
allows non-Rademacher laws only for the nuisance-effective statistic.

### Refining Mammen

The first result suggests three distinct mechanisms.

First, the zero-mean, unit-variance two-point family can be indexed by its third
moment $\gamma$. If

$$
b_\gamma = \frac{\gamma + \sqrt{\gamma^2+4}}{2}, \qquad
a_\gamma = -b_\gamma^{-1},
$$

and $P(w=a_\gamma)=b_\gamma^2/(1+b_\gamma^2)$, then
$E(w^3)=\gamma$ and $E(w^4)=1+\gamma^2$. This is the smallest possible fourth
moment at a given third moment. Rademacher is $\gamma=0$ and Mammen is
$\gamma=1$, so intermediate values test whether Mammen's remaining
conservatism is simply too much multiplier tail thickness.

Second, the restricted-fit scores have zero population mean under the null but
their tested-direction sample sum is the observed signal. A multiplier
empirical process can impose the null by replacing each row in information
stratum $s$ with
$\tilde u_{is}=\hat u_{is}-\bar u_s$ before multiplication. This retains
Mammen's moment matching and changes no observed numerator.

Third, a bootstrap-$t$ attempt replaces the fixed metric in each draw with
$\hat B^\ast=\sum_i w_i^2\tilde u_i\tilde u_i^\top$ and compares it with the
observed raw-OPG sandwich statistic. This is more invasive: it changes both
the resampled numerator law and its studentizer.

The same eight cells, observations, seeds, 200 replications, and 499 draws give:

```{r multiplier-refine-rank}
knitr::kable(refine_rank, digits = 3)
```

The mechanism is unusually clear. Moving along the two-point moment boundary
interpolates smoothly: $\gamma=.75$ remains liberal,
`r sprintf("%.3f", refine_overall[["two-point gamma=0.75"]])`, while
$\gamma=.90$ lands at
`r sprintf("%.3f", refine_overall[["two-point gamma=0.90"]])`. But this is a
tuned compromise and its cell rates still range from
`r sprintf("%.3f", min(refine_summary$rejection_two_point_0_90))` to
`r sprintf("%.3f", max(refine_summary$rejection_two_point_0_90))`.

Score recentering is the stronger hypothesis. Centered Mammen averages
`r sprintf("%.3f", refine_overall[["centered Mammen"]])` with cell rates
`r sprintf("%.3f", min(refine_summary$rejection_mammen_centered))`--`r sprintf("%.3f", max(refine_summary$rejection_mammen_centered))`;
it retains $\gamma=1$ rather than selecting a moment from these same cells.
The weighted-meat idea fails in this form: uncentered and centered variants
reject
`r sprintf("%.3f", refine_overall[["weighted-meat Mammen"]])` and
`r sprintf("%.3f", refine_overall[["centered weighted-meat"]])`. Their p-values
correlate only about
`r sprintf("%.2f", cor(refine_reps$p_two_point_1_00,
                       refine_reps$p_mammen_weighted_meat))`
with fixed-metric Mammen, confirming that this is not a small correction.

The independent gate therefore carries only raw Mammen, centered Mammen, direct
sandwich, and pEBA4 into a small normal/$t$/skew confirmation with more null
replications, followed by method-specific null thresholds for power. The
$\gamma=.90$ arm remains a mechanism check rather than a calibration constant
selected on the discovery cells.

### Independent confirmation and matched power

The confirmation uses a fresh seed base and 500 null replications in each of
12 cells: $q=8$, $N=60/200$, homogeneous or difficult information geometry,
and normal, $t_5$, or independently skewed innovations. The four alternatives
carried forward are deliberately distinct:

- **raw Mammen** multiplies the fitted efficient-score rows by the unmodified
  zero-mean, unit-variance Mammen two-point weights
  ($Ew^3=1$, $Ew^4=2$);
- **centered Mammen** first subtracts the score-row mean within each information
  stratum, imposing the multiplier null while leaving the observed numerator
  unchanged, then uses the same Mammen weights;
- **sandwich** uses the direct efficient-meat quadratic with a
  $\chi^2_q$ reference, so it is asymptotically pivotal but requires a noisy
  matrix inverse;
- **pEBA4** is the analytic score-spectrum approximation with order-four
  eigenvalue shrinkage; it is neither a multiplier test nor direct sandwich
  studentization.

```{r multiplier-confirm}
knitr::kable(confirm_rank, digits = 3)
```

Centered Mammen confirms at
`r sprintf("%.3f", mean(confirm_summary$rejection_mammen_centered))` overall,
with $N=60/200$ rates of
`r sprintf("%.3f", mean(confirm_summary$rejection_mammen_centered[
  confirm_summary$n_total == 60L]))` and
`r sprintf("%.3f", mean(confirm_summary$rejection_mammen_centered[
  confirm_summary$n_total == 200L]))`. Its cell range is
`r sprintf("%.3f", min(confirm_summary$rejection_mammen_centered))`--`r sprintf("%.3f", max(confirm_summary$rejection_mammen_centered))`.
Raw Mammen remains mildly conservative at
`r sprintf("%.3f", mean(confirm_summary$rejection_mammen))`; sandwich is more
conservative at `r sprintf("%.3f", mean(confirm_summary$rejection_sandwich))`;
and pEBA4 is
`r sprintf("%.3f", mean(confirm_summary$rejection_peba4))`. At 500
replications the single-cell Monte Carlo standard error is about 0.010 near
0.05, so the isolated 0.068 centered-Mammen cell is not evidence of a separate
failure regime.

The bounded power gate uses 300 fresh replications per matched cell. The
alternative increases the first tested group-2 loading by 0.50; it is therefore
a sparse loading-noninvariance alternative, not a broad power ranking. For each
method and cell, the critical p-value is the largest cutoff whose rejection
rate in the independent 500-replication null cell is no greater than 0.05.

```{r multiplier-matched-power}
knitr::kable(matched_power_overall, digits = 3)
```

After this forced size alignment, centered Mammen has power
`r sprintf("%.3f", matched_power_overall[["matched-null power"]][
  matched_power_overall$method == "centered Mammen"])`, versus
`r sprintf("%.3f", matched_power_overall[["matched-null power"]][
  matched_power_overall$method == "raw Mammen"])` for raw Mammen and
`r sprintf("%.3f", matched_power_overall[["matched-null power"]][
  matched_power_overall$method == "pEBA4"])` for pEBA4. Centering changes
cellwise matched power by
`r sprintf("%+.3f", mean(centered_raw_power_difference))` on average, with a
range of
`r sprintf("%+.3f", min(centered_raw_power_difference))` to
`r sprintf("%+.3f", max(centered_raw_power_difference))`. Thus its null-size
improvement is not purchased with a detectable power penalty in this gate.
The direct sandwich remains lower at
`r sprintf("%.3f", matched_power_overall[["matched-null power"]][
  matched_power_overall$method == "sandwich"])`.

This is enough to retain centered Mammen as the promising resampling candidate,
but not enough to call it generally valid. The empirical cutoffs themselves
have finite-null Monte Carlo error, all cells are complete-data affine ML at
$q=8$, and Mammen weighting has no exact sign-randomization interpretation.
The next useful step is theory and transport to FIML/ordinal score strata,
not another large complete-data null grid.

## Timing

Cost is a few milliseconds and linear in the number of flips. At 499 flips,
median core latency was about 2.3 ms at 1 df, 5.4 ms at 4 df, and 8.7 ms at 8 df,
of which flip-specific standardization is roughly 0.7, 2.7, and 4.2 ms. The FMG
transforms on the score eigenvalues are essentially free once the flip test has
run, since the eigenvalues are already computed.

```{r timing-df}
names(timing_499) <- c("df", "setup ms", "resampling ms", "standardization ms",
                       "asymptotic ms", "total core ms")
knitr::kable(timing_499, digits = 3)
```

## Recommendation

- Prefer the score/LM base to the Satorra-2000 LR base in this design, but do not
  infer that this makes every score transform correct. The LR routes are liberal;
  score-SB is trace shrinkage, and score-ALL is a conservative plug-in mixture.
- pEBA4 is the best provisional compromise. It stays closest to 0.05 across the
  main, severe-copula, and dimension probes by shrinking noisy eigenvalue shape
  without collapsing fully to SB's common eigenvalue.
- Keep direct sandwich score output as a theoretical benchmark and diagnostic,
  not the small-sample default. It is asymptotically pivotal, but null rejection
  falls to 0.024 at eight restrictions and its meat condition number becomes
  large in small or severe cells.
- Read the advantage correctly: much of the raw gap is the classic small-sample
  over-rejection of likelihood-ratio tests, which adequate $N$ removes. The
  genuine robustness margin under non-normality is smaller but real.
- Keep `p_standardized` as a cheap resampling option, but it is not better
  calibrated than the score-base asymptotics and is more strained by a severe
  copula.
- Retain centered Mammen as the experimental resampling candidate. Its
  independent null rate is 5.2% overall (5.4% at $N=60$, 4.9% at $N=200$),
  and its 35.6% matched-null power is indistinguishable from raw Mammen's 35.3%
  and pEBA4's 36.0%. This supports score recentering, not a tuned
  $\gamma=.90$ multiplier, but does not yet justify a default.
- Do not compare nominal power without comparing null size. The matched-null
  table is a forced size comparison, not evidence that a nominal reference is
  valid.
- Next: study shrinkage of the efficient meat itself, between SB's scalar trace
  and the unregularized sandwich inverse, while varying nuisance leverage and
  conditioning at fixed dimension and adequacy.

## Design and caveats

The null grid crosses 1/4/8 loading-equality restrictions, total $N=60/100/200$,
1:1 and 1:3 allocation, three group-information geometries, and normal, $t_5$,
and skewed innovations, at 200 replications and 499 flips per cell. Maximum cell
fit-failure rate was `r sprintf("%.3f", max(s$failure_rate))`; there were
`r sum(!r_all_ok)` failed fits overall. The power grid adds sparse and dense
alternatives at two effect sizes.

Two boundary probes sit outside the main grid so they do not unbalance it. The
severe-copula stress block crosses two copulas at fixed severe marginals with
1/8 restrictions, $N=60/100/200$, and two geometries. The dimension-scaling probe
crosses two copulas and two geometries, $p=10/20/30/40$ at fixed 8-df
restrictions, and two sample-to-parameter ratios (about 2.5 and 6), with $N$ set
from the ratio so it grows with $p$, at 1700 replications per cell; the ratios are
approximate because the parameter count is estimated. The direct-sandwich
dimension replay uses 200 replications per cell and one flip because only the
asymptotic comparator and conditioning diagnostics are consumed. The
independent-generator
copula was excluded because its marginal moment solve is infeasible at this
severity.

Monte Carlo and binomial uncertainty remain visible in cell-level 0.05 decisions.
The smaller sandwich dimension replay has standard errors around 0.015 within a
single cell; its aggregate trends are evidence for the next iteration, not a
fine ranking.
Score flipping here is a frontier method for affine, complete-data ML models. The
results do not automatically transfer to FIML, ordinal, estimated-weight,
clustered, boundary, or nonlinear-constraint tests.
The multiplier-law discovery replay is narrower still: it covers only
effective-score calibration at 8 df in the severe complete-data copula block.
Its 200 replications give a single-cell Monte Carlo standard error of about 1.5
percentage points near 0.05. The independent confirmation broadens the
generators but remains complete-data affine ML at 8 df; its 500 null
replications give a single-cell error of about 1.0 percentage point. The
matched-power cutoffs are estimated rather than known, so their null-sample
uncertainty is not reflected in the 300-replication alternative rates.

## Reproduce

```sh
Rscript experiments/62-standardized-flip-score/run_experiment.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --power
Rscript experiments/62-standardized-flip-score/run_expansion.R --power-sandwich
Rscript experiments/62-standardized-flip-score/run_expansion.R --stress
Rscript experiments/62-standardized-flip-score/run_dimension.R --big
Rscript experiments/62-standardized-flip-score/run_dimension.R --sandwich
Rscript experiments/62-standardized-flip-score/run_multiplier_probe.R --probe
Rscript experiments/62-standardized-flip-score/run_multiplier_probe.R --refine
Rscript experiments/62-standardized-flip-score/run_multiplier_probe.R --confirm
Rscript experiments/62-standardized-flip-score/run_multiplier_probe.R --confirm-power
Rscript experiments/62-standardized-flip-score/run_timing.R
quarto render experiments/62-standardized-flip-score/report.qmd
```
