---
title: "Score-flip standardization, FMG comparators, and power"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
files <- file.path("results", c(
  "summary.csv", "expansion_summary.csv", "expansion_replications.csv",
  "expansion_metadata.csv", "timing_summary.csv", "power_summary.csv",
  "power_replications.csv", "power_metadata.csv"))
if (any(!file.exists(files))) {
  stop(paste(
    "missing results; run run_experiment.R, run_expansion.R --probe,",
    "run_expansion.R --power, and run_timing.R"))
}
legacy <- read.csv(files[[1L]])
s <- read.csv(files[[2L]])
r <- read.csv(files[[3L]])
m <- read.csv(files[[4L]])
timing <- read.csv(files[[5L]])
power_summary <- read.csv(files[[6L]])
power <- read.csv(files[[7L]])
power_metadata <- read.csv(files[[8L]])
r <- r[r$ok, , drop = FALSE]
power <- power[power$ok, , drop = FALSE]
s$calibration_effective <- abs(s$rejection_effective - .05)
s$calibration_standardized <- abs(s$rejection_standardized - .05)
s$calibration_gain <- s$calibration_effective - s$calibration_standardized

means <- function(vars, by) {
  aggregate(s[vars], s[by], mean)
}
by_df <- means(c("rejection_effective", "rejection_standardized",
                 "changed_to_accept", "mean_abs_delta_p",
                 "mean_variance_relative_shift"), "df")
by_n <- means(c("rejection_effective", "rejection_standardized",
                "changed_to_accept", "mean_abs_delta_p",
                "mean_variance_relative_shift"), "n_total")
by_dist <- means(c("rejection_effective", "rejection_standardized"),
                 "distribution")
by_balance <- means(c("rejection_effective", "rejection_standardized",
                      "changed_to_accept"), "balance")
by_ndf <- aggregate(
  r[c("reject_effective", "reject_standardized", "changed_to_accept")],
  r[c("n_total", "df")], mean)

timing_ms <- timing
for (name in grep("seconds$", names(timing), value = TRUE)) {
  timing_ms[[sub("seconds$", "ms", name)]] <- 1000 * timing[[name]]
}
timing_499 <- subset(
  timing_ms, scenario == "easy" & n_flips == 499,
  select = c(df, median_setup_ms, median_resampling_score_ms,
             median_standardization_ms, median_asymptotic_ms,
             median_core_total_ms))
timing_scale <- subset(
  timing_ms, scenario == "easy" & df == 8,
  select = c(n_flips, median_resampling_score_ms,
             median_standardization_ms, median_core_total_ms))

method_labels <- c(
  p_effective = "score effective",
  p_standardized = "score standardized",
  p_lr_scaled = "LR SB scaled",
  p_lr_mv_adjusted = "LR MV adjusted",
  p_lr_scaled_shifted = "LR scaled-shifted",
  p_lr_mixture = "LR exact mixture",
  p_fmg_sb = "FMG SB",
  p_fmg_mv = "FMG MV",
  p_fmg_ss = "FMG SS",
  p_fmg_sf = "FMG SF",
  p_fmg_eba2 = "FMG EBA2",
  p_fmg_eba4 = "FMG EBA4",
  p_fmg_eba6 = "FMG EBA6",
  p_fmg_peba2 = "FMG pEBA2",
  p_fmg_peba4 = "FMG pEBA4",
  p_fmg_peba6 = "FMG pEBA6",
  p_fmg_pall = "FMG PALL",
  p_fmg_pols = "FMG pOLS",
  p_fmg_all = "FMG ALL")

method_size <- do.call(rbind, lapply(names(method_labels), function(v) {
  data.frame(
    method = unname(method_labels[[v]]),
    overall = mean(r[[v]] < .05, na.rm = TRUE),
    df1 = mean(r[[v]][r$df == 1L] < .05, na.rm = TRUE),
    df4 = mean(r[[v]][r$df == 4L] < .05, na.rm = TRUE),
    df8 = mean(r[[v]][r$df == 8L] < .05, na.rm = TRUE))
}))

power_methods <- names(method_labels)

raw_power <- do.call(rbind, lapply(power_methods, function(v) {
  out <- aggregate(power[[v]] < .05,
                   power[c("alternative", "effect")], mean, na.rm = TRUE)
  names(out)[[3L]] <- "power"
  out$method <- unname(method_labels[[v]])
  out
}))
raw_power <- reshape(raw_power, idvar = c("alternative", "effect"),
                     timevar = "method", direction = "wide")
names(raw_power) <- sub("^power\\.", "", names(raw_power))

# Use the largest strict critical value whose observed matched-cell null size
# does not exceed alpha. This respects the discreteness of the 499-flip tests.
empirical_critical <- function(x, alpha = .05) {
  x <- x[is.finite(x)]
  candidates <- sort(unique(x))
  rates <- vapply(candidates, function(cut) mean(x < cut), numeric(1L))
  valid <- candidates[rates <= alpha]
  if (!length(valid)) -Inf else max(valid)
}

keys <- c("df", "n_total", "balance", "heterogeneity", "distribution")
power_cells <- unique(power[c(keys, "alternative", "effect")])
calibrated <- do.call(rbind, lapply(seq_len(nrow(power_cells)), function(i) {
  cell <- power_cells[i, , drop = FALSE]
  null_match <- rep(TRUE, nrow(r))
  power_match <- rep(TRUE, nrow(power))
  for (key in keys) {
    null_match <- null_match & r[[key]] == cell[[key]]
    power_match <- power_match & power[[key]] == cell[[key]]
  }
  power_match <- power_match & power$alternative == cell$alternative &
    power$effect == cell$effect
  do.call(rbind, lapply(power_methods, function(v) {
    null_p <- r[[v]][null_match]
    alt_p <- power[[v]][power_match]
    critical <- empirical_critical(null_p)
    data.frame(
      cell, method = unname(method_labels[[v]]), critical = critical,
      achieved_size = mean(null_p < critical, na.rm = TRUE),
      calibrated_power = mean(alt_p < critical, na.rm = TRUE))
  }))
}))

calibrated_power <- aggregate(
  calibrated[c("achieved_size", "calibrated_power")],
  calibrated[c("alternative", "effect", "method")], mean)
calibrated_power <- reshape(
  calibrated_power, idvar = c("alternative", "effect"),
  timevar = "method", direction = "wide")
names(calibrated_power) <- sub("^calibrated_power\\.", "", names(calibrated_power))

overall_calibrated <- aggregate(
  calibrated[c("achieved_size", "calibrated_power")],
  calibrated[c("method")], mean)

power_by_df <- aggregate(
  calibrated$calibrated_power,
  calibrated[c("df", "alternative", "effect", "method")], mean)
names(power_by_df)[[5L]] <- "power"
power_by_df <- subset(
  power_by_df,
  method %in% c("score standardized", "LR MV adjusted", "FMG pEBA4", "FMG ALL"))

null_elapsed <- as.numeric(m$value[m$key == "elapsed_seconds"])
power_elapsed <- as.numeric(power_metadata$value[power_metadata$key == "elapsed_seconds"])
parallel_timing <- data.frame(
  stage = c("two fits", "499-flip score battery", "nested LR battery",
            "FMG battery"),
  median_ms = 1000 * c(
    median(r$fit_seconds, na.rm = TRUE),
    median(r$flip_elapsed_seconds, na.rm = TRUE),
    median(r$nested_seconds, na.rm = TRUE),
    median(r$fmg_seconds, na.rm = TRUE)))
```

## Bottom line

Standardization is cheap and directionally useful, but its practical value is
concentrated. Across `r format(nrow(r), big.mark = ",")` successful null
replications, it changed `r sum(r$changed_to_accept)` decisions from rejection
to acceptance and `r sum(r$changed_to_reject)` in the opposite direction. The
mean p-value increased from the effective reference by
`r sprintf("%.4f", mean(r$p_standardized - r$p_effective))`; no replication
had a smaller standardized p-value.

The larger comparator result is cautionary. At a nominal 5% cutoff, the
standardized score flip rejected `r sprintf("%.3f", method_size$overall[method_size$method == "score standardized"])`
of null replications. Mean/variance-adjusted LR and FMG ALL rejected about
`r sprintf("%.3f", method_size$overall[method_size$method == "LR MV adjusted"])`,
while SB and pEBA4 rejected
`r sprintf("%.3f", method_size$overall[method_size$method == "LR SB scaled"])`
and `r sprintf("%.3f", method_size$overall[method_size$method == "FMG pEBA4"])`.
Their larger nominal power is therefore mostly not a fair power advantage.
After each method was recalibrated to its matched empirical null cell, average
power was strikingly similar. The score tests were marginally ahead overall,
not behind FMG.

The strongest interaction was sample size by restriction dimension:

```{r n-df-table}
tab <- by_ndf
names(tab) <- c("total N", "df", "effective", "standardized",
                "effective-only rejection")
knitr::kable(tab, digits = 3)
```

At total $N=60$, standardization removes little for one restriction, but about
0.45 and 0.62 percentage points for 4 and 8 restrictions. It does not fully
repair the 8-df test, which remains liberal at 0.077. At $N=200$ the absolute
changes are only 0.06--0.17 percentage points. The answer is therefore not
"always standardize because the unstandardized test is broken". It is closer
to: **standardize by default because it is inexpensive, but expect a material
calibration difference mainly for small samples and moderate-to-large joint
tests.**

## Hypotheses and what happened

### H1: small samples should matter most — supported

The flip-specific covariance correction depends on group sign sums, whose
relative size shrinks with group sample size. Both the measured covariance
shift and the p-value movement decline rapidly with $N$.

```{r n-table}
names(by_n) <- c("total N", "effective", "standardized",
                 "effective-only rejection", "mean |delta p|",
                 "mean relative covariance shift")
knitr::kable(by_n, digits = 4)
```

### H2: more restrictions should increase the payoff — supported

The mean covariance shift itself is only slightly larger at 8 df, but the
multivariate quadratic aggregates those changes across more directions. The
mean absolute p-value movement rises from 0.0014 at 1 df to 0.0046 at 8 df.

```{r df-table}
names(by_df) <- c("df", "effective", "standardized",
                  "effective-only rejection", "mean |delta p|",
                  "mean relative covariance shift")
knitr::kable(by_df, digits = 4)
```

This makes "many restrictions" a better predictor than "large raw covariance
shift" alone. The replication-level correlation between mean relative
covariance shift and absolute p-value movement was
`r sprintf("%.3f", cor(r$mean_variance_relative_shift, r$abs_delta_p))`.

### H3: imbalance should increase the payoff — weakly supported

The 1:3 design produced more effective-only rejections, but did not have a
larger average covariance shift. Imbalance appears to make a given correction
more consequential near the rejection boundary rather than mechanically
making every flip variance more different.

```{r balance-table}
names(by_balance) <- c("balance", "effective", "standardized",
                       "effective-only rejection")
knitr::kable(by_balance, digits = 4)
```

### H4: heterogeneous populations should matter most — not as stated

Homogeneous, common-scale, and strongly geometry-heterogeneous populations had
similar effective-to-standardized rejection changes. The homogeneous cells
actually had the largest mean relative covariance shift. Raw variance
heterogeneity is therefore not a sufficient proxy: the relevant quantity is
the group-specific tested--nuisance information cross-geometry after the H0
restriction map is applied.

### H5: nonnormality should increase the need — consequential, not causal

The covariance-shift diagnostic was almost unchanged across normal, $t_5$, and
skewed generators. Nonnormality did make both flip tests more liberal, so the
same conservative standardization adjustment bought more calibration under
$t_5$ and skew. It did not eliminate the remaining misspecification effect.

```{r distribution-table}
names(by_dist) <- c("distribution", "effective", "standardized")
knitr::kable(by_dist, digits = 4)
```

## LR and FMG null calibration

The full comparator battery includes the basic FMG variants (`SB`, `MV`, `SS`,
and `SF`), EBA2/4/6 and pEBA2/4/6, the pooled approximations PALL and pOLS, and
the exact ALL mixture. `MV` is the requested mean/variance-adjusted comparator.
The FMG
SB/MV/SS rows are also a wiring check: they agree with the corresponding
nested-test scaled, mean/variance-adjusted, and scaled-shifted routes to a
maximum absolute p-value gap of
`r format(max(r$fmg_duplicate_max_gap, na.rm = TRUE), scientific = TRUE,
digits = 2)`.

```{r comparator-size}
names(method_size) <- c("method", "overall", "df = 1", "df = 4", "df = 8")
knitr::kable(method_size, digits = 3)
```

Three patterns matter:

- At one restriction most spectrum transformations collapse to nearly the
  same scalar test; the main exception is SF. The common LR/FMG rejection rate
  is already about 0.105, so scalar robust scaling does not solve this
  small-sample design.
- MV, exact mixture/ALL, and SF are less liberal than SB and pEBA/PALL/pOLS,
  especially at 4--8 df, but remain well above 0.05.
- Standardization improves the effective flip test most clearly at 8 df
  (0.073 to 0.070), but this is an adjustment, not complete calibration.

## Power

The independent power run contains
`r format(nrow(power), big.mark = ",")` successful fits from 81,000 attempts.
For `sparse`, the first tested loading departs by 0.25 or 0.50. For `dense`, the
same Euclidean violation norm is divided equally over all tested restrictions.
The df=1 dense cells are omitted because they duplicate sparse exactly.

### Nominal power is confounded by size

```{r raw-power}
raw_show <- raw_power[c(
  "alternative", "effect", "score standardized", "score effective",
  "LR SB scaled", "LR MV adjusted", "LR exact mixture", "FMG SF",
  "FMG pEBA4", "FMG ALL")]
knitr::kable(raw_show, digits = 3)
```

At face value SB and pEBA4 look substantially more powerful. Their null sizes
show why that ranking is misleading. For example, sparse 0.50 nominal power is
about 0.50 for SB, 0.49 for pEBA4, 0.46 for MV/ALL, and 0.38 for standardized
flips—but their overall null rejection rates are about 0.138, 0.130, 0.110,
and 0.063, respectively.

### Matched empirical-null calibration

For a fairer diagnostic, each method and each of the 162 matched null cells
gets its own critical p-value: the largest strict cutoff whose observed null
rejection rate does not exceed 0.05. Power is then evaluated on independent
alternative replications using that cutoff. This respects the discrete
499-flip distribution. It also consumes only 200 null replications per cell,
so differences of a few points should not be over-interpreted.

```{r calibrated-power}
calibrated_show <- calibrated_power[c(
  "alternative", "effect", "score standardized", "score effective",
  "LR SB scaled", "LR MV adjusted", "LR exact mixture", "FMG SF",
  "FMG pEBA4", "FMG ALL")]
knitr::kable(calibrated_show, digits = 3)
```

The dramatic nominal separation largely vanishes. Across all four alternative
families, standardized flips, MV-adjusted LR, pEBA4, and ALL are generally
within one or two percentage points. SF trails on sparse alternatives. Averaged
over all power cells, standardized/effective flips reach about 0.186 calibrated
power versus about 0.184 for MV, pEBA4, and ALL; that gap is too small to call a
general advantage.

The all-method average makes the comparison among the EBA orders and the basic
FMG variants explicit. `achieved_size` is the average rejection rate obtained
by the matched empirical critical values, not the original nominal size.

```{r all-method-calibrated}
knitr::kable(overall_calibrated, digits = 3)
```

### H6: sparse violations should be easier than dense equal-norm violations — supported

Concentrating the same Euclidean violation in one loading is much easier to
detect. At effect 0.50, calibrated standardized-flip power is about 0.350 for
sparse alternatives and 0.134 for dense alternatives. The comparison is not
available at df=1 because those constructions coincide, but the result persists
within both df=4 and df=8.

### H7: FMG's spectrum transforms should help more with many restrictions — not supported here

Before size calibration, pEBA4's apparent advantage grows with df, exactly
where its null over-rejection also grows. After calibration, pEBA4, MV, and ALL
are nearly indistinguishable. For the sparse 0.50 case, their calibrated power
is around 0.37 at df=4 and 0.28 at df=8. Standardized flips are about 0.39 and
0.30. This probe gives no evidence that a particular FMG transform dominates
once size is aligned.

### H8: more restrictions should always improve power — rejected

Restriction dimension and alternative geometry interact. Dense equal-norm
signals are diluted across coordinates, and the df=8 tests pay a larger
critical-value penalty. Sparse power also falls at df=8 because the joint test
adds seven null directions around one signal. "Many restrictions" predicts
when flip standardization changes calibration, but does not by itself predict
higher power.

## Timing

The controlled benchmark is single-process, warm-cache, and excludes model
fitting. At 499 flips, median core latency was about 2.3 ms at 1 df, 5.4 ms at
4 df, and 8.7 ms at 8 df. Flip-specific standardization accounts for roughly
0.7, 2.7, and 4.2 ms of those totals.

```{r timing-df}
names(timing_499) <- c("df", "setup ms", "basic/effective resampling ms",
                       "standardization ms", "asymptotic comparator ms",
                       "total core ms")
knitr::kable(timing_499, digits = 3)
```

Cost is essentially linear in the number of flips. For the 8-df easy case:

```{r timing-flips}
names(timing_scale) <- c("flips", "basic/effective resampling ms",
                         "standardization ms", "total core ms")
knitr::kable(timing_scale, digits = 3)
```

The two model fits took 3--5 ms in the warmed serial benchmark. In the full
parallel experiment, fit and score timings include worker contention. The
following medians average across restriction dimensions; the complete FMG
battery means all 13 requested transformations, not one p-value.

```{r parallel-timing}
knitr::kable(parallel_timing, digits = 1)
```

The 162-cell null run (200 replications per cell, 499 flips, 16 workers) took
`r sprintf("%.1f", null_elapsed)` seconds, or about
`r sprintf("%.0f", nrow(r) / null_elapsed)` successful replications per
wall-clock second. The 540-cell power run (150 replications per cell) took
`r sprintf("%.1f", power_elapsed)` seconds, or about
`r sprintf("%.0f", nrow(power) / power_elapsed)` per second. In a single call,
the FMG battery's median 8 ms is less than half the median score-flip battery;
one FMG transform would be cheaper still.

## Recommendation

- Keep `p_standardized` as the default. The serial incremental cost is only a
  few milliseconds at SEM-sized restriction dimensions, and every decision
  change in this probe went in the theoretically expected conservative
  direction.
- Expect little practical difference for a one-df release or total $N\geq200$,
  unless leverage or information conditioning is much more extreme than in
  this design.
- Expect the clearest benefit around total $N\leq100$ with 4--8 jointly tested
  restrictions, especially with group imbalance. Standardization helps there,
  but should not be sold as a complete small-sample correction.
- Do not decide from "heteroscedastic versus homogeneous" alone. Inspect the
  returned `mean_variance_relative_shift`; it is a more direct diagnostic of
  whether standardization is changing the reference distribution.
- Do not compare raw power for these robust LR/FMG transforms without also
  comparing null size. In this probe, the apparent pEBA/SB advantage disappears
  under matched empirical-null calibration; MV-adjusted LR, exact ALL, pEBA4,
  and standardized flips have broadly similar calibrated power.
- Prefer MV/ALL over SB or pEBA/PALL/pOLS if a non-flip comparator is needed in
  this design: they are materially less liberal. None is close enough to 0.05
  here to present as a small-sample solution.
- Next stress test: hold df fixed while varying nuisance dimension and leverage.
  This experiment held a 62-slot ambient model constant, so it cannot yet
  separate "many restrictions" from the ratio of tested dimension to effective
  residual information.

## Design and caveats

The null expansion crosses 1/4/8 loading-equality restrictions; total
$N=60/100/200$; 1:1/1:3 group allocation; homogeneous, common latent-scale,
and strong factor/residual-geometry heterogeneity; and normal, $t_5$, and
standardized skewed innovations. Each of 162 null cells has 200 attempted
replications and 499 random flips. There were
`r sum(!read.csv(files[[3L]])$ok)` failed fits (maximum cell failure rate
`r sprintf("%.3f", max(s$failure_rate))`). Monte Carlo and binomial uncertainty
remain visible, especially in cell-by-cell 0.05 decisions.

The power expansion adds sparse/dense alternatives and two effect sizes, for
540 cells and 150 attempts per cell. It had
`r 81000 - nrow(power)` fit failures. Comparator-specific failures are retained
as missing p-values rather than discarding otherwise valid flip results; they
occurred in about `r sprintf("%.3f", mean(nzchar(power$nested_error)))` of
successful power fits.

Score flipping here remains a frontier method for affine, complete-data ML
models. These results do not automatically transfer to FIML, ordinal,
estimated-weight, clustered, boundary, or nonlinear-constraint tests.

## Reproduce

```sh
Rscript experiments/62-standardized-flip-score/run_experiment.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --power
Rscript experiments/62-standardized-flip-score/run_timing.R
quarto render experiments/62-standardized-flip-score/report.qmd
```
