---
title: "When is standardized score flipping worth it?"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
files <- file.path("results", c(
  "summary.csv", "expansion_summary.csv", "expansion_replications.csv",
  "expansion_metadata.csv", "timing_summary.csv"))
if (any(!file.exists(files))) {
  stop("missing results; run run_experiment.R, run_expansion.R, and run_timing.R")
}
legacy <- read.csv(files[[1L]])
s <- read.csv(files[[2L]])
r <- read.csv(files[[3L]])
m <- read.csv(files[[4L]])
timing <- read.csv(files[[5L]])
r <- r[r$ok, , drop = FALSE]
s$calibration_effective <- abs(s$rejection_effective - .05)
s$calibration_standardized <- abs(s$rejection_standardized - .05)
s$calibration_gain <- s$calibration_effective - s$calibration_standardized

means <- function(vars, by) {
  aggregate(s[vars], s[by], mean)
}
by_df <- means(c("rejection_effective", "rejection_standardized",
                 "changed_to_accept", "mean_abs_delta_p",
                 "mean_variance_relative_shift"), "df")
by_n <- means(c("rejection_effective", "rejection_standardized",
                "changed_to_accept", "mean_abs_delta_p",
                "mean_variance_relative_shift"), "n_total")
by_dist <- means(c("rejection_effective", "rejection_standardized"),
                 "distribution")
by_balance <- means(c("rejection_effective", "rejection_standardized",
                      "changed_to_accept"), "balance")
by_ndf <- aggregate(
  r[c("reject_effective", "reject_standardized", "changed_to_accept")],
  r[c("n_total", "df")], mean)

timing_ms <- timing
for (name in grep("seconds$", names(timing), value = TRUE)) {
  timing_ms[[sub("seconds$", "ms", name)]] <- 1000 * timing[[name]]
}
timing_499 <- subset(
  timing_ms, scenario == "easy" & n_flips == 499,
  select = c(df, median_setup_ms, median_resampling_score_ms,
             median_standardization_ms, median_asymptotic_ms,
             median_core_total_ms))
timing_scale <- subset(
  timing_ms, scenario == "easy" & df == 8,
  select = c(n_flips, median_resampling_score_ms,
             median_standardization_ms, median_core_total_ms))
```

## Bottom line

Standardization is cheap and directionally useful, but its practical value is
concentrated. Across `r format(nrow(r), big.mark = ",")` successful null
replications, it changed `r sum(r$changed_to_accept)` decisions from rejection
to acceptance and `r sum(r$changed_to_reject)` in the opposite direction. The
mean p-value increased from the effective reference by
`r sprintf("%.4f", mean(r$p_standardized - r$p_effective))`; no replication
had a smaller standardized p-value.

The strongest interaction was sample size by restriction dimension:

```{r n-df-table}
tab <- by_ndf
names(tab) <- c("total N", "df", "effective", "standardized",
                "effective-only rejection")
knitr::kable(tab, digits = 3)
```

At total $N=60$, standardization removes little for one restriction, but about
0.45 and 0.62 percentage points for 4 and 8 restrictions. It does not fully
repair the 8-df test, which remains liberal at 0.077. At $N=200$ the absolute
changes are only 0.06--0.17 percentage points. The answer is therefore not
"always standardize because the unstandardized test is broken". It is closer
to: **standardize by default because it is inexpensive, but expect a material
calibration difference mainly for small samples and moderate-to-large joint
tests.**

## Hypotheses and what happened

### H1: small samples should matter most — supported

The flip-specific covariance correction depends on group sign sums, whose
relative size shrinks with group sample size. Both the measured covariance
shift and the p-value movement decline rapidly with $N$.

```{r n-table}
names(by_n) <- c("total N", "effective", "standardized",
                 "effective-only rejection", "mean |delta p|",
                 "mean relative covariance shift")
knitr::kable(by_n, digits = 4)
```

### H2: more restrictions should increase the payoff — supported

The mean covariance shift itself is only slightly larger at 8 df, but the
multivariate quadratic aggregates those changes across more directions. The
mean absolute p-value movement rises from 0.0014 at 1 df to 0.0046 at 8 df.

```{r df-table}
names(by_df) <- c("df", "effective", "standardized",
                  "effective-only rejection", "mean |delta p|",
                  "mean relative covariance shift")
knitr::kable(by_df, digits = 4)
```

This makes "many restrictions" a better predictor than "large raw covariance
shift" alone. The replication-level correlation between mean relative
covariance shift and absolute p-value movement was
`r sprintf("%.3f", cor(r$mean_variance_relative_shift, r$abs_delta_p))`.

### H3: imbalance should increase the payoff — weakly supported

The 1:3 design produced more effective-only rejections, but did not have a
larger average covariance shift. Imbalance appears to make a given correction
more consequential near the rejection boundary rather than mechanically
making every flip variance more different.

```{r balance-table}
names(by_balance) <- c("balance", "effective", "standardized",
                       "effective-only rejection")
knitr::kable(by_balance, digits = 4)
```

### H4: heterogeneous populations should matter most — not as stated

Homogeneous, common-scale, and strongly geometry-heterogeneous populations had
similar effective-to-standardized rejection changes. The homogeneous cells
actually had the largest mean relative covariance shift. Raw variance
heterogeneity is therefore not a sufficient proxy: the relevant quantity is
the group-specific tested--nuisance information cross-geometry after the H0
restriction map is applied.

### H5: nonnormality should increase the need — consequential, not causal

The covariance-shift diagnostic was almost unchanged across normal, $t_5$, and
skewed generators. Nonnormality did make both flip tests more liberal, so the
same conservative standardization adjustment bought more calibration under
$t_5$ and skew. It did not eliminate the remaining misspecification effect.

```{r distribution-table}
names(by_dist) <- c("distribution", "effective", "standardized")
knitr::kable(by_dist, digits = 4)
```

## Timing

The controlled benchmark is single-process, warm-cache, and excludes model
fitting. At 499 flips, median core latency was about 2.3 ms at 1 df, 5.4 ms at
4 df, and 8.7 ms at 8 df. Flip-specific standardization accounts for roughly
0.7, 2.7, and 4.2 ms of those totals.

```{r timing-df}
names(timing_499) <- c("df", "setup ms", "basic/effective resampling ms",
                       "standardization ms", "asymptotic comparator ms",
                       "total core ms")
knitr::kable(timing_499, digits = 3)
```

Cost is essentially linear in the number of flips. For the 8-df easy case:

```{r timing-flips}
names(timing_scale) <- c("flips", "basic/effective resampling ms",
                         "standardization ms", "total core ms")
knitr::kable(timing_scale, digits = 3)
```

The two model fits took 3--5 ms in the warmed serial benchmark. In the full
parallel experiment, fit and score timings include worker contention: median
fit time was about 13--15 ms and median 499-flip score-test time ranged from
9 ms at 1 df to 32--35 ms at 8 df. The complete 162-cell design (200
replications per cell, 499 flips, 16 workers) took
`r sprintf("%.1f", as.numeric(m$value[m$key == "elapsed_seconds"]))` seconds,
or about `r sprintf("%.0f", nrow(r) / as.numeric(m$value[m$key == "elapsed_seconds"]))`
successful replications per wall-clock second.

## Recommendation

- Keep `p_standardized` as the default. The serial incremental cost is only a
  few milliseconds at SEM-sized restriction dimensions, and every decision
  change in this probe went in the theoretically expected conservative
  direction.
- Expect little practical difference for a one-df release or total $N\geq200$,
  unless leverage or information conditioning is much more extreme than in
  this design.
- Expect the clearest benefit around total $N\leq100$ with 4--8 jointly tested
  restrictions, especially with group imbalance. Standardization helps there,
  but should not be sold as a complete small-sample correction.
- Do not decide from "heteroscedastic versus homogeneous" alone. Inspect the
  returned `mean_variance_relative_shift`; it is a more direct diagnostic of
  whether standardization is changing the reference distribution.
- Next stress test: hold df fixed while varying nuisance dimension and leverage.
  This experiment held a 62-slot ambient model constant, so it cannot yet
  separate "many restrictions" from the ratio of tested dimension to effective
  residual information.

## Design and caveats

The expansion crosses 1/4/8 loading-equality restrictions; total
$N=60/100/200$; 1:1/1:3 group allocation; homogeneous, common latent-scale,
and strong factor/residual-geometry heterogeneity; and normal, $t_5$, and
standardized skewed innovations. Each of 162 null cells has 200 attempted
replications and 499 random flips. There were
`r sum(!read.csv(files[[3L]])$ok)` failed fits (maximum cell failure rate
`r sprintf("%.3f", max(s$failure_rate))`). Monte Carlo and binomial uncertainty
remain visible, especially in cell-by-cell 0.05 decisions.

Score flipping here remains a frontier method for affine, complete-data ML
models. These results do not automatically transfer to FIML, ordinal,
estimated-weight, clustered, boundary, or nonlinear-constraint tests.

## Reproduce

```sh
Rscript experiments/62-standardized-flip-score/run_experiment.R --probe
Rscript experiments/62-standardized-flip-score/run_expansion.R --probe
Rscript experiments/62-standardized-flip-score/run_timing.R
quarto render experiments/62-standardized-flip-score/report.qmd
```
