---
title: "Can magmaan reproduce FMG Study 2?"
subtitle: "Weak-invariance Type-I calibration with semTests as a sentinel"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
roots <- file.path("results", c("full", "pilot", "smoke"))
available <- roots[vapply(roots, function(path)
  file.exists(file.path(path, "metadata.csv")), logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke --semtests-parity`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
meta_frame <- read_result("metadata.csv")
meta <- setNames(meta_frame$value, meta_frame$key)
calibration <- read_result("calibration_summary.csv")
generator_calibration <- read_result("generator_calibration.csv")
generator_summary <- read_result("generator_summary.csv")
cells <- read_result("cell_summary.csv")
timing <- read_result("timing_summary.csv")
failures <- read_result("failures.csv")
warnings <- read_result("warnings.csv")
parity <- read_result("semtests_parity.csv")
profile <- meta[["profile"]]
reps <- as.integer(meta[["reps"]])
n_cells <- as.integer(meta[["cells"]])
is_full <- identical(profile, "full") && n_cells == 189L && reps >= 2000L
```

## Question

Can native magmaan reproduce the 189-cell Satorra-2000 weak-invariance Type-I
comparison in Study 2 of Foldnes, Grønneberg and Moss (2026), now that the
complete-data path supports both the empirical and Browne/Du-Bentler unbiased
Gamma spectra? Can semTests be reduced from the simulation engine to a small
correctness sentinel?

## Short answer

`r if (is_full) {
  winner <- calibration$paper_label[[which.min(calibration$rmse_percentage_points)]]
  sprintf("Yes for the completed design: %s has the lowest magmaan cell-RMSE (%.3f percentage points), across %d cells with %d replications each. The run had %d failed and %d warning-bearing replications.", winner, min(calibration$rmse_percentage_points), n_cells, reps, nrow(failures), nrow(warnings))
} else {
  sprintf("The native pipeline is operational on the %s profile (%d cells, %d replications per cell), including all 38 method variants. This is sufficient as a translation/sentinel endpoint: the published study already supplies the paper-level Type-I evidence, so a new 189 x 2000 rerun is optional rather than a project requirement.", profile, n_cells, reps)
}`

`r if (nrow(parity) && any(is.finite(parity$abs_diff))) {
  comparable <- is.finite(parity$abs_diff)
  sprintf("The optional semTests sentinel passed %d/%d comparable p-values; max |delta p| = %.3g.", sum(parity$pass[comparable]), sum(comparable), max(parity$abs_diff[comparable]))
} else if (nrow(parity)) {
  "The optional semTests sentinel was requested, but it produced no comparable p-values; inspect semtests_parity.csv."
} else {
  "The optional semTests sentinel was not requested in this run."
}`

## Evidence

```{r calibration-table}
primary <- c(
  "pebad_ug_ml", "pebad_b_ml", "ss_b_rls", "peba6_ug_ml",
  "ss_b_ml", "sb_b_ml", "std_ml", "std_rls")
tab <- calibration[match(primary, calibration$method_id), c(
  "paper_label", "pooled_rejection", "rmse_percentage_points",
  "mad_percentage_points", "acceptable_percent",
  "below_2_5_percent", "above_7_5_percent")]
tab$pooled_rejection <- 100 * tab$pooled_rejection
knitr::kable(
  tab,
  digits = 3,
  row.names = FALSE,
  col.names = c(
    "Method", "Pooled rejection (%)", "Cell RMSE (pp)", "Cell MAD (pp)",
    "Acceptable cells (%)", "Below 2.5% (%)", "Above 7.5% (%)"),
  caption = paste(
    "Selected Study 2 methods.",
    if (is_full) "Comparable with the overlapping Satorra-2000 rows in published Table 9."
    else "Smoke/pilot rates are Monte Carlo diagnostics only.")
)
```

The design uses the paper's null loading sequence, one-factor weak-invariance
comparison, group-level sample sizes 400/800/2000, dimensions 5/10/20, and
2/4/8 groups. The seven generators are normal plus moderate and severe
Vale--Maurelli, independent-generator, and piecewise-linear conditions
(skewness/excess kurtosis 2/7 and 3/21). The full design therefore has 189
cells.

The 38 methods reproduced here are the Satorra (2000) subset of the published
comparison. They use the moment-tangent (`delta`) restriction used by semTests
and the published simulation: uncorrected ML/RLS plus nine robust families
crossed with ML/RLS bases and empirical/unbiased Gamma. The paper also retained
the Satorra--Bentler 2001 approximation as an additional comparator; it is
intentionally outside this replication.

```{r generator-table}
generator_view <- generator_summary[order(
  generator_summary$median_max_abs_excess_kurtosis_error,
  decreasing = TRUE), ]
generator_view <- head(generator_view, 8L)
knitr::kable(
  generator_view[, c(
    "p", "groups", "n_group", "distribution",
    "mean_achieved_skewness", "mean_achieved_excess_kurtosis",
    "median_max_abs_correlation_error",
    "median_max_relative_variance_error")],
  digits = 4, row.names = FALSE,
  col.names = c(
    "p", "Groups", "n/group", "Generator", "Mean skewness",
    "Mean excess kurtosis", "Median max |corr error|",
    "Median max relative variance error"),
  caption = "Most difficult observed generator cells by kurtosis error."
)
```

The generator calibration file separately records intermediate
correlation/root conditioning, setup cost, cache use, and analytic correlation
error where the generator exposes it. In this run,
`r sum(generator_calibration$calibration_ok)` of
`r nrow(generator_calibration)` distinct dimension-by-generator calibrations
passed their structural checks.

```{r parity-table}
if (nrow(parity) && any(is.finite(parity$abs_diff))) {
  comparable <- subset(parity, is.finite(abs_diff))
  by_cell <- do.call(rbind, lapply(split(comparable, comparable$cell_id),
    function(x) data.frame(
      p = x$p[[1L]], groups = x$groups[[1L]], n_group = x$n_group[[1L]],
      distribution = x$distribution[[1L]], df = x$df[[1L]],
      checks = nrow(x), passed = sum(x$pass),
      max_abs_delta_p = max(x$abs_diff))))
  knitr::kable(
    by_cell, digits = 7, row.names = FALSE,
    col.names = c("p", "Groups", "n/group", "Distribution", "df",
                  "Checks", "Passed", "Max |delta p|"),
    caption = "semTests sentinel: first replication in a spread of cells.")
}
```

```{r timing-table}
timing_primary <- subset(timing, phase == "replication_total")
timing_tab <- aggregate(
  timing_primary$median_seconds,
  list(p = timing_primary$p, groups = timing_primary$groups,
       distribution = timing_primary$distribution),
  median, na.rm = TRUE)
names(timing_tab)[[4L]] <- "median_seconds"
timing_tab <- timing_tab[order(
  timing_tab$median_seconds, decreasing = TRUE), ]
knitr::kable(
  head(timing_tab, 8L), digits = 3, row.names = FALSE,
  col.names = c("p", "Groups", "Distribution", "Median seconds/rep"),
  caption = "Slowest observed p x group x generator corners.")
```

## Caveats

`r if (is_full) {
  "This is a direct design replication, but exact agreement with the published rejection rates is not expected: magmaan replaces lavaan fitting and the original generator implementations while targeting the same population moments and test definitions. Monte Carlo uncertainty at 2000 replications is about 0.49 percentage points at a 5% rejection rate."
} else {
  sprintf("This %s run has only %d replications per cell. Rejection rates, RMSE rankings, and apparent winners are not inferential evidence. Its valid claims are that every design layer runs, the requested methods are finite, failures/warnings are recorded, and the optional oracle slice agrees.", profile, reps)
}`

The sentinel is deliberately sparse: it checks implementation parity, not
simulation validity. Generator calibration, achieved moments, convergence,
warning counts, cellwise Monte Carlo standard errors, and resumable raw
checkpoints remain separate diagnostics in `results/`.

## Reproduce

```bash
# Cheap structural run plus the independent R oracle
Rscript run_experiment.R --smoke --semtests-parity

# Speed/calibration pilot: all seven distributions at three geometry corners
Rscript run_experiment.R --pilot --cores 8 --semtests-parity

# Optional exact replication; not required for current magmaan hardening
Rscript run_experiment.R --full --cores 16 --semtests-parity
```

The target article is Foldnes, Grønneberg and Moss (2026), *Penalized
eigenvalue block averaging: Extension to nested model comparison and Monte
Carlo evaluations*, Behavior Research Methods 58:107,
[doi:10.3758/s13428-026-02968-4](https://doi.org/10.3758/s13428-026-02968-4).
The restriction-map construction follows Satorra, A. (2000), *Scaled and
adjusted restricted tests in multi-sample analysis of moment structures*, in
R. D. H. Heijmans, D. S. G. Pollock, and A. Satorra (Eds.), *Innovations in
Multivariate Statistical Analysis: A Festschrift for Heinz Neudecker*
(pp. 233--247), Kluwer Academic Publishers. The unbiased Gamma correction
follows Du, H., and Bentler, P. M. (2022), *40-Year Old Unbiased Distribution
Free Estimator Reliably Improves SEM Statistics for Nonnormal Data*,
*Structural Equation Modeling*, 29(6), 872--887,
[doi:10.1080/10705511.2022.2063870](https://doi.org/10.1080/10705511.2022.2063870).
