---
title: "Chen-Style Ordinal Missingness: Is It the Test or the Missing-Data Mechanism?"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- file.path("results", "summary.csv")
p_path <- file.path("results", "p_values.csv")
meta_path <- file.path("results", "metadata.csv")
if (!file.exists(summary_path) || !file.exists(p_path)) {
  stop("Run `Rscript run_experiment.R` first.", call. = FALSE)
}
S <- read.csv(summary_path, na.strings = "")
P <- read.csv(p_path, na.strings = "")
M <- read.csv(meta_path, na.strings = "")
meta <- function(key) M$value[match(key, M$key)]
fmt <- function(x, d = 3) ifelse(is.finite(x), sprintf(paste0("%.", d, "f"), x), "")

method_order <- c("naive", "SB", "adjusted", "scaled_shifted", "SS", "SF",
                  "EBA2", "EBA4", "EBA6", "pEBA2", "pEBA4", "pEBA6",
                  "pall", "pOLS", "all", "mixture")
focus_methods <- c("scaled_shifted", "pEBA4", "pOLS", "all", "mixture")
mech_label <- c(mar = "MAR", mcar = "MCAR")

cell_label <- function(rate, mech) {
  ifelse(rate <= 0, "0% (complete)",
         sprintf("%d%% %s", round(100 * rate), mech_label[mech]))
}
order_cells <- function(cells, rates, mechs) {
  key <- rates * 10 + ifelse(mechs == "mar", 0, 1)
  cells[order(key)]
}

S <- S[S$pd_gamma == "overlap", ]
S$cell <- cell_label(S$missing_rate, S$missing_mechanism)

mk_wide <- function(methods, frame) {
  x <- frame[frame$method %in% methods, ]
  x$rate <- sprintf("%.3f", x$rejection_rate)
  wide <- reshape(x[, c("cell", "method", "rate")], idvar = "cell",
                  timevar = "method", direction = "wide")
  names(wide) <- sub("^rate\\.", "", names(wide))
  for (m in methods) if (!m %in% names(wide)) wide[[m]] <- ""
  key <- x[!duplicated(x$cell), c("cell", "missing_rate", "missing_mechanism")]
  ord <- order_cells(key$cell, key$missing_rate, key$missing_mechanism)
  wide <- wide[match(ord, wide$cell), c("cell", methods)]
  names(wide)[1L] <- "Cell"
  wide
}

decomp_tab <- mk_wide(focus_methods, S)

battery <- S[S$missing_rate == 0.5 & S$method %in% method_order, ]
battery$rate <- sprintf("%.3f", battery$rejection_rate)
battery_tab <- reshape(battery[, c("method", "missing_mechanism", "rate")],
                       idvar = "method", timevar = "missing_mechanism",
                       direction = "wide")
names(battery_tab) <- sub("^rate\\.", "", names(battery_tab))
battery_tab <- battery_tab[match(method_order, battery_tab$method), ]
battery_tab <- data.frame(Method = battery_tab$method,
                          `MAR 50%` = battery_tab$mar,
                          `MCAR 50%` = battery_tab$mcar,
                          check.names = FALSE)

stat <- S[S$method == "scaled_shifted", ]
stat_tab <- data.frame(Cell = stat$cell,
                       `Realized Missing` = fmt(stat$realized_missing, 2),
                       `Mean Diff Statistic` = fmt(stat$mean_T_diff, 1),
                       `Mean Scale` = fmt(stat$mean_scale_c, 3),
                       check.names = FALSE)
ord <- order_cells(stat$cell, stat$missing_rate, stat$missing_mechanism)
stat_tab <- stat_tab[match(ord, stat_tab$Cell), ]

failures <- P[!is.finite(P$p_value) | !P$converged, ]
```

## Question

Experiment 33 showed WLSMV pairwise deletion inflates the ordinal scalar-invariance
difference test under missing data. Is that inflation a property of the *test
statistic* (which a better, full-spectrum p-value might repair), or of the
*missing-data mechanism* (which no p-value can repair)? We separate the two by
running the identical design under MAR and under MCAR.

## Short Answer

It is the mechanism, not the test. Under **MCAR**, every p-value variant holds its
nominal 5% even at 50% missing (.035 to .050). Under **MAR**, the same methods on
the same design inflate together to roughly **.30 to .35**. Pairwise-deletion
polychorics are consistent under MCAR but biased under MAR, so the damage is
upstream of the p-value: neither the scaled-shifted correction nor any
full-spectrum variant rescues it. The diagnostic confirms it: the difference
statistic's center is flat across MCAR (37.3 then 37.4) but climbs across MAR
(37.6, 41.4, 50.2) while the scale factor stays put near 0.98, so this is a
shifted center (a bias-driven non-centrality), not a miscalibrated scale.

The bottom line for practice is the same as Chen et al.'s "estimators *and*
missing-data techniques" split: under MAR the weak link is pairwise deletion, not
WLSMV. Full information (FIML / categorical links) or multiple imputation is the
fix.

This run uses no Mplus calls and writes only under this experiment's `results/`;
experiment 33's Mplus-parity files are untouched.

## Evidence

MAR drives missingness on group-B items `r meta("missing_items_group_b")` through
an auxiliary that correlates with the factor; MCAR drops the same items uniformly
at random. Complete data is the shared baseline.

```{r}
knitr::kable(decomp_tab, row.names = FALSE,
             caption = "Type-I rejection at nominal .05, by missingness cell.")
```

The split is not specific to one correction. At 50% missing, every method in the
battery is calibrated under MCAR and inflated under MAR:

```{r}
knitr::kable(battery_tab, row.names = FALSE,
             caption = "Full method battery at 50% missing: MAR vs MCAR.")
```

Why: the difference statistic acquires a non-centrality under MAR and none under
MCAR. The scale is unchanged throughout, so a calibration story is ruled out.

```{r}
knitr::kable(stat_tab, row.names = FALSE,
             caption = "Difference-statistic center by cell (reference df fixed).")
```

## Caveats

This is a Type-I study only (invariant population); it speaks to false-positive
control, not power. One cell (`N = r meta("n_total")`, symmetric thresholds,
overlap pairwise convention, `r meta("reps")` reps); Monte Carlo SE is about .015
at nominal and about .03 near .30. The MCAR result here keeps missingness
completely at random *within* the analyzed groups, which is the condition under
which pairwise deletion is consistent; group membership itself is a design
variable, not a missingness predictor.

```{r}
cat(nrow(failures), "failed p-value rows\n")
```

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 200 --n-total 1000 \
  --missing-rate 0,0.3,0.5 --thresholds symmetric --pd-gamma overlap \
  --missing-mechanism mar,mcar
quarto render report.qmd
```
