---
title: "Chen-Style Ordinal Missingness: Full-Spectrum Nested P-Values"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- file.path("results", "summary.csv")
p_path <- file.path("results", "p_values.csv")
meta_path <- file.path("results", "metadata.csv")
if (!file.exists(summary_path) || !file.exists(p_path)) {
  stop("Run `Rscript run_experiment.R` first.", call. = FALSE)
}
S <- read.csv(summary_path, na.strings = "")
P <- read.csv(p_path, na.strings = "")
M <- read.csv(meta_path, na.strings = "")
meta <- function(key) M$value[match(key, M$key)]
fmt <- function(x, d = 3) ifelse(is.finite(x), sprintf(paste0("%.", d, "f"), x), "")

method_order <- c("naive", "SB", "adjusted", "scaled_shifted", "SS", "SF",
                  "EBA2", "EBA4", "EBA6", "pEBA2", "pEBA4", "pEBA6",
                  "pall", "pOLS", "all", "mixture")
focus_methods <- c("scaled_shifted", "pEBA2", "pEBA4", "pEBA6",
                   "pall", "pOLS", "all", "mixture")

mk_wide <- function(methods) {
  x <- S[S$method %in% methods, ]
  x$rate <- sprintf("%.3f", x$rejection_rate)
  wide <- reshape(x[, c("missing_rate", "method", "rate")],
                  idvar = "missing_rate", timevar = "method",
                  direction = "wide")
  names(wide) <- sub("^rate\\.", "", names(wide))
  for (m in methods) if (!m %in% names(wide)) wide[[m]] <- ""
  wide <- wide[order(wide$missing_rate), c("missing_rate", methods)]
  names(wide)[1L] <- "Missing rate"
  wide
}

focus_tab <- mk_wide(focus_methods)
full_tab <- mk_wide(method_order)

stat_tab <- S[S$method == "scaled_shifted", c("missing_rate", "reps",
                                              "mean_T_diff", "mean_trace",
                                              "mean_scale_c")]
stat_tab$mean_T_diff <- fmt(stat_tab$mean_T_diff)
stat_tab$mean_trace <- fmt(stat_tab$mean_trace)
stat_tab$mean_scale_c <- fmt(stat_tab$mean_scale_c)
names(stat_tab) <- c("Missing rate", "Reps", "Mean Tdiff",
                     "Mean trace", "Mean scale")

failures <- P[!is.finite(P$p_value) | !P$converged, ]
```

## Question

If we drop Mplus from the loop and apply the ordinal FMG/full-spectrum p-values
directly to the Chen-style pairwise-missing scalar difference spectrum, do those
p-values repair the WLSMV pairwise-deletion Type-I inflation?

## Short Answer

No. In this run the full-spectrum methods inherit the missing-data problem in
the pairwise ordinal summary statistics. The exact mixture / `all` p-value is
slightly less liberal than scaled-shifted WLSMV at 50% missingness, but it still
rejects about 30% of invariant replications at nominal 5%. The penalized
pEBA/pOLS/pall family is more liberal here, around 35% at 50% missingness.

This is separate from experiment 33's Mplus-parity files. It uses no Mplus
calls and writes only under this experiment's `results/` directory.

## Design

The population is the Chen et al.-style one-factor ordinal CFA with ten
five-category indicators, equal groups, invariant loadings and thresholds, and
loading `r meta("loading")`. Missingness is rank-dependent on an auxiliary
variable and applied only to group-B `r meta("missing_items_group_b")`.

The fitted pair is configural versus Mplus-style scalar:

```text
H1: f =~ v1 + ... + v10
H0: group.equal = c("loadings", "thresholds")
    f ~ c(0, NA)*1
    v1 + ... + v10 observed intercepts fixed to c(0, 0)
```

The run shown here used `N = r meta("n_total")`, symmetric thresholds, overlap
pairwise Gamma, and `r meta("reps")` replications per missingness cell. The
overlap Gamma is used because experiment 33 established it as the Mplus-equivalent
pairwise convention for this DIFFTEST setting.

## Focus Methods

```{r}
knitr::kable(focus_tab, row.names = FALSE)
```

`scaled_shifted` is the WLSMV-style low-moment correction. `all` is the exact
weighted-mixture tail over the retained difference spectrum; `mixture` is the
same value through `robust_nested_lrt()`. `pEBA*`, `pOLS`, and `pall` are the
penalized full-spectrum approximations.

## Spectrum Summary

```{r}
knitr::kable(stat_tab, row.names = FALSE)
```

## Full Battery

```{r}
knitr::kable(full_tab, row.names = FALSE)
```

## Failures

```{r}
cat(nrow(failures), "failed p-value rows\n")
```

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --reps 200 --n-total 1000 \
  --missing-rate 0,0.3,0.5 --thresholds symmetric --pd-gamma overlap
quarto render report.qmd
```
