---
title: "Delta-method inference for the closed-form Guttman CFA estimator"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
need <- c("coverage", "risk", "gof", "difftest", "diagnostics", "metadata")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --full\n",
       "(or --smoke for a quick slice).")
}
coverage <- read.csv(paths[1]); risk <- read.csv(paths[2])
gof <- read.csv(paths[3]); difftest <- read.csv(paths[4])
diagnostics <- read.csv(paths[5]); metadata <- read.csv(paths[6])

meta <- stats::setNames(metadata$value, metadata$key)
gens <- intersect(c("normal", "ig", "ordinal"), unique(coverage$generator))
gen_label <- c(normal = "Normal", ig = "Independent generator",
               ordinal = "Ordinal (as continuous)")
rels <- sort(unique(coverage$reliability))
big_n <- max(coverage$n); small_n <- min(coverage$n)
lo_rel <- min(rels); hi_rel <- max(rels)
pct <- function(x) sprintf("%.1f%%", 100 * x)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)

rr <- function(gen, rel, group = "loadings", n = big_n) {
  s <- risk[risk$generator == gen & risk$reliability == rel & risk$group == group &
            risk$n == n & risk$spec == "correct", ]
  g <- s$rmse[s$estimator == "guttman"]; m <- s$rmse[s$estimator == "ml"]
  if (length(g) && length(m) && m > 0) g / m else NA_real_
}
cov_at <- function(gen, variant, rel, n = big_n) {
  v <- coverage$coverage[coverage$group == "all" & coverage$generator == gen &
                         coverage$variant == variant & coverage$reliability == rel &
                         coverage$n == n]
  if (length(v)) v[1] else NA_real_
}
```

## Question

magmaan's closed-form (non-iterative) CFA estimators now carry delta-method
inference: standard errors, a residual goodness-of-fit test, and a
difference / pseudo-LRT test. A first study established these are calibrated on
normal, non-normal, and ordinal data at 0.7 reliability, at a small efficiency
cost vs ML. This extension pushes into the hard regime, sweeping indicator
reliability down to `r fmt(lo_rel, 1)`, where the observed correlations shrink,
ML's optimal weighting should matter most, and the clamp-free Guttman map nears
its assumptions. Three questions: does the loadings efficiency gap widen, does
the inference stay calibrated, and how gracefully do the estimators degrade?

## Short answer

The efficiency gap **widens as reliability falls, spreading from the loadings to
the structural parameters generally, while residual variances stay at parity**.
On the independent generator the Guttman-to-ML RMSE ratio for loadings goes from
`r fmt(rr("ig", hi_rel), 2)` at reliability `r fmt(hi_rel,1)` to
`r fmt(rr("ig", lo_rel), 2)` at `r fmt(lo_rel,1)`; factor covariances follow the
same shape (`r fmt(rr("ig", hi_rel, "factor_cov"), 2)` to
`r fmt(rr("ig", lo_rel, "factor_cov"), 2)`); residual variances barely move off 1.
So ML's optimal weighting buys the most exactly where theory says it should, on
the parameters that carry the covariance structure under weak signal, and almost
nothing on the residual diagonal.

The inference **holds across the whole reliability range**: empirical-Gamma
coverage stays near 95% at every reliability (independent generator, n =
`r big_n`: `r pct(cov_at("ig","g_ntml_emp",lo_rel))` at reliability
`r fmt(lo_rel,1)` vs `r pct(cov_at("ig","g_ntml_emp",hi_rel))` at
`r fmt(hi_rel,1)`). The normal-theory Gamma undercovers on non-normal data at
every reliability, though, interestingly, *least* at low reliability
(`r pct(cov_at("ig","g_ntml_nt",lo_rel))` at `r fmt(lo_rel,1)` vs
`r pct(cov_at("ig","g_ntml_nt",hi_rel))` at `r fmt(hi_rel,1)`): weak loadings
dilute the non-normal common factor with normal residual noise, so the data look
more Gaussian. Low reliability does not break the delta method; non-normality
plus NT-Gamma does, as before.

The estimators **degrade gracefully, and the closed form has the edge**. At the
stress corner (reliability `r fmt(lo_rel,1)`, n = `r small_n`) *both* throw
Heywood cases, but ML produces more of them than Guttman and also occasionally
fails to converge, whereas the closed form is a formula that always returns a
finite, inspection-free answer. Guttman trades a few percent of structural-
parameter precision for that robustness.

## Evidence

### The efficiency gap is a loadings-only, low-reliability phenomenon

Guttman-to-ML RMSE ratio at n = `r big_n`, by reliability. A ratio of 1 means
parity with ML; above 1 is Guttman's efficiency loss.

```{r frontier-plot, fig.height=3.4, fig.width=7}
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2.2, 0.6), oma = c(0, 0, 0, 0))
groups <- c(loadings = "Loadings", factor_cov = "Factor (co)variances",
            residual = "Residual variances")
cols <- c(normal = "#333333", ig = "#0072b2", ordinal = "#c1272d")
for (grp in names(groups)) {
  plot(NA, xlim = range(rels), ylim = c(0.98, 1.28),
       xlab = "Reliability", ylab = if (grp == "loadings") "RMSE ratio Guttman / ML" else "",
       main = groups[[grp]])
  abline(h = 1, lty = 3, col = "grey55")
  for (g in gens) {
    y <- vapply(rels, function(r) rr(g, r, grp), numeric(1))
    lines(rels, y, col = cols[g], lwd = 2, type = "b", pch = 19)
  }
  if (grp == "loadings")
    legend("topright", legend = gen_label[gens], col = cols[gens], lwd = 2, pch = 19,
           bty = "n", cex = 0.85)
}
par(op)
```

*Loadings and factor (co)variances: the gap grows as reliability drops (worst on
the ordinal arm), reaching ~20-25% on the loadings at reliability
`r fmt(lo_rel,1)`. Residual variances: essentially at parity with ML across the
whole range.*

### Inference stays calibrated across reliability

95% interval coverage (all 21 parameters) at n = `r big_n`, empirical Gamma vs
normal-theory Gamma, by reliability.

```{r coverage-table}
variants <- c(g_ntml_emp = "Guttman, empirical Gamma",
              g_ntml_nt  = "Guttman, NT Gamma",
              ml_rob     = "ML, robust")
rows <- list()
for (g in gens) for (v in names(variants)) {
  r <- data.frame(Generator = gen_label[g], Method = variants[[v]], check.names = FALSE)
  for (rl in rels) r[[paste0("rel ", fmt(rl, 1))]] <- pct(cov_at(g, v, rl))
  rows[[paste(g, v)]] <- r
}
knitr::kable(do.call(rbind, rows), row.names = FALSE, align = "llccc",
             caption = sprintf("Coverage of 95%% intervals at n = %d.", big_n))
```

Empirical-Gamma coverage sits near nominal at every reliability and generator;
the normal-theory Gamma undercovers on the non-normal arms independently of
reliability. The goodness-of-fit test tells the same story it did at 0.7 (near
nominal under the empirical/Satorra-Bentler reference, over-rejecting under
NT-Gamma at small n), unchanged by the reliability sweep; the per-cell rates are
in `results/gof.csv`.

### Graceful degradation: Heywood cases and convergence

Improper-solution rate (any negative residual-variance estimate) and
convergence, pooled over generators, by reliability and n. Reliability
`r fmt(lo_rel,1)` at small n is the stress corner.

```{r degradation-table}
dcorr <- diagnostics[diagnostics$spec == "correct", ]
agg <- function(col) {
  a <- aggregate(dcorr[[col]], list(reliability = dcorr$reliability, n = dcorr$n), mean)
  names(a)[3] <- col; a
}
imp_g <- agg("improper_guttman"); imp_m <- agg("improper_ml"); cvm <- agg("conv_ml")
show_ns <- sort(unique(dcorr$n))
mk <- function(df, val, lab) {
  w <- reshape(df, idvar = "reliability", timevar = "n", direction = "wide")
  out <- data.frame(Reliability = fmt(w$reliability, 1), Quantity = lab, check.names = FALSE)
  for (nn in show_ns) out[[paste0("n=", nn)]] <- fmt(w[[paste0(val, ".", nn)]], 3)
  out
}
tbl <- rbind(mk(imp_g, "improper_guttman", "Guttman improper"),
             mk(imp_m, "improper_ml", "ML improper"),
             mk(cvm, "conv_ml", "ML converged"))
tbl <- tbl[order(tbl$Reliability, tbl$Quantity), ]
knitr::kable(tbl, row.names = FALSE, caption = "Improper-solution and ML convergence rates (Guttman convergence is 1.000 everywhere by construction).")
```

Both estimators throw Heywood cases at low reliability and small n, but ML throws
more of them than the closed form and, unlike the closed form, occasionally fails
to converge (Guttman convergence is 1.000 by construction). At the stress corner
the closed form is the better-behaved of the two, which is a structural advantage
on top of its inference being interchangeable with ML's.

### Efficiency and difference test at a glance

RMSE over all 21 parameters and the difference-test Type-I (true restriction
`f1 ~~ 0*f3`) at n = `r big_n`, lowest and highest reliability.

```{r summary-table}
d <- difftest[difftest$ptype == "mixture" & difftest$variant == "ntml_emp" &
              difftest$n == big_n, ]
rows <- list()
for (g in gens) for (rl in c(lo_rel, hi_rel)) {
  rows[[paste(g, rl)]] <- data.frame(
    Generator = gen_label[g], Reliability = fmt(rl, 1),
    `RMSE ratio (all)` = fmt(rr(g, rl, "all"), 3),
    `Diff-test Type-I` = fmt(d$reject_rate[d$generator == g & d$reliability == rl][1], 3),
    check.names = FALSE)
}
knitr::kable(do.call(rbind, rows), row.names = FALSE, align = "llrr",
             caption = sprintf("Overall efficiency and difference-test calibration at n = %d.", big_n))
```

## Caveats

- One population and model (balanced 3-factor CFA, simple structure); reliability
  is swept by scaling residual variances, holding the loading pattern and factor
  correlations fixed. This maps the machinery's behaviour, not a performance claim.
- The ordinal arm is Pearson-codes calibrated so the treated-as-continuous
  covariance is an *exact* factor model, isolating discrete skewed marginals'
  effect on the fourth-moment matrix; it is not categorization-induced
  misspecification.
- Rates are Monte Carlo estimates at `r meta[["reps"]]` replications per cell;
  full per-generator / per-n / per-parameter breakdowns are in `results/`.
- The empirical Gamma is mildly conservative for the goodness-of-fit test in
  small samples; the normal-theory Gamma errs the other way.

## Reproduce

```sh
# full grid: 3 generators x 2 specs x 3 reliabilities x 5 sample sizes, 1000 reps
Rscript run_experiment.R --full --reps 1000 --cores 10

# quick slice
Rscript run_experiment.R --smoke

quarto render report.qmd
```
