---
title: "Delta-method inference for the closed-form Guttman CFA estimator"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
need <- c("coverage", "risk", "gof", "difftest", "diagnostics", "metadata")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --full\n",
       "(or --smoke for a quick slice).")
}
coverage <- read.csv(paths[1]); risk <- read.csv(paths[2])
gof <- read.csv(paths[3]); difftest <- read.csv(paths[4])
diagnostics <- read.csv(paths[5]); metadata <- read.csv(paths[6])

gens <- intersect(c("normal", "ig", "ordinal"), unique(coverage$generator))
gen_label <- c(normal = "Normal", ig = "Independent generator",
               ordinal = "Ordinal (as continuous)")
regime_order <- c("bal.3", "bal.5", "bal.7", "het.mod", "het.str")
regimes <- intersect(regime_order, unique(coverage$regime))
regime_label <- c(bal.3 = "Uniform 0.3", bal.5 = "Uniform 0.5", bal.7 = "Uniform 0.7",
                  het.mod = "Het .65/.5/.35", het.str = "Het .8/.5/.2")
big_n <- max(coverage$n); small_n <- min(coverage$n)
pct <- function(x) sprintf("%.1f%%", 100 * x)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)

rr <- function(gen, regime, group = "loadings", n = big_n) {
  s <- risk[risk$generator == gen & risk$regime == regime & risk$group == group &
            risk$n == n & risk$spec == "correct", ]
  g <- s$rmse[s$estimator == "guttman"]; m <- s$rmse[s$estimator == "ml"]
  if (length(g) && length(m) && m > 0) g / m else NA_real_
}
cov_at <- function(gen, variant, regime, group = "all", n = big_n) {
  v <- coverage$coverage[coverage$group == group & coverage$generator == gen &
                         coverage$variant == variant & coverage$regime == regime &
                         coverage$n == n]
  if (length(v)) v[1] else NA_real_
}
imp <- function(col, regime, n) {
  d <- diagnostics[diagnostics$spec == "correct" & diagnostics$regime == regime &
                   diagnostics$n == n, ]
  mean(d[[col]])
}
```

## Question

magmaan's closed-form (non-iterative) CFA estimators carry delta-method
inference; earlier runs showed it is calibrated across data types and down a
uniform-reliability sweep, at a small efficiency cost vs ML. This extension asks
whether **weak loadings** are a distinct stress. They are not, on their own: at
matched reliability a weak loading with a small residual and a strong loading
with a large residual produce the same correlation matrix, and both estimators
are scale-equivariant, so "uniform weak loadings" is just the low end of the
reliability sweep. The manipulation that *is* distinct is **heterogeneity** -- a
weak indicator sitting next to strong ones at the same average reliability. We
run three balanced regimes (the reliability sweep) and two heterogeneous ones
(average reliability 0.5, one weak indicator per factor), tracking loadings by
within-factor position.

## Short answer

A single weak indicator per factor **inflates the closed form's efficiency gap
as much as making every indicator weak does**, at the same average reliability:
on the ordinal arm the all-parameter Guttman-to-ML RMSE ratio goes from
`r fmt(rr("ordinal","bal.5","all"),2)` (uniform 0.5) to
`r fmt(rr("ordinal","het.str","all"),2)` (one weak indicator, average still 0.5),
matching the uniform-0.3 regime.

The twist is **where the loss lands: on the weak indicator's factor-mates, not
on the weak indicator itself**. In the strongest heterogeneous regime the
loadings RMSE ratio for the *middle* indicator is
`r fmt(rr("ig","het.str","loadings_p2"),2)`-`r fmt(rr("ordinal","het.str","loadings_p2"),2)`,
but for the *weak* indicator only
`r fmt(rr("ig","het.str","loadings_p3"),2)`-`r fmt(rr("ordinal","het.str","loadings_p3"),2)`.
Guttman's communality for an indicator divides by the correlation between the
*other two* indicators; for the middle indicator that pair includes the weak one
(a small, noisy denominator), while the weak indicator's own denominator is the
strong marker-middle pair. ML pools optimally and has no such asymmetry.

The **inference stays calibrated throughout** -- empirical-Gamma coverage is near
95% in every regime, including for the weak indicator's own loading
(`r pct(cov_at("ig","g_ntml_emp","het.str","loadings_p3"))`-`r pct(cov_at("ordinal","g_ntml_emp","het.str","loadings_p3"))`).
And the **graceful-degradation story reverses**: under uniform low reliability
the closed form throws *fewer* Heywood cases than ML, but under heterogeneity the
weak indicator makes it throw *more* (n = `r small_n*2`:
`r pct(imp("improper_guttman","het.str",small_n*2))` vs ML's
`r pct(imp("improper_ml","het.str",small_n*2))`).

## Evidence

### The gap falls on the factor-mates, not the weak indicator

Guttman-to-ML loadings RMSE ratio at n = `r big_n`, at matched average
reliability 0.5, as the spread increases from balanced to strongly
heterogeneous. Separate lines for the middle indicator (position 2) and the weak
indicator (position 3).

```{r position-plot, fig.height=3.4, fig.width=7}
matched <- c("bal.5", "het.mod", "het.str")
matched <- intersect(matched, regimes)
xlab_at <- seq_along(matched)
op <- par(mfrow = c(1, 3), mar = c(5.5, 4, 2.2, 0.6))
for (g in gens) {
  plot(NA, xlim = c(0.8, length(matched) + 0.2), ylim = c(1.0, 1.42),
       xaxt = "n", xlab = "", ylab = if (g == gens[1]) "RMSE ratio Guttman / ML" else "",
       main = gen_label[g])
  axis(1, at = xlab_at, labels = regime_label[matched], las = 2, cex.axis = 0.8)
  abline(h = 1, lty = 3, col = "grey55")
  y2 <- vapply(matched, function(rg) rr(g, rg, "loadings_p2"), numeric(1))
  y3 <- vapply(matched, function(rg) rr(g, rg, "loadings_p3"), numeric(1))
  lines(xlab_at, y2, col = "#c1272d", lwd = 2, type = "b", pch = 19)
  lines(xlab_at, y3, col = "#0072b2", lwd = 2, type = "b", pch = 17)
  if (g == gens[1])
    legend("topleft", legend = c("Middle indicator (r=0.5)", "Weak indicator (r->0.2)"),
           col = c("#c1272d", "#0072b2"), lwd = 2, pch = c(19, 17), bty = "n", cex = 0.85)
}
par(op)
```

*Average reliability is 0.5 in all three regimes. As one indicator is made weak,
the closed form loses ground fastest on the middle indicator (whose triad
denominator is the marker-weak correlation), while the weak indicator's own
loading is comparatively spared. Uniform-reliability regimes (not shown here) sit
in `results/risk.csv`; the balanced sweep reproduces the earlier finding that the
gap widens as reliability drops.*

### Inference stays calibrated across every regime

95% interval coverage at n = `r big_n`, empirical Gamma, for all parameters and
for the weak indicator's loading alone.

```{r coverage-table}
rows <- list()
for (g in gens) {
  r_all <- data.frame(Generator = gen_label[g], Parameters = "All 21", check.names = FALSE)
  r_wk  <- data.frame(Generator = gen_label[g], Parameters = "Weak loading", check.names = FALSE)
  for (rg in regimes) {
    r_all[[regime_label[rg]]] <- pct(cov_at(g, "g_ntml_emp", rg, "all"))
    r_wk[[regime_label[rg]]]  <- pct(cov_at(g, "g_ntml_emp", rg, "loadings_p3"))
  }
  rows[[paste(g, "all")]] <- r_all; rows[[paste(g, "wk")]] <- r_wk
}
knitr::kable(do.call(rbind, rows), row.names = FALSE,
             caption = sprintf("Empirical-Gamma coverage at n = %d.", big_n))
```

Heterogeneity does not break the delta method. Even the weak indicator's loading,
which the closed form estimates least efficiently, has near-nominal coverage.

### Degradation reverses under heterogeneity

Improper-solution rate (any negative residual-variance estimate), pooled over
generators, at the two smallest sample sizes. Guttman convergence is 1.000
everywhere by construction.

```{r degradation-table}
show_ns <- sort(unique(diagnostics$n))[1:2]
rows <- list()
for (rg in regimes) {
  for (est in c("improper_guttman", "improper_ml")) {
    r <- data.frame(Regime = regime_label[rg],
                    Estimator = if (est == "improper_guttman") "Guttman" else "ML",
                    check.names = FALSE)
    for (nn in show_ns) r[[paste0("n=", nn)]] <- pct(imp(est, rg, nn))
    rows[[paste(rg, est)]] <- r
  }
}
knitr::kable(do.call(rbind, rows), row.names = FALSE,
             caption = "Improper-solution rate (Heywood cases).")
```

Under uniformly weak signal (`r regime_label[["bal.3"]]`) the closed form is the
steadier of the two: fewer Heywood cases than ML, and it cannot fail to converge.
Under a single weak indicator (`r regime_label[["het.str"]]`) that advantage
flips -- the weak indicator's triad communality overshoots into negative residual
variances more often than ML's, and the excess persists to larger n. The closed
form is more robust to weak *signal* and less robust to weak *balance*.

## Caveats

- One population and model (balanced 3-factor CFA, simple structure, three
  indicators per factor, factor pair f1-f3 uncorrelated). The three-indicator
  case is the most fragile for a triad-based communality; more indicators would
  average over triads and soften the heterogeneity effect.
- Regimes vary the within-factor reliability pattern with residual variances
  fixed at 1, so a low-reliability position is literally a weak loading; results
  are on the RMSE-ratio scale, which is invariant to that choice.
- The ordinal arm is Pearson-codes calibrated to an exact factor covariance;
  rates are Monte Carlo estimates at `r metadata$value[metadata$key=="reps"]`
  replications per cell. Full breakdowns are in `results/`.
- Goodness-of-fit and difference-test calibration track the earlier runs and are
  unchanged by the regime (per-cell rates in `results/gof.csv`,
  `results/difftest.csv`).

## Reproduce

```sh
# full grid: 3 generators x 2 specs x 5 regimes x 5 sample sizes, 1000 reps
Rscript run_experiment.R --full --reps 1000 --cores 10

# quick slice
Rscript run_experiment.R --smoke

quarto render report.qmd
```
