---
title: "Which robust tests did Savalei and Falk (2014) actually run?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
needed <- c("configuration_probe.csv", "implementation_map.csv",
            "published_design.csv", "missingness_sets.csv",
            "published_parameters.csv", "metadata.csv")
missing <- needed[!file.exists(file.path("results", needed))]
if (length(missing)) {
  stop("missing results; run `Rscript run_experiment.R --audit`")
}
probe <- utils::read.csv(file.path("results", "configuration_probe.csv"),
                         stringsAsFactors = FALSE, check.names = FALSE)
implementation <- utils::read.csv(file.path("results", "implementation_map.csv"),
                                  stringsAsFactors = FALSE, check.names = FALSE)
design <- utils::read.csv(file.path("results", "published_design.csv"),
                          stringsAsFactors = FALSE, check.names = FALSE)
missing_sets <- utils::read.csv(file.path("results", "missingness_sets.csv"),
                                stringsAsFactors = FALSE, check.names = FALSE)
parameters <- utils::read.csv(file.path("results", "published_parameters.csv"),
                              stringsAsFactors = FALSE, check.names = FALSE)
metadata_frame <- utils::read.csv(file.path("results", "metadata.csv"),
                                  stringsAsFactors = FALSE)
metadata <- setNames(metadata_frame$value, metadata_frame$key)
probe_stat <- function(route, implementation_prefix = NULL) {
  rows <- probe$route == route
  if (!is.null(implementation_prefix)) {
    rows <- rows & startsWith(probe$implementation, implementation_prefix)
  }
  stopifnot(sum(rows) == 1L)
  probe$corrected_statistic[rows]
}
```

## Question

Exactly which null and reference models, likelihood/discrepancy statistics, and
observed-versus-expected information choices produced the robust FIML and robust
two-stage tests in Savalei and Falk (2014)? Which current magmaan paths are the
same tests, and what must be verified before using them in the experiment research/44
design?

## Short answer

The paper ran **global exact-fit Type-I tests**, not nested restriction tests and
not a power study. In every generated sample the structured CFA was true
($H_0$); the comparison/reference model ($H_1$) was the saturated mean-and-
covariance model. Robust FIML scaled the ordinary observed-data saturated-versus-
structured likelihood ratio using the residual-space trace from **analytic
observed information evaluated at the structured fit**. Robust two-stage first
estimated saturated EM means/covariances and their observed-information sandwich,
then fit the structured CFA to those moments and scaled $(n-1)F_{ML}$ using the
corresponding complete-data observed-information residual projector. Both scaled
statistics were referred to $\chi^2_d$ and rejected at $.05$.

Neither current magmaan route in experiment research/44 is that FIML statistic. Its MLR
comparator implements the modern Mplus/Yuan--Bentler difference-of-traces
correction; its FMG-SB route uses a saturated-$H_1$ metric. Current ML2S, in turn,
matches current lavaan `robust.two.stage`, whose test metric is forced to the
unstructured/saturated side. These are legitimate nearby methods, but none is a
verification oracle for the 2014 EQS `SE=EXACT` configuration. Experiment research/44
should gain explicitly named paper-era statistics only after same-data matrix
parity has been established.

On the deterministic routing witness, the reconstructed paper-FIML proxy is
`r sprintf("%.3f", probe_stat("paper-FIML configuration proxy"))`, versus
`r sprintf("%.3f", probe_stat("current MLR", "magmaan"))` for current MLR and
`r sprintf("%.3f", probe_stat("FIML FMG-SB"))` for FMG-SB; the scalar difference
is large enough that these routes cannot be treated as aliases.

## Evidence

### The tests, reconstructed

Write

$$
\beta=(\operatorname{vech}\Sigma',\mu')',\qquad
\Omega=A^{-1}BA^{-1},\qquad
\Delta=\frac{\partial\beta(\theta)}{\partial\theta'}.
$$

Here $A$ is observed information and $B$ is the covariance of individual
likelihood scores. The article says explicitly that $A$ must be observed for
consistency under MAR, that all simulation computations used observed
information, and that Appendix B's `SE=EXACT` computed it analytically rather
than by numerical differentiation (article pp. 282, 285, 301--302).

| Item | Robust FIML | Robust two-stage |
|---|---|---|
| Generated $H_0$ | Structured CFA mean/covariance model, true in every cell | Same structured CFA, true in every cell |
| Saturated $H_1$ role | Supplies $\hat\beta$ and the maximized observed-data likelihood | Supplies Stage-1 EM $\hat\beta$ and $\hat\Omega$ |
| Structured estimate | Direct incomplete-data ML $\hat\theta$ | Complete-data ML discrepancy fit $\tilde\theta$ to $\hat\beta$ |
| Base statistic | $T_{ML}=2\{\ell(\hat\beta)-\ell(\hat\theta)\}$ | $(n-1)F_{ML}(\tilde\theta)$ |
| Residual projector | $W=A-A\Delta(\Delta'A\Delta)^{-1}\Delta'A$, with the Equation 2 quantities evaluated at the structured model | $U=H-H\Delta(\Delta'H\Delta)^{-1}\Delta'H$, with complete-data observed $H$ at the Stage-2 fit |
| Scale and test | $c_F=\operatorname{tr}(\Omega W)/d$; use $T_{ML}/c_F$ | $c_{TS}=\operatorname{tr}(\Omega U)/d$; use $(n-1)F_{ML}/c_{TS}$ |
| Reference | $\chi^2_d$, reject at $\alpha=.05$ | $\chi^2_d$, reject at $\alpha=.05$ |

This follows Equations 1--6 of the [published article](https://doi.org/10.1080/10705511.2014.882692).
The earlier open [two-stage derivation](https://escholarship.org/uc/item/89f8q6jc)
also makes the order explicit: saturated observed-data ML first, complete-data
discrepancy minimization second, then projection of the Stage-1 covariance
through the Stage-2 model derivatives. Savalei's separate paper on
[expected versus observed information](https://pubmed.ncbi.nlm.nih.gov/20853954/)
is the information-choice source cited by Appendix B.

The means are part of both $H_0$ and $H_1$, but were freely estimated and
therefore cancel from the model degrees of freedom. Thus Model 1 has $p=14$ and
$d=76$; Model 2 has $p=21$ and $d=186$. The paper's statement
$d=p(p+1)/2-q$ counts covariance restrictions, with $q$ the freely estimated
covariance-structure parameters.

### What “observed” does and does not mean

There are two different information objects:

1. **Stage-1/saturated incomplete-data information $A$.** It is the negative
   observed Hessian of the saturated likelihood. Together with the casewise
   score covariance $B$, it produces $\Omega=A^{-1}BA^{-1}$.
2. **The metric used to remove the structured tangent space.** For FIML this is
   the structured-model $A$ in $W$. For two-stage it is the complete-data $H$
   in $U$. The paper selected the observed form here as well.

`SE=EXACT` is therefore not “expected information,” and it is not EQS's
numerical observed-Hessian option: Appendix B says `SE=OBS` was the numerical
approximation, whereas `SE=EXACT` was the analytic observed-information formula.
The later computational review explains why these choices remain distinct in
modern software ([Savalei and Rosseel, 2022](https://doi.org/10.1080/10705511.2021.1877548)).

### Mapping to magmaan

```{r implementation-table}
show <- implementation[, c("route", "base_statistic", "correction_geometry",
                            "paper_target")]
names(show) <- c("Route", "Base", "Correction geometry", "2014 target?")
knitr::kable(show, row.names = FALSE)
```

The important distinction is not the base statistic: the FIML routes all begin
with the same saturated-versus-structured likelihood ratio, and the two-stage
routes all begin with the same Stage-2 discrepancy. The disagreement is the
metric and evaluation point inside the trace correction. Consequently, agreement
of point estimates or unscaled chi-square does not verify the paper statistic.

### A same-data routing witness

The audit runner fits one deterministic, nonnormal, MAR-incomplete, six-indicator
CFA. It is not a replication cell; it is a small engineering witness that the
three FIML correction routes are numerically distinguishable while magmaan still
matches the two modern lavaan routes it claims to implement.

```{r probe-table}
show <- probe[, c("route", "implementation", "base_statistic",
                  "corrected_statistic", "df", "scaling_factor", "p_value")]
names(show) <- c("Route", "Implementation", "Base", "Corrected", "df",
                 "Scale", "p")
for (field in c("Base", "Corrected", "Scale", "p")) {
  show[[field]] <- sprintf("%.6f", show[[field]])
}
knitr::kable(show, row.names = FALSE,
             caption = "Deterministic routing witness; not simulation evidence.")
```

The closest configurable lavaan reconstruction of the paper's FIML formula uses
Yuan--Bentler with observed information on the structured side. It is a useful
secondary sentinel, not an oracle for the original EQS implementation. The
original paper target must be checked from the displayed matrices and trace,
not blessed merely because this scalar agrees.

### The published simulation target

The exact-fit study crossed 2 models, 3 sample sizes, 2 kurtoses, 2 missing
proportions, 2 pattern regimes, and 4 missingness mechanisms: `r nrow(design)`
cells, 1,000 replications per cell, with both estimators applied to every sample
(`r format(nrow(design) * 1000L * 2L, big.mark = ",")` fits).

| Axis | Levels |
|---|---|
| CFA | 14 variables / 2 factors / 76 df; 21 variables / 3 factors / 186 df |
| $N$ | 200, 400, 600 |
| Margins | skewness 2; kurtosis 7 or 15; Vale--Maurelli/Fleishman in EQS 6.1 |
| Missing proportion | 15% or 30% on each selected variable set |
| Patterns | FP: at most 4; MP: at most 64 |
| Mechanism | MCAR, MAR-L ($x>0$), MAR-L2 ($x<0$), MAR-NL ($|x|>.54$) |
| Outcome for these tests | rejection proportion at $\alpha=.05$ under a true model |

Deletion proceeded set by set, repeatedly sampling rows until each set reached
its target missing proportion. A row could be chosen for multiple sets, creating
combined patterns. The exact conditioning/deletion map was:

```{r missing-sets}
show <- missing_sets
names(show) <- c("Model", "Pattern regime", "Condition on", "Delete together")
knitr::kable(show, row.names = FALSE)
```

For MCAR, the conditioning column was irrelevant. For MAR-L deletion was allowed
when the conditioning variable exceeded 0; for MAR-L2 when it was below 0; and
for MAR-NL when its absolute value exceeded .54.

Appendix A supplies all 21 standardized Model 2 loadings; residual variances are
$1-\lambda^2$, all observed means are zero and variances one, and factor
correlations are $.446$, $-.124$, and $-.085$ for F1--F2, F1--F3, and F2--F3.
Model 1 takes the F1 and F3 blocks, relabeling V15--V21 as V8--V14, with factor
correlation $-.124$. The machine-readable loading table is in
`results/published_parameters.csv`; `results/published_design.csv` contains the
complete 192-cell grid.

### Verification plan

1. **Freeze an independent equation-level reference.** Implement Equations 2--6
   directly for one fixed dataset. Export and compare $A$, $B$, $\Omega$,
   $\Delta$, $H$, the residual projector, its numerical rank, the trace, scale,
   corrected statistic, df, and p-value. Use the current lavaan reconstruction
   only as a secondary scalar sentinel.
2. **Add paper-specific routes, without changing current ones.** Give the 2014
   FIML and two-stage configurations explicit names. Preserve current MLR and
   ML2S because they already have different, documented lavaan targets.
3. **Gate identities before simulation.** Require symmetry and rank $d$ of the
   residual projector; direct equality of the independently computed trace;
   correct $(n-1)$ handling; mean rows cancelling from df; and convergence toward
   scale 1 under complete normal data. Require matrix agreement around $10^{-8}$
   and scalar agreement around $10^{-6}$ on the fixed reference case before
   interpreting Monte Carlo output.
4. **Run a focused replication first.** Start with Model 1 at $N=200$, both
   kurtoses and missing proportions, FP/MP, MCAR and the harsh MAR-L mechanism.
   Record achieved marginal moments, missing proportions/pattern counts,
   convergence, inadmissibility, and every excluded replication. Compare
   rejection rates with binomial 95% intervals.
5. **Expand only after parity.** Run all 192 cells only when the same-data gates
   pass. Bring the resulting explicitly labeled paper statistics into experiment
   77 as additional comparators; do not replace its current MLR baseline.

## Caveats

The statistic definitions, information choice, degrees of freedom, missingness
rules, population parameters, and factorial design are recoverable from the
article. Exact bitwise replication is not: the paper does not supply random
seeds, the precise EQS random-number/Fleishman implementation, raw cell outputs,
or complete failure/exclusion logs. The article says plotting code and results
for alternative statistics were available from the authors on request. A full
replication should therefore target equation-level parity and the published
qualitative/Monte Carlo pattern, not identical individual draws.

The current lavaan proxy can help diagnose the FIML route, but it postdates EQS
6.1 and cannot adjudicate the two-stage observed-$H$ implementation by itself.
Obtaining original EQS output or author code would strengthen the final oracle.

## Reproduce

```sh
Rscript run_experiment.R --audit
Rscript run_experiment.R --dry-run-plan
quarto render report.qmd
```

The audit uses seed `r metadata[["seed_base"]]`; package versions and the current
Git reference are recorded in `results/metadata.csv`.
