---
title: "Optimizer for the latent-determinacy barrier"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
profile <- Sys.getenv("E18_PROFILE", "full")
res <- file.path("results", profile)
need <- file.path(res, "fits.csv")
if (!file.exists(need)) {
  stop("Missing results. Run: Rscript run_experiment.R --full", call. = FALSE)
}
f <- read.csv(need, stringsAsFactors = FALSE)
f$fail <- f$status != "ok" | !f$converged
key <- c("study", "design", "chart", "n", "rep", "lambda")
best <- aggregate(pen_fmin ~ study + design + chart + n + rep + lambda,
                  f[!f$fail, ], min)
names(best)[7] <- "best"
f <- merge(f, best, by = key, all.x = TRUE)
f$gap <- f$pen_fmin - f$best
tol <- 1e-7
f$certified <- !f$fail & !is.na(f$gap) & f$gap <= tol
f$worse <- !f$fail & !is.na(f$gap) & f$gap > tol
opt_label <- c(`nlopt-lbfgs` = "L-BFGS (previous default)", port = "PORT",
               `nlopt-slsqp` = "SLSQP",
               `nlopt-lbfgs-slsqp-fallback` = "L-BFGS, SLSQP fallback")
opts <- names(opt_label)[names(opt_label) %in% unique(f$optimizer)]
f$group <- with(f, ifelse(study == "research47",
                          paste0("Research/47 populations, λ = ", lambda),
                  ifelse(study == "ernst", paste0("Ernst design, ", chart),
                  ifelse(study == "fold", paste0("Collinear predictors, ", chart),
                         "Textbook corpus cases"))))
groups <- unique(f$group[order(f$study != "research47", f$study != "ernst",
                               f$study != "fold", f$group)])
pct <- function(x) sprintf("%.1f", 100 * x)
pct2 <- function(x) sprintf("%.2f", 100 * x)
rate <- function(g, o, col) mean(f[[col]][f$group == g & f$optimizer == o])
n_in <- function(g) nrow(f[f$group == g & f$optimizer == opts[[1]], ])
overall <- function(o, col) mean(f[[col]][f$optimizer == o])
corp <- f[f$study == "corpus", ]
imp <- f[f$design == "imp_f1_p5", ]
lb_fail_imp <- mean(imp$fail[imp$optimizer == "nlopt-lbfgs" & imp$n == 400 & imp$lambda == 0.1])
stall <- f[f$fail & f$optimizer == "nlopt-lbfgs" & !is.na(f$fail_f) & !is.na(f$best), ]
secs <- tapply(f$seconds, f$optimizer, mean)
```

## Question

The barrier fitter `frontier_fit_ml_multiinfo(target = "determinacy")`
defaulted to NLopt's L-BFGS. On the Geiser quadratic growth model, one of the
covariance-domain paper's flagship examples, L-BFGS fails while PORT and SLSQP
converge. Which optimizer should be the default?

## Short answer

L-BFGS is the only optimizer that fails where the others succeed. Over all
`r format(sum(f$optimizer == opts[[1]]), big.mark = ",")` problems, it failed
`r pct2(overall("nlopt-lbfgs", "fail"))`% of the time, against
`r pct2(overall("port", "fail"))`% for PORT,
`r pct2(overall("nlopt-slsqp", "fail"))`% for SLSQP, and
`r pct2(overall("nlopt-lbfgs-slsqp-fallback", "fail"))`% for L-BFGS with an SLSQP
fallback. Its failures concentrate in two places. The first is populations
whose truth is improper, where the barrier optimum hugs a face
(`r pct(lb_fail_imp)`% at λ = 0.1 and N = 400). The second is the textbook
corpus cases, where it failed on `r sum(corp$fail & corp$optimizer == "nlopt-lbfgs")`
of `r sum(corp$optimizer == "nlopt-lbfgs")` fits. These are genuine stalls. A
failure is reported only when the final point fails magmaan's stationarity
audit, and the objective at the stall was above the optimum found by the other
optimizers in `r pct(mean(stall$fail_f > stall$best + 1e-6))`% of them. Where
the optimizers converge they agree. PORT, SLSQP, and the fallback reach the best
objective found in `r pct(overall("port", "certified"))`%,
`r pct(overall("nlopt-slsqp", "certified"))`%, and
`r pct(overall("nlopt-lbfgs-slsqp-fallback", "certified"))`% of problems. None
of them is slower than L-BFGS on average. The remaining failures shared by all
optimizers are in the Ernst marker-identification fits at N = 10, where the
data sit near a marker pole that no penalty can remove. There PORT is as
reliable as the others and lands on a worse local optimum least often
(`r pct(mean(f$worse[f$study == "ernst" & f$optimizer == "port"]))`% against
`r pct(mean(f$worse[f$study == "ernst" & f$optimizer == "nlopt-lbfgs"]))`% for
L-BFGS). PORT is now the barrier's default.

## Evidence

### Failure rate by optimizer

Share of fits that returned an error or failed the convergence audit (%).

```{r failures}
tab <- data.frame(Problems = groups, Fits = sapply(groups, n_in),
                  check.names = FALSE, row.names = NULL)
for (o in opts) tab[[opt_label[[o]]]] <- sapply(groups, function(g) pct(rate(g, o, "fail")))
knitr::kable(tab, align = paste0("lr", strrep("r", length(opts))))
```

### Agreement at convergence

Share of problems on which each optimizer reached the best penalized objective
found by any optimizer, within `r format(tol)` (%), and the share on which it
converged to a worse point.

```{r certified}
tab <- data.frame(Optimizer = unname(opt_label[opts]),
                  `Best objective reached` = sapply(opts, function(o) pct(overall(o, "certified"))),
                  `Converged to a worse point` = sapply(opts, function(o) pct2(overall(o, "worse"))),
                  `Mean seconds per fit` = sprintf("%.3f", secs[opts]),
                  check.names = FALSE, row.names = NULL)
knitr::kable(tab, align = "lrrr")
```

## Caveats

- The best objective is the best found by these four optimizers from one
  start each. A better optimum missed by all four would not show.
- The penalty is the latent-determinacy barrier only. The joint barrier and
  ordinary ML were not tested here.
- Default tolerances and budgets for every optimizer. The comparison is of
  the methods as magmaan configures them.
- The corpus cases need the optional textbook corpus and are skipped without
  it.

## Reproduce

```sh
just r-dev
Rscript experiments/engineering/18-barrier-optimizer/run_experiment.R --full --cores 4
quarto render experiments/engineering/18-barrier-optimizer/report.qmd
```

## Decision record

- Current default: PORT for `frontier_fit_ml_multiinfo()` and
  `frontier_fit_fiml_multiinfo()`, both targets (changed from NLopt L-BFGS on
  2026-09-25 on this evidence, which covers the determinacy target; the joint
  target has the same faces).
- Alternatives: L-BFGS, SLSQP, L-BFGS with an SLSQP fallback.
- Acceptance: no failure where another optimizer converges, the best
  objective reached at least as often as any alternative, and no material
  slowdown.
- Reopening trigger: a barrier fit that PORT fails and another optimizer
  certifies, or a change of the barrier's parametrization (for example
  log-Cholesky coordinates for the disturbance covariances, which remove the
  domain walls that defeat L-BFGS's line search).
