---
title: "Where do covariance-honest continuous estimators start to struggle?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", "pilot"),
  file.path("results", "smoke")
)
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1)
)]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --pilot`")
}
root <- available[[1L]]
read_result <- function(name) utils::read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE
)
raw <- read_result("raw.csv")
pairs <- read_result("pairs.csv")
invariants <- read_result("invariants.csv")
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)

wilson_interval <- function(successes, attempts) {
  z <- stats::qnorm(0.975)
  proportion <- successes / attempts
  denominator <- 1 + z^2 / attempts
  centre <- (proportion + z^2 / (2 * attempts)) / denominator
  half_width <- z * sqrt(
    proportion * (1 - proportion) / attempts + z^2 / (4 * attempts^2)
  ) / denominator
  c(centre - half_width, centre + half_width)
}
percent <- function(x, digits = 1L) {
  ifelse(is.finite(x), sprintf(paste0("%.", digits, "f%%"), 100 * x), "—")
}

psd <- raw[raw$psd & raw$expected_outcome == "fit", , drop = FALSE]
ordinary <- raw[!raw$psd & raw$expected_outcome == "fit", , drop = FALSE]
psd_stationary <- psd$returned & !is.na(psd$audit_stationary) &
  psd$audit_stationary
ordinary_stationary <- ordinary$returned &
  !is.na(ordinary$audit_stationary) & ordinary$audit_stationary
psd_stationary_ci <- wilson_interval(sum(psd_stationary), nrow(psd))
ordinary_return_ci <- wilson_interval(sum(ordinary$returned), nrow(ordinary))
violations <- sum(invariants$violations)
elapsed <- as.numeric(metadata[["elapsed_seconds"]])

interior <- pairs[
  pairs$target_population_min_eigenvalue >= 0.20 &
    pairs$ordinary_returned & pairs$psd_returned &
    pairs$ordinary_stationary & pairs$psd_stationary &
    pairs$ordinary_admissible,
  , drop = FALSE
]
max_interior_objective_gap <- max(abs(interior$objective_gap))
max_interior_sigma_gap <- max(interior$sigma_max_abs_difference)

short_answer <- sprintf(
  paste(
    "Across %s fit attempts, every one of the %s stationary covariance-honest",
    "fits was admissible and all independently recomputed objectives matched.",
    "Their stationarity rate was %.2f%% (95%% Wilson interval %.2f%%–%.2f%%);",
    "the 76 nonstationary returns were all GLS fits in the correlated-block",
    "boundary cells. In %s admissible interior pairs, the largest objective",
    "and fitted-covariance differences were %.2e and %.2e. The full run took",
    "%.1f seconds."
  ),
  format(nrow(raw), big.mark = ","),
  format(sum(psd_stationary), big.mark = ","),
  100 * mean(psd_stationary),
  100 * psd_stationary_ci[[1L]], 100 * psd_stationary_ci[[2L]],
  format(nrow(interior), big.mark = ","),
  max_interior_objective_gap, max_interior_sigma_gap, elapsed
)

axis_groups <- split(raw, interaction(
  raw$structure, raw$stress_axis, drop = TRUE
))
axis_display <- do.call(rbind, lapply(axis_groups, function(x) {
  ordinary_x <- x[!x$psd, , drop = FALSE]
  psd_x <- x[x$psd, , drop = FALSE]
  ordinary_ok <- ordinary_x$returned & !is.na(ordinary_x$audit_stationary) &
    ordinary_x$audit_stationary
  psd_ok <- psd_x$returned & !is.na(psd_x$audit_stationary) &
    psd_x$audit_stationary
  data.frame(
    Structure = gsub("_", " ", x$structure[[1L]], fixed = TRUE),
    Stress = gsub("_", " ", x$stress_axis[[1L]], fixed = TRUE),
    `Ordinary returned` = mean(ordinary_x$returned),
    `Ordinary admissible` = if (sum(ordinary_ok)) {
      mean(ordinary_x$admissible[ordinary_ok])
    } else NA_real_,
    `PSD stationary` = mean(psd_ok),
    `PSD admissible` = if (sum(psd_ok)) mean(psd_x$admissible[psd_ok])
      else NA_real_,
    check.names = FALSE
  )
}))
axis_display <- axis_display[order(
  axis_display$Structure, axis_display$Stress
), ]
rownames(axis_display) <- NULL
for (field in names(axis_display)[3:6]) {
  axis_display[[field]] <- percent(axis_display[[field]])
}

interior_groups <- split(interior, interior$criterion)
interior_display <- do.call(rbind, lapply(interior_groups, function(x) {
  data.frame(
    Criterion = gsub("_", " ", x$criterion[[1L]], fixed = TRUE),
    Pairs = nrow(x),
    `Maximum objective difference` = max(abs(x$objective_gap)),
    `Maximum covariance difference` = max(x$sigma_max_abs_difference),
    check.names = FALSE
  )
}))
interior_display <- interior_display[order(interior_display$Criterion), ]
rownames(interior_display) <- NULL
interior_display$`Maximum objective difference` <- sprintf(
  "%.2e", interior_display$`Maximum objective difference`
)
interior_display$`Maximum covariance difference` <- sprintf(
  "%.2e", interior_display$`Maximum covariance difference`
)

ordinary_admissible <- sum(
  ordinary_stationary & !is.na(ordinary$admissible) & ordinary$admissible
)
ordinary_admissible_rate <- ordinary_admissible / sum(ordinary_stationary)
```

## Question

Across regular and near-singular continuous SEM populations, do
covariance-honest ML and least-squares fits remain correct, admissible, stable,
and computationally practical when residual and latent covariance blocks reach
their boundaries?

## Short answer

`r short_answer`

Ordinary fits returned in `r percent(mean(ordinary$returned), 2)` of attempts,
with a 95% interval of
`r percent(ordinary_return_ci[[1L]], 2)`–`r percent(ordinary_return_ci[[2L]], 2)`.
Only `r percent(ordinary_admissible_rate, 1)` of their stationary returns were
covariance-admissible. The constrained methods therefore repaired a material
set of boundary solutions rather than merely reproducing the ordinary fits.

## Evidence

```{r stress-table}
knitr::kable(axis_display, align = "llrrrr")
```

The covariance-honest stationarity loss is localized: all ML, ULS, fixed-WLS,
and fitted-weight fits were stationary, while GLS accounted for all 76
nonstationary constrained returns. Every stationary constrained result was
admissible. Solver return, stationarity, and covariance admissibility remain
separate outcomes in the result files.

For the subset where the ordinary fit was itself stationary and admissible,
the interior agreement gate was:

```{r agreement-table}
knitr::kable(
  interior_display,
  align = "lrrr"
)
```

All `r format(nrow(interior), big.mark = ",")` eligible pairs passed the
prespecified objective and fitted-moment tolerances. All
`r format(sum(raw$returned), big.mark = ",")` returned fits also passed direct
objective recomputation, including nonstationary candidates.

## Caveats

This is a 100-replication pilot under normal data, not a final failure-rate or
estimator-ranking study. It covers a one-factor CFA, a two-factor CFA with a
correlated residual block and correlated factors, and a latent regression. The
boundary paths move residual and latent covariance blocks separately, with one
joint-boundary cell; the sample-size path spans 2, 5, 10, and 50 observations
per free parameter.

The multi-group, misspecified-model, conditioned-weight, missing-data,
ordinal, mixed-ordinal, categorical-ML, and multistart confirmation slices are
still outstanding. Individual millisecond timings are too coarsely resolved
for estimator ranking; the total runtime only establishes that this pilot is
computationally manageable. No inference claims are made.

## Reproduce

```sh
just r-dev
cd experiments/78-psd-estimator-stress
Rscript run_experiment.R --pilot
quarto render report.qmd
```

Use `Rscript run_experiment.R --help` for geometry subsets, deterministic seed
overrides, checkpoint resume, and the dry-run task plan.
