---
title: "Where do covariance-honest continuous estimators start to struggle?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", "pilot"),
  file.path("results", "smoke")
)
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1)
)]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --pilot`")
}
root <- available[[1L]]
read_result <- function(name) utils::read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE
)
raw <- read_result("raw.csv")
pairs <- read_result("pairs.csv")
invariants <- read_result("invariants.csv")
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)

wilson_interval <- function(successes, attempts) {
  z <- stats::qnorm(0.975)
  proportion <- successes / attempts
  denominator <- 1 + z^2 / attempts
  centre <- (proportion + z^2 / (2 * attempts)) / denominator
  half_width <- z * sqrt(
    proportion * (1 - proportion) / attempts + z^2 / (4 * attempts^2)
  ) / denominator
  c(centre - half_width, centre + half_width)
}
percent <- function(x, digits = 1L) {
  ifelse(is.finite(x), sprintf(paste0("%.", digits, "f%%"), 100 * x), "—")
}

psd <- raw[raw$psd & raw$expected_outcome == "fit", , drop = FALSE]
ordinary <- raw[!raw$psd & raw$expected_outcome == "fit", , drop = FALSE]
psd_lifted_stationary <- psd$returned & !is.na(psd$audit_stationary) &
  psd$audit_stationary
ordinary_driven_stationary <- ordinary$returned &
  !is.na(ordinary$audit_stationary) & ordinary$audit_stationary
psd_cone_stationary <- psd$returned & psd$geometric_checked &
  !is.na(psd$cone_stationary) & psd$cone_stationary
ordinary_cone_stationary <- ordinary$returned & ordinary$geometric_checked &
  !is.na(ordinary$cone_stationary) & ordinary$cone_stationary
psd_cone_stationary_ci <- wilson_interval(sum(psd_cone_stationary), nrow(psd))
ordinary_return_ci <- wilson_interval(sum(ordinary$returned), nrow(ordinary))
violations <- sum(invariants$violations)
elapsed <- as.numeric(metadata[["elapsed_seconds"]])

conditioning <- raw[
  raw$stress_axis == "weight_conditioning",
  , drop = FALSE
]
conditioning_groups <- split(conditioning, interaction(
  conditioning$weight_condition_target, conditioning$method, drop = TRUE
))
conditioning_display <- do.call(rbind, lapply(conditioning_groups, function(x) {
  lifted <- x$returned & !is.na(x$audit_stationary) & x$audit_stationary
  cone <- x$returned & x$geometric_checked &
    !is.na(x$cone_stationary) & x$cone_stationary
  data.frame(
    `Weight condition` = format(x$weight_condition_actual[[1L]], scientific = TRUE),
    Estimator = if (x$psd[[1L]]) "PSD WLS" else "ordinary WLS",
    Returned = percent(mean(x$returned)),
    `Driven/lifted stationary` = percent(mean(lifted)),
    `Cone stationary` = percent(mean(cone)),
    `Admissible among cone-stationary` = if (sum(cone)) {
      percent(mean(x$admissible[cone]))
    } else "—",
    check.names = FALSE
  )
}))
conditioning_display <- conditioning_display[order(
  as.numeric(conditioning_display$`Weight condition`),
  conditioning_display$Estimator
), ]
rownames(conditioning_display) <- NULL

two_group <- raw[raw$structure == "two_group_one_factor", , drop = FALSE]
misspecified <- raw[raw$misspecified, , drop = FALSE]
two_group_equality_max <- max(
  two_group$equality_violation_inf[two_group$returned]
)
misspecified_stationary <- misspecified$returned &
  misspecified$geometric_checked &
  !is.na(misspecified$cone_stationary) & misspecified$cone_stationary
misspecified_min_fmin <- min(misspecified$fmin[misspecified_stationary])

conditioning_cone_stationarity <- function(method, condition) {
  x <- conditioning[
    conditioning$method == method &
      conditioning$weight_condition_target == condition,
    , drop = FALSE
  ]
  mean(x$returned & x$geometric_checked &
         !is.na(x$cone_stationary) & x$cone_stationary)
}
conditioning_return <- function(method, condition) {
  x <- conditioning[
    conditioning$method == method &
      conditioning$weight_condition_target == condition,
    , drop = FALSE
  ]
  mean(x$returned)
}

interior <- pairs[
  pairs$target_population_min_eigenvalue >= 0.20 &
    pairs$ordinary_returned & pairs$psd_returned &
    pairs$ordinary_cone_stationary & pairs$psd_cone_stationary &
    pairs$ordinary_admissible,
  , drop = FALSE
]
max_interior_objective_gap <- max(abs(interior$objective_gap))
max_interior_sigma_gap <- max(interior$sigma_max_abs_difference)

short_answer <- sprintf(
  paste(
    "Across %s fit attempts, every one of the %s cone-stationary covariance-honest",
    "fits was admissible and every returned objective recomputed correctly.",
    "The grouped fits honored both equalities to %.2e, and all %s stationary",
    "misspecified-model fits retained nonzero discrepancy (minimum %.2e).",
    "Fixed-WLS conditioning was the clear limit: at condition numbers 1, 1e4,",
    "and 1e8, ordinary return rates were %.0f%%, %.0f%%, and %.0f%%, while PSD",
    "cone-stationarity rates were %.0f%%, %.0f%%, and %.0f%%."
  ),
  format(nrow(raw), big.mark = ","),
  format(sum(psd_cone_stationary), big.mark = ","),
  two_group_equality_max,
  format(sum(misspecified_stationary), big.mark = ","),
  misspecified_min_fmin,
  100 * conditioning_return("wls", 1),
  100 * conditioning_return("wls", 1e4),
  100 * conditioning_return("wls", 1e8),
  100 * conditioning_cone_stationarity("psd_wls", 1),
  100 * conditioning_cone_stationarity("psd_wls", 1e4),
  100 * conditioning_cone_stationarity("psd_wls", 1e8)
)

axis_groups <- split(raw, interaction(
  raw$structure, raw$stress_axis, drop = TRUE
))
axis_display <- do.call(rbind, lapply(axis_groups, function(x) {
  ordinary_x <- x[!x$psd, , drop = FALSE]
  psd_x <- x[x$psd, , drop = FALSE]
  ordinary_driven <- ordinary_x$returned &
    !is.na(ordinary_x$audit_stationary) &
    ordinary_x$audit_stationary
  ordinary_cone <- ordinary_x$returned & ordinary_x$geometric_checked &
    !is.na(ordinary_x$cone_stationary) & ordinary_x$cone_stationary
  psd_lifted <- psd_x$returned & !is.na(psd_x$audit_stationary) &
    psd_x$audit_stationary
  psd_cone <- psd_x$returned & psd_x$geometric_checked &
    !is.na(psd_x$cone_stationary) & psd_x$cone_stationary
  data.frame(
    Structure = gsub("_", " ", x$structure[[1L]], fixed = TRUE),
    Stress = gsub("_", " ", x$stress_axis[[1L]], fixed = TRUE),
    `Ordinary returned` = mean(ordinary_x$returned),
    `Ordinary driven stationary` = mean(ordinary_driven),
    `Ordinary cone stationary` = mean(ordinary_cone),
    `PSD returned` = mean(psd_x$returned),
    `PSD lifted stationary` = mean(psd_lifted),
    `PSD cone stationary` = mean(psd_cone),
    `PSD admissible` = if (sum(psd_cone)) {
      mean(psd_x$admissible[psd_cone])
    } else NA_real_,
    check.names = FALSE
  )
}))
axis_display <- axis_display[order(
  axis_display$Structure, axis_display$Stress
), ]
rownames(axis_display) <- NULL
for (field in names(axis_display)[3:9]) {
  axis_display[[field]] <- percent(axis_display[[field]])
}

interior_groups <- split(interior, interior$criterion)
interior_display <- if (length(interior_groups)) {
  do.call(rbind, lapply(interior_groups, function(x) {
    data.frame(
      Criterion = gsub("_", " ", x$criterion[[1L]], fixed = TRUE),
      Pairs = nrow(x),
      `Maximum objective difference` = max(abs(x$objective_gap)),
      `Maximum covariance difference` = max(x$sigma_max_abs_difference),
      check.names = FALSE
    )
  }))
} else {
  data.frame(
    Criterion = character(), Pairs = integer(),
    `Maximum objective difference` = numeric(),
    `Maximum covariance difference` = numeric(),
    check.names = FALSE
  )
}
interior_display <- interior_display[order(interior_display$Criterion), ,
                                     drop = FALSE]
rownames(interior_display) <- NULL
interior_display$`Maximum objective difference` <- sprintf(
  "%.2e", interior_display$`Maximum objective difference`
)
interior_display$`Maximum covariance difference` <- sprintf(
  "%.2e", interior_display$`Maximum covariance difference`
)

ordinary_admissible <- sum(
  ordinary_cone_stationary & !is.na(ordinary$admissible) &
    ordinary$admissible
)
ordinary_admissible_rate <- ordinary_admissible /
  sum(ordinary_cone_stationary)
psd_nonstationary <- psd[
  psd$returned & !is.na(psd$cone_stationary) & !psd$cone_stationary,
  , drop = FALSE
]
psd_lifted_only <- sum(
  psd$returned & psd_lifted_stationary & !psd_cone_stationary
)
psd_cone_only <- sum(
  psd$returned & !psd_lifted_stationary & psd_cone_stationary
)
psd_cone_not_ambient <- sum(
  psd_cone_stationary & !psd$ambient_stationary
)
nonstationary_by_method <- sort(table(psd_nonstationary$method), decreasing = TRUE)
nonstationary_description <- paste(
  paste0(names(nonstationary_by_method), " (", as.integer(nonstationary_by_method), ")"),
  collapse = " and "
)
```

## Question

Across regular, grouped, misspecified, and near-singular continuous SEM
populations, do covariance-honest ML and least-squares fits remain correct,
admissible, stable, and computationally practical as covariance blocks and
fixed weights become ill-conditioned?

## Short answer

`r short_answer`

Ordinary fits returned in `r percent(mean(ordinary$returned), 2)` of attempts,
with a 95% interval of
`r percent(ordinary_return_ci[[1L]], 2)`–`r percent(ordinary_return_ci[[2L]], 2)`.
Only `r percent(ordinary_admissible_rate, 1)` of their stationary returns were
covariance-admissible. The constrained methods therefore repaired a material
set of boundary solutions rather than merely reproducing the ordinary fits.
The overall covariance-honest cone-stationarity rate was
`r percent(mean(psd_cone_stationary), 2)` (95% Wilson interval
`r percent(psd_cone_stationary_ci[[1L]], 2)`–`r percent(psd_cone_stationary_ci[[2L]], 2)`).

## Evidence

```{r stress-table}
knitr::kable(axis_display, align = "llrrrrrrr")
```

The covariance-honest stationarity loss is localized: the
`r format(nrow(psd_nonstationary), big.mark = ",")` returned candidates that
failed the cone audit were `r nonstationary_description`. Every cone-stationary
constrained result was admissible. Solver return, the original driven/lifted
audit, common-coordinate cone stationarity, and covariance admissibility remain
separate outcomes in the result files.

The distinction is operational rather than cosmetic. The PSD normal cone made
`r format(psd_cone_not_ambient, big.mark = ",")` covariance-honest returns
stationary that would fail an equality/bound-only ambient test. Conversely,
`r format(psd_lifted_only, big.mark = ",")` returns passed the original
lifted-coordinate audit but failed the common cone audit; `r format(psd_cone_only,
big.mark = ",")` showed the reverse. All geometric gradients were finite and
both normal projections completed for every returned fit, so these differences
are stationarity results rather than diagnostic-solver failures.

For the subset where the ordinary fit was itself stationary and admissible,
the interior agreement gate was:

```{r agreement-table}
knitr::kable(
  interior_display,
  align = "lrrr"
)
```

All `r format(nrow(interior), big.mark = ",")` eligible pairs passed the
prespecified objective and fitted-moment tolerances. All
`r format(sum(raw$returned), big.mark = ",")` returned fits also passed direct
objective recomputation, including nonstationary candidates.

The targeted fixed-WLS slice separates return from stationarity sharply:

```{r conditioning-table}
knitr::kable(conditioning_display, align = "llrrrr")
```

All recorded weights had the requested condition number and full effective
rank. Thus the deterioration at `1e8` is optimizer/geometry behavior, not a
silently rank-truncated weight. The two-group panel contributed
`r format(nrow(two_group), big.mark = ",")` attempts at the anchor and residual
boundary; `r format(sum(two_group$returned), big.mark = ",")` returned, with a
maximum fitted equality violation of `r sprintf("%.2e", two_group_equality_max)`.
The misspecified panel returned and was stationary in all
`r format(sum(misspecified_stationary), big.mark = ",")` attempts, so ordinary
and constrained agreement is not confined to exact-fit populations.

## Caveats

This is a 100-replication pilot under normal data, not a final failure-rate or
estimator-ranking study. Its 29 cells cover a one-factor CFA, a two-factor CFA
with correlated residual and factor blocks, a latent regression, a deliberately
misspecified two-factor CFA, and a two-group CFA with a shared residual variance
plus one general linear equality. The boundary paths move residual and latent
covariance blocks separately, with one joint-boundary cell; the sample-size
path spans 2, 5, 10, and 50 observations per free parameter.

Cone stationarity is measured from the full model differential with the
product Frobenius metric on the model matrices. This makes ordinary and
covariance-honest fits directly comparable and is unchanged by orthogonal
changes of covariance-block basis, but it is not invariant to arbitrary
rescaling of observed variables. Its verdict uses the metric-dual L2 residual;
the coordinatewise infinity residual remains available for inspection but does
not drive the verdict. The original optimizer-coordinate audit is retained
separately for lavaan-compatible convergence inspection.

Missing-data, ordinal, mixed-ordinal, and categorical-ML paths currently have
deterministic smoke anchors; calibrated stress pilots and multistart
confirmation slices are still outstanding. The conditioning result is specific
to this diagonal fixed-weight spectrum and the current optimizers; it does not
establish a universal condition-number cutoff. Individual millisecond timings
are too coarsely resolved for estimator ranking; the total runtime of
`r sprintf("%.1f", elapsed)` seconds only establishes that this pilot is
computationally manageable. No inference claims are made.

## Reproduce

```sh
just r-dev
cd experiments/78-psd-estimator-stress
Rscript run_experiment.R --pilot
quarto render report.qmd
```

Use `Rscript run_experiment.R --help` for geometry subsets, deterministic seed
overrides, checkpoint resume, and the dry-run task plan.
