---
title: "Do the covariance-honest estimators survive a common stress harness?"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", "smoke")
)
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1)
)]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) utils::read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE
)
raw <- read_result("raw.csv")
summary <- read_result("summary.csv")
invariants <- read_result("invariants.csv")
metadata_frame <- read_result("metadata.csv")
metadata <- setNames(metadata_frame$value, metadata_frame$key)

violations <- sum(invariants$violations)
returned <- raw[raw$expected_outcome == "fit", ]
psd <- returned[returned$psd & returned$returned, ]
objective_errors <- returned$objective_abs_error[
  returned$returned & is.finite(returned$objective_abs_error)
]
max_objective_error <- if (length(objective_errors)) max(objective_errors) else NA_real_
short_answer <- sprintf(
  paste(
    "The smoke harness completed %d fit attempts across %d estimator families.",
    "%d hard invariant violations were found; %d/%d returned PSD fits were",
    "covariance-admissible, and the largest independently recomputed objective",
    "difference was %.2e. This validates the harness, not comparative estimator",
    "performance."
  ),
  nrow(raw), length(unique(raw$family)), violations,
  sum(psd$admissible, na.rm = TRUE), nrow(psd), max_objective_error
)

display <- summary[, c(
  "family", "fits", "returned_rate", "audit_stationary_rate",
  "psd_admissible_rate", "objective_pass_rate",
  "input_preservation_rate", "median_elapsed_ms"
)]
names(display) <- c(
  "Family", "Attempts", "Returned", "Audit-stationary",
  "PSD admissible", "Objective check", "Input preserved",
  "Median time (ms)"
)
display$Family <- gsub("_", " ", display$Family, fixed = TRUE)
for (field in c(
  "Returned", "Audit-stationary", "PSD admissible", "Objective check",
  "Input preserved"
)) {
  display[[field]] <- ifelse(
    is.finite(display[[field]]),
    sprintf("%.0f%%", 100 * display[[field]]),
    "—"
  )
}
display$`Median time (ms)` <- round(display$`Median time (ms)`, 2L)
```

## Question

Can one common, checkpointed validation harness exercise the covariance-honest
continuous, missing-data, ordinal, mixed, and categorical-ML estimators while
distinguishing optimizer failures, covariance failures, input-domain
rejections, and independently detected criterion errors?

## Short answer

`r short_answer`

## Evidence

```{r summary-table}
knitr::kable(display, align = "lrrrrrrr")
```

The non-positive-definite categorical-ML sentinel is an expected input-domain
rejection, not a failed fit. Missing-data two-stage fits additionally require
exact preservation of their shared first-stage object. The raw paired results,
constraint diagnostics, component eigenvalues, and timings remain in
`results/smoke/` rather than being expanded into the presentation table.

## Caveats

This run uses two deterministic replications per smoke cell, a four-indicator
one-factor model, one regular geometry, and one near-boundary residual-variance
geometry. It establishes that the design, independent calculations,
checkpointing, and reporting work together. It cannot estimate failure rates,
basin frequencies, boundary risk, or scaling costs. Mean-plus-multi-group,
correlated covariance blocks, sparse thresholds, conditioned weights, and the
larger missingness grid belong to the pilot expansion.

No standard errors, tests, confidence intervals, or model-selection claims are
part of this experiment.

## Reproduce

```sh
just r-dev
cd experiments/78-psd-estimator-stress
Rscript run_experiment.R --smoke
quarto render report.qmd
```

Use `Rscript run_experiment.R --help` for family/geometry subsets, deterministic
seed overrides, checkpoint resume, and the dry-run task plan.
