---
title: "Pairwise GLS Efficiency: Σ-only vs Γ_NT^pw Weight"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path), dirname(getwd())))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates,
                                                "summary.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
summary_path  <- file.path(results_dir, "summary.csv")
metadata_path <- file.path(results_dir, "metadata.csv")
if (!file.exists(summary_path)) {
  stop("Missing results/summary.csv. Run: Rscript run_experiment.R")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("The report requires ggplot2.")
}
library(ggplot2)
summary_df <- utils::read.csv(summary_path, check.names = FALSE)
summary_df$miss_label <- sprintf("%g%%", 100 * summary_df$miss_rate)
metadata <- if (file.exists(metadata_path))
  utils::read.csv(metadata_path, check.names = FALSE) else NULL
meta_lookup <- function(key, default = "") {
  if (is.null(metadata)) return(default)
  row <- metadata[metadata$key == key, , drop = FALSE]
  if (nrow(row) == 0L) default else as.character(row$value[[1L]])
}
n_reps     <- meta_lookup("reps", "?")
seed_base  <- meta_lookup("seed_base", "?")
```

## Question

Two pairwise GLS variants are available for fitting an SEM to incomplete
continuous data: the literature default with the Σ-only weight
`Γ_NT(Σ̂_pw)⁻¹` and the asymptotically efficient variant with the
`Γ_NT^pw⁻¹` weight. Does the asymptotic gain materialise at finite
samples and realistic missing rates?

## Short Answer

Asymptotic efficiency is **not the story**: under both Savalei-Bentler
2005 MCAR and the SB-2005 MAR design, the two weights produce nearly
identical fits at moderate-to-large n. Trace-MSE ratios stay in
[1.00, 1.10] for every cell except one (n = 200, MCAR, 30%) where the
Σ-only weight is occasionally catastrophically ill-conditioned.

**The interesting effect is that occasional catastrophe under MCAR, and
its absence under MAR**:

- Under MCAR-with-intact-x1, `Γ_NT(Ŝ_pw)` carries condition numbers ~2×
  to ~5× larger than `Γ_NT^pw`. The MCAR-with-intact Hadamard factor
  `Ω` has diagonal entries `1/(1−rate)` that pump up variance components
  and regularise the smallest eigendirections of `Γ_NT(Ŝ_pw)`. This is
  a happenstance of the missingness geometry, not a generic property.
- Under SB-2005 MAR, the conditioning gap collapses (ratio ≤ 1.2× across
  all cells). The 3-rule MAR design produces non-uniform Ω entries that
  don't preferentially regularise the noisy directions, so the two
  weights are essentially equivalent.

So `fit_gls_pairwise` is **not** a generally better-conditioned fit than
`fit_gls(Ŝ_pw)`. Its only practical advantage is under MCAR-with-intact
at small n and heavy missingness, where it avoids occasional blow-ups of
the literature path. Otherwise the two are interchangeable.

Complete-data fits agree to floating-point round-off, as expected.

## Evidence

```{r evidence-table}
tbl <- summary_df[summary_df$miss_rate > 0,
                  c("n", "miss_rate", "mechanism", "miss_label",
                    "trace_mse_sigma", "trace_mse_gamma", "eff_ratio_mse",
                    "median_cond_sigma", "median_cond_gamma")]
tbl$cond_ratio <- tbl$median_cond_sigma / tbl$median_cond_gamma
tbl <- tbl[order(tbl$n, tbl$miss_rate, tbl$mechanism), ]
tbl$trace_mse_sigma   <- round(tbl$trace_mse_sigma, 3)
tbl$trace_mse_gamma   <- round(tbl$trace_mse_gamma, 3)
tbl$eff_ratio_mse     <- round(tbl$eff_ratio_mse, 3)
tbl$median_cond_sigma <- round(tbl$median_cond_sigma)
tbl$median_cond_gamma <- round(tbl$median_cond_gamma)
tbl$cond_ratio        <- round(tbl$cond_ratio, 2)
tbl$mechanism         <- toupper(tbl$mechanism)
tbl$miss_rate         <- NULL
tbl <- tbl[, c("n", "mechanism", "miss_label",
               "trace_mse_sigma", "trace_mse_gamma", "eff_ratio_mse",
               "median_cond_sigma", "median_cond_gamma", "cond_ratio")]
names(tbl) <- c("n", "Mechanism", "Missing %",
                "Trace MSE Σ-only", "Trace MSE Γ_NT^pw", "MSE Ratio",
                "Median cond Σ", "Median cond Γ_NT^pw", "Cond Ratio")
knitr::kable(
  tbl,
  caption = paste(
    "Trace of per-parameter empirical MSE and median weight-matrix condition",
    "number across replications. MSE Ratio = trace MSE Σ-only ÷ trace MSE",
    "Γ_NT^pw; values > 1 mean Γ_NT^pw is more efficient. Cond Ratio is the",
    "matching ratio of median condition numbers; values > 1 mean Γ_NT^pw is",
    "better-conditioned. rate = 0 cells (degeneracy reference) are omitted",
    "from the table; see the metadata file for them."
  )
)
```

```{r evidence-plot, fig.width=7.5, fig.height=4.2}
plot_df <- summary_df[summary_df$miss_rate > 0, , drop = FALSE]
plot_df$mechanism <- toupper(plot_df$mechanism)
plot_df$n_label <- factor(paste0("n = ", plot_df$n),
                          levels = paste0("n = ", sort(unique(plot_df$n))))
plot_df$cond_ratio <- plot_df$median_cond_sigma / plot_df$median_cond_gamma
ggplot(plot_df,
       aes(x = miss_rate, y = cond_ratio,
           colour = mechanism, group = mechanism)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.6) +
  facet_wrap(~ n_label) +
  scale_x_continuous(labels = scales::percent) +
  scale_colour_manual(values = c(MCAR = "#1f78b4", MAR = "#e31a1c")) +
  labs(
    x = "Missing rate",
    y = "Conditioning ratio  cond(Γ_NT(Ŝ_pw)) / cond(Γ_NT^pw)",
    colour = "Mechanism"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "right")
```

Dashed line is the no-effect reference. Under MCAR the conditioning
ratio climbs to 2×–5× — `Γ_NT^pw` is meaningfully better conditioned.
Under MAR it stays in [1.05, 1.20] across every cell — the regularising
effect is gone.

## Caveats

- One model only — 1-factor / 9 indicators / loadings ≈ 0.7. The
  efficiency gap will scale with df (here 27) and with how strongly
  `Γ_NT^pw` differs from `Γ_NT(Σ̂_pw)`, which itself depends on the
  missing-pattern structure.
- Two missing mechanisms: SB-2005 MCAR (intact = x1) and SB-2005 MAR
  (predictors = x1, x2). Other realistic patterns — planned-monotone,
  attrition, multi-anchor MAR — could plausibly tip the conditioning
  ratio further in either direction. The reported gap is for these
  two designs specifically.
- Skipped: standard errors, chi-square tests, FIML / complete-data ML
  reference rows. The trace metric is a sufficient statistic for
  asymptotic efficiency under the correct-model assumption used here;
  finite-sample inference quality is a separate question.
- `r n_reps` replications per cell (seed base `r seed_base`). The
  results for the high-missing / small-n cells carry visible Monte
  Carlo noise; bump `--reps` if a particular cell deserves more weight.
- The n = 200 / rate = 30% MCAR cell's headline MSE ratio (≈ 3.7) is
  dominated by a handful of catastrophic Σ-only fits (one rep with
  summed squared error > 500 out of 200 reps; median is ~1.4). The
  blow-ups track the conditioning numbers exactly. The same cell under
  MAR gives a ratio of 1.08 with no comparable outliers.
- Singularity does propagate algebraically: `Ŝ_pw` singular implies
  `Γ_NT(Ŝ_pw)` singular implies `Γ_NT^pw` singular. The finite-sample
  conditioning advantage under MCAR is a happenstance of the Hadamard
  factor Ω, not a generic theorem. The MAR cells in this experiment
  confirm that the advantage is mechanism-specific, not estimator-
  generic.
- The Σ-only-weight fit is reached by passing `samp.S = Ŝ_pw` into the
  existing `magmaan_core$estimate_gls`; the Γ_NT^pw fit is
  `magmaan_core$estimate_gls_pairwise`. Both reduce to the same fit on
  complete data — the rate-0 rows in the table verify this at machine
  precision.

## Reproduce

```bash
cd experiments/08-pairwise-gls-efficiency
Rscript run_experiment.R          # full run (≈ minutes)
Rscript run_experiment.R --smoke  # quick check
quarto render report.qmd
```
