---
title: "FIML goodness-of-fit under non-normality: FMG/pEBA vs the MLR default"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
res_dir <- "results"
sum_path <- file.path(res_dir, "summary_rejection.csv")
if (!file.exists(sum_path)) {
  stop("No results yet. Run:\n",
       "  Rscript run_experiment.R --smoke         # single hardest cell\n",
       "  Rscript run_experiment.R --lavaan-parity # full grid + oracle\n")
}
S  <- read.csv(sum_path, stringsAsFactors = FALSE)
NC <- if (file.exists(file.path(res_dir, "noncentrality.csv")))
  read.csv(file.path(res_dir, "noncentrality.csv"), stringsAsFactors = FALSE) else NULL
P  <- if (file.exists(file.path(res_dir, "parity.csv")))
  read.csv(file.path(res_dir, "parity.csv"), stringsAsFactors = FALSE) else NULL
meta <- if (file.exists(file.path(res_dir, "metadata.csv")))
  read.csv(file.path(res_dir, "metadata.csv"), stringsAsFactors = FALSE) else NULL
reps <- if (!is.null(meta)) meta$value[meta$key == "reps"] else "?"
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)

gof <- S[S$outcome == "gof", ]
# Clean regime for the calibration claim: MCAR (FIML consistent), highest rate.
mcar_hi <- {
  r <- if ("MCAR" %in% gof$mech) max(gof$rate[gof$mech == "MCAR"]) else NA
  gof[gof$truth == "h0" & gof$mech == "MCAR" & gof$rate == r, ]
}
sev <- mcar_hi[mcar_hi$dist == "vm2", ]
mlr_sev <- sev$reject[sev$method == "MLR"]
all_sev <- sev$reject[sev$method == "all"]
ss_sev  <- sev$reject[sev$method == "SS"]
```

## Question

The dominant default for incomplete continuous SEM is lavaan's
`estimator = "MLR", missing = "ml"` — the Yuan-Bentler mean-scaled FIML test, a
first-moment-only correction. Under non-normality the test statistic's reference
law (a weighted sum of $\chi^2_1$) spreads its eigenvalues, so a mean-only
scaling should miscalibrate. Does the Foldnes-Moss-Gronneberg eigenvalue family
(scaled-shifted, EBA/pEBA, penalized-all, the full mixture) control the
goodness-of-fit Type-I rate where MLR does not, and does that survive missing
data?

## Short answer

```{r short, results="asis"}
if (length(mlr_sev) && length(all_sev)) {
  cat(sprintf(
    paste0("Under **MCAR** (where normal-theory FIML stays consistent), yes. At ",
           "severe Vale-Maurelli non-normality the MLR default rejects a true ",
           "model **%s** of the time at the .05 level, versus **%s** for the full ",
           "FMG mixture and **%s** for scaled-shifted — both essentially nominal. ",
           "The pattern holds across all non-normal families (%s reps).\n\n"),
    fmt(mlr_sev), fmt(all_sev), fmt(ss_sev), reps))
  cat(paste0("Under **MAR + non-normality the story is different and important**: ",
             "normal-theory FIML is itself biased there (it imputes missing values ",
             "with linear conditional means), so the statistic is non-central even ",
             "under a true model and *no* reference-law correction restores level. ",
             "The diagnostic below shows this directly."))
} else {
  cat("Run the full grid (`--lavaan-parity`) to populate this section.\n")
}
```

## Evidence

```{r tab-mcar}
if (nrow(mcar_hi)) {
  ord <- c("naive","MLR","SB","SS","SF","EBA4","pEBA4","pall","pOLS","all")
  w <- reshape(mcar_hi[, c("dist","method","reject")], idvar = "dist",
               timevar = "method", direction = "wide")
  names(w) <- sub("^reject\\.", "", names(w))
  w <- w[, c("dist", intersect(ord, names(w)))]
  num <- vapply(w, is.numeric, logical(1)); w[num] <- lapply(w[num], fmt)
  knitr::kable(w, align = "l", caption = sprintf(
    "GOF Type-I under MCAR %s (nominal .05). FIML is consistent here, so this isolates the reference-law correction.",
    fmt(max(mcar_hi$rate), 2)))
}
```

```{r tab-ncp}
if (!is.null(NC) && nrow(NC)) {
  n <- NC[order(NC$mech, NC$dist, NC$rate),
          c("dist","mech","rate","mean_base","mean_trace","ncp_hat")]
  n[c("mean_base","mean_trace","ncp_hat")] <-
    lapply(n[c("mean_base","mean_trace","ncp_hat")], fmt, d = 2)
  knitr::kable(n, align = "l", caption = paste(
    "FIML-bias diagnostic (truth = h0). With a consistent estimator E[base stat]",
    "= trace(UGamma), so ncp_hat = mean(base) - mean(trace) is ~0 under MCAR and",
    "grows under MAR + non-normality -- the bias no test correction can fix."))
}
```

```{r tab-oracle}
if (!is.null(P) && nrow(P)) {
  ag <- aggregate(abs_diff ~ metric, P, function(x) max(x, na.rm = TRUE))
  names(ag) <- c("Quantity", "Max |magmaan - lavaan|")
  ag[[2]] <- formatC(ag[[2]], format = "e", digits = 2)
  knitr::kable(ag, align = "l", caption =
    "Oracle: magmaan vs lavaan MLR (estimator = MLR, missing = ml, defaults).")
}
```

Full per-cell rejection (all truths, mechanisms, rates, methods; goodness-of-fit
and nested) is in `results/summary_rejection.csv`; per-replicate p-values, base
statistics, and UGamma spectra are in `results/raw/` (the latter in
`spectra.rds`) so any further test or size-adjustment is recomputable without
refitting.

## Caveats

- **Two regimes.** The MCAR table is the clean calibration claim. The MAR cells
  are reported to bound the method: there the over-rejection is FIML's bias, not
  the test's, as the diagnostic table shows.
- **Power** (in the full CSV) is not size-adjusted; MLR's higher raw power is
  partly its over-rejection leaking through.
- The MLR column is magmaan's `fiml_robust_mlr`, matched to lavaan's actual MLR
  defaults (see oracle). lavaan offers no mean-and-variance test under FIML, so
  `SS` is a magmaan-only construction.

## Reproduce

```sh
Rscript run_experiment.R --smoke                       # single hardest cell
Rscript run_experiment.R --lavaan-parity               # full grid + oracle
Rscript run_experiment.R --cells truth=h0,dist=vm2,mech=MCAR,rate=0.3 --reps 1000
```
