---
title: "FIML goodness-of-fit under non-normality: FMG/pEBA vs the MLR default"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
res_dir <- "results"
sum_path <- file.path(res_dir, "summary_rejection.csv")
if (!file.exists(sum_path)) {
  stop("No results yet. Run:\n",
       "  Rscript run_experiment.R --smoke         # single hardest cell\n",
       "  Rscript run_experiment.R --lavaan-parity # full grid + oracle\n")
}
S  <- read.csv(sum_path, stringsAsFactors = FALSE)
P  <- if (file.exists(file.path(res_dir, "parity.csv")))
  read.csv(file.path(res_dir, "parity.csv"), stringsAsFactors = FALSE) else NULL
meta <- if (file.exists(file.path(res_dir, "metadata.csv")))
  read.csv(file.path(res_dir, "metadata.csv"), stringsAsFactors = FALSE) else NULL
reps <- if (!is.null(meta)) meta$value[meta$key == "reps"] else "?"
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)

gof <- S[S$outcome == "gof", ]
# Clean regime for the calibration claim: MCAR (FIML consistent), highest rate.
mcar_hi <- {
  r <- if ("MCAR" %in% gof$mech) max(gof$rate[gof$mech == "MCAR"]) else NA
  gof[gof$truth == "h0" & gof$mech == "MCAR" & gof$rate == r, ]
}
sev <- mcar_hi[mcar_hi$dist == "vm2", ]
mlr_sev <- sev$reject[sev$method == "MLR"]
yb_exact_sev <- sev$reject[sev$method == "YB_exact"]
yb_exact_struct_sev <- sev$reject[sev$method == "YB_exact_structured"]
all_sev <- sev$reject[sev$method == "all"]
all_struct_sev <- sev$reject[sev$method == "all_structured"]
ss_sev  <- sev$reject[sev$method == "SS"]
ss_struct_sev  <- sev$reject[sev$method == "SS_structured"]
```

## Question

The dominant default for incomplete continuous SEM is lavaan's
`estimator = "MLR", missing = "ml"` — the Mplus variant of the Yuan-Bentler
mean-scaled FIML test. Under non-normality the test statistic's reference law
(a weighted sum of $\chi^2_1$) spreads its eigenvalues, so two questions are
separable: does MLR's finite-sample trace approximation match exact
saturated-space mean scaling, and does the Foldnes-Moss-Gronneberg eigenvalue
family (scaled-shifted, EBA/pEBA, penalized-all, the full mixture) control the
goodness-of-fit Type-I rate where MLR does not?

## Short answer

```{r short, results="asis"}
if (length(mlr_sev) && length(yb_exact_sev) &&
    length(yb_exact_struct_sev) && length(all_sev) &&
    length(all_struct_sev) && length(ss_sev) && length(ss_struct_sev)) {
  cat(sprintf(
    paste0("Under **MCAR** (where normal-theory FIML stays consistent), yes. At ",
           "severe Vale-Maurelli non-normality the MLR default rejects a true ",
           "model **%s** of the time at the .05 level. Exact UΓ mean scaling ",
           "(`YB_exact`, the same p-values as the FMG `SB` row) rejects **%s** ",
           "(**%s** with structured H1 information). The full FMG mixture rejects ",
           "**%s** (**%s** structured), and scaled-shifted rejects **%s** ",
           "(**%s** structured). ",
           "The pattern holds across all non-normal families (%s reps).\n\n"),
    fmt(mlr_sev), fmt(yb_exact_sev), fmt(yb_exact_struct_sev),
    fmt(all_sev), fmt(all_struct_sev), fmt(ss_sev), fmt(ss_struct_sev), reps))
  cat(paste0("Under this study's **MAR** the same conclusion holds. The ",
             "Savalei-Bentler mechanism here deletes indicators by coarse *sign* ",
             "rules on two always-observed indicators, which is mild enough that ",
             "FIML stays effectively unbiased: MAR rejection tracks MCAR for every ",
             "method at a matched rate, and the well-calibrated tests stay at level ",
             "under both. That is a property of *this weak* MAR, not of MAR in ",
             "general. A stronger mechanism (steep continuous selection, or a ",
             "genuinely nonlinear missing-given-observed mean) would bias FIML ",
             "itself and leave a residual no reference-law correction can remove; ",
             "this design does not probe that regime."))
} else {
  cat("Run the full grid (`--lavaan-parity`) to populate this section.\n")
}
```

## The model

One structural model is run throughout: a single common factor with six
indicators $x_1,\dots,x_6$ in two groups ($n = 500$ and $350$). The population is
**scalar-invariant** -- equal loadings and equal intercepts across groups -- with
a group-2 factor-variance and latent-mean shift, so the fitted models are
correctly specified under the null. Each replicate is fit at two nested
invariance levels, **configural** (loadings free across groups) and **metric**
(loadings tied). We report two likelihood-ratio tests, and the **metric model is
the null $H_0$ in both**; only the alternative differs:

- **goodness of fit** -- metric vs the *saturated* model ($\mathrm{df}=23$): does
  the one-factor metric model fit the data at all? This is the ordinary test of
  exact fit (the "model chi-square");
- **nested invariance** -- *configural* vs metric ($\mathrm{df}=5$): the
  chi-square-difference test of the five equal-loading constraints.

Both are likelihood-ratio tests of one model nested in a larger one (metric in
the saturated; metric in configural); the unqualified term "goodness of fit"
means the first. Power is created by two local misspecifications, one per test:
a residual covariance between $x_5$ and $x_6$ (0.13) that the one-factor model
omits (powers the goodness-of-fit test), and a group-2 loading inflated
$\times 1.35$ on one indicator (loading non-invariance; powers the nested test).
Non-normality (the VM/IG/PL families) and Savalei-Bentler MCAR/MAR missingness
act on the generated indicators; the factor model is otherwise unchanged.

## What is tested

Each fit maximizes the observed-data normal (FIML) log-likelihood over the
saturated moment vector $\eta = (\mu, \operatorname{vech}\Sigma)$ with score
$h_i(\eta)$. The saturated estimator $\hat\eta$ has the **sandwich covariance**
(the "$\Gamma$ matrix")
$$
  \Gamma_{\mathrm{mis}} = H^{-1} J H^{-1},
  \qquad
  H = -\mathbb{E}\!\left[\partial h_i/\partial\eta'\right],
  \quad
  J = \mathbb{E}\!\left[h_i h_i'\right],
$$
the missing-data analogue of Browne's ADF moment covariance. If the working
normal model is correct, information equality gives $J = H$, so
$\Gamma_{\mathrm{mis}} = H^{-1}$; under non-normality $J$ carries the third- and
fourth-moment information and $\Gamma_{\mathrm{mis}} \ne H^{-1}$.

Take the goodness-of-fit test first. The metric SEM implies $\eta(\theta)$ with
Jacobian $\Delta = \partial\eta(\theta)/\partial\theta'$. For a curvature $V$ the
residual operator is $U(V) = V - V\Delta(\Delta'V\Delta)^{-1}\Delta'V$, and the
FIML likelihood-ratio against the *saturated* fit,
$T = 2(\hat\ell_{\mathrm{sat}} - \hat\ell_{\mathrm{metric}})$, has the limit
$$
  T \;\Rightarrow\; \sum_{j=1}^{\mathrm{df}} \lambda_j\, \chi^2_{1j},
  \qquad
  \lambda_j = \operatorname{eig}_j\!\left\{ U(V)\,\Gamma_{\mathrm{mis}} \right\}.
$$
When the normal model is correct $\Gamma_{\mathrm{mis}} = V^{-1}$,
$U(V)\Gamma_{\mathrm{mis}}$ is a rank-$\mathrm{df}$ projector, every
$\lambda_j = 1$, and $T \Rightarrow \chi^2_{\mathrm{df}}$ -- the only case in
which the unscaled LRT is pivotal. The robust tests differ only in how they
reduce the spectrum $\{\lambda_j\}$:

- **MLR / SB** -- mean scaling, $c = \tfrac{1}{\mathrm{df}}\sum_j\lambda_j =
  \tfrac{1}{\mathrm{df}}\operatorname{tr}\{U(V)\Gamma_{\mathrm{mis}}\}$, refer
  $T/c$ to $\chi^2_{\mathrm{df}}$;
- **SS** -- match the mean *and* the variance (scaled-and-shifted);
- **FMG** (`EBA`/`pEBA`, `all`) -- keep the whole spectrum (Imhof / exact
  mixture).

A mean-only $c$ is exact only when the $\lambda_j$ are equal; non-normality
disperses them, so MLR must over-reject.

The nested configural-vs-metric test ($\mathrm{df}=5$) is the same construction
applied to the difference of the two models' residual operators (Satorra 2000);
the spectrum reduction above carries over verbatim, and its results are in the
`nested` rows of `results/summary_rejection.csv`.

**Structured ($V_S$) vs unstructured ($V_{US}$).** $V$ is the curvature
$U(V)$ is built from, and the two test variants differ only in where it is
evaluated:
$$
  V_{US} = H(\hat\eta) \quad\text{(saturated fit)},
  \qquad
  V_{S} = H(\eta(\hat\theta)) \quad\text{(model-implied moments)}.
$$
The default columns use $V_{US}$; the `_structured` columns use $V_S$. They
share a probability limit under the correct MCAR null -- so here they agree up to
finite-sample terms -- and separate only under misspecification or strong
non-normal MAR. Full derivations:
`docs/research/notes/non_normal_fiml_mlr_information`.

## Evidence

```{r tab-mcar}
if (nrow(mcar_hi)) {
  ord <- c("naive","MLR","YB_exact","YB_exact_structured","SS",
           "SS_structured","SF","EBA4","pEBA4","pEBA4_structured","pall",
           "pOLS","all","all_structured")
  dist_ord <- c("norm","vm1","vm2","ig1","ig2","pl1","pl2")
  w <- reshape(mcar_hi[, c("dist","method","reject")], idvar = "method",
               timevar = "dist", direction = "wide")
  names(w) <- sub("^reject\\.", "", names(w))
  w <- w[match(ord, w$method), c("method", intersect(dist_ord, names(w)))]
  w <- w[!is.na(w$method), ]
  names(w)[1] <- "Method"
  w$Method[w$Method == "YB_exact"] <- "YB_exact (= SB)"
  nn <- c("vm1", "vm2", "ig1", "ig2", "pl1", "pl2")
  w[["Non-normal mean"]] <-
    rowMeans(w[, intersect(nn, names(w)), drop = FALSE], na.rm = TRUE)
  pct <- function(x) formatC(100 * x, format = "f", digits = 1)
  num <- vapply(w, is.numeric, logical(1)); w[num] <- lapply(w[num], pct)
  knitr::kable(w, align = "l", row.names = FALSE, caption = sprintf(
    paste("GOF Type-I error (%%) under MCAR %s%%, nominal 5%%; rows are tests,",
          "columns the data families (norm; VM/IG/PL at moderate `1` and severe",
          "`2` skew/kurtosis). The final column averages the six non-normal",
          "families. SB is identical to YB_exact by construction and omitted.",
          "FIML is consistent here, so this isolates the reference-law correction."),
    formatC(100 * max(mcar_hi$rate), format = "d")))
}
```

```{r tab-power}
PW <- if (file.exists(file.path(res_dir, "power_adjusted.csv")))
  read.csv(file.path(res_dir, "power_adjusted.csv"), stringsAsFactors = FALSE) else NULL
if (!is.null(PW)) {
  nn <- c("vm1", "vm2", "ig1", "ig2", "pl1", "pl2")
  ord <- c("naive", "MLR", "YB_exact", "SS", "SF", "EBA4", "pEBA4",
           "pall", "pOLS", "all")
  ti <- gof[gof$truth == "h0" & gof$mech == "MCAR" &
            gof$rate == max(gof$rate[gof$mech == "MCAR"]) & gof$dist %in% nn, ]
  pw <- PW[PW$outcome == "gof" & PW$mech == "MCAR" &
           PW$rate == max(PW$rate[PW$mech == "MCAR"]) & PW$dist %in% nn, ]
  am <- function(df, col) vapply(ord, function(m)
    mean(df[[col]][df$method == m]), numeric(1))
  pct <- function(x) formatC(100 * x, format = "f", digits = 1)
  labs <- ifelse(ord == "YB_exact", "YB_exact (= SB)", ord)
  tab <- data.frame(Method = labs,
    `Type-I` = pct(am(ti, "reject")),
    `Power` = pct(am(pw, "power")),
    `Size-adjusted power` = pct(am(pw, "power_adj")),
    check.names = FALSE)
  knitr::kable(tab, align = "l", row.names = FALSE, caption = paste(
    "GOF detection (%) at the severe cell (MCAR 30%, mean over the 6 non-normal",
    "families). Ordinary power tracks the Type-I inflation; size-adjusted power",
    "(scored at each test's own empirical 5% cutoff) is essentially identical",
    "for every valid test, so MLR's raw-power edge is purely over-rejection."))
}
```

Size-adjusted power scores each test at the cutoff that *would* give a 5%
rejection rate of a true model in that cell, so an over-rejecting test cannot
borrow detection from a broken Type-I rate. On equal honest footing the valid
tests have the same power; `naive` and `MLR` are no better (slightly worse),
their apparent advantage being entirely the inflated size.

Full per-cell rejection (all truths, mechanisms, rates, methods; goodness-of-fit
and nested) is in `results/summary_rejection.csv`; per-replicate p-values, base
statistics, and UGamma spectra are in `results/raw/` (the latter in
`spectra.rds`) so any further test or size-adjustment is recomputable without
refitting. The FIML-bias non-centrality diagnostic is in
`results/noncentrality.csv` (it is a finite-sample non-normality artifact, equal
for MCAR and MAR, not a MAR signal; see the note for why).

## Caveats

- **One MAR, and a weak one.** The MAR cells use a mild sign-based
  Savalei-Bentler mechanism on always-observed indicators, so they behave like
  MCAR and do not probe the regime where MAR biases FIML itself. A stronger MAR
  is the obvious next run; MAR is not one mechanism.
- **Power is shown raw and size-adjusted.** The size-adjusted column is a
  diagnostic (its cutoff is read from the null simulation, unavailable in
  practice); it exists to compare the tests' intrinsic detection at equal size.
- The MLR column is magmaan's `fiml_robust_mlr`, matched to lavaan's
  `yuan.bentler.mplus` defaults (see oracle). `YB_exact` is the exact
  saturated-space trace scaling from the FIML UΓ spectrum; it duplicates `SB`
  by construction. lavaan offers no mean-and-variance test under FIML, so `SS`
  is a magmaan-only construction.

## Reproduce

```sh
Rscript run_experiment.R --smoke                       # single hardest cell
Rscript run_experiment.R --lavaan-parity               # full grid + oracle
Rscript run_experiment.R --cells truth=h0,dist=vm2,mech=MCAR,rate=0.3 --reps 1000
Rscript scripts/size_adjusted_power.R   # size-adjusted power from existing raw/, no refit
```

## Appendix: comparison to lavaan

The MLR over-rejection is genuine, not a magmaan computation error: on the worst
cells lavaan's *own* MLR, fit on the same data every replicate, gives the
identical rejection rate (agreeing per-replicate to ~1e-5), and the statistic
carries ~1.7-2.3x the variance of its reference chi-square, so a first-moment
scaling cannot calibrate it.

```{r tab-oracle}
if (!is.null(P) && nrow(P)) {
  ag <- aggregate(abs_diff ~ metric, P, function(x) max(x, na.rm = TRUE))
  names(ag) <- c("Quantity", "Max |magmaan - lavaan|")
  ag[[2]] <- formatC(ag[[2]], format = "e", digits = 2)
  knitr::kable(ag, align = "l", caption =
    "Oracle: magmaan vs lavaan MLR (estimator = MLR, missing = ml, defaults).")
}
```

```{r tab-mlr-validate}
V <- if (file.exists(file.path(res_dir, "mlr_validation.csv")))
  read.csv(file.path(res_dir, "mlr_validation.csv"), stringsAsFactors = FALSE) else NULL
if (!is.null(V) && nrow(V)) {
  v <- data.frame(
    dist = V$dist,
    `MLR reject (magmaan)` = fmt(V$mlr_reject_magmaan),
    `MLR reject (lavaan)`  = fmt(V$mlr_reject_lavaan),
    `max |Δp|` = formatC(V$max_abs_p_diff, format = "e", digits = 1),
    `variance inflation` = fmt(V$variance_inflation, 2),
    check.names = FALSE)
  knitr::kable(v, align = "l", caption = paste(
    "MLR is not a magmaan artifact: fitting lavaan's own MLR on the same data",
    "every replicate (not just rep 1) reproduces the rejection rate exactly. The",
    "variance-inflation factor (>1) is why a mean-scaling must over-reject."))
}
```
