---
title: "FIML goodness-of-fit under non-normality: FMG/pEBA vs the MLR default"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
res_dir <- "results"
sum_path <- file.path(res_dir, "summary_rejection.csv")
if (!file.exists(sum_path)) {
  stop("No results yet. Run:\n",
       "  Rscript run_experiment.R --smoke         # single hardest cell\n",
       "  Rscript run_experiment.R --lavaan-parity # full grid + oracle\n")
}
S  <- read.csv(sum_path, stringsAsFactors = FALSE)
NC <- if (file.exists(file.path(res_dir, "noncentrality.csv")))
  read.csv(file.path(res_dir, "noncentrality.csv"), stringsAsFactors = FALSE) else NULL
P  <- if (file.exists(file.path(res_dir, "parity.csv")))
  read.csv(file.path(res_dir, "parity.csv"), stringsAsFactors = FALSE) else NULL
meta <- if (file.exists(file.path(res_dir, "metadata.csv")))
  read.csv(file.path(res_dir, "metadata.csv"), stringsAsFactors = FALSE) else NULL
reps <- if (!is.null(meta)) meta$value[meta$key == "reps"] else "?"
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)

gof <- S[S$outcome == "gof", ]
# Clean regime for the calibration claim: MCAR (FIML consistent), highest rate.
mcar_hi <- {
  r <- if ("MCAR" %in% gof$mech) max(gof$rate[gof$mech == "MCAR"]) else NA
  gof[gof$truth == "h0" & gof$mech == "MCAR" & gof$rate == r, ]
}
sev <- mcar_hi[mcar_hi$dist == "vm2", ]
mlr_sev <- sev$reject[sev$method == "MLR"]
yb_exact_sev <- sev$reject[sev$method == "YB_exact"]
yb_exact_struct_sev <- sev$reject[sev$method == "YB_exact_structured"]
all_sev <- sev$reject[sev$method == "all"]
all_struct_sev <- sev$reject[sev$method == "all_structured"]
ss_sev  <- sev$reject[sev$method == "SS"]
ss_struct_sev  <- sev$reject[sev$method == "SS_structured"]
```

## Question

The dominant default for incomplete continuous SEM is lavaan's
`estimator = "MLR", missing = "ml"` — the Mplus variant of the Yuan-Bentler
mean-scaled FIML test. Under non-normality the test statistic's reference law
(a weighted sum of $\chi^2_1$) spreads its eigenvalues, so two questions are
separable: does MLR's finite-sample trace approximation match exact
saturated-space mean scaling, and does the Foldnes-Moss-Gronneberg eigenvalue
family (scaled-shifted, EBA/pEBA, penalized-all, the full mixture) control the
goodness-of-fit Type-I rate where MLR does not?

## Short answer

```{r short, results="asis"}
if (length(mlr_sev) && length(yb_exact_sev) &&
    length(yb_exact_struct_sev) && length(all_sev) &&
    length(all_struct_sev) && length(ss_sev) && length(ss_struct_sev)) {
  cat(sprintf(
    paste0("Under **MCAR** (where normal-theory FIML stays consistent), yes. At ",
           "severe Vale-Maurelli non-normality the MLR default rejects a true ",
           "model **%s** of the time at the .05 level. Exact UΓ mean scaling ",
           "(`YB_exact`, the same p-values as the FMG `SB` row) rejects **%s** ",
           "(**%s** with structured H1 information). The full FMG mixture rejects ",
           "**%s** (**%s** structured), and scaled-shifted rejects **%s** ",
           "(**%s** structured). ",
           "The pattern holds across all non-normal families (%s reps).\n\n"),
    fmt(mlr_sev), fmt(yb_exact_sev), fmt(yb_exact_struct_sev),
    fmt(all_sev), fmt(all_struct_sev), fmt(ss_sev), fmt(ss_struct_sev), reps))
  cat(paste0("Under this study's **MAR** the same conclusion holds. The ",
             "Savalei-Bentler mechanism here deletes indicators by coarse *sign* ",
             "rules on two always-observed indicators, which is mild enough that ",
             "FIML stays effectively unbiased: MAR rejection tracks MCAR for every ",
             "method at a matched rate, and the well-calibrated tests stay at level ",
             "under both. That is a property of *this weak* MAR, not of MAR in ",
             "general. A stronger mechanism (steep continuous selection, or a ",
             "genuinely nonlinear missing-given-observed mean) would bias FIML ",
             "itself and leave a residual no reference-law correction can remove; ",
             "this design does not probe that regime."))
} else {
  cat("Run the full grid (`--lavaan-parity`) to populate this section.\n")
}
```

## What is tested

The goodness-of-fit statistic is the ordinary FIML likelihood-ratio
$T = 2(\hat\ell_1 - \hat\ell_0)$ comparing the metric model to the saturated
observed-data normal model (lavaan's `missing = "ml"` test); the nested test is
the same statistic formed as a configural-vs-metric difference. Under a correct
normal model with ignorable missingness $T \Rightarrow \chi^2_{\mathrm{df}}$.
Under non-normality the working normal likelihood is misspecified, so $T$ limits
instead to a weighted mixture $\sum_j \lambda_j \chi^2_{1j}$ whose weights
$\lambda_j$ spread apart as the non-normality grows. Every robust test
references $T$ to that mixture rather than to $\chi^2_{\mathrm{df}}$: **MLR** and
**SB** rescale by the mean weight $\bar\lambda$ only, **scaled-shifted (SS)**
matches the mean and the variance, and the **FMG** family (EBA/pEBA, and the
full mixture `all`) keeps the entire spectrum. A mean-only scaling like MLR must
over-reject precisely when the weights are dispersed, which is what
non-normality produces. (The saturated vs `_structured` columns evaluate the
curvature at the saturated moments vs the fitted SEM; they share a limit under
the MCAR null. Full derivations:
`docs/research/notes/non_normal_fiml_mlr_information`.)

## Evidence

```{r tab-mcar}
if (nrow(mcar_hi)) {
  ord <- c("naive","MLR","YB_mplus","YB_exact","YB_exact_structured",
           "SB","SB_structured","SS","SS_structured","SF","EBA4","pEBA4",
           "pEBA4_structured","pall","pOLS","all","all_structured")
  w <- reshape(mcar_hi[, c("dist","method","reject")], idvar = "dist",
               timevar = "method", direction = "wide")
  names(w) <- sub("^reject\\.", "", names(w))
  w <- w[, c("dist", intersect(ord, names(w)))]
  num <- vapply(w, is.numeric, logical(1)); w[num] <- lapply(w[num], fmt)
  knitr::kable(w, align = "l", caption = sprintf(
    "GOF Type-I under MCAR %s (nominal .05). FIML is consistent here, so this isolates the reference-law correction.",
    fmt(max(mcar_hi$rate), 2)))
}
```

```{r tab-ncp}
if (!is.null(NC) && nrow(NC)) {
  n <- NC[order(NC$mech, NC$dist, NC$rate),
          c("dist","mech","rate","mean_base","mean_trace","ncp_hat")]
  n[c("mean_base","mean_trace","ncp_hat")] <-
    lapply(n[c("mean_base","mean_trace","ncp_hat")], fmt, d = 2)
  knitr::kable(n, align = "l", caption = paste(
    "Non-centrality diagnostic (truth = h0): ncp_hat = mean(base) -",
    "mean(trace(UGamma)). It is ~0 only under normal data; under non-normality it",
    "is already positive at complete data and grows with the missing rate, but is",
    "essentially equal for MCAR and MAR -- a finite-sample sandwich-trace bias,",
    "not the MAR estimator-target bias this weak mechanism does not trigger."))
}
```

```{r tab-oracle}
if (!is.null(P) && nrow(P)) {
  ag <- aggregate(abs_diff ~ metric, P, function(x) max(x, na.rm = TRUE))
  names(ag) <- c("Quantity", "Max |magmaan - lavaan|")
  ag[[2]] <- formatC(ag[[2]], format = "e", digits = 2)
  knitr::kable(ag, align = "l", caption =
    "Oracle: magmaan vs lavaan MLR (estimator = MLR, missing = ml, defaults).")
}
```

```{r tab-mlr-validate}
V <- if (file.exists(file.path(res_dir, "mlr_validation.csv")))
  read.csv(file.path(res_dir, "mlr_validation.csv"), stringsAsFactors = FALSE) else NULL
if (!is.null(V) && nrow(V)) {
  v <- data.frame(
    dist = V$dist,
    `MLR reject (magmaan)` = fmt(V$mlr_reject_magmaan),
    `MLR reject (lavaan)`  = fmt(V$mlr_reject_lavaan),
    `max |Δp|` = formatC(V$max_abs_p_diff, format = "e", digits = 1),
    `variance inflation` = fmt(V$variance_inflation, 2),
    check.names = FALSE)
  knitr::kable(v, align = "l", caption = paste(
    "MLR is not a magmaan artifact: fitting lavaan's own MLR on the same data",
    "every replicate (not just rep 1) reproduces the rejection rate exactly. The",
    "variance-inflation factor (>1) is why a mean-scaling must over-reject."))
}
```

The MLR over-rejection is genuine, not a computation error: on the worst cells
lavaan's *own* MLR gives the identical rejection rate, agreeing per-replicate to
~1e-5, and the statistic carries ~1.7-2.3x the variance of its reference
chi-square (so a first-moment scaling cannot calibrate it).

```{r tab-power}
PW <- if (file.exists(file.path(res_dir, "power_adjusted.csv")))
  read.csv(file.path(res_dir, "power_adjusted.csv"), stringsAsFactors = FALSE) else NULL
if (!is.null(PW)) {
  nn <- c("vm1", "vm2", "ig1", "ig2", "pl1", "pl2")
  ord <- c("naive", "MLR", "YB_exact", "SS", "SF", "EBA4", "pEBA4",
           "pall", "pOLS", "all")
  ti <- gof[gof$truth == "h0" & gof$mech == "MCAR" &
            gof$rate == max(gof$rate[gof$mech == "MCAR"]) & gof$dist %in% nn, ]
  pw <- PW[PW$outcome == "gof" & PW$mech == "MCAR" &
           PW$rate == max(PW$rate[PW$mech == "MCAR"]) & PW$dist %in% nn, ]
  am <- function(df, col) vapply(ord, function(m)
    mean(df[[col]][df$method == m]), numeric(1))
  tab <- data.frame(Method = ord,
    `Type-I` = fmt(am(ti, "reject"), 3),
    `Power` = fmt(am(pw, "power"), 3),
    `Size-adjusted power` = fmt(am(pw, "power_adj"), 3),
    check.names = FALSE)
  knitr::kable(tab, align = "l", row.names = FALSE, caption = paste(
    "GOF detection at the severe cell (MCAR 30%, mean over the 6 non-normal",
    "families). Ordinary power tracks the Type-I inflation; size-adjusted power",
    "(scored at each test's own empirical .05 cutoff) is essentially identical",
    "for every valid test, so MLR's raw-power edge is purely over-rejection."))
}
```

Size-adjusted power scores each test at the cutoff that *would* give a 5%
rejection rate of a true model in that cell, so an over-rejecting test cannot
borrow detection from a broken Type-I rate. On equal honest footing the valid
tests have the same power; `naive` and `MLR` are no better (slightly worse),
their apparent advantage being entirely the inflated size.

Full per-cell rejection (all truths, mechanisms, rates, methods; goodness-of-fit
and nested) is in `results/summary_rejection.csv`; per-replicate p-values, base
statistics, and UGamma spectra are in `results/raw/` (the latter in
`spectra.rds`) so any further test or size-adjustment is recomputable without
refitting.

## Caveats

- **One MAR, and a weak one.** The MAR cells use a mild sign-based
  Savalei-Bentler mechanism on always-observed indicators, so they behave like
  MCAR and do not probe the regime where MAR biases FIML itself. A stronger MAR
  is the obvious next run; MAR is not one mechanism.
- **Power is shown raw and size-adjusted.** The size-adjusted column is a
  diagnostic (its cutoff is read from the null simulation, unavailable in
  practice); it exists to compare the tests' intrinsic detection at equal size.
- The MLR column is magmaan's `fiml_robust_mlr`, matched to lavaan's
  `yuan.bentler.mplus` defaults (see oracle). `YB_exact` is the exact
  saturated-space trace scaling from the FIML UΓ spectrum; it duplicates `SB`
  by construction. lavaan offers no mean-and-variance test under FIML, so `SS`
  is a magmaan-only construction.

## Reproduce

```sh
Rscript run_experiment.R --smoke                       # single hardest cell
Rscript run_experiment.R --lavaan-parity               # full grid + oracle
Rscript run_experiment.R --cells truth=h0,dist=vm2,mech=MCAR,rate=0.3 --reps 1000
Rscript scripts/size_adjusted_power.R   # size-adjusted power from existing raw/, no refit
```
