---
title: "How far does ordinary Hotelling carry robust SEM goodness of fit?"
subtitle: "Residual rank, sample-to-rank ratio, and one severe nonnormal model"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
override <- Sys.getenv("MAGMAAN_EXPERIMENT_RESULTS", unset = "")
roots <- c(
  if (nzchar(override)) override else character(),
  file.path("results", c("full", "probe", "smoke")))
available <- roots[vapply(
  roots,
  function(path) file.exists(file.path(path, "metadata.csv")),
  logical(1))]
if (!length(available)) {
  stop("missing results; run `Rscript run_experiment.R --smoke`")
}
root <- available[[1L]]
read_result <- function(name) read.csv(
  file.path(root, name), stringsAsFactors = FALSE, check.names = FALSE)
meta_frame <- read_result("metadata.csv")
meta <- setNames(meta_frame$value, meta_frame$key)
methods <- read_result("method_summary.csv")
diagnostics <- read_result("diagnostics_summary.csv")
failures <- read_result("failures.csv")
profile <- meta[["profile"]]
reps <- as.integer(meta[["reps_per_cell"]])
is_evidence <- profile %in% c("probe", "full") && reps >= 300L

labels <- c(
  score_chisq_centered = "Centered robust score: chi-square",
  score_hotelling = "Centered robust score: Hotelling",
  score_chisq_raw = "Raw-OPG score: chi-square",
  rls_chisq = "Working RLS: chi-square",
  fmg_sb_rls = "Working RLS: SB",
  fmg_mv_rls = "Working RLS: mean + variance",
  fmg_peba4_rls = "Working RLS: pEBA4")
methods$label <- unname(labels[methods$method])
methods$distribution_label <- ifelse(
  methods$distribution == "vm", "Vale–Maurelli", "Normal")

summary_by_method <- do.call(rbind, lapply(
  split(
    methods,
    interaction(methods$distribution, methods$method, drop = TRUE)),
  function(x) data.frame(
    distribution = x$distribution[[1L]],
    method = x$method[[1L]],
    label = x$label[[1L]],
    mean_size = mean(x$rejection_rate),
    mean_absolute_error = mean(abs(x$rejection_rate - .05)),
    minimum = min(x$rejection_rate),
    maximum = max(x$rejection_rate),
    stringsAsFactors = FALSE)))
row.names(summary_by_method) <- NULL
summary_by_method$Distribution <- ifelse(
  summary_by_method$distribution == "vm", "Vale–Maurelli", "Normal")
summary_by_method$Method <- summary_by_method$label

pick <- function(method, distribution, column = "mean_absolute_error") {
  summary_by_method[
    summary_by_method$method == method &
      summary_by_method$distribution == distribution,
    column][[1L]]
}
hot_mae <- mean(summary_by_method[
  summary_by_method$method == "score_hotelling",
  "mean_absolute_error"])
chisq_mae <- mean(summary_by_method[
  summary_by_method$method == "score_chisq_centered",
  "mean_absolute_error"])
hot_better_cells <- with(
  reshape(
    methods[methods$method %in% c(
      "score_chisq_centered", "score_hotelling"), ],
    idvar = c(
      "cell_id", "p", "q", "n", "n_over_q", "distribution"),
    timevar = "method", direction = "wide"),
  sum(
    abs(rejection_rate.score_hotelling - .05) <
      abs(rejection_rate.score_chisq_centered - .05),
    na.rm = TRUE))
score_cells <- length(unique(methods$cell_id))
hot_normal_range <- range(methods$rejection_rate[
  methods$method == "score_hotelling" &
    methods$distribution == "normal"])
hot_vm_range <- range(methods$rejection_rate[
  methods$method == "score_hotelling" &
    methods$distribution == "vm"])
max_identity_error <- max(
  diagnostics$max_raw_opg_identity_error, na.rm = TRUE)
```

## Question

Does the ordinary Hotelling reference improve the already-pivotal centered
robust goodness-of-fit score beyond the original five-degree-of-freedom case,
as residual rank and the sample-to-rank ratio change? The practical competitor
is the established route: leave the working RLS score nonpivotal and correct
its weighted-chi-square limit with SB, mean-and-variance, or pEBA4.

## Short answer

`r if (!is_evidence) {
  sprintf("This is a %s run with only %d replications per cell; it verifies the grid and identities but cannot rank calibration.", profile, reps)
} else if (hot_mae < chisq_mae) {
  sprintf("Ordinary Hotelling generalized strongly in this grid: its average cellwise absolute size error was %.2f percentage points, versus %.2f for the centered-score chi-square, and it was closer to 5%% in %d of %d cells.", 100 * hot_mae, 100 * chisq_mae, hot_better_cells, score_cells)
} else {
  sprintf("Ordinary Hotelling did not generalize uniformly: its average cellwise absolute size error was %.2f percentage points, versus %.2f for the centered-score chi-square, and it was closer to 5%% in only %d of %d cells.", 100 * hot_mae, 100 * chisq_mae, hot_better_cells, score_cells)
}`
`r if (is_evidence) sprintf(" It is not exact: Hotelling size ranged from %.1f%% to %.1f%% under normality and %.1f%% to %.1f%% under severe Vale–Maurelli nonnormality. Its mean absolute errors were %.2f/%.2f points under normal/VM, compared with %.2f/%.2f for pEBA4-RLS.", 100 * hot_normal_range[[1L]], 100 * hot_normal_range[[2L]], 100 * hot_vm_range[[1L]], 100 * hot_vm_range[[2L]], 100 * pick("score_hotelling", "normal"), 100 * pick("score_hotelling", "vm"), 100 * pick("fmg_peba4_rls", "normal"), 100 * pick("fmg_peba4_rls", "vm"))`

## Evidence

Let \(\psi_i\in\mathbb R^q\) be the model-centred saturated covariance
contribution after projection off the fitted model tangent, and let

$$
\widehat B_c
=\frac{1}{n-1}\sum_{i=1}^n
  (\psi_i-\bar\psi)(\psi_i-\bar\psi)^\mathsf T,\qquad
Q_c=n\bar\psi^\mathsf T\widehat B_c^{-1}\bar\psi .
$$

The ordinary robust score uses \(Q_c\Rightarrow\chi_q^2\). Ordinary Hotelling
retains exactly the same numerator and centered sandwich denominator but uses

$$
\frac{n-q}{q(n-1)}Q_c\ \dot\sim\ F_{q,n-q}.
$$

The experiment crosses \(q=2,5,9,20\) with \(n=30,60,120,240\). The panel
therefore shows both an ordinary sample-size transition and the dangerous
low-\(n/q\) boundary.

```{r size-plot, fig.width=9, fig.height=5.6}
shown <- c(
  "score_chisq_centered", "score_hotelling",
  "fmg_sb_rls", "fmg_mv_rls", "fmg_peba4_rls")
plot_data <- methods[methods$method %in% shown, ]
plot_data$label <- factor(
  plot_data$label, levels = unname(labels[shown]))
colours <- c("#1B6CA8", "#D1495B", "#4D8B31", "#7A5195", "#C17C00")
old_par <- par(
  mfrow = c(2, length(unique(plot_data$q))),
  mar = c(3.2, 3.4, 2.1, .8),
  oma = c(2.2, 2.3, .3, .3), las = 1)
on.exit(par(old_par), add = TRUE)
ymax <- .20
for (distribution in c("normal", "vm")) {
  for (rank in sort(unique(plot_data$q))) {
    panel <- plot_data[
      plot_data$distribution == distribution & plot_data$q == rank, ]
    xlim <- range(panel$n_over_q)
    plot(
      xlim, c(0, ymax), type = "n", log = "x",
      xlab = "", ylab = "", axes = FALSE,
      main = sprintf("%s, q = %d",
        ifelse(distribution == "vm", "VM", "Normal"), rank))
    rect(
      xlim[[1L]], .025, xlim[[2L]], .075,
      col = adjustcolor("grey70", alpha.f = .22), border = NA)
    abline(h = .05, lty = 2, col = "grey35")
    axis(1)
    axis(2)
    box()
    for (k in seq_along(shown)) {
      rows <- panel[panel$method == shown[[k]], ]
      rows <- rows[order(rows$n_over_q), ]
      displayed <- pmin(rows$rejection_rate, ymax)
      lines(
        rows$n_over_q, displayed,
        type = "b", pch = 19, cex = .6, lwd = 1.1,
        col = colours[[k]])
      clipped <- rows$rejection_rate > ymax
      if (any(clipped)) {
        points(
          rows$n_over_q[clipped], displayed[clipped],
          pch = 17, cex = .8, col = colours[[k]])
      }
    }
    if (distribution == "normal" && rank == min(plot_data$q)) {
      legend(
        "topright", legend = unname(labels[shown]), col = colours,
        lty = 1, pch = 19, cex = .48, bty = "n")
    }
  }
}
mtext("Sample-to-score-rank ratio n/q (log scale)", side = 1, outer = TRUE)
mtext("Null rejection at 5%", side = 2, outer = TRUE)
```

Values above 20% are clipped to the panel ceiling and marked by triangles; the
summary table retains their actual values.

The cellwise summary keeps nonnormality separate and reports dispersion rather
than hiding opposite errors in one pooled rejection rate.

```{r method-summary}
show_summary <- summary_by_method[
  summary_by_method$method %in% shown,
  c(
    "Distribution", "Method", "mean_size",
    "mean_absolute_error", "minimum", "maximum")]
show_summary <- show_summary[
  order(
    match(show_summary$Distribution, c("Normal", "Vale–Maurelli")),
    match(show_summary$Method, unname(labels[shown]))), ]
names(show_summary) <- c(
  "Distribution", "Method", "Mean Size",
  "Mean Absolute Error", "Minimum", "Maximum")
knitr::kable(show_summary, digits = 3, row.names = FALSE)
```

### What every alternative is

| Alternative | Statistic | Reference |
|---|---:|---|
| Centered robust score: chi-square | \(Q_c\) | \(\chi_q^2\), the asymptotically pivotal baseline |
| Centered robust score: Hotelling | \(Q_c\) | \(F_{q,n-q}\) after the displayed finite-sample rescaling |
| Raw-OPG score: chi-square | \(Q_r=n\bar\psi^\mathsf T\widehat B_r^{-1}\bar\psi\) | \(\chi_q^2\), with \(\widehat B_r=n^{-1}\sum_i\psi_i\psi_i^\mathsf T\) |
| Working RLS: chi-square | \(T_{\mathrm{RLS}}=n\lVert\bar\psi\rVert^2\) | Naive \(\chi_q^2\) |
| Working RLS: SB | \(T_{\mathrm{RLS}}\) | Mean-scaled weighted-chi-square approximation |
| Working RLS: mean + variance | \(T_{\mathrm{RLS}}\) | First-two-moment Satterthwaite approximation |
| Working RLS: pEBA4 | \(T_{\mathrm{RLS}}\) | Four leading estimated eigenvalues plus a penalized tail |

The raw-OPG row is an identity audit, not independent corroboration:

$$
Q_r=\frac{nQ_c}{n-1+Q_c}.
$$

The largest numerical error in that identity was
`r format(max_identity_error, scientific = TRUE, digits = 3)`.

## Caveats

- `r if (is_evidence) sprintf("With %s replications per cell, a true 5%% rejection rate has Monte Carlo standard error about %.2f percentage points.", format(reps, big.mark = ","), 100 * sqrt(.05 * .95 / reps)) else sprintf("Only %d replications per cell were run; rejection rates are not evidence.", reps)`
- The nonnormal arm is one severe homogeneous Vale–Maurelli model with marginal
  skewness 3 and excess kurtosis 21. Directional kurtosis, misspecification,
  mean structures, missingness, and multigroup nested restrictions are outside
  this experiment.
- Hotelling is exact for iid Gaussian score vectors, not for quadratic SEM
  scores after fitting nuisance parameters. Here it is deliberately treated as
  a simple finite-sample benchmark.

## Reproduce

```sh
Rscript experiments/72-hotelling-robust-gof-grid/run_experiment.R --smoke
Rscript experiments/72-hotelling-robust-gof-grid/run_experiment.R --probe
quarto render experiments/72-hotelling-robust-gof-grid/report.qmd
```
