---
title: "magmaan vs lavaan: speed bench"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
  gfm: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"), candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "zoo.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")

zoo_path <- file.path(results_dir, "zoo.csv")
pipeline_path <- file.path(results_dir, "pipeline.csv")
geiser_path <- file.path(results_dir, "geiser.csv")
if (!file.exists(zoo_path)) {
  stop("Missing zoo.csv. Run: Rscript run_experiment.R")
}
if (!file.exists(pipeline_path)) {
  stop("Missing pipeline.csv. Run: Rscript run_experiment.R")
}

has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE) &&
  knitr::is_html_output()
if (has_ggplot2) {
  library(ggplot2)
}

zoo <- utils::read.csv(zoo_path, check.names = FALSE, stringsAsFactors = FALSE)
pipeline <- utils::read.csv(pipeline_path, check.names = FALSE,
                            stringsAsFactors = FALSE)
geiser <- if (file.exists(geiser_path)) {
  utils::read.csv(geiser_path, check.names = FALSE, stringsAsFactors = FALSE)
} else NULL

fmt_ms <- function(x) ifelse(is.na(x), "", sprintf("%.3f", x))
fmt_speedup <- function(x) ifelse(is.na(x), "", sprintf("%.1f×", x))
if (has_ggplot2) {
  theme_set(theme_minimal(base_size = 12))
}
```

How fast is magmaan compared to lavaan on the same models? Three complementary
slices below: a *headline* slice on five real published latent-variable models
from Geiser (2013), a small *zoo* slice on classic lavaan tutorial models that
the benchmark harness already validates against the lavaan oracle, and a
whole-pipeline slice that includes raw-data statistic construction and selected
post-fit reporting.

The headline and zoo slices time **estimate-only** workloads (`se = "none"`,
`test = "none"` in lavaan) at **single-threaded** BLAS / OpenMP. The pipeline
slice times broader calls: UGamma robust reporting, all-ordinal DWLS with
polychorics, and mixed ordinal/continuous DWLS with polyserial moments.

## Headline: Geiser (2013) corpus

Five real published latent-variable models from Geiser's textbook, drawn from
the SNLLS paper's `snlls_geiser_corpus_lavaan.csv` table (committed under
`papers/snlls-constrained/tables/`). magmaan's `full` and `snlls` columns are
the same fit reported at two backend choices; both are well under a
millisecond at this scale.

```{r}
if (!is.null(geiser)) {
  ge <- geiser
  ge$Speedup <- fmt_speedup(ge$speedup)
  out <- data.frame(
    Model = ge$model,
    `Lavaan ms` = fmt_ms(ge$lavaan_ms),
    `magmaan ms (full ML)` = fmt_ms(ge$magmaan_full_ms),
    `magmaan ms (SNLLS)`   = fmt_ms(ge$magmaan_snlls_ms),
    Speedup = ge$Speedup,
    check.names = FALSE
  )
  knitr::kable(out, row.names = FALSE,
               align = c("l", "r", "r", "r", "r"))
} else {
  cat("`results/geiser.csv` not present; run `Rscript run_experiment.R` first.\n")
}
```

```{r}
#| fig-width: 7
#| fig-height: 3.6
if (!is.null(geiser)) {
  if (has_ggplot2) {
    g <- geiser
    g$model <- factor(g$model, levels = g$model[order(g$speedup)])
    ggplot(g, aes(speedup, model)) +
      geom_col(fill = "#7a3b2c", width = 0.55) +
      geom_text(aes(label = sprintf("%.1f×", speedup)),
                hjust = -0.1, size = 3.6, color = "gray25") +
      scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
      labs(x = "lavaan / magmaan median wall time",
           y = NULL,
           title = "Geiser 2013 corpus: lavaan-to-magmaan speedup",
           subtitle = "estimate-only workload, single thread")
  }
}
```

## Zoo: classic lavaan tutorial models

Re-run live in this experiment off the `benchmarks/` harness, on lavaan's
in-package datasets:

- **Holzinger-Swineford 3-factor CFA** (`HolzingerSwineford1939`, n=301, 9
  indicators) at three estimators: ML, ULS, GLS.
- **Bollen industrialization/democracy SEM** (`PoliticalDemocracy`, n=75,
  11 observed variables, latent regressions + residual covariances).
- **`Demo.growth` linear latent growth curve** (n=400, 4 waves + 2
  time-varying covariates + 4 controls, with mean structure).

Speedup = `lavaan_ms_median / magmaan_ms_median`.

```{r}
zoo_out <- data.frame(
  Model = zoo$label,
  Estimator = zoo$estimator,
  `n` = zoo$n_obs,
  `Obs vars` = zoo$n_vars,
  Params = zoo$npar,
  `Lavaan ms` = fmt_ms(zoo$lavaan_ms_median),
  `magmaan ms` = fmt_ms(zoo$magmaan_ms_median),
  Speedup = fmt_speedup(zoo$speedup),
  `Max |Δθ|` = sprintf("%.1e", zoo$est_max_abs_diff),
  check.names = FALSE
)
knitr::kable(zoo_out, row.names = FALSE,
             align = c("l", "l", "r", "r", "r", "r", "r", "r", "r"))
```

```{r}
#| fig-width: 7
#| fig-height: 3.8
zoo$display <- paste0(zoo$label, " (", zoo$estimator, ")")
zoo$display <- factor(zoo$display, levels = zoo$display[order(zoo$speedup)])
if (has_ggplot2) {
  ggplot(zoo, aes(speedup, display)) +
    geom_col(fill = "#2c4e7a", width = 0.55) +
    geom_text(aes(label = sprintf("%.1f×", speedup)),
              hjust = -0.1, size = 3.6, color = "gray25") +
    scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
    labs(x = "lavaan / magmaan median wall time",
         y = NULL,
         title = "Tutorial-model zoo: lavaan-to-magmaan speedup",
         subtitle = "estimate-only workload, single thread, parity-validated to 1e-3")
}
```

## Whole-pipeline smoke

These rows are deliberately small. They ask whether the README headline still
has a representative path beyond continuous estimate-only fitting: raw data are
converted to the required sample statistics, the model is fit, and the named
post-fit report is materialized where magmaan exposes one.

```{r}
pipeline_out <- data.frame(
  Workflow = pipeline$label,
  Estimator = pipeline$estimator,
  `Data Path` = pipeline$data_path,
  `n` = pipeline$n_obs,
  `Obs vars` = pipeline$n_vars,
  `Lavaan ms` = fmt_ms(pipeline$lavaan_ms_median),
  `magmaan ms` = fmt_ms(pipeline$magmaan_ms_median),
  Speedup = fmt_speedup(pipeline$speedup),
  `Check` = sprintf("%.2g", pipeline$check_value),
  check.names = FALSE
)
knitr::kable(pipeline_out, row.names = FALSE,
             align = c("l", "l", "l", "r", "r", "r", "r", "r", "r"))
```

The `Check` column is the row-specific parity smoke: scaled chi-square/scale
factor drift for the ML UGamma row, unscaled chi-square drift for the ordinal
DWLS row, and free-parameter-count drift for the mixed row.

## Methodology

- **Estimate-only**: magmaan via `magmaan::magmaan()`; lavaan with
  `se = "none"`, `test = "none"`. The post-fit machinery (standard errors,
  test statistics, fit indices) is excluded — both libraries are timed on the
  *core fit loop* in the headline and zoo slices only.
- **Whole-pipeline**: the smoke rows include raw-data statistic construction
  and the named reporting path. The ML row includes UGamma spectra,
  Satorra-Bentler scaling, and robust SEs; the ordinal row includes
  polychorics, NACOV/weights, DWLS, and robust ordinal reporting; the mixed row
  includes threshold/polyserial/polychoric moment construction and DWLS fit.
- **Single-threaded**: `OMP_NUM_THREADS`, `OPENBLAS_NUM_THREADS`, and
  `MKL_NUM_THREADS` are set to `1` before R loads BLAS. The Geiser-corpus
  numbers were produced with the same protocol; see `papers/snlls-constrained`.
- **Iterations**: 30 per case. Times are medians of repeated elapsed-time
  measurements.
- **Correctness gate**: each zoo case's magmaan estimates must agree with the
  lavaan oracle to `1e-3` (max absolute over free parameters) before timing is
  reported. The pipeline rows use the row-specific smoke check shown in the
  table.

## Reproduce

```sh
Rscript experiments/05-lavaan-speed-bench/run_experiment.R
(cd experiments/05-lavaan-speed-bench && quarto render report.qmd)
```

## See also

- [`docs/validation/lavaan_tutorial_parity.md`](../../docs/validation/lavaan_tutorial_parity.md) — central lavaan-tutorial parity audit.
- [`benchmarks/README.md`](../../benchmarks/README.md) — the live magmaan-vs-lavaan harness this experiment is built on.
- The deeper Geiser-corpus study (estimator timing, convergence) lives in the
  separate SNLLS paper repository; its frozen `snlls_geiser_corpus_lavaan.csv`
  is copied here as [`results/geiser.csv`](results/geiser.csv).
