---
title: "Scalar parameter profiles: normal ML and nonnormal fixed-weight inference"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
read_result <- function(name) {
  path <- file.path("results", name)
  if (!file.exists(path)) stop("Missing retained output: ", path,
    "; see Reproduce and use a fresh --out-dir for new runs.")
  read.csv(path, stringsAsFactors=FALSE)
}
lrt <- read_result("profile_lrt_summary.csv")
ci_all <- read_result("profile_ci_summary.csv")
ci <- ci_all[ci_all$profile_weight == "fixed", ]
meta <- read_result("metadata.csv")
```

## Question

Do interior scalar parameter profiles calibrate under normal ML and under
nonnormal sampling with fixed GLS/WLS/ULS profile weights?

## Short answer

The retained normal ML loading test rejects 5.0–5.5% in four 200-attempt cells.
Nonnormal confidence intervals remain method-dependent: at $N=500$, the
complete-IJ references cover 91.3% for GLS and 88.8% for WLS on 80 usable
attempts each. This is an **active calibration study**, not a settled interval
policy. The fitted-weight ULS arm was retired; its historical rows are retained
as provenance and excluded from the current comparison.

## Evidence

Both lanes target the `x2` loading (.8) in the same four-indicator model.
Normal ML tests the true loading through a constrained likelihood ratio.
The CI lane uses multivariate $t_5$ draws scaled to the same population covariance.

```{r normal}
knitr::kable(data.frame(`N range`=paste(range(lrt$n),collapse="–"),
  Cells=nrow(lrt), `Usable tests`=sum(lrt$reps), Failures=sum(lrt$failures),
  `Rejection range (%)`=paste(sprintf("%.1f",100*range(lrt$type1_05)),collapse="–"),
  check.names=FALSE), row.names=FALSE)
```

The CI table is a selected $N=500$ slice. `ordinary` uses the unscaled profile;
`robust_fixed` uses the fixed-weight influence; `robust_estimated_weight`
uses the complete IJ for the estimator's weight influence. All displayed
profiles hold their weights fixed during inversion.

```{r nonnormal}
x <- ci[ci$n == 500, ]
knitr::kable(data.frame(Estimator=x$estimator, Reference=gsub("_"," ",x$reference),
  Usable=x$reps, Failures=x$failures, `Coverage (%)`=100*x$coverage,
  `Mean width`=x$mean_width, `Mean endpoint error`=x$mean_endpoint_error,
  check.names=FALSE), digits=3, row.names=FALSE)
```

The original summaries retain all $N=100,200,500$ cells and each reference's
failure count. Successful-output counts are conditional denominators, not all
attempts. `profile_ci_summary.csv` also retains the retired `profile_weight=fitted`
rows; do not use them to describe the current fixed-weight API. The raw failure
CSVs retain stage and error information.

## Caveats

With 200 usable tests, a 5% rejection rate has MCSE about 1.5 percentage points;
with 80 usable intervals, 95% coverage has MCSE about 2.4 points. These are
exploratory cells, and estimated references differ in width, failures and root
accuracy. The current metadata records 200 ML and 80 CI attempts per cell;
future runs must preserve that distinction. No Bartlett correction, boundary
law or generic misspecified-model guarantee is established here.

The next check is independent fixed-profile coverage with endpoint accuracy
and numerical availability reported separately; no correction should be chosen
from a pooled estimator/reference average.

## Reproduce

From this directory:

```bash
Rscript run_experiment.R --help
Rscript run_experiment.R --smoke --reps 2 --n-grid 50 --ci-reps 1 \
  --ci-n-grid 100 --out-dir /tmp/parameter_profile_check
quarto render report.qmd
```

A new verification run writes to its own output directory. Rendering reads only
this study's retained `results/` and never fits or simulates.
