---
title: "Ordinal Omega Target Audit"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- "results/target_summary.csv"
item_path <- "results/item_diagnostics.csv"
metadata_path <- "results/metadata.csv"
if (!file.exists(summary_path) || !file.exists(item_path) || !file.exists(metadata_path)) {
  stop("Missing results. Run: Rscript run_experiment.R --smoke")
}
summary <- read.csv(summary_path)
item <- read.csv(item_path)
metadata <- read.csv(metadata_path)
worst <- summary[which.max(summary$abs_difference), ]
close_enough <- max(summary$abs_difference) < 1e-3
```

## Question

Does the current ordinal observed-score covariance omega equal the direct
one-factor ordinal true-score target, or is it only on the same observed-score
metric?

## Short Answer

```{r, results='asis'}
if (close_enough) {
  cat(sprintf(
    "In this %d-cell population audit, the two targets are numerically close: the largest absolute difference is %.4f.",
    metadata$n_cells, max(summary$abs_difference)))
} else {
  cat(sprintf(
    "No. In this %d-cell population audit, the largest absolute difference is %.4f (%s), with the covariance target %.4f versus the direct target %.4f.",
    metadata$n_cells, max(summary$abs_difference), worst$cell,
    worst$current_covariance_omega, worst$direct_true_score_omega))
}
```

## Evidence

```{r}
shown <- summary[, c(
  "loading_pattern", "threshold_pattern", "min_category_prob",
  "current_covariance_omega", "direct_true_score_omega", "difference",
  "numerator_ratio", "max_abs_communality_gap"
)]
names(shown) <- c(
  "Loadings", "Thresholds", "Min Category Prob.",
  "Covariance Omega", "Direct Omega", "Difference",
  "Numerator Ratio", "Max Diagonal Gap"
)
shown[] <- lapply(shown, function(x) if (is.numeric(x)) round(x, 4) else x)
knitr::kable(shown)
```

The diagnostic gap is on the diagonal: the current covariance target recovers
item true-score variances from off-diagonal covariances, while the direct target
integrates the conditional expected category scores.

## Caveats

This is a population target audit under a one-factor probit response model with
local independence and integer category scores. It does not fit models, evaluate
sampling behavior, handle multi-group pooling, or choose an interval method.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R
quarto render report.qmd
```
