---
title: "Score versus LR: calibration and compute time"
params:
  run: primitives-2000
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
path <- file.path('results',params$run)
need <- c('metadata.csv','score_lrt_timing.csv','score_lrt_summary.csv',
          'replicates.csv','paired_vs_lr_sb.csv')
if(!all(file.exists(file.path(path,need)))) stop(
  'Missing completed run. Run: Rscript run_experiment.R --reps 2000 --seed-base 20260917 --output ',params$run)
tim <- read.csv(file.path(path,'score_lrt_timing.csv'))
meta <- read.csv(file.path(path,'metadata.csv'))
x <- read.csv(file.path(path,'replicates.csv'))
all_methods <- read.csv(file.path(path,'score_lrt_summary.csv'))
pairs <- read.csv(file.path(path,'paired_vs_lr_sb.csv'))
one <- x[!duplicated(paste(x$cell_id,x$rep)),]
primary <- c('sb_ml','score_sb','peba4_ml','score_peba4')
y <- x[x$method %in% primary,]
y$valid <- !nzchar(y$error) & is.finite(y$p_value)
common <- aggregate(valid~cell_id+rep,y,all)
names(common)[3] <- 'common_valid'
y <- merge(y,common,by=c('cell_id','rep'))
y <- y[y$common_valid,]
s <- do.call(rbind,lapply(split(y,interaction(y$cell_id,y$method,drop=TRUE)),function(z) {
  rate <- mean(z$p_value<.05)
  data.frame(z[1,c('cell_id','p','n','distribution','method')],valid=nrow(z),
    rejection_05=rate,mcse=sqrt(rate*(1-rate)/nrow(z)))
}))
stopifnot(nrow(s)==18*length(primary))
write.csv(s,file.path(path,'primary_common_valid.csv'),row.names=FALSE)
checks <- read.csv('results/validation/checks.csv')
stopifnot(all(checks$passed))
get_meta <- function(key) meta$value[match(key,meta$key)]
seconds <- tim$elapsed_seconds[tim$phase=='simulation_loop']
total <- tim$elapsed_seconds[tim$phase=='total_runner']
shared_fail <- sum(nzchar(one$error))
mae <- function(method) 100*mean(abs(s$rejection_05[s$method==method]-.05),na.rm=TRUE)
paired <- merge(s[s$method=='sb_ml',],s[s$method=='score_sb',],by='cell_id',suffixes=c('_lr','_score'))
score_closer <- sum(abs(paired$rejection_05_score-.05)<abs(paired$rejection_05_lr-.05))
fmt <- function(x) sprintf('%.2f',x)
```

## Question

Do score tests give better Type I error calibration than likelihood-ratio tests
with Satorra–Bentler correction, and how quickly can we compare them with pEBA?

## Short answer

**Mostly yes for SB calibration in this grid.** SB-calibrated score is closer to the nominal 5% rejection rate
in **`r score_closer` of 18 cells**. Its mean absolute deviation from 5% is
**`r fmt(mae('score_sb'))` percentage points**, versus
**`r fmt(mae('sb_ml'))` for LR–SB**. These are descriptive comparisons across
the prespecified cells, on draws valid for all four primary methods; the plot shows where the
difference occurs. With pEBA4 calibration, the corresponding deviations are
`r fmt(mae('score_peba4'))` for score and `r fmt(mae('peba4_ml'))` for LR.

This does not solve calibration uniformly: score–SB still rejects as often as
17.7% in the most difficult cell, and score–pEBA4 as often as 8.8%. These are
null rejection rates, not power gains.

**`r format(nrow(one),big.mark=',')` samples, `r get_meta('reps')` per cell,
took `r sprintf('%.1f',seconds/60)` minutes of simulation**, with one worker.
Total runner time, including setup, warmup and output summaries, was
`r sprintf('%.1f',total/60)` minutes. The laptop started on
`r get_meta('power_start')` and ended on `r get_meta('power_end')`.
This measures computation, not the time to develop or publish the experiment.

## Evidence

### Null calibration

```{r}
#| label: calibration
#| fig-width: 11
#| fig-height: 8
#| fig-cap: "Rejection at nominal 5%, on draws valid for all four displayed methods. Bars show pointwise ±1.96 Monte Carlo SE; comparisons share simulated samples."
plot_calibration <- function() {
  methods <- c('sb_ml','score_sb','peba4_ml','score_peba4')
  labels <- c('LR–SB','Score–SB','LR–pEBA4','Score–pEBA4')
  colors <- c('#333333','#d43b13','#286b8a','#8c4485')
  old <- par(mfrow=c(3,2),mar=c(3.4,4.3,2.4,1),oma=c(0,0,1,0),mgp=c(2,.7,0),cex=.85)
  on.exit(par(old))
  for(d in c('normal','t10','vm2')) for(p in c(10,20)) {
    z <- s[s$distribution==d & s$p==p & s$method %in% methods,]
    upper <- max(.08,z$rejection_05+1.96*z$mcse,na.rm=TRUE)*100
    plot(NA,xlim=c(.8,3.2),ylim=c(0,upper),xaxt='n',
      xlab='Sample size',ylab='Rejection (%)',
      main=paste(c(normal='Normal',t10='t(10)',vm2='Vale–Maurelli, severe')[d],
        '·',p,'indicators'))
    axis(1,at=1:3,labels=c(100,200,500));abline(h=5,lty=2,col='grey50')
    for(k in seq_along(methods)) {
      a <- z[z$method==methods[k],];a <- a[order(a$n),]
      xx <- match(a$n,c(100,200,500))+(k-2.5)*.045
      yy <- 100*a$rejection_05;se <- 196*a$mcse
      nz <- is.finite(se) & se>0
      arrows(xx[nz],pmax(0,yy[nz]-se[nz]),xx[nz],yy[nz]+se[nz],
        angle=90,code=3,length=.03,col=colors[k])
      lines(xx,yy,type='b',pch=c(16,17,1,2)[k],col=colors[k])
    }
    if(d=='normal' && p==10) legend('topright',labels,col=colors,pch=c(16,17,1,2),lty=1,bty='n',cex=.7)
  }
}
png(file.path(path,'calibration.png'),width=1800,height=1300,res=160)
plot_calibration()
invisible(dev.off())
plot_calibration()
```

Monte Carlo SE is approximately **0.49 percentage points** at a true rejection
rate of 5% with 2,000 valid draws. Paired rejection differences and their Monte
Carlo SEs are saved in `paired_vs_lr_sb.csv`; the bars above are pointwise,
not intervals for those paired differences. The full 12-method results,
including pEBA2, RLS and the sandwich score comparator, are in
`score_lrt_summary.csv`. The primary comparison uses between
`r min(s$valid)` and `r max(s$valid)` common valid draws per cell; these
rates and counts are saved in `primary_common_valid.csv`.

There were **`r shared_fail` shared fit failures among `r nrow(one)` samples**,
including convergence, covariance-admissibility and stationarity failures.
The sandwich score comparator is additionally unavailable with 20 indicators
and N = 100 because its estimated covariance lacks the required rank.
All unavailable results retain a reason in the raw output.

### Measured compute time

```{r}
display <- tim[,c('phase','elapsed_seconds')]
names(display) <- c('Phase','Seconds')
display$Phase <- c('Setup','Warmup','Simulation loop','Total runner')
display$Seconds <- round(display$Seconds,1)
knitr::kable(display)
```

The table gives realized wall time for the current implementation without
assuming a parallel speedup. The earlier run remains in `results/main-2000/`. The computation uses one worker and one BLAS thread. Hardware, software
fingerprints, power state, seeds and exact command are saved in `metadata.csv`.

All **`r nrow(checks)` numerical checks passed** on four separate datasets:
LR/SB agreement with lavaan, pEBA-family agreement with semTests, a separate
Gaussian score formula, and the normal-reference reduction. At an interior
stationary fit, the expected-information score statistic equals the fitted-weight
RLS statistic; the calibrations remain separate.

## Caveats

- This is a null-size study only; it does not yet compare power.
- The models are the source paper's two-factor, 10- and 20-indicator populations.
  N = 100, 200 and 500 are a smaller-sample extension of its design.
- Distributions are normal, the paper's severe Vale–Maurelli condition
  (skewness 3, excess kurtosis 21), and a symmetric standardized t(10)
  extension. The t(10) condition is **not** a paper replication; it has finite
  eighth moments. All three preserve the target covariance.
- Failed fits are retained, with rejection estimates conditioned on valid
  draws and paired comparisons restricted to common valid draws. Read failure
  rates alongside any later calibration result.
- Tests share data and a fitted null. Timing records distinguish the batched
  LR/RLS comparisons from score setup and calibrations. Individual LR-family
  post-fit times have not been separately measured. Shared inference preparation
  is included in the full simulation time, but excluded from the per-method
  pipeline-time field. The former nuisance-stationarity column is unavailable
  on the current path; the score/RLS identity check remains active.
- Question-to-report elapsed time and active researcher time are unmeasured.
  The compute timing must not be presented as total research-development time.

## Reproduce

From the repository root, with the installed magmaan, lavaan and semTests packages:

```sh
OMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  Rscript experiments/81-score-vs-lrt/validate.R

OMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  Rscript experiments/81-score-vs-lrt/run_experiment.R \
  --reps 2000 --workers 1 --budget-minutes 60 --seed-base 20260917 --output primitives-2000

quarto render experiments/81-score-vs-lrt/report.qmd
```

Use a new output name and `quarto render ... -P run:NEW_NAME` for a repeat run;
existing output directories are never overwritten. Rendering never starts a
simulation. Raw per-replicate tests, failures and timing live under
`results/<run>/`; `--help` describes the runner options.

The design follows [Foldnes, Moss & Grønneberg (2025; online 2024)](https://doi.org/10.1080/10705511.2024.2372028),
with population loadings from the [source supplement](https://osf.io/6trwu/).
