---
title: "Score versus LR: timing the first simulation"
params:
  run: timing-battery-final
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
path <- file.path('results',params$run)
if(!file.exists(file.path(path,'budget_projection.csv'))) stop(
  'Missing pilot results. Run: Rscript run_experiment.R --reps 50 --output ',params$run)
b <- read.csv(file.path(path,'budget_projection.csv'))
tim <- read.csv(file.path(path,'score_lrt_timing.csv'))
meta <- read.csv(file.path(path,'metadata.csv'))
x <- read.csv(file.path(path,'replicates.csv'))
one <- x[!duplicated(paste(x$cell_id,x$rep)),]
checks <- read.csv('results/validation/checks.csv')
get_meta <- function(key) meta$value[match(key,meta$key)]
seconds <- tim$elapsed_seconds[tim$phase=='simulation_loop']
project <- function(reps) (b$fixed_seconds+reps*b$seconds_per_grid_round)/60
shared_fail <- sum(nzchar(one$error))
```

## Question

Can we compare score and likelihood-ratio goodness-of-fit tests, including
Satorra–Bentler and pEBA calibrations, within an hour on this laptop?

## Short answer

**The timing pilot says yes.** `r nrow(one)` samples across 18 cells took
**`r round(seconds,1)` seconds** with one worker, on
`r get_meta('power_start')`. At that measured rate, **2,000 replications per cell
would take about `r round(project(2000))` minutes**, and **5,000 about
`r round(project(5000))` minutes**. The latter is a reasonable target within an
hour, allowing for slower fits and other activity.

This is a runtime result, not a verdict on which test is better. There are only
`r get_meta('reps')` replications per cell. The main simulation has not run.

## Evidence

```{r}
plan <- data.frame(
  `Replications per cell`=c(2000,5000),
  `Total samples`=18*c(2000,5000),
  `Projected minutes`=round(vapply(c(2000,5000),project,numeric(1)),1),
  `MC SE at 5% (percentage points)`=round(100*sqrt(.05*.95/c(2000,5000)),2),
  check.names=FALSE)
knitr::kable(plan)
```

The projections include measured setup and warmup, and retain the measured
single-worker throughput. They assume no parallel speedup. Compilation and
experiment development are excluded. Plugging in, CPU power settings, thermal
conditions and competing workloads can change the rate. A repeat pilot after
plugging in can update the estimate.

There were **`r shared_fail` shared fit failures among `r nrow(one)` samples**,
including convergence, covariance-admissibility and stationarity failures.
The sandwich score comparator is additionally unavailable with 20 indicators
and N = 100: there are too few observations for its covariance estimate to
have the required rank. This does not prevent SB and pEBA from using a
rank-deficient estimated spectrum. All unavailable results retain a reason.

All **`r nrow(checks)` numerical checks passed** on four separate datasets.
They cover LR/SB agreement with lavaan, pEBA-family agreement with semTests,
a separate Gaussian score formula, and the normal-reference reduction.
The expected-information score statistic equals the fitted-weight RLS statistic
at a stationary interior fit; the different calibrations remain separate.

## Caveats

- This is a null-size study only; it does not yet compare power.
- The models are the source paper's two-factor, 10- and 20-indicator populations.
  N = 100, 200 and 500 are a smaller-sample extension of its design.
- Distributions are normal, the paper's severe Vale–Maurelli condition
  (skewness 3, excess kurtosis 21), and a symmetric standardized t(10)
  extension. The t(10) condition is **not** a paper replication; it has finite
  eighth moments. All three preserve the target covariance.
- Failed fits are retained, with rejection estimates conditioned on valid
  draws and paired comparisons restricted to common valid draws. Read failure
  rates alongside any later calibration result.
- Tests share data and a fitted null. Timing records distinguish the batched
  LR/RLS comparisons from score setup and calibrations. Individual LR-family
  post-fit times have not been separately measured.
- Question-to-report elapsed time and active researcher time are unmeasured.
  The compute timing must not be presented as total research-development time.

## Reproduce

From the repository root, with the installed magmaan, lavaan and semTests packages:

```sh
OMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  Rscript experiments/81-score-vs-lrt/validate.R

# New output name preserves the battery pilot.
OMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  Rscript experiments/81-score-vs-lrt/run_experiment.R \
  --reps 50 --workers 1 --budget-minutes 60 --output timing-ac

quarto render experiments/81-score-vs-lrt/report.qmd -P run:timing-ac
```

A later main run can use `--reps 5000 --output main-5000` with the same thread
settings. Nothing in rendering starts a simulation, and no main run is started
automatically. Output folders must be new to preserve earlier runs.

Raw draws' test results, failures, per-cell timing, software fingerprints,
power state and hardware metadata are stored under `results/<run>/`. Use
`Rscript experiments/81-score-vs-lrt/run_experiment.R --help` for options.

The design follows [Foldnes, Moss & Grønneberg (2025; online 2024)](https://doi.org/10.1080/10705511.2024.2372028),
with population loadings from the [source supplement](https://osf.io/6trwu/).
