---
title: "Mplus Demo WLSMV DIFFTEST Probe"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
fit_path <- file.path("results", "model_fit.csv")
diff_path <- file.path("results", "difftest.csv")
meta_path <- file.path("results", "metadata.csv")
aux_path <- file.path("results", "auxiliary_check.csv")
if (!file.exists(fit_path) || !file.exists(diff_path)) {
  stop("Run `Rscript run_experiment.R` first.", call. = FALSE)
}
model_fit <- read.csv(fit_path, na.strings = "")
difftest <- read.csv(diff_path, na.strings = "")
metadata <- read.csv(meta_path, na.strings = "")
auxiliary <- read.csv(aux_path, na.strings = "")

fmt <- function(x, digits = 3) {
  ifelse(is.finite(x), sprintf(paste0("%.", digits, "f"), x), "")
}
fmt_p <- function(x) ifelse(is.finite(x), sprintf("%.4f", x), "")
meta_value <- function(key) metadata$value[match(key, metadata$key)]

main_methods <- rbind(
  difftest[difftest$engine == "mplus" &
             difftest$method == "DIFFTEST", ],
  difftest[difftest$engine == "lavaan" &
             difftest$method == "satorra.2000_delta_scaled_shifted" &
             difftest$restriction %in% c("metric", "scalar_mplus"), ],
  difftest[difftest$engine == "magmaan" &
             difftest$variant == "pd_gamma_overlap" &
             difftest$method == "restriction_map_delta_scaled_shifted", ],
  difftest[difftest$engine == "magmaan" &
             difftest$variant == "pd_gamma_overlap" &
             difftest$restriction %in% c("scalar_group_equal", "scalar_mplus") &
             !is.na(difftest$error), ]
)
main_methods$stat_fmt <- fmt(main_methods$stat)
main_methods$df_fmt <- fmt(main_methods$df)
main_methods$p_fmt <- fmt_p(main_methods$p_value)
main_methods$error[is.na(main_methods$error)] <- ""
diff_tab <- main_methods[, c("engine", "variant", "restriction", "method",
                             "stat_fmt", "df_fmt", "p_fmt", "error")]
names(diff_tab) <- c("Engine", "Variant", "Restriction", "Method",
                     "Statistic", "df", "p", "Error")

fit_focus <- model_fit[
  model_fit$model %in% c("configural", "metric", "scalar_mplus",
                         "scalar_group_equal") &
    (model_fit$engine != "magmaan" |
       model_fit$variant %in% c("pd_gamma_overlap", "pd_gamma_nominal")),
]
fit_focus$chisq_raw_fmt <- fmt(fit_focus$chisq_raw)
fit_focus$chisq_scaled_fmt <- fmt(fit_focus$chisq_scaled)
fit_focus$p_scaled_fmt <- fmt_p(fit_focus$pvalue_scaled)
fit_tab <- fit_focus[, c("engine", "variant", "model", "npar", "df",
                         "chisq_raw_fmt", "chisq_scaled_fmt", "p_scaled_fmt")]
names(fit_tab) <- c("Engine", "Variant", "Model", "npar", "df",
                    "Raw chi-square", "Adjusted chi-square", "Adjusted p")
```

## Question

Can the Mplus Demo be used to see what Mplus WLSMV pairwise-deletion DIFFTEST is
doing in the Chen et al. setting, and is that statistic essentially lavaan's
`satorra.2000` scaled-shifted difference test?

## Short Answer

For a demo-sized metric-invariance test, yes: Mplus DIFFTEST, lavaan's
`lavTestLRT(..., method = "satorra.2000", A.method = "delta",
scaled.shifted = TRUE)`, and magmaan with the overlap pairwise Gamma all give
the same result to rounding.

For the scalar-like Mplus default, lavaan can be made count-compatible with
Mplus by freeing the group-B latent mean and fixing the observed intercepts.
That count-compatible lavaan statistic is close but not identical to Mplus in
this one replicate. magmaan can fit the same 38-parameter Mplus-style scalar
model with the same explicit latent-mean/item-intercept syntax, and with the
overlap pairwise Gamma its raw least-squares objective and scaled-shifted
restriction-map statistic match Mplus to rounding: `22.365850` on 22 df versus
Mplus DIFFTEST `22.366000` on 22 df.

The auxiliary variable is only used to drive missingness. Mplus rejects
`AUXILIARY = aux (m)` with categorical variables in this WLSMV setup, so this is
a WLSMV_PD/pairwise probe rather than an auxiliary-variable probe.

## Nested Tests

```{r}
knitr::kable(diff_tab, row.names = FALSE)
```

## Model Fit Rows

The Mplus adjusted chi-square is the printed WLSMV model-fit statistic. The raw
chi-square is `2 * N * fmin`, included to compare the underlying least-squares
objective.

```{r}
knitr::kable(fit_tab, row.names = FALSE)
```

## Auxiliary Check

```{r}
knitr::kable(auxiliary, row.names = FALSE)
```

## Design

The data are a two-group, six-indicator, one-factor ordinal CFA, small enough
for Mplus Demo's six-dependent-variable limit:

```text
f =~ y1 + y2 + y3 + y4 + y5 + y6
```

Both groups use loading `r meta_value("loading")` and thresholds
`r meta_value("thresholds")`. The total sample size is
`r meta_value("n_total")`, split evenly by group. Missingness is applied only in
group B, on `y5` and `y6`, with target rate
`r meta_value("missing_rate_target")`; the realized group-B rate is
`r meta_value("missing_rate_realized_group_b_y5_y6")`.

The Mplus input/output files and `h1diff.dat` are kept in
`r meta_value("mplus_work_dir")`.

## Reproduce

```sh
Rscript run_experiment.R
quarto render report.qmd
```
