---
title: "Finite-sample goodness-of-fit references"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Question

Can Hotelling or projected Satterthwaite references improve finite-sample
calibration of centered robust SEM scores, and when does the improvement persist?

## Short answer

**Hotelling is promising in the available grids; the projected empirical
shape correction remains unproved.** The rank/sample-ratio grid extends the
five-degree-of-freedom pilot, but Hotelling is not exact for estimated SEM
covariance scores. In the projected lane's VM transition, replacing the
same-sample shape estimate with an independent replicate did not reveal a
consistent coupling benefit; both projected empirical versions remained
conservative. Higher-order moments and fitted nuisance effects remain unresolved.

This study is banked. Reopen for a named finite-sample GOF consumer with an
independent influence/moment validation. The lanes retain different ranks,
innovation moments and historical runs; they are not one pooled simulation.

## Evidence

```{r setup}
read_lane <- function(lane, candidates) {
  paths <- file.path("results", lane, candidates)
  good <- paths[file.exists(file.path(paths,"method_summary.csv"))]
  if (!length(good)) stop("Missing ",lane," results; run that lane's smoke command.")
  path <- good[[1L]]
  list(path=path, summary=read.csv(file.path(path,"method_summary.csv")))
}
p <- read_lane("projected",c("pilot","smoke"))
h <- read_lane("hotelling",c("full","probe","smoke"))
summarize <- function(z, methods) {
  z <- z[z$method %in% methods & is.finite(z$rejection_rate),]
  do.call(rbind,lapply(split(z,interaction(z$distribution,z$method)),function(x)
    data.frame(Distribution=x$distribution[1],Method=unname(c(score_chisq_centered="Centered chi-square",
        hotelling_gaussian="Hotelling",score_hotelling="Hotelling",
        satterthwaite_normal_projected="Normal-reference projected",
        satterthwaite_empirical_projected="Empirical projected",
        score_chisq_centered_matched="Centered chi-square",hotelling_gaussian_matched="Hotelling",
        satterthwaite_empirical_projected_matched="Same-sample projected",
        satterthwaite_independent_projected="Independent projected",
        fmg_peba4_rls="pEBA4 RLS")[x$method[1]]),Cells=nrow(x),
      `Mean absolute error (points)`=100*mean(abs(x$rejection_rate-.05)),
      `Minimum rejection (%)`=100*min(x$rejection_rate),
      `Maximum rejection (%)`=100*max(x$rejection_rate),
      `Minimum valid draws`=min(x$n_valid),check.names=FALSE)))
}
```

### Projected shape: fixed residual rank

The pilot fixes q=5 and varies N=30/60/120/240 under normal and Vale–Maurelli
innovations (skewness 2, excess kurtosis 7). The projected correction uses
normal-reference, DGP-oracle or empirical shape estimates. An oracle or
independent-replication shape is a diagnostic, not implementable single-sample
inference. The current report reads `r p$path`.

```{r}
knitr::kable(summarize(p$summary,c("score_chisq_centered","hotelling_gaussian",
  "satterthwaite_normal_projected","satterthwaite_empirical_projected")),
  digits=2,row.names=FALSE)
```

```{r}
path <- file.path("results","projected","transition","method_summary.csv")
if (file.exists(path)) {
  z <- read.csv(path)
  knitr::kable(summarize(z,c("score_chisq_centered_matched","hotelling_gaussian_matched",
    "satterthwaite_empirical_projected_matched","satterthwaite_independent_projected")),
    digits=2,row.names=FALSE,caption="VM transition: independent replicate isolates shape/numerator dependence.")
}
```

### Hotelling: rank and sample-to-rank ratio

The grid crosses q=2/5/9/20 and N=30/60/120/240 under normal and a more severe
Vale–Maurelli DGP (skewness 3, excess kurtosis 21). It reads `r h$path`.
Cellwise Wilson intervals, valid denominators and failure reasons remain
in each lane's CSVs. Descriptive mean absolute errors are averages of cell
errors, not pooled rejection estimates.

```{r}
knitr::kable(summarize(h$summary,c("score_chisq_centered","score_hotelling",
  "fmg_peba4_rls")),digits=2,row.names=FALSE)
```

## Caveats and methods

For fitted-model projected covariance contributions $\psi_i\in\mathbb R^q$,
use $\widehat B_c=(n-1)^{-1}\sum_i(\psi_i-\bar\psi)(\psi_i-\bar\psi)^\mathsf T$
and $Q_c=n\bar\psi^\mathsf T\widehat B_c^{-1}\bar\psi$.
The asymptotic reference is $\chi_q^2$; ordinary Hotelling replaces it by
$(n-q)Q_c/[q(n-1)]\dot\sim F_{q,n-q}$.

Hotelling's exact reference requires iid Gaussian vectors with suitable
mean/covariance assumptions. SEM covariance contributions are quadratic in
observations and use fitted nuisance parameters; normal observed data do not
establish exactness. The projected Satterthwaite calculation examines
denominator shape, without proving a complete nuisance-influence expansion.
Empirical shape involves noisy higher (up to eighth) moments; same-sample
and independent estimates are diagnostic controls. Raw OPG, centered
covariance and working-RLS spectra have different conventions.

Smoke verifies execution and identities, not rankings. The pilot/full artifacts
have substantive replication counts, but these exploratory grids have no
policy-selection or power claim. Different VM moment targets remain explicit.

## Reproduce

```sh
Rscript run_experiment.R --lane projected --help
Rscript run_experiment.R --lane projected --smoke --results-dir /tmp/gof-projected
Rscript run_experiment.R --lane hotelling --help
Rscript run_experiment.R --lane hotelling --smoke --results-dir /tmp/gof-hotelling
quarto render report.qmd
```

Former research/40 and /41 sources and outputs remain in their lanes. The
report reads only `results/{projected,hotelling}/`, preferring pilot/full to
smoke. Preserve seeds and method definitions; verify moves in fresh directories.
