---
title: "Total-variance scaling across SEM structures"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
if (!file.exists("results/fits.csv"))
  stop("Run Rscript run_experiment.R --smoke --stress first")
f <- read.csv("results/fits.csv")
v <- read.csv("results/validation.csv")
tm <- read.csv("results/timing.csv")
meta <- read.csv("results/metadata.csv")
mv <- function(key) meta$value[match(key, meta$key)]
```

## Question

Does fixing **total latent variance** improve SEM optimization across different
structural models, when all parameterizations start at the same model and search
the same admissible interior? Can we answer that in seconds rather than hours?

## Short answer

**Total-variance scaling is a viable candidate, not a universal winner.** The
initial stress pilot improves information conditioning substantially for mediation
with a highly predictable mediator and for a higher-order factor model. Disturbance
scaling remains better in the correlated-disturbance example. Structural variety
changes the answer even with nearly constant indicator count.

The current run completed **`r nrow(f)` fits in
`r round(as.numeric(mv("elapsed_seconds")), 2)` seconds**, across
`r length(unique(f$case))` structures. Completion status is
**`r mv("complete")`**. This is a small comparison prototype, not evidence of native
library speed or finite-sample failure rates. Ten-draw follow-ups also finished within a minute, but exposed near-boundary
stalls that the stationarity audit rejects. Practical starts and better boundary
handling are needed before drawing conclusions about failure rates.

## Evidence

### Different structures favor different scales

```{r}
w <- reshape(v[, c("case", "chart", "condition")], idvar = "case",
             timevar = "chart", direction = "wide")
names(w) <- sub("condition\\.", "", names(w))
knitr::kable(w, digits = 1, col.names = c("Model", "Marker", "Disturbance", "Total"),
  caption = "Expected-information condition numbers at the controlled population or empirical reference fit; smaller is better. Controlled models use the stress setting when requested. These are coordinate-dependent diagnostics, not an ordering of statistical models.")
```

All charts reproduce the same starting observed covariance to
`r format(max(v$covariance_error), scientific = TRUE, digits = 2)`; analytic covariance
derivatives agree with central differences to maximum relative error
`r format(max(v$derivative_error), scientific = TRUE, digits = 2)`.
The empirical reference objective is checked independently against lavaan before
fitting. The maximum attained objective discrepancy from the available population
or lavaan reference is
`r format(max(abs(f$objective - f$reference_objective), na.rm = TRUE), scientific = TRUE, digits = 2)`.

A fit is not accepted merely because its optimizer returns success. There are
`r sum(!f$stationary)` rows failing the common information-metric stationarity
threshold, and `r sum(!f$admissible)` failing the component covariance audit in this
run. The paired fit gaps are descriptive; neither their minimum nor agreement
between three charts proves a global optimum.

### Iteration costs seconds

```{r}
z <- aggregate(seconds ~ stage, tm, sum)
knitr::kable(z, digits = 3, col.names = c("Work", "Seconds"),
  caption = "Measured work in the current run. Total elapsed time additionally includes orchestration and checkpoint writes; process startup is excluded.")
```

The previous experiment's cost arm repeatedly timed model construction, fitting,
and back-conversion **inside every Monte Carlo replication**. Each timer requested
five batches of at least 50 ms, after calibration. Its default smoke cost grid has
960 rows: a nominal 12 minutes of measurement alone. Its full grid has 72,000 rows:
a nominal **15 hours**. The batch-count cap can shorten very cheap operations;
calibration, warmup, and other arms add time. These are workload calculations, not
measurements of a historical full run.

A fresh isolated probe on this machine, with one p=6 / loading=.3 / replication /
backend cell and all three charts, took **4.05 s** for cost, **1.80 s** for
back-conversion, **0.37 s** for geometry, and **0.34–0.39 s** for each of structural,
convergence, and inference. These timings include R startup. Thus repeated timing
is a demonstrated source of the long runtime; there is no saved historical profile
to attribute every second of the earlier run. Dense curvature calculations can
also grow expensive with model size, but they did not dominate this small probe.

The ten-draw follow-ups completed 198 fits each in **26.9 s** (ordinary) and
**31.0 s** (stress), with all planned chart triples present. These two runs were
launched concurrently, so these are iteration wall budgets, not comparative speed
measurements. Some Bollen draws and stressed controlled draws approach singular
disturbance covariances; the largest paired objective gap was .348, and every gap
above $10^{-7}$ was rejected by the stationarity audit. These are unresolved
near-boundary optimization cases, not evidence that any chart found a certified
optimum. They are retained in `results/ordinary` and `results/iteration`.

This experiment constructs each case once, validates first derivatives once per
chart, draws sample covariances directly, and fits each chart once per draw. It
prints and checkpoints each fit and stops between fits at a declared soft time
budget. It does not run repeated microbenchmarks or second-derivative curvature
calculations in the replication loop.

## Caveats

- Six selected structures and a few draws cannot establish a generally best chart.
  Population conditioning is deterministic; convergence frequency is not.
- All three charts use the same small R optimization prototype with analytic
  derivatives. Its timing includes R overhead and must not be used to claim a
  speedup in magmaan. Native backend comparisons follow only after choosing a
  promising map.
- Starts are perturbations of the controlled population or empirical reference fit,
  transformed exactly into each chart. This isolates coordinates but is
  deliberately reference-assisted. Native start heuristics need a separate arm.
- The comparison admits only strictly positive-definite disturbance and measurement
  covariance matrices. Boundary optima and unrestricted Heywood estimates are
  outside this pilot. Rejected steps are counted; rejection-based optimization can
  struggle near a boundary and is not a proposed production constraint method.
- An optimizer success flag, a small gradient in raw coordinates, and an admissible
  covariance are distinct properties. The report keeps them separate. The
  information-metric gradient threshold is a pilot diagnostic, not a proof of
  optimality or the library's authoritative fit verdict.
- The initial models are recursive, single-group, continuous, and covariance-only,
  without shared parameter labels. Feedback, observed exogenous predictors,
  multi-group equalities, fixed loading growth models, and mean structures require
  additional mapping and identification checks before inclusion.
- The private textbook corpus is not mounted in this checkout. The two empirical
  cases use shared benchmark syntax and public lavaan datasets; the other four are
  controlled structures. No sibling experiment or test fixture is consumed.
- The time limit is checked between fits, not a hard process timeout. A single fit
  is bounded by its iteration limit and can overrun the remaining wall budget.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --dry-run-plan --smoke --stress
Rscript run_experiment.R --smoke --stress
quarto render report.qmd
# Ordinary setting, kept separate from the rendered stress pilot:
Rscript run_experiment.R --smoke --results-dir results/ordinary
# Expand only the number of paired draws; no repeated timing batches:
Rscript run_experiment.R --reps 10 --stress --budget-sec 60 --results-dir results/iteration
```

The runner writes `fits.csv`, `validation.csv`, `timing.csv`, and `metadata.csv`.
The seed for each model/draw is stable when selecting a subset of cases.
A budget-limited run exits with status 2 and flags incomplete chart triples.
Changing `--results-dir` keeps separate runs; rendering reads the top-level
`results/` run. Failures remain in the output instead of being silently dropped.

The isolated legacy runtime probe can be repeated from the repository root:

```sh
# Repeat for geometry, cost, backconvert, structural, convergence, inference.
/usr/bin/time -p Rscript experiments/82-latent-metric-geometry/run_experiment.R \
  --arms cost --p 6 --curv-p 6 --bc-p 6 --lambda1 0.3 --struct-b 0.9 \
  --conv-n 50 --reps 1 --backends nlopt-lbfgs --results-dir /tmp/latent-metric-cost
```

## Design and next decisions

### Initial model selection

| Model | Distinctive feature | Stress change |
|---|---|---|
| Holzinger–Swineford | Three correlated CFA factors; control where disturbance and total scaling coincide | None; observed-data sentinel |
| Mediation | Direct and indirect paths through three factors | Mediator predictability about .94 |
| Correlated predictors | Two correlated latent causes of one latent outcome | Predictor correlation .95 |
| Higher-order | One higher-order cause of three measured first-order factors | Strong higher-order paths and small first-order disturbances |
| Correlated disturbances | Two endogenous factors with correlated innovations, plus correlated indicator errors | Innovation correlation .80 |
| Bollen political democracy | Empirical mediation with unequal observed scales and several correlated indicator errors | None; observed-data sentinel |

The controlled models all have nine observed variables. The higher-order model has
four latents rather than three; Bollen has eleven observed variables. This makes
structural differences central instead of disguising a size sweep as generality.
For each model, draw 0 uses its population or observed covariance. Additional draws
are Gaussian samples from its controlled or reference-fitted population, with the
same covariance draw shared across charts. The default sample size is 400 for
controlled cases and the original sample size for empirical cases.

### Three equivalent interior parameterizations

For $\eta=B\eta+\zeta$, let $A=(I-B)^{-1}$ and $C=A\Psi A^\top$.
The three conventions are marker loadings fixed to one, disturbance variances fixed
to one, and total latent variances fixed to one. Higher-order marker anchors are
loadings on lower-order factors. Latent signs use the same positive anchor
orientation at initialization.

Total scaling does not need a general equality-constrained optimizer for these
models. With $\Psi_{\mathrm{off}}$ containing only off-diagonal disturbances, solve

$$
(A\circ A)d=\mathbf 1-\operatorname{diag}
  (A\Psi_{\mathrm{off}}A^\top),\qquad
\Psi=\Psi_{\mathrm{off}}+\operatorname{diag}(d).
$$

Here $\circ$ denotes elementwise multiplication. This eliminates one disturbance
variance per latent while freeing its marker loading, preserving parameter count.
For recursive models the solve is nonsingular after a topological permutation.
The formula includes correlated disturbances; a shortcut such as $1-R^2$ without
the covariance terms would not suffice. Feedback models would additionally need
the invertibility of this elimination map checked, beyond invertibility of $I-B$.

All three maps retain the same structural zeros, the same observed covariance
likelihood, and the same positive-definite component domain. The full information
rank is checked; a singular matrix is never made to look well-conditioned by
silently dropping its smallest eigenvalues.

### What would justify a larger run?

First require covariance and derivative checks below $10^{-8}$ and $10^{-6}$,
respectively, full information rank, and agreement with empirical reference
objectives where the reference lies in the common domain. Record model-level
objective gaps, information-scaled gradient norms, rejected steps, evaluations,
and total elapsed time. An objective gap above $10^{-7}$ warrants investigation;
it is not automatically a chart failure or an oracle defect.

The first ten-draw follow-up in both settings is complete. Resolve its
near-boundary stalls before estimating convergence rates or declaring a winning
chart. Compare
optimizer work only among admissible, stationary complete pairs, while separately
reporting every failure and budget truncation. Do not pool model structures into
one headline failure rate. If a difference survives, add practical common starts
and a second optimizer before moving the candidate into the native library.

A later timing arm should use at most a few representative datasets per structure,
with a fixed per-operation timing budget, separately from replications. Any later
curvature arm should likewise be opt-in and evaluated once at selected reference
points. The previous CFA curvature ratio is not an optimality certificate for all
possible SEM parameterizations.
