---
title: "Total-variance scaling across SEM structures"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
if (!file.exists("results/fits.csv"))
  stop("Run Rscript run_experiment.R --smoke --stress first")
f <- read.csv("results/fits.csv")
v <- read.csv("results/validation.csv")
tm <- read.csv("results/timing.csv")
meta <- read.csv("results/metadata.csv")
mv <- function(key) meta$value[match(key, meta$key)]
```

## Question

Does fixing **total latent variance** improve SEM optimization across different
structural models, when all parameterizations start at the same model and search
the same admissible interior? Does native PSD fitting benefit from diagonal
information scaling with practical default starts? Can we answer this in seconds?

## Short answer

**Closed for engineering defaults (2026-09-20).** For supported complete-data
PSD ML, use the existing native fitter with its default starts, `nlopt-slsqp`,
and `preconditioning = "none"`. Preserve the supplied model identification,
zeros, fixed values and equalities, permit PSD boundary solutions, and retain
the existing objective, feasibility and cone-stationarity audits. Diagonal
scaling remains an explicit research option. Automatic chart selection,
preconditioning and targeted restarts are not adopted.

This is a decision to retain a fast, tested baseline, not a claim of global
optimality or a measured rare-failure guarantee. All 220 fits in the structural
pilot passed the audits; diagonal scaling increased median evaluations in all
ten settings, and the default unscaled fit reached the better endpoint in the
one discrepant pair. That evidence does not justify extra default machinery.
The singular-boundary limitation remains documented below.

**Keep native PSD ML unscaled by default.** Across six structures and ten settings,
all 220 fits passed the existing audits, but diagonal preconditioning increased
median evaluations in every setting. Of 110 paired datasets, 109 reached matching
objectives and boundary classifications. One stressed mediation draw reached two
different audited stationary endpoints; unchanged cross-restarts retained that
difference. A targeted boundary diagnostic now explains and escapes the worse one.
The full comparison took 2.83 seconds. The earlier benefit from reference-assisted
Bollen starts does not generalize to native starts in this pilot. **The discrepant mediation case is a boundary
identification problem:** a zero mediator disturbance makes direct and mediated
paths interchangeable. A score-guided move among equivalent parameters followed
by reopening that variance reaches the lower objective with both solver settings.
Random perturbations were less dependable in this single-case check.

**The Bollen follow-up separates coordinate scaling from PSD boundaries.** On the
original data and ten matched Gaussian draws, eight cases have matching interior
reference solutions; three have matching PSD-boundary reference solutions.
Information scaling resolves the interior audit failures. The existing magmaan
PSD fitter handles all eleven from two starts, including the three boundary
cases, using the appropriate cone-stationarity check. These are locally audited
reference candidates, not proofs of global optimality.

**Total-variance scaling is a viable candidate, not a universal winner.** The
initial stress pilot improves information conditioning substantially for mediation
with a highly predictable mediator and for a higher-order factor model. Disturbance
scaling remains better in the correlated-disturbance example. Structural variety
changes the answer even with nearly constant indicator count.

The current run completed **`r nrow(f)` fits in
`r round(as.numeric(mv("elapsed_seconds")), 2)` seconds**, across
`r length(unique(f$case))` structures. Completion status is
**`r mv("complete")`**. This is a small comparison prototype, not evidence of native
library speed or finite-sample failure rates. Ten-draw follow-ups also finished within a minute. Their near-boundary stalls
motivated the Bollen classification below; practical starts and more model
structures remain necessary before drawing conclusions about failure rates.

## Evidence

### Different structures favor different scales

```{r}
w <- reshape(v[, c("case", "chart", "condition")], idvar = "case",
             timevar = "chart", direction = "wide")
names(w) <- sub("condition\\.", "", names(w))
knitr::kable(w, digits = 1, col.names = c("Model", "Marker", "Disturbance", "Total"),
  caption = "Expected-information condition numbers at the controlled population or empirical reference fit; smaller is better. Controlled models use the stress setting when requested. These are coordinate-dependent diagnostics, not an ordering of statistical models.")
```

All charts reproduce the same starting observed covariance to
`r format(max(v$covariance_error), scientific = TRUE, digits = 2)`; analytic covariance
derivatives agree with central differences to maximum relative error
`r format(max(v$derivative_error), scientific = TRUE, digits = 2)`.
The empirical reference objective is checked independently against lavaan before
fitting. The maximum attained objective discrepancy from the available population
or lavaan reference is
`r format(max(abs(f$objective - f$reference_objective), na.rm = TRUE), scientific = TRUE, digits = 2)`.

A fit is not accepted merely because its optimizer returns success. There are
`r sum(!f$stationary)` rows failing the common information-metric stationarity
threshold, and `r sum(!f$admissible)` failing the component covariance audit in this
run. The paired fit gaps are descriptive; neither their minimum nor agreement
between three charts proves a global optimum.

### Native starts: diagonal scaling does not reduce optimizer work

```{r}
structure_path <- "results/native-structures/pairs.csv"
if (file.exists(structure_path)) {
  sp <- read.csv(structure_path)
  comparable <- sp[!is.na(sp$comparable) & sp$comparable, ]
  work <- aggregate(cbind(evaluations.none, evaluations.diagonal) ~ case, comparable, median)
  counts <- aggregate(comparable ~ case, sp, function(x) sprintf("%d/%d", sum(x), length(x)))
  work <- merge(work, counts, by="case")
  labels <- c(hs_3factor_cfa="HS CFA", mediation="Mediation",
    correlated_predictors="Correlated predictors", higher_order="Higher order",
    correlated_disturbances="Correlated disturbances", bollen_democracy_sem="Bollen SEM")
  work$case <- unname(labels[work$case])
  names(work) <- c("Structure", "Unscaled Evaluations", "Diagonal Evaluations", "Matched Pairs")
  knitr::kable(work, digits=1,
    caption="Median evaluations among audited pairs with matching objectives and nullities.")
}
```

Each empirical model uses its observed covariance and ten matched Gaussian draws.
Each synthetic model uses both ordinary and stressed settings, with population
moments and ten matched draws per setting. Both methods receive the same default
native starts and evaluation budget. The synthetic rows summarize both settings;
diagonal scaling increases median evaluations in each setting separately as well.
No reference estimates are supplied as optimizer starts. Population or fitted
reference moments define the simulated data only.

All 220 fits passed feasibility and cone-stationarity checks. The 109 matching
pairs differ in objective by less than $6\times10^{-12}$. Stressed mediation draw
4 differs by 0.00519: the unscaled endpoint has objective 0.05939 and the scaled
endpoint 0.06458. Both initially have covariance nullity one. Restarting either
method from either endpoint preserves its objective, so merely changing scaling
at termination does not resolve the discrepancy. This is evidence of distinct
stationary endpoints, not proof that either is globally optimal. One restart's
nullity changes from one to zero at essentially the same objective, emphasizing
that numerical rank classifications near the boundary depend on tolerance.

The reported counts exclude that unmatched pair from the work comparison but
retain it in the denominator. Single-pass timings and every fit, failure, and
paired objective gap remain in `results/native-structures/`. These timings are
sufficient to establish a cheap iteration loop, not stable speed ratios.

### Why the mediation endpoint can be escaped

Write the structural part as

$$
f_2=a f_1+\zeta_2,\qquad f_3=c f_1+b f_2+\zeta_3.
$$

At the worse endpoint, $\operatorname{Var}(\zeta_2)=0$. Only $c+ab$ matters:
changing $b$ by $\delta$ and $c$ by $-a\delta$ leaves the entire implied observed
covariance unchanged. The separate paths are unidentifiable there. The better
endpoint instead has a small positive mediator disturbance (0.002315) and zero
outcome disturbance, with large compensating paths ($b=17.95$, $c=-16.79$).
Neither solution is a regular interior point.

The derivative for opening the mediator variance depends on the otherwise
unidentified $b$. At the same implied covariance, it is positive at $b=0$
(0.1136) but negative at $b=1$ ($-0.1410$). The numerical check preserves covariance
to $9\times10^{-16}$, verifies the analytic derivative by finite differences,
and agrees with native objective evaluation before and after opening the variance.
Thus the worse covariance has an explicit admissible descent path after choosing
an equivalent parameter representation. Passing the first-order audit at the
original representation does not certify optimality over all nearby attainable
covariances at this singular point.

The diagnostic restart examines four compensated shifts, $\pm1$ and $\pm2$
in units of the fitted total latent standard-deviation ratio. It chooses the
most negative variance derivative, opens small variances to 1% of their marginal
variance, and refits. Selection uses the current fit and data only; the lower
endpoint is used solely to score the experiment's outcome afterward.

```{r}
mediation_path <- "results/mediation-boundary/fits.csv"
if (file.exists(mediation_path)) {
  md <- read.csv(mediation_path)
  lower <- min(md$objective[md$accepted])
  md <- md[md$endpoint == "diagonal" & !grepl("^ridge_", md$policy), ]
  md$policy[grepl("^jitter_", md$policy)] <- "Random perturbation"
  md$policy[md$policy == "unchanged"] <- "Unchanged restart"
  md$policy[md$policy == "reopen"] <- "Reopen variances only"
  md$policy[md$policy == "score_guided"] <- "Score-guided equivalent start"
  md$reached <- md$accepted & abs(md$objective-lower) < 1e-7
  tab <- aggregate(reached ~ policy + preconditioning, md,
                   function(x) sprintf("%d/%d",sum(x),length(x)))
  tab <- reshape(tab,idvar="policy",timevar="preconditioning",direction="wide")
  tab <- tab[c("policy", "reached.none", "reached.diagonal")]
  knitr::kable(tab, row.names=FALSE,
    col.names=c("Restart from Worse Endpoint", "Unscaled", "Diagonal"),
    caption="Restarts reaching the lower observed objective; all fits passed the existing audits.")
}
```

All 42 diagnostic fits passed the existing audits in about 0.6 seconds. Both
score-guided fits reached 0.05939. Random perturbations reached it in 1/6 unscaled
and 3/6 scaled attempts; reopening variances alone failed to escape. Perturbing
the better endpoint returned to its objective in all twelve random trials.
These are results for one selected difficult dataset, not success-rate estimates
for SEMs generally, and the lower objective is not proven globally optimal.

The **boundary identification diagnostic and targeted restart** remain an
experimental explanation and recovery tool. The compensation used here
requires both paths to be freely adjustable and the mediator disturbance to be
zero; it must not be applied blindly to other graphs, correlated disturbances,
or constrained paths. The procedure remains experiment-local, with no change
to native fitting defaults or its audit tolerances.

### Bollen: information scaling helps in the interior

```{r}
bollen_ready <- all(file.exists(file.path("results/bollen",
  c("fits.csv", "references.csv", "metadata.csv", "routing.csv"))))
if (bollen_ready) {
  bf <- read.csv("results/bollen/fits.csv")
  br <- read.csv("results/bollen/references.csv")
  bm <- read.csv("results/bollen/metadata.csv")
  bi <- bf[bf$reference_class == "interior_candidate", ]
  tab <- aggregate(evaluations ~ chart + method, bi, median)
  count <- aggregate(I(classification == "matched_interior") ~ chart + method, bi, sum)
  tab$accepted <- count[[3]][match(paste(tab$chart, tab$method), paste(count$chart, count$method))]
  knitr::kable(tab, digits = 1,
    col.names = c("Latent Scale", "Information Scaling", "Median Evaluations", "Interior Cases Accepted"))
} else cat("Run `Rscript scripts/bollen_probe.R` to produce the Bollen follow-up.")
```

All coordinate choices start from the same model and use the same objective,
analytic derivatives, and optimizer. Diagonal scaling removes differences in
parameter units. Full scaling also removes their local coupling, using information
at the starting point, held fixed throughout fitting. Both resolve all eight
interior cases in this small run; unscaled coordinates miss the stationarity
threshold in one or two cases per chart.

The prototype selector compares three diagonally equilibrated information
matrices at the start, chooses the least ill-conditioned, and applies full
information scaling in that chart. It selects **marker** for Bollen. This choice
uses no fitted outcomes. It is a heuristic, not an optimum certificate: total
scaling with full information scaling has slightly fewer median evaluations in
these draws. Most of the gain comes from information scaling; this does not yet
justify a complicated chart classifier.

### PSD boundaries require a different optimality check

```{r}
if (bollen_ready) {
  boundary <- unique(bf$replicate[bf$reference_class == "psd_boundary_candidate"])
  uu <- br[br$engine == "lavaan_unrestricted" & br$replicate %in% boundary, ]
  pp <- br[br$engine == "magmaan_psd" & br$start == "common" & br$replicate %in% boundary, ]
  pp <- pp[match(uu$replicate, pp$replicate), ]
  knitr::kable(data.frame(Draw = uu$replicate, Unrestricted = uu$objective,
      PSD = pp$objective, Nullity = pp$nullity, ConeStationary = pp$cone_stationary),
    digits = 6, col.names = c("Draw", "Unrestricted Objective", "PSD Objective",
      "Covariance Nullity", "Cone Stationary"))
}
```

The two native PSD starts agree within $10^{-7}$ in all eleven cases and pass
covariance admissibility and cone stationarity. The three boundary cases are draws
2, 5, and 8; their unrestricted references are inadmissible. Thus a higher PSD
objective is expected. Ordinary zero-gradient stationarity is not the appropriate
acceptance criterion at these constrained solutions. The open-interior fits can
approach the boundary without ever attaining it, or stall prematurely along the
way. Better coordinates alone do not fix that missing boundary treatment.

The native fitter represents each permitted covariance component as $LL^\top$,
with constraints linking it to the original model entries. Singular factors are
allowed. This preserves the specified covariance zeros, unlike simply freeing a
dense triangular factor. It also avoids treating a singular covariance as an
invalid statistical model merely because a strictly positive Cholesky diagonal
cannot represent it. The common cone audit is essential: stationarity in squared
factor coordinates alone is not sufficient at a zero factor.

An experimental routing replay uses only the selected fit's feasibility,
information-scaled gradient, and dimensionless boundary diagnostics. It accepts a
regular interior fit; otherwise it requests the existing PSD solver from the
original common start. Its decisions do not use the unrestricted or PSD reference
objectives. The resulting paths are recorded in `results/bollen/routing.csv`.
This is a prototype policy replay, not a new public fitting API or a justification
for changing the user's covariance domain.

### Iteration costs seconds

```{r}
z <- aggregate(seconds ~ stage, tm, sum)
knitr::kable(z, digits = 3, col.names = c("Work", "Seconds"),
  caption = "Measured work in the current run. Total elapsed time additionally includes orchestration and checkpoint writes; process startup is excluded.")
```

The previous experiment's cost arm repeatedly timed model construction, fitting,
and back-conversion **inside every Monte Carlo replication**. Each timer requested
five batches of at least 50 ms, after calibration. Its default smoke cost grid has
960 rows: a nominal 12 minutes of measurement alone. Its full grid has 72,000 rows:
a nominal **15 hours**. The batch-count cap can shorten very cheap operations;
calibration, warmup, and other arms add time. These are workload calculations, not
measurements of a historical full run.

A fresh isolated probe on this machine, with one p=6 / loading=.3 / replication /
backend cell and all three charts, took **4.05 s** for cost, **1.80 s** for
back-conversion, **0.37 s** for geometry, and **0.34–0.39 s** for each of structural,
convergence, and inference. These timings include R startup. Thus repeated timing
is a demonstrated source of the long runtime; there is no saved historical profile
to attribute every second of the earlier run. Dense curvature calculations can
also grow expensive with model size, but they did not dominate this small probe.

The ten-draw follow-ups completed 198 fits each in **26.9 s** (ordinary) and
**31.0 s** (stress), with all planned chart triples present. These two runs were
launched concurrently, so these are iteration wall budgets, not comparative speed
measurements. Some Bollen draws and stressed controlled draws approach singular
disturbance covariances; the largest paired objective gap was .348, and every gap
above $10^{-7}$ was rejected by the stationarity audit. The Bollen follow-up resolves its three difficult draws as matching,
cone-stationary PSD-boundary candidates; the other structures' boundary cases
remain open. They are retained in `results/ordinary` and `results/iteration`.

This experiment constructs each case once, validates first derivatives once per
chart, draws sample covariances directly, and fits each chart once per draw. It
prints and checkpoints each fit and stops between fits at a declared soft time
budget. It does not run repeated microbenchmarks or second-derivative curvature
calculations in the replication loop.

## Caveats

- Six selected structures and a few draws cannot establish a generally best chart.
  Population conditioning is deterministic; convergence frequency is not.
- All three charts use the same small R optimization prototype with analytic
  derivatives. Its timing includes R overhead and must not be used to claim a
  speedup in magmaan. Native backend comparisons follow only after choosing a
  promising map.
- Starts are perturbations of the controlled population or empirical reference fit,
  transformed exactly into each chart. This isolates coordinates but is
  deliberately reference-assisted. Native start heuristics need a separate arm.
- The original chart comparison admits only strictly positive-definite disturbance and measurement
  covariance matrices; the Bollen follow-up adds native PSD references. Boundary optima and unrestricted Heywood estimates are
  outside that interior arm. Rejected steps are counted; rejection-based optimization can
  struggle near a boundary and is not a proposed production constraint method.
- An optimizer success flag, a small gradient in raw coordinates, and an admissible
  covariance are distinct properties. The report keeps them separate. The
  information-metric gradient threshold is a pilot diagnostic, not a proof of
  optimality or the library's authoritative fit verdict.
- The initial models are recursive, single-group, continuous, and covariance-only,
  without shared parameter labels. Feedback, observed exogenous predictors,
  multi-group equalities, fixed loading growth models, and mean structures require
  additional mapping and identification checks before inclusion.
- The private textbook corpus is not mounted in this checkout. The two empirical
  cases use shared benchmark syntax and public lavaan datasets; the other four are
  controlled structures. No sibling experiment or test fixture is consumed.
- The time limit is checked between fits, not a hard process timeout. A single fit
  is bounded by its iteration limit and can overrun the remaining wall budget.

## Reproduce

From this directory:

```sh
Rscript run_experiment.R --help
Rscript run_experiment.R --dry-run-plan --smoke --stress
Rscript run_experiment.R --smoke --stress
Rscript scripts/bollen_probe.R --smoke
Rscript scripts/bollen_probe.R             # original data + ten draws, default 60s budget
quarto render report.qmd
# Ordinary setting, kept separate from the rendered stress pilot:
Rscript run_experiment.R --smoke --results-dir results/ordinary
# Expand only the number of paired draws; no repeated timing batches:
Rscript run_experiment.R --reps 10 --stress --budget-sec 60 --results-dir results/iteration
```

The runner writes `fits.csv`, `validation.csv`, `timing.csv`, and `metadata.csv`.
The seed for each model/draw is stable when selecting a subset of cases.
A budget-limited run exits with status 2 and flags incomplete chart triples.
Changing `--results-dir` keeps separate runs; rendering reads the top-level
`results/` run. Failures remain in the output instead of being silently dropped.

The isolated legacy runtime probe can be repeated from the repository root:

```sh
# Repeat for geometry, cost, backconvert, structural, convergence, inference.
/usr/bin/time -p Rscript experiments/82-latent-metric-geometry/run_experiment.R \
  --arms cost --p 6 --curv-p 6 --bc-p 6 --lambda1 0.3 --struct-b 0.9 \
  --conv-n 50 --reps 1 --backends nlopt-lbfgs --results-dir /tmp/latent-metric-cost
```

Run the native-start structural comparison and inspect any discrepant endpoints:

```sh
Rscript scripts/native_structure_probe.R --dry-run-plan
Rscript scripts/native_structure_probe.R
Rscript scripts/native_restart_probe.R
Rscript scripts/mediation_boundary_probe.R
```

The main comparison runs 220 fits under a 30-second soft budget, checked between
fits; each fit has a 1000-evaluation limit. `--smoke` runs 40 fits. The restart
runner requires saved discrepancies and runs four fits per discrepancy under its
own 30-second soft budget. The mediation diagnostic consumes the saved stressed
mediation draw-4 endpoints; it runs 42 fits (22 with `--smoke`) with six seeded
random perturbations per endpoint and a separate 30-second soft budget. Its
geometry checks and all restart results live in `results/mediation-boundary/`.
Native results are independent of the earlier R-chart
runs. Synthetic syntax is validated against its known population covariance
before fitting; validation discrepancies are below $2\times10^{-29}$ in ULS.

## Design and deferred alternatives

### Initial model selection

| Model | Distinctive feature | Stress change |
|---|---|---|
| Holzinger–Swineford | Three correlated CFA factors; control where disturbance and total scaling coincide | None; observed-data sentinel |
| Mediation | Direct and indirect paths through three factors | Mediator predictability about .94 |
| Correlated predictors | Two correlated latent causes of one latent outcome | Predictor correlation .95 |
| Higher-order | One higher-order cause of three measured first-order factors | Strong higher-order paths and small first-order disturbances |
| Correlated disturbances | Two endogenous factors with correlated innovations, plus correlated indicator errors | Innovation correlation .80 |
| Bollen political democracy | Empirical mediation with unequal observed scales and several correlated indicator errors | None; observed-data sentinel |

The controlled models all have nine observed variables. The higher-order model has
four latents rather than three; Bollen has eleven observed variables. This makes
structural differences central instead of disguising a size sweep as generality.
For each model, draw 0 uses its population or observed covariance. Additional draws
are Gaussian samples from its controlled or reference-fitted population, with the
same covariance draw shared across charts. The default sample size is 400 for
controlled cases and the original sample size for empirical cases.

### Three equivalent interior parameterizations

For $\eta=B\eta+\zeta$, let $A=(I-B)^{-1}$ and $C=A\Psi A^\top$.
The three conventions are marker loadings fixed to one, disturbance variances fixed
to one, and total latent variances fixed to one. Higher-order marker anchors are
loadings on lower-order factors. Latent signs use the same positive anchor
orientation at initialization.

Total scaling does not need a general equality-constrained optimizer for these
models. With $\Psi_{\mathrm{off}}$ containing only off-diagonal disturbances, solve

$$
(A\circ A)d=\mathbf 1-\operatorname{diag}
  (A\Psi_{\mathrm{off}}A^\top),\qquad
\Psi=\Psi_{\mathrm{off}}+\operatorname{diag}(d).
$$

Here $\circ$ denotes elementwise multiplication. This eliminates one disturbance
variance per latent while freeing its marker loading, preserving parameter count.
For recursive models the solve is nonsingular after a topological permutation.
The formula includes correlated disturbances; a shortcut such as $1-R^2$ without
the covariance terms would not suffice. Feedback models would additionally need
the invertibility of this elimination map checked, beyond invertibility of $I-B$.

All three maps retain the same structural zeros, the same observed covariance
likelihood, and the same positive-definite component domain. The full information
rank is checked; a singular matrix is never made to look well-conditioned by
silently dropping its smallest eigenvalues.

### Conditions for reopening

No further simulation or global-certificate project is required to close this
default decision. Reopen it if a supported downstream workload supplies a
reproducible, materially worse or unresolved unscaled PSD fit, or a measured
runtime bottleneck. Preserve that case and compare an explicit candidate against
the frozen baseline on fresh data. Count every failure and timeout, retain
original-model feasibility checks, and compare objective quality before speed.

A boundary-specific restart would additionally need graph and constraint
eligibility checks and evidence beyond the selected mediation dataset. A new
scaling default would need repeatable benefit from native starts across relevant
structures, including setup cost, without worse solution quality. Keep timing
batches separate from replications. These are promotion conditions, not queued
implementation tasks.

### Experimental parameterization algorithm, not adopted

The following prototype remains available for research. Given a supported model and a valid common starting point:

1. Construct equivalent marker, disturbance-unit, and total-unit coordinates.
   Reject unsupported constraints or a locally rank-deficient model explicitly.
2. Compute expected information $I_0$ in each chart. Compare
   $\operatorname{diag}(I_0)^{-1/2}I_0\operatorname{diag}(I_0)^{-1/2}$,
   so a chart does not win simply by changing individual parameter units.
3. Select the smallest condition number and hold that chart fixed. For
   $I_0=R^\top R$, optimize in $\theta=\theta_0+R^{-1}z$.
   The starting information in $z$ is identity. Diagonal scaling is a cheaper
   alternative and nearly as effective in the current Bollen run.
4. Preserve the requested PSD domain. An interior fast path needs a PSD fallback
   when it cannot pass its audit; an always-PSD solver can instead carry this
   scaling internally. A zero disturbance with positive total latent variance is
   representable in marker or total coordinates, but not by fixing that
   disturbance variance to one.
5. Classify the outcome using covariance feasibility and the appropriate interior
   or cone audit. Keep unresolved cases distinct from accepted boundary solutions.

`R/precondition.R` supplies the coordinate selector. The Bollen runner validates
its affine maps and replays the routing policy. These are separate from the
existing native marker-to-std.lv identification helper; no production policy has
changed. Full information scaling incurs a dense solve and is not necessarily the
best choice for large models. Its usefulness must be judged including setup cost.
In this R prototype, native PSD fits are already faster than the R optimization
loop, so the experiment does not establish a speed advantage over simply using
native PSD fitting throughout.

The prototype does not establish a universal mapping from a model diagram to
its best coordinates. It is retained as a reproducible alternative, with no
scheduled classifier or chart-selection work.

### Native PSD diagonal preconditioning

The complete-data native PSD ML fitter now exposes an optional frozen diagonal
information scale. It retains the existing model and covariance-factor links;
gradients and equality Jacobians transform together. Final feasibility and
cone-stationarity checks use the original coordinates. The default is unchanged.

```{r}
native_path <- "results/native/fits.csv"
if (file.exists(native_path)) {
  native <- read.csv(native_path)
  native_summary <- aggregate(cbind(evaluations, seconds) ~ start + preconditioning,
                              native, median)
  native_summary$milliseconds <- round(1000 * native_summary$seconds, 2)
  native_summary$seconds <- NULL
  knitr::kable(native_summary, caption = "Native Bollen fits: medians across eleven datasets.")
  native_pairs <- reshape(native[c("replicate", "start", "preconditioning", "objective")],
                          idvar=c("replicate", "start"), timevar="preconditioning",
                          direction="wide")
  cat(sprintf("\nAccepted fits: %d/%d. Maximum paired objective gap: %.2g.\n",
              sum(native$accepted), nrow(native),
              max(abs(native_pairs$objective.none-native_pairs$objective.diagonal))))
}
```

The first run accepted all 44 fits, including the same three boundary datasets,
and completed in 0.83 seconds. Common starts reduced median objective evaluations
from 108 to 31; native starts increased them from 114 to 152. Common starts here
are reference-assisted, so their improvement is not evidence for a default policy.
Single-pass timings include setup but are too short to establish a stable speed
ratio. Evaluation counts likewise omit constraint and derivative costs.

The implementation uses ordinary SEM information for the original parameter
block and lifted information for covariance-factor coordinates. Original
covariance coordinates have zero lifted-objective derivatives because they act
through the links; using those zeros as their scale would be inappropriate.
Zero-sensitivity coordinates retain unit scale, and nonzero inverse-square-root
scales are capped at $[10^{-4},10^4]$. No covariance eigenvalue floor is imposed
on the final solution by this scaling.

Reproduce with `Rscript scripts/native_psd_probe.R` after installing the updated
R bindings. It runs 44 fits with a 30-second soft budget and no timing batches;
`--smoke` runs 12 fits. Use
`frontier_fit_ml_psd(model, data, preconditioning = "diagonal")` to opt in.
The native-start comparison above covers all six structures and supports the
unscaled default. Loss of path identification at a zero mediator disturbance is
a known limitation of first-order acceptance at a single representation. It does
not justify installing the selected-case restart as a general default. Further
work is conditional on the reopening criteria above.
