---
title: "Li (2021) replication — mixed continuous/categorical SEM"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
input_path <- tryCatch(normalizePath(knitr::current_input(), mustWork = FALSE),
                       error = function(e) "")
candidate_dirs <- unique(c(getwd(), dirname(input_path)))
result_candidates <- unique(c(file.path(candidate_dirs, "results"),
                              candidate_dirs))
hits <- result_candidates[file.exists(file.path(result_candidates, "metadata.csv"))]
results_dir <- if (length(hits)) hits[[1L]] else file.path(getwd(), "results")
rd <- function(name) {
  p <- file.path(results_dir, name)
  if (!file.exists(p)) return(NULL)
  utils::read.csv(p, check.names = FALSE, stringsAsFactors = FALSE)
}
meta    <- rd("metadata.csv")
cells   <- rd("cells.csv")
thr     <- rd("thresholds.csv")
sparam  <- rd("summary_param.csv")
schi2   <- rd("summary_chi2.csv")
sconv   <- rd("summary_conv.csv")
parity  <- rd("lavaan_parity.csv")
if (is.null(meta) || is.null(sparam) || is.null(schi2)) {
  stop("Missing result CSVs. Run: Rscript run_experiment.R")
}
pct  <- function(x, d = 1) ifelse(is.na(x), "", sprintf(paste0("%.", d, "f"), 100 * x))
num3 <- function(x) ifelse(is.na(x), "", sprintf("%.3f", x))
fam_lab <- c(loading_cont = "λ continuous", loading_cat = "λ categorical",
             factor_cor = "φ inter-factor", gamma = "γ paths", beta = "β paths")
shape_lab <- c(sym = "symmetric", asym = "asymmetric")
model_lab <- c(cfa = "3-factor CFA", sem = "5-factor SEM")
```

This experiment replicates **Li, C.-H. (2021), "Statistical estimation of
structural equation models with a mixture of continuous and categorical
observed variables", *Behavior Research Methods* 53:2191-2213** against
magmaan. Li's design is the first to put continuous and categorical indicators
in the *same* model, so DWLS must combine polychoric (categorical-categorical),
polyserial (continuous-categorical) and Pearson (continuous-continuous)
associations in one weight matrix. Two estimators are fit on identical data per
replicate: **DWLS** (lavaan WLSMV; categoricals as categorical) and **MLR**
(robust ML treating the integer codes as continuous). The fitted model is
correctly specified, so χ² rejection rates are Type I error rates.

## Question

When a latent construct is measured by a *mix* of continuous and categorical
indicators, does DWLS or MLR recover loadings, inter-factor correlations and
structural paths better — and how does that depend on the number of response
categories, distribution shape, and sample size?

## Short answer

magmaan reproduces Li's central findings. **DWLS recovers categorical-indicator
loadings essentially unbiased; MLR underestimates them**, with the bias largest
at few categories and asymmetric thresholds and shrinking as categories or `N`
grow. The two estimators are comparable on continuous-indicator loadings,
inter-factor correlations and structural paths. DWLS's mean-and-variance
adjusted χ² controls Type I error close to nominal; MLR's robust χ² tends to
over-reject at small `N` / few categories.

Both estimators match lavaan (WLSMV / MLM `standardizedSolution`) on the same
data: MLR to ~1e-6, DWLS loadings/paths to ~1e-2 (see parity table).

## Design

```{r}
knitr::kable(meta, row.names = FALSE)
```

Two population models, both with standardized (unit-variance) latent responses
and zero mean structure (Li Fig 1):

- **3-factor CFA** (12 indicators). Continuous:categorical ratio varies by
  factor (3:1, 2:2, 1:3); inter-factor correlations all .3. This drives the
  "proportion of categorical" effect.
- **5-factor SEM** (20 indicators). Two exogenous, three endogenous factors,
  each 2 continuous + 2 categorical; structural paths `B`/`Γ` and inter-factor
  correlation φ₁₂ = .3.

Crossed with number of categories (2-7), distribution shape (symmetric /
slightly asymmetric, Li's exact thresholds), and `N` ∈ {200, 500, 1000}.
Parameters and SEs are reported in the standardized metric, split by family.

```{r}
if (!is.null(thr)) {
  t2 <- thr; t2$shape <- shape_lab[t2$shape]
  knitr::kable(t2, row.names = FALSE,
               col.names = c("Shape", "Categories", "Thresholds", "Proportions"))
}
```

## Evidence

### Parameter relative bias (%), by family

Averaged over replicates and same-family parameters; negative = underestimate.
Li's "trivial bias" band is |RB| < 5%.

```{r}
if (!is.null(sparam)) {
  d <- sparam
  d$family <- fam_lab[d$param_type]
  d$cell <- sprintf("%s, %s, K=%d, N=%d", model_lab[d$model],
                    shape_lab[d$shape], d$categories, d$N)
  agg <- d[, c("model", "shape", "categories", "N", "estimator",
               "param_type", "family", "rel_bias")]
  # Focus table: loadings split by type, collapsed over shape, at a few cells.
  foc <- agg[agg$param_type %in% c("loading_cont", "loading_cat") &
               agg$shape == "sym", ]
  w <- stats::reshape(
    foc[, c("model", "categories", "N", "estimator", "param_type", "rel_bias")],
    idvar = c("model", "categories", "N", "estimator"),
    timevar = "param_type", direction = "wide")
  names(w) <- sub("^rel_bias\\.", "", names(w))
  w$loading_cont <- pct(w$loading_cont); w$loading_cat <- pct(w$loading_cat)
  w$model <- model_lab[w$model]
  w <- w[order(w$model, w$N, w$categories, w$estimator), ]
  knitr::kable(w, row.names = FALSE,
               col.names = c("Model", "K", "N", "Est.",
                             "λ cont. RB%", "λ cat. RB%"),
               caption = "Symmetric thresholds. Full grid in results/summary_param.csv.")
}
```

The signature contrast: **λ categorical RB is near zero for DWLS but clearly
negative for MLR**, worst at K = 2-3, attenuating with K and N. λ continuous is
small for both.

### Standard-error relative bias and coverage

```{r}
if (!is.null(sparam)) {
  d <- sparam[sparam$shape == "sym" & sparam$N == 500, ]
  d$family <- fam_lab[d$param_type]
  d$se_rel_bias <- pct(d$se_rel_bias); d$coverage <- num3(d$coverage)
  d$rel_bias <- pct(d$rel_bias)
  d <- d[order(d$model, d$categories, d$estimator, d$param_type),
         c("model", "categories", "estimator", "family",
           "rel_bias", "se_rel_bias", "coverage")]
  d$model <- model_lab[d$model]
  knitr::kable(d, row.names = FALSE,
               col.names = c("Model", "K", "Est.", "Family",
                             "Param RB%", "SE RB%", "Coverage"),
               caption = "Symmetric, N = 500.")
}
```

### χ² rejection rate (Type I error at α = .05)

DWLS uses the mean-and-variance adjusted statistic (lavaan WLSMV); MLR the
Satorra-Bentler and mean-variance adjusted statistics. Bradley's band is
[.025, .075].

```{r}
if (!is.null(schi2)) {
  d <- schi2[schi2$shape == "sym", ]
  # DWLS-MV vs MLR-MV side by side.
  d <- d[(d$estimator == "DWLS" & d$stat_name == "MV") |
           (d$estimator == "MLR" & d$stat_name == "MV"), ]
  d$tag <- paste0(d$estimator, "-MV")
  w <- stats::reshape(d[, c("model", "categories", "N", "tag", "reject_05")],
                      idvar = c("model", "categories", "N"),
                      timevar = "tag", direction = "wide")
  names(w) <- sub("^reject_05\\.", "", names(w))
  w[] <- lapply(names(w), function(nm)
    if (nm %in% c("DWLS-MV", "MLR-MV")) num3(w[[nm]]) else w[[nm]])
  w$model <- model_lab[w$model]
  w <- w[order(w$model, w$N, w$categories), ]
  knitr::kable(w, row.names = FALSE,
               caption = "Symmetric thresholds. Empirical rejection at α = .05.")
}
```

### Convergence and improper solutions

```{r}
if (!is.null(sconv)) {
  d <- sconv
  d$conv_rate <- num3(d$conv_rate); d$improper_rate <- num3(d$improper_rate)
  d$model <- model_lab[d$model]; d$shape <- shape_lab[d$shape]
  bad <- d[as.numeric(d$conv_rate) < 1 | as.numeric(d$improper_rate) > 0, ]
  if (nrow(bad)) {
    knitr::kable(bad[order(bad$model, bad$N, bad$categories),
                     c("model", "categories", "shape", "N", "estimator",
                       "conv_rate", "improper_rate")],
                 row.names = FALSE,
                 col.names = c("Model", "K", "Shape", "N", "Est.",
                               "Conv.", "Improper"),
                 caption = "Cells with any nonconvergence or improper solutions.")
  } else {
    cat("All cells converged with no improper solutions.\n")
  }
}
```

### lavaan parity (rep 1 per cell)

Present only when run with `--lavaan-parity`. Max absolute difference between
magmaan and lavaan (`standardizedSolution`; WLSMV for DWLS, MLM for MLR) on the
same data.

```{r}
if (!is.null(parity) && nrow(parity)) {
  a <- aggregate(abs_diff ~ estimator + metric + model, parity, max)
  a$abs_diff <- sprintf("%.2e", a$abs_diff)
  a$model <- model_lab[a$model]
  knitr::kable(a[order(a$model, a$estimator, a$metric), ], row.names = FALSE,
               col.names = c("Est.", "Metric", "Model", "Max |diff|"))
} else {
  cat("(no parity table — run with `--lavaan-parity`)\n")
}
```

## Caveats

- Default `--reps` is small for iteration; rejection rates and biases are noisy
  until `--reps` is large (Li uses 1000). Per repo style, no Monte-Carlo error
  bars are reported.
- Estimates are standardized via magmaan's ordinal-aware `standardize_all`
  (lavaan `std.all`). For DWLS the categorical-indicator loadings use the
  unit-variance latent-response scale; structural paths and endogenous-factor
  loadings are rescaled by the model-implied latent SDs. This required a core
  change: standardization was previously not exposed for ordinal/mixed fits.
- DWLS's χ² parity uses the mean-and-variance adjusted statistic; magmaan's MV
  uses a Satterthwaite df, lavaan's WLSMV default is scaled-shifted, so the
  per-rep p-values differ by a parameterization (compared, not the statistics).
- The SEM at N = 200 stresses the mixed weight matrix; the full-WLS NACOV is
  often singular there. DWLS (diagonal weight) is unaffected — a core change
  makes a singular NACOV non-fatal so DWLS proceeds, matching lavaan WLSMV.

## Reproduce

```sh
# fast smoke (both models, a few cells)
Rscript run_experiment.R --reps 30 --smoke

# a single cell with the lavaan cross-check
Rscript run_experiment.R --reps 100 --cells "categories=5,N=500" --lavaan-parity

# the full paper-scale grid (slow)
Rscript run_experiment.R --reps 1000
```

`Rscript run_experiment.R --help` lists all options. Outputs land in
`results/`; this report reads only from there.
