---
title: "Ordinal Polychoric Omega Stress Probe"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- "results/simulation_summary.csv"
metadata_path <- "results/metadata.csv"
targets_path <- "results/pseudo_targets.csv"
if (!file.exists(summary_path) || !file.exists(metadata_path) ||
    !file.exists(targets_path)) {
  stop("Missing results. Run: Rscript run_experiment.R --smoke")
}
summary <- read.csv(summary_path)
metadata <- read.csv(metadata_path)
targets <- read.csv(targets_path)
worst <- summary[which.min(summary$coverage_pseudo_95), ]
```

## Question

When ordinal data are nonnormal or locally dependent, does the robust
delta-method interval for polychoric omega cover the pseudo-true polychoric
omega target?

## Short Answer

```{r, results='asis'}
cat(sprintf(
  "At %s reps per cell, pseudo-target coverage among successful draws ranges from %.3f to %.3f. The weakest cell is %s / %s at N = %d. The grid recorded %d construction failures.",
  metadata$reps, min(summary$coverage_pseudo_95), max(summary$coverage_pseudo_95),
  worst$dgp, worst$cuts, worst$n, sum(summary$failures)))
```

## Pseudo-Targets

```{r}
target_shown <- targets
target_shown[] <- lapply(target_shown, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})
knitr::kable(target_shown)
```

## Coverage

```{r}
shown <- summary[, c(
  "dgp", "cuts", "n", "failures", "target_pseudo",
  "pseudo_minus_continuous", "mean_value", "bias_pseudo", "sd_value",
  "mean_se", "median_se", "coverage_pseudo_95", "coverage_continuous_95",
  "mean_width_95", "median_width_95", "min_variable_category_count_p05"
)]
names(shown) <- c(
  "DGP", "Cuts", "N", "Failures", "Pseudo Target",
  "Pseudo - Continuous", "Mean Omega", "Bias", "SD", "Mean SE",
  "Median SE", "Coverage Pseudo", "Coverage Continuous", "Mean Width",
  "Median Width",
  "P05 Min Variable Category Count"
)
shown[] <- lapply(shown, function(x) if (is.numeric(x)) round(x, 4) else x)
knitr::kable(shown)
```

## Interpretation

The coverage target is the large-sample polychoric omega produced by the same
ordinal-statistics pipeline, not the continuous latent-response Pearson omega.
The continuous target is reported only as a bias diagnostic: when the polychoric
pseudo-target moves away from it, coverage against the continuous target is not
the relevant inferential criterion.

## Caveats

The pseudo-targets are Monte Carlo approximations from a large generated data set,
not analytic population roots of the polychoric likelihood. This probe still uses
the fit-free polychoric correlation-matrix coefficient; it does not test fitted
DWLS model-based omega, observed-score Green-Yang/Flora omega, profile-LR
inversion, or Bartlett calibration.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R
quarto render report.qmd
```
