---
title: "Can spectral estimation improve score and LRT calibration?"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Question

Can nonlinear shrinkage or cross-observation spectral moments improve on
pEBA-4 for global score and likelihood-ratio tests with non-normal data?

## Short answer

There is no consistent improvement over pEBA-4 in this 6,000-dataset null pilot.
For skewed factors and errors, nonlinear shrinkage rejects only 0.4–2.6% at
nominal 5%, across score/LRT and the four sample-size/model-size cells; pEBA-4
rejects 2.6–6.8%. Fourth-order direct-CGF rejection is 4.4–5.4% at n=400,
but the n=100, 12-indicator cell rejects 13.7% for LRT among usable calls and
fails on 15.2% of datasets. Sixth-order moments fail on 59.0% of datasets in
that same cell. The high-order moment route needs stabilization before it can
be a practical replacement. These are local research prototypes, not new
package defaults.

```{r}
path <- 'results/pilot500/summary.csv'
if(!file.exists(path)) stop('Run the pilot500 command under Reproduce first.')
s <- read.csv(path)
r <- read.csv('results/pilot500/replicates.csv')
```

## Evidence

The following plot shows rejection at nominal 5%, separately for each statistic
and distribution. Shape denotes sample size (circles: 100; triangles: 400). The complete CSV includes binomial intervals, method
failures, and paired differences from pEBA-4 with Monte Carlo standard errors.

```{r}
#| fig-width: 12
#| fig-height: 7
methods <- c('peba4','all','ans','mom4','mom6','cgf4','cgf6')
cols <- c('#222222','#999999','#0072B2','#009E73','#CC79A7','#D55E00','#E69F00')
par(mfrow=c(2,3),mar=c(4,4,3,1),oma=c(4,0,0,0))
for(kind in c('score','lrt')) for(dist in c('normal','skewed','heterogeneous')) {
  x <- s[s$statistic_kind==kind & s$distribution==dist,]
  plot(NA,xlim=c(0,.56),ylim=c(0,max(.15,x$rejection,na.rm=TRUE)),
       xlab='Degrees of freedom / sample size',ylab='Rejection proportion',
       main=paste(kind,dist))
  abline(h=.05,lty=2,col='grey50')
  for(j in seq_along(methods)) {
    z <- x[x$method==methods[j],]; z <- z[order(z$d_over_n),]
    lines(z$d_over_n,z$rejection,col=cols[j])
    points(z$d_over_n,z$rejection,col=cols[j],pch=ifelse(z$n==100,16,17))
  }
}
par(fig=c(0,1,0,.08),new=TRUE,mar=rep(0,4))
plot.new();legend('center',legend=methods,col=cols,lty=1,ncol=7,bty='n',cex=.8)
```

Failure rates are essential: a method that cannot decide is not counted as a
successful nonrejection. The table pools failures across design cells, not
rejection rates across different populations.

```{r}
f <- aggregate(cbind(failed,attempted)~method,s,sum)
f$`Failure (%)` <- round(100*f$failed/f$attempted,1)
knitr::kable(f[,c('method','failed','attempted','Failure (%)')],
             col.names=c('Method','Failed','Attempted','Failure (%)'))
```

## Caveats

This is a null-calibration pilot, not a power comparison. It uses one-factor
CFA with 8 or 12 indicators, loadings from .45 to .80, and unit population
observed variances. Independent latent factors and errors are either normal,
standardized chi-square(1), or heterogeneous: alternating normal and
chi-square components, with one standardized lognormal error (log SD .7).
All these distributions have finite moments of every order. The sample sizes
are 100 and 400, giving 20 or 54 degrees of freedom. These mechanisms preserve
the same null covariance and deliberately vary higher moments.

Five hundred replications give a Monte Carlo standard error near one percentage
point at a true 5% rate. Calibration alone does not establish superiority, and
these cells do not cover elliptical heavy tails, missing data, or nested tests.
No production inference policy changes.

Nonlinear shrinkage here means the analytical covariance-loss estimator of
[Ledoit and Wolf](https://www.econ.uzh.ch/apps/workingpapers/wp/econwp264.pdf),
used experimentally as mixture weights. It is **not** QuEST population-spectrum
recovery, and its usual random-matrix assumptions need not hold for transformed
quadratic observation contributions. Sample-size-dependent projection error is
also left in the primary comparison.

The moment methods use increasing-index cycles as in
[Kong and Valiant](https://arxiv.org/abs/1602.00061), after disjoint pair
differencing to remove unknown means. This halves the number of independent
rows. The cycle estimators are unbiased with an independently fixed projection;
the primary methods estimate that projection from the same data. The
`independent` diagnostic spends an **additional n observations** estimating the
reference moments with the original fitted projection held fixed. It is an
extra-data diagnostic, not an equal-cost competitor. Even there, the estimated
projection can differ from its population counterpart.

Orders 4 and 6 reconstruct a positive, equal-mass spectrum by constrained
relative-moment least squares, then use the existing numerical mixture tail.
This reconstruction is an approximation and can lose noisy higher moments;
its residual is saved per replication. Optimizer failures and nonpositive
moments remain explicit failures. Pair order is fixed by the reproducible draw,
not selected for favorable results.

The direct-CGF methods truncate the cumulant series at order 4 or 6 and use a
[Lugannani–Rice/Kuonen right-tail approximation](https://academic.oup.com/biomet/article-abstract/86/4/929/276253). A finite polynomial need not be a valid
CGF, so these are **experimental rejection rules**, not established p-values.
Below the estimated mean they record nonrejection with a missing p-value and
an explicit flag. Invalid right tails remain failures. `raw_saddle` separately
checks saddlepoint approximation using the unmodified positive spectrum.
A consistent resolvent estimator or tail-optimal shrinkage rule remains future
work; this pilot does not claim to implement either.

## Reproduce

From the repository root, with the installed `magmaanlab` package:

```sh
OPENBLAS_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/R/validate.R
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --reps 500 --output pilot500
quarto render experiments/research/49-spectral-tail-calibration/report.qmd
```

The output directory must be new. `--smoke` runs two replications per cell;
`--help` lists options. All fits and standard reference distributions use
`magmaanlab`; new estimators are local R research prototypes. Each replication
checks that projected rows reproduce the existing score and LRT spectra, and
that the independent-sample score transport agrees with the core on fitting
data. Results record package version and binary hash, source hashes, Git
revision and worktree status. This pilot uses the installed package without
rebuilding the repository's unrelated in-progress changes.
