---
title: "Can debiased spectral moments improve score and LRT calibration?"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Question

Can oracle population eigenvalues or a feasible second-moment constraint improve
expected-metric score and likelihood-ratio calibration? A compact independent-
generator (IG) study now tests this under both nearly flat and heterogeneous
spectra; a replay of the original latent-component pilot provides context.

## Short answer

**The IG follow-up changes the diagnosis: oracle calibration can work where
corrected MV fails.** In severe Cholesky IG with 12 indicators and n=500,
rejection is **4.2% for oracle score versus 9.8% for corrected MV**, and
**4.6% for oracle LRT versus 9.6% for corrected MV**, at nominal 5%.
pEBA-4 gives 8.2% and 8.4%. These comparisons use the same 498 usable datasets.
The oracle spectrum now ranges from 1 to 5.20, so this is a real test of
heterogeneous mixture weights.

**The lower-bound constraint is not enough.** It leaves those corrected-MV
rejection rates unchanged. Across the IG grid it remains close to SB, differing
on at most 1.0% of score decisions and 1.6% of LRT decisions per cell. In the
highlighted cell, the fitted score trace averages 92% of the oracle trace and
the corrected second moment only 61% of its population target. Fixed-transform
unbiasedness is not translating into an accurate feasible reference here.

The 10,000-dataset grid has 332 common fit/admissibility failures. In severe
Cholesky IG at n=100, 19–22% of fits fail, so rejection there is conditional
on substantial selection; at n=500 only 0.4–0.6% fail. Five hundred replications
per cell remain exploratory. The original latent-component generator had an
exactly flat oracle spectrum and also showed finite-sample statistic error;
its conclusions should not be generalized to IG. Expected weighting is retained,
and no package defaults change.

```{r}
path <- 'results/pilot500/summary.csv'
if(!file.exists(path)) stop('Run the pilot500 command under Reproduce first.')
s <- read.csv(path)
mv_path <- 'results/mv_pilot500/summary.csv'
if(!file.exists(mv_path)) stop('Run the --mv command under Reproduce first.')
mv <- read.csv(mv_path)
oracle_path <- 'results/oracle_pilot500/summary.csv'
if(!file.exists(oracle_path)) stop('Run --oracle --output oracle_pilot500 first.')
ora <- read.csv(oracle_path)
```

## Evidence

### Smaller IG follow-up

The IG grid uses the existing two-factor covariance model, 8 or 12 indicators,
and n=100 or 500. It has 500 replications per cell: 10,000 datasets across
normal data and four IG conditions. Moderate IG targets observed marginal
skewness 2 and excess kurtosis 7; severe IG targets 3 and 21. Each severity uses
both symmetric and Cholesky covariance roots with native Pearson generators.
These are the IG marginal targets used in the Foldnes showcase, on a smaller
model; this is not a replication of its exact five-factor population.

The symmetric root produces nearly flat oracle spectra (largest weight below
1.05). The Cholesky root produces weights up to 2.40 under moderate IG and
5.20 under severe IG. Changing the root preserves the specified covariance
and marginal moments but changes higher-order dependence. This therefore tests
more than simply making each individual variable more skewed.

```{r}
ig_path <- 'results/oracle_ig500/summary.csv'
if(!file.exists(ig_path)) stop('Run --mv --ig and its --oracle replay under Reproduce first.')
ig <- read.csv(ig_path)
```

```{r}
#| label: ig-calibration
#| fig-width: 16
#| fig-height: 8
#| fig-alt: "Score and LRT rejection under moderate and severe IG, with symmetric and Cholesky roots, comparing oracle, corrected MV, constrained MV, SB and pEBA-4."
ig_methods <- c('oracle_all','mv_tau2','mv_constrained','sb','peba4')
ig_colors <- c('#CC79A7','#D55E00','#0072B2','#999999','#222222')
ig_dists <- c('ig_moderate','ig_severe','ig_moderate_cholesky','ig_severe_cholesky')
ig_titles <- c('Symmetric: moderate','Symmetric: severe','Cholesky: moderate','Cholesky: severe')
ig_ylim <- c(0,max(.15,ig$rejection[ig$method%in%ig_methods],na.rm=TRUE)*1.1)
layout(rbind(matrix(1:8,2,byrow=TRUE),rep(9,4)),heights=c(1,1,.2))
par(mar=c(4,4,3,1))
for(kind in c('score','lrt')) for(k in seq_along(ig_dists)) {
  z <- ig[ig$statistic_kind==kind & ig$distribution==ig_dists[k],]
  plot(NA,xlim=c(.5,4.5),ylim=ig_ylim,xaxt='n',xlab='Indicators / sample size',
       ylab=paste(toupper(kind),'rejection'),main=ig_titles[k])
  axis(1,1:4,c('8/100','8/500','12/100','12/500'));abline(h=.05,lty=2,col='grey50')
  for(j in seq_along(ig_methods)) {
    a <- z[z$method==ig_methods[j],];a <- a[order(a$p,a$n),]
    points(1:4+(j-3)*.12,a$rejection,col=ig_colors[j],pch=j)
  }
}
par(mar=rep(0,4));plot.new()
legend('center',legend=c('Oracle spectrum','Corrected MV','Constrained MV','SB','pEBA-4'),
       col=ig_colors,pch=seq_along(ig_methods),ncol=5,bty='n')
```

The comparison uses 9,668 common usable fits, with no additional moment or
oracle-calibration failures. In severe Cholesky IG at n=500, oracle score
rejection is 4.2% for both dimensions; corrected MV gives 7.0% with 8 indicators
and 9.8% with 12. Oracle LRT gives 4.6% for both, versus 7.0% and 9.6% with
corrected MV. For the highlighted 12-indicator score cell, constraining MV still
rejects 1.6 percentage points more often than pEBA-4 (paired Monte Carlo SE
0.6 point). Thus the constraint is not a demonstrated replacement for pEBA-4.

In the 12-indicator severe-Cholesky n=500 cell, the empirical score mean is
95.4% of the oracle mean (Monte Carlo SE 1.2% of that mean), and its empirical
variance is 1.21 times the oracle variance. The first-moment estimate averages
91.9% of its target; the plug-in second moment averages 2.05 times its target,
while U4 averages 0.61 times its target. These diagnostics show errors on both
sides of the calibration problem, with opposite signs for plug-in and corrected
second moments. They do not identify fitted-projection bias separately from
all other finite-sample and fit-selection effects.

For symmetric-root IG, corrected score rejection ranges from 4.2% to 8.0%,
and the constraint reduces it to 3.7–5.6%. The much smaller oracle heterogeneity
in those cells makes their resemblance to the original pilot understandable.
Changing the covariance root changes the population target despite matching
marginal skewness and kurtosis, which is why the two roots are reported separately.

The additional numerical check uses 100,000 Gaussian quadratic forms for each
of ten population settings. Oracle critical values give rejection 4.873–5.121%,
including the nonflat mixtures. Population covariances, target marginal moments,
and projected spectra pass independent checks. The report retains all normal
and IG cell summaries, paired Monte Carlo errors, failures, bound frequencies,
and moment diagnostics in its CSV outputs.

### Oracle calibration in the original latent-component design

This is a paired replay of the original 24 cells (500 datasets each), with
expected information for both score weighting and nuisance projection. The
population covariance, model tangent, and fourth moments are calculated
analytically; no large fitted calibration sample or estimated population
spectrum is substituted. The fitted statistics themselves are unchanged.
All methods share the original 17 failed fits, leaving 11,983 usable datasets.

```{r}
#| label: oracle-calibration
#| fig-width: 12
#| fig-height: 8
#| fig-alt: "Oracle and estimated calibration for score and LRT in one-factor and two-factor skewed 12-indicator models."
omethods <- c('oracle_all','mv_tau2','mv_constrained','sb','peba4')
ocolors <- c('#CC79A7','#D55E00','#0072B2','#999999','#222222')
olabels <- c('Oracle population spectrum','Corrected MV','Constrained MV','SB','pEBA-4')
layout(rbind(matrix(1:4,2,byrow=TRUE),c(5,5)),heights=c(1,1,.25))
par(mar=c(4,4,3,1))
for(kind in c('score','lrt')) for(model in c('one_factor','two_factor')) {
  z <- ora[ora$statistic_kind==kind & ora$model==model & ora$p==12 &
             ora$distribution=='skewed',]
  plot(NA,xlim=c(.85,3.15),ylim=c(0,.16),xaxt='n',xlab='Sample size',
       ylab='Rejection proportion',main=paste(toupper(kind),gsub('_',' ',model),sep=': '))
  axis(1,1:3,c('100','200','500'));abline(h=.05,lty=2,col='grey50')
  for(j in seq_along(omethods)) {
    a <- z[z$method==omethods[j],];a <- a[order(a$n),]
    lines(1:3,a$rejection,col=ocolors[j],lty=j,pch=j,type='b')
  }
}
par(mar=rep(0,4));plot.new()
legend('center',legend=olabels,col=ocolors,lty=seq_along(omethods),
       pch=seq_along(omethods),ncol=3,bty='n')
```

The constrained score rates at n=100 improve on unconstrained corrected MV's
6.0–6.2%, but coincide with SB's rejection decisions in these two cells.
Across the full grid, the constraint binds on 62–91% of LRT fits, too. Detailed
paired differences and Monte Carlo errors are retained in the CSV outputs;
500 replications give about one percentage point of uncertainty at nominal 5%.
This replay is diagnostic rather than independent confirmation.

The oracle result exposes finite-sample errors beyond spectrum estimation.
For the skewed 12-indicator n=100 cells, the empirical score mean is 2.7–2.9%
above the oracle mean, whereas LRT is 6.8–7.8% above it. The Monte Carlo standard
errors of those means are about 1.0–1.1% of the oracle mean. Empirical variances
are also elevated: by 34–44% for score and 54–60% for LRT. These variance ratios
are descriptive estimates from 500 replications, not precise corrections.
A mean-only adjustment would therefore leave another discrepancy to investigate.

The estimated first moment averages close to the oracle trace, while the
plug-in second moment is roughly 3.7–4.1 times its population target in these
cells. U4 correction moves the second moment to 0.85–0.89 times that target.
This confirms that the correction removes substantial inflation but does not
remove all effects of fitting nuisance quantities. Dataset-specific estimated
calibration can also compensate for statistic fluctuations: being closer to the
population spectrum is not a guarantee of better finite-sample rejection.

The numerical benchmark generates 100,000 independent Gaussian quadratic forms
for each of eight population settings, with rejection 4.884–5.090%. Numerical
mixture tails agree with the chi-square reference within 4e-15 at checked
quantiles. Under normal data, oracle score rejection for 12 indicators at n=100
is 6.2–6.4%, while oracle LRT rejection is 10.2–11.8%. The issue is therefore
not confined to nonnormal data or to numerical evaluation of mixture tails.

### Earlier unconstrained MV pilot

The focused follow-up uses 12,000 null datasets: two CFA structures, 8 or 12
indicators, n=100/200/500, normal or skewed components, and 500 replications per
cell. Each dataset supplies both statistics and every comparison. Rejection
rates exclude failed calls; binomial intervals, paired Monte Carlo standard
errors against MV and pEBA-4, and moment estimates are in the CSV outputs.

```{r}
#| label: mv-calibration
#| fig-width: 14
#| fig-height: 8
#| fig-alt: "Eight panels compare corrected and standard MV with SB, full eigenvalues and pEBA-4 for score and LRT null rejection."
methods_mv <- c('sb','mv','mv_tau2','mv_both','all','peba4')
colors_mv <- c('#999999','#0072B2','#D55E00','#009E73','#CC79A7','#222222')
labels_mv <- c('SB','Standard MV','MV: corrected second moment',
               'MV: both moments corrected','Full eigenvalues','pEBA-4')
layout(rbind(matrix(1:8,nrow=2,byrow=TRUE),rep(9,4)),heights=c(1,1,.2))
par(mar=c(4,4,3,1))
for(kind in c('score','lrt')) for(model in c('one_factor','two_factor'))
  for(dist in c('normal','skewed')) {
    x <- mv[mv$statistic_kind==kind & mv$model==model & mv$distribution==dist &
              mv$method%in%methods_mv,]
    plot(NA,xlim=c(0,.56),ylim=c(0,max(.15,x$rejection,na.rm=TRUE)),
         xlab='Degrees of freedom / n',ylab=paste(toupper(kind),'rejection'),
         main=paste(if(model=='one_factor') 'One factor' else 'Two factors',dist,sep=': '))
    abline(h=.05,lty=2,col='grey50')
    for(j in seq_along(methods_mv)) {
      z <- x[x$method==methods_mv[j],];z <- z[order(z$d_over_n),]
      lines(z$d_over_n,z$rejection,col=colors_mv[j])
      points(z$d_over_n,z$rejection,col=colors_mv[j],pch=ifelse(z$p==8,16,17))
    }
  }
par(mar=rep(0,4));plot.new()
legend('center',legend=labels_mv,col=colors_mv,lty=1,ncol=3,bty='n',cex=.9)
```

For the one-factor, 12-indicator skewed cell at n=100, score correction raises
rejection by 5.4 percentage points relative to standard MV (paired Monte Carlo
SE 1.0 point). The corresponding LRT increase is 8.0 points (SE 1.2 points).
These are clear changes in rejection, but only the score change moves close
to nominal size. Both comparisons use the same fitted datasets.

Circles denote 8 indicators and triangles 12. The horizontal line is nominal
5%. These lines connect design cells, not a controlled experiment changing
only the dimension-to-sample-size ratio.

### Expected versus observed score weight

The original score uses expected Fisher information for both its weight and
nuisance projection. The paired extension changes only the weight to the
observed likelihood Hessian at the fitted null; nuisance projection stays
expected. The statistic, mixture spectrum, and corrected moments are all
recomputed for that weight. This is complete-data ML, not a missing-data FIML
experiment. Nonpositive observed metrics are recorded as failures, without repair.

```{r}
metric_path <- 'results/mv_metrics500/metric_pairs.csv'
if(!file.exists(metric_path)) stop('Run --mv --metrics --reps 500 --output mv_metrics500 first.')
mp <- read.csv(metric_path)
```

```{r}
#| label: metric-comparison
#| fig-width: 12
#| fig-height: 7
#| fig-alt: "Paired observed-minus-expected rejection differences with Monte Carlo intervals for MV, corrected MV, and pEBA-4."
metric_methods <- c('mv','mv_tau2','peba4')
metric_colors <- c('#0072B2','#D55E00','#222222')
layout(rbind(matrix(1:4,2,byrow=TRUE),c(5,5)),heights=c(1,1,.2))
par(mar=c(4,4,3,1))
shown <- mp[mp$method%in%metric_methods,]
metric_ylim <- range(0,100*(shown$difference-1.96*shown$mcse),
                       100*(shown$difference+1.96*shown$mcse),na.rm=TRUE)+c(-2,2)
for(model in c('one_factor','two_factor')) for(dist in c('normal','skewed')) {
  x <- mp[mp$model==model & mp$distribution==dist,]
  plot(NA,xlim=c(.5,6.5),ylim=metric_ylim,xaxt='n',
       xlab='Indicators / sample size',ylab='Observed minus expected (percentage points)',
       main=paste(if(model=='one_factor') 'One factor' else 'Two factors',dist,sep=': '))
  axis(1,1:6,labels=c('8/100','8/200','8/500','12/100','12/200','12/500'))
  abline(h=0,lty=2,col='grey50')
  for(j in seq_along(metric_methods)) {
    z <- x[x$method==metric_methods[j],];z <- z[order(z$p,z$n),]
    xx <- 1:6+(j-2)*.18;yy <- 100*z$difference;se <- 100*z$mcse
    segments(xx,yy-1.96*se,xx,yy+1.96*se,col=metric_colors[j])
    points(xx,yy,pch=16,col=metric_colors[j])
  }
}
par(mar=rep(0,4));plot.new()
legend('center',legend=c('Standard MV','MV: corrected second moment','pEBA-4'),
       col=metric_colors,pch=16,ncol=3,bty='n')
```

For the original skewed-component data with 12 indicators, observed-weight corrected MV still rejects
16.6–18.4% at n=200 and 8.8–11.0% at n=500. Ordinary MV with observed weighting
rejects 14.1–19.5%, 5.6–6.0%, and 3.4–4.0% respectively; pEBA-4 also becomes
liberal at small n. Thus changing the weight does not rescue the correction.
Normal data show the same small-sample problem: corrected MV rejects
36.9–45.7% with observed weighting for 12 indicators at n=100.

For the one-factor skewed 12-indicator n=100 cell, the corrected-MV increase is
36.3 percentage points (paired Monte Carlo SE 2.2 points). Of 12,000 attempted
datasets, 17 fail the common fit check and another 26 fail observed-metric
positive definiteness (all at n=100 with 12 indicators). Comparisons use the
11,957 datasets where both weights succeed. The expected-score and LRT results
reproduce the earlier run exactly. Changing the weight preserves the score
contributions and nuisance complement; direct quadratic and spectral checks
pass on every usable observed-metric fit. This is a replay of the original
simulation seeds, not independent confirmation.

Intervals show plus/minus 1.96 paired Monte Carlo standard errors on datasets
where both metrics succeed. Positive values mean more rejection with observed
weight; they do not by themselves mean better calibration. Full cell rates,
failures, and the both-moments correction are retained in the CSV outputs.

### Earlier spectral pilot

There is no consistent improvement over pEBA-4 in this 6,000-dataset null pilot.
For skewed factors and errors, nonlinear shrinkage rejects only 0.4–2.6% at
nominal 5%, across score/LRT and the four sample-size/model-size cells; pEBA-4
rejects 2.6–6.8%. Fourth-order direct-CGF rejection is 4.4–5.4% at n=400,
but the n=100, 12-indicator cell rejects 13.7% for LRT among usable calls and
fails on 15.2% of datasets. Sixth-order moments fail on 59.0% of datasets in
that same cell. The high-order moment route needs stabilization before it can
be a practical replacement. These are local research prototypes, not new
package defaults.

The following plot shows rejection at nominal 5%, separately for each statistic
and distribution. Shape denotes sample size (circles: 100; triangles: 400). The complete CSV includes binomial intervals, method
failures, and paired differences from pEBA-4 with Monte Carlo standard errors.

```{r}
#| fig-width: 12
#| fig-height: 7
methods <- c('peba4','all','ans','mom4','mom6','cgf4','cgf6')
cols <- c('#222222','#999999','#0072B2','#009E73','#CC79A7','#D55E00','#E69F00')
par(mfrow=c(2,3),mar=c(4,4,3,1),oma=c(4,0,0,0))
for(kind in c('score','lrt')) for(dist in c('normal','skewed','heterogeneous')) {
  x <- s[s$statistic_kind==kind & s$distribution==dist,]
  plot(NA,xlim=c(0,.56),ylim=c(0,max(.15,x$rejection,na.rm=TRUE)),
       xlab='Degrees of freedom / sample size',ylab='Rejection proportion',
       main=paste(kind,dist))
  abline(h=.05,lty=2,col='grey50')
  for(j in seq_along(methods)) {
    z <- x[x$method==methods[j],]; z <- z[order(z$d_over_n),]
    lines(z$d_over_n,z$rejection,col=cols[j])
    points(z$d_over_n,z$rejection,col=cols[j],pch=ifelse(z$n==100,16,17))
  }
}
par(fig=c(0,1,0,.08),new=TRUE,mar=rep(0,4))
plot.new();legend('center',legend=methods,col=cols,lty=1,ncol=7,bty='n',cex=.8)
```

Failure rates are essential: a method that cannot decide is not counted as a
successful nonrejection. The table pools failures across design cells, not
rejection rates across different populations.

```{r}
f <- aggregate(cbind(failed,attempted)~method,s,sum)
f$`Failure (%)` <- round(100*f$failed/f$attempted,1)
knitr::kable(f[,c('method','failed','attempted','Failure (%)')],
             col.names=c('Method','Failed','Attempted','Failure (%)'))
```

## Caveats

### What the oracle establishes in each generator

The independent latent components and independent errors have variance one
and excess kurtosis zero or twelve. Writing the generator as $X=Mz$, the
covariance of its covariance-moment vector is

$$
\Gamma=\Gamma_{\mathrm{normal}}+
\sum_r \kappa_r u_r u_r^\top,
\qquad u_r=\operatorname{vech}(m_r m_r^\top).
$$

For the original latent-component one-factor and two-factor generators, every $u_r$ lies
in the model's covariance tangent space. Projection removes the entire extra
term. The oracle mixture therefore has $d$ unit weights for both distributions.
The score and LRT share this asymptotic reference, although their finite-sample
laws differ. This is a property of this generator and model, not a general
claim that nonnormal SEM tests have chi-square limits.

The oracle checks include finite-difference tangent derivatives, an independent
moment-space residual projector, exact-covariance fitted CFA geometry, and
Monte Carlo checks of the nonnormal fourth moments before projection. The
oracle applies only to the 24-cell MV follow-up, not automatically to the earlier
heterogeneous-generator pilot. The Gaussian benchmark checks the reference
calculation, not the approximation of a fitted SEM statistic by that reference.

The new feasible version keeps the estimated first moment and uses

$$
\widehat\tau_{2,\mathrm{con}}=
\max\left(\widehat\tau_{2,U},\frac{\widehat\tau_1^2}{d}\right).
$$

When the bound binds, its Satterthwaite reference is exactly SB. Bound-active
p-values were checked against SB for every usable replication. The bound only
enforces this lower-moment constraint; it does not establish unbiasedness,
optimality, or exact finite-sample size. No Bartlett adjustment was fitted to
these same simulation results. Missing data and matched-null power remain
outside this experiment.

### What the debiased-MV estimator does

Let Y be the observation contributions after the fitted projection and metric
transformation. For four independent observations and a fixed transformation,

$$
\widehat\tau_{2,U} = \frac{1}{4(n)_4}
\sum_{i,j,k,l\;\mathrm{distinct}}
\{(Y_i-Y_j)^\top(Y_k-Y_l)\}^2,
\qquad E[\widehat\tau_{2,U}]=\operatorname{tr}(A^2).
$$

Here $(n)_4=n(n-1)(n-2)(n-3)$ and $A=\operatorname{Cov}(Y_i)$.
The two differences are independent, each with covariance $2A$ and zero mean,
so the squared inner product has expectation $4\operatorname{tr}(A^2)$.
Averaging all distinct quadruples removes arbitrary pairing and is exactly
translation invariant. Expanding the kernel reduces the calculation to a
cross-product matrix and sums of squared row lengths; no quadruple loop,
eigendecomposition, or leave-one-out fit is needed.

This uses established covariance-functional estimation machinery; see
[*A note on the unbiased estimator of Sigma squared*](https://www.sciencedirect.com/science/article/pii/S016771521730189X).
The conventional estimator targets the same population quantity consistently
at fixed dimension. Its defect here is finite-sample bias, not a different
asymptotic target.

The primary modification keeps each statistic's original first-moment estimate
and changes only the second moment. The second modification also replaces the
first moment with the trace of the sample covariance using denominator n−1.
Both feed the same Satterthwaite reference:

$$
a=\widehat\tau_2/\widehat\tau_1,\qquad
\nu=\widehat\tau_1^2/\widehat\tau_2,\qquad
p=P(\chi^2_\nu\ge T/a).
$$

The score reference originally uses uncentered contributions; the LRT reference
uses centered contributions. The U-statistic removes estimated contribution
means algebraically in either case. **It does not remove bias from estimating
the SEM projection or fitted covariance on the same observations.** A diagnostic
using an additional n independent observations holds that fitted transform
fixed and estimates only the second moment; it is not an equal-cost competitor.

The earlier unconstrained MV variants do not clip to force $\widehat\tau_2\ge\widehat\tau_1^2/d$.
Violations are recorded: such moments do not describe d nonnegative eigenvalues,
although the fractional-df gamma reference still exists. The simulation retains
these decisions rather than changing the estimator after seeing the results.
Failures and infeasibility fractions are in the summary CSV. The frequent
violations in this pilot are a reason to study moment feasibility and nuisance
estimation separately; they do not by themselves identify which effect is
responsible.

At fixed dimension, with consistent nuisance estimates and regularity ensuring
convergence of the empirical moment functionals, both versions converge to the
same first two spectral moments as ordinary MV. The continuous Satterthwaite
map therefore gives the same first-order MV reference. This is **not** a proof
of nominal asymptotic size for an arbitrary weighted mixture: two-moment gamma
matching remains an approximation. It is also not a growing-dimension theorem.
The fixed-transform unbiasedness identity is exact; extending it to a feasible
finite-sample SEM estimator is precisely the issue the simulation probes.

The follow-up uses one-factor and correlated two-factor CFA (factor correlation
.4), loadings .45–.80, and unit observed variances. Skewed components are
standardized chi-square(1); the second correlated factor is a standardized
linear combination of two such independent components. This preserves the
specified covariance. All conditions are under the null, with no power study.
The numerical validation includes brute-force quadruple averaging, translation,
rotation, permutation and scale identities, and normal/skewed Monte Carlo
checks at fixed projection. Five hundred replications give about a one-percentage-
point standard error at 5%, so modest differences remain uncertain.

### Scope of the earlier spectral pilot

This is a null-calibration pilot, not a power comparison. It uses one-factor
CFA with 8 or 12 indicators, loadings from .45 to .80, and unit population
observed variances. Independent latent factors and errors are either normal,
standardized chi-square(1), or heterogeneous: alternating normal and
chi-square components, with one standardized lognormal error (log SD .7).
All these distributions have finite moments of every order. The sample sizes
are 100 and 400, giving 20 or 54 degrees of freedom. These mechanisms preserve
the same null covariance and deliberately vary higher moments.

Five hundred replications give a Monte Carlo standard error near one percentage
point at a true 5% rate. Calibration alone does not establish superiority, and
these cells do not cover elliptical heavy tails, missing data, or nested tests.
No production inference policy changes.

Nonlinear shrinkage here means the analytical covariance-loss estimator of
[Ledoit and Wolf](https://www.econ.uzh.ch/apps/workingpapers/wp/econwp264.pdf),
used experimentally as mixture weights. It is **not** QuEST population-spectrum
recovery, and its usual random-matrix assumptions need not hold for transformed
quadratic observation contributions. Sample-size-dependent projection error is
also left in the primary comparison.

The moment methods use increasing-index cycles as in
[Kong and Valiant](https://arxiv.org/abs/1602.00061), after disjoint pair
differencing to remove unknown means. This halves the number of independent
rows. The cycle estimators are unbiased with an independently fixed projection;
the primary methods estimate that projection from the same data. The
`independent` diagnostic spends an **additional n observations** estimating the
reference moments with the original fitted projection held fixed. It is an
extra-data diagnostic, not an equal-cost competitor. Even there, the estimated
projection can differ from its population counterpart.

Orders 4 and 6 reconstruct a positive, equal-mass spectrum by constrained
relative-moment least squares, then use the existing numerical mixture tail.
This reconstruction is an approximation and can lose noisy higher moments;
its residual is saved per replication. Optimizer failures and nonpositive
moments remain explicit failures. Pair order is fixed by the reproducible draw,
not selected for favorable results.

The direct-CGF methods truncate the cumulant series at order 4 or 6 and use a
[Lugannani–Rice/Kuonen right-tail approximation](https://academic.oup.com/biomet/article-abstract/86/4/929/276253). A finite polynomial need not be a valid
CGF, so these are **experimental rejection rules**, not established p-values.
Below the estimated mean they record nonrejection with a missing p-value and
an explicit flag. Invalid right tails remain failures. `raw_saddle`{.literal-code} separately
checks saddlepoint approximation using the unmodified positive spectrum.
A consistent resolvent estimator or tail-optimal shrinkage rule remains future
work; this pilot does not claim to implement either.

## Reproduce

From the repository root, with the installed `magmaanlab` package, the focused follow-up is:

```{=html}
<pre><code class="literal-code">OPENBLAS_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/R/validate_mv.R
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --mv --reps 500 --output mv_pilot500
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --mv --metrics --reps 500 --output mv_metrics500
quarto render experiments/research/49-spectral-tail-calibration/report.qmd</code></pre>
```

The smaller IG follow-up and its analytic-oracle replay are:

```{=html}
<pre><code class="literal-code">TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --mv --ig --reps 500 --output mv_ig500
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --oracle --input mv_ig500 --output oracle_ig500
quarto render experiments/research/49-spectral-tail-calibration/report.qmd</code></pre>
```

The IG grid fixes expected weighting and omits the earlier extra-calibration-sample
diagnostic. Generator calibration is reused across sample sizes. All inference
methods are paired within each cell; the two root conditions use separate datasets.

After the focused follow-up has produced its saved statistics, the oracle replay is:

```{=html}
<pre><code class="literal-code">OPENBLAS_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/R/validate_oracle.R
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --oracle --input mv_pilot500 --output oracle_pilot500
quarto render experiments/research/49-spectral-tail-calibration/report.qmd</code></pre>
```

The replay records hashes of its input results and their metadata, along with
its own source and installed-package provenance. It adds no new fitted datasets;
its Gaussian benchmark has a separate seed and defaults to 100,000 draws per
population setting. `--oracle --smoke` uses two saved replications per cell and
5,000 benchmark draws. Outputs include population spectra, Gaussian benchmark
checks, paired calibration comparisons, bound frequencies, and fitted-statistic
moment diagnostics.

The earlier pilot is reproduced with:

```{=html}
<pre><code class="literal-code">OPENBLAS_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/R/validate.R
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --reps 500 --output pilot500
quarto render experiments/research/49-spectral-tail-calibration/report.qmd</code></pre>
```

The output directory must be new. `--smoke` runs two replications per cell;
`--help` lists options. All fits and standard reference distributions use
`magmaanlab`; new estimators are local R research prototypes. Each replication
checks that projected rows reproduce the existing score and LRT spectra, and
that the independent-sample score transport agrees with the core on fitting
data. Results record package version and binary hash, source hashes, Git
revision and worktree status. This pilot uses the installed package without
rebuilding the repository's unrelated in-progress changes.
