---
title: "Can debiased spectral moments improve score and LRT calibration?"
format:
  html:
    toc: true
    embed-resources: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Question

Does replacing the plug-in second spectral moment improve mean–variance (MV)
correction for global score and likelihood-ratio tests? The focused follow-up
compares it with ordinary MV, SB, full eigenvalues, and pEBA-4 under normal
and non-normal data.

## Short answer

**The score used expected information. Switching its weight to the observed
Hessian makes corrected MV substantially too liberal in this pilot.** In the
12-indicator skewed cells at n=100, corrected MV rejects **39.1–41.3%** with
observed weighting, versus **4.9–5.6%** with expected weighting on the same
usable datasets (nominal 5%). The observed metric also fails on 26 additional
datasets. Expected weighting remains the better candidate here; this result
concerns the weight alone, with nuisance projection held fixed.

With expected weighting, correcting only the second moment substantially reduces MV's conservatism for
score tests, but is not a general fix for LRT. Across the six skewed,
12-indicator cells, score rejection moves from **0.4–1.8% to 3.2–6.2%** at
nominal 5%; pEBA-4 gives 2.6–3.6%. For LRT at n=100, corrected MV overshoots
to **9.0–10.6%**, versus 5.8–7.0% for pEBA-4. Correcting both moments changes
little. The U-statistic has no moment-estimation failures on the 11,983 usable
fits; 17 of 12,000 datasets fail the common fitting/admissibility check.

This is more promising than the earlier high-order approach, especially for
score tests. However, in the skewed 12-indicator score cells, 86.0–93.0% of
corrected moment pairs cannot represent a positive spectrum with the nominal
degrees of freedom. Improved rejection rates alone do not justify a default
change or a claim that the fitted-projection bias has been removed.

```{r}
path <- 'results/pilot500/summary.csv'
if(!file.exists(path)) stop('Run the pilot500 command under Reproduce first.')
s <- read.csv(path)
mv_path <- 'results/mv_pilot500/summary.csv'
if(!file.exists(mv_path)) stop('Run the --mv command under Reproduce first.')
mv <- read.csv(mv_path)
```

## Evidence

The focused follow-up uses 12,000 null datasets: two CFA structures, 8 or 12
indicators, n=100/200/500, normal or skewed components, and 500 replications per
cell. Each dataset supplies both statistics and every comparison. Rejection
rates exclude failed calls; binomial intervals, paired Monte Carlo standard
errors against MV and pEBA-4, and moment estimates are in the CSV outputs.

```{r}
#| label: mv-calibration
#| fig-width: 14
#| fig-height: 8
#| fig-alt: "Eight panels compare corrected and standard MV with SB, full eigenvalues and pEBA-4 for score and LRT null rejection."
methods_mv <- c('sb','mv','mv_tau2','mv_both','all','peba4')
colors_mv <- c('#999999','#0072B2','#D55E00','#009E73','#CC79A7','#222222')
labels_mv <- c('SB','Standard MV','MV: corrected second moment',
               'MV: both moments corrected','Full eigenvalues','pEBA-4')
layout(rbind(matrix(1:8,nrow=2,byrow=TRUE),rep(9,4)),heights=c(1,1,.2))
par(mar=c(4,4,3,1))
for(kind in c('score','lrt')) for(model in c('one_factor','two_factor'))
  for(dist in c('normal','skewed')) {
    x <- mv[mv$statistic_kind==kind & mv$model==model & mv$distribution==dist &
              mv$method%in%methods_mv,]
    plot(NA,xlim=c(0,.56),ylim=c(0,max(.15,x$rejection,na.rm=TRUE)),
         xlab='Degrees of freedom / n',ylab=paste(toupper(kind),'rejection'),
         main=paste(if(model=='one_factor') 'One factor' else 'Two factors',dist,sep=': '))
    abline(h=.05,lty=2,col='grey50')
    for(j in seq_along(methods_mv)) {
      z <- x[x$method==methods_mv[j],];z <- z[order(z$d_over_n),]
      lines(z$d_over_n,z$rejection,col=colors_mv[j])
      points(z$d_over_n,z$rejection,col=colors_mv[j],pch=ifelse(z$p==8,16,17))
    }
  }
par(mar=rep(0,4));plot.new()
legend('center',legend=labels_mv,col=colors_mv,lty=1,ncol=3,bty='n',cex=.9)
```

For the one-factor, 12-indicator skewed cell at n=100, score correction raises
rejection by 5.4 percentage points relative to standard MV (paired Monte Carlo
SE 1.0 point). The corresponding LRT increase is 8.0 points (SE 1.2 points).
These are clear changes in rejection, but only the score change moves close
to nominal size. Both comparisons use the same fitted datasets.

Circles denote 8 indicators and triangles 12. The horizontal line is nominal
5%. These lines connect design cells, not a controlled experiment changing
only the dimension-to-sample-size ratio.

### Expected versus observed score weight

The original score uses expected Fisher information for both its weight and
nuisance projection. The paired extension changes only the weight to the
observed likelihood Hessian at the fitted null; nuisance projection stays
expected. The statistic, mixture spectrum, and corrected moments are all
recomputed for that weight. This is complete-data ML, not a missing-data FIML
experiment. Nonpositive observed metrics are recorded as failures, without repair.

```{r}
metric_path <- 'results/mv_metrics500/metric_pairs.csv'
if(!file.exists(metric_path)) stop('Run --mv --metrics --reps 500 --output mv_metrics500 first.')
mp <- read.csv(metric_path)
```

```{r}
#| label: metric-comparison
#| fig-width: 12
#| fig-height: 7
#| fig-alt: "Paired observed-minus-expected rejection differences with Monte Carlo intervals for MV, corrected MV, and pEBA-4."
metric_methods <- c('mv','mv_tau2','peba4')
metric_colors <- c('#0072B2','#D55E00','#222222')
layout(rbind(matrix(1:4,2,byrow=TRUE),c(5,5)),heights=c(1,1,.2))
par(mar=c(4,4,3,1))
shown <- mp[mp$method%in%metric_methods,]
metric_ylim <- range(0,100*(shown$difference-1.96*shown$mcse),
                       100*(shown$difference+1.96*shown$mcse),na.rm=TRUE)+c(-2,2)
for(model in c('one_factor','two_factor')) for(dist in c('normal','skewed')) {
  x <- mp[mp$model==model & mp$distribution==dist,]
  plot(NA,xlim=c(.5,6.5),ylim=metric_ylim,xaxt='n',
       xlab='Indicators / sample size',ylab='Observed minus expected (percentage points)',
       main=paste(if(model=='one_factor') 'One factor' else 'Two factors',dist,sep=': '))
  axis(1,1:6,labels=c('8/100','8/200','8/500','12/100','12/200','12/500'))
  abline(h=0,lty=2,col='grey50')
  for(j in seq_along(metric_methods)) {
    z <- x[x$method==metric_methods[j],];z <- z[order(z$p,z$n),]
    xx <- 1:6+(j-2)*.18;yy <- 100*z$difference;se <- 100*z$mcse
    segments(xx,yy-1.96*se,xx,yy+1.96*se,col=metric_colors[j])
    points(xx,yy,pch=16,col=metric_colors[j])
  }
}
par(mar=rep(0,4));plot.new()
legend('center',legend=c('Standard MV','MV: corrected second moment','pEBA-4'),
       col=metric_colors,pch=16,ncol=3,bty='n')
```

For skewed data with 12 indicators, observed-weight corrected MV still rejects
16.6–18.4% at n=200 and 8.8–11.0% at n=500. Ordinary MV with observed weighting
rejects 14.1–19.5%, 5.6–6.0%, and 3.4–4.0% respectively; pEBA-4 also becomes
liberal at small n. Thus changing the weight does not rescue the correction.
Normal data show the same small-sample problem: corrected MV rejects
36.9–45.7% with observed weighting for 12 indicators at n=100.

For the one-factor skewed 12-indicator n=100 cell, the corrected-MV increase is
36.3 percentage points (paired Monte Carlo SE 2.2 points). Of 12,000 attempted
datasets, 17 fail the common fit check and another 26 fail observed-metric
positive definiteness (all at n=100 with 12 indicators). Comparisons use the
11,957 datasets where both weights succeed. The expected-score and LRT results
reproduce the earlier run exactly. Changing the weight preserves the score
contributions and nuisance complement; direct quadratic and spectral checks
pass on every usable observed-metric fit. This is a replay of the original
simulation seeds, not independent confirmation.

Intervals show plus/minus 1.96 paired Monte Carlo standard errors on datasets
where both metrics succeed. Positive values mean more rejection with observed
weight; they do not by themselves mean better calibration. Full cell rates,
failures, and the both-moments correction are retained in the CSV outputs.

### Earlier spectral pilot

There is no consistent improvement over pEBA-4 in this 6,000-dataset null pilot.
For skewed factors and errors, nonlinear shrinkage rejects only 0.4–2.6% at
nominal 5%, across score/LRT and the four sample-size/model-size cells; pEBA-4
rejects 2.6–6.8%. Fourth-order direct-CGF rejection is 4.4–5.4% at n=400,
but the n=100, 12-indicator cell rejects 13.7% for LRT among usable calls and
fails on 15.2% of datasets. Sixth-order moments fail on 59.0% of datasets in
that same cell. The high-order moment route needs stabilization before it can
be a practical replacement. These are local research prototypes, not new
package defaults.

The following plot shows rejection at nominal 5%, separately for each statistic
and distribution. Shape denotes sample size (circles: 100; triangles: 400). The complete CSV includes binomial intervals, method
failures, and paired differences from pEBA-4 with Monte Carlo standard errors.

```{r}
#| fig-width: 12
#| fig-height: 7
methods <- c('peba4','all','ans','mom4','mom6','cgf4','cgf6')
cols <- c('#222222','#999999','#0072B2','#009E73','#CC79A7','#D55E00','#E69F00')
par(mfrow=c(2,3),mar=c(4,4,3,1),oma=c(4,0,0,0))
for(kind in c('score','lrt')) for(dist in c('normal','skewed','heterogeneous')) {
  x <- s[s$statistic_kind==kind & s$distribution==dist,]
  plot(NA,xlim=c(0,.56),ylim=c(0,max(.15,x$rejection,na.rm=TRUE)),
       xlab='Degrees of freedom / sample size',ylab='Rejection proportion',
       main=paste(kind,dist))
  abline(h=.05,lty=2,col='grey50')
  for(j in seq_along(methods)) {
    z <- x[x$method==methods[j],]; z <- z[order(z$d_over_n),]
    lines(z$d_over_n,z$rejection,col=cols[j])
    points(z$d_over_n,z$rejection,col=cols[j],pch=ifelse(z$n==100,16,17))
  }
}
par(fig=c(0,1,0,.08),new=TRUE,mar=rep(0,4))
plot.new();legend('center',legend=methods,col=cols,lty=1,ncol=7,bty='n',cex=.8)
```

Failure rates are essential: a method that cannot decide is not counted as a
successful nonrejection. The table pools failures across design cells, not
rejection rates across different populations.

```{r}
f <- aggregate(cbind(failed,attempted)~method,s,sum)
f$`Failure (%)` <- round(100*f$failed/f$attempted,1)
knitr::kable(f[,c('method','failed','attempted','Failure (%)')],
             col.names=c('Method','Failed','Attempted','Failure (%)'))
```

## Caveats

### What the debiased-MV estimator does

Let Y be the observation contributions after the fitted projection and metric
transformation. For four independent observations and a fixed transformation,

$$
\widehat\tau_{2,U} = \frac{1}{4(n)_4}
\sum_{i,j,k,l\;\mathrm{distinct}}
\{(Y_i-Y_j)^\top(Y_k-Y_l)\}^2,
\qquad E[\widehat\tau_{2,U}]=\operatorname{tr}(A^2).
$$

Here $(n)_4=n(n-1)(n-2)(n-3)$ and $A=\operatorname{Cov}(Y_i)$.
The two differences are independent, each with covariance $2A$ and zero mean,
so the squared inner product has expectation $4\operatorname{tr}(A^2)$.
Averaging all distinct quadruples removes arbitrary pairing and is exactly
translation invariant. Expanding the kernel reduces the calculation to a
cross-product matrix and sums of squared row lengths; no quadruple loop,
eigendecomposition, or leave-one-out fit is needed.

This uses established covariance-functional estimation machinery; see
[*A note on the unbiased estimator of Sigma squared*](https://www.sciencedirect.com/science/article/pii/S016771521730189X).
The conventional estimator targets the same population quantity consistently
at fixed dimension. Its defect here is finite-sample bias, not a different
asymptotic target.

The primary modification keeps each statistic's original first-moment estimate
and changes only the second moment. The second modification also replaces the
first moment with the trace of the sample covariance using denominator n−1.
Both feed the same Satterthwaite reference:

$$
a=\widehat\tau_2/\widehat\tau_1,\qquad
\nu=\widehat\tau_1^2/\widehat\tau_2,\qquad
p=P(\chi^2_\nu\ge T/a).
$$

The score reference originally uses uncentered contributions; the LRT reference
uses centered contributions. The U-statistic removes estimated contribution
means algebraically in either case. **It does not remove bias from estimating
the SEM projection or fitted covariance on the same observations.** A diagnostic
using an additional n independent observations holds that fitted transform
fixed and estimates only the second moment; it is not an equal-cost competitor.

There is no clipping to force $\widehat\tau_2\ge\widehat\tau_1^2/d$.
Violations are recorded: such moments do not describe d nonnegative eigenvalues,
although the fractional-df gamma reference still exists. The simulation retains
these decisions rather than changing the estimator after seeing the results.
Failures and infeasibility fractions are in the summary CSV. The frequent
violations in this pilot are a reason to study moment feasibility and nuisance
estimation separately; they do not by themselves identify which effect is
responsible.

At fixed dimension, with consistent nuisance estimates and regularity ensuring
convergence of the empirical moment functionals, both versions converge to the
same first two spectral moments as ordinary MV. The continuous Satterthwaite
map therefore gives the same first-order MV reference. This is **not** a proof
of nominal asymptotic size for an arbitrary weighted mixture: two-moment gamma
matching remains an approximation. It is also not a growing-dimension theorem.
The fixed-transform unbiasedness identity is exact; extending it to a feasible
finite-sample SEM estimator is precisely the issue the simulation probes.

The follow-up uses one-factor and correlated two-factor CFA (factor correlation
.4), loadings .45–.80, and unit observed variances. Skewed components are
standardized chi-square(1); the second correlated factor is a standardized
linear combination of two such independent components. This preserves the
specified covariance. All conditions are under the null, with no power study.
The numerical validation includes brute-force quadruple averaging, translation,
rotation, permutation and scale identities, and normal/skewed Monte Carlo
checks at fixed projection. Five hundred replications give about a one-percentage-
point standard error at 5%, so modest differences remain uncertain.

### Scope of the earlier spectral pilot

This is a null-calibration pilot, not a power comparison. It uses one-factor
CFA with 8 or 12 indicators, loadings from .45 to .80, and unit population
observed variances. Independent latent factors and errors are either normal,
standardized chi-square(1), or heterogeneous: alternating normal and
chi-square components, with one standardized lognormal error (log SD .7).
All these distributions have finite moments of every order. The sample sizes
are 100 and 400, giving 20 or 54 degrees of freedom. These mechanisms preserve
the same null covariance and deliberately vary higher moments.

Five hundred replications give a Monte Carlo standard error near one percentage
point at a true 5% rate. Calibration alone does not establish superiority, and
these cells do not cover elliptical heavy tails, missing data, or nested tests.
No production inference policy changes.

Nonlinear shrinkage here means the analytical covariance-loss estimator of
[Ledoit and Wolf](https://www.econ.uzh.ch/apps/workingpapers/wp/econwp264.pdf),
used experimentally as mixture weights. It is **not** QuEST population-spectrum
recovery, and its usual random-matrix assumptions need not hold for transformed
quadratic observation contributions. Sample-size-dependent projection error is
also left in the primary comparison.

The moment methods use increasing-index cycles as in
[Kong and Valiant](https://arxiv.org/abs/1602.00061), after disjoint pair
differencing to remove unknown means. This halves the number of independent
rows. The cycle estimators are unbiased with an independently fixed projection;
the primary methods estimate that projection from the same data. The
`independent` diagnostic spends an **additional n observations** estimating the
reference moments with the original fitted projection held fixed. It is an
extra-data diagnostic, not an equal-cost competitor. Even there, the estimated
projection can differ from its population counterpart.

Orders 4 and 6 reconstruct a positive, equal-mass spectrum by constrained
relative-moment least squares, then use the existing numerical mixture tail.
This reconstruction is an approximation and can lose noisy higher moments;
its residual is saved per replication. Optimizer failures and nonpositive
moments remain explicit failures. Pair order is fixed by the reproducible draw,
not selected for favorable results.

The direct-CGF methods truncate the cumulant series at order 4 or 6 and use a
[Lugannani–Rice/Kuonen right-tail approximation](https://academic.oup.com/biomet/article-abstract/86/4/929/276253). A finite polynomial need not be a valid
CGF, so these are **experimental rejection rules**, not established p-values.
Below the estimated mean they record nonrejection with a missing p-value and
an explicit flag. Invalid right tails remain failures. `raw_saddle`{.literal-code} separately
checks saddlepoint approximation using the unmodified positive spectrum.
A consistent resolvent estimator or tail-optimal shrinkage rule remains future
work; this pilot does not claim to implement either.

## Reproduce

From the repository root, with the installed `magmaanlab` package, the focused follow-up is:

```{=html}
<pre><code class="literal-code">OPENBLAS_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/R/validate_mv.R
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --mv --reps 500 --output mv_pilot500
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --mv --metrics --reps 500 --output mv_metrics500
quarto render experiments/research/49-spectral-tail-calibration/report.qmd</code></pre>
```

The earlier pilot is reproduced with:

```{=html}
<pre><code class="literal-code">OPENBLAS_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/R/validate.R
TZ=Europe/Oslo OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript experiments/research/49-spectral-tail-calibration/run_experiment.R --reps 500 --output pilot500
quarto render experiments/research/49-spectral-tail-calibration/report.qmd</code></pre>
```

The output directory must be new. `--smoke` runs two replications per cell;
`--help` lists options. All fits and standard reference distributions use
`magmaanlab`; new estimators are local R research prototypes. Each replication
checks that projected rows reproduce the existing score and LRT spectra, and
that the independent-sample score transport agrees with the core on fitting
data. Results record package version and binary hash, source hashes, Git
revision and worktree status. This pilot uses the installed package without
rebuilding the repository's unrelated in-progress changes.
