---
title: "Closed-form measurement invariance: is it calibrated and powered?"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
csv <- file.path(results_dir, "invariance_rejection.csv")
if (!file.exists(csv)) {
  stop("No results. Run:\n  Rscript run_experiment.R --full --reps 1000 --generators normal,ig")
}
res <- read.csv(csv, check.names = FALSE, stringsAsFactors = FALSE)
meta <- tryCatch(read.csv(file.path(results_dir, "metadata.csv"), stringsAsFactors = FALSE),
                 error = function(e) NULL)
alpha <- tryCatch(as.numeric(meta$value[meta$key == "alpha"]), error = function(e) 0.05)
if (length(alpha) != 1 || is.na(alpha)) alpha <- 0.05
fmt <- function(x) formatC(x, format = "f", digits = 3)
pick <- function(gen, scen, n, col) {
  r <- res[res$generator == gen & res$scenario == scen & res$n == n, ]
  if (!nrow(r)) NA_real_ else r[[col]][1]
}
gens <- unique(res$generator); ns <- sort(unique(res$n))
```

## Question

The closed-form Guttman CFA estimator gained two measurement-invariance tests: a
**metric** test (an exact-$\chi^2$ Wald from the minimum-distance projection onto
equal loadings) and a **scalar** test (a linearized Wald from the reference-group
mean map, which frees the latent means). Are they calibrated near the nominal
level when invariance holds, do they have power when it fails, and how do they
compare with the maximum-likelihood likelihood-ratio test, on normal and on
non-normal continuous data?

## Short answer

Both tests work, and on non-normal data the empirical fourth-moment matrix earns
its keep. On **normal data** the metric Wald and the ML likelihood-ratio test are
interchangeable: Type-I `r fmt(pick("normal","invariant",250,"rej_cf_metric_nt"))`
(closed form) versus `r fmt(pick("normal","invariant",250,"rej_ml_metric_nt"))`
(ML) at $n=250$ per group, and matched power against a broken loading
(`r fmt(pick("normal","metric_viol",250,"rej_cf_metric_nt"))` versus
`r fmt(pick("normal","metric_viol",250,"rej_ml_metric_nt"))`). The **scalar Wald
is exactly nominal on normal data** (`r fmt(pick("normal","invariant",250,"rej_cf_scalar_nt"))`
at $n=250$) with full power against an intercept shifted off the loading space
(`r fmt(pick("normal","scalar_viol",250,"rej_cf_scalar_nt"))`), and it delivers
true free-latent-mean scalar invariance in one closed-form step, which the ML
nested test cannot do here at all (freeing the latent means changes the
parameterization).

On **non-normal data the metric Wald needs the empirical $\Gamma$**: the
normal-theory version over-rejects at
`r fmt(pick("ig","invariant",250,"rej_cf_metric_nt"))`, and the empirical version
pulls it back to `r fmt(pick("ig","invariant",250,"rej_cf_metric_emp"))`. The ML
test shows the identical split, `r fmt(pick("ig","invariant",250,"rej_ml_metric_nt"))`
normal-theory versus `r fmt(pick("ig","invariant",250,"rej_ml_metric_rob"))`
robust. The **scalar Wald is far more robust** to non-normality, sitting at
`r fmt(pick("ig","invariant",250,"rej_cf_scalar_nt"))` (normal-theory) and
`r fmt(pick("ig","invariant",250,"rej_cf_scalar_emp"))` (empirical). Power is
retained throughout. The lesson matches the single-group goodness-of-fit study:
distribution-free fourth moments are what keep the closed-form inference honest
off normality.

## Evidence

### Metric invariance: the empirical $\Gamma$ under non-normality

Type-I under `invariant` for the closed-form Wald (normal-theory vs empirical
$\Gamma$) beside the ML LRT (normal-theory vs robust). On normal data every
column sits at $\alpha = `r alpha`$. On the non-normal generator the
normal-theory columns inflate and the distribution-free columns restore the
level. Power (rightmost, closed form, empirical $\Gamma$) is retained.

```{r metric-table}
mk_metric <- function() {
  rows <- list()
  for (g in gens) for (n in ns) {
    rows[[paste(g, n)]] <- data.frame(
      Generator = g, `N/group` = n,
      `CF NT` = fmt(pick(g, "invariant", n, "rej_cf_metric_nt")),
      `CF emp` = fmt(pick(g, "invariant", n, "rej_cf_metric_emp")),
      `ML NT` = fmt(pick(g, "invariant", n, "rej_ml_metric_nt")),
      `ML rob` = fmt(pick(g, "invariant", n, "rej_ml_metric_rob")),
      `CF power (emp)` = fmt(pick(g, "metric_viol", n, "rej_cf_metric_emp")),
      check.names = FALSE)
  }
  do.call(rbind, rows)
}
knitr::kable(mk_metric(), row.names = FALSE,
             caption = paste0("Metric-invariance Type-I (invariant population) and power. ",
                              "Target level ", alpha, "."))
```

### Scalar invariance: the closed-form reference-group map

Type-I under `invariant` (equal loadings and intercepts, group-2 latent means
free) and power under `scalar_viol` (one group-2 intercept shifted off the
loading space). There is no ML column: true scalar invariance frees the latent
means, and magmaan's restriction-map nested LRT does not accept that
parameterization change.

```{r scalar-table}
mk_scalar <- function() {
  rows <- list()
  for (g in gens) for (n in ns) {
    rows[[paste(g, n)]] <- data.frame(
      Generator = g, `N/group` = n,
      `Type-I (NT)`  = fmt(pick(g, "invariant", n, "rej_cf_scalar_nt")),
      `Type-I (emp)` = fmt(pick(g, "invariant", n, "rej_cf_scalar_emp")),
      `Power (NT)`   = fmt(pick(g, "scalar_viol", n, "rej_cf_scalar_nt")),
      `Power (emp)`  = fmt(pick(g, "scalar_viol", n, "rej_cf_scalar_emp")),
      check.names = FALSE)
  }
  do.call(rbind, rows)
}
knitr::kable(mk_scalar(), row.names = FALSE,
             caption = "Scalar-invariance rejection rate, normal-theory (NT) vs empirical (emp) Gamma.")
```

## Caveats

- Rejection rates are Monte Carlo estimates at the reps in `results/metadata.csv`;
  the smoke default is deliberately small and noisy. Read the full-run CSV for
  precise levels.
- Generators are normal and a non-normal continuous independent generator.
  Ordinal-as-continuous is deferred: a mean-structure invariance test on ordinal
  codes needs threshold modeling, not just treated-as-continuous moments.
- The scalar Type-I can sit a little above nominal at the smallest $N$; the test
  is a linearized (leading-order) Wald, and its mean statistic already sits close
  to its degrees of freedom on the true-invariance population.
- Under `metric_viol` the scalar test is not interpretable on its own (it assumes
  metric invariance), so its rate there is a conflated quantity, not power.

## Reproduce

```sh
# quick check (normal only, small reps)
Rscript run_experiment.R --smoke

# the full grid used here
Rscript run_experiment.R --full --reps 1000 --n 250,500 --generators normal,ig

# render this report
quarto render report.qmd
```
