---
title: "Guttman Communality Estimators"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
need <- c("summary", "complexity", "metadata")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --smoke\n",
       "or:\n  Rscript run_experiment.R --full")
}
summary <- read.csv(paths[1])
complexity <- read.csv(paths[2])
metadata <- read.csv(paths[3])
timing_path <- file.path(results_dir, "timing_summary.csv")
timing <- if (file.exists(timing_path)) read.csv(timing_path) else data.frame()
has_timing <- nrow(timing) > 0

method_label <- c(
  ar = "AR",
  rs = "RS",
  ilm = "ILM",
  nt_gls_block = "NT-GLS block",
  nt_gls_full = "NT-GLS full",
  gmm_block = "NT-GLS block",
  gmm_full = "NT-GLS full",
  nt_ml = "NT ML"
)
canonical_method <- function(x) {
  x[x == "gmm_block"] <- "nt_gls_block"
  x[x == "gmm_full"] <- "nt_gls_full"
  x
}
summary$method <- canonical_method(summary$method)
if (has_timing) timing$method <- canonical_method(timing$method)
method_order <- c("ar", "rs", "ilm", "nt_gls_block", "nt_gls_full", "nt_ml")
guttman_order <- setdiff(method_order, "nt_ml")
fmt <- function(x, digits = 2) {
  ifelse(is.na(x), "", formatC(x, format = "f", digits = digits))
}
pct <- function(x) paste0(fmt(100 * x, 1), "%")
meta <- function(key) {
  v <- metadata$value[metadata$key == key]
  if (length(v)) v[[1]] else ""
}
pick1 <- function(x) if (length(x)) x[[1]] else NA_real_

non_nt <- subset(summary, method != "nt_ml")
normal <- subset(summary, generator == "normal")
ordinal <- subset(summary, generator == "ordinal")

median_by_method <- function(data, value) {
  if (!nrow(data)) return(data.frame(method = character(), value = numeric()))
  out <- aggregate(data[[value]], list(method = data$method), median, na.rm = TRUE)
  names(out)[2] <- "value"
  out
}

normal_nt <- median_by_method(subset(normal, method != "nt_ml"), "common_rel_nt")
normal_rs <- median_by_method(subset(normal, method != "nt_ml"), "common_rel_rs")
ordinal_rs <- median_by_method(subset(ordinal, method != "nt_ml"), "common_rel_rs")
acc <- Reduce(function(x, y) merge(x, y, by = "method", all = TRUE),
              list(
                transform(normal_nt, normal_rel_nt = value)[c("method", "normal_rel_nt")],
                transform(normal_rs, normal_rel_rs = value)[c("method", "normal_rel_rs")],
                transform(ordinal_rs, ordinal_rel_rs = value)[c("method", "ordinal_rel_rs")]
              ))
acc$Method <- method_label[acc$method]
acc <- acc[match(guttman_order, acc$method),
           c("Method", "normal_rel_nt", "normal_rel_rs", "ordinal_rel_rs")]
names(acc) <- c("Method", "Normal / NT", "Normal / RS", "Ordinal / RS")
for (j in 2:ncol(acc)) acc[[j]] <- fmt(acc[[j]], 3)

rank_data <- subset(normal, method %in% c("rs", "ilm", "nt_gls_block", "nt_gls_full"))
best <- median_by_method(rank_data, "common_rel_nt")
best <- best[which.min(best$value), , drop = FALSE]
best_method <- method_label[best$method]
best_ratio <- best$value

ar_rel <- median(non_nt$common_rel_rs[non_nt$method == "ar"], na.rm = TRUE)
ilm_rel_rs <- median(non_nt$common_rel_rs[non_nt$method == "ilm"], na.rm = TRUE)

rt <- aggregate(cbind(elapsed_median_ms, elapsed_p90_ms, fail_rate) ~ method,
                summary,
                function(x) median(x, na.rm = TRUE))
rt$Method <- method_label[rt$method]
rt <- rt[match(method_order, rt$method),
         c("Method", "elapsed_median_ms", "elapsed_p90_ms", "fail_rate")]
names(rt) <- c("Method", "Median Time (ms)", "P90 Time (ms)", "Failure Rate")
rt$`Median Time (ms)` <- fmt(rt$`Median Time (ms)`, 3)
rt$`P90 Time (ms)` <- fmt(rt$`P90 Time (ms)`, 3)
rt$`Failure Rate` <- pct(rt$`Failure Rate`)

runtime_factor <- {
  nt <- median(summary$elapsed_median_ms[summary$method == "nt_ml"], na.rm = TRUE)
  ilm <- median(summary$elapsed_median_ms[summary$method == "ilm"], na.rm = TRUE)
  nt / ilm
}

small_n <- min(summary$n)
hey_small <- aggregate(improper_rate ~ method,
                       subset(summary, n == small_n),
                       median, na.rm = TRUE, na.action = na.pass)
hey_small <- stats::setNames(hey_small$improper_rate, hey_small$method)

bad_cols <- c("fail_rate", "improper_rate", "sample_singular_rate",
              "h_diag_bad_rate", "h_indef_rate", "score_nonpos_rate",
              "common_indef_rate")
bad <- aggregate(summary[bad_cols],
                 list(n = summary$n, method = summary$method),
                 median, na.rm = TRUE, na.action = na.pass)
bad <- subset(bad, n %in% head(sort(unique(summary$n)), 3))
bad$Method <- method_label[bad$method]
bad <- bad[order(bad$n, match(bad$method,
                              method_order)), ]
bad_show <- bad[c("n", "Method", bad_cols)]
names(bad_show) <- c("N", "Method", "Fail", "Heywood", "S Singular",
                     "Bad H Diag", "H Indef.", "Score Nonpos.",
                     "Common Indef.")
for (j in 3:ncol(bad_show)) bad_show[[j]] <- pct(bad_show[[j]])

if (has_timing) {
  timing_total <- subset(timing, component == "total")
  timing_by_n <- aggregate(cbind(elapsed_median_ms, fail_rate) ~ n + method,
                           timing_total, median, na.rm = TRUE,
                           na.action = na.pass)
  nt_by_n <- timing_by_n[timing_by_n$method == "nt_ml",
                         c("n", "elapsed_median_ms")]
  names(nt_by_n)[2] <- "nt_median_ms"
  timing_by_n <- merge(timing_by_n, nt_by_n, by = "n", all.x = TRUE)
  timing_by_n$nt_over_method <- timing_by_n$nt_median_ms /
    timing_by_n$elapsed_median_ms
  timing_by_n$Method <- method_label[timing_by_n$method]
  timing_by_n <- timing_by_n[order(timing_by_n$n,
                                   match(timing_by_n$method, method_order)), ]
  timing_n_show <- timing_by_n[c("n", "Method", "elapsed_median_ms",
                                 "nt_over_method", "fail_rate")]
  names(timing_n_show) <- c("N", "Method", "Median Time (ms)",
                            "NT / Method", "Failure Rate")
  timing_n_show$`Median Time (ms)` <- fmt(timing_n_show$`Median Time (ms)`, 3)
  timing_n_show$`NT / Method` <- fmt(timing_n_show$`NT / Method`, 2)
  timing_n_show$`Failure Rate` <- pct(timing_n_show$`Failure Rate`)

  timing_small_n <- min(timing_by_n$n)
  timing_factor <- median(timing_by_n$nt_over_method[
    timing_by_n$method == "ilm"], na.rm = TRUE)
  timing_small_factor <- pick1(timing_by_n$nt_over_method[
    timing_by_n$method == "ilm" & timing_by_n$n == timing_small_n])

  component_label <- c(
    sample_cov = "Covariance",
    guttman_h = "H diagonal",
    guttman_map = "Guttman map",
    sample_stats = "Sample stats",
    estimate_ml = "NT ML fit",
    extract_common = "Extract common"
  )
  component_order <- names(component_label)
  timing_comp <- subset(timing, component != "total" &
                          method %in% c("ilm", "nt_gls_block",
                                        "nt_gls_full", "nt_ml"))
  timing_comp <- aggregate(elapsed_median_ms ~ method + component,
                           timing_comp, median, na.rm = TRUE,
                           na.action = na.pass)
  timing_comp$Method <- method_label[timing_comp$method]
  timing_comp$Component <- component_label[timing_comp$component]
  timing_comp <- timing_comp[order(match(timing_comp$method,
                                         c("ilm", "nt_gls_block",
                                           "nt_gls_full", "nt_ml")),
                                   match(timing_comp$component,
                                         component_order)), ]
  timing_comp_show <- timing_comp[c("Method", "Component",
                                    "elapsed_median_ms")]
  names(timing_comp_show)[3] <- "Median Time (ms)"
  timing_comp_show$`Median Time (ms)` <- fmt(timing_comp_show$`Median Time (ms)`, 3)
} else {
  timing_n_show <- data.frame()
  timing_comp_show <- data.frame()
  timing_small_n <- NA_integer_
  timing_factor <- NA_real_
  timing_small_factor <- NA_real_
}

cx <- complexity[which.max(complexity$p), ]
```

## Question

Which closed-form communality estimator should the Guttman CFA paper recommend,
once the obvious Spearman variants are compared with normal-theory ML and with a
categorical data arm? The comparison uses the traditional incidence-composite
Guttman reconstruction throughout,
`H Z (Z' H Z)^-1 Z' H`, so the estimand is the common covariance matrix, not
only the diagonal communalities.

## Short answer

The current recommendation candidate is **ILM**: the identity-weighted linear
moment estimator. In this run its median common-covariance MISE is
`r fmt(ilm_rel_rs, 3)` times RS over the Guttman cells, while avoiding the
normal-theory triad covariance and inverse used by the NT-GLS variants. Under
normal data the best closed-form method versus magmaan NT ML is
`r best_method`, with median common-covariance MISE ratio
`r fmt(best_ratio, 2)`.

The NT-GLS variants do not buy much here. `nt_gls_block` needs only
within-factor correlation blocks, while `nt_gls_full` uses all within-factor
correlation pairs across factors, but neither uses the whole SEM Gamma. At the
largest displayed design (`r cx$factors` factors, `r cx$indicators` indicators
each), that means
`r cx$within_corr_pairs_total` within-block correlations and
`r cx$triad_rows_total` triad moments, versus `r cx$full_sem_corr_pairs`
off-diagonal correlations for full SEM ADF.

Runtime is consistent with the intended niche: NT ML is about
`r fmt(runtime_factor, 1)` times slower than ILM on this grid. Treat the exact
factor cautiously; the smoke grid is small and ML timing depends on optimizer
convergence. The timing scope is `r meta("timing_scope")`.
The dedicated timing pass gives NT ML / ILM median speedup
`r fmt(timing_factor, 1)` overall and `r fmt(timing_small_factor, 1)` at
`N = ` `r timing_small_n`.

The small-N arm is now deliberately rough. At `N = ` `r small_n`, the median
Heywood-like rate is `r pct(hey_small[["ilm"]])` for ILM,
`r pct(hey_small[["rs"]])` for RS, and `r pct(hey_small[["nt_ml"]])` for NT ML,
pooling over the displayed cells.

## Evidence

### Accuracy

Median common-covariance MISE ratios across the simulated cells. Lower is
better. The normal/NT column compares against magmaan normal-theory ML; the RS
columns compare against the ratio-of-sums Spearman rule.

```{r}
knitr::kable(acc, row.names = FALSE)
```

The point is not that AR is a serious baseline. It is useful as a warning:
individual ratio averaging is unstable when one of the other-indicator
correlations is small. RS already removes most of that pathology. ILM and the
NT-GLS variants then replace the ratio with a weighted linear moment pooling;
in this grid, the simple ILM pooling is as good as the heavier plug-in weights.

### Runtime

Median per-replication wall time by method, pooled across cells.

```{r}
knitr::kable(rt, row.names = FALSE)
```

The NT-GLS rows include construction and pseudo-inversion of the normal-theory
triad-moment covariance. The NT ML row is magmaan's iterative normal-theory fit.
Model syntax parsing and lavaanification are outside the per-replication clock;
the NT row still pays magmaan's current R-to-C++ partable conversion inside the
ML entry point.

### Detailed Timing

The dedicated timing pass uses `r meta("timing_reps")` recorded repetitions per
cell after `r meta("timing_warmup")` warmups. Each component timing is averaged
over `r meta("timing_inner")` repeated calls before taking medians, so sub-ms
Guttman steps are not rounded to zero by the elapsed-time clock. The timing pass
reuses the same model grid as the accuracy run, including the small-N cells,
but draws fresh data so that timing is not coupled to the MISE/improper
diagnostics.

```{r}
if (has_timing) {
  knitr::kable(timing_n_show, row.names = FALSE)
} else {
  cat("No dedicated timing pass was recorded.")
}
```

Median component timings, pooled over cells. For Guttman variants, `H diagonal`
is the communality estimator in `magmaan::guttman_h()` and `Guttman map` is the
traditional incidence-composite reconstruction.

```{r}
if (has_timing) {
  knitr::kable(timing_comp_show, row.names = FALSE)
}
```

### Improper Solutions

Heywood means at least one negative residual variance. For the Guttman variants,
that residual is `diag(S - H Z (Z' H Z)^-1 Z' H)`. `Bad H Diag` means the
estimated communality diagonal fell outside `[0, S_ii]`. `H Indef.` records when
the communality-filled matrix is indefinite; this is not automatically a
failed estimator, because the traditional map only needs the composite score
covariance to be invertible, but it marks cases where the literal covariance
interpretation of `H` is gone.

```{r}
knitr::kable(bad_show, row.names = FALSE)
```

### Complexity

The reduced NT-GLS covariance is block-local in the recommended form. It does
not need the full SEM Gamma matrix.

```{r}
cx_show <- complexity
names(cx_show) <- c("Factors", "Indicators", "p", "Pairs / Block",
                    "Triads / Block", "Within Pairs", "Triads Total",
                    "Full SEM Pairs", "Block Weight Dim", "Full Weight Dim")
knitr::kable(cx_show, row.names = FALSE)
```

For a factor with `m` indicators, block NT-GLS uses `choose(m, 2)` within-block
correlations and `m * choose(m - 1, 2)` triad equations. Full NT-GLS still
ignores cross-block correlations as moments, but its weight can couple the
within-block triads across factors.

## Caveats

- Results here are generated from `reps = ` `r meta("reps")`; use the full run
  before quoting numbers.
- The categorical arm is Pearson-code ordinal data generated by
  `magmaan::sim_ordcorr_calibrate()` / `sim_ordcorr_draw()` and then treated as
  continuous. It is a stress arm for the covariance estimator, not an ordinal
  latent-response estimator.
- The H-diagonal rules are computed by `magmaan::guttman_h()`. In the current
  C++ API, `nt_gls_block` and `nt_gls_full` are the experiment names for the
  existing `gmm_block` and `gmm_full` methods, which use the normal-theory
  covariance of sample correlations as the plug-in GLS weight. The full package
  frontier CFA estimator still uses the existing Guttman map; this experiment is
  evidence for which communality rule to promote into that map.
- Per-replication covariance/sample-statistics construction uses ML scaling
  (`1/n`) for both NT ML and the Guttman variants.

## Reproduce

```sh
cd experiments/55-guttman-communality-estimators
Rscript run_experiment.R --smoke
Rscript run_experiment.R --full --reps 200 --timing-reps 20
quarto render report.qmd
```
