---
title: "Guttman Communality Estimators"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
need <- c("summary", "complexity", "metadata")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --smoke\n",
       "or:\n  Rscript run_experiment.R --full")
}
summary <- read.csv(paths[1])
complexity <- read.csv(paths[2])
metadata <- read.csv(paths[3])

method_label <- c(
  ar = "AR",
  rs = "RS",
  ilm = "ILM",
  gmm_block = "GMM block",
  gmm_full = "GMM full",
  nt_ml = "NT ML"
)
fmt <- function(x, digits = 2) {
  ifelse(is.na(x), "", formatC(x, format = "f", digits = digits))
}
pct <- function(x) paste0(fmt(100 * x, 1), "%")
meta <- function(key) {
  v <- metadata$value[metadata$key == key]
  if (length(v)) v[[1]] else ""
}

non_nt <- subset(summary, method != "nt_ml")
normal <- subset(summary, generator == "normal")
ordinal <- subset(summary, generator == "ordinal")

median_by_method <- function(data, value) {
  if (!nrow(data)) return(data.frame(method = character(), value = numeric()))
  out <- aggregate(data[[value]], list(method = data$method), median, na.rm = TRUE)
  names(out)[2] <- "value"
  out
}

normal_nt <- median_by_method(subset(normal, method != "nt_ml"), "common_rel_nt")
normal_rs <- median_by_method(subset(normal, method != "nt_ml"), "common_rel_rs")
ordinal_rs <- median_by_method(subset(ordinal, method != "nt_ml"), "common_rel_rs")
acc <- Reduce(function(x, y) merge(x, y, by = "method", all = TRUE),
              list(
                transform(normal_nt, normal_rel_nt = value)[c("method", "normal_rel_nt")],
                transform(normal_rs, normal_rel_rs = value)[c("method", "normal_rel_rs")],
                transform(ordinal_rs, ordinal_rel_rs = value)[c("method", "ordinal_rel_rs")]
              ))
acc$Method <- method_label[acc$method]
acc <- acc[match(c("ar", "rs", "ilm", "gmm_block", "gmm_full"), acc$method),
           c("Method", "normal_rel_nt", "normal_rel_rs", "ordinal_rel_rs")]
names(acc) <- c("Method", "Normal / NT", "Normal / RS", "Ordinal / RS")
for (j in 2:ncol(acc)) acc[[j]] <- fmt(acc[[j]], 3)

rank_data <- subset(normal, method %in% c("rs", "ilm", "gmm_block", "gmm_full"))
best <- median_by_method(rank_data, "common_rel_nt")
best <- best[which.min(best$value), , drop = FALSE]
best_method <- method_label[best$method]
best_ratio <- best$value

ar_rel <- median(non_nt$common_rel_rs[non_nt$method == "ar"], na.rm = TRUE)
ilm_rel_rs <- median(non_nt$common_rel_rs[non_nt$method == "ilm"], na.rm = TRUE)

rt <- aggregate(cbind(elapsed_median_ms, elapsed_p90_ms, fail_rate) ~ method,
                summary,
                function(x) median(x, na.rm = TRUE))
rt$Method <- method_label[rt$method]
rt <- rt[match(c("ar", "rs", "ilm", "gmm_block", "gmm_full", "nt_ml"), rt$method),
         c("Method", "elapsed_median_ms", "elapsed_p90_ms", "fail_rate")]
names(rt) <- c("Method", "Median Time (ms)", "P90 Time (ms)", "Failure Rate")
rt$`Median Time (ms)` <- fmt(rt$`Median Time (ms)`, 3)
rt$`P90 Time (ms)` <- fmt(rt$`P90 Time (ms)`, 3)
rt$`Failure Rate` <- pct(rt$`Failure Rate`)

runtime_factor <- {
  nt <- median(summary$elapsed_median_ms[summary$method == "nt_ml"], na.rm = TRUE)
  ilm <- median(summary$elapsed_median_ms[summary$method == "ilm"], na.rm = TRUE)
  nt / ilm
}

cx <- complexity[which.max(complexity$p), ]
```

## Question

Which closed-form communality estimator should the Guttman CFA paper recommend,
once the obvious Spearman variants are compared with normal-theory ML and with a
categorical data arm? The comparison uses the loading-aligned Guttman
reconstruction throughout, so the estimand is the common covariance matrix, not
only the diagonal communalities.

## Short answer

The current recommendation candidate is **ILM**: the identity-weighted linear
moment estimator. In this run its median common-covariance MISE is
`r fmt(ilm_rel_rs, 3)` times RS over the Guttman cells, while avoiding the Gamma
estimation and inverse used by the plug-in GMM variants. Under normal data the
best closed-form method versus magmaan NT ML is `r best_method`, with median
common-covariance MISE ratio `r fmt(best_ratio, 2)`.

The GMM variants do not buy much here. `gmm_block` needs only within-factor
correlation blocks, while `gmm_full` uses all within-factor correlation pairs
across factors, but neither uses the whole SEM Gamma. At the largest displayed
design (`r cx$factors` factors, `r cx$indicators` indicators each), that means
`r cx$within_corr_pairs_total` within-block correlations and
`r cx$triad_rows_total` triad moments, versus `r cx$full_sem_corr_pairs`
off-diagonal correlations for full SEM ADF.

Runtime is consistent with the intended niche: NT ML is about
`r fmt(runtime_factor, 1)` times slower than ILM on this grid. Treat the exact
factor cautiously; the smoke grid is small and ML timing depends on optimizer
convergence.

## Evidence

### Accuracy

Median common-covariance MISE ratios across the simulated cells. Lower is
better. The normal/NT column compares against magmaan normal-theory ML; the RS
columns compare against the ratio-of-sums Spearman rule.

```{r}
knitr::kable(acc, row.names = FALSE)
```

The point is not that AR is a serious baseline. It is useful as a warning:
individual ratio averaging is unstable when one of the other-indicator
correlations is small. RS already removes most of that pathology. ILM and the
GMM variants then replace the ratio with a linear moment pooling; in this grid,
the simple ILM pooling is as good as the heavier plug-in weights.

### Runtime

Median per-replication wall time by method, pooled across cells.

```{r}
knitr::kable(rt, row.names = FALSE)
```

The GMM rows include construction and pseudo-inversion of their estimated
triad-moment covariance. The NT ML row is magmaan's iterative normal-theory fit.

### Complexity

The reduced GMM covariance is block-local in the recommended form. It does not
need the full SEM Gamma matrix.

```{r}
cx_show <- complexity
names(cx_show) <- c("Factors", "Indicators", "p", "Pairs / Block",
                    "Triads / Block", "Within Pairs", "Triads Total",
                    "Full SEM Pairs", "Block Weight Dim", "Full Weight Dim")
knitr::kable(cx_show, row.names = FALSE)
```

For a factor with `m` indicators, block GMM uses `choose(m, 2)` within-block
correlations and `m * choose(m - 1, 2)` triad equations. Full GMM still ignores
cross-block correlations as moments, but its weight can couple the within-block
triads across factors.

## Caveats

- Results here are generated from `reps = ` `r meta("reps")`; use the full run
  before quoting numbers.
- The categorical arm is Pearson-code ordinal data generated by
  `magmaan::sim_ordcorr_calibrate()` / `sim_ordcorr_draw()` and then treated as
  continuous. It is a stress arm for the covariance estimator, not an ordinal
  latent-response estimator.
- The Guttman variants are local to this experiment. The current package
  frontier estimator still exposes the existing Guttman map; this experiment is
  evidence for which communality rule to promote into that implementation.

## Reproduce

```sh
cd experiments/55-guttman-communality-estimators
Rscript run_experiment.R --smoke
Rscript run_experiment.R --full --reps 200
quarto render report.qmd
```
