---
title: "Guttman Communality Estimators"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
need <- c("summary", "complexity", "metadata")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --smoke\n",
       "or:\n  Rscript run_experiment.R --full")
}
summary <- read.csv(paths[1])
complexity <- read.csv(paths[2])
metadata <- read.csv(paths[3])
timing_path <- file.path(results_dir, "timing_summary.csv")
timing <- if (file.exists(timing_path)) read.csv(timing_path) else data.frame()
has_timing <- nrow(timing) > 0

method_label <- c(
  ar = "AR",
  rs = "RS",
  ilm = "ILM",
  anchor_ilm = "Anchor ILM",
  nt_gls_block = "NT-GLS block",
  nt_gls_full = "NT-GLS full",
  gmm_block = "NT-GLS block",
  gmm_full = "NT-GLS full",
  nt_ml = "NT ML"
)
map_label <- c(
  incidence = "incidence",
  aligned = "aligned",
  diag_aligned = "diag-aligned",
  block_aligned = "block-aligned",
  ml = "ML"
)
canonical_method <- function(x) {
  x[x == "gmm_block"] <- "nt_gls_block"
  x[x == "gmm_full"] <- "nt_gls_full"
  x
}
summary$method <- canonical_method(summary$method)
if (has_timing) timing$method <- canonical_method(timing$method)
if (!"map" %in% names(summary)) {
  summary$map <- ifelse(summary$method == "nt_ml", "ml", "incidence")
}
if (has_timing && !"map" %in% names(timing)) {
  timing$map <- ifelse(timing$method == "nt_ml", "ml", "incidence")
}
method_order <- c("ar", "rs", "ilm", "anchor_ilm", "nt_gls_block",
                  "nt_gls_full", "nt_ml")
guttman_order <- setdiff(method_order, "nt_ml")
map_order <- c("incidence", "aligned", "block_aligned", "diag_aligned")
maps_present <- intersect(map_order, unique(summary$map))
guttman_estimator_order <- unlist(lapply(
  guttman_order, function(m) paste(m, maps_present, sep = "|")))
estimator_order <- c(guttman_estimator_order, "nt_ml|ml")
estimator_label <- function(method, map) {
  ifelse(method == "nt_ml", "NT ML",
         paste(method_label[method], map_label[map], sep = " / "))
}
summary$estimator_key <- paste(summary$method, summary$map, sep = "|")
summary$Estimator <- estimator_label(summary$method, summary$map)
if (has_timing) {
  timing$estimator_key <- paste(timing$method, timing$map, sep = "|")
  timing$Estimator <- estimator_label(timing$method, timing$map)
}
focus_keys <- intersect(
  c("rs|incidence", "ilm|incidence", "ilm|aligned",
    "anchor_ilm|aligned", "nt_gls_block|aligned", "nt_ml|ml"),
  unique(summary$estimator_key)
)
guttman_focus_keys <- setdiff(focus_keys, "nt_ml|ml")
report_summary <- subset(summary, estimator_key %in% focus_keys)
report_timing <- if (has_timing) {
  subset(timing, estimator_key %in% focus_keys)
} else {
  data.frame()
}
probe_keys <- intersect(
  c("ilm|incidence", "ilm|aligned", "ilm|block_aligned",
    "anchor_ilm|incidence", "anchor_ilm|aligned",
    "anchor_ilm|block_aligned",
    "nt_gls_block|incidence", "nt_gls_block|aligned",
    "nt_gls_block|block_aligned"),
  unique(summary$estimator_key)
)
fmt <- function(x, digits = 2) {
  ifelse(is.na(x), "", formatC(x, format = "f", digits = digits))
}
fmt_g <- function(x, digits = 4) {
  ifelse(is.na(x), "", formatC(x, format = "fg", digits = digits))
}
pct <- function(x) paste0(fmt(100 * x, 1), "%")
meta <- function(key) {
  v <- metadata$value[metadata$key == key]
  if (length(v)) v[[1]] else ""
}
pick1 <- function(x) if (length(x)) x[[1]] else NA_real_

non_nt <- subset(report_summary, method != "nt_ml")
normal <- subset(report_summary, generator == "normal")
ordinal <- subset(report_summary, generator == "ordinal")

median_by_estimator <- function(data, value) {
  if (!nrow(data)) {
    return(data.frame(method = character(), map = character(),
                      estimator_key = character(), value = numeric()))
  }
  out <- aggregate(data[[value]],
                   list(method = data$method, map = data$map,
                        estimator_key = data$estimator_key),
                   median, na.rm = TRUE)
  names(out)[ncol(out)] <- "value"
  out
}

normal_nt <- median_by_estimator(subset(normal, method != "nt_ml"), "common_rel_nt")
normal_rs <- median_by_estimator(subset(normal, method != "nt_ml"), "common_rel_rs")
ordinal_rs <- median_by_estimator(subset(ordinal, method != "nt_ml"), "common_rel_rs")
acc <- Reduce(function(x, y) merge(x, y, by = c("method", "map", "estimator_key"), all = TRUE),
              list(
                transform(normal_nt, normal_rel_nt = value)[c("method", "map", "estimator_key", "normal_rel_nt")],
                transform(normal_rs, normal_rel_rs = value)[c("method", "map", "estimator_key", "normal_rel_rs")],
                transform(ordinal_rs, ordinal_rel_rs = value)[c("method", "map", "estimator_key", "ordinal_rel_rs")]
              ))
acc$Estimator <- estimator_label(acc$method, acc$map)
acc <- acc[match(guttman_focus_keys, acc$estimator_key),
           c("Estimator", "normal_rel_nt", "normal_rel_rs", "ordinal_rel_rs")]
names(acc) <- c("Estimator", "Normal / NT", "Normal / RS", "Ordinal / RS")
for (j in 2:ncol(acc)) acc[[j]] <- fmt(acc[[j]], 3)

rank_data <- subset(normal, method %in% c("rs", "ilm", "anchor_ilm",
                                          "nt_gls_block"))
best <- median_by_estimator(rank_data, "common_rel_nt")
best <- best[which.min(best$value), , drop = FALSE]
best_method <- estimator_label(best$method, best$map)
best_ratio <- best$value

ar_rel <- median(non_nt$common_rel_rs[non_nt$method == "ar" &
                                        non_nt$map == "incidence"], na.rm = TRUE)
ilm_rel_rs <- median(non_nt$common_rel_rs[non_nt$method == "ilm" &
                                            non_nt$map == "incidence"], na.rm = TRUE)
anchor_rel_rs <- median(non_nt$common_rel_rs[non_nt$method == "anchor_ilm" &
                                               non_nt$map == "aligned"], na.rm = TRUE)

rt <- aggregate(cbind(elapsed_median_ms, elapsed_p90_ms, fail_rate) ~
                  method + map + estimator_key + Estimator,
                report_summary,
                function(x) median(x, na.rm = TRUE))
rt <- rt[match(focus_keys, rt$estimator_key),
         c("Estimator", "elapsed_median_ms", "elapsed_p90_ms", "fail_rate")]
names(rt) <- c("Estimator", "Median Time (ms)", "P90 Time (ms)",
               "Failure Rate")
rt$`Median Time (ms)` <- fmt(rt$`Median Time (ms)`, 3)
rt$`P90 Time (ms)` <- fmt(rt$`P90 Time (ms)`, 3)
rt$`Failure Rate` <- pct(rt$`Failure Rate`)

runtime_factor <- {
  nt <- median(summary$elapsed_median_ms[summary$method == "nt_ml"], na.rm = TRUE)
  ilm <- median(summary$elapsed_median_ms[summary$method == "ilm" &
                                            summary$map == "incidence"],
                na.rm = TRUE)
  nt / ilm
}

small_n <- min(summary$n)
hey_small <- aggregate(improper_rate ~ estimator_key,
                       subset(report_summary, n == small_n),
                       median, na.rm = TRUE, na.action = na.pass)
hey_small <- stats::setNames(hey_small$improper_rate, hey_small$estimator_key)

bad_cols <- c("fail_rate", "improper_rate", "sample_singular_rate",
              "h_diag_bad_rate", "h_indef_rate", "score_nonpos_rate",
              "common_indef_rate")
bad <- aggregate(report_summary[bad_cols],
                 list(n = report_summary$n, method = report_summary$method,
                      map = report_summary$map,
                      estimator_key = report_summary$estimator_key,
                      Estimator = report_summary$Estimator),
                 median, na.rm = TRUE, na.action = na.pass)
bad <- subset(bad, n %in% head(sort(unique(summary$n)), 3))
bad <- bad[order(bad$n, match(bad$method,
                              method_order),
                 match(bad$map, c(map_order, "ml"))), ]
bad_show <- bad[c("n", "Estimator", bad_cols)]
names(bad_show) <- c("N", "Estimator", "Fail", "Heywood", "S Singular",
                     "Bad H Diag", "H Indef.", "Score Nonpos.",
                     "Common Indef.")
for (j in 3:ncol(bad_show)) bad_show[[j]] <- pct(bad_show[[j]])

if (has_timing) {
  timing_total <- subset(report_timing, component == "total")
  timing_by_n <- aggregate(cbind(elapsed_median_ms, fail_rate) ~
                             n + method + map + estimator_key + Estimator,
                           timing_total, median, na.rm = TRUE,
                           na.action = na.pass)
  nt_by_n <- timing_by_n[timing_by_n$method == "nt_ml",
                         c("n", "elapsed_median_ms")]
  names(nt_by_n)[2] <- "nt_median_ms"
  timing_by_n <- merge(timing_by_n, nt_by_n, by = "n", all.x = TRUE)
  timing_by_n$nt_over_method <- timing_by_n$nt_median_ms /
    timing_by_n$elapsed_median_ms
  timing_by_n <- timing_by_n[order(timing_by_n$n,
                                   match(timing_by_n$estimator_key,
                                         focus_keys)), ]
  timing_n_show <- timing_by_n[c("n", "Estimator", "elapsed_median_ms",
                                 "nt_over_method", "fail_rate")]
  names(timing_n_show) <- c("N", "Estimator", "Median Time (ms)",
                            "NT / Estimator", "Failure Rate")
  timing_n_show$`Median Time (ms)` <- fmt(timing_n_show$`Median Time (ms)`, 3)
  timing_n_show$`NT / Estimator` <- fmt(timing_n_show$`NT / Estimator`, 2)
  timing_n_show$`Failure Rate` <- pct(timing_n_show$`Failure Rate`)

  timing_small_n <- min(timing_by_n$n)
  timing_factor <- median(timing_by_n$nt_over_method[
    timing_by_n$method == "ilm" & timing_by_n$map == "incidence"],
    na.rm = TRUE)
  timing_small_factor <- pick1(timing_by_n$nt_over_method[
    timing_by_n$method == "ilm" & timing_by_n$map == "incidence" &
      timing_by_n$n == timing_small_n])

  component_label <- c(
    sample_cov = "Covariance",
    guttman_h = "H diagonal",
    guttman_map = "Guttman map",
    sample_stats = "Sample stats",
    estimate_ml = "NT ML fit",
    extract_common = "Extract common"
  )
  component_order <- names(component_label)
  timing_comp <- subset(report_timing, component != "total" &
                          method %in% c("ilm", "anchor_ilm", "nt_gls_block",
                                        "nt_ml"))
  timing_comp <- aggregate(elapsed_median_ms ~
                             method + map + estimator_key + Estimator +
                             component,
                           timing_comp, median, na.rm = TRUE,
                           na.action = na.pass)
  timing_comp$Component <- component_label[timing_comp$component]
  timing_comp <- timing_comp[order(match(timing_comp$estimator_key,
                                         focus_keys),
                                   match(timing_comp$component,
                                         component_order)), ]
  timing_comp_show <- timing_comp[c("Estimator", "Component",
                                    "elapsed_median_ms")]
  names(timing_comp_show)[3] <- "Median Time (ms)"
  timing_comp_show$`Median Time (ms)` <- fmt(timing_comp_show$`Median Time (ms)`, 3)
} else {
  timing_n_show <- data.frame()
  timing_comp_show <- data.frame()
  timing_small_n <- NA_integer_
  timing_factor <- NA_real_
  timing_small_factor <- NA_real_
}

cx <- complexity[which.max(complexity$p), ]

accuracy_methods <- subset(report_summary, method %in% method_order)

group_accuracy <- function(by) {
  cols <- c("common_mise", "diag_mise", "common_rel_rs", "common_rel_nt",
            "nt_joint_success_rate", "fail_rate", "improper_rate")
  out <- aggregate(accuracy_methods[cols],
                   accuracy_methods[c(by, "method", "map", "estimator_key",
                                      "Estimator")],
                   median, na.rm = TRUE, na.action = na.pass)
  ord <- do.call(order, c(out[by],
                          list(match(out$estimator_key, focus_keys))))
  out <- out[ord, , drop = FALSE]
  show <- out[c(by, "Estimator", cols)]
  names(show) <- c(by, "Estimator", "Common MISE",
                   "Diag MISE", "Common / RS", "Common / NT",
                   "NT Joint",
                   "Fail", "Heywood")
  show$`Common MISE` <- fmt_g(show$`Common MISE`)
  show$`Diag MISE` <- fmt_g(show$`Diag MISE`)
  show$`Common / RS` <- fmt(show$`Common / RS`, 3)
  show$`Common / NT` <- fmt(show$`Common / NT`, 3)
  show$`NT Joint` <- pct(show$`NT Joint`)
  show$Fail <- pct(show$Fail)
  show$Heywood <- pct(show$Heywood)
  show
}

factor_accuracy_show <- group_accuracy(c("generator", "factors"))
n_accuracy_show <- group_accuracy(c("generator", "n"))
rho_accuracy_show <- group_accuracy(c("generator", "rho"))

hs_accuracy_methods <- subset(report_summary, factors == 3 & indicators == 3 &
                                method %in% method_order)
if (nrow(hs_accuracy_methods)) {
  saved_accuracy_methods <- accuracy_methods
  accuracy_methods <- hs_accuracy_methods
  hs_accuracy_show <- group_accuracy(c("generator", "n"))
  accuracy_methods <- saved_accuracy_methods
} else {
  hs_accuracy_show <- data.frame()
}

winner_source <- subset(report_summary, method %in% guttman_order &
                          is.finite(common_mise))
if (nrow(winner_source)) {
  key_cols <- c("generator", "factors", "indicators", "loading", "rho", "n")
  winner_source$condition_id <- do.call(
    paste, c(winner_source[key_cols], sep = "|"))
  split_rows <- split(seq_len(nrow(winner_source)),
                      winner_source$condition_id)
  best_rows <- vapply(split_rows, function(ii) {
    ii[which.min(winner_source$common_mise[ii])]
  }, integer(1))
  winners <- winner_source[best_rows, , drop = FALSE]
  winner_counts <- aggregate(rep(1L, nrow(winners)),
                             winners[c("generator", "factors", "method", "map",
                                       "estimator_key", "Estimator")],
                             sum)
  names(winner_counts)[7] <- "wins"
  condition_counts <- unique(winner_source[c("condition_id", "generator",
                                             "factors")])
  condition_counts <- aggregate(condition_id ~ generator + factors,
                                condition_counts, length)
  names(condition_counts)[3] <- "conditions"
  winner_counts <- merge(winner_counts, condition_counts,
                         by = c("generator", "factors"), all.x = TRUE)
  winner_counts$share <- winner_counts$wins / winner_counts$conditions
  winner_counts <- winner_counts[order(winner_counts$generator,
                                       winner_counts$factors,
                                       match(winner_counts$estimator_key,
                                             focus_keys)), ]
  winner_show <- winner_counts[c("generator", "factors", "Estimator", "wins",
                                 "conditions", "share")]
  names(winner_show) <- c("Generator", "Factors", "Estimator", "Wins",
                          "Conditions", "Share")
  winner_show$Share <- pct(winner_show$Share)
} else {
  winner_show <- data.frame()
}

map_probe_source <- subset(summary, estimator_key %in% probe_keys)
if (nrow(map_probe_source)) {
  map_probe <- aggregate(cbind(common_mise, common_rel_nt, common_rel_rs,
                               nt_joint_success_rate, fail_rate,
                               improper_rate) ~
                           method + map + estimator_key + Estimator,
                         map_probe_source, median, na.rm = TRUE,
                         na.action = na.pass)
  map_probe <- map_probe[order(match(map_probe$estimator_key, probe_keys)), ]
  map_probe_show <- map_probe[c("Estimator", "common_mise", "common_rel_nt",
                                "common_rel_rs", "nt_joint_success_rate",
                                "fail_rate", "improper_rate")]
  names(map_probe_show) <- c("Estimator", "Common MISE", "Common / NT",
                             "Common / RS", "NT Joint", "Fail", "Heywood")
  map_probe_show$`Common MISE` <- fmt_g(map_probe_show$`Common MISE`)
  map_probe_show$`Common / NT` <- fmt(map_probe_show$`Common / NT`, 3)
  map_probe_show$`Common / RS` <- fmt(map_probe_show$`Common / RS`, 3)
  map_probe_show$`NT Joint` <- pct(map_probe_show$`NT Joint`)
  map_probe_show$Fail <- pct(map_probe_show$Fail)
  map_probe_show$Heywood <- pct(map_probe_show$Heywood)
} else {
  map_probe_show <- data.frame()
}
```

## Question

Which closed-form communality estimator should the Guttman CFA paper recommend,
once the obvious Spearman variants are compared with normal-theory ML and with a
categorical data arm? The comparison now crosses the communality rule with
several closed-form Guttman reconstruction maps, but the displayed report is
focused on a small candidate set. The baseline is the traditional
incidence-composite reconstruction, `H Z (Z' H Z)^-1 Z' H`. The main new map is
the loading-aligned score rule. The new communality rule is `anchor_ilm`, which
adds cross-block anchor-triad rows to the identity-weighted linear moment solve.
The estimand is the reconstructed common covariance matrix, not only the
diagonal communalities.

## Short answer

The old cheap baseline is **ILM / incidence**: the identity-weighted
within-block linear moment rule with the historical incidence map. In this run
its median common-covariance MISE is `r fmt(ilm_rel_rs, 3)` times RS /
incidence over the Guttman cells, while avoiding the normal-theory triad
covariance and inverse used by the NT-GLS variants. The new result is that the
extended anchor-triad probe, **Anchor ILM / aligned**, is the more interesting
candidate: its median common-covariance MISE is `r fmt(anchor_rel_rs, 3)` times
RS / aligned over the displayed Guttman cells. Under normal data the best
closed-form estimator versus magmaan NT ML on joint-success replications is
`r best_method`, with median common-covariance MISE ratio `r fmt(best_ratio, 2)`.

The NT-GLS variants do not buy much here. `nt_gls_block` needs only
within-factor correlation blocks, while `nt_gls_full` uses all within-factor
correlation pairs across factors, but neither uses the whole SEM Gamma. At the
largest displayed design (`r cx$factors` factors, `r cx$indicators` indicators
each), that means
`r cx$within_corr_pairs_total` within-block correlations and
`r cx$triad_rows_total` within-block triad moments. The anchor extension uses
`r cx$anchor_rows_total` anchor-triad rows, versus
`r cx$full_sem_corr_pairs` off-diagonal correlations for full SEM ADF.

Runtime is consistent with the intended niche: NT ML is about
`r fmt(runtime_factor, 1)` times slower than ILM / incidence on this grid.
Treat the exact factor cautiously; the smoke grid is small and ML timing
depends on optimizer convergence. The timing scope is `r meta("timing_scope")`.
The dedicated timing pass gives NT ML / ILM-incidence median speedup
`r fmt(timing_factor, 1)` overall and `r fmt(timing_small_factor, 1)` at
`N = ` `r timing_small_n`.

The small-N arm is now deliberately rough. At `N = ` `r small_n`, the median
Heywood-like rate is `r pct(hey_small[["ilm|incidence"]])` for ILM / incidence,
`r pct(hey_small[["rs|incidence"]])` for RS / incidence, and
`r pct(hey_small[["nt_ml|ml"]])` for NT ML, pooling over the displayed cells.

The invariance story is not solved yet. `aligned` means loading-aligned scores,
not diagonal standardization. It is a cleaner map once \(H\) has been supplied,
but the current triad \(H\)-estimators are still named-coordinate constructions.
They are invariant to ordinary rescaling through the correlation scale, not to
arbitrary within-block linear transforms. A genuinely block-invariant Guttman
class would need a block-equivariant \(H\) rule and a block-equivariant map.

## Evidence

### Accuracy

Median common-covariance MISE ratios across the simulated cells. Lower is
better. The normal/NT column compares against magmaan normal-theory ML on the
replications where both the displayed estimator and NT ML succeeded. The RS
columns compare against the ratio-of-sums Spearman rule using the same
reconstruction map.

```{r}
knitr::kable(acc, row.names = FALSE)
```

The point is not that AR is a serious baseline. It is useful as a warning:
individual ratio averaging is unstable when one of the other-indicator
correlations is small. RS already removes most of that pathology. ILM and the
NT-GLS variants then replace the ratio with a weighted linear moment pooling;
in this grid, the simple ILM pooling is as good as the heavier plug-in weights.
Anchor ILM keeps the same identity-weighted solve but stacks additional rows
\(c_i r_{jk}-r_{ij}r_{ik}=0\) where \(j\) is in \(i\)'s factor block and \(k\)
may be in another block. These rows should help mainly when the latent factors
are correlated; at \(\rho=0\) their population coefficient is zero, so they are
mostly a stress test for finite-sample noise.

The reconstruction maps are separate from the communality rule. `incidence` is
the historical \(Z\)-score map. `aligned` replaces \(Z\) by
\(B=\Gamma(\Gamma^\top\Gamma)^{-1}\), where \(\Gamma\) is the preliminary
incidence Guttman loading representative. `block-aligned` first whitens each
indicator block by its observed covariance and then applies the aligned score
rule in those canonical coordinates.

The next table is a compact map probe rather than part of the main candidate
set. It checks whether the more ambitious block-aligned map helps after the same
communality rule has been chosen.

```{r}
if (nrow(map_probe_show)) {
  knitr::kable(map_probe_show, row.names = FALSE)
}
```

### Invariance Target

The report is focused because the statistical target is now clearer. The cheap
candidate is ILM for the \(H\)-diagonal and loading-aligned scores for the
Guttman map. The efficiency-oriented candidate is NT-GLS block for the
\(H\)-diagonal and the same loading-aligned map.

The block-invariance target is stricter. The loading-aligned map is equivariant
if \(H\) and the preliminary loading representative are transformed together.
The present \(H\)-diagonal rules do not have that property under arbitrary
within-block transforms: triad identities rely on named pure indicators with
diagonal residual covariance, and a general block transform destroys that
diagonal residual structure. So the current experiment should be read as
evidence for the map-side improvement and for cheap/NT-weighted triad rules,
not as evidence that the full estimator is block invariant.

### Condition Breakdown

The pooled table above hides where an estimator is winning or failing. The next
tables keep the same common-covariance MISE target but aggregate over fewer
design dimensions. Ratios below one mean the estimator improved on the comparison
row; `Common / NT` uses the iterative normal-theory fit as the reference on
joint-success replications. `NT Joint` is the fraction of replications entering
that paired comparison.

By generator and number of factors:

```{r}
knitr::kable(factor_accuracy_show, row.names = FALSE)
```

By generator and sample size:

```{r}
knitr::kable(n_accuracy_show, row.names = FALSE)
```

By generator and latent correlation:

```{r}
knitr::kable(rho_accuracy_show, row.names = FALSE)
```

HS-style three-factor slice: `factors = 3` and `indicators = 3`, aggregated
over loading pattern and latent correlation. This is the smallest classical
simple-structure CFA layout with three indicators per factor, so the triad
systems have little redundancy; differences between weighting rules should be
read with that in mind.

```{r}
if (nrow(hs_accuracy_show)) {
  knitr::kable(hs_accuracy_show, row.names = FALSE)
} else {
  cat("No factors = 3, indicators = 3 cells were present in this run.")
}
```

Winner counts by condition, among the closed-form Guttman estimators. A
condition is one generator/factor/indicator/loading/rho/N cell, and the winner
is chosen over communality-rule/reconstruction-map combinations.

```{r}
if (nrow(winner_show)) {
  knitr::kable(winner_show, row.names = FALSE)
}
```

### Runtime

Median per-replication wall time by estimator, pooled across cells.

```{r}
knitr::kable(rt, row.names = FALSE)
```

The NT-GLS rows include construction and pseudo-inversion of the normal-theory
triad-moment covariance. The NT ML row is magmaan's iterative normal-theory fit.
Model syntax parsing and lavaanification are outside the per-replication clock;
the NT row still pays magmaan's current R-to-C++ partable conversion inside the
ML entry point.

### Detailed Timing

The dedicated timing pass uses `r meta("timing_reps")` recorded repetitions per
cell after `r meta("timing_warmup")` warmups. Each component timing is averaged
over `r meta("timing_inner")` repeated calls before taking medians, so sub-ms
Guttman steps are not rounded to zero by the elapsed-time clock. The timing pass
reuses the same model grid as the accuracy run, including the small-N cells,
but draws fresh data so that timing is not coupled to the MISE/improper
diagnostics.

```{r}
if (has_timing) {
  knitr::kable(timing_n_show, row.names = FALSE)
} else {
  cat("No dedicated timing pass was recorded.")
}
```

Median component timings, pooled over cells. For Guttman variants, `H diagonal`
is the communality estimator in `magmaan::guttman_h()` and `Guttman map` is the
chosen reconstruction map.

```{r}
if (has_timing) {
  knitr::kable(timing_comp_show, row.names = FALSE)
}
```

### Improper Solutions

Heywood means at least one negative residual variance. For the Guttman variants,
that residual is `diag(S - common_hat)`, where `common_hat` is the selected
Guttman reconstruction map. `Bad H Diag` means the
estimated communality diagonal fell outside `[0, S_ii]`. `H Indef.` records when
the communality-filled matrix is indefinite; this is not automatically a
failed estimator, because the selected map only needs the composite score
covariance to be invertible, but it marks cases where the literal covariance
interpretation of `H` is gone.

```{r}
knitr::kable(bad_show, row.names = FALSE)
```

### Complexity

The reduced NT-GLS covariance is block-local in the recommended form. It does
not need the full SEM Gamma matrix.

```{r}
cx_show <- complexity
names(cx_show) <- c("Factors", "Indicators", "p", "Pairs / Block",
                    "Triads / Block", "Anchor Rows / Item",
                    "Within Pairs", "Triads Total", "Anchor Rows Total",
                    "Full SEM Pairs", "Block Weight Dim", "Full Weight Dim")
knitr::kable(cx_show, row.names = FALSE)
```

For a factor with `m` indicators, block NT-GLS uses `choose(m, 2)` within-block
correlations and `m * choose(m - 1, 2)` triad equations. Full NT-GLS still
ignores cross-block correlations as moments, but its weight can couple the
within-block triads across factors. Anchor ILM adds
`choose(m - 1, 2) + (m - 1) * (p - m)` rows per indicator by pairing one
same-block anchor with either another same-block anchor or a cross-block anchor.

## Caveats

- Results here are generated from `reps = ` `r meta("reps")`; use the full run
  before quoting numbers.
- The default run is intentionally focused. The runner still accepts AR,
  NT-GLS-full, and `diag_aligned`, but the report centers on the small candidate
  set plus a compact block-aligned map probe.
- Ratios against NT ML are paired by replication and use only replications where
  both estimators succeeded. The raw marginal NT ratio is still written to
  `summary.csv` as `common_rel_nt_marginal` for auditing selective NT failure.
- The categorical arm is Pearson-code ordinal data generated by
  `magmaan::sim_ordcorr_calibrate()` / `sim_ordcorr_draw()` and then treated as
  continuous. It is a stress arm for the covariance estimator, not an ordinal
  latent-response estimator.
- Most H-diagonal rules are computed by `magmaan::guttman_h()`. In the current
  C++ API, `nt_gls_block` and `nt_gls_full` are the experiment names for the
  existing `gmm_block` and `gmm_full` methods, which use the normal-theory
  covariance of sample correlations as the plug-in GLS weight. `anchor_ilm` and
  the reconstruction-map variants are experiment-local R code for now. The full
  package frontier CFA estimator still uses the existing Guttman map; this
  experiment is evidence for which communality rule and map to promote.
- Per-replication covariance/sample-statistics construction uses ML scaling
  (`1/n`) for both NT ML and the Guttman variants.

## Reproduce

```sh
cd experiments/55-guttman-communality-estimators
Rscript run_experiment.R --smoke
Rscript run_experiment.R --full --reps 200 --timing-reps 20
quarto render report.qmd
```
