---
title: "DWLS policy calibration"
format:
  html:
    toc: true
    html-math-method: mathml
execute:
  echo: false
  warning: false
  message: false
---

## Default register

| Choice | Current value | Status | Evidence |
|---|---|---|---|
| Covariance | Estimated-weight IJ sandwich | Adopted by principle; calibrated | Production 2026-10-03: median coverage about 95% for every target |
| Global test | $nF$ on the `robust_ordinal` spectrum, exact weighted chi-square tail with every positive eigenvalue (All) | Chosen after exploration (user, 2026-10-03); confirmed on fresh draws | SB and PEBA4 over-reject at df ≥ 53; All 2.9–6.8% at confirmation |
| Nested test | $N(F_{H_0} - F_{H_1})$ with the $r$-term parameter-space law (observed Hessian, estimated-weight IJ meat, exact restriction map), SB and PEBA4 | Replaced the separate-point profile law after diagnosis; confirmed on fresh draws | Profile law 0.1–2% (SB) and 3–17% (PEBA4); parameter-space law 4.3–7.8% |
| Nested score test | Typed unavailable | Decision D4 | No joint least-squares nested score primitive |
| All-ordinal exact first stage | Exact sampling rows; unchanged fitting weights and references | Reconfirmation open | TASK-79 fresh-draw registration and local pricing pilot; production pending merger review |
| Mixed continuous/ordinal DWLS | Exact-first-stage IJ; global All; nested SB/PEBA4 | Calibration open | TASK-80 registration and pricing pilot; production pending |

## Question

How well calibrated are the all-ordinal DWLS policy components? The recipes'
ingredients are fixed by the misspecification-robust requirement
([scope](../../../project/scope.md)): IJ covariance with the estimated weight's
influence, the robust-ordinal spectrum for the global test, and an
estimated-weight law for the nested test. Which reference-law approximation to
use, and whether a law is correctly assembled, are empirical questions this
study answers. The registered design, its amendments and the confirmation
criteria are in `criteria/dwls_policy.md`.

## Short answer

The all-ordinal findings below are retained **OPG-first-stage** evidence. The
current exact-first-stage policy has a registered reconfirmation lane; its
local two-draw-per-cell pilot tests mechanics and price, not calibration. Mixed DWLS has a
registered lane and pricing pilot; its calibration is still open pending
reviewed production.

**The IJ covariance is calibrated.** Across 42 coverage cells (misspecification
0, .2 and .4; N 150 to 1000; binary and five categories; one and two groups)
its median coverage is about 95% for loadings, thresholds, correlations and
paths, with a minimum of 92.3%. The fixed-weight expected and observed
sandwiches fall to 89% for correlations and paths under misspecification.

**The global test needs the exact spectrum tail.** SB and PEBA4 on the robust
spectrum over-reject once df ≥ 53, at 8–12% even at N = 1000 and up to 20% at
N = 200. SB, EBA2 and every penalized law (pEBA, penalized All, pOLS) fail the
same way: they flatten the spectrum's spread. The laws that keep it (All,
scaled F, mean-variance, scaled-shifted, EBA6) stay near 5%. All, the
exact weighted chi-square tail of the estimated spectrum, was chosen and then
confirmed on fresh draws: 2.9–6.8% (median 5.1%), with no null cell outside
[3%, 7%] at N ≥ 500.

**The nested profile law had an assembly artifact; the parameter-space law
works.** The separate-point profile law produced 78–373 spectrum terms for
10–15 restrictions, because each model's profile is built at its own fitted
point and residual-dependent blocks cancel only up to $O_p(N^{-1/2})$. Its SB
tail rejected 0.1–2% and its PEBA4 tail 3–17%, worsening with N. The
first-order law

$$
\lambda = \operatorname{eig}\big\{(A H^{-1} A^\top)^{-1} A H^{-1} B H^{-1} A^\top\big\}
$$

has exactly $r$ terms ($H$ the observed Hessian at the larger fit, $B$ the
estimated-weight IJ meat, $A$ the exact restriction map). On fresh draws its
SB and PEBA4 tails reject 4.3–7.8%, with no cell outside [3%, 7%] at
N = 1000; the cells above 7% are all at N = 400. The fixed-weight
Satorra-2000 comparator rejects 4.8–11.4%. Size-adjusted power is the same
for every arm, so the choices cost no power.

```{r}
read_run <- function(run) {
  root <- file.path('results', 'dwls-policy', run)
  if (!file.exists(file.path(root, 'summary.csv')))
    stop('Frozen results for ', run, ' are absent; see criteria/dwls_policy.md for the run commands.')
  s <- read.csv(file.path(root, 'summary.csv'))
  merge(s, read.csv(file.path(root, 'cells.csv')), by = 'cell_id')
}
prod <- read_run('production-2026-10-03')
expl <- read_run('explore-2026-10-03')
cg <- read_run('confirm-global-2026-10-03')
cn <- read_run('confirm-nested-2026-10-03')
pct <- function(x) sprintf('%.1f', 100 * x)
arm_table <- function(z, arms, labels, judged = function(x) x$n >= 500) {
  do.call(rbind, lapply(seq_along(arms), function(i) {
    x <- z[z$role == 'null' & z$arm == arms[i], ]
    out <- x$rate < .03 | x$rate > .07
    data.frame(Reference = labels[i], Min = pct(min(x$rate)), Median = pct(median(x$rate)),
      Max = pct(max(x$rate)), `Outside [3, 7]` = sprintf('%d/%d', sum(out), nrow(x)),
      `Judged cells outside` = sprintf('%d/%d', sum(out & judged(x)), sum(judged(x))),
      check.names = FALSE)
  }))
}
```

## Covariance

Coverage of nominal 95% Wald intervals by target and covariance, over the
coverage cells. Pseudo-true values come from a population DWLS fit to one
100,000-row draw per group. The registered flag is IJ coverage below 93% at
N ≥ 300.

```{r}
cv <- prod[prod$family == 'coverage', ]
tab <- do.call(rbind, lapply(c('loading', 'threshold', 'correlation', 'path'), function(t)
  do.call(rbind, lapply(c('policy_ij', 'expected', 'observed'), function(a) {
    x <- cv[cv$target == t & cv$arm == a, ]
    data.frame(Target = t, Covariance = c(policy_ij = 'IJ (policy)', expected = 'Expected sandwich',
      observed = 'Observed sandwich')[[a]], Cells = nrow(x), Min = pct(min(x$rate)),
      Median = pct(median(x$rate)), Max = pct(max(x$rate)),
      `Below 93% at N >= 300` = sum(x$rate < .93 & x$n >= 300), check.names = FALSE)
  }))))
knitr::kable(tab, row.names = FALSE)
```

The three flagged IJ cells are factor correlations in two-group theta models at
N = 300 (92.8–93.0%); the fixed-weight sandwiches flag eight correlation cells each.

## Global test

The statistic is $nF$, equal to the score statistic of the least-squares fit,
with the `robust_ordinal` spectrum. Production evaluated four references.
An exploratory rerun with the production seeds, saving every spectrum,
evaluated the implemented family; it reproduced production's four arms exactly
(all 256 cell-by-arm counts). Rejection rates at nominal 5% over the 52 null
cells; judged cells are those with N ≥ 500.

```{r}
arms <- c('policy_sb', 'policy_peba4', 'scaled_shifted', 'mean_variance', 'scaled_f', 'all',
          'pall', 'eba2', 'eba4', 'eba6', 'peba2', 'peba6', 'pols')
labels <- c('SB (production policy)', 'PEBA4 (production policy)', 'Scaled-shifted',
            'Mean-variance', 'Scaled F', 'All (exact tail)', 'Penalized All', 'EBA2', 'EBA4',
            'EBA6', 'pEBA2', 'pEBA6', 'pOLS')
knitr::kable(arm_table(expl, arms, labels), row.names = FALSE)
```

The failures grow with df: at df = 9 every reference is near 5%. SB matches
only the spectrum's mean, and the penalized laws pull the eigenvalues toward
their mean, so all of them understate the spread of DWLS spectra at large df.

Confirmation of All on fresh draws (seed base 817160001), with the same
comparators recomputed:

```{r}
knitr::kable(arm_table(cg, c('policy_all', 'mean_variance', 'scaled_shifted', 'eba6', 'policy_sb', 'policy_peba4'),
  c('All (policy)', 'Mean-variance', 'Scaled-shifted', 'EBA6', 'SB', 'PEBA4')), row.names = FALSE)
```

The only null cell outside [3%, 7%] is binary, skewed, df = 132, N = 200
(2.95%), below the judged sample sizes.

## Nested test

Two groups with identical populations; metric versus configural and the
Wu–Estabrook step (thresholds versus thresholds plus loadings, theta); the
larger model correct or with an omitted .3 cross-loading in both groups, so
invariance holds at the pseudo-true values. Judged cells are those with
N = 1000.

```{r}
judged <- function(x) x$n >= 500
prod_nested <- arm_table(prod[prod$family == 'nested', ], c('policy_sb', 'policy_peba4', 'fixed_sb', 'fixed_peba4'),
  c('Profile law SB (production policy)', 'Profile law PEBA4 (production policy)',
    'Fixed-weight Satorra-2000 SB', 'Fixed-weight Satorra-2000 PEBA4'), judged)
conf_nested <- arm_table(cn[cn$family == 'nested', ], c('policy_sb', 'policy_peba4', 'all', 'profile_sb', 'profile_peba4', 'fixed_sb'),
  c('Parameter-space law SB (policy)', 'Parameter-space law PEBA4 (policy)', 'Parameter-space law All',
    'Profile law SB', 'Profile law PEBA4', 'Fixed-weight Satorra-2000 SB'), judged)
knitr::kable(prod_nested, row.names = FALSE, caption = 'Production (seed base 817150001)')
knitr::kable(conf_nested, row.names = FALSE, caption = 'Confirmation (seed base 817160001)')
```

The [diagnosis](diagnostics/nested_profile_law.md) replayed 100 production
draws in three cells. The profile law's positive trace exceeded the Monte Carlo
mean of $T$ while the parameter-space law matched it (for example 10.80 against
11.24 in the five-category N = 1000 cell); evaluated at one common point, the
profile law and the parameter-space law agree to $4 \times 10^{-14}$.

## Power

Size-adjusted power uses each arm's empirical 95th percentile in the matched
null cell. Within a run every arm has the same size-adjusted power, because the
arms are monotone in one statistic per draw.

```{r}
pw <- rbind(
  data.frame(Run = 'Global confirmation', x = cg[cg$role == 'power' & cg$arm == 'policy_all', ]),
  data.frame(Run = 'Nested confirmation', x = cn[cn$role == 'power' & cn$arm == 'policy_sb', ]))
tab <- do.call(rbind, lapply(split(pw, pw$Run), function(x)
  data.frame(Run = x$Run[1], Cells = nrow(x), Min = pct(min(x$x.size_adjusted_power)),
    Max = pct(max(x$x.size_adjusted_power)))))
knitr::kable(tab, row.names = FALSE)
```

## Design and limits

The population is a thresholded Gaussian copula: one to three factors with six
indicators each, two or five categories, symmetric or skewed thresholds, one or
two groups, and delta or theta parameterization. N is total across groups.
Production used 2,000 replicates per null and coverage cell and 1,000 per power
cell; the confirmations used the same counts on the fresh seed base. A smoke
run and a 20-draw pilot (frozen under `results/dwls-policy/`) checked the
runner and priced production at about 59 CPU-hours; production took 73. Library
convergence verdicts decide acceptance; failed draws stay in attempted counts
and leave the successful-arm denominators (312 failures in production, 3 in the
global and 42 in the nested confirmation, all nonconvergence verdicts).

The global reference was selected after seeing production and exploration data,
then confirmed on fresh draws under a criterion fixed beforehand. The nested
law was derived from first-order theory after a diagnosis, also confirmed on
fresh draws. A Gaussian copula does not stress the polychoric step itself;
heavier-tailed latent responses are outside this study. Threshold-invariance
tests that are not parameter-nested (thresholds versus loadings) are outside
it as well.

## Reproduce

Runs execute on Modal, one container per cell (`modal/app.py`); `combine.R`
reproduces the single-process summaries. From the study's `modal/` directory:

```sh
modal run app.py --mode production --run-id production-YYYY-MM-DD
modal run app.py --mode explore --run-id explore-YYYY-MM-DD
modal run app.py --mode confirm --family global --run-id confirm-global-YYYY-MM-DD
modal run app.py --mode confirm --family nested --run-id confirm-nested-YYYY-MM-DD
modal volume get magmaan-dwls-policy <run-id>/final <local-dir>
```

Single-process runs use `Rscript run_experiment.R --<mode> --workers 4`
(`--help` lists modes and options). Rendering reads only the frozen summaries
under `results/dwls-policy/`; raw rows and spectra stay untracked.


## Mixed continuous/ordinal lane

The mixed policy is composed, but its calibration remains open. The registered
160-cell design in `criteria/mixed_policy.md` covers Gaussian responses with
three continuous and three ordinal indicators on each of two factors, binary
and five-category symmetric/skewed thresholds, N 300–2000, and one/two groups
under delta/theta. A smoke checks global, metric, threshold moment nesting and
coverage availability. The full-grid pilot uses two draws per cell, with two
workers and one math thread per worker under `nice -n 10`. These draws price
production; they cannot resolve a 3–7% size rule or 93% coverage rule.

Policy inference uses the library convergence verdict. Failed or unavailable
arms remain in attempted denominators, with Wilson intervals and a separate
successful-arm rate. The policy's global All reference is compared with SB and
PEBA4 on the same spectrum and installed lavaan's default mixed WLSMV test.
Nested SB/PEBA4 is compared with `lavTestLRT`; five-category theta cells include
thresholds-equal versus configural moment nesting. Coverage targets are a
continuous loading, ordinal loading, correlation, threshold and continuous
intercept under omitted cross-loadings 0, .2 and .4. Population targets use
fixed 100,000-row Gaussian draws and population DWLS weights. The mixed lab
interface has no OPG-first-stage IJ comparator or mixed `evaluate_at` adapter;
lavaan arms use lavaan's convergence verdict. That difference needs review
before production and prevents claiming a common acceptance audit.

```{r}
mixed_root <- file.path('results','mixed','pilot-2026-10-05')
if (!file.exists(file.path(mixed_root,'cost.csv')))
  stop('Mixed pilot missing: Rscript run_experiment.R --pilot --lane mixed --workers 2 --run-id pilot-2026-10-05')
mixed_cost <- read.csv(file.path(mixed_root,'cost.csv'))
mixed_summary <- read.csv(file.path(mixed_root,'summary.csv'))
mixed_timing <- read.csv(file.path(mixed_root,'timing.csv'))
knitr::kable(data.frame(
  `Priced cells`=mixed_cost$priced_cells,
  `Attempted draws`=sum(mixed_timing$attempted),
  `Failed arm outcomes`=sum(mixed_summary$failed),
  `Production CPU hours`=round(mixed_cost$production_cpu_hours_priced,1),
  `Ideal four-worker hours`=round(mixed_cost$production_cpu_hours_priced/4,1),
  check.names=FALSE),row.names=FALSE)
```

The pilot completed all 320 draws in 662 seconds, plus 24 seconds for the
population targets. Four of 1,400 arm outcomes were unavailable: two policy
nested outcomes from one nonconverged larger fit, one lavaan nonconvergence,
and one lavaan empty-category error. The full-grid production estimate is
320.4 CPU-hours, or 80.1 hours at an ideal four-worker allocation.

The production price multiplies each cell's observed mean replicate CPU time by
2,000 (null/coverage) or 1,000 (power). It excludes the one-time population-target
cost, cloud startup, billing overhead and retry costs. Two draws give a noisy
price. Cell-level rates, Wilson intervals, failures and timing are frozen CSVs;
raw draws stay ignored. No pilot flag changes the recipe. Production uses fresh
seed base 1617000001 and requires merger review of the registration before any
launch. Power is size-adjusted using matched nulls, reported without gating.

The two-cell local fan-out reproduces single-process summaries exactly and
rejects mixed native-library provenance. Completed cells carry a marker written
after all draws and gates; remote attempts use separate directories and publish
atomically, so partial `raw.rds` checkpoints cannot count as completed cells.
Resuming a run ID reuses completed cells, and combine checks source/native
fingerprints, seeds, replicate counts and population targets before summarizing.
No Modal or production run was launched for TASK-80.

```sh
Rscript run_experiment.R --smoke --lane mixed --cell 1,49,113,129 --workers 1 --run-id mixed-smoke
Rscript run_experiment.R --pilot --lane mixed --workers 2 --run-id mixed-pilot
# After merger review and compute authorization only:
modal run modal/app.py --lane mixed --mode production --run-id mixed-production-YYYY-MM-DD
modal volume get magmaan-dwls-policy mixed/<run-id>/final <local-dir>
```


## Exact first-stage reconfirmation (lane A)

The policy now uses exact first-stage sampling influence while keeping the
OPG fitting weights. Earlier all-ordinal results therefore do not establish
calibration of the current policy. `criteria/exact_first_stage.md` registers
fresh Gaussian-copula draws for the same 146 cells and counts; latent
non-normality belongs to TASK-84. On each draw, exact and OPG inference share
fitted points, with paired rejection/coverage differences and bootstrap
intervals. Nested All is a comparator for TASK-81; policy SB/PEBA4 remain.

```{r}
exact_root <- file.path('results','exact-first-stage','pilot-2026-10-05')
if(!file.exists(file.path(exact_root,'summary.csv')))
  stop('Exact pilot missing: run run_experiment.R --pilot --lane exact-first-stage --run-id pilot-2026-10-05')
exact_cells <- read.csv(file.path(exact_root,'cells.csv'))
exact_summary <- read.csv(file.path(exact_root,'summary.csv'))
exact_timing <- read.csv(file.path(exact_root,'timing.csv'))
exact_cost <- read.csv(file.path(exact_root,'cost.csv'))
exact_failures <- read.csv(file.path(exact_root,'failures.csv'))
knitr::kable(data.frame(
  Cells=nrow(exact_cells), Draws=sum(exact_timing$attempted),
  `Failed arm rows`=nrow(exact_failures),
  `Projected production CPU hours`=round(exact_cost$production_cpu_hours_priced,1),
  check.names=FALSE))
```

The full-grid pilot completed 292 draws in 265.5 seconds, plus 9.6 seconds
for population targets, with no failed arm outcomes. The largest policy
reconstruction gap was 1.37e-14. Its projected production cost is 67.6 CPU-hours.
Two draws per cell cannot resolve rejection or coverage rates. `paired.csv` retains each cell/target/arm difference and jointly
available/excluded counts; `summary.csv` retains attempted-denominator rates,
Wilson intervals and successful-arm diagnostics. Each family and every loss
remains visible. Policy flags use [3%,7%] null rejection for N >= 500 and
93% coverage for N >= 300; a flag calls for finite-sample work, never OPG
restoration. Production pricing is a linear extrapolation from two draws,
not a runtime guarantee. Population pseudo-targets use a fresh fixed 100000-row
sample per group and remain Monte Carlo approximations.

Local two-cell fan-out (cells 1 and 53) reproduces full-pilot summaries and
paired intervals exactly. Completed cell outputs are published only after
all draws and policy-equivalence gates; resume and combine reject altered
sources, native packages, mode, seeds or replicate counts. Partial attempts
remain separate. Modal support is implemented but has not been launched.
Merger registration and a separate compute decision are required before
production.

```bash
R_LIBS=~/.cache/magmaan-rlib/lane-d OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  nice -n 10 Rscript run_experiment.R --smoke --lane exact-first-stage \
  --cell 1,53,85 --run-id smoke-fresh
R_LIBS=~/.cache/magmaan-rlib/lane-d OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  nice -n 10 Rscript run_experiment.R --pilot --lane exact-first-stage \
  --workers 1 --run-id pilot-fresh
# Only after registration/compute approval and Modal workspace restoration:
modal run modal/app.py --lane exact-first-stage --mode production --run-id production-fresh
```
