---
title: "DWLS policy calibration"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Default register

| Choice | Current value | Status | Evidence |
|---|---|---|---|
| Covariance | Estimated-weight IJ sandwich | Adopted by principle; calibrated | Production 2026-10-03: median coverage about 95% for every target |
| Global test | $nF$ on the `robust_ordinal` spectrum, exact weighted chi-square tail with every positive eigenvalue (All) | Chosen after exploration (user, 2026-10-03); confirmed on fresh draws | SB and PEBA4 over-reject at df ≥ 53; All 2.9–6.8% at confirmation |
| Nested test | $N(F_{H_0} - F_{H_1})$ with the $r$-term parameter-space law (observed Hessian, estimated-weight IJ meat, exact restriction map), SB and PEBA4 | Replaced the separate-point profile law after diagnosis; confirmed on fresh draws | Profile law 0.1–2% (SB) and 3–17% (PEBA4); parameter-space law 4.3–7.8% |
| Nested score test | Typed unavailable | Decision D4 | No joint least-squares nested score primitive |

## Question

How well calibrated are the all-ordinal DWLS policy components? The recipes'
ingredients are fixed by the misspecification-robust requirement
([scope](../../../project/scope.md)): IJ covariance with the estimated weight's
influence, the robust-ordinal spectrum for the global test, and an
estimated-weight law for the nested test. Which reference-law approximation to
use, and whether a law is correctly assembled, are empirical questions this
study answers. The registered design, its amendments and the confirmation
criteria are in `criteria/dwls_policy.md`.

## Short answer

**The IJ covariance is calibrated.** Across 42 coverage cells (misspecification
0, .2 and .4; N 150 to 1000; binary and five categories; one and two groups)
its median coverage is about 95% for loadings, thresholds, correlations and
paths, with a minimum of 92.3%. The fixed-weight expected and observed
sandwiches fall to 89% for correlations and paths under misspecification.

**The global test needs the exact spectrum tail.** SB and PEBA4 on the robust
spectrum over-reject once df ≥ 53, at 8–12% even at N = 1000 and up to 20% at
N = 200. SB, EBA2 and every penalized law (pEBA, penalized All, pOLS) fail the
same way: they flatten the spectrum's spread. The laws that keep it (All,
scaled F, mean-variance, scaled-shifted, EBA6) stay near 5%. All, the
exact weighted chi-square tail of the estimated spectrum, was chosen and then
confirmed on fresh draws: 2.9–6.8% (median 5.1%), with no null cell outside
[3%, 7%] at N ≥ 500.

**The nested profile law had an assembly artifact; the parameter-space law
works.** The separate-point profile law produced 78–373 spectrum terms for
10–15 restrictions, because each model's profile is built at its own fitted
point and residual-dependent blocks cancel only up to $O_p(N^{-1/2})$. Its SB
tail rejected 0.1–2% and its PEBA4 tail 3–17%, worsening with N. The
first-order law

$$
\lambda = \operatorname{eig}\big\{(A H^{-1} A^\top)^{-1} A H^{-1} B H^{-1} A^\top\big\}
$$

has exactly $r$ terms ($H$ the observed Hessian at the larger fit, $B$ the
estimated-weight IJ meat, $A$ the exact restriction map). On fresh draws its
SB and PEBA4 tails reject 4.3–7.8%, with no cell outside [3%, 7%] at
N = 1000; the cells above 7% are all at N = 400. The fixed-weight
Satorra-2000 comparator rejects 4.8–11.4%. Size-adjusted power is the same
for every arm, so the choices cost no power.

```{r}
read_run <- function(run) {
  root <- file.path('results', 'dwls-policy', run)
  if (!file.exists(file.path(root, 'summary.csv')))
    stop('Frozen results for ', run, ' are absent; see criteria/dwls_policy.md for the run commands.')
  s <- read.csv(file.path(root, 'summary.csv'))
  merge(s, read.csv(file.path(root, 'cells.csv')), by = 'cell_id')
}
prod <- read_run('production-2026-10-03')
expl <- read_run('explore-2026-10-03')
cg <- read_run('confirm-global-2026-10-03')
cn <- read_run('confirm-nested-2026-10-03')
pct <- function(x) sprintf('%.1f', 100 * x)
arm_table <- function(z, arms, labels, judged = function(x) x$n >= 500) {
  do.call(rbind, lapply(seq_along(arms), function(i) {
    x <- z[z$role == 'null' & z$arm == arms[i], ]
    out <- x$rate < .03 | x$rate > .07
    data.frame(Reference = labels[i], Min = pct(min(x$rate)), Median = pct(median(x$rate)),
      Max = pct(max(x$rate)), `Outside [3, 7]` = sprintf('%d/%d', sum(out), nrow(x)),
      `Judged cells outside` = sprintf('%d/%d', sum(out & judged(x)), sum(judged(x))),
      check.names = FALSE)
  }))
}
```

## Covariance

Coverage of nominal 95% Wald intervals by target and covariance, over the
coverage cells. Pseudo-true values come from a population DWLS fit to one
100,000-row draw per group. The registered flag is IJ coverage below 93% at
N ≥ 300.

```{r}
cv <- prod[prod$family == 'coverage', ]
tab <- do.call(rbind, lapply(c('loading', 'threshold', 'correlation', 'path'), function(t)
  do.call(rbind, lapply(c('policy_ij', 'expected', 'observed'), function(a) {
    x <- cv[cv$target == t & cv$arm == a, ]
    data.frame(Target = t, Covariance = c(policy_ij = 'IJ (policy)', expected = 'Expected sandwich',
      observed = 'Observed sandwich')[[a]], Cells = nrow(x), Min = pct(min(x$rate)),
      Median = pct(median(x$rate)), Max = pct(max(x$rate)),
      `Below 93% at N >= 300` = sum(x$rate < .93 & x$n >= 300), check.names = FALSE)
  }))))
knitr::kable(tab, row.names = FALSE)
```

The three flagged IJ cells are factor correlations in two-group theta models at
N = 300 (92.8–93.0%); the fixed-weight sandwiches flag eight correlation cells each.

## Global test

The statistic is $nF$, equal to the score statistic of the least-squares fit,
with the `robust_ordinal` spectrum. Production evaluated four references.
An exploratory rerun with the production seeds, saving every spectrum,
evaluated the implemented family; it reproduced production's four arms exactly
(all 256 cell-by-arm counts). Rejection rates at nominal 5% over the 52 null
cells; judged cells are those with N ≥ 500.

```{r}
arms <- c('policy_sb', 'policy_peba4', 'scaled_shifted', 'mean_variance', 'scaled_f', 'all',
          'pall', 'eba2', 'eba4', 'eba6', 'peba2', 'peba6', 'pols')
labels <- c('SB (production policy)', 'PEBA4 (production policy)', 'Scaled-shifted',
            'Mean-variance', 'Scaled F', 'All (exact tail)', 'Penalized All', 'EBA2', 'EBA4',
            'EBA6', 'pEBA2', 'pEBA6', 'pOLS')
knitr::kable(arm_table(expl, arms, labels), row.names = FALSE)
```

The failures grow with df: at df = 9 every reference is near 5%. SB matches
only the spectrum's mean, and the penalized laws pull the eigenvalues toward
their mean, so all of them understate the spread of DWLS spectra at large df.

Confirmation of All on fresh draws (seed base 817160001), with the same
comparators recomputed:

```{r}
knitr::kable(arm_table(cg, c('policy_all', 'mean_variance', 'scaled_shifted', 'eba6', 'policy_sb', 'policy_peba4'),
  c('All (policy)', 'Mean-variance', 'Scaled-shifted', 'EBA6', 'SB', 'PEBA4')), row.names = FALSE)
```

The only null cell outside [3%, 7%] is binary, skewed, df = 132, N = 200
(2.95%), below the judged sample sizes.

## Nested test

Two groups with identical populations; metric versus configural and the
Wu–Estabrook step (thresholds versus thresholds plus loadings, theta); the
larger model correct or with an omitted .3 cross-loading in both groups, so
invariance holds at the pseudo-true values. Judged cells are those with
N = 1000.

```{r}
judged <- function(x) x$n >= 500
prod_nested <- arm_table(prod[prod$family == 'nested', ], c('policy_sb', 'policy_peba4', 'fixed_sb', 'fixed_peba4'),
  c('Profile law SB (production policy)', 'Profile law PEBA4 (production policy)',
    'Fixed-weight Satorra-2000 SB', 'Fixed-weight Satorra-2000 PEBA4'), judged)
conf_nested <- arm_table(cn[cn$family == 'nested', ], c('policy_sb', 'policy_peba4', 'all', 'profile_sb', 'profile_peba4', 'fixed_sb'),
  c('Parameter-space law SB (policy)', 'Parameter-space law PEBA4 (policy)', 'Parameter-space law All',
    'Profile law SB', 'Profile law PEBA4', 'Fixed-weight Satorra-2000 SB'), judged)
knitr::kable(prod_nested, row.names = FALSE, caption = 'Production (seed base 817150001)')
knitr::kable(conf_nested, row.names = FALSE, caption = 'Confirmation (seed base 817160001)')
```

The [diagnosis](diagnostics/nested_profile_law.md) replayed 100 production
draws in three cells. The profile law's positive trace exceeded the Monte Carlo
mean of $T$ while the parameter-space law matched it (for example 10.80 against
11.24 in the five-category N = 1000 cell); evaluated at one common point, the
profile law and the parameter-space law agree to $4 \times 10^{-14}$.

## Power

Size-adjusted power uses each arm's empirical 95th percentile in the matched
null cell. Within a run every arm has the same size-adjusted power, because the
arms are monotone in one statistic per draw.

```{r}
pw <- rbind(
  data.frame(Run = 'Global confirmation', x = cg[cg$role == 'power' & cg$arm == 'policy_all', ]),
  data.frame(Run = 'Nested confirmation', x = cn[cn$role == 'power' & cn$arm == 'policy_sb', ]))
tab <- do.call(rbind, lapply(split(pw, pw$Run), function(x)
  data.frame(Run = x$Run[1], Cells = nrow(x), Min = pct(min(x$x.size_adjusted_power)),
    Max = pct(max(x$x.size_adjusted_power)))))
knitr::kable(tab, row.names = FALSE)
```

## Design and limits

The population is a thresholded Gaussian copula: one to three factors with six
indicators each, two or five categories, symmetric or skewed thresholds, one or
two groups, and delta or theta parameterization. N is total across groups.
Production used 2,000 replicates per null and coverage cell and 1,000 per power
cell; the confirmations used the same counts on the fresh seed base. A smoke
run and a 20-draw pilot (frozen under `results/dwls-policy/`) checked the
runner and priced production at about 59 CPU-hours; production took 73. Library
convergence verdicts decide acceptance; failed draws stay in attempted counts
and leave the successful-arm denominators (312 failures in production, 3 in the
global and 42 in the nested confirmation, all nonconvergence verdicts).

The global reference was selected after seeing production and exploration data,
then confirmed on fresh draws under a criterion fixed beforehand. The nested
law was derived from first-order theory after a diagnosis, also confirmed on
fresh draws. A Gaussian copula does not stress the polychoric step itself;
heavier-tailed latent responses are outside this study. Threshold-invariance
tests that are not parameter-nested (thresholds versus loadings) are outside
it as well.

## Reproduce

Runs execute on Modal, one container per cell (`modal/app.py`); `combine.R`
reproduces the single-process summaries. From the study's `modal/` directory:

```sh
modal run app.py --mode production --run-id production-YYYY-MM-DD
modal run app.py --mode explore --run-id explore-YYYY-MM-DD
modal run app.py --mode confirm --family global --run-id confirm-global-YYYY-MM-DD
modal run app.py --mode confirm --family nested --run-id confirm-nested-YYYY-MM-DD
modal volume get magmaan-dwls-policy <run-id>/final <local-dir>
```

Single-process runs use `Rscript run_experiment.R --<mode> --workers 4`
(`--help` lists modes and options). Rendering reads only the frozen summaries
under `results/dwls-policy/`; raw rows and spectra stay untracked.
