---
title: "DWLS policy calibration"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

The adopted register is estimated-weight IJ covariance, robust-ordinal global
fit-function SB/PEBA4, and estimated-weight nested profile-law SB/PEBA4.
These recipes are settled by the misspecification-robust requirement;
finite-sample calibration remains open. Comparators never select a replacement.

How well calibrated are these all-ordinal DWLS components? The smoke and pilot
check computation, policy equivalence, failures and cost. Twenty draws per cell
cannot establish calibration, coverage or power. The main tables below describe
the pilot; the linked nested-law diagnostic uses a bounded production-seed replay.

```{r}
root <- file.path('results','dwls-policy','pilot')
if (!file.exists(file.path(root,'cost.csv')))
  stop('Run Rscript run_experiment.R --pilot --workers 4 --run-id pilot before rendering.')
cost <- read.csv(file.path(root,'cost.csv'))
timing <- read.csv(file.path(root,'timing.csv'))
summary <- read.csv(file.path(root,'summary.csv'))
cells <- read.csv(file.path(root,'cells.csv'))
```

The pilot attempted `r cost$replicates` draws across `r cost$cells` cells, with
`r cost$failures` failed draws. Per-cell extrapolation gives
`r round(cost$production_cpu_hours,1)` worker-hours for production, or an ideal
`r round(cost$ideal_four_worker_hours,1)` hours with four workers. This includes
fitting and inference; population-target preparation is separate and performed
once per run. Elapsed timings include a brief scheduling pause for the shared
compute budget, recorded in `scheduling.csv`; CPU times exclude that pause. Detailed timings, Wilson intervals, target-specific coverage,
failure counts and policy gaps are in the frozen CSVs. These are pilot
engineering evidence, not confirmatory findings. The two failures are library
nonconvergence verdicts in the two-group theta coverage cell with binary
indicators, total N=300 and omitted cross-loading .4; all other draws completed.

```{r}
z <- merge(timing,cells[c('cell_id','family','role')],by='cell_id')
table <- do.call(rbind,lapply(split(z,paste(z$family,z$role)),function(x)
  data.frame(Family=x$family[1],Role=x$role[1],Cells=nrow(x),
    `Median CPU seconds per draw`=round(median(x$mean_cpu_seconds),3),
    `Production worker-hours`=round(sum(x$production_cpu_hours),2),check.names=FALSE)))
knitr::kable(table,row.names=FALSE)
```

The population is a thresholded Gaussian copula. Sample size is total across
groups. Population DWLS targets use one 100,000-row draw per group for each
coverage configuration, with seeds and target estimates recorded separately.
Intervals use the IJ policy covariance or the expected/observed fixed-weight
sandwich comparators, with a delta-method factor correlation. Global comparators
are scaled-shifted and mean-variance adjustments of the same robust-ordinal
spectrum. Nested comparators use the fixed-weight Satorra-2000 restriction map
at H1. Restriction degrees of freedom and reference-spectrum size are recorded
separately; the profile-law spectrum may contain more terms than restrictions.
Power uses matched null draws for arm-specific empirical size adjustment.

The original availability preflight is retained. The approved correction uses
H1 threshold equality and H0 threshold-plus-loading equality under theta,
the Wu–Estabrook metric step. The earlier loadings-to-thresholds-plus-loadings
pair changes fixed scales/intercepts and correctly returned `not_nested`.
The second amendment drops threshold-shift power: both corrected models impose
equal thresholds, so that perturbation would misspecify both models rather than
test threshold invariance. Loading power remains for metric versus configural.
A genuine threshold-invariance restriction map needs separate validation.

The 146-cell grid and both amendments are in `criteria/dwls_policy.md`.
Finite-sample flags use the registered thresholds, but pilot flags are only
nominations for follow-up. All convergence verdicts come from the library.
Failures remain in attempted counts and are excluded only from explicitly
reported successful-arm denominators. Monte Carlo error in population targets
and the small pilot limit interpretation. Production is pending a separate
compute decision; production evidence must use its distinct registered seeds.

Tiny pilot summaries, metadata, cells and population targets are frozen to
preserve timing and provenance; raw per-fit rows remain ignored. The native
library hash and source hashes identify the installed package and runner.

Reproduce with the lane library and one math thread per worker:

```sh
R_LIBS=~/.cache/magmaan-rlib/lane-d OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  nice -n 10 Rscript experiments/decisions/05-dwls-policy-calibration/run_experiment.R \
  --smoke --workers 1 --run-id fresh-smoke
R_LIBS=~/.cache/magmaan-rlib/lane-d OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 MKL_NUM_THREADS=1 \
  nice -n 10 Rscript experiments/decisions/05-dwls-policy-calibration/run_experiment.R \
  --pilot --workers 4 --run-id fresh-pilot
quarto render experiments/decisions/05-dwls-policy-calibration/report.qmd
```

Runner `--help` documents modes. Every run needs a fresh directory. Rendering
reads frozen results only. Production is never launched automatically.

## Nested reference-law diagnostic

The [production-seed replay](diagnostics/nested_profile_law.md) isolates the
separate-fitted-point cancellation artifact: 100 draws in each of three nested
cells yield 83–183 positive profile terms for ten restrictions. The common-point
profile spectrum agrees with the observed-Hessian parameter-space IJ law to
numerical precision and has exactly ten terms. Parameter-space SB/pEBA4 rejection
is 5–6%, versus profile SB 0–2% and pEBA4 5–18%. This exploratory diagnosis
recommends fresh-seed registered confirmation of the parameter-space law; it
does not change the current policy or establish release calibration.
