---
title: "Experiment 53: Ordinal Profile-LRT Finite-Sample Calibration"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- "results/simulation_summary.csv"
metadata_path <- "results/metadata.csv"
targets_path <- "results/pseudo_targets.csv"
if (!file.exists(summary_path) || !file.exists(metadata_path) ||
    !file.exists(targets_path)) {
  stop("Missing results. Run: Rscript run_experiment.R --smoke")
}
summary <- read.csv(summary_path)
metadata <- read.csv(metadata_path)
targets <- read.csv(targets_path)
if (!"stress" %in% names(metadata)) metadata$stress <- FALSE
ordinary <- summary[summary$arm == "dwls_ordinary", ]
profile_ref <- summary[summary$arm != "dwls_ordinary", ]
worst_ref <- profile_ref[which.min(profile_ref$reference_coverage_95), ]
worst_ordinary <- ordinary[which.min(ordinary$reference_coverage_95), ]
worst_wald <- summary[which.min(summary$omega_wald_coverage_95), ]
has_ci <- "ci_attempted" %in% names(summary) &&
  any(summary$ci_attempted > 0, na.rm = TRUE)
ci_summary <- if (has_ci) {
  summary[summary$ci_attempted > 0, ]
} else {
  summary[FALSE, ]
}
ci_success <- if (has_ci) {
  ci_summary[ci_summary$ci_attempted > ci_summary$ci_failures, ]
} else {
  ci_summary
}
```

## Question

For all-ordinal limited-information CFA, is small-sample miscoverage of
polychoric-omega profile-LRT intervals mainly LR inflation, robust reference
scaling, sparse ordinal summaries, point-functional bias, or working-model
misspecification?

This run intentionally keeps the grid DWLS-only; the WLS robust-scaled
comparison is deferred so the same compute budget buys more replications.
The baseline DGP grid includes the correctly specified one-factor ordinal
probit model and a locally dependent ordinal model fitted by the same one-factor
CFA. The stress grid replaces that with stronger residual local dependence and
a two-factor population fitted by the same one-factor CFA, with smaller sample
sizes and sparse thresholds. Targets are large-sample pseudo-true fitted-profile
omega values for each DGP and cut regime.

## Short Answer

```{r, results='asis'}
cat(sprintf(
  paste0(
    "At %s reps per cell%s, ordinary DWLS reference coverage ranges from %.3f ",
    "to %.3f; the weakest ordinary cell is %s / %s at N = %d. ",
    "Robust-scaled and misspec-scaled reference coverage ranges from %.3f ",
    "to %.3f; the weakest robust/misspec cell is %s / %s at N = %d. ",
    "The weakest Wald delta cell is %s at N = %d with coverage %.3f. ",
    "The grid recorded %d profile construction failures."
  ),
  metadata$reps,
  if (isTRUE(metadata$stress[[1]])) " in the stress grid" else "",
  min(ordinary$reference_coverage_95, na.rm = TRUE),
  max(ordinary$reference_coverage_95, na.rm = TRUE),
  paste(worst_ordinary$dgp, worst_ordinary$regime, sep = " / "),
  worst_ordinary$arm, worst_ordinary$n,
  min(profile_ref$reference_coverage_95, na.rm = TRUE),
  max(profile_ref$reference_coverage_95, na.rm = TRUE),
  paste(worst_ref$dgp, worst_ref$regime, sep = " / "), worst_ref$arm,
  worst_ref$n,
  paste(worst_wald$dgp, worst_wald$regime, sep = " / "),
  worst_wald$n, worst_wald$omega_wald_coverage_95,
  sum(summary$failures)))
if (has_ci && nrow(ci_success)) {
  cat(sprintf(
    paste0(
      " CI inversion was attempted for %s rows; CI coverage among successful ",
      "inversions ranges from %.3f to %.3f, with %d CI failures and %d ",
      "pointwise-LRT/CI coverage disagreements."
    ),
    sum(summary$ci_attempted),
    min(ci_success$ci_coverage_95, na.rm = TRUE),
    max(ci_success$ci_coverage_95, na.rm = TRUE),
    sum(summary$ci_failures),
    sum(summary$ci_lrt_disagreements, na.rm = TRUE)))
} else if (has_ci) {
  cat(sprintf(
    " CI inversion was attempted for %s rows, but no CI inversions succeeded.",
    sum(summary$ci_attempted)))
}
```

## Pseudo-Targets

```{r}
target_shown <- targets
names(target_shown) <- c(
  "DGP", "Regime", "Pseudo N", "Profile Target",
  "Sample Polychoric Target", "Profile - Sample"
)
target_shown[] <- lapply(target_shown, function(x) {
  if (is.numeric(x)) round(x, 5) else x
})
knitr::kable(target_shown)
```

## Profile References

```{r}
shown <- summary[, c(
  "dgp", "regime", "n", "arm", "reference", "failures", "mean_reference_stat",
  "reference_coverage_95", "oracle_constant_coverage_95",
  "mean_scaling_factor", "mean_misspec_scaling_factor",
  "min_variable_category_count_p05"
)]
names(shown) <- c(
  "DGP", "Regime", "N", "Arm", "Reference", "Failures", "Mean Ref Stat",
  "Reference Coverage", "Oracle Constant Coverage",
  "Mean Robust Scale", "Mean Misspec Scale", "P05 Min Category Count"
)
shown[] <- lapply(shown, function(x) if (is.numeric(x)) round(x, 4) else x)
knitr::kable(shown)
```

## Omega Point And Wald Baseline

```{r}
wald <- unique(summary[, c(
  "dgp", "regime", "n", "target", "sample_polychoric_target",
  "target_minus_sample_polychoric", "mean_omega_wald", "omega_bias",
  "omega_wald_coverage_95"
)])
names(wald) <- c(
  "DGP", "Regime", "N", "Profile Target", "Pseudo Sample Target",
  "Profile - Sample", "Mean Omega", "Bias", "Wald Coverage"
)
wald[] <- lapply(wald, function(x) if (is.numeric(x)) round(x, 4) else x)
knitr::kable(wald)
```

```{r, results='asis'}
if (has_ci) {
  cat("## Profile CI Inversion\n\n")
  ci_shown <- ci_summary[, c(
    "dgp", "regime", "n", "arm", "ci_attempted", "ci_failures",
    "ci_coverage_95", "mean_ci_width", "median_ci_width", "mean_ci_evals",
    "ci_bound_rate", "max_abs_ci_endpoint_gap", "ci_lrt_disagreements",
    "ci_lrt_disagreement_rate"
  )]
  names(ci_shown) <- c(
    "DGP", "Regime", "N", "Arm", "CI Attempts", "CI Failures",
    "CI Coverage", "Mean Width", "Median Width", "Mean Evals",
    "Bound Rate", "Max Endpoint Gap", "LRT/CI Disagreements",
    "Disagreement Rate"
  )
  ci_shown[] <- lapply(ci_shown, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  print(knitr::kable(ci_shown))
}
```

## Interpretation

The oracle-constant column asks a narrow diagnostic question: if the
reference-scaled statistic were divided by its cell mean, would coverage move
toward nominal? A large gap between reference coverage and oracle-constant
coverage points to LR inflation as a scalar-calibration problem. Poor Wald
coverage with near-nominal profile-reference coverage points instead to
point-functional bias or delta-method SE bias.

The scalar misspecification arm is reported as misspec-scaled. The underlying
weighted-$\chi^2$ reference has one eigenvalue for this one-degree-of-freedom
profile test, so the older misspecification-mixture label is only a legacy alias
for the same scaled reference in this experiment. In the misspecified DGPs the
target is the pseudo-true fitted one-factor omega, so any residual gap is
reference-law or sparse-summary behavior around that working-model target, not
bias relative to a generating-model reliability.

When CI inversion is enabled, the LRT/CI disagreement count is the first
connectedness diagnostic: it compares target inclusion in the inverted interval
with the pointwise profile test at the same target. A disagreement flags either
nonmonotone/disconnected profile behavior or numerical root trouble. The
misspec-scaled CI rows should be read as experimental endpoint diagnostics;
failures usually indicate that an endpoint robust profile quadratic is not
positive, even when the pointwise misspec-scaled test at the target is well
behaved.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R
Rscript run_experiment.R --dgp local_dependence --reps 500
Rscript run_experiment.R --stress --reps 300
Rscript run_experiment.R --smoke --ci
Rscript run_experiment.R --ci --reps 200
quarto render report.qmd
```
