---
title: "Experiment 53: Ordinal Profile-LRT Finite-Sample Calibration"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
summary_path <- "results/simulation_summary.csv"
metadata_path <- "results/metadata.csv"
if (!file.exists(summary_path) || !file.exists(metadata_path)) {
  stop("Missing results. Run: Rscript run_experiment.R --smoke")
}
summary <- read.csv(summary_path)
metadata <- read.csv(metadata_path)
worst_ref <- summary[which.min(summary$reference_coverage_95), ]
worst_wald <- summary[which.min(summary$omega_wald_coverage_95), ]
```

## Question

For all-ordinal limited-information CFA, is small-sample miscoverage of
polychoric-omega profile-LRT intervals mainly LR inflation, robust reference
scaling, sparse ordinal summaries, or point-functional bias?

## Short Answer

```{r, results='asis'}
cat(sprintf(
  paste0(
    "At %s reps per cell and target omega %.3f, profile-reference coverage ",
    "ranges from %.3f to %.3f. The weakest profile cell is %s / %s at N = %d. ",
    "The weakest Wald delta cell is %s at N = %d with coverage %.3f. ",
    "The grid recorded %d profile construction failures."
  ),
  metadata$reps, metadata$target, min(summary$reference_coverage_95, na.rm = TRUE),
  max(summary$reference_coverage_95, na.rm = TRUE),
  worst_ref$regime, worst_ref$arm, worst_ref$n,
  worst_wald$regime, worst_wald$n, worst_wald$omega_wald_coverage_95,
  sum(summary$failures)))
```

## Profile References

```{r}
shown <- summary[, c(
  "regime", "n", "arm", "failures", "mean_reference_stat",
  "reference_coverage_95", "oracle_constant_coverage_95",
  "mean_scaling_factor", "mean_misspec_scaling_factor",
  "min_variable_category_count_p05"
)]
names(shown) <- c(
  "Regime", "N", "Arm", "Failures", "Mean Ref Stat",
  "Reference Coverage", "Oracle Constant Coverage",
  "Mean Robust Scale", "Mean Misspec Scale", "P05 Min Category Count"
)
shown[] <- lapply(shown, function(x) if (is.numeric(x)) round(x, 4) else x)
knitr::kable(shown)
```

## Omega Point And Wald Baseline

```{r}
wald <- unique(summary[, c(
  "regime", "n", "target", "mean_omega_wald", "omega_bias",
  "omega_wald_coverage_95"
)])
names(wald) <- c("Regime", "N", "Target", "Mean Omega", "Bias", "Wald Coverage")
wald[] <- lapply(wald, function(x) if (is.numeric(x)) round(x, 4) else x)
knitr::kable(wald)
```

## Interpretation

The oracle-constant column asks a narrow diagnostic question: if the
reference-scaled statistic were divided by its cell mean, would coverage move
toward nominal? A large gap between reference coverage and oracle-constant
coverage points to LR inflation as a scalar-calibration problem. Poor Wald
coverage with near-nominal profile-reference coverage points instead to
point-functional bias or delta-method SE bias.

The misspecification-mixture arm is included for the reference-law taxonomy, not
because this correctly specified ordinal-probit grid needs a misspecification
target. Its value is to check whether the observed-bread reference differs
materially from the robust-scaled one in sparse ordinal cells.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R
quarto render report.qmd
```
