---
title: "Calibrating the Guttman communality clamp and score covariance"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

## Question

Can a hard or smooth repair of the aligned Guttman score covariance eliminate
ill-conditioned factor covariances after the communality clamp, without harming
loading RMSE, empirical-SE coverage, benign-cell accuracy, or runtime?

## Short answer

Production score conditioning remains `raw`. The screen is deliberately a
promotion gate, not an automatic default change: a candidate must pass every
success, PD-factor-covariance, coverage, stress-tail, benign-RMSE, and runtime
criterion before it is eligible for the larger confirmation run.

The completed 24-cell, 300-replication screen had no survivor. Raw-H arms
frequently failed at the existing improper-communality-split guard; when H was
clamped, some score-conditioned arms still had non-positive score variances.
That is outside the reach of the original Q repair, which deliberately preserves
the score diagonal. The follow-up `--h-feasibility` mode therefore tests a
separate fixed-diagonal PSD repair of H before scores are constructed. It is a
point-estimation diagnostic only: no SE or coverage claim is made until its map
derivative is implemented and validated.

That follow-up was run on all 18 weak/equal-scale stress cells at 100
replications. It restored 100% point-fit success and PD factor covariance, but
every tested repair worsened loading RMSE relative to its clamp-only counterpart.
The normalized-H shrinkage was necessarily large (median roughly 6--12, with
substantial upper tails), and hard and soft variants saturated to the same
practical repair. This is a second no-survivor result; no analytic-SE work or
default promotion is warranted for H repair.

```{r}
root <- dirname(knitr::current_input(dir = TRUE))
results <- file.path(root, "results")
need <- c("estimate_summary.csv", "estimate_joint_summary.csv",
          "diagnostic_summary.csv", "paired_timing_summary.csv")
paths <- file.path(results, need)
if (!all(file.exists(paths))) {
  stop("Missing results. Run: Rscript experiments/60-guttman-admissibility-clamp/run_experiment.R --smoke")
}
estimate <- read.csv(paths[[1L]])
joint <- read.csv(paths[[2L]])
diagnostic <- read.csv(paths[[3L]])
timing <- read.csv(paths[[4L]])
decision_path <- file.path(results, "screen_decisions.csv")
decision <- if (file.exists(decision_path)) read.csv(decision_path) else NULL
```

## Evidence

```{r}
loading <- estimate[estimate$kind == "loading", ]
loading <- aggregate(cbind(rmse, coverage, abs_error_p95, abs_error_p99,
                           abs_estimate_p99, ci_width_p99) ~ estimator,
                     loading, mean,
                     na.rm = TRUE)
joint_loading <- joint[joint$kind == "loading", ]
joint_loading <- aggregate(rmse_ratio ~ estimator + baseline, joint_loading,
                           mean, na.rm = TRUE)
diag_show <- aggregate(cbind(clamp_rate, fit_ok_rate, se_ok_rate, pd_phi_rate,
                             score_raw_normalized_min_eig_p01,
                             score_repaired_normalized_min_eig_p01,
                             phi_min_eig_min, phi_condition_p99,
                             marker_min_p01, shrinkage_rate) ~ estimator,
                       diagnostic, mean, na.rm = TRUE)
knitr::kable(merge(merge(loading, joint_loading, by = "estimator", all = TRUE),
                         diag_show, by = "estimator", all = TRUE), digits = 3)
```

## Promotion gate

```{r}
if (is.null(decision)) {
  cat("No 300-replication screen is present; there is no survivor decision.\n")
} else {
  show <- decision[order(!decision$survive, decision$stress_rmse), ]
  knitr::kable(head(show, 20), digits = 3)
  if (!any(decision$survive))
    cat("\nNo setting survived. Score conditioning remains raw.\n")
  else
    cat("\nEligible confirmation finalists (conditioning still raw in production): ",
        paste(head(decision$arm[decision$survive], 2L), collapse = ", "), "\n",
        sep = "")
}
```

## Caveats

The 24-cell screen uses 300 replications: 18 weak/equal-scale stress cells
(`m=3`, `rho=.8`, `n=50/100`) plus six benign cells (`m=5`, `rho=0`, `n=800`).
Only the best two survivors proceed to the 1,000-replication, 27-cell
normal/IG/ordinal confirmation. Smoke runs establish pipeline integrity only.

## Reproduce

```sh
Rscript experiments/60-guttman-admissibility-clamp/run_experiment.R --smoke \
  --results-dir experiments/60-guttman-admissibility-clamp/results
Rscript experiments/60-guttman-admissibility-clamp/run_experiment.R --screen \
  --results-dir experiments/60-guttman-admissibility-clamp/results
# Follow-up point-fit feasibility screen for fixed-diagonal H repair:
Rscript experiments/60-guttman-admissibility-clamp/run_experiment.R --h-feasibility \
  --results-dir /tmp/exp60-h-feasibility
# If survivors exist, pass the best two explicitly (or reuse screen_survivors.csv):
Rscript experiments/60-guttman-admissibility-clamp/run_experiment.R --confirm \
  --finalists ARM1,ARM2 --results-dir /tmp/exp60-confirm
quarto render experiments/60-guttman-admissibility-clamp/report.qmd
```
