---
title: "Calibrating the Guttman communality clamp"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

## Question

Which hard or soft admissibility setting best repairs out-of-box closed-form
communalities in weak-loading, small-sample CFA without sacrificing loading
RMSE or empirical-SE coverage relative to the raw map and NTML?

## Short answer

No calibration run is checked in. Run the experiment before interpreting a
policy cell; the report will summarize clamp activation, loading RMSE ratios,
coverage, failures, and score-system conditioning.

```{r}
root <- dirname(knitr::current_input(dir = TRUE))
results <- file.path(root, "results")
need <- c("estimate_summary.csv", "estimate_joint_summary.csv",
          "diagnostic_summary.csv")
paths <- file.path(results, need)
if (!all(file.exists(paths))) {
  stop("Missing results. Run: Rscript experiments/60-guttman-admissibility-clamp/run_experiment.R --smoke")
}
estimate <- read.csv(paths[[1L]])
joint <- read.csv(paths[[2L]])
diagnostic <- read.csv(paths[[3L]])
```

## Evidence

```{r}
loading <- estimate[estimate$kind == "loading", ]
loading <- aggregate(cbind(rmse, coverage) ~ estimator, loading, mean,
                     na.rm = TRUE)
joint_loading <- joint[joint$kind == "loading", ]
joint_loading <- aggregate(rmse_ratio ~ estimator + baseline, joint_loading,
                           mean, na.rm = TRUE)
diag_show <- aggregate(cbind(clamp_rate, fit_ok_rate, se_ok_rate,
                             zhz_min_eig_p05, bhb_min_eig_p05) ~ estimator,
                       diagnostic, mean, na.rm = TRUE)
knitr::kable(merge(merge(loading, joint_loading, by = "estimator", all = TRUE),
                         diag_show, by = "estimator", all = TRUE), digits = 3)
```

## Caveats

The experiment is a calibration benchmark, not a lavaan-parity gate. Smoke
runs establish pipeline integrity only; choosing production constants requires
the full replication count and attention to the stress-cell slices rather than
the pooled table alone.

## Reproduce

```sh
Rscript experiments/60-guttman-admissibility-clamp/run_experiment.R --smoke \
  --results-dir experiments/60-guttman-admissibility-clamp/results
quarto render experiments/60-guttman-admissibility-clamp/report.qmd
```
