---
title: "87 · Sphere chart sanity checks"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
res <- file.path("results", "full")
need <- c("recovery", "population", "population_reidentify", "invariance",
          "invariance_reidentify", "coverage")
paths <- file.path(res, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("results missing: run `Rscript run_experiment.R --full` first")
}
rd <- function(n) read.csv(file.path(res, paste0(n, ".csv")), stringsAsFactors = FALSE)
rec <- rd("recovery")
pop <- rd("population")
preid <- rd("population_reidentify")
inv <- rd("invariance")
ireid <- rd("invariance_reidentify")
cov <- rd("coverage")
sci <- function(x, d = 1) formatC(x, format = "e", digits = d)
preid$ok <- ifelse(preid$target_holds,
                   preid$translated & !is.na(preid$err_rel) & preid$err_rel < 1e-5,
                   !preid$translated)
ok_n <- function(route) {
  d <- preid[preid$route == route, ]
  paste(sum(d$ok), "of", nrow(d))
}
mx <- function(x) { x <- x[is.finite(x)]; if (length(x)) max(x) else NA_real_ }
inv_by_case <- tapply(inv$d_sphere_point_vs_marker, inv$case, max)
```

## Question

Before the sphere chart can become magmaan's default, it has to return the
standard answer everywhere the standard answer exists, across the model syntax
people actually write. Does it? And where a model's own identification cannot
hold the solution, does it say so instead of returning a wrong answer?

## Short answer

**Yes, on every deterministic check.** Across `r length(unique(rec$case))`
models and `r nrow(rec)` fits, the sphere route returned the ordinary result.
The models cover every identification convention, the measurement-invariance
ladder, equality and constraint syntax, and structural features, fitted by ML,
ULS, GLS, WLS, FIML and PSD-ML. The objectives agreed to
`r sci(mx(rec$d_fmin))` and the ML estimates to
`r sci(mx(rec$d_est[rec$estimator == "ML"]))`. Standard errors, robust
standard errors, the standardized solution, fit measures, modification indices
and defined parameters followed. Both routes were equally close to lavaan.
Each model's latents split between the sphere and the user's own
identification exactly as the design predicts, and every latent kept off the
sphere came with the stated reason.

**Known populations are recovered in every identification that contains
them, and the others are flagged.** The populations include a zero marker
loading, loadings that sum to zero, a zero or negative latent residual
variance, and one population that neither the marker, std.lv nor
effect-coding identification can hold.

- **Sphere ML:** recovered the population in all 24 (population,
  identification) cells whose identification contains it. It reported the
  other 8 as outside that identification, with a solution that re-expresses
  exactly in any identification that does contain it.
- **Ordinary ML:** recovered 20 of the 24, with three wrong points and one
  error. In the 8 cells that cannot hold the population, it three times
  reported convergence to a point that is not the optimum.
- **Least squares:** sphere ULS and GLS also recovered all 24 cells and
  flagged the other 8.

**The answer does not depend on how the user identified the model.** The
sphere route starts from a point computed on the model with its scale fixes
removed, which is the same whichever identification the user wrote. Four data
sets, each fitted under four identifications, gave the same sphere solution
within each data set: to `r sci(max(inv_by_case[names(inv_by_case) != "pd_free"]))`
on three of them, and to `r sci(inv_by_case[["pd_free"]])` on the flat
PoliticalDemocracy objective, whose objective values agree to
`r sci(max(inv$d_fmin_vs_marker))`. In
all 48 translations between identifications, the translated result matched a
direct ordinary fit to `r sci(mx(ireid$d_vs_ordinary))`. Ordinal data, ML2S,
two-level models and PSD for non-ML estimators are refused with a clear error.

The checks found one sphere bug, fixed before this run, and one design gap,
also fixed. A latent whose
variance is fixed by std.lv was not tolerance-checked against its pole. At a
population with a vanishing residual variance, the fit reported convergence
with a regression slope near 37,000 instead of flagging the identification.
The design gap was the start. In a first run, the sphere took its starting
values from the user's identification, and in 2 of 64 least-squares cells one
identification's start led to a non-global stationary point where the other
three recovered the population. A start computed without reference to the
identification removed both, with least-squares fits starting from the sphere
ML fit. The checks
also surfaced three pre-existing magmaan gaps unrelated to the sphere (see
Caveats).

## Evidence

### Standard results through the sphere

```{r recovery}
agg <- do.call(rbind, lapply(split(rec, rec$estimator), function(b) data.frame(
  Estimator = b$estimator[1], Fits = nrow(b),
  `Max Δ objective` = sci(mx(b$d_fmin)),
  `Max Δ estimate` = sci(mx(b$d_est)),
  `Max rel. Δ SE` = sci(mx(c(b$d_se, b$d_se_robust))),
  `Max rel. Δ fit measures` = sci(mx(b$d_fit_measures)),
  `Max Δ vs lavaan, ordinary` = sci(mx(b$lav_d_ord)),
  `Max Δ vs lavaan, sphere` = sci(mx(b$lav_d_sph)),
  check.names = FALSE)))
agg <- agg[match(c("ML", "FIML", "ULS", "GLS", "WLS", "PSD"), agg$Estimator), ]
agg[is.na(agg)] <- "-"
knitr::kable(agg, row.names = FALSE, align = "lrrrrrrr")
```

Differences are sphere minus ordinary. The largest estimate differences come
from flat objectives (FIML, and ULS on the N = 75 PoliticalDemocracy data),
where both routes stop within optimizer tolerance of each other and of
lavaan. PSD-ML has no lavaan counterpart. The per-fit table, including
standardized solutions, modification indices and defined parameters, is
`results/full/recovery.csv`.

### Which latents run on the sphere

```{r gauge}
first <- rec[!duplicated(rec$case), ]
reasons <- unlist(strsplit(first$pass_reasons[nzchar(first$pass_reasons) &
                                                !is.na(first$pass_reasons)], "; "))
fam <- aggregate(cbind(n_units, n_pass) ~ family, data = first, sum)
fam$models <- as.integer(table(first$family)[fam$family])
fam$matched <- as.integer(tapply(first$gauge_ok, first$family, sum)[fam$family])
knitr::kable(data.frame(
  Family = c(constraints = "Equality and constraint syntax",
             identification = "Identification conventions",
             invariance = "Measurement invariance",
             structure = "Structure")[fam$family],
  Models = fam$models, `On the sphere` = fam$n_units,
  `Kept in the user's identification` = fam$n_pass,
  `Classification as predicted` = paste0(fam$matched, "/", fam$models),
  check.names = FALSE), row.names = FALSE)
```

Counts are latents, with a latent tied across groups counted once. The
reasons for keeping a latent off the sphere were:
`r paste(sort(unique(reasons)), collapse = "; ")`.
Equal loadings within a factor, fixed non-unit markers, loading ratios and
inhomogeneous loading sums (`a + b == 1.2`) all stay on the sphere, as do
full metric, scalar and strict invariance.

### Known populations

```{r population}
lab <- function(d) {
  d$Cell <- ifelse(d$chart_holds, "identification holds the population",
                   "identification cannot hold it")
  d
}
p <- lab(pop)
p$route <- factor(p$route, c("sphere_ML", "sphere_PSD", "sphere_ULS", "sphere_GLS",
                             "ordinary_ML"))
tab <- function(d) {
  f <- function(x) sum(d$outcome %in% x)
  data.frame(Recovered = f("recovered"), `Flagged outside` = f("outside chart"),
             `Wrong point` = f("wrong point"),
             `Converged outside, not flagged` = f("returned outside chart"),
             `Error or not converged` = f(c("error", "not converged")),
             check.names = FALSE)
}
out <- do.call(rbind, lapply(split(p, list(p$route, p$Cell), drop = TRUE), function(d)
  cbind(Route = sub("_", " ", as.character(d$route[1])), Cell = d$Cell[1], tab(d))))
out <- out[order(out$Route, out$Cell, decreasing = c(FALSE, TRUE), method = "radix"), ]
knitr::kable(out, row.names = FALSE)
```

- **Populations.** A regular population, and populations with:
  - a marker loading of 0.02;
  - a marker loading of 0;
  - loadings summing to zero;
  - a zero endogenous residual variance;
  - a negative endogenous residual variance;
  - a latent outside the marker, std.lv and effect-coding identifications at
    once;
  - a two-group metric-invariance population with a zero marker loading.
- **Identifications:** the marker on the first or second indicator, std.lv,
  and effect coding.
- **PSD-ML** skips the population with a negative variance, which lies
  outside its domain.
- **Recovery** means every parameter within 1e-5 of the population value,
  relative to the largest parameter. The worst recovered case is
  `r sci(mx(pop$err_rel[pop$outcome == "recovered"]))`. The standardized
  solution matches too (`r sci(mx(pop$err_std[pop$outcome == "recovered"]))`).
- **Translation between identifications:** translating each sphere result
  into the other three identifications succeeded, or was refused, exactly as
  the population dictates:
  - ML: `r ok_n("sphere_ML")` translations correct;
  - PSD-ML: `r ok_n("sphere_PSD")`;
  - ULS: `r ok_n("sphere_ULS")`;
  - GLS: `r ok_n("sphere_GLS")`.

### The same answer under every identification

```{r invariance}
iv <- do.call(rbind, lapply(split(inv, inv$case), function(d) {
  r <- ireid[ireid$case == d$case[1], ]
  data.frame(Model = d$label[1],
             `Max Δ objective across identifications` = sci(mx(d$d_fmin_vs_marker)),
             `Max Δ sphere solution` = sci(mx(d$d_sphere_point_vs_marker)),
             `Translations matching the direct fit` = paste0(sum(r$translated), "/", nrow(r)),
             `Max Δ translated vs direct` = sci(mx(r$d_vs_ordinary)),
             check.names = FALSE)
}))
knitr::kable(iv, row.names = FALSE)
```

The sphere solution is compared up to the sign of each latent.

### What the sphere route refuses

```{r coverage}
knitr::kable(data.frame(Input = cov$feature,
                        `Ordinary route` = ifelse(cov$ordinary == "fits", "fits", "error"),
                        `Sphere route` = ifelse(cov$sphere == "fits", "fits", "clear error"),
                        check.names = FALSE), row.names = FALSE)
```

Composite (`<~`) syntax fails inside `fit_model()` on both routes. Composites
fit through the separate `magmaan_fcsem()` route, which has no sphere version.
Inequality constraints are unsupported everywhere.

## Caveats

- **Scope.** These are deterministic checks: fixed data sets and exact
  population moments, one fit each. They show the machinery is right. They
  say nothing about how often the sphere helps on sampled data. That is the
  next experiment (Ernst-type small-sample designs across estimators).
- **Least squares.** The sphere does not make least squares convex. On the
  sphere every indicator is equally reachable, including points where a
  factor collapses onto one indicator, which least squares (unlike ML) does
  not penalize. Starting least squares from the sphere ML fit avoided both
  first-run misses here. How often sphere and
  ordinary fits land in different basins on sampled data is for the next
  experiment.
- **Bounds.** `bounds = "standard"` bounds every loading, which keeps every
  latent off the sphere. A bounded fit therefore runs entirely in the user's
  identification.
- **lavaan comparisons.** ULS, GLS and WLS use the n - 1 covariance (lavaan's
  convention for them). WLS uses lavaan's own weight matrix.
- **Pre-existing magmaan gaps, unrelated to the sphere.** Both routes share
  these, so the comparisons work around them:
  - **Latent means under equal intercepts.** `group_equal = c("loadings",
    "intercepts")` leaves the group-2 latent means fixed at zero where lavaan
    frees them (df 63 against lavaan's 60). The scalar-type models here free
    them explicitly (`f ~ c(0, NA)*1`), which reproduces lavaan exactly.
  - **Effect coding with a mean structure.** `effect_coding = TRUE` codes the
    loadings only. lavaan's `effect.coding = TRUE` also codes the intercepts.
    The fit is identical but the mean parameters differ, so the comparison
    uses lavaan's `effect.coding = "loadings"`.
  - **Composites.** `fit_model()` fails on composite syntax with an optimizer
    error instead of pointing to `magmaan_fcsem()`.

## Reproduce

```sh
just r-dev                                  # install the current R package
cd experiments/engineering/14-sphere-chart-sanity
Rscript run_experiment.R --full             # about 10 seconds
quarto render report.qmd
```

`--smoke` runs a subset. `--sections` and `--cases` select parts, and `--help`
lists everything.
