---
title: "Standardized Guttman map on a Holzinger-Swineford CFA"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
need <- c("hs_snapshot", "population", "estimate_summary", "test_summary",
          "diagnostic_summary", "metadata")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --smoke\n",
       "or:\n  Rscript run_experiment.R --full --reps 500")
}
snapshot <- read.csv(paths[1])
population <- read.csv(paths[2])
est <- read.csv(paths[3])
tests <- read.csv(paths[4])
diag <- read.csv(paths[5])
metadata <- read.csv(paths[6])

meta <- function(key) metadata$value[match(key, metadata$key)]
fmt <- function(x, d = 3) ifelse(is.finite(x), formatC(x, format = "f", digits = d), "NA")
pct <- function(x) ifelse(is.finite(x), sprintf("%.1f%%", 100 * x), "NA")
label_est <- c(
  guttman_std = "Standardized Guttman map",
  guttman_std_ntml = "Guttman test: NTML metric",
  guttman_std_uls = "Guttman test: ULS metric",
  ml_nt = "Iterative NTML",
  uls_emp = "Iterative ULS"
)
largest_n <- max(est$n)

lookup <- function(df, estimator, group, scenario = "null", n = largest_n, col = "rmse") {
  z <- df[df$estimator == estimator & df$group == group &
          df$scenario == scenario & df$n == n, col]
  if (length(z)) z[[1]] else NA_real_
}
test_lookup <- function(estimator, test, scenario, n = largest_n, col = "rejection_rate") {
  z <- tests[tests$estimator == estimator & tests$test == test &
             tests$scenario == scenario & tests$n == n, col]
  if (length(z)) z[[1]] else NA_real_
}
ratio <- function(estimator, group, scenario = "null", n = largest_n) {
  lookup(est, estimator, group, scenario, n, "rmse") /
    lookup(est, "ml_nt", group, scenario, n, "rmse")
}
```

## Question

Does the standardized Guttman map behave like a practical CFA estimator on a
familiar three-factor Holzinger-Swineford problem? We compare it
with normal-theory ML and ULS on point estimates, Wald intervals, residual
goodness-of-fit, and the nested restriction `visual ~~ 0*speed`.

## Short answer

On this first Holzinger-Swineford-shaped run, the standardized Guttman
point estimates are close to ML but not ML-efficient. At n = `r largest_n`, its
RMSE is `r fmt(ratio("guttman_std", "loading"), 2)` times ML for loadings and
`r fmt(ratio("guttman_std", "residual_var"), 2)` times ML
for residual variances in the null cell.

Its Wald intervals are usable in the same broad sense: loading coverage is
`r pct(lookup(est, "guttman_std", "loading", col = "coverage"))` with empirical
Gamma, compared with
`r pct(lookup(est, "ml_nt", "loading", col = "coverage"))` for ML and
`r pct(lookup(est, "uls_emp", "loading", col = "coverage"))` for ULS.

For tests, the residual GOF route is available for Guttman and ML, while ULS is
included for the nested profile test. In the null cell, the nested rejection
rates at n = `r largest_n` are `r pct(test_lookup("guttman_std_ntml", "nested_visual_speed_zero", "null"))`
for Guttman with the NTML residual metric, `r pct(test_lookup("ml_nt", "nested_visual_speed_zero", "null"))`
for ML, and `r pct(test_lookup("uls_emp", "nested_visual_speed_zero", "null"))`
for ULS. In the alternative cell, the corresponding powers are
`r pct(test_lookup("guttman_std_ntml", "nested_visual_speed_zero", "alternative"))`,
`r pct(test_lookup("ml_nt", "nested_visual_speed_zero", "alternative"))`, and
`r pct(test_lookup("uls_emp", "nested_visual_speed_zero", "alternative"))`.

## Evidence

### Real-data snapshot

The table uses the actual Holzinger-Swineford data only as a sanity check. The
simulation below uses an ML-fitted population from the same data, with
`visual ~~ speed` either retained or set to zero.

```{r snapshot-table}
keep <- snapshot$param %in% c("visual =~ x2", "visual =~ x3",
                              "textual =~ x5", "speed =~ x8",
                              "visual ~~ speed", "x1 ~~ x1")
snap <- snapshot[keep, c("estimator", "param", "estimate", "se", "ci_lower", "ci_upper")]
snap$Estimator <- label_est[snap$estimator]
snap$Parameter <- snap$param
snap$Estimate <- fmt(snap$estimate)
snap$SE <- fmt(snap$se)
snap$CI <- paste0("[", fmt(snap$ci_lower), ", ", fmt(snap$ci_upper), "]")
knitr::kable(snap[, c("Estimator", "Parameter", "Estimate", "SE", "CI")],
             row.names = FALSE)
```

### Estimates and intervals

RMSE ratio relative to ML and 95% Wald interval coverage at the largest simulated
sample size. Values are grouped by free-parameter block.

```{r estimate-table}
rows <- list()
for (scenario in c("null", "alternative")) {
  for (group in c("loading", "factor_cov", "residual_var")) {
    for (estimator in c("guttman_std", "ml_nt", "uls_emp")) {
      r <- est[est$scenario == scenario & est$n == largest_n &
               est$group == group & est$estimator == estimator, ]
      if (!nrow(r)) next
      rows[[length(rows) + 1L]] <- data.frame(
        Scenario = scenario,
        Group = group,
        Estimator = label_est[estimator],
        `RMSE / ML` = fmt(r$rmse / lookup(est, "ml_nt", group, scenario), 2),
        Coverage = pct(r$coverage),
        `Avg SE` = fmt(r$avg_se),
        check.names = FALSE
      )
    }
  }
}
knitr::kable(do.call(rbind, rows), row.names = FALSE)
```

### Tests

Rejection rates for exact-fit GOF and the nested `visual ~~ 0*speed` restriction.
The null scenario sets that factor covariance to zero; the alternative scenario
uses the value from the ML fit to the observed Holzinger-Swineford data.

```{r test-table}
show_tests <- tests[tests$n == largest_n &
                    tests$test %in% c("gof", "nested_visual_speed_zero"), ]
show_tests <- show_tests[show_tests$estimator %in%
                           c("guttman_std_ntml", "guttman_std_uls", "ml_nt", "uls_emp"), ]
show_tests$Estimator <- label_est[show_tests$estimator]
show_tests$Test <- ifelse(show_tests$test == "gof", "GOF", "Nested")
show_tests$Rejection <- pct(show_tests$rejection_rate)
show_tests$`Mean p` <- fmt(show_tests$mean_p)
show_tests$`Negative stat` <- pct(show_tests$negative_stat_rate)
knitr::kable(show_tests[, c("scenario", "Estimator", "Test", "Rejection",
                            "Mean p", "Negative stat")],
             col.names = c("Scenario", "Estimator", "Test", "Rejection",
                           "Mean p", "Negative Stat"),
             row.names = FALSE)
```

### Diagnostics

```{r diag-table}
d <- diag[diag$n == largest_n & diag$stage == "H1", ]
d$Estimator <- label_est[d$estimator]
d$OK <- pct(d$ok_rate)
d$Converged <- pct(d$converged_rate)
d$Improper <- pct(d$improper_rate)
knitr::kable(d[, c("scenario", "Estimator", "OK", "Converged", "Improper")],
             col.names = c("Scenario", "Estimator", "OK", "Converged", "Improper"),
             row.names = FALSE)
```

## Caveats

- This is a normal-data, three-factor, three-indicator design. It is a first
  comparison of the new standardized Guttman path, not a general
  recommendation.
- The default `--smoke` render uses only `r meta("reps")` replications per cell.
  It is a plumbing check; use the `--full` command below before treating rates
  as Monte Carlo evidence.
- The population is seeded by fitting ML to the observed Holzinger-Swineford
  data. That makes the numbers interpretable, but it is one population.
- Guttman intervals use empirical Gamma. In this setup the NTML and ULS residual
  metrics give the same parameter covariance, so the NTML/ULS split is reported
  only for residual and nested tests. ULS intervals use the continuous-LS
  sandwich. ML intervals use the ordinary expected-information covariance.
- ULS exact-fit GOF is not reported here; the first pass uses ULS for estimates,
  CIs, and the nested continuous-LS profile test.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --full --reps 500 --cores 4
quarto render report.qmd
```
