---
title: "Standardized Guttman map on a Holzinger-Swineford CFA"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
results_dir <- "results"
need <- c("hs_snapshot", "population", "estimate_summary",
          "estimate_joint_summary", "whole_summary", "whole_joint_summary",
          "test_summary", "diagnostic_summary", "metadata")
paths <- file.path(results_dir, paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("Missing results. Run:\n  Rscript run_experiment.R --smoke\n",
       "or:\n  Rscript run_experiment.R --full --reps 500")
}
snapshot <- read.csv(paths[1])
population <- read.csv(paths[2])
est <- read.csv(paths[3])
est_joint <- read.csv(paths[4])
whole <- read.csv(paths[5])
whole_joint <- read.csv(paths[6])
tests <- read.csv(paths[7])
diag <- read.csv(paths[8])
metadata <- read.csv(paths[9])

meta <- function(key) metadata$value[match(key, metadata$key)]
fmt <- function(x, d = 3) ifelse(is.finite(x), formatC(x, format = "f", digits = d), "NA")
pct <- function(x) ifelse(is.finite(x), sprintf("%.1f%%", 100 * x), "NA")
label_est <- c(
  guttman_std = "Standardized Guttman map",
  guttman_std_ntml = "Guttman test: NTML metric",
  guttman_std_uls = "Guttman test: ULS metric",
  ml_nt = "Iterative NTML",
  uls_emp = "Iterative ULS"
)
largest_n <- max(est$n)

lookup <- function(df, estimator, group, scenario = "null", n = largest_n, col = "rmse") {
  z <- df[df$estimator == estimator & df$group == group &
          df$scenario == scenario & df$n == n, col]
  if (length(z)) z[[1]] else NA_real_
}
test_lookup <- function(estimator, test, scenario, n = largest_n, col = "rejection_rate") {
  z <- tests[tests$estimator == estimator & tests$test == test &
             tests$scenario == scenario & tests$n == n, col]
  if (length(z)) z[[1]] else NA_real_
}
ratio <- function(estimator, group, scenario = "null", n = largest_n) {
  z <- est_joint[est_joint$estimator == estimator & est_joint$group == group &
                 est_joint$scenario == scenario & est_joint$n == n, "rmse_ratio"]
  if (length(z)) z[[1]] else NA_real_
}
whole_ratio <- function(estimator, target, scenario = "null", n = largest_n) {
  col <- paste0(target, "_rmse_ratio")
  z <- whole_joint[whole_joint$estimator == estimator &
                   whole_joint$scenario == scenario & whole_joint$n == n, col]
  if (length(z)) z[[1]] else NA_real_
}
```

## Question

Does the standardized Guttman map behave like a practical CFA estimator on a
familiar three-factor Holzinger-Swineford problem? We compare it with
normal-theory ML and ULS on two targets: marker-chart free parameters and the
whole reconstructed covariance objects. Residual goodness-of-fit and the nested
restriction `visual ~~ 0*speed` are included as first test probes. The
`misspecified` scenario adds a hidden `x1 ~~ x2` residual covariance so GOF can
be checked for power/consistency, not just size.

## Short answer

On this first Holzinger-Swineford-shaped run, the standardized Guttman map is
being evaluated in the unconstrained H1 CFA: all factor covariances are free.
The H0 model only enters the nested-test rows. The free-parameter RMSEs are not
especially flattering. At n = `r largest_n`, the same-replication RMSE ratio
against ML is `r fmt(ratio("guttman_std", "loading"), 2)` for loadings and
`r fmt(ratio("guttman_std", "residual_var"), 2)` for residual variances in the
null cell.

The whole-estimator target is less harsh but still not ML-efficient. In the
same null cell, RMSE ratios against ML are
`r fmt(whole_ratio("guttman_std", "common"), 2)` for
`Lambda Phi Lambda'` and `r fmt(whole_ratio("guttman_std", "sigma"), 2)` for
the full implied `Sigma`. This is the target closest to experiment 55's
common-covariance MISE; the marker-parameter table is a different, chart-level
target.

Its Wald intervals are usable in the same broad sense: loading coverage is
`r pct(lookup(est, "guttman_std", "loading", col = "coverage"))` with empirical
Gamma, compared with
`r pct(lookup(est, "ml_nt", "loading", col = "coverage"))` for ML and
`r pct(lookup(est, "uls_emp", "loading", col = "coverage"))` for ULS.

For tests, the residual GOF route is available for Guttman and ML, while ULS is
included with Browne's residual NT statistic for exact-fit GOF and with the
continuous-LS profile test for the nested restriction. In the null cell, the
nested rejection rates at n = `r largest_n` are `r pct(test_lookup("guttman_std_ntml", "nested_visual_speed_zero", "null"))`
for Guttman with the NTML residual metric, `r pct(test_lookup("ml_nt", "nested_visual_speed_zero", "null"))`
for ML, and `r pct(test_lookup("uls_emp", "nested_visual_speed_zero", "null"))`
for ULS. In the alternative cell, the corresponding powers are
`r pct(test_lookup("guttman_std_ntml", "nested_visual_speed_zero", "alternative"))`,
`r pct(test_lookup("ml_nt", "nested_visual_speed_zero", "alternative"))`, and
`r pct(test_lookup("uls_emp", "nested_visual_speed_zero", "alternative"))`.

For exact-fit GOF, the relevant consistency check is the misspecified cell. At
n = `r largest_n`, rejection rates there are
`r pct(test_lookup("guttman_std_ntml", "gof", "misspecified"))` for Guttman
with the NTML metric, `r pct(test_lookup("guttman_std_uls", "gof", "misspecified"))`
for Guttman with the ULS metric, `r pct(test_lookup("ml_nt", "gof", "misspecified"))`
for ML, and `r pct(test_lookup("uls_emp", "gof", "misspecified"))` for ULS
Browne GOF. The smoke run is too small for stable rates; the table below is
mainly a plumbing and monotonicity check.

## Evidence

### Real-data snapshot

The table uses the actual Holzinger-Swineford data only as a sanity check. The
simulation below uses an ML-fitted population from the same data, with
`visual ~~ speed` retained, set to zero, or retained with an added hidden
`x1 ~~ x2` residual covariance.

```{r snapshot-table}
keep <- snapshot$param %in% c("visual =~ x2", "visual =~ x3",
                              "textual =~ x5", "speed =~ x8",
                              "visual ~~ speed", "x1 ~~ x1")
snap <- snapshot[keep, c("estimator", "param", "estimate", "se", "ci_lower", "ci_upper")]
snap$Estimator <- label_est[snap$estimator]
snap$Parameter <- snap$param
snap$Estimate <- fmt(snap$estimate)
snap$SE <- fmt(snap$se)
snap$CI <- paste0("[", fmt(snap$ci_lower), ", ", fmt(snap$ci_upper), "]")
knitr::kable(snap[, c("Estimator", "Parameter", "Estimate", "SE", "CI")],
             row.names = FALSE)
```

### Estimates and intervals

RMSE ratio relative to ML and 95% Wald interval coverage at the largest simulated
sample size. Values are grouped by free-parameter block. The ratio uses only
replications where both the listed estimator and ML produced finite estimates
for the same rows, so failures do not silently change the comparison target.

```{r estimate-table}
rows <- list()
for (scenario in c("null", "alternative")) {
  for (group in c("loading", "factor_cov", "residual_var")) {
    for (estimator in c("guttman_std", "ml_nt", "uls_emp")) {
      r <- est[est$scenario == scenario & est$n == largest_n &
               est$group == group & est$estimator == estimator, ]
      if (!nrow(r)) next
      j <- est_joint[est_joint$scenario == scenario & est_joint$n == largest_n &
                     est_joint$group == group & est_joint$estimator == estimator, ]
      rows[[length(rows) + 1L]] <- data.frame(
        Scenario = scenario,
        Group = group,
        Estimator = label_est[estimator],
        `RMSE / ML` = fmt(j$rmse_ratio, 2),
        Coverage = pct(r$coverage),
        `Avg SE` = fmt(r$avg_se),
        `Joint reps` = if (nrow(j)) j$n_joint_reps else NA_integer_,
        check.names = FALSE
      )
    }
  }
}
knitr::kable(do.call(rbind, rows), row.names = FALSE)
```

### Whole Estimator RMSE

The next table reconstructs the whole covariance objects from each fitted
partable before computing error. This is closer to experiment 55, which studied
common-covariance recovery rather than marker-chart free parameters. Ratios are
again computed on the same successful ML replications.

```{r whole-table}
rows <- list()
for (scenario in c("null", "alternative")) {
  for (estimator in c("guttman_std", "ml_nt", "uls_emp")) {
    r <- whole[whole$scenario == scenario & whole$n == largest_n &
               whole$estimator == estimator, ]
    j <- whole_joint[whole_joint$scenario == scenario & whole_joint$n == largest_n &
                     whole_joint$estimator == estimator, ]
    if (!nrow(r) || !nrow(j)) next
    rows[[length(rows) + 1L]] <- data.frame(
      Scenario = scenario,
      Estimator = label_est[estimator],
      `Common RMSE` = fmt(r$common_rmse),
      `Common / ML` = fmt(j$common_rmse_ratio, 2),
      `Sigma / ML` = fmt(j$sigma_rmse_ratio, 2),
      `Common diag / ML` = fmt(j$common_diag_rmse_ratio, 2),
      `Residual diag / ML` = fmt(j$residual_diag_rmse_ratio, 2),
      `Joint reps` = j$n_joint_reps,
      check.names = FALSE
    )
  }
}
knitr::kable(do.call(rbind, rows), row.names = FALSE)
```

### Tests

Rejection rates for exact-fit GOF by sample size. The null and alternative
scenarios are both correct for the H1 CFA, so GOF should stay near alpha there.
The misspecified scenario adds an unmodeled `x1 ~~ x2` residual covariance, so
GOF should reject more often as n increases.

```{r gof-table}
show_gof <- tests[tests$test == "gof", ]
show_gof <- show_gof[show_gof$estimator %in%
                       c("guttman_std_ntml", "guttman_std_uls", "ml_nt", "uls_emp"), ]
show_gof$Estimator <- label_est[show_gof$estimator]
show_gof$Rejection <- pct(show_gof$rejection_rate)
show_gof$`Mean p` <- fmt(show_gof$mean_p)
show_gof$`Finite` <- paste0(show_gof$n_finite, "/", show_gof$n_rows)
show_gof <- show_gof[order(show_gof$scenario, show_gof$n, show_gof$estimator), ]
knitr::kable(show_gof[, c("scenario", "n", "Estimator", "Rejection",
                          "Mean p", "Finite")],
             col.names = c("Scenario", "n", "Estimator", "Rejection",
                           "Mean p", "Finite"),
             row.names = FALSE)
```

Nested rejection rates for the `visual ~~ 0*speed` restriction at the largest
simulated sample size. The null scenario sets that factor covariance to zero;
the alternative and misspecified scenarios use the value from the ML fit to the
observed Holzinger-Swineford data.

```{r test-table}
show_tests <- tests[tests$n == largest_n &
                    tests$test == "nested_visual_speed_zero", ]
show_tests <- show_tests[show_tests$estimator %in%
                           c("guttman_std_ntml", "guttman_std_uls", "ml_nt", "uls_emp"), ]
show_tests$Estimator <- label_est[show_tests$estimator]
show_tests$Test <- "Nested"
show_tests$Rejection <- pct(show_tests$rejection_rate)
show_tests$`Mean p` <- fmt(show_tests$mean_p)
show_tests$`Negative stat` <- pct(show_tests$negative_stat_rate)
show_tests$`Finite` <- paste0(show_tests$n_finite, "/", show_tests$n_rows)
knitr::kable(show_tests[, c("scenario", "Estimator", "Test", "Rejection",
                            "Mean p", "Negative stat", "Finite")],
             col.names = c("Scenario", "Estimator", "Test", "Rejection",
                           "Mean p", "Negative Stat", "Finite"),
             row.names = FALSE)
```

### Diagnostics

```{r diag-table}
d <- diag[diag$n == largest_n & diag$stage == "H1", ]
d$Estimator <- label_est[d$estimator]
d$OK <- pct(d$ok_rate)
d$Converged <- pct(d$converged_rate)
d$Improper <- pct(d$improper_rate)
knitr::kable(d[, c("scenario", "Estimator", "OK", "Converged", "Improper")],
             col.names = c("Scenario", "Estimator", "OK", "Converged", "Improper"),
             row.names = FALSE)
```

## Caveats

- This is a normal-data, three-factor, three-indicator design. It is a first
  comparison of the new standardized Guttman path, not a general
  recommendation.
- The default `--smoke` render uses only `r meta("reps")` replications per cell.
  It is a plumbing check; use the `--full` command below before treating rates
  as Monte Carlo evidence.
- RMSE ratios in the estimate tables are apples-to-apples against ML on the
  same successful replications. The diagnostics table reports the marginal
  success, convergence, and improper-solution rates.
- The population is seeded by fitting ML to the observed Holzinger-Swineford
  data. That makes the numbers interpretable, but it is one population.
- Parameter and whole-estimator RMSE rows are omitted for the misspecified
  scenario, because the relevant large-sample target would be estimator-specific
  pseudo-true parameters.
- Guttman intervals use empirical Gamma. In this setup the NTML and ULS residual
  metrics give the same parameter covariance, so the NTML/ULS split is reported
  only for residual and nested tests. ULS intervals use the continuous-LS
  sandwich. ML intervals use the ordinary expected-information covariance.
- ULS exact-fit GOF uses Browne's residual NT statistic, matching magmaan's
  continuous-LS GOF convention for ULS.

## Reproduce

```sh
Rscript run_experiment.R --smoke
Rscript run_experiment.R --full --reps 500 --cores 4
quarto render report.qmd
```
