---
title: "Where do direct-FIML score flips stay calibrated?"
subtitle: "A 240-cell stress atlas with an independent hard-cell confirmation"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages(library(ggplot2))
script_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
read_result <- function(profile, name) read.csv(
  file.path(script_dir, "results", profile, paste0(name, ".csv")),
  stringsAsFactors = FALSE, check.names = FALSE)
required <- c("method_summary", "paired_summary", "timing_summary",
              "availability_summary", "failures", "metadata")
for (profile in c("screen", "confirm")) {
  paths <- file.path(script_dir, "results", profile,
                     paste0(required, ".csv"))
  if (!all(file.exists(paths))) stop("run the screen and confirmation before rendering")
}
screen <- lapply(required, read_result, profile = "screen")
confirmation <- lapply(required, read_result, profile = "confirm")
names(screen) <- names(confirmation) <- required
meta <- function(x) setNames(x$value, x$key)
screen_meta <- meta(screen$metadata)
confirmation_meta <- meta(confirmation$metadata)

method_label <- c(
  flip_basic = "flip: basic",
  flip_effective = "flip: effective",
  flip_standardized = "flip: standardized",
  score_sb = "score: SB",
  score_peba4 = "score: pEBA4",
  score_sandwich = "score: sandwich",
  score_all = "score: exact mixture",
  nested_sb_ml = "nested LR: SB",
  nested_peba4_ml = "nested LR: pEBA4",
  nested_all_ml = "nested LR: exact mixture")
cell_mean <- function(summary, method) mean(
  summary$rejection_rate[summary$method == method], na.rm = TRUE)
weighted <- function(x, field) {
  keep <- is.finite(x$n) & is.finite(x[[field]])
  stats::weighted.mean(x[[field]][keep], x$n[keep])
}
paired_row <- function(profile, x) data.frame(
  profile = profile,
  cells = nrow(x),
  paired_n = sum(x$n),
  effective = weighted(x, "rejection_effective"),
  standardized = weighted(x, "rejection_standardized"),
  disagreement = weighted(x, "disagreement_rate"),
  mean_abs_delta_p = weighted(x, "mean_abs_delta_p"))
paired_compare <- rbind(
  paired_row("Screen (199 signs)", screen$paired_summary),
  paired_row("Hard-cell confirmation (999 signs)", confirmation$paired_summary))

screen_effective <- cell_mean(screen$method_summary, "flip_effective")
screen_standardized <- cell_mean(screen$method_summary, "flip_standardized")
screen_peba4 <- cell_mean(screen$method_summary, "score_peba4")
screen_sb <- cell_mean(screen$method_summary, "score_sb")
screen_nested <- cell_mean(screen$method_summary, "nested_peba4_ml")
confirm_effective <- cell_mean(confirmation$method_summary, "flip_effective")
confirm_standardized <- cell_mean(confirmation$method_summary, "flip_standardized")
confirm_peba4 <- cell_mean(confirmation$method_summary, "score_peba4")
screen_flip_valid <- with(screen$availability_summary,
  availability[method == "flip_effective"])
screen_nested_valid <- with(screen$availability_summary,
  availability[method == "nested_peba4_ml"])
```

## Question

Across sample size, restriction rank, missingness mechanism, and four severe
distribution generators, do nuisance-effective or flip-standardized direct-FIML
score tests deliver dependable null p-values without choosing a chi-square
mixture approximation? When is the extra standardization worth its cost?

## Short answer

**Effective score flipping is a useful rescue of the unusably conservative
basic flip, but it is not uniformly calibrated in the hard nonnormal cells.**
Across 240 cells, effective and standardized flips rejected
`r sprintf("%.2f%%", 100 * screen_effective)` and
`r sprintf("%.2f%%", 100 * screen_standardized)`, respectively. The independent
11-cell confirmation deliberately concentrated on the difficult corners and
put them at `r sprintf("%.2f%%", 100 * confirm_effective)` and
`r sprintf("%.2f%%", 100 * confirm_standardized)`. Standardization helps in the
right direction, but too little to repair the worst VM, IG, and PL cells.

**Score pEBA4 is the clear calibration winner in this design.** It averaged
`r sprintf("%.2f%%", 100 * screen_peba4)` across the broad screen and
`r sprintf("%.2f%%", 100 * confirm_peba4)` in the independent hard-cell run.
Score SB also looks surprisingly good at `r sprintf("%.2f%%", 100 * screen_sb)`,
but that is an empirical result for these spectra, not evidence that a scalar
SB chi-square is generally the right reference law. The nested-LR pEBA4 arm
averaged `r sprintf("%.2f%%", 100 * screen_nested)` and remained badly liberal
under VM and PL.

**At ranks 1--5, standardization is rarely worth 999-sign use.** It changed
only `r sprintf("%.2f%%", 100 * paired_compare$disagreement[[2L]])` of confirmed
decisions and reduced rejection by about
`r sprintf("%.2f", 100 * (paired_compare$effective[[2L]] - paired_compare$standardized[[2L]]))`
percentage points, while its median covariance step cost about 95 ms per test.

## Evidence

The published two-group, six-indicator population is held fixed. The screen
crosses normal, VM, IG, and PL draws targeting skewness 3 and excess kurtosis
21; group-1 sample sizes 50, 100, 200, and 400; complete data, 15%/30% MCAR,
paper-style 30% MAR, and stronger 30% logistic MAR; and 1, 3, or 5 loading
equalities. Each screen cell has 500 attempts and 199 signs. The confirmation
uses an independent seed, 2,000 attempts per selected cell, and 999 signs.

### The broad atlas

```{r atlas, fig.width=12, fig.height=10}
atlas_methods <- c("flip_basic", "flip_effective", "flip_standardized",
                   "score_sb", "score_peba4", "nested_peba4_ml")
atlas <- screen$method_summary[screen$method_summary$method %in% atlas_methods, ]
atlas$method_label <- factor(unname(method_label[atlas$method]),
                             levels = unname(method_label[atlas_methods]))
atlas$cell_axis <- factor(
  paste0("n", atlas$n_group1, " / r", atlas$rank),
  levels = as.vector(outer(c(50, 100, 200, 400), c(1, 3, 5),
                           function(n, r) paste0("n", n, " / r", r))))
missing_label <- c(
  complete = "complete", mcar15 = "15% MCAR", mcar30 = "30% MCAR",
  mar30 = "30% paper MAR", strong_mar30 = "30% strong MAR")
atlas$missing_label <- factor(unname(missing_label[atlas$missingness]),
                              levels = unname(missing_label))
ggplot(atlas, aes(cell_axis, missing_label, fill = rejection_rate)) +
  geom_tile(colour = "white", linewidth = .25) +
  facet_grid(method_label ~ distribution) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b", midpoint = .05,
    limits = c(0, .20), oob = scales::squish,
    labels = scales::label_percent(accuracy = 1)) +
  labs(x = "group-1 sample size / restriction rank", y = NULL,
       fill = "null rejection",
       caption = "Each tile is a 500-attempt design cell; red above 5%, blue below 5%. Values above 20% share the darkest red.") +
  theme_minimal(base_size = 9) +
  theme(panel.grid = element_blank(), axis.text.x = element_text(
    angle = 55, hjust = 1), strip.text.y = element_text(angle = 0),
    legend.position = "bottom")
```

The heatmap makes the interaction plain. Normal cells are well behaved after
the nuisance correction. Under VM, IG, and PL, effective flips are most liberal
at small sample sizes and higher ranks; stronger MAR is not uniformly the worst
case. Basic flips remain severely conservative. Score pEBA4 stays close to 5%
through almost the entire grid, whereas the nested likelihood-ratio correction
does not.

### Independent hard-cell confirmation

```{r confirmation, fig.width=10, fig.height=6.5}
confirm_methods <- c("flip_effective", "flip_standardized", "score_peba4",
                     "nested_peba4_ml")
hard <- confirmation$method_summary[
  confirmation$method_summary$method %in% confirm_methods, ]
hard$method_label <- factor(unname(method_label[hard$method]),
                            levels = unname(method_label[confirm_methods]))
hard$cell_label <- with(hard, paste(
  distribution, paste0("n1=", n_group1), missing_label[missingness],
  paste0("r=", rank), sep = " / "))
cell_levels <- unique(hard$cell_label[order(
  hard$distribution, hard$n_group1, hard$missingness, hard$rank)])
hard$cell_label <- factor(hard$cell_label, levels = rev(cell_levels))
position <- position_dodge(width = .62)
ggplot(hard, aes(rejection_rate, cell_label, colour = method_label,
                 group = method_label)) +
  annotate("rect", xmin = .025, xmax = .075, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = .55) +
  geom_vline(xintercept = .05, linetype = 2, colour = "grey30") +
  geom_errorbar(aes(xmin = ci_lower, xmax = ci_upper), orientation = "y",
                width = .38, position = position, alpha = .55) +
  geom_point(position = position, size = 2) +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(x = "null rejection probability", y = NULL, colour = NULL,
       caption = "Independent seed; 2,000 attempts and 999 signs per cell. Bars are 95% Wilson intervals; grey is 2.5%--7.5%.") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_blank(), legend.position = "bottom")
```

The confirmation rejects a Monte Carlo accident explanation. Standardization
moves the worst flip cells down by roughly 0.2--0.7 percentage points, but VM
and PL at group-1 n = 50 remain around 11--12%. Score pEBA4 is not perfect—the
VM n = 50 cell is mildly liberal—but it is much more stable. The nested test is
liberal in nearly every severe nonnormal corner.

### Is standardization worth it?

```{r timing-table}
phase_median <- function(timing, phase) stats::median(
  timing$median_seconds[timing$phase == phase], na.rm = TRUE)
timing_compare <- paired_compare
timing_compare$fit_one_rank_ms <- 1000 * c(
  phase_median(screen$timing_summary, "fit_pair_equivalent"),
  phase_median(confirmation$timing_summary, "fit_pair_equivalent"))
timing_compare$effective_sum_ms <- 1000 * c(
  phase_median(screen$timing_summary, "flip_score"),
  phase_median(confirmation$timing_summary, "flip_score"))
timing_compare$standardization_ms <- 1000 * c(
  phase_median(screen$timing_summary, "flip_standardization"),
  phase_median(confirmation$timing_summary, "flip_standardization"))
timing_compare$wall_minutes <- as.numeric(c(
  screen_meta[["elapsed_seconds"]], confirmation_meta[["elapsed_seconds"]])) / 60
show <- timing_compare[c("profile", "cells", "paired_n", "effective",
  "standardized", "disagreement", "fit_one_rank_ms", "effective_sum_ms",
  "standardization_ms", "wall_minutes")]
knitr::kable(show, digits = c(0, 0, 0, 4, 4, 4, 1, 1, 1, 1),
  row.names = FALSE,
  col.names = c("run", "cells", "paired n", "effective reject",
    "standardized reject", "decision disagreement", "two fits (ms)",
    "effective sums (ms)", "standardization (ms)", "wall (min)"))
```

At 199 signs, standardization is a small addition to fitting. At 999 signs it
dominates the per-test work and still changes very few decisions. Its benefit
does increase with rank and with heterogeneous missing patterns, as expected,
but rank five is not enough to make it a default here. A sensible policy is to
keep the effective flip as the resampling baseline and reserve standardization
for diagnostics or much higher-rank tests.

Numerical availability also matters. Effective flips returned for
`r sprintf("%.2f%%", 100 * screen_flip_valid)` of the 120,000 screen attempts,
versus `r sprintf("%.2f%%", 100 * screen_nested_valid)` for the nested battery.
There were `r sum(!screen$failures$fit_ok)` row-level fit failures and only
`r sum(screen$failures$fit_ok & !screen$failures$flip_ok)` post-fit flip
failures. Most losses were concentrated at group-1 n = 50 with 30% missingness.

## Caveats

This is a focused null-calibration study, not a replication of the FIML--FMG
paper and not a power comparison. It uses one measurement model, one group-size
ratio, and four generators with the same severe marginal targets; there is no
Student-t condition. Calibration rates condition on a method returning a finite
p-value, so availability is reported separately.

The score-SB result deserves restraint. At rank one it coincides with the other
score references. At ranks three and five, its p-values differ from pEBA4 by
about 0.005--0.006 on average in the screen and change 0.35--0.54% of decisions.
Its 5.26% overall rate therefore says that scalar scaling happens to work well
for this collection of spectra; it does not turn a general chi-square mixture
into a scalar chi-square.

The screen's 199-sign p-values are discrete, and the stated rejection rule is
(p \le .05). The independent 999-sign confirmation is the basis for the claims
about difficult cells. More complex models, more restrictions, smaller
effective samples, or more heterogeneous pattern information could make
standardization matter more than it does here.

## Reproduce

```sh
# Cheap invariant check
Rscript experiments/67-fiml-flip-stress/run_experiment.R \
  --smoke --cores 4 --results-dir /tmp/magmaan-exp67-smoke

# Broad null screen (the measured run used 12 local workers)
Rscript experiments/67-fiml-flip-stress/run_experiment.R \
  --screen --cores 12

# Independent predefined hard-cell confirmation
Rscript experiments/67-fiml-flip-stress/run_experiment.R \
  --confirm --cores 12

quarto render experiments/67-fiml-flip-stress/report.qmd
```
